### 2023-02-22 paper/main copy.tex
init

% vim: ft=tex
\documentclass[a4,natbib=false]{article}
\usepackage[a4paper, total={8in, 10in}]{geometry}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{supertabular}
\usepackage[inline]{enumitem}
\usepackage{cleveref}

\newcommand{\niton}{\not\owns}

\newcommand{\ilyam}[1]{{\color{red} \texttt{Ilya:  #1}}}

\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}

\include{_build/_gen/included.tex}

\begin{document}

\begin{definition}[Context simplification]
  We simplify the context by applying it to itself.

  \ottfundefnsimpl{}
\end{definition}

\begin{lemma}
  Context is applicatively equivalent to its simplification.
  \[ [[ T* ]] \simeq [[ T ]] \]
\end{lemma}

\begin{definition}[Half-optimized unification algorithm]
  \label{def:half}
  I optimize the Ambroise's algorithm by simplifying the output context and
  removing the unused metavariables.

  \ottdefnun{}
\end{definition}

\begin{definition}[Fully-optimized unification algorithm]
  \label{def:full}
  I optimize the algorithm further by unrolling the UV\_F sub-inference into one step.

  \ottdefnaunTwo{}
\end{definition}

\begin{definition}[External terms]
  External terms are those whose metavariables do not have arguments bound by a
  constructor.

  \ottdefnaext{}
\end{definition}

\begin{definition}[External Contexts]
  External contexts are those whose assignments have external RHS.

  \ottdefnaextC{}
\end{definition}

\begin{lemma}
  \label{lemma:ext-app}
  if $ [[ t ext ]] $ and $ [[ T ext ]] $ then $ [[ [T]t ext ]] $.
\end{lemma}

\begin{lemma}
  Given a metavariable signature set  $\Sigma = [[ uc1[zs1] ]] \dots [[uc1[zsn] ]]$,
  an applied unification variable $[[ ua[xs] ]]$, an input context $[[ T1 ]]$,
  and a term $t \niton [[ua]]$ such that
  \begin{enumerate}
  \item $t$ and $[[ ua[xs] ]]$ are well-formed in $[[T1]]$
    % , i.e., $[[T1]]$ contains their unification variables and free variables.
  \item $[[T1]]$  is well-formed, external, and simplified ($ [[T1 ext]] $,  $[[T1*]] = [[T1]]$);
  \item $[[t]]$ is external ($[[t ext]]$);
  \item $[[ua : arn]] \in [[T1]]$
  \item $ [[T1]] $ is applied to $t$ ($[[ [T1] t ]] = [[t]]$);
  \item metavariable applications of $t$ are dictated by $\Sigma$
    (i.e. $UV(t) \subseteq \Sigma$), and $[[ ua[xs] ]] \in \Sigma$;
  \item for any $ [[ uci[zsi] ]] \in \Sigma $, the metavariable applications of
    $ [[ [T1](uci[zsi]) ]] $ are dictated by $\Sigma$
    (i.e. $UV([[ [T1]uci[zsi] ]]) \subseteq \Sigma$)
  \end{enumerate}
  if the half-optimized algorithm \ref{def:half} unifies $[[ ua[xs] ]]$ and
  $t$ in context $[[ T1 ]]$ with the resulting context $[[ T2 ]]$
  (i.e. $[[T1 ⊨ ua[xs] ≡ t : 0 ⫤ T2]]$),
  \textbf{then} the fully-optimized algorithm \ref{def:full}
  unifies them with certain resulting context $[[ T2' ]]$, equivalent to $[[ T2 ]]$
  (i.e. $[[T1 ⊫ ua[xs] ≡ t : 0 ⫥ T2']]$) and moreover,
  \begin{enumerate}
    \item $ [[ T2 ]] $ is well-formed, external and simplified;
    \item there exists $\Sigma' \supseteq \Sigma$ such that 
      $UV(\Sigma') \supseteq UV( [[ T2 ]] )$ and
      for any $ [[ uci[zsi] ]] \in \Sigma' $, the metavariable entries of 
      $ [[ [T2](uci[zsi]) ]] $ are dictated by $\Sigma'$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Induction on $t$.
  \begin{description}
  \item[$t = y$] In this case, $y = x_i$ and $UV(t) = \emptyset$.
    Then the output context of both algorithms is $[[(T1{ua:arn=xs.xi})*]]$.
    Let us prove that the output context satisfies the required properties.
    \begin{enumerate}
      \item This context is well-formed because $[[T1{ua:arn=xs.xi}]]$ is well-formed;
      \item This context is external because the input context $[[T1]]$ is
        external and the context update affects the part to the right from
        $[[ua:arn]]$ by applying $[[ua:arn=xs.xi]]$ (external!) to each
        instantiation (external!), which preserves the externality by
        \cref{lemma:ext-app}.
      \item This context is simplified by construction.
      \item Since no new metavariables are created, we keep the signature:
        $\Sigma' = \Sigma$. Since the input context $[[T1]]$ is simplified,
        the only instantiations that changed in the output context
        $[[(T1{ua:arn=xs.xi})*]]$ are $[[ua:arn]]$
        and those that contained $$[[ua]]$$. The substitution  
        $[[ua:arn=xs.xi]]$ only eliminates the entries of $[[ua]]$,
        and does not produces other entries. As such, the set of metavariable
        entries reduces, and each entry is still dictated by $\Sigma$ .
    \end{enumerate}

  \item[$t = [[ ub[ys] ]]$] 
    The output context of both algorithms is
    $[[T2]] = [[ (T1{uc:|zs|}{ua:arn=xs.uc[zs]}{ub:ark=ys.uc[zs]})* ]]$ where
    $[[zs = xs ∩ ys]]$ (a fresh metavariable $[[uc:|zs|]]$ is inserted just before 
    the earliest of two occurrences $[[ua]]$ and $[[ub]]$ in $[[T1]]$).
    Let us show that this context satisfies the required properties.
    \begin{enumerate}
      \item The context is well-formed because it was well-formed before
        simplification \ilyam{a lemma is required}.
      \item The context is external because so are the new instantiations, and
        context externality is preserved under simplification.
      \item The context is simplified by construction.
      \item We take $\Sigma' = \Sigma \cup \{ [[ uc[zs] ]] \}$.
        Let us take an entry from $\Sigma'$, which is affected by the context application.
        Then this entry is either 
        \begin{itemize}
        \item $[[ uai[xsi] ]]$ or $[[ ubi[ysi] ]]$, then the context application
          transforms it into $[[ uc[zs] ]] \in \Sigma'$;
        \item other $[[ udi[zsi] ]]$, then since $[[T1]]$ is simplified,
          $[[ [T2]udi[zsi] ]] = [[ [ ub:ark=ys.uc[zs] ][ ua:arn=xs.uc[zs]
          ]([T1]udi[zsi]) ]]$, which means that
          $UV([[ [T2]udi[zsi] ]]) \subseteq [[ [T1]udi[zsi] ]] \cup \{ [[ uc[zs] ]]\} \subseteq \Sigma'$. 
        \end{itemize}
    \end{enumerate}

  \item[$t = [[ f(ys1.t1,..,ysm.tm) ]]$] 
    blah

  \end{description}

\end{proof}

\begin{lemma}
  Given a signature of unification variables $\Sigma = [[ ua1[xs1] ]] \dots [[ ua1[xsn] ]]$,

  \textbf{if}
  \begin{enumerate}
    \item input context $ [[ T1 ]] $  is external and simplified ($ [[ T1 ext ]]
      $,  $ [[ T1 * ]] = [[ T1 ]] $);
    \item input terms $ [[ t1 ]] $ and $ [[ t2 ]] $ are external ($ [[ ti ext ]] $);
    \item $ [[ T1 ]] $ is applied to the input terms ($ [[ [T1] ti ]] = [[ ti ]] $);
    \item for any $ [[ uai ]] $, its entries in the terms are exactly as in
      $\Sigma$ ($UV(t_i) \subseteq \Sigma$);
    \item for any $ [[ uai[xsi] ]] \in \Sigma $, the metavariable entries of 
      $ [[ [T1](uai[xsi]) ]] $ are exactly as in $\Sigma$ ($UV([[ [T1]uai[xsi] ]]) \subseteq \Sigma$)
  \end{enumerate}

  \textbf{then} in result of the fully-optimized unification \ref{def:full},
    \begin{enumerate}
    \item the output context $ [[ T2 ]] $  is equivalent to those of the
      half-optimized unification \ref{def:half};
    \item $ [[ T2 ]] $ is external and simplified;
    \item output terms are external $ [[ [T2] ti ext ]] $;
    \item there exists $\Sigma' \supseteq \Sigma$ such that 
      $UV(\Sigma') \supseteq UV( [[ T2 ]] )$ and
      for any $ [[ uai[xsi] ]] \in \Sigma' $, the metavariable entries of 
      $ [[ [T2](uai[xsi]) ]] $ are exactly as in $\Sigma'$.
  \end{enumerate}

\end{lemma}

\begin{proof}

  % Induction on the inference tree corresponding to the step of the fully-optimized unification \ref{def:full}.
  % \begin{description}
  %   \item [$aV\_V$] In this case, the algorithms work identically, without
  %     changing the context. As such, the lemma conclusions hold almost immediately:
  %     \begin{enumerate}
  %       \item the rules of the two (deterministic!) algorithms are identical so are the output contexts;
  %       \item the output context is identical to the input context, which is
  %         external and simplified by the assumption;
  %       \item output terms are identical to the input terms, which are external;
  %       \item since the context has not changed, we chose $\Sigma' = \Sigma$.
  %     \end{enumerate}
  %   \item [$aF\_F$]
  %     Assume that $[[T0 ⊫ f(xs1.v1,..,xn.vn) ≡ f(xs1.w1,..,xsn.wn) : 0 ⫥ Tn]]$ and
  %     $[[ T0 ⊨ f(v1,..,vn) ≡ f(w1,..,wn) : 0 ⫤ Tn' ]]$. We want to show that
  %     $[[Tn]]$ is equivalent to $[[Tn']]$ and satisfies the required properties.
  %     By inversion, $[[T0, xs1 ⊫ v1 ≡ w1 : ark1 ⫥ T1, xs1]], \dots, [[Tn-1, xsn ⊫ [Tn-1]vn ≡ [Tn-1]wn : arkn ⫥ Tn, xsn]]$ and 
  %     $[[T0 ⊨ v1 ≡ w1 : ark1 ⫤ T1']], \dots, [[T'n-1 ⊨ [T'n-1]vn ≡ [T'n-1]wn : arkn ⫤ T'n]]$.
  %     Let us proceed by induction on $n$.

  %     For $n = 0$, the output context and the input context coincide, and the
  %     required property hold by the same reasons as in the previous case.

  %     If the  statement holds for $n-1$, then

  %     \begin{enumerate*}
  %     \item [(i)] $[[Tn-1]]$ is equivalent to $[[Tn-1']]$, and hence, they can be
  %       used interchangeably;
  %     \item [(ii)] $Tn-1$ is external and simplified;
  %     \item [(iii)] $T1$ are external
  %     \end{enumerate*}

  %      is equivalent to $[[T'n-1]]$ and

  %     Then by the IH,

  %     (and thus, inter)

  %     $[[T0 ⊨ v1 ≡ w1 : ark1 ⫤ T1]], \dots, [[Tn-1 ⊨ [Tn-1]vn ≡
  %     [Tn-1]wn : arkn ⫤ Tn]]$

  %     Induction on $n$.

  %     By inversion, $ [[T0, xs1 ⊫ v1 ≡ w1 : ark1 ⫥ T1, xs1 ]] $.
  %     First, let us show that the unification hypothesis is applicable to the
  %     context $ [[T0, xs1]] $, terms $ [[v1]]$ and $[[w1]]$, and the original $\Sigma$.
  %     \begin{enumerate}
  %       \item TODO 
  %     \end{enumerate}

  %     When unifying $ [[ xs1.v1 ]] $ and $ [[ xs1.w1 ]] $, the half-optimized
  %     unification ends up with $ [[ T0, xs1 ⊨ v1 ≡ w1 : ark1 ⫤ T1', xs1 ]] $.

  %     Then by the IH,
  %     \begin{enumerate*}
  %     \item [(i)] $[[T1]]$ is equivalent to $[[T1']]$, and hence, they can be
  %       used interchangeably;
  %     \item [(ii)] $T1$ is external and simplified;
  %     \item [(iii)] $T1$ is external and simplified;
  %     \end{enumerate*}

  %     $ [[ [T1]v1 ]] $ and $ [[ [T1]w1 ]] $ are external; there
  %     exists $\Sigma_1 \supseteq \Sigma$ stable under $ [[ T1' ]] $.

  %     Notice that the above is enough to apply the induction hypothesis to
  %     $ [[T1, xs2 ⊫ [T1]v2 ≡ [T1]w2 : ark2 ⫥ T2, xs2 ]] $ and $\Sigma_1$.
  %     Again, the half-optimized unification ends up with 
  %     $ [[ T1', xs2 ⊨ v2 ≡ w2 : ark2 ⫤ T2', xs1 ]] $ and since $[[T1]] ∼
  %     [[T2]]$, we apply the induction hypothesis and continue the process.

  %     % \begin{enumerate}

  %     %   \ref{def:half} will e
  %     %   By the induction hypothesis
  %     % \end{enumerate}
  % \end{description}
\end{proof}

\end{document}

### 2023-02-22 proofs/main.tex
init

% vim: ft=tex
\documentclass[a4,natbib=false]{article}
\usepackage[a4paper, total={8in, 10in}]{geometry}

\input{prelude.tex}

\input{../_build/_gen/included.tex}

\begin{document}

\section{Vanilla System}
\subsection{Grammar}
\ottgrammartabular{
  \ottP\ottinterrule
  \ottN\ottinterrule
}
\subsection{Declarative Subtyping}
\ottdefnsDZero

\section{Multi-Quantified System}
\subsection{Grammar}
\ottgrammartabular{
  \ottiP\ottinterrule
  \ottiN\ottinterrule
}
\subsection{Declarative Subtyping}
\ottdefnsDOne

\subsection{Declarative Equivalence}
\ottdefnsEOne

\section{Algorithm}
\subsection{Algorithmic Equivalence}
\ottdefnsEOneA

\subsection{Unification}
\ottdefnsU

\subsection{Algorithmic Subtyping}
\ottdefnsA

\subsection{Unification Solution Merge}
\ottdefnsSM

\subsection{Least Upper Bound}
\ottdefnsLUB

\subsection{Antiunification}
\ottdefnsAU

% \input{tex/grammar}
% \input{tex/declarative-subtyping}

\end{document}

### 2023-03-02 proofs/tex/ord-and-norm-aux-lemmas.tex
Add normalization proofs

\begin{lemma}
  \label{lemma:equiv-fv}
  Set of free variables is invariant under equivalence.
  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
\end{proof}

\begin{lemma}[Distributivity of renaming over variable ordering]
  \label{lemma:distr-mu-ord}
  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$, and $B$ is disjoint with $[[varset]]$.
  \begin{itemize}
  \item[$-$]
    If $B$ is disjoint with $[[fv iN]]$ then
    $[[ [mu] (ord varset in iN) ]] = [[ord ([mu] varset) in [mu] iN ]]$
  \item[$+$]
    If $B$ is disjoint with $[[fv iP]]$ then
    $[[ [mu] (ord varset in iP) ]] = [[ord ([mu] varset) in [mu] iP ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \todo[inline]{Proof (should be easy)}
\end{proof}

\begin{lemma}[Commutativity of normalization and renaming]
  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.

  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$. Then
  \begin{itemize}
    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
  \end{itemize}
  Here equality means alpha-equivalence.
\end{lemma}

\begin{proof}
  Mutual induction on $iN$ and $iP$.
  \todo[inline]{Write a little bit about the exists/forall case}
\end{proof}

### 2023-03-02 proofs/tex/ord-and-norm-completeness.tex
Add normalization proofs

\begin{lemma}[Completeness of variable ordering]
  \label{lemma:ord-completeness}
  Variable ordering is invariant under equivalence.

  \begin{itemize}
    \item[$-$] For $[[iN ≈ iM]]$ and any $[[varset]]$, if $[[ord varset in uN =
      ordVars1]]$ and $[[ord varset in uM = ordVars2]]$, then $[[ordVars1]]$ =
      $[[ordVars2]]$ (as lists)

    \item[$+$] For $[[iP ≈ iQ]]$ and any $[[varset]]$, if $[[ord varset in uP =
      ordVars1]]$ and $[[ord varset in uQ = ordVars2]]$, then $[[ordVars1]]$ =
      $[[ordVars2]]$ (as lists)
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
\end{proof}

\begin{lemma}[Completeness of quantified normalization]
  Normalization returns the same representative for equivalent types.

  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$
  \end{itemize}
  (Here equality means alpha-equivalence)
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
  \begin{itemize}
  \item {\nameref{\ottdruleEOneForallLabel}}

    From the definition of the normalization,
    \begin{itemize}
      \item $[[nf(∀pas.iN)]] = [[∀pas'.nf(iN)]]$ where $[[pas']]$ is $[[ord pas in nf(iN)]]$
      \item $[[nf(∀pbs.iM)]] = [[∀pbs'.nf(iM)]]$ where $[[pbs']]$ is $[[ord pbs in nf(iM)]]$
    \end{itemize}
    Let us take $[[mu : (pbs ∩ fv iM) ↔ (pas ∩ fv iN)]]$ from the
    inversion of the equivalence judgment. Notice that by
    \cref{lemma:ord-soundness,corollary:fv-nf}, the domain and codomain of $\mu$ can be written
    as $[[mu : pbs' ↔ pas']]$.

    To show the alpha-equivalence of $[[∀pas'.nf(iN)]]$ and $[[∀pbs'.nf(iM)]]$,
    it suffices to prove that
    \begin{enumerate*}
    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ and \newline
    \item[(ii)] $[[ [mu]pbs' ]] = [[pas']]$
    \end{enumerate*}.

    \begin{enumerate}
    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf([mu]iM)]] = [[nf(iN)]]$.
      The first equality holds by \cref{lemma:norm-subst-commute}, the second---by the induction hypothesis.

    \item[(ii)] $\begin{aligned}[t] [[ [mu]pbs' ]] &= [[ [mu] ord pbs in nf(iM) ]]
                                                  && \text{by the definition of $[[pbs']]$ } \\
                                                  &= [[ [mu] ord (pbs ∩ fv iM) in nf(iM) ]]
                                                  && \text{by \cref{corollary:ord-weakening,corollary:fv-nf} } \\
                                                  &= [[ ord [mu] (pbs ∩ fv iM) in [mu] nf(iM) ]]
                                                  && \text{by \cref{lemma:distr-mu-ord}, because
                                                     $[[pas' ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[pas ∩ fv iM ]]
                                                     = \emptyset$}\\
                                                  &
                                                  && \text{and $[[pas ∩ fv iN]] \cap [[(pbs ∩ fv iM)]] \subseteq
                                                     [[pas ∩ fv iM]] = \emptyset$} \\
                                                  &= [[ ord [mu] (pbs ∩ fv iM) in nf(iN) ]]
                                                  && \text{since $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ is proved } \\
                                                  &= [[ ord (pas ∩ fv iN) in nf(iN) ]]
                                                  && \text{because $\mu$ is a bijection between
                                                     $[[pas ∩ fv iN]]$ and $[[pbs ∩ fv iM]]$} \\
                                                  &= [[ ord pas in nf(iN) ]]
                                                  && \text{by \cref{corollary:ord-weakening,corollary:fv-nf} } \\
                                                  &= [[ pas' ]]
                                                  && \text{by the definition of $[[pas']]$} \\
      \end{aligned}$
    \end{enumerate}
  \item {\nameref{\ottdruleEOneExistsLabel}} Same as for \nameref{\ottdruleEOneForallLabel}.
  \item Other rules are congruent, and thus, proved by the corresponding congruent alpha-equivalence rule,
    which is applicable by the induction hypothesis. 
  \end{itemize}

\end{proof}

### 2023-03-06 proofs/tex/ord-and-norm-aux-lemmas.tex
Finish most of the normalization proofs

@@
 \end{lemma}
+
 \begin{proof}
   Mutual induction on $[[iN]]$ and $[[iP]]$.
-  \todo[inline]{Proof (should be easy)}
+  \begin{caseof}
+  \case[foo]{$[[iN]]$ = $[[na]]$}{ 
+    let us consider four cases:
+    \begin{enumerate}
+    \item $[[na]] \in A$ and $[[na]] \in [[varset]]$. Then
+      $
+      \begin{aligned}[t] [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in na)]] \\
+                                                             &= [[ [mu] na ]]
+                                                             && \text{by \ruleref{\ottdruleOPVarInLabel}}\\
+                                                             &= [[nb]]
+                                                             && \text{for some $[[nb]] \in B$ (notice that $[[nb]] \in [[ [mu]varset ]]$)} \\
+                                                             &= [[ ord [mu]varset in nb ]]
+                                                             && \text{by \ruleref{\ottdruleOPVarInLabel},
+                                                                because $[[nb]] \in [[ [mu]varset ]]$} \\
+                                                             &= [[ord [mu] varset in [mu] na ]]
+       \end{aligned}
+       $
+     \item $[[na]] \notin A$ and $[[na]] \notin [[varset]]$.
+       Notice that $[[na]] \notin B$, because $B$ is disjoint with $[[fv iN]]$.
+       Then $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[·]]$ by
+       \ruleref{\ottdruleOPVarNInLabel}.
+       On the other hand, $[[ ord [mu] varset in [mu] na = ord [mu] varset
+       in na ]] = [[·]]$ The latter equality is from
+       \ruleref{\ottdruleOPVarNInLabel}, because $[[na]] \notin B \cup
+       [[varset]] \supseteq [[ [mu] varset ]]$.
+     \item $[[na]] \in A$ but $[[na]] \notin [[varset]]$. Then
+       $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[·]]$
+       by \ruleref{\ottdruleOPVarNInLabel}.
+       To prove that $[[ ord [mu] varset in [mu] na ]] = [[·]]$, we apply
+       \ruleref{\ottdruleOPVarNInLabel}. Let us show that $ [[ [mu] na ]] \notin
+       [[ [mu] varset ]] $. If there is an element $x \in [[varset]]$ such that
+       $[[mu]] x = [[mu]] [[na]]$, then $x = [[na]]$ by bijectivity of
+       $[[mu]]$, which contradicts with $[[na]] \notin [[varset]]$. On the
+       other hand, $[[ [mu] na ]] \in B$, and hence, $ [[ [mu] na ]] \notin [[
+       varset ]]$.
+     \item $[[na]] \notin A$ but $[[na]] \in [[varset]]$.
+       $[[ ord [mu] varset in [mu] na ]] = [[ ord [mu] varset in na ]] = [[na]]$.
+       The latter is by \ruleref{\ottdruleOPVarNInLabel}, because
+       $[[na]] = [[ [mu] na ]] \in [[ [mu] varset ]]$ since $[[na]] \in [[varset]]$.
+       On the other hand, $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[ [mu] na ]] = [[na]]$.
+    \end{enumerate}
+  }
+  \case{$[[iN]] = [[↑iP]]$}{
+    $\begin{aligned}[t]
+       [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in ↑iP) ]] \\
+                                     &= [[ [mu] (ord varset in iP) ]]
+                                     && \text{by \ruleref{\ottdruleOShiftULabel}}\\
+                                     &= [[ ord [mu]varset in [mu]iP ]]
+                                     && \text{by the induction hypothesis}\\
+                                     &= [[ ord [mu]varset in  ↑[mu]iP ]]
+                                     && \text{by \ruleref{\ottdruleOShiftULabel}}\\
+                                     &= [[ ord [mu]varset in  [mu]↑iP ]]
+                                     && \text{by the definition of substitution}\\
+                                     &= [[ ord [mu]varset in  [mu]iN ]]
+            \end{aligned}$
+          }
+   \case[biz]{$[[iN]] = [[iP → iM]]$}{
+     $\begin{aligned}[t]
+        [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in iP → iM) ]] \\
+                                      &= [[ [mu] (ordVars1, (ordVars2 \ {ordVars1})) ]]
+                                      && \text{where } [[ord varset in iP = ordVars1]] \text{ and } [[ord varset in iM = ordVars2]] \\
+                                      &= [[ [mu] ordVars1, [mu](ordVars2 \ {ordVars1}) ]] \\
+                                      &= [[ [mu] ordVars1, ([mu]ordVars2 \ [mu]{ordVars1}) ]]
+                                      && \text{by induction on $[[ordVars2]]$;
+                                         the inductive step is similar to \cref{foo} }\\
+                                      &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
+      \end{aligned}$ \\
+    On the other hand,\\
+    $\begin{aligned}[t]
+       [[  (ord [mu] varset in [mu]iN) ]] &= [[ (ord [mu] varset in [mu]iP → [mu]iM) ]] \\
+                                     &= [[ (ordVarsb1, (ordVarsb2 \ {ordVarsb1})) ]]
+                                     && \text{where } [[ord [mu] varset in [mu] iP = ordVarsb1]] \text{ and } [[ord [mu] varset in [mu] iM = ordVarsb2]] \\
+                                          & && \text{then by the induction
+                                               hypothesis,
+                                               $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$,
+                                               $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$,
+                                               }\\
+                                     &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
+     \end{aligned}$
+   }
+   \case{$[[iN]] = [[∀ pas.iM]]$}{
+     $
+     \begin{aligned}[t]
+          [[ [mu] (ord varset in iN) ]] &= [[ [mu] ord varset in ∀pas.iM]] \\
+                                        &= [[ [mu] ord varset in iM]] \\
+                                        &= [[ ord [mu] varset in [mu] iM]]
+                                        && \text {by the induction hypothesis}\\
+     \end{aligned}\\
+     $
+     On the other hand,\\
+     $
+     \begin{aligned}[t]
+       [[ (ord [mu] varset in [mu] iN) ]] &= [[ ord [mu] varset in [mu] ∀pas.iM ]] \\
+                                          &= [[ ord [mu] varset in ∀pas.[mu]iM ]] \\
+                                          &= [[ ord [mu] varset in [mu] iM ]] \\
+     \end{aligned}
+     $
+    } 
+  \end{caseof}
 \end{proof}
@@
 \begin{proof}
-  Mutual induction on $iN$ and $iP$.
+  Mutual induction on $[[iN]]$ and $[[iP]]$.
   \todo[inline]{Write a little bit about the exists/forall case}

### 2023-03-08 proofs/main.tex
Add normalization
Add normalization proofs
Finish most of the normalization proofs
Add some missing proofs; restructure; cosmetics
Minor

@@
 \usepackage[a4paper, total={8in, 10in}]{geometry}
+\usepackage{hyperref}
+\usepackage{mathpartir}
 
@@
 
+% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
+% \newcommand{\ottdrulename}[1]{\textsc{#1}}
+
+% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
+%   }\ruleLabel{#4}{#4}}
+
+% \renewcommand{\ottdrule}[4][]{%
+%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
+%   \mpr@label{\textsc{#1}}%
+% }
+% \DeclareDocumentCommand \redrule {m m o}{%
+%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
+%   \mpr@label{\textsc{#3}}
+% }
+
+% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}
+
+% \renewcommand{\ottdruleEOneNVarName}[0]{foo }
+
+% \makeatletter
+% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
+%   }{
+%     % \protected@edef\@currentlabelname{foo}
+%     \def\@currentlabelname{foo}%
+%     \phantomsection
+%     \label{boo}
+%     \alpha ^{-}   \eqEOne   \alpha ^{-}
+%   }{%
+%     {\ottdruleEOneNVarName}{}%
+%   } }
+% \makeatother
+
+\newcommand{\ruleref}[1]{Rule \nameref{#1}}
+
+% ord varset in uN = varset'
+
+\renewcommand{\ottdruleONVarInName}[0]{(Var$_{\in}^-$)}
+\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{\notin}^-$)}
+\renewcommand{\ottdruleONUVarName}[0]{(UVar$^-$)}
+\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow$)}
+\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow$)}
+\renewcommand{\ottdruleOForallName}[0]{($\forall$)}
+
+% ord varset in uP = varset'
+
+\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{\in}^+$)}
+\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{\notin}^+$)}
+\renewcommand{\ottdruleOPUVarName}[0]{(UVar$^+$)}
+\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow$)}
+\renewcommand{\ottdruleOExistsName}[0]{($\exists$)}
+
+% nf(N) = M
+\renewcommand{\ottdruleNrmNVarName}[0]{(Var$^-$)}
+\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$^-$)}
+\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow$)}
+\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow$)}
+\renewcommand{\ottdruleNrmForallName}[0]{($\forall$)}
+
+% nf(P) = Q
+\renewcommand{\ottdruleNrmPVarName}[0]{(Var$^+$)}
+\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$^+$)}
+\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow$)}
+\renewcommand{\ottdruleNrmExistsName}[0]{($\exists$)}
+
+% N ≈ M
+
+\renewcommand{\ottdruleEOneNVarName}[0]{(Var$^-$$^{\eqEOne}$)}
+\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
+\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
+\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}
+
+% P ≈ Q
+\renewcommand{\ottdruleEOnePVarName}[0]{(Var$^+$)}
+\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
+\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}
+
+\renewcommand{\ottdruleSMESupSupName}[0]{$([[≥]]\&[[≥]])$}
+\renewcommand{\ottdruleSMEEqSupName}[0]{$([[≈]]\&[[≥]])$}
+\renewcommand{\ottdruleSMESupEqName}[0]{$([[≥]]\&[[≈]])$}
+\renewcommand{\ottdruleSMEPEqEqName}[0]{$([[≈]]\&[[≈]]^{+})$}
+\renewcommand{\ottdruleSMENEqEqName}[0]{$([[≈]]\&[[≈]]^{-})$}
+
 \begin{document}
 
-\section{Vanilla System}
+\section{The Vanilla System}
+
+First, we present the top-level system, which is easy to understand.
+
 \subsection{Grammar}
@@
 }
+
 \subsection{Declarative Subtyping}
@@
 \section{Algorithm}
-\subsection{Algorithmic Equivalence}
-\ottdefnsEOneA
+
+\subsection{Normalization}
+\subsubsection{Ordering}
+\ottdefnsOrder
+
+\subsubsection{Quantifier Normalization}
+\ottdefnsNrm
+
+% \subsection{Algorithmic Equivalence}
+% \ottdefnsEOneA
 
@@
 \subsection{Unification Solution Merge}
-\ottdefnsSM
+
+Unification solution is represented by a list of unification solution entries.
+Each entry restrict an unification variable in two possible ways: either stating
+that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua :≈
+iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥
+iP]]$).
+
+\begin{definition} [Matching Entries]
+  We call two entries matching if they are restricting the same unification variable.
+\end{definition}
+
+Two matching entries can be merged in the following way:
+\begin{definition} \hfill \\
+\ottdefnSME\\
+\end{definition}
+
+To merge two unification solution, we merge each pair of
+matching entries, and unite the results.
+
+\begin{definition}
+  $[[us1 & us2]] = \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
+  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with } [[usEntry2]] \}$
+\end{definition}
 
@@
 
-% \input{tex/grammar}
-% \input{tex/declarative-subtyping}
+\section{Proofs}
+
+\subsection{Variable Ordering}
+
+\input{../_build/_gen/tex/ord-lemmas.tex}
+
+\subsection{Normaliztaion}
+
+\input{../_build/_gen/tex/norm-lemmas.tex}
 

### 2023-03-13 proofs/main.tex
Add more contexts

@@
+\UseRawInputEncoding
 % vim: ft=tex
@@
 Each entry restrict an unification variable in two possible ways: either stating
-that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua :≈
-iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥
+that it must be equivalent to a certain type ($[[Δ ⊢ pua :≈ iP]]$ or $[[Δ ⊢ nua :≈
+iN]]$) or that it must be a (positive) supertype of a certain type ($[[Δ ⊢ pua :≥
 iP]]$).
@@
 
+\subsection{Subtyping}
+
+\input{../_build/_gen/tex/subtyping-lemmas.tex}
+
 \end{document}

### 2023-03-20 proofs/tex/norm-lemmas.tex
Add missing files + a couple of lemmas

\begin{lemma}
  \label{lemma:equiv-fv}
  Set of free variables is invariant under equivalence.
  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
\end{proof}

\begin{lemma}
  \label{lemma:fv-nf}
  Free variables are not changed by the normalization
  \begin{itemize}
  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  By straightforward induction on $[[nf(iN) = iM]]$.
\end{proof}

\begin{lemma}[Commutativity of normalization and renaming]
  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.

  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$. Then
  \begin{itemize}
    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
  \end{itemize}
  Here equality means alpha-equivalence.
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      $[[nf([mu]iN)]] = [[ nf([mu]na) ]] = [[ [mu]na ]] $.
      The latter follows from the fact that $[[ [mu]na ]]$ is a variable, and thus, 
      \ruleref{\ottdruleNrmNVarLabel} is applicable.

      $[[ [mu] nf(iN) ]] = [[ [mu] nf(na) ]] = [[ [mu] na ]] $.
   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
     the required equality follows from the congruence of the normalization and
     substitution, and the induction hypothesis.
     For example, if $[[iN]] = [[iP → iM]]$ then \\
     $\begin{aligned}[t]
        [[nf([mu] iN)]] &= [[ nf([mu] (iP → iM)) ]] \\
                        &= [[ nf([mu]iP → [mu]iM) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ nf([mu]iP) → nf([mu]iM) ]]
                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
                        &= [[ [mu]nf(iP) → [mu]nf(iM) ]]
                        && \text{By the induction hypothesis} \\
                        &= [[ [mu](nf(iP) → nf(iM)) ]]
                        && \text{By the congruence of substitution} \\
                        &= [[ [mu]nf(iP → iM) ]]
                        && \text{By the congruence of normalization} \\
                        &= [[ [mu]nf(iN) ]]
      \end{aligned}$ \\
    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
      $\begin{aligned}[t]
          [[ [mu] nf(iN) ]] &= [[ [mu] nf(∀pas.iM)]] \\
                            &= [[ [mu] ∀pas'.nf(iM) ]]
                            && \text {Where $[[ ord {pas} in nf(iM) = pas' ]]$}\\
        \end{aligned}$ \\

      $\begin{aligned}[t]
         [[ nf([mu]iN) ]] &= [[ nf([mu] ∀pas.iM)]] \\
                          &= [[ nf(∀pas.[mu]iM) ]]
                          && \text{Assuming $[[{pas}]] \cap A = \emptyset$
                             and $[[{pas}]] \cap B = \emptyset$}\\
                          &= [[ ∀pbs.nf([mu]iM) ]]
                          && \text {Where $[[ ord {pas} in nf([mu]iM) = pbs ]]$}\\
                          &= [[ ∀pas'.nf([mu]iM) ]]
                          && \text {As $[[pbs]] = [[pas']]$ (see below)}
         \end{aligned}$ \\

    Notice that $[[mu]]$ is free of collisions on $[[{pas}]]$ and $[[fv nf(iM)]]$ because
    \begin{enumerate}[nolistsep]
      \item[(i)] $[[mu]](A \cap [[{pas}]]) \cap [[fv nf(iM)]] = \emptyset \cap [[fv nf(iM)]] = \emptyset$ and
      \item[(ii)] $[[mu]](A \cap [[fv nf(iM)]]) \cap [[{pas}]] \subseteq B \cap
        [[{pas}]] = \emptyset$
    \end{enumerate}
    $\begin{aligned}[t]
       [[pbs]] &= [[ ord {pas} in nf([mu]iM) ]] \\
               &= [[ ord {pas} in [mu]nf(iM) ]]
               && \text{By the induction hypothesis} \\
               &= [[ ord {[mu]pas} in [mu]nf(iM) ]]
               && \text{Since $[[{pas}]] \cap A = \emptyset$} \\
               &= [[ [mu] ord {pas} in nf(iM) ]]
               && \text{by \cref{lemma:distr-mu-ord}} \\
               &= [[ ord {pas} in nf(iM) ]]
               && \text{Since $[[{ord {pas} in nf(iM)}]] \cap A \subseteq [[{pas}]] \cap A = \emptyset$} \\
               &= [[ pas' ]]
     \end{aligned}$ \\

     To show alpha-equivalence of 
     $[[ [mu] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.nf([mu]iM) ]]$,
     we can assume that $[[{pas'}]] \cap A = \emptyset$, and $[[{pas'}]] \cap B = \emptyset$.
     Then $[[ [mu] ∀pas'.nf(iM) ]] = [[∀pas'.[mu]nf(iM)]] = [[ ∀pas'.nf([mu]iM)
     ]]$, the latter follows from the induction hypothesis.

   \item $[[iP]] = [[∃ nas.iQ]]$ \\
     Same as for \cref{case:norm-subst-commute}.
  \end{caseof}
\end{proof}

\begin{lemma}[Soundness of quantifier normalization]
  \label{lemma:normalization-soundness}
  \hfill

  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}

\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

      To prove $[[∀pas.iN ≈ ∀pas'.iN']]$, it suffices to provide a bijection 
      $\mu : [[{pas'} ∩ fv iN']] \leftrightarrow [[{pas} ∩ fv iN]]$ such that
      $[[iN ≈ [mu]iN']]$. Since these sets are equal, we take $\mu = id$.
    \item{\nameref{\ottdruleNrmExistsLabel}} Same as for \cref{case:norm-soundness:forall}.
  \end{caseof}
\end{proof}

\begin{lemma}[Completeness of quantified normalization]
  Normalization returns the same representative for equivalent types.

  \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$
  \end{itemize}
  (Here equality means alpha-equivalence)
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
  \begin{caseof}
  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall} \\

    From the definition of the normalization,
    \begin{itemize}
      \item $[[nf(∀pas.iN)]] = [[∀pas'.nf(iN)]]$ where $[[pas']]$ is $[[ord {pas} in nf(iN)]]$
      \item $[[nf(∀pbs.iM)]] = [[∀pbs'.nf(iM)]]$ where $[[pbs']]$ is $[[ord {pbs} in nf(iM)]]$
    \end{itemize}
    Let us take $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$ from the
    inversion of the equivalence judgment. Notice that from
    \cref{lemma:fv-nf,lemma:ord-soundness}, the domain and the codomain of $\mu$ can be written
    as $[[mu : {pbs'} ↔ {pas'}]]$.

    To show the alpha-equivalence of $[[∀pas'.nf(iN)]]$ and $[[∀pbs'.nf(iM)]]$,
    it suffices to prove that
    \begin{enumerate*}
    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ and \newline
    \item[(ii)] $[[ [mu]pbs' ]] = [[pas']]$
    \end{enumerate*}.

    \begin{enumerate}
    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf([mu]iM)]] = [[nf(iN)]]$.
      The first equality holds by \cref{lemma:norm-subst-commute}, the second---by the induction hypothesis.

    \item[(ii)] $\begin{aligned}[t] [[ [mu]pbs' ]] &= [[ [mu] ord {pbs} in nf(iM) ]]
                                                  && \text{by the definition of $[[pbs']]$ } \\
                                                  &= [[ [mu] ord ({pbs} ∩ fv iM) in nf(iM) ]]
                                                  && \text{from \cref{lemma:fv-nf,corollary:ord-weakening} } \\
                                                  &= [[ ord [mu] ({pbs} ∩ fv iM) in [mu] nf(iM) ]]
                                                  && \text{by \cref{lemma:distr-mu-ord}, because
                                                     $[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]]
                                                     = \emptyset$}\\
                                                  &
                                                  && \text{and $[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq
                                                     [[{pas} ∩ fv iM]] = \emptyset$} \\
                                                  &= [[ ord [mu] ({pbs} ∩ fv iM) in nf(iN) ]]
                                                  && \text{since $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ is proved } \\
                                                  &= [[ ord ({pas} ∩ fv iN) in nf(iN) ]]
                                                  && \text{because $\mu$ is a bijection between
                                                     $[[{pas} ∩ fv iN]]$ and $[[{pbs} ∩ fv iM]]$} \\
                                                  &= [[ ord {pas} in nf(iN) ]]
                                                  && \text{from \cref{lemma:fv-nf,corollary:ord-weakening} } \\
                                                  &= [[ pas' ]]
                                                  && \text{by the definition of $[[pas']]$} \\
      \end{aligned}$
    \end{enumerate}
  \item {\nameref{\ottdruleEOneExistsLabel}} Same as for \cref{case:ord-completeness:forall}.
  \item Other rules are congruent, and thus, proved by the corresponding congruent alpha-equivalence rule,
    which is applicable by the induction hypothesis. 
  \end{caseof}
\end{proof}

### 2023-03-20 proofs/tex/ord-lemmas.tex
Add missing files + a couple of lemmas

\begin{definition} [Collision free bijection]
  We say that a bijection $\mu : A \leftrightarrow B$ between sets of
  variables is \textbf{collision free on sets} $P$ and $Q$ if and only if
  \begin{enumerate}
  \item $\mu(P \cap A) \cap Q = \emptyset$
  \item $\mu(Q \cap A) \cap P = \emptyset$
  \end{enumerate}
\end{definition}

\begin{lemma}[Soundness of variable ordering]
  \label{lemma:ord-soundness}
  Variable ordering extracts precisely used free variables.
  \begin{itemize}
    \item[$-$] $[[ {ord varset in uN} ]] \equiv [[varset ∩ fv uN]]$ (as sets)
    \item[$+$] $[[ {ord varset in uP} ]] \equiv [[varset ∩ fv uP]]$ (as sets)
  \end{itemize}
\end{lemma}
\begin{proof}
  Straightforward mutual induction on 
  $[[ ord varset in uN = ordVars ]]$ and $[[ ord varset in uP = ordVars ]]$
\end{proof}

\begin{corollary}[Additivity of ordering]
  \label{corollary:ord-additivity}
  Variable ordering is additive (in terms of set union) with respect to its first argument.
  \begin{itemize}
    \item[$-$] $[[ {ord (varset1 ∪ varset2) in iN} ]]
                \equiv
                [[{ord varset1 in iN} ∪ {ord varset2 in iN}]]$ (as sets)
    \item[$+$] $[[{ord (varset1 ∪ varset2) in iP}]]
                \equiv
                [[{ord varset1 in iP} ∪ {ord varset2 in iP}]]$ (as sets)

  \end{itemize}
\end{corollary}

\begin{corollary}[Weakening of ordering]
  \label{corollary:ord-weakening}
  Extending the first argument of the ordering with unused variables does not
  change the result.
  \begin{itemize}
  \item[$-$] $[[ ord (varset ∩ fv iN) in iN ]] = [[ ord varset in iN ]]$
  \item[$+$] $[[ ord (varset ∩ fv iP) in iP ]] = [[ ord varset in iP ]]$
  \end{itemize}
\end{corollary}

\begin{lemma}[Distributivity of renaming over variable ordering]
  \label{lemma:distr-mu-ord}
  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$.

  \begin{itemize}
  \item[$-$]
    If $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$ then
    $[[ [mu] (ord varset in iN) ]] = [[ord ([mu] varset) in [mu] iN ]]$
  \item[$+$]
    If $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iP]]$ then
    $[[ [mu] (ord varset in iP) ]] = [[ord ([mu] varset) in [mu] iP ]]$
  \end{itemize}
\end{lemma}

\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
  \item $[[iN]]$ = $[[na]]$ \label{case:distr-mu-ord:var} \\
    let us consider four cases:
    \begin{caseof}
    \item $[[na]] \in A$ and $[[na]] \in [[varset]]$\\ Then
      $
      \begin{aligned}[t] [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in na)]] \\
                                                             &= [[ [mu] na ]]
                                                             && \text{by \ruleref{\ottdruleOPVarInLabel}}\\
                                                             &= [[nb]]
                                                             && \text{for some $[[nb]] \in B$ (notice that $[[nb]] \in [[ [mu]varset ]]$)} \\
                                                             &= [[ ord [mu]varset in nb ]]
                                                             && \text{by \ruleref{\ottdruleOPVarInLabel},
                                                                because $[[nb]] \in [[ [mu]varset ]]$} \\
                                                             &= [[ord [mu] varset in [mu] na ]]
       \end{aligned}
       $
     \item $[[na]] \notin A$ and $[[na]] \notin [[varset]]$\\
       Notice that
       $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[·]]$ by
       \ruleref{\ottdruleOPVarNInLabel}.
       On the other hand, $[[ ord [mu] varset in [mu] na = ord [mu] varset
       in na ]] = [[·]]$ The latter equality is from
       \ruleref{\ottdruleOPVarNInLabel}, because
       $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$, so
       $[[fv iN]] \ni [[na]] \notin [[mu]](A \cap [[varset]]) \cup
       [[varset]] \supseteq [[ [mu] varset ]]$.
     \item $[[na]] \in A$ but $[[na]] \notin [[varset]]$\\ Then
       $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[·]]$
       by \ruleref{\ottdruleOPVarNInLabel}.
       To prove that $[[ ord [mu] varset in [mu] na ]] = [[·]]$, we apply
       \ruleref{\ottdruleOPVarNInLabel}. Let us show that
       $[[ [mu] na ]] \notin [[ [mu] varset ]]$.
       Since $[[ [mu] na ]] = [[mu]]([[na]])$ and
       $[[ [mu] varset ]] \subseteq [[mu]](A \cap [[varset]]) \cup [[varset]]$,
       it suffices to prove 
       $[[mu]]([[na]]) \notin [[mu]](A \cap [[varset]]) \cup [[varset]]$.

       \begin{enumerate}
       \item[(i)] If there is an element $x \in A \cap [[varset]]$ such that
         $[[mu]] x = [[mu]] [[na]]$, then $x = [[na]]$ by bijectivity of
         $[[mu]]$, which contradicts with $[[na]] \notin [[varset]]$. This way, 
         $[[mu]]([[na]]) \notin [[mu]](A \cap [[varset]])$.
       \item[(ii)]
         Since $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$,
         $[[mu]] (A \cap [[fv iN]]) \ni [[mu]]([[na]]) \notin [[varset]]$.
       \end{enumerate}

     \item $[[na]] \notin A$ but $[[na]] \in [[varset]]$\\
       $[[ ord [mu] varset in [mu] na ]] = [[ ord [mu] varset in na ]] = [[na]]$.
       The latter is by \ruleref{\ottdruleOPVarNInLabel}, because
       $[[na]] = [[ [mu] na ]] \in [[ [mu] varset ]]$ since $[[na]] \in [[varset]]$.
       On the other hand, $[[ [mu] (ord varset in iN) ]] = [[ [mu] (ord varset in na)]] = [[ [mu] na ]] = [[na]]$.
    \end{caseof}

  \item $[[iN]] = [[↑iP]]$ \\
    $\begin{aligned}[t]
       [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in ↑iP) ]] \\
                                     &= [[ [mu] (ord varset in iP) ]]
                                     && \text{by \ruleref{\ottdruleOShiftULabel}}\\
                                     &= [[ ord [mu]varset in [mu]iP ]]
                                     && \text{by the induction hypothesis}\\
                                     &= [[ ord [mu]varset in  ↑[mu]iP ]]
                                     && \text{by \ruleref{\ottdruleOShiftULabel}}\\
                                     &= [[ ord [mu]varset in  [mu]↑iP ]]
                                     && \text{by the definition of substitution}\\
                                     &= [[ ord [mu]varset in  [mu]iN ]]
            \end{aligned}$

   \item $[[iN]] = [[iP → iM]]$  \\
     $\begin{aligned}[t]
        [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in iP → iM) ]] \\
                                      &= [[ [mu] (ordVars1, (ordVars2 \ {ordVars1})) ]]
                                      && \text{where } [[ord varset in iP = ordVars1]] \text{ and } [[ord varset in iM = ordVars2]] \\
                                      &= [[ [mu] ordVars1, [mu](ordVars2 \ {ordVars1}) ]] \\
                                      &= [[ [mu] ordVars1, ([mu]ordVars2 \ [mu]{ordVars1}) ]]
                                      && \text{by induction on $[[ordVars2]]$;
                                         the inductive step is similar to \cref{case:distr-mu-ord:var}.
                                         Notice that $[[mu]]$ is} \\
                                      & && \text{collision free on $[[{ordVars1}]]$ and $[[{ordVars2}]]$
                                           since
                                           $[[{ordVars1}]] \subseteq [[varset]]$ and
                                           $[[{ordVars2}]] \subseteq [[fv iN]]$ }\\
                                      &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
      \end{aligned}$ \\
    $\begin{aligned}[t]
       [[  (ord [mu] varset in [mu]iN) ]] &= [[ (ord [mu] varset in [mu]iP → [mu]iM) ]] \\
                                     &= [[ (ordVarsb1, (ordVarsb2 \ {ordVarsb1})) ]]
                                     && \text{where } [[ord [mu] varset in [mu] iP = ordVarsb1]] \text{ and } [[ord [mu] varset in [mu] iM = ordVarsb2]] \\
                                          & && \text{then by the induction
                                               hypothesis,
                                               $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$,
                                               $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$,
                                               }\\
                                     &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
     \end{aligned}$

   \item $[[iN]] = [[∀ pas.iM]]$ \\
     $\begin{aligned}[t]
          [[ [mu] (ord varset in iN) ]] &= [[ [mu] ord varset in ∀pas.iM]] \\
                                        &= [[ [mu] ord varset in iM]] \\
                                        &= [[ ord [mu] varset in [mu] iM]]
                                        && \text {by the induction hypothesis}\\
     \end{aligned}$ \\
     $ 
     \begin{aligned}[t]
       [[ (ord [mu] varset in [mu] iN) ]] &= [[ ord [mu] varset in [mu] ∀pas.iM ]] \\
                                          &= [[ ord [mu] varset in ∀pas.[mu]iM ]] \\
                                          &= [[ ord [mu] varset in [mu] iM ]] \\
     \end{aligned}
     $

  \end{caseof}
\end{proof}

\begin{lemma}[Completeness of variable ordering]
  \label{lemma:ord-completeness}
  Variable ordering is invariant under equivalence.

  \begin{itemize}
  \item[$-$] For $[[iN ≈ iM]]$ and any $[[varset]]$,
    $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)

  \item[$+$] For $[[iP ≈ iQ]]$ and any $[[varset]]$,
    $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
\end{proof}

### 2023-03-20 proofs/main.tex
Add contexts
Add lemmas overview + Various fixes
minor
Fix the definition of merge

@@
 \begin{definition}
-  $[[us1 & us2]] = \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
-  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with } [[usEntry2]] \}$
+  foo
+
+  $\begin{aligned}[t]
+  [[us1 & us2]] &= \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
+  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with }
+                                   [[usEntry2]] \}\\
+                &\cup
+                                     \{ [[usEntry1]] \mid [[usEntry1]] \in
+                                     [[us1]], \text{ s.t. }
+                                     \forall [[usEntry2]]  \in [[us2]],
+                                     [[usEntry1]] \text{ does not match with }
+                                                     [[usEntry2]] \}\\
+        &\cup
+          \{ [[usEntry2]] \mid [[usEntry2]] \in
+          [[us2]], \text{ s.t. }
+          \forall [[usEntry1]]  \in [[us1]],
+          [[usEntry1]] \text{ does not match with }
+          [[usEntry2]] \}\\
+   \end{aligned}$
 \end{definition}
@@
 
+\subsection{Overview}
+\input{../_build/_gen/tex/lemmas-overview.tex}
+
 \subsection{Variable Ordering}
@@
 
-\subsection{Subtyping}
+\subsection{Upper Bounds}
 
-\input{../_build/_gen/tex/subtyping-lemmas.tex}
+\input{../_build/_gen/tex/lub-lemmas.tex}
 

### 2023-03-28 proofs/main.tex
Prove the UB lemma

@@
 
+% G ⊢ N ≤1 M
+
+\renewcommand{\ottdruleDOneNVarName}[0]{(Var$^-$$^{\subDOne}$)}
+\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
+\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
+\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}
+
+% G ⊢ P ≥1 Q
+\renewcommand{\ottdruleDOnePVarName}[0]{(Var$^+$$^{\supDOne}$)}
+\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
+\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}
+
 \begin{document}
@@
 \begin{definition}
-  foo
-
   $\begin{aligned}[t]
@@
 
+\subsection{Substitution}
+\input{../_build/_gen/tex/subst-lemmas.tex}
+
+\subsection{Type well-formedness}
+\input{../_build/_gen/tex/wf-lemmas.tex}
+
 \subsection{Overview}
@@
 \subsection{Variable Ordering}
-
 \input{../_build/_gen/tex/ord-lemmas.tex}
@@
 \subsection{Normaliztaion}
+\input{../_build/_gen/tex/norm-lemmas.tex}
 
-\input{../_build/_gen/tex/norm-lemmas.tex}
+\subsection{Equivalence}
+\input{../_build/_gen/tex/equiv-lemmas.tex}
 

### 2023-03-28 proofs/tex/norm-lemmas.tex
Prove the UB lemma

@@
 
-\begin{lemma}[Commutativity of normalization and renaming]
-  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
-
-  Suppose that $\mu$ is a bijection between two sets of variables
-  $\mu : A \leftrightarrow B$. Then
-  \begin{itemize}
-    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
-    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
-  \end{itemize}
-  Here equality means alpha-equivalence.
-\end{lemma}
-
-\begin{proof}
-  Mutual induction on $[[iN]]$ and $[[iP]]$.
-  \begin{caseof}
-    \item $[[iN]]$ = $[[na]]$ \\
-      $[[nf([mu]iN)]] = [[ nf([mu]na) ]] = [[ [mu]na ]] $.
-      The latter follows from the fact that $[[ [mu]na ]]$ is a variable, and thus, 
-      \ruleref{\ottdruleNrmNVarLabel} is applicable.
-
-      $[[ [mu] nf(iN) ]] = [[ [mu] nf(na) ]] = [[ [mu] na ]] $.
-   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
-     the required equality follows from the congruence of the normalization and
-     substitution, and the induction hypothesis.
-     For example, if $[[iN]] = [[iP → iM]]$ then \\
-     $\begin{aligned}[t]
-        [[nf([mu] iN)]] &= [[ nf([mu] (iP → iM)) ]] \\
-                        &= [[ nf([mu]iP → [mu]iM) ]]
-                        && \text{By the congruence of substitution} \\
-                        &= [[ nf([mu]iP) → nf([mu]iM) ]]
-                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
-                        &= [[ [mu]nf(iP) → [mu]nf(iM) ]]
-                        && \text{By the induction hypothesis} \\
-                        &= [[ [mu](nf(iP) → nf(iM)) ]]
-                        && \text{By the congruence of substitution} \\
-                        &= [[ [mu]nf(iP → iM) ]]
-                        && \text{By the congruence of normalization} \\
-                        &= [[ [mu]nf(iN) ]]
-      \end{aligned}$ \\
-    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
-      $\begin{aligned}[t]
-          [[ [mu] nf(iN) ]] &= [[ [mu] nf(∀pas.iM)]] \\
-                            &= [[ [mu] ∀pas'.nf(iM) ]]
-                            && \text {Where $[[ ord {pas} in nf(iM) = pas' ]]$}\\
-        \end{aligned}$ \\
-
-      $\begin{aligned}[t]
-         [[ nf([mu]iN) ]] &= [[ nf([mu] ∀pas.iM)]] \\
-                          &= [[ nf(∀pas.[mu]iM) ]]
-                          && \text{Assuming $[[{pas}]] \cap A = \emptyset$
-                             and $[[{pas}]] \cap B = \emptyset$}\\
-                          &= [[ ∀pbs.nf([mu]iM) ]]
-                          && \text {Where $[[ ord {pas} in nf([mu]iM) = pbs ]]$}\\
-                          &= [[ ∀pas'.nf([mu]iM) ]]
-                          && \text {As $[[pbs]] = [[pas']]$ (see below)}
-         \end{aligned}$ \\
-
-    Notice that $[[mu]]$ is free of collisions on $[[{pas}]]$ and $[[fv nf(iM)]]$ because
-    \begin{enumerate}[nolistsep]
-      \item[(i)] $[[mu]](A \cap [[{pas}]]) \cap [[fv nf(iM)]] = \emptyset \cap [[fv nf(iM)]] = \emptyset$ and
-      \item[(ii)] $[[mu]](A \cap [[fv nf(iM)]]) \cap [[{pas}]] \subseteq B \cap
-        [[{pas}]] = \emptyset$
-    \end{enumerate}
-    $\begin{aligned}[t]
-       [[pbs]] &= [[ ord {pas} in nf([mu]iM) ]] \\
-               &= [[ ord {pas} in [mu]nf(iM) ]]
-               && \text{By the induction hypothesis} \\
-               &= [[ ord {[mu]pas} in [mu]nf(iM) ]]
-               && \text{Since $[[{pas}]] \cap A = \emptyset$} \\
-               &= [[ [mu] ord {pas} in nf(iM) ]]
-               && \text{by \cref{lemma:distr-mu-ord}} \\
-               &= [[ ord {pas} in nf(iM) ]]
-               && \text{Since $[[{ord {pas} in nf(iM)}]] \cap A \subseteq [[{pas}]] \cap A = \emptyset$} \\
-               &= [[ pas' ]]
-     \end{aligned}$ \\
-
-     To show alpha-equivalence of 
-     $[[ [mu] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.nf([mu]iM) ]]$,
-     we can assume that $[[{pas'}]] \cap A = \emptyset$, and $[[{pas'}]] \cap B = \emptyset$.
-     Then $[[ [mu] ∀pas'.nf(iM) ]] = [[∀pas'.[mu]nf(iM)]] = [[ ∀pas'.nf([mu]iM)
-     ]]$, the latter follows from the induction hypothesis.
-
-   \item $[[iP]] = [[∃ nas.iQ]]$ \\
-     Same as for \cref{case:norm-subst-commute}.
-  \end{caseof}
-\end{proof}
-
 \begin{lemma}[Soundness of quantifier normalization]
@@
   \end{itemize}
-
 \end{lemma}
@@
 
+\begin{corollary}[Normalization preserves ordering]
+  \label{corollary:normalization-ord}
+  For any $[[varset]]$,
+  \begin{itemize}
+  \item[$-$] $[[ord varset in nf(uN)]] = [[ord varset in uM]]$
+  \item[$+$] $[[ord varset in nf(uP)]] = [[ord varset in uQ]]$
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  Immediately from \cref{lemma:ord-completeness,lemma:normalization-soundness}.
+\end{proof}
+
+\begin{lemma}[Distributivity of normalization over substitution]
+  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
+  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
+  taking free variables from $[[Γ2]]$.
+  Then
+  \begin{itemize}
+  \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
+  \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
+  \end{itemize}
+  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]
+  α⁻)]]$. %, and .
+\end{lemma}
+\begin{proof}
+  Mutual induction on $[[iN]]$ and $[[iP]]$.
+  \begin{caseof}
+    \item $[[iN]]$ = $[[na]]$ \\
+      \label{case:norm-subst-distr-var}
+      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.
+
+      $[[ [nf(σ)] nf(iN) ]] = [[ [nf(σ)] nf(na) ]] = [[ [nf(σ)] na ]] $.
+    \item $[[iP]]$ = $[[pa]]$ \\
+      Similar to \cref{case:norm-subst-distr-var}.
+   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
+     the required equality follows from the congruence of the normalization and
+     substitution, and the induction hypothesis.
+     For example, if $[[iN]] = [[iP → iM]]$ then \\
+     $\begin{aligned}[t]
+        [[nf([σ] iN)]] &= [[ nf([σ] (iP → iM)) ]] \\
+                        &= [[ nf([σ]iP → [σ]iM) ]]
+                        && \text{By the congruence of substitution} \\
+                        &= [[ nf([σ]iP) → nf([σ]iM) ]]
+                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
+                        &= [[ [nf(σ)]nf(iP) → [nf(σ)]nf(iM) ]]
+                        && \text{By the induction hypothesis} \\
+                        &= [[ [nf(σ)](nf(iP) → nf(iM)) ]]
+                        && \text{By the congruence of substitution} \\
+                        &= [[ [nf(σ)]nf(iP → iM) ]]
+                        && \text{By the congruence of normalization} \\
+                        &= [[ [nf(σ)]nf(iN) ]]
+      \end{aligned}$ \\
+    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
+      $\begin{aligned}[t]
+          [[ [nf(σ)] nf(iN) ]] &= [[ [nf(σ)] nf(∀pas.iM)]] \\
+                            &= [[ [nf(σ)] ∀pas'.nf(iM) ]]
+                            && \text {Where $[[pas']] = [[ ord {pas} in nf(iM)]]
+                               = [[ord {pas} in iM]]$
+                               (the latter is by
+                               \cref{corollary:normalization-ord})}\\
+        \end{aligned}$ \\
+
+      $\begin{aligned}[t]
+         [[ nf([σ]iN) ]] &= [[ nf([σ] ∀pas.iM)]] \\
+                          &= [[ nf(∀pas.[σ]iM) ]]
+                          && \text{Assuming $[[{pas} ∩ {Γ1}]] = \emptyset$
+                             and $[[{pas} ∩ {Γ2}]] = \emptyset$}\\
+                          &= [[ ∀pbs.nf([σ]iM) ]]
+                          && \text {Where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]
+                             = [[ord {pas} in [σ]iM]]$ (the latter is by
+                             \cref{corollary:normalization-ord})}\\
+                          &= [[ ∀pas'.nf([σ]iM) ]]
+                          && \text{By \cref{lemma:ord-sigma}, $[[pbs]] = [[pas']]$
+                             since $[[{pas}]]$ is disjoint with $[[Γ1]]$ and
+                             $[[Γ2]]$}\\
+                          &= [[ ∀pas'.[nf(σ)]nf(iM) ]]
+                          && \text {By the induction hypothesis}\\
+         \end{aligned}$ \\
+
+     To show alpha-equivalence of 
+     $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
+     we can assume that $[[{pas'} ∩ {Γ1}]] = \emptyset$, and $[[{pas'} ∩ {Γ2}]]
+     = \emptyset$.
+
+   \item $[[iP]] = [[∃ nas.iQ]]$ \\
+     Same as for \cref{case:norm-subst-commute}.
+  \end{caseof}
+\end{proof}
+
+\begin{corollary}[Commutativity of normalization and renaming]
+  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
+  Suppose that $\mu$ is a bijection between two sets of variables
+  $\mu : A \leftrightarrow B$. Then
+  \begin{itemize}
+    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
+    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  Immediately from \cref{lemma:norm-subst-distr},
+  after noticing that $[[nf(mu)]] = [[mu]]$.
+\end{proof}
+
 \begin{lemma}[Completeness of quantified normalization]
+  \label{lemma:normalization-completeness}
   Normalization returns the same representative for equivalent types.
@@
 \end{proof}
+
+\begin{lemma}[Idempotence of normalization]
+  \label{lemma:norm-idemp}
+  Normalization is idempotent
+  \begin{itemize}
+  \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
+  \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  By applying \cref{lemma:normalization-completeness} to \cref{lemma:normalization-soundness}.
+\end{proof}
+
+\begin{lemma}
+  \label{lemma:normal-after-subst}
+  The result of a substitution is normalized if and only if the initial type and
+  the substitution are normalized.
+
+  Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
+  $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
+  $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
+  \begin{itemize}
+  \item[$+$]
+    $[[ [σ]iP  ]] \text{ is normal} \iff
+    \begin{cases}
+      [[σ|fv(iP)]] &\text{is normal}\\
+      [[iP]]       &\text{is normal}\\
+    \end{cases} $
+  \item[$-$]
+    $[[ [σ]iN  ]] \text{is normal} \iff
+    \begin{cases}
+      [[σ|fv(iN)]] &\text{is normal}\\
+      [[iN]]       &\text{is normal}\\
+    \end{cases} $
+  \end{itemize}
+\end{lemma}
+
+\begin{proof}
+  Mutual induction on $[[Γ1 ⊢ iP]]$ and $[[Γ1 ⊢ iN]]$.
+  \begin{caseof}
+  \item $[[iN]] = [[na]]$\\
+    Then $[[iN]]$ is always normal, and
+    the normality of $[[σ|{na}]]$ by the definition means $[[ [σ]na ]]$ is normal.
+
+  \item $[[iN]] = [[iP → iM]]$\\
+    \label{case:normal-after-subst-arrow}
+    $
+    \begin{aligned}[t]
+      [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
+                                           && \text{by the substitution
+                                              congruence} \\
+                                           &\iff
+                                             \begin{cases}
+                                             [[ [σ]iP ]] &\text{is normal} \\
+                                             [[ [σ]iM ]] &\text{is normal} \\
+                                             \end{cases}
+                                           && \text{by congruence of normality \ilyam{lemma?}}\\
+                                           &\iff
+                                             \begin{cases}
+                                               [[ iP ]]       &\text{is normal} \\
+                                               [[ σ|fv(iP) ]] &\text{is normal} \\
+                                               [[ iM ]]       &\text{is normal} \\
+                                               [[ σ|fv(iM) ]] &\text{is normal} \\
+                                             \end{cases}
+                                           && \text{by the induction hypothesis}\\
+                                           &\iff
+                                             \begin{cases}
+                                               [[ iP → iM ]]  &\text{is normal} \\
+                                               [[ σ|fv(iP) ∪ fv(iM)]] &\text{is normal} \\
+                                             \end{cases}
+                                          \iff&&%
+                                             \begin{cases}
+                                               [[ iP → iM ]]  &\text{is normal} \\
+                                               [[ σ|fv(iP→iM)]] &\text{is normal} \\
+                                             \end{cases}
+    \end{aligned}
+    $
+  \item $[[iN]] = [[↑iP]]$\\
+    By congruence and the inductive hypothesis, similar to \cref{case:normal-after-subst-arrow}
+  \item $[[iN]] = [[∀pas.iM]]$\\
+    $
+    \begin{aligned}[t]
+      [[ [σ](∀pas.iM) ]] \text{ is normal} &\iff [[ (∀pas.[σ]iM) ]] \text{ is normal}
+                                           && \text{assuming $[[pas]] \cap [[Γ1]] = \emptyset$ and
+                                              $[[pas]] \cap [[Γ2]] = \emptyset$} \\
+                                           &\iff
+                                             \begin{cases}
+                                             [[ [σ]iM ]] \text{ is normal} \\
+                                             [[ord {pas} in [σ]iM = pas]] \\
+                                             \end{cases}
+                                           && \text{by the definition of normalization}\\
+                                           &\iff
+                                             \begin{cases}
+                                               [[ [σ]iM ]] \text{ is normal} \\
+                                               [[ord {pas} in iM = pas]] \\
+                                             \end{cases}
+                                           && \text{by \cref{lemma:ord-sigma}}\\
+                                           &\iff
+                                             \begin{cases}
+                                               [[ σ|fv(iM) ]] \text{ is normal} \\
+                                               [[ iM ]] \text{ is normal} \\
+                                               [[ord {pas} in iM = pas]] \\
+                                             \end{cases}
+                                           && \text{by the induction hypothesis}\\
+                                           &\iff
+                                             \begin{cases}
+                                               [[ σ|fv(∀pas.iM) ]] \text{ is normal} \\
+                                               [[ ∀pas.iM ]] \text{ is normal} \\
+                                             \end{cases}
+                                           &&
+                                              \begin{aligned}
+                                              &\text{since $[[fv(∀pas.iM) = fv(iM)]]$;}\\ &\text{by the definition of normalization}
+                                              \end{aligned}
+    \end{aligned}
+    $
+  \item $[[iP]] = \dots$\\
+    The positive cases are done in the same way as the negative ones.
+
+  \end{caseof}
+\end{proof}

### 2023-04-05 proofs/tex/equiv-lemmas.tex
Add soundness of ≈

\begin{lemma}[Type well-formedness is invariant under equivalence]
  \label{lemma:wf-equiv}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
  \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

\begin{corollary}[Normalization preserves well-formedness]
  \label{corollary:wf-nf}
  \hfill
  \begin{itemize}
  \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
  \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
\end{proof}

\begin{corollary}[Normalization preserves well-formedness of substitution]
  \label{corollary:wf-s-nf}
  \hfill \\
   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
\end{corollary}

\begin{lemma}[Soundness of equivalence]
  \label{lemma:equiv-soundness}
  Declarative equivalence implies mutual subtyping.
  \begin{itemize}
    \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
    \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by mutual induction on $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
  \begin{caseof}
    \item $[[a⁻ ≈ a⁻]]$\\
      Then $[[Γ ⊢ a⁻ ≤ a⁻]]$ by \ruleref{\ottdruleDOneNVarLabel},
      which immediately implies $[[Γ ⊢ a⁻ ≈ a⁻]]$ by \ruleref{\ottdruleDOneNDefLabel}.

    \item $[[↑iP ≈ ↑iQ]]$\\
      Then by inversion of \ruleref{\ottdruleDOneShiftULabel},
      $[[iP ≈ iQ]]$, and from the induction hypothesis, $[[Γ ⊢ iP ≈ iQ]]$,
      and (by symmetry) $[[Γ ⊢ iQ ≈ iP]]$.

      When \ruleref{\ottdruleDOneShiftULabel} is applied to $[[Γ ⊢ iP ≈ iQ]]$,
      it gives us $[[Γ ⊢ ↑iP ≤ ↑iQ]]$; when it is applied to $[[Γ ⊢ iQ ≈ iP]]$,
      we obtain $[[Γ ⊢ ↑iQ ≤ ↑iP]]$. Together, it implies $[[Γ ⊢ ↑iP ≈ ↑iQ]]$.

    \item $[[iP → iN ≈ iQ → iM]]$\\
      Then by inversion of \ruleref{\ottdruleDOneArrowLabel},
      $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$. By the induction hypothesis,
      $[[Γ ⊢ iP ≈ iQ]]$ and $[[Γ ⊢ iN ≈ iM]]$, which means by inversion:
      \begin{enumerate*}
        \item[(i)] [[Γ ⊢ iP ≥ iQ]],
        \item[(ii)] [[Γ ⊢ iQ ≥ iP]],
        \item[(iii)] [[Γ ⊢ iN ≤ iM]],
        \item[(iv)]  [[Γ ⊢ iM ≤ iN]].
      \end{enumerate*}
      Applying \ruleref{\ottdruleDOneArrowLabel} to (i) and (iii), we obtain
      $[[Γ ⊢ iP → iN ≤ iQ → iM]]$; applying it to (ii) and (iv), we have $[[Γ ⊢
      iQ → iM ≤ iP → iN]]$. Together, it implies $[[Γ ⊢ iP → iN ≈ iQ → iM]]$.
    \item $[[∀pas.iN ≈ ∀pbs.iM]]$\\
      Then by inversion, there exists bijection $[[mu : ({pbs} ∩ fv iM) ↔ ({pas}
      ∩ fv iN)]]$, such that $[[iN ≈ [mu] iM]]$. By the induction hypothesis,
      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$. From \cref{corollary:subst-pres-equiv} and
      the fact that $[[mu]]$ is bijective, we also have
      $[[Γ, pbs ⊢ [mu-1]iN ≈ iM]]$.

      Let us construct a subsitution $[[pas ⊢ iPs/pbs : pbs]]$ by
      extending $[[mu]]$ with arbitrary positive types on $[[{pbs} \ fv iM]]$.

      Notice that $[[ [mu]iM ]] = [[ [iPs/pbs]iM ]]$, and therefore,
      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$ implies $[[Γ, pas ⊢ [iPs/pbs]iM ≤ iN]]$. Then by
      \ruleref{\ottdruleDOneForallLabel}, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$.

      Analogously, we construct the substitution from $[[mu-1]]$, and use it to
      instantiate $[[pas]]$ in the application of
      \ruleref{\ottdruleDOneForallLabel} to infer $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$.

      This way, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$ and $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$
      gives us $[[Γ ⊢ ∀pbs.iM ≈ ∀pas.iN]]$.
  \end{caseof}
\end{proof}

\begin{lemma}[Completeness of equivalence]
  \label{lemma:equiv-completeness}
  Mutual subtyping implies declarative equivalence.
  \begin{itemize}
  \item[$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[iP ≈ iQ]]$,
  \item[$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[iN ≈ iM]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

### 2023-04-05 proofs/tex/subst-lemmas.tex
Add soundness of ≈

\begin{lemma}[Substitution strengthening]
  \label{lemma:subst-restr-fv}
  Restricting the substitution to the free variables of the
  substitution subject does not affect the result.
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
  \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

\begin{lemma}[Substitution preserves subtyping]
  \label{lemma:subst-pres-subt}

  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] if $[[Γ, Γ1 ⊢ iP]]$,~ $[[Γ, Γ1 ⊢ iQ]]$,~ and $[[Γ, Γ1 ⊢ iP ≥ iQ]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$
  \item[$-$] if $[[Γ, Γ1 ⊢ iN]]$,~ $[[Γ, Γ1 ⊢ iM]]$,~ and $[[Γ, Γ1 ⊢ iN ≤ iM]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iN ≤ [σ]iM]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

\begin{corollary}[Substitution preserves subtyping]
  \label{corollary:subst-pres-equiv}

  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] if $[[Γ, Γ1 ⊢ iP]]$,~ $[[Γ, Γ1 ⊢ iQ]]$,~ and $[[Γ, Γ1 ⊢ iP ≈ iQ]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iP ≈ [σ]iQ]]$
  \item[$-$] if $[[Γ, Γ1 ⊢ iN]]$,~ $[[Γ, Γ1 ⊢ iM]]$,~ and $[[Γ, Γ1 ⊢ iN ≈ iM]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iN ≈ [σ]iM]]$
  \end{itemize}
\end{lemma}

### 2023-04-05 proofs/tex/subtyping-lemmas.tex
Add soundness of ≈

\begin{lemma}[Free Variable Propagation]
  \label{lemma:fv-propagation}
  In the judgments of negative subtyping or positive supertyping,
  free variables propagate left-to-right. For a context $[[Γ]]$,
  \begin{itemize}
    \item $-$ if $[[Γ ⊢ iN ≤ iM]]$ then $[[fv(iN)]] \subseteq [[fv(iM)]]$
    \item $+$ if $[[Γ ⊢ iP ≥ iQ]]$ then $[[fv(iP)]] \subseteq [[fv(iQ)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iP ≥ iQ]]$.
  \begin{caseof}
  \item $[[G ⊢ a⁻ ≤ a⁻]]$\\
    It is self-evident that $[[{a⁻} ⊆ {a⁻}]]$.
  \item $[[G ⊢ ↑iP ≤ ↑iQ]]$
    From the inversion (and unfolding $[[G ⊢ iP ≈ iQ]]$ ), we have
    $[[G ⊢ iP ≥ iQ]]$. Then by the induction hypothesis,
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired inclusion
    inclusion holds, since $[[fv(↑iP)]] = [[fv(iP)]]$ and
    $[[fv(↑iQ)]] = [[fv(iQ)]]$.
  \item $[[G ⊢ iP → iN ≤ iQ → iM]]$
    The induction hypothesis applied to the premises gives:
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$ and
    $[[fv(iN)]] \subseteq [[fv(iM)]]$.
    Then $[[fv(iP → iN)]] = [[fv(iP) ∪ fv(iN)]] \subseteq
    [[fv(iQ) ∪ fv(iM)]] = [[fv(iQ → iM)]]$.

  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$\\
    $
    \begin{aligned}[t]
      [[fv ∀pas.iN ]] &\subseteq [[fv ([iPs/pas] iN) ]] ~\setminus~ [[{pbs}]] 
                      &&   \text{here $[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
                      &\subseteq [[fv iM]] ~\setminus~ [[{pbs}]]
                      &&   \text{by the induction hypothesis, } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
                      &\subseteq [[fv ∀pbs.iM]]
    \end{aligned}
    $
  \item The positive cases are symmetric.
  \end{caseof}
\end{proof}

### 2023-04-17 proofs/tex/equiv-lemmas.tex
WIP: equivalence lemmas

@@
       \begin{enumerate*}
-        \item[(i)] [[Γ ⊢ iP ≥ iQ]],
-        \item[(ii)] [[Γ ⊢ iQ ≥ iP]],
-        \item[(iii)] [[Γ ⊢ iN ≤ iM]],
-        \item[(iv)]  [[Γ ⊢ iM ≤ iN]].
+        \item[(i)] $[[Γ ⊢ iP ≥ iQ]]$,
+        \item[(ii)] $[[Γ ⊢ iQ ≥ iP]]$,
+        \item[(iii)] $[[Γ ⊢ iN ≤ iM]]$,
+        \item[(iv)]  $[[Γ ⊢ iM ≤ iN]]$.
       \end{enumerate*}
@@
       gives us $[[Γ ⊢ ∀pbs.iM ≈ ∀pas.iN]]$.
+
+    \item For the cases of the positive types, the proofs are symmetric.
   \end{caseof}
@@
 \end{proof}
+
+\begin{lemma}
+  Informally, this lemma says that if $[[Γ, Ord varset1 ⊢ [σ21]iP ≥ iQ]]$
+  and $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP]]$ holds for substitutions $[[σ12]]$
+  and $[[σ21]]$ then these substitutions are in fact mutually inverse bijections
+  between variables $[[varset1]]$ and $[[varset1]]$.
+
+  \begin{itemize}
+  \item[$+$]
+    For $[[Γ, Ord varset2 ⊢ iP]]$,~ $[[Γ, Ord varset1 ⊢ iQ]]$,~
+    $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,~
+    $[[Γ, Ord varset1 ⊢ σ21 : Ord varset2]]$,
+    suppose that:
+    \begin{enumerate}
+      \item $\{ [[coh]] \mid [[α±:coh]] \in [[varset1 ∩ fv iQ]] \}$ = $\{ [[coh]] \mid [[α±:coh]] \in [[varset2 ∩ fv iP]] \}$,
+      \item $[[Γ, Ord varset1 ⊢ [σ21]iP ≥ iQ]]$,
+      \item $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP]]$.
+    \end{enumerate}
+    Then there exists a bijection
+    $\mu : [[varset1 ∩ fv iQ]] \leftrightarrow [[varset2 ∩ fv iP]]$ such that:
+    \begin{enumerate}
+    \item $\mu$ preserves cohorts:
+      $\mu ([[α±:coh]]) = [[β±:coh]]$ ($[[α±:coh]]$ and $[[β±:coh]]$
+      have the same cohort label $\pm [[coh]]$).
+    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ   : Ord (varset1 ∩ fv iQ) ]]$ (the equivalence is pointwise)
+    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iP) ]]$ (the equivalence is pointwise)
+    \end{enumerate}
+
+  \item[$-$]
+    For $[[Γ, Ord varset2 ⊢ iN]]$,~ $[[Γ, Ord varset1 ⊢ iM]]$,~
+    $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,~
+    $[[Γ, Ord varset1 ⊢ σ21 : Ord varset2]]$,
+    suppose that:
+    \begin{enumerate}
+    \item $\{ [[coh]] \mid [[α±:coh]] \in [[varset1 ∩ fv iM]] \}$ = $\{ [[coh]] \mid [[β±:coh]] \in [[varset2 ∩ fv iN]] \}$,
+    \item $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$,
+    \item $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$.
+    \end{enumerate}
+    Then there exists a bijection
+    $\mu : [[varset1 ∩ fv iM]] \leftrightarrow [[varset2 ∩ fv iN]]$ such that:
+    \begin{enumerate}
+    \item $\mu$ preserves cohorts:
+      $\mu ([[α±:coh]]) = [[β±:coh]]$ ($[[α±:coh]]$ and $[[β±:coh]]$
+      have the same cohort label $\pm [[coh]]$).
+    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ : Ord (varset1 ∩ fv iM) ]]$ (the equivalence is pointwise)
+    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iN)]]$ (the equivalence is pointwise)
+    \end{enumerate}
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Mutual induction on the pair of sizes of inference trees: $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$
+  and $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ (or the corresponding trees in the
+  positive case).
+
+  \begin{caseof}
+  \item $[[iN]] = [[∀pds.β⁻:coh]]$, for $[[β⁻:coh]] \in
+    [[varset2]]$ and possibly empty $[[pds]]$\\
+    \label{case:double-subtyping-subst-happened}
+    Then by \cref{todo} \ilyam{lemma} $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ ∀pds.β⁻:coh]]$ means that $[[ [σ12]iM ]] =
+    [[∀pcs.β⁻:coh]]$, which (since $[[β⁻:coh]] \notin [[fv iM]]$) is only possible
+    when $[[iM]] = [[∀pcs'.α⁻:coh']]$ (where $[[α⁻:coh']] \in [[varset1]]$ and $[[pcs']]$ is possibly empty), and
+    $[[σ12]] ([[α⁻:coh']]) = [[∀pcs''.β⁻:coh]]$.
+    Notice that by the first condition, the sets of cohort labels of $[[varset1
+    ∩ fv iM]]$ and $[[varset2 ∩ fv iN]]$ must be equal, which means $[[coh]] = [[coh']]$.
+    Also notice that $[[Γ, Ord varset2 ⊢ ∀pcs''.β⁻:coh ≈ β⁻:coh]]$.
+
+    Then $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ becomes
+    $[[Γ, Ord varset1 ⊢ [σ21]∀pds.β⁻:coh ≤ ∀pcs'.α⁻:coh]]$,
+    which by \cref{todo} \ilyam{lemma} implies that $[[ [σ21] ∀pds.β⁻:coh]] =
+    [[∀pds'.α⁻:coh]]$, and thus,
+    $[[σ21]] ([[β⁻:coh]]) = [[∀pds''.α⁻:coh]]$.
+    Notice that $[[Γ, Ord varset1 ⊢ ∀pds''.α⁻:coh ≈ α⁻:coh]]$.
+
+    This way, we can take $\mu = [[α⁻:coh]] \mapsto [[β⁻:coh]] $, which by
+    construction is a bijection preserving cohorts. Moreover,
+    $[[Γ, Ord varset2 ⊢ ∀pcs''.β⁻:coh ≈ β⁻:coh]]$ means
+    $[[Γ, Ord varset2 ⊢ [σ12] α⁻:coh ≈ [Sub μ] α⁻:coh]]$ implying
+    $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ : Ord (varset1 ∩ fv iM) ]]$;
+    and
+    $[[Γ, Ord varset1 ⊢ ∀pds''.α⁻:coh ≈ α⁻:coh]]$ means
+    $[[Γ, Ord varset1 ⊢ [σ21] β⁻:coh ≈ [Sub μ-1] β⁻:coh]]$ implying
+    $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iN)]]$.
+
+  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
+    \ruleref{\ottdruleDOneNVarLabel}, i.e. $[[ [σ21]iN ]] = [[iM]] = [[γ⁻]]$.\\
+    Then the case when $[[γ⁻]] \in [[varset2]]$, has been covered by
+    \cref{case:double-subtyping-subst-happened}, so we assume
+    that $[[γ⁻]] \notin [[varset2]]$, and thus, $[[iN]] = [[γ⁻]]$.
+
+    Notice that $[[γ⁻]] \notin [[varset1]]$ because otherwise,
+    $[[varset1 ∩ fv iN]] \neq \emptyset$,
+    which would contradict with $[[Γ, Ord varset2 ⊢ iN]]$.
+
+    Then $[[fv iN ∩ varset2]] = [[{γ⁻} ∩ varset2]] = \emptyset$ and $[[fv iM ∩
+    varset1]] = [[{γ⁻} ∩ varset1]] = \emptyset$.
+    Hence, we take the empty $\mu : \emptyset \leftrightarrow \emptyset$, which
+    vacuously satisfies the required properties.
+
+  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
+    \ruleref{\ottdruleDOneShiftULabel}, i.e. $[[ [σ21]iN ]] = [[↑iP]]$, $[[ iM
+    ]] = [[↑iQ]]$, and $[[Γ, Ord varset1 ⊢ iP ≈ iQ]]$\\
+    Since $[[iN]]$ is not a variable from the domain of $[[σ21]]$ (which has
+    been covered by \cref{case:double-subtyping-subst-happened}),
+    the substitution applied to $[[iN]]$ must preserve its outer shape.
+    Specifically, $[[ [σ21]iN ]] = [[↑iP]]$ means
+    $[[ [σ21]iN ]] = [[ [σ21]↑iP']] = [[ ↑[σ21]iP']] =[[↑iP]]$,
+    i.e. $[[ iN ]] = [[↑iP']]$ and
+    $[[ [σ21]iP']] = [[iP]]$. In particular,
+    $[[Γ, Ord varset1 ⊢ iP ≈ iQ]]$ implies
+    $[[Γ, Ord varset1 ⊢ iP ≥ iQ]]$, i.e.
+    $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$.
+
+    In addition, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ becomes
+    $[[Γ, Ord varset2 ⊢ ↑[σ12]iQ ≤ ↑iP']]$, which is only inferable by
+    \ruleref{\ottdruleDOneShiftULabel}, meaning that
+    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≈ iP']]$, and in particular,
+    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$.
+
+    Notice that the tree inferring $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$ is a
+    proper subtree of $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$. Analogously,
+    $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$ is a proper subtree of
+    $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$. This way, we apply the induction
+    hypothesis to $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$ and $[[Γ, Ord varset2 ⊢
+    [σ12]iQ ≥ iP']]$ (notice that $[[varset1]]$, $[[varset2]]$, and the sets of free variables of
+    the types did not change) and obtain exactly what we aimed.
+
+  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
+    \ruleref{\ottdruleDOneForallLabel}, i.e. $[[ [σ21]iN ]] = [[∀pcs.iN']]$,
+    $[[iM]] = [[∀pds.iM']]$, and $[[Γ, Ord varset1, pds ⊢ [iPs/pcs]iN' ≤ iM']]$
+    for  $[[Γ, Ord varset1, pds ⊢ iPi]]$ \\
+
+    Since $[[iN]]$ does not have the shape of $[[∀pds.β⁻:coh]]$, for $[[β⁻:coh]]
+    \in [[varset2]]$ (which has been covered by
+    \cref{case:double-subtyping-subst-happened}), the substitution applied to
+    $[[iN]]$ must preserve its outer shape. Specifically,
+    $[[ [σ21]iN ]] = [[∀pcs.iN']]$
+    means that $[[ iN ]]$ ``starts with'' $\forall[[pcs]]$, i.e.
+    $[[ [σ21]iN ]] = [[ [σ21]∀pcs.iN'' ]] = [[ ∀pcs.[σ21]iN'' ]] = [[ ∀pcs.iN'
+    ]]$, where $[[iN]] = [[∀pcs.iN'']]$ and $[[ [σ21]iN'']] = [[iN']]$.
+
+    This way, $[[Γ, Ord varset1, pds ⊢ [iPs/pcs]iN' ≤ iM']]$ becomes
+    $[[Γ, Ord varset1, pds ⊢ [iPs/pcs][σ21]iN'' ≤ iM']]$.
+    Notice that the tree inferring this judgment is a proper subtree of
+    $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$.
+
+    On the other hand, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$
+    becomes $[[Γ, Ord varset2 ⊢ ∀pds.[σ12]iM' ≤ ∀pcs.iN'']]$
+    (where either $[[pds]]$ or $[[pcs]]$ is non-empty),
+    which is only inferable by \ruleref{\ottdruleDOneForallLabel},
+    meaning that
+    $[[Γ, Ord varset2, pcs ⊢ [iQs/pds][σ12]iM' ≤ iN'']]$ for some
+    $[[Γ, Ord varset2, pcs ⊢ iQi]]$.
+    Notice that the tree inferring this judgment is a proper subtree of
+    $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$.
+
+    Let us label $[[pcs]]$ and $[[pds]]$ with a sufficiently large cohort label
+    $[[mcoh]]$ such that $[[mcoh]] > [[coh]]$ for any $[[α⁺:coh]] \in [[varset1 ∪
+    varset2]]$. Then we merge $[[iQs/δ⁺:mcoh*]]$ and $[[σ12]]$ denoting the resulting
+    substitution as $[[σ12']]$ $([[(Γ, Ord varset2, γ⁺:mcoh*) ⊢ σ12' : (Ord varset1, δ⁺:mcoh*)]])$.
+
+    What do we mean by merging? The codomains of $[[iQs/δ⁺:mcoh*]]$ and
+    $[[σ12]]$ are $[[Γ, Ord varset2,  γ⁺:mcoh*]]$ and $[[Γ, Ord varset2]]$,
+    respectively, which are disjoint with both of their domains
+    ($[[δ⁺:mcoh*]]$ and $[[Ord varset1]]$, respectively). In turn, the domains
+    of $[[iQs/δ⁺:mcoh*]]$ and $[[σ12]]$ are themselves mutually disjoint.
+    This way, by merging we mean
+    $[[iQs/δ⁺:mcoh*]] \cup [[σ12]]$ (relationally) or
+    $[[iQs/δ⁺:mcoh* ○ σ12]]$ or
+    $[[σ12 ○ iQs/δ⁺:mcoh*]]$, since these three substitutions are equal.
+
+    Analogously, we merge $[[iPs/γ⁺:mcoh*]]$ and $[[σ21]]$ denoting the
+    resulting substitution as $[[σ21']]$ $([[(Γ, Ord varset1, δ⁺:mcoh*) ⊢ σ21' :
+    (Ord varset2, γ⁺:mcoh*)]])$.
+
+    We wish to apply the induction hypothesis to
+    $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$ and
+    $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12']iM' ≤ iN'']]$.
+    To do so, we need to show that the cohort labels of $[[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv
+    iN'']]$ coincide with those of $[[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]$ (as sets).
+    \begin{assertion*}
+      The set of cohorts of $[[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$
+      is equal to the set of cohorts of $[[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]$.
+    \end{assertion*}
+
+    \begin{proof}
+      $
+      \begin{aligned}[t]
+        [[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']] &= [[(varset2 ∪ {γ⁺:mcoh*}) ∩ (fv
+                                               iN ∪ ({γ⁺:mcoh*} ∩ fv iN''))]]\\
+                                             &= [[varset2 ∩ fv iN ∪ {γ⁺:mcoh*} ∩ fv iN'']]
+                                             && \text{because
+                                                $[[varset2]]$ and $[[fv iN]]$ are disjoint with $[[γ⁺:mcoh*]]$}\\
+        [[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]  &= [[varset1 ∩ fv iM ∪ {δ⁺:mcoh*} ∩ fv iM']]
+                                             && \text{analogously}\\
+      \end{aligned}\\
+      $
+      Since the cohort labels of $[[varset2 ∩ fv iN]]$ coincide with those of
+      $[[varset1 ∩ fv iM]]$ by the assumption, it suffices to prove that the cohort labels of
+      $[[{γ⁺:mcoh*} ∩ fv iN'']]$ and of $[[{δ⁺:mcoh*} ∩ fv iM']]$ coincide.
+      Note that these sets have the same
+      cohorts labels $[[mcoh]]$, i.e. it is required to show that these sets are
+      either both empty or both non-empty, i.e.
+      $[[{γ⁺:mcoh*} ∩ fv iN'']] = \emptyset \iff [[{δ⁺:mcoh*} ∩ fv iM']] = \emptyset$
+      \begin{itemize}
+      \item [$(\Leftarrow)$] Suppose that $[[{δ⁺:mcoh*} ∩ fv iM']] = \emptyset$.
+        Then by \cref{todo}, since
+        $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$, $[[fv [σ21']iN'']]
+        \subseteq [[fv iM']]$, implying that $[[{δ⁺:mcoh*} ∩ fv [σ21']iN'']] =
+        \emptyset$,
+        and by context strengthening \cref{todo}, $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'']]$ reduces to
+        $[[Γ, Ord varset1 ⊢ [σ21']iN'']]$.
+
+        Let us restrict $[[σ21']]$ to the set of free variables of $[[fv iN'']]$.
+        Then the domain and codomain of $[[σ21']]$ are the following:
+        $[[Γ, Ord varset1 ⊢ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩ fv iN]]$.
+        By \cref{todo}, $[[ [σ21']iN'' ]] = [[ [σ21'|fv iN'']iN'' ]]$,
+        and then
+        $[[Γ, Ord varset1 ⊢ [σ21']iN'' ≤ iM']]$, can be rewritten as
+        $[[Γ, Ord varset1 ⊢ [σ21'|fv iN'']iN'' ≤ iM']]$.
+
+        Let us apply substitution $[[σ12]]$ ($[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$)
+        to both sides of this judgment to obtain: $[[Γ, Ord varset2 ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ [σ12]iM']]$.
+        Using the transitivity \cref{todo}, let us compose this subtyping
+        judgment with $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12]iM' ≤ iN'']]$ to
+        form
+        $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ iN'']]$.
+
+        What is the codomain of $[[σ12 ○ σ21'|fv iN'']]$?
+        By composing
+        $[[Γ, Ord varset1 ⊢ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩ fv iN]]$
+        and
+        $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,
+        we have
+        $[[Γ, Ord varset2 ⊢ σ12 ○ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩
+        fv iN]]$, i.e. $[[σ12 ○ σ21'|fv iN'']]$ is a substitution contracting
+        $[[γ⁺:mcoh*]]$.
+        Then by \cref{todo},
+        $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ iN'']]$
+        implies that $[[{γ⁺:mcoh*} ∩ fv iN'']] = \emptyset$.
+        \item [$(\Rightarrow)$]
+          Analogous to the previous case.
+      \end{itemize}
+    \end{proof}
+    This way, we can apply the induction hypothesis to
+    $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$ and
+    $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12']iM' ≤ iN'']]$,
+    and obtain
+    $\mu : [[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']] \leftrightarrow [[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$ such that:
+    \begin{enumerate}
+    \item $\mu$ preserves cohorts,
+    \item $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ σ12' ≈ Sub μ : Ord (varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM' ]]$, and
+    \item $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ σ21' ≈ Sub μ-1 : Ord (varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$
+    \end{enumerate}
+
+    Let us decompose $\mu$ into the union of two bijections: $\mu_1 \sqcup
+    \mu_2$, where $\mu_1$ is defined on the variables whose cohort labels are less
+    than $[[mcoh]]$, and $\mu_2$ is defined on the variables whose cohort labels
+    are equal to $[[mcoh]]$. Notice that since $\mu$ preserves cohorts,
+    this union is disjoint: the \emph{range} of $\mu_1$ is the variables with cohorts
+    $< [[mcoh]]$ and the \emph{range} of $\mu_2$ is the variables with cohorts
+    $= [[mcoh]]$.
+
+    Recalling how $[[mcoh]]$ is chosen, notice that the signatures of $\mu_1$ and
+    $\mu_2$ are the following:
+    \begin{enumerate}
+      \item $\mu_1 : [[varset1 ∩ fv iM']] \leftrightarrow [[varset2 ∩ fv iN'']]$ and
+      \item $\mu_2 : [[{δ⁺:mcoh*} ∩ fv iM']] \leftrightarrow [[{γ⁺:mcoh*} ∩ fv iN'']]$.
+    \end{enumerate}
+
+    This way, $\mu_1$ is a cohort-preserving bijection with the required
+    signature, and what is left to show is the equivalences $[[μ1]] \eqDOne
+    [[σ12]]$, and $[[μ1-1]] \eqDOne [[σ21]]$:
+    \begin{enumerate}[nosep]
+    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ1 : Ord (varset1 ∩ fv iM) ]]$\\
+      Let us take an arbitrary $[[α±]] \in [[varset1 ∩ fv iM]]$.\\
+      $
+      \begin{aligned}
+        [[ [σ12] α± ]] &= [[ [σ12 ○ iQs/δ⁺:mcoh*] α± ]]
+                       && \text{because $[[α±]] \neq [[δi⁺:mcoh]]$} \\
+                       &= [[ [σ12'] α± ]]
+                       && \text{by the definition of $[[σ12']]$} \\
+                       &\eqDOne [[ [Sub μ] α± ]]
+                       && \text{since $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ σ12' ≈ Sub μ : Ord ((varset1 ∪ {δ⁺:mcoh*}) ∩ fv iN) ]]$} \\
+                       &= [[ [Sub μ1] α± ]]
+                       && \text{since $[[μ1]] = [[Sub μ | varset1 ∩ fv iM']]$
+                          and  $[[α±]] \in [[varset1 ∩ fv iM]] \subseteq [[varset1 ∩ fv iM']]$} \\
+      \end{aligned}
+      $\\
+      This way, $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12] α± ≈ [Sub μ1] α±  ]]$, which,
+      considering that the codomains of $[[σ12]]$ and $[[μ1]]$ are in $[[Γ, Ord
+      varset2]]$, can be strengthen to $[[Γ, Ord varset2 ⊢ [σ12] α± ≈ [Sub μ1] α± ]]$.
+
+    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ1-1 : Ord (varset2 ∩ fv iN)]]$ is proved analogously.
+
+    \end{enumerate}
+
+  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
+    \ruleref{\ottdruleDOneArrowLabel}, i.e. $[[ [σ21]iN ]] = [[iP → iN']]$, and
+    $[[iM]] = [[iQ → iM']]$; then by inverting this rule,
+    $[[Γ, Ord varset1 ⊢ iP ≥ iQ]]$ and $[[Γ, Ord varset1 ⊢ iN' ≤ iM']]$.
+
+    $[[ [σ21]iN ]] = [[iP → iN']]$ means that either $[[iN]]$ is a variable
+    from the domain of $[[σ21]]$ (which has been covered by
+    \cref{case:double-subtyping-subst-happened})
+    or $[[ [σ21]iN ]] = [[ [σ21] (iP' → iN'')]] = [[ [σ21] iP' → [σ21] iN'' ]] =
+    [[iP → iN']]$, where
+    $[[ iN ]] = [[iP' → iN'']]$, $[[ [σ21] iP' ]] = [[iP]]$, and $[[ [σ21] iN'' ]] = [[iN']]$.
+
+    This way, $[[Γ ⊢ iP ≥ iQ]]$ and $[[Γ ⊢ iN' ≤ iM']]$ can be rewritten as
+    $[[Γ ⊢ [σ21]iP' ≥ iQ]]$ and $[[Γ ⊢ [σ21] iN'' ≤ iM']]$.
+
+    In addition, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ becomes
+    $[[Γ, Ord varset2 ⊢ [σ12](iQ → iM') ≤ iP' → iN'']]$,
+    implying by inversion
+    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$ and
+    $[[Γ, Ord varset2 ⊢ [σ12]iM' ≤ iN'']]$.
+
+  \end{caseof}
+\end{proof}

### 2023-04-20 proofs/tex/au-lemmas.tex
Add equivalence soundness and completeness proofs

\begin{lemma}[Soundness of the anti-unification algorithm]
  \label{lemma:au-soundness}
\end{lemma}

\begin{lemma}[Completeness of the anti-unification algorithm]
  \label{lemma:au-completeness}
\end{lemma}

\begin{lemma}[Initiality of the anti-unification algorithm]
  \label{lemma:au-initiality}
\end{lemma}

### 2023-04-20 proofs/tex/decl-subtyping-lemmas.tex
Add equivalence soundness and completeness proofs

\begin{lemma}[Free Variable Propagation]
  \label{lemma:fv-propagation}
  In the judgments of negative subtyping or positive supertyping,
  free variables propagate left-to-right. For a context $[[Γ]]$,
  \begin{itemize}
    \item $-$ if $[[Γ ⊢ iN ≤ iM]]$ then $[[fv(iN)]] \subseteq [[fv(iM)]]$
    \item $+$ if $[[Γ ⊢ iP ≥ iQ]]$ then $[[fv(iP)]] \subseteq [[fv(iQ)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iP ≥ iQ]]$.
  \begin{caseof}
  \item $[[G ⊢ a⁻ ≤ a⁻]]$\\
    It is self-evident that $[[{a⁻} ⊆ {a⁻}]]$.
  \item $[[G ⊢ ↑iP ≤ ↑iQ]]$
    From the inversion (and unfolding $[[G ⊢ iP ≈ iQ]]$ ), we have
    $[[G ⊢ iP ≥ iQ]]$. Then by the induction hypothesis,
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired inclusion
    inclusion holds, since $[[fv(↑iP)]] = [[fv(iP)]]$ and
    $[[fv(↑iQ)]] = [[fv(iQ)]]$.
  \item $[[G ⊢ iP → iN ≤ iQ → iM]]$
    The induction hypothesis applied to the premises gives:
    $[[fv(iP)]] \subseteq [[fv(iQ)]]$ and
    $[[fv(iN)]] \subseteq [[fv(iM)]]$.
    Then $[[fv(iP → iN)]] = [[fv(iP) ∪ fv(iN)]] \subseteq
    [[fv(iQ) ∪ fv(iM)]] = [[fv(iQ → iM)]]$.

  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$\\
    $
    \begin{aligned}[t]
      [[fv ∀pas.iN ]] &\subseteq [[fv ([iPs/pas] iN) ]] ~\setminus~ [[{pbs}]] 
                      &&   \text{here $[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
                      &\subseteq [[fv iM]] ~\setminus~ [[{pbs}]]
                      &&   \text{by the induction hypothesis, } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
                      &\subseteq [[fv ∀pbs.iM]]
    \end{aligned}
    $
  \item The positive cases are symmetric.
  \end{caseof}
\end{proof}

\begin{corollary}[Free Variables of mutual subtypes] \label{corollary:fv-mut-sub}
  \hfill
  \begin{itemize}
    \item [$-$] If $[[Γ ⊢ iN ≈ iM]]$ then $[[fv iN]] = [[fv iM]]$, 
    \item [$+$] If $[[Γ ⊢ iP ≈ iQ]]$ then $[[fv iP]] = [[fv iQ]]$
  \end{itemize}
\end{corollary}

\begin{lemma}[Subtypes and supertypes of a variable]
  \label{lemma:var-subt}
  Assuming $[[Γ ⊢  α⁻]]$, $[[Γ ⊢ α⁺]]$, $[[Γ ⊢ iN]]$, and $[[Γ ⊢ iP]]$,
  \begin{itemize}
  \item[$+$] if $[[Γ ⊢ iP ≥ α⁺]]$ or $[[Γ ⊢ α⁺ ≥ iP ]]$ then $[[iP]] = [[∃nas.α⁺]]$ (for some potentially empty $[[nas]]$)
  \item[$-$] if $[[Γ ⊢ iN ≤ α⁻]]$ or $[[Γ ⊢ α⁻ ≤ iN ]]$ then $[[iN]] = [[∀pas.α⁻]]$ (for some potentially empty $[[pas]]$)
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove by induction on the tree
  inferring $[[Γ ⊢ iP ≥ α⁺]]$ or $[[Γ ⊢ α⁺ ≥ iP ]]$ or
  or $[[Γ ⊢ iN ≤ α⁻]]$ or $[[Γ ⊢ α⁻ ≤ iN ]]$.

  Let us consider which of these judgments the tree is inferring.
  \begin{caseof}
  \item $[[Γ ⊢ iP ≥ α⁺]]$\\
    If the size of the inference tree is $1$ then the only rule that can infer
    it is \ruleref{\ottdruleDOnePVarLabel}, which
    implies that $[[iP = α⁺]]$.

    If the size of the inference tree is $>1$ then the last rule inferring
    it must be \ruleref{\ottdruleDOneExistsLabel}. By inverting this rule,
    $[[iP = ∃nas.iP']]$ where $[[iP']]$ does not start with $\exists$ and
    $[[G ⊢ [iNs/nas] iP' ≥ α⁺]]$ for some $[[G, nbs ⊢ iNi]]$.

    By the induction hypothesis, $[[ [iNs/nas] iP' = ∃nbs.α⁺]]$.
    Notice that $[[iP']]$ must be a variable, because
    $[[iP']]$ does not start with $\exists$, nor does it start with
    $\uparrow$ (otherwise, $[[ [iNs/nas] iP' ]]$ would also
    started with $\uparrow$ and would not be equal to $[[∃nbs.α⁺]]$).
    Since $[[iP']]$ is a \emph{positive} variable, $[[ [iNs/nas] iP' = iP']]$,
    and then $[[iP' = ∃nbs.α⁺]]$ means that $[[iP' = α⁺]]$.
    This way, $[[iP]] = [[∃nas.iP']] = [[∃nas.α⁺]]$, as required.

  \item $[[Γ ⊢ α⁺ ≥ iP]]$\\
    If the size of the inference tree is $1$ then the only rule that can infer
    it is \ruleref{\ottdruleDOnePVarLabel}, which
    implies that $[[iP = α⁺]]$.

    If the size of the inference tree is $>1$ then the last rule inferring
    it must be \ruleref{\ottdruleDOneExistsLabel}. By inverting this rule,
    $[[iP = ∃nbs.iQ]]$ where and $[[G, nbs ⊢ α⁺ ≥ iQ]]$.

    By the induction hypothesis, $[[iQ = ∃nbs'.α⁺]]$.
    This way, $[[iP]] = [[∃nbs.iQ]] = [[∃nbs.∃nbs'.α⁺]]$, as required.

  \item The negative cases ($[[Γ ⊢ iN ≤ α⁻]]$ and $[[Γ ⊢ α⁻ ≤ iN ]]$)
    are proved analogously.
  \end{caseof}
\end{proof}

\begin{corollary}[Variables have no proper subtypes and supertypes]
  \label{corollary:vars-no-proper-subtypes}
  Assuming that all mentioned types are well-formed in $[[Γ]]$,
  \begin{align*}
    [[Γ ⊢ iP ≥ α⁺]] ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
    [[Γ ⊢ α⁺≥ iP]]  ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
    [[Γ ⊢ iN ≤ α⁻]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
    [[Γ ⊢ α⁻ ≤ iN]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
  \end{align*}
\end{corollary}
\begin{proof}
  Notice that $[[Γ ⊢ ∃nas.α⁺ ≈ α⁺]]$ and $[[∃nas.α⁺ ≈ α⁺]]$ and apply
  \cref{lemma:var-subt}.
  \ilyam{fix}
\end{proof}

### 2023-04-20 proofs/tex/equiv-lemmas.tex
Add equivalence soundness and completeness proofs

--- 2023-04-05 proofs/tex/equiv-lemmas.tex
@@
 
-\begin{lemma}[Completeness of equivalence]
-  \label{lemma:equiv-completeness}
+\begin{lemma}[Subtyping induced by disjoint substitutions]
+  \label{lemma:subt-ind-disj-subst}
+  If two disjoint substitutions induce subtyping, they are degenerate (so is the
+  subtyping).
+  Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
+  where $[[{Γi} ⊆ {Γ}]]$ and $[[{Γ1} ∩ {Γ2}= ∅]]$. Then
+  \begin{itemize}
+  \item[$-$] assuming $[[Γ ⊢ iN]]$,~
+    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id : Ord fv iN]]$
+  \item[$+$] assuming $[[Γ ⊢ iP]]$,~
+    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id : Ord fv iP]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Proof by induciton on $[[Γ ⊢ iN]]$ (and mutually on $[[Γ ⊢ iP]]$).
+  \begin{caseof}
+    \item $[[iN]] = [[α⁻]]$\\
+      Then $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]α⁻ ≤ [σ2]α⁻]]$.
+      Let us consider the following cases:
+      \begin{caseof}
+      \item $[[α⁻ ∉ {Γ1}]]$ and $[[α⁻ ∉ {Γ2}]]$ \label{case:var-not-in-ctxts}\\
+        Then $[[Γ ⊢ σi ≈ id : α⁻]]$ holds immediately,
+        since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
+        $[[Γ ⊢ α⁻ ≈ α⁻]]$.
+      \item $[[α⁻ ∊ {Γ1}]]$ and $[[α⁻ ∊ {Γ2}]]$\\
+        This case is not possible by assumption: $[[{Γ1} ∩ {Γ2}= ∅]]$.
+      \item $[[α⁻ ∊ {Γ1}]]$ and $[[α⁻ ∉ {Γ2}]]$\\
+        Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
+        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
+        and hence, $[[Γ ⊢ σ1 ≈ id : α⁻]]$.
+
+        $[[Γ ⊢ σ2 ≈ id : α⁻]]$ holds since $[[ [σ2]α⁻ ]] = [[α⁻]]$,
+        similarly to \cref{case:var-not-in-ctxts}.
+
+      \item $[[α⁻ ∉ {Γ1}]]$ and $[[α⁻ ∊ {Γ2}]]$\\
+        Then we have $[[Γ ⊢ α⁻ ≤ [σ2]α⁻]]$,
+        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ α⁻ ≈ [σ2]α⁻]]$,
+        and hence, $[[Γ ⊢ σ2 ≈ id : α⁻]]$.
+
+        $[[Γ ⊢ σ1 ≈ id : α⁻]]$ holds since $[[ [σ1]α⁻ ]] = [[α⁻]]$,
+        similarly to \cref{case:var-not-in-ctxts}.
+      \end{caseof}
+  \item $[[iN]] = [[∀pas.iM]]$\\
+    Then by inversion, $[[Γ, pas ⊢ iM]]$.
+    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]∀pas.iM ≤ [σ2]∀pas.iM]]$.
+    By the congruence of substitution and by the inversion of
+    \ruleref{\ottdruleDOneForallLabel}, $[[Γ, pas ⊢ [iQs/pas][σ1]iM ≤ [σ2]iM]]$,
+    where $[[Γ, pas ⊢ iQi]]$.
+    Let us denote the (Kleisli) composition of $[[σ1]]$ and $[[iQs/pas]]$ as
+    $[[σ1']]$, noting that $[[Γ, pas ⊢ σ1' : Γ1, pas]]$,
+    and $[[{Γ1, pas} ∩ {Γ2} = ∅]]$.
+
+    Let us apply the induction hypothesis to $[[iM]]$ and the
+    substitutions $[[σ1']]$ and $[[σ2]]$ with
+    $[[Γ, pas ⊢ [σ1']iM ≤ [σ2]iM]]$ to obtain:
+    \begin{align}
+      [[Γ, pas ⊢ σ1' ≈ id : Ord fv iM]] \label{fact:subs-proper-sub:forall-ih}\\
+      [[Γ, pas ⊢ σ2 ≈ id : Ord fv iM]]  \label{fact:subs-proper-sub:forall-ih2}
+    \end{align}
+
+    Then $[[Γ ⊢ σ2 ≈ id : Ord fv ∀pas.iM]]$ holds by strengthening of
+    \ref{fact:subs-proper-sub:forall-ih2}:
+    for any $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$,
+    $[[Γ, pas ⊢ [σ2]β± ≈ β±]]$ is strengthened to $[[Γ ⊢ [σ2]β± ≈ β±]]$, because
+    $[[fv [σ2]β±]] = [[fv β±]] = \{[[β±]]\} \subseteq [[{Γ}]]$.
+
+    To show that $[[Γ ⊢ σ1 ≈ id : Ord fv ∀pas.iM]]$, let us take an arbitrary
+    $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.
+
+    $
+    \begin{aligned}[t]
+      [[β±]] &= [[ [id]β± ]]
+             && \text{by definition of $[[id]]$}\\
+             &\eqDOne [[ [σ1']β± ]]
+             && \text{by \ref{fact:subs-proper-sub:forall-ih}}\\
+             &= [[ [iQs/pas][σ1]β±]]
+             && \text{by definition of $[[σ1']]$}\\
+             &= [[ [σ1]β± ]]
+             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ {Γ}]] = \emptyset$}
+    \end{aligned}
+    $\\
+    This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
+    $[[Γ ⊢ σ1 ≈ id : Ord fv ∀pas.iM]]$.
+
+  \item $[[iN]] = [[iP → iM]]$\\
+    Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
+    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as
+    $[[Γ ⊢ [σ1](iP → iM) ≤ [σ2](iP → iM)]]$,
+    then by congruence of substitution,
+    $[[Γ ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$,
+    then by inversion
+    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
+    and
+    $[[Γ ⊢ [σ1]iM ≤ [σ2]iM]]$.
+
+    Applying the induction hypothesis to $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
+    and to $[[Γ ⊢ [σ1]iM ≤ [σ2]iM]]$, we obtain (respectively):
+    \begin{align}
+      &[[Γ ⊢ σi ≈ id : Ord fv iP]] \label{fact:subs-proper-sub:arrow-ih1}\\
+      &[[Γ ⊢ σi ≈ id : Ord fv iM]] \label{fact:subs-proper-sub:arrow-ih2}
+    \end{align}
+
+    Noting that $[[fv (iP → iM)]] = [[fv iP ∪ fv iM]]$,
+    we combine
+    \cref{fact:subs-proper-sub:arrow-ih1,fact:subs-proper-sub:arrow-ih2}
+    to conclude:
+    $[[Γ ⊢ σi ≈ id : Ord fv (iP → iM)]]$.
+
+  \item $[[iN]] = [[↑iP]]$\\
+    Then by inversion, $[[Γ ⊢ iP]]$.
+    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as
+    $[[Γ ⊢ [σ1]↑iP ≤ [σ2]↑iP]]$,
+    then by congruence of substitution and by inversion,
+    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
+
+    Applying the induction hypothesis to $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$, we obtain
+    $[[Γ ⊢ σi ≈ id : Ord fv iP]]$. Since $[[fv ↑iP]] = [[fv iP]]$, we can
+    conclude: $[[Γ ⊢ σi ≈ id : Ord fv ↑iP]]$.
+  \item The positive cases are proved symmetrically.
+  \end{caseof}
+\end{proof}
+
+\begin{corollary}[Substitution cannot induce proper subtypes or supertypes] \label{corollary:subst-proper-subt}
+  Assuming all mentioned types are well-formed in $[[Γ]]$ and $[[σ]]$ is a
+  substitution $[[Γ ⊢ σ : Γ]]$,
+  \begin{align*}
+    [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
+                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iN]] \\
+    [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
+                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iN]] \\
+    [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
+                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iP]] \\
+    [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
+                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iP]] \\
+  \end{align*}
+\end{corollary}
+
+\begin{lemma} \label{lemma:mutual-subst-subtyping}
+  Asssuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
+  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
+  \begin{itemize}
+  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
+    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
+    $[[Γ ⊢ σ1 ≈ Sub μ : Ord fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 : Ord fv iQ]]$;
+  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
+    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
+    $[[Γ ⊢ σ1 ≈ Sub μ : Ord fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 : Ord fv iM]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  \hfill
+  \begin{itemize}
+  \item[$+$]
+    Applying $[[σ2]]$ to both sides of
+    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{todo}),
+    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
+    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ (by transitivity \cref{todo}),
+    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
+    Then by \cref{corollary:subst-proper-subt},
+    $[[Γ ⊢ σ2 ○ σ1 ≈ id : Ord fv iP]]$.
+
+    % Applying $[[σ1]]$ to both sides of
+    % $[[Γ ⊢ [σ2]iQ ≥ iP]]$ (by \cref{todo}),
+    % we have: $[[Γ ⊢ [σ1 ○ σ2] iQ ≥ [σ1]iP]]$.
+    % Composing it with $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by transitivity \cref{todo}),
+    % we have $[[Γ ⊢ [σ1 ○ σ2] iQ ≥ iQ]]$.
+    % Then by \cref{corollary:subst-proper-subt},
+    By a symmetric argument, we also have:
+    $[[Γ ⊢ σ1 ○ σ2 ≈ id : Ord fv iQ]]$.
+
+    Now, we prove that
+    $[[Γ ⊢ σ2 ○ σ1 ≈ id : Ord fv iP]]$ and
+    $[[Γ ⊢ σ1 ○ σ2 ≈ id : Ord fv iQ]]$
+    implies that $[[σ1]]$ and $[[σ1]]$
+    are (equivalent to) mutually inverse bijections.
+
+    To do so, it suffices to prove that
+    \begin{enumerate}
+    \item[(i)] for any $[[α± ∊ fv iP]]$ there exists $[[β± ∊ fv iQ]]$
+        such that $[[ Γ ⊢ [σ1] α± ≈ β± ]]$ and
+        $[[ Γ ⊢ [σ2] β± ≈ α± ]]$; and
+    \item[(ii)] for any $[[β± ∊ fv iQ]]$ there exists $[[α± ∊ fv iP]]$
+        such that $[[ Γ ⊢ [σ2] β± ≈ α± ]]$ and
+        $[[ Γ ⊢ [σ1] α± ≈ β± ]]$.
+    \end{enumerate}
+    Then the these correspondences between $[[fv iP]]$ and
+    $[[fv iQ]]$ are mutually inverse functions,
+    since for any $[[β±]]$ there can be at most one $[[α±]]$
+    such that $[[ Γ ⊢ [σ2] β± ≈ α± ]]$ (and vice versa).
+
+    \begin{enumerate}
+    \item[(i)] Let us take $[[α± ∊ fv iP]]$.
+      \begin{enumerate}
+      \item if $[[α±]]$ is positive ($[[α± = α⁺]]$),
+        from $[[ Γ ⊢ [σ2][σ1]α⁺ ≈ α⁺ ]]$,
+        by \cref{corollary:vars-no-proper-subtypes},
+        we have
+        $[[ [σ2][σ1]α⁺ = ∃nbs.α⁺ ]]$.
+
+        What shape can $[[ [σ1]α⁺ ]]$ have? It cannot be $[[∃nas.↓iN]]$ (for
+        potentially empty $[[nas]]$), because the outer constructor $\downarrow$
+        would remain after the substitution $[[σ2]]$, whereas $[[∃nbs.α⁺]]$ does
+        not have $[[↓]]$. The only case left is $[[ [σ1]α⁺ = ∃nas.γ⁺ ]]$.
+
+        Notice that $[[Γ ⊢ ∃nas.γ⁺ ≈ γ⁺]]$, meaning that $[[Γ ⊢ [σ1]α⁺ ≈ γ⁺]]$.
+        Also notice that $[[ [σ2]∃nas.γ⁺ = ∃nbs.α⁺ ]]$ implies
+        $[[Γ ⊢ [σ2]γ⁺ ≈ α⁺]]$.
+
+      \item if $[[α±]]$ is negative ($[[α± = α⁻]]$) from $[[ Γ ⊢ [σ2][σ1]α⁻ ≈ α⁻
+        ]]$, by \cref{corollary:vars-no-proper-subtypes}, we have
+        $[[ [σ2][σ1]α⁻ = ∀pbs.α⁻ ]]$.
+
+        What shape can $[[ [σ1]α⁻ ]]$ have? It cannot be $[[∀pas.↑iP]]$
+        nor $[[∀pas.iP → iM]]$ (for potentially empty $[[pas]]$),
+        because the outer constructor ($[[→]]$ or $[[↑]]$), remaining
+        after the substitution $[[σ2]]$, is however absent in the resulting
+        $[[∀pbs.α⁻]]$. Hence, the only case left is $[[ [σ1]α⁻ = ∀pas.γ⁻ ]]$
+        Notice that $[[Γ ⊢ γ⁻ ≈ ∀pas.γ⁻]]$, meaning that $[[Γ ⊢ [σ1]α⁻ ≈ γ⁻]]$.
+        Also notice that $[[ [σ2]∀pas.γ⁻ = ∀pbs.α⁻ ]]$ implies
+        $[[Γ ⊢ [σ2]γ⁻ ≈ α⁻]]$.
+      \end{enumerate}
+    \item[(ii)] The proof is symmetric:
+      We swap $[[iP]]$ and $[[iQ]]$,
+      $[[σ1]]$ and $[[σ2]]$,
+      and exploit $[[ Γ ⊢ [σ1][σ2]α± ≈ α± ]]$ instead of
+      $[[ Γ ⊢ [σ2][σ1]α± ≈ α± ]]$.
+
+    \end{enumerate}
+
+  \item[$-$] The proof is symmetric to the positive case.
+  \end{itemize}
+\end{proof}
+
+\begin{lemma}[Equivalence of polymorphic types]
+  \label{lemma:poly-types-equivalence}
+  \hfill
+  \begin{itemize}
+    \item[$-$] For $[[Γ ⊢ ∀pas.iN]]$ and $[[Γ ⊢ ∀pbs.iM]]$,\\ if $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$
+    then there exists a bijection $[[μ : {pbs} ∩ fv iM ↔ {pas} ∩ fv iN]]$
+    such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iN ]]$,
+    \item[$+$] For $[[Γ ⊢ ∃nas.iP]]$ and $[[Γ ⊢ ∃nbs.iQ]]$,\\  if $[[Γ ⊢ ∃nas.iP ≈ ∃nbs.iQ ]]$
+    then there exists a bijection $[[μ : {nbs} ∩ fv iQ ↔ {nas} ∩ fv iP]]$
+    such that $[[ Γ, nbs ⊢ iP ≈ [Sub μ] iQ ]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+    \hfill
+  \begin{itemize}
+    \item[$-$]
+    First, by $\alpha$-conversion, we ensure $[[{pas} ∩ fv iM = ∅]]$ and $[[{pbs} ∩ fv iN = ∅]]$.
+    By inversion, $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$ implies 
+    \begin{enumerate} 
+      \item $[[Γ,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pbs ⊢ σ1 : pas ]]$ and 
+      \item $[[Γ,pas ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas ⊢ σ2 : pbs ]]$.
+    \end{enumerate}
+    To apply \cref{lemma:mutual-subst-subtyping}, we weaken 
+    and rearrange the contexts, and extend the substitutions to act as identity
+    outside of their initial domain:
+    \begin{enumerate} 
+      \item $[[Γ,pas,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pas,pbs ⊢ σ1 : Γ,pas,pbs ]]$ and 
+      \item $[[Γ,pas,pbs ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas,pbs ⊢ σ2 : Γ,pas,pbs ]]$.
+    \end{enumerate}
+    Then from \cref{lemma:mutual-subst-subtyping}, 
+    there exists a bijection $[[μ : fv iM ↔ fv iN]]$ such that 
+    $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ : Ord fv iM]]$ and 
+    $[[Γ,pas,pbs ⊢ σ1 ≈ Sub μ-1 : Ord fv iN]]$.
+
+    Let us show that if we restrict the domain of $[[μ]]$ to 
+    $[[pbs]]$, its range will be contained in $[[pas]]$.
+    Let us take $[[γ⁺ ∊ {pbs} ∩ fv iM]]$ and 
+    assume $[[ [μ]γ⁺]] \notin [[pas]]$.
+    Then since $[[ Γ,pbs ⊢ σ1 : pas ]]$, 
+    $[[σ1]]$ acts as identity outside of $[[pas]]$, i.e.
+    $[[ [σ1][Sub μ]γ⁺ = [Sub μ]γ⁺ ]]$.
+    Since
+    $[[Γ,pas,pbs ⊢ σ1 ≈ Sub μ-1 : Ord fv iN]]$, 
+    application of $[[σ1]]$ is equivalent to application of $[[Sub μ-1]]$,
+    then 
+    $[[ Γ,pas,pbs ⊢ [Sub μ-1][Sub μ]γ⁺ ≈ [Sub μ]γ⁺ ]]$, i.e.
+    $[[Γ,pas,pbs ⊢ γ⁺ ≈ [Sub μ]γ⁺]]$, 
+    which means $[[γ⁺ ∊ fv [Sub μ]γ⁺]] \subseteq [[fv iN]]$.
+    By assumption, $[[γ⁺ ∊ {pbs} ∩ fv iM]]$, i.e. $[[{pbs} ∩ fv iN]] \neq \emptyset$, hence contradiction.
+
+    By \cref{todo}, 
+    $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ|{pbs} : Ord fv iM]]$ implies
+    $[[Γ,pas,pbs ⊢ [σ2]iM ≈ [Sub μ|{pbs}]iM]]$.
+    By similar reasoning, $[[Γ,pas,pbs ⊢ [σ1]iN ≈ [Sub μ-1|{pas}]iN]]$.
+
+    This way,
+    \begin{align} 
+      [[Γ,pas,pbs ⊢ [Sub μ-1|{pas}]iN ≤ iM]] \label{fact:mu-inv-n-sub-m}\\
+      [[Γ,pas,pbs ⊢ [Sub μ|{pbs}]iM ≤ iN]] \label{fact:mu-m-subt-n}
+    \end{align}
+
+    By applying $[[μ|_{pbs}]]$ to both sides of \ref{fact:mu-inv-n-sub-m} (\cref{todo})
+    and contracting $[[μ-1|_{pas} ○ μ|_{pbs}]] = [[μ|_{pbs}-1 ○ μ|_{pbs}]] = [[id]]$,
+    we have: $[[Γ,pas,pbs ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
+    means $[[Γ,pas,pbs ⊢ iN ≈ [Sub μ|{pbs}]iM]]$, and by strengthening, $[[Γ,pas⊢ iN ≈ [Sub μ|{pbs}]iM]]$.
+    Symmetrically, $[[Γ,pbs ⊢ iM ≈ [Sub μ|_{pbs}-1]iN]]$.
+    \item{$+$} The proof is symmetric to the proof of the negative case.
+  \end{itemize}
+
+\end{proof}
+
+\begin{lemma}[Completeness of equivalence] \label{lemma:equiv-completeness}
   Mutual subtyping implies declarative equivalence.
+  Assuming all the types below are well-formed in $[[Γ]]$: 
   \begin{itemize}
@@
 \begin{proof}
-  \ilyam{todo}
+  \begin{itemize}
+    \item[$-$] 
+    Induction on the sum of sizes of  $[[iN]]$ and $[[iM]]$. 
+    By inversion, $[[Γ ⊢ iN ≈ iM]]$ means $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iM ≤ iN ]]$.
+    Let us consider the last rule that forms $[[Γ ⊢ iN ≤ iM]]$:
+    \begin{caseof}
+      \item \ruleref{\ottdruleDOneNVarLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ α⁻ ≤ α⁻]]$\\
+      Then $[[iN ≈ iM]]$ (i.e. $[[α⁻ ≈ α⁻]]$) holds immediately by \ruleref{\ottdruleEOneNVarLabel}.
+
+      \item \ruleref{\ottdruleDOneShiftULabel} i.e. 
+      $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ ↑iP ≤ ↑iQ]]$\\
+      Then by inversion, $[[Γ ⊢ iP ≈ iQ]]$, 
+      and by induction hypothesis, $[[iP ≈ iQ]]$.
+      Then $[[iN ≈ iM]]$ (i.e. $[[↑iP ≈ ↑iQ]]$) holds 
+      by \ruleref{\ottdruleEOneShiftULabel}.
+
+      \item \ruleref{\ottdruleDOneArrowLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ iP → iN' ≤ iQ → iM']]$\\
+      Then by inversion, $[[Γ ⊢ iP ≥ iQ]]$ and $[[Γ ⊢ iN' ≤ iM']]$.
+      Notice that $[[Γ ⊢ iM ≤ iN]]$ is of the form $[[Γ ⊢ iQ → iM' ≤ iP → iN']]$, 
+      which by inversion means $[[Γ ⊢ iQ ≥ iP]]$ and $[[Γ ⊢ iM' ≤ iN']]$.
+
+      This way, $[[Γ ⊢ iQ ≈ iP]]$ and $[[Γ ⊢ iM' ≈ iN']]$. 
+      Then by induction hypothesis, $[[iQ ≈ iP]]$ and $[[iM' ≈ iN']]$.
+      Then $[[iN ≈ iM]]$ (i.e. $[[iP → iN' ≈ iQ → iM']]$) holds by \ruleref{\ottdruleEOneArrowLabel}.
+
+      \item \ruleref{\ottdruleDOneForallLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.iM']]$\\
+      Then by \cref{lemma:poly-type-equivalence}, $[[Γ ⊢ ∀pas.iN' ≈ ∀pbs.iM']]$ means that 
+      there exists a bijection $[[μ : {pbs} ∩ fv iM' ↔ {pas} ∩ fv iN']]$ such that  
+      $[[Γ,pas ⊢ [Sub μ]iM' ≈ iN']]$.
+
+      Notice that the application of bijection $[[μ]]$ to $[[iM']]$ does
+      not change its size (which is less than the size of $[[iM]]$), hence the induction hypothesis applies.
+      This way, $[[ [Sub μ]iM' ≈ iN']]$ (and by symmetry, $[[iN' ≈ [Sub μ]iM']]$) holds by induction. 
+      Then we apply \ruleref{\ottdruleEOneForallLabel} to get $[[∀pas.iN' ≈ ∀pbs.iM']]$, i.e. $[[iN ≈ iM]]$.
+    \end{caseof}
+
+\item[$+$] The proof is symmetric to the proof of the negative case.
+  \end{itemize}
 \end{proof}

### 2023-04-20 proofs/tex/old-proof.tex
Add equivalence soundness and completeness proofs

\begin{lemma}
  Informally, this lemma says that if $[[Γ, Ord varset1 ⊢ [σ21]iP ≥ iQ]]$
  and $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP]]$ holds for substitutions $[[σ12]]$
  and $[[σ21]]$ then these substitutions are in fact mutually inverse bijections
  between variables $[[varset1]]$ and $[[varset1]]$.

  \begin{itemize}
  \item[$+$]
    For $[[Γ, Ord varset2 ⊢ iP]]$,~ $[[Γ, Ord varset1 ⊢ iQ]]$,~
    $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,~
    $[[Γ, Ord varset1 ⊢ σ21 : Ord varset2]]$,
    suppose that:
    \begin{enumerate}
      \item $\{ [[coh]] \mid [[α±:coh]] \in [[varset1 ∩ fv iQ]] \}$ = $\{ [[coh]] \mid [[α±:coh]] \in [[varset2 ∩ fv iP]] \}$,
      \item $[[Γ, Ord varset1 ⊢ [σ21]iP ≥ iQ]]$,
      \item $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP]]$.
    \end{enumerate}
    Then there exists a bijection
    $\mu : [[varset1 ∩ fv iQ]] \leftrightarrow [[varset2 ∩ fv iP]]$ such that:
    \begin{enumerate}
    \item $\mu$ preserves cohorts:
      $\mu ([[α±:coh]]) = [[β±:coh]]$ ($[[α±:coh]]$ and $[[β±:coh]]$
      have the same cohort label $\pm [[coh]]$).
    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ   : Ord (varset1 ∩ fv iQ) ]]$ (the equivalence is pointwise)
    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iP) ]]$ (the equivalence is pointwise)
    \end{enumerate}

  \item[$-$]
    For $[[Γ, Ord varset2 ⊢ iN]]$,~ $[[Γ, Ord varset1 ⊢ iM]]$,~
    $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,~
    $[[Γ, Ord varset1 ⊢ σ21 : Ord varset2]]$,
    suppose that:
    \begin{enumerate}
    \item $\{ [[coh]] \mid [[α±:coh]] \in [[varset1 ∩ fv iM]] \}$ = $\{ [[coh]] \mid [[β±:coh]] \in [[varset2 ∩ fv iN]] \}$,
    \item $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$,
    \item $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$.
    \end{enumerate}
    Then there exists a bijection
    $\mu : [[varset1 ∩ fv iM]] \leftrightarrow [[varset2 ∩ fv iN]]$ such that:
    \begin{enumerate}
    \item $\mu$ preserves cohorts:
      $\mu ([[α±:coh]]) = [[β±:coh]]$ ($[[α±:coh]]$ and $[[β±:coh]]$
      have the same cohort label $\pm [[coh]]$).
    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ : Ord (varset1 ∩ fv iM) ]]$ (the equivalence is pointwise)
    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iN)]]$ (the equivalence is pointwise)
    \end{enumerate}
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on the pair of sizes of inference trees: $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$
  and $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ (or the corresponding trees in the
  positive case).

  \begin{caseof}
  \item $[[iN]] = [[∀pds.β⁻:coh]]$, for $[[β⁻:coh]] \in
    [[varset2]]$ and possibly empty $[[pds]]$\\
    \label{case:double-subtyping-subst-happened}
    Then by \cref{todo} \ilyam{lemma} $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ ∀pds.β⁻:coh]]$ means that $[[ [σ12]iM ]] =
    [[∀pcs.β⁻:coh]]$, which (since $[[β⁻:coh]] \notin [[fv iM]]$) is only possible
    when $[[iM]] = [[∀pcs'.α⁻:coh']]$ (where $[[α⁻:coh']] \in [[varset1]]$ and $[[pcs']]$ is possibly empty), and
    $[[σ12]] ([[α⁻:coh']]) = [[∀pcs''.β⁻:coh]]$.
    Notice that by the first condition, the sets of cohort labels of $[[varset1
    ∩ fv iM]]$ and $[[varset2 ∩ fv iN]]$ must be equal, which means $[[coh]] = [[coh']]$.
    Also notice that $[[Γ, Ord varset2 ⊢ ∀pcs''.β⁻:coh ≈ β⁻:coh]]$.

    Then $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ becomes
    $[[Γ, Ord varset1 ⊢ [σ21]∀pds.β⁻:coh ≤ ∀pcs'.α⁻:coh]]$,
    which by \cref{todo} \ilyam{lemma} implies that $[[ [σ21] ∀pds.β⁻:coh]] =
    [[∀pds'.α⁻:coh]]$, and thus,
    $[[σ21]] ([[β⁻:coh]]) = [[∀pds''.α⁻:coh]]$.
    Notice that $[[Γ, Ord varset1 ⊢ ∀pds''.α⁻:coh ≈ α⁻:coh]]$.

    This way, we can take $\mu = [[α⁻:coh]] \mapsto [[β⁻:coh]] $, which by
    construction is a bijection preserving cohorts. Moreover,
    $[[Γ, Ord varset2 ⊢ ∀pcs''.β⁻:coh ≈ β⁻:coh]]$ means
    $[[Γ, Ord varset2 ⊢ [σ12] α⁻:coh ≈ [Sub μ] α⁻:coh]]$ implying
    $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ : Ord (varset1 ∩ fv iM) ]]$;
    and
    $[[Γ, Ord varset1 ⊢ ∀pds''.α⁻:coh ≈ α⁻:coh]]$ means
    $[[Γ, Ord varset1 ⊢ [σ21] β⁻:coh ≈ [Sub μ-1] β⁻:coh]]$ implying
    $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ-1 : Ord (varset2 ∩ fv iN)]]$.

  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
    \ruleref{\ottdruleDOneNVarLabel}, i.e. $[[ [σ21]iN ]] = [[iM]] = [[γ⁻]]$.\\
    Then the case when $[[γ⁻]] \in [[varset2]]$, has been covered by
    \cref{case:double-subtyping-subst-happened}, so we assume
    that $[[γ⁻]] \notin [[varset2]]$, and thus, $[[iN]] = [[γ⁻]]$.

    Notice that $[[γ⁻]] \notin [[varset1]]$ because otherwise,
    $[[varset1 ∩ fv iN]] \neq \emptyset$,
    which would contradict with $[[Γ, Ord varset2 ⊢ iN]]$.

    Then $[[fv iN ∩ varset2]] = [[{γ⁻} ∩ varset2]] = \emptyset$ and $[[fv iM ∩
    varset1]] = [[{γ⁻} ∩ varset1]] = \emptyset$.
    Hence, we take the empty $\mu : \emptyset \leftrightarrow \emptyset$, which
    vacuously satisfies the required properties.

  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
    \ruleref{\ottdruleDOneShiftULabel}, i.e. $[[ [σ21]iN ]] = [[↑iP]]$, $[[ iM
    ]] = [[↑iQ]]$, and $[[Γ, Ord varset1 ⊢ iP ≈ iQ]]$\\
    Since $[[iN]]$ is not a variable from the domain of $[[σ21]]$ (which has
    been covered by \cref{case:double-subtyping-subst-happened}),
    the substitution applied to $[[iN]]$ must preserve its outer shape.
    Specifically, $[[ [σ21]iN ]] = [[↑iP]]$ means
    $[[ [σ21]iN ]] = [[ [σ21]↑iP']] = [[ ↑[σ21]iP']] =[[↑iP]]$,
    i.e. $[[ iN ]] = [[↑iP']]$ and
    $[[ [σ21]iP']] = [[iP]]$. In particular,
    $[[Γ, Ord varset1 ⊢ iP ≈ iQ]]$ implies
    $[[Γ, Ord varset1 ⊢ iP ≥ iQ]]$, i.e.
    $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$.

    In addition, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ becomes
    $[[Γ, Ord varset2 ⊢ ↑[σ12]iQ ≤ ↑iP']]$, which is only inferable by
    \ruleref{\ottdruleDOneShiftULabel}, meaning that
    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≈ iP']]$, and in particular,
    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$.

    Notice that the tree inferring $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$ is a
    proper subtree of $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$. Analogously,
    $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$ is a proper subtree of
    $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$. This way, we apply the induction
    hypothesis to $[[Γ, Ord varset1 ⊢ [σ21]iP' ≥ iQ]]$ and $[[Γ, Ord varset2 ⊢
    [σ12]iQ ≥ iP']]$ (notice that $[[varset1]]$, $[[varset2]]$, and the sets of free variables of
    the types did not change) and obtain exactly what we aimed.

  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
    \ruleref{\ottdruleDOneForallLabel}, i.e. $[[ [σ21]iN ]] = [[∀pcs.iN']]$,
    $[[iM]] = [[∀pds.iM']]$, and $[[Γ, Ord varset1, pds ⊢ [iPs/pcs]iN' ≤ iM']]$
    for  $[[Γ, Ord varset1, pds ⊢ iPi]]$ \\

    Since $[[iN]]$ does not have the shape of $[[∀pds.β⁻:coh]]$, for $[[β⁻:coh]]
    \in [[varset2]]$ (which has been covered by
    \cref{case:double-subtyping-subst-happened}), the substitution applied to
    $[[iN]]$ must preserve its outer shape. Specifically,
    $[[ [σ21]iN ]] = [[∀pcs.iN']]$
    means that $[[ iN ]]$ ``starts with'' $\forall[[pcs]]$, i.e.
    $[[ [σ21]iN ]] = [[ [σ21]∀pcs.iN'' ]] = [[ ∀pcs.[σ21]iN'' ]] = [[ ∀pcs.iN'
    ]]$, where $[[iN]] = [[∀pcs.iN'']]$ and $[[ [σ21]iN'']] = [[iN']]$.

    This way, $[[Γ, Ord varset1, pds ⊢ [iPs/pcs]iN' ≤ iM']]$ becomes
    $[[Γ, Ord varset1, pds ⊢ [iPs/pcs][σ21]iN'' ≤ iM']]$.
    Notice that the tree inferring this judgment is a proper subtree of
    $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$.

    On the other hand, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$
    becomes $[[Γ, Ord varset2 ⊢ ∀pds.[σ12]iM' ≤ ∀pcs.iN'']]$
    (where either $[[pds]]$ or $[[pcs]]$ is non-empty),
    which is only inferable by \ruleref{\ottdruleDOneForallLabel},
    meaning that
    $[[Γ, Ord varset2, pcs ⊢ [iQs/pds][σ12]iM' ≤ iN'']]$ for some
    $[[Γ, Ord varset2, pcs ⊢ iQi]]$.
    Notice that the tree inferring this judgment is a proper subtree of
    $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$.

    Let us label $[[pcs]]$ and $[[pds]]$ with a sufficiently large cohort label
    $[[mcoh]]$ such that $[[mcoh]] > [[coh]]$ for any $[[α⁺:coh]] \in [[varset1 ∪
    varset2]]$. Then we merge $[[iQs/δ⁺:mcoh*]]$ and $[[σ12]]$ denoting the resulting
    substitution as $[[σ12']]$ $([[(Γ, Ord varset2, γ⁺:mcoh*) ⊢ σ12' : (Ord varset1, δ⁺:mcoh*)]])$.

    What do we mean by merging? The codomains of $[[iQs/δ⁺:mcoh*]]$ and
    $[[σ12]]$ are $[[Γ, Ord varset2,  γ⁺:mcoh*]]$ and $[[Γ, Ord varset2]]$,
    respectively, which are disjoint with both of their domains
    ($[[δ⁺:mcoh*]]$ and $[[Ord varset1]]$, respectively). In turn, the domains
    of $[[iQs/δ⁺:mcoh*]]$ and $[[σ12]]$ are themselves mutually disjoint.
    This way, by merging we mean
    $[[iQs/δ⁺:mcoh*]] \cup [[σ12]]$ (relationally) or
    $[[iQs/δ⁺:mcoh* ○ σ12]]$ or
    $[[σ12 ○ iQs/δ⁺:mcoh*]]$, since these three substitutions are equal.

    Analogously, we merge $[[iPs/γ⁺:mcoh*]]$ and $[[σ21]]$ denoting the
    resulting substitution as $[[σ21']]$ $([[(Γ, Ord varset1, δ⁺:mcoh*) ⊢ σ21' :
    (Ord varset2, γ⁺:mcoh*)]])$.

    We wish to apply the induction hypothesis to
    $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$ and
    $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12']iM' ≤ iN'']]$.
    To do so, we need to show that the cohort labels of $[[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv
    iN'']]$ coincide with those of $[[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]$ (as sets).
    \begin{assertion*}
      The set of cohorts of $[[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$
      is equal to the set of cohorts of $[[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]$.
    \end{assertion*}

    \begin{proof}
      $
      \begin{aligned}[t]
        [[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']] &= [[(varset2 ∪ {γ⁺:mcoh*}) ∩ (fv
                                               iN ∪ ({γ⁺:mcoh*} ∩ fv iN''))]]\\
                                             &= [[varset2 ∩ fv iN ∪ {γ⁺:mcoh*} ∩ fv iN'']]
                                             && \text{because
                                                $[[varset2]]$ and $[[fv iN]]$ are disjoint with $[[γ⁺:mcoh*]]$}\\
        [[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']]  &= [[varset1 ∩ fv iM ∪ {δ⁺:mcoh*} ∩ fv iM']]
                                             && \text{analogously}\\
      \end{aligned}\\
      $
      Since the cohort labels of $[[varset2 ∩ fv iN]]$ coincide with those of
      $[[varset1 ∩ fv iM]]$ by the assumption, it suffices to prove that the cohort labels of
      $[[{γ⁺:mcoh*} ∩ fv iN'']]$ and of $[[{δ⁺:mcoh*} ∩ fv iM']]$ coincide.
      Note that these sets have the same
      cohorts labels $[[mcoh]]$, i.e. it is required to show that these sets are
      either both empty or both non-empty, i.e.
      $[[{γ⁺:mcoh*} ∩ fv iN'']] = \emptyset \iff [[{δ⁺:mcoh*} ∩ fv iM']] = \emptyset$
      \begin{itemize}
      \item [$(\Leftarrow)$] Suppose that $[[{δ⁺:mcoh*} ∩ fv iM']] = \emptyset$.
        Then by \cref{todo}, since
        $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$, $[[fv [σ21']iN'']]
        \subseteq [[fv iM']]$, implying that $[[{δ⁺:mcoh*} ∩ fv [σ21']iN'']] =
        \emptyset$,
        and by context strengthening \cref{todo}, $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'']]$ reduces to
        $[[Γ, Ord varset1 ⊢ [σ21']iN'']]$.

        Let us restrict $[[σ21']]$ to the set of free variables of $[[fv iN'']]$.
        Then the domain and codomain of $[[σ21']]$ are the following:
        $[[Γ, Ord varset1 ⊢ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩ fv iN]]$.
        By \cref{todo}, $[[ [σ21']iN'' ]] = [[ [σ21'|fv iN'']iN'' ]]$,
        and then
        $[[Γ, Ord varset1 ⊢ [σ21']iN'' ≤ iM']]$, can be rewritten as
        $[[Γ, Ord varset1 ⊢ [σ21'|fv iN'']iN'' ≤ iM']]$.

        Let us apply substitution $[[σ12]]$ ($[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$)
        to both sides of this judgment to obtain: $[[Γ, Ord varset2 ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ [σ12]iM']]$.
        Using the transitivity \cref{todo}, let us compose this subtyping
        judgment with $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12]iM' ≤ iN'']]$ to
        form
        $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ iN'']]$.

        What is the codomain of $[[σ12 ○ σ21'|fv iN'']]$?
        By composing
        $[[Γ, Ord varset1 ⊢ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩ fv iN]]$
        and
        $[[Γ, Ord varset2 ⊢ σ12 : Ord varset1]]$,
        we have
        $[[Γ, Ord varset2 ⊢ σ12 ○ σ21'|fv iN'' : Ord {Ord varset2, γ⁺:mcoh*} ∩
        fv iN]]$, i.e. $[[σ12 ○ σ21'|fv iN'']]$ is a substitution contracting
        $[[γ⁺:mcoh*]]$.
        Then by \cref{todo},
        $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12 ○ σ21'|fv iN'']iN'' ≤ iN'']]$
        implies that $[[{γ⁺:mcoh*} ∩ fv iN'']] = \emptyset$.
        \item [$(\Rightarrow)$]
          Analogous to the previous case.
      \end{itemize}
    \end{proof}
    This way, we can apply the induction hypothesis to
    $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ [σ21']iN'' ≤ iM']]$ and
    $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12']iM' ≤ iN'']]$,
    and obtain
    $\mu : [[(varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM']] \leftrightarrow [[(varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$ such that:
    \begin{enumerate}
    \item $\mu$ preserves cohorts,
    \item $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ σ12' ≈ Sub μ : Ord (varset1 ∪ {δ⁺:mcoh*}) ∩ fv iM' ]]$, and
    \item $[[Γ, Ord varset1, δ⁺:mcoh* ⊢ σ21' ≈ Sub μ-1 : Ord (varset2 ∪ {γ⁺:mcoh*}) ∩ fv iN'']]$
    \end{enumerate}

    Let us decompose $\mu$ into the union of two bijections: $\mu_1 \sqcup
    \mu_2$, where $\mu_1$ is defined on the variables whose cohort labels are less
    than $[[mcoh]]$, and $\mu_2$ is defined on the variables whose cohort labels
    are equal to $[[mcoh]]$. Notice that since $\mu$ preserves cohorts,
    this union is disjoint: the \emph{range} of $\mu_1$ is the variables with cohorts
    $< [[mcoh]]$ and the \emph{range} of $\mu_2$ is the variables with cohorts
    $= [[mcoh]]$.

    Recalling how $[[mcoh]]$ is chosen, notice that the signatures of $\mu_1$ and
    $\mu_2$ are the following:
    \begin{enumerate}
      \item $\mu_1 : [[varset1 ∩ fv iM']] \leftrightarrow [[varset2 ∩ fv iN'']]$ and
      \item $\mu_2 : [[{δ⁺:mcoh*} ∩ fv iM']] \leftrightarrow [[{γ⁺:mcoh*} ∩ fv iN'']]$.
    \end{enumerate}

    This way, $\mu_1$ is a cohort-preserving bijection with the required
    signature, and what is left to show is the equivalences $[[μ1]] \eqDOne
    [[σ12]]$, and $[[μ1-1]] \eqDOne [[σ21]]$:
    \begin{enumerate}[nosep]
    \item $[[Γ, Ord varset2 ⊢ σ12 ≈ Sub μ1 : Ord (varset1 ∩ fv iM) ]]$\\
      Let us take an arbitrary $[[α±]] \in [[varset1 ∩ fv iM]]$.\\
      $
      \begin{aligned}
        [[ [σ12] α± ]] &= [[ [σ12 ○ iQs/δ⁺:mcoh*] α± ]]
                       && \text{because $[[α±]] \neq [[δi⁺:mcoh]]$} \\
                       &= [[ [σ12'] α± ]]
                       && \text{by the definition of $[[σ12']]$} \\
                       &\eqDOne [[ [Sub μ] α± ]]
                       && \text{since $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ σ12' ≈ Sub μ : Ord ((varset1 ∪ {δ⁺:mcoh*}) ∩ fv iN) ]]$} \\
                       &= [[ [Sub μ1] α± ]]
                       && \text{since $[[μ1]] = [[Sub μ | varset1 ∩ fv iM']]$
                          and  $[[α±]] \in [[varset1 ∩ fv iM]] \subseteq [[varset1 ∩ fv iM']]$} \\
      \end{aligned}
      $\\
      This way, $[[Γ, Ord varset2, γ⁺:mcoh* ⊢ [σ12] α± ≈ [Sub μ1] α±  ]]$, which,
      considering that the codomains of $[[σ12]]$ and $[[μ1]]$ are in $[[Γ, Ord
      varset2]]$, can be strengthen to $[[Γ, Ord varset2 ⊢ [σ12] α± ≈ [Sub μ1] α± ]]$.

    \item $[[Γ, Ord varset1 ⊢ σ21 ≈ Sub μ1-1 : Ord (varset2 ∩ fv iN)]]$ is proved analogously.

    \end{enumerate}

  \item The last rule to infer $[[Γ, Ord varset1 ⊢ [σ21]iN ≤ iM]]$ was
    \ruleref{\ottdruleDOneArrowLabel}, i.e. $[[ [σ21]iN ]] = [[iP → iN']]$, and
    $[[iM]] = [[iQ → iM']]$; then by inverting this rule,
    $[[Γ, Ord varset1 ⊢ iP ≥ iQ]]$ and $[[Γ, Ord varset1 ⊢ iN' ≤ iM']]$.

    $[[ [σ21]iN ]] = [[iP → iN']]$ means that either $[[iN]]$ is a variable
    from the domain of $[[σ21]]$ (which has been covered by
    \cref{case:double-subtyping-subst-happened})
    or $[[ [σ21]iN ]] = [[ [σ21] (iP' → iN'')]] = [[ [σ21] iP' → [σ21] iN'' ]] =
    [[iP → iN']]$, where
    $[[ iN ]] = [[iP' → iN'']]$, $[[ [σ21] iP' ]] = [[iP]]$, and $[[ [σ21] iN'' ]] = [[iN']]$.

    This way, $[[Γ ⊢ iP ≥ iQ]]$ and $[[Γ ⊢ iN' ≤ iM']]$ can be rewritten as
    $[[Γ ⊢ [σ21]iP' ≥ iQ]]$ and $[[Γ ⊢ [σ21] iN'' ≤ iM']]$.

    In addition, $[[Γ, Ord varset2 ⊢ [σ12]iM ≤ iN]]$ becomes
    $[[Γ, Ord varset2 ⊢ [σ12](iQ → iM') ≤ iP' → iN'']]$,
    implying by inversion
    $[[Γ, Ord varset2 ⊢ [σ12]iQ ≥ iP']]$ and
    $[[Γ, Ord varset2 ⊢ [σ12]iM' ≤ iN'']]$.

  \end{caseof}
\end{proof}

### 2023-04-20 proofs/tex/wf-lemmas.tex
Add equivalence soundness and completeness proofs

\begin{lemma}[Well-formedness agrees with substitution]
  \label{lemma:wf-subst}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ, Γ1 ⊢ iP]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iP]]$
  \item[$-$] $[[Γ, Γ1 ⊢ iN]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iN]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

\begin{corollary}
  \label{lemma:wf-subst}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ1, Γ2 ⊢ iP]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iP]]$
  \item[$-$] $[[Γ1, Γ2 ⊢ iN]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iN]]$
  \end{itemize}
\end{corollary}

\begin{lemma}[Equivalent Contexts]
  \label{lemma:wf-ctxt-equiv}
  In the well-formedness judgment, only used variables matter:
  \begin{itemize}
  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$ then
    $[[Γ1 ⊢ iP]] \iff [[Γ2 ⊢ iP]]$,
  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$ then
    $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
\end{proof}

\begin{corollary}
  \label{lemma:mut-sub-types-wf-equiv}
  Suppose that all the types below are well-formed in $[[Γ]]$ and
  $[[{Γ'} ⊆ {Γ}]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ ⊢ iP ≈ iQ]]$ implies $[[Γ' ⊢ iP]] \iff [[Γ' ⊢ iQ]]$
  \item[$-$] $[[Γ ⊢ iN ≈ iM]]$ implies $[[Γ' ⊢ iN]] \iff [[Γ' ⊢ iM]]$
  \end{itemize}
\end{corollary}
\begin{proof}
  From \cref{lemma:wf-ctxt-equiv,corollary:fv-mut-sub}.
\end{proof}

### 2023-04-24 proofs/tex/decl-subtyping-lemmas.tex
Add equivalence and upgrade lemmas

@@
-\begin{lemma}[Free Variable Propagation]
-  \label{lemma:fv-propagation}
+\begin{lemma}[Free Variable Propagation] \label{lemma:fv-propagation}
   In the judgments of negative subtyping or positive supertyping,

### 2023-04-27 proofs/main.tex
WIP
finish completeness of lub; add initiality of lub
Add soundness of ≈
Add equivalence soundness and completeness proofs
Add equivalence and upgrade lemmas
Subtyping: soundness is done, completeness WIP

@@
 % P ≈ Q
-\renewcommand{\ottdruleEOnePVarName}[0]{(Var$^+$)}
+\renewcommand{\ottdruleEOnePVarName}[0]{(Var$^+$$^{\eqEOne}$)}
 \renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
@@
 \renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}
+
+% G ⊢ N ≈1 M
+\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne^{-}$)}
+
+% G ⊢ P ≈1 Q
+\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne^{+}$)}
+
+% G ⊨ iP1 ∨ iP2 = iQ
+\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
+\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
+\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
+\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}
+
+% G ; Θ ⊨ uN ≤ iM ⫤ us
+\renewcommand{\ottdruleANVarName}[0]{(Var$^-$$^{\subA}$)}
+\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
+\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
+\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}
+
+% G ; Θ ⊨ iP ≥ uQ ⫤ us
+\renewcommand{\ottdruleAPVarName}[0]{(Var$^+$$^{\supA}$)}
+\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
+\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
+\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}
 
@@
 Unification solution is represented by a list of unification solution entries.
-Each entry restrict an unification variable in two possible ways: either stating
+Each entry restricts an unification variable in two possible ways: either stating
 that it must be equivalent to a certain type ($[[Δ ⊢ pua :≈ iP]]$ or $[[Δ ⊢ nua :≈
@@
 \end{definition}
+Notice that in case of equivalence, the assigned types
+must be equal (i.e. alpha-equivalent) to be merged. This is because
+the unification algorithm assumes that every type is normalized,
+and hence, equivalence is alpha-equivalence 
+(\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
 
@@
 
+\subsection{Declarative Subtyping}
+\input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
+
 \subsection{Substitution}
@@
 
+\subsection{Upgrade}
+\input{../_build/_gen/tex/upgrade-lemmas.tex}
+
 \subsection{Upper Bounds}
@@
 
+\subsection{Unification}
+\input{../_build/_gen/tex/unification-lemmas.tex}\
+
+\subsection{Anti-unification}
+\input{../_build/_gen/tex/anti-unification-lemmas.tex}
+
+\subsection{Solution Merge}
+\input{../_build/_gen/tex/solution-merge-lemmas.tex}
+
+\subsection{Subtyping Algorithm}
+\input{../_build/_gen/tex/alg-subtyping-lemmas.tex}
+
 \end{document}

### 2023-04-27 proofs/tex/alg-subtyping-lemmas.tex
Subtyping: soundness is done, completeness WIP

\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
    \begin{itemize}
        \item [$-$] 
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
        implies $[[us : Θ ∩ uv uN]]$ and 
        for any $[[us']]$ such that $[[us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uN ≤ iM ]]$

        \item [$+$] 
        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
        implies $[[us : Θ ∩ uv uP]]$ and 
        for any $[[us']]$ such that $[[us' ⇒ us]]$,
        $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
     \end{itemize}
\end{theorem}
\begin{proof}
    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ (and mutually, on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$).
    Let us consider the last rule to infer this judgment. 
    \begin{caseof}
        \item $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ ∩ ∅]]$.
        Since $[[uv a⁻ = ∅]]$, any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
        $[[ [us']a⁻ = a⁻ ]]$. Therefore, $[[Γ ⊢ [us']a⁻ ≤ a⁻]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.

        \item $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
        Then the next step of the algorithm is the unification of $[[nf(uP)]]$ and $[[nf(iQ)]]$.
        By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
        it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uP]]$.
        By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[us' ⇒ us]]$ means 
        $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uP ]]$ as substitutions.

        $[[ [us]nf(uP) = nf(iQ) ]]$ implies $[[Γ ⊢ [us]nf(uP) ≈ nf(iQ)]]$, and then 
        $[[Γ ⊢ [us']nf(uP) ≈ nf(iQ)]]$. \ilyam{add lemmas}

        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uP) ≈ nf(iQ)]]$ by 
        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
        $[[Γ ⊢ [us']nf(uP) ≈ [us']uP ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
        $[[Γ ⊢ nf(iQ) ≈ iQ ]]$. 
        This way, we have $[[Γ ⊢ [us']uP ≈ iQ]]$.
        Then by \ruleref{\ottdruleDOneShiftULabel}
        and congruence of substitution, $[[Γ ⊢ [us']↑uP ≤ ↑iQ]]$.

        \item $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
        The next step of the algorithm is two recursive calls:
        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
        By the induction hypothesis, 
        \begin{enumerate}
            \item $[[us1 : Θ ∩ uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1' ⇒ us1]]$
            \item $[[us2 : Θ ∩ uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2' ⇒ us2]]$
        \end{enumerate}

        Then the algorithm merges two unification solutions $[[us1]]$ and $[[us2]]$.
        By \cref{lemma:merge-soundness}, since $[[uv uP ∪ uv uN' = uv (uP → uN')]]$, 
        we have $[[us1 & us2 : Θ ∩ uv (uP → uN')]]$, and also
        $[[us1 & us2 ⇒ us1]]$ and $[[us1 & us2 ⇒ us2]]$.
        By the transitivity of solution weakening (\cref{todo}),
         $[[us' ⇒ us1 & us2]]$ implies $[[us' ⇒ us1]]$ and $[[us' ⇒ us2]]$.

% \[\begin{tikzcd}[arrows=Rightarrow]
% 	& {[[us']]} \\
% 	{[[us1]]} & {[[us1 & us2]]} & {[[us2]]}
% 	\arrow[from=2-2, to=2-3]
% 	\arrow[from=2-2, to=2-1]
% 	\arrow[from=1-2, to=2-2]
% \end{tikzcd}\]

        The application of the induction hypothesis gives us 
        $[[Γ ⊢ [us']uP ≥ iQ ]]$ and $[[ Γ ⊢ [us']uN' ≤ iM' ]]$.
        Finally, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [us'](uP → uN') ≤ iQ → iM']]$.

        \item $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
        Then the algorithm creates fresh unification variables $[[â⁺*[Γ,pbs] ]]$, 
        substitutes the old $[[pas]]$ with them in $[[uN']]$, and makes the recursive call:
        $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ [â⁺*/pas] uN ≤ iM ⫤ us']]$, returning as the resulting solution 
        $[[us]] = [[us' \ {α̂⁺*}]]$.

        Notice that $[[Γ, pbs ⊢ Θ, â⁺*[G, pbs] ]]$, $[[Γ,pbs ⊢ iM']]$, and 
        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN ]]$, so the induction hypothesis is applicable,
        that is $[[us' : Θ ∩ uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any $[[us'2 ⇒ us']]$.

        Since the domain of $[[us']]$ is $[[uv [â⁺*/pas]uN']]$, the domain of 
        $[[us]] = [[us' \  {α̂⁺*}]]$ is $[[uv [â⁺*/pas]uN \ {â⁺*} = uv uN']] = [[uv ∀pas.uN']]$,
        this way, $[[us : Θ ∩ uv ∀pas.uN']]$, as required.

        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2 ⇒ us]]$.
        Let us consider an arbitrary $[[us2 ⇒ us]]$. Let us construct $[[us'2]]$, 
        extending $[[us2]]$ to the domain $[[Ord uv [â⁺*/pas]uN']]$ with the values of $[[us']]$,
        i.e.  $[[us'2 : Θ ∩ uv [â⁺*/pas]uN']]$, 
        and
        \[
            [[us'2(β̂±)]]  = 
            \begin{cases}
               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']] \\
               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁺*]]
            \end{cases}
        \] 
        , where the application of the unification solution to a variable returns the 
        corresponding \emph{unification entry}. 
        Notice that $[[Sub us'2|uv uN']] = [[us2]]$.
    It is easy to see that $[[us'2 ⇒ us']]$: 
    \begin{enumerate}
        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±) ⇒ us(β̂±)]] = [[us'(β̂±)]]$;
        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±) ⇒ us'(β̂±)]]$,
    \end{enumerate}
    which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
    $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$.\\
    Notice that
    $
    \begin{aligned}[t]
                 [[ [us'2][â⁺*/pas]uN' ]] &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][Sub us'2|uv uN']uN' ]]
                                          && \text{by substitution properties \ilyam{todo}}\\
                                          &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ]]
                                          && \text{Since $[[Sub us'2|uv uN']] = [[us2]]$}.
    \end{aligned}
    $\\
    Also notice that the domain of $[[Sub us'2|{â⁺*} ○ â⁺*/pas]]$ is $[[pas]]$,
    and the range is $[[Γ, pbs]]$, i.e. $[[Γ, pbs ⊢ Sub us'2|{â⁺*} ○ â⁺*/pas : pas]]$, 
    which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
    $[[ Γ, pbs ⊢ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ≤ iM' ]]$
    to get $[[ Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM' ]]$, as required.

    \item $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (Δ ⊢ â⁺ :≥ iQ')]]$ where
    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$\\

    Notice that $[[â⁺[Δ] ∊ Θ]]$ and $[[Γ ⊢ Θ]]$ 
    implies $[[Γ = Δ, pnas]]$ for some $[[pnas]]$, hence, the
    soundness of upgrade (\cref{lemma:upgrade-soundness}) is applicable:
    \begin{enumerate}
        \item $[[Δ ⊢ iQ']]$ and
        \item $[[Γ ⊢ iQ ≥ iP]]$.
    \end{enumerate}

    Since $[[â⁺[Δ] ∊ Θ]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(Δ ⊢ â⁺ :≥ iQ') : Θ ∩ uv â⁺ ]]$.

    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us' ⇒ Δ ⊢ â⁺ :≥ iQ']]$.
    The latter weakening statement means that either $[[us']] \ni [[Δ ⊢ â⁺ :≥ iQ'']]$ or
    $[[us']] \ni [[(Δ ⊢ â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
    transitivity (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
    as required.

    \item For the other positive cases, the proof is symmetric to the corresponding negative cases.
    \end{caseof}
\end{proof}

Notice that the induction statement in following theorem is not symmetric for the positive and negative cases.
\begin{theorem}[Completeness and Initiality of the Subtyping Algorithm]
    \hfill
    \begin{itemize}
        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$ and $[[Γ ; Θ ⊢ uN]] \neq [[α̂⁻]]$ and there exists $[[us : Θ ∩ uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
        Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
        and $[[us ⇒ us']]$.

        \item [$+$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
        there exists $[[us : Θ ∩ uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
        Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
        and $[[us ⇒ us']]$.
    \end{itemize}
\end{theorem}
\begin{proof}
    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$ (and mutually, on $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$ or $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
    but first consider the base case for the substitution $[[ [us]uP ]]$:
    \begin{caseof}
        \item $[[uP]] = [[ α̂⁺ ]]$ is a unification variable\\
        Then by assumption, there exists $[[us : Θ ∩ uv uP]]$ such that $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$.
        It means that $[[us]]$ sends $[[α̂⁺]]$ to $[[iP']]$ such that $[[Γ ⊢ iP' ≥ iQ]]$.
        This way, $[[us]]$ is either $[[ (Δ ⊢ α̂⁺ :≈ iP') ]]$ or $[[ (Δ ⊢ α̂⁺ :≥ iP') ]]$,
        where $[[α̂⁺[Δ] ∊ Θ]]$.

        In this case, the algorithm tries to apply \ruleref{\ottdruleAPUVarLabel}
        and the resulting solution is $[[us']] = [[(Δ ⊢ α̂⁺ :≥ iQ')]]$ where
        $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.

        Why does the upgrade procedure terminates?
        Because $[[iP']]$ satisfies the pre-conditions of the completeness of the upgrade
        (\cref{lemma:upgrade-completeness})
        :
        \begin{itemize}
            \item $[[Δ ⊢ iP']]$ because $[[iP' = [us]α̂⁺]]$, $[[us : Θ ∩ uv uP]]$ and 
            $[[α̂⁺[Δ] ∊ Θ ∩ uv uP]]$,
            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted before
        \end{itemize}
        Moreover, completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
        and further, we strengthen it to $[[Δ ⊢ iP' ≥ iQ']]$
        (since by the soundness of the upgrade (\cref{lemma:upgrade-soundness}),
        $[[Δ ⊢ iQ']]$).

        It means that $[[ (Δ ⊢ α̂⁺ :≈ iP') ⇒ (Δ ⊢ α̂⁺ :≥ iQ') ]]$ and 
        $[[ (Δ ⊢ α̂⁺ :≥ iP') ⇒ (Δ ⊢ α̂⁺ :≥ iQ') ]]$, which means that in any case, 
        $[[us ⇒ us']]$.
    \end{caseof}
\end{proof}

### 2023-04-27 proofs/tex/anti-unification-lemmas.tex
Subtyping: soundness is done, completeness WIP

\begin{lemma}[Soundness of Anti-Unification] \label{lemma:anti-unification-soundness}
    \hfill
    \begin{itemize}
        \item [$+$] 
        \item [$-$] 
    \end{itemize}
\end{lemma}

\begin{lemma}[Completeness of Anti-Unification] \label{lemma:anti-unification-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] 
        \item [$-$] 
    \end{itemize}
\end{lemma}

### 2023-04-27 proofs/tex/equiv-lemmas.tex
Add equivalence and upgrade lemmas
Subtyping: soundness is done, completeness WIP

@@
   \end{caseof}
+\end{proof}
+
+\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] \label{corollary:nf-sound-wrt-subt-equiv}
+  \hfill
+  \begin{itemize}
+    \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
+    \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
 \end{proof}
@@
 \end{proof}
+
+\begin{corollary}[Normalization is complete w.r.t. subtyping-induced equivalence]
+  \label{corollary:nf-complete-wrt-subt-equiv}
+  Assuming all the types below are well-formed in $[[Γ]]$:
+  \begin{itemize}
+    \item [$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[nf(iP) = nf(iQ)]]$,
+    \item [$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[nf(iN) = nf(iM)]]$.
+  \end{itemize}
+\end{corollary}  
+\begin{proof}
+  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
+\end{proof}

### 2023-04-27 proofs/tex/unification-lemmas.tex
Subtyping: soundness is done, completeness WIP

\begin{lemma}[Soundness of Unification] \label{lemma:unification-soundness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$ then $[[us : Θ ∩ uv uP]]$, 
        $[[us]]$ is equivalence-only, and $[[ [us]uP = iQ ]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$ then $[[us : Θ ∩ uv uN]]$,
        $[[us]]$ is equivalence-only, and $[[ [us]uN = iM ]]$.
    \end{itemize}
\end{lemma}

\begin{lemma}[Completeness of Unification] \label{lemma:unification-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
        assume there exists $[[us' : Θ ∩ uv uP]]$ such that
        $[[ [us']uP = iQ ]]$. Then there exists $[[us]]$ such that
        $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
        assume there exists $[[us' : Θ ∩ uv uN]]$ such that
        $[[ [us']uN = iM ]]$. Then there exists $[[us]]$ such that
        $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
   \end{itemize}
\end{lemma}

### 2023-04-27 proofs/tex/upgrade-lemmas.tex
Subtyping: soundness is done, completeness WIP

Let us consider a type $[[iP]]$ well-formed in $[[Γ]]$.
Some of its $[[Γ]]$-supertypes are also well-formed in a smaller context $[[{Δ} ⊆ {Γ}]]$.
The upgrade is the operation that returns the least of such supertypes.

\begin{lemma}[Soundness of Upgrade]\label{lemma:upgrade-soundness}
    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]],$\\
    if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
    then
    \begin{enumerate}
        \item $[[Δ ⊢ iQ]]$
        \item $[[Γ ⊢ iQ ≥ iP]]$
    \end{enumerate}
\end{lemma}
\begin{proof}
    By inversion, $[[upgrade Γ ⊢ iP to Δ = iQ]]$ means that 
    for fresh $[[pnbs]]$ and $[[pncs]]$,
    $[[Δ, pnbs, pncs ⊨ [pnbs/pnas]iP ∨ [pncs/pnas]iP = iQ]]$.
    Then by the soundness of the least upper bound (\cref{lemma:lub-soundness}),
    \begin{enumerate}
        \item $[[Δ, pnbs, pncs ⊢ iQ]]$, 
        \item $[[Δ, pnbs, pncs ⊢ iQ ≥ [pnbs/pnas]iP]]$, and 
        \item $[[Δ, pnbs, pncs ⊢ iQ ≥ [pncs/pnas]iP]]$.
    \end{enumerate}

    $ 
    \begin{aligned}
        [[fv iQ]] &\subseteq [[fv [pnbs/pnas]iP ∩ fv [pncs/pnas]iP]]
                  &&\text{Since by \cref{lemma:fv-propagation}, 
                         $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$ and
                         $[[fv iQ ⊆ fv [pncs/pnas]iP]]$}\\
                  &\subseteq [[ ((fv iP \ {pnas}) ∪ {pnbs}) ∩ ((fv iP \ {pnas}) ∪ {pncs})]]\\
                  &= [[ (fv iP \ {pnas}) ∩ (fv iP \ {pnas}) ]]
                  &&\text{since $[[pnbs]]$ and $[[pncs]]$ are fresh}\\
                  &= [[ fv iP \ {pnas} ]]\\
                  &\subseteq [[ {Γ} \ {pnas} ]]
                  &&\text{since $[[iP]]$ is well-formed in $[[Γ]]$}\\
                  &\subseteq [[ {Δ} ]]\\
    \end{aligned}
    $\\
    This way, by \cref{lemma:wf-ctxt-equiv}, $[[Δ ⊢ iQ]]$.

    Let us apply $[[pnas/pnbs]]$---the inverse of the substitution $[[ pnbs/pnas ]]$ to 
    both sides of $[[Δ, pnbs, pncs ⊢ iQ ≥ [pnbs/pnas]iP]]$ and 
    by \cref{todo}, get $[[Δ, pnas, pncs ⊢ [pnas/pnbs]iQ ≥ iP]]$.
    Notice that $[[Δ ⊢ iQ]]$ implies that $[[fv iQ ∩ {pnbs} = ∅]]$, 
    then by \cref{todo}, $[[ [pnas/pnbs]iQ = iQ]]$, and thus $[[Δ, pnas, pncs ⊢ iQ ≥ iP]]$.
    By context strengthening, $[[Δ, pnas ⊢ iQ ≥ iP]]$.
\end{proof}

\begin{lemma}[Completeness and Initiality of Upgrade] \label{lemma:upgrade-completeness}
    The upgrade returns the least $[[Γ]]$-supertype of $[[iP]]$ well-formed in $[[Δ]]$.
    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]],$\\
    For any $[[iQ']]$ such that 
    \begin{enumerate}
        \item $[[Δ ⊢ iQ']]$ and
        \item $[[Γ ⊢ iQ' ≥ iP]]$,
    \end{enumerate}

    The result of the upgrade algorithm $[[iQ]]$ exists
    ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
\end{lemma}
\begin{proof}

    Let us consider fresh (not intersecting with $[[Γ]]$) $[[pnbs]]$ and $[[pncs]]$.

    If we apply substitution $[[pnbs/pnas]]$ to both sides of $[[Δ, pnas ⊢ iQ' ≥ iP]]$,
    we have $[[Δ, pnbs ⊢ [pnbs/pnas]iQ' ≥ [pnbs/pnas]iP]]$, which by  
    \cref{todo}, since $[[iQ']]$ is well-formed in $[[Δ]]$, simplifies to 
     $[[Δ, pnbs ⊢ iQ' ≥ [pnbs/pnas]iP]]$.

    Analogously, if we apply substitution $[[pncs/pnas]]$ to both sides of $[[Δ, pnas ⊢ iQ' ≥ iP]]$,
    we have $[[Δ, pncs ⊢ iQ' ≥ [pncs/pnas]iP]]$.

    This way, $[[iQ']]$ is a common supertype of $[[ [pnbs/pnas]iP ]]$ and $[[ [pncs/pnas]iP ]]$ in
    context $[[Δ, pnbs, pncs]]$. It means that we can apply the completeness of the least upper bound
    (\cref{lemma:lub-completeness}):
    \begin{enumerate}
        \item there exists $[[iQ]]$ s.t. $[[Γ ⊨ [pnbs/pnas]iP ∨ [pncs/pnas]iP = iQ]]$ 
        \item $[[Γ ⊢ iQ' ≥ iQ]]$.
    \end{enumerate}
    The former means that the upgrade algorithm terminates and returns $[[iQ]]$.
    The latter means that since both $[[iQ']]$ and $[[iQ]]$ are well-formed in $[[Δ]]$,
    by \cref{tood}, $[[Δ ⊢ iQ' ≥ iQ]]$.
\end{proof}

### 2023-04-27 proofs/main.tex
Move context from the solution entries; fix a bit

@@
 Each entry restricts an unification variable in two possible ways: either stating
-that it must be equivalent to a certain type ($[[Δ ⊢ pua :≈ iP]]$ or $[[Δ ⊢ nua :≈
-iN]]$) or that it must be a (positive) supertype of a certain type ($[[Δ ⊢ pua :≥
+that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua :≈
+iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥
 iP]]$).

### 2023-04-27 proofs/tex/alg-subtyping-lemmas.tex
Move context from the solution entries; fix a bit

@@
         $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
-        implies $[[us : Θ ∩ uv uN]]$ and 
-        for any $[[us']]$ such that $[[us' ⇒ us]]$,
+        implies $[[us : Θ|uv uN]]$ and 
+        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
         $[[ Γ ⊢ [us']uN ≤ iM ]]$
@@
         $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
-        implies $[[us : Θ ∩ uv uP]]$ and 
-        for any $[[us']]$ such that $[[us' ⇒ us]]$,
+        implies $[[us : Θ|uv uP]]$ and 
+        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
         $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
@@
         \item $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
-        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ ∩ ∅]]$.
-        Since $[[uv a⁻ = ∅]]$, any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
+        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
+        Since $[[uv a⁻ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
         $[[ [us']a⁻ = a⁻ ]]$. Therefore, $[[Γ ⊢ [us']a⁻ ≤ a⁻]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
@@
         it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uP]]$.
-        By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[us' ⇒ us]]$ means 
+        By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
         $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uP ]]$ as substitutions.
@@
         \begin{enumerate}
-            \item $[[us1 : Θ ∩ uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1' ⇒ us1]]$
-            \item $[[us2 : Θ ∩ uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2' ⇒ us2]]$
+            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
+            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
         \end{enumerate}
@@
         By \cref{lemma:merge-soundness}, since $[[uv uP ∪ uv uN' = uv (uP → uN')]]$, 
-        we have $[[us1 & us2 : Θ ∩ uv (uP → uN')]]$, and also
-        $[[us1 & us2 ⇒ us1]]$ and $[[us1 & us2 ⇒ us2]]$.
+        we have $[[us1 & us2 : Θ | uv (uP → uN')]]$, and also
+        $[[Θ ⊢ us1 & us2 ⇒ us1]]$ and $[[Θ ⊢ us1 & us2 ⇒ us2]]$.
         By the transitivity of solution weakening (\cref{todo}),
-         $[[us' ⇒ us1 & us2]]$ implies $[[us' ⇒ us1]]$ and $[[us' ⇒ us2]]$.
+         $[[Θ ⊢ us' ⇒ us1 & us2]]$ implies $[[Θ ⊢ us' ⇒ us1]]$ and $[[Θ ⊢ us' ⇒ us2]]$.
 
@@
         $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN ]]$, so the induction hypothesis is applicable,
-        that is $[[us' : Θ ∩ uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any $[[us'2 ⇒ us']]$.
+        that is $[[us' : Θ | uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any
+        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.
 
@@
         $[[us]] = [[us' \  {α̂⁺*}]]$ is $[[uv [â⁺*/pas]uN \ {â⁺*} = uv uN']] = [[uv ∀pas.uN']]$,
-        this way, $[[us : Θ ∩ uv ∀pas.uN']]$, as required.
+        this way, $[[us : Θ | uv ∀pas.uN']]$, as required.
 
-        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2 ⇒ us]]$.
-        Let us consider an arbitrary $[[us2 ⇒ us]]$. Let us construct $[[us'2]]$, 
+        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
+        Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
         extending $[[us2]]$ to the domain $[[Ord uv [â⁺*/pas]uN']]$ with the values of $[[us']]$,
-        i.e.  $[[us'2 : Θ ∩ uv [â⁺*/pas]uN']]$, 
+        i.e.  $[[us'2 : Θ | uv [â⁺*/pas]uN']]$, 
         and
@@
         Notice that $[[Sub us'2|uv uN']] = [[us2]]$.
-    It is easy to see that $[[us'2 ⇒ us']]$: 
+    It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
     \begin{enumerate}
-        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±) ⇒ us(β̂±)]] = [[us'(β̂±)]]$;
-        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±) ⇒ us'(β̂±)]]$,
+        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
+        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
     \end{enumerate}
@@
                                           &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ]]
-                                          && \text{Since $[[Sub us'2|uv uN']] = [[us2]]$}.
+                                          && \text{since $[[Sub us'2|uv uN']] = [[us2]]$}.
     \end{aligned}
@@
 
-    \item $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (Δ ⊢ â⁺ :≥ iQ')]]$ where
+    \item $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
     $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$\\
@@
 
-    Since $[[â⁺[Δ] ∊ Θ]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(Δ ⊢ â⁺ :≥ iQ') : Θ ∩ uv â⁺ ]]$.
+    Since $[[â⁺[Δ] ∊ Θ|uv â⁺]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(â⁺ :≥ iQ') : Θ | uv â⁺ ]]$.
 
-    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us' ⇒ Δ ⊢ â⁺ :≥ iQ']]$.
-    The latter weakening statement means that either $[[us']] \ni [[Δ ⊢ â⁺ :≥ iQ'']]$ or
-    $[[us']] \ni [[(Δ ⊢ â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
+    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
+    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
+    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
     $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
@@
     \begin{itemize}
-        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$ and $[[Γ ; Θ ⊢ uN]] \neq [[α̂⁻]]$ and there exists $[[us : Θ ∩ uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$ and $[[Γ ; Θ ⊢ uN]] \neq [[α̂⁻]]$ and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
         Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
-        and $[[us ⇒ us']]$.
+        and $[[Θ ⊢ us ⇒ us']]$.
 
         \item [$+$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
-        there exists $[[us : Θ ∩ uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
+        there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
         Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
-        and $[[us ⇒ us']]$.
+        and $[[Θ ⊢ us ⇒ us']]$.
     \end{itemize}
@@
         \item $[[uP]] = [[ α̂⁺ ]]$ is a unification variable\\
-        Then by assumption, there exists $[[us : Θ ∩ uv uP]]$ such that $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$.
-        It means that $[[us]]$ sends $[[α̂⁺]]$ to $[[iP']]$ such that $[[Γ ⊢ iP' ≥ iQ]]$.
-        This way, $[[us]]$ is either $[[ (Δ ⊢ α̂⁺ :≈ iP') ]]$ or $[[ (Δ ⊢ α̂⁺ :≥ iP') ]]$,
-        where $[[α̂⁺[Δ] ∊ Θ]]$.
+        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$.
+        $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
+        where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
+        $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$.
 
-        In this case, the algorithm tries to apply \ruleref{\ottdruleAPUVarLabel}
-        and the resulting solution is $[[us']] = [[(Δ ⊢ α̂⁺ :≥ iQ')]]$ where
+        The algorithm tries to apply \ruleref{\ottdruleAPUVarLabel}
+        and the resulting solution is $[[us']] = [[(α̂⁺ :≥ iQ')]]$ where
         $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.
@@
         \begin{itemize}
-            \item $[[Δ ⊢ iP']]$ because $[[iP' = [us]α̂⁺]]$, $[[us : Θ ∩ uv uP]]$ and 
-            $[[α̂⁺[Δ] ∊ Θ ∩ uv uP]]$,
+            \item $[[Δ ⊢ iP']]$ because $[[iP' = [us]α̂⁺]]$, $[[us : Θ | uv uP]]$ and 
+            $[[α̂⁺[Δ] ∊ Θ | uv uP]]$,
             \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted before
@@
 
-        It means that $[[ (Δ ⊢ α̂⁺ :≈ iP') ⇒ (Δ ⊢ α̂⁺ :≥ iQ') ]]$ and 
-        $[[ (Δ ⊢ α̂⁺ :≥ iP') ⇒ (Δ ⊢ α̂⁺ :≥ iQ') ]]$, which means that in any case, 
-        $[[us ⇒ us']]$.
+        It means that $[[Θ ⊢ (α̂⁺ :≈ iP') ⇒ (α̂⁺ :≥ iQ') ]]$ and 
+        $[[ Θ ⊢ (α̂⁺ :≥ iP') ⇒ (α̂⁺ :≥ iQ') ]]$, which means that in any case, 
+        $[[ Θ ⊢ us ⇒ us']]$.
     \end{caseof}

### 2023-04-27 proofs/tex/unification-lemmas.tex
Move context from the solution entries; fix a bit

@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$ then $[[us : Θ ∩ uv uP]]$, 
+        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$ then $[[us : Θ | uv uP]]$, 
         $[[us]]$ is equivalence-only, and $[[ [us]uP = iQ ]]$.
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$ then $[[us : Θ ∩ uv uN]]$,
+        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$ then $[[us : Θ | uv uN]]$,
         $[[us]]$ is equivalence-only, and $[[ [us]uN = iM ]]$.
@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        assume there exists $[[us' : Θ ∩ uv uP]]$ such that
+        assume there exists $[[us' : Θ | uv uP]]$ such that
         $[[ [us']uP = iQ ]]$. Then there exists $[[us]]$ such that
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        assume there exists $[[us' : Θ ∩ uv uN]]$ such that
+        assume there exists $[[us' : Θ | uv uN]]$ such that
         $[[ [us']uN = iM ]]$. Then there exists $[[us]]$ such that

### 2023-05-02 proofs/tex/alg-subtyping-lemmas.tex
Add soundness and completeness of subtyping

@@
         $[[Θ ⊢ us1 & us2 ⇒ us1]]$ and $[[Θ ⊢ us1 & us2 ⇒ us2]]$.
-        By the transitivity of solution weakening (\cref{todo}),
+        By the transitivity of solution weakening (\cref{lemma:weakening-transitivity}),
          $[[Θ ⊢ us' ⇒ us1 & us2]]$ implies $[[Θ ⊢ us' ⇒ us1]]$ and $[[Θ ⊢ us' ⇒ us2]]$.
@@
     \begin{itemize}
-        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$ and $[[Γ ; Θ ⊢ uN]] \neq [[α̂⁻]]$ and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
+        $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
+        and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
         Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
@@
     \begin{caseof}
-        \item $[[uP]] = [[ α̂⁺ ]]$ is a unification variable\\
-        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$.
+        \item \label{case:subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
+        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
         $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
         where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
-        $[[ Γ ⊢ [us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$.
+        $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$
+        because multiple inversions of \ruleref{\ottdruleDOneExistsLabel} 
+        gives us $[[Γ ⊢ [us]α̂⁺ ≥ iQ]]$ since $[[ {nbs} ∩ fv [us]α̂⁺ = ∅]]$.
 
-        The algorithm tries to apply \ruleref{\ottdruleAPUVarLabel}
+        In the algorithm, after multiple applications of \ruleref{\ottdruleAExistsLabel}
+        the type $[[∃nbs.α̂⁺]]$ is reduced to $[[α̂⁺]]$.
+        Next, the algorithm tries to apply
+        \ruleref{\ottdruleAPUVarLabel}
         and the resulting solution is $[[us']] = [[(α̂⁺ :≥ iQ')]]$ where
@@
         $[[ Θ ⊢ us ⇒ us']]$.
+
+        \item \label{case:subt-complete-nvar}
+        $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}, 
+        i.e. $[[ [us]uN ]] = [[iM]] = [[ α⁻ ]] = [[iN]]$ 
+        (the latter equality holds because $[[uN]] \neq [[α̂⁻]]$).
+
+        Notice that $[[us : Θ | uv α⁻]]$ = $[[us : Θ | ∅]]$ = $[[·]]$.
+
+        The algorithm applies \ruleref{\ottdruleANVarLabel} and 
+        infers $[[us']] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$.
+
+        Since $[[Θ ⊢ · ⇒ ·]]$, we have $[[Θ ⊢ us ⇒ us']]$.
+
+        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel},
+        \label{case:subt-complete-upshift}
+        i.e. $[[ [us]uN ]] = [[ ↑[us]uP ]]$ (since the substitution $[[ [us]uN ]]$ must preserve the 
+        top-level constructor of $[[uN]]\neq [[α̂⁻]]$), and $[[uM]] = [[ ↑iQ ]]$,
+        and by inversion, $[[ Γ ⊢ [us]uP ≈ iQ ]]$.
+
+        Since both types start with $[[↑]]$, 
+        the algorithm tries to apply \ruleref{\ottdruleAShiftULabel}: 
+        $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us']]$. The premise of this rule is the
+        unification of $[[nf(uP)]]$ and $[[nf(iQ)]]$:
+        $[[Γ;Θ ⊨ nf(uP) ≈u nf(iQ) ⫤ us']]$. Let us show that the unification successfully 
+        terminates and returns $[[us']] = [[nf(us)]]$.
+
+        To demonstrate that the unification terminates, we apply the completeness 
+        of the unification algorithm (\cref{lemma:unification-completeness}). 
+        In order to do that, we need to provide a unifier of 
+        $[[nf(uP)]]$ and $[[nf(iQ)]]$. Thankfully, $[[nf(us)]]$ does it.
+
+        \begin{itemize}
+            \item $[[nf(uP)]]$ and $[[nf(iQ)]]$ are normalized 
+            \item $[[Γ ; Θ ⊢ nf(uP)]]$ because $[[Γ ; Θ ⊢ uP]]$ (\cref{todo})
+            \item $[[Γ ⊢ nf(iQ)]]$ because $[[Γ ⊢ iQ]]$ (\cref{corollary:wf-nf})
+            \item $[[nf(us) : Θ | uv nf(uP)]]$ because $[[us: Θ | uv uP]]$ (\cref{todo})
+            \item $ \begin{aligned}[t]
+                    [[ Γ ⊢ [us]uP ≈ iQ ]] &\Rightarrow [[ [us]uP ≈ iQ ]]
+                                          && \text {by \cref{lemma:equiv-completeness}}\\
+                                          &\Rightarrow [[ nf([us]uP) = nf(iQ) ]]
+                                          && \text {by \cref{lemma:normalization-completeness}}\\
+                                          &\Rightarrow [[ [nf(us)]nf(uP) = nf(iQ) ]]
+                                          && \text {by \cref{lemma:norm-subst-distr}}\\
+                    \end{aligned}
+                  $
+        \end{itemize}
+        Then by the completeness of the unification,
+        $[[Γ ; Θ ⊨ nf(uP) ≈u nf(iQ) ⫤ nf(us)]]$.
+        This way, the subtyping algorithm terminates and the resulting solution is
+        $[[nf(us)]]$.
+
+        It is left to note that $[[Θ ⊢ us ⇒ nf(us)]]$, by \cref{todo}, since $[[Θ ⊢ us ≈ nf(us)]]$
+
+        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneArrowLabel}, 
+        i.e. $[[ [us]uN ]] = [[ [us]uP → [us]uN' ]]$ and $[[iM]] = [[iQ → iM']]$, 
+        and by inversion, $[[ Γ ⊢ [us]uP ≥ iQ ]]$ and $[[ Γ ⊢ [us]uN' ≤ iM' ]]$.
+
+        Let us consider restrictions of $[[us]]$ to 
+        the set of unification variables in $[[uP]]$ and $[[uN']]$:
+        $[[us | uv uP : Θ | uv uP]]$ and $[[us | uv uN' : Θ | uv uN']]$.
+        Notice that 
+        $[[ [us]uP = [us | uv uP]uP ]]$ and 
+        $[[ [us]uN' = [us | uv uN']uN' ]]$.
+
+       Let us apply the induction hypothesis to
+       $[[ Γ ⊢ [us | uv uP]uP ≥ iQ ]]$ and
+       $[[ Γ ⊢ [us | uv uN']uN' ≤ iM' ]]$ (notice that
+       since $[[uv uN' ⊆ uv uN]]$, there are no negative unification variables in $[[uN']]$)
+       to obtain $[[us'1]]$ and $[[us'2]]$ such that
+       \begin{enumerate}
+        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'1]]$ and $[[Θ ⊢ us | uv uP ⇒ us'1]]$
+        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'2]]$ and $[[Θ ⊢ us | uv uN' ⇒ us'2]]$ 
+       \end{enumerate}
+
+       This way, the algorithm applies \ruleref{\ottdruleAArrowLabel}
+       and terminates returning $[[us'1 & us'2]]$ as the result.
+
+       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
+       Since $[[us | uv uP ⊆ us]]$, 
+       by \cref{lemma:weakening-monotonicity}, 
+       $[[Θ ⊢ us ⇒ us | uv uP]]$,
+       and then by \cref{lemma:weakening-transitivity}, 
+       $[[Θ ⊢ us | uv uP ⇒ us'1]]$.
+       Analogously, $[[Θ ⊢ us | uv uN' ⇒ us'2]]$.
+       Then since by \cref{lemma:merge-completeness}, 
+       $[[us'1 & us'2]]$ is the weakest restriction  
+       implying both $[[us'1]]$ and $[[us'2]]$,
+       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.
+
+       \item \label{case:subt-complete-forall}
+       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel},
+       i.e. $[[ [us]uN ]] = [[ ∀pas.[us]uN' ]]$
+       (assuming $[[pas]]$ does not intersect with the 
+       range of $[[us]]$)
+       and 
+       $[[iM]] = [[∀pbs.iM']]$, where either $[[pas]]$ or $[[pbs]]$
+       is non-empty.
+
+       By inversion, 
+       $[[ Γ, pbs ⊢ [σ][us]uN' ≤ iM' ]]$ for 
+       $[[Γ, pbs ⊢ σ : pas ]]$.
+       Since $[[σ]]$ and $[[us]]$ have disjoint domains,
+       and the range of one does not intersect with the domain of the other,
+       they commute, i.e. $[[ Γ, pbs ⊢ [us][σ]uN' ≤ iM' ]]$
+       (notice that the tree inferring this judgement is 
+       a proper subtree of the tree inferring 
+       $[[ Γ ⊢ [us]uN ≤ iM ]]$).
+
+       On the next step, 
+       the algorithm creates fresh (disjoint with $[[uv uN']]$) 
+       unification variables $[[â⁺*]]$, replaces $[[pas]]$ with them in $[[ [us]uN' ]]$,
+       and makes the recursive call:
+       $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ uN0 ≤ iM ⫤ us1]]$,
+       (where $[[uN0]] = [[ [â⁺*/pas]uN' ]]$),
+       returning $[[us1 \ {â⁺*}]]$ as the result.
+
+       Notice that $[[ [us][σ][pas/â⁺*]uN0 = [us][σ]uN' ]]$,
+       and thus, $[[ Γ, pbs ⊢ [us][σ][pas/â⁺*]uN0 ≤ iM' ]]$.
+       Let us combine $[[us]]$ and $[[σ  ○ pas/â⁺*]]$ into one 
+       $[[us0]]$ an unification solution for $[[uN0]]$:
+        \[
+            [[us0(β̂±)]]  = 
+            \begin{cases}
+               [[ β̂± :≈ [σ]αi⁺ ]] & \text{if } [[β̂±]] = [[αî⁺]] \in [[â⁺*]] \\
+               [[us(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']]
+            \end{cases}
+       \]
+
+       Let us show that the induction hypothesis is applicable to 
+       $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$.
+       \begin{itemize}
+        \item $[[ us0 : Θ|uv uN', â⁺*[Γ, pbs] ]]$. Notice that for every $[[αî⁺]]$, 
+            the type corresponding to the entry $[[us0(αî⁺)]]$ 
+            is well-formed in $[[Γ, pbs]]$ since $[[Γ, pbs ⊢ σ : pas ]]$;
+            and for every $[[β̂±]] \in [[uv uN']]$, 
+            the type corresponding to the entry $[[us0(β̂±)]]$ is well-formed in
+            context $[[Θ(β̂±)]]$ since $[[us : Θ | uv uN']]$.
+        \item $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$. This is because
+            $[[ [us0]uN0 ]] = [[ [us][σ][pas/â⁺*]uN0 ]] = [[ [us][σ]uN' ]]$.
+        \item $[[ Γ, pbs ⊢ Θ, â⁺*[Γ, pbs] ]]$ since
+            $[[Γ, pbs ⊢ Θ]]$ and $[[ {Γ, pbs} ⊆ {Γ, pbs} ]]$.
+        \item $[[Γ, pbs ; Θ, â⁺*[Γ, pbs] ⊢ uN0 ]]$
+        \item There are no negative unification variables in $[[uN0]]$ because
+            $[[ uv uN0 ⊆ uv uN ∪ {â⁺*} ]]$.
+       \end{itemize}
+
+       This way, we apply the induction hypothesis to $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$ and 
+       infer that there exists\\
+        $[[us0' :  (Θ, â⁺*[Γ, pbs]) | uv uN0]]$ such that
+       $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ uN0 ≤ iM ⫤ us0']]$,
+       and $[[Θ, â⁺*[G, pbs] ⊢ us0' ⇒ us0]]$.
+
+       This way, the algorithm terminates with the result $[[us0' \ {â⁺*}]]$
+       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁺*} ]]$. Notice that
+       $[[us0' \ {â⁺*} : Θ | uv uN' ]]$. This way, to show $[[ Θ ⊢ us ⇒ us0' \
+       {â⁺*} ]]$. it suffices to consider an unification variable $[[β̂±]] ∊
+       [[uv uN']]$ and show that $[[Γ, pbs ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
+       reflexivity of weakening (\cref{lemma:weakening-reflexivity}) since
+       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
+
+       \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+       \ruleref{\ottdruleDOneShiftDLabel} is symmetric to the corresponding negative
+        \cref{case:subt-complete-upshift}. Notice that the algorithm does not 
+        make a recursive call in this case but rather invoke the unification.
+        This way, the proof does not apply the induction hypothesis, and hence, 
+        the absence of negative unification variables in a negative type
+        ($[[α̂⁻]] \notin [[uN]]$ for $[[↓uN]] =[[uP]]$) is not required.
+
+      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+      \ruleref{\ottdruleDOneExistsLabel} is symmetric to the corresponding negative
+      \cref{case:subt-complete-forall}. Notice that we only consider the case
+      when the substitution $[[ [us]uP ]]$ results in the existential type 
+      $[[∃nas.iP']]$ (for $[[iP']] \neq [[∃]]\dots$) by congruence, 
+      i.e. $[[uP = ∃nas.uP'']]$ (for $[[uP'']] \neq [[∃]]\dots$) and $[[ [us]uP'' = iP' ]]$.
+      This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
+      (\cref{case:subt-complete-base}), and thus, the substitution $[[us]]$ must
+      preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
+
+      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
+      \cref{case:subt-complete-nvar}.
+      Notice that $[[ [us]uP = β⁺ ]]$ 
+      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
+      (\cref{case:subt-complete-base}).
     \end{caseof}

### 2023-05-02 proofs/tex/solution-merge-lemmas.tex
Add soundness and completeness of subtyping

\begin{lemma} [Soundness of Solution Merge] \label{lemma:merge-soundness}
    Suppose that $[[us1 : Θ1]]$ and $[[us2 : Θ2]]$ 
    and $[[us1 & us2]]$ is defined.
    Then 
    \begin{enumerate}
        \item $[[us1 & us2 : Θ1 ∪ Θ2]]$,
        \item $[[Θ1 ⊢ us1 & us2 ⇒ us1]]$, and
        \item $[[Θ2 ⊢ us1 & us2 ⇒ us2]]$.
    \end{enumerate}
\end{lemma}

\begin{lemma} [Completeness and Initiality of Solution Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[us1 : Θ1]]$ and $[[us2 : Θ2]]$
    and there exists $[[us]]$
    such that $[[Θ1 ⊢ us1 ⇒ us]]$ and $[[ Θ2 ⊢ us2 ⇒ us]]$.
    Then $[[us1 & us2]]$ is defined and 
    $[[ Θ1 ∪ Θ2 ⊢ us1 & us2 ⇒ us]]$.
\end{lemma}

\begin{lemma} [Solution Weakening is Monotonous]
    \label{lemma:weakening-monotonicity}
    if $[[us1 : Θ]]$ and $[[us2 : Θ]]$ and 
    $[[us1 ⊆ us2]]$ then $[[Θ ⊢ us2 ⇒ us1]]$.
\end{lemma}

\begin{lemma} [Solution Weakening is transitive]
    \label{lemma:weakening-transitivity}
    Assuming $[[us1 : Θ]]$, $[[us2 : Θ]]$ and $[[us3 : Θ]]$,
    if $[[Θ ⊢ us1 ⇒ us2]]$ and $[[Θ ⊢ us2 ⇒ us3]]$ then $[[Θ ⊢ us1 ⇒ us3]]$.
\end{lemma}

### 2023-05-02 proofs/tex/unification-lemmas.tex
Add soundness and completeness of subtyping

@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        assume there exists $[[us' : Θ | uv uP]]$ such that
-        $[[ [us']uP = iQ ]]$. Then there exists $[[us]]$ such that
-        $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$.
+        assume there exists 
+        $[[us : Θ | uv uP]]$ such that $[[ [us]uP = iQ ]]$.
+        Then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
 
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        assume there exists $[[us' : Θ | uv uN]]$ such that
-        $[[ [us']uN = iM ]]$. Then there exists $[[us]]$ such that
-        $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+        assume there exists $[[us : Θ | uv uN]]$        
+        such that $[[ [us]uN = iM ]]$.
+        Then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
    \end{itemize}

### 2023-05-18 paper/tex/abstract.tex
Add soundness of solution merge.

My work focuses on bidirectional type inference. This involves combining the
checking and inference into one algorithm. During my first year, I utilized an
approach called polarization, also known as call-by-push-value. This method
separates the system into two subsystems: positive (values) and negative
(computations). This approach crystallizes the dichotomy and unveils the
underlying dualities not only between values and computations, but also between
polymorphism and existentialism, and algorithmically, unification and
anti-unification.

### 2023-05-18 proofs/main.tex
Add soundness of solution merge.

@@
 
-\renewcommand{\ottdruleSMESupSupName}[0]{$([[≥]]\&[[≥]])$}
-\renewcommand{\ottdruleSMEEqSupName}[0]{$([[≈]]\&[[≥]])$}
-\renewcommand{\ottdruleSMESupEqName}[0]{$([[≥]]\&[[≈]])$}
-\renewcommand{\ottdruleSMEPEqEqName}[0]{$([[≈]]\&[[≈]]^{+})$}
-\renewcommand{\ottdruleSMENEqEqName}[0]{$([[≈]]\&[[≈]]^{-})$}
-
 % G ⊢ N ≤1 M
@@
 \renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}
+
+% Γ ⊢ usEntry1 & usEntry2 = usEntry3
+\renewcommand{\ottdruleSMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
+\renewcommand{\ottdruleSMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
+\renewcommand{\ottdruleSMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
+\renewcommand{\ottdruleSMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
+\renewcommand{\ottdruleSMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}
+
+% Γ ⊢ usEntry1 ⇒ usEntry2
+\renewcommand{\ottdruleSImpESupSupName}[0]{$([[≥]]\Rightarrow^{+}[[≥]])$}
+\renewcommand{\ottdruleSImpEEqSupName}[0]{$([[≈]]\Rightarrow^{+}[[≥]])$}
+\renewcommand{\ottdruleSImpEPEqEqName}[0]{$([[≈]]\Rightarrow^{+}[[≈]])$}
+\renewcommand{\ottdruleSImpENEqEqName}[0]{$([[≈]]\Rightarrow^{-}[[≈]])$}
 
@@
 
-\subsection{Unification Solution Merge}
+\subsection{Unification Solution Weakening}
 
 Unification solution is represented by a list of unification solution entries.
-Each entry restricts an unification variable in two possible ways: either stating
-that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua :≈
-iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥
-iP]]$).
+Each entry restricts an unification variable in two possible ways: either
+stating that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua
+:≈ iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥ iP]]$).
 
@@
 
+Unification solutions are preordered by the weakening relation.
+First, let us define the weakening on the unification solution entries.\\
+
+\begin{definition} \hfill \\
+\ottdefnSImpE{}\\
+\end{definition}
+Notice that  $[[Γ ⊢ usEntry1 ⇒ usEntry2]]$ means that $[[usEntry1]]$ and $[[usEntry2]]$ are matching. And 
+matching is a equivalence relation on the set of unification solutions.
+
+Next, we lift the weakening relation to unification solutions.
+Informally, $[[Θ ⊢ us1 ⇒ us2]]$ means that for any entry of $[[us2]]$,
+there is a matching entry in $[[us1]]$ that is stronger than it.
+
+\begin{definition}
+  Assuming $[[us2 : Θ]]$ and $[[us1 : Θ']] \supseteq [[Θ]]$,
+  we say $[[Θ ⊢ us1 ⇒ us2]]$ iff
+  $\forall [[usEntry2]] \in [[us2]]$, $\exists [[usEntry1]] \in [[us1]]$ s.t.
+  \begin{enumerate}
+    \item $[[usEntry1]]$ and $[[usEntry2]]$ are matching, i.e. restricting the same unification variable $[[α̂±]]$;
+    \item $[[Θ(α̂±) ⊢ usEntry1 ⇒ usEntry2]]$ (where $[[Θ(α̂±)]]$ is the context corresponding to $[[α̂±]]$ in $[[Θ]]$).
+  \end{enumerate}
+\end{definition}
+
+\subsection{Unification Solution Merge}
+
 Two matching entries can be merged in the following way:
@@
 \end{definition}
-Notice that in case of equivalence, the assigned types
-must be equal (i.e. alpha-equivalent) to be merged. This is because
-the unification algorithm assumes that every type is normalized,
-and hence, equivalence is alpha-equivalence 
-(\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
+% Notice that in case of equivalence, the assigned types
+% must be equal (i.e. alpha-equivalent) to be merged. This is because
+% the unification algorithm assumes that every type is normalized,
+% and hence, equivalence is alpha-equivalence 
+% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
 

### 2023-05-18 proofs/tex/alg-subtyping-lemmas.tex
Add soundness of solution merge.

@@
-\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
-    \begin{itemize}
-        \item [$-$] 
+\begin{lemma}[Soundness of Positive Subtyping] \label{lemma:pos-subt-soundness}
+    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
+    implies $[[us : Θ|uv uP]]$ and 
+    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
+    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
+\end{lemma}
+\begin{proof} 
+    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
+    \begin{caseof}
+    \item \ruleref{\ottdruleAPUVarLabel} then
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
+    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.
+
+    Notice that $[[â⁺[Δ] ∊ Θ]]$ and $[[Γ ⊢ Θ]]$ 
+    implies $[[Γ = Δ, pnas]]$ for some $[[pnas]]$, hence, the
+    soundness of upgrade (\cref{lemma:upgrade-soundness}) is applicable:
+    \begin{enumerate}
+        \item $[[Δ ⊢ iQ']]$ and
+        \item $[[Γ ⊢ iQ ≥ iP]]$.
+    \end{enumerate}
+
+    Since $[[â⁺[Δ] ∊ Θ|uv â⁺]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(â⁺ :≥ iQ') : Θ | uv â⁺ ]]$.
+
+    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
+    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
+    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
+    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
+    transitively (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
+    as required.
+
+    \item  \ruleref{\ottdruleAPVarLabel} then
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
+    Then $[[uv a⁺ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
+    Since $[[uv a⁺ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁺]]$, i.e.
+    $[[ [us'] a⁺ = a⁺]]$. Therefore, $[[Γ ⊢ [us']a⁺ ≥ a⁺]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
+
+    \item \label{case:pos-subt-soundness:arrow} 
+     \ruleref{\ottdruleAShiftDLabel} then
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ↓uN ≥ ↓iM ⫤ us]]$.\\
+    Then the next step of the algorithm is the unification of $[[nf(uN)]]$ and $[[nf(iM)]]$.
+    By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
+    it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uN]]$.
+    By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
+    $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uN ]]$ as substitutions.
+    $[[ [us]nf(uN) = nf(iM) ]]$ implies $[[Γ ⊢ [us]nf(uN) ≈ nf(iM)]]$, and then 
+        $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$. \ilyam{add lemmas}
+
+        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$ by 
+        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
+        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
+        $[[Γ ⊢ [us']nf(uN) ≈ [us']uN ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
+        $[[Γ ⊢ nf(iM) ≈ iM ]]$. 
+        This way, we have $[[Γ ⊢ [us']uN ≈ iM]]$.
+        Then by \ruleref{\ottdruleDOneShiftULabel}
+        and congruence of substitution, $[[Γ ⊢ [us']↓uN ≥ ↓iM]]$.
+
+    \item \label{case:pos-subt-soundness:exists}
+       \ruleref{\ottdruleAExistsLabel} then
+        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ∃nas.uP' ≥ ∃nbs.iQ' ⫤ us]]$ s.t. either $[[nas]]$ or $[[nbs]]$ is not empty.\\
+        Then the algorithm creates fresh unification variables $[[â⁻*[Γ,nbs] ]]$, 
+        substitutes the old $[[nas]]$ with them in $[[uP']]$, and makes the recursive call:
+        $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP ≥ iQ ⫤ us']]$, returning as the result
+        $[[us]] = [[us' \ {α̂⁻*}]]$.
+
+        Notice that $[[Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$, $[[Γ,nbs ⊢ iQ']]$, and 
+        $[[Γ,nbs; Θ, â⁻*[Γ, nbs] ⊢ [â⁻*/nas] uP' ]]$, so the induction hypothesis is applicable,
+        that is $[[us' : Θ | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
+        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.
+
+        Since the domain of $[[us']]$ is $[[uv [â⁺*/pas]uN']]$, the domain of 
+        $[[us]] = [[us' \  {α̂⁺*}]]$ is $[[uv [â⁺*/pas]uN \ {â⁺*} = uv uN']] = [[uv ∀pas.uN']]$,
+        this way, $[[us : Θ | uv ∀pas.uN']]$, as required.
+
+        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
+        Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
+        extending $[[us2]]$ to the domain $[[Ord uv [â⁺*/pas]uN']]$ with the values of $[[us']]$,
+        i.e.  $[[us'2 : Θ | uv [â⁺*/pas]uN']]$, 
+        and
+        \[
+            [[us'2(β̂±)]]  = 
+            \begin{cases}
+               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']] \\
+               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁺*]]
+            \end{cases}
+        \] 
+        , where the application of the unification solution to a variable returns the 
+        corresponding \emph{unification entry}. 
+        Notice that $[[Sub us'2|uv uN']] = [[us2]]$.
+    It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
+    \begin{enumerate}
+        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
+        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
+    \end{enumerate}
+    which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
+    $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$.\\
+    Notice that
+    $
+    \begin{aligned}[t]
+                 [[ [us'2][â⁺*/pas]uN' ]] &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][Sub us'2|uv uN']uN' ]]
+                                          && \text{by substitution properties \ilyam{todo}}\\
+                                          &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ]]
+                                          && \text{since $[[Sub us'2|uv uN']] = [[us2]]$}.
+    \end{aligned}
+    $\\
+    Also notice that the domain of $[[Sub us'2|{â⁺*} ○ â⁺*/pas]]$ is $[[pas]]$,
+    and the range is $[[Γ, pbs]]$, i.e. $[[Γ, pbs ⊢ Sub us'2|{â⁺*} ○ â⁺*/pas : pas]]$, 
+    which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
+    $[[ Γ, pbs ⊢ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ≤ iM' ]]$
+    to get $[[ Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM' ]]$, as required.
+
+    \end{caseof}
+\end{proof}
+
+\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
         If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
@@
         $[[ Γ ⊢ [us']uN ≤ iM ]]$
-
-        \item [$+$] 
-        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
-        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
-        implies $[[us : Θ|uv uP]]$ and 
-        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
-        $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
-     \end{itemize}
-\end{theorem}
+\end{lemma}
 \begin{proof}
-    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ (and mutually, on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$).
+    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$.
     Let us consider the last rule to infer this judgment. 
@@
         $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
-        By the induction hypothesis, 
+        By the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}) and induction hypothesis, respectively 
         \begin{enumerate}
@@
         Notice that $[[Γ, pbs ⊢ Θ, â⁺*[G, pbs] ]]$, $[[Γ,pbs ⊢ iM']]$, and 
-        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN ]]$, so the induction hypothesis is applicable,
+        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN' ]]$, so the induction hypothesis is applicable,
         that is $[[us' : Θ | uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any
@@
 
-    \item $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
-    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$\\
-
-    Notice that $[[â⁺[Δ] ∊ Θ]]$ and $[[Γ ⊢ Θ]]$ 
-    implies $[[Γ = Δ, pnas]]$ for some $[[pnas]]$, hence, the
-    soundness of upgrade (\cref{lemma:upgrade-soundness}) is applicable:
-    \begin{enumerate}
-        \item $[[Δ ⊢ iQ']]$ and
-        \item $[[Γ ⊢ iQ ≥ iP]]$.
-    \end{enumerate}
-
-    Since $[[â⁺[Δ] ∊ Θ|uv â⁺]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(â⁺ :≥ iQ') : Θ | uv â⁺ ]]$.
-
-    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
-    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
-    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
-    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
-    transitivity (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
-    as required.
-
-    \item For the other positive cases, the proof is symmetric to the corresponding negative cases.
 \end{caseof}
@@
 
-Notice that the induction statement in following theorem is not symmetric for the positive and negative cases.
-\begin{theorem}[Completeness and Initiality of the Subtyping Algorithm]
-    \hfill
-    \begin{itemize}
-        \item [$-$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
-        $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
-        and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-        Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
-        and $[[Θ ⊢ us ⇒ us']]$.
+\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
+\end{theorem}
+\begin{proof}
+\end{proof}
 
-        \item [$+$] Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
+\begin{lemma}[Completeness of the Positive Subtyping]
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
     there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
@@
     and $[[Θ ⊢ us ⇒ us']]$.
-    \end{itemize}
-\end{theorem}
+\end{lemma}
 \begin{proof}
-    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$ (and mutually, on $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
-    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$ or $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
+    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
+    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
     but first consider the base case for the substitution $[[ [us]uP ]]$:
+
     \begin{caseof}
@@
 
+       \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+       \ruleref{\ottdruleDOneShiftDLabel} is symmetric to the corresponding negative
+        \cref{case:subt-complete-upshift}. Notice that the algorithm does not 
+        make a recursive call in this case but rather invoke the unification.
+        This way, the proof does not apply the induction hypothesis, and hence, 
+        the absence of negative unification variables in a negative type
+        ($[[α̂⁻]] \notin [[uN]]$ for $[[↓uN]] =[[uP]]$) is not required.
+
+      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+      \ruleref{\ottdruleDOneExistsLabel} is symmetric to the corresponding negative
+      \cref{case:subt-complete-forall}. Notice that we only consider the case
+      when the substitution $[[ [us]uP ]]$ results in the existential type 
+      $[[∃nas.iP']]$ (for $[[iP']] \neq [[∃]]\dots$) by congruence, 
+      i.e. $[[uP = ∃nas.uP'']]$ (for $[[uP'']] \neq [[∃]]\dots$) and $[[ [us]uP'' = iP' ]]$.
+      This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
+      (\cref{case:subt-complete-base}), and thus, the substitution $[[us]]$ must
+      preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
+
+      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
+      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
+      \cref{case:subt-complete-nvar}.
+      Notice that $[[ [us]uP = β⁺ ]]$ 
+      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
+      (\cref{case:subt-complete-base}).
+
+    \end{caseof}
+\end{proof}
+
+\begin{lemma}[Completeness of the Negative Subtyping]
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
+    $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
+    and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
+    and $[[Θ ⊢ us ⇒ us']]$.
+\end{lemma}
+\begin{proof}
+    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+    \begin{caseof}
         \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel},
@@
        $[[Θ ⊢ us ⇒ us | uv uP]]$,
-       and then by \cref{lemma:weakening-transitivity}, 
-       $[[Θ ⊢ us | uv uP ⇒ us'1]]$.
-       Analogously, $[[Θ ⊢ us | uv uN' ⇒ us'2]]$.
+       and then by transitivity (\cref{lemma:weakening-transitivity}), 
+       $[[Θ ⊢ us ⇒ us'1]]$.
+       Analogously, $[[Θ ⊢ us ⇒ us'2]]$.
        Then since by \cref{lemma:merge-completeness}, 
@@
        $[[us0'(β̂±)]] = [[us(β̂±)]]$.
-
-       \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-       \ruleref{\ottdruleDOneShiftDLabel} is symmetric to the corresponding negative
-        \cref{case:subt-complete-upshift}. Notice that the algorithm does not 
-        make a recursive call in this case but rather invoke the unification.
-        This way, the proof does not apply the induction hypothesis, and hence, 
-        the absence of negative unification variables in a negative type
-        ($[[α̂⁻]] \notin [[uN]]$ for $[[↓uN]] =[[uP]]$) is not required.
-
-      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-      \ruleref{\ottdruleDOneExistsLabel} is symmetric to the corresponding negative
-      \cref{case:subt-complete-forall}. Notice that we only consider the case
-      when the substitution $[[ [us]uP ]]$ results in the existential type 
-      $[[∃nas.iP']]$ (for $[[iP']] \neq [[∃]]\dots$) by congruence, 
-      i.e. $[[uP = ∃nas.uP'']]$ (for $[[uP'']] \neq [[∃]]\dots$) and $[[ [us]uP'' = iP' ]]$.
-      This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
-      (\cref{case:subt-complete-base}), and thus, the substitution $[[us]]$ must
-      preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
-
-      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
-      \cref{case:subt-complete-nvar}.
-      Notice that $[[ [us]uP = β⁺ ]]$ 
-      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
-      (\cref{case:subt-complete-base}).
     \end{caseof}

### 2023-05-18 proofs/tex/equiv-lemmas.tex
Add soundness of solution merge.

@@
+\begin{lemma}[Declarative equivalence is transitive]
+  \hfill
+  \label{lemma:decl-equiv-transitivity}
+  \begin{itemize}
+  \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
+  \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  \ilyam{should be easy to do by induction since the types are getting smaller}
+\end{proof}
+\begin{lemma}[Algorithmization of declarative equivalence]
+  \label{lemma:decl-equiv-algorithmization}
+  Declarative equivalence is equality of normal forms. 
+  \begin{itemize}
+    \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
+    \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof} \hfill
+  \begin{itemize}
+    \item[$+$] Let us prove both directions separately.
+    \begin{itemize}
+      \item[$\Rightarrow$] 
+        exactly by \cref{lemma:normalization-completeness},
+      \item[$\Leftarrow$] 
+        from \cref{lemma:normalization-soundness}, we know
+        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
+        $[[iP ≈ iQ]]$.
+    \end{itemize}
+    \item[$-$] The proof is exactly the same.
+  \end{itemize}
+\end{proof}
+
 \begin{lemma}[Type well-formedness is invariant under equivalence]
@@
 \end{proof}
+
+\begin{lemma}[Algorithmization of subtyping-induced equivalence]
+  \label{lemma:subt-equiv-algorithmization}
+  Mutual subtyping is equality of normal forms.
+ Assuming all the types below are well-formed in $[[Γ]]$:
+  \begin{itemize}
+    \item [$+$] $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
+    \item [$-$] $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Let us prove the positive case, the negative case is symmetric.
+  We prove both directions of $\iff$ separately:
+  \begin{itemize}
+    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
+    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
+  \end{itemize}
+\end{proof}

### 2023-05-18 proofs/tex/solution-merge-lemmas.tex
Add soundness of solution merge.

@@
+\begin{lemma} \label{lemma:entry-weakening-preorder}
+    Given a fixed context $[[Γ]]$, weakening forms a preorder on the set of entries well-formed in $[[Γ]]$.
+\end{lemma}
+\begin{proof} \hfill 
+  \begin{itemize}
+    \item Reflexivity: $[[Γ ⊢ usEntry ≈ usEntry]]$ then $[[Γ ⊢ usEntry ⇒ usEntry]]$.
+    Let us consider the shape of $[[usEntry]]$. In all cases, there is a rule inferring
+     $[[Γ ⊢ usEntry ⇒ usEntry]]$, since $[[Γ ⊢ iP ≈ iQ]]$ implies $[[Γ ⊢ iP ≥ iQ]]$ by inversion.
+    \item Transitivity: $[[Γ ⊢ usEntry1 ⇒ usEntry2]]$ and $[[Γ ⊢ usEntry2 ⇒ usEntry3]]$ implies $[[Γ ⊢ usEntry1 ⇒ usEntry3]]$.
+    It follows immediately by considering the rules inferring 
+    $[[Γ ⊢ usEntry1 ⇒ usEntry2]]$ and $[[Γ ⊢ usEntry2 ⇒ usEntry3]]$, and applying the 
+    transitivity of subtyping (\cref{corollary:subtyping-transitivity}).
+  \end{itemize}
+\end{proof}
+
+\begin{lemma} [Solution Weakening forms a preorder]
+    \label{lemma:solution-weakening-preorder}
+    Let us consider a set of pairs $([[Θ]], [[us]])$ such that $[[us : Θ]]$.
+    Then the relation defined as 
+    $([[Θ1]], [[us1]]) \Rightarrow ([[Θ2]], [[us2]])$ iff $[[Θ1]] \supseteq [[Θ2]]$ 
+    and $[[Θ2 ⊢ us1 ⇒ us2]]$ forms a preorder.
+\end{lemma}
+\begin{proof} \hfill
+    \begin{itemize}
+        \item Reflexivity: $([[Θ]], [[us]]) \Rightarrow ([[Θ]], [[us]])$.\\ It is clear that $[[Θ]] \supseteq [[Θ]]$.
+        $[[Θ ⊢ us ⇒ us]]$ holds since for any $[[usEntry]]$ in $[[us]]$ restricting $[[α̂±]]$, $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry]]$  
+        by \cref{lemma:entry-weakening-preorder}
+        \item Transitivity: If $([[Θ1]], [[us1]]) \Rightarrow ([[Θ2]], [[us2]])$ and $([[Θ2]], [[us2]]) \Rightarrow ([[Θ3]], [[us3]])$
+        then $([[Θ1]], [[usEntry1]]) \Rightarrow ([[Θ3]], [[usEntry3]])$.\\
+        It is clear that since $[[Θ1]] \subseteq [[Θ2]]$ and $[[Θ2]] \subseteq
+        [[Θ3]]$ then $[[Θ1]] \subseteq [[Θ3]]$. Let us consider $[[usEntry3]]
+        \in [[us3]]$. Then there exists $[[usEntry2]] \in [[us2]]$ such that
+        $[[Θ3 ⊢ usEntry2 ⇒ usEntry3]]$. Also there exists $[[usEntry1]] \in [[us1]]$ 
+        such that $[[Θ2 ⊢ usEntry1 ⇒ usEntry2]]$, and by weakening,
+        $[[Θ3 ⊢ usEntry1 ⇒ usEntry2]]$. Then by transitivity of solution entry
+        weakening (\cref{lemma:entry-weakening-preorder}), $[[Θ3 ⊢ usEntry1 ⇒
+        usEntry3]]$.
+    \end{itemize}
+\end{proof}
+
+\begin{corollary} [Solution Weakening is transitive]
+    \label{lemma:weakening-transitivity}
+    If $[[Θ ⊢ us1 ⇒ us2]]$ and $[[Θ ⊢ us2 ⇒ us3]]$ then $[[Θ ⊢ us1 ⇒ us3]]$.
+\end{corollary}
+
+\begin{lemma} [Soundness of Merge of Unification Solution Entries]
+    \label{lemma:unif-sol-ent-merge-soundness}
+    For a fixed context $[[Γ]]$,
+    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$,
+    where $[[Γ ⊢ usEntryi]]$ is an equivalence-shaped restriction
+    (i.e. it has shape $[[α̂⁺ :≈ iP]]$ or $[[α̂⁻ :≈ iN]]$ but not $[[α̂⁺ :≥ iP]]$).
+    If $[[usEntry1 & usEntry2]]$ is defined then
+    \begin{enumerate}
+    \item $[[Γ ⊢ usEntry1 & usEntry2]]$
+    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry1]]$
+    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry2]]$
+    \end{enumerate}
+\end{lemma}
+\begin{proof}
+    Let us consider the rule forming $[[Γ ⊢ usEntry1 & usEntry2 = usEntry]]$.
+    \begin{caseof}
+        \item \ruleref{\ottdruleSMEPEqEqLabel}, i.e. 
+        $[[Γ ⊢ usEntry1 & usEntry2 = usEntry]]$
+        has form $[[Γ ⊢ (pua :≈ iP) & (pua :≈ iP') = (pua :≈ iP)]]$
+        and $[[nf(iP) = nf(iP')]]$. Then
+         \begin{enumerate}
+            \item $[[Γ ⊢ usEntry]]$, i.e. $[[Γ ⊢ pua :≈ iP]]$ holds by assumption;
+            \item $[[Γ ⊢ (pua :≈ iP) ⇒ (pua :≈ iP)]]$ holds by reflexivity (\cref{lemma:entry-weakening-preorder});
+            \item $[[Γ ⊢ (pua :≈ iP) ⇒ (pua :≈ iP')]]$ holds
+            because $[[nf(iP) = nf(iP')]]$ implies $[[Γ ⊢ iP ≈ iP']]$
+            (by \cref{lemma:subt-equiv-algorithmization}), and thus, 
+            \ruleref{\ottdruleSImpEPEqEqLabel} applies.
+         \end{enumerate}
+        \item \ruleref{\ottdruleSMENEqEqLabel}. The negative case is proved in exactly the same way.
+    \end{caseof}
+\end{proof}
+
+\begin{lemma} [Soundness of Merge of Unification Solutions]
+    \label{lemma:unif-sol-merge-soundness}
+    Suppose that $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$ 
+    are unification solutions (i.e. $[[us1]]$ and $[[us2]]$ can only have equivalence-shaped restrictions).
+    If $[[us1 & us2]]$ is defined then
+    \begin{enumerate}
+        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$,
+        \item $[[Θ|varset1 ⊢ us1 & us2 ⇒ us1]]$, and
+        \item $[[Θ|varset2 ⊢ us1 & us2 ⇒ us2]]$.
+    \end{enumerate}
+\end{lemma}
+\begin{proof}
+    Let us prove the properties separately:
+    \begin{enumerate}
+        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$.
+        It suffices to prove the following two properties:
+        \begin{itemize}
+            \item The set of variables of the entries of $[[us1 & us2]]$ 
+            coincides with $[[varset1 ∪ varset2]]$\\
+            By definition, $[[us1 & us2]]$ consists of three parts:
+            entries of $[[us1]]$ that do not have matching entries of $[[us2]]$,
+            entries of $[[us2]]$ that do not have matching entries of $[[us1]]$,
+            and the merge of matching entries.
+            It means that $[[dom(us1 & us2)]] = [[dom(us1) \ dom(us2) ∪ dom(us2) \ dom(us1) ∪ 
+            dom(us1) ∩ dom(us2)]] = [[dom(us1) ∪ dom(us2)]]$, which, since 
+            $[[usi : Θ | varseti]]$, is equal to $[[varset1 ∪ varset2]]$.
+
+            \item Each entry of $[[us1 & us2]]$ restricting $[[α̂±]]$ is well-formed in
+            the corresponding context $[[Θ(α̂±)]]$.\\
+            Let us consider an arbitrary entry $[[usEntry]]$ of $[[us1 & us2]]$ restricting
+            $[[α̂±]]$. Then there are three cases:
+            \begin{enumerate}
+                \item $[[usEntry]]$ the entry is from $[[us1]]$ and does not have a matching entry 
+                in $[[us2]]$, i.e. $[[α̂±]] \in [[dom(us1) \ dom(us2)]]$.
+                Then $[[usEntry]]$ is well-formed in $[[Θ | varset1]]$ by assumption.
+                \item $[[usEntry]]$ the entry is from $[[us2]]$ and does not
+                have a matching entry in $[[us1]]$. This case is symmetric.
+                \item $[[usEntry]]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
+                 and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
+                 Since $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$,
+                 $[[α̂±]] \in [[dom(Θ|varset1) ∩ dom(Θ|varset2)]]$, i.e. there 
+                 is an entry $[[ α̂±[Γ] ]] \in [[Θ]]$, and 
+                 $[[usEntry1]]$ and $[[usEntry2]]$ are well-formed in $[[Γ]]$.
+                 Then by \cref{lemma:unif-sol-ent-merge-soundness}, 
+                 $[[Γ ⊢ usEntry1 & usEntry2]]$, where $[[Γ]] = [[Θ(α̂±)]]$.
+            \end{enumerate}
+        \end{itemize}
+        \item $[[Θ|varset1 ⊢ us1 & us2 ⇒ us1]]$
+        We need to show that for every entry $[[usEntry]]$ from $[[us1]]$,
+        there is an entry $[[usEntry']]$ from $[[us1 & us2]]$ such that
+        $[[Θ|varset1 ⊢ usEntry' ⇒ usEntry]]$. 
+        Let us consider an arbitrary $[[usEntry]]$ from $[[us1]]$ restricting $[[α̂±]]$.
+        Then there are two cases:
+        \begin{itemize}
+            \item $[[usEntry]]$ does not have a matching entry in $[[us2]]$.
+            Then $[[usEntry]]$ is also in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry ⇒ usEntry]]$ 
+            by reflexivity (\cref{lemma:entry-weakening-preorder}).
+            \item $[[usEntry]]$ has a matching entry $[[usEntry']]$ in $[[us2]]$.
+            Then $[[usEntry & usEntry']]$ is in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry & usEntry' ⇒ usEntry]]$ by \cref{lemma:unif-sol-ent-merge-soundness}.
+        \end{itemize}
+        \item $[[Θ|varset2 ⊢ us1 & us2 ⇒ us2]]$ is proved analogously.
+    \end{enumerate}
+\end{proof}
+
+\begin{lemma} [Soundness of Solution Entry Merge]
+\label{lemma:entry-merge-soundness}
+For a fixed context $[[Γ]]$,
+suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$. 
+If $[[usEntry1 & usEntry2]]$ is defined then
+\begin{enumerate}
+    \item $[[Γ ⊢ usEntry1 & usEntry2]]$
+    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry1]]$
+    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry2]]$
+\end{enumerate}
+\end{lemma}
+\begin{proof}
+    Let us consider the rule forming $[[Γ ⊢ usEntry1 & usEntry2 = usEntry]]$.
+    \begin{caseof}
+        \item \ruleref{\ottdruleSMEPEqEqLabel} or \ruleref{\ottdruleSMENEqEqLabel}
+        The proof is analogous to the proof of \cref{lemma:unif-sol-ent-merge-soundness}.
+        \item \ruleref{\ottdruleSMESupSupLabel} 
+        Then $[[usEntry1]]$ is $[[pua :≥ iP1]]$, $[[usEntry2]]$ is $[[pua :≥ iP2]]$,
+        and $[[usEntry1 & usEntry2]]$ is $[[pua :≥ iQ]]$ where $[[iQ]]$ is the least upper bound of $[[iP1]]$ and $[[iP2]]$.
+        Then by \cref{lemma:lub-soundness},
+        \begin{itemize}
+            \item $[[Γ ⊢ iQ]]$, and hence $[[Γ ⊢ pua :≥ iQ]]$, that is $[[Γ ⊢ usEntry1 & usEntry2]]$,
+            \item $[[Γ ⊢ iQ ≥ iP1]]$, and hence $[[Γ ⊢ pua :≥ iQ ⇒ pua :≥ iP1]]$, that is $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry1]]$,
+            \item $[[Γ ⊢ iQ ≥ iQ1]]$, and hence $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry2]]$.
+        \end{itemize}
+        \item \ruleref{\ottdruleSMESupEqLabel}
+        Then $[[usEntry1]]$ is $[[pua :≥ iP]]$, $[[usEntry2]]$ is $[[pua :≈ iQ]]$, 
+        where $[[Γ;· ⊨ uQ ≥ iP ⫤ us']]$, and the resulting   
+        $[[usEntry1 & usEntry2]]$ is equal to $[[usEntry2]]$, that is $[[pua :≈ iQ]]$.
+
+        \begin{itemize}
+            \item By assumption, $[[Γ ⊢ iQ]]$, and hence $[[Γ ⊢ pua :≈ iQ]]$, that is $[[Γ ⊢ usEntry1 & usEntry2]]$.
+            \item Since $[[uv(uQ) = ∅]]$, 
+                $[[Γ;· ⊨ uQ ≥ iP ⫤ us']]$ implies $[[Γ ⊢ iQ ≥ iP]]$
+                by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), 
+                which means $[[Γ ⊢ pua :≈ iQ ⇒ pua :≥ iP]]$, that is $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry1]]$.
+            \item  $[[Γ ⊢ pua :≈ iQ ⇒ pua :≈ iQ]]$ by reflexivity
+             (\cref{lemma:entry-weakening-preorder}), and hence
+              $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry2]]$.
+        \end{itemize}
+        \item \ruleref{\ottdruleSMEEqSupLabel} Thee proof is analogous to the previous case.
+    \end{caseof}
+\end{proof}
+
 \begin{lemma} [Soundness of Solution Merge] \label{lemma:merge-soundness}
@@
 \end{lemma}
+\begin{proof}
+    The proof repeats the proof of \cref{lemma:unif-sol-merge-soundness},
+    but uses \cref{lemma:entry-merge-soundness} instead of 
+    \cref{lemma:unif-sol-ent-merge-soundness}.
+\end{proof}
 
@@
 \end{lemma}
-
-\begin{lemma} [Solution Weakening is transitive]
-    \label{lemma:weakening-transitivity}
-    Assuming $[[us1 : Θ]]$, $[[us2 : Θ]]$ and $[[us3 : Θ]]$,
-    if $[[Θ ⊢ us1 ⇒ us2]]$ and $[[Θ ⊢ us2 ⇒ us3]]$ then $[[Θ ⊢ us1 ⇒ us3]]$.
-\end{lemma}

### 2023-05-19 proofs/tex/alg-subtyping-lemmas.tex
Add merge completeness

@@
 
-\begin{lemma}[Completeness of the Positive Subtyping]
+\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and

### 2023-05-19 proofs/tex/solution-merge-lemmas.tex
Add merge completeness

@@
 \end{corollary}
+
+\begin{lemma} [Solution Weakening is Monotonous]
+    \label{lemma:weakening-monotonicity}
+    if $[[us1 : Θ1]]$, $[[us2 : Θ2]]$,
+    and $[[us1 ⊆ us2]]$ then $[[Θ1 ⊢ us2 ⇒ us1]]$.
+\end{lemma}
+\begin{proof}
+    We need to show that for any $[[usEntry]] \in [[us1]]$,
+    there exists $[[usEntry']] \in [[us2]]$ such that 
+    $[[Θ1 ⊢ usEntry' ⇒ usEntry]]$.
+    Let us consider an arbitrary $[[usEntry]] \in [[us1]] \subseteq [[us2]]$
+    restricting $[[α̂±]]$. 
+    Notice that $[[Θ1(α̂±) ⊢ usEntry ]]$ since 
+    $[[usEntry]] \in [[us1 : Θ1]]$.
+    By \cref{lemma:entry-weakening-preorder}, $[[Θ1 ⊢ usEntry ⇒ usEntry]]$. 
+\end{proof}
 
@@
 
-\begin{lemma} [Completeness and Initiality of Solution Merge] 
+\begin{lemma} [Completeness of Unification Solution Entry Merge]
+    \label{lemma:unif-entry-merge-completeness}
+    For a fixed context $[[Γ]]$,
+    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$, 
+    where $[[Γ ⊢ usEntryi]]$ are equivalence-shaped restrictions.
+    Suppose there exists an equivalence-shaped restriction 
+    $[[Γ ⊢ usEntry]]$ such that $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$.
+    Then $[[usEntry1 & usEntry2]]$ is defined and
+    $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
+\end{lemma}
+\begin{proof}
+    By the soundness of unification solution entry merge (\cref{lemma:unif-sol-ent-merge-soundness}),
+    Since the restrictions are equivalence-shaped, 
+    the only possible rules inferring the weakening judgments are
+    the positive \ruleref{\ottdruleSImpEPEqEqLabel} and the negative \ruleref{\ottdruleSImpENEqEqLabel}.
+    Since $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$ 
+    have the same polarity, we consider the positive case,
+    and the negative case is proved analogously.
+
+    Let us assume that $[[usEntry1]]$ is $[[pua :≈ iP1]]$,
+    $[[usEntry2]]$ is $[[pua :≈ iP2]]$, $[[usEntry]]$ is $[[pua :≈ iQ]]$.
+    Then by inversion of \ruleref{\ottdruleSImpEPEqEqLabel},
+    \begin{itemize}
+        \item $[[Γ ⊢ iQ ≈ iP1]]$, and hence $[[nf(iQ) = nf(iP1)]]$ by \cref{lemma:subt-equiv-algorithmization},
+        \item $[[Γ ⊢ iQ ≈ iP2]]$, and hence $[[nf(iQ) = nf(iP2)]]$ by \cref{lemma:subt-equiv-algorithmization},
+    \end{itemize}
+    which implies $[[nf(iP1) = nf(iP2)]]$.
+    It means that the merge of $[[usEntry1]]$ and $[[usEntry2]]$ is defined by \ruleref{\ottdruleSMEPEqEqLabel}:
+    $[[Γ ⊢ (pua :≈ iP1) & (pua :≈ iP2) = (pua :≈ iP1)]]$.
+
+    To show that $[[Γ ⊢ (pua :≈ iQ) ⇒ (pua :≈ iP1)]]$, we apply \ruleref{\ottdruleSImpEPEqEqLabel}:
+    $[[Γ ⊢ iP1 ≈ iQ]]$ is obtained from $[[Γ ⊢ iQ ≈ iP1]]$ by symmetry of equivalence.
+\end{proof}
+
+\begin{lemma} [Completeness of Unification Solution Merge] 
+    \label{lemma:unif-sol-merge-completeness}
+    Suppose that $[[us1 : Θ|varset1]]$ and $[[us2 : Θ|varset2]]$ 
+    are equivalence-only unification solutions
+    and there exists $[[us]]$
+    such that $[[Θ ⊢ us ⇒ us1]]$ and $[[ Θ ⊢ us ⇒ us2]]$.
+    Then $[[us1 & us2]]$ is defined and 
+    $[[ Θ ⊢ us ⇒ us1 & us2 ]]$.
+\end{lemma}
+\begin{proof}
+    By  definition, $[[us1 & us2]]$ is a union of entries of $[[us1]]$, which do not have matching entries in $[[us2]]$,
+    entries of $[[us2]]$, which do not have matching entries in $[[us1]]$, and the merge of matching entries.
+    It is clear that the first two components of this union exist. Let us show that the third component exists, 
+    i.e. that for every two entries $[[usEntry1]] \in [[us1]]$ and $[[usEntry2]] \in [[us2]]$
+    restricting the same variable $[[α̂±]]$, $[[usEntry1 & usEntry2]]$ is defined.
+
+    $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ restricting $[[α̂±]]$ such that
+    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1]]$. Similarly, since there can be only one entry in $[[usEntry]]$ restricting 
+    the same variable$[[α̂±]]$, $[[Θ2 ⊢ us ⇒ us2]]$ means $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry2]]$.
+    Then by \cref{lemma:unif-sol-ent-merge-soundness}, $[[usEntry1 & usEntry2]]$ is defined and 
+    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
+
+    Let us show: $[[ Θ ⊢ us ⇒ us1 & us2 ]]$, i.e. for every $[[usEntry']] \in [[us1 & us2]]$ restricting 
+    $[[α̂±]]$, there exists $[[usEntry]] \in [[us]]$ such that $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$.
+    Let us consider three cases:
+    \begin{caseof}
+        \item $[[usEntry']]$ is from $[[us1]]$ and does not have a matching entry in $[[us2]]$.
+        Then $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ such that
+        $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$, as required.
+        \item $[[usEntry']]$ is from $[[us2]]$ and does not have a matching entry in $[[us1]]$.
+        This case is symmetric to the previous one.
+        \item $[[usEntry']]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
+        and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
+        Then as noted previously, $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]] = [[usEntry']]$.
+    \end{caseof}
+\end{proof}
+
+\begin{lemma} [Completeness of Solution Entry Merge]
+    \label{lemma:entry-merge-completeness}
+    For a fixed context $[[Γ]]$,
+    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$.
+    Suppose there exists an entry $[[Γ ⊢ usEntry]]$ such that $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$.
+    Then $[[usEntry1 & usEntry2]]$ is defined and
+    $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
+\end{lemma}
+\begin{proof}
+    Notice that $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$
+    restrict the same variable and have the same polarity. If the polarity is negative,
+    then the proof is analogous to the proof of \cref{lemma:unif-entry-merge-completeness}.
+
+    Let us consider the shape of $[[usEntry]]$.
+    \begin{caseof}
+        \item $[[usEntry]]$ is $[[pua :≥ iP]]$.
+        \label{case:entry-merge-completeness-sup-sup}
+         Then the only rule inferring 
+        $[[Γ ⊢ usEntry ⇒ usEntryi]]$ is \ruleref{\ottdruleSImpESupSupLabel}.
+        This way, $[[usEntry1]]$ is $[[pua :≥ iP1]]$ and $[[usEntry2]]$ is $[[pua :≥ iP2]]$,
+        and $[[iP]]$ is a common supertype of $[[iP1]]$ and $[[iP2]]$, i.e. $[[Γ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.
+        Then by \cref{lemma:lub-completeness}, $[[iP1 ∨ iP2]]$ is defined and $[[Γ ⊢ iP ≥ iP1 ∨ iP2]]$.
+        Then $[[usEntry1 & usEntry2]]$, that is $[[pua :≥ iP1 & pua :≥ iP1]]$, is defined as 
+        $[[pua :≥ iP1 ∨ iP2]]$ by \ruleref{\ottdruleSMESupSupLabel}.
+        Moreover, $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$, that is $[[Γ ⊢ pua :≥ iP ⇒ pua :≥ iP1 ∨ iP2]]$,
+        holds by \ruleref{\ottdruleSImpESupSupLabel}.
+
+        \item $[[usEntry]]$ is $[[pua :≈ iP]]$. Then let us consider the rules inferring
+        $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$ respectively:
+        \begin{caseof}
+            \item \ruleref{\ottdruleSImpEPEqEqLabel} and \ruleref{\ottdruleSImpEPEqEqLabel}.
+            Then $[[usEntry1]]$ is $[[pua :≈ iP1]]$, $[[usEntry2]]$ is $[[pua :≈ iP2]]$, 
+            where $[[Γ ⊢ iP ≈ iP1]]$ and $[[Γ ⊢ iP ≈ iP2]]$.
+            Then $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≈ iP1]]$ by \ruleref{\ottdruleSMEPEqEqLabel}, 
+            and $[[Γ ⊢ pua :≈ iP ⇒ pua :≈ iP1]]$ holds by \ruleref{\ottdruleSImpEPEqEqLabel},
+            since $[[Γ ⊢ iP ≈ iP1]]$.
+
+            \item \ruleref{\ottdruleSImpEPEqEqLabel} and \ruleref{\ottdruleSImpEEqSupLabel}.
+            Then $[[usEntry1]]$ is $[[pua :≈ iP1]]$, $[[usEntry2]]$ is $[[pua :≥ iP2]]$, 
+            where $[[Γ ⊢ iP ≈ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.
+
+            Then by transitivity of subtyping (\cref{corollary:subtyping-transitivity}),
+            $[[Γ ⊢ iP1 ≥ iP2]]$. The completeness of positive subtyping algorithm 
+            (\cref{lemma:pos-subt-completeness}) implies that
+            $[[Γ;· ⊨ uP1 ≥ iP2 ⫤ us']]$.
+            Then $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≈ iP1]]$ by \ruleref{\ottdruleSMEPEqEqLabel}.
+            Moreover, $[[Γ ⊢ pua :≈ iP ⇒ pua :≈ iP1]]$ holds by \ruleref{\ottdruleSImpEPEqEqLabel}, 
+            as required.
+
+            \item \ruleref{\ottdruleSImpEEqSupLabel} and \ruleref{\ottdruleSImpEPEqEqLabel}.
+            This case is symmetric to the previous one.
+
+            \item \ruleref{\ottdruleSImpEEqSupLabel} and \ruleref{\ottdruleSImpEEqSupLabel}.
+            Then $[[usEntry1]]$ is $[[pua :≥ iP1]]$, $[[usEntry2]]$ is $[[pua :≥ iP2]]$, 
+            where $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$.
+            And similarly to \cref{case:entry-merge-completeness-sup-sup},
+            $[[usEntry1 & usEntry2]]$ is defined as $[[pua :≥ iP1 ∨ iP2]]$.
+            Moreover, $[[Γ ⊢ pua :≈ iP ⇒ pua :≥ iP1 ∨ iP2]]$ holds by \ruleref{\ottdruleSImpEEqSupLabel},
+            since $[[Γ ⊢ iP ≥ iP1 ∨ iP2]]$ by completeness of the least upper bound algorithm (\cref{lemma:lub-completeness}).
+        \end{caseof}
+    \end{caseof}
+\end{proof}
+
+\begin{lemma} [Completeness of Solution Merge] 
     \label{lemma:merge-completeness}
-    Suppose that $[[us1 : Θ1]]$ and $[[us2 : Θ2]]$
+    Suppose that $[[us1 : Θ|varset1]]$ and $[[us2 : Θ|varset2]]$
     and there exists $[[us]]$
-    such that $[[Θ1 ⊢ us1 ⇒ us]]$ and $[[ Θ2 ⊢ us2 ⇒ us]]$.
+    such that $[[Θ ⊢ us ⇒ us1]]$ and $[[Θ ⊢ us ⇒ us2]]$.
     Then $[[us1 & us2]]$ is defined and 
-    $[[ Θ1 ∪ Θ2 ⊢ us1 & us2 ⇒ us]]$.
+    $[[ Θ ⊢ us ⇒ us1 & us2]]$.
 \end{lemma}
-
-\begin{lemma} [Solution Weakening is Monotonous]
-    \label{lemma:weakening-monotonicity}
-    if $[[us1 : Θ]]$ and $[[us2 : Θ]]$ and 
-    $[[us1 ⊆ us2]]$ then $[[Θ ⊢ us2 ⇒ us1]]$.
-\end{lemma}
+\begin{proof}
+    The proof is analogous to the proof of \cref{lemma:unif-sol-merge-completeness}.
+    The only difference is that we use \cref{lemma:entry-merge-completeness} instead of
+    \cref{lemma:unif-entry-merge-completeness}.
+\end{proof}

### 2023-05-22 proofs/tex/alg-subtyping-lemmas.tex
Soundness and completeness of subtyping

@@
 
-    \item  \ruleref{\ottdruleAPVarLabel} then
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
+    \item \label{case:pos-subt-soundness:var} \ruleref{\ottdruleAPVarLabel}  
+    then $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
     Then $[[uv a⁺ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
@@
 
-    \item \label{case:pos-subt-soundness:arrow} 
+    \item \label{case:pos-subt-soundness:shift} 
      \ruleref{\ottdruleAShiftDLabel} then
@@
         $[[Γ,nbs; Θ, â⁻*[Γ, nbs] ⊢ [â⁻*/nas] uP' ]]$, so the induction hypothesis is applicable,
-        that is $[[us' : Θ | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
-        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.
+        that is $[[us' : Θ, â⁻*[Γ, nbs] | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
+        $[[us'2]]$ s.t. $[[Θ, â⁻*[Γ, nbs] ⊢ us'2 ⇒ us']]$.
 
-        Since the domain of $[[us']]$ is $[[uv [â⁺*/pas]uN']]$, the domain of 
-        $[[us]] = [[us' \  {α̂⁺*}]]$ is $[[uv [â⁺*/pas]uN \ {â⁺*} = uv uN']] = [[uv ∀pas.uN']]$,
-        this way, $[[us : Θ | uv ∀pas.uN']]$, as required.
+        Since the domain of $[[us']]$ is $[[uv [â⁻*/nas]uP']]$, the domain of 
+        $[[us]] = [[us' \  {α̂⁻*}]]$ is $[[uv [â⁻*/nas]uP' \ {â⁻*}]] = [[uv ∃nas.uP']]$,
+        this way, $[[us : Θ | uv ∃nas.uP']]$, as required.
 
-        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
+        It is left to show that $[[Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
         Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
-        extending $[[us2]]$ to the domain $[[Ord uv [â⁺*/pas]uN']]$ with the values of $[[us']]$,
-        i.e.  $[[us'2 : Θ | uv [â⁺*/pas]uN']]$, 
+        extending $[[us2]]$ to the domain $[[uv [â⁻*/nas]uP']]$ with the values of $[[us']]$,
+        i.e.  $[[us'2 : Θ | uv [â⁻*/nas]uP']]$, 
         and
@@
             \begin{cases}
-               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']] \\
-               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁺*]]
+               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']] \\
+               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁻*]]
             \end{cases}
@@
         corresponding \emph{unification entry}. 
-        Notice that $[[Sub us'2|uv uN']] = [[us2]]$.
+        Notice that $[[Sub us'2|uv uP']] = [[us2]]$.
     It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
     \begin{enumerate}
-        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
-        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
+        \item if $[[β̂±]] \in [[uv uP']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
+        \item it $[[β̂±]] \in [[â⁻*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
     \end{enumerate}
     which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
-    $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$.\\
+    $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$.\\
     Notice that
@@
     \begin{aligned}[t]
-                 [[ [us'2][â⁺*/pas]uN' ]] &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][Sub us'2|uv uN']uN' ]]
+                 [[ [us'2][â⁻*/nas]uP' ]] &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][Sub us'2|uv uP']uP' ]]
                                           && \text{by substitution properties \ilyam{todo}}\\
-                                          &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ]]
-                                          && \text{since $[[Sub us'2|uv uN']] = [[us2]]$}.
+                                          &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uP' ]]
+                                          && \text{since $[[Sub us'2|uv uP']] = [[us2]]$}.
     \end{aligned}
     $\\
-    Also notice that the domain of $[[Sub us'2|{â⁺*} ○ â⁺*/pas]]$ is $[[pas]]$,
-    and the range is $[[Γ, pbs]]$, i.e. $[[Γ, pbs ⊢ Sub us'2|{â⁺*} ○ â⁺*/pas : pas]]$, 
+    Also notice that the domain of $[[Sub us'2|{â⁻*} ○ â⁻*/nas]]$ is $[[nas]]$,
+    and the range is $[[Γ, nbs]]$, i.e. $[[Γ, nbs ⊢ Sub us'2|{â⁻*} ○ â⁻*/nas : nas]]$, 
     which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
-    $[[ Γ, pbs ⊢ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ≤ iM' ]]$
-    to get $[[ Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM' ]]$, as required.
-
+    $[[ Γ, nbs ⊢ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uN' ≤ iM' ]]$
+    to obtain $[[ Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ' ]]$, as required.
     \end{caseof}
@@
     \begin{caseof}
-        \item $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
-        Then $[[uv a⁻ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
-        Since $[[uv a⁻ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁻]]$, i.e.
-        $[[ [us']a⁻ = a⁻ ]]$. Therefore, $[[Γ ⊢ [us']a⁻ ≤ a⁻]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
-
-        \item $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
-        Then the next step of the algorithm is the unification of $[[nf(uP)]]$ and $[[nf(iQ)]]$.
-        By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
-        it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uP]]$.
-        By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
-        $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uP ]]$ as substitutions.
-
-        $[[ [us]nf(uP) = nf(iQ) ]]$ implies $[[Γ ⊢ [us]nf(uP) ≈ nf(iQ)]]$, and then 
-        $[[Γ ⊢ [us']nf(uP) ≈ nf(iQ)]]$. \ilyam{add lemmas}
-
-        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uP) ≈ nf(iQ)]]$ by 
-        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
-        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
-        $[[Γ ⊢ [us']nf(uP) ≈ [us']uP ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
-        $[[Γ ⊢ nf(iQ) ≈ iQ ]]$. 
-        This way, we have $[[Γ ⊢ [us']uP ≈ iQ]]$.
-        Then by \ruleref{\ottdruleDOneShiftULabel}
-        and congruence of substitution, $[[Γ ⊢ [us']↑uP ≤ ↑iQ]]$.
-
-        \item $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
-        The next step of the algorithm is two recursive calls:
+        \item \ruleref{\ottdruleAArrowLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape
+        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
+        On the next step, the the algorithm makes two recursive calls:
         $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
@@
 
-        \item $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
-        Then the algorithm creates fresh unification variables $[[â⁺*[Γ,pbs] ]]$, 
-        substitutes the old $[[pas]]$ with them in $[[uN']]$, and makes the recursive call:
-        $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ [â⁺*/pas] uN ≤ iM ⫤ us']]$, returning as the resulting solution 
-        $[[us]] = [[us' \ {α̂⁺*}]]$.
+        \item \ruleref{\ottdruleANVarLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
+        This case is symmetric to \cref{case:pos-subt-soundness:var} of \cref{lemma:pos-subt-soundness}.
 
-        Notice that $[[Γ, pbs ⊢ Θ, â⁺*[G, pbs] ]]$, $[[Γ,pbs ⊢ iM']]$, and 
-        $[[Γ,pbs; Θ, â⁺*[G, pbs] ⊢ [â⁺*/pas] uN' ]]$, so the induction hypothesis is applicable,
-        that is $[[us' : Θ | uv [â⁺*/pas]uN']]$ and $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$ for any
-        $[[us'2]]$ s.t. $[[Θ ⊢ us'2 ⇒ us']]$.
+        \item \ruleref{\ottdruleAShiftULabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
+        This case is symmetric to \cref{case:pos-subt-soundness:shift} of \cref{lemma:pos-subt-soundness}.
 
-        Since the domain of $[[us']]$ is $[[uv [â⁺*/pas]uN']]$, the domain of 
-        $[[us]] = [[us' \  {α̂⁺*}]]$ is $[[uv [â⁺*/pas]uN \ {â⁺*} = uv uN']] = [[uv ∀pas.uN']]$,
-        this way, $[[us : Θ | uv ∀pas.uN']]$, as required.
-
-        It is left to show that $[[Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
-        Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
-        extending $[[us2]]$ to the domain $[[Ord uv [â⁺*/pas]uN']]$ with the values of $[[us']]$,
-        i.e.  $[[us'2 : Θ | uv [â⁺*/pas]uN']]$, 
-        and
-        \[
-            [[us'2(β̂±)]]  = 
-            \begin{cases}
-               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']] \\
-               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁺*]]
-            \end{cases}
-        \] 
-        , where the application of the unification solution to a variable returns the 
-        corresponding \emph{unification entry}. 
-        Notice that $[[Sub us'2|uv uN']] = [[us2]]$.
-    It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
-    \begin{enumerate}
-        \item if $[[β̂±]] \in [[uv uN']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
-        \item it $[[β̂±]] \in [[â⁺*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
-    \end{enumerate}
-    which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
-    $[[ Γ, pbs ⊢ [us'2][â⁺*/pas]uN' ≤ iM' ]]$.\\
-    Notice that
-    $
-    \begin{aligned}[t]
-                 [[ [us'2][â⁺*/pas]uN' ]] &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][Sub us'2|uv uN']uN' ]]
-                                          && \text{by substitution properties \ilyam{todo}}\\
-                                          &= [[ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ]]
-                                          && \text{since $[[Sub us'2|uv uN']] = [[us2]]$}.
-    \end{aligned}
-    $\\
-    Also notice that the domain of $[[Sub us'2|{â⁺*} ○ â⁺*/pas]]$ is $[[pas]]$,
-    and the range is $[[Γ, pbs]]$, i.e. $[[Γ, pbs ⊢ Sub us'2|{â⁺*} ○ â⁺*/pas : pas]]$, 
-    which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
-    $[[ Γ, pbs ⊢ [Sub us'2|{â⁺*} ○ â⁺*/pas][us2]uN' ≤ iM' ]]$
-    to get $[[ Γ ⊢ [us2]∀pas.uN' ≤ ∀pbs.iM' ]]$, as required.
+        \item \ruleref{\ottdruleAForallLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
+        This case is symmetric to \cref{case:pos-subt-soundness:exists} of \cref{lemma:pos-subt-soundness}.
 
@@
     \begin{caseof}
-        \item \label{case:subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
+        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
         Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
@@
 
-        \item \label{case:subt-complete-nvar}
-        $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}, 
-        i.e. $[[ [us]uN ]] = [[iM]] = [[ α⁻ ]] = [[iN]]$ 
-        (the latter equality holds because $[[uN]] \neq [[α̂⁻]]$).
+        \item \label{case:pos-subt-complete-pvar}
+        $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
+        i.e. $[[iP]] = [[ [us]uP ]] = [[ α⁺ ]] = [[iQ]]$.
+        Here the first equality holds because $[[uP]]$ is not a unification variable:
+        this case has been covered by \cref{case:pos-subt-complete-base}.
+        The second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
 
-        Notice that $[[us : Θ | uv α⁻]]$ = $[[us : Θ | ∅]]$ = $[[·]]$.
+        Notice that $[[us : Θ | uv α⁺]]$ = $[[us : Θ | ∅]]$ = $[[·]]$.
 
-        The algorithm applies \ruleref{\ottdruleANVarLabel} and 
-        infers $[[us']] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$.
+        The algorithm applies \ruleref{\ottdruleAPVarLabel} and 
+        infers $[[us']] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
 
@@
 
-       \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-       \ruleref{\ottdruleDOneShiftDLabel} is symmetric to the corresponding negative
-        \cref{case:subt-complete-upshift}. Notice that the algorithm does not 
-        make a recursive call in this case but rather invoke the unification.
-        This way, the proof does not apply the induction hypothesis, and hence, 
-        the absence of negative unification variables in a negative type
-        ($[[α̂⁻]] \notin [[uN]]$ for $[[↓uN]] =[[uP]]$) is not required.
+        \item \label{case:pos-subt-complete-upshift} $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneShiftDLabel},
 
-      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-      \ruleref{\ottdruleDOneExistsLabel} is symmetric to the corresponding negative
-      \cref{case:subt-complete-forall}. Notice that we only consider the case
+        Then $[[ uP ]] = [[ ↓uN ]]$, since the substitution $[[ [us]uP ]]$ must preserve the 
+        top-level constructor of $[[uP]]\neq [[α̂⁺]]$ (the case $[[uP]] = [[α̂⁺]]$ has been covered
+        by \cref{case:pos-subt-complete-base}), and $[[uQ]] = [[ ↓iM ]]$,
+        and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
+
+        Since both types start with $[[↓]]$, 
+        the algorithm tries to apply \ruleref{\ottdruleAShiftDLabel}: 
+        $[[G;Θ ⊨ ↓uN ≥ ↓iM ⫤ us']]$. The premise of this rule is the
+        unification of $[[nf(uN)]]$ and $[[nf(iM)]]$:
+        $[[Γ;Θ ⊨ nf(uN) ≈u nf(iM) ⫤ us']]$. Let us show that the unification successfully 
+        terminates and returns $[[us']] = [[nf(us)]]$.
+
+        To demonstrate that the unification terminates, we apply the completeness 
+        of the unification algorithm (\cref{lemma:unification-completeness}). 
+        In order to do that, we need to provide a unifier of 
+        $[[nf(uN)]]$ and $[[nf(iM)]]$. Thankfully, $[[nf(us)]]$ does it.
+
+        \begin{itemize}
+            \item $[[nf(uN)]]$ and $[[nf(iM)]]$ are normalized 
+            \item $[[Γ ; Θ ⊢ nf(uN)]]$ because $[[Γ ; Θ ⊢ uN]]$ (\cref{todo})
+            \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
+            \item $[[nf(us) : Θ | uv nf(uN)]]$ because $[[us: Θ | uv uN]]$ (\cref{todo})
+            \item $ \begin{aligned}[t]
+                    [[ Γ ⊢ [us]uN ≈ iM ]] &\Rightarrow [[ [us]uN ≈ iM ]]
+                                          && \text {by \cref{lemma:equiv-completeness}}\\
+                                          &\Rightarrow [[ nf([us]uN) = nf(iM) ]]
+                                          && \text {by \cref{lemma:normalization-completeness}}\\
+                                          &\Rightarrow [[ [nf(us)]nf(uN) = nf(iM) ]]
+                                          && \text {by \cref{lemma:norm-subst-distr}}\\
+                    \end{aligned}
+                  $
+        \end{itemize}
+        Then by the completeness of the unification,
+        $[[Γ ; Θ ⊨ nf(uN) ≈u nf(iM) ⫤ nf(us)]]$.
+        This way, the subtyping algorithm terminates and the resulting solution is
+        $[[nf(us)]]$.
+
+        It is left to note that $[[Θ ⊢ us ⇒ nf(us)]]$, by \cref{todo}, since $[[Θ ⊢ us ≈ nf(us)]]$
+
+      \item $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
+      We should only consider the case
       when the substitution $[[ [us]uP ]]$ results in the existential type 
-      $[[∃nas.iP']]$ (for $[[iP']] \neq [[∃]]\dots$) by congruence, 
-      i.e. $[[uP = ∃nas.uP'']]$ (for $[[uP'']] \neq [[∃]]\dots$) and $[[ [us]uP'' = iP' ]]$.
+      $[[∃nas.iP'']]$ (for $[[iP'']] \neq [[∃]]\dots$) by congruence, 
+      i.e. $[[uP = ∃nas.uP']]$ (for $[[uP']] \neq [[∃]]\dots$) and $[[ [us]uP' = iP'' ]]$.
       This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
-      (\cref{case:subt-complete-base}), and thus, the substitution $[[us]]$ must
+      (\cref{case:pos-subt-complete-base}), and thus, the substitution $[[us]]$ must
       preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
+
+      This way, $[[uP]] = [[∃nas.uP']]$, $[[ [us]uP ]] = [[ ∃nas.[us]uP' ]]$ 
+      (assuming $[[nas]]$ does not intersect with the range of $[[us]]$)
+      and $[[iQ]] = [[ ∃nbs.iQ' ]]$, where either $[[nas]]$ or $[[nbs]]$ is not empty.
+
+      By inversion, $[[ Γ ⊢ [σ][us]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ : nas]]$.
+      Since $[[σ]]$ and $[[us]]$ have disjoint domains,
+      and the range of one does not intersect with the domain of the other,
+      they commute, i.e. $[[ Γ, nbs ⊢ [us][σ]uP' ≥ iQ' ]]$
+      (notice that the tree inferring this judgement is 
+      a proper subtree of the tree inferring 
+      $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
+
+      At the next step, 
+      the algorithm creates fresh (disjoint with $[[uv uP']]$) 
+      unification variables $[[â⁻*]]$, replaces $[[nas]]$ with them in $[[ uP' ]]$,
+      and makes the recursive call:
+      $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us1]]$,
+      (where $[[uP0]] = [[ [â⁻*/nas]uP' ]]$),
+      returning $[[us1 \ {â⁻*}]]$ as the result.
+
+       Notice that $[[ [us][σ][nas/â⁻*]uP0 = [us][σ]uP' ]]$,
+       and thus, $[[ Γ, nbs ⊢ [us][σ][nas/â⁻*]uP0 ≥ iQ' ]]$.
+       Let us combine $[[us]]$ and $[[σ  ○ nas/â⁻*]]$ into one 
+       $[[us0]]$---a unification solution for $[[uP0]]$:
+        \[
+            [[us0(β̂±)]]  = 
+            \begin{cases}
+               [[ β̂± :≈ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[â⁻*]] \\
+               [[us(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']]
+            \end{cases}
+       \]
+
+       Let us show that the induction hypothesis is applicable to 
+       $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$, with the meta-context $[[ Θ, â⁻*[Γ, nbs] ]]$
+       \begin{itemize}
+        \item $[[ us0 : (Θ, â⁻*[Γ, nbs])|uv uP0 ]]$. Notice that for every $[[αî⁻]] \in [[â⁻*]]$, 
+            the type corresponding to the entry $[[us0(αî⁻)]]$ 
+            is well-formed in $[[Γ, nbs]]$ since $[[Γ, nbs ⊢ σ : nas ]]$;
+            and for every $[[β̂±]] \in [[uv uP']]$, 
+            the type corresponding to the entry $[[us0(β̂±)]]$ is well-formed in
+            context $[[Θ(β̂±)]]$ since $[[us : Θ | uv uP']]$.
+        \item $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$. This is because
+            $[[ [us0]uP0 ]] = [[ [us][σ][nas/â⁻*]uP0 ]] = [[ [us][σ]uP' ]]$.
+        \item $[[ Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$ since
+            $[[Γ, nbs ⊢ Θ]]$ and $[[ {Γ, nbs} ⊆ {Γ, nbs} ]]$.
+        \item $[[Γ, nbs ; Θ, â⁻*[Γ, nbs] ⊢ uP0 ]]$
+       \end{itemize}
+
+       This way, we apply the induction hypothesis to $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$ and 
+       infer that there exists\\
+        $[[us0' :  (Θ, â⁻*[Γ, nbs]) | uv uP0]]$ such that
+       $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us0']]$,
+       and $[[Θ, â⁻*[G, nbs] ⊢ us0' ⇒ us0]]$.
+
+       This way, the algorithm terminates with the result $[[us0' \ {â⁻*}]]$
+       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁻*} ]]$. Notice that
+       $[[us0' \ {â⁻*} : Θ | uv uP' ]]$. To show $[[ Θ ⊢ us ⇒ us0' \
+       {â⁻*} ]]$, it suffices to consider a unification variable $[[β̂±]] \in 
+       [[uv uP']]$ and show that $[[Γ ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
+       reflexivity of weakening (\cref{lemma:entry-weakening-preorder}) since
+       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
 
@@
     \begin{caseof}
-        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel},
-        \label{case:subt-complete-upshift}
-        i.e. $[[ [us]uN ]] = [[ ↑[us]uP ]]$ (since the substitution $[[ [us]uN ]]$ must preserve the 
-        top-level constructor of $[[uN]]\neq [[α̂⁻]]$), and $[[uM]] = [[ ↑iQ ]]$,
-        and by inversion, $[[ Γ ⊢ [us]uP ≈ iQ ]]$.
-
-        Since both types start with $[[↑]]$, 
-        the algorithm tries to apply \ruleref{\ottdruleAShiftULabel}: 
-        $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us']]$. The premise of this rule is the
-        unification of $[[nf(uP)]]$ and $[[nf(iQ)]]$:
-        $[[Γ;Θ ⊨ nf(uP) ≈u nf(iQ) ⫤ us']]$. Let us show that the unification successfully 
-        terminates and returns $[[us']] = [[nf(us)]]$.
-
-        To demonstrate that the unification terminates, we apply the completeness 
-        of the unification algorithm (\cref{lemma:unification-completeness}). 
-        In order to do that, we need to provide a unifier of 
-        $[[nf(uP)]]$ and $[[nf(iQ)]]$. Thankfully, $[[nf(us)]]$ does it.
-
-        \begin{itemize}
-            \item $[[nf(uP)]]$ and $[[nf(iQ)]]$ are normalized 
-            \item $[[Γ ; Θ ⊢ nf(uP)]]$ because $[[Γ ; Θ ⊢ uP]]$ (\cref{todo})
-            \item $[[Γ ⊢ nf(iQ)]]$ because $[[Γ ⊢ iQ]]$ (\cref{corollary:wf-nf})
-            \item $[[nf(us) : Θ | uv nf(uP)]]$ because $[[us: Θ | uv uP]]$ (\cref{todo})
-            \item $ \begin{aligned}[t]
-                    [[ Γ ⊢ [us]uP ≈ iQ ]] &\Rightarrow [[ [us]uP ≈ iQ ]]
-                                          && \text {by \cref{lemma:equiv-completeness}}\\
-                                          &\Rightarrow [[ nf([us]uP) = nf(iQ) ]]
-                                          && \text {by \cref{lemma:normalization-completeness}}\\
-                                          &\Rightarrow [[ [nf(us)]nf(uP) = nf(iQ) ]]
-                                          && \text {by \cref{lemma:norm-subst-distr}}\\
-                    \end{aligned}
-                  $
-        \end{itemize}
-        Then by the completeness of the unification,
-        $[[Γ ; Θ ⊨ nf(uP) ≈u nf(iQ) ⫤ nf(us)]]$.
-        This way, the subtyping algorithm terminates and the resulting solution is
-        $[[nf(us)]]$.
-
-        It is left to note that $[[Θ ⊢ us ⇒ nf(us)]]$, by \cref{todo}, since $[[Θ ⊢ us ≈ nf(us)]]$
+        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
+        Then $[[ uN ]] = [[ ↑uP ]]$, since
+        the substitution $[[ [us]uN ]]$ must preserve the 
+        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
+        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
+        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
+         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
+        and the difference in the induction statement for the positive and 
+        the negative case here does not matter.
 
@@
        \item \label{case:subt-complete-forall}
-       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel},
-       i.e. $[[ [us]uN ]] = [[ ∀pas.[us]uN' ]]$
-       (assuming $[[pas]]$ does not intersect with the 
-       range of $[[us]]$)
-       and 
-       $[[iM]] = [[∀pbs.iM']]$, where either $[[pas]]$ or $[[pbs]]$
-       is non-empty.
+       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel}.
+       Since $[[uN]]$ does not contain negative unification variables,
+       $[[uN]]$ must be of the form $[[∀pas.uN']]$,
+       such that $[[ [us]uN = ∀pas.[us]uN' ]]$ and $[[ [us]uN']] \neq [[∀]]\dots$
+       (assuming $[[pas]]$ does not intersect with the range of $[[us]]$).
+       Also, $[[iM]] = [[∀pbs.iM']]$ and either $[[pas]]$ or $[[pbs]]$ is non-empty.
 
-       By inversion, 
-       $[[ Γ, pbs ⊢ [σ][us]uN' ≤ iM' ]]$ for 
-       $[[Γ, pbs ⊢ σ : pas ]]$.
-       Since $[[σ]]$ and $[[us]]$ have disjoint domains,
-       and the range of one does not intersect with the domain of the other,
-       they commute, i.e. $[[ Γ, pbs ⊢ [us][σ]uN' ≤ iM' ]]$
-       (notice that the tree inferring this judgement is 
-       a proper subtree of the tree inferring 
-       $[[ Γ ⊢ [us]uN ≤ iM ]]$).
+       The rest of the proof is symmetric to \cref{case:pos-subt-complete-exists} of
+       \cref{lemma:pos-subt-completeness}.
+       To apply the induction hypothesis, we need to show additionally that
+       there are no negative unification variables in $[[uN0]] = [[ [â⁺*/pas]uN' ]]$.
+       This is because $[[ uv uN0 ⊆ uv uN ∪ {â⁺*} ]]$, and $[[uN]]$ is free of negative
+       unification variables by assumption.
 
-       On the next step, 
-       the algorithm creates fresh (disjoint with $[[uv uN']]$) 
-       unification variables $[[â⁺*]]$, replaces $[[pas]]$ with them in $[[ [us]uN' ]]$,
-       and makes the recursive call:
-       $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ uN0 ≤ iM ⫤ us1]]$,
-       (where $[[uN0]] = [[ [â⁺*/pas]uN' ]]$),
-       returning $[[us1 \ {â⁺*}]]$ as the result.
+       \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}.\\
+       Then $[[iN]] = [[ [us]uN ]] = [[ α⁻ ]] = [[iM]]$. 
+       Here the first equality holds because $[[uN]]$ is not a unification variable:
+       by assumption, $[[uN]]$ is free of negative unification variables.
+       The second and the third equations hold because \ruleref{\ottdruleDOneNVarLabel}
+       was applied.
 
-       Notice that $[[ [us][σ][pas/â⁺*]uN0 = [us][σ]uN' ]]$,
-       and thus, $[[ Γ, pbs ⊢ [us][σ][pas/â⁺*]uN0 ≤ iM' ]]$.
-       Let us combine $[[us]]$ and $[[σ  ○ pas/â⁺*]]$ into one 
-       $[[us0]]$ an unification solution for $[[uN0]]$:
-        \[
-            [[us0(β̂±)]]  = 
-            \begin{cases}
-               [[ β̂± :≈ [σ]αi⁺ ]] & \text{if } [[β̂±]] = [[αî⁺]] \in [[â⁺*]] \\
-               [[us(β̂±)]] & \text{if } [[β̂±]] \in [[uv uN']]
-            \end{cases}
-       \]
+       The rest of the proof is symmetric to \cref{case:pos-subt-complete-pvar} of
+       \cref{lemma:pos-subt-completeness}.
 
-       Let us show that the induction hypothesis is applicable to 
-       $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$.
-       \begin{itemize}
-        \item $[[ us0 : Θ|uv uN', â⁺*[Γ, pbs] ]]$. Notice that for every $[[αî⁺]]$, 
-            the type corresponding to the entry $[[us0(αî⁺)]]$ 
-            is well-formed in $[[Γ, pbs]]$ since $[[Γ, pbs ⊢ σ : pas ]]$;
-            and for every $[[β̂±]] \in [[uv uN']]$, 
-            the type corresponding to the entry $[[us0(β̂±)]]$ is well-formed in
-            context $[[Θ(β̂±)]]$ since $[[us : Θ | uv uN']]$.
-        \item $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$. This is because
-            $[[ [us0]uN0 ]] = [[ [us][σ][pas/â⁺*]uN0 ]] = [[ [us][σ]uN' ]]$.
-        \item $[[ Γ, pbs ⊢ Θ, â⁺*[Γ, pbs] ]]$ since
-            $[[Γ, pbs ⊢ Θ]]$ and $[[ {Γ, pbs} ⊆ {Γ, pbs} ]]$.
-        \item $[[Γ, pbs ; Θ, â⁺*[Γ, pbs] ⊢ uN0 ]]$
-        \item There are no negative unification variables in $[[uN0]]$ because
-            $[[ uv uN0 ⊆ uv uN ∪ {â⁺*} ]]$.
-       \end{itemize}
-
-       This way, we apply the induction hypothesis to $[[Γ, pbs ⊢ [us0]uN0 ≤ iM' ]]$ and 
-       infer that there exists\\
-        $[[us0' :  (Θ, â⁺*[Γ, pbs]) | uv uN0]]$ such that
-       $[[G, pbs; Θ, â⁺*[G, pbs] ⊨ uN0 ≤ iM ⫤ us0']]$,
-       and $[[Θ, â⁺*[G, pbs] ⊢ us0' ⇒ us0]]$.
-
-       This way, the algorithm terminates with the result $[[us0' \ {â⁺*}]]$
-       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁺*} ]]$. Notice that
-       $[[us0' \ {â⁺*} : Θ | uv uN' ]]$. This way, to show $[[ Θ ⊢ us ⇒ us0' \
-       {â⁺*} ]]$. it suffices to consider an unification variable $[[β̂±]] ∊
-       [[uv uN']]$ and show that $[[Γ, pbs ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
-       reflexivity of weakening (\cref{lemma:weakening-reflexivity}) since
-       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
     \end{caseof}

### 2023-05-23 proofs/tex/unification-lemmas.tex
Add soundness and completeness of unification Unification solution merge needs to be updated

@@
 \end{lemma}
+\begin{proof}
+    We prove by induction on the derivation of 
+    $[[ Γ ; Θ ⊨ uN ≈u iM ⫤ us ]]$ and mutually $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$.
+    Let us consider the last rule forming this derivation. 
+    \begin{caseof}
+        \item \ruleref{\ottdruleUNVarLabel}, then $[[uN]] = [[α⁻]] = [[iM]]$.
+        The resulting unification solution is empty: $[[us]] = [[·]]$.
+        Since $[[uv uN]] = [[∅]]$, it satisfies $[[us : Θ | uv uP]]$.
+        Vacuously, it is equivalence-only, and $[[ [us]α⁻ = α⁻ ]]$, that is $[[ [us]uN = iM ]]$.
+
+        \item \ruleref{\ottdruleUShiftULabel}, then $[[uN]] = [[↑uP]]$ and $[[iM]] = [[↑iQ]]$.
+        The algorithm makes a recursive call to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$ returning $[[us']]$.
+        By induction hypothesis, $[[us' : Θ | uv uP]]$,
+        and since $[[uv uN]] = [[uv ↑uP]] = [[uv uP]]$,  $[[us' : Θ | uv uN]]$.
+        Finally, $[[ [us]uN ]] = [[ [us]↑uP ]] = [[ ↑[us]uP ]] = [[ ↑iQ ]] = [[ iM ]]$.
+
+        \item \ruleref{\ottdruleUArrowLabel}, then $[[uN]] = [[uP → uN']]$ and $[[iM]] = [[iQ → iM']]$.
+        The algorithm makes two recursive calls to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us1]]$ and
+        $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ us2]]$ returning $[[us1 & us2]]$ as the result.
+
+        It is clear that $[[uP]]$, $[[uN']]$, $[[iQ]]$, and $[[iM']]$ are normalized,
+        and that $[[Γ ; Θ ⊢ uP]]$, $[[Γ ; Θ ⊢ uN']]$, $[[Γ ⊢ iQ]]$, and $[[Γ ⊢ iM']]$.
+        This way, the induction hypothesis is applicable to both recursive calls.
+        By applying the induction hypothesis to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us1]]$,
+        we have:
+        \begin{itemize}
+            \item $[[us1 : Θ | uv uP]]$,
+            \item $[[us1]]$ is equivalence-only, and
+            \item $[[ [us1]uP = iQ ]]$.
+        \end{itemize}
+        By applying it to $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ us2]]$, we have:
+        \begin{itemize}
+            \item $[[us2 : Θ | uv uN']]$,
+            \item $[[us2]]$ is equivalence-only, and
+            \item $[[ [us2]uN' = iM' ]]$.
+        \end{itemize}
+
+        By the soundness of the unification solution merge (\cref{lemma:unif-sol-merge-soundness}), we have:
+        \begin{itemize}
+            \item $[[us1 & us2 : Θ | uv uP ∪ uv uN']]$, and hence, $[[us1 & us2 : Θ | uv uN]]$; 
+            \item $[[us1 & us2]]$ is equivalence-only; 
+            \item $[[us1 & us2]]$ is normalized;
+            \item $[[us1 & us2 | uv uP]] = [[us1]]$ and hence, $[[ [us1 & us2]uP ]] = [[ [us1]uP ]] = [[iQ]]$;
+            \item $[[us1 & us2 | uv uN']] = [[us2]]$ and hence, $[[ [us1 & us2]uN' ]] = [[ [us2]uN' ]] = [[iM']]$.
+        \end{itemize}
+        This way, $[[ [us1 & us2]uN ]] = [[ [us1 & us2]uP → [us1 & us2]uN' ]] = [[ iQ → iM' ]] = [[ iM ]]$.
+
+        \item \ruleref{\ottdruleUForallLabel}, then $[[uN]] = [[∀pas.uN']]$ and $[[iM]] = [[∀pas.iM']]$.
+        The algorithm makes a recursive call to $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ us]]$
+        returning $[[us]]$ as the result.
+
+        The induction hypothesis is applicable: $[[Γ,pas ; Θ ⊢ uN']]$ and $[[Γ,pas ⊢ iM']]$ hold
+        by inversion, and $[[uN']]$ and $[[iM']]$ are normalized, since $[[uN]]$ and $[[iM]]$ are.
+
+        By the induction hypothesis, $[[us : Θ | uv uN']]$, then, since
+        $[[uv uN]] = [[uv ∀pas.uN']] = [[uv uN']]$, we have $[[us : Θ | uv uN]]$.
+        Finally, $[[ [us]uN ]] = [[ [us]∀pas.uN' ]] = [[ ∀pas.[us]uN' ]] = [[ ∀pas.iM' ]] = [[ iM ]]$.
+
+        \item \ruleref{\ottdruleUNUVarLabel}, then $[[uN]] = [[α̂⁻]]$, $[[â⁻[Δ] ∊ Θ]]$, and $[[Δ ⊢ iM]]$.
+        As the result, the algorithm returns $[[us]] = [[ (â⁻ :≈ iM) ]]$.
+
+        It is clear that $[[(â⁻ :≈ iM) : (α̂⁻[Δ])]]$, since $[[Δ ⊢ iM]]$.
+        Notice that since $[[Θ|uv uN]] = [[Θ|{α̂⁻}]] = [[(α̂⁻[Δ])]]$, we can rewrite 
+        $[[(â⁻ :≈ iM) : (α̂⁻[Δ])]]$ as $[[us : Θ | uv uN]]$, as required.
+
+        It is also clear that $[[us]] = [[ (â⁻ :≈ iM) ]]$ is equivalence-only, 
+        and that $[[ [us]uN ]] = [[ [â⁻ :≈ iM]α̂⁻ ]] = [[ iM ]]$, as required.
+
+        \item The positive cases are proved symmetrically.
+
+    \end{caseof}
+
+\end{proof}
 
@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        assume there exists 
+        assume there exists an equivalence-only unification solution
         $[[us : Θ | uv uP]]$ such that $[[ [us]uP = iQ ]]$.
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        assume there exists $[[us : Θ | uv uN]]$        
+        assume there exists an equivalence-only unification solution
+        $[[us : Θ | uv uN]]$        
         such that $[[ [us]uN = iM ]]$.
@@
 \end{lemma}
+\begin{proof}
+    We prove it by induction on the structure of $[[uP]]$ and mutually, $[[uN]]$.
+    \begin{caseof}
+        \item $[[uN]] = [[α̂⁻]]$\\
+            Since $[[uv uN]]$ = $[[α̂⁻]]$, 
+            $[[us : Θ | {α̂⁻}]]$ means that $[[ α̂⁻[Δ] ]] \in [[Θ]]$, 
+            and since $[[us]]$ is equivalence only, $[[us]] = [[ (α̂⁻ :≈ iM') ]]$
+            for some $[[Δ ⊢ iM']]$.
+            $[[ [us]uN = iM ]]$ means $[[iM' = iM]]$.
+            This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 
+            $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
+        \item $[[uN]] = [[α⁻]]$\\
+            Then $[[us : Θ | {α⁻}]]$ means that $[[ us ]] = [[·]]$, 
+            and $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
+            This way, \ruleref{\ottdruleUNVarLabel} infers 
+            $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+
+        \item $[[uN]] = [[↑uP]]$\\
+            Notice that $[[ [us]uN = iM ]]$ means $[[ ↑[us]uP = iM ]]$, i.e.
+            $[[iM]] = [[↑iQ]]$ for some $[[iQ]]$ and $[[ [us]uP = iQ ]]$.
+
+            Let us show that the induction hypothesis is applicable to $[[ [us]uP = iQ ]]$.
+            Notice that $[[uP]]$ is normalized, since $[[uN]] = [[↑uP]]$ is normalized,
+            $[[Γ ; Θ ⊢ uP]]$ follows from inversion of $[[Γ ; Θ ⊢ ↑uP]]$, 
+            and $[[Γ ⊢ iQ]]$ follows from inversion of $[[Γ ⊢ ↑iQ]]$,
+            $[[us]]$ is equivalence-only by assumption, and
+            $[[us : Θ | uv uP]]$, since $[[uv uP]] = [[uv ↑uP]] = [[uv uN]]$.
+
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$.
+            Hence, \ruleref{\ottdruleUShiftULabel} is applicable to infer
+            $[[Γ ; Θ ⊨ ↑uP ≈u ↑iQ ⫤ us]]$, that is $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+
+        \item $[[uN]] = [[uP → uN']]$\\
+            Notice that $[[ [us]uN = iM ]]$ means $[[ [us]uP → [us]uN' = iM ]]$, i.e.
+            $[[iM]] = [[iQ → iM']]$ for some $[[iQ]]$ and $[[iM']]$, 
+            such that $[[ [us]uP = iQ ]]$ and $[[ [us]uN' = iM' ]]$.
+            Then $[[ [us | uv uP]uP = iQ ]]$ and $[[ [us | uv uN']uN' = iM' ]]$.
+
+            Let us show that the induction hypothesis is applicable to $[[ [us | uv uP]uP = iQ ]]$ 
+            and $[[ [us | uv uN']uN' = iM' ]]$.
+            Notice that $[[uP]]$ and $[[uN']]$ are normalized, since $[[uN]] = [[uP → uN']]$ is normalized,
+            $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$ follow from inversion of $[[Γ ; Θ ⊢ uP → uN']]$,
+            $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$ follow from inversion of $[[Γ ⊢ iQ → iM']]$,
+            $[[us|uv uP]]$ and $[[us|uv uN']]$ are equivalence-only since $[[us]]$ is equivalence-only,
+            $[[us|uv uP : Θ | uv uP]]$ and $[[us|uv uN' : Θ | uv uN']]$, 
+            since $[[us : Θ | uv uP ∪ uv uN']]$ (\cref{todo}).
+
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us|uv uP]]$ and
+            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ us|uv uN']]$.
+            To apply \ruleref{\ottdruleUArrowLabel} and infer the required
+            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$, we need to show that
+            $[[us|uv uP & us|uv uN']] = [[us]]$. 
+            By the completeness of the unification solution merge 
+            (\cref{lemma:unif-sol-merge-completeness}), 
+            $[[us|uv uP & us|uv uN']] = [[us|uv uP ∪ uv uN']]$,
+            and $[[us|uv uP ∪ uv uN']] = [[us|uv uN]] = [[us]]$, since $[[us : Θ | uv uN]]$.
+
+        \item $[[uN]] = [[∀pas.uN']]$\\
+            Notice that $[[ [us]uN = iM ]]$ means $[[ ∀pas.[us]uN' = iM ]]$, i.e.
+            $[[iM]] = [[∀pas.iM']]$ for some $[[iM']]$ such that $[[ [us]uN' = iM' ]]$.
+
+            Let us show that the induction hypothesis is applicable to $[[ [us]uN' = iM' ]]$.
+            Notice that $[[uN']]$ is normalized, since $[[uN]] = [[∀pas.uN']]$ is normalized,
+            $[[Γ,pas ; Θ ⊢ uN']]$ follows from inversion of $[[Γ ; Θ ⊢ ∀pas.uN']]$,
+            $[[Γ,pas ⊢ iM']]$ follows from inversion of $[[Γ ⊢ ∀pas.iM']]$,
+            $[[us]]$ is equivalence-only by assumption, and
+            $[[us : Θ | uv uN']]$ since $[[uv uN']] = [[uv ∀pas.uN']] = [[uv uN]]$.
+
+            By the induction hypothesis, $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ us]]$.
+            Hence, \ruleref{\ottdruleUForallLabel} is applicable to infer
+            $[[Γ ; Θ ⊨ ∀pas.uN' ≈u ∀pas.iM' ⫤ us]]$, that is $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+
+        \item The positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}

### 2023-06-05 proofs/tex/anti-unification-lemmas.tex
WIP: finish merge lemmas

@@
     \begin{itemize}
-        \item [$+$] 
-        \item [$-$] 
+        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then 
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uQ]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+            \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
+        \end{enumerate}
+
+        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
+        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+        then
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uM]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
+        \end{enumerate}
     \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove it by induction on 
+    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+    and mutually, $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$.
+    Let us consider the last rule applied to infer this judgement.
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUNVarLabel}, then $[[iN1]] = [[α⁻]] = [[iN2]]$,
+              $[[Ξ]] = [[·]]$, $[[uM]] = [[α⁻]]$, and $[[aus1]] = [[aus2]] = [[·]]$.
+            \begin{enumerate}
+                \item $[[Γ ; · ⊢ α⁻]]$ follows from the assumption $[[Γ ⊢ α⁻]]$,
+                \item $[[Γ ; · ⊢ · : ·]]$ holds trivially, and
+                \item $[[ [·] α⁻ = α⁻ ]]$ holds trivially.
+            \end{enumerate}
+        \item \label{case:anti-unification-soundness:shiftu}
+         \ruleref{\ottdruleAUShiftULabel}, then $[[iN1]] = [[↑iP1]]$,
+                $[[iN2]] = [[↑iP2]]$, and the algorithm makes the recursive call:
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
+                returning $[[(Ξ, ↑uQ, aus1, aus2)]]$ as the result.
+
+                Since $[[iN1]] = [[↑iP1]]$ and $[[iN2]] = [[↑iP2]]$ are normalized, 
+                so are $[[iP1]]$ and $[[iP2]]$, and thus, the induction hypothesis 
+                is applicable to $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$:
+                \begin{enumerate}
+                    \item $[[Γ ; Ξ ⊢ uQ]]$, and hence, $[[Γ ; Ξ ⊢ ↑uQ]]$,
+                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+                    \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$, and then by 
+                    the definition of the substitution, $[[ [ausi] ↑uQ = ↑iPi ]]$ for $i \in \{1,2\}$.
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUArrowLabel}, then $[[iN1]] = [[iP1 → iN'1]]$,
+                $[[iN2]] = [[iP2 → iN'2]]$, and the algorithm makes two recursive calls:
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ', uM, aus'1, aus'2)]]$ and
+                and returns $[[(Ξ ∪ Ξ', uQ → uM, aus1 ∪ aus'1, aus2 ∪ aus'2)]]$ as the result.
+
+                Notice, that the induction hypothesis is applicable to 
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$:
+                $[[iP1]]$ and $[[iP2]]$ are normalized, since $[[iN1]] = [[iP1 → iN'1]]$
+                and $[[iN2]] = [[iP2 → iN'2]]$ are normalized.
+                Similarly, the induction hypothesis is applicable to
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ', uM, aus'1, aus'2)]]$.
+
+                This way, by the induction hypothesis:
+                \begin{enumerate}
+                    \item $[[Γ ; Ξ ⊢ uQ]]$ and $[[Γ ; Ξ' ⊢ uM]]$. 
+                    Then by weakening (\cref{todo}), $[[Γ ; Ξ ∪ Ξ' ⊢ uQ]]$ and 
+                    $[[Γ ; Ξ ∪ Ξ' ⊢ uM]]$, which implies $[[Γ ; Ξ ∪ Ξ' ⊢ uQ → uM]]$;
+
+                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ and $[[Γ ; · ⊢ aus'i : Ξ']]$, 
+                           and hence, $[[Γ ; · ⊢ ausi ∪ aus'i : Ξ ∪ Ξ']]$ (\cref{todo});
+
+                    \item $[[ [ausi] uQ = iPi ]]$ and $[[ [aus'i] uM = iN'i ]]$.
+                    Since $[[ausi ∪ aus'i]]$ restricted to $[[Ξ]]$ is $[[ausi]]$,
+                    we have $[[ [ausi ∪ aus'i] uQ = iPi ]]$ and 
+                    $[[ [ausi ∪ aus'i] uM = iN'i ]]$, and thus, 
+                    $[[ [ausi ∪ aus'i] uQ → uM = iP1 → iN'1 ]]$
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUForallLabel}, then $[[iN1]] = [[∀pas.iN'1]]$,
+                $[[iN2]] = [[∀pas.iN'2]]$, and the algorithm makes a recursive call:
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
+                returns $[[(Ξ, ∀pas.uM, aus1, aus2)]]$ as the result.
+
+                Similarly to \cref{case:anti-unification-soundness:shiftu}, 
+                we apply the induction hypothesis to
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ, uM, aus1, aus2)]]$ to obtain:
+                \begin{enumerate}
+                    \item $[[Γ; Ξ ⊢ uM]]$, and hence, $[[Γ ; Ξ ⊢ ∀pas.uM]]$;
+                    \item $[[Γ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+                    \item $[[ [ausi] uM = iN'i ]]$ for $i \in \{1,2\}$,
+                        and then by the definition of the substitution,
+                        $[[ [ausi] ∀pas.uM = ∀pas.iN'i ]]$ for $i \in \{1,2\}$. 
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUAULabel}, which applies 
+        when other rules do not, and $[[G ⊢ iNi]]$,
+        returning as the result $[[(Ξ, uM, aus1, aus2)]] = $
+        $[[(â⁻_{iN1, iN2}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} :≈ iN1) ,  (â⁻_{iN1, iN2} :≈ iN2))]]$.
+
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uM]]$ is rewritten as $[[Γ ; â⁻_{iN1, iN2} ⊢ â⁻_{iN1, iN2}]]$,
+                which holds trivially;
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} :≈ iNi) : â⁻_{iN1, iN2}]]$,
+                which holds since $[[Γ ⊢ iNi]]$ by the premise of the rule;
+            \item $[[ [ausi] uM = iNi ]]$ is rewritten as $[[ [â⁻_{iN1, iN2} :≈ iNi] â⁻_{iN1, iN2} = iNi ]]$,
+                which holds trivially by the definition of substitution.
+        \end{enumerate}
+
+        \item Positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}
+
+\begin{observation} \label{obs:names-defined-by-mapping}
+    Names of the anti-unification variables are uniquely defined by
+    the types they are mapped to by the resulting substitutions.
+
+    \begin{itemize}
+        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+        \item [$-$]  Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
+        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+    \end{itemize}
+\end{observation}
+\begin{proof}
+    By simple induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+    Let us consider tha last rule applied to infer this judgment.
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUPVarLabel} or \ruleref{\ottdruleAUNVarLabel},
+        then $[[Ξ]] = [[·]]$, and the property holds vacuously.
+
+        \item \ruleref{\ottdruleAUAULabel}
+        Then  $[[Ξ]] = [[â⁻_{iN1, iN2}]]$,
+        $[[aus1]] = [[â⁻_{iN1, iN2} :≈ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} :≈ iN2]]$.
+        So the property holds trivially.
+
+        \item \ruleref{ottdruleAUArrowLabel}
+        In this case, $[[Ξ]] = [[Ξ' ∪ Ξ'']]$, $[[aus1]] = [[aus'1 & aus''1]]$, and 
+        $[[aus2]] = [[aus'2 & aus''2]]$,
+        where the property holds for ($[[Ξ']]$, $[[aus'1]]$, $[[aus'2]]$) and 
+        ($[[Ξ'']]$, $[[aus''1]]$, $[[aus''2]]$) by the induction hypothesis.
+        Then since the merge of solutions does not change the types the variables are mapped to,
+        the required property holds for $[[Ξ]]$, $[[aus1]]$, and $[[aus2]]$.
+
+        \item For the other rules, the resulting $[[Ξ]]$ is taken from the recursive call
+        and the required property holds immediately by the induction hypothesis.
+    \end{caseof}
+\end{proof}
+
+\begin{observation}[Anti-unification algorithm is deterministic]
+    \label{obs:anti-unification-deterministic}
+    \hfill\\
+\begin{itemize}
+    \item [$+$] If $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and 
+    $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ', uQ', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uQ]] = [[uQ']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+    \item [$-$] If $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
+    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ', uM', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uM]] = [[uM']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+\end{itemize}
+\end{observation}
+\begin{proof}
+    By trivial induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+\end{proof}
+
+% \renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
+% \renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
+
+% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
+% \renewcommand{\ottdruleAUNVarName}[0]{(Var$^{-[[≈au]]}$)}
+% \renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUAUName}[0]{(AU$^{-}$)}
 
@@
         \item [$+$] 
+            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
+            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uQ']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
+            \end{enumerate}
+
+            Then the anti-unification algorithm terminates, that is there exists
+            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+
         \item [$-$] 
+            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
+            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uM']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
+            \end{enumerate}
+
+            Then the anti-unification algorithm succeeds, that is 
+            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
     \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove it by the induction on $[[uM']]$ and mutually on $[[uQ']]$.
+    \begin{caseof}
+        \item $[[uM']] = [[â⁻]]$ 
+            Then since $[[Γ ; · ⊢ aus'i : Ξ']]$,
+            $[[Γ ⊢ [aus'i] uM']] = [[ iNi ]]$. 
+            This way, \ruleref{\ottdruleAUAULabel} is always applicable
+            if other rules are not.
+
+        \item $[[uM']] = [[α⁻]]$
+            Then $[[α⁻]] = [[ [aus'i] α⁻]] = [[ iNi ]]$, which means
+            that \ruleref{\ottdruleAUNVarLabel} is applicable.
+
+        \item $[[uM']] = [[↑uQ']]$
+            Then $[[ ↑[aus'i]uQ']] = [[ [aus'i]↑uQ']] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[↑iP1]]$ and $[[↑iP2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uQ' = iPi ]]$, which means that 
+            $[[(Ξ', uQ', aus'1, aus'2)]]$ is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+            Then by the induction hypothesis, there exists $[[(Ξ, uQ, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, and hence, 
+            $[[Γ ⊨ ↑iP1 ≈au ↑iP2 ⫤ (Ξ, ↑uQ, aus1, aus2)]]$ by \ruleref{\ottdruleAUShiftULabel}.
+        \item $[[uM']] = [[∀pas.uM'']]$ This case is similar to the previous one:
+            we consider $\forall[[pas]]$ as a constructor. 
+            Notice that $[[∀pas.[aus'i]uM'']] = [[ [aus'i]∀pas.uM'']] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[∀pas.iN''1]]$ and $[[∀pas.iN''2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uM'' = iN''i ]]$, which means that
+            $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+            Then by the induction hypothesis, there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ, uM, aus1, aus2)]]$, and hence,
+            $[[Γ ⊨ ∀pas.iN''1 ≈au ∀pas.iN''2 ⫤ (Ξ, ∀pas.uM, aus1, aus2)]]$ by 
+            \ruleref{\ottdruleAUForallLabel}.
+        \item $[[uM']] = [[uQ' → uM'']]$
+            Then $[[ [aus'i]uQ' → [aus'i]uM'']] = [[ [aus'i](uQ' → uM'')]] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[uP1 → uN'1]]$ and $[[uP2 → uN'2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uQ' = iPi ]]$ and $[[ [aus'i]uM'' = iN''i ]]$, which means that
+            $[[(Ξ', uQ', aus'1, aus'2)]]$ is an anti-unifier of $[[iP1]]$ and $[[iP2]]$,
+            and $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+            Then by the induction hypothesis, 
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ1, uQ, aus1, aus2)]]$ and 
+            $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ2, uM, aus3, aus4)]]$ succeed.
+            The result of the algorithm is $[[(Ξ1 ∪ Ξ2, uQ → uM, aus1 & aus3, aus2 & aus4)]]$,
+            and it is left to show that $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined.
+
+            From \cref{obs:names-defined-by-mapping}, we know that 
+            if $[[β̂⁻ ∊ dom(aus1) ∩ dom(aus3)]]$ then $[[ [aus1]β̂⁻ = [aus3]β̂⁻ ]]$,
+            and if $[[β̂⁻ ∊ dom(aus2) ∩ dom(aus4)]]$ then $[[ [aus2]β̂⁻ = [aus4]β̂⁻ ]]$.
+            This way, $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined.
+
+        \item $[[uQ']] = [[â⁺]]$ 
+            This case if not possible, since $[[Γ ; Ξ' ⊢ uQ']]$ means 
+            $[[â⁺]] \in [[Ξ']]$, but $[[Ξ']]$ can only contain negative variables.
+
+        \item Other positive cases are proved symmetrically to the corresponding negative ones.
+    \end{caseof}
+\end{proof}
+
+\begin{lemma}[Initiality of Anti-Unification] \label{lemma:anti-unification-initial}
+    \hfill
+    \begin{itemize}
+        \item [$+$] 
+            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
+            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
+            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
+            if 
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uQ']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
+            \end{enumerate}
+            then there exists $[[ausr]]$ such that
+            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
+            Moreover, $[[ [ausr]β̂⁻]]$ can
+            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
+            $[[Γ]]$.
+        \item [$-$] 
+            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
+            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
+            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
+            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
+            if
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uM']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
+            \end{enumerate}
+            then there exists $[[ausr]]$ such that
+            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
+            Moreover, $[[ [ausr]β̂⁻]]$ can
+            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
+            $[[Γ]]$.
+    \end{itemize}
+\end{lemma}
+\begin{proof}
+    First, let us assume that $[[uM']]$ is a metavariable $[[α̂⁻]]$. 
+    Then we can take $[[ausr]] = [[α̂⁻]] \mapsto [[uM]]$, which satisfies the required properties:
+    \begin{itemize}
+        \item $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ holds since 
+        $[[Ξ' | uv uM']] = [[{α̂⁻}]]$ and $[[Γ ; Ξ ⊢ uM]]$ by the soundness of anti-unification (\cref{lemma:anti-unification-soundness});
+        \item $[[ [ausr] uM' = uM ]]$ holds by construction
+        \item $[[ [ausr]α̂⁻]] = [[uM]]$ is the anti-unifier of 
+            $[[iN1]] = [[ [aus'1] α̂⁻]]$ and $[[iN2]] = [[ [aus'2] α̂⁻]]$
+            in context $[[Γ]]$, and hence, it is uniquely determined by them (\cref{obs:anti-unification-deterministic}).
+    \end{itemize}
+
+    Now, we can assume that $[[uM']]$ is not a metavariable. 
+    We prove by induction on the derivation of $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on the derivation of $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+
+    Since $[[uM']]$ is not a metavariable, 
+    the substitution acting on $[[uM']]$ preserves its outer constructor. 
+    In other words, 
+    $[[ [aus'i] uM' = iNi ]]$ means that $[[uM']]$, 
+    $[[iN1]]$ and $[[iN2]]$ have the same outer constructor. 
+    Let us consider the algorithmic anti-unification rule corresponding to this constructor, 
+    and show that it was successfully applied to anti-unify $[[iN1]]$ and $[[iN2]]$ 
+    (or $[[iP1]]$ and $[[iP2]]$).
+
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUNVarLabel}, i.e. $[[iN1]] = [[α⁻]] = [[iN2]]$.
+        \label{case:anti-unification-initial:nvar}
+        This rule is applicable since it has no premises.
+
+        Then $[[Ξ]] = [[·]]$, $[[uM]] = [[α⁻]]$, 
+        and $[[aus1]] = [[aus2]] = [[·]]$.
+        Since $[[ [aus'i] uM' = iNi ]] = [[α⁻]]$
+        and $[[uM']]$ is not a metavariable, $[[uM']] = [[α⁻]]$.
+        Then we can take $[[ausr]] = [[·]]$, which satisfies the required properties:
+        \begin{itemize}
+            \item $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ holds vacuously since 
+            $[[Ξ' | uv uM']] = [[∅]]$; 
+            \item $[[ [ausr] uM' = uM ]]$, that is $[[ [·] α⁻ = α⁻ ]]$
+            holds by substitution properties;
+            \item the unique determination of $[[ [ausr]α̂⁻]]$ for $[[α̂⁻]] \in [[Ξ' | uv uM']] = [[∅]]$ holds vacuously.
+        \end{itemize}
+
+        \item \ruleref{\ottdruleAUShiftULabel}, i.e. 
+        $[[iN1]] = [[↑iP1]]$ and $[[iN2]] = [[↑iP2]]$.
+        \label{case:anti-unification-initial:shiftu}
+
+        Then since $[[ [aus'i] uM' = iNi ]] = [[↑iPi]]$ and $[[uM']]$ is not a metavariable,
+        $[[uM']] = [[↑uQ']]$, where $[[ [aus'i] uQ' = iPi ]]$. 
+        Let us show that $[[(Ξ', uQ', aus'1, aus'2)]]$ 
+        is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+        \begin{enumerate}
+            \item $[[Γ ; Ξ' ⊢ uQ']]$ holds by inversion of $[[Γ ; Ξ' ⊢ ↑uQ']]$;
+            \item $[[Γ ; · ⊢ aus'i : Ξ']]$ holds by assumption;
+            \item $[[ [aus'i] uQ' = iPi ]]$ holds by assumption.
+        \end{enumerate}
+
+        This way, by the completeness of anti-unification 
+        (\cref{lemma:anti-unification-completeness}),
+        the anti-unification algorithm succeeds on $[[iP1]]$ and $[[iP2]]$:
+        $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$,
+        which means that \ruleref{\ottdruleAUShiftULabel} is applicable to infer 
+        $[[Γ ⊨ ↑iP1 ≈au ↑iP2 ⫤ (Ξ, ↑uQ, aus1, aus2)]]$.
+
+        Moreover, by the induction hypothesis,
+        $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than $[[(Ξ', uQ', aus'1, aus'2)]]$,
+        which immediately implies that $[[(Ξ, ↑uQ, aus1, aus2)]]$ is more specific than
+        $[[(Ξ', ↑uQ', aus'1, aus'2)]]$ (we keep the same $[[ausr]]$).
+
+        \item \ruleref{\ottdruleAUForallLabel}, i.e. 
+        $[[iN1]] = [[∀pas.iN'1]]$ and $[[iN2]] = [[∀pas.iN'2]]$.
+        \label{case:anti-unification-initial:forall}
+        The proof is symmetric to the previous case.
+        Notice that the context $[[Γ]]$ is not changed in \ruleref{\ottdruleAUForallLabel}, 
+        as it represents the context in which the anti-unification variables must be instantiated,
+        rather than the context forming the types that are being anti-unified.
+
+        \item \ruleref{\ottdruleAUArrowLabel}, i.e.
+        $[[iN1]] = [[iP1 → iN'1]]$ and $[[iN2]] = [[iP2 → iN'2]]$.
+
+        Then since $[[ [aus'i] uM' = iNi ]] = [[iPi → iN'i]]$ and $[[uM']]$ is not a metavariable,
+        $[[uM']] = [[uQ' → uM'']]$, where $[[ [aus'i] uQ' = iPi ]]$ and $[[ [aus'i] uM'' = iN''i ]]$.
+
+        Let us show that $[[(Ξ', uQ', aus'1, aus'2)]]$
+        is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+        \begin{enumerate}
+            \item $[[Γ ; Ξ' ⊢ uQ']]$ holds by inversion of $[[Γ ; Ξ' ⊢ uQ' → uM'']]$;
+            \item $[[Γ ; · ⊢ aus'i : Ξ']]$ holds by assumption;
+            \item $[[ [aus'i] uQ' = iPi ]]$ holds by assumption.
+        \end{enumerate}
+
+        Similarly, $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+
+        Then by the completeness of anti-unification (\cref{lemma:anti-unification-completeness}),
+        the anti-unification algorithm succeeds on $[[iP1]]$ and $[[iP2]]$:
+        $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ1, uQ, aus1, aus2)]]$;
+        and on $[[iN'1]]$ and $[[iN'2]]$:
+        $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ2, uM''', aus3, aus4)]]$.
+        Notice that $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined, 
+        in other words, for any $[[β̂⁻]] \in [[Ξ1]] \cap [[Ξ2]]$,
+        $[[ [aus1] β̂⁻ = [aus2] β̂⁻ ]]$ and $[[ [aus3] β̂⁻ = [aus4] β̂⁻ ]]$,
+        which follows immediately from \cref{obs:names-defined-by-mapping}.
+        This way, the algorithm proceeds by applying \ruleref{\ottdruleAUArrowLabel} and returns
+        $[[(Ξ1 ∪ Ξ2, uQ → uM''', aus1 & aus3, aus2 & aus4)]]$.
+
+        It is left to construct $[[ausr]]$ such that $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
+        By the induction hypothesis, there exist $[[ausr1]]$ and $[[ausr2]]$ such that
+        $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$, 
+        $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$,
+        $[[ [ausr1] uQ' = uQ ]]$, and $[[ [ausr2] uM'' = uM''' ]]$.
+
+        Let us show that $[[ausr]] = [[ausr1 & ausr2]]$ is defined, 
+        in other words, for any $[[β̂⁻]] \in [[(Ξ' | uv uQ')]] \cap [[(Ξ' | uv uM'')]]$,
+        $[[ [ausr1] β̂⁻ = [ausr2] β̂⁻ ]]$.It holds because by the induction hypothesis,
+        $[[ [ausri] β̂⁻ ]]$ is uniquely determined by $[[ [aus'1] β̂⁻ ]]$, $[[ [aus'2] β̂⁻ ]]$, and $[[Γ]]$,
+        none of which depends on $i$.
+
+        Let us show that $[[ausr]] = [[ausr1 & ausr2]]$ satisfies the required properties:
+        \begin{itemize}
+            \item $[[Γ ; Ξ1 ∪ Ξ2 ⊢ ausr1 & ausr2 : (Ξ' | uv uM')]]$ holds since 
+            $[[Ξ' | uv uM']] = [[Ξ' | uv uQ' → uM'']] = [[(Ξ' | uv uQ') ∪ (Ξ' | uv uM'')]]$,
+            $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$ and $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$
+            \ilyam{add a lemma?};
+            \item $[[ [ausr] uM' ]] = [[ [ausr] (uQ' → uM'') ]] = [[ [ausr | uv uQ'] uQ' → [ausr | uv uM''] uM'' ]] =
+            [[ [ausr1] uQ' → [ausr2] uM'' ]] = [[ uQ → uM''']] = [[ uM ]]$;
+            \item Since $[[ [ausr]β̂⁻]]$ is either equal to  $[[ [ausr1]β̂⁻]]$ or $[[ [ausr2]β̂⁻]]$,
+            it inherits their property that it is uniquely determined by $[[ [aus'1]β̂⁻]]$, $[[ [aus'2]β̂⁻]]$, and $[[Γ]]$.
+        \end{itemize}
+
+        \item $[[iP1]] = [[iP2]] = [[a⁺]]$. This case is symmetric to \cref{case:anti-unification-initial:nvar}.
+        \item $[[iP1]] = [[↓iN1]]$ and $[[iP2]] = [[↓iN2]]$. This case is symmetric to \cref{case:anti-unification-initial:shiftu}
+        \item $[[iP1]] = [[∃nas.iP'1]]$ and $[[iP2]] = [[∃nas.iP'2]]$. This case is symmetric to \cref{case:anti-unification-initial:forall}
+        \end{caseof}
+
+\end{proof}

### 2023-06-05 proofs/tex/solution-merge-lemmas.tex
WIP: finish merge lemmas

@@
 
-\begin{lemma} [Soundness of Merge of Unification Solution Entries]
-    \label{lemma:unif-sol-ent-merge-soundness}
+\begin{lemma} [Soundness of Solution Entry Merge]
+\label{lemma:entry-merge-soundness}
 For a fixed context $[[Γ]]$,
-    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$,
-    where $[[Γ ⊢ usEntryi]]$ is an equivalence-shaped restriction
-    (i.e. it has shape $[[α̂⁺ :≈ iP]]$ or $[[α̂⁻ :≈ iN]]$ but not $[[α̂⁺ :≥ iP]]$).
+suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$. 
 If $[[usEntry1 & usEntry2]]$ is defined then
@@
          \end{enumerate}
-        \item \ruleref{\ottdruleSMENEqEqLabel}. The negative case is proved in exactly the same way.
-    \end{caseof}
-\end{proof}
-
-\begin{lemma} [Soundness of Merge of Unification Solutions]
-    \label{lemma:unif-sol-merge-soundness}
-    Suppose that $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$ 
-    are unification solutions (i.e. $[[us1]]$ and $[[us2]]$ can only have equivalence-shaped restrictions).
-    If $[[us1 & us2]]$ is defined then
-    \begin{enumerate}
-        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$,
-        \item $[[Θ|varset1 ⊢ us1 & us2 ⇒ us1]]$, and
-        \item $[[Θ|varset2 ⊢ us1 & us2 ⇒ us2]]$.
-    \end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us prove the properties separately:
-    \begin{enumerate}
-        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$.
-        It suffices to prove the following two properties:
-        \begin{itemize}
-            \item The set of variables of the entries of $[[us1 & us2]]$ 
-            coincides with $[[varset1 ∪ varset2]]$\\
-            By definition, $[[us1 & us2]]$ consists of three parts:
-            entries of $[[us1]]$ that do not have matching entries of $[[us2]]$,
-            entries of $[[us2]]$ that do not have matching entries of $[[us1]]$,
-            and the merge of matching entries.
-            It means that $[[dom(us1 & us2)]] = [[dom(us1) \ dom(us2) ∪ dom(us2) \ dom(us1) ∪ 
-            dom(us1) ∩ dom(us2)]] = [[dom(us1) ∪ dom(us2)]]$, which, since 
-            $[[usi : Θ | varseti]]$, is equal to $[[varset1 ∪ varset2]]$.
-
-            \item Each entry of $[[us1 & us2]]$ restricting $[[α̂±]]$ is well-formed in
-            the corresponding context $[[Θ(α̂±)]]$.\\
-            Let us consider an arbitrary entry $[[usEntry]]$ of $[[us1 & us2]]$ restricting
-            $[[α̂±]]$. Then there are three cases:
-            \begin{enumerate}
-                \item $[[usEntry]]$ the entry is from $[[us1]]$ and does not have a matching entry 
-                in $[[us2]]$, i.e. $[[α̂±]] \in [[dom(us1) \ dom(us2)]]$.
-                Then $[[usEntry]]$ is well-formed in $[[Θ | varset1]]$ by assumption.
-                \item $[[usEntry]]$ the entry is from $[[us2]]$ and does not
-                have a matching entry in $[[us1]]$. This case is symmetric.
-                \item $[[usEntry]]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
-                 and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
-                 Since $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$,
-                 $[[α̂±]] \in [[dom(Θ|varset1) ∩ dom(Θ|varset2)]]$, i.e. there 
-                 is an entry $[[ α̂±[Γ] ]] \in [[Θ]]$, and 
-                 $[[usEntry1]]$ and $[[usEntry2]]$ are well-formed in $[[Γ]]$.
-                 Then by \cref{lemma:unif-sol-ent-merge-soundness}, 
-                 $[[Γ ⊢ usEntry1 & usEntry2]]$, where $[[Γ]] = [[Θ(α̂±)]]$.
-            \end{enumerate}
-        \end{itemize}
-        \item $[[Θ|varset1 ⊢ us1 & us2 ⇒ us1]]$
-        We need to show that for every entry $[[usEntry]]$ from $[[us1]]$,
-        there is an entry $[[usEntry']]$ from $[[us1 & us2]]$ such that
-        $[[Θ|varset1 ⊢ usEntry' ⇒ usEntry]]$. 
-        Let us consider an arbitrary $[[usEntry]]$ from $[[us1]]$ restricting $[[α̂±]]$.
-        Then there are two cases:
-        \begin{itemize}
-            \item $[[usEntry]]$ does not have a matching entry in $[[us2]]$.
-            Then $[[usEntry]]$ is also in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry ⇒ usEntry]]$ 
-            by reflexivity (\cref{lemma:entry-weakening-preorder}).
-            \item $[[usEntry]]$ has a matching entry $[[usEntry']]$ in $[[us2]]$.
-            Then $[[usEntry & usEntry']]$ is in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry & usEntry' ⇒ usEntry]]$ by \cref{lemma:unif-sol-ent-merge-soundness}.
-        \end{itemize}
-        \item $[[Θ|varset2 ⊢ us1 & us2 ⇒ us2]]$ is proved analogously.
-    \end{enumerate}
-\end{proof}
-
-\begin{lemma} [Soundness of Solution Entry Merge]
-\label{lemma:entry-merge-soundness}
-For a fixed context $[[Γ]]$,
-suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$. 
-If $[[usEntry1 & usEntry2]]$ is defined then
-\begin{enumerate}
-    \item $[[Γ ⊢ usEntry1 & usEntry2]]$
-    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry1]]$
-    \item $[[Γ ⊢ usEntry1 & usEntry2 ⇒ usEntry2]]$
-\end{enumerate}
-\end{lemma}
-\begin{proof}
-    Let us consider the rule forming $[[Γ ⊢ usEntry1 & usEntry2 = usEntry]]$.
-    \begin{caseof}
-        \item \ruleref{\ottdruleSMEPEqEqLabel} or \ruleref{\ottdruleSMENEqEqLabel}
-        The proof is analogous to the proof of \cref{lemma:unif-sol-ent-merge-soundness}.
+        \item \ruleref{\ottdruleSMENEqEqLabel} the negative case is proved in exactly the same way as the positive one.
         \item \ruleref{\ottdruleSMESupSupLabel} 
@@
 \begin{proof}
-    The proof repeats the proof of \cref{lemma:unif-sol-merge-soundness},
-    but uses \cref{lemma:entry-merge-soundness} instead of 
-    \cref{lemma:unif-sol-ent-merge-soundness}.
-\end{proof}
 
-\begin{lemma} [Completeness of Unification Solution Entry Merge]
-    \label{lemma:unif-entry-merge-completeness}
-    For a fixed context $[[Γ]]$,
-    suppose that  $[[Γ ⊢ usEntry1]]$ and $[[Γ ⊢ usEntry2]]$, 
-    where $[[Γ ⊢ usEntryi]]$ are equivalence-shaped restrictions.
-    Suppose there exists an equivalence-shaped restriction 
-    $[[Γ ⊢ usEntry]]$ such that $[[Γ ⊢ usEntry ⇒ usEntry1]]$ and $[[Γ ⊢ usEntry ⇒ usEntry2]]$.
-    Then $[[usEntry1 & usEntry2]]$ is defined and
-    $[[Γ ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
-\end{lemma}
-\begin{proof}
-    By the soundness of unification solution entry merge (\cref{lemma:unif-sol-ent-merge-soundness}),
-    Since the restrictions are equivalence-shaped, 
-    the only possible rules inferring the weakening judgments are
-    the positive \ruleref{\ottdruleSImpEPEqEqLabel} and the negative \ruleref{\ottdruleSImpENEqEqLabel}.
-    Since $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$ 
-    have the same polarity, we consider the positive case,
-    and the negative case is proved analogously.
+    Let us prove the properties separately:
+    \begin{enumerate}
+        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$.
+        It suffices to prove the following two properties:
+        \begin{itemize}
+            \item The set of variables of the entries of $[[us1 & us2]]$ 
+            coincides with $[[varset1 ∪ varset2]]$\\
+            By definition, $[[us1 & us2]]$ consists of three parts:
+            entries of $[[us1]]$ that do not have matching entries of $[[us2]]$,
+            entries of $[[us2]]$ that do not have matching entries of $[[us1]]$,
+            and the merge of matching entries.
+            It means that $[[dom(us1 & us2)]] = [[dom(us1) \ dom(us2) ∪ dom(us2) \ dom(us1) ∪ 
+            dom(us1) ∩ dom(us2)]] = [[dom(us1) ∪ dom(us2)]]$, which, since 
+            $[[usi : Θ | varseti]]$, is equal to $[[varset1 ∪ varset2]]$.
 
-    Let us assume that $[[usEntry1]]$ is $[[pua :≈ iP1]]$,
-    $[[usEntry2]]$ is $[[pua :≈ iP2]]$, $[[usEntry]]$ is $[[pua :≈ iQ]]$.
-    Then by inversion of \ruleref{\ottdruleSImpEPEqEqLabel},
+            \item Each entry of $[[us1 & us2]]$ restricting $[[α̂±]]$ is well-formed in
+            the corresponding context $[[Θ(α̂±)]]$.\\
+            Let us consider an arbitrary entry $[[usEntry]]$ of $[[us1 & us2]]$ restricting
+            $[[α̂±]]$. Then there are three cases:
+            \begin{enumerate}
+                \item $[[usEntry]]$ the entry is from $[[us1]]$ and does not have a matching entry 
+                in $[[us2]]$, i.e. $[[α̂±]] \in [[dom(us1) \ dom(us2)]]$.
+                Then $[[usEntry]]$ is well-formed in $[[Θ | varset1]]$ by assumption.
+                \item $[[usEntry]]$ the entry is from $[[us2]]$ and does not
+                have a matching entry in $[[us1]]$. This case is symmetric.
+                \item $[[usEntry]]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
+                 and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
+                 Since $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$,
+                 $[[α̂±]] \in [[dom(Θ|varset1) ∩ dom(Θ|varset2)]]$, i.e. there 
+                 is an entry $[[ α̂±[Γ] ]] \in [[Θ]]$, and 
+                 $[[usEntry1]]$ and $[[usEntry2]]$ are well-formed in $[[Γ]]$.
+                 Then by \cref{lemma:entry-merge-soundness}, 
+                 $[[Γ ⊢ usEntry1 & usEntry2]]$, where $[[Γ]] = [[Θ(α̂±)]]$.
+            \end{enumerate}
+        \end{itemize}
+        \item $[[Θ|varset1 ⊢ us1 & us2 ⇒ us1]]$
+        We need to show that for every entry $[[usEntry]]$ from $[[us1]]$,
+        there is an entry $[[usEntry']]$ from $[[us1 & us2]]$ such that
+        $[[Θ|varset1 ⊢ usEntry' ⇒ usEntry]]$. 
+        Let us consider an arbitrary $[[usEntry]]$ from $[[us1]]$ restricting $[[α̂±]]$.
+        Then there are two cases:
         \begin{itemize}
-        \item $[[Γ ⊢ iQ ≈ iP1]]$, and hence $[[nf(iQ) = nf(iP1)]]$ by \cref{lemma:subt-equiv-algorithmization},
-        \item $[[Γ ⊢ iQ ≈ iP2]]$, and hence $[[nf(iQ) = nf(iP2)]]$ by \cref{lemma:subt-equiv-algorithmization},
+            \item $[[usEntry]]$ does not have a matching entry in $[[us2]]$.
+            Then $[[usEntry]]$ is also in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry ⇒ usEntry]]$ 
+            by reflexivity (\cref{lemma:entry-weakening-preorder}).
+            \item $[[usEntry]]$ has a matching entry $[[usEntry']]$ in $[[us2]]$.
+            Then $[[usEntry & usEntry']]$ is in $[[us1 & us2]]$ and $[[Θ|varset1 ⊢ usEntry & usEntry' ⇒ usEntry]]$ 
+            by \cref{lemma:entry-merge-soundness}.
         \end{itemize}
-    which implies $[[nf(iP1) = nf(iP2)]]$.
-    It means that the merge of $[[usEntry1]]$ and $[[usEntry2]]$ is defined by \ruleref{\ottdruleSMEPEqEqLabel}:
-    $[[Γ ⊢ (pua :≈ iP1) & (pua :≈ iP2) = (pua :≈ iP1)]]$.
-
-    To show that $[[Γ ⊢ (pua :≈ iQ) ⇒ (pua :≈ iP1)]]$, we apply \ruleref{\ottdruleSImpEPEqEqLabel}:
-    $[[Γ ⊢ iP1 ≈ iQ]]$ is obtained from $[[Γ ⊢ iQ ≈ iP1]]$ by symmetry of equivalence.
-\end{proof}
-
-\begin{lemma} [Completeness of Unification Solution Merge] 
-    \label{lemma:unif-sol-merge-completeness}
-    Suppose that $[[us1 : Θ|varset1]]$ and $[[us2 : Θ|varset2]]$ 
-    are equivalence-only unification solutions
-    and there exists $[[us]]$
-    such that $[[Θ ⊢ us ⇒ us1]]$ and $[[ Θ ⊢ us ⇒ us2]]$.
-    Then $[[us1 & us2]]$ is defined and 
-    $[[ Θ ⊢ us ⇒ us1 & us2 ]]$.
-\end{lemma}
-\begin{proof}
-    By  definition, $[[us1 & us2]]$ is a union of entries of $[[us1]]$, which do not have matching entries in $[[us2]]$,
-    entries of $[[us2]]$, which do not have matching entries in $[[us1]]$, and the merge of matching entries.
-    It is clear that the first two components of this union exist. Let us show that the third component exists, 
-    i.e. that for every two entries $[[usEntry1]] \in [[us1]]$ and $[[usEntry2]] \in [[us2]]$
-    restricting the same variable $[[α̂±]]$, $[[usEntry1 & usEntry2]]$ is defined.
-
-    $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ restricting $[[α̂±]]$ such that
-    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1]]$. Similarly, since there can be only one entry in $[[usEntry]]$ restricting 
-    the same variable$[[α̂±]]$, $[[Θ2 ⊢ us ⇒ us2]]$ means $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry2]]$.
-    Then by \cref{lemma:unif-sol-ent-merge-soundness}, $[[usEntry1 & usEntry2]]$ is defined and 
-    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
-
-    Let us show: $[[ Θ ⊢ us ⇒ us1 & us2 ]]$, i.e. for every $[[usEntry']] \in [[us1 & us2]]$ restricting 
-    $[[α̂±]]$, there exists $[[usEntry]] \in [[us]]$ such that $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$.
-    Let us consider three cases:
-    \begin{caseof}
-        \item $[[usEntry']]$ is from $[[us1]]$ and does not have a matching entry in $[[us2]]$.
-        Then $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ such that
-        $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$, as required.
-        \item $[[usEntry']]$ is from $[[us2]]$ and does not have a matching entry in $[[us1]]$.
-        This case is symmetric to the previous one.
-        \item $[[usEntry']]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
-        and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
-        Then as noted previously, $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]] = [[usEntry']]$.
-    \end{caseof}
+        \item $[[Θ|varset2 ⊢ us1 & us2 ⇒ us2]]$ is proved analogously.
+    \end{enumerate}
 \end{proof}
@@
 \begin{proof}
-    Notice that $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$
-    restrict the same variable and have the same polarity. If the polarity is negative,
-    then the proof is analogous to the proof of \cref{lemma:unif-entry-merge-completeness}.
 
@@
     \begin{caseof}
+
+        \item $[[usEntry]]$ is $[[nua :≈ iM]]$
+        Then since $[[usEntry1]]$, $[[usEntry2]]$, and $[[usEntry]]$
+        restrict the same variable, they have the same polarity. 
+        So $[[usEntry1]]$ is $[[nua :≈ iN1]]$, $[[usEntry2]]$ is $[[nua :≈ iN2]]$, 
+        Then by inversion of \ruleref{\ottdruleSImpENEqEqLabel},
+        \begin{itemize}
+            \item $[[Γ ⊢ iM ≈ iN1]]$, and hence $[[nf(iM) = nf(iN1)]]$ by \cref{lemma:subt-equiv-algorithmization},
+            \item $[[Γ ⊢ iM ≈ iN2]]$, and hence $[[nf(iM) = nf(iN2)]]$ by \cref{lemma:subt-equiv-algorithmization},
+        \end{itemize}
+        which implies $[[nf(iN1) = nf(iN2)]]$.
+        It means that the merge of $[[usEntry1]]$ and $[[usEntry2]]$ is defined by \ruleref{\ottdruleSMENEqEqLabel}:
+        $[[Γ ⊢ (nua :≈ iN1) & (nua :≈ iN2) = (nua :≈ iN1)]]$.
+
+        To show that $[[Γ ⊢ (nua :≈ iM) ⇒ (nua :≈ iN1)]]$, we apply \ruleref{\ottdruleSImpENEqEqLabel}:
+        $[[Γ ⊢ iN1 ≈ iM]]$ is obtained from $[[Γ ⊢ iM ≈ iN1]]$ by symmetry of equivalence.
+
         \item $[[usEntry]]$ is $[[pua :≥ iP]]$.
@@
 \begin{proof}
-    The proof is analogous to the proof of \cref{lemma:unif-sol-merge-completeness}.
-    The only difference is that we use \cref{lemma:entry-merge-completeness} instead of
-    \cref{lemma:unif-entry-merge-completeness}.
+    By  definition, $[[us1 & us2]]$ is a union of entries of $[[us1]]$, which do not have matching entries in $[[us2]]$,
+    entries of $[[us2]]$, which do not have matching entries in $[[us1]]$, and the merge of matching entries.
+    It is clear that the first two components of this union exist. Let us show that the third component exists, 
+    i.e. that for every two entries $[[usEntry1]] \in [[us1]]$ and $[[usEntry2]] \in [[us2]]$
+    restricting the same variable $[[α̂±]]$, $[[usEntry1 & usEntry2]]$ is defined.
+
+    $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ restricting $[[α̂±]]$ such that
+    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1]]$. Similarly, since there can be only one entry in $[[usEntry]]$ restricting 
+    the same variable$[[α̂±]]$, $[[Θ2 ⊢ us ⇒ us2]]$ means $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry2]]$.
+    Then by \cref{lemma:entry-merge-completeness}, $[[usEntry1 & usEntry2]]$ is defined and 
+    $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]]$.
+
+    Let us show: $[[ Θ ⊢ us ⇒ us1 & us2 ]]$, i.e. for every $[[usEntry']] \in [[us1 & us2]]$ restricting 
+    $[[α̂±]]$, there exists $[[usEntry]] \in [[us]]$ such that $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$.
+    Let us consider three cases:
+    \begin{caseof}
+        \item $[[usEntry']]$ is from $[[us1]]$ and does not have a matching entry in $[[us2]]$.
+        Then $[[Θ|varset1 ⊢ us ⇒ us1]]$ means that there exists $[[usEntry]] \in [[us]]$ such that
+        $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry']]$, as required.
+        \item $[[usEntry']]$ is from $[[us2]]$ and does not have a matching entry in $[[us1]]$.
+        This case is symmetric to the previous one.
+        \item $[[usEntry']]$ is the merge of two matching entries $[[usEntry1]] \in [[us1]]$
+        and $[[usEntry2]] \in [[us2]]$ restricting $[[α̂±]]$.
+        Then as noted previously, $[[Θ(α̂±) ⊢ usEntry ⇒ usEntry1 & usEntry2]] = [[usEntry']]$.
+    \end{caseof}
 \end{proof}

### 2023-06-05 proofs/tex/unif-solution-merge-lemmas.tex
WIP: finish merge lemmas

\begin{lemma} [Soundness of Merge of Unification Solutions]
    \label{lemma:unif-sol-merge-soundness}
    Suppose that $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$ 
    are normalized unification solutions 
    (i.e. $[[us1]]$ and $[[us2]]$ can only have equivalence-shaped restrictions, 
    which types are normalized),
    If $[[us1 & us2]]$ is defined then
    $[[us1 & us2]] = [[us1]] ∪ [[us2]]$, where
\end{lemma}
\begin{proof}
    \hfill
    \begin{itemize}
        \item $[[us1 & us2]] \subseteq [[us1]] \cup [[us2]]$\\
        By definition, 
        $[[us1 & us2]]$ consists of three parts:
        entries of $[[us1]]$ that do not have matching entries of $[[us2]]$,
        entries of $[[us2]]$ that do not have matching entries of $[[us1]]$,
        and the merge of matching entries.

        If $[[usEntry]]$ is from the first or the second part, 
        then $[[usEntry]] \in [[us1]] \cup [[us2]]$.

        If $[[usEntry]]$ is from the third part,
        then $[[usEntry]]$ is the merge of two matching entries
        $[[usEntry1]] \in [[us1]]$ and $[[usEntry2]] \in [[us2]]$.
        Since $[[us1]]$ and $[[us2]]$ are normalized unification solutions, 
        $[[usEntry1]]$ and $[[usEntry2]]$ have one of the following forms:
        \begin{itemize}
            \item $[[α1̂⁺ :≈ iP1]]$ and $[[α2̂⁺ :≈ iP2]]$, 
                where $[[iP1]]$ and $[[iP2]]$ are normalized,
                and then since $[[usEntry1 & usEntry2]]$ exists, 
                \ruleref{\ottdruleSMEPEqEqLabel} was applied to infer it.
                It means that 
                $[[usEntry]] = ([[usEntry1 & usEntry2]]) = [[usEntry1]] = [[usEntry2]]$;
            \item $[[α1̂⁻ :≈ iN1]]$ and $[[α2̂⁻ :≈ iN2]]$, 
               then symmetrically, 
               $[[usEntry]] = ([[usEntry1 & usEntry2]]) = [[usEntry1]] = [[usEntry2]]$
        \end{itemize}
        In both cases, $[[usEntry]] \in [[us1]] \cup [[us2]]$.
        \item $[[us1]] \cup [[us2]] \subseteq [[us1 & us2]]$\\
        Let us take 
        an arbitrary $[[usEntry]] \in [[us1]]$.
        Then since $[[us1]]$ is a unification solution,
         $[[usEntry]]$ has one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP]]$ where $[[iP]]$ is normalized.
            If $[[α̂⁺]] \notin [[dom(us2)]]$, then $[[usEntry]] \in [[us1 & us2]]$.
            Otherwise there is a normalized
            $[[usEntry]] = [[(α̂⁺ :≈ iP')]] \in [[us2]]$ and then
            since $[[us1 & us2]]$ exists, 
            \ruleref{\ottdruleSMEPEqEqLabel} was applied to construct
            $[[usEntry & usEntry']] \in [[us1 & us2]]$.
            By inversion of \ruleref{\ottdruleSMEPEqEqLabel},
            $[[usEntry & usEntry']] = [[usEntry]]$, and
            $[[nf(iP) = nf(iP')]]$, which since $[[iP]]$
            and $[[iP']]$ are normalized, implies that $[[iP = iP']]$, 
            that is $[[usEntry]] = [[usEntry']]$.
            This way, $[[usEntry']] = [[usEntry]] \in [[us1 & us2]]$.
            \item $[[α̂⁻ :≈ iN]]$ where $[[iN]]$ is normalized.
            Then symmetrically, $[[usEntry]] \in [[us1 & us2]]$.
        \end{itemize}
        Similarly, if we take an arbitrary $[[usEntry']] \in [[us2]]$,
        then $[[usEntry']] \in [[us1 & us2]]$. In fact, this is 
        the case where normalization is important, since
        \ruleref{\ottdruleSMEPEqEqLabel} returns the left-hand
        operand of $\&$, but as noted above, $[[nf(iP) = nf(iP')]]$
        implies $[[iP = iP']]$ for normalized types.
    \end{itemize}
\end{proof}

\begin{corollary}
    \label{corollary:unif-sol-merge-soundness}
    Suppose that $[[us1 : Θ | varset1]]$ and $[[us2 : Θ | varset2]]$ 
    are normalized unification solutions 
    If $[[us1 & us2]]$ is defined then
    \begin{enumerate}
        \item $[[us1 & us2 : Θ|varset1 ∪ varset2]]$,
        \item $[[us1 & us2]]$ is a normalized unification solution, 
        \item $[[us1 & us2 | varseti]] = [[usi]]$ for $i = 1, 2$,
    \end{enumerate}
\end{corollary}
\begin{proof}
    The first two properties follow immediately from
    \cref{lemma:unif-sol-merge-soundness}.

    To prove the third property, first notice that
    $[[us1 & us2 | varseti]] \supseteq [[usi]]$ follows immediately from 
    $[[us1 & us2]] = [[us1]] \cup [[us2]]$.
    For the other inclusion $[[us1 & us2 | varset1]] \subseteq [[us1]]$,
    let us take an arbitrary $[[usEntry]] \in [[us1 & us2 | varset1]]$.
    Then $[[usEntry]]$ is either from $[[us1]]$ (if it does not have a matching entry in
    $[[us2]]$) or it is a result of merge of two matching entries, 
    but since $[[us1 & us2]]$ exist and $[[us1]]$ and $[[us2]]$ 
    are normalized unification solutions, the matching
    entries of $[[us1]]$ and $[[us2]]$ are equal, and they are equal to their merge.
    In both cases, $[[usEntry]] \in [[us1]]$.
    Similarly, $[[us1 & us2 | varset2]] \subseteq [[us2]]$.
\end{proof}

\begin{lemma} [Completeness of Unification Solution Merge] 
    \label{lemma:unif-sol-merge-completeness}
    Suppose that $[[us]]$ is a unification solution, 
    and $[[varset1]]$, $[[varset2]]$ are sets of variables. 
    Then $[[us | varset1 & us | varset2]]$ is defined and
    equal to $[[us | varset1 ∪ varset2]]$.
\end{lemma}
\begin{proof}
    $[[us | varset1 & us | varset2]]$ is defined as the union of three parts:
    entries of $[[us | varset1]]$ that do not have matching entries of $[[us | varset2]]$,
    entries of $[[us | varset2]]$ that do not have matching entries of $[[us | varset1]]$,
    and the merge of matching entries.
    The first two parts are defined. The merge of matching entries is defined
    by \ruleref{\ottdruleSMEPEqEqLabel}, since the matching entries must be equal
    if they both belong to $[[us]]$.

    It remains to show that $[[us | varset1 & us | varset2]] = [[us | varset1 ∪ varset2]]$.
    It is easy to see that the three parts comprising $[[us | varset1 & us | varset2]]$ 
    correspond to the three parts comprising 
    $[[us | varset1 ∪ varset2]] = [[us | (varset1 \ varset2)]] \cup [[us | (varset2 \ varset1)]]
    \cup [[us | varset1 ∩ varset2]]$. 
\end{proof}

### 2023-06-05 proofs/tex/unification-lemmas.tex
WIP: finish merge lemmas

@@
 
-        By the soundness of the unification solution merge (\cref{lemma:unif-sol-merge-soundness}), we have:
+        By the soundness of the unification solution merge (\cref{corollary:unif-sol-merge-soundness}), we have:
         \begin{itemize}
@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        assume there exists an equivalence-only unification solution
+        assume there exists a normalized unification solution
         $[[us : Θ | uv uP]]$ such that $[[ [us]uP = iQ ]]$.
-        Then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+        Then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$.
 
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        assume there exists an equivalence-only unification solution
+        assume there exists a normalized unification solution
         $[[us : Θ | uv uN]]$        
@@
             $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$, we need to show that
-            $[[us|uv uP & us|uv uN']] = [[us]]$. 
-            By the completeness of the unification solution merge 
-            (\cref{lemma:unif-sol-merge-completeness}), 
-            $[[us|uv uP & us|uv uN']] = [[us|uv uP ∪ uv uN']]$,
-            and $[[us|uv uP ∪ uv uN']] = [[us|uv uN]] = [[us]]$, since $[[us : Θ | uv uN]]$.
+            $[[us|uv uP & us|uv uN']]$ is defined and equal to $[[us]]$. 
+            It holds by the completeness  of the unification solution merge 
+            (\cref{lemma:unif-sol-merge-completeness}):
+            \begin{itemize}
+                \item $[[us|uv uP : Θ | uv uP]]$,
+                \item $[[us|uv uN' : Θ | uv uN']]$,
+                \item $[[us|uv uP]]$ and  $[[us|uv uN']]$ are normalized unification solution since so is $[[us]]$,
+                \item $[[us : Θ | uv uP ∪ uv uN']]$ since $[[uv uP ∪ uv uN']] = [[uv uN]]$
+            \end{itemize}
 

### 2023-06-06 proofs/tex/anti-unification-lemmas.tex
anti-unification: & → ∪

@@
+\begin{observation}[Anti-unification algorithm is deterministic]
+    \label{obs:anti-unification-deterministic}
+    \hfill\\
+\begin{itemize}
+    \item [$+$] If $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and 
+    $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ', uQ', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uQ]] = [[uQ']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+    \item [$-$] If $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
+    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ', uM', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uM]] = [[uM']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+\end{itemize}
+\end{observation}
+\begin{proof}
+    By trivial induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+\end{proof}
+
+\begin{observation} \label{obs:names-defined-by-mapping}
+    Names of the anti-unification variables are uniquely defined by
+    the types they are mapped to by the resulting substitutions.
+
+    \begin{itemize}
+        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+        \item [$-$]  Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
+        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+    \end{itemize}
+\end{observation}
+\begin{proof}
+    By simple induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+    Let us consider tha last rule applied to infer this judgment.
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUPVarLabel} or \ruleref{\ottdruleAUNVarLabel},
+        then $[[Ξ]] = [[·]]$, and the property holds vacuously.
+
+        \item \ruleref{\ottdruleAUAULabel}
+        Then  $[[Ξ]] = [[â⁻_{iN1, iN2}]]$,
+        $[[aus1]] = [[â⁻_{iN1, iN2} :≈ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} :≈ iN2]]$.
+        So the property holds trivially.
+
+        \item \ruleref{ottdruleAUArrowLabel}
+        In this case, $[[Ξ]] = [[Ξ' ∪ Ξ'']]$, $[[aus1]] = [[aus'1 ∪ aus''1]]$, and 
+        $[[aus2]] = [[aus'2 ∪ aus''2]]$,
+        where the property holds for ($[[Ξ']]$, $[[aus'1]]$, $[[aus'2]]$) and 
+        ($[[Ξ'']]$, $[[aus''1]]$, $[[aus''2]]$) by the induction hypothesis.
+        Then since the union of solutions does not change the types the variables are mapped to,
+        the required property holds for $[[Ξ]]$, $[[aus1]]$, and $[[aus2]]$.
+
+        \item For the other rules, the resulting $[[Ξ]]$ is taken from the recursive call
+        and the required property holds immediately by the induction hypothesis.
+    \end{caseof}
+\end{proof}
+
 \begin{lemma}[Soundness of Anti-Unification] \label{lemma:anti-unification-soundness}
@@
             \item $[[Γ ; Ξ ⊢ uQ]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
+            are anti-unification solutions, and
             \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
@@
             \item $[[Γ ; Ξ ⊢ uM]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
+            are anti-unification solutions, and
             \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
@@
 
-                Notice, that the induction hypothesis is applicable to 
+                Notice that the induction hypothesis is applicable to 
                 $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$:
@@
 
-                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ and $[[Γ ; · ⊢ aus'i : Ξ']]$, 
-                           and hence, $[[Γ ; · ⊢ ausi ∪ aus'i : Ξ ∪ Ξ']]$ (\cref{todo});
+                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ and $[[Γ ; · ⊢ aus'i : Ξ']]$
+                        Then $[[Γ ; · ⊢ ausi ∪ aus'i : Ξ ∪ Ξ']]$
+                        are well-defined anti-unification solution.
+                        Let us take an arbitrary $[[β̂⁻]] \in [[Ξ ∪ Ξ']]$.
+                        If $[[β̂⁻]] \in [[Ξ]]$.
+                        then $[[Γ ; · ⊢ ausi : Ξ]]$ implies that $[[ausi]]$, and hence,
+                        $[[ausi ∪ aus'i]]$ contains an entry well-formed in $[[Γ]]$.
+                        If $[[β̂⁻]] \in [[Ξ']]$, the reasoning is symmetric.
+
+                        $[[ausi ∪ aus'i]]$ is a well-defined anti-unification solution:
+                        any anti-unification variable occurs 
+                        uniquely $[[ausi ∪ aus'i]]$, since by \cref{obs:names-defined-by-mapping},
+                        the name of the variable is in one-to-one correspondence with 
+                        the pair of types it is mapped to
+                        by $[[aus1]]$ and $[[aus2]]$, 
+                        an is in one-to-one correspondence with the pair of types it is mapped to
+                        by $[[aus'1]]$ and $[[aus'2]]$ 
+                        i.e.  if $[[β̂⁻]] \in [[Ξ ∩ Ξ']]$ then $[[ [aus1]β̂⁻ = [aus'1]β̂⁻ ]]$,
+                        and $[[ [aus2]β̂⁻ = [aus'2]β̂⁻ ]]$.
 
@@
                     Since $[[ausi ∪ aus'i]]$ restricted to $[[Ξ]]$ is $[[ausi]]$,
+                    and $[[ausi ∪ aus'i]]$ restricted to $[[Ξ']]$ is $[[aus'i]]$,
                     we have $[[ [ausi ∪ aus'i] uQ = iPi ]]$ and 
@@
 
-\begin{observation} \label{obs:names-defined-by-mapping}
-    Names of the anti-unification variables are uniquely defined by
-    the types they are mapped to by the resulting substitutions.
-
-    \begin{itemize}
-        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
-        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-        then for any $[[β̂⁻]] \in [[Ξ]]$,
-        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
-        \item [$-$]  Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
-        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
-        then for any $[[β̂⁻]] \in [[Ξ]]$,
-        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
-    \end{itemize}
-\end{observation}
-\begin{proof}
-    By simple induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
-    Let us consider tha last rule applied to infer this judgment.
-    \begin{caseof}
-        \item \ruleref{\ottdruleAUPVarLabel} or \ruleref{\ottdruleAUNVarLabel},
-        then $[[Ξ]] = [[·]]$, and the property holds vacuously.
-
-        \item \ruleref{\ottdruleAUAULabel}
-        Then  $[[Ξ]] = [[â⁻_{iN1, iN2}]]$,
-        $[[aus1]] = [[â⁻_{iN1, iN2} :≈ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} :≈ iN2]]$.
-        So the property holds trivially.
-
-        \item \ruleref{ottdruleAUArrowLabel}
-        In this case, $[[Ξ]] = [[Ξ' ∪ Ξ'']]$, $[[aus1]] = [[aus'1 & aus''1]]$, and 
-        $[[aus2]] = [[aus'2 & aus''2]]$,
-        where the property holds for ($[[Ξ']]$, $[[aus'1]]$, $[[aus'2]]$) and 
-        ($[[Ξ'']]$, $[[aus''1]]$, $[[aus''2]]$) by the induction hypothesis.
-        Then since the merge of solutions does not change the types the variables are mapped to,
-        the required property holds for $[[Ξ]]$, $[[aus1]]$, and $[[aus2]]$.
-
-        \item For the other rules, the resulting $[[Ξ]]$ is taken from the recursive call
-        and the required property holds immediately by the induction hypothesis.
-    \end{caseof}
-\end{proof}
-
-\begin{observation}[Anti-unification algorithm is deterministic]
-    \label{obs:anti-unification-deterministic}
-    \hfill\\
-\begin{itemize}
-    \item [$+$] If $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and 
-    $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ', uQ', aus'1, aus'2)]]$,
-    then $[[Ξ]] = [[Ξ']]$, $[[uQ]] = [[uQ']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
-    \item [$-$] If $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
-    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ', uM', aus'1, aus'2)]]$,
-    then $[[Ξ]] = [[Ξ']]$, $[[uM]] = [[uM']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
-\end{itemize}
-\end{observation}
-\begin{proof}
-    By trivial induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
-\end{proof}
-
 % \renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
@@
                 \item $[[Γ ; Ξ' ⊢ uQ']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
+                are anti-unification solutions, and
                 \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
@@
                 \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
+                are anti-unification solutions, and
                 \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
@@
             $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ2, uM, aus3, aus4)]]$ succeed.
-            The result of the algorithm is $[[(Ξ1 ∪ Ξ2, uQ → uM, aus1 & aus3, aus2 & aus4)]]$,
-            and it is left to show that $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined.
-
-            From \cref{obs:names-defined-by-mapping}, we know that 
-            if $[[β̂⁻ ∊ dom(aus1) ∩ dom(aus3)]]$ then $[[ [aus1]β̂⁻ = [aus3]β̂⁻ ]]$,
-            and if $[[β̂⁻ ∊ dom(aus2) ∩ dom(aus4)]]$ then $[[ [aus2]β̂⁻ = [aus4]β̂⁻ ]]$.
-            This way, $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined.
+            The result of the algorithm is $[[(Ξ1 ∪ Ξ2, uQ → uM, aus1 ∪ aus3, aus2 ∪ aus4)]]$.
 
@@
                 \item $[[Γ ; Ξ' ⊢ uQ']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
+                are anti-unification solutions , and
                 \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
@@
                 \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
+                are anti-unification solutions, and
                 \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
@@
         This way, the algorithm proceeds by applying \ruleref{\ottdruleAUArrowLabel} and returns
-        $[[(Ξ1 ∪ Ξ2, uQ → uM''', aus1 & aus3, aus2 & aus4)]]$.
+        $[[(Ξ1 ∪ Ξ2, uQ → uM''', aus1 ∪ aus3, aus2 ∪ aus4)]]$.
 
@@
 
-        Let us show that $[[ausr]] = [[ausr1 & ausr2]]$ is defined, 
-        in other words, for any $[[β̂⁻]] \in [[(Ξ' | uv uQ')]] \cap [[(Ξ' | uv uM'')]]$,
-        $[[ [ausr1] β̂⁻ = [ausr2] β̂⁻ ]]$.It holds because by the induction hypothesis,
-        $[[ [ausri] β̂⁻ ]]$ is uniquely determined by $[[ [aus'1] β̂⁻ ]]$, $[[ [aus'2] β̂⁻ ]]$, and $[[Γ]]$,
-        none of which depends on $i$.
-
-        Let us show that $[[ausr]] = [[ausr1 & ausr2]]$ satisfies the required properties:
+        Let us show that $[[ausr]] = [[ausr1 ∪ ausr2]]$ satisfies the required properties:
         \begin{itemize}
-            \item $[[Γ ; Ξ1 ∪ Ξ2 ⊢ ausr1 & ausr2 : (Ξ' | uv uM')]]$ holds since 
+            \item $[[Γ ; Ξ1 ∪ Ξ2 ⊢ ausr1 ∪ ausr2 : (Ξ' | uv uM')]]$ holds since 
             $[[Ξ' | uv uM']] = [[Ξ' | uv uQ' → uM'']] = [[(Ξ' | uv uQ') ∪ (Ξ' | uv uM'')]]$,
-            $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$ and $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$
-            \ilyam{add a lemma?};
+            $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$ and $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$;
             \item $[[ [ausr] uM' ]] = [[ [ausr] (uQ' → uM'') ]] = [[ [ausr | uv uQ'] uQ' → [ausr | uv uM''] uM'' ]] =
@@
         \end{caseof}
-
 \end{proof}

### 2023-06-06 proofs/tex/alg-neg-subtyping-lemmas.tex
Restructure

--- 2023-05-22 proofs/tex/alg-subtyping-lemmas.tex
@@
-\begin{lemma}[Soundness of Positive Subtyping] \label{lemma:pos-subt-soundness}
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
-    implies $[[us : Θ|uv uP]]$ and 
-    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
-    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
-\end{lemma}
-\begin{proof} 
-    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
-    \begin{caseof}
-    \item \ruleref{\ottdruleAPUVarLabel} then
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
-    $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.
-
-    Notice that $[[â⁺[Δ] ∊ Θ]]$ and $[[Γ ⊢ Θ]]$ 
-    implies $[[Γ = Δ, pnas]]$ for some $[[pnas]]$, hence, the
-    soundness of upgrade (\cref{lemma:upgrade-soundness}) is applicable:
-    \begin{enumerate}
-        \item $[[Δ ⊢ iQ']]$ and
-        \item $[[Γ ⊢ iQ ≥ iP]]$.
-    \end{enumerate}
-
-    Since $[[â⁺[Δ] ∊ Θ|uv â⁺]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(â⁺ :≥ iQ') : Θ | uv â⁺ ]]$.
-
-    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
-    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
-    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
-    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
-    transitively (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
-    as required.
-
-    \item \label{case:pos-subt-soundness:var} \ruleref{\ottdruleAPVarLabel}  
-    then $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
-    Then $[[uv a⁺ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
-    Since $[[uv a⁺ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁺]]$, i.e.
-    $[[ [us'] a⁺ = a⁺]]$. Therefore, $[[Γ ⊢ [us']a⁺ ≥ a⁺]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
-
-    \item \label{case:pos-subt-soundness:shift} 
-     \ruleref{\ottdruleAShiftDLabel} then
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ↓uN ≥ ↓iM ⫤ us]]$.\\
-    Then the next step of the algorithm is the unification of $[[nf(uN)]]$ and $[[nf(iM)]]$.
-    By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
-    it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uN]]$.
-    By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
-    $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uN ]]$ as substitutions.
-    $[[ [us]nf(uN) = nf(iM) ]]$ implies $[[Γ ⊢ [us]nf(uN) ≈ nf(iM)]]$, and then 
-        $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$. \ilyam{add lemmas}
-
-        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$ by 
-        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
-        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
-        $[[Γ ⊢ [us']nf(uN) ≈ [us']uN ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
-        $[[Γ ⊢ nf(iM) ≈ iM ]]$. 
-        This way, we have $[[Γ ⊢ [us']uN ≈ iM]]$.
-        Then by \ruleref{\ottdruleDOneShiftULabel}
-        and congruence of substitution, $[[Γ ⊢ [us']↓uN ≥ ↓iM]]$.
-
-    \item \label{case:pos-subt-soundness:exists}
-       \ruleref{\ottdruleAExistsLabel} then
-        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ∃nas.uP' ≥ ∃nbs.iQ' ⫤ us]]$ s.t. either $[[nas]]$ or $[[nbs]]$ is not empty.\\
-        Then the algorithm creates fresh unification variables $[[â⁻*[Γ,nbs] ]]$, 
-        substitutes the old $[[nas]]$ with them in $[[uP']]$, and makes the recursive call:
-        $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP ≥ iQ ⫤ us']]$, returning as the result
-        $[[us]] = [[us' \ {α̂⁻*}]]$.
-
-        Notice that $[[Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$, $[[Γ,nbs ⊢ iQ']]$, and 
-        $[[Γ,nbs; Θ, â⁻*[Γ, nbs] ⊢ [â⁻*/nas] uP' ]]$, so the induction hypothesis is applicable,
-        that is $[[us' : Θ, â⁻*[Γ, nbs] | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
-        $[[us'2]]$ s.t. $[[Θ, â⁻*[Γ, nbs] ⊢ us'2 ⇒ us']]$.
-
-        Since the domain of $[[us']]$ is $[[uv [â⁻*/nas]uP']]$, the domain of 
-        $[[us]] = [[us' \  {α̂⁻*}]]$ is $[[uv [â⁻*/nas]uP' \ {â⁻*}]] = [[uv ∃nas.uP']]$,
-        this way, $[[us : Θ | uv ∃nas.uP']]$, as required.
-
-        It is left to show that $[[Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
-        Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
-        extending $[[us2]]$ to the domain $[[uv [â⁻*/nas]uP']]$ with the values of $[[us']]$,
-        i.e.  $[[us'2 : Θ | uv [â⁻*/nas]uP']]$, 
-        and
-        \[
-            [[us'2(β̂±)]]  = 
-            \begin{cases}
-               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']] \\
-               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁻*]]
-            \end{cases}
-        \] 
-        , where the application of the unification solution to a variable returns the 
-        corresponding \emph{unification entry}. 
-        Notice that $[[Sub us'2|uv uP']] = [[us2]]$.
-    It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
-    \begin{enumerate}
-        \item if $[[β̂±]] \in [[uv uP']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
-        \item it $[[β̂±]] \in [[â⁻*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
-    \end{enumerate}
-    which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
-    $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$.\\
-    Notice that
-    $
-    \begin{aligned}[t]
-                 [[ [us'2][â⁻*/nas]uP' ]] &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][Sub us'2|uv uP']uP' ]]
-                                          && \text{by substitution properties \ilyam{todo}}\\
-                                          &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uP' ]]
-                                          && \text{since $[[Sub us'2|uv uP']] = [[us2]]$}.
-    \end{aligned}
-    $\\
-    Also notice that the domain of $[[Sub us'2|{â⁻*} ○ â⁻*/nas]]$ is $[[nas]]$,
-    and the range is $[[Γ, nbs]]$, i.e. $[[Γ, nbs ⊢ Sub us'2|{â⁻*} ○ â⁻*/nas : nas]]$, 
-    which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
-    $[[ Γ, nbs ⊢ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uN' ≤ iM' ]]$
-    to obtain $[[ Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ' ]]$, as required.
-    \end{caseof}
-\end{proof}
-
 \begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
@@
         This case is symmetric to \cref{case:pos-subt-soundness:exists} of \cref{lemma:pos-subt-soundness}.
-
-\end{caseof}
-\end{proof}
-
-\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
-\end{theorem}
-\begin{proof}
-\end{proof}
-
-\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
-    there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
-    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
-    and $[[Θ ⊢ us ⇒ us']]$.
-\end{lemma}
-\begin{proof}
-    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
-    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
-    but first consider the base case for the substitution $[[ [us]uP ]]$:
-
-    \begin{caseof}
-        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
-        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
-        $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
-        where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
-        $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$
-        because multiple inversions of \ruleref{\ottdruleDOneExistsLabel} 
-        gives us $[[Γ ⊢ [us]α̂⁺ ≥ iQ]]$ since $[[ {nbs} ∩ fv [us]α̂⁺ = ∅]]$.
-
-        In the algorithm, after multiple applications of \ruleref{\ottdruleAExistsLabel}
-        the type $[[∃nbs.α̂⁺]]$ is reduced to $[[α̂⁺]]$.
-        Next, the algorithm tries to apply
-        \ruleref{\ottdruleAPUVarLabel}
-        and the resulting solution is $[[us']] = [[(α̂⁺ :≥ iQ')]]$ where
-        $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.
-
-        Why does the upgrade procedure terminates?
-        Because $[[iP']]$ satisfies the pre-conditions of the completeness of the upgrade
-        (\cref{lemma:upgrade-completeness})
-        :
-        \begin{itemize}
-            \item $[[Δ ⊢ iP']]$ because $[[iP' = [us]α̂⁺]]$, $[[us : Θ | uv uP]]$ and 
-            $[[α̂⁺[Δ] ∊ Θ | uv uP]]$,
-            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted before
-        \end{itemize}
-        Moreover, completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
-        and further, we strengthen it to $[[Δ ⊢ iP' ≥ iQ']]$
-        (since by the soundness of the upgrade (\cref{lemma:upgrade-soundness}),
-        $[[Δ ⊢ iQ']]$).
-
-        It means that $[[Θ ⊢ (α̂⁺ :≈ iP') ⇒ (α̂⁺ :≥ iQ') ]]$ and 
-        $[[ Θ ⊢ (α̂⁺ :≥ iP') ⇒ (α̂⁺ :≥ iQ') ]]$, which means that in any case, 
-        $[[ Θ ⊢ us ⇒ us']]$.
-
-        \item \label{case:pos-subt-complete-pvar}
-        $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
-        i.e. $[[iP]] = [[ [us]uP ]] = [[ α⁺ ]] = [[iQ]]$.
-        Here the first equality holds because $[[uP]]$ is not a unification variable:
-        this case has been covered by \cref{case:pos-subt-complete-base}.
-        The second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
-
-        Notice that $[[us : Θ | uv α⁺]]$ = $[[us : Θ | ∅]]$ = $[[·]]$.
-
-        The algorithm applies \ruleref{\ottdruleAPVarLabel} and 
-        infers $[[us']] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
-
-        Since $[[Θ ⊢ · ⇒ ·]]$, we have $[[Θ ⊢ us ⇒ us']]$.
-
-        \item \label{case:pos-subt-complete-upshift} $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneShiftDLabel},
-
-        Then $[[ uP ]] = [[ ↓uN ]]$, since the substitution $[[ [us]uP ]]$ must preserve the 
-        top-level constructor of $[[uP]]\neq [[α̂⁺]]$ (the case $[[uP]] = [[α̂⁺]]$ has been covered
-        by \cref{case:pos-subt-complete-base}), and $[[uQ]] = [[ ↓iM ]]$,
-        and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
-
-        Since both types start with $[[↓]]$, 
-        the algorithm tries to apply \ruleref{\ottdruleAShiftDLabel}: 
-        $[[G;Θ ⊨ ↓uN ≥ ↓iM ⫤ us']]$. The premise of this rule is the
-        unification of $[[nf(uN)]]$ and $[[nf(iM)]]$:
-        $[[Γ;Θ ⊨ nf(uN) ≈u nf(iM) ⫤ us']]$. Let us show that the unification successfully 
-        terminates and returns $[[us']] = [[nf(us)]]$.
-
-        To demonstrate that the unification terminates, we apply the completeness 
-        of the unification algorithm (\cref{lemma:unification-completeness}). 
-        In order to do that, we need to provide a unifier of 
-        $[[nf(uN)]]$ and $[[nf(iM)]]$. Thankfully, $[[nf(us)]]$ does it.
-
-        \begin{itemize}
-            \item $[[nf(uN)]]$ and $[[nf(iM)]]$ are normalized 
-            \item $[[Γ ; Θ ⊢ nf(uN)]]$ because $[[Γ ; Θ ⊢ uN]]$ (\cref{todo})
-            \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
-            \item $[[nf(us) : Θ | uv nf(uN)]]$ because $[[us: Θ | uv uN]]$ (\cref{todo})
-            \item $ \begin{aligned}[t]
-                    [[ Γ ⊢ [us]uN ≈ iM ]] &\Rightarrow [[ [us]uN ≈ iM ]]
-                                          && \text {by \cref{lemma:equiv-completeness}}\\
-                                          &\Rightarrow [[ nf([us]uN) = nf(iM) ]]
-                                          && \text {by \cref{lemma:normalization-completeness}}\\
-                                          &\Rightarrow [[ [nf(us)]nf(uN) = nf(iM) ]]
-                                          && \text {by \cref{lemma:norm-subst-distr}}\\
-                    \end{aligned}
-                  $
-        \end{itemize}
-        Then by the completeness of the unification,
-        $[[Γ ; Θ ⊨ nf(uN) ≈u nf(iM) ⫤ nf(us)]]$.
-        This way, the subtyping algorithm terminates and the resulting solution is
-        $[[nf(us)]]$.
-
-        It is left to note that $[[Θ ⊢ us ⇒ nf(us)]]$, by \cref{todo}, since $[[Θ ⊢ us ≈ nf(us)]]$
-
-      \item $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
-      We should only consider the case
-      when the substitution $[[ [us]uP ]]$ results in the existential type 
-      $[[∃nas.iP'']]$ (for $[[iP'']] \neq [[∃]]\dots$) by congruence, 
-      i.e. $[[uP = ∃nas.uP']]$ (for $[[uP']] \neq [[∃]]\dots$) and $[[ [us]uP' = iP'' ]]$.
-      This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
-      (\cref{case:pos-subt-complete-base}), and thus, the substitution $[[us]]$ must
-      preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
-
-      This way, $[[uP]] = [[∃nas.uP']]$, $[[ [us]uP ]] = [[ ∃nas.[us]uP' ]]$ 
-      (assuming $[[nas]]$ does not intersect with the range of $[[us]]$)
-      and $[[iQ]] = [[ ∃nbs.iQ' ]]$, where either $[[nas]]$ or $[[nbs]]$ is not empty.
-
-      By inversion, $[[ Γ ⊢ [σ][us]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ : nas]]$.
-      Since $[[σ]]$ and $[[us]]$ have disjoint domains,
-      and the range of one does not intersect with the domain of the other,
-      they commute, i.e. $[[ Γ, nbs ⊢ [us][σ]uP' ≥ iQ' ]]$
-      (notice that the tree inferring this judgement is 
-      a proper subtree of the tree inferring 
-      $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
-
-      At the next step, 
-      the algorithm creates fresh (disjoint with $[[uv uP']]$) 
-      unification variables $[[â⁻*]]$, replaces $[[nas]]$ with them in $[[ uP' ]]$,
-      and makes the recursive call:
-      $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us1]]$,
-      (where $[[uP0]] = [[ [â⁻*/nas]uP' ]]$),
-      returning $[[us1 \ {â⁻*}]]$ as the result.
-
-       Notice that $[[ [us][σ][nas/â⁻*]uP0 = [us][σ]uP' ]]$,
-       and thus, $[[ Γ, nbs ⊢ [us][σ][nas/â⁻*]uP0 ≥ iQ' ]]$.
-       Let us combine $[[us]]$ and $[[σ  ○ nas/â⁻*]]$ into one 
-       $[[us0]]$---a unification solution for $[[uP0]]$:
-        \[
-            [[us0(β̂±)]]  = 
-            \begin{cases}
-               [[ β̂± :≈ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[â⁻*]] \\
-               [[us(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']]
-            \end{cases}
-       \]
-
-       Let us show that the induction hypothesis is applicable to 
-       $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$, with the meta-context $[[ Θ, â⁻*[Γ, nbs] ]]$
-       \begin{itemize}
-        \item $[[ us0 : (Θ, â⁻*[Γ, nbs])|uv uP0 ]]$. Notice that for every $[[αî⁻]] \in [[â⁻*]]$, 
-            the type corresponding to the entry $[[us0(αî⁻)]]$ 
-            is well-formed in $[[Γ, nbs]]$ since $[[Γ, nbs ⊢ σ : nas ]]$;
-            and for every $[[β̂±]] \in [[uv uP']]$, 
-            the type corresponding to the entry $[[us0(β̂±)]]$ is well-formed in
-            context $[[Θ(β̂±)]]$ since $[[us : Θ | uv uP']]$.
-        \item $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$. This is because
-            $[[ [us0]uP0 ]] = [[ [us][σ][nas/â⁻*]uP0 ]] = [[ [us][σ]uP' ]]$.
-        \item $[[ Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$ since
-            $[[Γ, nbs ⊢ Θ]]$ and $[[ {Γ, nbs} ⊆ {Γ, nbs} ]]$.
-        \item $[[Γ, nbs ; Θ, â⁻*[Γ, nbs] ⊢ uP0 ]]$
-       \end{itemize}
-
-       This way, we apply the induction hypothesis to $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$ and 
-       infer that there exists\\
-        $[[us0' :  (Θ, â⁻*[Γ, nbs]) | uv uP0]]$ such that
-       $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us0']]$,
-       and $[[Θ, â⁻*[G, nbs] ⊢ us0' ⇒ us0]]$.
-
-       This way, the algorithm terminates with the result $[[us0' \ {â⁻*}]]$
-       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁻*} ]]$. Notice that
-       $[[us0' \ {â⁻*} : Θ | uv uP' ]]$. To show $[[ Θ ⊢ us ⇒ us0' \
-       {â⁻*} ]]$, it suffices to consider a unification variable $[[β̂±]] \in 
-       [[uv uP']]$ and show that $[[Γ ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
-       reflexivity of weakening (\cref{lemma:entry-weakening-preorder}) since
-       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
-
-      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
-      \cref{case:subt-complete-nvar}.
-      Notice that $[[ [us]uP = β⁺ ]]$ 
-      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
-      (\cref{case:subt-complete-base}).
 

### 2023-06-06 proofs/tex/alg-pos-subtyping-lemmas.tex
Restructure

--- 2023-05-22 proofs/tex/alg-subtyping-lemmas.tex
@@
-\begin{lemma}[Soundness of Positive Subtyping] \label{lemma:pos-subt-soundness}
+\begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}
     If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
@@
     \end{caseof}
-\end{proof}
-
-\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
-        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
-        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
-        implies $[[us : Θ|uv uN]]$ and 
-        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
-        $[[ Γ ⊢ [us']uN ≤ iM ]]$
-\end{lemma}
-\begin{proof}
-    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$.
-    Let us consider the last rule to infer this judgment. 
-    \begin{caseof}
-        \item \ruleref{\ottdruleAArrowLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape
-        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
-        On the next step, the the algorithm makes two recursive calls:
-        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
-        By the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}) and induction hypothesis, respectively 
-        \begin{enumerate}
-            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
-            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
-        \end{enumerate}
-
-        Then the algorithm merges two unification solutions $[[us1]]$ and $[[us2]]$.
-        By \cref{lemma:merge-soundness}, since $[[uv uP ∪ uv uN' = uv (uP → uN')]]$, 
-        we have $[[us1 & us2 : Θ | uv (uP → uN')]]$, and also
-        $[[Θ ⊢ us1 & us2 ⇒ us1]]$ and $[[Θ ⊢ us1 & us2 ⇒ us2]]$.
-        By the transitivity of solution weakening (\cref{lemma:weakening-transitivity}),
-         $[[Θ ⊢ us' ⇒ us1 & us2]]$ implies $[[Θ ⊢ us' ⇒ us1]]$ and $[[Θ ⊢ us' ⇒ us2]]$.
-
-% \[\begin{tikzcd}[arrows=Rightarrow]
-% 	& {[[us']]} \\
-% 	{[[us1]]} & {[[us1 & us2]]} & {[[us2]]}
-% 	\arrow[from=2-2, to=2-3]
-% 	\arrow[from=2-2, to=2-1]
-% 	\arrow[from=1-2, to=2-2]
-% \end{tikzcd}\]
-
-        The application of the induction hypothesis gives us 
-        $[[Γ ⊢ [us']uP ≥ iQ ]]$ and $[[ Γ ⊢ [us']uN' ≤ iM' ]]$.
-        Finally, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [us'](uP → uN') ≤ iQ → iM']]$.
-
-        \item \ruleref{\ottdruleANVarLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
-        This case is symmetric to \cref{case:pos-subt-soundness:var} of \cref{lemma:pos-subt-soundness}.
-
-        \item \ruleref{\ottdruleAShiftULabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
-        This case is symmetric to \cref{case:pos-subt-soundness:shift} of \cref{lemma:pos-subt-soundness}.
-
-        \item \ruleref{\ottdruleAForallLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
-        This case is symmetric to \cref{case:pos-subt-soundness:exists} of \cref{lemma:pos-subt-soundness}.
-
-\end{caseof}
-\end{proof}
-
-\begin{theorem}[Soundness of Subtyping Algorithm] \hfill
-\end{theorem}
-\begin{proof}
 \end{proof}
@@
 \end{proof}
-
-\begin{lemma}[Completeness of the Negative Subtyping]
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
-    $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
-    and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
-    and $[[Θ ⊢ us ⇒ us']]$.
-\end{lemma}
-\begin{proof}
-    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-    \begin{caseof}
-        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
-        Then $[[ uN ]] = [[ ↑uP ]]$, since
-        the substitution $[[ [us]uN ]]$ must preserve the 
-        top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
-        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
-        The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
-         \cref{lemma:pos-subt-completeness}: notice that the algorithm does not make a recursive call, 
-        and the difference in the induction statement for the positive and 
-        the negative case here does not matter.
-
-        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneArrowLabel}, 
-        i.e. $[[ [us]uN ]] = [[ [us]uP → [us]uN' ]]$ and $[[iM]] = [[iQ → iM']]$, 
-        and by inversion, $[[ Γ ⊢ [us]uP ≥ iQ ]]$ and $[[ Γ ⊢ [us]uN' ≤ iM' ]]$.
-
-        Let us consider restrictions of $[[us]]$ to 
-        the set of unification variables in $[[uP]]$ and $[[uN']]$:
-        $[[us | uv uP : Θ | uv uP]]$ and $[[us | uv uN' : Θ | uv uN']]$.
-        Notice that 
-        $[[ [us]uP = [us | uv uP]uP ]]$ and 
-        $[[ [us]uN' = [us | uv uN']uN' ]]$.
-
-       Let us apply the induction hypothesis to
-       $[[ Γ ⊢ [us | uv uP]uP ≥ iQ ]]$ and
-       $[[ Γ ⊢ [us | uv uN']uN' ≤ iM' ]]$ (notice that
-       since $[[uv uN' ⊆ uv uN]]$, there are no negative unification variables in $[[uN']]$)
-       to obtain $[[us'1]]$ and $[[us'2]]$ such that
-       \begin{enumerate}
-        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'1]]$ and $[[Θ ⊢ us | uv uP ⇒ us'1]]$
-        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'2]]$ and $[[Θ ⊢ us | uv uN' ⇒ us'2]]$ 
-       \end{enumerate}
-
-       This way, the algorithm applies \ruleref{\ottdruleAArrowLabel}
-       and terminates returning $[[us'1 & us'2]]$ as the result.
-
-       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
-       Since $[[us | uv uP ⊆ us]]$, 
-       by \cref{lemma:weakening-monotonicity}, 
-       $[[Θ ⊢ us ⇒ us | uv uP]]$,
-       and then by transitivity (\cref{lemma:weakening-transitivity}), 
-       $[[Θ ⊢ us ⇒ us'1]]$.
-       Analogously, $[[Θ ⊢ us ⇒ us'2]]$.
-       Then since by \cref{lemma:merge-completeness}, 
-       $[[us'1 & us'2]]$ is the weakest restriction  
-       implying both $[[us'1]]$ and $[[us'2]]$,
-       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.
-
-       \item \label{case:subt-complete-forall}
-       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel}.
-       Since $[[uN]]$ does not contain negative unification variables,
-       $[[uN]]$ must be of the form $[[∀pas.uN']]$,
-       such that $[[ [us]uN = ∀pas.[us]uN' ]]$ and $[[ [us]uN']] \neq [[∀]]\dots$
-       (assuming $[[pas]]$ does not intersect with the range of $[[us]]$).
-       Also, $[[iM]] = [[∀pbs.iM']]$ and either $[[pas]]$ or $[[pbs]]$ is non-empty.
-
-       The rest of the proof is symmetric to \cref{case:pos-subt-complete-exists} of
-       \cref{lemma:pos-subt-completeness}.
-       To apply the induction hypothesis, we need to show additionally that
-       there are no negative unification variables in $[[uN0]] = [[ [â⁺*/pas]uN' ]]$.
-       This is because $[[ uv uN0 ⊆ uv uN ∪ {â⁺*} ]]$, and $[[uN]]$ is free of negative
-       unification variables by assumption.
-
-       \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}.\\
-       Then $[[iN]] = [[ [us]uN ]] = [[ α⁻ ]] = [[iM]]$. 
-       Here the first equality holds because $[[uN]]$ is not a unification variable:
-       by assumption, $[[uN]]$ is free of negative unification variables.
-       The second and the third equations hold because \ruleref{\ottdruleDOneNVarLabel}
-       was applied.
-
-       The rest of the proof is symmetric to \cref{case:pos-subt-complete-pvar} of
-       \cref{lemma:pos-subt-completeness}.
-
-    \end{caseof}
-\end{proof}

### 2023-06-09 proofs/tex/alg-pos-subtyping-lemmas.tex
Add algorithmic typing

@@
     By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
-    it returns an equivalence-only solution $[[us]]$ such that $[[us : Ord uv uN]]$.
+    it returns an equivalence-only solution $[[us]]$ such that $[[us : Θ | uv uN]]$.
     By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 

### 2023-06-11 proofs/tex/decl-subtyping-lemmas.tex
Subtyping: soundness is done, completeness WIP
Add algorithmic typing
WIP: misc subtyping lemmas

@@
 \end{proof}
+
+\begin{lemma}[Reflexivity of subtyping] \label{lemma:subtyping-reflexivity}
+  Assuming all the types are well-formed in $[[Γ]]$,
+  \begin{itemize}
+    \item [$-$] $[[Γ ⊢ iN ≤ iN]]$
+    \item [$+$] $[[Γ ⊢ iP ≥ iP]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Let us prove it by the size of $[[iN]]$ and mutually, $[[iP]]$.
+  \begin{caseof}
+    \item $[[iN]] = [[α⁻]]$\\
+      Then $[[Γ ⊢ α⁻ ≤ α⁻]]$ is inferred immediately by \ruleref{\ottdruleDOneNVarLabel}.
+    \item $[[iN]] = [[∀pas.iN']]$ where $[[pas]]$ is not empty\\
+      First, we rename $[[pas]]$ to fresh $[[pbs]]$ in $[[∀pas.iN']]$ to avoid
+      name clashes: $[[∀pas.iN']] = [[∀pbs.[pas/pbs]iN']]$.
+      Then to infer $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.[pas/pbs]iN']]$ we can apply 
+      \ruleref{\ottdruleDOneForallLabel}, instantiating $[[pas]]$ with $[[pbs]]$:
+      \begin{itemize}
+        \item $[[fv iN ∩ {pbs} = ∅ ]]$ by choice of $[[pbs]]$,
+        \item $[[G, pbs ⊢ pbi]]$,
+        \item $[[G, pbs ⊢ [pbs/pas] iN' ≤ [pbs/pas] iN']]$ by the induction hypothesis,
+        since the size of $[[ [pbs/pas]iN' ]]$ is equal to the size of $[[iN']]$,
+        which is smaller than the size of $[[iN]] = [[∀pas.iN']]$.
+      \end{itemize}
+    \item $[[iN]] = [[iP → iM]]$\\
+      Then $[[Γ ⊢ iP → iM ≤ iP → iM]]$ is inferred by \ruleref{\ottdruleDOneArrowLabel},
+      since $[[Γ ⊢ iP ≥ iP]]$ and $[[Γ ⊢ iM ≤ iM]]$ hold the induction hypothesis. 
+    \item $[[iN]] = [[↑iP]]$\\
+      Then $[[Γ ⊢ ↑iP ≤ ↑iP]]$ is inferred by \ruleref{\ottdruleDOneShiftULabel},
+      since $[[Γ ⊢ iP ≥ iP]]$ holds by the induction hypothesis.
+    \item The positive cases are symmetric to the negative ones.
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Substitution preserves subtyipng]
+  \label{lemma:subst-pres-subt}
+  Assuming that all mentioned types are well-formed in $[[Γ]]$,
+  and $[[Γ' ⊢ σ : Γ]]$, where $[[Γ']]$ is disjoint from $[[Γ]]$,
+  \begin{itemize}
+    \item $-$ If $[[Γ ⊢ iN ≤ iM]]$ then $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$
+    \item $+$ If $[[Γ ⊢ iP ≥ iQ]]$ then $[[Γ' ⊢ [σ]iP ≥ [σ]iQ]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on the size of the derivation of $[[Γ ⊢ iN ≤ iM]]$
+  and mutually, $[[Γ ⊢ iP ≥ iQ]]$. Let us consider the last rule 
+  used in the derivation:
+  \begin{caseof}
+    \item \ruleref{\ottdruleDOneNVarLabel}. Then by inversion, 
+      $[[iN = α⁻]]$ and $[[iM = α⁻]]$. By reflexivity of subtyping
+      (\cref{lemma:subtyping-reflexivity}),
+      we have $[[Γ' ⊢ [σ]α⁻ ≤ [σ]α⁻]]$, i.e. $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$,
+      as required.
+    \item  \ruleref{\ottdruleDOneForallLabel}. Then by inversion,
+      $[[iN = ∀pas.iN']]$, $[[iM = ∀pbs.iM']]$, where $[[pas]]$ or $[[pbs]]$ is not empty.
+      Moreover, $[[Γ,pbs ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ, pbs ⊢ iPs]]$, and 
+      $[[fv iN ∩ {pbs} = ∅ ]]$.
+
+      Notice that since the derivation of $[[Γ,pbs ⊢ [iPs/pas]iN' ≤ iM']]$ is
+      a subderivation of the derivation of $[[Γ ⊢ iN ≤ iM]]$, its size is smaller, 
+      and hence, the induction hypothesis applies:
+      $[[Γ', pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$.
+
+      First, let us assume that $[[{pas} ∩ {Γ'} = ∅]]$ and $[[{pbs} ∩ {Γ'} = ∅]]$
+      (otherwise, we rename $[[pas]]$ and $[[pbs]]$ to fresh $[[pas']]$ and $[[pbs']]$).
+      Then $[[ [σ]∀pas.iN' ]] = [[ ∀pas.[σ]iN' ]]$ and $[[ [σ]∀pbs.iM' ]] = [[ ∀pbs.[σ]iM' ]]$, 
+      which means that the required $[[Γ' ⊢ [σ]∀pas.iN' ≤ [σ]∀pbs.iM']]$ is rewritten as
+      $[[Γ' ⊢ ∀pas.[σ]iN' ≤ ∀pbs.[σ]iM']]$.
+
+      To infer it, we apply \ruleref{\ottdruleDOneForallLabel}, 
+      instantiating $[[pai]]$ with $[[ [σ]iPi ]]$:
+      \begin{itemize}
+        \item $[[fv iN ∩ {pbs} = ∅ ]]$ as noted before, from the inversion;
+        \item $[[Γ', pbs ⊢ [σ]iPi]]$, by \cref{lemma:wf-subst} since from the inversion,
+        $[[Γ, pbs ⊢ iPi]]$;
+        \item $[[Γ, pbs ⊢ [ [σ]iPs/pas ][σ]iN' ≤ [σ]iM']]$ holds because 
+        $[[ [ [σ]iPs/pas ][σ]iN' ]] = [[ [σ][iPs/pas]iN ]]$ (since $[[{pas} ∩ {Γ} = ∅]]$), 
+        and $[[Γ', pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$ holds by the induction hypothesis.
+      \end{itemize}
+    \item \ruleref{\ottdruleDOneArrowLabel}. Then by inversion,
+      $[[iN = iP → iN1]]$, $[[iM = iQ → iM1]]$, $[[Γ ⊢ iP ≥ iQ]]$, and $[[Γ ⊢ iN1 ≤ iM1]]$.
+      And by the induction hypothesis, $[[Γ' ⊢ [σ]iP ≥ [σ]iQ]]$ and $[[Γ' ⊢ [σ]iN1 ≤ [σ]iM1]]$.
+      Then $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$, i.e. $[[Γ' ⊢ [σ]iP → [σ]iN1 ≤ [σ]iQ → [σ]iM1]]$,
+      is inferred by \ruleref{\ottdruleDOneArrowLabel}.
+    \item \ruleref{\ottdruleDOneShiftULabel}. Then by inversion,
+      $[[iN = ↑iP]]$, $[[iM = ↑iQ]]$, and $[[Γ ⊢ iP ≈ iQ]]$,
+      which by inversion means that $[[Γ ⊢ iP ≥ iQ]]$ and $[[Γ ⊢ iQ ≥ iP]]$.
+      Then the induction hypothesis applies, and we have $[[Γ' ⊢ [σ]iP ≥ [σ]iQ]]$
+      and $[[Γ' ⊢ [σ]iQ ≥ [σ]iP]]$. 
+      Then by sequential application of \ruleref{\ottdruleDOneNDefLabel} 
+      and \ruleref{\ottdruleDOneShiftULabel} to these judgments,
+      we have $[[Γ' ⊢ ↑[σ]iP ≤ ↑[σ]iQ]]$, i.e.
+      $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$, as required.
+    \item The positive cases are proved symmetrically.
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Strong transitivity of subtyping] \label{lemma:subtyping-transitivity}
+  Assuming all the types are well-formed in $[[Γ]]$,
+  \begin{itemize}
+    \item $-$ if $[[Γ ⊢ iN ≤ iM1]]$, $[[Γ ⊢ iM2 ≤ iK]]$, and for 
+    $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iM1 = [σ]iM2 ]]$ then $[[Γ' ⊢ [σ]iN ≤ [σ]iK]]$
+    \item $+$ if $[[Γ ⊢ iP ≥ iQ1]]$, $[[Γ ⊢ iQ2 ≥ iR]]$, and for
+    $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iQ1 = [σ]iQ2 ]]$ then $[[Γ' ⊢ [σ]iP ≥ [σ]iR]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on $\depth{[[Γ ⊢ iN ≤ iM1]]} + \depth{[[Γ ⊢ iM2 ≤ iK]]}$ and mutually, 
+  on $\depth{[[Γ ⊢ iP ≥ iQ1]]} + \depth{[[Γ ⊢ iQ2 ≥ iR]]}$.
+
+  \begin{caseof}
+    \item Firs, let us consider the case when the last rule applied to infer 
+    $[[Γ ⊢ iN ≤ iM1]]$ is \ruleref{\ottdruleDOneNVarLabel}.
+    Notice that this case covers the base of the induction: 
+    the sum of the depths is minimal when both derivations are inferred by
+    the non-recursive rules (i.e. \ruleref{\ottdruleDOneNVarLabel}).
+
+    By inverting the rule, $[[iN = α⁻]]$ and $[[iM1 = α⁻]]$
+    Then $[[ [σ]iN ]] = [[ [σ]α⁻ ]] = [[ [σ]iM1 ]] = [[ [σ]iM2 ]]$. 
+    And $[[Γ' ⊢ [σ]iM2 ≤ [σ]iK]]$ by \cref{}
+
+  \end{caseof}
+
+\end{proof}
+
+\begin{corollary}[Transitivity of subtyping] \label{corollary:subtyping-transitivity}
+  Assuming the types are well-formed in $[[Γ]]$,
+  \begin{itemize}
+    \item[$-$] if $[[Γ ⊢ iN1 ≤ iN2]]$ and $[[Γ ⊢ iN2 ≤ iN3]]$ then $[[Γ ⊢ iN1 ≤ iN3]]$,
+    \item[$+$] if $[[Γ ⊢ iP1 ≥ iP2]]$ and $[[Γ ⊢ iP2 ≥ iP3]]$ then $[[Γ ⊢ iP1 ≥ iP3]]$.
+  \end{itemize}
+\end{corollary}
+
+\begin{corollary}[Transitivity of equivalence] \label{corollary:equivalence-transitivity}
+  Assuming the types are well-formed in $[[Γ]]$,
+  \begin{itemize}
+    \item[$-$] if $[[Γ ⊢ iN1 ≈ iN2]]$ and $[[Γ ⊢ iN2 ≈ iN3]]$ then $[[Γ ⊢ iN1 ≈ iN3]]$,
+    \item[$+$] if $[[Γ ⊢ iP1 ≈ iP2]]$ and $[[Γ ⊢ iP2 ≈ iP3]]$ then $[[Γ ⊢ iP1 ≈ iP3]]$.
+  \end{itemize}
+\end{corollary}

### 2023-06-29 proofs/main.tex
Add soundness and completeness of unification Unification solution merge needs to be updated
WIP: finish merge lemmas
Restructure
Add term grammar
Add declarative typing
Add algorithmic typing
WIP: misc subtyping lemmas
wip

--- 2023-04-27 proofs/main.tex
@@
 % \makeatother
+
+\newcommand{\depth}[1]{\mathsf{depth}(#1)}
 
@@
 
+% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
+\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
+\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
+\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
+\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
+\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}
+
+% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
+\renewcommand{\ottdruleUNVarName}[0]{(Var$^{-[[≈u]]}$)}
+\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
+\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
+\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
+\renewcommand{\ottdruleUNUVarName}[0]{(UVar$^{-[[≈u]]}$)}
+
+% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
+\renewcommand{\ottdruleUPVarName}[0]{(Var$^{+[[≈u]]}$)}
+\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
+\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
+\renewcommand{\ottdruleUPUVarName}[0]{(UVar$^{+[[≈u]]}$)}
+
+% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
+\renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
+\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
+\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
+
+% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
+\renewcommand{\ottdruleAUNVarName}[0]{(Var$^{-[[≈au]]}$)}
+\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
+\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
+\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
+\renewcommand{\ottdruleAUAUName}[0]{(AU$^{-}$)}
+
 \begin{document}
@@
   \ottN\ottinterrule
+  \ottv\ottinterrule
+  \ottc\ottinterrule
 }
+
+\subsection{Declarative Typing}
+\ottdefnsDT
 
@@
 }
-\subsection{Declarative Subtyping}
+\subsection{Declarative Multiquantified Subtyping}
 \ottdefnsDOne
@@
 \subsection{Normalization}
+
 \subsubsection{Ordering}
@@
 \ottdefnsNrm
+
+\subsection{Singularity}
+\ottdefnsSING
 
@@
 
-\subsection{Unification Solution Merge}
+\subsection{Constraint Merge}
 
-Unification solution is represented by a list of unification solution entries.
-Each entry restricts an unification variable in two possible ways: either stating
-that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua :≈
-iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥
-iP]]$).
+Unification and subtyping constraints is are by a list of constraint entries.
+Each entry restricts an unification variable in two possible ways: either
+stating that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua
+:≈ iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥ iP]]$).
 
@@
 \begin{definition} \hfill \\
-\ottdefnSME\\
+\ottdefnSCME\\
 \end{definition}
-Notice that in case of equivalence, the assigned types
-must be equal (i.e. alpha-equivalent) to be merged. This is because
-the unification algorithm assumes that every type is normalized,
-and hence, equivalence is alpha-equivalence 
-(\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
+% Notice that in case of equivalence, the assigned types
+% must be equal (i.e. alpha-equivalent) to be merged. This is because
+% the unification algorithm assumes that every type is normalized,
+% and hence, equivalence is alpha-equivalence 
+% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
 
-To merge two unification solution, we merge each pair of
+To merge two constraints, we merge each pair of
 matching entries, and unite the results.
-
+\ilyam{add contexts}
 \begin{definition}
-  $\begin{aligned}[t]
-  [[us1 & us2]] &= \{ [[usEntry1 & usEntry2]] \mid [[usEntry1]] \in [[us1]],
-  [[usEntry2]]  \in [[us2]], \text{s.t. } [[usEntry1]] \text{ matches with }
-                                   [[usEntry2]] \}\\
+  $\begin{aligned}[t] [[SC1 & SC2]] &= 
+    \{ [[scE1 & scE2]] \mid [[scE1]] \in [[SC1]], [[scE2]]  \in [[SC2]], \text{s.t. } [[scE1]] \text{ matches with } [[scE2]] \}\\
     &\cup 
-                                     \{ [[usEntry1]] \mid [[usEntry1]] \in
-                                     [[us1]], \text{ s.t. }
-                                     \forall [[usEntry2]]  \in [[us2]],
-                                     [[usEntry1]] \text{ does not match with }
-                                                     [[usEntry2]] \}\\
+    \{ [[scE1]] \mid [[scE1]] \in [[SC1]], \text{ s.t. } \forall [[scE2]]  \in [[SC2]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
     &\cup 
-          \{ [[usEntry2]] \mid [[usEntry2]] \in
-          [[us2]], \text{ s.t. }
-          \forall [[usEntry1]]  \in [[us1]],
-          [[usEntry1]] \text{ does not match with }
-          [[usEntry2]] \}\\
+    \{ [[scE2]] \mid [[scE2]] \in [[SC2]], \text{ s.t. } \forall [[scE1]]  \in [[SC1]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
    \end{aligned}$
 \end{definition}
+
+\subsection{Constraint Satisfaction}
+\ottdefnsSATSCE
 
@@
 
+\subsection{Typing}
+\ottdefnsAT
+
 \section{Proofs}
+
+\subsection{Substitution}
+\input{../_build/_gen/tex/subst-lemmas.tex}
 
@@
 \input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Substitution}
-\input{../_build/_gen/tex/subst-lemmas.tex}
 
@@
 
-\subsection{Upgrade}
-\input{../_build/_gen/tex/upgrade-lemmas.tex}
-
-\subsection{Upper Bounds}
-
-\input{../_build/_gen/tex/lub-lemmas.tex}
+\subsection{Unification Constraint Merge}
+\input{../_build/_gen/tex/unif-constraint-merge-lemmas.tex}
 
@@
 
-\subsection{Solution Merge}
-\input{../_build/_gen/tex/solution-merge-lemmas.tex}
+\subsection{Upper Bounds}
+\input{../_build/_gen/tex/lub-lemmas.tex}
 
-\subsection{Subtyping Algorithm}
-\input{../_build/_gen/tex/alg-subtyping-lemmas.tex}
+\subsection{Upgrade}
+\input{../_build/_gen/tex/upgrade-lemmas.tex}
+
+\subsection{Positive Subtyping}
+\input{../_build/_gen/tex/alg-pos-subtyping-lemmas.tex}
+
+\subsection{Subtyping Constraint Merge}
+\input{../_build/_gen/tex/constraint-merge-lemmas.tex}
+
+\subsection{Constraint Satisfaction}
+\input{../_build/_gen/tex/constraint-sat-lemmas.tex}
+
+\subsection{Negative Subtyping}
+\input{../_build/_gen/tex/alg-neg-subtyping-lemmas.tex}
+
+\subsection{Singularity}
+\input{../_build/_gen/tex/singularity-lemmas.tex}
+
+\subsection{Typing}
+\input{../_build/_gen/tex/alg-typing-lemmas.tex}
 

### 2023-06-29 proofs/tex/alg-neg-subtyping-lemmas.tex
wip

@@
 \begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
-        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, and $[[Γ ; Θ ⊢ uN]]$ then\\ 
-        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$
-        implies $[[us : Θ|uv uN]]$ and 
-        for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
-        $[[ Γ ⊢ [us']uN ≤ iM ]]$
+        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$ and 
+        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
+        $[[Θ ⊢ SC]]$ and 
+        for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : SC]]$,
+        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
 \end{lemma}
 \begin{proof}
-    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$.
+    We prove it by induction on $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
+
+    Suppose that $[[uσ]]$ is normalized and $[[Θ ⊢ uσ : SC]]$,
     Let us consider the last rule to infer this judgment. 
     \begin{caseof}
-        \item \ruleref{\ottdruleAArrowLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape
-        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ us]]$\\
+        \item \ruleref{\ottdruleAArrowLabel}. Then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
+        $[[G;Θ ⊨ uP → uN' ≤ iQ → iM' ⫤ SC]]$\\
         On the next step, the the algorithm makes two recursive calls:
-        $[[G;Θ ⊨ uP ≥ iQ ⫤ us1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ us2]]$.
-        By the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}) and induction hypothesis, respectively 
-        \begin{enumerate}
-            \item $[[us1 : Θ | uv uP]]$ and $[[ Γ ⊢ [us1']uP ≥ iQ ]]$ for any $[[us1']]$ s.t. $[[Θ ⊢ us1' ⇒ us1]]$
-            \item $[[us2 : Θ | uv uN']]$ and $[[ Γ ⊢ [us2']uN' ≤ iM' ]]$ for any $[[us2']]$ s.t. $[[Θ ⊢ us2' ⇒ us2]]$
-        \end{enumerate}
+        $[[G;Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[G;Θ ⊨ uN' ≤ iM' ⫤ SC2]]$
+        and returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.
 
-        Then the algorithm merges two unification solutions $[[us1]]$ and $[[us2]]$.
-        By \cref{lemma:merge-soundness}, since $[[uv uP ∪ uv uN' = uv (uP → uN')]]$, 
-        we have $[[us1 & us2 : Θ | uv (uP → uN')]]$, and also
-        $[[Θ ⊢ us1 & us2 ⇒ us1]]$ and $[[Θ ⊢ us1 & us2 ⇒ us2]]$.
-        By the transitivity of solution weakening (\cref{lemma:weakening-transitivity}),
-         $[[Θ ⊢ us' ⇒ us1 & us2]]$ implies $[[Θ ⊢ us' ⇒ us1]]$ and $[[Θ ⊢ us' ⇒ us2]]$.
+        By the soundness of constraint merge (\cref{lemma:merge-soundness}),
+        $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
+        and by the induction hypothesis, $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
+        This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [uσ](uP → uN') ≤ iQ → iM']]$.
 
-% \[\begin{tikzcd}[arrows=Rightarrow]
-% 	& {[[us']]} \\
-% 	{[[us1]]} & {[[us1 & us2]]} & {[[us2]]}
-% 	\arrow[from=2-2, to=2-3]
-% 	\arrow[from=2-2, to=2-1]
-% 	\arrow[from=1-2, to=2-2]
-% \end{tikzcd}\]
-
-        The application of the induction hypothesis gives us 
-        $[[Γ ⊢ [us']uP ≥ iQ ]]$ and $[[ Γ ⊢ [us']uN' ≤ iM' ]]$.
-        Finally, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ ⊢ [us'](uP → uN') ≤ iQ → iM']]$.
-
-        \item \ruleref{\ottdruleANVarLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
+        \item \ruleref{\ottdruleANVarLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape $[[G;Θ ⊨ a⁻ ≤ a⁻ ⫤ ·]]$\\
         This case is symmetric to \cref{case:pos-subt-soundness:var} of \cref{lemma:pos-subt-soundness}.
 
-        \item \ruleref{\ottdruleAShiftULabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ us]]$\\
+        \item \ruleref{\ottdruleAShiftULabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape $[[G;Θ ⊨ ↑uP ≤ ↑iQ ⫤ SC]]$\\
         This case is symmetric to \cref{case:pos-subt-soundness:shift} of \cref{lemma:pos-subt-soundness}.
 
-        \item \ruleref{\ottdruleAForallLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us]]$ has shape $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ us]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
+        \item \ruleref{\ottdruleAForallLabel}, and then $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ has shape
+         $[[G;Θ ⊨ ∀pas.uN' ≤ ∀pbs.iM' ⫤ SC]]$ s.t. either $[[pas]]$ or $[[pbs]]$ is not empty\\
         This case is symmetric to \cref{case:pos-subt-soundness:exists} of \cref{lemma:pos-subt-soundness}.
@@
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
-    $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$)
-    and there exists $[[us : Θ | uv uN]]$ such that $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uN ≤ iM ⫤ us']]$
-    and $[[Θ ⊢ us ⇒ us']]$.
+    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
+    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
+    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
 \end{lemma}
 \begin{proof}
-    We prove it by induction on $[[ Γ ⊢ [us]uN ≤ iM ]]$.
-    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [us]uN ≤ iM ]]$.
+    We prove it by induction on $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
+    Let us consider the last rule used in the derivation of $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
     \begin{caseof}
-        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
+        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneShiftULabel}\\
         Then $[[ uN ]] = [[ ↑uP ]]$, since
-        the substitution $[[ [us]uN ]]$ must preserve the 
+        the substitution $[[ [uσ]uN ]]$ must preserve the 
         top-level constructor of $[[uN]] \neq [[α̂⁻]]$ (since by assumption, $[[α̂⁻]] \notin [[uv uN]]$), 
-        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
+        and $[[uQ]] = [[ ↓iM ]]$, and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
         The rest of the proof is symmetric to \cref{case:pos-subt-complete-upshift} of
@@
 
-        \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneArrowLabel}, 
-        i.e. $[[ [us]uN ]] = [[ [us]uP → [us]uN' ]]$ and $[[iM]] = [[iQ → iM']]$, 
-        and by inversion, $[[ Γ ⊢ [us]uP ≥ iQ ]]$ and $[[ Γ ⊢ [us]uN' ≤ iM' ]]$.
+        \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneArrowLabel}, 
+        i.e. $[[ [uσ]uN ]] = [[ [uσ]uP → [uσ]uN' ]]$ and $[[iM]] = [[iQ → iM']]$, 
+        and by inversion, $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ and $[[ Γ ⊢ [uσ]uN' ≤ iM' ]]$.
 
-        Let us consider restrictions of $[[us]]$ to 
-        the set of unification variables in $[[uP]]$ and $[[uN']]$:
-        $[[us | uv uP : Θ | uv uP]]$ and $[[us | uv uN' : Θ | uv uN']]$.
-        Notice that 
-        $[[ [us]uP = [us | uv uP]uP ]]$ and 
-        $[[ [us]uN' = [us | uv uN']uN' ]]$.
+        The algorithm makes two recursive calls: $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ and $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$,
+        and then returns $[[Θ ⊢ SC1 & SC2 = SC]]$ as the result.
+        Let us show that these recursive calls are successful and the returning constraints 
+        are fulfilled by $[[uσ]]$.
 
-       Let us apply the induction hypothesis to
-       $[[ Γ ⊢ [us | uv uP]uP ≥ iQ ]]$ and
-       $[[ Γ ⊢ [us | uv uN']uN' ≤ iM' ]]$ (notice that
-       since $[[uv uN' ⊆ uv uN]]$, there are no negative unification variables in $[[uN']]$)
-       to obtain $[[us'1]]$ and $[[us'2]]$ such that
-       \begin{enumerate}
-        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'1]]$ and $[[Θ ⊢ us | uv uP ⇒ us'1]]$
-        \item $[[Γ; Θ ⊨ uP ≥ iQ ⫤ us'2]]$ and $[[Θ ⊢ us | uv uN' ⇒ us'2]]$ 
-       \end{enumerate}
+        Notice that from the inversion of $[[Γ ⊢ iM]]$, we have: $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$;
+        from the inversion of $[[Γ ; Θ ⊢ uN]]$, we have: $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$;
+        and since $[[uN]]$ does not contain negative unification variables,
+        $[[uN']]$ does not contain negative unification variables either.
 
-       This way, the algorithm applies \ruleref{\ottdruleAArrowLabel}
-       and terminates returning $[[us'1 & us'2]]$ as the result.
+        This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
+        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
+        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ uσ : SC1]]$.
 
-       It is left to show that $[[Θ ⊢ us ⇒ us'1 & us'2]]$.
-       Since $[[us | uv uP ⊆ us]]$, 
-       by \cref{lemma:weakening-monotonicity}, 
-       $[[Θ ⊢ us ⇒ us | uv uP]]$,
-       and then by transitivity (\cref{lemma:weakening-transitivity}), 
-       $[[Θ ⊢ us ⇒ us'1]]$.
-       Analogously, $[[Θ ⊢ us ⇒ us'2]]$.
-       Then since by \cref{lemma:merge-completeness}, 
-       $[[us'1 & us'2]]$ is the weakest restriction  
-       implying both $[[us'1]]$ and $[[us'2]]$,
-       we have $[[Θ ⊢ us ⇒ us'1 & us'2]]$, as required.
+        Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
+        satisfies the required properties.
+        To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
+        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[Θ ⊢ uσ : SC]]$, as required.
 
        \item \label{case:subt-complete-forall}
-       $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel}.
+       $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneForallLabel}.
        Since $[[uN]]$ does not contain negative unification variables,
        $[[uN]]$ must be of the form $[[∀pas.uN']]$,
-       such that $[[ [us]uN = ∀pas.[us]uN' ]]$ and $[[ [us]uN']] \neq [[∀]]\dots$
-       (assuming $[[pas]]$ does not intersect with the range of $[[us]]$).
+       such that $[[ [uσ]uN = ∀pas.[uσ]uN' ]]$ and $[[ [uσ]uN']] \neq [[∀]]\dots$
+       (assuming $[[pas]]$ does not intersect with the range of $[[uσ]]$).
        Also, $[[iM]] = [[∀pbs.iM']]$ and either $[[pas]]$ or $[[pbs]]$ is non-empty.
@@
 
-       \item $[[ Γ ⊢ [us]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}.\\
-       Then $[[iN]] = [[ [us]uN ]] = [[ α⁻ ]] = [[iM]]$. 
+       \item $[[ Γ ⊢ [uσ]uN ≤ iM ]]$ is derived by \ruleref{\ottdruleDOneNVarLabel}.\\
+       Then $[[iN]] = [[ [uσ]uN ]] = [[ α⁻ ]] = [[iM]]$. 
        Here the first equality holds because $[[uN]]$ is not a unification variable:

### 2023-06-29 proofs/tex/alg-pos-subtyping-lemmas.tex
wip

@@
 \begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, and $[[Γ ; Θ ⊢ uP]]$ then\\
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$
-    implies $[[us : Θ|uv uP]]$ and 
-    for any $[[us']]$ such that $[[Θ ⊢ us' ⇒ us]]$,
-    $[[ Γ ⊢ [us']uP ≥ iQ ]]$.
+    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; Θ ⊢ uP]]$, and 
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
+    then $[[Θ ⊢ SC]]$ and
+    for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : SC]]$,
+    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
 \end{lemma}
 \begin{proof} 
-    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$. Let us consider the last rule to infer this judgment.
+    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$. 
+    Let us consider the last rule to infer this judgment.
     \begin{caseof}
     \item \ruleref{\ottdruleAPUVarLabel} then
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ has shape $[[Γ;Θ ⊨ â⁺ ≥ iP' ⫤ (â⁺ :≥ iQ')]]$ where
     $[[â⁺[Δ] ∊ Θ]]$ and $[[upgrade G ⊢ iP' to Δ = iQ']]$.
@@
         \item $[[Δ ⊢ iQ']]$ and
-        \item $[[Γ ⊢ iQ ≥ iP]]$.
+        \item $[[Γ ⊢ iQ' ≥ iP]]$.
     \end{enumerate}
 
-    Since $[[â⁺[Δ] ∊ Θ|uv â⁺]]$ and $[[Δ ⊢ iQ']]$, it is clear that $[[(â⁺ :≥ iQ') : Θ | uv â⁺ ]]$.
+    Since $[[â⁺[Δ] ∊ Θ]]$ and $[[Δ ⊢ iQ']]$, 
+    it is clear that $[[ Θ ⊢ (â⁺ :≥ iQ') ]]$.
 
-    It is left to show that $[[Γ ⊢ [us']â⁺ ≥ iP']]$ for any $[[us']]$ s.t. $[[Θ ⊢ us' ⇒ (â⁺ :≥ iQ')]]$.
-    The latter weakening statement means that either $[[us']] \ni [[â⁺ :≥ iQ'']]$ or
-    $[[us']] \ni [[(â⁺ :≈ iQ'')]]$ for $[[Δ ⊢ iQ'' ≥ iQ']]$. In any case,
-    $[[Δ ⊢ [us']â⁺ ≥ iQ]]$. By weakening the context to $[[Γ]]$ and combining this judgment
-    transitively (\cref{todo}) with $[[Γ ⊢ iQ ≥ iP]]$, we have $[[Γ ⊢ [us']â⁺ ≥ iP]]$,
+    It is left to show that $[[Γ ⊢ [uσ]â⁺ ≥ iP']]$ for any normalized $[[uσ]]$ 
+    s.t. $[[Θ ⊢ uσ : (â⁺ :≥ iQ')]]$.
+    The latter means that $[[ Θ(â⁺) ⊢ [uσ]â⁺ ≥ iQ' ]]$, i.e. $[[Δ ⊢ [uσ]â⁺ ≥ iQ']]$. 
+    By weakening the context to $[[Γ]]$ and combining this judgment
+    transitively with $[[Γ ⊢ iQ' ≥ iP]]$, we have $[[Γ ⊢ [uσ]â⁺ ≥ iP]]$,
     as required.
@@
     \item \label{case:pos-subt-soundness:var} \ruleref{\ottdruleAPVarLabel}  
-    then $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
-    Then $[[uv a⁺ = ∅]]$, and $[[us]] = [[· : ·]]$ satisfies $[[us : Θ|∅]]$.
-    Since $[[uv a⁺ = ∅]]$, application of any unification solution $[[us']]$ does not change $[[a⁺]]$, i.e.
-    $[[ [us'] a⁺ = a⁺]]$. Therefore, $[[Γ ⊢ [us']a⁺ ≥ a⁺]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
+    then $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
+    Then $[[uv a⁺ = ∅]]$, and $[[SC]] = [[·]]$ satisfies $[[Θ ⊢ ·]]$.
+    Since $[[uv a⁺ = ∅]]$, application of any substitution $[[uσ]]$ 
+     does not change $[[a⁺]]$, i.e. $[[ [uσ] a⁺ = a⁺]]$.
+      Therefore, $[[Γ ⊢ [uσ]a⁺ ≥ a⁺]]$ holds by \ruleref{\ottdruleDOneNVarLabel}.
 
@@
      \ruleref{\ottdruleAShiftDLabel} then
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ↓uN ≥ ↓iM ⫤ us]]$.\\
-    Then the next step of the algorithm is the unification of $[[nf(uN)]]$ and $[[nf(iM)]]$.
-    By the soundness of the unification algorithm (\cref{lemma:unification-soundness}),
-    it returns an equivalence-only solution $[[us]]$ such that $[[us : Θ | uv uN]]$.
-    By \cref{todo}, since $[[us]]$ is equivalence-only and $[[Γ ⊢ Θ]]$, $[[Θ ⊢ us' ⇒ us]]$ means 
-    $[[ Γ ⊢ Sub us' ≈ Sub us : Ord uv uN ]]$ as substitutions.
-    $[[ [us]nf(uN) = nf(iM) ]]$ implies $[[Γ ⊢ [us]nf(uN) ≈ nf(iM)]]$, and then 
-        $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$. \ilyam{add lemmas}
-
-        Let us rewrite the left-hand side and the right-hand side of $[[Γ ⊢ [us']nf(uN) ≈ nf(iM)]]$ by 
-        transitivity of equivalence (\cref{corollary:equivalence-transitivity}).
-        By \cref{corollary:nf-sound-wrt-subt-equiv,corollary:subst-pres-equiv},
-        $[[Γ ⊢ [us']nf(uN) ≈ [us']uN ]]$. By \cref{corollary:nf-sound-wrt-subt-equiv}, 
-        $[[Γ ⊢ nf(iM) ≈ iM ]]$. 
-        This way, we have $[[Γ ⊢ [us']uN ≈ iM]]$.
-        Then by \ruleref{\ottdruleDOneShiftULabel}
-        and congruence of substitution, $[[Γ ⊢ [us']↓uN ≥ ↓iM]]$.
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ has shape $[[Γ;Θ ⊨ ↓uN ≥ ↓iM ⫤ SC]]$.\\
+    Then the next step of the algorithm is the unification of $[[nf(uN)]]$ and $[[nf(iM)]]$,
+    and it returns the resulting unification constraint $[[lift UC]] = [[SC]]$ as the result.
+    By the soundness of unification (\cref{lemma:unification-soundness}),
+    $[[Θ ⊢ SC]]$ and for any normalized $[[uσ]]$, $[[Θ ⊢ uσ : SC]]$
+    implies $[[ [uσ]nf(uN) = nf(iM) ]]$, 
+    then we rewrite the left-hand side by \cref{lemma:norm-subst-distr}:
+    $[[ nf([uσ]uN) = nf(iM) ]]$ and apply \cref{lemma:subt-equiv-algorithmization}:
+    $[[Γ ⊢ [uσ]uN ≈ iM]]$, then by \ruleref{\ottdruleDOneShiftULabel},
+    $[[Γ ⊢ ↓[uσ]uN ≥ ↓iM]]$.
 
@@
        \ruleref{\ottdruleAExistsLabel} then
-        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us]]$ has shape $[[Γ;Θ ⊨ ∃nas.uP' ≥ ∃nbs.iQ' ⫤ us]]$ s.t. either $[[nas]]$ or $[[nbs]]$ is not empty.\\
+        $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ has shape $[[Γ;Θ ⊨ ∃nas.uP' ≥ ∃nbs.iQ' ⫤ SC]]$ s.t. either 
+        $[[nas]]$ or $[[nbs]]$ is not empty.\\
         Then the algorithm creates fresh unification variables $[[â⁻*[Γ,nbs] ]]$, 
         substitutes the old $[[nas]]$ with them in $[[uP']]$, and makes the recursive call:
-        $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP ≥ iQ ⫤ us']]$, returning as the result
-        $[[us]] = [[us' \ {α̂⁻*}]]$.
+        $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP' ≥ iQ' ⫤ SC']]$, returning as the result
+        $[[SC]] = [[SC' \ {α̂⁻*}]]$.
 
-        Notice that $[[Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$, $[[Γ,nbs ⊢ iQ']]$, and 
-        $[[Γ,nbs; Θ, â⁻*[Γ, nbs] ⊢ [â⁻*/nas] uP' ]]$, so the induction hypothesis is applicable,
-        that is $[[us' : Θ, â⁻*[Γ, nbs] | uv [â⁻*/nas]uP']]$ and $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$ for any
-        $[[us'2]]$ s.t. $[[Θ, â⁻*[Γ, nbs] ⊢ us'2 ⇒ us']]$.
+        Let us take an arbitrary normalized $[[uσ]]$ s.t. $[[Θ ⊢ uσ : SC' \ {α̂⁻*}]]$.
+        We wish to show $[[Γ ⊢ [uσ]uP ≥ iQ]]$, i.e. $[[Γ ⊢ ∃nas.[uσ]uP' ≥ ∃nbs.iQ']]$.
+        To do that, we apply \ruleref{\ottdruleDOneExistsLabel}, and what is left to show is
+        $[[Γ, nbs ⊢ [iNs/nas][uσ]uP' ≥ iQ']]$ for some $[[iNs]]$.
+        If we construct a normalized $[[uσ']]$ such that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']]$
+        and for some $[[iNs]]$, $[[ [iNs/nas][uσ]uP' ]] = [[ [uσ'][â⁻*/nas]uP' ]]$,
+        we can apply the induction hypothesis to 
+        $[[Γ, nbs; Θ, â⁻*[G, nbs] ⊨ [â⁻*/nas] uP ≥ iQ ⫤ SC']]$ and infer 
+        the required subtyping.
 
-        Since the domain of $[[us']]$ is $[[uv [â⁻*/nas]uP']]$, the domain of 
-        $[[us]] = [[us' \  {α̂⁻*}]]$ is $[[uv [â⁻*/nas]uP' \ {â⁻*}]] = [[uv ∃nas.uP']]$,
-        this way, $[[us : Θ | uv ∃nas.uP']]$, as required.
+        Let us construct such $[[uσ']]$ by extending $[[uσ]]$ with $[[â⁻*]]$
+        mapped to the corresponding types in $[[SC']]$:
+        $$
+        [[ [uσ']β̂± ]]  = 
+            \begin{cases}
+               [[ [uσ]β̂± ]] & \text{if } [[β̂±]] \in [[dom(SC') \ {α̂⁻*}]]  \\
+               [[ nf(iN) ]] & \text{if } [[β̂±]] \in [[α̂⁻*]] \text{ and } [[(β̂± :≈ iN)]] \in SC' \\
+            \end{cases}
+        $$
 
-        It is left to show that $[[Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ']]$ for any $[[us2]]$ s.t. $[[Θ ⊢ us2 ⇒ us]]$.
-        Let us consider an arbitrary such $[[us2]]$. Let us construct $[[us'2]]$, 
-        extending $[[us2]]$ to the domain $[[uv [â⁻*/nas]uP']]$ with the values of $[[us']]$,
-        i.e.  $[[us'2 : Θ | uv [â⁻*/nas]uP']]$, 
-        and
-        \[
-            [[us'2(β̂±)]]  = 
-            \begin{cases}
-               [[us2(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']] \\
-               [[us'(β̂±)]] & \text{if } [[β̂±]] \in [[â⁻*]]
-            \end{cases}
-        \] 
-        , where the application of the unification solution to a variable returns the 
-        corresponding \emph{unification entry}. 
-        Notice that $[[Sub us'2|uv uP']] = [[us2]]$.
-    It is easy to see that $[[Θ ⊢ us'2 ⇒ us']]$: 
+        It is easy to see that $[[uσ']]$ is normalized.
+        Let us show that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']]$.
+        Let us take an arbitrary entry $[[scE]]$ from $[[SC']]$ restricting a variable $[[β̂±]]$.
+        Suppose $[[β̂±]] \in [[dom(SC') \ {α̂⁻*}]]$. Then
+        $[[ (Θ, â⁻*[G, nbs])(β̂±) ⊢ [uσ']β̂± : scE ]]$ is
+        rewritten as $[[ Θ(β̂±) ⊢ [uσ]β̂± : scE ]]$, which holds since $[[Θ ⊢ uσ : SC']]$.
+        Suppose $[[β̂±]] = [[αî⁻]] \in [[α̂⁻*]]$. Then
+        $[[scE]] = [[(αî⁻ :≈ iN)]]$ for some $[[iN]]$, 
+        $[[ [uσ']αî⁻ ]] = [[ nf(iN) ]]$ by the definition,
+        and $[[ Γ, nbs ⊢ nf(iN) : (αî⁻ :≈ iN) ]]$ by \ruleref{\ottdruleSATSCENEqLabel},
+        since $[[Γ ⊢ nf(iN) ≈ iN]]$ by \cref{lemma:subt-equiv-algorithmization}.
+
+        Finally, let us show that $[[ [iNs/nas][uσ]uP' ]] = [[ [uσ'][â⁻*/nas]uP' ]]$.
+        For $[[iNi]]$, we take the \emph{normalized} type restricting $[[αî⁻]]$ in $[[SC']]$.
+        Let us take an arbitrary variable from $[[uP]]$.
         \begin{enumerate}
-        \item if $[[β̂±]] \in [[uv uP']]$ then $[[us'2(β̂±)]] = [[us2(β̂±)]] \Rightarrow [[us(β̂±)]] = [[us'(β̂±)]]$;
-        \item it $[[β̂±]] \in [[â⁻*]]$ then $[[us'2(β̂±)]] = [[us'(β̂±)]] \Rightarrow [[us'(β̂±)]]$,
+            \item If this variable is a unification variable $[[β̂±]]$, then
+                $[[ [iNs/nas][uσ] β̂± ]] = [[ [uσ]β̂± ]] $, since $[[Θ ⊢ uσ : SC' \ {α̂⁻*}]]$ and 
+                $[[ dom(Θ) ∩ {nas} = ∅ ]]$.
+
+                Notice that $[[β̂±]] \in [[dom(Θ)]]$, which is disjoint from $[[{α̂⁻*}]]$, 
+                that is $[[β̂±]] \in [[dom(SC') \ {α̂⁻*}]]$. This way,
+                $[[ [uσ'][â⁻*/nas]β̂± ]] = [[  [uσ']β̂± ]] = [[ [uσ]β̂± ]]$ by the definition 
+                of $[[uσ']]$,
+            \item If this variable is a regular variable $[[β±]] \notin [[nas]]$, then 
+                $[[ [iNs/nas][uσ] β± ]] = [[ β± ]] $ and $[[ [uσ'][â⁻*/nas]β± ]] = [[ β± ]]$. 
+            \item If this variable is a regular variable $[[αi⁻]] \in [[nas]]$, then 
+                $[[ [iNs/nas][uσ] αi⁻ ]] = [[ iNi ]] = [[ nf(iNi) ]]$
+                (the latter equality holds since $[[iNi]]$ is normalized)
+                and $[[ [uσ'][â⁻*/nas]αi⁻ ]] = [[  [uσ']αî⁻ ]] = [[ nf(iNi) ]]$.
         \end{enumerate}
-    which means that the induction hypothesis can be applied to $[[us'2]]$, i.e.
-    $[[ Γ, nbs ⊢ [us'2][â⁻*/nas]uP' ≥ iQ' ]]$.\\
-    Notice that
-    $
-    \begin{aligned}[t]
-                 [[ [us'2][â⁻*/nas]uP' ]] &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][Sub us'2|uv uP']uP' ]]
-                                          && \text{by substitution properties \ilyam{todo}}\\
-                                          &= [[ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uP' ]]
-                                          && \text{since $[[Sub us'2|uv uP']] = [[us2]]$}.
-    \end{aligned}
-    $\\
-    Also notice that the domain of $[[Sub us'2|{â⁻*} ○ â⁻*/nas]]$ is $[[nas]]$,
-    and the range is $[[Γ, nbs]]$, i.e. $[[Γ, nbs ⊢ Sub us'2|{â⁻*} ○ â⁻*/nas : nas]]$, 
-    which means that we can apply \ruleref{\ottdruleDOneForallLabel} to 
-    $[[ Γ, nbs ⊢ [Sub us'2|{â⁻*} ○ â⁻*/nas][us2]uN' ≤ iM' ]]$
-    to obtain $[[ Γ ⊢ [us2]∃nas.uP' ≥ ∃nbs.iQ' ]]$, as required.
     \end{caseof}
@@
 \begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$ and
-    there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
-    Then there exists $[[us']]$ such that $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ us']]$
-    and $[[Θ ⊢ us ⇒ us']]$.
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$.
+    Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
+    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
 \end{lemma}
 \begin{proof}
-    We prove it by induction on $[[ Γ ⊢ [us]uP ≥ iQ ]]$.
-    Let us consider the last rule used in the derivation of  $[[ Γ ⊢ [us]uP ≥ iQ ]]$,
-    but first consider the base case for the substitution $[[ [us]uP ]]$:
+    Let us prove this lemma by induction on $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
+    Let us consider the last rule used in the derivation,
+    but first, consider the base case for the substitution $[[ [uσ]uP ]]$:
+    \begin{caseof}
+        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[nbs]]$)\\
+        Then by assumption, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ (where $[[ {nbs} ∩ fv [uσ]α̂⁺ = ∅]]$).
+        By \cref{todo}, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ means 
+        $[[ Γ ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ ]]$, and hence $[[ Γ ⊢ [uσ]α̂⁺ ≥ iQ ]]$.
+        By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some $[[Δ]]$.
 
-    \begin{caseof}
-        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[pbs]]$)\\
-        Then by assumption, there exists $[[us : Θ | uv uP]]$ such that $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$.
-        $[[us : Θ | uv α̂⁺]]$ means that  $[[us]]$ is either $[[ (α̂⁺ :≈ iP') ]]$ or $[[ (α̂⁺ :≥ iP') ]]$,
-        where $[[Δ ⊢ iP']]$ and $[[α̂⁺[Δ] ∊ Θ]]$.
-        $[[ Γ ⊢ ∃nbs.[us]α̂⁺ ≥ iQ ]]$ means that $[[Γ ⊢ iP' ≥ iQ]]$
-        because multiple inversions of \ruleref{\ottdruleDOneExistsLabel} 
-        gives us $[[Γ ⊢ [us]α̂⁺ ≥ iQ]]$ since $[[ {nbs} ∩ fv [us]α̂⁺ = ∅]]$.
-
-        In the algorithm, after multiple applications of \ruleref{\ottdruleAExistsLabel}
+        In the algorithm trying to infer the subtyping 
+        $[[Γ; Θ ⊨ ∃nbs.α̂⁺ ≥ iQ ⫤ SC]]$,
+        after multiple applications of \ruleref{\ottdruleAExistsLabel}
         the type $[[∃nbs.α̂⁺]]$ is reduced to $[[α̂⁺]]$.
@@
         \ruleref{\ottdruleAPUVarLabel}
-        and the resulting solution is $[[us']] = [[(α̂⁺ :≥ iQ')]]$ where
+        and the resulting restriction is $[[SC']] = [[(α̂⁺ :≥ iQ')]]$ where
         $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.
@@
         Why does the upgrade procedure terminates?
-        Because $[[iP']]$ satisfies the pre-conditions of the completeness of the upgrade
-        (\cref{lemma:upgrade-completeness})
-        :
+        Because $[[ [uσ]α̂⁺ ]]$ satisfies the pre-conditions of the completeness of the upgrade
+        (\cref{lemma:upgrade-completeness}):
         \begin{itemize}
-            \item $[[Δ ⊢ iP']]$ because $[[iP' = [us]α̂⁺]]$, $[[us : Θ | uv uP]]$ and 
-            $[[α̂⁺[Δ] ∊ Θ | uv uP]]$,
-            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted before
+            \item $[[Δ ⊢ iP']]$ because $[[iP' = [uσ]α̂⁺]]$ and $[[Θ ⊢ uσ]]$ and 
+            $[[α̂⁺[Δ] ∊ Θ]]$,
+            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted above
         \end{itemize}
-        Moreover, completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
+        Moreover, the completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
         and further, we strengthen it to $[[Δ ⊢ iP' ≥ iQ']]$
         (since by the soundness of the upgrade (\cref{lemma:upgrade-soundness}),
-        $[[Δ ⊢ iQ']]$).
-
-        It means that $[[Θ ⊢ (α̂⁺ :≈ iP') ⇒ (α̂⁺ :≥ iQ') ]]$ and 
-        $[[ Θ ⊢ (α̂⁺ :≥ iP') ⇒ (α̂⁺ :≥ iQ') ]]$, which means that in any case, 
-        $[[ Θ ⊢ us ⇒ us']]$.
+        $[[Δ ⊢ iQ']]$).  It means that $[[Δ ⊢ iP' : (α̂⁺ :≥ iQ') ]]$, that is 
+        $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ') ]]$, as required.
 
         \item \label{case:pos-subt-complete-pvar}
-        $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
-        i.e. $[[iP]] = [[ [us]uP ]] = [[ α⁺ ]] = [[iQ]]$.
+        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
+        i.e. $[[iP]] = [[ [uσ]uP ]] = [[ α⁺ ]] = [[iQ]]$.
         Here the first equality holds because $[[uP]]$ is not a unification variable:
-        this case has been covered by \cref{case:pos-subt-complete-base}.
+        it has been covered by \cref{case:pos-subt-complete-base}.
         The second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
 
-        Notice that $[[us : Θ | uv α⁺]]$ = $[[us : Θ | ∅]]$ = $[[·]]$.
+        The algorithm applies \ruleref{\ottdruleAPVarLabel} and 
+        infers $[[SC]] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
+        Then $[[Θ ⊢ uσ : ·]]$ holds trivially.
 
-        The algorithm applies \ruleref{\ottdruleAPVarLabel} and 
-        infers $[[us']] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
+        \item \label{case:pos-subt-complete-upshift} $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ 
+        is derived by \ruleref{\ottdruleDOneShiftDLabel},
 
-        Since $[[Θ ⊢ · ⇒ ·]]$, we have $[[Θ ⊢ us ⇒ us']]$.
-
-        \item \label{case:pos-subt-complete-upshift} $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneShiftDLabel},
-
-        Then $[[ uP ]] = [[ ↓uN ]]$, since the substitution $[[ [us]uP ]]$ must preserve the 
+        Then $[[ uP ]] = [[ ↓uN ]]$, since the substitution $[[ [uσ]uP ]]$ must preserve the 
         top-level constructor of $[[uP]]\neq [[α̂⁺]]$ (the case $[[uP]] = [[α̂⁺]]$ has been covered
         by \cref{case:pos-subt-complete-base}), and $[[uQ]] = [[ ↓iM ]]$,
-        and by inversion, $[[ Γ ⊢ [us]uN ≈ iM ]]$.
+        and by inversion, $[[ Γ ⊢ [uσ]uN ≈ iM ]]$.
 
@@
         the algorithm tries to apply \ruleref{\ottdruleAShiftDLabel}: 
-        $[[G;Θ ⊨ ↓uN ≥ ↓iM ⫤ us']]$. The premise of this rule is the
+        $[[G;Θ ⊨ ↓uN ≥ ↓iM ⫤ SC]]$. The premise of this rule is the
         unification of $[[nf(uN)]]$ and $[[nf(iM)]]$:
-        $[[Γ;Θ ⊨ nf(uN) ≈u nf(iM) ⫤ us']]$. Let us show that the unification successfully 
-        terminates and returns $[[us']] = [[nf(us)]]$.
+        $[[Γ;Θ ⊨ nf(uN) ≈u nf(iM) ⫤ UC]]$. And the algorithm 
+        returns it as a subtyping constraint $[[SC]] = [[lift UC]]$.
 
-        To demonstrate that the unification terminates, we apply the completeness 
+        To demonstrate that the unification terminates
+        ant $[[uσ]]$ satisfies the resulting constraints, 
+        we apply the completeness 
         of the unification algorithm (\cref{lemma:unification-completeness}). 
-        In order to do that, we need to provide a unifier of 
-        $[[nf(uN)]]$ and $[[nf(iM)]]$. Thankfully, $[[nf(us)]]$ does it.
+        In order to do that, we need to provide a substitution unifying  
+        $[[nf(uN)]]$ and $[[nf(iM)]]$. 
+        Let us show that $[[nf(uσ)]]$ is such a substitution.
 
@@
             \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
-            \item $[[nf(us) : Θ | uv nf(uN)]]$ because $[[us: Θ | uv uN]]$ (\cref{todo})
+            \item $[[ Θ ⊢ nf(uσ) ]]$ because $[[Θ ⊢ uσ ]]$ (\cref{todo})
             \item $ \begin{aligned}[t]
-                    [[ Γ ⊢ [us]uN ≈ iM ]] &\Rightarrow [[ [us]uN ≈ iM ]]
+                    [[ Γ ⊢ [uσ]uN ≈ iM ]] &\Rightarrow [[ [uσ]uN ≈ iM ]]
                                           && \text {by \cref{lemma:equiv-completeness}}\\
-                                          &\Rightarrow [[ nf([us]uN) = nf(iM) ]]
+                                          &\Rightarrow [[ nf([uσ]uN) = nf(iM) ]]
                                           && \text {by \cref{lemma:normalization-completeness}}\\
-                                          &\Rightarrow [[ [nf(us)]nf(uN) = nf(iM) ]]
+                                          &\Rightarrow [[ [nf(uσ)]nf(uN) = nf(iM) ]]
                                           && \text {by \cref{lemma:norm-subst-distr}}\\
@@
         Then by the completeness of the unification,
-        $[[Γ ; Θ ⊨ nf(uN) ≈u nf(iM) ⫤ nf(us)]]$.
-        This way, the subtyping algorithm terminates and the resulting solution is
-        $[[nf(us)]]$.
+        $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ exists, and
+        $[[Θ ⊢ nf(uσ) : lift UC]]$.  Then by \cref{todo}, $[[Θ ⊢ uσ : lift UC]]$.
 
-        It is left to note that $[[Θ ⊢ us ⇒ nf(us)]]$, by \cref{todo}, since $[[Θ ⊢ us ≈ nf(us)]]$
-
-      \item $[[ Γ ⊢ [us]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
+      \item $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
       We should only consider the case
-      when the substitution $[[ [us]uP ]]$ results in the existential type 
+      when the substitution $[[ [uσ]uP ]]$ results in the existential type 
       $[[∃nas.iP'']]$ (for $[[iP'']] \neq [[∃]]\dots$) by congruence, 
@@
       This is because the case when $[[uP = ∃nbs.α̂⁺]]$ has been covered
-      (\cref{case:pos-subt-complete-base}), and thus, the substitution $[[us]]$ must
+      (\cref{case:pos-subt-complete-base}), and thus, the substitution $[[uσ]]$ must
       preserve all the outer quantifiers of $[[uP]]$ and does not generate any new ones.
 
-      This way, $[[uP]] = [[∃nas.uP']]$, $[[ [us]uP ]] = [[ ∃nas.[us]uP' ]]$ 
-      (assuming $[[nas]]$ does not intersect with the range of $[[us]]$)
+      This way, $[[uP]] = [[∃nas.uP']]$, $[[ [uσ]uP ]] = [[ ∃nas.[uσ]uP' ]]$ 
+      (assuming $[[nas]]$ does not intersect with the range of $[[uσ]]$)
       and $[[iQ]] = [[ ∃nbs.iQ' ]]$, where either $[[nas]]$ or $[[nbs]]$ is not empty.
 
-      By inversion, $[[ Γ ⊢ [σ][us]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ : nas]]$.
-      Since $[[σ]]$ and $[[us]]$ have disjoint domains,
+      By inversion, $[[ Γ ⊢ [σ][uσ]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ : nas]]$.
+      Since $[[σ]]$ and $[[uσ]]$ have disjoint domains,
       and the range of one does not intersect with the domain of the other,
-      they commute, i.e. $[[ Γ, nbs ⊢ [us][σ]uP' ≥ iQ' ]]$
+      they commute, i.e. $[[ Γ, nbs ⊢ [uσ][σ]uP' ≥ iQ' ]]$
       (notice that the tree inferring this judgement is 
       a proper subtree of the tree inferring 
-      $[[ Γ ⊢ [us]uP ≥ iQ ]]$).
+      $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$).
 
@@
       and makes the recursive call:
-      $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us1]]$,
+      $[[Γ, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ' ⫤ SC1]]$,
       (where $[[uP0]] = [[ [â⁻*/nas]uP' ]]$),
-      returning $[[us1 \ {â⁻*}]]$ as the result.
+      returning $[[SC1 \ {â⁻*}]]$ as the result.
 
-       Notice that $[[ [us][σ][nas/â⁻*]uP0 = [us][σ]uP' ]]$,
-       and thus, $[[ Γ, nbs ⊢ [us][σ][nas/â⁻*]uP0 ≥ iQ' ]]$.
-       Let us combine $[[us]]$ and $[[σ  ○ nas/â⁻*]]$ into one 
-       $[[us0]]$---a unification solution for $[[uP0]]$:
+      To show that the recursive call terminates and that 
+      $[[Θ ⊢ uσ : SC1 \ {â⁻*}]]$,
+      it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0]]$---an extension of $[[uσ]]$ with
+      $[[â⁻*]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
+      Then by the induction hypothesis, $[[Θ, â⁻*[G, nbs] ⊢ us0 : SC1]]$,
+      and hence, $[[Θ ⊢ uσ : SC1 \ {â⁻*}]]$, as required.
+
+      Let us construct such a substitution $[[us0]]$:
         \[
-            [[us0(β̂±)]]  = 
+            [[ [uσ0]β̂± ]]  = 
             \begin{cases}
-               [[ β̂± :≈ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[â⁻*]] \\
-               [[us(β̂±)]] & \text{if } [[β̂±]] \in [[uv uP']]
+               [[ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[â⁻*]] \\
+               [[ [uσ]β̂± ]] & \text{if } [[β̂±]] \in [[ uv(uP') ]]
             \end{cases}
        \]
+       It is easy to see $[[Θ, â⁻*[Γ, nbs] ⊢ uσ0]]$:
+       \begin{enumerate}
+            \item for $[[αî⁻]] \in [[â⁻*]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
+            i.e. $[[ Γ, nbs ⊢ [σ]αi⁻ ]]$ holds since $[[Γ, nbs ⊢ σ : nas]]$,
+            \item for $[[β̂±]] \in [[ uv(uP') ]] \subseteq [[dom(Θ)]]$, $[[ (Θ, â⁻*[Γ, nbs])(β̂±) ⊢ [uσ0] β̂± ]]$,
+            i.e. $[[Θ(β̂±) ⊢ [uσ] β̂± ]]$ holds since $[[Θ ⊢ uσ]]$.
+       \end{enumerate}
 
-       Let us show that the induction hypothesis is applicable to 
-       $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$, with the meta-context $[[ Θ, â⁻*[Γ, nbs] ]]$
-       \begin{itemize}
-        \item $[[ us0 : (Θ, â⁻*[Γ, nbs])|uv uP0 ]]$. Notice that for every $[[αî⁻]] \in [[â⁻*]]$, 
-            the type corresponding to the entry $[[us0(αî⁻)]]$ 
-            is well-formed in $[[Γ, nbs]]$ since $[[Γ, nbs ⊢ σ : nas ]]$;
-            and for every $[[β̂±]] \in [[uv uP']]$, 
-            the type corresponding to the entry $[[us0(β̂±)]]$ is well-formed in
-            context $[[Θ(β̂±)]]$ since $[[us : Θ | uv uP']]$.
-        \item $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$. This is because
-            $[[ [us0]uP0 ]] = [[ [us][σ][nas/â⁻*]uP0 ]] = [[ [us][σ]uP' ]]$.
-        \item $[[ Γ, nbs ⊢ Θ, â⁻*[Γ, nbs] ]]$ since
-            $[[Γ, nbs ⊢ Θ]]$ and $[[ {Γ, nbs} ⊆ {Γ, nbs} ]]$.
-        \item $[[Γ, nbs ; Θ, â⁻*[Γ, nbs] ⊢ uP0 ]]$
-       \end{itemize}
-
-       This way, we apply the induction hypothesis to $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ' ]]$ and 
-       infer that there exists\\
-        $[[us0' :  (Θ, â⁻*[Γ, nbs]) | uv uP0]]$ such that
-       $[[G, nbs; Θ, â⁻*[G, nbs] ⊨ uP0 ≥ iQ ⫤ us0']]$,
-       and $[[Θ, â⁻*[G, nbs] ⊢ us0' ⇒ us0]]$.
-
-       This way, the algorithm terminates with the result $[[us0' \ {â⁻*}]]$
-       and it is left to show $[[ Θ ⊢ us ⇒ us0' \ {â⁻*} ]]$. Notice that
-       $[[us0' \ {â⁻*} : Θ | uv uP' ]]$. To show $[[ Θ ⊢ us ⇒ us0' \
-       {â⁻*} ]]$, it suffices to consider a unification variable $[[β̂±]] \in 
-       [[uv uP']]$ and show that $[[Γ ⊢ us(β̂±) ⇒ us0'(β̂±)]]$. It holds by the
-       reflexivity of weakening (\cref{lemma:entry-weakening-preorder}) since
-       $[[us0'(β̂±)]] = [[us(β̂±)]]$.
-
-      \item The positive case when $[[Γ ⊢ [us]uP ≥ iQ]]$ is derived by 
-      \ruleref{\ottdruleDOnePVarLabel} is symmetric to the corresponding negative
-      \cref{case:subt-complete-nvar}.
-      Notice that $[[ [us]uP = β⁺ ]]$ 
-      implies $[[uP = β⁺]]$ because the case when $[[uP = α̂⁺]]$ has been covered 
-      (\cref{case:subt-complete-base}).
+       Now, let us show that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
+       To do that, we notice that $[[ [uσ0]uP0 ]] = [[ [uσ][σ][nas/â⁻*]uP0 ]]$:
+       let us consider an arbitrary variable appearing freely in $[[uP0]]$:
+       \begin{enumerate}
+        \item if this variable is a metavariable $[[αî⁻]] \in [[â⁻*]]$, then
+        $[[ [uσ0]αî⁻ ]] = [[ [σ]αi⁻ ]]$ and 
+        $[[ [uσ][σ][nas/â⁻*]αî⁻ ]] = [[ [uσ][σ]αi⁻ ]] = [[ [σ]αi⁻ ]]$,
+        \item if this variable is a metavariable $[[β̂±]] \in [[ uv(uP0) \ {â⁻*} ]] = [[uv(uP')]]$, then
+        $[[ [uσ0]β̂± ]] = [[ [uσ]β̂± ]]$ and $[[ [uσ][σ][nas/â⁻*]β̂± ]] = [[ [uσ][σ]β̂± ]] = [[ [uσ]β̂± ]]$,
+        \item if this variable is a regular variable from $[[fv(uP0)]]$, both substitutions do not change it:
+        $[[ uσ0 ]]$, $[[ uσ ]]$ and $[[ nas / â⁻* ]]$ act on metavariables, 
+        and $[[σ]]$ is defined on $[[nas]]$, however, $[[{nas} ∩ fv(uP0) = ∅]]$.
+       \end{enumerate}
+       This way, $[[ [uσ0]uP0 ]] = [[ [uσ][σ][nas/â⁻*]uP0 ]] = [[ [uσ][σ]uP' ]]$,
+       and thus, $[[ Γ, nbs ⊢ [uσ0]uP0 ≥ iQ' ]]$.
 

### 2023-06-29 proofs/tex/constraint-merge-lemmas.tex
wip

\begin{lemma} [Soundness of Constraint Entry Merge]
\label{lemma:entry-merge-soundness}
For a fixed context $[[Γ]]$,
suppose that  $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$. 
If $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined then
\begin{enumerate}
    \item $[[Γ ⊢ scE]]$
    \item For any $[[Γ ⊢ iP]]$, $[[Γ ⊢ iP : scE]]$ implies $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ 
\end{enumerate}
\end{lemma}
\begin{proof}
    Let us consider the rule forming $[[Γ ⊢ scE1 & scE2 = scE]]$.
    \begin{caseof}
        \item \ruleref{\ottdruleSCMEPEqEqLabel}, i.e. 
        $[[Γ ⊢ scE1 & scE2 = scE]]$
        has form $[[Γ ⊢ (pua :≈ iQ) & (pua :≈ iQ') = (pua :≈ iQ)]]$
        and $[[nf(iQ) = nf(iQ')]]$. The latter implies $[[Γ ⊢ iQ ≈ iQ']]$ by
        \cref{lemma:subt-equiv-algorithmization}.
        Then
        \begin{enumerate}
            \item $[[Γ ⊢ scE]]$, i.e. $[[Γ ⊢ pua :≈ iQ]]$ holds by assumption;
            \item by inversion, $[[Γ ⊢ iP : (pua :≈ iQ)]]$ means $[[Θ ⊢ iP ≈ iQ]]$,
            and by transitivity of equivalence (\cref{corollary:equivalence-transitivity}), 
            $[[Θ ⊢ iP ≈ iQ']]$. Thus, $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ hold
            by \ruleref{\ottdruleSATSCEPEqLabel}.
        \end{enumerate}
        \item \ruleref{\ottdruleSCMENEqEqLabel} the negative case is proved in exactly the same way as the positive one.
        \item \ruleref{\ottdruleSCMESupSupLabel} 
        Then $[[scE1]]$ is $[[pua :≥ iQ1]]$, $[[scE2]]$ is $[[pua :≥ iQ2]]$,
        and $[[scE1 & scE2]] = [[scE]]$ is $[[pua :≥ iQ]]$ where $[[iQ]]$ is the least upper bound of $[[iQ1]]$ and $[[iQ2]]$.
        Then by \cref{lemma:lub-soundness},
        \begin{itemize}
            \item $[[Γ ⊢ iQ]]$,
            \item $[[Γ ⊢ iQ ≥ iQ1]]$,
            \item $[[Γ ⊢ iQ ≥ iQ2]]$.
        \end{itemize}

        Let us show the required properties.
        \begin{itemize}
            \item $[[Γ ⊢ scE]]$ holds from $[[Γ ⊢ iQ]]$,
            \item Assuming $[[Γ ⊢ iP : scE]]$, by inversion, we have $[[Γ ⊢ iP ≥ iQ]]$.
                Combining it transitively with $[[Γ ⊢ iQ ≥ iQ1]]$, we have $[[Γ ⊢ iP ≥ iQ1]]$.
                Analogously, $[[Γ ⊢ iP ≥ iQ2]]$.
                Then $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ hold by \ruleref{\ottdruleSATSCESupLabel}.
        \end{itemize}

        \item \ruleref{\ottdruleSCMESupEqLabel}
        Then $[[scE1]]$ is $[[pua :≥ iQ1]]$, $[[scE2]]$ is $[[pua :≈ iQ2]]$, 
        where $[[Γ;· ⊨ uQ2 ≥ iQ1 ⫤ ·]]$, and the resulting   
        $[[scE1 & scE2]] = [[scE]]$ is equal to $[[scE2]]$, that is $[[pua :≈ iQ2]]$.

        Let us show the required properties.
        \begin{itemize}
            \item By assumption, $[[Γ ⊢ iQ]]$, and hence $[[Γ ⊢ scE]]$.
            \item Since $[[uv(uQ2) = ∅]]$, 
                $[[Γ;· ⊨ uQ2 ≥ iQ1 ⫤ ·]]$ implies $[[Γ ⊢ iQ2 ≥ iQ1]]$
                by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}).
                Then let us take an arbitrary $[[Γ ⊢ iP]]$ such that $[[Γ ⊢ iP : scE]]$.
                Since $[[scE2]] = [[scE]]$, $[[Γ ⊢ iP : scE2]]$ holds immediately.

                By inversion, $[[Γ ⊢ iP : (pua :≈ iQ2)]]$ means $[[Γ ⊢ iP ≈ iQ2]]$, 
                and then by transitivity of subtyping (\cref{lemma:subtyping-transitivity}),
                $[[Γ⊢ iP ≥ iQ1]]$.  Then $[[Γ ⊢ iP : scE1]]$ holds by \ruleref{\ottdruleSATSCESupLabel}.
        \end{itemize}
        \item \ruleref{\ottdruleSCMEEqSupLabel} Thee proof is analogous to the previous case.
    \end{caseof}
\end{proof}

\begin{lemma} [Soundness of Constraint Merge] \label{lemma:merge-soundness}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$ 
    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
    Then 
    \begin{enumerate}
        \item $[[Θ ⊢ SC]]$,
        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    By definition, $[[SC1 & SC2]] = [[SC]]$ consists of three parts:
    entries of $[[SC1]]$ that do not have matching entries of $[[SC]]$,
    entries of $[[SC2]]$ that do not have matching entries of $[[SC1]]$,
    and the merge of matching entries.

    Let us show $[[Θ ⊢ SC]]$.
    First, let us assume that an entry $[[scE]] \in [[SC]]$ belongs to the first group, 
    i.e. $[[scE]] \in [[SC1]]$.  Let us denote the variable $[[scE]]$ as $[[α̂±]]$. 
    Then $[[Θ(α̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC1]] \ni [[scE]]$.
    Analogously, if $[[scE]]$ belongs to the second group, then $[[Θ(α̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC2]] \ni [[scE]]$.
    Finally, if $[[scE]]$ belongs to the third group, then $[[scE]]$ is a merge of two entries $[[Θ(α̂±) ⊢ scE1]]$ 
    and $[[Θ(α̂±) ⊢ scE2]]$.  Then $[[Θ(α̂±) ⊢ scE]]$ holds by \cref{lemma:entry-merge-soundness}.

    Let us show the second property.
    We take an arbitrary $[[uσ]]$ such that $[[Θ ⊢ uσ]]$ and $[[Θ ⊢ uσ : SC]]$.
    To prove $[[Θ ⊢ uσ : SC1]]$, we need to show that for any $[[scE1]] \in [[SC1]]$, restricting $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ holds.

    Let us assume that $[[α̂±]] \notin [[dom(SC2)]]$. It means that $[[SC]] \ni [[scE1]]$, 
    and then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.

    Otherwise, $[[SC2]]$ contains an entry $[[scE2]]$ restricting $[[α̂±]]$,
    and $[[SC]] \ni [[scE]]$ where $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$.
    Then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
    and by \cref{lemma:entry-merge-soundness}, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.

    The proof of $[[Θ ⊢ uσ : SC2]]$ is symmetric.
\end{proof}

\begin{lemma} [Completeness of Constraint Entry Merge]
    \label{lemma:entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iP : scE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : scE1]]$ and $[[Γ ⊢ iN : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iN : scE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us consider the shape of $[[scE1]]$ and $[[scE2]]$.
    \begin{caseof}
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≈ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≈ iQ2]]$.
            Then by transitivity of equivalence (\cref{corollary:equivalence-transitivity}),
            $[[Γ ⊢ iQ1 ≈ iQ2]]$, which means $[[nf(iQ1) = nf(iQ2)]]$ by
            \cref{lemma:subt-equiv-algorithmization}.
            Hence, \ruleref{\ottdruleSCMEPEqEqLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = scE2]]$, and $[[Γ ⊢ iP : scE2]]$ holds by assumption.
        \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            Then by transitivity of subtyping, $[[Γ ⊢ iQ1 ≥ iQ2]]$,
            which means $[[Γ ; · ⊨ uQ1 ≥ iQ2 ⫤ ·]]$ by \cref{lemma:pos-subt-completeness}.
            This way, \ruleref{\ottdruleSCMEEqSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = scE1]]$, and $[[Γ ⊢ iP : scE1]]$ holds by assumption.
        \item $[[scE1]]$ is $[[pua :≥ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≥ iQ1]]$, 
            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≥ iQ2]]$.
            By the completeness of the least upper bound (\cref{lemma:lub-completeness}), 
            $[[Γ ⊨ iQ1 ∨ iQ2 = iQ]]$, and $[[Γ ⊢ iP ≥ iQ]]$. 
            This way, \ruleref{\ottdruleSCMESupSupLabel} applies to infer
            $[[Γ ⊢ scE1 & scE2 = (pua :≥ iQ)]]$, 
            and $[[Γ ⊢ iP : (pua :≥ iQ)]]$ holds by \ruleref{\ottdruleSATSCESupLabel}.
        \item The negative cases are proved symmetrically.
    \end{caseof}
\end{proof}

\begin{lemma} [Completeness of Constraint Merge] 
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
    \begin{enumerate}
        \item $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
        \item $[[Θ ⊢ SC]]$, and
        \item $[[Θ ⊢ uσ : SC]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    By  definition, $[[SC1 & SC2]]$ is a union of
    \begin{enumerate}
        \item entries of $[[SC1]]$, which do not have matching entries in $[[SC2]]$,
        \item entries of $[[SC2]]$, which do not have matching entries in $[[SC1]]$, and 
        \item the merge of matching entries.
    \end{enumerate}

    This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that each of these components is defined.
    To prove $[[Θ ⊢ uσ : SC]]$ we need to show that for every entry $[[scE]]$ of $[[SC]]$ restricting a variable $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ holds.

    It is clear that the first two components of this union exist, and that if $[[scE]]$ is an entry 
    restricting $[[α̂±]]$ $[[dom(SC1) \ dom(SC2)]]$ then then $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ because  
    $[[Θ ⊢ uσ : SC1]]$ and $[[scE]] \in [[SC1]]$. Analogously, if $[[scE]]$ is an entry
    from the second component of the union, the property holds for it.  Let us show that the third component exists, and each of its entries satisfies the property.  Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.  $[[Θ ⊢ uσ : SC1]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[Θ ⊢ uσ : SC2]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
    Then by \cref{lemma:entry-merge-completeness}, $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ is defined and $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$.
\end{proof}

\begin{lemma}[Substitution existence]
    \label{lemma:substitution-existence}
    If $[[Θ ⊢ SC]]$ then there exists $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC]]$.
\end{lemma}
\begin{proof}

\end{proof}

### 2023-06-29 proofs/tex/equiv-lemmas.tex
wip

@@
   Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
+\end{proof}
+
+\begin{corollary}[Normalization preserves subtyping] 
+  \label{corollary:nf-pres-subt}
+  Assuming all the types are well-formed in context $[[Γ]]$,
+  \begin{itemize}
+    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
+    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  \hfill
+  \begin{itemize}
+    \item [$+$]  
+    \begin{itemize}
+      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
+        By \cref{corollary:nf-sound-wrt-subt-equiv},
+        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
+        in particular, by inversion, 
+        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
+        Then by the transitivity of subtyping 
+        (\cref{corollary:subtyping-transitivity}), 
+        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
+      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
+        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
+        and inversion, 
+        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
+        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
+    \end{itemize}
+    \item [$-$] The negative case is proved symmetrically.
+  \end{itemize}
 \end{proof}

### 2023-06-29 proofs/tex/singularity-lemmas.tex
wip

\begin{lemma}[Soundness of Entry Singularity]
    \label{lemma:entry-singularity-soundness}
    Suppose $[[scE singular]]$.
    Then 
    \begin{itemize}
        \item [$+$] for any $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
            if $[[Γ ⊢ iP1 : scE]]$ and $[[Γ ⊢ iP2 : scE]]$ then $[[Γ ⊢ iP1 ≈ iP2]]$;
        \item [$-$] for any $[[iN1]]$ and $[[iN2]]$ well-formed in $[[Γ]]$,
            if $[[Γ ⊢ iN1 : scE]]$ and $[[Γ ⊢ iN2 : scE]]$ then $[[Γ ⊢ iN1 ≈ iN2]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us consider how $[[scE singular]]$ is formed.
    \begin{caseof}
        \item \ruleref{\ottdruleSINGNEqLabel}, that is $[[scE]] = [[α̂⁻ :≈ iM]]$.
            Then $[[Γ ⊢ iNi : scE]]$ means $[[Γ ⊢ iNi ≈ iM]]$ 
            (by inversion of \ruleref{\ottdruleSATSCENEqLabel}).
            Then by symmetry and transitivity of equivalence (\cref{lemma:subtyping-reflexivity,corollary:equivalence-transitivity}), 
            $[[Γ ⊢ iN1 ≈ iN2]]$.
        \item \ruleref{\ottdruleSINGPEqLabel}. This case is symmetric to the previous one.

        \item \ruleref{\ottdruleSINGSupVarLabel}, that is 
            $[[scE]] = [[α̂⁺ :≥ ∃nas.β⁺]]$.
            Then $[[Γ ⊢ iPi : scE]]$ means $[[Γ ⊢ iPi ≥ ∃nas.β⁺]]$.

            Let us show that it implies $[[Γ ⊢ iPi ≈ ∃nas.β⁺]]$.
            By applying \cref{lemma:shape-of-supertypes} once, 
            we have $[[Γ, nas ⊢ iPi ≥ β⁺]]$.
            By applying it again, we notice that
            $[[Γ, nas ⊢ iPi ≥ β⁺]]$ implies $[[iPi = ∃nas'.β⁺]]$.
            Finally, it is easy to see that $[[Γ ⊢ ∃nas'.β⁺ ≈ ∃nas.β⁺]]$

            This way, by symmetry and transitivity of equivalence (\cref{lemma:subtyping-reflexivity,corollary:equivalence-transitivity}),
            $[[Γ ⊢ iP1 ≈ ∃nas.β⁺]]$ and $[[Γ ⊢ iP2 ≈ ∃nas.β⁺]]$ implies $[[Γ ⊢ iP1 ≈ iP2]]$.

        \item \ruleref{\ottdruleSINGSupShiftLabel},
            that is $[[scE]] = [[α̂⁺ :≥ ∃nbs.↓iN]]$, 
            where $[[iN ≈ nbj]]$.
            Then $[[Γ ⊢ iPi : scE]]$ means $[[Γ ⊢ iPi ≥ ∃nbs.↓iN]]$

            Let us show that it implies $[[Γ ⊢ iPi ≈ ∃nbs.↓iN]]$.

            $
            \begin{aligned}[h]
            [[Γ ⊢ iPi ≥ ∃nbs.↓iN]] &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nf(iN)]] \text{ where } [[ord {nbs} in iN' = nbs']] 
                                   && \text{by \cref{corollary:nf-pres-subt}} \\
                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nf(nbj)]]  
                                   && \text{by \cref{lemma:normalization-completeness}}\\
                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nbn]]  
                                   && \text{by definition of normalization}\\
                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbj.↓nbj]]  
                                   && \text{since $[[ord {nbs} in nf(iN')]] = [[nbj]]$}\\
                                   &\Rightarrow [[Γ, nbj ⊢ nf(iP) ≥ ↓nbj]] \text { and } [[nbj ∉ fv(nf(iPi))]]
                                   && \text{by \cref{lemma:shape-supertypes-norm}}\\
            \end{aligned}
            $

            By \cref{lemma:shape-supertypes-norm}, 
            the last subtyping means that $[[nf(iPi) = ∃nas.↓iN']]$,
            such that
            \begin{enumerate}
                \item $[[Γ, nbj, nas ⊢ iN']]$
                \item $[[ord {nas} in iN' = nas]]$
                \item for some substitution $[[Γ, nbj ⊢ σ : nas]]$, 
                    $[[ [σ]iN' = nbj ]]$.
            \end{enumerate}
            Since $[[nbj ∉ fv(nf(iPi))]]$,
            the latter means that $[[iN' = na]]$, and then 
            $[[nf(iPi) = ∃na.↓na]]$ for some $[[na]]$.
            Finally, notice that all the types of shape
            $[[∃na.↓na]]$ are equal, and hence,
            $[[nf(iP1) = nf(iP2)]]$, which implies
            $[[Γ ⊢ iP1 ≈ iP2]]$ by \cref{lemma:subt-equiv-algorithmization}.
   \end{caseof}

\end{proof}

\begin{lemma}[Completeness of Entry Singularity]
    \label{lemma:entry-singularity-completeness}
    \begin{itemize}
        \item [$-$] Suppose that for any $[[iN1]]$ and $[[iN2]]$ well-formed in $[[Γ]]$
            such that $[[Γ ⊢ iN1 : scE]]$ and $[[Γ ⊢ iN2 : scE]]$,
            $[[Γ ⊢ iN1 ≈ iN2]]$ holds. Then $[[scE singular]]$ holds.
        \item [$+$] Suppose that for any $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$
            such that $[[Γ ⊢ iP1 : scE]]$ and $[[Γ ⊢ iP2 : scE]]$,
            $[[Γ ⊢ iP1 ≈ iP2]]$ holds. Then $[[scE singular]]$ holds.
    \end{itemize}
\end{lemma}
\begin{proof}
    \begin{itemize}
        \item [$-$] Since $[[iNi]]$ are negative, by inversion of
            $[[Γ ⊢ iN1 : scE]]$, $[[scE]]$ has shape $[[α̂⁻ :≈ iM]]$.
            Then $[[scE singular]]$ by \ruleref{\ottdruleSINGNEqLabel}.
        \item [$+$]
            Let us consider the shape of $[[scE]]$:
            \begin{caseof}
                \item $[[scE]] = [[(α̂⁺ :≈ iP)]]$ then $[[scE]]$ is 
                    singular by \ruleref{\ottdruleSINGPEqLabel};
                \item $[[scE]] = [[(α̂⁺ :≥ iP)]]$. Let us consider the shape 
                    of the positive type $[[iP]]$:
                    \begin{caseof}
                        \item $[[iP]] = [[∃nbs.β⁺]]$ (for potentially empty $[[nbs]]$) then
                            $[[scE]]$ is singular by \ruleref{\ottdruleSINGSupVarLabel};
                        \item $[[iP]] = [[∃nbs.↓iN]]$ (for potentially empty $[[nbs]]$).
                            Notice since $[[Γ ⊢ iP ≥ iP]]$, $[[Γ ⊢ iP : scE]]$ holds.

                            Notice that $[[Γ ⊢ ∃γ⁻.↓γ⁻ ≥ ∃nbs.↓iN]]$ 
                            (by \ruleref{\ottdruleDOneExistsLabel}, 
                            with substitution $[[iN / γ⁻]]$), and thus, 
                            $[[Γ ⊢ ∃γ⁻.↓γ⁻ : scE]]$ by \ruleref{\ottdruleSATSCESupLabel}.

                            This way, by assumption, $[[Γ ⊢ ∃γ⁻.↓γ⁻ ≈ ∃nbs.↓iN]]$, which implies
                            $[[nf(∃nbs.↓iN) = ∃γ⁻.↓γ⁻]]$ (by \cref{lemma:subt-equiv-algorithmization}), 
                            which by definition of normalization means
                            $[[∃nbs'.↓nf(iN) = ∃γ⁻.↓γ⁻]]$, where $[[ord {nbs} in iN' = nbs']]$.
                            This way, $[[nbs']]$ is a variable $[[β⁻]]$, and $[[ nf(iN) = β⁻ ]]$.
                            Notice that $[[β⁻]] \in [[nbs']] \subseteq [[nbs]]$ by \cref{lemma:ord-soundness}.

                            This way, $[[Γ ⊢ iN ≈ β⁻]]$ for $[[β⁻]] \in [[nbs]]$ (by \cref{lemma:subt-equiv-algorithmization}),
                            and hence, $[[scE]]$ is singular by \ruleref{\ottdruleSINGSupShiftLabel}.
                    \end{caseof}
            \end{caseof}
    \end{itemize}
\end{proof}

\begin{lemma}[Soundness of Singularity]
    \label{lemma:singularity-soundness}
    Suppose $[[Θ ⊢ SC]]$, and $[[SC singular]]$. 
    Then for any $[[uσ1]]$ and $[[uσ2]]$ such that $[[Θ ⊢ uσ1 : SC]]$ and $[[Θ ⊢ uσ2 : SC]]$,
    $[[Θ ⊢ uσ1 ≈ uσ2 : dom(SC)]]$.
\end{lemma}
\begin{proof}
    Suppose that $[[Θ ⊢ uσ : SC]]$.
    It means that for any $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ holds.
    $[[SC singular]]$ means $[[scE singular]]$,
    and hence, by \cref{lemma:entry-singularity-completeness},
    $[[ [uσ]α̂± ]]$ is unique up-to equivalence.

    Since the uniqueness holds for every variable from $[[dom(SC)]]$,
    $[[uσ]]$ is defined uniquely on this set.
\end{proof}

\begin{lemma}[Completeness of Singularity]
\end{lemma}

### 2023-06-29 proofs/tex/subst-lemmas.tex
WIP: equivalence lemmas
Subtyping: soundness is done, completeness WIP
WIP: misc subtyping lemmas
wip

@@
 
-\begin{lemma}[Substitution preserves subtyping]
-  \label{lemma:subst-pres-subt}
+\begin{corollary}[Substitution preserves equivalence]
+  \label{corollary:subst-pres-equiv}
 
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
+  Suppose that $[[Γ ⊢ σ : Γ1]]$. Then
   \begin{itemize}
-  \item[$+$] if $[[Γ, Γ1 ⊢ iP]]$,~ $[[Γ, Γ1 ⊢ iQ]]$,~ and $[[Γ, Γ1 ⊢ iP ≥ iQ]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$
-  \item[$-$] if $[[Γ, Γ1 ⊢ iN]]$,~ $[[Γ, Γ1 ⊢ iM]]$,~ and $[[Γ, Γ1 ⊢ iN ≤ iM]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iN ≤ [σ]iM]]$
+  \item[$+$] if $[[Γ1 ⊢ iP]]$,~ $[[Γ1 ⊢ iQ]]$,~ and $[[Γ1 ⊢ iP ≈ iQ]]$ ~ then $[[Γ ⊢ [σ]iP ≈ [σ]iQ]]$
+  \item[$-$] if $[[Γ1 ⊢ iN]]$,~ $[[Γ1 ⊢ iM]]$,~ and $[[Γ1 ⊢ iN ≈ iM]]$ ~ then $[[Γ ⊢ [σ]iN ≈ [σ]iM]]$
   \end{itemize}
+\end{corollary}
+
+\begin{lemma}[]
+  Suppose that $[[{Γ'} ⊆ {Γ}]]$,
+  $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
+  Then 
+  \begin{enumerate}
+    \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
+    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$;
+    \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
+    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ {Γ'}]]$.
+  \end{enumerate}
 \end{lemma}
 \begin{proof}
-  \ilyam{todo}
+  Let us make an additional assumption that $[[σ1]]$, $[[σ2]]$, 
+  and the mentioned types are normalized. If they are not,
+  we normalize them first.
+
+  Notice that the normalization preserves
+  the set of free variables (\cref{lemma:fv-nf}),
+  well-formedness (\cref{corollary:wf-nf}), 
+  and equivalence (\cref{lemma:subt-equiv-algorithmization}), 
+  and distributes over substitution (\cref{lemma:norm-subst-distr}). 
+  This way, the assumed and desired properties are equivalent to their 
+  normalized versions.
+
+  We prove it by induction on the structure of $[[iP]]$ and mutually, $[[iN]]$.
+  Let us consider the shape of this type.
+  \begin{caseof}
+    \item $[[iP]] = [[α⁺]] \in [[Γ']]$.
+      Then $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$ means $[[Γ ⊢ σ1 ≈ σ2 : {α⁺}]]$, 
+      i.e. $[[Γ ⊢ [σ1]α⁺ ≈ [σ2]α⁺ ]]$, which holds by assumption.
+    \item $[[iP]] = [[α⁺]] \in [[{Γ} \ {Γ'}]]$.
+      Then $[[fv iP ∩ {Γ'}]] = [[∅]]$, 
+      so $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$ holds vacuously.
+    \item $[[iP]] = [[↓iN]]$.
+      Then the induction hypothesis is applicable to type $[[iN]]$:
+      \begin{enumerate}
+        \item $[[iN]]$ is normalized,
+        \item $[[Γ ⊢ iN]]$ by inversion of $[[Γ ⊢ ↓iN]]$,
+        \item $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ holds by inversion of 
+          $[[Γ ⊢ [σ1]↓iN ≈ [σ2]↓iN]]$, i.e. $[[Γ ⊢ ↓[σ1]iN ≈ ↓[σ2]iN]]$.
+      \end{enumerate}
+      This way, we obtain $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ {Γ'}]]$, 
+      which implies the required equivalence since 
+      $[[fv iP ∩ {Γ'}]] = [[fv ↓iN ∩ {Γ'}]] = [[fv iN ∩ {Γ'}]]$.
+    \item $[[iP]] = [[∃nas.iQ]]$
+      Then the induction hypothesis is applicable to type $[[iQ]]$ 
+      well-formed in context $[[Γ, nas]]$:
+      \begin{enumerate}
+        \item $[[{Γ'} ⊆ {Γ, nas}]]$ since $[[{Γ'} ⊆ {Γ}]]$,
+        \item $[[Γ, nas ⊢ σi : Γ']]$ by weakening,
+        \item $[[iQ]]$ is normalized,
+        \item $[[Γ, nas ⊢ iQ]]$ by inversion of $[[Γ ⊢ ∃nas.iQ]]$,
+        \item Notice that $[[ [σi]∃nas.iQ ]]$ is normalized, and thus, 
+          $[[ [σ1]∃nas.iQ ≈ [σ2]∃nas.iQ]]$ implies 
+          $[[ [σ1]∃nas.iQ = [σ2]∃nas.iQ ]]$
+          (by \cref{lemma:subt-equiv-algorithmization}).).
+          This equality means $[[ [σ1]iQ = [σ2]iQ ]]$, 
+          which implies $[[Γ ⊢ [σ1]iQ ≈ [σ2]iQ]]$.
+      \end{enumerate}
+    \item $[[iN]] = [[iP → iM]]$
+  \end{caseof}
 \end{proof}
-
-\begin{corollary}[Substitution preserves subtyping]
-  \label{corollary:subst-pres-equiv}
-
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-  \item[$+$] if $[[Γ, Γ1 ⊢ iP]]$,~ $[[Γ, Γ1 ⊢ iQ]]$,~ and $[[Γ, Γ1 ⊢ iP ≈ iQ]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iP ≈ [σ]iQ]]$
-  \item[$-$] if $[[Γ, Γ1 ⊢ iN]]$,~ $[[Γ, Γ1 ⊢ iM]]$,~ and $[[Γ, Γ1 ⊢ iN ≈ iM]]$ ~ then $[[Γ, Γ2 ⊢ [σ]iN ≈ [σ]iM]]$
-  \end{itemize}
-\end{lemma}

### 2023-06-29 proofs/tex/unif-constraint-merge-lemmas.tex
wip

\begin{lemma} [Soundness of Unification Constraint Merge]
    \label{lemma:unif-merge-soundness}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ 
    are normalized unification constraints.
    If $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined then
    $[[UC = UC1 ∪ UC2]]$.
\end{lemma}
\begin{proof}
    \hfill
    \begin{itemize}
        \item $[[UC1 & UC2]] \subseteq [[UC1]] \cup [[UC2]]$\\
        By definition, 
        $[[UC1 & UC2]]$ consists of three parts:
        entries of $[[UC1]]$ that do not have matching entries of $[[UC2]]$,
        entries of $[[UC2]]$ that do not have matching entries of $[[UC1]]$,
        and the merge of matching entries.

        If $[[ucE]]$ is from the first or the second part, 
        then $[[ucE]] \in [[UC1]] \cup [[UC2]]$ holds immediately.
        If $[[ucE]]$ is from the third part,
        then $[[ucE]]$ is the merge of two matching entries
        $[[ucE1]] \in [[UC1]]$ and $[[ucE2]] \in [[UC2]]$.
        Since $[[UC1]]$ and $[[UC2]]$ are normalized unification , 
        $[[ucE1]]$ and $[[ucE2]]$ have one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP1]]$ and $[[α̂⁺ :≈ iP2]]$, 
                where $[[iP1]]$ and $[[iP2]]$ are normalized,
                and then since $[[Θ(α̂⁺) ⊢ ucE1 & ucE2 = ucE]]$ exists, 
                \ruleref{\ottdruleUCMEPEqEqLabel} was applied to infer it.
                It means that $[[ucE]] = [[ucE1]] = [[ucE2]]$;
            \item $[[α̂⁻ :≈ iN1]]$ and $[[α̂⁻ :≈ iN2]]$, 
               then symmetrically, 
               $[[Θ(α̂⁻) ⊢ ucE1 & ucE2 = ucE]] = [[ucE1]] = [[ucE2]]$
        \end{itemize}
        In both cases, $[[ucE]] \in [[UC1]] \cup [[UC2]]$.

        \item $[[UC1]] \cup [[UC2]] \subseteq [[UC1 & UC2]]$\\
        Let us take 
        an arbitrary $[[ucE1]] \in [[UC1]]$.
        Then since $[[UC1]]$ is a unification constraint,
         $[[ucE1]]$ has one of the following forms:
        \begin{itemize}
            \item $[[α̂⁺ :≈ iP]]$ where $[[iP]]$ is normalized.
            If $[[α̂⁺]] \notin [[dom(UC2)]]$, then $[[ucE1]] \in [[UC1 & UC2]]$.
            Otherwise, there is a normalized matching
            $[[ucE2]] = [[(α̂⁺ :≈ iP')]] \in [[UC2]]$ and then
            since $[[UC1 & UC2]]$ exists, 
            \ruleref{\ottdruleUCMEPEqEqLabel} was applied to construct
            $[[ucE1 & ucE2]] \in [[UC1 & UC2]]$.
            By inversion of \ruleref{\ottdruleUCMEPEqEqLabel},
            $[[ucE1 & ucE2]] = [[ucE1]]$, and
            $[[nf(iP) = nf(iP')]]$, which since $[[iP]]$
            and $[[iP']]$ are normalized, implies that $[[iP = iP']]$, 
            that is $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
            \item $[[α̂⁻ :≈ iN]]$ where $[[iN]]$ is normalized.
            Then symmetrically, $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$.
        \end{itemize}
        Similarly, if we take an arbitrary $[[ucE2]] \in [[UC2]]$,
        then $[[ucE1]] = [[ucE2]] \in [[UC1 & UC2]]$. 
    \end{itemize}
\end{proof}

\begin{corollary}
    \label{corollary:unif-merge-soundness}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ 
    are normalized unification constraints.
    If $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined then
    \begin{enumerate}
        \item $[[Θ ⊢ UC]]$ is normalized unification constraint,
        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : lift UC]]$ implies 
        $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
    \end{enumerate}
\end{corollary}
\begin{proof}
    It is clear that since $[[UC = UC1 ∪ UC2]]$ (by \cref{lemma:unif-merge-soundness}),
    and being normalized means that all entries are normalized,
    $[[UC]]$ is a normalized unification constraint.
    Analogously, $[[Θ ⊢ UC]] = [[UC1 ∪ UC2]]$ holds immediately, 
    since $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.

    Let us take an arbitrary substitution $[[Θ ⊢ uσ]]$ and assume that 
    $[[Θ ⊢ uσ : lift UC]]$.
    Then $[[Θ ⊢ uσ : lift UCi]]$ holds by definition:
    If $[[ucE]] \in [[lift UCi]] \subseteq [[lift UC1 ∪ lift UC2]] = [[lift UC]]$.
    So $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$ holds.
\end{proof}

% \begin{lemma} [Completeness of Unification Constraint Merge]
%     \label{lemma:unif-merge-completeness}
%     Suppose that $[[Θ ⊢ UC|varset1]]$ and $[[Θ ⊢ UC|varset2]]$.
%     Then $[[Θ ⊢ UC|varset1 & UC|varset2 = UC']]$ exists and
%     $[[UC' = UC|varset1 ∪ varset2]]$.
% \end{lemma}
% \begin{proof}
%     $[[Θ ⊢ UC|varset1 & UC|varset2 = UC']]$ is defined as the union of three parts:
%     entries of $[[UC|varset1]]$ that do not have matching entries of $[[UC|varset2]]$,
%     entries of $[[UC|varset2]]$ that do not have matching entries of $[[UC|varset1]]$,
%     and the merge of matching entries.
%     The first two parts are defined. The merge of matching entries is defined
%     by \ruleref{\ottdruleUCMEPEqEqLabel}, since the matching entries must be equal
%     if they both belong to $[[UC]]$.

%     It remains to show that $[[UC' = UC|varset1 ∪ varset2]]$.
%     It is easy to see that the three parts comprising $[[UC']]$ 
%     correspond to the three parts comprising 
%     $[[UC|varset1 ∪ varset2]] = [[UC | (varset1 \ varset2) ∪ UC | (varset2 \ varset1) ∪ UC | varset1 ∩ varset2]]$. 
% \end{proof}

\begin{lemma} [Completeness of Unification Constraint Entry Merge]
    \label{lemma:unif-entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : ucE1]]$ and $[[Γ ⊢ iP : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iP : ucE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : ucE1]]$ and $[[Γ ⊢ iN : ucE2]]$,
        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iN : ucE]]$.
    \end{itemize}
\end{lemma}
\begin{proof}
    The proof repeats the one of \cref{lemma:entry-merge-completeness}
    and is done by the case analysis on the shape of $[[ucE1]]$ and $[[ucE2]]$.
    However, it only needs to consider two cases.
    \begin{caseof}
        \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$.
        \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iM2]]$.
    \end{caseof}
    The proof of these cases is based only on \cref{lemma:subt-equiv-algorithmization}
    and \cref{corollary:equivalence-transitivity}, and does not require the properties of 
    the least upper bound or subtyping.
\end{proof}

\begin{lemma} [Completeness of Unification Constraint Merge] 
    \label{lemma:unif-merge-completeness}
    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$, 
    \begin{enumerate}
        \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
        \item $[[Θ ⊢ uσ : UC]]$.
    \end{enumerate}
\end{lemma}
\begin{proof}
    The proof repeats the proof of \cref{lemma:merge-completeness} but 
    uses \cref{lemma:unif-entry-merge-completeness} instead of \cref{lemma:entry-merge-completeness}.
\end{proof}

### 2023-06-29 proofs/tex/unification-lemmas.tex
wip

--- 2023-05-02 proofs/tex/unification-lemmas.tex
@@
         $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ us]]$ then $[[us : Θ | uv uP]]$, 
-        $[[us]]$ is equivalence-only, and $[[ [us]uP = iQ ]]$.
+        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
+        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : lift UC]]$,
+        $[[ [uσ]uP = iQ ]]$.
 
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$ then $[[us : Θ | uv uN]]$,
-        $[[us]]$ is equivalence-only, and $[[ [us]uN = iM ]]$.
+        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
+        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : lift UC]]$,
+        $[[ [uσ]uN = iM ]]$.
     \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove by induction on the derivation of 
+    $[[ Γ ; Θ ⊨ uN ≈u iM ⫤ UC ]]$ and mutually $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$.
+    Let us consider the last rule forming this derivation. 
+    \begin{caseof}
+        \item \ruleref{\ottdruleUNVarLabel}, then $[[uN]] = [[α⁻]] = [[iM]]$.
+        The resulting unification constraint is empty: $[[UC]] = [[·]]$.
+        It satisfies $[[Θ ⊢ UC]]$ vacuously, and $[[ [us]α⁻ = α⁻ ]]$, that is $[[ [us]uN = iM ]]$.
+
+        \item \ruleref{\ottdruleUShiftULabel}, then $[[uN]] = [[↑uP]]$ and $[[iM]] = [[↑iQ]]$.
+        The algorithm makes a recursive call to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ returning $[[UC]]$.
+        By induction hypothesis, $[[Θ ⊢ UC]]$ and for any $[[Θ ⊢ uσ : lift UC]]$,
+        $[[ [uσ]uN ]] = [[ [uσ]↑uP ]] = [[ ↑[uσ]uP ]] = [[ ↑iQ ]] = [[ iM ]]$, as 
+        required.
+
+        \item \ruleref{\ottdruleUArrowLabel}, then $[[uN]] = [[uP → uN']]$ and $[[iM]] = [[iQ → iM']]$.
+        The algorithm makes two recursive calls to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and
+        $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ returning $[[Θ ⊢ UC1 & UC2 = UC]]$ as the result.
+
+        It is clear that $[[uP]]$, $[[uN']]$, $[[iQ]]$, and $[[iM']]$ are normalized,
+        and that $[[Γ ; Θ ⊢ uP]]$, $[[Γ ; Θ ⊢ uN']]$, $[[Γ ⊢ iQ]]$, and $[[Γ ⊢ iM']]$.
+        This way, the induction hypothesis is applicable to both recursive calls.
+
+        By applying the induction hypothesis to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$,
+        we have:
+        \begin{itemize}
+            \item $[[Θ ⊢ UC1]]$,
+            \item for any $[[Θ ⊢ uσ' : lift UC1]]$, $[[ [uσ']uP = iQ ]]$.
+        \end{itemize}
+        By applying it to $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$, we have:
+        \begin{itemize}
+            \item $[[Θ ⊢ UC2]]$,
+            \item for any $[[Θ ⊢ uσ' : lift UC2]]$, $[[ [uσ']uN' = iM' ]]$.
+        \end{itemize}
+
+        Let us take an arbitrary $[[Θ ⊢ uσ : lift UC]]$.
+        By the soundness of the constraint merge (\cref{lemma:merge-soundness}), 
+        $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$ implies
+        $[[Θ ⊢ uσ : lift UC1 ]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+
+        Applying the induction hypothesis to $[[Θ ⊢ uσ : lift UC1]]$, we have
+        $[[ [uσ]uP = iQ ]]$; applying it to $[[Θ ⊢ uσ : lift UC2]]$, we have
+        $[[ [uσ]uN' = iM' ]]$.
+        This way, $[[ [uσ]uN ]] = [[ [uσ]uP → [uσ]uN' ]] = [[ iQ → iM' ]] = [[ iM ]]$.
+
+        \item \ruleref{\ottdruleUForallLabel}, then $[[uN]] = [[∀pas.uN']]$ and $[[iM]] = [[∀pas.iM']]$.
+        The algorithm makes a recursive call to $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ UC]]$
+        returning $[[UC]]$ as the result.
+
+        The induction hypothesis is applicable: $[[Γ,pas ; Θ ⊢ uN']]$ and $[[Γ,pas ⊢ iM']]$ hold
+        by inversion, and $[[uN']]$ and $[[iM']]$ are normalized, since $[[uN]]$ and $[[iM]]$ are.
+        Let us take an arbitrary $[[Θ ⊢ uσ : lift UC]]$.
+        By the induction hypothesis, $[[ [uσ]uN' ]] = [[ iM' ]]$. 
+        Then $[[ [uσ]uN ]] = [[ [uσ]∀pas.uN' ]] = [[ ∀pas.[uσ]uN' ]] = [[ ∀pas.iM' ]] = [[ iM ]]$.
+
+        \item \ruleref{\ottdruleUNUVarLabel}, then $[[uN]] = [[α̂⁻]]$, $[[â⁻[Δ] ∊ Θ]]$, and $[[Δ ⊢ iM]]$.
+        As the result, the algorithm returns $[[UC]] = [[ (â⁻ :≈ iM) ]]$.
+
+        It is clear that $[[α̂⁻[Δ] ⊢ (â⁻ :≈ iM) ]]$, since $[[Δ ⊢ iM]]$, 
+        meaning that $[[Θ ⊢ UC]]$.
+
+        Let us take an arbitrary $[[uσ]]$ such that  $[[Θ ⊢ uσ : lift UC]]$.
+        Since $[[UC]] = [[ (â⁻ :≈ iM) ]]$, $[[Θ ⊢ uσ : lift UC]]$ implies 
+        $[[Θ(â⁻) ⊢ [uσ]â⁻ : (â⁻ :≈ iM) ]]$.
+        By inversion of \ruleref{\ottdruleSATSCENEqLabel}, it  means $[[Θ(â⁻) ⊢ [uσ]â⁻ ≈ iM]]$.
+        This way, $[[Θ(â⁻) ⊢ [uσ]uN ≈ iM]]$. 
+        Notice that $[[uσ]]$ and $[[uN]]$ are normalized, and by \cref{todo}, so is $[[ [uσ]uN ]]$.
+        Since both sides of $[[Θ(â⁻) ⊢ [uσ]uN ≈ iM]]$ are normalized,
+        by \cref{lemma:subt-equiv-algorithmization}, we have $[[ [uσ]uN = iM ]]$.
+
+        \item The positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}
 
@@
         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
-        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        assume there exists 
-        $[[us : Θ | uv uP]]$ such that $[[ [us]uP = iQ ]]$.
-        Then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
+        for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uP = iQ ]]$,
+        there exists $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$,
+        and $[[Θ ⊢ uσ : lift UC]]$.
 
@@
         $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        assume there exists $[[us : Θ | uv uN]]$        
-        such that $[[ [us]uN = iM ]]$.
-        Then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ us]]$.
+        for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uN = iM ]]$,
+        there exists $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$,
+        and $[[Θ ⊢ uσ : lift UC]]$.
    \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove it by induction on the structure of $[[uP]]$ and mutually, $[[uN]]$.
+    \begin{caseof}
+        \item $[[uN]] = [[α̂⁻]]$\\
+            $[[Γ ; Θ ⊢ α̂⁻]]$ means that $[[ α̂⁻[Δ] ]] \in [[Θ]]$ for some $[[Δ]]$.
+
+            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
+            $[[Θ ⊢ uσ]]$ means that $[[Δ ⊢ iM]]$.
+            This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 
+            $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
+            $[[Θ ⊢ uσ : lift (â⁻ :≈ iM)]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
+
+        \item $[[uN]] = [[α⁻]]$\\
+            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]α⁻ = iM ]]$.
+            The latter means $[[iM = α⁻]]$.
+
+            Then $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
+            This way, \ruleref{\ottdruleUNVarLabel} infers 
+            $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ ·]]$, 
+            and $[[Θ ⊢ uσ : lift ·]]$ holds trivially.
+
+        \item $[[uN]] = [[↑uP]]$\\
+            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]↑uP = iM ]]$.
+            The latter means $[[ ↑[uσ]uP = iM ]]$, i.e.
+            $[[iM]] = [[↑iQ]]$ for some $[[iQ]]$ and $[[ [uσ]uP = iQ ]]$.
+
+            Let us show that the induction hypothesis is applicable to $[[ [uσ]uP = iQ ]]$.
+            Notice that $[[uP]]$ is normalized, since $[[uN]] = [[↑uP]]$ is normalized,
+            $[[Γ ; Θ ⊢ uP]]$ holds by inversion of $[[Γ ; Θ ⊢ ↑uP]]$, 
+            and $[[Γ ⊢ iQ]]$ holds by inversion of $[[Γ ⊢ ↑iQ]]$.
+
+            This way, by the induction hypothesis there exists $[[UC]]$ such that
+            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[Θ ⊢ uσ : lift UC]]$.
+
+        \item $[[uN]] = [[uP → uN']]$\\
+            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ](uP → uN') = iM ]]$.
+            The latter means $[[ [uσ]uP → [uσ]uN' = iM ]]$, i.e.
+            $[[iM]] = [[iQ → iM']]$ for some $[[iQ]]$ and $[[iM']]$, 
+            such that $[[ [uσ]uP = iQ ]]$ and $[[ [uσ]uN' = iM' ]]$.
+
+            Let us show that the induction hypothesis is applicable to 
+            $[[ [uσ]uP = iQ ]]$ and to $[[ [uσ]uN' = iM' ]]$:
+            \begin{itemize}
+                \item $[[uP]]$ and $[[uN']]$ are normalized, since $[[uN]] = [[uP → uN']]$ is normalized
+                \item $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$ follow from the inversion of $[[Γ ; Θ ⊢ uP → uN']]$,
+                \item $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$ follow from inversion of $[[Γ ⊢ iQ → iM']]$.
+            \end{itemize}
+
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[Θ ⊢ uσ : lift UC1]]$,
+            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+            To apply \ruleref{\ottdruleUArrowLabel} and infer the required
+            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$, we need to show that
+            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[Θ ⊢ uσ : lift UC]]$.
+            It holds by the completeness of the unification constraint merge 
+            (\cref{lemma:merge-completeness}):
+            \begin{itemize}
+                \item $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ holds by the soundness of unification (\cref{lemma:unification-soundness})
+                \item $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$ holds as noted above 
+            \end{itemize}.
+
+        \item $[[uN]] = [[∀pas.uN']]$\\
+            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]∀pas.uN' = iM ]]$.
+            The latter means $[[ ∀pas.[uσ]uN' = iM ]]$, i.e.
+            $[[iM]] = [[∀pas.iM']]$ for some $[[iM']]$ such that $[[ [uσ]uN' = iM' ]]$.
+
+            Let us show that the induction hypothesis is applicable to $[[ [uσ]uN' = iM' ]]$.
+            Notice that $[[uN']]$ is normalized, since $[[uN]] = [[∀pas.uN']]$ is normalized,
+            $[[Γ,pas ; Θ ⊢ uN']]$ follows from inversion of $[[Γ ; Θ ⊢ ∀pas.uN']]$,
+            $[[Γ,pas ⊢ iM']]$ follows from inversion of $[[Γ ⊢ ∀pas.iM']]$, and
+            $[[Θ ⊢ uσ]]$ by assumption.
+
+            This way, by the induction hypothesis, $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ UC]]$ exists and 
+            moreover, $[[Θ ⊢ uσ : lift UC]]$.
+            Hence, \ruleref{\ottdruleUForallLabel} is applicable to infer
+            $[[Γ ; Θ ⊨ ∀pas.uN' ≈u ∀pas.iM' ⫤ UC]]$, that is $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$.
+
+        \item The positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}

### 2023-07-10 proofs/tex/decl-typing-lemmas.tex
wip

\begin{lemma} \label{lemma:app-inf-equ-stable}
    If $[[Γ; Φ ⊢ iN1 ● args ⇒> iM]]$ and $[[Γ ⊢ iN1 ≈ iN2]]$ 
    then $[[Γ; Φ ⊢ iN2 ● args ⇒> iM]]$.
\end{lemma}
\begin{proof}
    By \cref{lemma:equiv-completeness}, 
    $[[Γ ⊢ iN1 ≈ iN2]]$ implies $[[iN1 ≈ iN2]]$.
    Let us prove the required judgement by induction on $[[iN1 ≈ iN2]]$.
    Let us consider the last rule used in the derivation.
    \begin{caseof}
        \item \ruleref{\ottdruleEOneNVarLabel}.
            It means that $[[iN1]]$ is $[[α⁻]]$ and $[[iN2]]$ is $[[α⁻]]$.
            Then the required property coincides with the assumption. 
        \item \ruleref{\ottdruleEOneShiftULabel}. 
            It means that $[[iN1]]$ is $[[↑iP1]]$ and $[[iN2]]$ is $[[↑iP2]]$.
            where $[[iP1 ≈ iP2]]$.

            Then the only rule applicable to infer $[[Γ; Φ ⊢ ↑iP1 ● args ⇒> iM]]$
            is \ruleref{\ottdruleDTEmptyAppLabel},
            meaning that $[[args = ·]]$ and $[[Γ ⊢ ↑iP1 ≈ iM]]$.
            Then by transitivity of equivalence \cref{corollary:equivalence-transitivity},
            $[[Γ ⊢ ↑iP2 ≈ iM]]$, and then \ruleref{\ottdruleDTEmptyAppLabel} is applicable to infer
            $[[Γ; Φ ⊢ ↑iP2 ● · ⇒> iM]]$.

        \item \ruleref{\ottdruleEOneArrowLabel}.
            Then we are proving that  
            $[[Γ; Φ ⊢ (iQ1 → iN1) ● v, args ⇒> iM]]$ and $[[iQ1 → iN1 ≈ iQ2 → iN2]]$
            imply $[[Γ; Φ ⊢ (iQ2 → iN2) ● v, args ⇒> iM]]$.

            By inversion, $[[(iQ1 → iN1) ≈ (iQ2 → iN2)]]$
            means $[[iQ1 ≈ iQ2]]$ and $[[iN1 ≈ iN2]]$.

            By inversion of $[[Γ; Φ ⊢ (iQ1 → iN1) ● v, args ⇒> iM]]$:
            \begin{enumerate}
                \item $[[Γ ; Φ ⊢ v : iP]]$
                \item $[[Γ ⊢ iQ1 ≥ iP]]$,
                    and then by transitivity \cref{corollary:subtyping-transitivity},
                    $[[Γ ⊢ iQ2 ≥ iP]]$;
                \item $[[Γ ; Φ ⊢ iN1 ● args ⇒> iM]]$, 
                    and then by induction hypothesis, $[[Γ ; Φ ⊢ iN2 ● args ⇒> iM]]$.
            \end{enumerate}

            Since we have $[[Γ ; Φ ⊢ v : iP]]$, $[[Γ ⊢ iQ2 ≥ iP]]$ and 
            $[[Γ ; Φ ⊢ iN2 ● args ⇒> iM]]$, we can apply \ruleref{\ottdruleDTArrowAppLabel}
            to infer $[[Γ; Φ ⊢ (iQ2 → iN2) ● v, args ⇒> iM]]$.

        \item \ruleref{\ottdruleEOneForallLabel}
            Then we are proving that 
            $[[Γ ; Φ ⊢ ∀pas1.iN1' ● args ⇒> iM]]$ and $[[∀pas1.iN1' ≈ ∀pas2.iN2']]$
            imply $[[Γ ; Φ ⊢ ∀pas2.iN2' ● args ⇒> iM]]$.

            By inversion of $[[∀pas1.iN1' ≈ ∀pas2.iN2']]$:
            \begin{enumerate}
                \item $[[{pas2} ∩ fv iN1 = ∅]]$,
                \item there exists a bijection 
                    $[[mu : ({pas2} ∩ fv iN2') ↔ ({pas1} ∩ fv iN1')]]$
                    such that $[[iN1' ≈ [mu] iN2']]$.
            \end{enumerate}

            By inversion of $[[Γ ; Φ ⊢ ∀pas1.iN1' ● args ⇒> iM]]$:
            \begin{enumerate}
                \item $[[Γ ⊢ σ : pas1]]$        
                \item $[[Γ ; Φ ⊢ [σ]iN1' ● args ⇒> iM]]$
                \item $[[args ≠ ·]]$
            \end{enumerate}

            Let us construct $[[Γ ⊢ σ0 : pas2]]$ in the following way:
            $$
            \begin{cases}
                [[ [σ0]α⁺ =  [σ][mu]α⁺ ]] & \text{if } [[α⁺]] \in [[ {pas2} ∩ fv iN2' ]] \\
                [[ [σ0]α⁺ =  ∃β⁻.↓β⁻ ]] & \text{otherwise (the type does not matter here)} \\
            \end{cases}
            $$

            Then to infer $[[Γ ; Φ ⊢ iN2 ● args ⇒> iM]]$, we 
            apply \ruleref{\ottdruleDTArrowAppLabel} with $[[σ0]]$. 
            Let us show the required premises:
            \begin{enumerate}
                \item $[[Γ ⊢ σ0 : pas2]]$ by construction;
                \item $[[args ≠ ·]]$ as noted above;
                \item To show $[[Γ ; Φ ⊢ [σ0]iN2' ● args ⇒> iM]]$,
                Notice that $[[ [σ0]iN2' = [σ][mu]iN2' ]]$   
                and since $[[ [mu]iN2' ≈ iN1' ]]$, $[[ [σ][mu]iN2' ≈ [σ]iN1' ]]$.
                This way, by \cref{lemma:equiv-soundness}, $[[Γ ⊢ [σ]iN1' ≈ [σ0]iN2']]$.
                Then the required judgement holds by the induction hypothesis
                applied to $[[Γ ; Φ ⊢ [σ]iN1' ● args ⇒> iM]]$.
            \end{enumerate}
    \end{caseof}
\end{proof}

\newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
\newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement 
    (i.e. the one of the shape
    $[[Γ ; Φ ⊢ v : iP]]$, 
    $[[Γ ; Φ ⊢ c : iN]]$, or 
    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$),
    let us define a metrics $\size{T}$
    as the number of nodes in $T$ that 
    have one of these shapes
    (i.e., $[[Γ ; Φ ⊢ v : iP]]$, 
    $[[Γ ; Φ ⊢ c : iN]]$, or 
    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$).
\end{definition}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement,
    let us a function $\eqNodes{T}$
    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
\end{definition}

\begin{definition}
    For a tree $T$ inferring
    a declarative typing judgement 
    let us define a metrics $\pureSize{T}$
    as the number of nodes in $T$ except 
    the ones labeled with \ruleref{\ottdruleDTPEquivLabel} or 
    \ruleref{\ottdruleDTNEquivLabel}.
    In other words, $\pureSize{T} = \size{T} - \eqNodes{T}$. 
\end{definition}

\begin{lemma}[Declarative typing is preserved under context equivalence]
    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
    \begin{itemize}
        \item [$+$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
        \item [$-$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
        \item [$\bullet$] 
            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$ such that 
            $\pureSize{T_2} = \pureSize{T_1}$.
    \end{itemize}
\end{lemma}
\begin{proof}
    Let us prove it by induction on a pair $(\pureSize{T_1}, \eqNodes{T_1})$.
    Let us consider the last rule applied in $T_1$ (i.e., its root node).
    \begin{caseof}
        \item \ruleref{\ottdruleDTVarLabel}\\
            Then we are proving 
            that $[[Γ ; Φ1 ⊢ x : iP]]$ implies $[[Γ ; Φ2 ⊢ x : iP]]$.
            By inversion, $[[x : iP ∊ Φ1]]$, and 
            since $[[Γ ⊢ Φ1 ≈ Φ2]]$, $[[x : iP' ∊ Φ2]]$ for some $[[iP']]$ 
            such that $[[Γ ⊢ iP ≈ iP']]$.
            Then we infer $[[Γ ; Φ2 ⊢ x : iP']]$ by \ruleref{\ottdruleDTVarLabel},
            and next, $[[Γ ; Φ2 ⊢ x : iP]]$ by \ruleref{\ottdruleDTPEquivLabel}.

            Notice that 
            to infer $T_1$, only \ruleref{\ottdruleDTVarLabel} was applied, 
            to infer $T_2$, only \ruleref{\ottdruleDTVarLabel} and
            \ruleref{\ottdruleDTPEquivLabel} were applied.
            This way, $\pureSize{T_1} = \pureSize{T_2} = 1$.

        \item For \ruleref{\ottdruleDTThunkLabel},
              \ruleref{\ottdruleDTPAnnotLabel}, 
              \ruleref{\ottdruleDTTLamLabel},
              \ruleref{\ottdruleDTReturnLabel},
              \ruleref{\ottdruleDTNAnnotLabel}, and
              the proof is analogous. We
              apply the induction hypothesis to the premise of the rule
              to substitute $[[Φ1]]$ for $[[Φ2]]$ in it. 
              The induction is applicable as the pure size of the 
              premises is less than the pure size of the conclusion by one.
              And after that, we apply the same rule to infer the required judgement.

              By induction hypothesis, 
              for each of the premises (i.e. the children $T_1'$ of $T_1$),
              the tree $T_2'$ inferring the corresponding judgement
              (with $[[Φ1]]$ substituted for $[[Φ2]]$) has the same pure size,
              i.e., $\pureSize{T_1'} = \pureSize{T_2'}$.
              For each of these rules, the pure size of $T_1$ 
              and the pure size $T_2$ is a sum of the pure sizes of their children
              plus one. This way, 
              $ 
                \pureSize{T_1} =
                1 + \sum_{T'_1}{\pureSize{T'_1}} =
                1 + \sum_{T'_2}{\pureSize{T'_2}} =
                \pureSize{T_2} 
              $.

        \item \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel}
            In these cases, the induction hypothesis is also applicable to the premise:
            although the pure size of the premise is the same as the pure size of the conclusion,
            but the number of equivalence nodes is less by one.
            After this note, the proof is analogous to the previous case.

        \item \ruleref{\ottdruleDTtLamLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ λx:iP.c : iP → iN]]$ implies $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$.
            Analogously to the previous case, 
            we apply the induction hypothesis to the
            equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
            and the premise $[[Γ ; Φ1, x:iP ⊢ c : iN]]$
            inferred by $T'_1$ to obtain
            $T'_2$ inferring $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
            Then we construct $T_2$ inferring $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$ by \ruleref{\ottdruleDTtLamLabel}
            and notice that $\pureSize{T_1} = 1 + \pureSize{T'_1} = 1 + \pureSize{T'_2} = \pureSize{T_2}$.

        \item \ruleref{\ottdruleDTVarLetLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ let x = v; c : iN]]$ implies $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$.
            First, we apply the induction hypothesis to 
            $[[Γ; Φ1 ⊢ v : iP]]$ to obtain $[[Γ; Φ2 ⊢ v : iP]]$ 
            of the same pure size.

            Then we apply the induction hypothesis to
            the equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
            and the premise $[[Γ ; Φ1, x:iP ⊢ c : iN]]$ to obtain
            $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
            Then we infer $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$ by \ruleref{\ottdruleDTVarLetLabel}.

            The pure size of the resulting tree $T_2$ and of $T_1$ is
            one plus the sum of sizes of the premise trees.
            The premises of $T_1$ and of $T_2$ have the same pure size by 
            the induction hypothesis. This way, $\pureSize{T_1} = \pureSize{T_2}$.

        \item \ruleref{\ottdruleDTAppLetLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ let x = v(args); c : iN]]$ implies 
            $[[Γ ; Φ2 ⊢ let x = v(args); c : iN]]$.

            We apply the induction hypothesis to each of the premises.
            to rewrite:
            \begin{itemize}
                \item $[[Γ ; Φ1 ⊢ v : ↓iM]]$ into $[[Γ ; Φ2 ⊢ v : ↓iM]]$,
                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ]]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.
                \item $[[Γ ; Φ1, x:iQ ⊢ c : iN]]$ into $[[Γ ; Φ2, x:iQ ⊢ c : iN]]$
                (notice that $[[Γ ⊢ Φ1, x:iQ ≈ Φ2, x:iQ]]$).
            \end{itemize}

            It is left to show the uniqueness of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.

            Then we infer $[[Γ ; Φ2 ⊢ let x = v(args); c : iN]]$ by \ruleref{\ottdruleDTAppLetLabel}.

        \item \ruleref{\ottdruleDTAppLetAnnLabel}
            Then we are proving that
            $[[Γ ; Φ1 ⊢ let x:iP = v(args); c : iN]]$ implies
            $[[Γ ; Φ2 ⊢ let x:iP = v(args); c : iN]]$.

            As in the previous case, we apply the induction hypothesis to each of the premises
            and rewrite:
            \begin{itemize}
                \item $[[Γ ; Φ1 ⊢ v : ↓iM]]$ into $[[Γ ; Φ2 ⊢ v : ↓iM]]$,
                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ]]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$, 
                    and
                \item $[[Γ ; Φ1, x:iP ⊢ c : iN]]$ into $[[Γ ; Φ2, x:iP ⊢ c : iN]]$
                (notice that $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$).
            \end{itemize}

            Notice that $[[Γ ⊢ iP]]$ and $[[Γ ⊢ ↑iQ ≤ ↑iP]]$ 
            do not depend on the variable context, and hold by assumption.
            Then we infer $[[Γ ; Φ2 ⊢ let x:iP = v(args); c : iN]]$ by \ruleref{\ottdruleDTAppLetAnnLabel}.

        \item \ruleref{\ottdruleDTUnpackLabel}, \ruleref{\ottdruleDTNAnnotLabel}, and \ruleref{\ottdruleDTNEquivLabel}
            are proved in the same way.

        \item \ruleref{\ottdruleDTEmptyAppLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$ (inferred by \ruleref{\ottdruleDTEmptyAppLabel})
            implies $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$,
            and if there is only one $[[iN']]$ (up-to-equivalence in $[[Γ]]$) 
            that satisfies $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$,
            then there is only one $[[iN']]$ satisfying $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$.

            To infer $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$, 
            we apply \ruleref{\ottdruleDTEmptyAppLabel}, noting that 
            $[[Γ ⊢ iN ≈ iN']]$ holds by assumption.

            To show the uniqueness $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN' uniq]]$, 
            let us assume $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN'']]$.
            Then the only rule applicable to infer it is \ruleref{\ottdruleDTEmptyAppLabel},
            which, by inversion, means $[[Γ ⊢ iN ≈ iN'']]$, and by transitivity, 
            $[[Γ ⊢ iN' ≈ iN'']]$.

        \item \ruleref{\ottdruleDTArrowAppLabel}
            Then we are proving that 
            $[[Γ ; Φ1 ⊢ iQ → iN ● v, args ⇒> iM]]$ (inferred by \ruleref{\ottdruleDTArrowAppLabel})
            implies $[[Γ ; Φ2 ⊢ iQ → iN ● v, args ⇒> iM]]$.
            And uniqueness of the $[[iM]]$ in the first case implies uniqueness in the second case.

            By induction, we rewrite $[[Γ ; Φ1 ⊢ v : iP]]$ into $[[Γ ; Φ2 ⊢ v : iP]]$, 
            and $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$ into $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
            Then we infer $[[Γ ; Φ2 ⊢ iQ → iN ● v, args ⇒> iM]]$ by \ruleref{\ottdruleDTArrowAppLabel}.

            Now, let us show the uniqueness.
            The only rule that can infer $[[Γ ; Φ1 ⊢ iQ → iN ● v, args ⇒> iM]]$
            is \ruleref{\ottdruleDTArrowAppLabel}.
            Then by inversion, 
            uniqueness of $[[Γ ; Φ1 ⊢ iQ → iN ● v, args ⇒> iM]]$ implies
            uniqueness of $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$. By 
            the induction hypothesis, it implies the uniqueness of 
            $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.

            Suppose that 
            $[[Γ ; Φ2 ⊢ iQ → iN ● v, args ⇒> iM']]$.
            By inversion, $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM']]$, 
            which by uniqueness of $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$ implies
            $[[Γ ⊢ iM ≈ iM']]$.

        \item \ruleref{\ottdruleDTForallAppLabel}
    \end{caseof}

\end{proof}

### 2023-07-10 proofs/tex/singularity-lemmas.tex
wip

@@
     \label{lemma:entry-singularity-soundness}
-    Suppose $[[scE singular]]$.
-    Then 
     \begin{itemize}
-        \item [$+$] for any $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
-            if $[[Γ ⊢ iP1 : scE]]$ and $[[Γ ⊢ iP2 : scE]]$ then $[[Γ ⊢ iP1 ≈ iP2]]$;
-        \item [$-$] for any $[[iN1]]$ and $[[iN2]]$ well-formed in $[[Γ]]$,
-            if $[[Γ ⊢ iN1 : scE]]$ and $[[Γ ⊢ iN2 : scE]]$ then $[[Γ ⊢ iN1 ≈ iN2]]$.
+        \item [$+$] Suppose $[[scE singular with iP]]$ for $[[iP]]$ well-formed in $[[Γ]]$.
+            Then $[[ Γ ⊢ iP : scE ]]$
+            and for any $[[ Γ ⊢ iP']]$ such that $[[Γ ⊢ iP' : scE]]$, $[[Γ ⊢ iP' ≈ iP]]$;
+        \item [$-$] Suppose $[[scE singular with iN]]$ for $[[iN]]$ well-formed in $[[Γ]]$.
+            Then $[[ Γ ⊢ iN : scE ]]$
+            and for any $[[ Γ ⊢ iN']]$ such that $[[Γ ⊢ iN' : scE]]$, $[[Γ ⊢ iN' ≈ iN]]$.
     \end{itemize}
@@
 \begin{proof}
-    Let us consider how $[[scE singular]]$ is formed.
+    Let us consider how $[[scE singular with iP]]$ or $[[scE singular with iN]]$ is formed.
     \begin{caseof}
-        \item \ruleref{\ottdruleSINGNEqLabel}, that is $[[scE]] = [[α̂⁻ :≈ iM]]$.
-            Then $[[Γ ⊢ iNi : scE]]$ means $[[Γ ⊢ iNi ≈ iM]]$ 
-            (by inversion of \ruleref{\ottdruleSATSCENEqLabel}).
-            Then by symmetry and transitivity of equivalence (\cref{lemma:subtyping-reflexivity,corollary:equivalence-transitivity}), 
-            $[[Γ ⊢ iN1 ≈ iN2]]$.
+        \item \ruleref{\ottdruleSINGNEqLabel}, that is $[[scE]] = [[α̂⁻ :≈ iN0]]$.
+            and $[[iN]]$ is $[[nf(iN0)]]$.
+            Then $[[Γ ⊢ iN' : scE]]$ means $[[Γ ⊢ iN' ≈ iN0]]$, 
+            (by inversion of \ruleref{\ottdruleSATSCENEqLabel}),
+            which by transitivity, using \cref{corollary:nf-sound-wrt-subt-equiv},
+            means $[[Γ ⊢ iN' ≈ nf(iN0)]]$, 
+            as required.
         \item \ruleref{\ottdruleSINGPEqLabel}. This case is symmetric to the previous one.
@@
         \item \ruleref{\ottdruleSINGSupVarLabel}, that is 
-            $[[scE]] = [[α̂⁺ :≥ ∃nas.β⁺]]$.
-            Then $[[Γ ⊢ iPi : scE]]$ means $[[Γ ⊢ iPi ≥ ∃nas.β⁺]]$.
+            $[[scE]] = [[α̂⁺ :≥ ∃nas.β⁺]]$, and $[[iP = β⁺]]$.
 
-            Let us show that it implies $[[Γ ⊢ iPi ≈ ∃nas.β⁺]]$.
+            Since $[[Γ ⊢ β⁺ ≥  ∃nas.β⁺]]$, we have $[[Γ ⊢ β⁺ : scE ]]$, 
+            as required.
+
+            Notice that $[[Γ ⊢ iP' : scE]]$ means $[[Γ ⊢ iP' ≥ ∃nas.β⁺]]$.
+            Let us show that it implies $[[Γ ⊢ iP' ≈ β⁺]]$.
             By applying \cref{lemma:shape-of-supertypes} once, 
-            we have $[[Γ, nas ⊢ iPi ≥ β⁺]]$.
+            we have $[[Γ, nas ⊢ iP' ≥ β⁺]]$.
             By applying it again, we notice that
-            $[[Γ, nas ⊢ iPi ≥ β⁺]]$ implies $[[iPi = ∃nas'.β⁺]]$.
-            Finally, it is easy to see that $[[Γ ⊢ ∃nas'.β⁺ ≈ ∃nas.β⁺]]$
-
-            This way, by symmetry and transitivity of equivalence (\cref{lemma:subtyping-reflexivity,corollary:equivalence-transitivity}),
-            $[[Γ ⊢ iP1 ≈ ∃nas.β⁺]]$ and $[[Γ ⊢ iP2 ≈ ∃nas.β⁺]]$ implies $[[Γ ⊢ iP1 ≈ iP2]]$.
+            $[[Γ, nas ⊢ iP' ≥ β⁺]]$ implies $[[iPi = ∃nas'.β⁺]]$.
+            Finally, it is easy to see that $[[Γ ⊢ ∃nas'.β⁺ ≈ β⁺]]$
 
         \item \ruleref{\ottdruleSINGSupShiftLabel},
-            that is $[[scE]] = [[α̂⁺ :≥ ∃nbs.↓iN]]$, 
-            where $[[iN ≈ nbj]]$.
-            Then $[[Γ ⊢ iPi : scE]]$ means $[[Γ ⊢ iPi ≥ ∃nbs.↓iN]]$
+            that is $[[scE]] = [[α̂⁺ :≥ ∃nbs.↓iN1]]$, 
+            where $[[iN1 ≈ nbj]]$, and $[[iP = ∃α⁻.↓α⁻]]$.
 
-            Let us show that it implies $[[Γ ⊢ iPi ≈ ∃nbs.↓iN]]$.
+            Since $[[Γ ⊢ ∃α⁻.↓α⁻ ≥ ∃nbs.↓iN1]]$ 
+            (by \ruleref{\ottdruleDOneExistsLabel}, with substitution $[[iN1 / α⁻]]$),
+            we have $[[Γ ⊢ ∃α⁻.↓α⁻ : scE ]]$, as required.
+
+            Notice $[[Γ ⊢ iP' : scE]]$ means $[[Γ ⊢ iP' ≥ ∃nbs.↓iN1]]$.
+            Let us show that it implies $[[Γ ⊢ iP' ≈ ∃α⁻.↓α⁻]]$.
 
@@
             \begin{aligned}[h]
-            [[Γ ⊢ iPi ≥ ∃nbs.↓iN]] &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nf(iN)]] \text{ where } [[ord {nbs} in iN' = nbs']] 
+            [[Γ ⊢ iP' ≥ ∃nbs.↓iN1]] &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nf(iN1)]] \text{ where } [[ord {nbs} in iN' = nbs']] 
                                    && \text{by \cref{corollary:nf-pres-subt}} \\
-                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nf(nbj)]]  
+                                   &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nf(nbj)]]  
                                    && \text{by \cref{lemma:normalization-completeness}}\\
-                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbs'.↓nbn]]  
+                                   &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nbn]]  
                                    && \text{by definition of normalization}\\
-                                   &\Rightarrow [[Γ ⊢ nf(iPi) ≥ ∃nbj.↓nbj]]  
-                                   && \text{since $[[ord {nbs} in nf(iN')]] = [[nbj]]$}\\
-                                   &\Rightarrow [[Γ, nbj ⊢ nf(iP) ≥ ↓nbj]] \text { and } [[nbj ∉ fv(nf(iPi))]]
+                                   &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbj.↓nbj]]  
+                                   && \text{since $[[ord {nbs} in nf(iN1)]] = [[nbj]]$}\\
+                                   &\Rightarrow [[Γ, nbj ⊢ nf(iP') ≥ ↓nbj]] \text { and } [[nbj ∉ fv(nf(iP'))]]
                                    && \text{by \cref{lemma:shape-supertypes-norm}}\\
@@
             By \cref{lemma:shape-supertypes-norm}, 
-            the last subtyping means that $[[nf(iPi) = ∃nas.↓iN']]$,
+            the last subtyping means that $[[nf(iP') = ∃nas.↓iN']]$,
             such that
@@
             \end{enumerate}
-            Since $[[nbj ∉ fv(nf(iPi))]]$,
+            Since $[[nbj ∉ fv(nf(iP'))]]$,
             the latter means that $[[iN' = na]]$, and then 
-            $[[nf(iPi) = ∃na.↓na]]$ for some $[[na]]$.
+            $[[nf(iP') = ∃na.↓na]]$ for some $[[na]]$.
             Finally, notice that all the types of shape
-            $[[∃na.↓na]]$ are equal, and hence,
-            $[[nf(iP1) = nf(iP2)]]$, which implies
-            $[[Γ ⊢ iP1 ≈ iP2]]$ by \cref{lemma:subt-equiv-algorithmization}.
+            $[[∃na.↓na]]$ are equal.
    \end{caseof}
@@
     \label{lemma:entry-singularity-completeness}
+    \hfill
     \begin{itemize}
-        \item [$-$] Suppose that for any $[[iN1]]$ and $[[iN2]]$ well-formed in $[[Γ]]$
-            such that $[[Γ ⊢ iN1 : scE]]$ and $[[Γ ⊢ iN2 : scE]]$,
-            $[[Γ ⊢ iN1 ≈ iN2]]$ holds. Then $[[scE singular]]$ holds.
-        \item [$+$] Suppose that for any $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$
-            such that $[[Γ ⊢ iP1 : scE]]$ and $[[Γ ⊢ iP2 : scE]]$,
-            $[[Γ ⊢ iP1 ≈ iP2]]$ holds. Then $[[scE singular]]$ holds.
+        \item [$-$] Suppose that there exists $[[iN]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iN']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iN' : scE]]$ implies $[[Γ ⊢ iN' ≈ iN]]$. 
+            Then $[[scE singular with nf(iN)]]$.
+        \item [$+$] Suppose that there exists $[[iP]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iP']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iP' : scE]]$ implies $[[Γ ⊢ iP' ≈ iP]]$. 
+            Then $[[scE singular with nf(iP)]]$ .
     \end{itemize}
@@
 \begin{proof}
+    \hfill
     \begin{itemize}
-        \item [$-$] Since $[[iNi]]$ are negative, by inversion of
-            $[[Γ ⊢ iN1 : scE]]$, $[[scE]]$ has shape $[[α̂⁻ :≈ iM]]$.
-            Then $[[scE singular]]$ by \ruleref{\ottdruleSINGNEqLabel}.
+        \item [$-$] 
+            By \cref{lemma:constraint-sat},
+            there exists $[[Γ ⊢ iN' : scE]]$.
+            Since $[[iN']]$ is negative, by inversion of
+            $[[Γ ⊢ iN' : scE]]$, $[[scE]]$ has shape $[[α̂⁻ :≈ iM]]$, 
+            where $[[Γ ⊢ iN' ≈ iM]]$, and transitively, $[[Γ ⊢ iN ≈ iM]]$.
+            Then $[[nf(iM) = nf(iN)]]$, 
+            and $[[scE singular with nf(iM)]]$ (by \ruleref{\ottdruleSINGNEqLabel})
+            is rewritten as $[[scE singular with nf(iN)]]$.
         \item [$+$]
+            By \cref{lemma:constraint-sat}, there exists $[[Γ ⊢ iP' : scE]]$, 
+            then by assumption, $[[Γ ⊢ iP' ≈ iP]]$,
+            which by \cref{lemma:entry-sat-equiv} implies $[[Γ ⊢ iP : scE]]$.
+
             Let us consider the shape of $[[scE]]$:
             \begin{caseof}
-                \item $[[scE]] = [[(α̂⁺ :≈ iP)]]$ then $[[scE]]$ is 
-                    singular by \ruleref{\ottdruleSINGPEqLabel};
-                \item $[[scE]] = [[(α̂⁺ :≥ iP)]]$. Let us consider the shape 
-                    of the positive type $[[iP]]$:
+                \item $[[scE]] = [[(α̂⁺ :≈ iQ)]]$ then 
+                    inversion of $[[Γ ⊢ iP : scE]]$
+                    implies $[[Γ ⊢ iP ≈ iQ]]$, and hence, $[[nf(iP) = nf(iQ)]]$
+                    (by \cref{lemma:subt-equiv-algorithmization}).
+                    Then $[[scE singular with nf(iQ)]]$, 
+                    which holds by \ruleref{\ottdruleSINGPEqLabel}, 
+                    is rewritten as $[[scE singular with nf(iP)]]$.
+
+                \item $[[scE]] = [[(α̂⁺ :≥ iQ)]]$.
+                    Then the inversion of $[[Γ ⊢ iP : scE]]$ 
+                    implies $[[Γ ⊢ iP ≥ iQ]]$.
+                    Let us consider the shape of $[[iQ]]$:
                     \begin{caseof}
-                        \item $[[iP]] = [[∃nbs.β⁺]]$ (for potentially empty $[[nbs]]$) then
-                            $[[scE]]$ is singular by \ruleref{\ottdruleSINGSupVarLabel};
-                        \item $[[iP]] = [[∃nbs.↓iN]]$ (for potentially empty $[[nbs]]$).
-                            Notice since $[[Γ ⊢ iP ≥ iP]]$, $[[Γ ⊢ iP : scE]]$ holds.
+                        \item $[[iQ]] = [[∃nbs.β⁺]]$ (for potentially empty $[[nbs]]$).
+                            Then $[[Γ ⊢ iP ≥ ∃nbs.β⁺]]$ 
+                            implies $[[Γ ⊢ iP ≈ β⁺]]$ by 
+                            \cref{lemma:shape-of-supertypes}, 
+                            as was noted in the proof of 
+                            \cref{lemma:entry-singularity-soundness},
+                            and hence, $[[nf(iP) = β⁺]]$.
 
+                            Then $[[scE singular with β⁺]]$, which holds by
+                            \ruleref{\ottdruleSINGSupVarLabel},
+                            can be rewritten as $[[scE singular with nf(iP)]]$.
+
+                        \item $[[iQ]] = [[∃nbs.↓iN]]$ (for potentially empty $[[nbs]]$).
                             Notice that $[[Γ ⊢ ∃γ⁻.↓γ⁻ ≥ ∃nbs.↓iN]]$ 
@@
 
-                            This way, by assumption, $[[Γ ⊢ ∃γ⁻.↓γ⁻ ≈ ∃nbs.↓iN]]$, which implies
+                            Then by assumption, $[[Γ ⊢ ∃γ⁻.↓γ⁻ ≈ iP]]$,
+                            that is $[[nf(iP) = ∃γ⁻.↓γ⁻]]$.
+                            To apply \ruleref{\ottdruleSINGSupShiftLabel}
+                            to infer $[[(α̂⁺ :≥ ∃nbs.↓iN) singular with ∃γ⁻.↓γ⁻]]$,
+                            it is left to show that $[[iN ≈ nbi]]$ for some $i$.
+
+                            Since $[[Γ ⊢ iQ : scE]]$, by assumption,
+                            $[[Γ ⊢ iQ ≈ iP]]$, and by transitivity, 
+                            $[[Γ ⊢ iQ ≈ ∃γ⁻.↓γ⁻]]$.
+                            It implies
                             $[[nf(∃nbs.↓iN) = ∃γ⁻.↓γ⁻]]$ (by \cref{lemma:subt-equiv-algorithmization}), 
@@
                             Notice that $[[β⁻]] \in [[nbs']] \subseteq [[nbs]]$ by \cref{lemma:ord-soundness}.
-
-                            This way, $[[Γ ⊢ iN ≈ β⁻]]$ for $[[β⁻]] \in [[nbs]]$ (by \cref{lemma:subt-equiv-algorithmization}),
-                            and hence, $[[scE]]$ is singular by \ruleref{\ottdruleSINGSupShiftLabel}.
+                            This way, $[[iN ≈ β⁻]]$ for $[[β⁻]] \in [[nbs]]$ (by \cref{lemma:subt-equiv-algorithmization}),
                     \end{caseof}
@@
     \label{lemma:singularity-soundness}
-    Suppose $[[Θ ⊢ SC]]$, and $[[SC singular]]$. 
-    Then for any $[[uσ1]]$ and $[[uσ2]]$ such that $[[Θ ⊢ uσ1 : SC]]$ and $[[Θ ⊢ uσ2 : SC]]$,
-    $[[Θ ⊢ uσ1 ≈ uσ2 : dom(SC)]]$.
+    Suppose $[[Θ ⊢ SC]]$, and $[[SC singular with uσ]]$. 
+    Then $[[Θ ⊢ uσ : SC]]$ and for any 
+    $[[uσ']]$ such that $[[Θ ⊢ uσ' : SC]]$,
+    $[[Θ ⊢ uσ' ≈ uσ : dom(SC)]]$.
 \end{lemma}
 \begin{proof}
-    Suppose that $[[Θ ⊢ uσ : SC]]$.
-    It means that for any $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
-    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ holds.
-    $[[SC singular]]$ means $[[scE singular]]$,
+    Suppose that $[[Θ ⊢ uσ' : SC]]$.
+    It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
+    $[[Θ(α̂±) ⊢ [uσ']α̂± : scE]]$ holds.
+    $[[SC singular with uσ]]$ means $[[scE singular with [uσ]α̂± ]]$,
     and hence, by \cref{lemma:entry-singularity-completeness},
-    $[[ [uσ]α̂± ]]$ is unique up-to equivalence.
+    $[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂±  ]]$ holds.
 
     Since the uniqueness holds for every variable from $[[dom(SC)]]$,
-    $[[uσ]]$ is defined uniquely on this set.
+    $[[uσ]]$ is equivalent to $[[uσ']]$ on this set.
 \end{proof}
@@
 \end{lemma}
+
+\begin{lemma}[Completeness of Singularity]
+    \label{lemma:singularity-completeness}
+    Suppose there exists $[[Θ ⊢ uσ1]]$ such that
+    for any $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ ≈ uσ1 : varset]]$.
+    Then 
+    \begin{itemize}
+        \item $[[SC|varset singular with uσ0]]$ for some $[[uσ0]]$, and
+        \item $[[varset ⊆ dom(SC)]]$.
+    \end{itemize} 
+\end{lemma}
+\begin{proof}
+\end{proof}

### 2023-07-11 proofs/tex/decl-typing-lemmas.tex
Context Equivalence Lemma

@@
 \newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
+\newcommand{\metric}[1]{\ensuremath{\mathsf{metric}(#1)}}
 \newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
 \newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
+\newcommand{\npq}[1]{\ensuremath{\mathsf{npq}(#1)}}
 
-\begin{definition}
-    For a tree $T$ inferring
-    a declarative typing judgement 
-    (i.e. the one of the shape
-    $[[Γ ; Φ ⊢ v : iP]]$, 
-    $[[Γ ; Φ ⊢ c : iN]]$, or 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$),
-    let us define a metrics $\size{T}$
-    as the number of nodes in $T$ that 
-    have one of these shapes
-    (i.e., $[[Γ ; Φ ⊢ v : iP]]$, 
-    $[[Γ ; Φ ⊢ c : iN]]$, or 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$).
+\begin{definition}[Number of prenex quantifiers]
+    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
+    \begin{itemize}
+        \item [$+$] $\npq{[[∃nas.iP]]} = |nas|$, if $[[iP ≠ ∃nbs.iP']]$,
+        \item [$-$] $\npq{[[∀pas.iN]]} = |pas|$, if $[[iN ≠ ∀pbs.iN']]$.
+    \end{itemize}
 \end{definition}
 
-\begin{definition}
+\begin{definition}[Size of a Judgement]
+    \label{def:decl-typing-size}
+    For a declarative typing judgement $J$
+    let us define a metrics $\size{J}$ as a pair of numbers 
+    in the following way:
+    \begin{itemize}
+        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
+            (\size{[[args]]}, \npq{[[iN]]})$)
+    \end{itemize}
+    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
+    syntax tree of the term $[[v]]$ or $[[c]]$
+    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
+\end{definition}
+
+\begin{definition}[Number of Equivalence Nodes]
     For a tree $T$ inferring
@@
 
-\begin{definition}
+\begin{definition}[Metric]
     For a tree $T$ inferring
-    a declarative typing judgement 
-    let us define a metrics $\pureSize{T}$
-    as the number of nodes in $T$ except 
-    the ones labeled with \ruleref{\ottdruleDTPEquivLabel} or 
-    \ruleref{\ottdruleDTNEquivLabel}.
-    In other words, $\pureSize{T} = \size{T} - \eqNodes{T}$. 
+    a declarative typing judgement $J$,
+    let us define a metric $\metric{T}$
+    as a pair $(\size{J}, \eqNodes{T})$.
 \end{definition}
@@
             for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$ such that 
-            $\pureSize{T_2} = \pureSize{T_1}$.
+            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
         \item [$-$] 
             for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$ such that 
-            $\pureSize{T_2} = \pureSize{T_1}$.
+            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
         \item [$\bullet$] 
             for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$ such that 
-            $\pureSize{T_2} = \pureSize{T_1}$.
+            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
     \end{itemize}
@@
 \begin{proof}
-    Let us prove it by induction on a pair $(\pureSize{T_1}, \eqNodes{T_1})$.
+    Let us prove it by induction on the $\metric{T_1}$.
     Let us consider the last rule applied in $T_1$ (i.e., its root node).
@@
 
-            Notice that 
-            to infer $T_1$, only \ruleref{\ottdruleDTVarLabel} was applied, 
-            to infer $T_2$, only \ruleref{\ottdruleDTVarLabel} and
-            \ruleref{\ottdruleDTPEquivLabel} were applied.
-            This way, $\pureSize{T_1} = \pureSize{T_2} = 1$.
-
         \item For \ruleref{\ottdruleDTThunkLabel},
@@
               \ruleref{\ottdruleDTTLamLabel},
-              \ruleref{\ottdruleDTReturnLabel},
-              \ruleref{\ottdruleDTNAnnotLabel}, and
+              \ruleref{\ottdruleDTReturnLabel}, and
+              \ruleref{\ottdruleDTNAnnotLabel}
               the proof is analogous. We
@@
               to substitute $[[Φ1]]$ for $[[Φ2]]$ in it. 
-              The induction is applicable as the pure size of the 
-              premises is less than the pure size of the conclusion by one.
+              The induction is applicable because 
+              the metric of the 
+              premises is less than the metric of the conclusion:
+              the term in the premise is a syntactic subterm of the
+              term in the conclusion.
+
               And after that, we apply the same rule to infer the required judgement.
-
-              By induction hypothesis, 
-              for each of the premises (i.e. the children $T_1'$ of $T_1$),
-              the tree $T_2'$ inferring the corresponding judgement
-              (with $[[Φ1]]$ substituted for $[[Φ2]]$) has the same pure size,
-              i.e., $\pureSize{T_1'} = \pureSize{T_2'}$.
-              For each of these rules, the pure size of $T_1$ 
-              and the pure size $T_2$ is a sum of the pure sizes of their children
-              plus one. This way, 
-              $ 
-                \pureSize{T_1} =
-                1 + \sum_{T'_1}{\pureSize{T'_1}} =
-                1 + \sum_{T'_2}{\pureSize{T'_2}} =
-                \pureSize{T_2} 
-              $.
 
@@
             In these cases, the induction hypothesis is also applicable to the premise:
-            although the pure size of the premise is the same as the pure size of the conclusion,
-            but the number of equivalence nodes is less by one.
-            After this note, the proof is analogous to the previous case.
+            although the first component of the metric 
+            is the same for the premise and the conclusion:
+            $\size{[[Γ ; Φ ⊢ c : iN']]} = \size{[[Γ ; Φ ⊢ c : iN]]} = \size{[[c]]}$,
+            the second component of the metric is less for the premise by one,
+            since the equivalence rule was applied to turn the premise tree into
+            $T1$.
+            Having made this note, we continue the proof in the same way as in the previous case.
 
@@
             $[[Γ ; Φ1 ⊢ λx:iP.c : iP → iN]]$ implies $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$.
-            Analogously to the previous case, 
+            Analogously to the previous cases, 
             we apply the induction hypothesis to the
@@
             and the premise $[[Γ ; Φ1, x:iP ⊢ c : iN]]$
-            inferred by $T'_1$ to obtain
-            $T'_2$ inferring $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
-            Then we construct $T_2$ inferring $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$ by \ruleref{\ottdruleDTtLamLabel}
-            and notice that $\pureSize{T_1} = 1 + \pureSize{T'_1} = 1 + \pureSize{T'_2} = \pureSize{T_2}$.
+            to obtain $[[Γ ; Φ2, x:iP ⊢ c : iN]]$.
+            Notice that $[[c]]$ is a subterm of $[[λx:iP.c]]$,
+            i.e., the metric of the premise tree is less than the metric of the conclusion, 
+            and the induction hypothesis is applicable.
+            Then we infer $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$ by \ruleref{\ottdruleDTtLamLabel}.
 
@@
             Then we infer $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$ by \ruleref{\ottdruleDTVarLetLabel}.
-
-            The pure size of the resulting tree $T_2$ and of $T_1$ is
-            one plus the sum of sizes of the premise trees.
-            The premises of $T_1$ and of $T_2$ have the same pure size by 
-            the induction hypothesis. This way, $\pureSize{T_1} = \pureSize{T_2}$.
 
@@
             It is left to show the uniqueness of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.
+            Let us assume that this judgement holds for other $[[iQ']]$, 
+            i.e.  there exists a tree $T_0$ inferring 
+            $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ']]$.
+            Then notice that the induction hypothesis is applicable to
+            $T_0$: the first component of the first component of $\metric{T_0}$
+            is $S = \sum_{[[v]] \in [[args]]} \size{[[v]]}$, and it is less than
+            the corresponding component of $\metric{T_1}$, which is
+            $\size{[[let x = v(args); c]]} = 1 + \size{[[v]]} + \size{[[c]]} + S$.
+            This way, $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ']]$ holds by the induction hypothesis,
+            but since $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ uniq]]$, we have $[[Γ ⊢ iQ' ≈ iQ]]$.
 
@@
 
-        \item \ruleref{\ottdruleDTUnpackLabel}, \ruleref{\ottdruleDTNAnnotLabel}, and \ruleref{\ottdruleDTNEquivLabel}
+        \item \ruleref{\ottdruleDTUnpackLabel}, and \ruleref{\ottdruleDTNAnnotLabel}
             are proved in the same way.
@@
             $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$ (inferred by \ruleref{\ottdruleDTEmptyAppLabel})
-            implies $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$,
-            and if there is only one $[[iN']]$ (up-to-equivalence in $[[Γ]]$) 
-            that satisfies $[[Γ ; Φ1 ⊢ iN ● · ⇒> iN']]$,
-            then there is only one $[[iN']]$ satisfying $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$.
+            implies $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN']]$.
 
@@
             $[[Γ ⊢ iN ≈ iN']]$ holds by assumption.
-
-            To show the uniqueness $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN' uniq]]$, 
-            let us assume $[[Γ ; Φ2 ⊢ iN ● · ⇒> iN'']]$.
-            Then the only rule applicable to infer it is \ruleref{\ottdruleDTEmptyAppLabel},
-            which, by inversion, means $[[Γ ⊢ iN ≈ iN'']]$, and by transitivity, 
-            $[[Γ ⊢ iN' ≈ iN'']]$.
 
@@
         \item \ruleref{\ottdruleDTForallAppLabel}
+            Then we are proving that
+            $[[Γ ; Φ1 ⊢ ∀pas.iN ● args ⇒> iM]]$ (inferred by \ruleref{\ottdruleDTForallAppLabel})
+            implies $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$.
+
+            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ : pas]]$ and
+            $[[Γ ; Φ1 ⊢ [σ]iN ● args ⇒> iM]]$ is inferred.
+            Let us denote the inference tree as $T_1'$.
+            Notice that the induction hypothesis is applicable to $T_1'$:
+            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ (for some $x$) is less than 
+            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ (for some $y$),
+            since $|[[pas]]| > 0$ by inversion.
+
+            This way, by the induction hypothesis, 
+            there exists a tree $T_2'$ inferring
+            $[[Γ ; Φ2 ⊢ [σ]iN ● args ⇒> iM]]$.
+            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ : pas]]$,
+            and $[[pas ≠ ·]]$ do not depend on the variable context,
+            and hold by inversion.
+            Then we infer $[[Γ ; Φ2 ⊢ ∀pas.iN ● args ⇒> iM]]$ by \ruleref{\ottdruleDTForallAppLabel}.
     \end{caseof}

### 2023-07-17 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: completeness of typing

@@
 \begin{lemma}[Completeness of the Negative Subtyping]
+    \label{lemma:neg-subt-completeness}
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,

### 2023-07-17 proofs/tex/constraint-merge-lemmas.tex
WIP: completeness of typing

@@
     Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
-    \begin{enumerate}
-        \item $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
-        \item $[[Θ ⊢ SC]]$, and
-        \item $[[Θ ⊢ uσ : SC]]$.
-    \end{enumerate}
+    $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
 \end{lemma}
@@
     This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that each of these components is defined.
-    To prove $[[Θ ⊢ uσ : SC]]$ we need to show that for every entry $[[scE]]$ of $[[SC]]$ restricting a variable $[[α̂±]]$,
-    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ holds.
 
-    It is clear that the first two components of this union exist, and that if $[[scE]]$ is an entry 
-    restricting $[[α̂±]]$ $[[dom(SC1) \ dom(SC2)]]$ then then $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$ because  
-    $[[Θ ⊢ uσ : SC1]]$ and $[[scE]] \in [[SC1]]$. Analogously, if $[[scE]]$ is an entry
-    from the second component of the union, the property holds for it.
-Let us show that the third component exists, and each of its entries satisfies the property.
+    It is clear that the first two components of this union exist. 
+    Let us show that the third component exists.  
     Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.

### 2023-07-17 proofs/tex/decl-typing-lemmas.tex
WIP: completeness of typing

@@
 
-\begin{definition}[Size of a Judgement]
+\begin{definition}[Size of a Declarative Judgement]
     \label{def:decl-typing-size}
@@
 \begin{definition}[Metric]
+    \label{def:decl-typing-metric}
     For a tree $T$ inferring
@@
 \begin{lemma}[Declarative typing is preserved under context equivalence]
+    \label{lemma:decl-typing-context-equiv}
     Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
@@
             Notice that the induction hypothesis is applicable to $T_1'$:
-            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ (for some $x$) is less than 
-            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ (for some $y$),
+            $\metric{T_1'} = ((\size{[[args]]}, 0), x)$ is less than 
+            $\metric{T_1} = ((\size{[[args]]}, |[[pas]]|), y)$ for any $x$ and $y$,
             since $|[[pas]]| > 0$ by inversion.
@@
     \end{caseof}
-
 \end{proof}

### 2023-07-27 proofs/tex/decl-subtyping-lemmas.tex
wip
Transitivity of ≤

@@
     $[[G ⊢ iP ≥ iQ]]$. Then by the induction hypothesis,
-    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired inclusion
+    $[[fv(iP)]] \subseteq [[fv(iQ)]]$. The desired 
     inclusion holds, since $[[fv(↑iP)]] = [[fv(iP)]]$ and
@@
 
+\begin{lemma}[Decomposition of quantifier rules]
+  \label{lemma:quant-rule-decomposition}
+  Assuming that $[[pas]]$, $[[pbs]]$, $[[nas]]$, and $[[nas]]$ are disjoint from $[[Γ]]$,
+  \begin{itemize}
+    \item [$-_{R}$] $[[Γ ⊢ iN ≤ ∀pbs.iM]]$ holds if and only if $[[Γ, pbs ⊢ iN ≤ iM]]$;
+    \item [$+_{R}$] $[[Γ ⊢ iP ≥ ∃nbs.iQ]]$ holds if and only if $[[Γ, nbs ⊢ iP ≥ iQ]]$;
+    \item [$-_{L}$] suppose $[[iM]] \neq [[∀]]\dots$
+      then $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$
+      for some $[[Γ ⊢ iPs]]$;
+    \item [$+_{L}$] suppose $[[iQ]] \neq [[∃]]\dots$
+      then $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ [iNs/nas]iP ≥ iQ]]$
+      for some $[[Γ ⊢ iNs]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  \hfill
+  \begin{itemize}
+    \item [$-_{R}$] Let us prove both directions. 
+      \begin{itemize}
+        \item [\Rightarrow] Let us assume $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
+          $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
+          Let us decompose $[[iM]]$ as $[[∀pbs'.iM']]$ where $[[iM']]$ does not start with $[[∀]]$, 
+          and decompose $[[iN]]$ as $[[∀pas.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
+          If $[[pbs]]$ is empty, then $[[Γ, pbs ⊢ iN ≤ iM]]$ holds by assumption.
+          Otherwise, $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.∀pbs'.iM]]$ is inferred by
+          \ruleref{\ottdruleDOneForallLabel}, and by inversion:
+          $[[Γ,pbs,pbs' ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ,pbs,pbs' ⊢ iPs]]$.
+          Then again by \ruleref{\ottdruleDOneForallLabel} with the same $[[iPs]]$,
+          $[[Γ,pbs ⊢ ∀pas.iN' ≤ ∀pbs'.iM']]$, that is $[[Γ,pbs ⊢ iN ≤ iM]]$.
+        \item [\Leftarrow] let us assume $[[Γ, pbs ⊢ iN ≤ iM]]$, and let us decompose 
+          $[[iN]]$ as $[[∀pas.iN']]$ where $[[iN']]$ does not start with $[[∀]]$, 
+          and $[[iM]]$ as $[[∀pbs'.iM']]$ where $[[iM']]$ does not start with $[[∀]]$.
+          if $[[pas]]$ and $[[pbs']]$ are empty then $[[Γ, pbs ⊢ iN ≤ iM]]$
+          is turned into $[[Γ ⊢ iN ≤ ∀pbs.iM]]$ by \ruleref{\ottdruleDOneForallLabel}.
+          Otherwise, $[[Γ, pbs ⊢ ∀pas.iN' ≤ ∀pbs'.iM']]$ is inferred by
+          \ruleref{\ottdruleDOneForallLabel}, that is $[[Γ, pbs, pbs' ⊢ [iPs/pas]iN' ≤ iM']]$
+          for some $[[Γ, pbs, pbs' ⊢ iPs]]$.
+          Then by \ruleref{\ottdruleDOneForallLabel} again,
+          $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs,pbs'.iM']]$, in other words, $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.∀pbs'.iM']]$, 
+          that is $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
+
+      \end{itemize}
+    \item [$-_{L}$] Suppose $[[iM]] \neq [[∀]]\dots$. Let us prove both directions.
+      \begin{itemize}
+        \item [\Rightarrow] Let us assume $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
+          then if $[[pas = ·]]$, $[[Γ ⊢ iN ≤ iM]]$ holds immediately.
+          Otherwise, let us decompose  $iN$ as $[[∀pas'.iN']]$ where 
+          $[[iN']]$ does not start with $[[∀]]$.
+          Then $[[Γ ⊢ ∀pas.∀pas'.iN' ≤ iM']]$ is inferred by
+          \ruleref{\ottdruleDOneForallLabel},
+          and by inversion, 
+          there exist $[[Γ ⊢ iPs,iPs']]$ 
+          such that $[[Γ ⊢ [iPs/pas][iPs'/pas']iN' ≤ iM']]$ 
+          (the decomposition of substitutions is possible since $[[{pas} ∩ {Γ} = ∅]]$).
+          Then by \ruleref{\ottdruleDOneForallLabel} again,
+          $[[Γ ⊢ ∀pas'.[iPs'/pas']iN' ≤ iM']]$ (notice that $[[ [iPs'/pas']iN' ]]$ cannot
+          start with $[[∀]]$).
+        \item [\Leftarrow] Let us assume 
+          $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$ for some $[[Γ ⊢ iPs]]$.
+          let us decompose $iN$ as $[[∀pas'.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
+          Then $[[Γ ⊢ [iPs/pas]∀pas'.iN' ≤ iM']]$ or, equivalently,
+          $[[Γ ⊢ ∀pas'.[iPs/pas]iN' ≤ iM']]$ is inferred by \ruleref{\ottdruleDOneForallLabel}
+          (notice that $[[ [iPs/pas]iN' ]]$ cannot start with $[[∀]]$).
+          By inversion, there exist $[[Γ ⊢ iPs']]$ such that 
+          $[[Γ ⊢ [iPs'/pas'][iPs/pas]iN' ≤ iM']]$. Since $[[pas']]$ is disjoint
+          from the free variables of $[[iPs]]$ and from $[[pas]]$, the composition of 
+          $[[iPs'/pas']]$ and $[[iPs/pas]]$ can be joined into a single substitution
+          well-formed in $[[Γ]]$. Then by \ruleref{\ottdruleDOneForallLabel} again,
+          $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
+      \end{itemize}
+      \item [$+$] The positive cases are proved symmetrically.
+  \end{itemize}
+\end{proof}
+
+\begin{corollary}[Redundant quantifier elimination]
+  \label{corollary:red-quant-elim}
+  \hfill
+  \begin{itemize}
+    \item [$-_{L}$] Suppose that $[[ {pas} ∩ fv(iN) = ∅]]$ then 
+      $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
+    \item [$-_{R}$] Suppose that $[[ {pas} ∩ fv(iM) = ∅]]$ then 
+      $[[Γ ⊢ iN ≤ ∀pas.iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
+    \item [$+_{L}$] Suppose that $[[ {nas} ∩ fv(iP) = ∅]]$ then
+      $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
+    \item [$+_{R}$] Suppose that $[[ {nas} ∩ fv(iQ) = ∅]]$ then 
+      $[[Γ ⊢ iP ≥ ∃nas.iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  \begin{itemize}
+    \item [$-_{R}$] Suppose that $[[ {pas} ∩ fv(iM) = ∅]]$ then 
+      by \cref{lemma:quant-rule-decomposition},
+      $[[Γ ⊢ iN ≤ ∀pas.iM]]$ 
+      is equivalent to $[[Γ, pas ⊢ iN ≤ iM]]$,
+      By \label{lemma:wf-ctxt-equiv},
+      since $[[{pas} ∩ fv(iN) = ∅]]$ and $[[{pas} ∩ fv(iM) = ∅]]$,
+      $[[Γ, pas ⊢ iN ≤ iM]]$ is equivalent to $[[Γ ⊢ iN ≤ iM]]$.
+
+    \item [$-_{L}$] Suppose that $[[ {pas} ∩ fv(iN) = ∅]]$.
+      Let us decompose $[[iM]]$ as $[[∀pbs.iM']]$ 
+      where $[[iM']]$ does not start with $[[∀]]$.
+      By \cref{lemma:quant-rule-decomposition},
+      $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM']]$ is equivalent to
+      $[[Γ,pbs ⊢ ∀pas.iN ≤ iM']]$, 
+      which is equivalent to 
+      existence of $[[Γ,pbs ⊢ iPs]]$ such that 
+      $[[Γ,pbs ⊢ [iPs/pas]iN ≤ iM']]$.
+      Since $[[ [iPs/pas]iN  = iN]]$, the latter is equivalent to 
+      $[[Γ,pbs ⊢ iN ≤ iM']]$,
+      which is equivalent to $[[Γ ⊢ iN ≤ ∀pbs.iM']]$.
+      $[[Γ,pbs ⊢ iPs]]$ can be chosen arbitrary, for example, $[[iPsi]] = [[∃α⁻.↓α⁻]]$.
+    \item [$+$] The positive cases are proved symmetrically.
+  \end{itemize}
+\end{proof}
+
 \begin{lemma}[Subtypes and supertypes of a variable]
@@
   \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ iP ≥ α⁺]]$ or $[[Γ ⊢ α⁺ ≥ iP ]]$ then $[[iP]] = [[∃nas.α⁺]]$ (for some potentially empty $[[nas]]$)
-  \item[$-$] if $[[Γ ⊢ iN ≤ α⁻]]$ or $[[Γ ⊢ α⁻ ≤ iN ]]$ then $[[iN]] = [[∀pas.α⁻]]$ (for some potentially empty $[[pas]]$)
+  \item[$+$] if $[[Γ ⊢ iP ≥ ∃nas.α⁺]]$ or $[[Γ ⊢ ∃nas.α⁺ ≥ iP ]]$ then $[[iP]] = [[∃nbs.α⁺]]$ (for some potentially empty $[[nbs]]$)
+  \item[$-$] if $[[Γ ⊢ iN ≤ ∀pas.α⁻]]$ or $[[Γ ⊢ ∀pas.α⁻ ≤ iN ]]$ then $[[iN]] = [[∀pbs.α⁻]]$ (for some potentially empty $[[pbs]]$)
   \end{itemize}
@@
   We prove by induction on the tree
-  inferring $[[Γ ⊢ iP ≥ α⁺]]$ or $[[Γ ⊢ α⁺ ≥ iP ]]$ or
-  or $[[Γ ⊢ iN ≤ α⁻]]$ or $[[Γ ⊢ α⁻ ≤ iN ]]$.
+  inferring $[[Γ ⊢ iP ≥ ∃nas.α⁺]]$ or $[[Γ ⊢ ∃nas.α⁺ ≥ iP ]]$ or
+  or $[[Γ ⊢ iN ≤ ∀pas.α⁻]]$ or $[[Γ ⊢ ∀pas.α⁻ ≤ iN ]]$.
 
-  Let us consider which of these judgments the tree is inferring.
+  Let us consider which one of these judgments is inferred.
   \begin{caseof}
-  \item $[[Γ ⊢ iP ≥ α⁺]]$\\
+  \item $[[Γ ⊢ iP ≥ ∃nas.α⁺]]$\\
     If the size of the inference tree is $1$ then the only rule that can infer
     it is \ruleref{\ottdruleDOnePVarLabel}, which
-    implies that $[[iP = α⁺]]$.
+    implies that $[[nas]]$ is empty and $[[iP = α⁺]]$.
 
@@
     it must be \ruleref{\ottdruleDOneExistsLabel}. By inverting this rule,
-    $[[iP = ∃nas.iP']]$ where $[[iP']]$ does not start with $\exists$ and
-    $[[G ⊢ [iNs/nas] iP' ≥ α⁺]]$ for some $[[G, nbs ⊢ iNi]]$.
+    $[[iP = ∃nbs.iP']]$ where $[[iP']]$ does not start with $\exists$ and
+    $[[Γ, nas ⊢ [iNs/nbs] iP' ≥ α⁺]]$ for some $[[G, nas ⊢ iNi]]$.
 
-    By the induction hypothesis, $[[ [iNs/nas] iP' = ∃nbs.α⁺]]$.
-    Notice that $[[iP']]$ must be a variable, because
-    $[[iP']]$ does not start with $\exists$, nor does it start with
+    By the induction hypothesis, $[[ [iNs/nbs] iP' = ∃ncs.α⁺]]$.
+    What shape can $[[iP']]$ have?
+    As mentioned, it does not start with $\exists$, and it cannot start with
     $\uparrow$ (otherwise, $[[ [iNs/nas] iP' ]]$ would also
-    started with $\uparrow$ and would not be equal to $[[∃nbs.α⁺]]$).
-    Since $[[iP']]$ is a \emph{positive} variable, $[[ [iNs/nas] iP' = iP']]$,
-    and then $[[iP' = ∃nbs.α⁺]]$ means that $[[iP' = α⁺]]$.
-    This way, $[[iP]] = [[∃nas.iP']] = [[∃nas.α⁺]]$, as required.
+    start with $\uparrow$ and would not be equal to $[[∃nbs.α⁺]]$).
+    This way, $[[iP']]$ is a \emph{positive} variable. 
+    As such, $[[ [iNs/nas] iP' = iP']]$,
+    and then $[[iP' = ∃ncs.α⁺]]$ meaning that $[[ncs]]$ is empty and $[[iP' = α⁺]]$.
+    This way, $[[iP]] = [[∃nbs.iP']] = [[∃nbs.α⁺]]$, as required.
 
-  \item $[[Γ ⊢ α⁺ ≥ iP]]$\\
+  \item $[[Γ ⊢ ∃nas.α⁺ ≥ iP]]$\\
     If the size of the inference tree is $1$ then the only rule that can infer
     it is \ruleref{\ottdruleDOnePVarLabel}, which
-    implies that $[[iP = α⁺]]$.
+    implies that $[[nas]]$ is empty and $[[iP = α⁺]]$.
 
@@
     it must be \ruleref{\ottdruleDOneExistsLabel}. By inverting this rule,
-    $[[iP = ∃nbs.iQ]]$ where and $[[G, nbs ⊢ α⁺ ≥ iQ]]$.
+    $[[iP = ∃nbs.iQ]]$ where $[[G, nbs ⊢ [iNs/nas]α⁺ ≥ iQ]]$ and $[[iQ]]$ 
+    does not start with $\exists$.
+    Notice that since $[[α⁺]]$ is positive, $[[ [iNs/nas]α⁺ = α⁺]]$, 
+    i.e. $[[G, nbs ⊢ α⁺ ≥ iQ]]$.
 
-    By the induction hypothesis, $[[iQ = ∃nbs'.α⁺]]$.
-    This way, $[[iP]] = [[∃nbs.iQ]] = [[∃nbs.∃nbs'.α⁺]]$, as required.
+    By the induction hypothesis, $[[iQ = ∃nbs'.α⁺]]$,
+    and since $[[iQ]]$ does not start with $\exists$, $[[nbs']]$ is empty
+    This way, $[[iP]] = [[∃nbs.iQ]] = [[∃nbs.α⁺]]$, as required.
 
-  \item The negative cases ($[[Γ ⊢ iN ≤ α⁻]]$ and $[[Γ ⊢ α⁻ ≤ iN ]]$)
+  \item The negative cases ($[[Γ ⊢ iN ≤ ∀pas.α⁻]]$ and $[[Γ ⊢ ∀pas.α⁻ ≤ iN ]]$)
     are proved analogously.
@@
 \begin{proof}
-  Notice that $[[Γ ⊢ ∃nas.α⁺ ≈ α⁺]]$ and $[[∃nas.α⁺ ≈ α⁺]]$ and apply
+  Notice that $[[Γ ⊢ ∃nbs.α⁺ ≈ α⁺]]$ and $[[∃nbs.α⁺ ≈ α⁺]]$ and apply
   \cref{lemma:var-subt}.
-  \ilyam{fix}
 \end{proof}
@@
   \label{lemma:subst-pres-subt}
-  Assuming that all mentioned types are well-formed in $[[Γ]]$,
-  and $[[Γ' ⊢ σ : Γ]]$, where $[[Γ']]$ is disjoint from $[[Γ]]$,
+  Suppose that all mentioned types are well-formed in $[[Γ1 , Γ]]$,
+  and $[[σ]]$ is a substitution $[[Γ2 ⊢ σ : Γ1]]$.
   \begin{itemize}
-    \item $-$ If $[[Γ ⊢ iN ≤ iM]]$ then $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$
-    \item $+$ If $[[Γ ⊢ iP ≥ iQ]]$ then $[[Γ' ⊢ [σ]iP ≥ [σ]iQ]]$
+    \item $-$ If $[[Γ1 , Γ ⊢ iN ≤ iM]]$ then $[[Γ2 , Γ ⊢ [σ]iN ≤ [σ]iM]]$
+    \item $+$ If $[[Γ1 , Γ ⊢ iP ≥ iQ]]$ then $[[Γ2 , Γ ⊢ [σ]iP ≥ [σ]iQ]]$
   \end{itemize}
@@
       (\cref{lemma:subtyping-reflexivity}),
-      we have $[[Γ' ⊢ [σ]α⁻ ≤ [σ]α⁻]]$, i.e. $[[Γ' ⊢ [σ]iN ≤ [σ]iM]]$,
+      we have $[[Γ2, Γ ⊢ [σ]α⁻ ≤ [σ]α⁻]]$, i.e. $[[Γ2, Γ ⊢ [σ]iN ≤ [σ]iM]]$,
       as required.
@@
       $[[iN = ∀pas.iN']]$, $[[iM = ∀pbs.iM']]$, where $[[pas]]$ or $[[pbs]]$ is not empty.
-      Moreover, $[[Γ,pbs ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ, pbs ⊢ iPs]]$, and 
+      Moreover, $[[Γ1 , Γ, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ1 , Γ, pbs ⊢ iPs]]$, and 
       $[[fv iN ∩ {pbs} = ∅ ]]$.
 
-      Notice that since the derivation of $[[Γ,pbs ⊢ [iPs/pas]iN' ≤ iM']]$ is
+      Notice that since the derivation of $[[Γ1, Γ, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ is
       a subderivation of the derivation of $[[Γ ⊢ iN ≤ iM]]$, its size is smaller, 
       and hence, the induction hypothesis applies:
-      $[[Γ', pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$.
+      $[[Γ2 , Γ, pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$.
 
-      First, let us assume that $[[{pas} ∩ {Γ'} = ∅]]$ and $[[{pbs} ∩ {Γ'} = ∅]]$
-      (otherwise, we rename $[[pas]]$ and $[[pbs]]$ to fresh $[[pas']]$ and $[[pbs']]$).
-      Then $[[ [σ]∀pas.iN' ]] = [[ ∀pas.[σ]iN' ]]$ and $[[ [σ]∀pbs.iM' ]] = [[ ∀pbs.[σ]iM' ]]$, 
-      which means that the required $[[Γ' ⊢ [σ]∀pas.iN' ≤ [σ]∀pbs.iM']]$ is rewritten as
-      $[[Γ' ⊢ ∀pas.[σ]iN' ≤ ∀pbs.[σ]iM']]$.
+      Notice that by convention, $[[pas]]$ and $[[pbs]]$ are fresh, and thus,  
+      $[[ [σ]∀pas.iN' ]] = [[ ∀pas.[σ]iN' ]]$ and $[[ [σ]∀pbs.iM' ]] = [[ ∀pbs.[σ]iM' ]]$, 
+      which means that the required $[[Γ2, Γ ⊢ [σ]∀pas.iN' ≤ [σ]∀pbs.iM']]$ is rewritten as
+      $[[Γ2 , Γ ⊢ ∀pas.[σ]iN' ≤ ∀pbs.[σ]iM']]$.
 
@@
       \begin{itemize}
-        \item $[[fv iN ∩ {pbs} = ∅ ]]$ as noted before, from the inversion;
-        \item $[[Γ', pbs ⊢ [σ]iPi]]$, by \cref{lemma:wf-subst} since from the inversion,
-        $[[Γ, pbs ⊢ iPi]]$;
-        \item $[[Γ, pbs ⊢ [ [σ]iPs/pas ][σ]iN' ≤ [σ]iM']]$ holds because 
-        $[[ [ [σ]iPs/pas ][σ]iN' ]] = [[ [σ][iPs/pas]iN ]]$ (since $[[{pas} ∩ {Γ} = ∅]]$), 
-        and $[[Γ', pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$ holds by the induction hypothesis.
+        \item $[[fv [σ]iN ∩ {pbs} = ∅ ]]$;
+        \item $[[Γ2, Γ,pbs⊢ [σ]iPi]]$, by \cref{lemma:wf-subst} since from the inversion,
+          $[[Γ1, Γ, pbs ⊢ iPi]]$;
+        \item $[[Γ, pbs ⊢ [ [σ]iPs/pas ][σ]iN' ≤ [σ]iM']]$ holds
+          by \cref{lemma:subst-composition}:
+          Since $[[pas]]$ is fresh, it is disjoint with the domain and the codomain of $[[σ]]$
+          ($[[Γ1]]$ and $[[Γ2]]$), and thus, 
+          $[[ [σ][iPs/pas]iN' ]] = [[ [ σ <=< iPs/pas ][σ]iN' ]] = [[ [ [σ]iPs/pas ][σ]iN' ]]$.
+          Then $[[Γ2, Γ, pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$ holds by the induction hypothesis.
       \end{itemize}
+
     \item \ruleref{\ottdruleDOneArrowLabel}. Then by inversion,
@@
 
+\begin{corollary}[Substitution preserves subtyping induced equivalence]
+  \label{corollary:subst-pres-equiv}
+  Suppose that $[[Γ ⊢ σ : Γ1]]$. Then
+    \begin{itemize}
+      \item[$+$] if $[[Γ1 ⊢ iP]]$,~ $[[Γ1 ⊢ iQ]]$,~ and $[[Γ1 ⊢ iP ≈ iQ]]$ ~ 
+        then $[[Γ ⊢ [σ]iP ≈ [σ]iQ]]$
+      \item[$-$] if $[[Γ1 ⊢ iN]]$,~ $[[Γ1 ⊢ iM]]$,~ and $[[Γ1 ⊢ iN ≈ iM]]$ ~ 
+        then $[[Γ ⊢ [σ]iN ≈ [σ]iM]]$
+    \end{itemize}
+\end{corollary}
+
 \begin{lemma}[Strong transitivity of subtyping] \label{lemma:subtyping-transitivity}
@@
 
+  First, let us consider the 3 important cases.
   \begin{caseof}
-    \item Firs, let us consider the case when the last rule applied to infer 
-    $[[Γ ⊢ iN ≤ iM1]]$ is \ruleref{\ottdruleDOneNVarLabel}.
-    Notice that this case covers the base of the induction: 
-    the sum of the depths is minimal when both derivations are inferred by
-    the non-recursive rules (i.e. \ruleref{\ottdruleDOneNVarLabel}).
+    \item Let us consider the case when $[[iM1 = ∀pbs1.α⁻]]$. 
+      Then by \cref{lemma:var-subt},
+       $[[Γ ⊢ iN ≤ iM1]]$ means that $[[iN = ∀pas.α⁻]]$. 
+      $[[ [σ]iM1 = [σ]iM2 ]]$ means that $[[ ∀pbs1.[σ]α⁻ = [σ]iM2 ]]$.
+      Applying $[[σ]]$ to both sides of $[[Γ ⊢ iM2 ≤ iK]]$ (by \cref{lemma:subst-pres-subt}),
+      we obtain $[[Γ' ⊢ [σ]iM2 ≤ [σ]iK]]$, that is $[[Γ' ⊢  ∀pbs1.[σ]α⁻ ≤ [σ]iK]]$.
+      Since $[[ fv([σ]α⁻) ⊆ {Γ,α⁻} ]]$, it is disjoint from $[[pas]]$ and $[[pbs1]]$,
+      This way, by \cref{corollary:red-quant-elim}, 
+      $[[Γ' ⊢  ∀pbs1.[σ]α⁻ ≤ [σ]iK]]$ is equivalent to 
+      $[[Γ' ⊢  [σ]α⁻ ≤ [σ]iK]]$, which is equivalent to $[[Γ' ⊢  ∀pas.[σ]α⁻ ≤ [σ]iK]]$,
+      that is $[[Γ' ⊢  [σ]iN ≤ [σ]iK]]$.
+    \item Let us consider the case when $[[iM2 = ∀pbs2.α⁻]]$.
+      This case is symmetric to the previous one. Notice that 
+      \cref{lemma:var-subt,corollary:red-quant-elim} are agnostic to the 
+      side on which the the quantifiers occur, and thus, 
+      the proof stays the same. 
+    \item Let us decompose the types, by extracting the outer quantifiers:
+      \begin{itemize}
+        \item $[[iN = ∀pas.iN']]$, where $[[iN']] \neq [[∀]]\dots$,
+        \item $[[iM1 = ∀pbs1.iM1']]$, where $[[iM1']] \neq [[∀]]\dots$,
+        \item $[[iM2 = ∀pbs2.iM2']]$, where $[[iM2']] \neq [[∀]]\dots$,
+        \item $[[iK = ∀pcs.iK']]$, where $[[iK']] \neq [[∀]]\dots$.
+      \end{itemize}
+      and assume that at least one of $[[pas]]$, $[[pbs1]]$, $[[pbs2]]$, and $[[pcs]]$ is not empty.
+      Since $[[ [σ]iM1 = [σ]iM2 ]]$, we have $[[ ∀pbs1.[σ]iM1' = ∀pbs2.[σ]iM2' ]]$,
+      and since $[[iMi']]$ are not variables 
+      (which was covered by the previous cases) and do not start with $\forall$,
+      $[[ [σ]iMi' ]]$ do not start with $\forall$ either,
+      which means $[[pbs1]] = [[pbs2]]$ and $[[ [σ]iM1' = [σ]iM2' ]]$.
+      Let us rename $[[pbs1]]$ and $[[pbs2]]$ to $[[pbs]]$.
+      Then $[[iM1 = ∀pbs.iM1']]$ and $[[iM2 = ∀pbs.iM2']]$.
 
-    By inverting the rule, $[[iN = α⁻]]$ and $[[iM1 = α⁻]]$
-    Then $[[ [σ]iN ]] = [[ [σ]α⁻ ]] = [[ [σ]iM1 ]] = [[ [σ]iM2 ]]$. 
-    And $[[Γ' ⊢ [σ]iM2 ≤ [σ]iK]]$ by \cref{}
+      By \cref{lemma:quant-rule-decomposition} applied twice
+      to $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.iM1']]$ and to $[[Γ ⊢ ∀pbs.iM2' ≤ ∀pcs.iK']]$,
+      we have the following:
+      \begin{enumerate}
+        \item $[[Γ, pbs ⊢ [iPs/pas]iN' ≤ iM1']]$ for some $[[Γ, pbs ⊢ iPs]]$;
+        \item $[[Γ, pcs ⊢ [iQs/pbs]iM2' ≤ iK']]$ for some $[[Γ, pcs ⊢ iQs]]$.
+      \end{enumerate}
+      And since at least one of 
+      $[[pas]]$, $[[pbs]]$, and $[[pcs]]$ is not empty,
+      either $[[Γ ⊢ iN ≤ iM1]]$ or $[[Γ ⊢ iM2 ≤ iK]]$ is inferred 
+      by \ruleref{\ottdruleDOneForallLabel}, meaning that either 
+      $[[Γ, pbs ⊢ [iPs/pas]iN' ≤ iM1']]$ is a proper subderivation of $[[Γ ⊢ iN ≤ iM1]]$ or
+      $[[Γ, pcs ⊢ [iQs/pbs]iM2' ≤ iK']]$ is a proper subderivation of $[[Γ ⊢ iM2 ≤ iK]]$.
 
+      Notice that we can weaken and rearrange the contexts without changing the depths of the 
+      derivations: $[[Γ, pbs, pcs ⊢ [iPs/pas]iN' ≤ iM1']]$
+      and $[[Γ, pbs, pcs ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
+      the sum of the depths of these derivations is smaller than the sum of the depths of
+      $[[Γ ⊢ iN ≤ iM1]]$ and $[[Γ ⊢ iM2 ≤ iK]]$.
+      Let us apply the induction hypothesis to these derivations, 
+      with the substitution $[[ Γ', pcs ⊢ σ ○ (iQs/pbs) : Γ, pbs, pcs  ]]$
+      (\cref{lemma:subst-domain-weakening}).
+      To apply the induction hypothesis, it is left to show that 
+      $[[ σ ○ (iQs/pbs) ]]$ unifies $[[iM1']]$ and $[[ [iQs/pbs]iM2']]$:
+      $$
+      \begin{aligned}[t]
+        [[ [σ ○ iQs/pbs]iM1' ]] &= [[ [σ][iQs/pbs]iM1' ]]\\
+                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
+                                && \text{by \cref{lemma:subst-composition}}\\
+                                &= [[ [ [σ]iQs/pbs ][σ]iM2' ]]
+                                && \text{Since $[[ [σ]iM1' = [σ]iM2' ]]$}\\
+                                &= [[  [σ][iQs/pbs]iM2' ]]
+                                && \text{by \cref{lemma:subst-composition}}\\
+                                &= [[  [σ][iQs/pbs][iQs/pbs]iM2' ]]
+                                && \text{Since $[[Γ, pcs ⊢ iQs]]$, and $[[{(Γ, pcs)} ∩ {pbs} = ∅]]$ }\\
+                                &= [[  [σ ○ iQs/pbs][iQs/pbs]iM2' ]]
+      \end{aligned}
+      $$
+      This way the induction hypothesis gives us
+      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ][iQs/pbs]iK' ]]$,
+      and since $[[Γ, pcs ⊢ iK']]$, $[[ [iQs/pbs]iK' = iK' ]]$, that is 
+      $[[ Γ', pcs ⊢ [σ][iQs/pbs][iPs/pas]iN' ≤  [σ]iK' ]]$.
+      Let us rewrite the substitution that we apply to $[[iN']]$:
+      $$
+      \begin{aligned}[t]
+        [[ [σ ○ iQs/pbs ○ iPs/pas]iN' ]] &= [[ [ (σ <=< iQs/pbs) ○ σ ○ iPs/pas]iN' ]]
+                                       && \text{by \cref{lemma:subst-composition}}\\
+                                       &= [[ [(σ <=< iQs/pbs) ○ (σ <=< iPs/pas) ○ σ] iN' ]]
+                                       && \text{by \cref{lemma:subst-composition}}\\
+                                       &= [[ [(((σ <=< iQs/pbs) ○ σ) <=< iPs/pas) ○ σ] iN' ]]
+                                       && \text{Since $[[fv([σ]iN') ∩ {pbs} = ∅]]$}\\
+                                       &= [[ [((σ ○ iQs/pbs) <=< iPs/pas) ○ σ] iN' ]]
+                                       && \text{by \cref{lemma:subst-composition}}\\
+                                       &= [[ [(σ ○ iQs/pbs) <=< iPs/pas][σ] iN' ]]
+      \end{aligned}
+      $$
+      Notice that $[[(σ ○ iQs/pbs) <=< iPs/pas]]$
+      is a substitution that turns $[[pai]]$ into [[ [σ ○ iQs/pbs]iPi ]], 
+      where $[[ Γ',pcs ⊢ [σ ○ iQs/pbs]iPi]]$.
+      This way, 
+      $[[ Γ', pcs ⊢ [(σ ○ iQs/pbs) <=< iPs/pas][σ]iN' ≤  [σ]iK' ]]$
+      means $[[Γ ⊢ ∀pas.[σ]iN' ≤ ∀pcs.[σ]iK']]$
+      by \cref{lemma:quant-rule-decomposition}, that is
+      $[[Γ ⊢ [σ]iN ≤ [σ]iK]]$, as required.
   \end{caseof}
 
+  Now, we can assume that neither $[[Γ ⊢ iN ≤ iM1]]$ nor $[[Γ ⊢ iM2 ≤ iK]]$ 
+  is inferred by \ruleref{\ottdruleDOneForallLabel}, and that neither $[[iM1]]$ nor $[[iM2]]$
+  is equivalent to a variable.  Because of that, $[[ [σ]iM1 = [σ]iM2 ]]$ means that 
+  $[[iM1]]$ and $[[iM2]]$ have the same outer constructor. Let us consider the shape of $[[iM1]]$.
+
+  \begin{caseof}
+    \item $[[iM1 = α⁻]]$ this case has been considered;
+    \item $[[iM1 = ∀pbs.iM1']]$ this case has been considered;
+    \item $[[iM1 = ↑iQ1]]$. Then as noted above, 
+      $[[ [σ]iM1 = [σ]iM2 ]]$ means that $[[iM2 = ↑iQ2]]$ and $[[ [σ]iQ1 = [σ]iQ2 ]]$.
+      Moreover, $[[Γ ⊢ iN ≤ ↑iQ1]]$ can only be inferred by \ruleref{\ottdruleDOneShiftULabel},
+      and thus, $[[iN = ↑iP]]$, and by inversion, $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iQ1 ≥ iP]]$.
+      Analogously, $[[Γ ⊢ ↑iQ2 ≤ iK]]$ means that $[[iK = ↑iR]]$, $[[Γ ⊢ iQ2 ≥ iR]]$, and $[[Γ ⊢ iR ≥ iQ2]]$.
+
+      Notice that the derivations of $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iQ1 ≥ iP]]$ are proper sub-derivations of 
+      $[[Γ ⊢ iN ≤ iM1]]$, and the derivations of $[[Γ ⊢ iQ2 ≥ iR]]$ and $[[Γ ⊢ iR ≥ iQ2]]$ are proper sub-derivations of
+      $[[Γ ⊢ iM2 ≤ iK]]$. This way, the induction hypothesis is applicable:
+      \begin{itemize}
+        \item applying the induction hypothesis to $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iQ2 ≥ iR]]$ 
+          with $[[Γ' ⊢ σ : Γ]]$ unifying $[[iQ1]]$ and $[[iQ2]]$, we obtain $[[Γ' ⊢ [σ]iP ≥ [σ]iR]]$;
+        \item applying the induction hypothesis to $[[Γ ⊢ iR ≥ iQ2]]$ and $[[Γ ⊢ iQ1 ≥ iP]]$ 
+          with $[[Γ' ⊢ σ : Γ]]$ unifying $[[iQ2]]$ and $[[iQ1]]$, we obtain $[[Γ' ⊢ [σ]iR ≥ [σ]iP]]$.
+      \end{itemize}
+      This way, by \ruleref{\ottdruleDOneShiftULabel}, $[[Γ' ⊢ [σ]iN ≤ [σ]iK]]$, as required.
+
+    \item $[[iM1 = iQ1 → iM1']]$. Then as noted above, 
+      $[[ [σ]iM1 = [σ]iM2 ]]$ means that $[[iM2 = iQ2 → iM2']]$, $[[ [σ]iQ1 = [σ]iQ2 ]]$, and $[[ [σ]iM1' = [σ]iM2' ]]$.
+      Moreover, $[[Γ ⊢ iN ≤ iQ1 → iM1']]$ can only be inferred by \ruleref{\ottdruleDOneArrowLabel},
+      and thus, $[[iN = iP → iN']]$, and by inversion, $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iN' ≤ iM1']]$.
+      Analogously, $[[Γ ⊢ iQ2 → iM2' ≤ iK]]$ means that $[[iK = iR → iK']]$, $[[Γ ⊢ iQ2 ≥ iR]]$, and $[[Γ ⊢ iM2' ≤ iK']]$.
+
+      Notice that the derivations of $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iN' ≤ iM1']]$ are proper sub-derivations of
+      $[[Γ ⊢ iP → iN' ≤ iQ1 → iM1']]$, and the derivations of $[[Γ ⊢ iQ2 ≥ iR]]$ and $[[Γ ⊢ iM2' ≤ iK']]$ are proper sub-derivations of
+      $[[Γ ⊢ iQ2 → iM2' ≤ iR → iK']]$. This way, the induction hypothesis is applicable:
+      \begin{itemize}
+        \item applying the induction hypothesis to $[[Γ ⊢ iP ≥ iQ1]]$ and $[[Γ ⊢ iQ2 ≥ iR]]$ 
+          with $[[Γ' ⊢ σ : Γ]]$ unifying $[[iQ1]]$ and $[[iQ2]]$, we obtain $[[Γ' ⊢ [σ]iP ≥ [σ]iR]]$;
+        \item applying the induction hypothesis to $[[Γ ⊢ iN' ≤ iM1']]$ and $[[Γ ⊢ iM2' ≤ iK']]$ 
+          with $[[Γ' ⊢ σ : Γ]]$ unifying $[[iM1']]$ and $[[iM2']]$, we obtain $[[Γ' ⊢ [σ]iN' ≤ [σ]iK']]$.
+      \end{itemize}
+      This way, by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ]iP → [σ]iN' ≤ [σ]iR → [σ]iK']]$,
+      that is $[[Γ' ⊢ [σ]iN ≤ [σ]iK]]$, as required.
+  \end{caseof}
+
+  After that we consider all the 
+  analogous positive cases, and prove them symmetrically.
 \end{proof}

### 2023-07-27 proofs/tex/equiv-lemmas.tex
WIP: completeness of typing
Typing correctness
Transitivity of ≤

@@
 \end{proof}
+
 \begin{lemma}[Algorithmization of declarative equivalence]
@@
 
-\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] \label{corollary:nf-sound-wrt-subt-equiv}
+\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] 
+  \label{corollary:nf-sound-wrt-subt-equiv}
   \hfill
@@
 \end{proof}
+
+\begin{lemma}[Substitution preserves declarative equivalence]
+  \label{corollary:subst-pres-equiv}
+  Suppose that $[[σ]]$ is a substitution. Then
+    \begin{itemize}
+      \item[$+$] $[[iP ≈ iQ]]$ implies $[[ [σ]iP ≈ [σ]iQ ]]$
+      \item[$-$] $[[iN ≈ iM]]$ implies $[[ [σ]iN ≈ [σ]iM ]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+
+  $
+  \begin{aligned}[t] 
+    [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
+                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
+                  &\Rightarrow [[ [nf(σ)]nf(iP) = [nf(σ)]nf(iQ)]]\\
+                  &\Rightarrow [[ nf([σ]iP) = nf([σ]iQ)]]
+                  && \text{by \cref{lemma:norm-subst-distr}}\\ 
+                  &\Rightarrow        [[ [σ]iP ≈ [σ]iQ ]]
+                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
+  \end{aligned} 
+  $
+\end{proof}

### 2023-07-27 proofs/tex/subst-lemmas.tex
wip
Transitivity of ≤

@@
 \end{proof}
-
-\begin{corollary}[Substitution preserves equivalence]
-  \label{corollary:subst-pres-equiv}
-
-  Suppose that $[[Γ ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-  \item[$+$] if $[[Γ1 ⊢ iP]]$,~ $[[Γ1 ⊢ iQ]]$,~ and $[[Γ1 ⊢ iP ≈ iQ]]$ ~ then $[[Γ ⊢ [σ]iP ≈ [σ]iQ]]$
-  \item[$-$] if $[[Γ1 ⊢ iN]]$,~ $[[Γ1 ⊢ iM]]$,~ and $[[Γ1 ⊢ iN ≈ iM]]$ ~ then $[[Γ ⊢ [σ]iN ≈ [σ]iM]]$
-  \end{itemize}
-\end{corollary}
 
@@
 \end{proof}
+
+\begin{lemma}[Substitutions equivalent on the metavariables]
+  \label{lemma:subst-equiv-metavar}
+  Suppose that $[[Γ ⊢ Θ]]$, $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
+  of signature $[[Θ ⊢ uσi]]$.
+  Then 
+  \begin{enumerate}
+    \item [$+$] for a type $[[Γ; Θ ⊢ uP]]$, if $[[Γ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
+      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
+    \item [$-$] for a type $[[Γ; Θ ⊢ uN]]$, if $[[Γ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
+      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  The proof is a trivial structural induction on 
+  $[[Γ; Θ ⊢ uP]]$ and mutually, on $[[Γ; Θ ⊢ uN]]$.
+\end{proof}
+
+\begin{lemma}[Substitution composition well-formedness]
+  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
+  then $[[Γ'1, Γ'2 ⊢ σ2 ○ σ1 : Γ1, Γ2]]$.
+\end{lemma}
+
+\begin{lemma}[Substitution monadic composition well-formedness]
+  \label{lemma:subst-monad-composition-wf}
+  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
+  then $[[Γ'2 ⊢ σ2 <=< σ1 : Γ1]]$.
+\end{lemma}
+
+\begin{lemma}[Substitution composition]
+  \label{lemma:subst-composition}
+    If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
+    $[[{Γ1} ∩ {Γ'2} = ∅ ]]$ and $[[ {Γ1} ∩ {Γ2} = ∅ ]]$ then 
+    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
+\end{lemma}
+
+\begin{corollary}[Substitution composition commutativity]
+  \label{corollary:subst-composition-commutativity}
+  If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
+  $[[ {Γ1} ∩ {Γ2} = ∅ ]]$, 
+  $[[ {Γ1} ∩ {Γ'2} = ∅ ]]$, and
+  $[[ {Γ'1} ∩ {Γ2} = ∅ ]]$ then 
+  $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
+\end{corollary}
+\begin{proof}
+  by \cref{lemma:subst-composition},
+    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
+    Since the codomain of $[[σ1]]$ is $[[Γ'1]]$,
+    and it is disjoint with the domain of $[[σ2]]$,
+    $[[σ2 <=< σ1]] = [[σ1]]$.
+\end{proof}
+
+\begin{lemma}[Substitution domain weakening]
+  \label{lemma:subst-domain-weakening}
+  If $[[Γ2 ⊢ σ : Γ1]]$ and $[[ {Γ2'} ⊆ {Γ} ]]$ then $[[Γ2 ⊢ σ : Γ1, Γ2']]$
+\end{lemma}
+\begin{proof}
+  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption.
+  If the variable $[[α±]]$ is in $[[{Γ2'} \ {Γ1}]]$ then $[[ [σ]α± = α± ]] \in [[Γ2']] ⊆ [[Γ2]]$, 
+  and thus, $[[Γ2 ⊢ α± ]]$.
+\end{proof}

### 2023-07-31 proofs/main.tex
wip
wip: leftovers

@@
 
-\section{The Vanilla System}
+\tableofcontents
 
+\newpage
+
+\section{Declarative System}
 First, we present the top-level system, which is easy to understand.
@@
 \ottgrammartabular{
-  \ottP\ottinterrule
-  \ottN\ottinterrule
+  \ottiP\ottinterrule
+  \ottiN\ottinterrule
   \ottv\ottinterrule
@@
 \subsection{Declarative Subtyping}
-\ottdefnsDZero
-
-\section{Multi-Quantified System}
-\subsection{Grammar}
-\ottgrammartabular{
-  \ottiP\ottinterrule
-  \ottiN\ottinterrule
-}
-\subsection{Declarative Multiquantified Subtyping}
 \ottdefnsDOne
@@
 
-\subsection{Constraint Merge}
+\subsection{Constraints}
 
-Unification and subtyping constraints is are by a list of constraint entries.
-Each entry restricts an unification variable in two possible ways: either
-stating that it must be equivalent to a certain type ($[[pua :≈ iP]]$ or $[[nua
-:≈ iN]]$) or that it must be a (positive) supertype of a certain type ($[[pua :≥ iP]]$).
+Unification and subtyping algorithms are based on the constraint generation.
+The constraints are represented by set of constraint entries.
+
+\begin{definition}[Unification Constraint]
+  \hfill
+  \begin{itemize}
+    \item Unification entry (denoted as $[[ucE]]$) is an expression of shape 
+      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
+    \item unification constraint (denoted as $[[UC]]$) is a set of 
+      unification constraint entries.
+  \end{itemize}
+
+\end{definition}
+
+\begin{definition}[Subtyping Constraint]
+  \hfill
+  \begin{itemize}
+    \item Subtyping entry (denoted as $[[scE]]$) is an expression of shape 
+      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
+    \item subtyping constraint (denoted as $[[SC]]$) is a set of subtyping constraint entries.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Well-formed Constraint Entry]
+  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
+  the type it restricts the unification variable to is well-formed in $[[Γ]]$.
+  \begin{itemize}
+    \item $[[Γ ⊢ pua :≥ iP]]$ iff $[[Γ ⊢ iP]]$;
+    \item $[[Γ ⊢ pua :≈ iP]]$ iff $[[Γ ⊢ iP]]$;
+    \item $[[Γ ⊢ nua :≈ iN]]$ iff $[[Γ ⊢ iN]]$.
+  \end{itemize}
+\end{definition}
 
 \begin{definition} [Matching Entries]
-  We call two entries matching if they are restricting the same unification variable.
+  We call two unification constraint entries 
+  or two subtyping constraint entries matching 
+  if they are restricting the same unification variable.
 \end{definition}
 
-Two matching entries can be merged in the following way:
-\begin{definition} \hfill \\
+Two matching entries formed in the same context $[[Γ]]$ 
+can be merged in the following way:
+\begin{definition}[Merge of Matching Constraint Entries]
+   \hfill \\
 \ottdefnSCME\\
@@
 
-To merge two constraints, we merge each pair of
-matching entries, and unite the results.
-\ilyam{add contexts}
-\begin{definition}
-  $\begin{aligned}[t] [[SC1 & SC2]] &= 
-    \{ [[scE1 & scE2]] \mid [[scE1]] \in [[SC1]], [[scE2]]  \in [[SC2]], \text{s.t. } [[scE1]] \text{ matches with } [[scE2]] \}\\
-    &\cup 
-    \{ [[scE1]] \mid [[scE1]] \in [[SC1]], \text{ s.t. } \forall [[scE2]]  \in [[SC2]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
-    &\cup 
-    \{ [[scE2]] \mid [[scE2]] \in [[SC2]], \text{ s.t. } \forall [[scE1]]  \in [[SC1]], [[scE1]] \text{ does not match with } [[scE2]] \}\\
-   \end{aligned}$
+\begin{definition}[Constraint Context]
+  Constraint context (denoted as $[[Θ]]$) is a set of entries of shape $[[ pua[Γ] ]]$ 
+  and $[[ nua[Γ] ]]$, specifying that the unification variable ($[[pua]]$ or $[[nua]]$)
+  must be instantiated in a type well-formed in context $[[Γ]]$.
+  We assume that for each $[[α̂±]]$, at most one entry of shape $[[ α̂±[Γ] ]]$ is in $[[Θ]]$.
+  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
+\end{definition}
+
+\begin{definition}[Well-formed Constraint]
+  We say that a constraint is well-formed in a
+  constraint context $[[Θ]]$ if all its entries are well-formed in
+  the corresponding elements of $[[Θ]]$.
+  More formally, 
+  $[[Θ ⊢ UC]]$ iff for every $[[ucE]] \in $ $[[UC]]$,
+  such that $[[ucE]]$ restricts $[[α̂±]]$,
+  there exists $[[ α̂±[Γ] ]] \in [[Θ]]$
+  and $[[Γ ⊢ ucE]]$.
+\end{definition}
+
+\begin{definition}[Merge of Subtyping Constraints]
+  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
+  Then $[[Θ ⊢ SC1 & SC2 = SC]]$
+  defines a set such that $[[ucE]] \in [[SC]]$ iff either
+  \begin{itemize}
+    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
+    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
+    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
+      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
+      $[[α̂±]]$. 
+  \end{itemize}
 \end{definition}
@@
 
-\subsection{Typing}
+\subsection{Declarative Typing}
+\input{../_build/_gen/tex/decl-typing-lemmas.tex}
+
+\subsection{Algorithmic Typing}
 \input{../_build/_gen/tex/alg-typing-lemmas.tex}

### 2023-07-31 proofs/tex/alg-pos-subtyping-lemmas.tex
wip: leftovers

@@
     Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
-    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
+    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
 \end{lemma}
@@
     \begin{caseof}
-        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ (for potentially empty $[[nbs]]$)\\
+        \item \label{case:pos-subt-complete-base} $[[uP]] = [[ ∃nbs.α̂⁺ ]]$ 
+            (for potentially empty $[[nbs]]$)\\
             Then by assumption, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ (where $[[ {nbs} ∩ fv [uσ]α̂⁺ = ∅]]$).
-        By \cref{todo}, $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ iQ ]]$ means 
-        $[[ Γ ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ ]]$, and hence $[[ Γ ⊢ [uσ]α̂⁺ ≥ iQ ]]$.
-        By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some $[[Δ]]$.
+            Let us decompose $[[iQ]]$ as $[[iQ]] = [[∃ncs.iQ0]]$, where $[[iQ0]]$ does
+            not start with $[[∃]]$.
 
-        In the algorithm trying to infer the subtyping 
-        $[[Γ; Θ ⊨ ∃nbs.α̂⁺ ≥ iQ ⫤ SC]]$,
-        after multiple applications of \ruleref{\ottdruleAExistsLabel}
-        the type $[[∃nbs.α̂⁺]]$ is reduced to $[[α̂⁺]]$.
+            By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some 
+            $[[{Δ} ⊆ {Γ}]]$.
+
+            By \cref{lemma:quant-rule-decomposition} applied twice, 
+            $[[ Γ ⊢ ∃nbs.[uσ]α̂⁺ ≥ ∃ncs.iQ0 ]]$ implies
+            $[[ Γ,ncs  ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ0 ]]$ for some $[[iN]]$, 
+            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ {Γ} = ∅ ]]$,
+            $[[ [iNs/nbs][uσ]α̂⁺ = [uσ]α̂⁺ ]]$, that is $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$.
+
+            When algorithm tires to infer the subtyping 
+            $[[Γ; Θ ⊨ ∃nbs.α̂⁺ ≥ ∃ncs.iQ0 ⫤ SC]]$,
+            it applies \ruleref{\ottdruleAExistsLabel},
+            which reduces the problem to
+            $[[Γ, ncs; Θ, β̂⁻*[Γ, ncs] ⊨ [β̂⁻*/nbs] α̂⁺ ≥ iQ0 ⫤ SC]]$, 
+            which is equivalent to 
+            $[[Γ, ncs; Θ, β̂⁻*[Γ, ncs] ⊨ α̂⁺ ≥ iQ0 ⫤ SC]]$.
+
             Next, the algorithm tries to apply
             \ruleref{\ottdruleAPUVarLabel}
-        and the resulting restriction is $[[SC']] = [[(α̂⁺ :≥ iQ')]]$ where
-        $[[upgrade Γ ⊢ iQ to Δ = iQ']]$.
+            and the resulting restriction is $[[SC]] = [[(α̂⁺ :≥ iQ0')]]$ where
+            $[[upgrade Γ, ncs ⊢ iQ0 to Δ = iQ0']]$.
 
-        Why does the upgrade procedure terminates?
+            Why does the upgrade procedure terminate?
             Because $[[ [uσ]α̂⁺ ]]$ satisfies the pre-conditions of the completeness of the upgrade
             (\cref{lemma:upgrade-completeness}):
-        \begin{itemize}
-            \item $[[Δ ⊢ iP']]$ because $[[iP' = [uσ]α̂⁺]]$ and $[[Θ ⊢ uσ]]$ and 
-            $[[α̂⁺[Δ] ∊ Θ]]$,
-            \item $[[Γ ⊢ iP' ≥ iQ]]$ as noted above
-        \end{itemize}
-        Moreover, the completeness of the upgrade also gives us $[[Γ ⊢ iP' ≥ iQ']]$
-        and further, we strengthen it to $[[Δ ⊢ iP' ≥ iQ']]$
-        (since by the soundness of the upgrade (\cref{lemma:upgrade-soundness}),
-        $[[Δ ⊢ iQ']]$).  It means that $[[Δ ⊢ iP' : (α̂⁺ :≥ iQ') ]]$, that is 
-        $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ') ]]$, as required.
+            \begin{enumerate}
+                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
+                \item $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$ as noted above
+            \end{enumerate}
+
+            Moreover, the completeness of upgrade also says that $[[iQ0']]$ is 
+            \emph{the least} supertype of $[[iQ0]]$ among types well-formed in $[[Δ]]$, 
+            that is $[[Δ ⊢ [uσ]α̂⁺ ≥ iQ0']]$, which means 
+            $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ0')]]$, that is $[[Θ ⊢ uσ : SC]]$.
 
         \item \label{case:pos-subt-complete-pvar}
-        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}, 
-        i.e. $[[iP]] = [[ [uσ]uP ]] = [[ α⁺ ]] = [[iQ]]$.
-        Here the first equality holds because $[[uP]]$ is not a unification variable:
-        it has been covered by \cref{case:pos-subt-complete-base}.
-        The second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
+        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOnePVarLabel}\\
+        Then $[[iP]] = [[ [uσ]uP ]] = [[ α⁺ ]] = [[iQ]]$, where
+        the first equality holds because $[[uP]]$ is not a unification variable:
+        it has been covered by \cref{case:pos-subt-complete-base}; and
+        the second equality hold because \ruleref{\ottdruleDOnePVarLabel} was applied.
 

### 2023-07-31 proofs/tex/decl-subtyping-lemmas.tex
wip: leftovers

@@
       \begin{itemize}
-        \item [\Rightarrow] Let us assume $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
+        \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
           $[[Γ ⊢ iN ≤ ∀pbs.iM]]$.
@@
           $[[Γ,pbs ⊢ ∀pas.iN' ≤ ∀pbs'.iM']]$, that is $[[Γ,pbs ⊢ iN ≤ iM]]$.
-        \item [\Leftarrow] let us assume $[[Γ, pbs ⊢ iN ≤ iM]]$, and let us decompose 
+        \item [$\Leftarrow$] let us assume $[[Γ, pbs ⊢ iN ≤ iM]]$, and let us decompose 
           $[[iN]]$ as $[[∀pas.iN']]$ where $[[iN']]$ does not start with $[[∀]]$, 
@@
       \begin{itemize}
-        \item [\Rightarrow] Let us assume $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
+        \item [$\Rightarrow$] Let us assume $[[Γ ⊢ ∀pas.iN ≤ iM]]$.
           then if $[[pas = ·]]$, $[[Γ ⊢ iN ≤ iM]]$ holds immediately.
@@
           start with $[[∀]]$).
-        \item [\Leftarrow] Let us assume 
+        \item [$\Leftarrow$] Let us assume 
           $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$ for some $[[Γ ⊢ iPs]]$.
@@
   \end{itemize}
-\end{lemma}
+\end{corollary}
 \begin{proof}
@@
   \label{lemma:subst-pres-subt}
-  Suppose that all mentioned types are well-formed in $[[Γ1 , Γ]]$,
+  Suppose that all mentioned types are well-formed in $[[Γ1]]$,
   and $[[σ]]$ is a substitution $[[Γ2 ⊢ σ : Γ1]]$.
   \begin{itemize}
-    \item $-$ If $[[Γ1 , Γ ⊢ iN ≤ iM]]$ then $[[Γ2 , Γ ⊢ [σ]iN ≤ [σ]iM]]$
-    \item $+$ If $[[Γ1 , Γ ⊢ iP ≥ iQ]]$ then $[[Γ2 , Γ ⊢ [σ]iP ≥ [σ]iQ]]$
+    \item $-$ If $[[Γ1 ⊢ iN ≤ iM]]$ then $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$
+    \item $+$ If $[[Γ1 ⊢ iP ≥ iQ]]$ then $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$
   \end{itemize}
@@
 \begin{proof}
-  We prove it by induction on the size of the derivation of $[[Γ ⊢ iN ≤ iM]]$
-  and mutually, $[[Γ ⊢ iP ≥ iQ]]$. Let us consider the last rule 
+  We prove it by induction on the size of the derivation of $[[Γ1 ⊢ iN ≤ iM]]$
+  and mutually, $[[Γ1 ⊢ iP ≥ iQ]]$. Let us consider the last rule 
   used in the derivation:
@@
       (\cref{lemma:subtyping-reflexivity}),
-      we have $[[Γ2, Γ ⊢ [σ]α⁻ ≤ [σ]α⁻]]$, i.e. $[[Γ2, Γ ⊢ [σ]iN ≤ [σ]iM]]$,
+      we have $[[Γ2 ⊢ [σ]α⁻ ≤ [σ]α⁻]]$, i.e. $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$,
       as required.
@@
       $[[iN = ∀pas.iN']]$, $[[iM = ∀pbs.iM']]$, where $[[pas]]$ or $[[pbs]]$ is not empty.
-      Moreover, $[[Γ1 , Γ, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ1 , Γ, pbs ⊢ iPs]]$, and 
+      Moreover, $[[Γ1, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ1, pbs ⊢ iPs]]$, and 
       $[[fv iN ∩ {pbs} = ∅ ]]$.
 
-      Notice that since the derivation of $[[Γ1, Γ, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ is
+      Notice that since the derivation of $[[Γ1, pbs ⊢ [iPs/pas]iN' ≤ iM']]$ is
       a subderivation of the derivation of $[[Γ ⊢ iN ≤ iM]]$, its size is smaller, 
-      and hence, the induction hypothesis applies:
-      $[[Γ2 , Γ, pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$.
+      and hence, the induction hypothesis applies
+      ($[[Γ1, pbs ⊢ σ : Γ1, pbs]]$ by  \label{lemma:subst-domain-weakening})
+      :
+      $[[Γ2, pbs ⊢ [σ][iPs/pas]iN' ≤ [σ]iM']]$.
 
@@
       Notice that $[[(σ ○ iQs/pbs) <=< iPs/pas]]$
-      is a substitution that turns $[[pai]]$ into [[ [σ ○ iQs/pbs]iPi ]], 
+      is a substitution that turns $[[pai]]$ into $[[ [σ ○ iQs/pbs]iPi ]]$, 
       where $[[ Γ',pcs ⊢ [σ ○ iQs/pbs]iPi]]$.

### 2023-07-31 proofs/tex/subst-lemmas.tex
wip: leftovers

@@
   \label{lemma:subst-domain-weakening}
-  If $[[Γ2 ⊢ σ : Γ1]]$ and $[[ {Γ2'} ⊆ {Γ} ]]$ then $[[Γ2 ⊢ σ : Γ1, Γ2']]$
+  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2, Γ' ⊢ σ : Γ1, Γ']]$
 \end{lemma}
 \begin{proof}
-  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption.
-  If the variable $[[α±]]$ is in $[[{Γ2'} \ {Γ1}]]$ then $[[ [σ]α± = α± ]] \in [[Γ2']] ⊆ [[Γ2]]$, 
-  and thus, $[[Γ2 ⊢ α± ]]$.
+  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption,
+  and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
+  If the variable $[[α±]]$ is in $[[{Γ'} \ {Γ1}]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
+  and thus, $[[Γ2, Γ' ⊢ α± ]]$.
 \end{proof}

### 2023-07-31 proofs/tex/upgrade-lemmas.tex
wip: leftovers

@@
     both sides of $[[Δ, pnbs, pncs ⊢ iQ ≥ [pnbs/pnas]iP]]$ and 
-    by \cref{todo}, get $[[Δ, pnas, pncs ⊢ [pnas/pnbs]iQ ≥ iP]]$.
+    by \cref{lemma:subst-pres-subt} 
+    (since $[[pnbs/pnas]]$ can be specified as 
+    $[[Δ,pnbs,pncs ⊢ pnbs/pnas : Δ, pnas, pncs]]$ by \label{lemma:subst-domain-weakening})
+    obtain $[[Δ, pnas, pncs ⊢ [pnas/pnbs]iQ ≥ iP]]$.
     Notice that $[[Δ ⊢ iQ]]$ implies that $[[fv iQ ∩ {pnbs} = ∅]]$, 

### 2023-08-08 proofs/main.tex
wip: leftovers

@@
 \newcommand{\depth}[1]{\mathsf{depth}(#1)}
+\newcommand{\size}[1]{\mathsf{size}(#1)}
 
@@
 \subsection{Grammar}
+We assume that there is an infinite set of positive and 
+negative \emph{type} variables. Positive type variables are denoted as 
+$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
+Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
+We assume there is an infinite set of \emph{term} variables,
+which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
+A list of objects (variables, types or terms) is denoted by
+an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
+$[[nbs]]$ is a list of negative type variables, 
+$[[args]]$ is a list of values, which are arguments of a function.
+$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
+in a type $[[iP]]$ and $[[iN]]$, respectively.
+
+\bigskip
+
 \ottgrammartabular{
@@
 
+\subsection{Well-Formedness}
+\ottdefnsWFT
+\ottdefnsWFAT
+
 \section{Algorithm}
+
+\subsection{Grammar}
+
+\subsubsection{Variables and Types}
+In the algorithmic system, we extend the grammar of types
+by adding positive and negative \emph{algorithmic variables}
+($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
+They represent the unknown types, which will be inferred by the algorithm.
+This way, we add two base cases to the grammar of 
+positive and negative types, and use highlight to denote that the type
+can potentially contain algorithmic variables.
+
+\begin{definition}[Algorithmic Types]
+  \hfill \\
+\ottgrammartabular{
+  \ottuP\ottinterrule
+  \ottuN\ottinterrule
+}
+\end{definition}
+
+\subsubsection{Contexts}
+
+Unannotated set of algorithmic variables is denoted as $[[Ξ]]$,
+and it is used in the anti-unification algorithm.
+$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
+that the algorithmic type is well-formed in the contexts
+$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
+of the type is contained in $[[Ξ]]$, and each free non-algorithmic type variable
+of the type is contained in $[[Γ]]$.
+
+Typically each algorithmic variable is associated with a
+context it must be instantiated in 
+(i.e. the context in which the type replacing the variable must be well-formed).
+This association is represented by an \emph{algorithmic variable context} $[[Θ]]$.
+\begin{definition}[Algorithmic Context]
+  \hfill \\
+  Algorithmic Context $[[Θ]]$ is represented by a set of 
+  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
+  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
+  and $[[Γ]]$ is a context in which they must be instantiated.
+  We assume that no two entries associating the same variable
+  appear in $[[Θ]]$.
+
+  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
+  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
+\end{definition}
+
+\subsubsection{Subsitutions}
+
+Substitution that operates on algorithmic type variables is denoted as
+$[[uσ]]$. It is defined as a total function from algorithmic 
+type variables to non-algorithmic types, preserving the polarity.
+
+The signature $[[Θ ⊢ uσ]]$ means that $[[uσ]]$ maps each algorithmic 
+variable appearing in $[[Θ]]$ to a type well-formed in the corresponding
+context; and for each variable not appearing in $[[dom(Θ)]]$, acts as identity.
+
+\begin{definition}[Signature of Algorithmic Substitution]
+  $[[Θ ⊢ uσ]]$ means that
+  \begin{enumerate}
+    \item for any $[[ α̂±[Γ] ∊ Θ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$ and
+    \item for any $[[ α̂± ∉ dom(Θ)]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
+  \end{enumerate}
+\end{definition}
+
+Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.
+In contrast to algorithmic substitution $[[uσ]]$,
+it only defined on the negative algorithmic variables, 
+and it allows mapping algorithmic variables to
+\emph{algorithmic} terms.
+
+The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
+in which the results of the anti-unification substitutions 
+are formed, is fixed for the whole substitution.
+This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
+variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.
+
+\begin{definition}[Signature of Anti-unification substitution]
+  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
+  \begin{enumerate}
+    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
+    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
+  \end{enumerate}
+\end{definition}
 
@@
 \subsection{Substitution}
+
+\begin{definition}[Substitution]
+  Substitutions (denoted as $[[σ]]$) 
+  are represented by total functions form variables to types, preserving the polarity. 
+\end{definition}
+
+\begin{definition}[Substitution Application]
+  Substitution application (denoted as $[[ [σ]iP ]]$ or $[[ [σ]iN ]]$) 
+  is defined congruently as follows:
+  \begin{itemize}
+    \item $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
+    \item $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
+    \item $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
+    \item $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
+    \item $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
+    \item $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$(here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
+      that is the variable capture never happens);
+    \item $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Substitution Signature]
+  The signature $[[Γ' ⊢ σ : Γ]]$ means that
+  \begin{enumerate}
+    \item for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ]α± ]]$; and
+    \item for any $[[α± ∉ {Γ'}]], [[ [σ]α± = α± ]]$.
+  \end{enumerate}
+\end{definition}
+
+A substitution can be restricted to a set of variables. 
+The restricted substitution is define as expected. 
+\begin{definition}[Subsitution Restriction]
+  The specification $[[σ  | varset]]$ is defined as
+  a function such that 
+  \begin{enumerate}
+    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
+    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
+  \end{enumerate}
+\end{definition}
+
+Two substitutions can be composed in two ways:
+$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
+while $[[σ2 <=< σ1]]$
+depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
+$[[σ2]]$ to its results on the domain.
+\begin{definition}[Substitution Composition]
+  $[[σ2 ○ σ1]]$ is defined as a function such that
+  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
+\end{definition}
+
+\begin{definition}[Monadic Substitution Composition]
+  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
+  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
+\end{definition}
+
+\begin{definition}[Equivalent Substitutions]
+  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
+  indicates that on the domain $[[Γ]]$, 
+  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
+  Formally, for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
+\end{definition}
+
 \input{../_build/_gen/tex/subst-lemmas.tex}

### 2023-08-08 proofs/tex/decl-subtyping-lemmas.tex
wip: leftovers

@@
 
-\begin{lemma}[Strong transitivity of subtyping] \label{lemma:subtyping-transitivity}
-  Assuming all the types are well-formed in $[[Γ]]$,
+\begin{lemma}[Transitivity of subtyping] 
+  \label{lemma:subtyping-transitivity}
+  Assuming the types are well-formed in $[[Γ]]$,
   \begin{itemize}
-    \item $-$ if $[[Γ ⊢ iN ≤ iM1]]$, $[[Γ ⊢ iM2 ≤ iK]]$, and for 
-      $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iM1 = [σ]iM2 ]]$ then $[[Γ' ⊢ [σ]iN ≤ [σ]iK]]$
-    \item $+$ if $[[Γ ⊢ iP ≥ iQ1]]$, $[[Γ ⊢ iQ2 ≥ iR]]$, and for
-      $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iQ1 = [σ]iQ2 ]]$ then $[[Γ' ⊢ [σ]iP ≥ [σ]iR]]$
+    \item[$-$] if $[[Γ ⊢ iN1 ≤ iN2]]$ and $[[Γ ⊢ iN2 ≤ iN3]]$ then $[[Γ ⊢ iN1 ≤ iN3]]$,
+    \item[$+$] if $[[Γ ⊢ iP1 ≥ iP2]]$ and $[[Γ ⊢ iP2 ≥ iP3]]$ then $[[Γ ⊢ iP1 ≥ iP3]]$.
   \end{itemize}
@@
 \begin{proof}
-  We prove it by induction on $\depth{[[Γ ⊢ iN ≤ iM1]]} + \depth{[[Γ ⊢ iM2 ≤ iK]]}$ and mutually, 
-  on $\depth{[[Γ ⊢ iP ≥ iQ1]]} + \depth{[[Γ ⊢ iQ2 ≥ iR]]}$.
+  To prove it, we formulate a stronger property, 
+  which will imply the required one, taking $[[σ]] = [[Γ ⊢ id : Γ]]$.
+
+    Assuming all the types are well-formed in $[[Γ]]$,
+    \begin{itemize}
+      \item[$-$] if $[[Γ ⊢ iN ≤ iM1]]$, $[[Γ ⊢ iM2 ≤ iK]]$, and for 
+        $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iM1 = [σ]iM2 ]]$ then $[[Γ' ⊢ [σ]iN ≤ [σ]iK]]$
+      \item[$+$] if $[[Γ ⊢ iP ≥ iQ1]]$, $[[Γ ⊢ iQ2 ≥ iR]]$, and for
+        $[[Γ' ⊢ σ : Γ]]$, $[[ [σ]iQ1 = [σ]iQ2 ]]$ then $[[Γ' ⊢ [σ]iP ≥ [σ]iR]]$
+    \end{itemize}
+
+  We prove it by induction on $\size{[[Γ ⊢ iN ≤ iM1]]} + \size{[[Γ ⊢ iM2 ≤ iK]]}$ and mutually, 
+  on $\size{[[Γ ⊢ iP ≥ iQ1]]} + \size{[[Γ ⊢ iQ2 ≥ iR]]}$.
 
@@
 
-      Notice that we can weaken and rearrange the contexts without changing the depths of the 
+      Notice that we can weaken and rearrange the contexts without changing the sizes of the 
       derivations: $[[Γ, pbs, pcs ⊢ [iPs/pas]iN' ≤ iM1']]$
       and $[[Γ, pbs, pcs ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
-      the sum of the depths of these derivations is smaller than the sum of the depths of
+      the sum of the sizes of these derivations is smaller than the sum of the sizes of
       $[[Γ ⊢ iN ≤ iM1]]$ and $[[Γ ⊢ iM2 ≤ iK]]$.
@@
   \end{caseof}
-
   After that we consider all the 
@@
 \end{proof}
-
-\begin{corollary}[Transitivity of subtyping] \label{corollary:subtyping-transitivity}
-  Assuming the types are well-formed in $[[Γ]]$,
-  \begin{itemize}
-    \item[$-$] if $[[Γ ⊢ iN1 ≤ iN2]]$ and $[[Γ ⊢ iN2 ≤ iN3]]$ then $[[Γ ⊢ iN1 ≤ iN3]]$,
-    \item[$+$] if $[[Γ ⊢ iP1 ≥ iP2]]$ and $[[Γ ⊢ iP2 ≥ iP3]]$ then $[[Γ ⊢ iP1 ≥ iP3]]$.
-  \end{itemize}
-\end{corollary}
 

### 2023-08-08 proofs/tex/equiv-lemmas.tex
wip: leftovers

@@
+\begin{lemma} [Declarative Equivalence is invariant under bijections]
+  \label{lemma:decl-equiv-bijection}
+  Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
+  \begin{itemize}
+    \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
+      and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
+      with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
+    \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
+      and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
+      with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on $[[iP1 ≈ iP2]]$ and mutually, on $[[iN1 ≈ iN2]]$.
+  Let us consider the last rule used in the derivation.
+  \begin{caseof}
+    \item \ruleref{\ottdruleEOneForallLabel}\\
+      Then we decompose $[[iN1]]$ as $[[∀pas1.iM1]]$ and $[[iN2]]$ as $[[∀pas2.iM2]]$,
+      where $[[iM1]]$ and $[[iM2]]$ do not start with $[[∀]]$-quantifiers.
+      where $[[|pas1| + |pas2| > 0]]$.
+      By convention, let us assume that
+      $[[pas1]]$ and $[[pas2]]$ are disjoint form 
+      $[[varset2]]$ and $[[varset1]]$.
+
+      By inversion, $[[{pas1} ∩ fv iM2 = ∅]]$ and 
+      $[[iM1 ≈ [μ'] iM2]]$ for some bijection
+      $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$.
+      Then let us apply the induction hypothesis to
+       $[[iM1 ≈ [μ'] iM2]]$ to obtain $[[ [μ]iM1 ≈ [μ] [μ'] iM2 ]]$ 
+      inferred by the tree of the same shape as 
+      $[[ iM1 ≈ [μ'] iM2 ]]$.
+
+      Notice that $[[ [μ]iM1 ]]$ and $[[ [μ]iM2 ]]$ 
+      do not start with $[[∀]]$, 
+      That is $[[ [μ]∀pas1.iM1 ≈ [μ]∀pas2.iM2]]$, 
+      rewritten as $[[ ∀pas1.[μ]iM1 ≈ ∀pas2.[μ]iM2]]$,
+      can be inferred by \ruleref{\ottdruleEOneForallLabel}:
+      \begin{enumerate}
+        \item $[[{pas1}]]$ is disjoint from 
+        $[[varset2 ∪ fv iM2 ⊆ fv [μ]iM2 ]]$;
+        \item $[[ [μ]iM1 ≈ [μ'] [μ] iM2 ]]$
+          because $[[ [μ'] [μ] iM2 = [μ] [μ'] iM2 ]]$
+          (by \cref{corollary:subst-composition-commutativity}:
+          $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$, 
+          $[[μ : varset1 ↔ varset2]]$,
+          $[[varset1]]$ is disjoint from 
+          $[[pas2]]$ and $[[pas1]]$; 
+          $[[pas2]]$ is disjoint from $[[varset1]]$ and $[[varset2]]$)
+      \end{enumerate}
+      Notice that it is the same rule as the one inferring 
+      $[[iN1 ≈ iN2]]$, and thus, the shapes of the trees are the same.
+
+    \item \ruleref{\ottdruleEOneNVarLabel}\\
+      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and the required
+      $[[ [μ]α⁻ = [μ]α⁻ ]]$ is also inferred by
+      \ruleref{\ottdruleEOneNVarLabel}, since 
+      $[[ [μ]α⁻ ]]$ is a variable.
+    \item \ruleref{\ottdruleEOneArrowLabel}\\
+      Then we are proving 
+      that $[[iP1 → iM1 ≈ iP2 → iM2]]$ implies
+      $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$ 
+      (preserving the tree structure).
+
+      By inversion, we have $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$,
+      and thus, by the induction hypothesis,
+      $[[ [μ]iP1 ≈ [μ]iP2 ]]$ and $[[ [μ]iM1 ≈ [μ]iM2 ]]$.
+      Then $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$, 
+      or in other words, $[[ [μ]iP1 → [μ]iM1 ≈ [μ]iP2 → [μ]iM2 ]]$,
+      is inferred by the same rule---\ruleref{\ottdruleEOneArrowLabel}.
+    \item \ruleref{\ottdruleEOneShiftULabel}
+      This case is done by similar congruent arguments as
+      the previous one.
+    \item The positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}
+
 \begin{lemma}[Declarative equivalence is transitive]
@@
 \begin{proof}
-  \ilyam{should be easy to do by induction since the types are getting smaller}
+  We prove it by
+  $\size{[[iP1 ≈ iP2]]} + \size{[[iP2 ≈ iP3]]}$
+  and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where
+  by size we mean the size of the nodes in the corresponding inference tree.
+
+  \begin{caseof}
+    \item First, let us consider the case when either 
+      $[[iN1 ≈ iN2]]$ or $[[iN2 ≈ iN3]]$ is inferred by 
+      \ruleref{\ottdruleEOneForallLabel}.
+      Let us decompose $[[iN1]]$, $[[iN2]]$, and $[[iN3]]$ as follows:
+      $[[iN1 = ∀pas1.iM1]]$,
+      $[[iN2 = ∀pas2.iM2]]$, and
+      $[[iN3 = ∀pas3.iM3]]$.
+
+      Then by inversion of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
+      (or if $[[pas1]]$ and $[[pas2]]$ are both empty, 
+      by assumption) : 
+      \begin{enumerate}
+        \item $[[{pas1} ∩ fv iM2 = ∅]]$ and
+        \item there exists a bijection on variables 
+          $[[mu1 : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$
+          such that $[[iM1 ≈ [mu1] iM2]]$.
+      \end{enumerate}
+
+      Analogously, $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
+      implies:
+      \begin{enumerate}
+        \item $[[{pas2} ∩ fv iM3 = ∅]]$ and
+        \item $[[iM2 ≈ [mu2] iM3]]$ for some bijection 
+          $[[mu2 : ({pas3} ∩ fv iM3) ↔ ({pas2} ∩ fv iM2)]]$.
+      \end{enumerate}
+
+      Notice that either $[[iM1 ≈ [mu1] iM2]]$
+      is inferred by a proper sub-tree of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
+      or $[[iM2 ≈ [mu2] iM3]]$ is inferred by a proper sub-tree of 
+      $[[∀pas2.iM2 ≈ ∀pas3.iM3]]$.
+
+      Then by \cref{lemma:decl-equiv-bijection}, 
+      $[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]$ and moreover, 
+      $\size{[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]} = \size{[[iM2 ≈ [mu2] iM3]]}$.
+
+      Since at least one of the trees inferring 
+      $[[iM1 ≈ [mu1] iM2]]$ and $[[iM2 ≈ [mu2] iM3]]$
+      is a proper sub-tree of the corresponding original tree,
+      $\size{[[iM1 ≈ [mu1] iM2]]} + \size{[[iM2 ≈ [mu2] iM3]]} <
+      \size{[[∀pas1.iM1 ≈ ∀pas2.iM2]]} + \size{[[∀pas2.iM2 ≈ ∀pas3.iM3]]}$,
+      i.e., the induction hypothesis is applicable.
+
+      By the induction hypothesis, $[[iM1 ≈ [mu1 ○ mu2] iM3]]$.
+      Where $[[mu1 ○ mu2]]$ is a bijection on variables
+      $[[mu1 ○ mu2 : ({pas3} ∩ fv iM3) ↔ ({pas1} ∩ fv iM1)]]$.
+      Then $[[∀pas1.iM1 ≈ ∀pas3.iM3]]$ by \ruleref{\ottdruleEOneForallLabel}.
+
+      Once this case has been considered, 
+      we can assume that neither $[[iN1 ≈ iN2]]$ nor $[[iN2 ≈ iN3]]$
+      is inferred by \ruleref{\ottdruleEOneForallLabel}.
+
+    \item $[[iN1 ≈ iN2]]$ is inferred by 
+      \ruleref{\ottdruleEOneNVarLabel}\\
+      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and thus,  
+      $[[iN1 ≈ iN3]]$ holds since $[[iN2 ≈ iN3]]$.
+
+    \item $[[iN1 ≈ iN2]]$ is inferred by 
+      \ruleref{\ottdruleEOneArrowLabel}\\
+      Then $[[iN1]] = [[iP1 → iM1]]$ and $[[iN2]] = [[iP2 → iM2]]$,
+      and by inversion, $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$.
+
+      Moreover, since $[[iN3]]$ does not start with $[[∀]]$,
+      $[[iN2 ≈ iN3]]$ is also inferred by 
+      \ruleref{\ottdruleEOneArrowLabel},
+      which means that $[[iN3]] = [[iP3 → iM3]]$,
+      $[[iP2 ≈ iP3]]$, and $[[iM2 ≈ iM3]]$.
+
+      Then by the induction hypothesis, 
+      $[[iP1 ≈ iP3]]$ and $[[iM1 ≈ iM3]]$, and thus, 
+      $[[iP1 → iM1 ≈ iP3 → iM3]]$ by \ruleref{\ottdruleEOneArrowLabel}.
+    \item $[[iN1 ≈ iN2]]$ is inferred by
+      \ruleref{\ottdruleEOneArrowLabel}\\
+      For this case, the reasoning is the same as for the previous one.
+    \item The positive cases are proved symmetrically.
+  \end{caseof}
 \end{proof}
@@
 \begin{proof}
-  \ilyam{todo}
+  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
+  Let us consider the last rule used in the derivation.
+  \begin{caseof}
+    \item \ruleref{\ottdruleEOneNVarLabel}, 
+      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
+      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
+      which holds trivially.
+    \item \ruleref{\ottdruleEOneShiftULabel},
+      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
+      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
+      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
+      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
+      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
+      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
+    \item \ruleref{\ottdruleEOneArrowLabel},
+      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
+      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
+      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
+      $
+      \begin{aligned}[t]
+        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
+                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
+                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
+                          && \text{as noted above}\\
+                          &\iff [[Γ ⊢ iQ → iM']]
+                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
+      \end{aligned}
+      $\\
+    \item \ruleref{\ottdruleEOneForallLabel},
+      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
+      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
+      $[[{pas} ∩ fv iM = ∅]]$ and 
+      that there exists a bijection on variables 
+      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
+      such that $[[iN' ≈ [mu] iM']]$.
+
+      By inversion and \ruleref{\ottdruleWFTForallLabel},
+      $[[Γ ⊢ ∀pas.iN']]$ is equivalent to $[[Γ, pas ⊢ iN']]$,
+      and by \cref{lemma:wf-ctxt-equiv}, it is equivalent to 
+      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ iN']]$,
+      which, by the induction hypothesis, is equivalent to
+      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ [mu]iM']]$.
+
+      Analogously, $[[Γ ⊢ ∀pbs.iM']]$ is equivalent to
+      $[[Γ, (Ord {pbs} ∩ fv iM') ⊢ iM']]$.
+      By \cref{lemma:wf-subst}, it implies 
+      $[[Γ, Ord ({pas} ∩ fv iM') ⊢ [mu]iM']]$. And vice versa, 
+      $[[Γ, Ord ({pas} ∩ fv iM') ⊢ [mu]iM']]$ implies  
+      $[[Γ, Ord ({pbs} ∩ fv iM') ⊢ [mu-1][mu]iM']]$.
+
+      This way, both $[[Γ ⊢ ∀pas.iN']]$ and $[[Γ ⊢ ∀pbs.iM']]$ are equivalent to
+      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ [mu]iM']]$. 
+    \item For the cases of the positive types, the proofs are symmetric.
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
+  \label{lemma:wf-equiv-algo}
+  Mutual subtyping implies declarative equivalence.
+  \begin{itemize}
+  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Θ ⊢ uP]] \iff [[Γ; Θ ⊢ uQ]]$,
+  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Θ ⊢ uN]] \iff [[Γ; Θ ⊢ uM]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The proof coincides with the proof of \cref{lemma:wf-equiv},
+  and adds two cases for equating two positive or two negative metavariables,
+  which must be equal by inversion, and thus, 
+  $[[Γ; Θ ⊢ α̂±]] \iff [[Γ;  ⊢ α̂±]]$ holds trivially.
 \end{proof}
@@
     Applying $[[σ2]]$ to both sides of
-    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{todo}),
+    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
     we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
-    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ (by transitivity \cref{todo}),
+    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
+    (\cref{lemma:subtyping-transitivity}),
     we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
@@
     \end{enumerate}
-    Then the these correspondences between $[[fv iP]]$ and
+    Then these correspondences between $[[fv iP]]$ and
     $[[fv iQ]]$ are mutually inverse functions,
@@
 
+\begin{lemma}[Equivalent substitution act equivalently]
+  \label{lemma:equiv-subst-on-same-term}
+  Suppose that $[[Γ' ⊢ σ1 : Γ]]$ and $[[Γ' ⊢ σ2 : Γ]]$
+  are substitutions equivalent on their domain, that is $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
+  Then
+  \begin{itemize}
+    \item[$+$] for any $[[Γ ⊢ iP]]$, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$;
+    \item[$-$] for any $[[Γ ⊢ iN]]$, $[[Γ' ⊢ [σ1]iN ≈ [σ2]iN]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on $[[iP]]$ (and mutually on $[[N]]$).
+  \begin{caseof}
+    \item $[[iN = α⁻]]$\\
+      Then since by inversion, $[[α⁻ ∊ {Γ}]]$, 
+      $[[Γ' ⊢ [σ1]α⁻ ≈ [σ2]α⁻]]$ holds by definition of $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
+    \item $[[iN = ↑iP]]$\\
+      Then by inversion, $[[Γ ⊢ iP]]$.
+      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$, 
+      Then by \ruleref{\ottdruleDOneShiftULabel}, 
+      $[[Γ' ⊢ ↑[σ1]iP ≤ ↑[σ2]iP]]$, and symmetrically, $[[Γ' ⊢ ↑[σ2]iP ≤ ↑[σ1]iP]]$,
+      together meaning that $[[Γ' ⊢ ↑[σ1]iP ≈ ↑[σ2]iP]]$,
+      or equivalently, $[[Γ' ⊢ [σ1]↑iP ≈ [σ2]↑iP]]$.
+    \item $[[iN = iP → iM]]$\\
+      Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
+      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$ and
+      $[[Γ' ⊢ [σ1]iM ≈ [σ2]iM]]$,
+      that is $[[Γ' ⊢ [σ1]iP ≥ [σ2]iP]]$, $[[Γ' ⊢ [σ2]iP ≥ [σ1]iP]]$,
+      $[[Γ' ⊢ [σ1]iM ≤ [σ2]iM]]$, and $[[Γ' ⊢ [σ2]iM ≤ [σ1]iM]]$.
+      Then by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$, 
+      and again by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ2]iP → [σ2]iM ≤ [σ1]iP → [σ1]iM]]$.
+      This way, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≈ [σ2]iP → [σ2]iM]]$,
+      or equivalently, $[[Γ' ⊢ [σ1](iP → iM) ≈ [σ2](iP → iM)]]$.
+    \item $[[iN = ∀pas.iM]]$
+      We can assume that $[[pas]]$ is disjoint from $[[Γ]]$ and $[[Γ']]$.
+      By inversion, $[[Γ ⊢ ∀pas.iM]]$ implies $[[Γ, pas ⊢ iM]]$.
+      Notice that $[[Γ' ⊢ σi : Γ]]$ and 
+      $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$
+      can be extended to 
+      $[[Γ', pas ⊢ σi : Γ, pas]]$
+      and 
+      $[[Γ', pas ⊢ σ1 ≈ σ2 : Γ, pas]]$
+      by \cref{lemma:subst-domain-weakening}.
+      Then by the induction hypothesis,
+      $[[Γ', pas ⊢ [σ1]iM ≈ [σ2]iM]]$,
+      meaning by inversion that
+      $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM]]$ and 
+      $[[Γ', pas ⊢ [σ2]iM ≤ [σ1]iM]]$.
+
+      To infer
+      $[[Γ' ⊢ ∀pas.[σ1]iM ≤ ∀pas.[σ2]iM]]$,
+      we apply \ruleref{\ottdruleDOneForallLabel}
+      with the substitution $[[Γ', pas ⊢ id : pas]]$, 
+      noting that $[[Γ', pas ⊢ [id][σ1]iM ≤ [σ2]iM ]]$
+      holds since $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM ]]$, 
+      as noted above.
+
+      Symmetrically, we infer
+      $[[Γ' ⊢ ∀pas.[σ2]iM ≤ ∀pas.[σ1]iM]]$,
+      which together with
+      $[[Γ' ⊢ ∀pas.[σ1]iM ≤ ∀pas.[σ2]iM]]$
+      means
+      $[[Γ' ⊢ ∀pas.[σ1]iM ≈ ∀pas.[σ2]iM]]$,
+      or equivalently, 
+      $[[Γ' ⊢ [σ1]∀pas.iM ≈ [σ2]∀pas.iM]]$.
+
+    \item The positive cases are proved symmetrically.
+  \end{caseof}
+\end{proof}
+
 \begin{lemma}[Equivalence of polymorphic types]
@@
     then there exists a bijection $[[μ : {pbs} ∩ fv iM ↔ {pas} ∩ fv iN]]$
-    such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iN ]]$,
+    such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iM ]]$,
     \item[$+$] For $[[Γ ⊢ ∃nas.iP]]$ and $[[Γ ⊢ ∃nbs.iQ]]$,\\  if $[[Γ ⊢ ∃nas.iP ≈ ∃nbs.iQ ]]$
@@
 
-    Let us show that if we restrict the domain of $[[μ]]$ to 
+      Let us show that $[[Sub μ|{pbs}]]$ is the appropriate candidate.
+
+      First, we show that if we restrict the domain of $[[μ]]$ to 
       $[[pbs]]$, its range will be contained in $[[pas]]$.
+
       Let us take $[[γ⁺ ∊ {pbs} ∩ fv iM]]$ and 
@@
       $[[σ1]]$ acts as identity outside of $[[pas]]$, i.e.
-    $[[ [σ1][Sub μ]γ⁺ = [Sub μ]γ⁺ ]]$.
+      $[[ [σ1][Sub μ]γ⁺ = [Sub μ]γ⁺ ]]$ (notice that $[[γ⁺]]$ is in the domain of $[[μ]]$).
       Since
       $[[Γ,pas,pbs ⊢ σ1 ≈ Sub μ-1 : Ord fv iN]]$, 
-    application of $[[σ1]]$ is equivalent to application of $[[Sub μ-1]]$,
-    then 
+      application of $[[σ1]]$ to $[[ [μ]γ⁺ ]] \in [[fv iN]]$
+      is equivalent to application of $[[Sub μ-1]]$, then 
       $[[ Γ,pas,pbs ⊢ [Sub μ-1][Sub μ]γ⁺ ≈ [Sub μ]γ⁺ ]]$, i.e.
@@
 
-    By \cref{todo}, 
+      Second, 
+      we will show $[[ Γ, pas ⊢ iN ≈ [Sub μ|{pbs}] iM ]]$.
+
+      Since $[[ Γ,pas ⊢ σ2 : pbs ]]$ and 
+      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ : Ord fv iM]]$, 
+      we have
+      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ|{pbs} : Ord fv iM]]$:
+      for any $[[α± ∊ fv iM \ {pbs}]]$, $[[ [σ2]α± = α± ]]$
+      since $[[ Γ,pas ⊢ σ2 : pbs ]]$, and
+      $[[ [Sub μ|{pbs}]α± = α± ]]$ by definition of substitution restriction;
+      for $[[β⁺ ∊ {pbs}]]$, $[[ [Sub μ|{pbs}]β⁺ = [μ]β⁺ ]]$,
+      and thus, $[[Γ,pas,pbs ⊢ [σ2]β⁺ ≈ [Sub μ]β⁺ ]]$
+      can be rewritten to $[[Γ,pas,pbs ⊢ [σ2]β⁺ ≈ [Sub μ|{pbs}]β⁺ ]]$.
+
+      By \cref{lemma:equiv-subst-on-same-term},
       $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ|{pbs} : Ord fv iM]]$ implies
@@
 
-    This way,
+      This way, by transitivity of subtyping (\cref{lemma:subtyping-transitivity}),
       \begin{align} 
@@
 
-    By applying $[[μ|_{pbs}]]$ to both sides of \ref{fact:mu-inv-n-sub-m} (\cref{todo})
-    and contracting $[[μ-1|_{pas} ○ μ|_{pbs}]] = [[μ|_{pbs}-1 ○ μ|_{pbs}]] = [[id]]$,
-    we have: $[[Γ,pas,pbs ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
+      By applying $[[Sub μ|{pbs}]]$ to both sides of \ref{fact:mu-inv-n-sub-m} 
+      (\cref{lemma:subst-pres-subt}), we have
+      $[[Γ,pas,pbs ⊢ [Sub μ|{pbs}][Sub μ-1|{pas}]iN ≤ [Sub μ|{pbs}]iM]]$. 
+      By contracting $[[μ-1|_{pas} ○ μ|_{pbs}]] = [[μ|_{pbs}-1 ○ μ|_{pbs}]]$
+      (notice that $[[fv iN ∩ {pbs} = ∅]]$), 
+      we have $[[Γ,pas,pbs ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
       means $[[Γ,pas,pbs ⊢ iN ≈ [Sub μ|{pbs}]iM]]$, and by strengthening, $[[Γ,pas⊢ iN ≈ [Sub μ|{pbs}]iM]]$.
-    Symmetrically, $[[Γ,pbs ⊢ iM ≈ [Sub μ|_{pbs}-1]iN]]$.
-    \item{$+$} The proof is symmetric to the proof of the negative case.
+      % Symmetrically, $[[Γ,pbs ⊢ iM ≈ [Sub μ|_{pbs}-1]iN]]$.
+    \item[$+$] The proof is symmetric to the proof of the negative case.
   \end{itemize}
@@
       \item \ruleref{\ottdruleDOneForallLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.iM']]$\\
-      Then by \cref{lemma:poly-type-equivalence}, $[[Γ ⊢ ∀pas.iN' ≈ ∀pbs.iM']]$ means that 
+      Then by \cref{lemma:poly-types-equivalence}, $[[Γ ⊢ ∀pas.iN' ≈ ∀pbs.iM']]$ means that 
       there exists a bijection $[[μ : {pbs} ∩ fv iM' ↔ {pas} ∩ fv iN']]$ such that  
@@
 
+  \label{lemma:poly-types-equivalence}
       Notice that the application of bijection $[[μ]]$ to $[[iM']]$ does

### 2023-08-08 proofs/tex/norm-lemmas.tex
wip
Transitivity of ≤
wip: leftovers

@@
 
-\begin{lemma}[Soundness of quantifier normalization]
+\begin{lemma}[Soundness of normalization]
   \label{lemma:normalization-soundness}
   \hfill
-
   \begin{itemize}
@@
 
+\begin{lemma}[Soundness of normalization of algorithmic types]
+  \label{lemma:normalization-soundness-alg}
+  \hfill
+  \begin{itemize}
+    \item[$-$] $[[uN ≈ nf(uN)]]$
+    \item[$+$] $[[uP ≈ nf(uP)]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The proof coincides with the proof of \cref{lemma:normalization-soundness}.
+\end{proof}
+
 \begin{corollary}[Normalization preserves ordering]
@@
   \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
-  taking free variables from $[[Γ2]]$.
-  Then
+  Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
     \begin{itemize}
@@
   \end{itemize}
-  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]
-  α⁻)]]$. %, and .
+  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
 \end{lemma}

### 2023-08-08 proofs/tex/ord-lemmas.tex
Prove the UB lemma
wip: leftovers

@@
 
+\begin{lemma}[Ordering is not affected by independent substitutions]
+  \label{lemma:ord-sigma}
+  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
+  taking free variables from $[[Γ2]]$, and $[[varset]]$ is a set of variables
+  disjoint with both $[[Γ1]]$ and $[[Γ2]]$, 
+  $[[iN]]$ and $[[iP]]$ are types. Then
+    \begin{itemize}
+  \item[$-$] $[[ ord varset in [σ]iN ]] = [[ord varset in iN ]]$
+  \item[$+$] $[[ ord varset in [σ]iP ]] = [[ord varset in iP ]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  Mutual induction on $[[iN]]$ and $[[iP]]$.
+  \begin{caseof}
+    \item $[[iN = na]]$ \\
+      If $[[na ∉ {Γ1}]]$ then $[[ [σ]na = na ]]$ and $[[ ord varset in [σ]na ]] = [[ ord varset in na ]]$, 
+      as requried.
+      If $[[na ∊ {Γ1}]]$ then $[[na ∉ varset]]$, so $[[ ord varset in na ]] = [[·]]$.
+      Moreover, $[[Γ2 ⊢ σ : Γ1]]$ means $[[ fv([σ]na) ⊆ {Γ2} ]]$, and thus, 
+      as a set, $[[ ord varset in [σ]na ]] = [[varset ∩ fv([σ]na)]] \subseteq [[varset ∩ {Γ2}]] = [[·]]$.
+    \item $[[iN = ∀pas.iM]]$\\
+      We can assume $[[{pas} ∩ {Γ1} = ∅]]$
+      and $[[{pas} ∩ varset = ∅]]$. Then
+
+      $\begin{aligned}[t]
+         [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]∀pas.iM ]] \\
+                                   &= [[ ord varset in ∀pas.[σ]iM ]]\\
+                                   &= [[ ord varset in [σ]iM ]]
+                                   && \text{by the induction hypothesis}\\
+                                   &= [[ ord varset in iM ]]\\
+                                   &= [[ ord varset in ∀pas.iM ]]\\
+                                   &= [[ ord varset in iN ]]
+       \end{aligned}$
+    \item $[[iN = ↑iP]]$\\
+       $\begin{aligned}[t]
+        [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]↑iP ]] \\
+                                   &= [[ ord varset in ↑[σ]iP ]]
+                                   && \text{by the definition of substitution}\\
+                                   &= [[ ord varset in [σ]iP ]]
+                                   && \text{by the induction hypothesis}\\
+                                   &= [[ ord varset in iP ]]
+                                   && \text{by the definition of substitution}\\
+                                   &= [[ ord varset in ↑iP ]]
+                                   && \text{by the definition of ordering}\\
+                                   &= [[ ord varset in iN ]]
+       \end{aligned} $
+
+    \item $[[iN = iP → iM]]$\\
+       $ \begin{aligned}
+        [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ](iP → iM) ]] \\
+                                   &= [[ ord varset in ([σ]iP → [σ]iM) ]]
+                                   && \text{by the definition of substitution}\\
+                                   &= [[ ord varset in [σ]iP, (ord varset in [σ]iM \ {ord varset in [σ]iP}) ]]
+                                   && \text{by the definition of ordering}\\
+                                   &= [[ ord varset in iP, (ord varset in iM \ {ord varset in iP}) ]]
+                                   && \text{by the induction hypothesis}\\
+                                   &= [[ ord varset in iP → iM ]]
+                                   && \text{by the definition of ordering}\\
+                                   &= [[ ord varset in iN ]]
+       \end{aligned} $
+    \item The proofs of the positive cases are symmetric.
+  \end{caseof}
+\end{proof}
+
 \begin{lemma}[Completeness of variable ordering]
   \label{lemma:ord-completeness}
-  Variable ordering is invariant under equivalence.
-
+  Variable ordering is invariant under equivalence. For arbitrary $[[varset]]$,
    \begin{itemize}
-  \item[$-$] For $[[iN ≈ iM]]$ and any $[[varset]]$,
-    $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)
-
-  \item[$+$] For $[[iP ≈ iQ]]$ and any $[[varset]]$,
-    $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
+  \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)
+  \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
   \end{itemize}

### 2023-08-08 proofs/tex/subst-lemmas.tex
wip: leftovers

@@
   substitution subject does not affect the result.
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
+  Suppose that $[[σ]]$ is a substitution,
+  $[[iP]]$ and $[[iN]]$ are types. Then
  \begin{itemize}
-  \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
-  \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
+    \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
+    \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
   \end{itemize}
@@
 \begin{proof}
-  \ilyam{todo}
+  First, we strengthen the statement by saying that
+  one can restrict the substitution to an arbitrary
+  superset of the free variables of the substitution subject:
+ \begin{itemize}
+    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
+    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
+  \end{itemize}
+  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
+  For the base cases: 
+  \begin{caseof}
+    \item $[[iN = α⁻]]$\\
+      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
+      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
+    \item $[[iN = iP → iM]]$\\
+      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
+      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
+      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
+      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
+      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
+      Then 
+      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
+       = [[ [σ|varset](iP → iM) ]]$.
+    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
+    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
+      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
+      Let us assume $[[pas]]$ are fresh variables, 
+      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
+      $[[α± ]] \in [[pas]]$, and thus, 
+      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
+      immediately from the definition.
+
+      Since $[[varset ⊆ fv(∀pas.iM)]] = [[fv iM \ {pas}]]$,
+      $[[varset ∪ {pas} ⊆ fv(iM)]]$.
+      Then by the induction hypothesis, 
+      $[[ [σ]iM =  [σ|(varset ∪ {pas})]iM ]]$.
+      Finally, 
+      $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ|(varset ∪ {pas})]iM ]] = 
+      [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
+    \item The positive cases are proves symmetrically. 
+  \end{caseof}
 \end{proof}

### 2023-08-08 proofs/tex/wf-lemmas.tex
wip: leftovers

@@
-\begin{lemma}[Well-formedness agrees with substitution]
-  \label{lemma:wf-subst}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-  \item[$+$] $[[Γ, Γ1 ⊢ iP]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iP]]$
-  \item[$-$] $[[Γ, Γ1 ⊢ iN]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iN]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  \ilyam{todo}
-\end{proof}
-
-\begin{corollary}
-  \label{lemma:wf-subst}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-  \item[$+$] $[[Γ1, Γ2 ⊢ iP]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iP]]$
-  \item[$-$] $[[Γ1, Γ2 ⊢ iN]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iN]]$
-  \end{itemize}
-\end{corollary}
-
 \begin{lemma}[Equivalent Contexts]
   \label{lemma:wf-ctxt-equiv}
+
   In the well-formedness judgment, only used variables matter:
@@
     $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
+\end{proof}
+
+\begin{lemma}[Well-formedness Context Weakening]
+  \label{lemma:wf-weakening}
+  Suppose that $[[{Γ1} ⊆ {Γ2}]]$, then
+  \begin{itemize}
+    \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[Γ2 ⊢ iP]]$,
+    \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[Γ2 ⊢ iN]]$.
   \end{itemize}
@@
 \end{proof}
+
+\begin{lemma}[Well-formedness agrees with substitution]
+  \label{lemma:wf-subst}
+  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
+  \begin{itemize}
+  \item[$+$] $[[Γ, Γ1 ⊢ iP]]$ implies $[[Γ, Γ2 ⊢ [σ]iP]]$, and
+  \item[$-$] $[[Γ, Γ1 ⊢ iN]]$ implies $[[Γ, Γ2 ⊢ [σ]iN]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on $[[Γ, Γ1 ⊢ iP]]$ and mutually, on $[[Γ, Γ1 ⊢ iN]]$.
+  Let us consider the last rule used in the derivation.
+  \begin{caseof}
+    \item \ruleref{\ottdruleWFTPVarLabel}, 
+      i.e. $[[iP]]$ is $[[α⁺]]$.\\
+      By inversion, $[[α⁺ ∊ {Γ, Γ1}]]$, then
+      \begin{itemize}
+        \item if $[[α⁺ ∊ {Γ1}]]$ then $[[ Γ2 ⊢ [σ]α⁺ ]]$, 
+          and by weakening (\cref{lemma:wf-weakening}),
+          $[[ Γ, Γ2 ⊢ [σ]α⁺ ]]$;
+        \item if $[[α⁺ ∊ {Γ} \ {Γ1}]]$ then $[[ [σ]α⁺ = α⁺ ]]$,
+          and by \ruleref{\ottdruleWFTPVarLabel}, $[[ Γ, Γ2 ⊢ α⁺ ]]$.
+      \end{itemize}
+
+    \item \ruleref{\ottdruleWFTShiftULabel},
+      i.e. $[[iP]]$ is $[[↓iN]]$.\\
+      Then $[[Γ, Γ1 ⊢ ↓iN]]$ means $[[Γ, Γ1 ⊢ iN]]$ by inversion,
+      and by the induction hypothesis, $[[Γ, Γ2 ⊢ [σ]iN]]$.
+      Then by  \ruleref{\ottdruleWFTShiftULabel}, $[[Γ, Γ2 ⊢ ↓[σ]iN]]$, 
+      which by definition of substitution is rewritten as $[[Γ, Γ2 ⊢ [σ]↓iN]]$.
+
+    \item \ruleref{\ottdruleWFTExistsLabel},
+      i.e. $[[iP]]$ is $[[∃nas.iQ]]$.\\
+      Then $[[Γ, Γ1 ⊢ ∃nas.iQ]]$ means $[[Γ, nas, Γ1 ⊢ iQ]]$ 
+      by inversion, and by the induction hypothesis, 
+      $[[Γ, nas, Γ2 ⊢ [σ]iQ]]$.
+      Then by  \ruleref{\ottdruleWFTExistsLabel}, 
+      $[[Γ, nas, Γ2 ⊢ ∃nas.[σ]iQ]]$, 
+      which by definition of substitution is rewritten as 
+      $[[Γ, Γ2 ⊢ [σ]∃nas.iQ]]$.
+
+    \item The negative cases are proved symmetrically.
+  \end{caseof}
+
+\end{proof}

### 2023-08-15 proofs/tex/decl-subtyping-lemmas.tex
WIP: leftovers

@@
 
+\begin{lemma}[Subtyping context irrelevance]
+  \label{lemma:subt-ctxt-irrelevance}
+  Suppose that all the mentioned types are well-formed in $[[Γ1]]$ and $[[Γ2]]$. 
+  Then 
+  \begin{enumerate}
+    \item[$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ is equivalent to $[[Γ2 ⊢ iP ≥ iQ]]$;
+    \item[$-$] $[[Γ1 ⊢ iN ≤ iM]]$ is equivalent to $[[Γ2 ⊢ iN ≤ iM]]$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  We prove it by induction on 
+  the size of $[[Γ1 ⊢ iP ≥ iQ]]$ and mutually,
+  the size of $[[Γ1 ⊢ iN ≤ iM]]$.
+
+  All the cases except \ruleref{\ottdruleDOneExistsLabel} and \ruleref{\ottdruleDOneForallLabel}
+  are proven congruently:
+  first, we apply the inversion to $[[Γ1 ⊢ iP ≥ iQ]]$ to obtain the premises of the
+  corresponding rule $X$, then we apply the induction hypothesis to each premise,
+  and build the inference tree (with $[[Γ2]]$) by the same rule $X$.
+
+  Suppose that the judgement is inferred by \ruleref{\ottdruleDOneExistsLabel}.
+  Then we are proving that $[[Γ1 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$ implies 
+  $[[Γ2 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$ (the other implication is proven symmetrically).
+
+  By inversion  of $[[Γ1 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$, 
+  we obtain $[[σ]]$ such that $[[ Γ1, nbs ⊢ σ : nas]]$
+  and $[[Γ1, nbs ⊢ [σ]iP ≥ iQ]]$.
+  By \cref{lemma:fv-propagation}, $[[fv([σ]iP) ⊆ fv(iQ)]]$.
+
+  From the well-formedness statements 
+  $[[Γi ⊢ ∃nas.iP]]$ and $[[Γi ⊢ ∃nbs.iQ]]$ we have:
+  \begin{itemize}
+    \item $[[Γ1, nas ⊢ iP]]$, which also means $[[Γ1, nbs ⊢ [σ]iP]]$
+      by \cref{lemma:wf-subst};
+    \item $[[Γ2, nas ⊢ iP]]$;
+    \item $[[Γ1, nbs ⊢ iQ]]$; and 
+    \item $[[Γ2, nbs ⊢ iQ]]$, which means $[[ fv(iQ) ⊆ {Γ2, nbs} ]]$
+      by \cref{lemma:wf-soundness},
+      and  combining it with $[[fv([σ]iP) ⊆ fv(iQ)]]$, 
+      we have $[[fv([σ]iP) ⊆ {Γ2, nbs}]]$.
+  \end{itemize}
+
+  Let us construct a substitution $[[σ0]]$ in the following way:
+  $$
+  \begin{cases}
+      [[ [σ0]αi⁻ = [σ]αi⁻  ]] & \text{for $[[αi⁻ ∊ {nas} ∩ fv(iP)]]$ }\\
+      [[ [σ0]αi⁻ = ∀γ⁺.↑γ⁺ ]] & \text{for $[[αi⁻ ∊ {nas} \ fv(iP)]]$ }\\
+      [[ [σ0]γ±  = γ± ]]      & \text{for any other $[[γ±]]$ }\\
+  \end{cases}
+  $$
+  Notice that 
+  \begin{enumerate}
+    \item $[[ [σ0]iP = [σ]iP ]]$.
+      Since $[[σ0|fv(iP)]] = [[σ|fv(iP)]]$ as functions
+      (which follows from the construction of 
+      $[[σ0]]$ and the signature of $[[σ]]$), 
+      $[[ [σ0]iP = [σ0|fv(iP)]iP = [σ|fv(iP)]iP = [σ]iP]]$
+      (where the first and the last equalities are by
+       \cref{lemma:subst-restr-fv}).
+
+    \item $[[ Ord fv([σ]iP) ⊢ σ0 : nas]]$.
+      To show that, let us consider $[[αi⁻]]$ 
+      \begin{itemize}
+        \item if $[[αi⁻ ∊ {nas} \ fv(iP)]]$ then
+          $[[· ⊢ [σ0]αi⁻]]$, which can be weakened to 
+          $[[Ord fv([σ]iP) ⊢ [σ0]αi⁻]]$;
+        \item if $[[αi⁻ ∊ {nas} ∩ fv(iP)]]$, 
+          we have $[[ [σ0]αi⁻ = [σ]αi⁻ ]]$, and thus, by
+          specification of $[[σ]]$, $[[Γ1, pbs ⊢ [σ0]αi⁻ ]]$. 
+          By \cref{corollary:wf-ctxt-strengthening}, 
+          it means $[[Ord fv([σ0]αi⁻) ⊢ [σ0]αi⁻]]$,
+          which we weaken (\cref{lemma:wf-weakening}) 
+          to $[[Ord fv([σ]iP) ⊢ [σ0]αi⁻]]$ 
+          (since $[[fv([σ0]αi⁻) ⊆ fv([σ0]iP)]]$ by \cref{lemma:subst-fv-image},
+          and $[[ [σ0]iP = [σ]iP]]$, as noted above).
+      \end{itemize} 
+  \end{enumerate}
+
+  By \cref{corollary:wf-ctxt-strengthening}, 
+  $[[Γ1, nbs ⊢ [σ]iP]]$ implies 
+  $[[ Ord fv([σ]iP) ⊢ [σ]iP ]]$, which,
+  since $[[fv([σ]iP) ⊆ {Γ2, nbs}]]$,
+  is weakened to $[[Γ2, nbs ⊢ [σ]iP]]$.
+  and rewritten as $[[Γ2, nbs ⊢ [σ0]iP]]$.
+
+  Notice that the premises of the induction hold:
+  \begin{enumerate}
+    \item $[[Γi, nbs ⊢ [σ0]iP]]$, 
+    \item $[[Γi, nbs ⊢ iQ]]$, and
+    \item $[[Γ1, nbs ⊢ [σ0]iP ≥ iQ ]]$, 
+      notice that the tree inferring this judgement 
+      is the same tree inferring $[[Γ1, nbs ⊢ [σ]iP ≥ iQ ]]$ 
+      (since $[[ [σ0]iP = [σ]iP]]$), i.e., 
+      it is a subtree of $[[Γ1 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$.
+  \end{enumerate}
+  This way, by the induction hypothesis,
+  $[[Γ2, nbs ⊢ [σ0]iP ≥ iQ ]]$.
+  Combining it with $[[ Γ2, nbs ⊢ σ0 : nas]]$
+  by \ruleref{\ottdruleDOneExistsLabel},
+  we obtain $[[Γ2 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$.
+
+  The case of $[[Γ1 ⊢ ∀pas.iN ≤ ∀pbs.iM]]$ is symmetric.
+\end{proof}
+
+\begin{lemma}[Weakening of subtyping context]
+  \label{lemma:subt-weakening}
+  Suppose $[[Γ1]]$ and $[[Γ2]]$ are contexts 
+  and $[[ {Γ1} ⊆ {Γ2} ]]$.
+  Then 
+  \begin{enumerate}
+    \item [$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ implies $[[Γ2 ⊢ iP ≥ iQ]]$;
+    \item [$-$] $[[Γ1 ⊢ iN ≤ iM]]$ implies $[[Γ2 ⊢ iN ≤ iM]]$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+\end{proof}
+
 \begin{lemma}[Reflexivity of subtyping] \label{lemma:subtyping-reflexivity}
@@
   \begin{itemize}
-    \item $-$ If $[[Γ1 ⊢ iN ≤ iM]]$ then $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$
-    \item $+$ If $[[Γ1 ⊢ iP ≥ iQ]]$ then $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$
+    \item [$-$] If $[[Γ1 ⊢ iN ≤ iM]]$ then $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$.
+    \item [$+$] If $[[Γ1 ⊢ iP ≥ iQ]]$ then $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$.
   \end{itemize}

### 2023-08-15 proofs/tex/equiv-lemmas.tex
WIP: leftovers

@@
 
+\begin{lemma}[Normalization preserves substitution signature]
+  \label{lemma:norm-subst-sig}
+  Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
+  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
+\end{lemma}
+\being{proof}
+  Suppose that $[[α± ∊ {Γ1}]]$. 
+  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
+  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.
+
+  Suppose that $[[α± ∉ {Γ1}]]$. 
+  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
+  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
+\end{proof}
+
+\begin{corollary}[Normalization preserves well-formedness of algorithmic types]
+  \label{corollary:wf-nf-algo}
+  \begin{itemize}
+    \item[$+$] $[[Γ; Θ ⊢ uP]] \iff [[Γ; Θ ⊢ nf(uP)]]$,
+    \item[$-$] $[[Γ; Θ ⊢ uN]] \iff [[Γ; Θ ⊢ nf(uN)]]$
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  Immediately from \cref{lemma:wf-equiv-algo,lemma:normalization-soundness-alg}.
+\end{proof}
+
 \begin{corollary}[Normalization preserves well-formedness of substitution]
@@
 \end{corollary}
+\begin{proof}
+  Let us prove the forward direction.
+  Suppose that $[[α± ∊ {Γ1}]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
+  By the definition of substitution normalization,
+  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf-algo},
+  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
+  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
+
+  The backward direction is proved analogously.
+\end{proof}
+
+\begin{corollary}
+  \label{corollary:wf-s-nf-algo}
+  \hfill \\
+  $[[Θ ⊢ uσ]] \iff [[Θ ⊢ nf(σ)]]$
+\end{corollary}
+\begin{proof}
+  The proof is analogous to \cref{corollary:wf-s-nf}.
+\end{proof}
 
@@
     $[[Γ ⊢ σ2 ○ σ1 ≈ id : Ord fv iP]]$.
-
-    % Applying $[[σ1]]$ to both sides of
-    % $[[Γ ⊢ [σ2]iQ ≥ iP]]$ (by \cref{todo}),
-    % we have: $[[Γ ⊢ [σ1 ○ σ2] iQ ≥ [σ1]iP]]$.
-    % Composing it with $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by transitivity \cref{todo}),
-    % we have $[[Γ ⊢ [σ1 ○ σ2] iQ ≥ iQ]]$.
-    % Then by \cref{corollary:subst-proper-subt},
     By a symmetric argument, we also have:

### 2023-08-15 proofs/tex/subst-lemmas.tex
WIP: leftovers

@@
   \end{caseof}
+\end{proof}
+
+\begin{lemma}[Signature of a restricted substitution]
+  \label{lemma:subst-restr-sig}
+  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset : Ord {Γ1} ∩ varset]]$.
+\end{lemma}
+\begin{proof}
+  Let us take an arbitrary $[[α± ∊ {Γ1} ∩ varset]]$.
+  Since $[[α± ∊ {Γ1}]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.
+
+  Let us take an arbitrary $[[α± ∉ {Γ1} ∩ varset]]$.
+  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α± ]] = [[α±]]$ by definition
+  of restriction. 
+  If $[[α± ∊ varset \ {Γ1}]]$ then $[[ [σ|varset]α± ]] = [[ [σ]α± ]]$ by definition
+  and $[[ [σ]α± ]] = [[α±]]$ by the signature of $[[σ]]$.
+\end{proof}
+
+\begin{lemma}
+  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. 
+  Then if $[[varset]]$ is disjoint from $[[Γ1]]$, 
+  then $[[ σ|varset]] = [[ id ]]$. 
+\end{lemma}
+\begin{proof}
+  Let us take an arbitrary $[[α±]]$.
+  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α±]] = [[α±]]$ by definition.
+
+  If $[[α± ∊ varset]]$ then $[[α± ∉ {Γ1}]]$ by assumption.
+  Then $[[ [σ|varset]α± = [σ]α±]]$ by definition of restricted substitution, and
+  since $[[Γ2 ⊢ σ : Γ1]]$, we have $[[ [σ]α± = α± ]]$.
+\end{proof}
+
+\begin{corollary}[Application of a disjoint substitution]
+  \label{corollary:subst-disj}
+  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. Then
+  \begin{itemize}
+    \item [$+$] if $[[{Γ1} ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
+    \item [$-$] if $[[{Γ1} ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
+  \end{itemize}
+\end{corollary}
+
+\begin{lemma}[Substitution range weakening]
+  \label{lemma:subst-range-weakening}
+  Suppose that $[[{Γ2} ⊆ {Γ'2}]]$ are contexts and 
+  $[[σ]]$ is a substitution.
+  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ'2 ⊢ σ : Γ1]]$.
+\end{lemma}
+\begin{proof}
+  For any $[[α± ∊ {Γ1}]]$, 
+  $[[Γ2 ⊢ σ : Γ1]]$ gives us  $[[ Γ2 ⊢ [σ]α± ]]$, which can be
+  weakened to $[[ Γ'2 ⊢ [σ]α± ]]$ by \cref{lemma:wf-weakening}. 
+  This way, $[[Γ'2 ⊢ σ : Γ1]]$.
 \end{proof}
@@
 \end{proof}
+
+\begin{lemma}[Free variables after substitution]
+  \label{lemma:subst-fv}
+  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, then
+  \begin{enumerate}
+    \item [$+$] for a type $[[iP]]$,
+      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
+     $[[fv(iP) \ {Γ1} ⊆ fv([σ]iP) ⊆ (fv(iP) \ {Γ1}) ∪ {Γ2}]]$;
+    \item [$-$] for a type $[[iN]]$, 
+      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
+      $[[fv(iN) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iN) \ {Γ1}) ∪ {Γ2}]]$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  We prove it by structural induction on 
+  $[[iP]]$ and mutually, on $[[iN]]$.
+  \begin{caseof}
+    \item $[[iP = α⁺]]$\\
+      If $[[α⁺ ∊ {Γ1}]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
+      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ {Γ2}]]$.
+      $[[fv(α⁺) \ {Γ1}]] = [[∅]]$, so $[[fv([σ]iP) \ {Γ1}  ⊆ fv([σ]α⁺)]]$
+      vacuously.
+
+      If $[[α⁺ ∉ {Γ1}]]$ then $[[ [σ]α⁺ = α⁺ ]]$, and 
+      $[[fv([σ]α⁺) = {α⁺}]] = [[{α⁺} \ {Γ1}]]$.
+
+    \item $[[iP = ∃nas.iQ]]$\\
+      Then we need to show that $[[fv([σ]iP) = fv([σ]iQ) \ {nas}]]$
+      is a subset of $[[(fv(iP) \ {Γ1}) ∪ {Γ2}]]$ and a superset of
+      $[[fv(iP) \ {Γ1}]]$.
+      Notice that $[[fv(iP) = fv(iQ) \ {nas}]]$ by definition.
+      This way, we need to show that 
+      $[[fv(iQ) \ {nas} \ {Γ1} ⊆ fv([σ]iQ) \ {nas} ⊆ (fv(iQ) \ {nas} \ {Γ1}) ∪ {Γ2}]]$,
+
+      By the induction hypothesis, 
+      $[[fv([σ]iQ) ⊆ (fv(iQ) \ {Γ1}) ∪ {Γ2}]]$.
+      So for the second inclusion, it suffices to show that
+      $[[((fv(iQ) \ {Γ1}) ∪ {Γ2}) \ {nas} ⊆ (fv(iQ) \ {nas} \ {Γ1}) ∪ {Γ2}]]$,
+      which holds by set theoretical reasoning.
+
+      Also by the induction hypothesis,
+      $[[fv(iQ) \ {Γ1} ⊆ fv([σ]iQ)]]$,
+      and thus, by subtracting $[[{nas}]]$ from both sides,
+      $[[fv(iQ) \ {nas} \ {Γ1} ⊆ fv([σ]iQ) \ {nas}]]$.
+
+    \item The case $[[iN = ∀pas.iM]]$ is proved analogously.
+    \item $[[iN = iP → iM]]$\\
+      Then $[[fv([σ]iN) = fv([σ]iP) ∪ fv([σ]iM)]]$.
+      By the induction hypothesis, 
+      \begin{enumerate}
+        \item $[[ fv(iP) \ {Γ1} ⊆ fv([σ]iP) ⊆ (fv(iP) \ {Γ1}) ∪ {Γ2}]]$ and
+        \item $[[ fv(iM) \ {Γ1} ⊆ fv([σ]iM) ⊆ (fv(iM) \ {Γ1}) ∪ {Γ2}]]$.
+      \end{enumerate}
+      We unite these inclusions vertically and obtain
+      $[[ fv(iP) \ {Γ1} ∪ fv(iM) \ {Γ1} ⊆ fv([σ]iN) ⊆ ((fv(iP) \ {Γ1}) ∪ {Γ2}) ∪ ((fv(iM) \ {Γ1}) ∪ {Γ2})]]$,
+      which is equivalent to 
+      $[[ (fv(iP) ∪ fv(iM)) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iP) ∪ fv(iM)) \ {Γ1} ∪ {Γ2}]]$.
+      Since $[[fv(iP) ∪ fv(iM) = fv(iN)]]$, 
+      $[[fv(iN) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iN) \ {Γ1}) ∪ {Γ2}]]$.
+    \item The cases when $[[iP = ↓iM]]$ and $[[iN = ↑iQ]]$ are proved
+      analogously
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Free variables of a variable image]
+  \label{lemma:subst-fv-image}
+  Suppose that $[[σ]]$ is an arbitrary substitution, 
+  Then 
+  \begin{itemize}
+    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
+    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  By straightforward mutual induction on $[[iP]]$ and $[[iN]]$.
+\end{proof}

### 2023-08-15 proofs/tex/upgrade-lemmas.tex
WIP: leftovers

@@
     Notice that $[[Δ ⊢ iQ]]$ implies that $[[fv iQ ∩ {pnbs} = ∅]]$, 
-    then by \cref{todo}, $[[ [pnas/pnbs]iQ = iQ]]$, and thus $[[Δ, pnas, pncs ⊢ iQ ≥ iP]]$.
+    then by \cref{corollary:subst-disj}, $[[ [pnas/pnbs]iQ = iQ]]$,
+     and thus $[[Δ, pnas, pncs ⊢ iQ ≥ iP]]$.
     By context strengthening, $[[Δ, pnas ⊢ iQ ≥ iP]]$.
@@
     we have $[[Δ, pnbs ⊢ [pnbs/pnas]iQ' ≥ [pnbs/pnas]iP]]$, which by  
-    \cref{todo}, since $[[iQ']]$ is well-formed in $[[Δ]]$, simplifies to 
-     $[[Δ, pnbs ⊢ iQ' ≥ [pnbs/pnas]iP]]$.
+    \cref{corollary:subst-disj}, since $[[pnas]]$ is disjoint from $[[fv(iQ')]]$
+    (because $[[Δ ⊢ iQ']]$), simplifies to $[[Δ, pnbs ⊢ iQ' ≥ [pnbs/pnas]iP]]$.
 

### 2023-08-15 proofs/tex/wf-lemmas.tex
WIP: leftovers

@@
-\begin{lemma}[Equivalent Contexts]
+\begin{lemma}[Soundness of well-formedness]
+    \label{lemma:wf-soundness}
+    \hfill
+
+  \begin{itemize}
+    \item[$+$] if $[[Γ ⊢ iP]]$ then $[[fv(iP) ⊆ {Γ}]]$;
+    \item[$-$] if $[[Γ ⊢ iN]]$ then $[[fv(iN) ⊆ {Γ}]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The proof is done by a simple structural induction on 
+  $[[Γ⊢ iP]]$ and mutually, $[[Γ ⊢ iN]]$.
+  \begin{caseof}
+    \item $[[Γ ⊢ α±]]$ means by inversion that $[[α±]] \in [[Γ]]$,
+      that is, $[[α±]] = [[fv(α±) ⊆ {Γ}]]$.
+    \item $[[Γ ⊢ iQ → iM]]$ means by inversion that
+      $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM]]$.
+      Then by the induction hypothesis, $[[fv(iQ) ⊆ {Γ}]]$ and $[[fv(iM) ⊆ {Γ}]]$,
+      and hence, $[[fv(iQ → iM) = fv(iQ) ∪ fv(iM)]] \subseteq [[{Γ}]]$.
+    \item the cases when $[[iP = ↓iN']]$ or $[[iN = ↑iP']]$ are proven analogously.
+    \item $[[Γ ⊢ ∀pas.iM]]$ means by inversion that $[[Γ, pas ⊢ iM]]$.
+      Then by the induction hypothesis, $[[fv(iM) ⊆ {Γ, pas}]]$, and hence,
+      $[[fv(∀pas.iM) = fv(iM) \ {pas}]] \subseteq [[{Γ, pas} \ {pas} = {Γ}]]$.
+    \item The case $[[iP = ∃nas.iQ]]$ is proven analogously.
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Completeness of well-formedness]
   \label{lemma:wf-ctxt-equiv}
@@
 \begin{proof}
-  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
+  By simple mutual induction on $[[iP]]$ and $[[iN]]$.
 \end{proof}
 
-\begin{lemma}[Well-formedness Context Weakening]
+\begin{corollary}[Context Strengthening]
+  \label{corollary:wf-ctxt-strengthening}
+  \hfill
+
+  \begin{itemize}
+    \item [$+$] If $[[Γ ⊢ iP]]$ then $[[Ord fv(iP) ⊢ iP]]$;
+    \item [$-$] If $[[Γ ⊢ iN]]$ then $[[Ord fv(iN) ⊢ iN]]$.
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  It follows from \cref{lemma:wf-ctxt-equiv} and \cref{lemma:wf-soundness}.
+  \begin{itemize}
+    \item [$+$] 
+      By \cref{lemma:wf-soundness}, $[[fv(iP) ⊆ {Γ}]]$, and hence,
+      $[[{Γ} ∩ fv iP = fv iP ]]$, which makes \cref{lemma:wf-ctxt-equiv}
+      applicable fore contexts $[[Γ]]$ and $[[Ord fv(iP)]]$.
+    \item[$-$] The negative case is proven analogously.
+  \end{itemize}
+
+\begin{corollary}[Well-formedness Context Weakening]
   \label{lemma:wf-weakening}
@@
   \end{itemize}
-\end{lemma}
+\end{corollary}
 \begin{proof}
-  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
+  By \cref{lemma:wf-soundness},
+  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ {Γ1}]]$,
+  which means that $[[fv(iP) ⊆ {Γ2}]]$,
+  and thus, $[[fv(iP)]] = [[fv(iP) ∩ {Γ1}]] = [[fv(iP) ∩ {Γ2}]]$.
+  Then by \cref{lemma:wf-ctxt-equiv}, $[[Γ2 ⊢ iP]]$. 
+  The negative case is symmetric.
 \end{proof}

### 2023-08-21 proofs/main.tex
WIP: leftovers
WIP: description, notation

@@
 
+\subsection{Equalities}
+For simplicity, assume alpha-equivalent terms equal. 
+This way, we assume that substitutions do not capture bound variables.
+Besides, we equate
+$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
+as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
+and lift these equations transitively and congruently 
+to the whole system.
+
+\subsection{Contexts}
+
+\begin{definition}[Declarative Type Context]
+  \hfill \\
+  Declarative type context $[[Γ]]$ is represented by a set of 
+  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
+  union of two contexts  $[[{Γ1} ∪ {Γ2}]]$.
+\end{definition}
+
+\subsection{Substitutions}
+
+\begin{definition}[Substitution]
+  Substitutions (denoted as $[[σ]]$) 
+  are represented by total functions form variables to types, preserving the polarity. 
+\end{definition}
+
+\begin{definition}[Substitution Application]
+  Substitution application (denoted as $[[ [σ]iP ]]$ or $[[ [σ]iN ]]$) 
+  is defined congruently as follows:
+  \begin{itemize}
+    \item $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
+    \item $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
+    \item $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
+    \item $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
+    \item $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
+    \item $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$(here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
+      that is the variable capture never happens);
+    \item $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Substitution Signature]
+  The signature $[[Γ' ⊢ σ : Γ]]$ means that
+  \begin{enumerate}
+    \item for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ]α± ]]$; and
+    \item for any $[[α± ∉ {Γ'}]], [[ [σ]α± = α± ]]$.
+  \end{enumerate}
+\end{definition}
+
+A substitution can be restricted to a set of variables. 
+The restricted substitution is define as expected. 
+\begin{definition}[Subsitution Restriction]
+  The specification $[[σ  | varset]]$ is defined as
+  a function such that 
+  \begin{enumerate}
+    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
+    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
+  \end{enumerate}
+\end{definition}
+
+Two substitutions can be composed in two ways:
+$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
+while $[[σ2 <=< σ1]]$
+depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
+$[[σ2]]$ to its results on the domain.
+\begin{definition}[Substitution Composition]
+  $[[σ2 ○ σ1]]$ is defined as a function such that
+  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
+\end{definition}
+
+\begin{definition}[Monadic Substitution Composition]
+  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
+  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
+\end{definition}
+Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
+specification of $[[σ1]]$, which is not unique. 
+However, we assume that the used specification clear from the 
+context of the proof.
+
+\begin{definition}[Equivalent Substitutions]
+  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
+  indicates that on the domain $[[Γ]]$, 
+  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
+  Formally, for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
+\end{definition}
+
+Sometimes it is convenient to construct substitution 
+explicitly mapping each variable from a list (or a set)
+to a type. Such substitutions are denoted as $[[iPs / pas]]$
+and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
+the corresponding types.
+\begin{definition}[Explicit Substitution]
+  \hfill
+  \begin{itemize}
+    \item [$-$]
+      Suppose that $[[nas]]$ is a list of negative type variables,
+      and $[[iNs]]$ is a list of negative types of the same length 
+      Then $[[iNs / nas]]$ denotes a substitution such that 
+      \begin{enumerate}
+        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
+        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
+      \end{enumerate}
+    \item [$+$]
+      Positive explicit substitution $[[iPs / pas]]$
+      is defined symmetrically.
+  \end{itemize}
+\end{definition}
+
 \subsection{Declarative Typing}
@@
 
-\subsubsection{Variables and Types}
 In the algorithmic system, we extend the grammar of types
@@
 
-\subsubsection{Contexts}
+\subsection{Algorithmic Renaming}
 
-Unannotated set of algorithmic variables is denoted as $[[Ξ]]$,
-and it is used in the anti-unification algorithm.
+In several places of our algorithm, 
+we turn a declarative type into the algorithmic one
+via replacing certain type variables with fresh algorithmic variables.
+This is done via algorithmic renaming
+
+\begin{definition}
+  Suppose that $[[nas]]$ is a list of negative type variables,
+
+\subsection{Contexts}
+
+\begin{definition}[Algorithmic Type Context]
+  \hfill \\
+  Algorithmic type context $[[Ξ]]$ is represented by a set of 
+  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
+\end{definition}
+
 $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
@@
 $[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
-of the type is contained in $[[Ξ]]$, and each free non-algorithmic type variable
+of the type is contained in $[[Ξ]]$, and each free declarative type variable
 of the type is contained in $[[Γ]]$.
 
-Typically each algorithmic variable is associated with a
+Algorithmic Type Context are used in the unification algorithm.
+In the subtyping algorithm, 
+the context needs to remember additional information.
+In the subtyping context, each algorithmic variable is associated with a
 context it must be instantiated in 
 (i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by an \emph{algorithmic variable context} $[[Θ]]$.
-\begin{definition}[Algorithmic Context]
+This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
+\begin{definition}[Algorithmic Subtyping Context]
   \hfill \\
-  Algorithmic Context $[[Θ]]$ is represented by a set of 
+  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
   entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
@@
 
-\subsubsection{Subsitutions}
+\subsection{Subsitutions}
 
@@
 \ottdefnsNrm
+
+We also define normalization of a substitution pointwise:
+\begin{definition}[Substitution Normalization]
+  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
+  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
+\end{definition}
 
@@
 
+\subsection{Type well-formedness}
+\input{../_build/_gen/tex/wf-lemmas.tex}
+
 \subsection{Substitution}
-
-\begin{definition}[Substitution]
-  Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions form variables to types, preserving the polarity. 
-\end{definition}
-
-\begin{definition}[Substitution Application]
-  Substitution application (denoted as $[[ [σ]iP ]]$ or $[[ [σ]iN ]]$) 
-  is defined congruently as follows:
-  \begin{itemize}
-    \item $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
-    \item $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
-    \item $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
-    \item $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
-    \item $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
-    \item $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$(here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
-      that is the variable capture never happens);
-    \item $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
-  \end{itemize}
-\end{definition}
-
-\begin{definition}[Substitution Signature]
-  The signature $[[Γ' ⊢ σ : Γ]]$ means that
-  \begin{enumerate}
-    \item for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ {Γ'}]], [[ [σ]α± = α± ]]$.
-  \end{enumerate}
-\end{definition}
-
-A substitution can be restricted to a set of variables. 
-The restricted substitution is define as expected. 
-\begin{definition}[Subsitution Restriction]
-  The specification $[[σ  | varset]]$ is defined as
-  a function such that 
-  \begin{enumerate}
-    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
-    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
-  \end{enumerate}
-\end{definition}
-
-Two substitutions can be composed in two ways:
-$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
-while $[[σ2 <=< σ1]]$
-depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
-$[[σ2]]$ to its results on the domain.
-\begin{definition}[Substitution Composition]
-  $[[σ2 ○ σ1]]$ is defined as a function such that
-  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
-\end{definition}
-
-\begin{definition}[Monadic Substitution Composition]
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
-\end{definition}
-
-\begin{definition}[Equivalent Substitutions]
-  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
-  indicates that on the domain $[[Γ]]$, 
-  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
-\end{definition}
 
@@
 \input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Type well-formedness}
-\input{../_build/_gen/tex/wf-lemmas.tex}
 
@@
 
+\subsection{Constraint Satisfaction}
+\input{../_build/_gen/tex/constraint-sat-lemmas.tex}
+
 \subsection{Positive Subtyping}
@@
 \input{../_build/_gen/tex/constraint-merge-lemmas.tex}
-
-\subsection{Constraint Satisfaction}
-\input{../_build/_gen/tex/constraint-sat-lemmas.tex}
 

### 2023-08-21 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: description, notation

@@
 \begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
-        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$ and 
+        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
         $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
         $[[Θ ⊢ SC]]$ and 
-        for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : SC]]$,
+        for any normalized $[[uσ]]$ such that $[[(Θ ⊢ SC) ⊢ uσ]]$,
         $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
@@
 
-    Suppose that $[[uσ]]$ is normalized and $[[Θ ⊢ uσ : SC]]$,
+    Suppose that $[[uσ]]$ is normalized and $[[(Θ ⊢ SC) ⊢ uσ]]$,
     Let us consider the last rule to infer this judgment. 
@@
         By the soundness of constraint merge (\cref{lemma:merge-soundness}),
-        $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        $[[(Θ ⊢ SC1) ⊢ uσ]]$ and $[[(Θ ⊢ SC2) ⊢ uσ]]$.
         Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
@@
     \label{lemma:neg-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; Θ ⊢ uN]]$,
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
     and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
     Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
-    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
+    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
 \end{lemma}
@@
         Notice that from the inversion of $[[Γ ⊢ iM]]$, we have: $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$;
-        from the inversion of $[[Γ ; Θ ⊢ uN]]$, we have: $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$;
+        from the inversion of $[[Γ ; dom(Θ) ⊢ uN]]$, we have: $[[Γ ; dom( Θ) ⊢  uP]]$ and $[[Γ ; dom( Θ) ⊢  uN']]$;
         and since $[[uN]]$ does not contain negative unification variables,
@@
         This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
-        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$.
         Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
-        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ uσ : SC1]]$.
+        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[(Θ  ⊢  SC1) ⊢ uσ]]$.
 
@@
         To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
-        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[Θ ⊢ uσ : SC]]$, as required.
+        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[(Θ  ⊢  SC) ⊢ uσ]]$, as required.
 

### 2023-08-21 proofs/tex/alg-pos-subtyping-lemmas.tex
WIP: leftovers
WIP: description, notation

@@
 \begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; Θ ⊢ uP]]$, and 
+    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
     $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
     then $[[Θ ⊢ SC]]$ and
-    for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : SC]]$,
+    for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  SC) ⊢ uσ]]$,
     $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
@@
     It is left to show that $[[Γ ⊢ [uσ]â⁺ ≥ iP']]$ for any normalized $[[uσ]]$ 
-    s.t. $[[Θ ⊢ uσ : (â⁺ :≥ iQ')]]$.
+    s.t. $[[(Θ  ⊢  (â⁺ :≥ iQ')) ⊢ uσ]]$.
     The latter means that $[[ Θ(â⁺) ⊢ [uσ]â⁺ ≥ iQ' ]]$, i.e. $[[Δ ⊢ [uσ]â⁺ ≥ iQ']]$. 
@@
     By the soundness of unification (\cref{lemma:unification-soundness}),
-    $[[Θ ⊢ SC]]$ and for any normalized $[[uσ]]$, $[[Θ ⊢ uσ : SC]]$
+    $[[Θ ⊢ SC]]$ and for any normalized $[[uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$
     implies $[[ [uσ]nf(uN) = nf(iM) ]]$, 
@@
 
-        Let us take an arbitrary normalized $[[uσ]]$ s.t. $[[Θ ⊢ uσ : SC' \ {α̂⁻*}]]$.
+        Let us take an arbitrary normalized $[[uσ]]$ s.t. $[[(Θ  ⊢  SC' \ {α̂⁻*}) ⊢ uσ]]$.
         We wish to show $[[Γ ⊢ [uσ]uP ≥ iQ]]$, i.e. $[[Γ ⊢ ∃nas.[uσ]uP' ≥ ∃nbs.iQ']]$.
@@
         $[[Γ, nbs ⊢ [iNs/nas][uσ]uP' ≥ iQ']]$ for some $[[iNs]]$.
-        If we construct a normalized $[[uσ']]$ such that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']]$
+        If we construct a normalized $[[uσ']]$ such that $[[(Θ, â⁻*[G, nbs] ⊢ SC') ⊢ uσ']]$
         and for some $[[iNs]]$, $[[ [iNs/nas][uσ]uP' ]] = [[ [uσ'][â⁻*/nas]uP' ]]$,
@@
 
-        It is easy to see that $[[uσ']]$ is normalized.
-        Let us show that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']]$.
+        It is easy to see that $[[uσ']]$ is normalized: it inherits this property from 
+        $[[uσ]]$.
+        Let us show that $[[(Θ, â⁻*[G, nbs]  ⊢ SC') ⊢ uσ' ]] $.
         Let us take an arbitrary entry $[[scE]]$ from $[[SC']]$ restricting a variable $[[β̂±]]$.
@@
         $[[ (Θ, â⁻*[G, nbs])(β̂±) ⊢ [uσ']β̂± : scE ]]$ is
-        rewritten as $[[ Θ(β̂±) ⊢ [uσ]β̂± : scE ]]$, which holds since $[[Θ ⊢ uσ : SC']]$.
+        rewritten as $[[ Θ(β̂±) ⊢ [uσ]β̂± : scE ]]$, which holds since $[[(Θ  ⊢  SC') ⊢ uσ]]$.
         Suppose $[[β̂±]] = [[αî⁻]] \in [[α̂⁻*]]$. Then
@@
             \item If this variable is a unification variable $[[β̂±]]$, then
-                $[[ [iNs/nas][uσ] β̂± ]] = [[ [uσ]β̂± ]] $, since $[[Θ ⊢ uσ : SC' \ {α̂⁻*}]]$ and 
+                $[[ [iNs/nas][uσ] β̂± ]] = [[ [uσ]β̂± ]] $, since $[[(Θ  ⊢  SC' \ {α̂⁻*}) ⊢ uσ]]$ and 
                 $[[ dom(Θ) ∩ {nas} = ∅ ]]$.
@@
 \begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; Θ ⊢ uP]]$.
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
     Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
-    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[Θ ⊢ uσ : SC]]$.
+    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
 \end{lemma}
@@
 
-            By inversion, $[[Γ ; Θ ⊢ ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some 
+            By inversion, $[[Γ ; dom(Θ) ⊢  ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some 
             $[[{Δ} ⊆ {Γ}]]$.
@@
             that is $[[Δ ⊢ [uσ]α̂⁺ ≥ iQ0']]$, which means 
-            $[[Θ ⊢ uσ : (α̂⁺ :≥ iQ0')]]$, that is $[[Θ ⊢ uσ : SC]]$.
+            $[[(Θ  ⊢  (α̂⁺ :≥ iQ0')) ⊢ uσ]]$, that is $[[(Θ  ⊢  SC) ⊢ uσ]]$.
 
@@
         infers $[[SC]] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
-        Then $[[Θ ⊢ uσ : ·]]$ holds trivially.
+        Then $[[(Θ  ⊢  ·) ⊢ uσ]]$ holds trivially.
 
@@
             \item $[[nf(uN)]]$ and $[[nf(iM)]]$ are normalized 
-            \item $[[Γ ; Θ ⊢ nf(uN)]]$ because $[[Γ ; Θ ⊢ uN]]$ (\cref{todo})
+            \item $[[Γ ; dom(Θ) ⊢  nf(uN)]]$ because $[[Γ ; dom(Θ) ⊢  uN]]$ (\cref{corollary:wf-nf-algo})
             \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
-            \item $[[ Θ ⊢ nf(uσ) ]]$ because $[[Θ ⊢ uσ ]]$ (\cref{todo})
+            \item $[[ Θ ⊢ nf(uσ) ]]$ because $[[Θ ⊢ uσ ]]$ (\cref{corollary:norm-subst-sig-algo})
             \item $ \begin{aligned}[t]
@@
         $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ exists, and
-        $[[Θ ⊢ nf(uσ) : lift UC]]$.  Then by \cref{todo}, $[[Θ ⊢ uσ : lift UC]]$.
+        $(Θ  ⊢ lift UC) ⊢ nf(uσ)$.  Then by \cref{corollary:sat-equiv}, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
 
@@
       To show that the recursive call terminates and that 
-      $[[Θ ⊢ uσ : SC1 \ {â⁻*}]]$,
+      $[[(Θ  ⊢  SC1 \ {â⁻*}) ⊢ uσ]]$,
       it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0]]$---an extension of $[[uσ]]$ with
       $[[â⁻*]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
-      Then by the induction hypothesis, $[[Θ, â⁻*[G, nbs] ⊢ us0 : SC1]]$,
-      and hence, $[[Θ ⊢ uσ : SC1 \ {â⁻*}]]$, as required.
+      Then by the induction hypothesis, $[[(Θ, â⁻*[G, nbs] ⊢ SC1) ⊢ us0]]$,
+      and hence, $[[(Θ  ⊢  SC1 \ {â⁻*}) ⊢ uσ]]$, as required.
 

### 2023-08-21 proofs/tex/constraint-merge-lemmas.tex
WIP: description, notation

@@
         \item $[[Θ ⊢ SC]]$,
-        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        \item for any substitution $[[Θ ⊢ uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$ implies $[[(Θ  ⊢  SC1) ⊢ uσ]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$.
     \end{enumerate}
@@
     Let us show the second property.
-    We take an arbitrary $[[uσ]]$ such that $[[Θ ⊢ uσ]]$ and $[[Θ ⊢ uσ : SC]]$.
-    To prove $[[Θ ⊢ uσ : SC1]]$, we need to show that for any $[[scE1]] \in [[SC1]]$, restricting $[[α̂±]]$,
+    We take an arbitrary $[[uσ]]$ such that $[[Θ ⊢ uσ]]$ and $[[(Θ  ⊢  SC) ⊢ uσ]]$.
+    To prove $[[(Θ  ⊢  SC1) ⊢ uσ]]$, we need to show that for any $[[scE1]] \in [[SC1]]$, restricting $[[α̂±]]$,
     $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ holds.
@@
     Let us assume that $[[α̂±]] \notin [[dom(SC2)]]$. It means that $[[SC]] \ni [[scE1]]$, 
-    and then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
+    and then since $[[(Θ  ⊢  SC) ⊢ uσ]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
 
@@
     and $[[SC]] \ni [[scE]]$ where $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$.
-    Then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
+    Then since $[[(Θ  ⊢  SC) ⊢ uσ]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
     and by \cref{lemma:entry-merge-soundness}, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
 
-    The proof of $[[Θ ⊢ uσ : SC2]]$ is symmetric.
+    The proof of $[[(Θ  ⊢  SC2) ⊢ uσ]]$ is symmetric.
 \end{proof}
@@
     Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
-    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
+    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[(Θ  ⊢  SC1) ⊢ uσ]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$, 
     $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
@@
     Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.
-$[[Θ ⊢ uσ : SC1]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[Θ ⊢ uσ : SC2]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
+$[[(Θ  ⊢  SC1) ⊢ uσ]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
     Then by \cref{lemma:entry-merge-completeness}, $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ is defined and $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$.
@@
     \label{lemma:substitution-existence}
-    If $[[Θ ⊢ SC]]$ then there exists $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC]]$.
+    If $[[Θ ⊢ SC]]$ then there exists $[[Θ ⊢ uσ]]$ such that $[[(Θ  ⊢  SC) ⊢ uσ]]$.
 \end{lemma}

### 2023-08-21 proofs/tex/equiv-lemmas.tex
WIP: description, notation

@@
   and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where
-  by size we mean the size of the nodes in the corresponding inference tree.
+  by size, we mean the size of the nodes in the corresponding inference tree.
 
@@
   \label{lemma:decl-equiv-algorithmization}
-  Declarative equivalence is equality of normal forms. 
+  Declarative equivalence is the equality of normal forms. 
   \begin{itemize}
@@
   \begin{itemize}
-  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Θ ⊢ uP]] \iff [[Γ; Θ ⊢ uQ]]$,
-  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Θ ⊢ uN]] \iff [[Γ; Θ ⊢ uM]]$
+  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
+  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
   \end{itemize}
@@
   which must be equal by inversion, and thus, 
-  $[[Γ; Θ ⊢ α̂±]] \iff [[Γ;  ⊢ α̂±]]$ holds trivially.
+  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
 \end{proof}
@@
 \end{lemma}
-\being{proof}
+\begin{proof}
   Suppose that $[[α± ∊ {Γ1}]]$. 
@@
   \begin{itemize}
-    \item[$+$] $[[Γ; Θ ⊢ uP]] \iff [[Γ; Θ ⊢ nf(uP)]]$,
-    \item[$-$] $[[Γ; Θ ⊢ uN]] \iff [[Γ; Θ ⊢ nf(uN)]]$
+    \item[$+$] $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ nf(uP)]]$,
+    \item[$-$] $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ nf(uN)]]$
   \end{itemize}
@@
 
-\begin{corollary}
-  \label{corollary:wf-s-nf-algo}
+\begin{corollary}[Normalization preserves the signature of the algorithmic substitution]
+  \label{corollary:norm-subst-sig-algo}
   \hfill \\
-  $[[Θ ⊢ uσ]] \iff [[Θ ⊢ nf(σ)]]$
+  $[[Θ ⊢ uσ]] \iff [[Θ ⊢ nf(uσ)]]$ and
+  $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
 \end{corollary}
@@
 
+\begin{corollary}[Algorithmic substitution equivalence becomes equality after normalization]
+  \label{corollary:subst-subt-equiv-algorithmization}
+  Suppose that $[[Θ ⊢ uσ1]]$ and $[[Θ ⊢ uσ2]]$ are algorithmic substitutions.
+  Then
+  $[[Θ ⊢ uσ1 ≈ uσ2]] \iff [[nf(uσ1)|dom(Θ) = nf(uσ2)|dom(Θ)]]$
+  and 
+  $[[Γ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
+\end{corollary}
+
 \begin{lemma}[Substitution preserves declarative equivalence]

### 2023-08-21 proofs/tex/singularity-lemmas.tex
wip: leftovers
WIP: description, notation

@@
     Suppose $[[Θ ⊢ SC]]$, and $[[SC singular with uσ]]$. 
-    Then $[[Θ ⊢ uσ : SC]]$ and for any 
-    $[[uσ']]$ such that $[[Θ ⊢ uσ' : SC]]$,
-    $[[Θ ⊢ uσ' ≈ uσ : dom(SC)]]$.
+    Then $[[(Θ  ⊢  SC) ⊢ uσ]]$ and for any 
+    $[[uσ']]$ such that $(Θ  ⊢ SC) ⊢ uσ'$,
+    $[[Θ|dom(SC) ⊢ uσ' ≈ uσ]]$.
 \end{lemma}
 \begin{proof}
-    Suppose that $[[Θ ⊢ uσ' : SC]]$.
+    Suppose that $(Θ  ⊢ SC) ⊢ uσ'$.
     It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
@@
 \begin{lemma}[Completeness of Singularity]
-\end{lemma}
-
-\begin{lemma}[Completeness of Singularity]
     \label{lemma:singularity-completeness}
-    Suppose there exists $[[Θ ⊢ uσ1]]$ such that
-    for any $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ ≈ uσ1 : varset]]$.
+    For a given $[[Θ ⊢ SC]]$
+    and $[[varset ⊆ dom(Θ)]]$,
+    suppose that all the substitutions satisfying $[[SC]]$ are equivalent,
+    in other words, suppose that there exists $[[Θ ⊢ uσ1]]$ such that
+    for any $[[Θ ⊢ uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$ implies $[[Θ|varset ⊢ uσ ≈ uσ1]]$.
     Then 
@@
 \begin{proof}
+    First, let us suppose that $[[varset ⊆ dom(SC)]]$ does not hold.
+    It means that there exists $[[ α̂±[Γ] ∊ Θ]]$ and $[[α̂± ∉ dom(SC)]]$
+
+    By \cref{lemma:constraint-sat}, let us take an arbitrary 
+    $[[uσ']]$ satisfying $[[SC]]$ (that is $[[(Θ  ⊢  SC) ⊢ uσ']]$).
+    Let us extend $[[uσ']]$ to $[[uσ'']]$ so that $[[Θ ⊢ uσ'']]$.
+
+    Although $[[uσ']]$ might not be defined on
+
+    such that for any $[[Θ ⊢ uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$ implies 
+    $[[ Γ ⊢ [uσ]α̂± ≈ [uσ1]α̂± ]]$.
+    Let us take
+
+    for any $[[Θ ⊢ uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$ implies $[[Θ|varset ⊢ uσ ≈ uσ1]]$.
+
 \end{proof}
+
+        \item [$-$] Suppose that there exists $[[iN]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iN']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iN' : scE]]$ implies $[[Γ ⊢ iN' ≈ iN]]$. 
+            Then $[[scE singular with nf(iN)]]$.
+        \item [$+$] Suppose that there exists $[[iP]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iP']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iP' : scE]]$ implies $[[Γ ⊢ iP' ≈ iP]]$. 
+            Then $[[scE singular with nf(iP)]]$ .

### 2023-08-21 proofs/tex/subst-lemmas.tex
WIP: description, notation

@@
 
-\begin{lemma}[Substitutions equivalent on the metavariables]
+\begin{lemma}[Substitutions equal on the metavariables]
+  \label{lemma:subst-eq-metavar}
+  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
+  normalized substitutions of signature $[[Θ ⊢ uσi]]$.
+  Then 
+  \begin{enumerate}
+    \item [$+$] for a normalized type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
+      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
+    \item [$-$] for a normalized type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
+      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
+  \end{enumerate}
+\end{lemma}
+\begin{proof}
+  The proof is a simple structural induction on 
+  $[[Γ; dom(Θ) ⊢ uP]]$ and mutually, on $[[Γ; dom(Θ) ⊢ uN]]$.
+  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
+  \begin{caseof}
+    \item $[[Γ; dom(Θ) ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
+      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
+    \item $[[Γ; dom(Θ) ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
+      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
+    \item $[[Γ; dom(Θ) ⊢ ∀pas.uN]]$. Then we are proving that
+      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
+      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
+      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
+      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; dom(Θ) ⊢ uN]]$, 
+      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
+      as required.
+    \item $[[Γ; dom(Θ) ⊢ uP → uN]]$. Then we are proving that
+      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
+      By definition of substitution and congruence of equality, 
+      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
+      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
+      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
+      and well-formed in the same contexts.
+      This way, by the induction hypothesis, 
+      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
+      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
+      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
+    \item $[[Γ; dom(Θ) ⊢ ↑uP]]$. The proof is similar to the previous case:
+      we apply congruence of substitution, equality, and normalization,
+      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
+  \end{caseof}
+\end{proof}
+
+\begin{corollary}[Substitutions equivalent on the metavariables]
   \label{lemma:subst-equiv-metavar}
-  Suppose that $[[Γ ⊢ Θ]]$, $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
+  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
   of signature $[[Θ ⊢ uσi]]$.
@@
   \begin{enumerate}
-    \item [$+$] for a type $[[Γ; Θ ⊢ uP]]$, if $[[Γ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
-    \item [$-$] for a type $[[Γ; Θ ⊢ uN]]$, if $[[Γ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
+    \item [$+$] for a type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
+      $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$;
+    \item [$-$] for a type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
+      $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
   \end{enumerate}
-\end{lemma}
+\end{corollary}
 \begin{proof}
-  The proof is a trivial structural induction on 
-  $[[Γ; Θ ⊢ uP]]$ and mutually, on $[[Γ; Θ ⊢ uN]]$.
+  First, let us normalize the types and the substitutions, and show that
+  the given equivalences and well-formedness properties are preserved. 
+  $[[Γ; dom(Θ) ⊢ uP]]$ implies $[[Γ; dom(Θ) ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
+  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
+  \cref{lemma:subt-equiv-algorithmization}.
+  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
+  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
+  Notice that by \cref{corollary:norm-subst-sig-algo}
+  $[[Θ ⊢ uσi]]$ implies $[[Θ ⊢ nf(uσi)]]$.
+
+  This way, by \cref{lemma:subst-eq-metavar}, 
+  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
+  $[[nf(uσ1)|(uv nf(uP)) = nf(uσ2)|(uv nf(uP))]]$.
+  Then by \cref{lemma:uv-nf}, 
+  $[[nf(uσ1)|(uv uP) = nf(uσ2)|(uv uP)]]$,
+  and by ${{corollary:subst-subt-equiv-algorithmization}}$,
+  $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$.
+
+  Symmetrically, 
+  $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
+  $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
 \end{proof}

### 2023-08-21 proofs/tex/unif-constraint-merge-lemmas.tex
WIP: description, notation

@@
         \item $[[Θ ⊢ UC]]$ is normalized unification constraint,
-        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : lift UC]]$ implies 
-        $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+        \item for any substitution $[[Θ ⊢ uσ]]$, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$ implies 
+        $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$.
     \end{enumerate}
@@
     Let us take an arbitrary substitution $[[Θ ⊢ uσ]]$ and assume that 
-    $[[Θ ⊢ uσ : lift UC]]$.
-    Then $[[Θ ⊢ uσ : lift UCi]]$ holds by definition:
+    $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+    Then $[[(Θ  ⊢  lift UCi) ⊢ uσ]]$ holds by definition:
     If $[[ucE]] \in [[lift UCi]] \subseteq [[lift UC1 ∪ lift UC2]] = [[lift UC]]$.
@@
     Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
-    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$, 
+    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$, 
     \begin{enumerate}
         \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
-        \item $[[Θ ⊢ uσ : UC]]$.
+        \item $[[(Θ  ⊢  UC) ⊢ uσ]]$.
     \end{enumerate}

### 2023-08-21 proofs/tex/unification-lemmas.tex
WIP: leftovers
WIP: description, notation

@@
         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
-        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
+        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
         if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
-        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : lift UC]]$,
+        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
         $[[ [uσ]uP = iQ ]]$.
@@
         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
+        $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
         if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
-        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[Θ ⊢ uσ : lift UC]]$,
+        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
         $[[ [uσ]uN = iM ]]$.
@@
         The algorithm makes a recursive call to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ returning $[[UC]]$.
-        By induction hypothesis, $[[Θ ⊢ UC]]$ and for any $[[Θ ⊢ uσ : lift UC]]$,
+        By induction hypothesis, $[[Θ ⊢ UC]]$ and for any $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
         $[[ [uσ]uN ]] = [[ [uσ]↑uP ]] = [[ ↑[uσ]uP ]] = [[ ↑iQ ]] = [[ iM ]]$, as 
@@
         It is clear that $[[uP]]$, $[[uN']]$, $[[iQ]]$, and $[[iM']]$ are normalized,
-        and that $[[Γ ; Θ ⊢ uP]]$, $[[Γ ; Θ ⊢ uN']]$, $[[Γ ⊢ iQ]]$, and $[[Γ ⊢ iM']]$.
+        and that $[[Γ ; dom(Θ) ⊢ uP]]$, $[[Γ ; dom(Θ) ⊢  uN']]$, $[[Γ ⊢ iQ]]$, and $[[Γ ⊢ iM']]$.
         This way, the induction hypothesis is applicable to both recursive calls.
@@
             \item $[[Θ ⊢ UC1]]$,
-            \item for any $[[Θ ⊢ uσ' : lift UC1]]$, $[[ [uσ']uP = iQ ]]$.
+            \item for any $(Θ  ⊢ lift UC1) ⊢ uσ'$, $[[ [uσ']uP = iQ ]]$.
         \end{itemize}
@@
             \item $[[Θ ⊢ UC2]]$,
-            \item for any $[[Θ ⊢ uσ' : lift UC2]]$, $[[ [uσ']uN' = iM' ]]$.
+            \item for any $[[(Θ ⊢ lift UC2) ⊢ uσ']]$, $[[ [uσ']uN' = iM' ]]$.
         \end{itemize}
 
-        Let us take an arbitrary $[[Θ ⊢ uσ : lift UC]]$.
+        Let us take an arbitrary $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
         By the soundness of the constraint merge (\cref{lemma:merge-soundness}), 
         $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$ implies
-        $[[Θ ⊢ uσ : lift UC1 ]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+        $[[(Θ  ⊢  lift UC1 ) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$.
 
-        Applying the induction hypothesis to $[[Θ ⊢ uσ : lift UC1]]$, we have
-        $[[ [uσ]uP = iQ ]]$; applying it to $[[Θ ⊢ uσ : lift UC2]]$, we have
+        Applying the induction hypothesis to $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$, we have
+        $[[ [uσ]uP = iQ ]]$; applying it to $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$, we have
         $[[ [uσ]uN' = iM' ]]$.
@@
 
-        The induction hypothesis is applicable: $[[Γ,pas ; Θ ⊢ uN']]$ and $[[Γ,pas ⊢ iM']]$ hold
+        The induction hypothesis is applicable: $[[Γ,pas ; dom(Θ) ⊢  uN']]$ and $[[Γ,pas ⊢ iM']]$ hold
         by inversion, and $[[uN']]$ and $[[iM']]$ are normalized, since $[[uN]]$ and $[[iM]]$ are.
-        Let us take an arbitrary $[[Θ ⊢ uσ : lift UC]]$.
+        Let us take an arbitrary $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
         By the induction hypothesis, $[[ [uσ]uN' ]] = [[ iM' ]]$. 
@@
 
-        Let us take an arbitrary $[[uσ]]$ such that  $[[Θ ⊢ uσ : lift UC]]$.
-        Since $[[UC]] = [[ (â⁻ :≈ iM) ]]$, $[[Θ ⊢ uσ : lift UC]]$ implies 
+        Let us take an arbitrary $[[uσ]]$ such that  $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+        Since $[[UC]] = [[ (â⁻ :≈ iM) ]]$, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$ implies 
         $[[Θ(â⁻) ⊢ [uσ]â⁻ : (â⁻ :≈ iM) ]]$.
@@
         This way, $[[Θ(â⁻) ⊢ [uσ]uN ≈ iM]]$. 
-        Notice that $[[uσ]]$ and $[[uN]]$ are normalized, and by \cref{todo}, so is $[[ [uσ]uN ]]$.
+        Notice that $[[uσ]]$ and $[[uN]]$ are normalized, and by \cref{lemma:norm-subst-distr}, 
+        so is $[[ [uσ]uN ]]$.
         Since both sides of $[[Θ(â⁻) ⊢ [uσ]uN ≈ iM]]$ are normalized,
@@
         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
-        $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
+        $[[Γ ; dom(Θ) ⊢  uP]]$ and $[[Γ ⊢ iQ]]$, 
         for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uP = iQ ]]$,
         there exists $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$,
-        and $[[Θ ⊢ uσ : lift UC]]$.
+        and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
 
         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; Θ ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
+        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,\\
         for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uN = iM ]]$,
         there exists $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$,
-        and $[[Θ ⊢ uσ : lift UC]]$.
+        and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
    \end{itemize}
@@
         \item $[[uN]] = [[α̂⁻]]$\\
-            $[[Γ ; Θ ⊢ α̂⁻]]$ means that $[[ α̂⁻[Δ] ]] \in [[Θ]]$ for some $[[Δ]]$.
+            $[[Γ ; dom(Θ) ⊢  α̂⁻]]$ means that $[[ α̂⁻[Δ] ]] \in [[Θ]]$ for some $[[Δ]]$.
 
@@
             $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
-            $[[Θ ⊢ uσ : lift (â⁻ :≈ iM)]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
+            $[[(Θ  ⊢  lift (â⁻ :≈ iM)) ⊢ uσ]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
 
@@
             $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ ·]]$, 
-            and $[[Θ ⊢ uσ : lift ·]]$ holds trivially.
+            and $[[(Θ  ⊢  lift ·) ⊢ uσ]]$ holds trivially.
 
@@
             Notice that $[[uP]]$ is normalized, since $[[uN]] = [[↑uP]]$ is normalized,
-            $[[Γ ; Θ ⊢ uP]]$ holds by inversion of $[[Γ ; Θ ⊢ ↑uP]]$, 
+            $[[Γ ; dom(Θ) ⊢  uP]]$ holds by inversion of $[[Γ ; dom(Θ) ⊢  ↑uP]]$, 
             and $[[Γ ⊢ iQ]]$ holds by inversion of $[[Γ ⊢ ↑iQ]]$.
@@
             This way, by the induction hypothesis there exists $[[UC]]$ such that
-            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[Θ ⊢ uσ : lift UC]]$.
+            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
 
@@
                 \item $[[uP]]$ and $[[uN']]$ are normalized, since $[[uN]] = [[uP → uN']]$ is normalized
-                \item $[[Γ ; Θ ⊢ uP]]$ and $[[Γ ; Θ ⊢ uN']]$ follow from the inversion of $[[Γ ; Θ ⊢ uP → uN']]$,
+                \item $[[Γ ; dom(Θ) ⊢  uP]]$ and $[[Γ ; dom(Θ) ⊢  uN']]$ follow from the inversion of $[[Γ ; dom(Θ) ⊢  uP → uN']]$,
                 \item $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$ follow from inversion of $[[Γ ⊢ iQ → iM']]$.
@@
 
-            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[Θ ⊢ uσ : lift UC1]]$,
-            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$,
+            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$.
             To apply \ruleref{\ottdruleUArrowLabel} and infer the required
             $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$, we need to show that
-            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[Θ ⊢ uσ : lift UC]]$.
+            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
             It holds by the completeness of the unification constraint merge 
@@
                 \item $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ holds by the soundness of unification (\cref{lemma:unification-soundness})
-                \item $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$ holds as noted above 
+                \item $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$ holds as noted above 
             \end{itemize}.
@@
             Notice that $[[uN']]$ is normalized, since $[[uN]] = [[∀pas.uN']]$ is normalized,
-            $[[Γ,pas ; Θ ⊢ uN']]$ follows from inversion of $[[Γ ; Θ ⊢ ∀pas.uN']]$,
+            $[[Γ,pas ; dom(Θ) ⊢  uN']]$ follows from inversion of $[[Γ ; dom(Θ) ⊢  ∀pas.uN']]$,
             $[[Γ,pas ⊢ iM']]$ follows from inversion of $[[Γ ⊢ ∀pas.iM']]$, and
@@
             This way, by the induction hypothesis, $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ UC]]$ exists and 
-            moreover, $[[Θ ⊢ uσ : lift UC]]$.
+            moreover, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
             Hence, \ruleref{\ottdruleUForallLabel} is applicable to infer

### 2023-08-29 proofs/main.tex
WIP: update typing lemmas

@@
 
-The signature $[[Θ ⊢ uσ]]$ means that $[[uσ]]$ maps each algorithmic 
-variable appearing in $[[Θ]]$ to a type well-formed in the corresponding
-context; and for each variable not appearing in $[[dom(Θ)]]$, acts as identity.
+The signature $[[Θ ⊢ uσ : Ξ]]$ means that
+$[[Ξ ⊆ dom(Θ)]]$ and 
+$[[uσ]]$ maps each algorithmic variable 
+from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
+and for each variable not appearing in $[[dom(Θ)]]$, 
+it acts as identity.
 
 \begin{definition}[Signature of Algorithmic Substitution]
-  $[[Θ ⊢ uσ]]$ means that
+  $[[Θ ⊢ uσ : Θ]]$ means that
   \begin{enumerate}
-    \item for any $[[ α̂±[Γ] ∊ Θ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$ and
-    \item for any $[[ α̂± ∉ dom(Θ)]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
+    \item for any $[[α̂± ∊ Ξ]]$,
+      there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
+      and $[[ Γ ⊢ [uσ]α̂± ]]$; 
+    \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
   \end{enumerate}
@@
 In contrast to algorithmic substitution $[[uσ]]$,
-it only defined on the negative algorithmic variables, 
-and it allows mapping algorithmic variables to
+it allows mapping algorithmic variables to
 \emph{algorithmic} terms.

### 2023-08-29 proofs/tex/alg-neg-subtyping-lemmas.tex
WIP: update typing lemmas

@@
         $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
-        $[[Θ ⊢ SC]]$ and 
-        for any normalized $[[uσ]]$ such that $[[(Θ ⊢ SC) ⊢ uσ]]$,
+        $[[Θ ⊢ SC : uv(uN)]]$ and 
+        for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
         $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
@@
 
-    Suppose that $[[uσ]]$ is normalized and $[[(Θ ⊢ SC) ⊢ uσ]]$,
+    Suppose that $[[uσ]]$ is normalized and $[[ Θ  ⊢ uσ : SC ]]$,
     Let us consider the last rule to infer this judgment. 
@@
         By the soundness of constraint merge (\cref{lemma:merge-soundness}),
-        $[[(Θ ⊢ SC1) ⊢ uσ]]$ and $[[(Θ ⊢ SC2) ⊢ uσ]]$.
+        $[[ Θ  ⊢ uσ : SC1 ]]$ and $[[ Θ  ⊢ uσ : SC2 ]]$.
         Then by the soundness of positive subtyping (\cref{lemma:pos-subt-soundness}), $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$; 
@@
     and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
-    Then for any $[[Θ ⊢ uσ]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
-    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, such that $[[Θ ⊢ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
+    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
+    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
 \end{lemma}
@@
             This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
-        obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2]]$ and $[[(Θ  ⊢  SC2) ⊢ uσ]]$.
+            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv(uN')]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
             Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
-        $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1]]$ and $[[(Θ  ⊢  SC1) ⊢ uσ]]$.
+            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv(uP)]]$
+            and $[[ Θ ⊢ uσ : SC1 ]]$.
 
@@
             satisfies the required properties.
-        To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness}).
-        This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, $[[Θ ⊢ SC]]$, and $[[(Θ  ⊢  SC) ⊢ uσ]]$, as required.
+            To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness})
+            (notice that $[[Θ ⊢ uσ : uv(uP → uN')]]$ means 
+            $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$).
+            This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined and $[[ Θ ⊢ uσ : SC ]]$ holds.
 

### 2023-08-29 proofs/tex/alg-pos-subtyping-lemmas.tex
WIP: update typing lemmas

@@
     $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
-    then $[[Θ ⊢ SC]]$ and
-    for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  SC) ⊢ uσ]]$,
+    then $[[Θ ⊢ SC : uv uP]]$ and
+    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
     $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
@@
         Since $[[â⁺[Δ] ∊ Θ]]$ and $[[Δ ⊢ iQ']]$, 
-    it is clear that $[[ Θ ⊢ (â⁺ :≥ iQ') ]]$.
+        it is clear that $[[ Θ ⊢ (â⁺ :≥ iQ') : â⁺]]$.
 
         It is left to show that $[[Γ ⊢ [uσ]â⁺ ≥ iP']]$ for any normalized $[[uσ]]$ 
-    s.t. $[[(Θ  ⊢  (â⁺ :≥ iQ')) ⊢ uσ]]$.
+        s.t. $[[ Θ   ⊢ uσ : (â⁺ :≥ iQ') ]]$.
         The latter means that $[[ Θ(â⁺) ⊢ [uσ]â⁺ ≥ iQ' ]]$, i.e. $[[Δ ⊢ [uσ]â⁺ ≥ iQ']]$. 
@@
         then $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ has shape $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$ .
-    Then $[[uv a⁺ = ∅]]$, and $[[SC]] = [[·]]$ satisfies $[[Θ ⊢ ·]]$.
+        Then $[[uv a⁺ = ∅]]$, and $[[SC]] = [[·]]$ satisfies $[[Θ ⊢ SC : ·]]$.
         Since $[[uv a⁺ = ∅]]$, application of any substitution $[[uσ]]$ 
@@
         By the soundness of unification (\cref{lemma:unification-soundness}),
-    $[[Θ ⊢ SC]]$ and for any normalized $[[uσ]]$, $[[(Θ  ⊢  SC) ⊢ uσ]]$
+        $[[Θ ⊢ SC : uv(uN)]]$ and for any normalized $[[uσ]]$, $[[ Θ ⊢ uσ : SC ]]$
         implies $[[ [uσ]nf(uN) = nf(iM) ]]$, 
@@
 
-        Let us take an arbitrary normalized $[[uσ]]$ s.t. $[[(Θ  ⊢  SC' \ {α̂⁻*}) ⊢ uσ]]$.
+        Let us take an arbitrary normalized $[[uσ]]$ s.t. $[[ Θ   ⊢ uσ : SC' \ {α̂⁻*} ]]$.
         We wish to show $[[Γ ⊢ [uσ]uP ≥ iQ]]$, i.e. $[[Γ ⊢ ∃nas.[uσ]uP' ≥ ∃nbs.iQ']]$.
@@
         $[[Γ, nbs ⊢ [iNs/nas][uσ]uP' ≥ iQ']]$ for some $[[iNs]]$.
-        If we construct a normalized $[[uσ']]$ such that $[[(Θ, â⁻*[G, nbs] ⊢ SC') ⊢ uσ']]$
+        If we construct a normalized $[[uσ']]$ such that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']]$
         and for some $[[iNs]]$, $[[ [iNs/nas][uσ]uP' ]] = [[ [uσ'][â⁻*/nas]uP' ]]$,
@@
         $[[uσ]]$.
-        Let us show that $[[(Θ, â⁻*[G, nbs]  ⊢ SC') ⊢ uσ' ]] $.
+        Let us show that $[[Θ, â⁻*[G, nbs] ⊢ uσ' : SC']] $.
         Let us take an arbitrary entry $[[scE]]$ from $[[SC']]$ restricting a variable $[[β̂±]]$.
@@
         $[[ (Θ, â⁻*[G, nbs])(β̂±) ⊢ [uσ']β̂± : scE ]]$ is
-        rewritten as $[[ Θ(β̂±) ⊢ [uσ]β̂± : scE ]]$, which holds since $[[(Θ  ⊢  SC') ⊢ uσ]]$.
+        rewritten as $[[ Θ(β̂±) ⊢ [uσ]β̂± : scE ]]$, which holds since $[[ Θ   ⊢ uσ : SC' ]]$.
         Suppose $[[β̂±]] = [[αî⁻]] \in [[α̂⁻*]]$. Then
@@
             \item If this variable is a unification variable $[[β̂±]]$, then
-                $[[ [iNs/nas][uσ] β̂± ]] = [[ [uσ]β̂± ]] $, since $[[(Θ  ⊢  SC' \ {α̂⁻*}) ⊢ uσ]]$ and 
+                $[[ [iNs/nas][uσ] β̂± ]] = [[ [uσ]β̂± ]] $, since $[[ Θ   ⊢ uσ : SC' \ {α̂⁻*} ]]$ and 
                 $[[ dom(Θ) ∩ {nas} = ∅ ]]$.
@@
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
-    Then for any $[[Θ ⊢ uσ]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
-    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[(Θ  ⊢  SC) ⊢ uσ]]$.
+    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
+    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
 \end{lemma}
@@
             \begin{enumerate}
-                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
+                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ : α̂⁺]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
                 \item $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$ as noted above
@@
             that is $[[Δ ⊢ [uσ]α̂⁺ ≥ iQ0']]$, which means 
-            $[[(Θ  ⊢  (α̂⁺ :≥ iQ0')) ⊢ uσ]]$, that is $[[(Θ  ⊢  SC) ⊢ uσ]]$.
+            $[[ Θ   ⊢ uσ : (α̂⁺ :≥ iQ0') ]]$, that is $[[ Θ   ⊢ uσ : SC ]]$.
 
@@
             infers $[[SC]] = [[·]]$, i.e. $[[Γ;Θ ⊨ a⁺ ≥ a⁺ ⫤ ·]]$.
-        Then $[[(Θ  ⊢  ·) ⊢ uσ]]$ holds trivially.
+            Then $[[ Θ   ⊢ uσ : · ]]$ holds trivially.
 
-        \item \label{case:pos-subt-complete-upshift} $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ 
+        \item \label{case:pos-subt-complete-upshift} 
+            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ 
             is derived by \ruleref{\ottdruleDOneShiftDLabel},
@@
                 \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
-            \item $[[ Θ ⊢ nf(uσ) ]]$ because $[[Θ ⊢ uσ ]]$ (\cref{corollary:norm-subst-sig-algo})
+                \item $[[ Θ ⊢ nf(uσ) : uv(uP) ]]$ because $[[Θ ⊢ uσ : uv(uP) ]]$ (\cref{corollary:norm-subst-sig-algo})
                 \item $ \begin{aligned}[t]
@@
             \end{itemize}
-        Then by the completeness of the unification,
+            By the completeness of the unification,
             $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ exists, and
-        $(Θ  ⊢ lift UC) ⊢ nf(uσ)$.  Then by \cref{corollary:sat-equiv}, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+            $[[Θ ⊢ nf(uσ) : lift UC]]$,
+            and by \cref{corollary:sat-equiv}, $[[ Θ ⊢ uσ : lift UC ]]$.
 
-      \item $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
+        \item \label{case:pos-subt-complete-exists}
+            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ is derived by \ruleref{\ottdruleDOneExistsLabel}.\\
             We should only consider the case
@@
             To show that the recursive call terminates and that 
-      $[[(Θ  ⊢  SC1 \ {â⁻*}) ⊢ uσ]]$,
-      it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0]]$---an extension of $[[uσ]]$ with
-      $[[â⁻*]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
-      Then by the induction hypothesis, $[[(Θ, â⁻*[G, nbs] ⊢ SC1) ⊢ us0]]$,
-      and hence, $[[(Θ  ⊢  SC1 \ {â⁻*}) ⊢ uσ]]$, as required.
+            $[[ Θ ⊢ uσ : SC1 \ {â⁻*} ]]$,
+            it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0 : uv(uP0)]]$---an extension of $[[uσ]]$ with
+            $[[{â⁻*} ∩ uv(uP0)]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
+            Then by the induction hypothesis, $[[Θ, â⁻*[G, nbs] ⊢ us0 : SC1]]$,
+            and hence, $[[ Θ  ⊢ uσ : SC1 \ {â⁻*} ]]$, as required.
 
@@
                     \begin{cases}
-               [[ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[â⁻*]] \\
+                    [[ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[{â⁻*} ∩ uv(uP0)]] \\
                     [[ [uσ]β̂± ]] & \text{if } [[β̂±]] \in [[ uv(uP') ]]
@@
             \]
-       It is easy to see $[[Θ, â⁻*[Γ, nbs] ⊢ uσ0]]$:
+            It is easy to see $[[Θ, â⁻*[Γ, nbs] ⊢ uσ0 : uv(uP0)]]$:
+            $[[uv(uP0)]] = [[ uv([â⁻*/nas]uP') ]] = [[{â⁻*} ∩ uv(uP0) ∪ uv(uP')]]$. Then
             \begin{enumerate}
-            \item for $[[αî⁻]] \in [[â⁻*]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
+                    \item for $[[αî⁻]] \in [[{â⁻*} ∩ uv(uP0)]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
                     i.e. $[[ Γ, nbs ⊢ [σ]αi⁻ ]]$ holds since $[[Γ, nbs ⊢ σ : nas]]$,
                     \item for $[[β̂±]] \in [[ uv(uP') ]] \subseteq [[dom(Θ)]]$, $[[ (Θ, â⁻*[Γ, nbs])(β̂±) ⊢ [uσ0] β̂± ]]$,
-            i.e. $[[Θ(β̂±) ⊢ [uσ] β̂± ]]$ holds since $[[Θ ⊢ uσ]]$.
+                    i.e. $[[Θ(β̂±) ⊢ [uσ] β̂± ]]$ holds since $[[Θ ⊢ uσ : uv(uP)]]$ and $[[β̂±]] \in [[ uv(uP') ]] = [[uv(uP)]]$.
             \end{enumerate}
@@
             and thus, $[[ Γ, nbs ⊢ [uσ0]uP0 ≥ iQ' ]]$.
-
     \end{caseof}

### 2023-08-29 proofs/tex/constraint-merge-lemmas.tex
WIP: update typing lemmas

--- 2023-07-17 proofs/tex/constraint-merge-lemmas.tex
@@
                 \item $[[Γ ⊢ scE]]$, i.e. $[[Γ ⊢ pua :≈ iQ]]$ holds by assumption;
-            \item by inversion, $[[Γ ⊢ iP : (pua :≈ iQ)]]$ means $[[Θ ⊢ iP ≈ iQ]]$,
+                \item by inversion, $[[Γ ⊢ iP : (pua :≈ iQ)]]$ means $[[Γ ⊢ iP ≈ iQ]]$,
                 and by transitivity of equivalence (\cref{corollary:equivalence-transitivity}), 
-            $[[Θ ⊢ iP ≈ iQ']]$. Thus, $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ hold
+                $[[Γ ⊢ iP ≈ iQ']]$. Thus, $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ hold
                 by \ruleref{\ottdruleSATSCEPEqLabel}.
@@
 \begin{lemma} [Soundness of Constraint Merge] \label{lemma:merge-soundness}
-    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$ 
+    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
     and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
@@
     \begin{enumerate}
-        \item $[[Θ ⊢ SC]]$,
-        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$.
+        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
+        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
+            $[[ Θ ⊢ uσ : SC ]]$
+            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
     \end{enumerate}
@@
 \begin{proof}
-    By definition, $[[SC1 & SC2]] = [[SC]]$ consists of three parts:
-    entries of $[[SC1]]$ that do not have matching entries of $[[SC]]$,
+    By definition, $[[Θ ⊢ SC1 & SC2 = SC]]$ consists of three parts:
+    entries of $[[SC1]]$ that do not have matching entries of $[[SC2]]$,
     entries of $[[SC2]]$ that do not have matching entries of $[[SC1]]$,
@@
 
-    Let us show $[[Θ ⊢ SC]]$.
-    First, let us assume that an entry $[[scE]] \in [[SC]]$ belongs to the first group, 
-    i.e. $[[scE]] \in [[SC1]]$.  Let us denote the variable $[[scE]]$ as $[[α̂±]]$. 
-    Then $[[Θ(α̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC1]] \ni [[scE]]$.
-    Analogously, if $[[scE]]$ belongs to the second group, then $[[Θ(α̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC2]] \ni [[scE]]$.
-    Finally, if $[[scE]]$ belongs to the third group, then $[[scE]]$ is a merge of two entries $[[Θ(α̂±) ⊢ scE1]]$ 
-    and $[[Θ(α̂±) ⊢ scE2]]$.  Then $[[Θ(α̂±) ⊢ scE]]$ holds by \cref{lemma:entry-merge-soundness}.
+    Notice that $[[α̂± ∊ Ξ1 \ Ξ2]]$
+    if and only if there is an entry $[[scE]]$ in $[[SC1]]$ 
+    restricting $[[α̂±]]$, but there is no such entry in $[[SC2]]$.
+    Therefore, for any $[[α̂± ∊ Ξ1 \ Ξ2]]$,
+    there is an entry $[[scE]]$ in $[[SC]]$ restricting $[[α̂±]]$.
+    Notice that $[[Θ(α̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC1 : Ξ1]]$.
+
+    Analogously, for any $[[β̂± ∊ Ξ2 \ Ξ1]]$,
+    there is an entry $[[scE]]$ in $[[SC]]$ restricting $[[β̂±]]$.
+    Notice that  $[[Θ(β̂±) ⊢ scE]]$ holds since $[[Θ ⊢ SC2 : Ξ2]]$.
+
+    Finally, for any $[[γ̂± ∊ Ξ1 ∩ Ξ2]]$,
+    there is an entry $[[scE1]]$ in $[[SC1]]$ restricting $[[γ̂±]]$
+    and an entry $[[scE2]]$ in $[[SC2]]$ restricting $[[γ̂±]]$.
+    Since $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
+    $[[Θ(γ̂±) ⊢ scE1 & scE2 = scE]]$ restricting $[[γ̂±]]$ is
+    defined and belongs to $[[SC]]$,
+    moreover, $[[Θ(γ̂±) ⊢ scE]]$ by \cref{lemma:entry-merge-soundness}.
+    This way, $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$.
 
     Let us show the second property.
-    We take an arbitrary $[[uσ]]$ such that $[[Θ ⊢ uσ]]$ and $[[Θ ⊢ uσ : SC]]$.
-    To prove $[[Θ ⊢ uσ : SC1]]$, we need to show that for any $[[scE1]] \in [[SC1]]$, restricting $[[α̂±]]$,
-    $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ holds.
+    We take an arbitrary $[[uσ]]$ such that $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ 
+    and $[[ Θ ⊢ uσ : SC ]]$.
+    To prove $[[ Θ ⊢ uσ : SC1 ]]$, 
+    we need to show that for any $[[scE1]] \in [[SC1]]$, 
+    restricting $[[α̂±]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ holds.
 
     Let us assume that $[[α̂±]] \notin [[dom(SC2)]]$. It means that $[[SC]] \ni [[scE1]]$, 
-    and then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
+    and then since $[[ Θ ⊢ uσ : SC ]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
 
@@
     and $[[SC]] \ni [[scE]]$ where $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$.
-    Then since $[[Θ ⊢ uσ : SC]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
+    Then since $[[ Θ ⊢ uσ : SC ]]$, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
     and by \cref{lemma:entry-merge-soundness}, $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$.
 
-    The proof of $[[Θ ⊢ uσ : SC2]]$ is symmetric.
+    The proof of $[[ Θ ⊢ uσ : SC2 ]]$ is symmetric.
 \end{proof}
@@
     \label{lemma:merge-completeness}
-    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
-    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC1]]$ and $[[Θ ⊢ uσ : SC2]]$, 
-    $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
+    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
+    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
+    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
+    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
+    and $[[ Θ ⊢ uσ : SC ]]$.
 \end{lemma}
@@
 
-    This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that each of these components is defined.
+    This way, to show that $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined, we need to demonstrate that 
+    each of these components is defined and satisfies 
+    the required property 
+    (that the result of $[[uσ]]$ satisfies the corresponding constraint entry).
 
     It is clear that the first two components of this union exist. 
+    Moreover, if $[[scE]]$ is an entry of $[[SCi]]$
+    restricting $[[α̂± ∉ dom(SC2)]]$,
+    then $[[ Θ ⊢ uσ : SCi ]]$ implies $[[ Θ(α̂±) ⊢ [uσ]α̂± : scE]]$,
+
     Let us show that the third component exists.  
     Let us take two entries $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$ restricting the same variable $[[α̂±]]$.
-$[[Θ ⊢ uσ : SC1]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[Θ ⊢ uσ : SC2]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
+$[[ Θ   ⊢ uσ : SC1 ]]$ means that $[[Θ(α̂±) ⊢ [uσ]α̂± : scE1]]$ and $[[ Θ   ⊢ uσ : SC2 ]]$ means $[[Θ(α̂±) ⊢ [uσ]α̂± : scE2]]$.
     Then by \cref{lemma:entry-merge-completeness}, $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ is defined and $[[Θ(α̂±) ⊢ [uσ]α̂± : scE]]$.
-\end{proof}
-
-\begin{lemma}[Substitution existence]
-    \label{lemma:substitution-existence}
-    If $[[Θ ⊢ SC]]$ then there exists $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : SC]]$.
-\end{lemma}
-\begin{proof}
 

### 2023-08-29 proofs/tex/singularity-lemmas.tex
WIP: update typing lemmas

--- 2023-07-10 proofs/tex/singularity-lemmas.tex
@@
     \label{lemma:singularity-soundness}
-    Suppose $[[Θ ⊢ SC]]$, and $[[SC singular with uσ]]$. 
-    Then $[[Θ ⊢ uσ : SC]]$ and for any 
-    $[[uσ']]$ such that $[[Θ ⊢ uσ' : SC]]$,
-    $[[Θ ⊢ uσ' ≈ uσ : dom(SC)]]$.
+    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
+    Then $[[ Θ ⊢ uσ : Ξ ]]$,
+     $[[ Θ ⊢ uσ : SC ]]$ and for any 
+    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
+    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
 \end{lemma}
 \begin{proof}
-    Suppose that $[[Θ ⊢ uσ' : SC]]$.
+    Suppose that $(Θ  ⊢ SC) ⊢ uσ'$.
     It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
@@
 \begin{lemma}[Completeness of Singularity]
-\end{lemma}
-
-\begin{lemma}[Completeness of Singularity]
     \label{lemma:singularity-completeness}
-    Suppose there exists $[[Θ ⊢ uσ1]]$ such that
-    for any $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ ≈ uσ1 : varset]]$.
+    For a given $[[Θ ⊢ SC : Ξ]]$,
+    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
+    on $[[Ξ']] \supseteq [[Ξ]]$.
+    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ']]$ such that
+    for any $[[Θ ⊢ uσ : Ξ']]$, $[[Θ ⊢ uσ : SC]]$ implies 
+    $[[Θ ⊢ uσ ≈ uσ1 : Ξ']]$.
     Then 
     \begin{itemize}
-        \item $[[SC|varset singular with uσ0]]$ for some $[[uσ0]]$, and
-        \item $[[varset ⊆ dom(SC)]]$.
+        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
+        \item $[[Ξ']] = [[Ξ]]$.
     \end{itemize} 
@@
 \begin{proof}
+    First, let us suppose that $[[varset ⊆ dom(SC)]]$ does not hold.
+    It means that there exists $[[ α̂±[Γ] ∊ Θ]]$ and $[[α̂± ∉ dom(SC)]]$
+
+    By \cref{lemma:constraint-sat}, let us take an arbitrary 
+    $[[uσ']]$ satisfying $[[SC]]$ (that is $[[ Θ   ⊢ uσ' : SC ]]$).
+    Let us extend $[[uσ']]$ to $[[uσ'']]$ so that $[[Θ ⊢ uσ'']]$.
+
+    Although $[[uσ']]$ might not be defined on
+
+    such that for any $[[Θ ⊢ uσ]]$, $[[ Θ   ⊢ uσ : SC ]]$ implies 
+    $[[ Γ ⊢ [uσ]α̂± ≈ [uσ1]α̂± ]]$.
+    Let us take
+
+    for any $[[Θ ⊢ uσ]]$, $[[ Θ   ⊢ uσ : SC ]]$ implies $[[Θ|varset ⊢ uσ ≈ uσ1]]$.
+
 \end{proof}
+
+        \item [$-$] Suppose that there exists $[[iN]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iN']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iN' : scE]]$ implies $[[Γ ⊢ iN' ≈ iN]]$. 
+            Then $[[scE singular with nf(iN)]]$.
+        \item [$+$] Suppose that there exists $[[iP]]$ well-formed in $[[Γ]]$ 
+            such that for any $[[iP']]$ well-formed in $[[Γ]]$,
+            $[[Γ ⊢ iP' : scE]]$ implies $[[Γ ⊢ iP' ≈ iP]]$. 
+            Then $[[scE singular with nf(iP)]]$ .

### 2023-08-29 proofs/tex/subst-lemmas.tex
WIP: update typing lemmas

@@
   \end{caseof}
+\end{proof}
+
+\begin{lemma}[Algorithmic Substitution strengthening]
+  \label{lemma:subst-restr-uv}
+  Restricting the substitution to the free algorithmic 
+  variables of the
+  substitution subject does not affect the result.
+  Suppose that $[[uσ]]$ is an algorithmic substitution,
+  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
+ \begin{itemize}
+    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
+    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv iN]uN ]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
 \end{proof}

### 2023-08-29 proofs/tex/unif-constraint-merge-lemmas.tex
WIP: update typing lemmas

--- 2023-06-29 proofs/tex/unif-constraint-merge-lemmas.tex
@@
         \item $[[Θ ⊢ UC]]$ is normalized unification constraint,
-        \item for any substitution $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : lift UC]]$ implies 
-        $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
+        \item for any substitution $[[Θ ⊢ uσ]]$, $[[ Θ   ⊢ uσ : lift UC ]]$ implies 
+        $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$.
     \end{enumerate}
@@
     Let us take an arbitrary substitution $[[Θ ⊢ uσ]]$ and assume that 
-    $[[Θ ⊢ uσ : lift UC]]$.
-    Then $[[Θ ⊢ uσ : lift UCi]]$ holds by definition:
+    $[[ Θ   ⊢ uσ : lift UC ]]$.
+    Then $[[ Θ   ⊢ uσ : lift UCi ]]$ holds by definition:
     If $[[ucE]] \in [[lift UCi]] \subseteq [[lift UC1 ∪ lift UC2]] = [[lift UC]]$.
@@
     Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
-    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[Θ ⊢ uσ : lift UC1]]$ and $[[Θ ⊢ uσ : lift UC2]]$, 
+    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
     \begin{enumerate}
         \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
-        \item $[[Θ ⊢ uσ : UC]]$.
+        \item $[[ Θ   ⊢ uσ : UC ]]$.
     \end{enumerate}

### 2023-08-29 proofs/tex/unification-lemmas.tex
WIP: update typing lemmas

@@
         if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
-        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
-        $[[ [uσ]uP = iQ ]]$.
+        $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
+        such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.
 
@@
         if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
-        $[[Θ ⊢ UC]]$ and for any normalized $[[uσ]]$ such that $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
-        $[[ [uσ]uN = iM ]]$.
+        $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
+        $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
     \end{itemize}
@@
         The resulting unification constraint is empty: $[[UC]] = [[·]]$.
-        It satisfies $[[Θ ⊢ UC]]$ vacuously, and $[[ [us]α⁻ = α⁻ ]]$, that is $[[ [us]uN = iM ]]$.
+        It satisfies $[[Θ ⊢ UC : ·]]$ vacuously, and $[[ [us]α⁻ = α⁻ ]]$, that is $[[ [us]uN = iM ]]$.
 
@@
         The algorithm makes a recursive call to $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ returning $[[UC]]$.
-        By induction hypothesis, $[[Θ ⊢ UC]]$ and for any $[[(Θ  ⊢  lift UC) ⊢ uσ]]$,
+        By induction hypothesis, $[[Θ ⊢ UC : uv uP]]$
+        and thus, $[[Θ ⊢ UC : uv ↑uP]]$,
+        and for any $[[ Θ ⊢ uσ : lift UC ]]$,
         $[[ [uσ]uN ]] = [[ [uσ]↑uP ]] = [[ ↑[uσ]uP ]] = [[ ↑iQ ]] = [[ iM ]]$, as 
@@
         \begin{itemize}
-            \item $[[Θ ⊢ UC1]]$,
+            \item $[[Θ ⊢ UC1 : uv uP]]$,
             \item for any $(Θ  ⊢ lift UC1) ⊢ uσ'$, $[[ [uσ']uP = iQ ]]$.
@@
         \begin{itemize}
-            \item $[[Θ ⊢ UC2]]$,
-            \item for any $[[(Θ ⊢ lift UC2) ⊢ uσ']]$, $[[ [uσ']uN' = iM' ]]$.
+            \item $[[Θ ⊢ UC2 : uv uN']]$,
+            \item for any $[[ Θ  ⊢ uσ' : lift UC2 ]]$, $[[ [uσ']uN' = iM' ]]$.
         \end{itemize}
 
-        Let us take an arbitrary $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+        Let us take an arbitrary $[[ Θ ⊢ uσ : lift UC ]]$.
         By the soundness of the constraint merge (\cref{lemma:merge-soundness}), 
         $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$ implies
-        $[[(Θ  ⊢  lift UC1 ) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$.
+        $[[ Θ   ⊢ uσ : lift UC1  ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$.
 
-        Applying the induction hypothesis to $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$, we have
-        $[[ [uσ]uP = iQ ]]$; applying it to $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$, we have
+        Applying the induction hypothesis to $[[ Θ   ⊢ uσ : lift UC1 ]]$, we have
+        $[[ [uσ]uP = iQ ]]$; applying it to $[[ Θ   ⊢ uσ : lift UC2 ]]$, we have
         $[[ [uσ]uN' = iM' ]]$.
@@
         by inversion, and $[[uN']]$ and $[[iM']]$ are normalized, since $[[uN]]$ and $[[iM]]$ are.
-        Let us take an arbitrary $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+        Let us take an arbitrary $[[ Θ   ⊢ uσ : lift UC ]]$.
         By the induction hypothesis, $[[ [uσ]uN' ]] = [[ iM' ]]$. 
@@
 
-        Let us take an arbitrary $[[uσ]]$ such that  $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
-        Since $[[UC]] = [[ (â⁻ :≈ iM) ]]$, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$ implies 
+        Let us take an arbitrary $[[uσ]]$ such that  $[[ Θ   ⊢ uσ : lift UC ]]$.
+        Since $[[UC]] = [[ (â⁻ :≈ iM) ]]$, $[[ Θ   ⊢ uσ : lift UC ]]$ implies 
         $[[Θ(â⁻) ⊢ [uσ]â⁻ : (â⁻ :≈ iM) ]]$.
@@
         there exists $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$,
-        and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+        and $[[ Θ   ⊢ uσ : lift UC ]]$.
 
@@
         there exists $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$,
-        and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+        and $[[ Θ   ⊢ uσ : lift UC ]]$.
    \end{itemize}
@@
             $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
-            $[[(Θ  ⊢  lift (â⁻ :≈ iM)) ⊢ uσ]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
+            $[[ Θ   ⊢ uσ : lift (â⁻ :≈ iM) ]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
 
@@
             $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ ·]]$, 
-            and $[[(Θ  ⊢  lift ·) ⊢ uσ]]$ holds trivially.
+            and $[[ Θ   ⊢ uσ : lift · ]]$ holds trivially.
 
@@
             This way, by the induction hypothesis there exists $[[UC]]$ such that
-            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[ Θ   ⊢ uσ : lift UC ]]$.
 
@@
 
-            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$,
-            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$.
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[ Θ   ⊢ uσ : lift UC1 ]]$,
+            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$.
             To apply \ruleref{\ottdruleUArrowLabel} and infer the required
             $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$, we need to show that
-            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[ Θ   ⊢ uσ : lift UC ]]$.
             It holds by the completeness of the unification constraint merge 
@@
                 \item $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ holds by the soundness of unification (\cref{lemma:unification-soundness})
-                \item $[[(Θ  ⊢  lift UC1) ⊢ uσ]]$ and $[[(Θ  ⊢  lift UC2) ⊢ uσ]]$ holds as noted above 
+                \item $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$ holds as noted above 
             \end{itemize}.
@@
             This way, by the induction hypothesis, $[[Γ,pas ; Θ ⊨ uN' ≈u iM' ⫤ UC]]$ exists and 
-            moreover, $[[(Θ  ⊢  lift UC) ⊢ uσ]]$.
+            moreover, $[[ Θ   ⊢ uσ : lift UC ]]$.
             Hence, \ruleref{\ottdruleUForallLabel} is applicable to infer

### 2023-08-30 proofs/tex/singularity-lemmas.tex
WIP: add completeness of singularity

--- 2023-07-10 proofs/tex/singularity-lemmas.tex
@@
 \begin{lemma}[Completeness of Singularity]
-\end{lemma}
-
-\begin{lemma}[Completeness of Singularity]
     \label{lemma:singularity-completeness}
-    Suppose there exists $[[Θ ⊢ uσ1]]$ such that
-    for any $[[Θ ⊢ uσ]]$, $[[Θ ⊢ uσ : SC]]$ implies $[[Θ ⊢ uσ ≈ uσ1 : varset]]$.
+    For a given $[[Θ ⊢ SC]]$,
+    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
+    on $[[Ξ]] \supseteq [[dom(SC)]]$.
+    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
+    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
+    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
     Then 
     \begin{itemize}
-        \item $[[SC|varset singular with uσ0]]$ for some $[[uσ0]]$, and
-        \item $[[varset ⊆ dom(SC)]]$.
+        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
+        \item $[[Ξ]] = [[dom(SC)]]$.
     \end{itemize} 
@@
 \begin{proof}
+
+    First, let us assume $[[Ξ]] \neq [[dom(SC)]]$. 
+    Then there exists $[[α̂± ∊ Ξ \ dom(SC)]]$.
+    Let us take $[[Θ ⊢ uσ1 : Ξ]]$ such that 
+    any other substitution $[[Θ ⊢ uσ : Ξ]]$ satisfying $[[SC]]$
+    is equivalent to $[[uσ1]]$ on $[[Ξ]]$.
+
+    Notice that $[[Θ ⊢ uσ1 : SC]]$: 
+    by \cref{lemma:constraint-sat}, there exists $[[uσ']]$
+    such that $[[Θ ⊢ uσ' : Ξ]]$ and $[[Θ ⊢ uσ' : SC]]$,
+    and by assumption, $[[Θ ⊢ uσ' ≈ uσ1 : Ξ]]$,
+    implying  $[[Θ ⊢ uσ' ≈ uσ1 : dom(SC)]]$.
+
+    Let us construct $[[uσ2]]$ such that $[[Θ ⊢ uσ2 : Ξ]]$ as
+    follows:
+    $$
+    \begin{cases}
+        [[ [uσ2]β̂± = [uσ1]β̂±  ]] & \text{if } [[β̂± ≠ α̂±]] \\
+        [[ [uσ2]α̂± ]] = T & \text{where $T$ is any closed type not equivalent to $[[ [uσ1]α̂± ]]$} \\
+    \end{cases}
+    $$
+    It is easy to see that $[[Θ ⊢ uσ2 : SC]]$ since
+    $[[uσ1|dom(SC) = uσ2|dom(SC)]]$, and $[[Θ ⊢ uσ1 : SC]]$.
+    However, $[[Θ ⊢ uσ2 ≈ uσ1 : Ξ]]$ does not hold because
+    by construction, $[[Θ(α̂±) ⊢ [uσ2]α̂± ≈ [uσ1]α̂±]]$ does not hold. 
+    This way, we have a contradiction.
+
+    Second, let us show $[[SC singular with uσ0]]$.
+    Let us take arbitrary $[[scE ∊ SC]]$ restricting $[[β̂±]]$.
+    We need to show that $[[scE]]$ is singular.
+    Notice that $[[Θ ⊢ uσ1 : SC]]$ implies $[[Θ(β̂±) ⊢ [uσ1]β̂±]]$ and $[[Θ(β̂±) ⊢ [uσ1]β̂± : scE ]]$.
+    We will show that any other type satisfying $[[scE]]$ is equivalent to $[[ [uσ1]β̂± ]]$,
+    then by \cref{lemma:entry-singularity-completeness}, $[[scE singular with [uσ1]β̂± ]]$.
+    \begin{itemize}
+        \item if $[[β̂±]]$ is positive, 
+            let us take any type $[[Θ(β̂±) ⊢ iP']]$
+            and assume $[[Θ(β̂±) ⊢ iP' : scE]]$.
+            We will show that $[[Θ(β̂±) ⊢ iP' ≈ [uσ1]β̂±]]$,
+            which by \cref{lemma:subt-equiv-algorithmization} will
+            imply $[[scE singular with nf([uσ1]β̂±)]]$.
+
+            Let us construct $[[uσ2]]$ such that $[[Θ ⊢ uσ2 : Ξ]]$ as
+            follows:
+            $$
+            \begin{cases}
+                [[ [uσ2]γ̂± = [uσ1]γ̂±  ]] & \text{if } [[γ̂± ≠ β̂±]] \\
+                [[ [uσ2]β̂± = iP' ]]
+            \end{cases}
+            $$
+            It is easy to see that $[[Θ ⊢ uσ2 : SC]]$:
+            for $[[scE]]$, $[[Θ(β̂±) ⊢ [uσ2]β̂± : scE ]]$ by construction,
+            since $[[Θ(β̂±) ⊢ iP' : scE]]$; for any other $[[scE' ∊ SC]]$
+            restricting $[[γ̂±]]$, $[[ [uσ2]γ̂± = [uσ1]γ̂± ]]$, 
+            and $[[Θ(γ̂±) ⊢ [uσ1]γ̂± : scE' ]]$ since $[[Θ ⊢ uσ1 : SC]]$.
+
+            Then by assumption, $[[Θ ⊢ uσ2 ≈ uσ1 : Ξ]]$,
+            which in particular means $[[Θ(β̂±) ⊢ [uσ2]β̂± ≈ [uσ1]β̂±]]$,
+            that is $[[Θ(β̂±) ⊢ iP' ≈ [uσ1]β̂±]]$.
+        \item if $[[β̂±]]$ is negative, the proof is analogous.
+    \end{itemize}
 \end{proof}

### 2023-08-30 proofs/tex/constraint-merge-lemmas.tex
WIP: fix unification lemmas

@@
     $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
-    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined,
-    and $[[ Θ ⊢ uσ : SC ]]$.
+    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
 \end{lemma}
@@
 \end{proof}
+
+\begin{corollary} [Completeness of Unification Constraint Merge] 
+    \label{corollary:merge-completeness-unif}
+    Suppose that $[[Θ ⊢ UC1 : Ξ1]]$ and $[[Θ ⊢ UC2 : Ξ2]]$.
+    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
+    $[[ Θ ⊢ uσ : UC1 ]]$ and $[[ Θ ⊢ uσ : UC2 ]]$
+    then $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined.
+\end{corollary}

### 2023-08-30 proofs/tex/unification-lemmas.tex
WIP: fix unification lemmas

@@
         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
-        for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uP = iQ ]]$,
-        there exists $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$,
-        and $[[ Θ   ⊢ uσ : lift UC ]]$.
+        suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
+        then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.
 
         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,\\
-        for any $[[Θ ⊢ uσ]]$ such that $[[ [uσ]uN = iM ]]$,
-        there exists $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$,
-        and $[[ Θ   ⊢ uσ : lift UC ]]$.
+        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
+        suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
+        then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
    \end{itemize}
@@
 
-            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
-            $[[Θ ⊢ uσ]]$ means that $[[Δ ⊢ iM]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : α̂⁻ ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
+            $[[Θ ⊢ uσ : α̂⁻]]$ means that $[[Δ ⊢ iM]]$.
             This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 
             $[[Γ ; Θ ⊨ â⁻ ≈u iM ⫤ (â⁻ :≈ iM)]]$.
-            $[[ Θ   ⊢ uσ : lift (â⁻ :≈ iM) ]]$ holds by \ruleref{\ottdruleSATSCENEqLabel}.
 
         \item $[[uN]] = [[α⁻]]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]α⁻ = iM ]]$.
-            The latter means $[[iM = α⁻]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv(α⁻)]]$ such that $[[ [uσ]α⁻ = iM ]]$.
+            Since $[[uv(α⁻) = ∅]]$, $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
 
-            Then $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
             This way, \ruleref{\ottdruleUNVarLabel} infers 
-            $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ ·]]$, 
-            and $[[ Θ   ⊢ uσ : lift · ]]$ holds trivially.
+            $[[Γ; Θ ⊨ a⁻ ≈u a⁻ ⫤ ·]]$, which is rewritten as $[[Γ; Θ ⊨ uN ≈u iM ⫤ ·]]$.
 
         \item $[[uN]] = [[↑uP]]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]↑uP = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]↑uP = iM ]]$.
             The latter means $[[ ↑[uσ]uP = iM ]]$, i.e.
@@
             This way, by the induction hypothesis there exists $[[UC]]$ such that
-            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$, and moreover, $[[ Θ   ⊢ uσ : lift UC ]]$.
+            $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$.
 
         \item $[[uN]] = [[uP → uN']]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ](uP → uN') = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uP → uN')]]$ such that $[[ [uσ](uP → uN') = iM ]]$.
             The latter means $[[ [uσ]uP → [uσ]uN' = iM ]]$, i.e.
@@
             Let us show that the induction hypothesis is applicable to 
-            $[[ [uσ]uP = iQ ]]$ and to $[[ [uσ]uN' = iM' ]]$:
+            $[[Θ ⊢ uσ|uv(uP) : uv(uP)]]$ and $[[ [uσ|uv(uP)]uP = iQ ]]$ 
+            (the latter holds since $[[ [uσ|uv(uP)]uP = [uσ]uP ]]$ by \cref{lemma:subst-restr-uv}),
             \begin{itemize}
-                \item $[[uP]]$ and $[[uN']]$ are normalized, since $[[uN]] = [[uP → uN']]$ is normalized
-                \item $[[Γ ; dom(Θ) ⊢  uP]]$ and $[[Γ ; dom(Θ) ⊢  uN']]$ follow from the inversion of $[[Γ ; dom(Θ) ⊢  uP → uN']]$,
-                \item $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM']]$ follow from inversion of $[[Γ ⊢ iQ → iM']]$.
+                \item $[[uP]]$ is normalized, since $[[uN]] = [[uP → uN']]$ is normalized
+                \item $[[Γ ; dom(Θ) ⊢ uP]]$ follows from the inversion of $[[Γ ; dom(Θ) ⊢  uP → uN']]$,
+                \item $[[Γ ⊢ iQ]]$.
             \end{itemize}
+            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$.
+            Analogously, the induction hypothesis is applicable to $[[ [uσ|uv uN']uN' = iM' ]]$,
+            and thus, $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$.
 
-            Then by the induction hypothesis, $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC1]]$ and $[[ Θ   ⊢ uσ : lift UC1 ]]$,
-            $[[Γ ; Θ ⊨ uN' ≈u iM' ⫤ UC2]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$.
             To apply \ruleref{\ottdruleUArrowLabel} and infer the required
-            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$, we need to show that
-            $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and $[[ Θ   ⊢ uσ : lift UC ]]$.
-            It holds by the completeness of the unification constraint merge 
-            (\cref{lemma:merge-completeness}):
-            \begin{itemize}
-                \item $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ holds by the soundness of unification (\cref{lemma:unification-soundness})
-                \item $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$ holds as noted above 
-            \end{itemize}.
+            $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$,
+            it is left to show that $[[Θ ⊢ UC1 & UC2 = UC]]$.
+            It holds by completeness of the unification constraint merge 
+            (\cref{lemma:unif-merge-completeness})
+            for $[[Θ ⊢ lift UC1 : uv uP]]$, 
+                $[[Θ ⊢ lift UC2 : uv uN']]$ (which hold by soundness),
+                and $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$, 
+                which holds since $[[uv(uP) ∪ uv(uN') = uv(uP → uN')]]$.
+            Notice that by soundness, $[[Θ ⊢ uσ|uv(uP) : lift UC1]]$, 
+            which implies $[[Θ ⊢ uσ : lift UC1]]$.
+            Analogously, $[[Θ ⊢ uσ : lift UC2]]$.
 
         \item $[[uN]] = [[∀pas.uN']]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ]]$ such that $[[ [uσ]∀pas.uN' = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uN')]]$ such that $[[ [uσ]∀pas.uN' = iM ]]$.
             The latter means $[[ ∀pas.[uσ]uN' = iM ]]$, i.e.
@@
             $[[Γ,pas ⊢ iM']]$ follows from inversion of $[[Γ ⊢ ∀pas.iM']]$, and
-            $[[Θ ⊢ uσ]]$ by assumption.
+            $[[Θ ⊢ uσ : uv(uN')]]$ by assumption.
 

### 2023-09-01 proofs/tex/subst-lemmas.tex
Finish syntax update

@@
 
-\begin{lemma}[Algorithmic Substitution strengthening]
+\begin{lemma}[Algorithmic Substitution Strengthening]
   \label{lemma:subst-restr-uv}
@@
     \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
-    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv iN]uN ]]$
+    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv uN]uN ]]$
   \end{itemize}
@@
   Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
-  normalized substitutions of signature $[[Θ ⊢ uσi]]$.
+  normalized substitutions of signature $[[Θ ⊢ uσi : Ξ]]$.
   Then 
   \begin{enumerate}
-    \item [$+$] for a normalized type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
+    \item [$+$] for a normalized type $[[Γ; Ξ ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
       $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
-    \item [$-$] for a normalized type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
+    \item [$-$] for a normalized type $[[Γ; Ξ ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
       $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
@@
   The proof is a simple structural induction on 
-  $[[Γ; dom(Θ) ⊢ uP]]$ and mutually, on $[[Γ; dom(Θ) ⊢ uN]]$.
+  $[[Γ; Ξ ⊢ uP]]$ and mutually, on $[[Γ; Ξ ⊢ uN]]$.
   Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
   \begin{caseof}
-    \item $[[Γ; dom(Θ) ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
+    \item $[[Γ; Ξ ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
       implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
-    \item $[[Γ; dom(Θ) ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
+    \item $[[Γ; Ξ ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
       $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
-    \item $[[Γ; dom(Θ) ⊢ ∀pas.uN]]$. Then we are proving that
+    \item $[[Γ; Ξ ⊢ ∀pas.uN]]$. Then we are proving that
       $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
@@
       $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
-      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; dom(Θ) ⊢ uN]]$, 
+      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; Ξ ⊢ uN]]$, 
       hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
       as required.
-    \item $[[Γ; dom(Θ) ⊢ uP → uN]]$. Then we are proving that
+    \item $[[Γ; Ξ ⊢ uP → uN]]$. Then we are proving that
       $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
@@
       $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
-    \item $[[Γ; dom(Θ) ⊢ ↑uP]]$. The proof is similar to the previous case:
+    \item $[[Γ; Ξ ⊢ ↑uP]]$. The proof is similar to the previous case:
       we apply congruence of substitution, equality, and normalization,
@@
   Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
-  of signature $[[Θ ⊢ uσi]]$.
+  of signature $[[Θ ⊢ uσi : Ξ]]$.
   Then 
   \begin{enumerate}
-    \item [$+$] for a type $[[Γ; dom(Θ) ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
-      $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$;
-    \item [$-$] for a type $[[Γ; dom(Θ) ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
-      $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
+    \item [$+$] for a type $[[Γ; Ξ ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
+      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
+    \item [$-$] for a type $[[Γ; Ξ ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
+      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
   \end{enumerate}
@@
   the given equivalences and well-formedness properties are preserved. 
-  $[[Γ; dom(Θ) ⊢ uP]]$ implies $[[Γ; dom(Θ) ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
+  $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
   $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
@@
   Notice that by \cref{corollary:norm-subst-sig-algo}
-  $[[Θ ⊢ uσi]]$ implies $[[Θ ⊢ nf(uσi)]]$.
+  $[[Θ ⊢ uσi : Ξ]]$ implies $[[Θ ⊢ nf(uσi) : Ξ]]$.
 
@@
   and by ${{corollary:subst-subt-equiv-algorithmization}}$,
-  $[[Θ|uv uP ⊢ uσ1 ≈ uσ2]]$.
+  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$.
 
@@
   $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
-  $[[Θ|uv uN ⊢ uσ1 ≈ uσ2]]$.
+  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
 \end{proof}

### 2023-09-01 proofs/main.tex
Finish syntax update

@@
 
-\subsection{Algorithmic Renaming}
+\subsection{Variable Algorithmization}
 
@@
 via replacing certain type variables with fresh algorithmic variables.
-This is done via algorithmic renaming
 
-\begin{definition}
-  Suppose that $[[nas]]$ is a list of negative type variables,
+\begin{definition}[Variable Algorithmization]
+  Suppose that $[[nas]]$ is a list of negative type variables
+  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
+  Then $[[ nuas/nas ]]$ is a substitution turning each $[[αi⁻]]$ from $[[nas]]$ into 
+  $[[αî⁻]]$ from $[[nuas]]$.
+  Positive algorithmization $[[ puas/pas ]]$
+  is defined symmetrically.
+\end{definition}
 
@@
   $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
+
+  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
 \end{definition}
@@
 \begin{definition}[Signature of Algorithmic Substitution]
-  $[[Θ ⊢ uσ : Θ]]$ means that
+  \hfill
+  \begin{itemize}
+    \item $[[Θ ⊢ uσ : Ξ]]$ means that
       \begin{enumerate}
@@
       \end{enumerate}
+    \item $[[Γ ⊢ uσ : Ξ]]$ means that
+      \begin{enumerate}
+        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
+        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
+      \end{enumerate}
+  \end{itemize}
 \end{definition}
@@
   \end{itemize}
+\end{definition}
 
-\end{definition}
+However, in subtyping we need to consider more general
+kind of constraints. Specifically,
+subtyping constraint entries can restrict a variable
+not only to be equivalent to a certain type, but
+also to be a supertype of a positive type.
 
@@
 
+\begin{definition}[Well-formed Constraint]
+  We say that a constraint is well-formed in a
+  subtyping context $[[Θ]]$ if all its entries are well-formed in
+  the corresponding elements of $[[Θ]]$.
+  More formally, 
+  $[[Θ ⊢ UC]]$ iff for every $[[ucE]] \in $ $[[UC]]$,
+  such that $[[ucE]]$ restricts $[[α̂±]]$,
+  there exists $[[ α̂±[Γ] ]] \in [[Θ]]$
+  and $[[Γ ⊢ ucE]]$.
+\end{definition}
+
+Next, define the least upper bound 
+for two subtyping constraints.
+First, we define the least upper bound for entries,
+and then extend it to the set of entries.
+
 \begin{definition} [Matching Entries]
@@
 \begin{definition}[Merge of Matching Constraint Entries]
+  \label{definition:merge-matching-entries}
    \hfill \\
@@
 \end{definition}
+
+Unification constraint entries are a special case of subtyping constraint
+entries. They are merged using the same algorithm 
+(\cref{definition:merge-matching-entries}).
+Notice that the merge of two matching unification constraint entries
+is a unification constraint entry.
+\begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
+  Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
+  are unification constraint entries. 
+  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
+  $[[Γ ⊢ ucE1 & ucE2 = ucE]]$
+  according to \cref{definition:merge-matching-entries},
+  is a unification constraint entry.
+\end{lemma}
+\begin{proof}
+  Since $[[ucE1]]$ and $[[ucE2]]$ are matching unification constraint entries,
+  they have the shape $([[pua :≈ iP1]], [[pua :≈ iP2]])$ or
+  $([[nua :≈ iN1]], [[nua :≈ iN2]])$.
+  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
+  can only be defined by \ruleref{\ottdruleSCMEPEqEqLabel} or
+  \ruleref{\ottdruleSCMENEqEqLabel}.
+  In both cases the result, if it exists, 
+  is a unification constraint entry:
+  in the first case, the result has shape $[[pua :≈ iP1]]$,
+  in the second case, the result has shape $[[nua :≈ iN1]]$.
+\end{proof}
+
 % Notice that in case of equivalence, the assigned types
@@
 
-\begin{definition}[Constraint Context]
-  Constraint context (denoted as $[[Θ]]$) is a set of entries of shape $[[ pua[Γ] ]]$ 
-  and $[[ nua[Γ] ]]$, specifying that the unification variable ($[[pua]]$ or $[[nua]]$)
-  must be instantiated in a type well-formed in context $[[Γ]]$.
-  We assume that for each $[[α̂±]]$, at most one entry of shape $[[ α̂±[Γ] ]]$ is in $[[Θ]]$.
-  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
-\end{definition}
-
-\begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  constraint context $[[Θ]]$ if all its entries are well-formed in
-  the corresponding elements of $[[Θ]]$.
-  More formally, 
-  $[[Θ ⊢ UC]]$ iff for every $[[ucE]] \in $ $[[UC]]$,
-  such that $[[ucE]]$ restricts $[[α̂±]]$,
-  there exists $[[ α̂±[Γ] ]] \in [[Θ]]$
-  and $[[Γ ⊢ ucE]]$.
-\end{definition}
-
 \begin{definition}[Merge of Subtyping Constraints]
+  \label{definition:merge-subtyping-constraints}
   Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
@@
 \end{definition}
+
+Unification constraints can be considered 
+as a special case of subtyping constraints,
+and the merge of unification constraints
+is defined as the merge of subtyping constraints.
+Then it is easy to see that the merge of two 
+unification constraints is a unification constraint.
+
+\begin{lemma}[Merge of Unification Constraints is well-defined]
+  Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
+  are unification constraints. 
+  Then the merge of $[[UC1]]$ and $[[UC2]]$ 
+  $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$
+  according to \cref{definition:merge-subtyping-constraints},
+  is a unification constraint.
+\end{lemma}
+\begin{proof}
+  $[[UC]]$ consists of unmatched entries of $[[UC1]]$ and $[[UC2]]$,
+  which are \emph{unification} constraint entries by assumption,
+  and merge of matching entries, which also are  
+  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-is-well-defined}.
+\end{proof}
 

### 2023-09-01 proofs/tex/equiv-lemmas.tex
WIP: update typing lemmas
Finish syntax update

@@
   \hfill \\
-  $[[Θ ⊢ uσ]] \iff [[Θ ⊢ nf(uσ)]]$ and
+  $[[Θ ⊢ uσ : Ξ]] \iff [[Θ ⊢ nf(uσ) : Ξ]]$,
   $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
@@
 \begin{proof}
-  We prove it by induction on $[[iP]]$ (and mutually on $[[N]]$).
+  We prove it by induction on $[[iP]]$ (and mutually on $[[iN]]$).
   \begin{caseof}
@@
   \label{corollary:subst-subt-equiv-algorithmization}
-  Suppose that $[[Θ ⊢ uσ1]]$ and $[[Θ ⊢ uσ2]]$ are algorithmic substitutions.
-  Then
-  $[[Θ ⊢ uσ1 ≈ uσ2]] \iff [[nf(uσ1)|dom(Θ) = nf(uσ2)|dom(Θ)]]$
-  and 
-  $[[Γ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
+  Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
+  and $[[Ξ ⊆ Ξ']]$. 
+  Then $[[Γ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
 \end{corollary}

### 2023-09-01 proofs/tex/unif-constraint-merge-lemmas.tex
Finish syntax update

@@
         \item $[[Θ ⊢ UC]]$ is normalized unification constraint,
-        \item for any substitution $[[Θ ⊢ uσ]]$, $[[ Θ   ⊢ uσ : lift UC ]]$ implies 
-        $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$.
+        \item for any substitution $[[Θ ⊢ uσ : dom(UC)]]$, $[[ Θ  ⊢ uσ : lift UC ]]$ implies 
+        $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
     \end{enumerate}
@@
 
-    Let us take an arbitrary substitution $[[Θ ⊢ uσ]]$ and assume that 
+    Let us take an arbitrary substitution $[[Θ ⊢ uσ : dom(UC)]]$ and assume that 
     $[[ Θ   ⊢ uσ : lift UC ]]$.
     Then $[[ Θ   ⊢ uσ : lift UCi ]]$ holds by definition:
-    If $[[ucE]] \in [[lift UCi]] \subseteq [[lift UC1 ∪ lift UC2]] = [[lift UC]]$.
-    So $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$ holds.
+    If $[[ucE]] \in [[lift UCi]] \subseteq [[lift UC1 ∪ lift UC2]] = [[lift UC]]$ then
+    $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$ (where $[[ucE]]$ restricts $[[α̂±]]$) holds since $[[Θ ⊢ uσ : dom(UC)]]$.
 \end{proof}
-
-% \begin{lemma} [Completeness of Unification Constraint Merge]
-%     \label{lemma:unif-merge-completeness}
-%     Suppose that $[[Θ ⊢ UC|varset1]]$ and $[[Θ ⊢ UC|varset2]]$.
-%     Then $[[Θ ⊢ UC|varset1 & UC|varset2 = UC']]$ exists and
-%     $[[UC' = UC|varset1 ∪ varset2]]$.
-% \end{lemma}
-% \begin{proof}
-%     $[[Θ ⊢ UC|varset1 & UC|varset2 = UC']]$ is defined as the union of three parts:
-%     entries of $[[UC|varset1]]$ that do not have matching entries of $[[UC|varset2]]$,
-%     entries of $[[UC|varset2]]$ that do not have matching entries of $[[UC|varset1]]$,
-%     and the merge of matching entries.
-%     The first two parts are defined. The merge of matching entries is defined
-%     by \ruleref{\ottdruleUCMEPEqEqLabel}, since the matching entries must be equal
-%     if they both belong to $[[UC]]$.
-
-%     It remains to show that $[[UC' = UC|varset1 ∪ varset2]]$.
-%     It is easy to see that the three parts comprising $[[UC']]$ 
-%     correspond to the three parts comprising 
-%     $[[UC|varset1 ∪ varset2]] = [[UC | (varset1 \ varset2) ∪ UC | (varset2 \ varset1) ∪ UC | varset1 ∩ varset2]]$. 
-% \end{proof}
 
@@
     Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
-    Then for any substitution $[[Θ ⊢ uσ]]$ such that $[[ Θ   ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
+    Then for any $[[Ξ]] \supseteq [[dom(UC1) ∪ dom(UC2)]]$ 
+    and substitution $[[Θ ⊢ uσ : Ξ]]$ such that $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
     \begin{enumerate}

### 2023-09-04 proofs/tex/wf-algo-lemmas.tex
WIP: separate the algorithmic proofs

\begin{lemma}[AASoundness of algorithmic type well-formedness]
    \label{lemma:wf-algo-soundness}
    \hfill
  \begin{itemize}
    \item[$+$] if $[[Γ ; Ξ ⊢ uP]]$ then $[[fv(uP) ⊆ {Γ}]]$ and $[[uv(uP) ⊆ Ξ]]$;
    \item[$-$] if $[[Γ ; Ξ ⊢ uN]]$ then $[[fv(uN) ⊆ {Γ}]]$ and $[[uv(uN) ⊆ Ξ]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof is analogous to \cref{lemma:wf-soundness}.
  The additional base case is when $[[Γ ; Ξ ⊢ uP]]$ is derived by \ruleref{\ottdruleWFATPUvarLabel},
  and the symmetric negative case.
  In this case, $[[uP]] = [[α̂⁺]]$, and $[[uv(uP)]] = [[{α̂⁺} ⊆ Ξ]]$ by inversion; $[[fv(uP)]] = [[∅ ⊆ {Γ}]]$ vacuously.
\end{proof}

\begin{lemma}[Completeness of algorithmic type well-formedness]
  \label{lemma:wf-algo-ctxt-equiv}
  In the well-formedness judgment, only used variables matter:
  \begin{itemize}
  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$
    and $[[Ξ1 ∩ uv uP]] = [[Ξ2 ∩ uv uP]]$ then
    $[[Γ1 ; Ξ1 ⊢ uP]] \iff [[Γ2 ; Ξ2 ⊢ uP]]$, and
  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$
    and $[[Ξ1 ∩ uv uN]] = [[Ξ2 ∩ uv uN]]$ then
    $[[Γ1 ; Ξ1 ⊢ uN]] \iff [[Γ2 ; Ξ2 ⊢ uN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By mutual structural induction on $[[uP]]$ and $[[uN]]$.
\end{proof}

\begin{lemma}[Variable algorithmization agrees with well-formedness]
  \hfill
  \begin{itemize}
    \item[$+$]  $[[Γ, nas ⊢ iP]]$ implies $[[Γ; nuas ⊢ [nuas/nas]iP]]$;
    \item[$-$]  $[[Γ, nas ⊢ iN]]$ implies $[[Γ; nuas ⊢ [nuas/nas]iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof is a simple structural induction on $[[Γ, nas ⊢ iP]]$ and mutually, on $[[Γ, nas ⊢ iN]]$.
\end{proof}

\begin{corollary}[Well-formedness Algorithmic Context Weakening]
  \label{lemma:wf-weakening-algo}
  Suppose that $[[{Γ1} ⊆ {Γ2}]]$,
  and $[[Ξ1 ⊆ Ξ2]]$. Then
  \begin{itemize}
    \item[$+$] if $[[Γ1 ; Ξ1 ⊢ uP]]$ implies $[[Γ2 ; Ξ2 ⊢ uP]]$,
    \item [$-$] if $[[Γ1 ; Ξ1 ⊢ uN]]$ implies $[[Γ2 ; Ξ2 ⊢ uN]]$.
  \end{itemize}
\end{corollary}
\begin{proof}
  By \cref{lemma:wf-soundness},
  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ {Γ1}]]$,
  which means that $[[fv(iP) ⊆ {Γ2}]]$,
  and thus, $[[fv(iP)]] = [[fv(iP) ∩ {Γ1}]] = [[fv(iP) ∩ {Γ2}]]$.
  Then by \cref{lemma:wf-ctxt-equiv}, $[[Γ2 ⊢ iP]]$. 
  The negative case is symmetric.
\end{proof}

### 2023-09-04 proofs/tex/alg-pos-subtyping-lemmas.tex
Grammar update: Γ = set

@@
         Since $[[â⁺[Δ] ∊ Θ]]$ and $[[Δ ⊢ iQ']]$, 
-        it is clear that $[[ Θ ⊢ (â⁺ :≥ iQ') : â⁺]]$.
+        it is clear that $[[ Θ ⊢ (â⁺ :≥ iQ') : {â⁺}]]$.
 
@@
             By inversion, $[[Γ ; dom(Θ) ⊢  ∃nbs.α̂⁺]]$ implies $[[â⁺[Δ] ∊ Θ]]$ for some 
-            $[[{Δ} ⊆ {Γ}]]$.
+            $[[{Δ} ⊆ Γ]]$.
 
@@
             $[[ Γ,ncs  ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ0 ]]$ for some $[[iN]]$, 
-            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ {Γ} = ∅ ]]$,
+            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ Γ = ∅ ]]$,
             $[[ [iNs/nbs][uσ]α̂⁺ = [uσ]α̂⁺ ]]$, that is $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$.
@@
             \begin{enumerate}
-                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ : α̂⁺]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
+                \item $[[Δ ⊢ [uσ]α̂⁺ ]]$ because $[[Θ ⊢ uσ : {α̂⁺}]]$ and $[[α̂⁺[Δ] ∊ Θ]]$,
                 \item $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$ as noted above
@@
 
-            By inversion, $[[ Γ ⊢ [σ][uσ]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ : nas]]$.
+            By inversion, $[[ Γ ⊢ [σ][uσ]uP' ≥ iQ' ]]$ for some $[[Γ, nbs ⊢ σ :{nas}]]$.
             Since $[[σ]]$ and $[[uσ]]$ have disjoint domains,
@@
                     \item for $[[αî⁻]] \in [[{â⁻*} ∩ uv(uP0)]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
-                    i.e. $[[ Γ, nbs ⊢ [σ]αi⁻ ]]$ holds since $[[Γ, nbs ⊢ σ : nas]]$,
+                    i.e. $[[ Γ, nbs ⊢ [σ]αi⁻ ]]$ holds since $[[Γ, nbs ⊢ σ :{nas}]]$,
                     \item for $[[β̂±]] \in [[ uv(uP') ]] \subseteq [[dom(Θ)]]$, $[[ (Θ, â⁻*[Γ, nbs])(β̂±) ⊢ [uσ0] β̂± ]]$,

### 2023-09-04 proofs/tex/anti-unification-lemmas.tex
Grammar update: Γ = set

@@
         returning as the result $[[(Ξ, uM, aus1, aus2)]] = $
-        $[[(â⁻_{iN1, iN2}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} :≈ iN1) ,  (â⁻_{iN1, iN2} :≈ iN2))]]$.
+        $[[({â⁻_{iN1, iN2}}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} :≈ iN1) ,  (â⁻_{iN1, iN2} :≈ iN2))]]$.
 
         \begin{enumerate}
-            \item $[[Γ ; Ξ ⊢ uM]]$ is rewritten as $[[Γ ; â⁻_{iN1, iN2} ⊢ â⁻_{iN1, iN2}]]$,
+            \item $[[Γ ; Ξ ⊢ uM]]$ is rewritten as $[[Γ ; { â⁻_{iN1, iN2} } ⊢ â⁻_{iN1, iN2}]]$,
                 which holds trivially;
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} :≈ iNi) : â⁻_{iN1, iN2}]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} :≈ iNi) : {â⁻_{iN1, iN2}}]]$,
                 which holds since $[[Γ ⊢ iNi]]$ by the premise of the rule;

### 2023-09-04 proofs/tex/decl-subtyping-lemmas.tex
Grammar update: Γ = set

@@
           \ruleref{\ottdruleDOneForallLabel}, and by inversion:
-          $[[Γ,pbs,pbs' ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ,pbs,pbs' ⊢ iPs]]$.
+          $[[Γ, {pbs}, {pbs'} ⊢ [iPs/pas]iN' ≤ iM']]$ for some $[[Γ, {pbs}, {pbs'} ⊢ iPs]]$.
           Then again by \ruleref{\ottdruleDOneForallLabel} with the same $[[iPs]]$,
@@
           Otherwise, $[[Γ, pbs ⊢ ∀pas.iN' ≤ ∀pbs'.iM']]$ is inferred by
-          \ruleref{\ottdruleDOneForallLabel}, that is $[[Γ, pbs, pbs' ⊢ [iPs/pas]iN' ≤ iM']]$
-          for some $[[Γ, pbs, pbs' ⊢ iPs]]$.
+          \ruleref{\ottdruleDOneForallLabel}, that is $[[Γ, {pbs}, {pbs'} ⊢ [iPs/pas]iN' ≤ iM']]$
+          for some $[[Γ, {pbs}, {pbs'} ⊢ iPs]]$.
           Then by \ruleref{\ottdruleDOneForallLabel} again,
@@
           such that $[[Γ ⊢ [iPs/pas][iPs'/pas']iN' ≤ iM']]$ 
-          (the decomposition of substitutions is possible since $[[{pas} ∩ {Γ} = ∅]]$).
+          (the decomposition of substitutions is possible since $[[{pas} ∩ Γ = ∅]]$).
           Then by \ruleref{\ottdruleDOneForallLabel} again,
@@
   By inversion  of $[[Γ1 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$, 
-  we obtain $[[σ]]$ such that $[[ Γ1, nbs ⊢ σ : nas]]$
+  we obtain $[[σ]]$ such that $[[ Γ1, nbs ⊢ σ :{nas}]]$
   and $[[Γ1, nbs ⊢ [σ]iP ≥ iQ]]$.
@@
     \item $[[Γ1, nbs ⊢ iQ]]$; and 
-    \item $[[Γ2, nbs ⊢ iQ]]$, which means $[[ fv(iQ) ⊆ {Γ2, nbs} ]]$
+    \item $[[Γ2, nbs ⊢ iQ]]$, which means $[[ fv(iQ) ⊆ Γ2, nbs ]]$
       by \cref{lemma:wf-soundness},
       and  combining it with $[[fv([σ]iP) ⊆ fv(iQ)]]$, 
-      we have $[[fv([σ]iP) ⊆ {Γ2, nbs}]]$.
+      we have $[[fv([σ]iP) ⊆ Γ2, nbs]]$.
   \end{itemize}
@@
 
-    \item $[[ Ord fv([σ]iP) ⊢ σ0 : nas]]$.
+    \item $[[  fv([σ]iP) ⊢ σ0 :{nas}]]$.
       To show that, let us consider $[[αi⁻]]$ 
@@
           $[[· ⊢ [σ0]αi⁻]]$, which can be weakened to 
-          $[[Ord fv([σ]iP) ⊢ [σ0]αi⁻]]$;
+          $[[ fv([σ]iP) ⊢ [σ0]αi⁻]]$;
         \item if $[[αi⁻ ∊ {nas} ∩ fv(iP)]]$, 
@@
           By \cref{corollary:wf-ctxt-strengthening}, 
-          it means $[[Ord fv([σ0]αi⁻) ⊢ [σ0]αi⁻]]$,
+          it means $[[ fv([σ0]αi⁻) ⊢ [σ0]αi⁻]]$,
           which we weaken (\cref{lemma:wf-weakening}) 
-          to $[[Ord fv([σ]iP) ⊢ [σ0]αi⁻]]$ 
+          to $[[ fv([σ]iP) ⊢ [σ0]αi⁻]]$ 
           (since $[[fv([σ0]αi⁻) ⊆ fv([σ0]iP)]]$ by \cref{lemma:subst-fv-image},
@@
   $[[Γ1, nbs ⊢ [σ]iP]]$ implies 
-  $[[ Ord fv([σ]iP) ⊢ [σ]iP ]]$, which,
-  since $[[fv([σ]iP) ⊆ {Γ2, nbs}]]$,
+  $[[  fv([σ]iP) ⊢ [σ]iP ]]$, which,
+  since $[[fv([σ]iP) ⊆ Γ2, nbs]]$,
   is weakened to $[[Γ2, nbs ⊢ [σ]iP]]$.
@@
   $[[Γ2, nbs ⊢ [σ0]iP ≥ iQ ]]$.
-  Combining it with $[[ Γ2, nbs ⊢ σ0 : nas]]$
+  Combining it with $[[ Γ2, nbs ⊢ σ0 :{nas}]]$
   by \ruleref{\ottdruleDOneExistsLabel},
@@
   Suppose $[[Γ1]]$ and $[[Γ2]]$ are contexts 
-  and $[[ {Γ1} ⊆ {Γ2} ]]$.
+  and $[[ Γ1 ⊆ Γ2 ]]$.
   Then 
@@
       we obtain $[[Γ' ⊢ [σ]iM2 ≤ [σ]iK]]$, that is $[[Γ' ⊢  ∀pbs1.[σ]α⁻ ≤ [σ]iK]]$.
-      Since $[[ fv([σ]α⁻) ⊆ {Γ,α⁻} ]]$, it is disjoint from $[[pas]]$ and $[[pbs1]]$,
+      Since $[[ fv([σ]α⁻) ⊆ Γ,α⁻ ]]$, it is disjoint from $[[pas]]$ and $[[pbs1]]$,
       This way, by \cref{corollary:red-quant-elim}, 
@@
       Notice that we can weaken and rearrange the contexts without changing the sizes of the 
-      derivations: $[[Γ, pbs, pcs ⊢ [iPs/pas]iN' ≤ iM1']]$
-      and $[[Γ, pbs, pcs ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
+      derivations: $[[Γ, {pbs}, {pcs} ⊢ [iPs/pas]iN' ≤ iM1']]$
+      and $[[Γ, {pbs}, {pcs} ⊢ [iQs/pbs]iM2' ≤ iK']]$. This way, 
       the sum of the sizes of these derivations is smaller than the sum of the sizes of
@@
       Let us apply the induction hypothesis to these derivations, 
-      with the substitution $[[ Γ', pcs ⊢ σ ○ (iQs/pbs) : Γ, pbs, pcs  ]]$
+      with the substitution $[[ Γ', pcs ⊢ σ ○ (iQs/pbs) : Γ, {pbs}, {pcs}  ]]$
       (\cref{lemma:subst-domain-weakening}).

### 2023-09-04 proofs/tex/decl-typing-lemmas.tex
WIP: separate the algorithmic proofs
Grammar update: Γ = set

@@
             \begin{enumerate}
-                \item $[[Γ ⊢ σ : pas1]]$        
+                \item $[[Γ ⊢ σ :{pas1}]]$        
                 \item $[[Γ ; Φ ⊢ [σ]iN1' ● args ⇒> iM]]$
@@
 
-            Let us construct $[[Γ ⊢ σ0 : pas2]]$ in the following way:
+            Let us construct $[[Γ ⊢ σ0 :{pas2}]]$ in the following way:
             $$
@@
             \begin{enumerate}
-                \item $[[Γ ⊢ σ0 : pas2]]$ by construction;
+                \item $[[Γ ⊢ σ0 :{pas2}]]$ by construction;
                 \item $[[args ≠ ·]]$ as noted above;
@@
 \newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
-\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
 \newcommand{\npq}[1]{\ensuremath{\mathsf{npq}(#1)}}
@@
 
-            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ : pas]]$ and
+            By inversion, we have $[[σ]]$ such that $[[Γ ⊢ σ :{pas}]]$ and
             $[[Γ ; Φ1 ⊢ [σ]iN ● args ⇒> iM]]$ is inferred.
@@
             $[[Γ ; Φ2 ⊢ [σ]iN ● args ⇒> iM]]$.
-            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ : pas]]$,
+            Notice that the premises $[[args ≠ ·]]$, $[[Γ ⊢ σ :{pas}]]$,
             and $[[pas ≠ ·]]$ do not depend on the variable context,

### 2023-09-04 proofs/tex/equiv-lemmas.tex
Grammar update: Γ = set

@@
       and by \cref{lemma:wf-ctxt-equiv}, it is equivalent to 
-      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ iN']]$,
+      $[[Γ,  ({pas} ∩ fv iN') ⊢ iN']]$,
       which, by the induction hypothesis, is equivalent to
-      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ [mu]iM']]$.
+      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$.
 
       Analogously, $[[Γ ⊢ ∀pbs.iM']]$ is equivalent to
-      $[[Γ, (Ord {pbs} ∩ fv iM') ⊢ iM']]$.
+      $[[Γ, ( {pbs} ∩ fv iM') ⊢ iM']]$.
       By \cref{lemma:wf-subst}, it implies 
-      $[[Γ, Ord ({pas} ∩ fv iM') ⊢ [mu]iM']]$. And vice versa, 
-      $[[Γ, Ord ({pas} ∩ fv iM') ⊢ [mu]iM']]$ implies  
-      $[[Γ, Ord ({pbs} ∩ fv iM') ⊢ [mu-1][mu]iM']]$.
+      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$. And vice versa, 
+      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$ implies  
+      $[[Γ,  ({pbs} ∩ fv iM') ⊢ [mu-1][mu]iM']]$.
 
       This way, both $[[Γ ⊢ ∀pas.iN']]$ and $[[Γ ⊢ ∀pbs.iM']]$ are equivalent to
-      $[[Γ, Ord ({pas} ∩ fv iN') ⊢ [mu]iM']]$. 
+      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$. 
     \item For the cases of the positive types, the proofs are symmetric.
@@
 \begin{proof}
-  Suppose that $[[α± ∊ {Γ1}]]$. 
+  Suppose that $[[α± ∊ Γ1]]$. 
   Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
@@
 
-  Suppose that $[[α± ∉ {Γ1}]]$. 
+  Suppose that $[[α± ∉ Γ1]]$. 
   $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
@@
   Let us prove the forward direction.
-  Suppose that $[[α± ∊ {Γ1}]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
+  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
   By the definition of substitution normalization,
@@
 
-      Let us construct a subsitution $[[pas ⊢ iPs/pbs : pbs]]$ by
+      Let us construct a substitution $[[{pas} ⊢ iPs/pbs : {pbs}]]$ by
       extending $[[mu]]$ with arbitrary positive types on $[[{pbs} \ fv iM]]$.
@@
   Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
-  where $[[{Γi} ⊆ {Γ}]]$ and $[[{Γ1} ∩ {Γ2}= ∅]]$. Then
+  where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
   \begin{itemize}
   \item[$-$] assuming $[[Γ ⊢ iN]]$,~
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id : Ord fv iN]]$
+    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id :  fv iN]]$
   \item[$+$] assuming $[[Γ ⊢ iP]]$,~
-    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id : Ord fv iP]]$
+    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id :  fv iP]]$
   \end{itemize}
@@
       \begin{caseof}
-      \item $[[α⁻ ∉ {Γ1}]]$ and $[[α⁻ ∉ {Γ2}]]$ \label{case:var-not-in-ctxts}\\
-        Then $[[Γ ⊢ σi ≈ id : α⁻]]$ holds immediately,
+      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∉ Γ2]]$ \label{case:var-not-in-ctxts}\\
+        Then $[[Γ ⊢ σi ≈ id : {α⁻}]]$ holds immediately,
         since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
         $[[Γ ⊢ α⁻ ≈ α⁻]]$.
-      \item $[[α⁻ ∊ {Γ1}]]$ and $[[α⁻ ∊ {Γ2}]]$\\
-        This case is not possible by assumption: $[[{Γ1} ∩ {Γ2}= ∅]]$.
-      \item $[[α⁻ ∊ {Γ1}]]$ and $[[α⁻ ∉ {Γ2}]]$\\
+      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
+        This case is not possible by assumption: $[[Γ1 ∩ Γ2= ∅]]$.
+      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∉ Γ2]]$\\
         Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
         which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
-        and hence, $[[Γ ⊢ σ1 ≈ id : α⁻]]$.
+        and hence, $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$.
 
-        $[[Γ ⊢ σ2 ≈ id : α⁻]]$ holds since $[[ [σ2]α⁻ ]] = [[α⁻]]$,
+        $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$ holds since $[[ [σ2]α⁻ ]] = [[α⁻]]$,
         similarly to \cref{case:var-not-in-ctxts}.
 
-      \item $[[α⁻ ∉ {Γ1}]]$ and $[[α⁻ ∊ {Γ2}]]$\\
+      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
         Then we have $[[Γ ⊢ α⁻ ≤ [σ2]α⁻]]$,
         which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ α⁻ ≈ [σ2]α⁻]]$,
-        and hence, $[[Γ ⊢ σ2 ≈ id : α⁻]]$.
+        and hence, $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$.
 
-        $[[Γ ⊢ σ1 ≈ id : α⁻]]$ holds since $[[ [σ1]α⁻ ]] = [[α⁻]]$,
+        $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$ holds since $[[ [σ1]α⁻ ]] = [[α⁻]]$,
         similarly to \cref{case:var-not-in-ctxts}.
@@
     $[[σ1']]$, noting that $[[Γ, pas ⊢ σ1' : Γ1, pas]]$,
-    and $[[{Γ1, pas} ∩ {Γ2} = ∅]]$.
+    and $[[(Γ1, pas) ∩ Γ2 = ∅]]$.
 
@@
     \begin{align}
-      [[Γ, pas ⊢ σ1' ≈ id : Ord fv iM]] \label{fact:subs-proper-sub:forall-ih}\\
-      [[Γ, pas ⊢ σ2 ≈ id : Ord fv iM]]  \label{fact:subs-proper-sub:forall-ih2}
+      [[Γ, pas ⊢ σ1' ≈ id :  fv iM]] \label{fact:subs-proper-sub:forall-ih}\\
+      [[Γ, pas ⊢ σ2 ≈ id :  fv iM]]  \label{fact:subs-proper-sub:forall-ih2}
     \end{align}
 
-    Then $[[Γ ⊢ σ2 ≈ id : Ord fv ∀pas.iM]]$ holds by strengthening of
+    Then $[[Γ ⊢ σ2 ≈ id :  fv ∀pas.iM]]$ holds by strengthening of
     \ref{fact:subs-proper-sub:forall-ih2}:
@@
     $[[Γ, pas ⊢ [σ2]β± ≈ β±]]$ is strengthened to $[[Γ ⊢ [σ2]β± ≈ β±]]$, because
-    $[[fv [σ2]β±]] = [[fv β±]] = \{[[β±]]\} \subseteq [[{Γ}]]$.
+    $[[fv [σ2]β±]] = [[fv β±]] = \{[[β±]]\} \subseteq [[Γ]]$.
 
-    To show that $[[Γ ⊢ σ1 ≈ id : Ord fv ∀pas.iM]]$, let us take an arbitrary
+    To show that $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$, let us take an arbitrary
     $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.
@@
              &= [[ [σ1]β± ]]
-             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ {Γ}]] = \emptyset$}
+             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
     \end{aligned}
@@
     This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
-    $[[Γ ⊢ σ1 ≈ id : Ord fv ∀pas.iM]]$.
+    $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$.
 
@@
     \begin{align}
-      &[[Γ ⊢ σi ≈ id : Ord fv iP]] \label{fact:subs-proper-sub:arrow-ih1}\\
-      &[[Γ ⊢ σi ≈ id : Ord fv iM]] \label{fact:subs-proper-sub:arrow-ih2}
+      &[[Γ ⊢ σi ≈ id :  fv iP]] \label{fact:subs-proper-sub:arrow-ih1}\\
+      &[[Γ ⊢ σi ≈ id :  fv iM]] \label{fact:subs-proper-sub:arrow-ih2}
     \end{align}
@@
     to conclude:
-    $[[Γ ⊢ σi ≈ id : Ord fv (iP → iM)]]$.
+    $[[Γ ⊢ σi ≈ id :  fv (iP → iM)]]$.
 
@@
     Applying the induction hypothesis to $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$, we obtain
-    $[[Γ ⊢ σi ≈ id : Ord fv iP]]$. Since $[[fv ↑iP]] = [[fv iP]]$, we can
-    conclude: $[[Γ ⊢ σi ≈ id : Ord fv ↑iP]]$.
+    $[[Γ ⊢ σi ≈ id :  fv iP]]$. Since $[[fv ↑iP]] = [[fv iP]]$, we can
+    conclude: $[[Γ ⊢ σi ≈ id :  fv ↑iP]]$.
   \item The positive cases are proved symmetrically.
@@
     [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iN]] \\
+                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
     [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iN]] \\
+                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
     [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iP]] \\
+                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
     [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id : Ord fv iP]] \\
+                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
   \end{align*}
@@
     then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ : Ord fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 : Ord fv iQ]]$;
+    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
   \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
     then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ : Ord fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 : Ord fv iM]]$.
+    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
   \end{itemize}
@@
     Then by \cref{corollary:subst-proper-subt},
-    $[[Γ ⊢ σ2 ○ σ1 ≈ id : Ord fv iP]]$.
+    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
     By a symmetric argument, we also have:
-    $[[Γ ⊢ σ1 ○ σ2 ≈ id : Ord fv iQ]]$.
+    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.
 
     Now, we prove that
-    $[[Γ ⊢ σ2 ○ σ1 ≈ id : Ord fv iP]]$ and
-    $[[Γ ⊢ σ1 ○ σ2 ≈ id : Ord fv iQ]]$
+    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$ and
+    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$
     implies that $[[σ1]]$ and $[[σ1]]$
@@
     \item $[[iN = α⁻]]$\\
-      Then since by inversion, $[[α⁻ ∊ {Γ}]]$, 
+      Then since by inversion, $[[α⁻ ∊ Γ]]$, 
       $[[Γ' ⊢ [σ1]α⁻ ≈ [σ2]α⁻]]$ holds by definition of $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
@@
       we apply \ruleref{\ottdruleDOneForallLabel}
-      with the substitution $[[Γ', pas ⊢ id : pas]]$, 
+      with the substitution $[[Γ', pas ⊢ id :{pas}]]$, 
       noting that $[[Γ', pas ⊢ [id][σ1]iM ≤ [σ2]iM ]]$
@@
       \begin{enumerate} 
-        \item $[[Γ,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pbs ⊢ σ1 : pas ]]$ and 
-        \item $[[Γ,pas ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas ⊢ σ2 : pbs ]]$.
+        \item $[[Γ,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pbs ⊢ σ1 :{pas} ]]$ and 
+        \item $[[Γ,pas ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas ⊢ σ2 : {pbs} ]]$.
       \end{enumerate}
@@
       \begin{enumerate} 
-        \item $[[Γ,pas,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pas,pbs ⊢ σ1 : Γ,pas,pbs ]]$ and 
-        \item $[[Γ,pas,pbs ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas,pbs ⊢ σ2 : Γ,pas,pbs ]]$.
+        \item $[[Γ, {pas}, {pbs} ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ, {pas}, {pbs} ⊢ σ1 : Γ, {pas}, {pbs} ]]$ and 
+        \item $[[Γ, {pas}, {pbs} ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ, {pas}, {pbs} ⊢ σ2 : Γ, {pas}, {pbs} ]]$.
       \end{enumerate}
@@
       there exists a bijection $[[μ : fv iM ↔ fv iN]]$ such that 
-      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ : Ord fv iM]]$ and 
-      $[[Γ,pas,pbs ⊢ σ1 ≈ Sub μ-1 : Ord fv iN]]$.
+      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ :  fv iM]]$ and 
+      $[[Γ, {pas}, {pbs} ⊢ σ1 ≈ Sub μ-1 :  fv iN]]$.
 
@@
       assume $[[ [μ]γ⁺]] \notin [[pas]]$.
-      Then since $[[ Γ,pbs ⊢ σ1 : pas ]]$, 
+      Then since $[[ Γ,pbs ⊢ σ1 :{pas} ]]$, 
       $[[σ1]]$ acts as identity outside of $[[pas]]$, i.e.
@@
       Since
-      $[[Γ,pas,pbs ⊢ σ1 ≈ Sub μ-1 : Ord fv iN]]$, 
+      $[[Γ, {pas}, {pbs} ⊢ σ1 ≈ Sub μ-1 :  fv iN]]$, 
       application of $[[σ1]]$ to $[[ [μ]γ⁺ ]] \in [[fv iN]]$
       is equivalent to application of $[[Sub μ-1]]$, then 
-      $[[ Γ,pas,pbs ⊢ [Sub μ-1][Sub μ]γ⁺ ≈ [Sub μ]γ⁺ ]]$, i.e.
-      $[[Γ,pas,pbs ⊢ γ⁺ ≈ [Sub μ]γ⁺]]$, 
+      $[[ Γ, {pas}, {pbs} ⊢ [Sub μ-1][Sub μ]γ⁺ ≈ [Sub μ]γ⁺ ]]$, i.e.
+      $[[Γ, {pas}, {pbs} ⊢ γ⁺ ≈ [Sub μ]γ⁺]]$, 
       which means $[[γ⁺ ∊ fv [Sub μ]γ⁺]] \subseteq [[fv iN]]$.
@@
 
-      Since $[[ Γ,pas ⊢ σ2 : pbs ]]$ and 
-      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ : Ord fv iM]]$, 
+      Since $[[ Γ,pas ⊢ σ2 : {pbs} ]]$ and 
+      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ :  fv iM]]$, 
       we have
-      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ|{pbs} : Ord fv iM]]$:
+      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ|{pbs} :  fv iM]]$:
       for any $[[α± ∊ fv iM \ {pbs}]]$, $[[ [σ2]α± = α± ]]$
-      since $[[ Γ,pas ⊢ σ2 : pbs ]]$, and
+      since $[[ Γ,pas ⊢ σ2 : {pbs} ]]$, and
       $[[ [Sub μ|{pbs}]α± = α± ]]$ by definition of substitution restriction;
       for $[[β⁺ ∊ {pbs}]]$, $[[ [Sub μ|{pbs}]β⁺ = [μ]β⁺ ]]$,
-      and thus, $[[Γ,pas,pbs ⊢ [σ2]β⁺ ≈ [Sub μ]β⁺ ]]$
-      can be rewritten to $[[Γ,pas,pbs ⊢ [σ2]β⁺ ≈ [Sub μ|{pbs}]β⁺ ]]$.
+      and thus, $[[Γ, {pas}, {pbs} ⊢ [σ2]β⁺ ≈ [Sub μ]β⁺ ]]$
+      can be rewritten to $[[Γ, {pas}, {pbs} ⊢ [σ2]β⁺ ≈ [Sub μ|{pbs}]β⁺ ]]$.
 
       By \cref{lemma:equiv-subst-on-same-term},
-      $[[Γ,pas,pbs ⊢ σ2 ≈ Sub μ|{pbs} : Ord fv iM]]$ implies
-      $[[Γ,pas,pbs ⊢ [σ2]iM ≈ [Sub μ|{pbs}]iM]]$.
-      By similar reasoning, $[[Γ,pas,pbs ⊢ [σ1]iN ≈ [Sub μ-1|{pas}]iN]]$.
+      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ|{pbs} :  fv iM]]$ implies
+      $[[Γ, {pas}, {pbs} ⊢ [σ2]iM ≈ [Sub μ|{pbs}]iM]]$.
+      By similar reasoning, $[[Γ, {pas}, {pbs} ⊢ [σ1]iN ≈ [Sub μ-1|{pas}]iN]]$.
 
@@
       \begin{align} 
-        [[Γ,pas,pbs ⊢ [Sub μ-1|{pas}]iN ≤ iM]] \label{fact:mu-inv-n-sub-m}\\
-        [[Γ,pas,pbs ⊢ [Sub μ|{pbs}]iM ≤ iN]] \label{fact:mu-m-subt-n}
+        [[Γ, {pas}, {pbs} ⊢ [Sub μ-1|{pas}]iN ≤ iM]] \label{fact:mu-inv-n-sub-m}\\
+        [[Γ, {pas}, {pbs} ⊢ [Sub μ|{pbs}]iM ≤ iN]] \label{fact:mu-m-subt-n}
       \end{align}
@@
       (\cref{lemma:subst-pres-subt}), we have
-      $[[Γ,pas,pbs ⊢ [Sub μ|{pbs}][Sub μ-1|{pas}]iN ≤ [Sub μ|{pbs}]iM]]$. 
+      $[[Γ, {pas}, {pbs} ⊢ [Sub μ|{pbs}][Sub μ-1|{pas}]iN ≤ [Sub μ|{pbs}]iM]]$. 
       By contracting $[[μ-1|_{pas} ○ μ|_{pbs}]] = [[μ|_{pbs}-1 ○ μ|_{pbs}]]$
       (notice that $[[fv iN ∩ {pbs} = ∅]]$), 
-      we have $[[Γ,pas,pbs ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
-      means $[[Γ,pas,pbs ⊢ iN ≈ [Sub μ|{pbs}]iM]]$, and by strengthening, $[[Γ,pas⊢ iN ≈ [Sub μ|{pbs}]iM]]$.
+      we have $[[Γ, {pas}, {pbs} ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
+      means $[[Γ, {pas}, {pbs} ⊢ iN ≈ [Sub μ|{pbs}]iM]]$, and by strengthening, $[[Γ,pas⊢ iN ≈ [Sub μ|{pbs}]iM]]$.
       % Symmetrically, $[[Γ,pbs ⊢ iM ≈ [Sub μ|_{pbs}-1]iN]]$.

### 2023-09-04 proofs/tex/norm-lemmas.tex
WIP: leftovers
WIP: description, notation
Grammar update: Γ = set

@@
 \begin{proof}
-  By straightforward induction on $[[nf(iN) = iM]]$.
+  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
+\end{proof}
+
+\begin{lemma}
+  \label{lemma:uv-nf}
+  Algorithmic variables are not changed by the normalization
+  \begin{itemize}
+  \item[$-$] $[[uv uN]] \equiv [[uv nf(uN)]]$
+  \item[$+$] $[[uv uP]] \equiv [[uv nf(uP)]]$
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  By straightforward induction on $[[uN]]$ and mutually on $[[uP]]$.
 \end{proof}
@@
                           &= [[ nf(∀pas.[σ]iM) ]]
-                          && \text{Assuming $[[{pas} ∩ {Γ1}]] = \emptyset$
-                             and $[[{pas} ∩ {Γ2}]] = \emptyset$}\\
+                          && \text{Assuming $[[{pas} ∩ Γ1]] = \emptyset$
+                             and $[[{pas} ∩ Γ2]] = \emptyset$}\\
                           &= [[ ∀pbs.nf([σ]iM) ]]
@@
      $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
-     we can assume that $[[{pas'} ∩ {Γ1}]] = \emptyset$, and $[[{pas'} ∩ {Γ2}]]
+     we can assume that $[[{pas'} ∩ Γ1]] = \emptyset$, and $[[{pas'} ∩ Γ2]]
      = \emptyset$.
@@
 \end{lemma}
-
 \begin{proof}
@@
                                              [[ [σ]iM ]] &\text{is normal} \\
-                                             \end{cases}
-                                           && \text{by congruence of normality \ilyam{lemma?}}\\
+                                             \end{cases}\\
                                            &\iff
@@
                                                [[ σ|fv(iP) ∪ fv(iM)]] &\text{is normal} \\
-                                             \end{cases}
-                                          \iff&&%
+                                             \end{cases}\\
+                                           &\iff
                                              \begin{cases}

### 2023-09-04 proofs/tex/ord-lemmas.tex
WIP: leftovers
Grammar update: Γ = set

@@
 \end{corollary}
+
+\begin{corollary}[Idempotency of ordering]
+  \label{corollary:ord-idemp}
+  \hfill
+  \begin{itemize}
+    \item[$-$] If $[[ ord varset in iN = ordVars ]]$ then 
+      $[[ ord {ordVars} in iN = ordVars ]]$,
+    \item[$+$] If $[[ ord varset in iP = ordVars ]]$ then 
+      $[[ ord {ordVars} in iP = ordVars ]]$;
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  By \cref{lemma:ord-soundness,corollary:ord-weakening}.
+\end{proof}
+
+\end{proof}
 
@@
     \item $[[iN = na]]$ \\
-      If $[[na ∉ {Γ1}]]$ then $[[ [σ]na = na ]]$ and $[[ ord varset in [σ]na ]] = [[ ord varset in na ]]$, 
+      If $[[na ∉ Γ1]]$ then $[[ [σ]na = na ]]$ and $[[ ord varset in [σ]na ]] = [[ ord varset in na ]]$, 
       as requried.
-      If $[[na ∊ {Γ1}]]$ then $[[na ∉ varset]]$, so $[[ ord varset in na ]] = [[·]]$.
-      Moreover, $[[Γ2 ⊢ σ : Γ1]]$ means $[[ fv([σ]na) ⊆ {Γ2} ]]$, and thus, 
-      as a set, $[[ ord varset in [σ]na ]] = [[varset ∩ fv([σ]na)]] \subseteq [[varset ∩ {Γ2}]] = [[·]]$.
+      If $[[na ∊ Γ1]]$ then $[[na ∉ varset]]$, so $[[ ord varset in na ]] = [[·]]$.
+      Moreover, $[[Γ2 ⊢ σ : Γ1]]$ means $[[ fv([σ]na) ⊆ Γ2 ]]$, and thus, 
+      as a set, $[[ ord varset in [σ]na ]] = [[varset ∩ fv([σ]na)]] \subseteq [[varset ∩ Γ2]] = [[·]]$.
     \item $[[iN = ∀pas.iM]]$\\
-      We can assume $[[{pas} ∩ {Γ1} = ∅]]$
+      We can assume $[[{pas} ∩ Γ1 = ∅]]$
       and $[[{pas} ∩ varset = ∅]]$. Then

### 2023-09-04 proofs/tex/subst-lemmas.tex
Grammar update: Γ = set

@@
   \label{lemma:subst-restr-sig}
-  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset : Ord {Γ1} ∩ varset]]$.
+  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset :  Γ1 ∩ varset]]$.
 \end{lemma}
 \begin{proof}
-  Let us take an arbitrary $[[α± ∊ {Γ1} ∩ varset]]$.
-  Since $[[α± ∊ {Γ1}]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.
+  Let us take an arbitrary $[[α± ∊ Γ1 ∩ varset]]$.
+  Since $[[α± ∊ Γ1]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.
 
-  Let us take an arbitrary $[[α± ∉ {Γ1} ∩ varset]]$.
+  Let us take an arbitrary $[[α± ∉ Γ1 ∩ varset]]$.
   If $[[α± ∉ varset]]$ then $[[ [σ|varset]α± ]] = [[α±]]$ by definition
   of restriction. 
-  If $[[α± ∊ varset \ {Γ1}]]$ then $[[ [σ|varset]α± ]] = [[ [σ]α± ]]$ by definition
+  If $[[α± ∊ varset \ Γ1]]$ then $[[ [σ|varset]α± ]] = [[ [σ]α± ]]$ by definition
   and $[[ [σ]α± ]] = [[α±]]$ by the signature of $[[σ]]$.
@@
 
-  If $[[α± ∊ varset]]$ then $[[α± ∉ {Γ1}]]$ by assumption.
+  If $[[α± ∊ varset]]$ then $[[α± ∉ Γ1]]$ by assumption.
   Then $[[ [σ|varset]α± = [σ]α±]]$ by definition of restricted substitution, and
@@
   \begin{itemize}
-    \item [$+$] if $[[{Γ1} ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
-    \item [$-$] if $[[{Γ1} ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
+    \item [$+$] if $[[Γ1 ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
+    \item [$-$] if $[[Γ1 ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
   \end{itemize}
@@
   \label{lemma:subst-range-weakening}
-  Suppose that $[[{Γ2} ⊆ {Γ'2}]]$ are contexts and 
+  Suppose that $[[Γ2 ⊆ Γ'2]]$ are contexts and 
   $[[σ]]$ is a substitution.
@@
 \begin{proof}
-  For any $[[α± ∊ {Γ1}]]$, 
+  For any $[[α± ∊ Γ1]]$, 
   $[[Γ2 ⊢ σ : Γ1]]$ gives us  $[[ Γ2 ⊢ [σ]α± ]]$, which can be
@@
 \begin{lemma}[]
-  Suppose that $[[{Γ'} ⊆ {Γ}]]$,
+  Suppose that $[[Γ' ⊆ Γ]]$,
   $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
@@
     \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$;
+    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$;
     \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ {Γ'}]]$.
+    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$.
   \end{enumerate}
@@
     \item $[[iP]] = [[α⁺]] \in [[Γ']]$.
-      Then $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$ means $[[Γ ⊢ σ1 ≈ σ2 : {α⁺}]]$, 
+      Then $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$ means $[[Γ ⊢ σ1 ≈ σ2 : {α⁺}]]$, 
       i.e. $[[Γ ⊢ [σ1]α⁺ ≈ [σ2]α⁺ ]]$, which holds by assumption.
-    \item $[[iP]] = [[α⁺]] \in [[{Γ} \ {Γ'}]]$.
-      Then $[[fv iP ∩ {Γ'}]] = [[∅]]$, 
-      so $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ {Γ'}]]$ holds vacuously.
+    \item $[[iP]] = [[α⁺]] \in [[Γ \ Γ']]$.
+      Then $[[fv iP ∩ Γ']] = [[∅]]$, 
+      so $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$ holds vacuously.
     \item $[[iP]] = [[↓iN]]$.
@@
       \end{enumerate}
-      This way, we obtain $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ {Γ'}]]$, 
+      This way, we obtain $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$, 
       which implies the required equivalence since 
-      $[[fv iP ∩ {Γ'}]] = [[fv ↓iN ∩ {Γ'}]] = [[fv iN ∩ {Γ'}]]$.
+      $[[fv iP ∩ Γ']] = [[fv ↓iN ∩ Γ']] = [[fv iN ∩ Γ']]$.
     \item $[[iP]] = [[∃nas.iQ]]$
@@
       \begin{enumerate}
-        \item $[[{Γ'} ⊆ {Γ, nas}]]$ since $[[{Γ'} ⊆ {Γ}]]$,
+        \item $[[Γ' ⊆ Γ, nas]]$ since $[[Γ' ⊆ Γ]]$,
         \item $[[Γ, nas ⊢ σi : Γ']]$ by weakening,
@@
     If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
-    $[[{Γ1} ∩ {Γ'2} = ∅ ]]$ and $[[ {Γ1} ∩ {Γ2} = ∅ ]]$ then 
+    $[[Γ1 ∩ Γ'2 = ∅ ]]$ and $[[ Γ1 ∩ Γ2 = ∅ ]]$ then 
     $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
@@
   If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
-  $[[ {Γ1} ∩ {Γ2} = ∅ ]]$, 
-  $[[ {Γ1} ∩ {Γ'2} = ∅ ]]$, and
-  $[[ {Γ'1} ∩ {Γ2} = ∅ ]]$ then 
+  $[[ Γ1 ∩ Γ2 = ∅ ]]$, 
+  $[[ Γ1 ∩ Γ'2 = ∅ ]]$, and
+  $[[ Γ'1 ∩ Γ2 = ∅ ]]$ then 
   $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
@@
   and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
-  If the variable $[[α±]]$ is in $[[{Γ'} \ {Γ1}]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
+  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
   and thus, $[[Γ2, Γ' ⊢ α± ]]$.
@@
       the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-     $[[fv(iP) \ {Γ1} ⊆ fv([σ]iP) ⊆ (fv(iP) \ {Γ1}) ∪ {Γ2}]]$;
+     $[[fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$;
     \item [$-$] for a type $[[iN]]$, 
       the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-      $[[fv(iN) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iN) \ {Γ1}) ∪ {Γ2}]]$.
+      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
   \end{enumerate}
@@
     \item $[[iP = α⁺]]$\\
-      If $[[α⁺ ∊ {Γ1}]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
-      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ {Γ2}]]$.
-      $[[fv(α⁺) \ {Γ1}]] = [[∅]]$, so $[[fv([σ]iP) \ {Γ1}  ⊆ fv([σ]α⁺)]]$
+      If $[[α⁺ ∊ Γ1]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
+      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ Γ2]]$.
+      $[[fv(α⁺) \ Γ1]] = [[∅]]$, so $[[fv([σ]iP) \ Γ1  ⊆ fv([σ]α⁺)]]$
       vacuously.
 
-      If $[[α⁺ ∉ {Γ1}]]$ then $[[ [σ]α⁺ = α⁺ ]]$, and 
-      $[[fv([σ]α⁺) = {α⁺}]] = [[{α⁺} \ {Γ1}]]$.
+      If $[[α⁺ ∉ Γ1]]$ then $[[ [σ]α⁺ = α⁺ ]]$, and 
+      $[[fv([σ]α⁺) = {α⁺}]] = [[{α⁺} \ Γ1]]$.
 
@@
       Then we need to show that $[[fv([σ]iP) = fv([σ]iQ) \ {nas}]]$
-      is a subset of $[[(fv(iP) \ {Γ1}) ∪ {Γ2}]]$ and a superset of
-      $[[fv(iP) \ {Γ1}]]$.
+      is a subset of $[[(fv(iP) \ Γ1) ∪ Γ2]]$ and a superset of
+      $[[fv(iP) \ Γ1]]$.
       Notice that $[[fv(iP) = fv(iQ) \ {nas}]]$ by definition.
       This way, we need to show that 
-      $[[fv(iQ) \ {nas} \ {Γ1} ⊆ fv([σ]iQ) \ {nas} ⊆ (fv(iQ) \ {nas} \ {Γ1}) ∪ {Γ2}]]$,
+      $[[fv(iQ) \ {nas} \ Γ1 ⊆ fv([σ]iQ) \ {nas} ⊆ (fv(iQ) \ {nas} \ Γ1) ∪ Γ2]]$,
 
       By the induction hypothesis, 
-      $[[fv([σ]iQ) ⊆ (fv(iQ) \ {Γ1}) ∪ {Γ2}]]$.
+      $[[fv([σ]iQ) ⊆ (fv(iQ) \ Γ1) ∪ Γ2]]$.
       So for the second inclusion, it suffices to show that
-      $[[((fv(iQ) \ {Γ1}) ∪ {Γ2}) \ {nas} ⊆ (fv(iQ) \ {nas} \ {Γ1}) ∪ {Γ2}]]$,
+      $[[((fv(iQ) \ Γ1) ∪ Γ2) \ {nas} ⊆ (fv(iQ) \ {nas} \ Γ1) ∪ Γ2]]$,
       which holds by set theoretical reasoning.
@@
       Also by the induction hypothesis,
-      $[[fv(iQ) \ {Γ1} ⊆ fv([σ]iQ)]]$,
+      $[[fv(iQ) \ Γ1 ⊆ fv([σ]iQ)]]$,
       and thus, by subtracting $[[{nas}]]$ from both sides,
-      $[[fv(iQ) \ {nas} \ {Γ1} ⊆ fv([σ]iQ) \ {nas}]]$.
+      $[[fv(iQ) \ {nas} \ Γ1 ⊆ fv([σ]iQ) \ {nas}]]$.
 
@@
       \begin{enumerate}
-        \item $[[ fv(iP) \ {Γ1} ⊆ fv([σ]iP) ⊆ (fv(iP) \ {Γ1}) ∪ {Γ2}]]$ and
-        \item $[[ fv(iM) \ {Γ1} ⊆ fv([σ]iM) ⊆ (fv(iM) \ {Γ1}) ∪ {Γ2}]]$.
+        \item $[[ fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$ and
+        \item $[[ fv(iM) \ Γ1 ⊆ fv([σ]iM) ⊆ (fv(iM) \ Γ1) ∪ Γ2]]$.
       \end{enumerate}
       We unite these inclusions vertically and obtain
-      $[[ fv(iP) \ {Γ1} ∪ fv(iM) \ {Γ1} ⊆ fv([σ]iN) ⊆ ((fv(iP) \ {Γ1}) ∪ {Γ2}) ∪ ((fv(iM) \ {Γ1}) ∪ {Γ2})]]$,
+      $[[ fv(iP) \ Γ1 ∪ fv(iM) \ Γ1 ⊆ fv([σ]iN) ⊆ ((fv(iP) \ Γ1) ∪ Γ2) ∪ ((fv(iM) \ Γ1) ∪ Γ2)]]$,
       which is equivalent to 
-      $[[ (fv(iP) ∪ fv(iM)) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iP) ∪ fv(iM)) \ {Γ1} ∪ {Γ2}]]$.
+      $[[ (fv(iP) ∪ fv(iM)) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iP) ∪ fv(iM)) \ Γ1 ∪ Γ2]]$.
       Since $[[fv(iP) ∪ fv(iM) = fv(iN)]]$, 
-      $[[fv(iN) \ {Γ1} ⊆ fv([σ]iN) ⊆ (fv(iN) \ {Γ1}) ∪ {Γ2}]]$.
+      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
     \item The cases when $[[iP = ↓iM]]$ and $[[iN = ↑iQ]]$ are proved

### 2023-09-04 proofs/tex/upgrade-lemmas.tex
Grammar update: Γ = set

@@
 Let us consider a type $[[iP]]$ well-formed in $[[Γ]]$.
-Some of its $[[Γ]]$-supertypes are also well-formed in a smaller context $[[{Δ} ⊆ {Γ}]]$.
+Some of its $[[Γ]]$-supertypes are also well-formed in a smaller context $[[{Δ} ⊆ Γ]]$.
 The upgrade is the operation that returns the least of such supertypes.
@@
                   &= [[ fv iP \ {pnas} ]]\\
-                  &\subseteq [[ {Γ} \ {pnas} ]]
+                  &\subseteq [[ Γ \ {pnas} ]]
                   &&\text{since $[[iP]]$ is well-formed in $[[Γ]]$}\\

### 2023-09-04 proofs/tex/wf-algo-lemmas.tex
minor
Grammar update: Γ = set

@@
   \begin{itemize}
-    \item[$+$] if $[[Γ ; Ξ ⊢ uP]]$ then $[[fv(uP) ⊆ {Γ}]]$ and $[[uv(uP) ⊆ Ξ]]$;
-    \item[$-$] if $[[Γ ; Ξ ⊢ uN]]$ then $[[fv(uN) ⊆ {Γ}]]$ and $[[uv(uN) ⊆ Ξ]]$.
+    \item[$+$] if $[[Γ ; Ξ ⊢ uP]]$ then $[[fv(uP) ⊆ Γ]]$ and $[[uv(uP) ⊆ Ξ]]$;
+    \item[$-$] if $[[Γ ; Ξ ⊢ uN]]$ then $[[fv(uN) ⊆ Γ]]$ and $[[uv(uN) ⊆ Ξ]]$.
   \end{itemize}
@@
   The proof is analogous to \cref{lemma:wf-soundness}.
-  The additional base case is when $[[Γ ; Ξ ⊢ uP]]$ is derived by \ruleref{\ottdruleWFATPUvarLabel},
+  The additional base case is when $[[Γ ; Ξ ⊢ uP]]$ is derived by \ruleref{\ottdruleWFATPUVarLabel},
   and the symmetric negative case.
-  In this case, $[[uP]] = [[α̂⁺]]$, and $[[uv(uP)]] = [[{α̂⁺} ⊆ Ξ]]$ by inversion; $[[fv(uP)]] = [[∅ ⊆ {Γ}]]$ vacuously.
+  In this case, $[[uP]] = [[α̂⁺]]$, and $[[uv(uP)]] = [[{α̂⁺} ⊆ Ξ]]$ by inversion; $[[fv(uP)]] = [[∅ ⊆ Γ]]$ vacuously.
 \end{proof}
@@
   \begin{itemize}
-  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$
+  \item[$+$] if $[[Γ1 ∩ fv iP]] = [[Γ2 ∩ fv iP]]$
     and $[[Ξ1 ∩ uv uP]] = [[Ξ2 ∩ uv uP]]$ then
     $[[Γ1 ; Ξ1 ⊢ uP]] \iff [[Γ2 ; Ξ2 ⊢ uP]]$, and
-  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$
+  \item[$-$] if $[[Γ1 ∩ fv iN]] = [[Γ2 ∩ fv iN]]$
     and $[[Ξ1 ∩ uv uN]] = [[Ξ2 ∩ uv uN]]$ then
@@
   \begin{itemize}
-    \item[$+$]  $[[Γ, nas ⊢ iP]]$ implies $[[Γ; nuas ⊢ [nuas/nas]iP]]$;
-    \item[$-$]  $[[Γ, nas ⊢ iN]]$ implies $[[Γ; nuas ⊢ [nuas/nas]iN]]$.
+    \item[$+$]  $[[Γ, nas ⊢ iP]]$ implies $[[Γ; {nuas} ⊢ [nuas/nas]iP]]$;
+    \item[$-$]  $[[Γ, nas ⊢ iN]]$ implies $[[Γ; {nuas} ⊢ [nuas/nas]iN]]$.
   \end{itemize}
@@
   \label{lemma:wf-weakening-algo}
-  Suppose that $[[{Γ1} ⊆ {Γ2}]]$,
+  Suppose that $[[Γ1 ⊆ Γ2]]$,
   and $[[Ξ1 ⊆ Ξ2]]$. Then
@@
   By \cref{lemma:wf-soundness},
-  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ {Γ1}]]$,
-  which means that $[[fv(iP) ⊆ {Γ2}]]$,
-  and thus, $[[fv(iP)]] = [[fv(iP) ∩ {Γ1}]] = [[fv(iP) ∩ {Γ2}]]$.
+  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ Γ1]]$,
+  which means that $[[fv(iP) ⊆ Γ2]]$,
+  and thus, $[[fv(iP)]] = [[fv(iP) ∩ Γ1]] = [[fv(iP) ∩ Γ2]]$.
   Then by \cref{lemma:wf-ctxt-equiv}, $[[Γ2 ⊢ iP]]$. 

### 2023-09-04 proofs/tex/wf-lemmas.tex
WIP: description, notation
WIP: separate the algorithmic proofs
Grammar update: Γ = set

--- 2023-08-08 proofs/tex/wf-lemmas.tex
@@
-\begin{lemma}[Equivalent Contexts]
+\begin{lemma}[Soundness of type well-formedness]
+    \label{lemma:wf-soundness}
+    \hfill
+
+  \begin{itemize}
+    \item[$+$] if $[[Γ ⊢ iP]]$ then $[[fv(iP) ⊆ Γ]]$;
+    \item[$-$] if $[[Γ ⊢ iN]]$ then $[[fv(iN) ⊆ Γ]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  The proof is done by a simple structural induction on 
+  $[[Γ⊢ iP]]$ and mutually, $[[Γ ⊢ iN]]$.
+  \begin{caseof}
+    \item $[[Γ ⊢ α±]]$ means by inversion that $[[α±]] \in [[Γ]]$,
+      that is, $[[α±]] = [[fv(α±) ⊆ Γ]]$.
+    \item $[[Γ ⊢ iQ → iM]]$ means by inversion that
+      $[[Γ ⊢ iQ]]$ and $[[Γ ⊢ iM]]$.
+      Then by the induction hypothesis, $[[fv(iQ) ⊆ Γ]]$ and $[[fv(iM) ⊆ Γ]]$,
+      and hence, $[[fv(iQ → iM) = fv(iQ) ∪ fv(iM)]] \subseteq [[Γ]]$.
+    \item the cases when $[[iP = ↓iN']]$ or $[[iN = ↑iP']]$ are proven analogously.
+    \item $[[Γ ⊢ ∀pas.iM]]$ means by inversion that $[[Γ, pas ⊢ iM]]$.
+      Then by the induction hypothesis, $[[fv(iM) ⊆ Γ, pas]]$, and hence,
+      $[[fv(∀pas.iM) = fv(iM) \ {pas}]] \subseteq [[{Γ, pas} \ {pas} = Γ]]$.
+    \item The case $[[iP = ∃nas.iQ]]$ is proven analogously.
+  \end{caseof}
+\end{proof}
+
+\begin{lemma}[Completeness of type well-formedness]
   \label{lemma:wf-ctxt-equiv}
@@
   \begin{itemize}
-  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$ then
+  \item[$+$] if $[[Γ1 ∩ fv iP]] = [[Γ2 ∩ fv iP]]$ then
     $[[Γ1 ⊢ iP]] \iff [[Γ2 ⊢ iP]]$,
-  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$ then
+  \item[$-$] if $[[Γ1 ∩ fv iN]] = [[Γ2 ∩ fv iN]]$ then
     $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
@@
 \begin{proof}
-  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
+  By simple mutual induction on $[[iP]]$ and $[[iN]]$.
 \end{proof}
 
-\begin{lemma}[Well-formedness Context Weakening]
+\begin{corollary}[Context Strengthening]
+  \label{corollary:wf-ctxt-strengthening}
+  \hfill
+
+  \begin{itemize}
+    \item [$+$] If $[[Γ ⊢ iP]]$ then $[[fv(iP) ⊢ iP]]$;
+    \item [$-$] If $[[Γ ⊢ iN]]$ then $[[fv(iN) ⊢ iN]]$.
+  \end{itemize}
+\end{corollary}
+\begin{proof}
+  It follows from \cref{lemma:wf-ctxt-equiv} and \cref{lemma:wf-soundness}.
+  \begin{itemize}
+    \item [$+$] 
+      By \cref{lemma:wf-soundness}, $[[fv(iP) ⊆ Γ]]$, and hence,
+      $[[Γ ∩ fv iP = fv iP ]]$, which makes \cref{lemma:wf-ctxt-equiv}
+      applicable fore contexts $[[Γ]]$ and $[[fv(iP)]]$.
+    \item[$-$] The negative case is proven analogously.
+  \end{itemize}
+
+\begin{corollary}[Well-formedness Context Weakening]
   \label{lemma:wf-weakening}
-  Suppose that $[[{Γ1} ⊆ {Γ2}]]$, then
+  Suppose that $[[Γ1 ⊆ Γ2]]$, then
   \begin{itemize}
@@
   \end{itemize}
-\end{lemma}
+\end{corollary}
 \begin{proof}
-  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
+  By \cref{lemma:wf-soundness},
+  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ Γ1]]$,
+  which means that $[[fv(iP) ⊆ Γ2]]$,
+  and thus, $[[fv(iP)]] = [[fv(iP) ∩ Γ1]] = [[fv(iP) ∩ Γ2]]$.
+  Then by \cref{lemma:wf-ctxt-equiv}, $[[Γ2 ⊢ iP]]$. 
+  The negative case is symmetric.
 \end{proof}
@@
   Suppose that all the types below are well-formed in $[[Γ]]$ and
-  $[[{Γ'} ⊆ {Γ}]]$. Then
+  $[[Γ' ⊆ Γ]]$. Then
   \begin{itemize}
@@
       \begin{itemize}
-        \item if $[[α⁺ ∊ {Γ1}]]$ then $[[ Γ2 ⊢ [σ]α⁺ ]]$, 
+        \item if $[[α⁺ ∊ Γ1]]$ then $[[ Γ2 ⊢ [σ]α⁺ ]]$, 
           and by weakening (\cref{lemma:wf-weakening}),
           $[[ Γ, Γ2 ⊢ [σ]α⁺ ]]$;
-        \item if $[[α⁺ ∊ {Γ} \ {Γ1}]]$ then $[[ [σ]α⁺ = α⁺ ]]$,
+        \item if $[[α⁺ ∊ Γ \ Γ1]]$ then $[[ [σ]α⁺ = α⁺ ]]$,
           and by \ruleref{\ottdruleWFTPVarLabel}, $[[ Γ, Γ2 ⊢ α⁺ ]]$.
@@
       by inversion, and by the induction hypothesis, 
-      $[[Γ, nas, Γ2 ⊢ [σ]iQ]]$.
+      $[[Γ, {nas}, Γ2 ⊢ [σ]iQ]]$.
       Then by  \ruleref{\ottdruleWFTExistsLabel}, 
-      $[[Γ, nas, Γ2 ⊢ ∃nas.[σ]iQ]]$, 
+      $[[Γ, {nas}, Γ2 ⊢ ∃nas.[σ]iQ]]$, 
       which by definition of substitution is rewritten as 
@@
   \end{caseof}
-
 \end{proof}

### 2023-09-05 proofs/main.tex
WIP: separate the algorithmic proofs
Grammar update: Γ = set
Rearrange proofs

@@
 
-\newcommand{\depth}[1]{\mathsf{depth}(#1)}
-\newcommand{\size}[1]{\mathsf{size}(#1)}
+\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
+\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
 
@@
 
-\section{Declarative System}
-First, we present the top-level system, which is easy to understand.
+\section{Declarative Type Systems}
 
@@
   \ottiN\ottinterrule
-  \ottv\ottinterrule
-  \ottc\ottinterrule
 }
@@
 
-\subsection{Contexts}
+\subsection{Contexts and Well-formedness}
 
@@
   type variables. The concatenation $[[Γ1, Γ2]]$ means the 
-  union of two contexts  $[[{Γ1} ∪ {Γ2}]]$.
+  union of two contexts $[[Γ1 ∪ Γ2]]$.
 \end{definition}
+
+$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
+which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
+(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
+
+\hfill\\
+\ottdefnsWFT
 
@@
   \begin{enumerate}
-    \item for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ {Γ'}]], [[ [σ]α± = α± ]]$.
+    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
+    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
   \end{enumerate}
@@
   Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|{Γ}]]$.
+  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
 \end{definition}
@@
   the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ {Γ}]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
+  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
 \end{definition}
@@
 
-\subsection{Declarative Typing}
-\ottdefnsDT
-
 \subsection{Declarative Subtyping}
@@
 
-\subsection{Well-Formedness}
-\ottdefnsWFT
-\ottdefnsWFAT
-
-\section{Algorithm}
+\section{Algorithmic Type System}
 
@@
 
-In several places of our algorithm, 
+In several places of our algorithm, in particular, during anti-unification,
 we turn a declarative type into the algorithmic one
@@
   and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution turning each $[[αi⁻]]$ from $[[nas]]$ into 
-  $[[αî⁻]]$ from $[[nuas]]$.
-  Positive algorithmization $[[ puas/pas ]]$
-  is defined symmetrically.
+  Then $[[ nuas/nas ]]$ is a substitution-like procedure turning each $[[αi⁻ ∊ {nas}]]$ 
+  into $[[αî⁻ ∊ {nuas}]]$.
 \end{definition}
 
-\subsection{Contexts}
+Conversely, we can turn algorithmic type variables
+into declarative type variables via \emph{dealgorithmization}.
+
+\begin{definition}[Variable Dealgorithmization]
+  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
+  and $[[nas]]$ is a list of negative type variables of the same length. 
+  Then $[[ nas/nuas ]]$ is a substitution-like procedure turning each
+  $[[αî⁻ ∊ {nuas}]]$ into $[[αi⁻ ∊ {nas}]]$.
+\end{definition}
+
+\subsection{Contexts and Well-formedness}
 
@@
   \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
+  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
 \end{definition}
@@
 of the type is contained in $[[Γ]]$.
+
+\hfill\\
+\ottdefnsWFAT
 
@@
 \end{definition}
-
-\subsection{Singularity}
-\ottdefnsSING
-
-% \subsection{Algorithmic Equivalence}
-% \ottdefnsEOneA
 
@@
 
-\subsection{Typing}
+\section{Declarative Typing}
+
+\subsection{Grammar}
+
+\bigskip
+\ottgrammartabular{
+  \ottv\ottinterrule
+  \ottc\ottinterrule
+}
+
+\subsection{Declarative Type Inference}
+\ottdefnsDT
+
+\section{Algorithmic Typing}
+
+\subsection{Algorithmic Type Inference}
 \ottdefnsAT
 
-\section{Proofs}
+\subsection{Singularity}
+\ottdefnsSING
 
-\subsection{Type well-formedness}
+\section{Properties of the Declarative Type System}
+
+\subsection{Type Well-formedness}
 \input{../_build/_gen/tex/wf-lemmas.tex}
@@
 
-\subsection{Overview}
-\input{../_build/_gen/tex/lemmas-overview.tex}
-
 \subsection{Variable Ordering}
@@
 \input{../_build/_gen/tex/equiv-lemmas.tex}
+
+\section{Properties of the Algorithmic Type System}
+
+\subsection{Algorithmic Type Well-formedness}
+\input{../_build/_gen/tex/wf-algo-lemmas.tex}
+
+\subsection{Substitution}
+\input{../_build/_gen/tex/subst-algo-lemmas.tex}
+
+\subsection{Normalization}
+\input{../_build/_gen/tex/norm-alg-lemmas.tex}
+
+\subsection{Equivalence}
+\input{../_build/_gen/tex/equiv-alg-lemmas.tex}
 
@@
 
+\section{Properties of the Declarative Typing}
+\input{../_build/_gen/tex/decl-typing-lemmas.tex}
+
+\section{Properties of the Algorithmic Typing}
+
 \subsection{Singularity}
@@
 
-\subsection{Declarative Typing}
-\input{../_build/_gen/tex/decl-typing-lemmas.tex}
-
-\subsection{Algorithmic Typing}
+\subsection{Correctness of the Typing Algorithm}
 \input{../_build/_gen/tex/alg-typing-lemmas.tex}

### 2023-09-05 proofs/tex/alg-pos-subtyping-lemmas.tex
Rearrange proofs

@@
             $[[ Γ,ncs  ⊢ [iNs/nbs][uσ]α̂⁺ ≥ iQ0 ]]$ for some $[[iN]]$, 
-            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ Γ = ∅ ]]$,
+            and since $[[ {nbs} ∩ fv([uσ]α̂⁺) ⊆ {nbs} ∩ {Θ(α̂⁺)} ⊆ {nbs} ∩ Γ]] = [[ ∅ ]]$,
             $[[ [iNs/nbs][uσ]α̂⁺ = [uσ]α̂⁺ ]]$, that is $[[ Γ,ncs ⊢ [uσ]α̂⁺ ≥ iQ0]]$.
@@
             \begin{enumerate}
-                \item if this variable is a metavariable $[[αî⁻]] \in [[â⁻*]]$, then
+                \item if this variable is a algorithmic variable $[[αî⁻]] \in [[â⁻*]]$, then
                 $[[ [uσ0]αî⁻ ]] = [[ [σ]αi⁻ ]]$ and 
                 $[[ [uσ][σ][nas/â⁻*]αî⁻ ]] = [[ [uσ][σ]αi⁻ ]] = [[ [σ]αi⁻ ]]$,
-                \item if this variable is a metavariable $[[β̂±]] \in [[ uv(uP0) \ {â⁻*} ]] = [[uv(uP')]]$, then
+                \item if this variable is a algorithmic variable $[[β̂±]] \in [[ uv(uP0) \ {â⁻*} ]] = [[uv(uP')]]$, then
                 $[[ [uσ0]β̂± ]] = [[ [uσ]β̂± ]]$ and $[[ [uσ][σ][nas/â⁻*]β̂± ]] = [[ [uσ][σ]β̂± ]] = [[ [uσ]β̂± ]]$,
                 \item if this variable is a regular variable from $[[fv(uP0)]]$, both substitutions do not change it:
-                $[[ uσ0 ]]$, $[[ uσ ]]$ and $[[ nas / â⁻* ]]$ act on metavariables, 
+                $[[ uσ0 ]]$, $[[ uσ ]]$ and $[[ nas / â⁻* ]]$ act on algorithmic variables, 
                 and $[[σ]]$ is defined on $[[nas]]$, however, $[[{nas} ∩ fv(uP0) = ∅]]$.

### 2023-09-05 proofs/tex/anti-unification-lemmas.tex
Rearrange proofs

@@
                     \item $[[Γ ; Ξ ⊢ uQ]]$ and $[[Γ ; Ξ' ⊢ uM]]$. 
-                    Then by weakening (\cref{todo}), $[[Γ ; Ξ ∪ Ξ' ⊢ uQ]]$ and 
+                    Then by weakening (\cref{lemma:wf-weakening-algo}), 
+                    $[[Γ ; Ξ ∪ Ξ' ⊢ uQ]]$ and 
                     $[[Γ ; Ξ ∪ Ξ' ⊢ uM]]$, which implies $[[Γ ; Ξ ∪ Ξ' ⊢ uQ → uM]]$;
@@
 \begin{proof}
-    First, let us assume that $[[uM']]$ is a metavariable $[[α̂⁻]]$. 
+    First, let us assume that $[[uM']]$ is a algorithmic variable $[[α̂⁻]]$. 
     Then we can take $[[ausr]] = [[α̂⁻]] \mapsto [[uM]]$, which satisfies the required properties:
@@
 
-    Now, we can assume that $[[uM']]$ is not a metavariable. 
+    Now, we can assume that $[[uM']]$ is not a algorithmic variable. 
     We prove by induction on the derivation of $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
@@
 
-    Since $[[uM']]$ is not a metavariable, 
+    Since $[[uM']]$ is not a algorithmic variable, 
     the substitution acting on $[[uM']]$ preserves its outer constructor. 
@@
         Since $[[ [aus'i] uM' = iNi ]] = [[α⁻]]$
-        and $[[uM']]$ is not a metavariable, $[[uM']] = [[α⁻]]$.
+        and $[[uM']]$ is not a algorithmic variable, $[[uM']] = [[α⁻]]$.
         Then we can take $[[ausr]] = [[·]]$, which satisfies the required properties:
@@
 
-        Then since $[[ [aus'i] uM' = iNi ]] = [[↑iPi]]$ and $[[uM']]$ is not a metavariable,
+        Then since $[[ [aus'i] uM' = iNi ]] = [[↑iPi]]$ and $[[uM']]$ is not a algorithmic variable,
         $[[uM']] = [[↑uQ']]$, where $[[ [aus'i] uQ' = iPi ]]$. 
@@
 
-        Then since $[[ [aus'i] uM' = iNi ]] = [[iPi → iN'i]]$ and $[[uM']]$ is not a metavariable,
+        Then since $[[ [aus'i] uM' = iNi ]] = [[iPi → iN'i]]$ and $[[uM']]$ is not a algorithmic variable,
         $[[uM']] = [[uQ' → uM'']]$, where $[[ [aus'i] uQ' = iPi ]]$ and $[[ [aus'i] uM'' = iN''i ]]$.

### 2023-09-05 proofs/tex/equiv-alg-lemmas.tex
Rearrange proofs

--- 2023-09-04 proofs/tex/equiv-lemmas.tex
@@
-\begin{lemma} [Declarative Equivalence is invariant under bijections]
-  \label{lemma:decl-equiv-bijection}
-  Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
-  \begin{itemize}
-    \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
-      and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
-      with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
-    \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
-      and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
-      with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  We prove it by induction on $[[iP1 ≈ iP2]]$ and mutually, on $[[iN1 ≈ iN2]]$.
-  Let us consider the last rule used in the derivation.
-  \begin{caseof}
-    \item \ruleref{\ottdruleEOneForallLabel}\\
-      Then we decompose $[[iN1]]$ as $[[∀pas1.iM1]]$ and $[[iN2]]$ as $[[∀pas2.iM2]]$,
-      where $[[iM1]]$ and $[[iM2]]$ do not start with $[[∀]]$-quantifiers.
-      where $[[|pas1| + |pas2| > 0]]$.
-      By convention, let us assume that
-      $[[pas1]]$ and $[[pas2]]$ are disjoint form 
-      $[[varset2]]$ and $[[varset1]]$.
-
-      By inversion, $[[{pas1} ∩ fv iM2 = ∅]]$ and 
-      $[[iM1 ≈ [μ'] iM2]]$ for some bijection
-      $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$.
-      Then let us apply the induction hypothesis to
-       $[[iM1 ≈ [μ'] iM2]]$ to obtain $[[ [μ]iM1 ≈ [μ] [μ'] iM2 ]]$ 
-      inferred by the tree of the same shape as 
-      $[[ iM1 ≈ [μ'] iM2 ]]$.
-
-      Notice that $[[ [μ]iM1 ]]$ and $[[ [μ]iM2 ]]$ 
-      do not start with $[[∀]]$, 
-      That is $[[ [μ]∀pas1.iM1 ≈ [μ]∀pas2.iM2]]$, 
-      rewritten as $[[ ∀pas1.[μ]iM1 ≈ ∀pas2.[μ]iM2]]$,
-      can be inferred by \ruleref{\ottdruleEOneForallLabel}:
-      \begin{enumerate}
-        \item $[[{pas1}]]$ is disjoint from 
-        $[[varset2 ∪ fv iM2 ⊆ fv [μ]iM2 ]]$;
-        \item $[[ [μ]iM1 ≈ [μ'] [μ] iM2 ]]$
-          because $[[ [μ'] [μ] iM2 = [μ] [μ'] iM2 ]]$
-          (by \cref{corollary:subst-composition-commutativity}:
-          $[[μ' : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$, 
-          $[[μ : varset1 ↔ varset2]]$,
-          $[[varset1]]$ is disjoint from 
-          $[[pas2]]$ and $[[pas1]]$; 
-          $[[pas2]]$ is disjoint from $[[varset1]]$ and $[[varset2]]$)
-      \end{enumerate}
-      Notice that it is the same rule as the one inferring 
-      $[[iN1 ≈ iN2]]$, and thus, the shapes of the trees are the same.
-
-    \item \ruleref{\ottdruleEOneNVarLabel}\\
-      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and the required
-      $[[ [μ]α⁻ = [μ]α⁻ ]]$ is also inferred by
-      \ruleref{\ottdruleEOneNVarLabel}, since 
-      $[[ [μ]α⁻ ]]$ is a variable.
-    \item \ruleref{\ottdruleEOneArrowLabel}\\
-      Then we are proving 
-      that $[[iP1 → iM1 ≈ iP2 → iM2]]$ implies
-      $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$ 
-      (preserving the tree structure).
-
-      By inversion, we have $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$,
-      and thus, by the induction hypothesis,
-      $[[ [μ]iP1 ≈ [μ]iP2 ]]$ and $[[ [μ]iM1 ≈ [μ]iM2 ]]$.
-      Then $[[ [μ](iP1 → iM1) ≈ [μ](iP2 → iM2) ]]$, 
-      or in other words, $[[ [μ]iP1 → [μ]iM1 ≈ [μ]iP2 → [μ]iM2 ]]$,
-      is inferred by the same rule---\ruleref{\ottdruleEOneArrowLabel}.
-    \item \ruleref{\ottdruleEOneShiftULabel}
-      This case is done by similar congruent arguments as
-      the previous one.
-    \item The positive cases are proved symmetrically.
-    \end{caseof}
-\end{proof}
-
-\begin{lemma}[Declarative equivalence is transitive]
-  \hfill
-  \label{lemma:decl-equiv-transitivity}
-  \begin{itemize}
-    \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
-    \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  We prove it by
-  $\size{[[iP1 ≈ iP2]]} + \size{[[iP2 ≈ iP3]]}$
-  and mutually, $\size{[[iN1 ≈ iN2]]} + \size{[[iN2 ≈ iN3]]}$, where 
-  by size, we mean the size of the nodes in the corresponding inference tree.
-
-  \begin{caseof}
-    \item First, let us consider the case when either 
-      $[[iN1 ≈ iN2]]$ or $[[iN2 ≈ iN3]]$ is inferred by 
-      \ruleref{\ottdruleEOneForallLabel}.
-      Let us decompose $[[iN1]]$, $[[iN2]]$, and $[[iN3]]$ as follows:
-      $[[iN1 = ∀pas1.iM1]]$,
-      $[[iN2 = ∀pas2.iM2]]$, and
-      $[[iN3 = ∀pas3.iM3]]$.
-
-      Then by inversion of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
-      (or if $[[pas1]]$ and $[[pas2]]$ are both empty, 
-      by assumption) : 
-      \begin{enumerate}
-        \item $[[{pas1} ∩ fv iM2 = ∅]]$ and
-        \item there exists a bijection on variables 
-          $[[mu1 : ({pas2} ∩ fv iM2) ↔ ({pas1} ∩ fv iM1)]]$
-          such that $[[iM1 ≈ [mu1] iM2]]$.
-      \end{enumerate}
-
-      Analogously, $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
-      implies:
-      \begin{enumerate}
-        \item $[[{pas2} ∩ fv iM3 = ∅]]$ and
-        \item $[[iM2 ≈ [mu2] iM3]]$ for some bijection 
-          $[[mu2 : ({pas3} ∩ fv iM3) ↔ ({pas2} ∩ fv iM2)]]$.
-      \end{enumerate}
-
-      Notice that either $[[iM1 ≈ [mu1] iM2]]$
-      is inferred by a proper sub-tree of $[[∀pas1.iM1 ≈ ∀pas2.iM2]]$
-      or $[[iM2 ≈ [mu2] iM3]]$ is inferred by a proper sub-tree of 
-      $[[∀pas2.iM2 ≈ ∀pas3.iM3]]$.
-
-      Then by \cref{lemma:decl-equiv-bijection}, 
-      $[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]$ and moreover, 
-      $\size{[[ [mu1]iM2 ≈ [mu1 ○ mu2] iM3]]} = \size{[[iM2 ≈ [mu2] iM3]]}$.
-
-      Since at least one of the trees inferring 
-      $[[iM1 ≈ [mu1] iM2]]$ and $[[iM2 ≈ [mu2] iM3]]$
-      is a proper sub-tree of the corresponding original tree,
-      $\size{[[iM1 ≈ [mu1] iM2]]} + \size{[[iM2 ≈ [mu2] iM3]]} <
-      \size{[[∀pas1.iM1 ≈ ∀pas2.iM2]]} + \size{[[∀pas2.iM2 ≈ ∀pas3.iM3]]}$,
-      i.e., the induction hypothesis is applicable.
-
-      By the induction hypothesis, $[[iM1 ≈ [mu1 ○ mu2] iM3]]$.
-      Where $[[mu1 ○ mu2]]$ is a bijection on variables
-      $[[mu1 ○ mu2 : ({pas3} ∩ fv iM3) ↔ ({pas1} ∩ fv iM1)]]$.
-      Then $[[∀pas1.iM1 ≈ ∀pas3.iM3]]$ by \ruleref{\ottdruleEOneForallLabel}.
-
-      Once this case has been considered, 
-      we can assume that neither $[[iN1 ≈ iN2]]$ nor $[[iN2 ≈ iN3]]$
-      is inferred by \ruleref{\ottdruleEOneForallLabel}.
-
-    \item $[[iN1 ≈ iN2]]$ is inferred by 
-      \ruleref{\ottdruleEOneNVarLabel}\\
-      Then $[[iN1]] = [[iN2]] = [[α⁻]]$, and thus,  
-      $[[iN1 ≈ iN3]]$ holds since $[[iN2 ≈ iN3]]$.
-
-    \item $[[iN1 ≈ iN2]]$ is inferred by 
-      \ruleref{\ottdruleEOneArrowLabel}\\
-      Then $[[iN1]] = [[iP1 → iM1]]$ and $[[iN2]] = [[iP2 → iM2]]$,
-      and by inversion, $[[iP1 ≈ iP2]]$ and $[[iM1 ≈ iM2]]$.
-
-      Moreover, since $[[iN3]]$ does not start with $[[∀]]$,
-      $[[iN2 ≈ iN3]]$ is also inferred by 
-      \ruleref{\ottdruleEOneArrowLabel},
-      which means that $[[iN3]] = [[iP3 → iM3]]$,
-      $[[iP2 ≈ iP3]]$, and $[[iM2 ≈ iM3]]$.
-
-      Then by the induction hypothesis, 
-      $[[iP1 ≈ iP3]]$ and $[[iM1 ≈ iM3]]$, and thus, 
-      $[[iP1 → iM1 ≈ iP3 → iM3]]$ by \ruleref{\ottdruleEOneArrowLabel}.
-    \item $[[iN1 ≈ iN2]]$ is inferred by
-      \ruleref{\ottdruleEOneArrowLabel}\\
-      For this case, the reasoning is the same as for the previous one.
-    \item The positive cases are proved symmetrically.
-  \end{caseof}
-\end{proof}
-
 \begin{lemma}[Algorithmization of declarative equivalence]
@@
 
-\begin{lemma}[Type well-formedness is invariant under equivalence]
-  \label{lemma:wf-equiv}
-  Mutual subtyping implies declarative equivalence.
-  \begin{itemize}
-  \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
-  \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  We prove it by induction on $[[iP ≈ iQ]]$ and mutually, on $[[iN ≈ iM]]$.
-  Let us consider the last rule used in the derivation.
-  \begin{caseof}
-    \item \ruleref{\ottdruleEOneNVarLabel}, 
-      that is $[[iN ≈ iM]]$ has shape $[[α⁻ ≈ α⁻]]$.\\
-      Than $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ is rewritten as $[[Γ ⊢ α⁻]] \iff [[Γ ⊢ α⁻]]$,
-      which holds trivially.
-    \item \ruleref{\ottdruleEOneShiftULabel},
-      that is $[[iN ≈ iM]]$ has shape $[[↑iP ≈ ↑iQ]]$.\\
-      By inversion, $[[iP ≈ iQ]]$, and by the induction hypothesis,
-      $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$. 
-      Also notice that $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]]$ 
-      and $[[Γ ⊢ ↑iQ]] \iff [[Γ ⊢ iQ]]$ by inversion and \ruleref{\ottdruleWFTShiftULabel}.
-      This way, $[[Γ ⊢ ↑iP]] \iff [[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]] \iff [[Γ ⊢ ↑iQ]]$.
-    \item \ruleref{\ottdruleEOneArrowLabel},
-      that is $[[iN ≈ iM]]$ has shape $[[iP → iN' ≈ iQ → iM']]$.\\
-      Then by inversion, $[[iP ≈ iQ]]$ and $[[iN' ≈ iM']]$, 
-      and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
-      $
-      \begin{aligned}[t]
-        [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
-                          && \text{by inversion and \ruleref{\ottdruleWFTArrowLabel}}\\
-                          &\iff [[Γ ⊢ iQ]] \text{ and } [[Γ ⊢ iM']]
-                          && \text{as noted above}\\
-                          &\iff [[Γ ⊢ iQ → iM']]
-                          && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
-      \end{aligned}
-      $\\
-    \item \ruleref{\ottdruleEOneForallLabel},
-      that is $[[iN ≈ iM]]$ has shape $[[∀pas.iN' ≈ ∀pbs.iM']]$.\\
-      By inversion, $[[∀pas.iN' ≈ ∀pbs.iM']]$ means that
-      $[[{pas} ∩ fv iM = ∅]]$ and 
-      that there exists a bijection on variables 
-      $[[mu : ({pbs} ∩ fv iM') ↔ ({pas} ∩ fv iN')]]$
-      such that $[[iN' ≈ [mu] iM']]$.
-
-      By inversion and \ruleref{\ottdruleWFTForallLabel},
-      $[[Γ ⊢ ∀pas.iN']]$ is equivalent to $[[Γ, pas ⊢ iN']]$,
-      and by \cref{lemma:wf-ctxt-equiv}, it is equivalent to 
-      $[[Γ,  ({pas} ∩ fv iN') ⊢ iN']]$,
-      which, by the induction hypothesis, is equivalent to
-      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$.
-
-      Analogously, $[[Γ ⊢ ∀pbs.iM']]$ is equivalent to
-      $[[Γ, ( {pbs} ∩ fv iM') ⊢ iM']]$.
-      By \cref{lemma:wf-subst}, it implies 
-      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$. And vice versa, 
-      $[[Γ,  ({pas} ∩ fv iM') ⊢ [mu]iM']]$ implies  
-      $[[Γ,  ({pbs} ∩ fv iM') ⊢ [mu-1][mu]iM']]$.
-
-      This way, both $[[Γ ⊢ ∀pas.iN']]$ and $[[Γ ⊢ ∀pbs.iM']]$ are equivalent to
-      $[[Γ,  ({pas} ∩ fv iN') ⊢ [mu]iM']]$. 
-    \item For the cases of the positive types, the proofs are symmetric.
-  \end{caseof}
-\end{proof}
-
 \begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
@@
   The proof coincides with the proof of \cref{lemma:wf-equiv},
-  and adds two cases for equating two positive or two negative metavariables,
+  and adds two cases for equating two positive or two negative algorithmic variables,
   which must be equal by inversion, and thus, 
   $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
-\end{proof}
-
-\begin{corollary}[Normalization preserves well-formedness]
-  \label{corollary:wf-nf}
-  \hfill
-  \begin{itemize}
-  \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
-  \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
-\end{proof}
-
-\begin{lemma}[Normalization preserves substitution signature]
-  \label{lemma:norm-subst-sig}
-  Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
-  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
-\end{lemma}
-\begin{proof}
-  Suppose that $[[α± ∊ Γ1]]$. 
-  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
-  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.
-
-  Suppose that $[[α± ∉ Γ1]]$. 
-  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
-  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
 \end{proof}
@@
 
-\begin{corollary}[Normalization preserves well-formedness of substitution]
-  \label{corollary:wf-s-nf}
-  \hfill \\
-   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
-\end{corollary}
-\begin{proof}
-  Let us prove the forward direction.
-  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
-  By the definition of substitution normalization,
-  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf-algo},
-  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
-  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
-
-  The backward direction is proved analogously.
-\end{proof}
-
 \begin{corollary}[Normalization preserves the signature of the algorithmic substitution]
@@
 
-\begin{lemma}[Soundness of equivalence]
-  \label{lemma:equiv-soundness}
-  Declarative equivalence implies mutual subtyping.
-  \begin{itemize}
-    \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
-    \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  We prove it by mutual induction on $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
-  \begin{caseof}
-    \item $[[a⁻ ≈ a⁻]]$\\
-      Then $[[Γ ⊢ a⁻ ≤ a⁻]]$ by \ruleref{\ottdruleDOneNVarLabel},
-      which immediately implies $[[Γ ⊢ a⁻ ≈ a⁻]]$ by \ruleref{\ottdruleDOneNDefLabel}.
-
-    \item $[[↑iP ≈ ↑iQ]]$\\
-      Then by inversion of \ruleref{\ottdruleDOneShiftULabel},
-      $[[iP ≈ iQ]]$, and from the induction hypothesis, $[[Γ ⊢ iP ≈ iQ]]$,
-      and (by symmetry) $[[Γ ⊢ iQ ≈ iP]]$.
-
-      When \ruleref{\ottdruleDOneShiftULabel} is applied to $[[Γ ⊢ iP ≈ iQ]]$,
-      it gives us $[[Γ ⊢ ↑iP ≤ ↑iQ]]$; when it is applied to $[[Γ ⊢ iQ ≈ iP]]$,
-      we obtain $[[Γ ⊢ ↑iQ ≤ ↑iP]]$. Together, it implies $[[Γ ⊢ ↑iP ≈ ↑iQ]]$.
-
-    \item $[[iP → iN ≈ iQ → iM]]$\\
-      Then by inversion of \ruleref{\ottdruleDOneArrowLabel},
-      $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$. By the induction hypothesis,
-      $[[Γ ⊢ iP ≈ iQ]]$ and $[[Γ ⊢ iN ≈ iM]]$, which means by inversion:
-      \begin{enumerate*}
-        \item[(i)] $[[Γ ⊢ iP ≥ iQ]]$,
-        \item[(ii)] $[[Γ ⊢ iQ ≥ iP]]$,
-        \item[(iii)] $[[Γ ⊢ iN ≤ iM]]$,
-        \item[(iv)]  $[[Γ ⊢ iM ≤ iN]]$.
-      \end{enumerate*}
-      Applying \ruleref{\ottdruleDOneArrowLabel} to (i) and (iii), we obtain
-      $[[Γ ⊢ iP → iN ≤ iQ → iM]]$; applying it to (ii) and (iv), we have $[[Γ ⊢
-      iQ → iM ≤ iP → iN]]$. Together, it implies $[[Γ ⊢ iP → iN ≈ iQ → iM]]$.
-    \item $[[∀pas.iN ≈ ∀pbs.iM]]$\\
-      Then by inversion, there exists bijection $[[mu : ({pbs} ∩ fv iM) ↔ ({pas}
-      ∩ fv iN)]]$, such that $[[iN ≈ [mu] iM]]$. By the induction hypothesis,
-      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$. From \cref{corollary:subst-pres-equiv} and
-      the fact that $[[mu]]$ is bijective, we also have
-      $[[Γ, pbs ⊢ [mu-1]iN ≈ iM]]$.
-
-      Let us construct a substitution $[[{pas} ⊢ iPs/pbs : {pbs}]]$ by
-      extending $[[mu]]$ with arbitrary positive types on $[[{pbs} \ fv iM]]$.
-
-      Notice that $[[ [mu]iM ]] = [[ [iPs/pbs]iM ]]$, and therefore,
-      $[[Γ, pas ⊢ iN ≈ [mu] iM]]$ implies $[[Γ, pas ⊢ [iPs/pbs]iM ≤ iN]]$. Then by
-      \ruleref{\ottdruleDOneForallLabel}, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$.
-
-      Analogously, we construct the substitution from $[[mu-1]]$, and use it to
-      instantiate $[[pas]]$ in the application of
-      \ruleref{\ottdruleDOneForallLabel} to infer $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$.
-
-      This way, $[[Γ ⊢ ∀pbs.iM ≤ ∀pas.iN]]$ and $[[Γ ⊢ ∀pas.iN ≤ ∀pbs.iM]]$
-      gives us $[[Γ ⊢ ∀pbs.iM ≈ ∀pas.iN]]$.
-
-    \item For the cases of the positive types, the proofs are symmetric.
-  \end{caseof}
-\end{proof}
-
-\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] 
-  \label{corollary:nf-sound-wrt-subt-equiv}
-  \hfill
-  \begin{itemize}
-    \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
-    \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
-\end{proof}
-
-\begin{corollary}[Normalization preserves subtyping] 
-  \label{corollary:nf-pres-subt}
-  Assuming all the types are well-formed in context $[[Γ]]$,
-  \begin{itemize}
-    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
-    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  \hfill
-  \begin{itemize}
-    \item [$+$]  
-    \begin{itemize}
-      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
-        By \cref{corollary:nf-sound-wrt-subt-equiv},
-        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
-        in particular, by inversion, 
-        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
-        Then by the transitivity of subtyping 
-        (\cref{corollary:subtyping-transitivity}), 
-        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
-      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
-        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
-        and inversion, 
-        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
-        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
-    \end{itemize}
-    \item [$-$] The negative case is proved symmetrically.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Subtyping induced by disjoint substitutions]
-  \label{lemma:subt-ind-disj-subst}
-  If two disjoint substitutions induce subtyping, they are degenerate (so is the
-  subtyping).
-  Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
-  where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
-  \begin{itemize}
-  \item[$-$] assuming $[[Γ ⊢ iN]]$,~
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id :  fv iN]]$
-  \item[$+$] assuming $[[Γ ⊢ iP]]$,~
-    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id :  fv iP]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Proof by induciton on $[[Γ ⊢ iN]]$ (and mutually on $[[Γ ⊢ iP]]$).
-  \begin{caseof}
-    \item $[[iN]] = [[α⁻]]$\\
-      Then $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]α⁻ ≤ [σ2]α⁻]]$.
-      Let us consider the following cases:
-      \begin{caseof}
-      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∉ Γ2]]$ \label{case:var-not-in-ctxts}\\
-        Then $[[Γ ⊢ σi ≈ id : {α⁻}]]$ holds immediately,
-        since $[[ [σi] α⁻]] = [[ [id] α⁻]] = [[α⁻]]$ and
-        $[[Γ ⊢ α⁻ ≈ α⁻]]$.
-      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
-        This case is not possible by assumption: $[[Γ1 ∩ Γ2= ∅]]$.
-      \item $[[α⁻ ∊ Γ1]]$ and $[[α⁻ ∉ Γ2]]$\\
-        Then we have $[[Γ ⊢ [σ1]α⁻ ≤ α⁻]]$,
-        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ [σ1]α⁻ ≈ α⁻]]$,
-        and hence, $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$.
-
-        $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$ holds since $[[ [σ2]α⁻ ]] = [[α⁻]]$,
-        similarly to \cref{case:var-not-in-ctxts}.
-
-      \item $[[α⁻ ∉ Γ1]]$ and $[[α⁻ ∊ Γ2]]$\\
-        Then we have $[[Γ ⊢ α⁻ ≤ [σ2]α⁻]]$,
-        which by \cref{corollary:vars-no-proper-subtypes} means $[[Γ ⊢ α⁻ ≈ [σ2]α⁻]]$,
-        and hence, $[[Γ ⊢ σ2 ≈ id : {α⁻}]]$.
-
-        $[[Γ ⊢ σ1 ≈ id : {α⁻}]]$ holds since $[[ [σ1]α⁻ ]] = [[α⁻]]$,
-        similarly to \cref{case:var-not-in-ctxts}.
-      \end{caseof}
-  \item $[[iN]] = [[∀pas.iM]]$\\
-    Then by inversion, $[[Γ, pas ⊢ iM]]$.
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as $[[Γ ⊢ [σ1]∀pas.iM ≤ [σ2]∀pas.iM]]$.
-    By the congruence of substitution and by the inversion of
-    \ruleref{\ottdruleDOneForallLabel}, $[[Γ, pas ⊢ [iQs/pas][σ1]iM ≤ [σ2]iM]]$,
-    where $[[Γ, pas ⊢ iQi]]$.
-    Let us denote the (Kleisli) composition of $[[σ1]]$ and $[[iQs/pas]]$ as
-    $[[σ1']]$, noting that $[[Γ, pas ⊢ σ1' : Γ1, pas]]$,
-    and $[[(Γ1, pas) ∩ Γ2 = ∅]]$.
-
-    Let us apply the induction hypothesis to $[[iM]]$ and the
-    substitutions $[[σ1']]$ and $[[σ2]]$ with
-    $[[Γ, pas ⊢ [σ1']iM ≤ [σ2]iM]]$ to obtain:
-    \begin{align}
-      [[Γ, pas ⊢ σ1' ≈ id :  fv iM]] \label{fact:subs-proper-sub:forall-ih}\\
-      [[Γ, pas ⊢ σ2 ≈ id :  fv iM]]  \label{fact:subs-proper-sub:forall-ih2}
-    \end{align}
-
-    Then $[[Γ ⊢ σ2 ≈ id :  fv ∀pas.iM]]$ holds by strengthening of
-    \ref{fact:subs-proper-sub:forall-ih2}:
-    for any $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$,
-    $[[Γ, pas ⊢ [σ2]β± ≈ β±]]$ is strengthened to $[[Γ ⊢ [σ2]β± ≈ β±]]$, because
-    $[[fv [σ2]β±]] = [[fv β±]] = \{[[β±]]\} \subseteq [[Γ]]$.
-
-    To show that $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$, let us take an arbitrary
-    $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.
-
-    $
-    \begin{aligned}[t]
-      [[β±]] &= [[ [id]β± ]]
-             && \text{by definition of $[[id]]$}\\
-             &\eqDOne [[ [σ1']β± ]]
-             && \text{by \ref{fact:subs-proper-sub:forall-ih}}\\
-             &= [[ [iQs/pas][σ1]β±]]
-             && \text{by definition of $[[σ1']]$}\\
-             &= [[ [σ1]β± ]]
-             && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
-    \end{aligned}
-    $\\
-    This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,
-    $[[Γ ⊢ σ1 ≈ id :  fv ∀pas.iM]]$.
-
-  \item $[[iN]] = [[iP → iM]]$\\
-    Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as
-    $[[Γ ⊢ [σ1](iP → iM) ≤ [σ2](iP → iM)]]$,
-    then by congruence of substitution,
-    $[[Γ ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$,
-    then by inversion
-    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
-    and
-    $[[Γ ⊢ [σ1]iM ≤ [σ2]iM]]$.
-
-    Applying the induction hypothesis to $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
-    and to $[[Γ ⊢ [σ1]iM ≤ [σ2]iM]]$, we obtain (respectively):
-    \begin{align}
-      &[[Γ ⊢ σi ≈ id :  fv iP]] \label{fact:subs-proper-sub:arrow-ih1}\\
-      &[[Γ ⊢ σi ≈ id :  fv iM]] \label{fact:subs-proper-sub:arrow-ih2}
-    \end{align}
-
-    Noting that $[[fv (iP → iM)]] = [[fv iP ∪ fv iM]]$,
-    we combine
-    \cref{fact:subs-proper-sub:arrow-ih1,fact:subs-proper-sub:arrow-ih2}
-    to conclude:
-    $[[Γ ⊢ σi ≈ id :  fv (iP → iM)]]$.
-
-  \item $[[iN]] = [[↑iP]]$\\
-    Then by inversion, $[[Γ ⊢ iP]]$.
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ is rewritten as
-    $[[Γ ⊢ [σ1]↑iP ≤ [σ2]↑iP]]$,
-    then by congruence of substitution and by inversion,
-    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$
-
-    Applying the induction hypothesis to $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$, we obtain
-    $[[Γ ⊢ σi ≈ id :  fv iP]]$. Since $[[fv ↑iP]] = [[fv iP]]$, we can
-    conclude: $[[Γ ⊢ σi ≈ id :  fv ↑iP]]$.
-  \item The positive cases are proved symmetrically.
-  \end{caseof}
-\end{proof}
-
-\begin{corollary}[Substitution cannot induce proper subtypes or supertypes] \label{corollary:subst-proper-subt}
-  Assuming all mentioned types are well-formed in $[[Γ]]$ and $[[σ]]$ is a
-  substitution $[[Γ ⊢ σ : Γ]]$,
-  \begin{align*}
-    [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
-    [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
-    [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
-    [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
-  \end{align*}
-\end{corollary}
-
-\begin{lemma} \label{lemma:mutual-subst-subtyping}
-  Asssuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
-  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
-  \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
-    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
-  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
-    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  \hfill
-  \begin{itemize}
-  \item[$+$]
-    Applying $[[σ2]]$ to both sides of
-    $[[Γ ⊢ [σ1] iP ≥ iQ]]$ (by \cref{lemma:subst-pres-subt}),
-    we have: $[[Γ ⊢ [σ2 ○ σ1] iP ≥ [σ2]iQ]]$.
-    Composing it with $[[Γ ⊢ [σ2] iQ ≥ iP]]$ by transitivity 
-    (\cref{lemma:subtyping-transitivity}),
-    we have $[[Γ ⊢ [σ2 ○ σ1] iP ≥ iP]]$.
-    Then by \cref{corollary:subst-proper-subt},
-    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$.
-    By a symmetric argument, we also have:
-    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$.
-
-    Now, we prove that
-    $[[Γ ⊢ σ2 ○ σ1 ≈ id :  fv iP]]$ and
-    $[[Γ ⊢ σ1 ○ σ2 ≈ id :  fv iQ]]$
-    implies that $[[σ1]]$ and $[[σ1]]$
-    are (equivalent to) mutually inverse bijections.
-
-    To do so, it suffices to prove that
-    \begin{enumerate}
-    \item[(i)] for any $[[α± ∊ fv iP]]$ there exists $[[β± ∊ fv iQ]]$
-        such that $[[ Γ ⊢ [σ1] α± ≈ β± ]]$ and
-        $[[ Γ ⊢ [σ2] β± ≈ α± ]]$; and
-    \item[(ii)] for any $[[β± ∊ fv iQ]]$ there exists $[[α± ∊ fv iP]]$
-        such that $[[ Γ ⊢ [σ2] β± ≈ α± ]]$ and
-        $[[ Γ ⊢ [σ1] α± ≈ β± ]]$.
-    \end{enumerate}
-    Then these correspondences between $[[fv iP]]$ and
-    $[[fv iQ]]$ are mutually inverse functions,
-    since for any $[[β±]]$ there can be at most one $[[α±]]$
-    such that $[[ Γ ⊢ [σ2] β± ≈ α± ]]$ (and vice versa).
-
-    \begin{enumerate}
-    \item[(i)] Let us take $[[α± ∊ fv iP]]$.
-      \begin{enumerate}
-      \item if $[[α±]]$ is positive ($[[α± = α⁺]]$),
-        from $[[ Γ ⊢ [σ2][σ1]α⁺ ≈ α⁺ ]]$,
-        by \cref{corollary:vars-no-proper-subtypes},
-        we have
-        $[[ [σ2][σ1]α⁺ = ∃nbs.α⁺ ]]$.
-
-        What shape can $[[ [σ1]α⁺ ]]$ have? It cannot be $[[∃nas.↓iN]]$ (for
-        potentially empty $[[nas]]$), because the outer constructor $\downarrow$
-        would remain after the substitution $[[σ2]]$, whereas $[[∃nbs.α⁺]]$ does
-        not have $[[↓]]$. The only case left is $[[ [σ1]α⁺ = ∃nas.γ⁺ ]]$.
-
-        Notice that $[[Γ ⊢ ∃nas.γ⁺ ≈ γ⁺]]$, meaning that $[[Γ ⊢ [σ1]α⁺ ≈ γ⁺]]$.
-        Also notice that $[[ [σ2]∃nas.γ⁺ = ∃nbs.α⁺ ]]$ implies
-        $[[Γ ⊢ [σ2]γ⁺ ≈ α⁺]]$.
-
-      \item if $[[α±]]$ is negative ($[[α± = α⁻]]$) from $[[ Γ ⊢ [σ2][σ1]α⁻ ≈ α⁻
-        ]]$, by \cref{corollary:vars-no-proper-subtypes}, we have
-        $[[ [σ2][σ1]α⁻ = ∀pbs.α⁻ ]]$.
-
-        What shape can $[[ [σ1]α⁻ ]]$ have? It cannot be $[[∀pas.↑iP]]$
-        nor $[[∀pas.iP → iM]]$ (for potentially empty $[[pas]]$),
-        because the outer constructor ($[[→]]$ or $[[↑]]$), remaining
-        after the substitution $[[σ2]]$, is however absent in the resulting
-        $[[∀pbs.α⁻]]$. Hence, the only case left is $[[ [σ1]α⁻ = ∀pas.γ⁻ ]]$
-        Notice that $[[Γ ⊢ γ⁻ ≈ ∀pas.γ⁻]]$, meaning that $[[Γ ⊢ [σ1]α⁻ ≈ γ⁻]]$.
-        Also notice that $[[ [σ2]∀pas.γ⁻ = ∀pbs.α⁻ ]]$ implies
-        $[[Γ ⊢ [σ2]γ⁻ ≈ α⁻]]$.
-      \end{enumerate}
-    \item[(ii)] The proof is symmetric:
-      We swap $[[iP]]$ and $[[iQ]]$,
-      $[[σ1]]$ and $[[σ2]]$,
-      and exploit $[[ Γ ⊢ [σ1][σ2]α± ≈ α± ]]$ instead of
-      $[[ Γ ⊢ [σ2][σ1]α± ≈ α± ]]$.
-
-    \end{enumerate}
-
-  \item[$-$] The proof is symmetric to the positive case.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Equivalent substitution act equivalently]
-  \label{lemma:equiv-subst-on-same-term}
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$ and $[[Γ' ⊢ σ2 : Γ]]$
-  are substitutions equivalent on their domain, that is $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
-  Then
-  \begin{itemize}
-    \item[$+$] for any $[[Γ ⊢ iP]]$, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$;
-    \item[$-$] for any $[[Γ ⊢ iN]]$, $[[Γ' ⊢ [σ1]iN ≈ [σ2]iN]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  We prove it by induction on $[[iP]]$ (and mutually on $[[iN]]$).
-  \begin{caseof}
-    \item $[[iN = α⁻]]$\\
-      Then since by inversion, $[[α⁻ ∊ Γ]]$, 
-      $[[Γ' ⊢ [σ1]α⁻ ≈ [σ2]α⁻]]$ holds by definition of $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
-    \item $[[iN = ↑iP]]$\\
-      Then by inversion, $[[Γ ⊢ iP]]$.
-      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$, 
-      Then by \ruleref{\ottdruleDOneShiftULabel}, 
-      $[[Γ' ⊢ ↑[σ1]iP ≤ ↑[σ2]iP]]$, and symmetrically, $[[Γ' ⊢ ↑[σ2]iP ≤ ↑[σ1]iP]]$,
-      together meaning that $[[Γ' ⊢ ↑[σ1]iP ≈ ↑[σ2]iP]]$,
-      or equivalently, $[[Γ' ⊢ [σ1]↑iP ≈ [σ2]↑iP]]$.
-    \item $[[iN = iP → iM]]$\\
-      Then by inversion, $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM]]$.
-      By the induction hypothesis, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$ and
-      $[[Γ' ⊢ [σ1]iM ≈ [σ2]iM]]$,
-      that is $[[Γ' ⊢ [σ1]iP ≥ [σ2]iP]]$, $[[Γ' ⊢ [σ2]iP ≥ [σ1]iP]]$,
-      $[[Γ' ⊢ [σ1]iM ≤ [σ2]iM]]$, and $[[Γ' ⊢ [σ2]iM ≤ [σ1]iM]]$.
-      Then by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≤ [σ2]iP → [σ2]iM]]$, 
-      and again by \ruleref{\ottdruleDOneArrowLabel}, $[[Γ' ⊢ [σ2]iP → [σ2]iM ≤ [σ1]iP → [σ1]iM]]$.
-      This way, $[[Γ' ⊢ [σ1]iP → [σ1]iM ≈ [σ2]iP → [σ2]iM]]$,
-      or equivalently, $[[Γ' ⊢ [σ1](iP → iM) ≈ [σ2](iP → iM)]]$.
-    \item $[[iN = ∀pas.iM]]$
-      We can assume that $[[pas]]$ is disjoint from $[[Γ]]$ and $[[Γ']]$.
-      By inversion, $[[Γ ⊢ ∀pas.iM]]$ implies $[[Γ, pas ⊢ iM]]$.
-      Notice that $[[Γ' ⊢ σi : Γ]]$ and 
-      $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$
-      can be extended to 
-      $[[Γ', pas ⊢ σi : Γ, pas]]$
-      and 
-      $[[Γ', pas ⊢ σ1 ≈ σ2 : Γ, pas]]$
-      by \cref{lemma:subst-domain-weakening}.
-      Then by the induction hypothesis,
-      $[[Γ', pas ⊢ [σ1]iM ≈ [σ2]iM]]$,
-      meaning by inversion that
-      $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM]]$ and 
-      $[[Γ', pas ⊢ [σ2]iM ≤ [σ1]iM]]$.
-
-      To infer
-      $[[Γ' ⊢ ∀pas.[σ1]iM ≤ ∀pas.[σ2]iM]]$,
-      we apply \ruleref{\ottdruleDOneForallLabel}
-      with the substitution $[[Γ', pas ⊢ id :{pas}]]$, 
-      noting that $[[Γ', pas ⊢ [id][σ1]iM ≤ [σ2]iM ]]$
-      holds since $[[Γ', pas ⊢ [σ1]iM ≤ [σ2]iM ]]$, 
-      as noted above.
-
-      Symmetrically, we infer
-      $[[Γ' ⊢ ∀pas.[σ2]iM ≤ ∀pas.[σ1]iM]]$,
-      which together with
-      $[[Γ' ⊢ ∀pas.[σ1]iM ≤ ∀pas.[σ2]iM]]$
-      means
-      $[[Γ' ⊢ ∀pas.[σ1]iM ≈ ∀pas.[σ2]iM]]$,
-      or equivalently, 
-      $[[Γ' ⊢ [σ1]∀pas.iM ≈ [σ2]∀pas.iM]]$.
-
-    \item The positive cases are proved symmetrically.
-  \end{caseof}
-\end{proof}
-
-\begin{lemma}[Equivalence of polymorphic types]
-  \label{lemma:poly-types-equivalence}
-  \hfill
-  \begin{itemize}
-    \item[$-$] For $[[Γ ⊢ ∀pas.iN]]$ and $[[Γ ⊢ ∀pbs.iM]]$,\\ if $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$
-    then there exists a bijection $[[μ : {pbs} ∩ fv iM ↔ {pas} ∩ fv iN]]$
-    such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iM ]]$,
-    \item[$+$] For $[[Γ ⊢ ∃nas.iP]]$ and $[[Γ ⊢ ∃nbs.iQ]]$,\\  if $[[Γ ⊢ ∃nas.iP ≈ ∃nbs.iQ ]]$
-    then there exists a bijection $[[μ : {nbs} ∩ fv iQ ↔ {nas} ∩ fv iP]]$
-    such that $[[ Γ, nbs ⊢ iP ≈ [Sub μ] iQ ]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-    \hfill
-  \begin{itemize}
-    \item[$-$]
-      First, by $\alpha$-conversion, we ensure $[[{pas} ∩ fv iM = ∅]]$ and $[[{pbs} ∩ fv iN = ∅]]$.
-      By inversion, $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$ implies 
-      \begin{enumerate} 
-        \item $[[Γ,pbs ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ,pbs ⊢ σ1 :{pas} ]]$ and 
-        \item $[[Γ,pas ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ,pas ⊢ σ2 : {pbs} ]]$.
-      \end{enumerate}
-      To apply \cref{lemma:mutual-subst-subtyping}, we weaken 
-      and rearrange the contexts, and extend the substitutions to act as identity
-      outside of their initial domain:
-      \begin{enumerate} 
-        \item $[[Γ, {pas}, {pbs} ⊢ [σ1]iN ≤ iM]]$ for $[[ Γ, {pas}, {pbs} ⊢ σ1 : Γ, {pas}, {pbs} ]]$ and 
-        \item $[[Γ, {pas}, {pbs} ⊢ [σ2]iM ≤ iN]]$ for $[[ Γ, {pas}, {pbs} ⊢ σ2 : Γ, {pas}, {pbs} ]]$.
-      \end{enumerate}
-      Then from \cref{lemma:mutual-subst-subtyping}, 
-      there exists a bijection $[[μ : fv iM ↔ fv iN]]$ such that 
-      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ :  fv iM]]$ and 
-      $[[Γ, {pas}, {pbs} ⊢ σ1 ≈ Sub μ-1 :  fv iN]]$.
-
-      Let us show that $[[Sub μ|{pbs}]]$ is the appropriate candidate.
-
-      First, we show that if we restrict the domain of $[[μ]]$ to 
-      $[[pbs]]$, its range will be contained in $[[pas]]$.
-
-      Let us take $[[γ⁺ ∊ {pbs} ∩ fv iM]]$ and 
-      assume $[[ [μ]γ⁺]] \notin [[pas]]$.
-      Then since $[[ Γ,pbs ⊢ σ1 :{pas} ]]$, 
-      $[[σ1]]$ acts as identity outside of $[[pas]]$, i.e.
-      $[[ [σ1][Sub μ]γ⁺ = [Sub μ]γ⁺ ]]$ (notice that $[[γ⁺]]$ is in the domain of $[[μ]]$).
-      Since
-      $[[Γ, {pas}, {pbs} ⊢ σ1 ≈ Sub μ-1 :  fv iN]]$, 
-      application of $[[σ1]]$ to $[[ [μ]γ⁺ ]] \in [[fv iN]]$
-      is equivalent to application of $[[Sub μ-1]]$, then 
-      $[[ Γ, {pas}, {pbs} ⊢ [Sub μ-1][Sub μ]γ⁺ ≈ [Sub μ]γ⁺ ]]$, i.e.
-      $[[Γ, {pas}, {pbs} ⊢ γ⁺ ≈ [Sub μ]γ⁺]]$, 
-      which means $[[γ⁺ ∊ fv [Sub μ]γ⁺]] \subseteq [[fv iN]]$.
-      By assumption, $[[γ⁺ ∊ {pbs} ∩ fv iM]]$, i.e. $[[{pbs} ∩ fv iN]] \neq \emptyset$, hence contradiction.
-
-      Second, 
-      we will show $[[ Γ, pas ⊢ iN ≈ [Sub μ|{pbs}] iM ]]$.
-
-      Since $[[ Γ,pas ⊢ σ2 : {pbs} ]]$ and 
-      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ :  fv iM]]$, 
-      we have
-      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ|{pbs} :  fv iM]]$:
-      for any $[[α± ∊ fv iM \ {pbs}]]$, $[[ [σ2]α± = α± ]]$
-      since $[[ Γ,pas ⊢ σ2 : {pbs} ]]$, and
-      $[[ [Sub μ|{pbs}]α± = α± ]]$ by definition of substitution restriction;
-      for $[[β⁺ ∊ {pbs}]]$, $[[ [Sub μ|{pbs}]β⁺ = [μ]β⁺ ]]$,
-      and thus, $[[Γ, {pas}, {pbs} ⊢ [σ2]β⁺ ≈ [Sub μ]β⁺ ]]$
-      can be rewritten to $[[Γ, {pas}, {pbs} ⊢ [σ2]β⁺ ≈ [Sub μ|{pbs}]β⁺ ]]$.
-
-      By \cref{lemma:equiv-subst-on-same-term},
-      $[[Γ, {pas}, {pbs} ⊢ σ2 ≈ Sub μ|{pbs} :  fv iM]]$ implies
-      $[[Γ, {pas}, {pbs} ⊢ [σ2]iM ≈ [Sub μ|{pbs}]iM]]$.
-      By similar reasoning, $[[Γ, {pas}, {pbs} ⊢ [σ1]iN ≈ [Sub μ-1|{pas}]iN]]$.
-
-      This way, by transitivity of subtyping (\cref{lemma:subtyping-transitivity}),
-      \begin{align} 
-        [[Γ, {pas}, {pbs} ⊢ [Sub μ-1|{pas}]iN ≤ iM]] \label{fact:mu-inv-n-sub-m}\\
-        [[Γ, {pas}, {pbs} ⊢ [Sub μ|{pbs}]iM ≤ iN]] \label{fact:mu-m-subt-n}
-      \end{align}
-
-      By applying $[[Sub μ|{pbs}]]$ to both sides of \ref{fact:mu-inv-n-sub-m} 
-      (\cref{lemma:subst-pres-subt}), we have
-      $[[Γ, {pas}, {pbs} ⊢ [Sub μ|{pbs}][Sub μ-1|{pas}]iN ≤ [Sub μ|{pbs}]iM]]$. 
-      By contracting $[[μ-1|_{pas} ○ μ|_{pbs}]] = [[μ|_{pbs}-1 ○ μ|_{pbs}]]$
-      (notice that $[[fv iN ∩ {pbs} = ∅]]$), 
-      we have $[[Γ, {pas}, {pbs} ⊢ iN ≤ [Sub μ|{pbs}]iM]]$, which together with \ref{fact:mu-m-subt-n}
-      means $[[Γ, {pas}, {pbs} ⊢ iN ≈ [Sub μ|{pbs}]iM]]$, and by strengthening, $[[Γ,pas⊢ iN ≈ [Sub μ|{pbs}]iM]]$.
-      % Symmetrically, $[[Γ,pbs ⊢ iM ≈ [Sub μ|_{pbs}-1]iN]]$.
-    \item[$+$] The proof is symmetric to the proof of the negative case.
-  \end{itemize}
-
-\end{proof}
-
-\begin{lemma}[Completeness of equivalence] \label{lemma:equiv-completeness}
-  Mutual subtyping implies declarative equivalence.
-  Assuming all the types below are well-formed in $[[Γ]]$: 
-  \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[iP ≈ iQ]]$,
-  \item[$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[iN ≈ iM]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  \begin{itemize}
-    \item[$-$] 
-    Induction on the sum of sizes of  $[[iN]]$ and $[[iM]]$. 
-    By inversion, $[[Γ ⊢ iN ≈ iM]]$ means $[[Γ ⊢ iN ≤ iM]]$ and $[[Γ ⊢ iM ≤ iN ]]$.
-    Let us consider the last rule that forms $[[Γ ⊢ iN ≤ iM]]$:
-    \begin{caseof}
-      \item \ruleref{\ottdruleDOneNVarLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ α⁻ ≤ α⁻]]$\\
-      Then $[[iN ≈ iM]]$ (i.e. $[[α⁻ ≈ α⁻]]$) holds immediately by \ruleref{\ottdruleEOneNVarLabel}.
-
-      \item \ruleref{\ottdruleDOneShiftULabel} i.e. 
-      $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ ↑iP ≤ ↑iQ]]$\\
-      Then by inversion, $[[Γ ⊢ iP ≈ iQ]]$, 
-      and by induction hypothesis, $[[iP ≈ iQ]]$.
-      Then $[[iN ≈ iM]]$ (i.e. $[[↑iP ≈ ↑iQ]]$) holds 
-      by \ruleref{\ottdruleEOneShiftULabel}.
-
-      \item \ruleref{\ottdruleDOneArrowLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ iP → iN' ≤ iQ → iM']]$\\
-      Then by inversion, $[[Γ ⊢ iP ≥ iQ]]$ and $[[Γ ⊢ iN' ≤ iM']]$.
-      Notice that $[[Γ ⊢ iM ≤ iN]]$ is of the form $[[Γ ⊢ iQ → iM' ≤ iP → iN']]$, 
-      which by inversion means $[[Γ ⊢ iQ ≥ iP]]$ and $[[Γ ⊢ iM' ≤ iN']]$.
-
-      This way, $[[Γ ⊢ iQ ≈ iP]]$ and $[[Γ ⊢ iM' ≈ iN']]$. 
-      Then by induction hypothesis, $[[iQ ≈ iP]]$ and $[[iM' ≈ iN']]$.
-      Then $[[iN ≈ iM]]$ (i.e. $[[iP → iN' ≈ iQ → iM']]$) holds by \ruleref{\ottdruleEOneArrowLabel}.
-
-      \item \ruleref{\ottdruleDOneForallLabel} i.e. $[[Γ ⊢ iN ≤ iM]]$ is of the form $[[Γ ⊢ ∀pas.iN' ≤ ∀pbs.iM']]$\\
-      Then by \cref{lemma:poly-types-equivalence}, $[[Γ ⊢ ∀pas.iN' ≈ ∀pbs.iM']]$ means that 
-      there exists a bijection $[[μ : {pbs} ∩ fv iM' ↔ {pas} ∩ fv iN']]$ such that  
-      $[[Γ,pas ⊢ [Sub μ]iM' ≈ iN']]$.
-
-  \label{lemma:poly-types-equivalence}
-      Notice that the application of bijection $[[μ]]$ to $[[iM']]$ does
-      not change its size (which is less than the size of $[[iM]]$), hence the induction hypothesis applies.
-      This way, $[[ [Sub μ]iM' ≈ iN']]$ (and by symmetry, $[[iN' ≈ [Sub μ]iM']]$) holds by induction. 
-      Then we apply \ruleref{\ottdruleEOneForallLabel} to get $[[∀pas.iN' ≈ ∀pbs.iM']]$, i.e. $[[iN ≈ iM]]$.
-    \end{caseof}
-
-\item[$+$] The proof is symmetric to the proof of the negative case.
-  \end{itemize}
-\end{proof}
-
-\begin{corollary}[Normalization is complete w.r.t. subtyping-induced equivalence]
-  \label{corollary:nf-complete-wrt-subt-equiv}
-  Assuming all the types below are well-formed in $[[Γ]]$:
-  \begin{itemize}
-    \item [$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[nf(iP) = nf(iQ)]]$,
-    \item [$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{corollary}  
-\begin{proof}
-  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
-\end{proof}
-
-\begin{lemma}[Algorithmization of subtyping-induced equivalence]
-  \label{lemma:subt-equiv-algorithmization}
-  Mutual subtyping is equality of normal forms.
- Assuming all the types below are well-formed in $[[Γ]]$:
-  \begin{itemize}
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Let us prove the positive case, the negative case is symmetric.
-  We prove both directions of $\iff$ separately:
-  \begin{itemize}
-    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
-    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
-  \end{itemize}
-\end{proof}
-
-\begin{corollary}[Algorithmic substitution equivalence becomes equality after normalization]
+\begin{lemma}[Algorithmic substitution equivalence becomes equality after normalization]
   \label{corollary:subst-subt-equiv-algorithmization}
@@
   and $[[Ξ ⊆ Ξ']]$. 
-  Then $[[Γ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
-\end{corollary}
-
-\begin{lemma}[Substitution preserves declarative equivalence]
-  \label{corollary:subst-pres-equiv}
-  Suppose that $[[σ]]$ is a substitution. Then
-    \begin{itemize}
-      \item[$+$] $[[iP ≈ iQ]]$ implies $[[ [σ]iP ≈ [σ]iQ ]]$
-      \item[$-$] $[[iN ≈ iM]]$ implies $[[ [σ]iN ≈ [σ]iM ]]$
-  \end{itemize}
+  Then $[[Θ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
 \end{lemma}
 \begin{proof}
-
-  $
-  \begin{aligned}[t] 
-    [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
-                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
-                  &\Rightarrow [[ [nf(σ)]nf(iP) = [nf(σ)]nf(iQ)]]\\
-                  &\Rightarrow [[ nf([σ]iP) = nf([σ]iQ)]]
-                  && \text{by \cref{lemma:norm-subst-distr}}\\ 
-                  &\Rightarrow        [[ [σ]iP ≈ [σ]iQ ]]
-                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
-  \end{aligned} 
-  $
+  Follows immediately from \cref{lemma:subt-equiv-algorithmization}:
+  \begin{itemize}
+    \item [$\Rightarrow$]
+      If $[[α̂± ∉ Ξ]]$, then $[[ [nf(uσ1)|Ξ]α̂± = [nf(uσ2)|Ξ]α̂± ]] = [[α̂±]]$ by 
+      definition. 
+      For any $[[α̂± ∊ Ξ]]$, 
+      $[[ [nf(uσ1)|Ξ]α̂± = nf([uσ1]α̂±)]]$ and 
+      $[[ [nf(uσ2)|Ξ]α̂± = nf([uσ2]α̂±) ]]$; 
+      $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ implies
+      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by \cref{lemma:decl-equiv-algorithmization}.
+    \item [$\Leftarrow$]
+      If $[[α̂± ∊ Ξ]]$, then 
+      $[[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$ implies
+      $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ by definition 
+      of substitution restriction and normalization.
+      In turn, $[[ nf([uσ1]α̂±) = nf([uσ2]α̂±) ]]$ means
+       $[[Θ(α̂±) ⊢ [uσ1]α̂± ≈ [uσ2]α̂±]]$ by \cref{lemma:decl-equiv-algorithmization}.
+  \end{itemize}
 \end{proof}

### 2023-09-05 proofs/tex/equiv-lemmas.tex
Rearrange proofs

@@
 
-\begin{lemma}[Algorithmization of declarative equivalence]
-  \label{lemma:decl-equiv-algorithmization}
-  Declarative equivalence is the equality of normal forms. 
-  \begin{itemize}
-    \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
-    \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof} \hfill
-  \begin{itemize}
-    \item[$+$] Let us prove both directions separately.
-    \begin{itemize}
-      \item[$\Rightarrow$] 
-        exactly by \cref{lemma:normalization-completeness},
-      \item[$\Leftarrow$] 
-        from \cref{lemma:normalization-soundness}, we know
-        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
-        $[[iP ≈ iQ]]$.
-    \end{itemize}
-    \item[$-$] The proof is exactly the same.
-  \end{itemize}
-\end{proof}
-
 \begin{lemma}[Type well-formedness is invariant under equivalence]
@@
 
-\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
-  \label{lemma:wf-equiv-algo}
-  Mutual subtyping implies declarative equivalence.
-  \begin{itemize}
-  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
-  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  The proof coincides with the proof of \cref{lemma:wf-equiv},
-  and adds two cases for equating two positive or two negative metavariables,
-  which must be equal by inversion, and thus, 
-  $[[Γ; Ξ ⊢ α̂±]] \iff [[Γ; Ξ  ⊢ α̂±]]$ holds trivially.
-\end{proof}
-
 \begin{corollary}[Normalization preserves well-formedness]
@@
   Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
+\end{proof}
+
+\begin{corollary}[Normalization preserves well-formedness of substitution]
+  \label{corollary:wf-s-nf}
+  \hfill \\
+   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
+\end{corollary}
+\begin{proof}
+  Let us prove the forward direction.
+  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
+  By the definition of substitution normalization,
+  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf},
+  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
+  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
+
+  The backward direction is proved analogously.
 \end{proof}
@@
   and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
-\end{proof}
-
-\begin{corollary}[Normalization preserves well-formedness of algorithmic types]
-  \label{corollary:wf-nf-algo}
-  \begin{itemize}
-    \item[$+$] $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ nf(uP)]]$,
-    \item[$-$] $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ nf(uN)]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:wf-equiv-algo,lemma:normalization-soundness-alg}.
-\end{proof}
-
-\begin{corollary}[Normalization preserves well-formedness of substitution]
-  \label{corollary:wf-s-nf}
-  \hfill \\
-   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
-\end{corollary}
-\begin{proof}
-  Let us prove the forward direction.
-  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
-  By the definition of substitution normalization,
-  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf-algo},
-  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
-  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
-
-  The backward direction is proved analogously.
-\end{proof}
-
-\begin{corollary}[Normalization preserves the signature of the algorithmic substitution]
-  \label{corollary:norm-subst-sig-algo}
-  \hfill \\
-  $[[Θ ⊢ uσ : Ξ]] \iff [[Θ ⊢ nf(uσ) : Ξ]]$,
-  $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
-\end{corollary}
-\begin{proof}
-  The proof is analogous to \cref{corollary:wf-s-nf}.
 \end{proof}
@@
   \label{lemma:subt-equiv-algorithmization}
-  Mutual subtyping is equality of normal forms.
+  Mutual subtyping is the equality of normal forms.
  Assuming all the types below are well-formed in $[[Γ]]$:
@@
 \end{proof}
-
-\begin{corollary}[Algorithmic substitution equivalence becomes equality after normalization]
-  \label{corollary:subst-subt-equiv-algorithmization}
-  Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
-  and $[[Ξ ⊆ Ξ']]$. 
-  Then $[[Γ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
-\end{corollary}
 

### 2023-09-05 proofs/tex/norm-alg-lemmas.tex
Rearrange proofs

--- 2023-09-04 proofs/tex/norm-lemmas.tex
@@
-\begin{lemma}
-  \label{lemma:equiv-fv}
-  Set of free variables is invariant under equivalence.
-  \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
-\end{proof}
-
-\begin{lemma}
-  \label{lemma:fv-nf}
-  Free variables are not changed by the normalization
-  \begin{itemize}
-  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
-  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
-\end{proof}
-
 \begin{lemma}
@@
 \begin{proof}
-  By straightforward induction on $[[uN]]$ and mutually on $[[uP]]$.
-\end{proof}
-
-\begin{lemma}[Soundness of normalization]
-  \label{lemma:normalization-soundness}
-  \hfill
-  \begin{itemize}
-    \item[$-$] $[[iN ≈ nf(iN)]]$
-    \item[$+$] $[[iP ≈ nf(iP)]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
-  Let us consider how this judgment is formed:
-  \begin{caseof}
-    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
-      the corresponding equivalence rules.
-    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
-        and \nameref{\ottdruleNrmArrowLabel}}\\
-      By the induction hypothesis and the corresponding congruent equivalence rules.
-    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
-      From the induction hypothesis, we
-      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
-        iN]] \equiv [[fv iN']]$.
-      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
-      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
-      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.
-
-      To prove $[[∀pas.iN ≈ ∀pas'.iN']]$, it suffices to provide a bijection 
-      $\mu : [[{pas'} ∩ fv iN']] \leftrightarrow [[{pas} ∩ fv iN]]$ such that
-      $[[iN ≈ [mu]iN']]$. Since these sets are equal, we take $\mu = id$.
-    \item{\nameref{\ottdruleNrmExistsLabel}} Same as for \cref{case:norm-soundness:forall}.
-  \end{caseof}
+  By straightforward induction on $[[uN]]$ and mutually on $[[uP]]$, 
+  similar to the proof of \cref{lemma:fv-nf}.
 \end{proof}
@@
 \end{proof}
-
-\begin{corollary}[Normalization preserves ordering]
-  \label{corollary:normalization-ord}
-  For any $[[varset]]$,
-  \begin{itemize}
-  \item[$-$] $[[ord varset in nf(uN)]] = [[ord varset in uM]]$
-  \item[$+$] $[[ord varset in nf(uP)]] = [[ord varset in uQ]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:ord-completeness,lemma:normalization-soundness}.
-\end{proof}
-
-\begin{lemma}[Distributivity of normalization over substitution]
-  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
-  Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
-    \begin{itemize}
-      \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
-      \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
-  \end{itemize}
-  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
-\end{lemma}
-\begin{proof}
-  Mutual induction on $[[iN]]$ and $[[iP]]$.
-  \begin{caseof}
-    \item $[[iN]]$ = $[[na]]$ \\
-      \label{case:norm-subst-distr-var}
-      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.
-
-      $[[ [nf(σ)] nf(iN) ]] = [[ [nf(σ)] nf(na) ]] = [[ [nf(σ)] na ]] $.
-    \item $[[iP]]$ = $[[pa]]$ \\
-      Similar to \cref{case:norm-subst-distr-var}.
-   \item If the type is formed by $[[→]]$, $[[↑]]$, or $[[↓]]$, 
-     the required equality follows from the congruence of the normalization and
-     substitution, and the induction hypothesis.
-     For example, if $[[iN]] = [[iP → iM]]$ then \\
-     $\begin{aligned}[t]
-        [[nf([σ] iN)]] &= [[ nf([σ] (iP → iM)) ]] \\
-                        &= [[ nf([σ]iP → [σ]iM) ]]
-                        && \text{By the congruence of substitution} \\
-                        &= [[ nf([σ]iP) → nf([σ]iM) ]]
-                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
-                        &= [[ [nf(σ)]nf(iP) → [nf(σ)]nf(iM) ]]
-                        && \text{By the induction hypothesis} \\
-                        &= [[ [nf(σ)](nf(iP) → nf(iM)) ]]
-                        && \text{By the congruence of substitution} \\
-                        &= [[ [nf(σ)]nf(iP → iM) ]]
-                        && \text{By the congruence of normalization} \\
-                        &= [[ [nf(σ)]nf(iN) ]]
-      \end{aligned}$ \\
-    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
-      $\begin{aligned}[t]
-          [[ [nf(σ)] nf(iN) ]] &= [[ [nf(σ)] nf(∀pas.iM)]] \\
-                            &= [[ [nf(σ)] ∀pas'.nf(iM) ]]
-                            && \text {Where $[[pas']] = [[ ord {pas} in nf(iM)]]
-                               = [[ord {pas} in iM]]$
-                               (the latter is by
-                               \cref{corollary:normalization-ord})}\\
-        \end{aligned}$ \\
-
-      $\begin{aligned}[t]
-         [[ nf([σ]iN) ]] &= [[ nf([σ] ∀pas.iM)]] \\
-                          &= [[ nf(∀pas.[σ]iM) ]]
-                          && \text{Assuming $[[{pas} ∩ Γ1]] = \emptyset$
-                             and $[[{pas} ∩ Γ2]] = \emptyset$}\\
-                          &= [[ ∀pbs.nf([σ]iM) ]]
-                          && \text {Where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]
-                             = [[ord {pas} in [σ]iM]]$ (the latter is by
-                             \cref{corollary:normalization-ord})}\\
-                          &= [[ ∀pas'.nf([σ]iM) ]]
-                          && \text{By \cref{lemma:ord-sigma}, $[[pbs]] = [[pas']]$
-                             since $[[{pas}]]$ is disjoint with $[[Γ1]]$ and
-                             $[[Γ2]]$}\\
-                          &= [[ ∀pas'.[nf(σ)]nf(iM) ]]
-                          && \text {By the induction hypothesis}\\
-         \end{aligned}$ \\
-
-     To show alpha-equivalence of 
-     $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
-     we can assume that $[[{pas'} ∩ Γ1]] = \emptyset$, and $[[{pas'} ∩ Γ2]]
-     = \emptyset$.
-
-   \item $[[iP]] = [[∃ nas.iQ]]$ \\
-     Same as for \cref{case:norm-subst-commute}.
-  \end{caseof}
-\end{proof}
-
-\begin{corollary}[Commutativity of normalization and renaming]
-  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
-  Suppose that $\mu$ is a bijection between two sets of variables
-  $\mu : A \leftrightarrow B$. Then
-  \begin{itemize}
-    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
-    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:norm-subst-distr},
-  after noticing that $[[nf(mu)]] = [[mu]]$.
-\end{proof}
-
-\begin{lemma}[Completeness of quantified normalization]
-  \label{lemma:normalization-completeness}
-  Normalization returns the same representative for equivalent types.
-
-  \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$
-  \end{itemize}
-  (Here equality means alpha-equivalence)
-\end{lemma}
-
-\begin{proof}
-  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
-  \begin{caseof}
-  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall} \\
-
-    From the definition of the normalization,
-    \begin{itemize}
-      \item $[[nf(∀pas.iN)]] = [[∀pas'.nf(iN)]]$ where $[[pas']]$ is $[[ord {pas} in nf(iN)]]$
-      \item $[[nf(∀pbs.iM)]] = [[∀pbs'.nf(iM)]]$ where $[[pbs']]$ is $[[ord {pbs} in nf(iM)]]$
-    \end{itemize}
-    Let us take $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$ from the
-    inversion of the equivalence judgment. Notice that from
-    \cref{lemma:fv-nf,lemma:ord-soundness}, the domain and the codomain of $\mu$ can be written
-    as $[[mu : {pbs'} ↔ {pas'}]]$.
-
-    To show the alpha-equivalence of $[[∀pas'.nf(iN)]]$ and $[[∀pbs'.nf(iM)]]$,
-    it suffices to prove that
-    \begin{enumerate*}
-    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ and \newline
-    \item[(ii)] $[[ [mu]pbs' ]] = [[pas']]$
-    \end{enumerate*}.
-
-    \begin{enumerate}
-    \item[(i)] $[[ [mu] nf(iM) ]] = [[nf([mu]iM)]] = [[nf(iN)]]$.
-      The first equality holds by \cref{lemma:norm-subst-commute}, the second---by the induction hypothesis.
-
-    \item[(ii)] $\begin{aligned}[t] [[ [mu]pbs' ]] &= [[ [mu] ord {pbs} in nf(iM) ]]
-                                                  && \text{by the definition of $[[pbs']]$ } \\
-                                                  &= [[ [mu] ord ({pbs} ∩ fv iM) in nf(iM) ]]
-                                                  && \text{from \cref{lemma:fv-nf,corollary:ord-weakening} } \\
-                                                  &= [[ ord [mu] ({pbs} ∩ fv iM) in [mu] nf(iM) ]]
-                                                  && \text{by \cref{lemma:distr-mu-ord}, because
-                                                     $[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]]
-                                                     = \emptyset$}\\
-                                                  &
-                                                  && \text{and $[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq
-                                                     [[{pas} ∩ fv iM]] = \emptyset$} \\
-                                                  &= [[ ord [mu] ({pbs} ∩ fv iM) in nf(iN) ]]
-                                                  && \text{since $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ is proved } \\
-                                                  &= [[ ord ({pas} ∩ fv iN) in nf(iN) ]]
-                                                  && \text{because $\mu$ is a bijection between
-                                                     $[[{pas} ∩ fv iN]]$ and $[[{pbs} ∩ fv iM]]$} \\
-                                                  &= [[ ord {pas} in nf(iN) ]]
-                                                  && \text{from \cref{lemma:fv-nf,corollary:ord-weakening} } \\
-                                                  &= [[ pas' ]]
-                                                  && \text{by the definition of $[[pas']]$} \\
-      \end{aligned}$
-    \end{enumerate}
-  \item {\nameref{\ottdruleEOneExistsLabel}} Same as for \cref{case:ord-completeness:forall}.
-  \item Other rules are congruent, and thus, proved by the corresponding congruent alpha-equivalence rule,
-    which is applicable by the induction hypothesis. 
-  \end{caseof}
-\end{proof}
-
-\begin{lemma}[Idempotence of normalization]
-  \label{lemma:norm-idemp}
-  Normalization is idempotent
-  \begin{itemize}
-  \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
-  \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  By applying \cref{lemma:normalization-completeness} to \cref{lemma:normalization-soundness}.
-\end{proof}
-
-\begin{lemma}
-  \label{lemma:normal-after-subst}
-  The result of a substitution is normalized if and only if the initial type and
-  the substitution are normalized.
-
-  Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
-  $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
-  $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
-  \begin{itemize}
-  \item[$+$]
-    $[[ [σ]iP  ]] \text{ is normal} \iff
-    \begin{cases}
-      [[σ|fv(iP)]] &\text{is normal}\\
-      [[iP]]       &\text{is normal}\\
-    \end{cases} $
-  \item[$-$]
-    $[[ [σ]iN  ]] \text{is normal} \iff
-    \begin{cases}
-      [[σ|fv(iN)]] &\text{is normal}\\
-      [[iN]]       &\text{is normal}\\
-    \end{cases} $
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Mutual induction on $[[Γ1 ⊢ iP]]$ and $[[Γ1 ⊢ iN]]$.
-  \begin{caseof}
-  \item $[[iN]] = [[na]]$\\
-    Then $[[iN]]$ is always normal, and
-    the normality of $[[σ|{na}]]$ by the definition means $[[ [σ]na ]]$ is normal.
-
-  \item $[[iN]] = [[iP → iM]]$\\
-    \label{case:normal-after-subst-arrow}
-    $
-    \begin{aligned}[t]
-      [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
-                                           && \text{by the substitution
-                                              congruence} \\
-                                           &\iff
-                                             \begin{cases}
-                                             [[ [σ]iP ]] &\text{is normal} \\
-                                             [[ [σ]iM ]] &\text{is normal} \\
-                                             \end{cases}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ iP ]]       &\text{is normal} \\
-                                               [[ σ|fv(iP) ]] &\text{is normal} \\
-                                               [[ iM ]]       &\text{is normal} \\
-                                               [[ σ|fv(iM) ]] &\text{is normal} \\
-                                             \end{cases}
-                                           && \text{by the induction hypothesis}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ iP → iM ]]  &\text{is normal} \\
-                                               [[ σ|fv(iP) ∪ fv(iM)]] &\text{is normal} \\
-                                             \end{cases}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ iP → iM ]]  &\text{is normal} \\
-                                               [[ σ|fv(iP→iM)]] &\text{is normal} \\
-                                             \end{cases}
-    \end{aligned}
-    $
-  \item $[[iN]] = [[↑iP]]$\\
-    By congruence and the inductive hypothesis, similar to \cref{case:normal-after-subst-arrow}
-  \item $[[iN]] = [[∀pas.iM]]$\\
-    $
-    \begin{aligned}[t]
-      [[ [σ](∀pas.iM) ]] \text{ is normal} &\iff [[ (∀pas.[σ]iM) ]] \text{ is normal}
-                                           && \text{assuming $[[pas]] \cap [[Γ1]] = \emptyset$ and
-                                              $[[pas]] \cap [[Γ2]] = \emptyset$} \\
-                                           &\iff
-                                             \begin{cases}
-                                             [[ [σ]iM ]] \text{ is normal} \\
-                                             [[ord {pas} in [σ]iM = pas]] \\
-                                             \end{cases}
-                                           && \text{by the definition of normalization}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ [σ]iM ]] \text{ is normal} \\
-                                               [[ord {pas} in iM = pas]] \\
-                                             \end{cases}
-                                           && \text{by \cref{lemma:ord-sigma}}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ σ|fv(iM) ]] \text{ is normal} \\
-                                               [[ iM ]] \text{ is normal} \\
-                                               [[ord {pas} in iM = pas]] \\
-                                             \end{cases}
-                                           && \text{by the induction hypothesis}\\
-                                           &\iff
-                                             \begin{cases}
-                                               [[ σ|fv(∀pas.iM) ]] \text{ is normal} \\
-                                               [[ ∀pas.iM ]] \text{ is normal} \\
-                                             \end{cases}
-                                           &&
-                                              \begin{aligned}
-                                              &\text{since $[[fv(∀pas.iM) = fv(iM)]]$;}\\ &\text{by the definition of normalization}
-                                              \end{aligned}
-    \end{aligned}
-    $
-  \item $[[iP]] = \dots$\\
-    The positive cases are done in the same way as the negative ones.
-
-  \end{caseof}
-\end{proof}

### 2023-09-05 proofs/tex/norm-lemmas.tex
Rearrange proofs

@@
   By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
-\end{proof}
-
-\begin{lemma}
-  \label{lemma:uv-nf}
-  Algorithmic variables are not changed by the normalization
-  \begin{itemize}
-  \item[$-$] $[[uv uN]] \equiv [[uv nf(uN)]]$
-  \item[$+$] $[[uv uP]] \equiv [[uv nf(uP)]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  By straightforward induction on $[[uN]]$ and mutually on $[[uP]]$.
 \end{proof}
@@
 
-\begin{lemma}[Soundness of normalization of algorithmic types]
-  \label{lemma:normalization-soundness-alg}
-  \hfill
-  \begin{itemize}
-    \item[$-$] $[[uN ≈ nf(uN)]]$
-    \item[$+$] $[[uP ≈ nf(uP)]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  The proof coincides with the proof of \cref{lemma:normalization-soundness}.
-\end{proof}
-
 \begin{corollary}[Normalization preserves ordering]
@@
   \begin{itemize}
-  \item[$-$] $[[ord varset in nf(uN)]] = [[ord varset in uM]]$
-  \item[$+$] $[[ord varset in nf(uP)]] = [[ord varset in uQ]]$
+  \item[$-$] $[[ord varset in nf(iN)]] = [[ord varset in iM]]$
+  \item[$+$] $[[ord varset in nf(iP)]] = [[ord varset in iQ]]$
   \end{itemize}
@@
   \end{itemize}
-  (Here equality means alpha-equivalence)
 \end{lemma}

### 2023-09-05 proofs/tex/ord-lemmas.tex
Rearrange proofs

@@
   \label{lemma:ord-soundness}
-  Variable ordering extracts precisely used free variables.
+  Variable ordering extracts used free variables.
   \begin{itemize}
-    \item[$-$] $[[ {ord varset in uN} ]] \equiv [[varset ∩ fv uN]]$ (as sets)
-    \item[$+$] $[[ {ord varset in uP} ]] \equiv [[varset ∩ fv uP]]$ (as sets)
+    \item[$-$] $[[ {ord varset in iN} ]] = [[varset ∩ fv iN]]$ (as sets)
+    \item[$+$] $[[ {ord varset in iP} ]] = [[varset ∩ fv iP]]$ (as sets)
   \end{itemize}
@@
   Straightforward mutual induction on 
-  $[[ ord varset in uN = ordVars ]]$ and $[[ ord varset in uP = ordVars ]]$
+  $[[ ord varset in iN = ordVars ]]$ and $[[ ord varset in iP = ordVars ]]$
 \end{proof}
@@
    \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
+  \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in iN]] = [[ord varset in iM]]$ (as lists)
+  \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in iP]] = [[ord varset in iQ]]$ (as lists)
   \end{itemize}

### 2023-09-05 proofs/tex/subst-algo-lemmas.tex
Rearrange proofs

--- 2023-09-04 proofs/tex/subst-lemmas.tex
@@
-\begin{lemma}[Substitution strengthening]
-  \label{lemma:subst-restr-fv}
-  Restricting the substitution to the free variables of the
-  substitution subject does not affect the result.
-  Suppose that $[[σ]]$ is a substitution,
-  $[[iP]]$ and $[[iN]]$ are types. Then
- \begin{itemize}
-    \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
-    \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  First, we strengthen the statement by saying that
-  one can restrict the substitution to an arbitrary
-  superset of the free variables of the substitution subject:
- \begin{itemize}
-    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
-    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
-  \end{itemize}
-  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
-  For the base cases: 
-  \begin{caseof}
-    \item $[[iN = α⁻]]$\\
-      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
-      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
-    \item $[[iN = iP → iM]]$\\
-      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
-      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
-      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
-      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
-      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
-      Then 
-      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
-       = [[ [σ|varset](iP → iM) ]]$.
-    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
-    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
-      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
-      Let us assume $[[pas]]$ are fresh variables, 
-      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
-      $[[α± ]] \in [[pas]]$, and thus, 
-      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
-      immediately from the definition.
-
-      Since $[[varset ⊆ fv(∀pas.iM)]] = [[fv iM \ {pas}]]$,
-      $[[varset ∪ {pas} ⊆ fv(iM)]]$.
-      Then by the induction hypothesis, 
-      $[[ [σ]iM =  [σ|(varset ∪ {pas})]iM ]]$.
-      Finally, 
-      $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ|(varset ∪ {pas})]iM ]] = 
-      [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
-    \item The positive cases are proves symmetrically. 
-  \end{caseof}
-\end{proof}
-
 \begin{lemma}[Algorithmic Substitution Strengthening]
   \label{lemma:subst-restr-uv}
-  Restricting the substitution to the free algorithmic 
+  Restricting the substitution to the algorithmic 
   variables of the
@@
 
-\begin{lemma}[Signature of a restricted substitution]
-  \label{lemma:subst-restr-sig}
-  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset :  Γ1 ∩ varset]]$.
-\end{lemma}
-\begin{proof}
-  Let us take an arbitrary $[[α± ∊ Γ1 ∩ varset]]$.
-  Since $[[α± ∊ Γ1]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.
-
-  Let us take an arbitrary $[[α± ∉ Γ1 ∩ varset]]$.
-  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α± ]] = [[α±]]$ by definition
-  of restriction. 
-  If $[[α± ∊ varset \ Γ1]]$ then $[[ [σ|varset]α± ]] = [[ [σ]α± ]]$ by definition
-  and $[[ [σ]α± ]] = [[α±]]$ by the signature of $[[σ]]$.
-\end{proof}
-
-\begin{lemma}
-  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. 
-  Then if $[[varset]]$ is disjoint from $[[Γ1]]$, 
-  then $[[ σ|varset]] = [[ id ]]$. 
-\end{lemma}
-\begin{proof}
-  Let us take an arbitrary $[[α±]]$.
-  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α±]] = [[α±]]$ by definition.
-
-  If $[[α± ∊ varset]]$ then $[[α± ∉ Γ1]]$ by assumption.
-  Then $[[ [σ|varset]α± = [σ]α±]]$ by definition of restricted substitution, and
-  since $[[Γ2 ⊢ σ : Γ1]]$, we have $[[ [σ]α± = α± ]]$.
-\end{proof}
-
-\begin{corollary}[Application of a disjoint substitution]
-  \label{corollary:subst-disj}
-  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-    \item [$+$] if $[[Γ1 ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
-    \item [$-$] if $[[Γ1 ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
-  \end{itemize}
-\end{corollary}
-
-\begin{lemma}[Substitution range weakening]
-  \label{lemma:subst-range-weakening}
-  Suppose that $[[Γ2 ⊆ Γ'2]]$ are contexts and 
-  $[[σ]]$ is a substitution.
-  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ'2 ⊢ σ : Γ1]]$.
-\end{lemma}
-\begin{proof}
-  For any $[[α± ∊ Γ1]]$, 
-  $[[Γ2 ⊢ σ : Γ1]]$ gives us  $[[ Γ2 ⊢ [σ]α± ]]$, which can be
-  weakened to $[[ Γ'2 ⊢ [σ]α± ]]$ by \cref{lemma:wf-weakening}. 
-  This way, $[[Γ'2 ⊢ σ : Γ1]]$.
-\end{proof}
-
-\begin{lemma}[]
-  Suppose that $[[Γ' ⊆ Γ]]$,
-  $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$;
-    \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  Let us make an additional assumption that $[[σ1]]$, $[[σ2]]$, 
-  and the mentioned types are normalized. If they are not,
-  we normalize them first.
-
-  Notice that the normalization preserves
-  the set of free variables (\cref{lemma:fv-nf}),
-  well-formedness (\cref{corollary:wf-nf}), 
-  and equivalence (\cref{lemma:subt-equiv-algorithmization}), 
-  and distributes over substitution (\cref{lemma:norm-subst-distr}). 
-  This way, the assumed and desired properties are equivalent to their 
-  normalized versions.
-
-  We prove it by induction on the structure of $[[iP]]$ and mutually, $[[iN]]$.
-  Let us consider the shape of this type.
-  \begin{caseof}
-    \item $[[iP]] = [[α⁺]] \in [[Γ']]$.
-      Then $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$ means $[[Γ ⊢ σ1 ≈ σ2 : {α⁺}]]$, 
-      i.e. $[[Γ ⊢ [σ1]α⁺ ≈ [σ2]α⁺ ]]$, which holds by assumption.
-    \item $[[iP]] = [[α⁺]] \in [[Γ \ Γ']]$.
-      Then $[[fv iP ∩ Γ']] = [[∅]]$, 
-      so $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$ holds vacuously.
-    \item $[[iP]] = [[↓iN]]$.
-      Then the induction hypothesis is applicable to type $[[iN]]$:
-      \begin{enumerate}
-        \item $[[iN]]$ is normalized,
-        \item $[[Γ ⊢ iN]]$ by inversion of $[[Γ ⊢ ↓iN]]$,
-        \item $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ holds by inversion of 
-          $[[Γ ⊢ [σ1]↓iN ≈ [σ2]↓iN]]$, i.e. $[[Γ ⊢ ↓[σ1]iN ≈ ↓[σ2]iN]]$.
-      \end{enumerate}
-      This way, we obtain $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$, 
-      which implies the required equivalence since 
-      $[[fv iP ∩ Γ']] = [[fv ↓iN ∩ Γ']] = [[fv iN ∩ Γ']]$.
-    \item $[[iP]] = [[∃nas.iQ]]$
-      Then the induction hypothesis is applicable to type $[[iQ]]$ 
-      well-formed in context $[[Γ, nas]]$:
-      \begin{enumerate}
-        \item $[[Γ' ⊆ Γ, nas]]$ since $[[Γ' ⊆ Γ]]$,
-        \item $[[Γ, nas ⊢ σi : Γ']]$ by weakening,
-        \item $[[iQ]]$ is normalized,
-        \item $[[Γ, nas ⊢ iQ]]$ by inversion of $[[Γ ⊢ ∃nas.iQ]]$,
-        \item Notice that $[[ [σi]∃nas.iQ ]]$ is normalized, and thus, 
-          $[[ [σ1]∃nas.iQ ≈ [σ2]∃nas.iQ]]$ implies 
-          $[[ [σ1]∃nas.iQ = [σ2]∃nas.iQ ]]$
-          (by \cref{lemma:subt-equiv-algorithmization}).).
-          This equality means $[[ [σ1]iQ = [σ2]iQ ]]$, 
-          which implies $[[Γ ⊢ [σ1]iQ ≈ [σ2]iQ]]$.
-      \end{enumerate}
-    \item $[[iN]] = [[iP → iM]]$
-  \end{caseof}
-\end{proof}
-
-\begin{lemma}[Substitutions equal on the metavariables]
-  \label{lemma:subst-eq-metavar}
+\begin{lemma}[Substitutions equal on the algorithmic variables]
+  \label{lemma:subst-eq-algovar}
   Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
@@
 
-\begin{corollary}[Substitutions equivalent on the metavariables]
-  \label{lemma:subst-equiv-metavar}
+\begin{corollary}[Substitutions equivalent on the algorithmic variables]
+  \label{lemma:subst-equiv-algovar}
   Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
@@
 
-  This way, by \cref{lemma:subst-eq-metavar}, 
+  This way, by \cref{lemma:subst-eq-algovar}, 
   $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
@@
 \end{proof}
-
-\begin{lemma}[Substitution composition well-formedness]
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
-  then $[[Γ'1, Γ'2 ⊢ σ2 ○ σ1 : Γ1, Γ2]]$.
-\end{lemma}
-
-\begin{lemma}[Substitution monadic composition well-formedness]
-  \label{lemma:subst-monad-composition-wf}
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
-  then $[[Γ'2 ⊢ σ2 <=< σ1 : Γ1]]$.
-\end{lemma}
-
-\begin{lemma}[Substitution composition]
-  \label{lemma:subst-composition}
-    If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
-    $[[Γ1 ∩ Γ'2 = ∅ ]]$ and $[[ Γ1 ∩ Γ2 = ∅ ]]$ then 
-    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
-\end{lemma}
-
-\begin{corollary}[Substitution composition commutativity]
-  \label{corollary:subst-composition-commutativity}
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
-  $[[ Γ1 ∩ Γ2 = ∅ ]]$, 
-  $[[ Γ1 ∩ Γ'2 = ∅ ]]$, and
-  $[[ Γ'1 ∩ Γ2 = ∅ ]]$ then 
-  $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
-\end{corollary}
-\begin{proof}
-  by \cref{lemma:subst-composition},
-    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
-    Since the codomain of $[[σ1]]$ is $[[Γ'1]]$,
-    and it is disjoint with the domain of $[[σ2]]$,
-    $[[σ2 <=< σ1]] = [[σ1]]$.
-\end{proof}
-
-\begin{lemma}[Substitution domain weakening]
-  \label{lemma:subst-domain-weakening}
-  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2, Γ' ⊢ σ : Γ1, Γ']]$
-\end{lemma}
-\begin{proof}
-  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption,
-  and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
-  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
-  and thus, $[[Γ2, Γ' ⊢ α± ]]$.
-\end{proof}
-
-\begin{lemma}[Free variables after substitution]
-  \label{lemma:subst-fv}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, then
-  \begin{enumerate}
-    \item [$+$] for a type $[[iP]]$,
-      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-     $[[fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$;
-    \item [$-$] for a type $[[iN]]$, 
-      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  We prove it by structural induction on 
-  $[[iP]]$ and mutually, on $[[iN]]$.
-  \begin{caseof}
-    \item $[[iP = α⁺]]$\\
-      If $[[α⁺ ∊ Γ1]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
-      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ Γ2]]$.
-      $[[fv(α⁺) \ Γ1]] = [[∅]]$, so $[[fv([σ]iP) \ Γ1  ⊆ fv([σ]α⁺)]]$
-      vacuously.
-
-      If $[[α⁺ ∉ Γ1]]$ then $[[ [σ]α⁺ = α⁺ ]]$, and 
-      $[[fv([σ]α⁺) = {α⁺}]] = [[{α⁺} \ Γ1]]$.
-
-    \item $[[iP = ∃nas.iQ]]$\\
-      Then we need to show that $[[fv([σ]iP) = fv([σ]iQ) \ {nas}]]$
-      is a subset of $[[(fv(iP) \ Γ1) ∪ Γ2]]$ and a superset of
-      $[[fv(iP) \ Γ1]]$.
-      Notice that $[[fv(iP) = fv(iQ) \ {nas}]]$ by definition.
-      This way, we need to show that 
-      $[[fv(iQ) \ {nas} \ Γ1 ⊆ fv([σ]iQ) \ {nas} ⊆ (fv(iQ) \ {nas} \ Γ1) ∪ Γ2]]$,
-
-      By the induction hypothesis, 
-      $[[fv([σ]iQ) ⊆ (fv(iQ) \ Γ1) ∪ Γ2]]$.
-      So for the second inclusion, it suffices to show that
-      $[[((fv(iQ) \ Γ1) ∪ Γ2) \ {nas} ⊆ (fv(iQ) \ {nas} \ Γ1) ∪ Γ2]]$,
-      which holds by set theoretical reasoning.
-
-      Also by the induction hypothesis,
-      $[[fv(iQ) \ Γ1 ⊆ fv([σ]iQ)]]$,
-      and thus, by subtracting $[[{nas}]]$ from both sides,
-      $[[fv(iQ) \ {nas} \ Γ1 ⊆ fv([σ]iQ) \ {nas}]]$.
-
-    \item The case $[[iN = ∀pas.iM]]$ is proved analogously.
-    \item $[[iN = iP → iM]]$\\
-      Then $[[fv([σ]iN) = fv([σ]iP) ∪ fv([σ]iM)]]$.
-      By the induction hypothesis, 
-      \begin{enumerate}
-        \item $[[ fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$ and
-        \item $[[ fv(iM) \ Γ1 ⊆ fv([σ]iM) ⊆ (fv(iM) \ Γ1) ∪ Γ2]]$.
-      \end{enumerate}
-      We unite these inclusions vertically and obtain
-      $[[ fv(iP) \ Γ1 ∪ fv(iM) \ Γ1 ⊆ fv([σ]iN) ⊆ ((fv(iP) \ Γ1) ∪ Γ2) ∪ ((fv(iM) \ Γ1) ∪ Γ2)]]$,
-      which is equivalent to 
-      $[[ (fv(iP) ∪ fv(iM)) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iP) ∪ fv(iM)) \ Γ1 ∪ Γ2]]$.
-      Since $[[fv(iP) ∪ fv(iM) = fv(iN)]]$, 
-      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
-    \item The cases when $[[iP = ↓iM]]$ and $[[iN = ↑iQ]]$ are proved
-      analogously
-  \end{caseof}
-\end{proof}
-
-\begin{lemma}[Free variables of a variable image]
-  \label{lemma:subst-fv-image}
-  Suppose that $[[σ]]$ is an arbitrary substitution, 
-  Then 
-  \begin{itemize}
-    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
-    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  By straightforward mutual induction on $[[iP]]$ and $[[iN]]$.
-\end{proof}

### 2023-09-05 proofs/tex/subst-lemmas.tex
Rearrange proofs

@@
   \end{caseof}
-\end{proof}
-
-\begin{lemma}[Algorithmic Substitution Strengthening]
-  \label{lemma:subst-restr-uv}
-  Restricting the substitution to the free algorithmic 
-  variables of the
-  substitution subject does not affect the result.
-  Suppose that $[[uσ]]$ is an algorithmic substitution,
-  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
- \begin{itemize}
-    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
-    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv uN]uN ]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
 \end{proof}
@@
   \end{caseof}
-\end{proof}
-
-\begin{lemma}[Substitutions equal on the metavariables]
-  \label{lemma:subst-eq-metavar}
-  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
-  normalized substitutions of signature $[[Θ ⊢ uσi : Ξ]]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a normalized type $[[Γ; Ξ ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
-      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
-    \item [$-$] for a normalized type $[[Γ; Ξ ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
-      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
-  \end{enumerate}
-\end{lemma}
-\begin{proof}
-  The proof is a simple structural induction on 
-  $[[Γ; Ξ ⊢ uP]]$ and mutually, on $[[Γ; Ξ ⊢ uN]]$.
-  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
-  \begin{caseof}
-    \item $[[Γ; Ξ ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
-      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
-    \item $[[Γ; Ξ ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
-      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
-    \item $[[Γ; Ξ ⊢ ∀pas.uN]]$. Then we are proving that
-      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
-      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
-      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
-      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; Ξ ⊢ uN]]$, 
-      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
-      as required.
-    \item $[[Γ; Ξ ⊢ uP → uN]]$. Then we are proving that
-      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
-      By definition of substitution and congruence of equality, 
-      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
-      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
-      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
-      and well-formed in the same contexts.
-      This way, by the induction hypothesis, 
-      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
-      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
-      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
-    \item $[[Γ; Ξ ⊢ ↑uP]]$. The proof is similar to the previous case:
-      we apply congruence of substitution, equality, and normalization,
-      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
-  \end{caseof}
-\end{proof}
-
-\begin{corollary}[Substitutions equivalent on the metavariables]
-  \label{lemma:subst-equiv-metavar}
-  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
-  of signature $[[Θ ⊢ uσi : Ξ]]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a type $[[Γ; Ξ ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
-    \item [$-$] for a type $[[Γ; Ξ ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
-  \end{enumerate}
-\end{corollary}
-\begin{proof}
-  First, let us normalize the types and the substitutions, and show that
-  the given equivalences and well-formedness properties are preserved. 
-  $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
-  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
-  \cref{lemma:subt-equiv-algorithmization}.
-  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
-  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
-  Notice that by \cref{corollary:norm-subst-sig-algo}
-  $[[Θ ⊢ uσi : Ξ]]$ implies $[[Θ ⊢ nf(uσi) : Ξ]]$.
-
-  This way, by \cref{lemma:subst-eq-metavar}, 
-  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
-  $[[nf(uσ1)|(uv nf(uP)) = nf(uσ2)|(uv nf(uP))]]$.
-  Then by \cref{lemma:uv-nf}, 
-  $[[nf(uσ1)|(uv uP) = nf(uσ2)|(uv uP)]]$,
-  and by ${{corollary:subst-subt-equiv-algorithmization}}$,
-  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$.
-
-  Symmetrically, 
-  $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
-  $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
 \end{proof}

### 2023-09-05 proofs/tex/wf-algo-lemmas.tex
Rearrange proofs

@@
-\begin{lemma}[AASoundness of algorithmic type well-formedness]
+\begin{lemma}[Soundness of algorithmic type well-formedness]
     \label{lemma:wf-algo-soundness}
@@
 \begin{lemma}[Variable algorithmization agrees with well-formedness]
+  \label{lemma:var-algo-wf}
   \hfill
@@
 \begin{proof}
-  The proof is a simple structural induction on $[[Γ, nas ⊢ iP]]$ and mutually, on $[[Γ, nas ⊢ iN]]$.
+  The proof is a structural induction on $[[Γ, nas ⊢ iP]]$ and mutually, on $[[Γ, nas ⊢ iN]]$.
+  Notice that the substitutions commute with all the constructors,
+  providing the step of the induction. 
+\end{proof}
+
+\begin{lemma}[Variable dealgorithmization agrees with well-formedness]
+  \label{lemma:var-dealgo-wf}
+  \hfill
+  \begin{itemize}
+    \item[$+$]  $[[Γ; {nuas} ⊢ uP]]$ implies $[[Γ, nas ⊢ [nas/nuas]iP]]$;
+    \item[$-$]  $[[Γ; {nuas} ⊢ uN]]$ implies $[[Γ, nas ⊢ [nas/nuas]iN]]$.
+  \end{itemize}
+\end{lemma}
+\begin{proof}
+  As for \cref{lemma:var-algo-wf}, the proof is a structural induction on 
+  $[[Γ; {nuas} ⊢ uP]]$ and mutually, on $[[Γ; {nuas} ⊢ uN]]$.
 \end{proof}
@@
 \begin{proof}
-  By \cref{lemma:wf-soundness},
-  $[[Γ1 ⊢ iP]]$ implies $[[fv(iP) ⊆ Γ1]]$,
-  which means that $[[fv(iP) ⊆ Γ2]]$,
-  and thus, $[[fv(iP)]] = [[fv(iP) ∩ Γ1]] = [[fv(iP) ∩ Γ2]]$.
-  Then by \cref{lemma:wf-ctxt-equiv}, $[[Γ2 ⊢ iP]]$. 
+  By \cref{lemma:wf-algo-soundness},
+  $[[Γ1; Ξ1 ⊢ uP]]$ implies $[[fv(uP) ⊆ Γ1 ⊆ Γ2]]$
+  and $[[uv(uP) ⊆ Ξ1 ⊆ Ξ2]]$,
+  and thus, $[[fv(uP)]] = [[fv(uP) ∩ Γ1]] = [[fv(uP) ∩ Γ2]]$, 
+  and $[[uv(uP)]] = [[uv(uP) ∩ Ξ1]] = [[uv(uP) ∩ Ξ2]]$.
+  Then by \cref{lemma:wf-algo-ctxt-equiv}, $[[Γ2; Ξ2 ⊢ uP]]$. 
   The negative case is symmetric.

### 2023-09-06 proofs/main.tex
Add deterinicity lemmas, get rid of (??)

@@
 
+\subsection{Fresh Variable Selection}
+\label{sec:fresh-selection}
+Both the subtyping and the type inference algorithm
+rely on the ability to select fresh, unused variables.
+In the inference rules, it is indicated as 
+$[[{varset} are fresh]]$. 
+We assume that the selection subroutine always succeeds and is 
+deterministic. In other words, whenever it is called in 
+an algorithmic inference rule, it returns the same result, 
+uniquely determined by the input of this rule.
+
 \subsection{Variable Algorithmization}
@@
 \begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
+  \label{lemma:merge-matching-entries-welldef}
   Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
@@
   and merge of matching entries, which also are  
-  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-is-well-defined}.
+  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-welldef}.
 \end{proof}
@@
 \subsection{Anti-unification}
-\input{../_build/_gen/tex/anti-unification-lemmas.tex}
+\input{../_build/_gen/tex/au-lemmas.tex}
 

### 2023-09-06 proofs/tex/alg-neg-subtyping-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
+\begin{observation}[Negative Algorithmic Subtyping is Deterministic]
+    \label{obs:neg-subt-deterministic}
+    For fixed $[[Γ]]$, $[[Θ]]$, $[[iM]]$, and $[[uN]]$, 
+    if $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ 
+    and $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC']]$
+    then $[[SC]]$ = $[[SC']]$.
+\end{observation}
+\begin{proof}
+    First, notice that the shape of
+    the input uniquely determines the rule applied to infer
+    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$,
+    which is consequently, the same rule used to
+    infer $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC']]$.
+
+    Second, notice that for each of the inference rules, 
+    the premises are deterministic on the input.
+    Specifically, 
+    \begin{itemize}
+        \item \ruleref{\ottdruleAShiftULabel} relies on unification,
+            which is deterministic by \cref{obs:unif-deterministic};
+        \item \ruleref{\ottdruleAForallLabel} relies on
+            the choice of fresh algorithmic variables,
+            which is deterministic as discussed in \cref{sec:fresh-selection},
+            and on the negative subtyping, which is deterministic by
+            the induction hypothesis;
+        \item \ruleref{\ottdruleAArrowLabel} uses 
+            the negative subtyping 
+            (deterministic by the induction hypothesis),
+            the positive subtyping 
+            (\cref{obs:pos-subt-deterministic}),
+            and the merge of subtyping constraints
+            (\cref{obs:subt-merge-deterministic});
+    \end{itemize}
+\end{proof}
+
 \begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}

### 2023-09-06 proofs/tex/alg-pos-subtyping-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
+\begin{observation}[Positive Subtyping is Deterministic]
+    \label{obs:pos-subt-deterministic}
+    For fixed $[[Γ]]$, $[[Θ]]$, $[[uP]]$, and $[[iQ]]$,
+    if $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC']]$
+    then $[[SC]] = [[SC']]$.
+\end{observation}
+\begin{proof}
+    We prove it by induction on $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
+    First, it is easy to see that the rule applied to infer 
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ uniquely depends on the input, 
+    and those, it is the same rule that is inferring
+    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC']]$.
+    Second, the premises of each rule are deterministic on the input: 
+    unification is deterministic by \cref{obs:unif-deterministic},
+    upgrade is deterministic by \cref{obs:upgrade-deterministic},
+    the choice of the fresh algorithmic variables is deterministic by
+    convention, as discussed in \cref{sec:fresh-selection},
+    positive subtyping by the induction hypothesis.
+\end{proof}
+
 \begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}

### 2023-09-06 proofs/tex/au-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
-\begin{lemma}[Soundness of the anti-unification algorithm]
-  \label{lemma:au-soundness}
+\begin{observation}[Anti-unification algorithm is deterministic]
+    \label{obs:au-deterministic}
+    \hfill\\
+\begin{itemize}
+    \item [$+$] If $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and 
+    $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ', uQ', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uQ]] = [[uQ']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+    \item [$-$] If $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
+    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ', uM', aus'1, aus'2)]]$,
+    then $[[Ξ]] = [[Ξ']]$, $[[uM]] = [[uM']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
+\end{itemize}
+\end{observation}
+\begin{proof}
+    By trivial induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+\end{proof}
+
+\begin{observation} \label{obs:names-defined-by-mapping}
+    Names of the anti-unification variables are uniquely defined by
+    the types they are mapped to by the resulting substitutions.
+
+    \begin{itemize}
+        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+        \item [$-$]  Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
+        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
+    \end{itemize}
+\end{observation}
+\begin{proof}
+    By simple induction on $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+    Let us consider tha last rule applied to infer this judgment.
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUPVarLabel} or \ruleref{\ottdruleAUNVarLabel},
+        then $[[Ξ]] = [[·]]$, and the property holds vacuously.
+
+        \item \ruleref{\ottdruleAUAULabel}
+        Then  $[[Ξ]] = [[â⁻_{iN1, iN2}]]$,
+        $[[aus1]] = [[â⁻_{iN1, iN2} :≈ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} :≈ iN2]]$.
+        So the property holds trivially.
+
+        \item \ruleref{\ottdruleAUArrowLabel}
+        In this case, $[[Ξ]] = [[Ξ' ∪ Ξ'']]$, $[[aus1]] = [[aus'1 ∪ aus''1]]$, and 
+        $[[aus2]] = [[aus'2 ∪ aus''2]]$,
+        where the property holds for ($[[Ξ']]$, $[[aus'1]]$, $[[aus'2]]$) and 
+        ($[[Ξ'']]$, $[[aus''1]]$, $[[aus''2]]$) by the induction hypothesis.
+        Then since the union of solutions does not change the types the variables are mapped to,
+        the required property holds for $[[Ξ]]$, $[[aus1]]$, and $[[aus2]]$.
+
+        \item For the other rules, the resulting $[[Ξ]]$ is taken from the recursive call
+        and the required property holds immediately by the induction hypothesis.
+    \end{caseof}
+\end{proof}
+
+\begin{lemma}[Soundness of Anti-Unification] \label{lemma:au-soundness}
+    \hfill
+    \begin{itemize}
+        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then 
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uQ]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
+            are anti-unification solutions, and
+            \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
+        \end{enumerate}
+
+        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
+        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+        then
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uM]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
+            are anti-unification solutions, and
+            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
+        \end{enumerate}
+    \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove it by induction on 
+    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
+    and mutually, $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$.
+    Let us consider the last rule applied to infer this judgement.
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUNVarLabel}, then $[[iN1]] = [[α⁻]] = [[iN2]]$,
+              $[[Ξ]] = [[·]]$, $[[uM]] = [[α⁻]]$, and $[[aus1]] = [[aus2]] = [[·]]$.
+            \begin{enumerate}
+                \item $[[Γ ; · ⊢ α⁻]]$ follows from the assumption $[[Γ ⊢ α⁻]]$,
+                \item $[[Γ ; · ⊢ · : ·]]$ holds trivially, and
+                \item $[[ [·] α⁻ = α⁻ ]]$ holds trivially.
+            \end{enumerate}
+        \item \label{case:anti-unification-soundness:shiftu}
+         \ruleref{\ottdruleAUShiftULabel}, then $[[iN1]] = [[↑iP1]]$,
+                $[[iN2]] = [[↑iP2]]$, and the algorithm makes the recursive call:
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
+                returning $[[(Ξ, ↑uQ, aus1, aus2)]]$ as the result.
 
-\begin{lemma}[Completeness of the anti-unification algorithm]
-  \label{lemma:au-completeness}
+                Since $[[iN1]] = [[↑iP1]]$ and $[[iN2]] = [[↑iP2]]$ are normalized, 
+                so are $[[iP1]]$ and $[[iP2]]$, and thus, the induction hypothesis 
+                is applicable to $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$:
+                \begin{enumerate}
+                    \item $[[Γ ; Ξ ⊢ uQ]]$, and hence, $[[Γ ; Ξ ⊢ ↑uQ]]$,
+                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+                    \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$, and then by 
+                    the definition of the substitution, $[[ [ausi] ↑uQ = ↑iPi ]]$ for $i \in \{1,2\}$.
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUArrowLabel}, then $[[iN1]] = [[iP1 → iN'1]]$,
+                $[[iN2]] = [[iP2 → iN'2]]$, and the algorithm makes two recursive calls:
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ', uM, aus'1, aus'2)]]$ and
+                and returns $[[(Ξ ∪ Ξ', uQ → uM, aus1 ∪ aus'1, aus2 ∪ aus'2)]]$ as the result.
+
+                Notice that the induction hypothesis is applicable to 
+                $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$:
+                $[[iP1]]$ and $[[iP2]]$ are normalized, since $[[iN1]] = [[iP1 → iN'1]]$
+                and $[[iN2]] = [[iP2 → iN'2]]$ are normalized.
+                Similarly, the induction hypothesis is applicable to
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ', uM, aus'1, aus'2)]]$.
+
+                This way, by the induction hypothesis:
+                \begin{enumerate}
+                    \item $[[Γ ; Ξ ⊢ uQ]]$ and $[[Γ ; Ξ' ⊢ uM]]$. 
+                    Then by weakening (\cref{lemma:wf-weakening-algo}), 
+                    $[[Γ ; Ξ ∪ Ξ' ⊢ uQ]]$ and 
+                    $[[Γ ; Ξ ∪ Ξ' ⊢ uM]]$, which implies $[[Γ ; Ξ ∪ Ξ' ⊢ uQ → uM]]$;
+
+                    \item $[[Γ ; · ⊢ ausi : Ξ]]$ and $[[Γ ; · ⊢ aus'i : Ξ']]$
+                        Then $[[Γ ; · ⊢ ausi ∪ aus'i : Ξ ∪ Ξ']]$
+                        are well-defined anti-unification solution.
+                        Let us take an arbitrary $[[β̂⁻]] \in [[Ξ ∪ Ξ']]$.
+                        If $[[β̂⁻]] \in [[Ξ]]$.
+                        then $[[Γ ; · ⊢ ausi : Ξ]]$ implies that $[[ausi]]$, and hence,
+                        $[[ausi ∪ aus'i]]$ contains an entry well-formed in $[[Γ]]$.
+                        If $[[β̂⁻]] \in [[Ξ']]$, the reasoning is symmetric.
+
+                        $[[ausi ∪ aus'i]]$ is a well-defined anti-unification solution:
+                        any anti-unification variable occurs 
+                        uniquely $[[ausi ∪ aus'i]]$, since by \cref{obs:names-defined-by-mapping},
+                        the name of the variable is in one-to-one correspondence with 
+                        the pair of types it is mapped to
+                        by $[[aus1]]$ and $[[aus2]]$, 
+                        an is in one-to-one correspondence with the pair of types it is mapped to
+                        by $[[aus'1]]$ and $[[aus'2]]$ 
+                        i.e.  if $[[β̂⁻]] \in [[Ξ ∩ Ξ']]$ then $[[ [aus1]β̂⁻ = [aus'1]β̂⁻ ]]$,
+                        and $[[ [aus2]β̂⁻ = [aus'2]β̂⁻ ]]$.
+
+                    \item $[[ [ausi] uQ = iPi ]]$ and $[[ [aus'i] uM = iN'i ]]$.
+                    Since $[[ausi ∪ aus'i]]$ restricted to $[[Ξ]]$ is $[[ausi]]$,
+                    and $[[ausi ∪ aus'i]]$ restricted to $[[Ξ']]$ is $[[aus'i]]$,
+                    we have $[[ [ausi ∪ aus'i] uQ = iPi ]]$ and 
+                    $[[ [ausi ∪ aus'i] uM = iN'i ]]$, and thus, 
+                    $[[ [ausi ∪ aus'i] uQ → uM = iP1 → iN'1 ]]$
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUForallLabel}, then $[[iN1]] = [[∀pas.iN'1]]$,
+                $[[iN2]] = [[∀pas.iN'2]]$, and the algorithm makes a recursive call:
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
+                returns $[[(Ξ, ∀pas.uM, aus1, aus2)]]$ as the result.
+
+                Similarly to \cref{case:anti-unification-soundness:shiftu}, 
+                we apply the induction hypothesis to
+                $[[Γ ⊨ iN'1 ≈au iN'2 ⫤ (Ξ, uM, aus1, aus2)]]$ to obtain:
+                \begin{enumerate}
+                    \item $[[Γ; Ξ ⊢ uM]]$, and hence, $[[Γ ; Ξ ⊢ ∀pas.uM]]$;
+                    \item $[[Γ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$, and
+                    \item $[[ [ausi] uM = iN'i ]]$ for $i \in \{1,2\}$,
+                        and then by the definition of the substitution,
+                        $[[ [ausi] ∀pas.uM = ∀pas.iN'i ]]$ for $i \in \{1,2\}$. 
+                \end{enumerate}
+
+        \item \ruleref{\ottdruleAUAULabel}, which applies 
+        when other rules do not, and $[[G ⊢ iNi]]$,
+        returning as the result $[[(Ξ, uM, aus1, aus2)]] = $
+        $[[({â⁻_{iN1, iN2}}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} :≈ iN1) ,  (â⁻_{iN1, iN2} :≈ iN2))]]$.
+
+        \begin{enumerate}
+            \item $[[Γ ; Ξ ⊢ uM]]$ is rewritten as $[[Γ ; { â⁻_{iN1, iN2} } ⊢ â⁻_{iN1, iN2}]]$,
+                which holds trivially;
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} :≈ iNi) : {â⁻_{iN1, iN2}}]]$,
+                which holds since $[[Γ ⊢ iNi]]$ by the premise of the rule;
+            \item $[[ [ausi] uM = iNi ]]$ is rewritten as $[[ [â⁻_{iN1, iN2} :≈ iNi] â⁻_{iN1, iN2} = iNi ]]$,
+                which holds trivially by the definition of substitution.
+        \end{enumerate}
+
+        \item Positive cases are proved symmetrically.
+    \end{caseof}
+\end{proof}
+
+% \renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
+% \renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
+
+% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
+% \renewcommand{\ottdruleAUNVarName}[0]{(Var$^{-[[≈au]]}$)}
+% \renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
+% \renewcommand{\ottdruleAUAUName}[0]{(AU$^{-}$)}
+
+\begin{lemma}[Completeness of Anti-Unification] \label{lemma:au-completeness}
+    \hfill
+    \begin{itemize}
+        \item [$+$] 
+            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
+            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uQ']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
+                are anti-unification solutions, and
+                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
+            \end{enumerate}
+
+            Then the anti-unification algorithm terminates, that is there exists
+            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+
+        \item [$-$] 
+            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
+            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uM']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
+                are anti-unification solutions, and
+                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
+            \end{enumerate}
+
+            Then the anti-unification algorithm succeeds, that is 
+            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+    \end{itemize}
 \end{lemma}
+\begin{proof}
+    We prove it by the induction on $[[uM']]$ and mutually on $[[uQ']]$.
+    \begin{caseof}
+        \item $[[uM']] = [[â⁻]]$ 
+            Then since $[[Γ ; · ⊢ aus'i : Ξ']]$,
+            $[[Γ ⊢ [aus'i] uM']] = [[ iNi ]]$. 
+            This way, \ruleref{\ottdruleAUAULabel} is always applicable
+            if other rules are not.
 
-\begin{lemma}[Initiality of the anti-unification algorithm]
-  \label{lemma:au-initiality}
+        \item $[[uM']] = [[α⁻]]$
+            Then $[[α⁻]] = [[ [aus'i] α⁻]] = [[ iNi ]]$, which means
+            that \ruleref{\ottdruleAUNVarLabel} is applicable.
+
+        \item $[[uM']] = [[↑uQ']]$
+            Then $[[ ↑[aus'i]uQ']] = [[ [aus'i]↑uQ']] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[↑iP1]]$ and $[[↑iP2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uQ' = iPi ]]$, which means that 
+            $[[(Ξ', uQ', aus'1, aus'2)]]$ is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+            Then by the induction hypothesis, there exists $[[(Ξ, uQ, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, and hence, 
+            $[[Γ ⊨ ↑iP1 ≈au ↑iP2 ⫤ (Ξ, ↑uQ, aus1, aus2)]]$ by \ruleref{\ottdruleAUShiftULabel}.
+        \item $[[uM']] = [[∀pas.uM'']]$ This case is similar to the previous one:
+            we consider $\forall[[pas]]$ as a constructor. 
+            Notice that $[[∀pas.[aus'i]uM'']] = [[ [aus'i]∀pas.uM'']] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[∀pas.iN''1]]$ and $[[∀pas.iN''2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uM'' = iN''i ]]$, which means that
+            $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+            Then by the induction hypothesis, there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
+            $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ, uM, aus1, aus2)]]$, and hence,
+            $[[Γ ⊨ ∀pas.iN''1 ≈au ∀pas.iN''2 ⫤ (Ξ, ∀pas.uM, aus1, aus2)]]$ by 
+            \ruleref{\ottdruleAUForallLabel}.
+        \item $[[uM']] = [[uQ' → uM'']]$
+            Then $[[ [aus'i]uQ' → [aus'i]uM'']] = [[ [aus'i](uQ' → uM'')]] = [[ iNi ]]$, that is
+            $[[iN1]]$ and $[[iN2]]$ have form $[[uP1 → uN'1]]$ and $[[uP2 → uN'2]]$ respectively.
+
+            Moreover, $[[ [aus'i]uQ' = iPi ]]$ and $[[ [aus'i]uM'' = iN''i ]]$, which means that
+            $[[(Ξ', uQ', aus'1, aus'2)]]$ is an anti-unifier of $[[iP1]]$ and $[[iP2]]$,
+            and $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+            Then by the induction hypothesis, 
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ1, uQ, aus1, aus2)]]$ and 
+            $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ2, uM, aus3, aus4)]]$ succeed.
+            The result of the algorithm is $[[(Ξ1 ∪ Ξ2, uQ → uM, aus1 ∪ aus3, aus2 ∪ aus4)]]$.
+
+        \item $[[uQ']] = [[â⁺]]$ 
+            This case if not possible, since $[[Γ ; Ξ' ⊢ uQ']]$ means 
+            $[[â⁺]] \in [[Ξ']]$, but $[[Ξ']]$ can only contain negative variables.
+
+        \item Other positive cases are proved symmetrically to the corresponding negative ones.
+    \end{caseof}
+\end{proof}
+
+\begin{lemma}[Initiality of Anti-Unification] \label{lemma:au-initial}
+    \hfill
+    \begin{itemize}
+        \item [$+$] 
+            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
+            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
+            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
+            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
+            if 
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uQ']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
+                are anti-unification solutions , and
+                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
+            \end{enumerate}
+            then there exists $[[ausr]]$ such that
+            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
+            Moreover, $[[ [ausr]β̂⁻]]$ can
+            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
+            $[[Γ]]$.
+        \item [$-$] 
+            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
+            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
+            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
+            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
+            if
+            \begin{enumerate}
+                \item $[[Γ ; Ξ' ⊢ uM']]$,
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
+                are anti-unification solutions, and
+                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
+            \end{enumerate}
+            then there exists $[[ausr]]$ such that
+            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
+            Moreover, $[[ [ausr]β̂⁻]]$ can
+            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
+            $[[Γ]]$.
+    \end{itemize}
 \end{lemma}
+\begin{proof}
+    First, let us assume that $[[uM']]$ is a algorithmic variable $[[α̂⁻]]$. 
+    Then we can take $[[ausr]] = [[α̂⁻]] \mapsto [[uM]]$, which satisfies the required properties:
+    \begin{itemize}
+        \item $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ holds since 
+        $[[Ξ' | uv uM']] = [[{α̂⁻}]]$ and $[[Γ ; Ξ ⊢ uM]]$ by the soundness of anti-unification (\cref{lemma:au-soundness});
+        \item $[[ [ausr] uM' = uM ]]$ holds by construction
+        \item $[[ [ausr]α̂⁻]] = [[uM]]$ is the anti-unifier of 
+            $[[iN1]] = [[ [aus'1] α̂⁻]]$ and $[[iN2]] = [[ [aus'2] α̂⁻]]$
+            in context $[[Γ]]$, and hence, it is uniquely determined by them (\cref{obs:au-deterministic}).
+    \end{itemize}
+
+    Now, we can assume that $[[uM']]$ is not a algorithmic variable. 
+    We prove by induction on the derivation of $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+    and mutually on the derivation of $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
+
+    Since $[[uM']]$ is not a algorithmic variable, 
+    the substitution acting on $[[uM']]$ preserves its outer constructor. 
+    In other words, 
+    $[[ [aus'i] uM' = iNi ]]$ means that $[[uM']]$, 
+    $[[iN1]]$ and $[[iN2]]$ have the same outer constructor. 
+    Let us consider the algorithmic anti-unification rule corresponding to this constructor, 
+    and show that it was successfully applied to anti-unify $[[iN1]]$ and $[[iN2]]$ 
+    (or $[[iP1]]$ and $[[iP2]]$).
+
+    \begin{caseof}
+        \item \ruleref{\ottdruleAUNVarLabel}, i.e. $[[iN1]] = [[α⁻]] = [[iN2]]$.
+        \label{case:anti-unification-initial:nvar}
+        This rule is applicable since it has no premises.
+
+        Then $[[Ξ]] = [[·]]$, $[[uM]] = [[α⁻]]$, 
+        and $[[aus1]] = [[aus2]] = [[·]]$.
+        Since $[[ [aus'i] uM' = iNi ]] = [[α⁻]]$
+        and $[[uM']]$ is not a algorithmic variable, $[[uM']] = [[α⁻]]$.
+        Then we can take $[[ausr]] = [[·]]$, which satisfies the required properties:
+        \begin{itemize}
+            \item $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ holds vacuously since 
+            $[[Ξ' | uv uM']] = [[∅]]$; 
+            \item $[[ [ausr] uM' = uM ]]$, that is $[[ [·] α⁻ = α⁻ ]]$
+            holds by substitution properties;
+            \item the unique determination of $[[ [ausr]α̂⁻]]$ for $[[α̂⁻]] \in [[Ξ' | uv uM']] = [[∅]]$ holds vacuously.
+        \end{itemize}
+
+        \item \ruleref{\ottdruleAUShiftULabel}, i.e. 
+        $[[iN1]] = [[↑iP1]]$ and $[[iN2]] = [[↑iP2]]$.
+        \label{case:anti-unification-initial:shiftu}
+
+        Then since $[[ [aus'i] uM' = iNi ]] = [[↑iPi]]$ and $[[uM']]$ is not a algorithmic variable,
+        $[[uM']] = [[↑uQ']]$, where $[[ [aus'i] uQ' = iPi ]]$. 
+        Let us show that $[[(Ξ', uQ', aus'1, aus'2)]]$ 
+        is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+        \begin{enumerate}
+            \item $[[Γ ; Ξ' ⊢ uQ']]$ holds by inversion of $[[Γ ; Ξ' ⊢ ↑uQ']]$;
+            \item $[[Γ ; · ⊢ aus'i : Ξ']]$ holds by assumption;
+            \item $[[ [aus'i] uQ' = iPi ]]$ holds by assumption.
+        \end{enumerate}
+
+        This way, by the completeness of anti-unification 
+        (\cref{lemma:au-completeness}),
+        the anti-unification algorithm succeeds on $[[iP1]]$ and $[[iP2]]$:
+        $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$,
+        which means that \ruleref{\ottdruleAUShiftULabel} is applicable to infer 
+        $[[Γ ⊨ ↑iP1 ≈au ↑iP2 ⫤ (Ξ, ↑uQ, aus1, aus2)]]$.
+
+        Moreover, by the induction hypothesis,
+        $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than $[[(Ξ', uQ', aus'1, aus'2)]]$,
+        which immediately implies that $[[(Ξ, ↑uQ, aus1, aus2)]]$ is more specific than
+        $[[(Ξ', ↑uQ', aus'1, aus'2)]]$ (we keep the same $[[ausr]]$).
+
+        \item \ruleref{\ottdruleAUForallLabel}, i.e. 
+        $[[iN1]] = [[∀pas.iN'1]]$ and $[[iN2]] = [[∀pas.iN'2]]$.
+        \label{case:anti-unification-initial:forall}
+        The proof is symmetric to the previous case.
+        Notice that the context $[[Γ]]$ is not changed in \ruleref{\ottdruleAUForallLabel}, 
+        as it represents the context in which the anti-unification variables must be instantiated,
+        rather than the context forming the types that are being anti-unified.
+
+        \item \ruleref{\ottdruleAUArrowLabel}, i.e.
+        $[[iN1]] = [[iP1 → iN'1]]$ and $[[iN2]] = [[iP2 → iN'2]]$.
+
+        Then since $[[ [aus'i] uM' = iNi ]] = [[iPi → iN'i]]$ and $[[uM']]$ is not a algorithmic variable,
+        $[[uM']] = [[uQ' → uM'']]$, where $[[ [aus'i] uQ' = iPi ]]$ and $[[ [aus'i] uM'' = iN''i ]]$.
+
+        Let us show that $[[(Ξ', uQ', aus'1, aus'2)]]$
+        is an anti-unifier of $[[iP1]]$ and $[[iP2]]$.
+        \begin{enumerate}
+            \item $[[Γ ; Ξ' ⊢ uQ']]$ holds by inversion of $[[Γ ; Ξ' ⊢ uQ' → uM'']]$;
+            \item $[[Γ ; · ⊢ aus'i : Ξ']]$ holds by assumption;
+            \item $[[ [aus'i] uQ' = iPi ]]$ holds by assumption.
+        \end{enumerate}
+
+        Similarly, $[[(Ξ', uM'', aus'1, aus'2)]]$ is an anti-unifier of $[[iN''1]]$ and $[[iN''2]]$.
+
+        Then by the completeness of anti-unification (\cref{lemma:au-completeness}),
+        the anti-unification algorithm succeeds on $[[iP1]]$ and $[[iP2]]$:
+        $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ1, uQ, aus1, aus2)]]$;
+        and on $[[iN'1]]$ and $[[iN'2]]$:
+        $[[Γ ⊨ iN''1 ≈au iN''2 ⫤ (Ξ2, uM''', aus3, aus4)]]$.
+        Notice that $[[aus1 & aus3]]$ and $[[aus2 & aus4]]$ are defined, 
+        in other words, for any $[[β̂⁻]] \in [[Ξ1]] \cap [[Ξ2]]$,
+        $[[ [aus1] β̂⁻ = [aus2] β̂⁻ ]]$ and $[[ [aus3] β̂⁻ = [aus4] β̂⁻ ]]$,
+        which follows immediately from \cref{obs:names-defined-by-mapping}.
+        This way, the algorithm proceeds by applying \ruleref{\ottdruleAUArrowLabel} and returns
+        $[[(Ξ1 ∪ Ξ2, uQ → uM''', aus1 ∪ aus3, aus2 ∪ aus4)]]$.
+
+        It is left to construct $[[ausr]]$ such that $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
+        By the induction hypothesis, there exist $[[ausr1]]$ and $[[ausr2]]$ such that
+        $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$, 
+        $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$,
+        $[[ [ausr1] uQ' = uQ ]]$, and $[[ [ausr2] uM'' = uM''' ]]$.
+
+        Let us show that $[[ausr]] = [[ausr1 ∪ ausr2]]$ satisfies the required properties:
+        \begin{itemize}
+            \item $[[Γ ; Ξ1 ∪ Ξ2 ⊢ ausr1 ∪ ausr2 : (Ξ' | uv uM')]]$ holds since 
+            $[[Ξ' | uv uM']] = [[Ξ' | uv uQ' → uM'']] = [[(Ξ' | uv uQ') ∪ (Ξ' | uv uM'')]]$,
+            $[[Γ ; Ξ1 ⊢ ausr1 : (Ξ' | uv uQ')]]$ and $[[Γ ; Ξ2 ⊢ ausr2 : (Ξ' | uv uM'')]]$;
+            \item $[[ [ausr] uM' ]] = [[ [ausr] (uQ' → uM'') ]] = [[ [ausr | uv uQ'] uQ' → [ausr | uv uM''] uM'' ]] =
+            [[ [ausr1] uQ' → [ausr2] uM'' ]] = [[ uQ → uM''']] = [[ uM ]]$;
+            \item Since $[[ [ausr]β̂⁻]]$ is either equal to  $[[ [ausr1]β̂⁻]]$ or $[[ [ausr2]β̂⁻]]$,
+            it inherits their property that it is uniquely determined by $[[ [aus'1]β̂⁻]]$, $[[ [aus'2]β̂⁻]]$, and $[[Γ]]$.
+        \end{itemize}
+
+        \item $[[iP1]] = [[iP2]] = [[a⁺]]$. This case is symmetric to \cref{case:anti-unification-initial:nvar}.
+        \item $[[iP1]] = [[↓iN1]]$ and $[[iP2]] = [[↓iN2]]$. This case is symmetric to \cref{case:anti-unification-initial:shiftu}
+        \item $[[iP1]] = [[∃nas.iP'1]]$ and $[[iP2]] = [[∃nas.iP'2]]$. This case is symmetric to \cref{case:anti-unification-initial:forall}
+        \end{caseof}
+\end{proof}

### 2023-09-06 proofs/tex/constraint-merge-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
+\begin{observation}[Constraint Entry Merge is Deterministic]
+    \label{obs:entry-merge-deterministic}
+    For fixed $[[Γ]]$, $[[scE1]]$, $[[scE2]]$,
+    if $[[Γ ⊢ scE1 & scE2 = scE]]$ and $[[Γ ⊢ scE1 & scE2 = scE']]$ 
+    then $[[scE]] = [[scE']]$.
+\end{observation}
+\begin{proof}
+    First, notice that the shape of $[[scE1]]$ and $[[scE2]]$
+    uniquely determines the rule applied to infer  
+    $[[Γ ⊢ scE1 & scE2 = scE]]$,
+    which is consequently, the same rule used to 
+    infer $[[Γ ⊢ scE1 & scE2 = scE']]$.
+    Second, notice that the premises of each rule are deterministic
+    on the input:
+    the positive subtyping is deterministic by \cref{obs:pos-subt-deterministic},
+    and the least upper bound is deterministic by \cref{obs:lub-deterministic}.
+\end{proof}
+
+\begin{observation}[Subgyping Constraint Merge is Deterministic]
+    \label{obs:subt-merge-deterministic}
+    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$
+    If $[[Θ ⊢ SC1 & SC2 = UC]]$ and $[[Θ ⊢ SC1 & SC2 = UC']]$ are defined then 
+    $[[SC]] = [[SC']]$.
+\end{observation}
+\begin{proof}
+    The proof is analogous to the proof of \cref{obs:unif-merge-deterministic} 
+    but uses \cref{obs:entry-merge-deterministic} to show 
+    that the merge of the matching constraint entries is fixed.
+\end{proof}
+
 \begin{lemma} [Soundness of Constraint Entry Merge]
@@
         \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≈ iQ2]]$.
-            Then $[[Γ ⊢ iP : scE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
-            and $[[Γ ⊢ iP : scE2]]$ means $[[Γ ⊢ iP ≈ iQ2]]$.
-            Then by transitivity of equivalence (\cref{corollary:equivalence-transitivity}),
-            $[[Γ ⊢ iQ1 ≈ iQ2]]$, which means $[[nf(iQ1) = nf(iQ2)]]$ by
-            \cref{lemma:subt-equiv-algorithmization}.
-            Hence, \ruleref{\ottdruleSCMEPEqEqLabel} applies to infer
-            $[[Γ ⊢ scE1 & scE2 = scE2]]$, and $[[Γ ⊢ iP : scE2]]$ holds by assumption.
+            The proof repeats the corresponding case of \cref{lemma:unif-entry-merge-completeness}
         \item $[[scE1]]$ is $[[pua :≈ iQ1]]$ and $[[scE2]]$ is $[[pua :≥ iQ2]]$.
@@
 \end{proof}
-
-\begin{corollary} [Completeness of Unification Constraint Merge] 
-    \label{corollary:merge-completeness-unif}
-    Suppose that $[[Θ ⊢ UC1 : Ξ1]]$ and $[[Θ ⊢ UC2 : Ξ2]]$.
-    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
-    $[[ Θ ⊢ uσ : UC1 ]]$ and $[[ Θ ⊢ uσ : UC2 ]]$
-    then $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined.
-\end{corollary}

### 2023-09-06 proofs/tex/singularity-lemmas.tex
Grammar update: Γ = set
Add deterinicity lemmas, get rid of (??)

@@
                 \item $[[ord {nas} in iN' = nas]]$
-                \item for some substitution $[[Γ, nbj ⊢ σ : nas]]$, 
+                \item for some substitution $[[Γ, nbj ⊢ σ :{nas}]]$, 
                     $[[ [σ]iN' = nbj ]]$.
@@
 \begin{proof}
-    Suppose that $(Θ  ⊢ SC) ⊢ uσ'$.
+    Suppose that $[[Θ ⊢ uσ' : SC]]$.
     It means that for every $[[scE]] \in [[SC]]$ restricting $[[α̂±]]$,
@@
     $[[uσ]]$ is equivalent to $[[uσ']]$ on this set.
+\end{proof}
+
+\begin{observation}[Singularity is Deterministiic]
+    For a fixed $[[SC]]$ such that $[[Θ ⊢ SC : Ξ]]$,
+    if $[[SC singular with uσ]]$ and $[[SC singular with uσ']]$,
+    then $[[uσ = uσ']]$.
+\end{observation}
+\begin{proof}
+    By \label{lemma:singularity-soundness},
+    $[[Θ ⊢ uσ : Ξ]]$ and $[[Θ ⊢ uσ' : Ξ]]$.
+    It means that both $[[uσ]]$ and $[[uσ']]$
+    act as identity outside of $[[Ξ]]$.
+
+    Moreover, for any $[[α̂± ∊ Ξ]]$,
+    $[[Θ ⊢ SC : Ξ]]$ means that 
+    there is a unique $[[scE ∊ SC]]$ restricting $[[α̂±]]$.
+    Then $[[SC singular with uσ]]$ means 
+    that $[[scE singular with [uσ]α̂± ]]$.
+    By looking at the inference rules, it is easy to see that
+    $[[ [uσ]α̂± ]]$ is uniquely determined by $[[scE]]$, which, 
+    Similarly, $[[ [uσ']α̂± ]]$ is also uniquely determined by $[[scE]]$,  
+    in the same way, and hence, $[[ [uσ]α̂± = [uσ']α̂± ]]$.
 \end{proof}

### 2023-09-06 proofs/tex/unif-constraint-merge-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
+\begin{observation} [Unification Constraint Merge is Deterministic]
+    \label{obs:unif-merge-deterministic}
+    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
+    If $[[Θ ⊢ UC1 & UC2 = UC]]$ and $[[Θ ⊢ UC1 & UC2 = UC']]$ are defined then 
+    $[[UC]] = [[UC']]$.
+\end{observation}
+\begin{proof}
+    $[[UC]]$ and $[[UC']]$ both consists of three parts: 
+    Entries of $[[UC1]]$ that do not have matching entries in $[[UC2]]$,
+    entries of $[[UC2]]$ that do not have matching entries in $[[UC1]]$,
+    and the merge of matching entries.
+
+    The parts corresponding to unmatched entries of $[[UC1]]$ and $[[UC2]]$ coincide, 
+    since $[[UC1]]$ and $[[UC2]]$ are fixed.
+    To show that the merge of matching entries coincide,
+    let us take any pair of matching $[[ucE1 ∊ UC1]]$ and $[[ucE2 ∊ UC2]]$
+    and consider their shape.
+    \begin{caseof}
+        \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$
+            then the result, if it exists, is always $[[ucE1]]$,
+            by inversion of \ruleref{\ottdruleSCMEPEqEqLabel}.
+        \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iN2]]$
+            then analogously, the result, if it exists, is always $[[ucE1]]$,
+            by inversion of \ruleref{\ottdruleSCMENEqEqLabel}.
+    \end{caseof}
+    This way, the third group of entries coincide as well.
+\end{proof}
+
 \begin{lemma} [Soundness of Unification Constraint Merge]
@@
                 and then since $[[Θ(α̂⁺) ⊢ ucE1 & ucE2 = ucE]]$ exists, 
-                \ruleref{\ottdruleUCMEPEqEqLabel} was applied to infer it.
+                \ruleref{\ottdruleSCMEPEqEqLabel} was applied to infer it.
                 It means that $[[ucE]] = [[ucE1]] = [[ucE2]]$;
@@
             since $[[UC1 & UC2]]$ exists, 
-            \ruleref{\ottdruleUCMEPEqEqLabel} was applied to construct
+            \ruleref{\ottdruleSCMEPEqEqLabel} was applied to construct
             $[[ucE1 & ucE2]] \in [[UC1 & UC2]]$.
-            By inversion of \ruleref{\ottdruleUCMEPEqEqLabel},
+            By inversion of \ruleref{\ottdruleSCMEPEqEqLabel},
             $[[ucE1 & ucE2]] = [[ucE1]]$, and
@@
 \begin{proof}
-    The proof repeats the one of \cref{lemma:entry-merge-completeness}
-    and is done by the case analysis on the shape of $[[ucE1]]$ and $[[ucE2]]$.
-    However, it only needs to consider two cases.
+    Let us consider the shape of $[[ucE1]]$ and $[[ucE2]]$.
     \begin{caseof}
         \item $[[ucE1]]$ is $[[pua :≈ iQ1]]$ and $[[ucE2]]$ is $[[pua :≈ iQ2]]$.
+            Then $[[Γ ⊢ iP : ucE1]]$ means $[[Γ ⊢ iP ≈ iQ1]]$, 
+            and $[[Γ ⊢ iP : ucE2]]$ means $[[Γ ⊢ iP ≈ iQ2]]$.
+            Then by transitivity of equivalence (\cref{corollary:equivalence-transitivity}),
+            $[[Γ ⊢ iQ1 ≈ iQ2]]$, which means $[[nf(iQ1) = nf(iQ2)]]$ by
+            \cref{lemma:subt-equiv-algorithmization}.
+            Hence, \ruleref{\ottdruleSCMEPEqEqLabel} applies to infer
+            $[[Γ ⊢ ucE1 & ucE2 = ucE2]]$, and $[[Γ ⊢ iP : ucE2]]$ holds by assumption.
         \item $[[ucE1]]$ is $[[nua :≈ iN1]]$ and $[[ucE2]]$ is $[[nua :≈ iM2]]$.
+            The proof is symmetric.
     \end{caseof}
-    The proof of these cases is based only on \cref{lemma:subt-equiv-algorithmization}
-    and \cref{corollary:equivalence-transitivity}, and does not require the properties of 
-    the least upper bound or subtyping.
 \end{proof}
@@
 \begin{proof}
-    The proof repeats the proof of \cref{lemma:merge-completeness} but 
+    The proof repeats the proof of \cref{lemma:merge-completeness}
+    for cases 
     uses \cref{lemma:unif-entry-merge-completeness} instead of \cref{lemma:entry-merge-completeness}.

### 2023-09-06 proofs/tex/unification-lemmas.tex
Grammar update: Γ = set
Add deterinicity lemmas, get rid of (??)

@@
+\begin{observation}[Unification is deterministic]
+    \label{obs:unif-deterministic}
+    \hfill
+    \begin{itemize}
+        \item [$+$] If $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$
+            and $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC']]$ then $[[UC = UC']]$.
+        \item [$-$] If $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$
+            and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$ then $[[UC = UC']]$.
+    \end{itemize}
+\end{observation}
+\begin{proof}
+    We prove it by mutual structural induction on 
+    $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$.
+    Let us consider the positive case, since the negative case is symmetric.
+
+    First, notice that the rule applied the last is uniquely determined by the
+    shape of $[[uP]]$ and $[[iQ]]$.
+    Second, the premises of each rule are
+    deterministic on the input either by the induction 
+    hypothesis or by \cref{obs:unif-merge-deterministic}.
+\end{proof}
+
 \begin{lemma}[Soundness of Unification] \label{lemma:unification-soundness}
@@
 
-            Let us take an arbitrary $[[Θ ⊢ uσ : α̂⁻ ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
-            $[[Θ ⊢ uσ : α̂⁻]]$ means that $[[Δ ⊢ iM]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : {α̂⁻} ]]$ such that $[[ [uσ]α̂⁻ = iM ]]$.
+            $[[Θ ⊢ uσ : {α̂⁻}]]$ means that $[[Δ ⊢ iM]]$.
             This way, \ruleref{\ottdruleUNUVarLabel} is applicable to infer 

### 2023-09-06 proofs/tex/upgrade-lemmas.tex
Add deterinicity lemmas, get rid of (??)

@@
 
+\begin{observation}[Upgrade is deterministic]
+    \label{obs:upgrade-deterministic}
+    Assuming $[[iP]]$ is well-formed in $[[Γ ⊆ Δ]],$\\
+    if $[[upgrade Γ ⊢ iP to Δ = iQ]]$ and $[[upgrade Γ ⊢ iP to Δ = iQ']]$ are defined 
+    then $[[iQ = iQ']]$.
+\end{observation}
+\begin{proof}
+    It follows directly from \cref{obs:lub-deterministic},
+    and the convention that the fresh variables are chosen by a fixed deterministic algorithm
+    (\cref{sec:fresh-selection}).
+\end{proof}
+
 \begin{lemma}[Soundness of Upgrade]\label{lemma:upgrade-soundness}
-    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]],$\\
+    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
     if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
@@
         [[fv iQ]] &\subseteq [[fv [pnbs/pnas]iP ∩ fv [pncs/pnas]iP]]
-                  &&\text{Since by \cref{lemma:fv-propagation}, 
+                  &&\text{since by \cref{lemma:fv-propagation}, 
                          $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$ and
@@
     The former means that the upgrade algorithm terminates and returns $[[iQ]]$.
-    The latter means that since both $[[iQ']]$ and $[[iQ]]$ are well-formed in $[[Δ]]$,
-    by \cref{tood}, $[[Δ ⊢ iQ' ≥ iQ]]$.
+    The latter means that since both 
+    $[[iQ']]$ and $[[iQ]]$ are well-formed in $[[Δ]]$ and $[[Γ]]$,
+    by \cref{lemma:subt-ctxt-irrelevance}, $[[Δ ⊢ iQ' ≥ iQ]]$.
 \end{proof}

### 2023-09-07 proofs/tex/decl-subtyping-lemmas.tex
Add rule labels

@@
           then if $[[pas = ·]]$, $[[Γ ⊢ iN ≤ iM]]$ holds immediately.
-          Otherwise, let us decompose  $iN$ as $[[∀pas'.iN']]$ where 
+          Otherwise, let us decompose  $[[iN]]$ as $[[∀pas'.iN']]$ where 
           $[[iN']]$ does not start with $[[∀]]$.
@@
           $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$ for some $[[Γ ⊢ iPs]]$.
-          let us decompose $iN$ as $[[∀pas'.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
+          let us decompose $[[iN]]$ as $[[∀pas'.iN']]$ where $[[iN']]$ does not start with $[[∀]]$.
           Then $[[Γ ⊢ [iPs/pas]∀pas'.iN' ≤ iM']]$ or, equivalently,

### 2023-09-10 proofs/tex/ord-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Add some missing proofs

@@
-\begin{definition} [Collision free bijection]
-  We say that a bijection $\mu : A \leftrightarrow B$ between sets of
-  variables is \textbf{collision free on sets} $P$ and $Q$ if and only if
-  \begin{enumerate}
-    \item $\mu(P \cap A) \cap Q = \emptyset$
-    \item $\mu(Q \cap A) \cap P = \emptyset$
-  \end{enumerate}
-\end{definition}
+\begin{observation}[Ordering is deterministic]
+  \label{obs:ord-deterministic}
+  If $[[ord varset in iN = ordVars1]]$ and $[[ord varset in iN = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
+  If $[[ord varset in iP = ordVars1]]$ and $[[ord varset in iP = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
+  This way, we can use $[[ord varset in iN]]$ and as a function on $[[iN]]$,
+  and $[[ord varset in iP]]$ as a function on $[[iP]]$.
+\end{observation}
+\begin{proof}
+  By mutual structural induction on $[[iN]]$ and $[[iP]]$.
+  Notice that the shape of the term $[[iN]]$ or $[[iP]]$
+  uniquely determines the last used inference rule,
+  and all the premises are deterministic on the input.
+\end{proof}
 
@@
 \begin{proof}
-  Straightforward mutual induction on 
-  $[[ ord varset in iN = ordVars ]]$ and $[[ ord varset in iP = ordVars ]]$
+  We prove it by mutual induction on 
+  $[[ ord varset in iN = ordVars ]]$ and $[[ ord varset in iP = ordVars ]]$.
+  The only non-trivial cases are 
+  \ruleref{\ottdruleOArrowLabel} and 
+  \ruleref{\ottdruleOForallLabel}.  
+  \begin{caseof}
+    \item \ruleref{\ottdruleOArrowLabel}  
+      Then the inferred ordering judgement has shape
+      $[[ord varset in iP → iN = ordVars1, (ordVars2 \ {ordVars1})]]$
+      and by inversion, 
+      $[[ord varset in iP = ordVars1]]$   
+      and 
+      $[[ord varset in iN = ordVars2]]$.
+
+      By definition of free variables, 
+      $[[varset ∩ fv iP → iN = varset ∩ fv iP ∪ varset ∩ fv iN]]$,
+      and since by the induction hypothesis 
+      $[[varset ∩ fv iP = {ordVars1}]]$ and
+      $[[varset ∩ fv iN = {ordVars2}]]$,
+      we have
+      $[[varset ∩ fv iP → iN = {ordVars1} ∪ {ordVars2}]]$.
+
+      On the other hand, 
+      As a set, $[[{ordVars1} ∪ {ordVars2}]]$
+      is equal to $[[ordVars1, (ordVars2 \ {ordVars1})]]$. 
+    \item  \ruleref{\ottdruleOForallLabel}.
+      Then  the inferred ordering judgement has shape
+      $[[ord varset in ∀pas.iN = ordVars]]$,
+      and by inversion, 
+      $[[varset ∩ {pas} = ∅]]$    
+      $[[ord varset in iN = ordVars]]$.
+      The latter implies that $[[varset ∩ fv iN = {ordVars}]]$.
+      We need to show that $[[varset ∩ fv ∀pas.iN = {ordVars}]]$,
+      or equivalently, that
+      $[[varset ∩ (fv iN \ {pas}) = varset ∩ fv iN ]]$,
+      which holds since $[[varset ∩ {pas} = ∅]]$.
 \end{proof}
@@
   \begin{itemize}
-    \item[$-$] $[[ {ord (varset1 ∪ varset2) in iN} ]]
-                \equiv
-                [[{ord varset1 in iN} ∪ {ord varset2 in iN}]]$ (as sets)
-    \item[$+$] $[[{ord (varset1 ∪ varset2) in iP}]]
-                \equiv
-                [[{ord varset1 in iP} ∪ {ord varset2 in iP}]]$ (as sets)
+    \item[$-$] $[[ {ord (varset1 ∪ varset2) in iN} 
+                = 
+                {ord varset1 in iN} ∪ {ord varset2 in iN}]]$ (as sets)
+    \item[$+$] $[[{ord (varset1 ∪ varset2) in iP}
+                =
+                {ord varset1 in iP} ∪ {ord varset2 in iP}]]$ (as sets)
 
@@
 
-\begin{corollary}[Weakening of ordering]
+\begin{lemma}[Weakening of ordering]
   \label{corollary:ord-weakening}
-  Extending the first argument of the ordering with unused variables does not
-  change the result.
+  Only used variables matter in the first argument of the ordering,
   \begin{itemize}
@@
   \end{itemize}
-\end{corollary}
+\end{lemma}
+\begin{proof}
+  Mutual structural induction on $[[iN]]$ and $[[iP]]$.
+
+  \begin{caseof}
+    \item If $[[iN]]$ is a variable $[[na]]$,
+      we notice that $[[na ∊ varset]]$ 
+      is equivalent to $[[na ∊ varset ∩ {na}]]$.
+    \item If $[[iN]]$ has shape $[[↑iP]]$, then
+      the required property holds immediately by the 
+      induction hypothesis, since 
+      $[[fv(↑iP) = fv(iP)]]$.
+    \item If the term has shape $[[iP → iN]]$ then
+      \ruleref{\ottdruleOArrowLabel} was applied
+      to infer $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iP → iN ]]$
+      and $[[ ord varset in iP → iN]]$. 
+      By inversion, the result of 
+      $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iP → iN ]]$
+      depends on 
+      $A = [[ ord (varset ∩ (fv iP ∪ fv iN)) in iP]]$
+      and 
+      $B = [[ ord (varset ∩ (fv iP ∪ fv iN)) in iN]]$.
+      The result of
+       $[[ ord varset in iP → iN]]$ 
+       depends on 
+      $X = [[ord varset in iP]]$ and
+      $Y = [[ord varset in iN]]$.
+
+      Let us show that that $A = B$ and $X = Y$, so the results are equal. 
+      By the induction hypothesis and set properties,
+      $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iP ]] = 
+       [[ ord (varset ∩ (fv iP ∪ fv iN)) ∩ fv(iP) in iP ]] = 
+       [[ ord varset ∩ fv(iP) in iP ]] = 
+       [[ ord varset in iP ]]$.
+      Analogously, 
+      $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iN ]] = 
+       [[ ord varset in iN ]]$.
+    \item If the term has shape $[[∀pas.iN]]$,
+      we can assume that $[[pas]]$ is disjoint
+      from $[[varset]]$,
+      since we operate on alpha-equivalence classes.
+      Then using the induction hypothesis,
+      set properties and \ruleref{\ottdruleOForallLabel}: 
+      $[[ord varset ∩ (fv(∀pas.iN)) in ∀pas.iN]] =
+       [[ord varset ∩ (fv(iN) \ {pas}) in iN]] =
+       [[ord varset ∩ (fv(iN) \ {pas}) ∩ fv(iN) in iN]] =
+       [[ord varset ∩ fv(iN) in iN]] =
+       [[ord varset in iN]]$.
+  \end{caseof}
+\end{proof}
 
@@
 
-\end{proof}
+Next, we define the notion of a collision free bijection.
+It allows us to deal with the situations when
+the domain and the range of the bijection overlap.
+Intuitively,
+
+\begin{definition} [Collision free bijection]
+  We say that a bijection $\mu : A \leftrightarrow B$ between sets of
+  variables is \textbf{collision free on sets} $P$ and $Q$ if and only if
+  \begin{enumerate}
+    \item $\mu(P \cap A) \cap Q = \emptyset$
+    \item $\mu(Q \cap A) \cap P = \emptyset$
+  \end{enumerate}
+\end{definition}
 
@@
   Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
+  Let us consider the rule inferring $[[iN ≈ iM]]$. 
+  \begin{caseof}
+    \item \ruleref{\ottdruleEOneNVarLabel}
+    \item \ruleref{\ottdruleEOneShiftULabel}
+    \item \ruleref{\ottdruleEOneArrowLabel}
+      Then the equivalence has shape $[[iP → iN ≈ iQ → iM]]$,
+      and by inversion, $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
+      They by the induction hypothesis,
+      $[[ord varset in iP]] = [[ord varset in iQ]]$ 
+      and $[[ord varset in iN]] = [[ord varset in iM]]$.
+      Since the resulting ordering for $[[iP → iN]]$ and $[[iQ → iM]]$
+      depend on the ordering of the corresponding components, 
+      which are equal, the results are equal.
+    \item \ruleref{\ottdruleEOneForallLabel}
+      Then the equivalence has shape $[[∀pas.iN ≈ ∀pbs.iM]]$.
+      and by inversion there exists 
+      $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$ such that
+      \begin{itemize}
+        \item $[[{pas} ∩ fv iM = ∅]]$ and 
+        \item $[[iN ≈ [mu] iM]]$
+      \end{itemize}
+
+      Let us assume that $[[varset]]$ is disjoint from 
+      $[[pas]]$ and $[[pbs]]$ 
+      (we can always alpha-rename the bound variables).
+      Then $[[ord varset in ∀pas.iN = ord varset in iN]]$, 
+      $[[ord varset in ∀pas.iM = ord varset in iM]]$
+      and by the induction hypothesis,
+      $[[ord varset in iN]] = [[ord varset in [mu]iM]]$.
+      This way, it suffices tho show  that 
+      $[[ord varset in [mu]iM = ord varset in iM]]$.
+      It holds by \cref{lemma:ord-sigma} since
+      $[[varset]]$ is disjoint form 
+      the domain and the codomain of 
+      $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$ 
+      by assumption.
+
+    \item The positive cases are proved symmetrically.
+  \end{caseof}
 \end{proof}

### 2023-09-10 proofs/tex/ord-lemmas.tex
Collision-free bijection

@@
 
-Next, we define the notion of a collision free bijection.
-It allows us to deal with the situations when
-the domain and the range of the bijection overlap.
-Intuitively,
+Next we make a set-theoretical observation
+that will be useful further.
+In general, any injective function (its image)
+distributes over set intersection.
+However, for convenience we allow the bijections
+on variables to be applied
+\emph{outside of their domains}
+(as identities), which may violate
+the injectivity. To deal with these cases, 
+we define a special notion of
+bijections collision-free on certain sets
+in such a way that
+a bijection that is collision-free on $P$ and $Q$,
+distributes over intersection of $P$ and $Q$.
 
@@
 \end{definition}
+
+\begin{observation}
+  Suppose that $\mu : A \leftrightarrow B$ is a bijection between two sets of variables,
+  and $\mu$ is collision free on $P$ and $Q$.
+  Then $\mu(P \cap Q) = \mu(P) \cap \mu(Q)$.
+\end{observation}
 

### 2023-09-10 proofs/tex/norm-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Add "straightforward" proofs

@@
+\begin{observation}[Normalization is deterministic]
+  \label{obs:norm-deterministic}
+  If $[[nf(iN) = iM]]$ and $[[nf(iN) = iM']]$ then $[[iM = iM']]$.
+  If $[[nf(iP) = iQ]]$ and $[[nf(iP) = iQ']]$ then $[[iQ = iQ']]$.
+  This way, we can use normalization as a function.
+\end{observation}
+\begin{proof}
+  By straightforward induction using \cref{obs:ord-deterministic}.
+\end{proof}
+
 \begin{lemma}
@@
   \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
+  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN = fv iM]]$ (as sets)
+  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP = fv iQ]]$ (as sets)
   \end{itemize}
@@
 \begin{proof}
-  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
+  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
+  The base cases 
+  (\ruleref{\ottdruleEOneNVarLabel} and \ruleref{\ottdruleEOnePVarLabel})
+  are trivial.
+  So are \ruleref{\ottdruleEOneShiftULabel}, 
+  \ruleref{\ottdruleEOneShiftDLabel}, and
+  \ruleref{\ottdruleEOneArrowLabel}, where
+  the required property follows from the induction hypothesis.
+
+  Let us consider the case when the equivalence is formed by
+  \ruleref{\ottdruleEOneForallLabel}, 
+  that is the equivalence has shape 
+  $[[∀pas.iN ≈ ∀pbs.iM]]$, 
+  and by inversion,
+  there is a bijection 
+  $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$
+  such that $[[iN ≈ [mu] iM]]$, which 
+  by the induction hypothesis means
+  $[[ fv iN = fv [mu]iM ]] = [[ [mu] fv iM]] $.
+
+  Let us ensure by alpha-equivalence that 
+  $[[pas]]$ is disjoint from $[[fv iM]]$.
+  Then $[[ (fv ∀pbs.iM) \ {pas}]] = [[fv ∀pbs.iM]]$.
+  Then we apply the following chain of equalities:
+  $[[fv ∀pas.iN = fv iN \ {pas}]] = 
+  [[ ([mu] fv iM) \ {pas}]] = 
+  [[ [mu] (fv ∀pbs.iM ∪ ({pbs} ∩ fv iM)) \ {pas}]] = 
+  [[ ([mu] fv ∀pbs.iM ∪ [mu]({pbs} ∩ fv iM)) \ {pas}]] = 
+  [[ ([mu] fv ∀pbs.iM) \ {pas}]] = 
+  [[ (fv ∀pbs.iM) \ {pas}]] = 
+  [[  fv ∀pbs.iM ]]$.
+
+  Symmetrically, we prove the case when the equivalence is formed by
+  \ruleref{\ottdruleEOneExistsLabel}.
 \end{proof}
@@
   \begin{itemize}
-  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
-  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
+  \item[$-$] $[[fv iN = fv nf(iN)]]$
+  \item[$+$] $[[fv iP = fv nf(iP)]]$
   \end{itemize}
@@
 \begin{proof}
-  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
+  By mutual induction on $[[iN]]$ and $[[iP]]$.
+  The base cases 
+  (\ruleref{\ottdruleNrmNVarLabel} and \ruleref{\ottdruleNrmPVarLabel})
+  are trivial; the congruent cases
+  (\ruleref{\ottdruleNrmShiftULabel},
+  \ruleref{\ottdruleNrmShiftDLabel}, and
+  \ruleref{\ottdruleNrmArrowLabel}) are proved by the induction hypothesis.
+
+  Let us consider the case when the term is formed by $[[∀]]$,
+  that is the normalization judgement has shape 
+  $[[nf(∀pas.iN) = ∀pas'.iN']]$,
+  where by inversion $[[nf(iN) = iN']]$
+  and $[[ord {pas} in iN' = pas']]$.
+  By the induction hypothesis, $[[fv iN = fv iN']]$.
+  Since $[[fv(∀pas.iN) = fv iN \ {pas}]]$,
+  and $[[fv(∀pas'.iN') = fv iN' \ {pas'}]]$,
+  it is left to show that $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
+  By \cref{lemma:ord-completeness}, 
+  $[[{pas'}]] = [[{pas} ∩ fv iN']] = [[{pas} ∩ fv iN]]$.
+  Then
+  $[[fv iN \ {pas} = fv iN \ ({pas} ∪ fv iN)]]$ by
+  set-theoretic properties, and
+  thus, $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
+
+  The case when the term is positive and formed by $[[∃]]$ is symmetric.
 \end{proof}

### 2023-09-10 proofs/tex/subst-lemmas.tex
Add "straightforward" proofs

@@
 \begin{proof}
-  By straightforward mutual induction on $[[iP]]$ and $[[iN]]$.
+  By mutual induction on $[[iP]]$ and $[[iN]]$.
+  The base cases (when $[[iP]]$ or $[[iN]]$ is a variable) 
+  are trivial, since then $[[α± ∊ fv(iP)]]$ means 
+  $[[α± = iP]]$ (and symmetrically for $[[iN]]$).
+  The congruent cases 
+  (when the type is formed by $[[↓]]$, $[[↑]]$, or $[[→]]$)
+  hold since $[[α±]]$ occurs in type means that it occurs
+  in one of its parts, to which we apply the induction hypothesis.
+
+  Let us suppose that te type is $[[∃nas.iQ]]$.
+  Then $[[α± ∊ fv(∃nas.iQ)]]$ means
+  $[[α± ∊ fv(iQ)]]$ and $[[α± ∉ {nas}]]$.
+  Then by the induction hypothesis, 
+  $[[ fv([σ]α±) ⊆ fv([σ]iQ) ]]$,
+  and it is left to notice that $[[ fv([σ]α±) ∩ {nas} = ∅ ]]$,
+  which we can ensure by alpha-equivalence.
 \end{proof}

### 2023-09-14 proofs/main.tex
Add rule labels
WIP: add comments

@@
 
-\renewcommand{\ottdruleONVarInName}[0]{(Var$_{\in}^-$)}
-\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{\notin}^-$)}
-\renewcommand{\ottdruleONUVarName}[0]{(UVar$^-$)}
-\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow$)}
-\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow$)}
-\renewcommand{\ottdruleOForallName}[0]{($\forall$)}
+\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
+\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
+\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
+\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
+\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
+\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}
 
@@
 
-\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{\in}^+$)}
-\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{\notin}^+$)}
-\renewcommand{\ottdruleOPUVarName}[0]{(UVar$^+$)}
-\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow$)}
-\renewcommand{\ottdruleOExistsName}[0]{($\exists$)}
+\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
+\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
+\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
+\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
+\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}
 
 % nf(N) = M
-\renewcommand{\ottdruleNrmNVarName}[0]{(Var$^-$)}
-\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$^-$)}
-\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow$)}
-\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow$)}
-\renewcommand{\ottdruleNrmForallName}[0]{($\forall$)}
+\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}
 
 % nf(P) = Q
-\renewcommand{\ottdruleNrmPVarName}[0]{(Var$^+$)}
-\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$^+$)}
-\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow$)}
-\renewcommand{\ottdruleNrmExistsName}[0]{($\exists$)}
+\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
+\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}
 
@@
 
-\renewcommand{\ottdruleEOneNVarName}[0]{(Var$^-$$^{\eqEOne}$)}
+\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
 \renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
@@
 % P ≈ Q
-\renewcommand{\ottdruleEOnePVarName}[0]{(Var$^+$$^{\eqEOne}$)}
+\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
 \renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
@@
 
-\renewcommand{\ottdruleDOneNVarName}[0]{(Var$^-$$^{\subDOne}$)}
+\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
 \renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
@@
 % G ⊢ P ≥1 Q
-\renewcommand{\ottdruleDOnePVarName}[0]{(Var$^+$$^{\supDOne}$)}
+\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
 \renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
@@
 % G ⊢ N ≈1 M
-\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne^{-}$)}
+\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}
 
 % G ⊢ P ≈1 Q
-\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne^{+}$)}
+\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}
 
@@
 % G ; Θ ⊨ uN ≤ iM ⫤ us
-\renewcommand{\ottdruleANVarName}[0]{(Var$^-$$^{\subA}$)}
+\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
 \renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
@@
 % G ; Θ ⊨ iP ≥ uQ ⫤ us
-\renewcommand{\ottdruleAPVarName}[0]{(Var$^+$$^{\supA}$)}
+\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
 \renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
@@
 % Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
-\renewcommand{\ottdruleUNVarName}[0]{(Var$^{-[[≈u]]}$)}
+\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
 \renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
@@
 \renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
-\renewcommand{\ottdruleUNUVarName}[0]{(UVar$^{-[[≈u]]}$)}
+\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}
 
 % Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
-\renewcommand{\ottdruleUPVarName}[0]{(Var$^{+[[≈u]]}$)}
+\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
 \renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
 \renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
-\renewcommand{\ottdruleUPUVarName}[0]{(UVar$^{+[[≈u]]}$)}
+\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}
 
 % G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
-\renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
+\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
 \renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
@@
 % G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
-\renewcommand{\ottdruleAUNVarName}[0]{(Var$^{-[[≈au]]}$)}
+\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
 \renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
@@
 \renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUAUName}[0]{(AU$^{-}$)}
+\renewcommand{\ottdruleAUAUName}[0]{(AU$_{-}$)}
+
+% Γ ⊢ iN 
+\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}
+
+% Γ ⊢ iP
+\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}
+
+% Γ ; Ξ ⊢ uN
+\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}
+
+% Γ ; Ξ ⊢ uP
+\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
+\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}
+
+% Γ ⊢ iP : scE
+\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
+\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}
+
+% Γ ⊢ iN : scE
+\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}
+
+% Γ ; Φ ⊢ v : iP 
+\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
+\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
+\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
+\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}
+
+% Γ ; Φ ⊢ c : iN 
+\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
+\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
+\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
+\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
+\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
+\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
+\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
+\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
+\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}
+
+% Γ ; Φ ⊢ iN ● args ⇒> iM 
+\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
+\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
+\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
+
+% Γ ; Φ ⊨ v : iP  
+\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
+\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
+\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
+
+% Γ ; Φ ⊨ c : iN 
+\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
+\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
+\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
+\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
+\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
+\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
+\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
+\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
+
+% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
+\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
+\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
+\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
+
+% scE1 singular with iP
+\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
+\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
+\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}
+
+% scE1 singular with iN
+\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}
 
@@
 
+% \ottgrammartabular{
+%   \ottiP\ottinterrule
+%   \ottiN\ottinterrule
+% }
+
+\begin{definition}[Declarative Types]
+  \hfill
+  \begin{multicols}{2}
     \ottgrammartabular{
-  \ottiP\ottinterrule
       \ottiN\ottinterrule
@@
 
+    \ottgrammartabular{
+      \ottiP\ottinterrule
+    }
+    \columnbreak
+  \end{multicols}
+\end{definition}
+
 \subsection{Equalities}
-For simplicity, assume alpha-equivalent terms equal. 
+For simplicity, we assume alpha-equivalent terms equal.
 This way, we assume that substitutions do not capture bound variables.
@@
 
-\hfill\\
-\ottdefnsWFT
+Notice that checking the well-formedness of a type
+is an \emph{algorithmic} procedure, in which 
+both the context and the type are considered inputs.
+In other words, it is syntax-directed and mode-correct 
+(according to \cite{dunfieldBidirectionalTyping2020}), 
+which means that
+
+\begin{algorithm}[Type Well-formedness]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnWFTNiWF{}
+  \columnbreak
+
+  \ottdefnWFTPiWF{}
+  \end{multicols}
+
+\end{algorithm}
 
@@
 
-\begin{definition}[Substitution Application]
-  Substitution application (denoted as $[[ [σ]iP ]]$ or $[[ [σ]iN ]]$) 
-  is defined congruently as follows:
+\begin{algorithm}[Substitution Application]
+  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
+  It is defined naturally as follows:
+    \begin{multicols}{2}
       \begin{itemize}
-    \item $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
-    \item $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
-    \item $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
-    \item $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
-    \item $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
-    \item $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$(here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
+        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
+        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
+        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
+        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
+        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
+        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
+        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
           that is the variable capture never happens);
-    \item $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
       \end{itemize}
-\end{definition}
+    \end{multicols}
+\end{algorithm}
 
@@
       Suppose that $[[nas]]$ is a list of negative type variables,
-      and $[[iNs]]$ is a list of negative types of the same length 
+      and $[[iNs]]$ is a list of negative types of the same length.
       Then $[[iNs / nas]]$ denotes a substitution such that 
@@
 \subsection{Declarative Subtyping}
-\ottdefnsDOne
+Subtyping is one of the key mechanism of our system. 
+It realizes the polymorphism: abstract $[[∀]]$ and 
+$[[∃]]$ types can be used where concrete types are expected,
+exactly because the subtyping relation between them.
 
-\subsection{Declarative Equivalence}
-\ottdefnsEOne
+\begin{definition} 
+  \label{def:subDOne}
+  \hfill
+
+  \begin{multicols}{2}
+    \ottdefnDOneNsub{}
+    \columnbreak\\
+    \ottdefnDOnePsup{}
+  \end{multicols}
+
+  \begin{multicols}{2}
+    \ottdefnDOneNeq{}
+    \columnbreak\\
+    \ottdefnDOnePeq{}
+  \end{multicols}
+\end{definition}
+
+The following observations about the declarative subtyping are worth noting:
+\begin{itemize}
+  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
+    make the subtyping reflexive on variables (and further, on any type).
+  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
+    resulting type and contravariant on the argument type.
+  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
+    the subtyping is \emph{invariant} for shifts. 
+    This way, the subtyping of shifted types in one direction implies the subtyping
+    in the opposite direction.
+    Although this rule restricts the
+    subtyping relation, it makes the system decidable.
+  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
+    non-algorithmic rules: the substitution for the quantified variable is
+    not specified, those, these rules `drive' the subtyping relation.
+\end{itemize}
+
+In the next section, we present the sound and complete algorithm
+checking whether one type is a subtype of another according to \cref{def:subDOne}.
 
@@
 \begin{definition}[Algorithmic Types]
-  \hfill \\
+  \label{def:algo-types}
+  \hfill\\
+  \begin{multicols}{2}
+    \ottgrammartabular{
+      \ottuN\ottinterrule
+    }
+
     \ottgrammartabular{
       \ottuP\ottinterrule
-  \ottuN\ottinterrule
     }
+    \columnbreak
+  \end{multicols}
 \end{definition}
@@
 rely on the ability to select fresh, unused variables.
-In the inference rules, it is indicated as 
-$[[{varset} are fresh]]$. 
+For a set of variables $[[varset]]$, it is indicated as 
+$[[varset are fresh]]$ in the inference rules.
 We assume that the selection subroutine always succeeds and is 
@@
 \subsection{Variable Algorithmization}
+\label{sec:variable-algorithmization}
 
-In several places of our algorithm, in particular, during anti-unification,
+In several places of our algorithm, in particular, during
+algorithmic subtyping,
 we turn a declarative type into the algorithmic one
 via replacing certain type variables with fresh algorithmic variables.
+We call this procedure \emph{variable algorithmization}, and define it as follows.
 
@@
   and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution-like procedure turning each $[[αi⁻ ∊ {nas}]]$ 
-  into $[[αî⁻ ∊ {nuas}]]$.
+  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
+  in a type for $[[αî⁻ ∊ {nuas}]]$.
 \end{definition}
 
-Conversely, we can turn algorithmic type variables
+Conversely, we have the opposite procedure turning algorithmic type variables
 into declarative type variables via \emph{dealgorithmization}.
@@
   and $[[nas]]$ is a list of negative type variables of the same length. 
-  Then $[[ nas/nuas ]]$ is a substitution-like procedure turning each
-  $[[αî⁻ ∊ {nuas}]]$ into $[[αi⁻ ∊ {nas}]]$.
+  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
+  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
 \end{definition}
@@
 
-\begin{definition}[Algorithmic Type Context]
+\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
   \hfill \\
@@
 
-\hfill\\
-\ottdefnsWFAT
+\begin{algorithm}[Algorithmic Type Well-formedness]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnWFATNauWF{}
+  \columnbreak
+
+  \ottdefnWFATPauWF{}
+  \end{multicols}
+
+\end{algorithm}
 
@@
 This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context]
+\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
   \hfill \\
@@
   $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
-
   If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
@@
 $[[uσ]]$. It is defined as a total function from algorithmic 
-type variables to non-algorithmic types, preserving the polarity.
+type variables to \emph{declarative} types, preserving the polarity.
 
@@
 \begin{definition}[Signature of Algorithmic Substitution]
+  \label{def:algo-subst-sig}
   \hfill
@@
 
-Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.
+In the anti-unification algorithm, we use another kind of substitution.
 In contrast to algorithmic substitution $[[uσ]]$,
 it allows mapping algorithmic variables to
-\emph{algorithmic} terms.
+\emph{algorithmic} types.
+Additionally, anti-unification substitution is restricted to the
+\emph{negative} segment of the language.
+Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a
 
 The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
-in which the results of the anti-unification substitutions 
-are formed, is fixed for the whole substitution.
+in which the results of an anti-unification substitution 
+are formed, is fixed for this substitution.
 This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
@@
 
-\subsection{Normalization}
+\subsection{Equivalence and Normalization}
 
-\subsubsection{Ordering}
-\ottdefnsOrder
+The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
+there are types that are subtypes of each other but not equal. 
+For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
+and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
+although these types are not alpha-equivalent.
+For the subtyping algorithm, it is crucial to be able to check whether
+two types are equivalent, without checking mutual subtyping. 
+For this purpose we define the normalization procedure, 
+which allows us to uniformly choose the representative type of the equivalence class.
+This way, the equivalence checking is reduced to normalization and equality checking.
 
-\subsubsection{Quantifier Normalization}
-\ottdefnsNrm
+For clarification of the proofs and better understanding of the system, 
+we introduce an intermediate relation---\emph{declarative equivalence}. 
+As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
+this relation is equivalent to the subtyping-induced equivalence, but does not 
+depend on it. Although this relation is not defined algorithmically, 
+it gives the intuition of what types our system considers equivalent.
+Specifically, in addition to \emph{alpha-equivalence}, 
+our system allows for \emph{reordering of adjacent quantifiers},
+and \emph{introduction/elimination of unused quantifiers}.
 
-We also define normalization of a substitution pointwise:
-\begin{definition}[Substitution Normalization]
+The non-trivial rules of the declarative equivalence are
+\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
+Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
+the recursive call on the body of the quantified type. 
+It will be covered formally in \cref{sec:decl-equiv-lemmas}.
+
+\begin{definition}[Declarative Type Equivalence]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnEOneNeq{}
+  \columnbreak\\
+  \ottdefnEOnePeq{}
+  \end{multicols}
+
+\end{definition}
+
+As the equivalence includes arbitrary reordering of quantified variables,
+the normalization procedure is needed to choose the canonical order.
+For this purpose, we introduce an auxiliary procedure---variable ordering. 
+Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
+in the order they appear in $[[iN]]$.
+
+\begin{algorithm}[Variable Ordering]
+  \label{alg:var-ordering}
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnONVar{}
+  \columnbreak\\
+  \ottdefnOPVar{}
+  \end{multicols}
+
+  Analogously, the variable can be ordered in 
+  an \emph{algorithmic} type ($[[ord varset in uP]]$ and 
+  $[[ord varset in uN]]$). In these cases, we treat the algorithmic variables
+  as if they were declarative variables.
+
+\end{algorithm}
+
+Next, we use the variable ordering in the normalization procedure. 
+Specifically, normalization recursively traverses the type, 
+and for each quantified case reorders the quantified variables in a 
+canonical order dictated by \cref{alg:var-ordering}, removing unused ones.
+
+\begin{algorithm}[Type Normalization]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnNrmNNorm{}
+  \columnbreak\\
+  \ottdefnNrmPNorm{}
+  \end{multicols}
+
+  Analogously,  we define normalization of algorithmic types 
+  by adding base cases:
+
+  \begin{multicols}{2}
+  \ottdefnNrmuNNorm{}
+  \columnbreak\\
+  \ottdefnNrmuPNorm{}
+  \end{multicols}
+
+\end{algorithm}
+
+\begin{algorithm}[Substitution Normalization]
   For a substitution $[[σ]]$, we define $[[nf(σ)]]$
   as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
-\end{definition}
+\end{algorithm}
 
-\subsection{Unification}
-\ottdefnsU
+The rest of this chapter is devoted to the
+central algorithm of the type system---the subtyping algorithm. 
+\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
+The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
+$A$ uses $B$ as a sub-procedure.
 
-\subsection{Algorithmic Subtyping}
-\ottdefnsA
+\begin{figure}[h]
+  \centering
+  \begin{tikzpicture}
+    [>=stealth',node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
+
+    % Define nodes
+    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
+    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
+    \node[left=2cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[right=2cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
+    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
+    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
+    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
+    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
+
+    % Define edges
+    \draw[->] (1) to (2);
+    \draw[->] (1) to (3);
+    \draw[->] (1) to (5);
+    \draw[->] (2) to (3);
+    \draw[->] (2) to (6);
+    \draw[->] (3) to (4);
+    \draw[->] (3) to (5);
+    \draw[->] (4) to (6);
+    \draw[->] (5) to (8);
+    \draw[->] (6) to (7);
+  \end{tikzpicture}  
+  \caption{Dependency graph of the subtyping algorithm}
+  \label{fig:subtyping-algo}
+\end{figure}
+
+\subsection{Subtyping}
+\label{sec:subtyping}
+
+Now, we present the subtyping algorithm itself.
+Although the algorithm is presented as a single procedure,
+is important for the structure of the proof that the positive subtyping algorithm
+does not invoke the negative one. This way, the correctness of the positive 
+subtyping will be proved independently, and used afterwards to prove the
+correctness of the negative subtyping.
+
+\begin{algorithm}[Subtyping]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnANsub{}
+  \columnbreak\\
+  \ottdefnAPsup{}
+  \end{multicols}
+\end{algorithm}
+
+The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
+the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
+must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
+and $[[uP]]$ and $[[iQ]]$ for the positive case. 
+As one of the invariants, we require
+$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
+The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
+which will be discussed in the next section.
+
+Let us overview the inference rules of the subtyping algorithm.
+\begin{itemize}
+  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
+    are the base cases. They copy the corresponding declarative rules and
+    ensure the reflexivity.
+  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
+    subtyping constraints. In this case, we must ensure
+    that the resulting constrains guarantees that the instantiation of 
+    $[[â⁺]]$ is a supertype of $[[iP]]$.
+    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
+    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
+    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
+    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
+
+    Notice that this rule does not have a negative counterpart. This is
+    because one one of the important invariants of the algorithm: 
+    in the negative subtyping, only positive algorithmic variables
+    can occur in the types.
+
+  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
+    \emph{shift} rules. According to the declarative system,
+    shifted subtyping requires equivalence. In presence of the algorithmic 
+    variables, it means that the left and the right-hand sides of the subtyping
+    must be unified. Hence, the shift rules invoke the unification algorithm, 
+    which will be discussed in \cref{sec:unification}. The unification 
+    returns the minimal set of constraints $[[UC]]$, which is necessary
+    and sufficient for the subtyping.
+
+  \item \ruleref{\ottdruleAArrowLabel}.
+    In this case, the algorithm makes two calls:
+    a recursive call to the negative subtyping algorithm for the argument types,
+    and a call to the positive subtyping algorithm for the result types.
+    After that, the resulting constraints are merged using the
+    \emph{subtyping constraint merge} procedure, 
+    which is discussed in \cref{sec:constraint-merge}.
+  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
+    are symmetric. These are the only places where 
+    the algorithmic variables are introduced.
+    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
+    of the quantified variables: these variables are replaced by 
+    fresh algorithmic variables in the body of the quantified type,
+    the algorithmic variables are added to the subtyping context $[[Θ]]$,
+    after that the recursive call is made. Notice that the declarative context
+    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
+    which matches the declarative system.
+\end{itemize}
+
+Then soundness lemma (\cref{lemma:pos-subt-soundness,lemma:neg-subt-soundness}) 
+and completeness (\cref{lemma:pos-subt-completeness,lemma:neg-subt-completeness})
+of the algorithm together give us the following simplified theorem:
+
+\begin{theorempreview}[Correctness of subtyping algorithm]
+  \hfill
+  \begin{itemize}
+    \item [$-$]  $[[ Γ ; · ⊨ uN ≤ iM ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iN ≤ iM ]]$;
+    \item [$+$] $[[ Γ ; · ⊨ uP ≥ iQ ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iP ≥ iQ ]]$.
+  \end{itemize}
+\end{theorempreview}
 
@@
   \hfill
-  \begin{itemize}
-    \item Unification entry (denoted as $[[ucE]]$) is an expression of shape 
+  \begin{description}
+    \item[unification entry] (denoted as $[[ucE]]$) is an expression of shape 
       $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
-    \item unification constraint (denoted as $[[UC]]$) is a set of 
+    \item[unification constraint] (denoted as $[[UC]]$) is a set of 
       unification constraint entries.
-  \end{itemize}
+      We denote $\{[[α̂±]] \mid [[ucE ∊ UC]] \text{ restricting $[[α̂±]]$ }\}$ 
+      as $[[dom(UC)]]$.
+  \end{description}
 \end{definition}
 
-However, in subtyping we need to consider more general
+However, in the subtyping, we need to consider more general
 kind of constraints. Specifically,
@@
   \hfill
-  \begin{itemize}
-    \item Subtyping entry (denoted as $[[scE]]$) is an expression of shape 
+  \begin{description}
+    \item[subtyping entry] (denoted as $[[scE]]$) is an expression of shape 
       $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
-    \item subtyping constraint (denoted as $[[SC]]$) is a set of subtyping constraint entries.
-  \end{itemize}
+    \item[subtyping constraint] (denoted as $[[SC]]$) is a set of subtyping constraint entries.
+      We denote $\{[[α̂±]] \mid [[scE ∊ SC]] \text{ restricting $[[α̂±]]$ }\}$ 
+      as $[[dom(SC)]]$.
+  \end{description}
 \end{definition}
@@
   We say that a constraint entry is well-formed in a context $[[Γ]]$ if
-  the type it restricts the unification variable to is well-formed in $[[Γ]]$.
+  its associated type is well-formed in $[[Γ]]$.
   \begin{itemize}
-    \item $[[Γ ⊢ pua :≥ iP]]$ iff $[[Γ ⊢ iP]]$;
-    \item $[[Γ ⊢ pua :≈ iP]]$ iff $[[Γ ⊢ iP]]$;
-    \item $[[Γ ⊢ nua :≈ iN]]$ iff $[[Γ ⊢ iN]]$.
+    \item[] $[[Γ ⊢ pua :≥ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
+    \item[] $[[Γ ⊢ pua :≈ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
+    \item[] $[[Γ ⊢ nua :≈ iN]]$ ~iff~ $[[Γ ⊢ iN]]$.
   \end{itemize}
@@
   More formally, 
-  $[[Θ ⊢ UC]]$ iff for every $[[ucE]] \in $ $[[UC]]$,
-  such that $[[ucE]]$ restricts $[[α̂±]]$,
-  there exists $[[ α̂±[Γ] ]] \in [[Θ]]$
-  and $[[Γ ⊢ ucE]]$.
+  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
+  such that $[[scE]]$ restricts $[[α̂±]]$,
+  we have $[[Θ(α̂±) ⊢ scE]]$.
+
+  We write $[[Θ ⊢ SC : Ξ]]$ to denote
+  that $[[Θ ⊢ SC]]$ and $[[dom(SC) = Ξ]]$.
+
+  $[[Θ ⊢ UC]]$ and $[[Θ ⊢ UC : Ξ]]$ are defined analogously.
 \end{definition}
 
-Next, define the least upper bound 
-for two subtyping constraints.
-First, we define the least upper bound for entries,
+\subsubsection{Constraint Satisfaction}
+
+A constraint entry restricts a type that can be assigned to a variable.
+We say that a type satisfies a constraint entry if it can be assigned
+to the variable restricted by the entry.
+
+\begin{definition}[Type Satisfying a Constraint Entry]
+  \hfill\\
+  \begin{multicols}{2}
+  \ottdefnSATSCEN{}
+  \columnbreak\\
+  \ottdefnSATSCEP{}
+  \end{multicols}
+\end{definition}
+
+We say that a substitution satisfies a constraint---a set of constraint 
+entries if each entry is satisfied by the type assigned to the variable
+by the substitution.
+
+\begin{definition}[Substitution Satisfying a Constraint]
+  We write $[[Θ ⊢ uσ : SC]]$ to denote that
+  a substitution $[[uσ]]$ satisfies a constraint $[[SC]]$ in a context $[[Θ]]$.
+  It presumes that $[[Θ ⊢ SC]]$ and 
+   means that for any $[[ucE]] \in [[SC]]$, if $[[ucE]]$ restricts $[[α̂±]]$,
+  then $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$.
+
+  Unification constraint satisfaction $[[Θ ⊢ uσ : UC]]$ 
+  is defined analogously as a special case of subtyping constraint satisfaction.
+
+\end{definition}
+
+Notice that $[[Θ ⊢ uσ : SC]]$ does not 
+imply the signature $[[Θ ⊢ uσ : dom(SC)]]$, because 
+the latter also specifies $[[uσ]]$ outside of the domain $[[dom(SC)]]$
+(see \cref{def:algo-subst-sig}).
+
+\subsubsection{Constraint Merge}
+\label{sec:constraint-merge}
+
+In this section, define the least upper bound 
+for constraints, which we call \emph{merge}.
+Intuitively, the merge of two constraints is the least
+constraint such that any substitution satisfying both constraints
+satisfies the merge as well.
+First, we define the merge of entries,
 and then extend it to the set of entries.
@@
 can be merged in the following way:
-\begin{definition}[Merge of Matching Constraint Entries]
+\begin{algorithm}[Merge of Matching Constraint Entries]
   \label{definition:merge-matching-entries}
-   \hfill \\
+   \hfill
+
   \ottdefnSCME\\
-\end{definition}
+\end{algorithm}
 
@@
 
-\begin{definition}[Merge of Subtyping Constraints]
+\begin{algorithm}[Merge of Subtyping Constraints]
   \label{definition:merge-subtyping-constraints}
@@
   \end{itemize}
-\end{definition}
+\end{algorithm}
 
@@
 
-\subsection{Constraint Satisfaction}
-\ottdefnsSATSCE
+\Cref{lemma:merge-soundness,lemma:merge-completeness} 
+show the correctness and initiality of the merge operation,
+which can be expressed in the following simplified theorem:
+\begin{theorempreview}[Correctness of Constraint Merge]
+  A substitution $[[uσ]]$ satisfying both constraints
+  $[[SC1]]$ and $[[SC2]]$ 
+  if and only if it satisfies their merge.
+\end{theorempreview}
+
+The unification constraint merge satisfies the same theorem,
+however, because the merge of unification constraint entries 
+$[[ucE1]]$ and $[[ucE2]]$ always results in one of them, 
+a stronger soundness property holds (see \cref{lemma:unif-merge-soundness}):
+\begin{theorempreview}[Soundness of Unification Constraint Merge]
+  If $[[Θ ⊢ UC1 & UC2 = UC]]$ then $[[UC = UC1 ∪ UC2]]$.
+\end{theorempreview}
+
+\subsection{Unification}
+\label{sec:unification}
+
+The subtyping algorithm calls the following subtask:
+given two algorithmic types, we need to find the most general substitution 
+for the algorithmic variables in these types, such that the resulting 
+types are equivalent. This problem is known as \emph{unification}.
+
+In our case, the unification is restricted in the following way:
+first, before unifying the types, we normalize them, which 
+allows us to reduce (non-trivial) equivalence to (trivial) equality;
+second, we preserve invariants which guarantee that
+one side of the unification is always declarative, which in fact, 
+reduces the unification to the \emph{matching} problem.
+
+The unification procedure
+returns a set of minimal constraints,
+that must be satisfied by a substitution
+unifying the input types.
+
+\begin{algorithm}[Unification]
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnUNUnif{}
+  \columnbreak\\
+  \ottdefnUPUnif{}
+  \end{multicols}
+\end{algorithm}
 
 \subsection{Least Upper Bound}
+\label{sec:lub}
 \ottdefnsLUB
@@
 \subsection{Antiunification}
+\label{sec:antiunification}
 \ottdefnsAU
@@
 \subsection{Substitution}
-
 \input{../_build/_gen/tex/subst-lemmas.tex}
@@
 \input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
+
+\subsection{Equivalence}
+\label{sec:decl-equiv-lemmas}
+\input{../_build/_gen/tex/equiv-lemmas.tex}
 
@@
 \input{../_build/_gen/tex/norm-lemmas.tex}
-
-\subsection{Equivalence}
-\input{../_build/_gen/tex/equiv-lemmas.tex}
 
@@
 
+\printbibliography
+
 \end{document}

### 2023-09-14 proofs/tex/au-lemmas.tex
WIP: add comments

@@
         Then  $[[Ξ]] = [[â⁻_{iN1, iN2}]]$,
-        $[[aus1]] = [[â⁻_{iN1, iN2} :≈ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} :≈ iN2]]$.
+        $[[aus1]] = [[â⁻_{iN1, iN2} ↦ iN1]]$, and $[[aus2]] = [[â⁻_{iN1, iN2} ↦ iN2]]$.
         So the property holds trivially.
@@
         returning as the result $[[(Ξ, uM, aus1, aus2)]] = $
-        $[[({â⁻_{iN1, iN2}}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} :≈ iN1) ,  (â⁻_{iN1, iN2} :≈ iN2))]]$.
+        $[[({â⁻_{iN1, iN2}}, â⁻_{iN1, iN2}, (â⁻_{iN1, iN2} ↦ iN1) ,  (â⁻_{iN1, iN2} ↦ iN2))]]$.
 
@@
                 which holds trivially;
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} :≈ iNi) : {â⁻_{iN1, iN2}}]]$,
+            \item $[[Γ ; · ⊢ ausi : Ξ]]$ is rewritten as $[[Γ ; · ⊢ (â⁻_{iN1, iN2} ↦ iNi) : {â⁻_{iN1, iN2}}]]$,
                 which holds since $[[Γ ⊢ iNi]]$ by the premise of the rule;
-            \item $[[ [ausi] uM = iNi ]]$ is rewritten as $[[ [â⁻_{iN1, iN2} :≈ iNi] â⁻_{iN1, iN2} = iNi ]]$,
+            \item $[[ [ausi] uM = iNi ]]$ is rewritten as $[[ [â⁻_{iN1, iN2} ↦ iNi] â⁻_{iN1, iN2} = iNi ]]$,
                 which holds trivially by the definition of substitution.

### 2023-09-14 proofs/tex/decl-subtyping-lemmas.tex
WIP: add comments

@@
       is equivalent to $[[Γ, pas ⊢ iN ≤ iM]]$,
-      By \label{lemma:wf-ctxt-equiv},
+      By \cref{lemma:wf-ctxt-equiv},
       since $[[{pas} ∩ fv(iN) = ∅]]$ and $[[{pas} ∩ fv(iM) = ∅]]$,

### 2023-09-18 proofs/main.tex
Finish commenting subtyping

@@
 \renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUAUName}[0]{(AU$_{-}$)}
+\renewcommand{\ottdruleAUAUName}[0]{(AU)}
 
@@
 \subsection{Equivalence and Normalization}
+\label{sec:equivalence-normalization}
 
@@
 
+\Cref{lemma:subt-equiv-algorithmization}
+demonstrates that the equivalence of types is the same
+as the equality of their normal forms.
+\begin{theorempreview}[Correctness of Normalization]
+  Assuming the types are well-formed in $[[Γ]]$, 
+  \begin{itemize}
+    \item [$-$] $[[Γ ⊢ iN ≈ iM]]$ if and only if $[[nf(iN) = nf(iM)]]$;
+    \item [$+$] $[[Γ ⊢ iP ≈ iQ]]$ if and only if $[[nf(iP) = nf(iQ)]]$.
+  \end{itemize}
+\end{theorempreview}
+
 \begin{algorithm}[Substitution Normalization]
@@
   \begin{tikzpicture}
-    [>=stealth',node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
+    [>={Stealth[scale=2]},node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
 
@@
 \end{algorithm}
+
+\begin{itemize}
+  \item \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}
+    are symmetric cases. To merge two matching entries restricting
+    a variable to be equivalent to certain types, we check
+    that these types are equivalent to each other.
+    To do so, it suffices to check for \emph{equality} of their normal forms,
+    as discussed in \cref{sec:equivalence-normalization}. 
+    After that, we return the left-hand entry.
+
+  \item \ruleref{\ottdruleSCMEEqSupLabel} and \ruleref{\ottdruleSCMESupEqLabel}
+    are also symmetric. 
+    In this case,
+    since one of the entries requires the variable to be equal to 
+    a type, the resulting entry must also imply that.
+    However, for the soundness, it is needed to ensure that
+    the equating restriction is stronger than the subtyping restriction.
+    For this purpose, the premise invokes the positive subtyping.
+
+  \item \ruleref{\ottdruleSCMESupSupLabel} 
+    In this case, we find the least upper bound of the types from the input
+    restrictions, 
+    and as the output, restrict the variable to be a supertype of the result.
+    The least upper bound procedure will be discussed in \cref{sec:lub}.
+\end{itemize}
 
@@
   \hfill
-
   \begin{multicols}{2}
@@
 
+\begin{itemize}
+  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
+    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
+    are defined congruently. In the shift rules, the algorithm
+    removes the outermost constructor. In the
+    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
+    adding the quantified variables to the context $[[Γ]]$.
+    Notice that $[[Θ]]$, which specifies
+    the contexts in which the algorithmic variables must be instantiated,
+    is not changed.
+  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
+    are the base cases. 
+    Since the sides are equal and free from algorithmic variables,
+    the unification returns an empty constraint. 
+  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
+    are symmetric cases constructing the constraints. 
+    When an algorithmic variable is unified with a type, 
+    we must check that the type is well-formed in the required context,
+    and if it is, we return a constraint restricting the variable
+    to be equivalent to that type.
+  \item \ruleref{\ottdruleUArrowLabel}.
+    In this case, the algorithm makes two recursive calls:
+    it unifies the arguments and the results of the arrows.
+    After that, the resulting constraints are merged using the
+    \emph{unification constraint merge} procedure, 
+    which is discussed in \cref{sec:constraint-merge}.
+    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
+    \emph{unification} constraints, not arbitrary \emph{subtyping} 
+    constraints: it is important for modularizing the proofs, 
+    since the properties of the \emph{unification} constraint merge
+    can be proved independently from the \emph{subtyping} constraint merge.
+\end{itemize}
+
 \subsection{Least Upper Bound}
 \label{sec:lub}
-\ottdefnsLUB
-
-\subsection{Antiunification}
+
+In this section, we present
+the algorithm finding the least common supertype of two positive types. 
+It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
+and indirectly, through the type upgrade by positive subtyping
+(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
+intuitive part of the algorithm, and its correctness will be covered
+in \cref{sec:alg-upper-bounds-proofs}.
+
+\begin{algorithm}[The Least Upper Bound Algorithm]
+  \hfill\\
+  \ottdefnLUBNsub{}
+\end{algorithm}
+
+\begin{itemize}
+  \item \ruleref{\ottdruleLUBVarLabel}
+    The base case is trivial: 
+    the least upper bound of to equal variables
+    is the variable itself.
+  \item \ruleref{\ottdruleLUBShiftLabel}
+    In case both sides of the least upper bound are shifted,
+    the algorithm needs to find the anti-unifier of them. 
+    Intuitively, this is because in general, the upper bounds of
+    $[[↓iN]]$ are $[[∃nas.iP]]$ such that 
+    $[[nas]]$ can be instantiated with some $[[iMs]]$ so that
+    $[[ Γ ⊢ [iMs/nas]iP ≈ ↓iN ]]$ (see \cref{lemma:shape-of-supertypes}).
+  \item \ruleref{\ottdruleLUBExistsLabel}
+    In this case, we move the quantified variables to the context $[[Γ]]$, 
+    and make a recursive call. 
+    It is important to make sure that $[[nas]]$ and $[[nbs]]$ are disjoint.
+    In this case, it is guaranteed that the resulting 
+    $[[fv(iQ)]]$ will be free of $[[nas]]$ and $[[nbs]]$,
+    and thus, the resulting type will be a supertype of both sides
+    (it will be discussed in \cref{lemma:shape-of-supertypes}).
+\end{itemize}
+
+In the positive subtyping algorithm (\cref{sec:subtyping}),
+\ruleref{\ottdruleAPUVarLabel} 
+generates a restriction of a variable $[[α̂⁺]]$.
+On the one hand, this restriction must imply 
+$[[â⁺ :≥ iP]]$ for the subtyping to hold.
+On the other hand, the type used in this restriction 
+must be well-formed in a potentially stronger (smaller) 
+context than $[[iP]]$.
+
+To resolve this problem, we define the \emph{upgrade} procedure,
+which for given $[[Δ]]$, $[[pnas]]$, and $[[Δ, pnas ⊢ iP]]$,
+finds $[[Δ ⊢ iQ]]$---the least supertype of $[[iP]]$ 
+among the types well-formed in $[[Δ]]$.
+
+The trick is to make sure that the `forbidden' variables
+$[[pnas]]$ are not used explicitly in the supertypes
+of $[[iP]]$. For this purpose, we
+construct new types $[[iP1]]$ and $[[iP2]]$,
+in each of them replacing the forbidden variables
+with fresh variables $[[pnbs]]$ and $[[pncs]]$,
+and then find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
+It turns out that this renaming forces the common types of 
+$[[iP1]]$ and $[[iP2]]$ to be agnostic to $[[pnas]]$,
+and thus, the supertypes of $[[iP]]$ well-formed in $[[Δ]]$
+are exactly the common supertypes of $[[iP1]]$ and $[[iP2]]$.
+These properties are considered in more details in \cref{sec:upgrade-lemmas}.
+
+\begin{algorithm}[Type Upgrade]
+  \hfill\\
+  \ottdefnLUBUp{}
+\end{algorithm}
+
+\begin{paragraph}{Note on the Greatest Lower Bound}
+  In contrast to the least upper bound, 
+  the general greatest lower bound does not exist in our system.
+  For instance, consider a positive type $[[iP]]$, 
+  together with its non-equivalent
+  supertypes $[[iP1]]$ and $[[iP2]] \not\simeq [[iP1]]$
+  (for example, $[[iP = ↓↑↓γ⁻]]$, $[[iP1 = ∃α⁻.↓↑↓α⁻]]$, 
+  and $[[iP2 = ∃α⁻.↓α⁻]]$).
+  Then for arbitrary $[[iQ]]$ and $[[iN]]$, 
+  let us consider the common subtypes of 
+  $A = [[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $B = [[iP → ↓↑iP1 → ↓↑iP2 → iN]]$.
+  It is easy to see that $[[∀α⁺.∀β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and 
+  $[[∀α⁺.∀β⁺. α⁺ → ↓↑β⁺ → ↓↑α⁺ → iN]]$ are
+  both \emph{maximal} common subtypes of $A$ and $B$,
+  and since they are not equivalent, none of them is 
+  the \emph{greatest} one.
+
+  However, we designed the subtyping system in such a way 
+  that the greatest lower bound is not needed:
+  the negative variables are always `protected'
+  by \emph{invariant} shifts ($[[↑]]$ and $[[↓]]$), 
+  and thus, the algorithm can only require
+  a substitution of a negative variable to be 
+  \emph{equivalent} to some type but never 
+  to be a \emph{subtype}.
+\end{paragraph}
+
+\subsection{Anti-unification}
 \label{sec:antiunification}
-\ottdefnsAU
+
+Next, we define the anti-unification procedure,
+also known as the \emph{most specific generalization}.
+As an input, it takes two declarative types
+(e.g., in the positive case $[[iP1]]$ and $[[iP2]]$)
+and a context $[[Γ]]$.
+and returns a type $[[uQ]]$---the generalizer,
+containing negative placeholders 
+(represented by algorithmic variables) from 
+$[[Ξ]]$ and two substitutions $[[uτ1]]$ and $[[uτ2]]$.
+The substitutions replace the placeholders with 
+declarative types well-formed in $[[Γ]]$,
+such that $[[ [uτ1]uQ = iP1 ]]$ and $[[ [uτ2]uQ = iP2 ]]$.
+Moreover, the algorithm guarantees that 
+$[[uQ]]$ is the most specific type with this property: 
+any other generalizer can be turned into $[[uQ]]$ by some substitution
+$[[uρ]]$.
+
+It is important to note the differences between 
+the standard anti-unification and our version.
+First, we only allow the placeholders at \emph{negative} positions,
+which means, for example, that $[[α⁺]]$ and $[[β⁺]]$ cannot be
+generalized. Second, the generated pair of substitutions 
+$[[uτ1]]$ and $[[uτ2]]$ must replace the placeholders with 
+types well-formed in a specified context $[[Γ]]$.
+
+The anti-unification algorithm assumes that the input types 
+are normalized. This way, anti-unification up-to-equality rather than 
+ anti-unification up-to-equivalence is sufficient.
+
+\begin{algorithm}[Anti-unification]
+  \hfill
+
+  \ottdefnAUAUP{}
+  \ottdefnAUAUN{}
+
+\end{algorithm}
+
+\begin{itemize}
+  \item \ruleref{\ottdruleAUPVarLabel} and \ruleref{\ottdruleAUNVarLabel}
+    are the base cases. 
+    In this case, since the input types are equal, 
+    the algorithm returns this type as a generalizer,
+    without generating any placeholders.
+
+  \item \ruleref{\ottdruleAUShiftDLabel}, \ruleref{\ottdruleAUShiftULabel},
+    \ruleref{\ottdruleAUForallLabel}, and \ruleref{\ottdruleAUExistsLabel}
+    are defined congruently. In the shift rules, the algorithm
+    removes the outermost constructor. In the
+    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers.
+    Notice that the algorithm does not add the removed variables to
+    the context $[[Γ]]$. This is because $[[Γ]]$
+    is used to restrict the resulting anti-unification substitutions, 
+    and is fixed throughout the algorithm.
+
+  \item \ruleref{\ottdruleAUAULabel} is the most important rule, 
+    since it generates the placeholders. 
+    This rule only applies if other negative rules failed.
+    Because of that, the anti-unification procedure is 
+    \emph{not} syntax-directed.
+
+    The generated placeholder is indexed with a pair of 
+    types it is mapped to. It allows the algorithm to 
+    automatically unite the anti-unification solutions 
+    generated by the different branches of 
+    \ruleref{\ottdruleAUArrowLabel}.
+
+    Notice that this rule does not have a positive counterpart,
+    since we only allow negative placeholders.
+
+  \item \ruleref{\ottdruleAUArrowLabel}
+    makes two recursive calls to the anti-unification procedure,
+    and unites the results. Suppose that
+    $[[uτ1]]$  and $[[uτ2]]$ are the substitutions generated by
+    anti-unification of \emph{argument} types of the arrow,
+    and $[[uτ1']]$ and $[[uτ2']]$ are the substitutions generated by 
+    anti-unification of \emph{result} types of the arrow.
+    It is important that if ($[[uτ1]]$, $[[uτ2]]$)
+    and ($[[uτ1']]$, $[[uτ2']]$) send some variables to the same pair of types,
+    i.e., $[[ [uτ1]α̂⁻ = [uτ1']β̂⁻]]$ and $[[ [uτ2]α̂⁻ = [uτ2']β̂⁻]]$,
+    then these variables are equal, i.e., $[[α̂⁻ = β̂⁻]]$.
+    This property is guaranteed by \ruleref{\ottdruleAUAULabel}:
+    the name of the placeholder is determined by the pair of 
+    types it is mapped to.
+\end{itemize}
+
+\newpage
 
@@
 \subsection{Upper Bounds}
+\label{sec:alg-upper-bounds-proofs}
 \input{../_build/_gen/tex/lub-lemmas.tex}
@@
 \subsection{Upgrade}
+\label{sec:upgrade-lemmas}
 \input{../_build/_gen/tex/upgrade-lemmas.tex}

### 2023-09-18 proofs/tex/au-lemmas.tex
Finish commenting subtyping

@@
             \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification solutions, and
+            are anti-unification substitutions, and
             \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
@@
             \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification solutions, and
+            are anti-unification substitutions, and
             \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
@@
                         Then $[[Γ ; · ⊢ ausi ∪ aus'i : Ξ ∪ Ξ']]$
-                        are well-defined anti-unification solution.
+                        are well-defined anti-unification substitutions.
                         Let us take an arbitrary $[[β̂⁻]] \in [[Ξ ∪ Ξ']]$.
@@
 
-                        $[[ausi ∪ aus'i]]$ is a well-defined anti-unification solution:
+                        $[[ausi ∪ aus'i]]$ is a well-defined anti-unification substitution:
                         any anti-unification variable occurs 
@@
                 \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
-                are anti-unification solutions, and
+                are anti-unification substitutions, and
                 \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
@@
                 \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
-                are anti-unification solutions, and
+                are anti-unification substitutions, and
                 \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
@@
                 \item $[[Γ ; Ξ' ⊢ uQ']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
-                are anti-unification solutions , and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                 \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
@@
                 \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$
-                are anti-unification solutions, and
+                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                 \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$

### 2023-09-19 proofs/main.tex
Add Declarative Typing comments

@@
 
+In the previous section, we presented the 
+type system together with subtyping specification 
+and the algorithm. In this section, 
+we describe the language under this type system, 
+together with the type inference specification
+and algorithm.
+
 \subsection{Grammar}
 
-\bigskip
+First, we define the syntax of the language.
+The language combines System F with call-by-push-value 
+style. 
+\begin{definition}[Language Grammar]
+  \hfill
+  \begin{multicols}{2}
+    \ottgrammartabular{
+      \ottc\ottinterrule
+    }
+
     \ottgrammartabular{
       \ottv\ottinterrule
-  \ottc\ottinterrule
     }
+  \end{multicols}
+\end{definition}
+Notice that the language does not have
+first-class applications: instead, 
+we use applicative let bindings---
+constructions that bind a result of
+a fully applied function to a (positive) 
+variable.
+In the call-by-push-value paradigm, 
+it corresponds to monadic bind or do-notation. 
+Typewise, these let-binders come in two forms:
+annotated and unannotated. The annotated let-binders
+$[[let x:iP = v(args); c]]$ 
+requires the application to infer the annotated
+$[[iP]]$, whereas the unannotated 
+$[[let x = v(args); c]]$ 
+is used when the inferred type is unique.
+
+A computation of a polymorphic type is constructed 
+using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
+is implicit. Conversely, the existential types
+are constructed implicitly and eliminated 
+using the standard unpack mechanism:
+$[[let∃ (nas, x) = v; c]]$.
+
+Another dual pair of constructions are 
+$[[return v]]$ and $[[{c}]]$. 
+The former allows us to embed a value in a pure 
+computations. The latter, on the contrary,
+encapsulates a thunk of computation in a value.
+
+Finally, the language has a number standard constructions:
+lambda-abstractions $[[λx:iP.c]]$,
+standard let-bindings $[[let x = v; c]]$,
+and type annotations that can be added to any value or computation:
+$[[(v:iP)]]$ and $[[(c:iN)]]$.
 
 \subsection{Declarative Type Inference}
-\ottdefnsDT
+
+Next, we define the specification of the type 
+inference for our language.
+The judgement $[[Γ ; Φ ⊢ v : iP]]$
+means that under the type context $[[Γ]]$ and
+the variable context $[[Φ]]$, for the value $[[v]]$,
+type $[[iP]]$ is inferrable. The negative form of the judgement is symmetric.
+The application inference judgement $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ means that if
+a head of type $[[iN]]$ is applied to list of values $[[args]]$,
+then the resulting computation can be typed as $[[iM]]$.
+
+\begin{definition}[Variable Context]
+  The variable typing context $[[Φ]]$
+  is represented by a set of entries of the form
+  $[[x : iP]]$. 
+\end{definition}
+
+\begin{definition}[Declarative Type Inference]
+  \hfill
+  \begin{multicols}{2}
+  \ottdefnDTNInf{}
+  \ottdefnDTPInf{}
+  \ottdefnDTSpinInf{}
+  \end{multicols}
+\end{definition}
+
+Let us discuss selected rules of the declarative system:
+\begin{itemize}
+  \item \ruleref{\ottdruleDTVarLabel}
+    says that the type of a variable is inferred from the context.
+  \item \ruleref{\ottdruleDTThunkLabel} says
+    that the type of a thunk is inferred by shifting up the type of the 
+    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
+    infers the type of a return by shifting down the type of the
+    contained value.
+  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
+    They allow the inferred type to be refined by annotating it with a supertype.
+  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
+    mean that th declarative system allows to infer any type from the equivalence class.
+  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
+    and its first premise inferring the existential type of the value being unpacked.
+    It is important however that the inferred existential type is normalized. 
+    This is because there might be multiple equivalent existential types 
+    with different order or even number of quantified variables, 
+    and to bind them, the algorithm needs to fix the canonical one.
+  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
+    with annotated let-bindings. The first premise infers the type of the head of the application,
+    which must be a thunked computation. Then if after applying it 
+    to tha arguments, the resulting type can be equated to the annotated one,
+    we infer the body of the let-binding in the context extended with the bound variable.
+  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
+    but it is used when the type of the application is unique, and thus, the annotation
+    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
+    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
+\end{itemize}
+
+Let us discuss the rules of the application inference:
+\begin{itemize}
+  \item \ruleref{\ottdruleDTEmptyAppLabel} 
+    is the base case. If the list of arguments is empty, 
+    the inferred type is the type of the head. 
+    However, we relax this specification by allowing it to 
+    infer any other equivalent type. 
+    The relaxation of this rule is enough to guarantee 
+    this property for the whole judgement:
+    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
+    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
+    $[[iM']]$.
+  \item \ruleref{\ottdruleDTArrowAppLabel}
+    is where the application type is inferred: 
+    if the head has an arrow type $[[iQ → iN]]$,
+     we are allowed to apply it as soon as 
+    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
+  \item \ruleref{\ottdruleDTForallAppLabel}
+    is the rule ensuring the implicit elimination of the universal quantifiers. 
+    If we are applying a polymorphic computation, 
+    we can instantiate its quantified variables with any types,
+    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
+\end{itemize}
 
@@
 \subsection{Algorithmic Type Inference}
-\ottdefnsAT
-
-\subsection{Singularity}
-\ottdefnsSING
+\begin{algorithm}
+  \hfill
+  \begin{multicols}{2}
+  \ottdefnATNInf{}
+
+  \columnbreak
+  \ottdefnATPInf{}
+  \ottdefnATSpinInf{}
+  \end{multicols}
+\end{algorithm}
+
+\subsection{Constraint Singularity}
+
+\begin{algorithm}
+  \hfill
+  \begin{multicols}{2}
+    \ottdefnSINGscEN{}
+
+    \columnbreak
+    \ottdefnSINGscEP{}
+  \end{multicols}
+\end{algorithm}
+
+\newpage
 

### 2023-09-19 proofs/tex/equiv-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Add Declarative Typing comments

@@
         $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
-        Then by the transitivity of subtyping 
-        (\cref{corollary:subtyping-transitivity}), 
+        Then by transitivity of subtyping 
+        (\cref{lemma:subtyping-transitivity}), 
         $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
@@
 \begin{lemma} \label{lemma:mutual-subst-subtyping}
-  Asssuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
+  Assuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
   are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,

### 2023-09-20 proofs/main.tex
Elaborate on the algorithmic typing

@@
 \begin{algorithm}[Type Well-formedness]
+  \label{alg:wf}
   \hfill
@@
 \begin{algorithm}[Type Normalization]
+  \label{alg:type-nf}
   \hfill
@@
 \begin{algorithm}[Subtyping]
+  \label{alg:subtyping}
   \hfill
@@
 Next, we define the specification of the type 
-inference for our language.
-The judgement $[[Γ ; Φ ⊢ v : iP]]$
-means that under the type context $[[Γ]]$ and
-the variable context $[[Φ]]$, for the value $[[v]]$,
-type $[[iP]]$ is inferrable. The negative form of the judgement is symmetric.
-The application inference judgement $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ means that if
-a head of type $[[iN]]$ is applied to list of values $[[args]]$,
-then the resulting computation can be typed as $[[iM]]$.
+inference for our language. First, we introduce 
+variable context specifying the types of the variables 
+in the scope of the current rule.
 
@@
 
+The specification is represented by an inference system of
+three mutually recursive judgments:
+positive inference $[[Γ ; Φ ⊢ v : iP]]$,
+negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
+application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
+In the premises, the inference rules also refer to 
+the declarative subtyping (\cref{def:subDOne}),
+type well-formedness (\cref{alg:wf}), 
+and normalization (\cref{alg:alg:type-nf}) judgements.
+\begin{itemize}
+  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
+    means that under the type context $[[Γ]]$ and
+    the variable context $[[Φ]]$, for the value $[[v]]$,
+    type $[[iP]]$ is inferrable. It guarantees that 
+    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
+    the standard sense.
+  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
+    judgement. It means that if a head of type $[[iN]]$ 
+    is applied to list of values $[[args]]$, 
+    then the resulting computation can be typed as $[[iM]]$.
+\end{itemize}
+
 \begin{definition}[Declarative Type Inference]
+  \label{def:declarative-typing}
   \hfill
@@
 
+Next, we present the type inference algorithm, 
+which is sound and complete with respect to the declarative specification
+(\cref{def:declarative-typing}).
+
 \subsection{Algorithmic Type Inference}
+
+Mirroring the declarative typing, 
+the algorithm is represented by an inference system of three mutually recursive
+judgments:
+\begin{itemize}
+  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
+    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
+    In contrast with the declarative counterparts, they are deterministic,
+    and  guarantee that the inferred type is normalized. 
+  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
+    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
+    Notice that $[[uN]]$ contains algorithmic variables, 
+    which are specified by the context $[[Θ1]]$.
+    Moreover, the inferred type $[[uM]]$ is also algorithmic,
+    and can have several non-equivalent instantiations. To accommodate that, 
+    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
+    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
+    in which the variables must be instantiated, and $[[SC]]$
+    imposes restrictions on the variables. 
+\end{itemize}
+As subroutines, the algorithm calls
+subtyping (\cref{alg:subtyping}),
+type well-formedness (\cref{alg:wf}),
+constraint merge (\cref{sec:constraint-merge}),
+normalization (\cref{alg:type-nf}),
+and constraint singularity which will be defined later in \cref{todo}.
+It also relies on basic set operations and the ability to 
+deterministically choose fresh variables.
+
 \begin{algorithm}
@@
 
+Let us discuss the inference rules of the algorithm:
+\begin{itemize}
+  \item \ruleref{\ottdruleATVarLabel} 
+    infers the type of a variable by looking it up in the context
+    and normalizing the result.
+  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
+    are similar to the declarative rules: they make a recursive call
+    to type the body of the thunk or the return expression and
+    put the shift on top of the result.
+  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
+    are symmetric. They make a recursive call to infer the type of the annotated
+    expression, check that the inferred type is a subtype of the annotation,
+    and return the normalized annotation.
+  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
+    It makes a recursive call to infer the type of the body in the extended context,
+    and returns the corresponding arrow type. Notice that the algorithm also
+    normalizes the result, which is because the annotation type $[[iP]]$
+    is allowed to be non-normalized.
+  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
+    Similarly to the previous case, it makes a recursive call to infer the type
+    of the body in the extended \emph{type} context. 
+    After that, it returns the corresponding universal type. 
+    It is also required to normalize the result, because, 
+    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
+    in which case the $[[∀]]$ must be removed. 
+  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
+    it makes a recursive call to infer the type of the bound value,
+    and then returns the type of the body in the extended context.
+  \item \ruleref{\ottdruleATAppLetAnnLabel}
+    is interpreted as follows.
+    First, it infers the type of the head of the application,
+    ensuring that it is a thunked computation $[[↓iM]]$;
+    after that, it makes a recursive call
+    to the application inference procedure,
+    which returns the algorithmic type, whose
+    instantiation to a declarative type must be associated with the bound variable 
+    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
+    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
+    check whether the instantiation to the annotated type $[[iP]]$ is possible,
+    and if it is, the algorithm infers the type of the body in the extended context,
+    and returns it as the result. 
+  \item \ruleref{\ottdruleATAppLetAnnLabel}
+    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
+    However, since there is no annotation, 
+    instead of checking the instantiation to it, 
+    the algorithm checks that the inferred type
+    $[[↑uQ]]$ is unique.
+    It is the case if all the algorithmic variables of 
+    $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$,
+    which is checked by the combination of
+    $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
+    Together, these two premises guarantee that the only 
+    possible instantiation of $[[↑uQ]]$ is $[[ [uσ]uQ ]]$.
+  \item \ruleref{\ottdruleATUnpackLabel}
+    works in the expected way. First, it infers the 
+    existential type $[[∃nas.iP]]$ of the value being unpacked,
+    and since the type is guaranteed to be normalized, binds 
+    the quantified variables with $[[nas]]$.
+    Then it infers the type of the body in the appropriately extended context,
+    and checks that the inferred type does not depend on $[[nas]]$
+    by checking well-formedness $[[Γ ⊢ iN]]$.
+\end{itemize}
+
+Finally, let us discuss the algorithmic rules of the application inference:
+\begin{itemize}
+  \item \ruleref{\ottdruleATEmptyAppLabel}
+    is the base case. If the list of arguments is empty, 
+    the inferred type is the type of the head,
+    and the algorithm returns it after normalizing.
+  \item \ruleref{\ottdruleATArrowAppLabel}
+    is the main rule of the algorithmic application inference.
+    If the head has an arrow type $[[uQ → uN]]$,
+    we find $[[SC1]]$---the minimal constraint ensuring that 
+    $[[uQ]]$ is a supertype of the first argument's type.
+    Then we make a recursive call applying $[[uN]]$ to the rest of the arguments,
+    and merge the resulting constraint with $[[SC1]]$
+  \item \ruleref{\ottdruleATForallAppLabel},
+    analogously to the declarative case,
+    is the rule ensuring the implicit elimination of the universal quantifiers. 
+    This is the place where the algorithmic variables are generated.
+    The algorithm simply replaces the quantified variables 
+    $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
+    and makes a recursive call in the extended context. 
+\end{itemize}
+
+The correctness of the algorithm consists of its soundness and 
+completeness, which is by mutual
+induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
+The simplified result is the following.
+\begin{theorempreview}
+  \hfill
+  \begin{itemize}
+    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
+      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
+    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
+      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
+  \end{itemize}
+\end{theorempreview}
+
 \subsection{Constraint Singularity}
 
-\begin{algorithm}
+The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
+of the algorithmic typing check wether the constraint $[[SC]]$
+is uniquely defines a substitution satisfying it, and if it does
+returns such a substitution as the result.
+To do that, we define a partial function $[[SC singular with uσ]]$,
+taking a subtyping constraint $[[SC]]$ and returning a substitution 
+$[[uσ]]$---the only possible solution of $[[SC]]$.
+
+First, we define the notion of singularity on constraint entries. 
+$[[scE singular with iP]]$ and $[[scE singular with iN]]$
+are considered partial functions taking a constraint entry $[[scE]]$
+and returning the type satisfying $[[scE]]$ if such a type is unique.
+
+\begin{algorithm}[Singular Constraint Entry]
   \hfill
@@
 
+\begin{itemize}
+  \item \ruleref{\ottdruleSINGNEqLabel} and \ruleref{\ottdruleSINGPEqLabel} 
+    are symmetric. If the constraint entry says that a variable must be equivalent to 
+    a type $T$, then it is evidently singular, and the only (up-to-equivalence) type
+    instantiating this variable could be $T$. This way, we return its normal form. 
+  \item \ruleref{\ottdruleSINGSupVarLabel}
+    implies that the only (normalized) solution of $[[pua :≥ ∃nas.pa]]$ is 
+    $[[pa]]$ (it will be shown in \cref{lemma:var-subt}).
+  \item \ruleref{\ottdruleSINGSupShiftLabel}
+    is perhaps the least obvious rule.
+    In type $[[∃nas.↓iN]]$, if $[[iN]]$ is anything different (non-equivalent) to 
+    $[[nai ∊ {nas}]]$, there are at least 
+    one proper supertype of $[[∃nas.↓iN]]$, since $[[iN]]$ can be abstracted over
+    by an existential quantifier. Otherwise, any supertype of $[[∃nas.↓nai]]$
+    is equivalent to it, and thus, the solution is unique. 
+\end{itemize}
+
+Next, we extrapolate the singularity function on constraints---sets of constraint entries. 
+We require $[[SC]]$ to be a set of singular constraints, and the resulting substitution
+sends each variable from $[[dom(SC)]]$ to the unique type satisfying the corresponding constraint.
+
+\begin{algorithm}
+  $[[SC singular with uσ]]$
+  means that 
+  \begin{enumerate}
+    \item for any positive $[[scE ∊ SC]]$,
+      there exists $[[iP]]$ such that $[[scE singular with iP]]$, 
+      and for any negative $[[scE ∊ SC]]$,
+      there exists $[[iN]]$ such that $[[scE singular with iN]]$;
+    \item $[[uσ]]$ is defined as follows:
+      $$
+      [[ [uσ]β̂⁺ ]]  = 
+          \begin{cases}
+              [[ iP ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁺]] 
+                         \text{ and } [[scE singular with iP]] \\
+              [[ β̂⁺ ]] & \text{otherwise}  \\
+          \end{cases}
+      $$
+      $$
+      [[ [uσ]β̂⁻ ]]  = 
+          \begin{cases}
+              [[ iN ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁻]] 
+                         \text{ and } [[scE singular with iN]]\\
+              [[ β̂⁻ ]] & \text{otherwise}  \\
+          \end{cases}
+      $$
+  \end{enumerate} 
+\end{algorithm}
+
+The correctness of the singularity algorithm is formulated as follows:
+\begin{theorempreview}
+  Suppose that $[[SC]]$ is a subtyping constraint.
+  Then $[[SC singular with uσ]]$ holds if and only if 
+  $[[uσ]]$ is the only (up-to-equivalence on $[[dom(SC)]]$) 
+  normalized substitution satisfying $[[SC]]$.
+\end{theorempreview}
+
 \newpage

### 2023-09-22 oopsla24/paper.tex
Start of paper

%%
%% This is file `sample-acmsmall-submission.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-submission')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-submission.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen,review]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Local Type Inference for Polarised System F with Existentials}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Neel Krishnaswami}
\email{nk480@cl.cam.ac.uk}
\orcid{0000-0003-2838-5865}

\author{Ilia Kaisin}
\email{ik404@cl.cam.ac.uk}
\orcid{0000-1111-2222-3333}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Type Inference, System F, Call-by-Push-Value, Polarized Typing, Focalisation, Subtypingr}

\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\texttt{acmsmall}}: The default journal template style.
\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
\item {\texttt{acmtog}}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\texttt{acmconf}}: The default proceedings template style.
\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \texttt{\acmSubmissionID} command to print the
  submission's unique ID on each page of the work.
\item{\texttt{authorversion}}: Produces a version of the work suitable
  for posting by the author.
\item{\texttt{screen}}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall,screen,review]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification.  As an exception, multiple authors may share one
affiliation. Authors' names should not be abbreviated; use full first
names wherever possible. Include authors' e-mail addresses whenever
possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
  Artifacts: \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{Multi-language papers}

Papers may be written in languages other than English or include
titles, subtitles, keywords and abstracts in different languages (as a
rule, a paper in a language other than English should include an
English title and an English abstract).  Use \verb|language=...| for
every language used in the paper.  The last language indicated is the
main language of the paper.  For example, a French paper with
additional titles and abstracts in English and German may start with
the following command
\begin{verbatim}
\documentclass[sigconf, language=english, language=german,
               language=french]{acmart}
\end{verbatim}

The title, subtitle, keywords and abstract will be typeset in the main
language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
begin title, subtitle and keywords, can be used to set these elements
in the other languages.  The environment \verb|translatedabstract| is
used to set the translation of the abstract.  These commands and
environment have a mandatory first argument: the language of the
second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
of their usage.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{description}
\item[\texttt{sidebar}:]  Place formatted text in the margin.
\item[\texttt{marginfigure}:] Place a figure in the margin.
\item[\texttt{margintable}:] Place a table in the margin.
\end{description}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall-submission.tex'.

### 2023-09-27 oopsla24/paper.tex
Update oopsla24

@@
 %%
-\documentclass[acmsmall,screen,review]{acmart}
+% \documentclass[acmsmall,screen,review]{acmart}
+\documentclass[acmsmall,review,anonymous]{acmart}
 
@@
 
-\author{Ilia Kaisin}
-\email{ik404@cl.cam.ac.uk}
-\orcid{0000-1111-2222-3333}
+\author{Ilya Kaysin}
+\email{ik404@cam.ac.uk}
+\orcid{0000-0002-6301-152X}
 
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Type Inference, System F, Call-by-Push-Value, Polarized Typing, Focalisation, Subtypingr}
+\keywords{Type Inference, System F, Call-by-Push-Value, Polarized Typing, Focalisation, Subtyping}
 
@@
 \section{Introduction}
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
+\input{tex/01-introduction.tex}
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
+\section{Overview}
+\input{tex/02-overview.tex}
 
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
+\section{Declarative System}
+\input{tex/03-declarative-system.tex}
 
-\section{Template Overview}
-As noted in the introduction, the ``\verb|acmart|'' document class can
-be used to prepare many different kinds of documentation --- a
-double-blind initial submission of a full-length technical paper, a
-two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
-journal article, a SIGCHI Extended Abstract, and more --- all by
-selecting the appropriate {\itshape template style} and {\itshape
-  template parameters}.
+\section{Algorithm}
+\input{tex/04-algorithm.tex}
 
-This document will explain the major features of the document
-class. For further information, the {\itshape \LaTeX\ User's Guide} is
-available from
-\url{https://www.acm.org/publications/proceedings-template}.
+\section{Proof}
+\input{tex/05-proof.tex}
+
+\section{Extensions}
+\input{tex/06-extensions.tex}
+
+\section{Conclusion}
+\input{tex/07-conclusion.tex}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
+  % \includegraphics[width=\linewidth]{sample-franklin}
   \caption{1907 Franklin Model D roadster. Photograph by Harris \&

### 2023-09-27 oopsla24/tex/01-introduction.tex
Update oopsla24

ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

### 2023-09-28 proofs/main.tex
Fix file dirs

@@
 
-\input{../_build/_gen/included.tex}
+\newcommand{\ottDir}{../ott/_gen}
+\newcommand{\genDir}{_gen}
+\input{\ottDir/included.tex}
 
@@
 \subsection{Type Well-formedness}
-\input{../_build/_gen/tex/wf-lemmas.tex}
+\input{\genDir/tex/wf-lemmas.tex}
 
 \subsection{Substitution}
-\input{../_build/_gen/tex/subst-lemmas.tex}
+\input{\genDir/tex/subst-lemmas.tex}
 
 \subsection{Declarative Subtyping}
-\input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
+\input{\genDir/tex/decl-subtyping-lemmas.tex}
 
@@
 \label{sec:decl-equiv-lemmas}
-\input{../_build/_gen/tex/equiv-lemmas.tex}
+\input{\genDir/tex/equiv-lemmas.tex}
 
 \subsection{Variable Ordering}
-\input{../_build/_gen/tex/ord-lemmas.tex}
+\input{\genDir/tex/ord-lemmas.tex}
 
 \subsection{Normaliztaion}
-\input{../_build/_gen/tex/norm-lemmas.tex}
+\input{\genDir/tex/norm-lemmas.tex}
 
@@
 \subsection{Algorithmic Type Well-formedness}
-\input{../_build/_gen/tex/wf-algo-lemmas.tex}
+\input{\genDir/tex/wf-algo-lemmas.tex}
 
 \subsection{Substitution}
-\input{../_build/_gen/tex/subst-algo-lemmas.tex}
+\input{\genDir/tex/subst-algo-lemmas.tex}
 
 \subsection{Normalization}
-\input{../_build/_gen/tex/norm-alg-lemmas.tex}
+\input{\genDir/tex/norm-alg-lemmas.tex}
 
 \subsection{Equivalence}
-\input{../_build/_gen/tex/equiv-alg-lemmas.tex}
+\input{\genDir/tex/equiv-alg-lemmas.tex}
 
 \subsection{Unification Constraint Merge}
-\input{../_build/_gen/tex/unif-constraint-merge-lemmas.tex}
+\input{\genDir/tex/unif-constraint-merge-lemmas.tex}
 
 \subsection{Unification}
-\input{../_build/_gen/tex/unification-lemmas.tex}\
+\input{\genDir/tex/unification-lemmas.tex}\
 
 \subsection{Anti-unification}
-\input{../_build/_gen/tex/au-lemmas.tex}
+\input{\genDir/tex/au-lemmas.tex}
 
@@
 \label{sec:alg-upper-bounds-proofs}
-\input{../_build/_gen/tex/lub-lemmas.tex}
+\input{\genDir/tex/lub-lemmas.tex}
 
@@
 \label{sec:upgrade-lemmas}
-\input{../_build/_gen/tex/upgrade-lemmas.tex}
+\input{\genDir/tex/upgrade-lemmas.tex}
 
 \subsection{Constraint Satisfaction}
-\input{../_build/_gen/tex/constraint-sat-lemmas.tex}
+\input{\genDir/tex/constraint-sat-lemmas.tex}
 
 \subsection{Positive Subtyping}
-\input{../_build/_gen/tex/alg-pos-subtyping-lemmas.tex}
+\input{\genDir/tex/alg-pos-subtyping-lemmas.tex}
 
 \subsection{Subtyping Constraint Merge}
-\input{../_build/_gen/tex/constraint-merge-lemmas.tex}
+\input{\genDir/tex/constraint-merge-lemmas.tex}
 
 \subsection{Negative Subtyping}
-\input{../_build/_gen/tex/alg-neg-subtyping-lemmas.tex}
+\input{\genDir/tex/alg-neg-subtyping-lemmas.tex}
 
 \section{Properties of the Declarative Typing}
-\input{../_build/_gen/tex/decl-typing-lemmas.tex}
+\input{\genDir/tex/decl-typing-lemmas.tex}
 
@@
 \subsection{Singularity}
-\input{../_build/_gen/tex/singularity-lemmas.tex}
+\input{\genDir/tex/singularity-lemmas.tex}
 
 \subsection{Correctness of the Typing Algorithm}
-\input{../_build/_gen/tex/alg-typing-lemmas.tex}
+\input{\genDir/tex/alg-typing-lemmas.tex}
 

### 2023-09-28 oopsla24/tex/03-declarative-system.tex
minor

--- 2023-09-19 proofs/main.tex
@@
-\UseRawInputEncoding
-% vim: ft=tex
-\documentclass[a4,natbib=false]{article}
-\usepackage[a4paper, total={8in, 10in}]{geometry}
-\usepackage{hyperref}
-\usepackage{mathpartir}
-
-\input{prelude.tex}
-
-\input{../_build/_gen/included.tex}
-
-% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
-% \newcommand{\ottdrulename}[1]{\textsc{#1}}
-
-% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
-%   }\ruleLabel{#4}{#4}}
-
-% \renewcommand{\ottdrule}[4][]{%
-%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
-%   \mpr@label{\textsc{#1}}%
-% }
-% \DeclareDocumentCommand \redrule {m m o}{%
-%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
-%   \mpr@label{\textsc{#3}}
-% }
-
-% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}
-
-% \renewcommand{\ottdruleEOneNVarName}[0]{foo }
-
-% \makeatletter
-% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
-%   }{
-%     % \protected@edef\@currentlabelname{foo}
-%     \def\@currentlabelname{foo}%
-%     \phantomsection
-%     \label{boo}
-%     \alpha ^{-}   \eqEOne   \alpha ^{-}
-%   }{%
-%     {\ottdruleEOneNVarName}{}%
-%   } }
-% \makeatother
-
-\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
-\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
-
-\newcommand{\ruleref}[1]{Rule \nameref{#1}}
-
-% ord varset in uN = varset'
-
-\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}
-
-% ord varset in uP = varset'
-
-\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}
-
-% nf(N) = M
-\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}
-
-% nf(P) = Q
-\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}
-
-% N ≈ M
-
-\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}
-
-% P ≈ Q
-\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}
-
-% G ⊢ N ≤1 M
-
-\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
-\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}
-
-% G ⊢ P ≥1 Q
-\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
-\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
-\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}
-
-% G ⊢ N ≈1 M
-\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}
-
-% G ⊢ P ≈1 Q
-\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}
-
-% G ⊨ iP1 ∨ iP2 = iQ
-\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
-\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
-\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
-\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}
-
-% G ; Θ ⊨ uN ≤ iM ⫤ us
-\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
-\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
-\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
-\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}
-
-% G ; Θ ⊨ iP ≥ uQ ⫤ us
-\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
-\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
-\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
-\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}
-
-% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
-\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}
-
-% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
-\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
-\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}
-
-% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
-\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
-\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}
-
-% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
-\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
-
-% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
-\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUAUName}[0]{(AU)}
-
-% Γ ⊢ iN 
-\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ⊢ iP
-\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uN
-\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uP
-\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ⊢ iP : scE
-\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
-\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}
-
-% Γ ⊢ iN : scE
-\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}
-
-% Γ ; Φ ⊢ v : iP 
-\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ c : iN 
-\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ iN ● args ⇒> iM 
-\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ v : iP  
-\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ c : iN 
-\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-
-% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
-\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% scE1 singular with iP
-\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}
-
-% scE1 singular with iN
-\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}
-
-\begin{document}
-
-\tableofcontents
-
-\newpage
-
-\section{Declarative Type Systems}
-
-\subsection{Grammar}
-We assume that there is an infinite set of positive and 
-negative \emph{type} variables. Positive type variables are denoted as 
-$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
-Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
-We assume there is an infinite set of \emph{term} variables,
-which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
-A list of objects (variables, types or terms) is denoted by
-an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
-$[[nbs]]$ is a list of negative type variables, 
-$[[args]]$ is a list of values, which are arguments of a function.
-$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
-in a type $[[iP]]$ and $[[iN]]$, respectively.
-
-\bigskip
-
-% \ottgrammartabular{
-%   \ottiP\ottinterrule
-%   \ottiN\ottinterrule
-% }
+\subsection{The Language}
 
@@
 \end{definition}
-
-\subsection{Equalities}
-For simplicity, we assume alpha-equivalent terms equal.
-This way, we assume that substitutions do not capture bound variables.
-Besides, we equate
-$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
-and lift these equations transitively and congruently 
-to the whole system.
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Declarative Type Context]
-  \hfill \\
-  Declarative type context $[[Γ]]$ is represented by a set of 
-  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
-  union of two contexts $[[Γ1 ∪ Γ2]]$.
-\end{definition}
-
-$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
-which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
-(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
-
-Notice that checking the well-formedness of a type
-is an \emph{algorithmic} procedure, in which 
-both the context and the type are considered inputs.
-In other words, it is syntax-directed and mode-correct 
-(according to \cite{dunfieldBidirectionalTyping2020}), 
-which means that
-
-\begin{algorithm}[Type Well-formedness]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFTNiWF{}
-  \columnbreak
-
-  \ottdefnWFTPiWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-\subsection{Substitutions}
-
-\begin{definition}[Substitution]
-  Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions form variables to types, preserving the polarity. 
-\end{definition}
-
-\begin{algorithm}[Substitution Application]
-  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
-  It is defined naturally as follows:
-    \begin{multicols}{2}
-      \begin{itemize}
-        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
-        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
-        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
-        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
-        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
-        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
-        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
-          that is the variable capture never happens);
-      \end{itemize}
-    \end{multicols}
-\end{algorithm}
-
-\begin{definition}[Substitution Signature]
-  The signature $[[Γ' ⊢ σ : Γ]]$ means that
-  \begin{enumerate}
-    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
-  \end{enumerate}
-\end{definition}
-
-A substitution can be restricted to a set of variables. 
-The restricted substitution is define as expected. 
-\begin{definition}[Subsitution Restriction]
-  The specification $[[σ  | varset]]$ is defined as
-  a function such that 
-  \begin{enumerate}
-    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
-    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
-  \end{enumerate}
-\end{definition}
-
-Two substitutions can be composed in two ways:
-$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
-while $[[σ2 <=< σ1]]$
-depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
-$[[σ2]]$ to its results on the domain.
-\begin{definition}[Substitution Composition]
-  $[[σ2 ○ σ1]]$ is defined as a function such that
-  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
-\end{definition}
-
-\begin{definition}[Monadic Substitution Composition]
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
-\end{definition}
-Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
-specification of $[[σ1]]$, which is not unique. 
-However, we assume that the used specification clear from the 
-context of the proof.
-
-\begin{definition}[Equivalent Substitutions]
-  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
-  indicates that on the domain $[[Γ]]$, 
-  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
-\end{definition}
-
-Sometimes it is convenient to construct substitution 
-explicitly mapping each variable from a list (or a set)
-to a type. Such substitutions are denoted as $[[iPs / pas]]$
-and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
-the corresponding types.
-\begin{definition}[Explicit Substitution]
-  \hfill
-  \begin{itemize}
-    \item [$-$]
-      Suppose that $[[nas]]$ is a list of negative type variables,
-      and $[[iNs]]$ is a list of negative types of the same length.
-      Then $[[iNs / nas]]$ denotes a substitution such that 
-      \begin{enumerate}
-        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
-        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
-      \end{enumerate}
-    \item [$+$]
-      Positive explicit substitution $[[iPs / pas]]$
-      is defined symmetrically.
-  \end{itemize}
-\end{definition}
-
-\subsection{Declarative Subtyping}
-Subtyping is one of the key mechanism of our system. 
-It realizes the polymorphism: abstract $[[∀]]$ and 
-$[[∃]]$ types can be used where concrete types are expected,
-exactly because the subtyping relation between them.
-
-\begin{definition} 
-  \label{def:subDOne}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNsub{}
-    \columnbreak\\
-    \ottdefnDOnePsup{}
-  \end{multicols}
-
-  \begin{multicols}{2}
-    \ottdefnDOneNeq{}
-    \columnbreak\\
-    \ottdefnDOnePeq{}
-  \end{multicols}
-\end{definition}
-
-The following observations about the declarative subtyping are worth noting:
-\begin{itemize}
-  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
-    make the subtyping reflexive on variables (and further, on any type).
-  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
-    resulting type and contravariant on the argument type.
-  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
-    the subtyping is \emph{invariant} for shifts. 
-    This way, the subtyping of shifted types in one direction implies the subtyping
-    in the opposite direction.
-    Although this rule restricts the
-    subtyping relation, it makes the system decidable.
-  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
-    non-algorithmic rules: the substitution for the quantified variable is
-    not specified, those, these rules `drive' the subtyping relation.
-\end{itemize}
-
-In the next section, we present the sound and complete algorithm
-checking whether one type is a subtype of another according to \cref{def:subDOne}.
-
-\section{Algorithmic Type System}
-
-\subsection{Grammar}
-
-In the algorithmic system, we extend the grammar of types
-by adding positive and negative \emph{algorithmic variables}
-($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
-They represent the unknown types, which will be inferred by the algorithm.
-This way, we add two base cases to the grammar of 
-positive and negative types, and use highlight to denote that the type
-can potentially contain algorithmic variables.
-
-\begin{definition}[Algorithmic Types]
-  \label{def:algo-types}
-  \hfill\\
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottuN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottuP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Fresh Variable Selection}
-\label{sec:fresh-selection}
-Both the subtyping and the type inference algorithm
-rely on the ability to select fresh, unused variables.
-For a set of variables $[[varset]]$, it is indicated as 
-$[[varset are fresh]]$ in the inference rules.
-We assume that the selection subroutine always succeeds and is 
-deterministic. In other words, whenever it is called in 
-an algorithmic inference rule, it returns the same result, 
-uniquely determined by the input of this rule.
-
-\subsection{Variable Algorithmization}
-\label{sec:variable-algorithmization}
-
-In several places of our algorithm, in particular, during
-algorithmic subtyping,
-we turn a declarative type into the algorithmic one
-via replacing certain type variables with fresh algorithmic variables.
-We call this procedure \emph{variable algorithmization}, and define it as follows.
-
-\begin{definition}[Variable Algorithmization]
-  Suppose that $[[nas]]$ is a list of negative type variables
-  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
-  in a type for $[[αî⁻ ∊ {nuas}]]$.
-\end{definition}
-
-Conversely, we have the opposite procedure turning algorithmic type variables
-into declarative type variables via \emph{dealgorithmization}.
-
-\begin{definition}[Variable Dealgorithmization]
-  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
-  and $[[nas]]$ is a list of negative type variables of the same length. 
-  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
-  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
-\end{definition}
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
-  \hfill \\
-  Algorithmic type context $[[Ξ]]$ is represented by a set of 
-  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
-  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
-\end{definition}
-
-$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
-that the algorithmic type is well-formed in the contexts
-$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
-of the type is contained in $[[Ξ]]$, and each free declarative type variable
-of the type is contained in $[[Γ]]$.
-
-\begin{algorithm}[Algorithmic Type Well-formedness]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFATNauWF{}
-  \columnbreak
-
-  \ottdefnWFATPauWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-Algorithmic Type Context are used in the unification algorithm.
-In the subtyping algorithm, 
-the context needs to remember additional information.
-In the subtyping context, each algorithmic variable is associated with a
-context it must be instantiated in 
-(i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
-  \hfill \\
-  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
-  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
-  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
-  and $[[Γ]]$ is a context in which they must be instantiated.
-  We assume that no two entries associating the same variable
-  appear in $[[Θ]]$.
-
-  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
-  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
-  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
-\end{definition}
-
-\subsection{Subsitutions}
-
-Substitution that operates on algorithmic type variables is denoted as
-$[[uσ]]$. It is defined as a total function from algorithmic 
-type variables to \emph{declarative} types, preserving the polarity.
-
-The signature $[[Θ ⊢ uσ : Ξ]]$ means that
-$[[Ξ ⊆ dom(Θ)]]$ and 
-$[[uσ]]$ maps each algorithmic variable 
-from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
-and for each variable not appearing in $[[dom(Θ)]]$, 
-it acts as identity.
-
-\begin{definition}[Signature of Algorithmic Substitution]
-  \label{def:algo-subst-sig}
-  \hfill
-  \begin{itemize}
-    \item $[[Θ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$,
-          there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
-          and $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-    \item $[[Γ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-  \end{itemize}
-\end{definition}
-
-In the anti-unification algorithm, we use another kind of substitution.
-In contrast to algorithmic substitution $[[uσ]]$,
-it allows mapping algorithmic variables to
-\emph{algorithmic} types.
-Additionally, anti-unification substitution is restricted to the
-\emph{negative} segment of the language.
-Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a
-
-The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
-in which the results of an anti-unification substitution 
-are formed, is fixed for this substitution.
-This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
-variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.
-
-\begin{definition}[Signature of Anti-unification substitution]
-  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
-  \begin{enumerate}
-    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
-    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
-  \end{enumerate}
-\end{definition}
-
-\subsection{Equivalence and Normalization}
-\label{sec:equivalence-normalization}
-
-The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
-there are types that are subtypes of each other but not equal. 
-For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
-and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
-although these types are not alpha-equivalent.
-For the subtyping algorithm, it is crucial to be able to check whether
-two types are equivalent, without checking mutual subtyping. 
-For this purpose we define the normalization procedure, 
-which allows us to uniformly choose the representative type of the equivalence class.
-This way, the equivalence checking is reduced to normalization and equality checking.
-
-For clarification of the proofs and better understanding of the system, 
-we introduce an intermediate relation---\emph{declarative equivalence}. 
-As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
-this relation is equivalent to the subtyping-induced equivalence, but does not 
-depend on it. Although this relation is not defined algorithmically, 
-it gives the intuition of what types our system considers equivalent.
-Specifically, in addition to \emph{alpha-equivalence}, 
-our system allows for \emph{reordering of adjacent quantifiers},
-and \emph{introduction/elimination of unused quantifiers}.
-
-The non-trivial rules of the declarative equivalence are
-\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
-Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
-the recursive call on the body of the quantified type. 
-It will be covered formally in \cref{sec:decl-equiv-lemmas}.
-
-\begin{definition}[Declarative Type Equivalence]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnEOneNeq{}
-  \columnbreak\\
-  \ottdefnEOnePeq{}
-  \end{multicols}
-
-\end{definition}
-
-As the equivalence includes arbitrary reordering of quantified variables,
-the normalization procedure is needed to choose the canonical order.
-For this purpose, we introduce an auxiliary procedure---variable ordering. 
-Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
-in the order they appear in $[[iN]]$.
-
-\begin{algorithm}[Variable Ordering]
-  \label{alg:var-ordering}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnONVar{}
-  \columnbreak\\
-  \ottdefnOPVar{}
-  \end{multicols}
-
-  Analogously, the variable can be ordered in 
-  an \emph{algorithmic} type ($[[ord varset in uP]]$ and 
-  $[[ord varset in uN]]$). In these cases, we treat the algorithmic variables
-  as if they were declarative variables.
-
-\end{algorithm}
-
-Next, we use the variable ordering in the normalization procedure. 
-Specifically, normalization recursively traverses the type, 
-and for each quantified case reorders the quantified variables in a 
-canonical order dictated by \cref{alg:var-ordering}, removing unused ones.
-
-\begin{algorithm}[Type Normalization]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnNrmNNorm{}
-  \columnbreak\\
-  \ottdefnNrmPNorm{}
-  \end{multicols}
-
-  Analogously,  we define normalization of algorithmic types 
-  by adding base cases:
-
-  \begin{multicols}{2}
-  \ottdefnNrmuNNorm{}
-  \columnbreak\\
-  \ottdefnNrmuPNorm{}
-  \end{multicols}
-
-\end{algorithm}
-
-\Cref{lemma:subt-equiv-algorithmization}
-demonstrates that the equivalence of types is the same
-as the equality of their normal forms.
-\begin{theorempreview}[Correctness of Normalization]
-  Assuming the types are well-formed in $[[Γ]]$, 
-  \begin{itemize}
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]]$ if and only if $[[nf(iN) = nf(iM)]]$;
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]]$ if and only if $[[nf(iP) = nf(iQ)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\begin{algorithm}[Substitution Normalization]
-  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
-  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
-\end{algorithm}
-
-The rest of this chapter is devoted to the
-central algorithm of the type system---the subtyping algorithm. 
-\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
-The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
-$A$ uses $B$ as a sub-procedure.
-
-\begin{figure}[h]
-  \centering
-  \begin{tikzpicture}
-    [>={Stealth[scale=2]},node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
-
-    % Define nodes
-    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[left=2cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[right=2cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
-    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
-    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
-
-    % Define edges
-    \draw[->] (1) to (2);
-    \draw[->] (1) to (3);
-    \draw[->] (1) to (5);
-    \draw[->] (2) to (3);
-    \draw[->] (2) to (6);
-    \draw[->] (3) to (4);
-    \draw[->] (3) to (5);
-    \draw[->] (4) to (6);
-    \draw[->] (5) to (8);
-    \draw[->] (6) to (7);
-  \end{tikzpicture}  
-  \caption{Dependency graph of the subtyping algorithm}
-  \label{fig:subtyping-algo}
-\end{figure}
-
-\subsection{Subtyping}
-\label{sec:subtyping}
-
-Now, we present the subtyping algorithm itself.
-Although the algorithm is presented as a single procedure,
-is important for the structure of the proof that the positive subtyping algorithm
-does not invoke the negative one. This way, the correctness of the positive 
-subtyping will be proved independently, and used afterwards to prove the
-correctness of the negative subtyping.
-
-\begin{algorithm}[Subtyping]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnANsub{}
-  \columnbreak\\
-  \ottdefnAPsup{}
-  \end{multicols}
-\end{algorithm}
-
-The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
-the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
-must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
-and $[[uP]]$ and $[[iQ]]$ for the positive case. 
-As one of the invariants, we require
-$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
-The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
-which will be discussed in the next section.
-
-Let us overview the inference rules of the subtyping algorithm.
-\begin{itemize}
-  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
-    are the base cases. They copy the corresponding declarative rules and
-    ensure the reflexivity.
-  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
-    subtyping constraints. In this case, we must ensure
-    that the resulting constrains guarantees that the instantiation of 
-    $[[â⁺]]$ is a supertype of $[[iP]]$.
-    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
-    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
-    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
-    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
-
-    Notice that this rule does not have a negative counterpart. This is
-    because one one of the important invariants of the algorithm: 
-    in the negative subtyping, only positive algorithmic variables
-    can occur in the types.
-
-  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
-    \emph{shift} rules. According to the declarative system,
-    shifted subtyping requires equivalence. In presence of the algorithmic 
-    variables, it means that the left and the right-hand sides of the subtyping
-    must be unified. Hence, the shift rules invoke the unification algorithm, 
-    which will be discussed in \cref{sec:unification}. The unification 
-    returns the minimal set of constraints $[[UC]]$, which is necessary
-    and sufficient for the subtyping.
-
-  \item \ruleref{\ottdruleAArrowLabel}.
-    In this case, the algorithm makes two calls:
-    a recursive call to the negative subtyping algorithm for the argument types,
-    and a call to the positive subtyping algorithm for the result types.
-    After that, the resulting constraints are merged using the
-    \emph{subtyping constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
-    are symmetric. These are the only places where 
-    the algorithmic variables are introduced.
-    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
-    of the quantified variables: these variables are replaced by 
-    fresh algorithmic variables in the body of the quantified type,
-    the algorithmic variables are added to the subtyping context $[[Θ]]$,
-    after that the recursive call is made. Notice that the declarative context
-    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
-    which matches the declarative system.
-\end{itemize}
-
-Then soundness lemma (\cref{lemma:pos-subt-soundness,lemma:neg-subt-soundness}) 
-and completeness (\cref{lemma:pos-subt-completeness,lemma:neg-subt-completeness})
-of the algorithm together give us the following simplified theorem:
-
-\begin{theorempreview}[Correctness of subtyping algorithm]
-  \hfill
-  \begin{itemize}
-    \item [$-$]  $[[ Γ ; · ⊨ uN ≤ iM ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iN ≤ iM ]]$;
-    \item [$+$] $[[ Γ ; · ⊨ uP ≥ iQ ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iP ≥ iQ ]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraints}
-
-Unification and subtyping algorithms are based on the constraint generation.
-The constraints are represented by set of constraint entries.
-
-\begin{definition}[Unification Constraint]
-  \hfill
-  \begin{description}
-    \item[unification entry] (denoted as $[[ucE]]$) is an expression of shape 
-      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
-    \item[unification constraint] (denoted as $[[UC]]$) is a set of 
-      unification constraint entries.
-      We denote $\{[[α̂±]] \mid [[ucE ∊ UC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(UC)]]$.
-  \end{description}
-\end{definition}
-
-However, in the subtyping, we need to consider more general
-kind of constraints. Specifically,
-subtyping constraint entries can restrict a variable
-not only to be equivalent to a certain type, but
-also to be a supertype of a positive type.
-
-\begin{definition}[Subtyping Constraint]
-  \hfill
-  \begin{description}
-    \item[subtyping entry] (denoted as $[[scE]]$) is an expression of shape 
-      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
-    \item[subtyping constraint] (denoted as $[[SC]]$) is a set of subtyping constraint entries.
-      We denote $\{[[α̂±]] \mid [[scE ∊ SC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(SC)]]$.
-  \end{description}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint Entry]
-  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
-  its associated type is well-formed in $[[Γ]]$.
-  \begin{itemize}
-    \item[] $[[Γ ⊢ pua :≥ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ pua :≈ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ nua :≈ iN]]$ ~iff~ $[[Γ ⊢ iN]]$.
-  \end{itemize}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  subtyping context $[[Θ]]$ if all its entries are well-formed in
-  the corresponding elements of $[[Θ]]$.
-  More formally, 
-  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
-  such that $[[scE]]$ restricts $[[α̂±]]$,
-  we have $[[Θ(α̂±) ⊢ scE]]$.
-
-  We write $[[Θ ⊢ SC : Ξ]]$ to denote
-  that $[[Θ ⊢ SC]]$ and $[[dom(SC) = Ξ]]$.
-
-  $[[Θ ⊢ UC]]$ and $[[Θ ⊢ UC : Ξ]]$ are defined analogously.
-\end{definition}
-
-\subsubsection{Constraint Satisfaction}
-
-A constraint entry restricts a type that can be assigned to a variable.
-We say that a type satisfies a constraint entry if it can be assigned
-to the variable restricted by the entry.
-
-\begin{definition}[Type Satisfying a Constraint Entry]
-  \hfill\\
-  \begin{multicols}{2}
-  \ottdefnSATSCEN{}
-  \columnbreak\\
-  \ottdefnSATSCEP{}
-  \end{multicols}
-\end{definition}
-
-We say that a substitution satisfies a constraint---a set of constraint 
-entries if each entry is satisfied by the type assigned to the variable
-by the substitution.
-
-\begin{definition}[Substitution Satisfying a Constraint]
-  We write $[[Θ ⊢ uσ : SC]]$ to denote that
-  a substitution $[[uσ]]$ satisfies a constraint $[[SC]]$ in a context $[[Θ]]$.
-  It presumes that $[[Θ ⊢ SC]]$ and 
-   means that for any $[[ucE]] \in [[SC]]$, if $[[ucE]]$ restricts $[[α̂±]]$,
-  then $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$.
-
-  Unification constraint satisfaction $[[Θ ⊢ uσ : UC]]$ 
-  is defined analogously as a special case of subtyping constraint satisfaction.
-
-\end{definition}
-
-Notice that $[[Θ ⊢ uσ : SC]]$ does not 
-imply the signature $[[Θ ⊢ uσ : dom(SC)]]$, because 
-the latter also specifies $[[uσ]]$ outside of the domain $[[dom(SC)]]$
-(see \cref{def:algo-subst-sig}).
-
-\subsubsection{Constraint Merge}
-\label{sec:constraint-merge}
-
-In this section, define the least upper bound 
-for constraints, which we call \emph{merge}.
-Intuitively, the merge of two constraints is the least
-constraint such that any substitution satisfying both constraints
-satisfies the merge as well.
-First, we define the merge of entries,
-and then extend it to the set of entries.
-
-\begin{definition} [Matching Entries]
-  We call two unification constraint entries 
-  or two subtyping constraint entries matching 
-  if they are restricting the same unification variable.
-\end{definition}
-
-Two matching entries formed in the same context $[[Γ]]$ 
-can be merged in the following way:
-\begin{algorithm}[Merge of Matching Constraint Entries]
-  \label{definition:merge-matching-entries}
-   \hfill
-
-  \ottdefnSCME\\
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}
-    are symmetric cases. To merge two matching entries restricting
-    a variable to be equivalent to certain types, we check
-    that these types are equivalent to each other.
-    To do so, it suffices to check for \emph{equality} of their normal forms,
-    as discussed in \cref{sec:equivalence-normalization}. 
-    After that, we return the left-hand entry.
-
-  \item \ruleref{\ottdruleSCMEEqSupLabel} and \ruleref{\ottdruleSCMESupEqLabel}
-    are also symmetric. 
-    In this case,
-    since one of the entries requires the variable to be equal to 
-    a type, the resulting entry must also imply that.
-    However, for the soundness, it is needed to ensure that
-    the equating restriction is stronger than the subtyping restriction.
-    For this purpose, the premise invokes the positive subtyping.
-
-  \item \ruleref{\ottdruleSCMESupSupLabel} 
-    In this case, we find the least upper bound of the types from the input
-    restrictions, 
-    and as the output, restrict the variable to be a supertype of the result.
-    The least upper bound procedure will be discussed in \cref{sec:lub}.
-\end{itemize}
-
-Unification constraint entries are a special case of subtyping constraint
-entries. They are merged using the same algorithm 
-(\cref{definition:merge-matching-entries}).
-Notice that the merge of two matching unification constraint entries
-is a unification constraint entry.
-\begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
-  \label{lemma:merge-matching-entries-welldef}
-  Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
-  are unification constraint entries. 
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  $[[Γ ⊢ ucE1 & ucE2 = ucE]]$
-  according to \cref{definition:merge-matching-entries},
-  is a unification constraint entry.
-\end{lemma}
-\begin{proof}
-  Since $[[ucE1]]$ and $[[ucE2]]$ are matching unification constraint entries,
-  they have the shape $([[pua :≈ iP1]], [[pua :≈ iP2]])$ or
-  $([[nua :≈ iN1]], [[nua :≈ iN2]])$.
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  can only be defined by \ruleref{\ottdruleSCMEPEqEqLabel} or
-  \ruleref{\ottdruleSCMENEqEqLabel}.
-  In both cases the result, if it exists, 
-  is a unification constraint entry:
-  in the first case, the result has shape $[[pua :≈ iP1]]$,
-  in the second case, the result has shape $[[nua :≈ iN1]]$.
-\end{proof}
-
-% Notice that in case of equivalence, the assigned types
-% must be equal (i.e. alpha-equivalent) to be merged. This is because
-% the unification algorithm assumes that every type is normalized,
-% and hence, equivalence is alpha-equivalence 
-% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
-
-\begin{algorithm}[Merge of Subtyping Constraints]
-  \label{definition:merge-subtyping-constraints}
-  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
-  Then $[[Θ ⊢ SC1 & SC2 = SC]]$
-  defines a set such that $[[ucE]] \in [[SC]]$ iff either
-  \begin{itemize}
-    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
-    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
-    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
-      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
-      $[[α̂±]]$. 
-  \end{itemize}
-\end{algorithm}
-
-Unification constraints can be considered 
-as a special case of subtyping constraints,
-and the merge of unification constraints
-is defined as the merge of subtyping constraints.
-Then it is easy to see that the merge of two 
-unification constraints is a unification constraint.
-
-\begin{lemma}[Merge of Unification Constraints is well-defined]
-  Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
-  are unification constraints. 
-  Then the merge of $[[UC1]]$ and $[[UC2]]$ 
-  $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$
-  according to \cref{definition:merge-subtyping-constraints},
-  is a unification constraint.
-\end{lemma}
-\begin{proof}
-  $[[UC]]$ consists of unmatched entries of $[[UC1]]$ and $[[UC2]]$,
-  which are \emph{unification} constraint entries by assumption,
-  and merge of matching entries, which also are  
-  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-welldef}.
-\end{proof}
-
-\Cref{lemma:merge-soundness,lemma:merge-completeness} 
-show the correctness and initiality of the merge operation,
-which can be expressed in the following simplified theorem:
-\begin{theorempreview}[Correctness of Constraint Merge]
-  A substitution $[[uσ]]$ satisfying both constraints
-  $[[SC1]]$ and $[[SC2]]$ 
-  if and only if it satisfies their merge.
-\end{theorempreview}
-
-The unification constraint merge satisfies the same theorem,
-however, because the merge of unification constraint entries 
-$[[ucE1]]$ and $[[ucE2]]$ always results in one of them, 
-a stronger soundness property holds (see \cref{lemma:unif-merge-soundness}):
-\begin{theorempreview}[Soundness of Unification Constraint Merge]
-  If $[[Θ ⊢ UC1 & UC2 = UC]]$ then $[[UC = UC1 ∪ UC2]]$.
-\end{theorempreview}
-
-\subsection{Unification}
-\label{sec:unification}
-
-The subtyping algorithm calls the following subtask:
-given two algorithmic types, we need to find the most general substitution 
-for the algorithmic variables in these types, such that the resulting 
-types are equivalent. This problem is known as \emph{unification}.
-
-In our case, the unification is restricted in the following way:
-first, before unifying the types, we normalize them, which 
-allows us to reduce (non-trivial) equivalence to (trivial) equality;
-second, we preserve invariants which guarantee that
-one side of the unification is always declarative, which in fact, 
-reduces the unification to the \emph{matching} problem.
-
-The unification procedure
-returns a set of minimal constraints,
-that must be satisfied by a substitution
-unifying the input types.
-
-\begin{algorithm}[Unification]
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnUNUnif{}
-  \columnbreak\\
-  \ottdefnUPUnif{}
-  \end{multicols}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
-    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
-    adding the quantified variables to the context $[[Γ]]$.
-    Notice that $[[Θ]]$, which specifies
-    the contexts in which the algorithmic variables must be instantiated,
-    is not changed.
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
-    are the base cases. 
-    Since the sides are equal and free from algorithmic variables,
-    the unification returns an empty constraint. 
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
-    are symmetric cases constructing the constraints. 
-    When an algorithmic variable is unified with a type, 
-    we must check that the type is well-formed in the required context,
-    and if it is, we return a constraint restricting the variable
-    to be equivalent to that type.
-  \item \ruleref{\ottdruleUArrowLabel}.
-    In this case, the algorithm makes two recursive calls:
-    it unifies the arguments and the results of the arrows.
-    After that, the resulting constraints are merged using the
-    \emph{unification constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
-    \emph{unification} constraints, not arbitrary \emph{subtyping} 
-    constraints: it is important for modularizing the proofs, 
-    since the properties of the \emph{unification} constraint merge
-    can be proved independently from the \emph{subtyping} constraint merge.
-\end{itemize}
-
-\subsection{Least Upper Bound}
-\label{sec:lub}
-
-In this section, we present
-the algorithm finding the least common supertype of two positive types. 
-It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
-and indirectly, through the type upgrade by positive subtyping
-(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
-intuitive part of the algorithm, and its correctness will be covered
-in \cref{sec:alg-upper-bounds-proofs}.
-
-\begin{algorithm}[The Least Upper Bound Algorithm]
-  \hfill\\
-  \ottdefnLUBNsub{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleLUBVarLabel}
-    The base case is trivial: 
-    the least upper bound of to equal variables
-    is the variable itself.
-  \item \ruleref{\ottdruleLUBShiftLabel}
-    In case both sides of the least upper bound are shifted,
-    the algorithm needs to find the anti-unifier of them. 
-    Intuitively, this is because in general, the upper bounds of
-    $[[↓iN]]$ are $[[∃nas.iP]]$ such that 
-    $[[nas]]$ can be instantiated with some $[[iMs]]$ so that
-    $[[ Γ ⊢ [iMs/nas]iP ≈ ↓iN ]]$ (see \cref{lemma:shape-of-supertypes}).
-  \item \ruleref{\ottdruleLUBExistsLabel}
-    In this case, we move the quantified variables to the context $[[Γ]]$, 
-    and make a recursive call. 
-    It is important to make sure that $[[nas]]$ and $[[nbs]]$ are disjoint.
-    In this case, it is guaranteed that the resulting 
-    $[[fv(iQ)]]$ will be free of $[[nas]]$ and $[[nbs]]$,
-    and thus, the resulting type will be a supertype of both sides
-    (it will be discussed in \cref{lemma:shape-of-supertypes}).
-\end{itemize}
-
-In the positive subtyping algorithm (\cref{sec:subtyping}),
-\ruleref{\ottdruleAPUVarLabel} 
-generates a restriction of a variable $[[α̂⁺]]$.
-On the one hand, this restriction must imply 
-$[[â⁺ :≥ iP]]$ for the subtyping to hold.
-On the other hand, the type used in this restriction 
-must be well-formed in a potentially stronger (smaller) 
-context than $[[iP]]$.
-
-To resolve this problem, we define the \emph{upgrade} procedure,
-which for given $[[Δ]]$, $[[pnas]]$, and $[[Δ, pnas ⊢ iP]]$,
-finds $[[Δ ⊢ iQ]]$---the least supertype of $[[iP]]$ 
-among the types well-formed in $[[Δ]]$.
-
-The trick is to make sure that the `forbidden' variables
-$[[pnas]]$ are not used explicitly in the supertypes
-of $[[iP]]$. For this purpose, we
-construct new types $[[iP1]]$ and $[[iP2]]$,
-in each of them replacing the forbidden variables
-with fresh variables $[[pnbs]]$ and $[[pncs]]$,
-and then find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
-It turns out that this renaming forces the common types of 
-$[[iP1]]$ and $[[iP2]]$ to be agnostic to $[[pnas]]$,
-and thus, the supertypes of $[[iP]]$ well-formed in $[[Δ]]$
-are exactly the common supertypes of $[[iP1]]$ and $[[iP2]]$.
-These properties are considered in more details in \cref{sec:upgrade-lemmas}.
-
-\begin{algorithm}[Type Upgrade]
-  \hfill\\
-  \ottdefnLUBUp{}
-\end{algorithm}
-
-\begin{paragraph}{Note on the Greatest Lower Bound}
-  In contrast to the least upper bound, 
-  the general greatest lower bound does not exist in our system.
-  For instance, consider a positive type $[[iP]]$, 
-  together with its non-equivalent
-  supertypes $[[iP1]]$ and $[[iP2]] \not\simeq [[iP1]]$
-  (for example, $[[iP = ↓↑↓γ⁻]]$, $[[iP1 = ∃α⁻.↓↑↓α⁻]]$, 
-  and $[[iP2 = ∃α⁻.↓α⁻]]$).
-  Then for arbitrary $[[iQ]]$ and $[[iN]]$, 
-  let us consider the common subtypes of 
-  $A = [[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $B = [[iP → ↓↑iP1 → ↓↑iP2 → iN]]$.
-  It is easy to see that $[[∀α⁺.∀β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and 
-  $[[∀α⁺.∀β⁺. α⁺ → ↓↑β⁺ → ↓↑α⁺ → iN]]$ are
-  both \emph{maximal} common subtypes of $A$ and $B$,
-  and since they are not equivalent, none of them is 
-  the \emph{greatest} one.
-
-  However, we designed the subtyping system in such a way 
-  that the greatest lower bound is not needed:
-  the negative variables are always `protected'
-  by \emph{invariant} shifts ($[[↑]]$ and $[[↓]]$), 
-  and thus, the algorithm can only require
-  a substitution of a negative variable to be 
-  \emph{equivalent} to some type but never 
-  to be a \emph{subtype}.
-\end{paragraph}
-
-\subsection{Anti-unification}
-\label{sec:antiunification}
-
-Next, we define the anti-unification procedure,
-also known as the \emph{most specific generalization}.
-As an input, it takes two declarative types
-(e.g., in the positive case $[[iP1]]$ and $[[iP2]]$)
-and a context $[[Γ]]$.
-and returns a type $[[uQ]]$---the generalizer,
-containing negative placeholders 
-(represented by algorithmic variables) from 
-$[[Ξ]]$ and two substitutions $[[uτ1]]$ and $[[uτ2]]$.
-The substitutions replace the placeholders with 
-declarative types well-formed in $[[Γ]]$,
-such that $[[ [uτ1]uQ = iP1 ]]$ and $[[ [uτ2]uQ = iP2 ]]$.
-Moreover, the algorithm guarantees that 
-$[[uQ]]$ is the most specific type with this property: 
-any other generalizer can be turned into $[[uQ]]$ by some substitution
-$[[uρ]]$.
-
-It is important to note the differences between 
-the standard anti-unification and our version.
-First, we only allow the placeholders at \emph{negative} positions,
-which means, for example, that $[[α⁺]]$ and $[[β⁺]]$ cannot be
-generalized. Second, the generated pair of substitutions 
-$[[uτ1]]$ and $[[uτ2]]$ must replace the placeholders with 
-types well-formed in a specified context $[[Γ]]$.
-
-The anti-unification algorithm assumes that the input types 
-are normalized. This way, anti-unification up-to-equality rather than 
- anti-unification up-to-equivalence is sufficient.
-
-\begin{algorithm}[Anti-unification]
-  \hfill
-
-  \ottdefnAUAUP{}
-  \ottdefnAUAUN{}
-
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleAUPVarLabel} and \ruleref{\ottdruleAUNVarLabel}
-    are the base cases. 
-    In this case, since the input types are equal, 
-    the algorithm returns this type as a generalizer,
-    without generating any placeholders.
-
-  \item \ruleref{\ottdruleAUShiftDLabel}, \ruleref{\ottdruleAUShiftULabel},
-    \ruleref{\ottdruleAUForallLabel}, and \ruleref{\ottdruleAUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers.
-    Notice that the algorithm does not add the removed variables to
-    the context $[[Γ]]$. This is because $[[Γ]]$
-    is used to restrict the resulting anti-unification substitutions, 
-    and is fixed throughout the algorithm.
-
-  \item \ruleref{\ottdruleAUAULabel} is the most important rule, 
-    since it generates the placeholders. 
-    This rule only applies if other negative rules failed.
-    Because of that, the anti-unification procedure is 
-    \emph{not} syntax-directed.
-
-    The generated placeholder is indexed with a pair of 
-    types it is mapped to. It allows the algorithm to 
-    automatically unite the anti-unification solutions 
-    generated by the different branches of 
-    \ruleref{\ottdruleAUArrowLabel}.
-
-    Notice that this rule does not have a positive counterpart,
-    since we only allow negative placeholders.
-
-  \item \ruleref{\ottdruleAUArrowLabel}
-    makes two recursive calls to the anti-unification procedure,
-    and unites the results. Suppose that
-    $[[uτ1]]$  and $[[uτ2]]$ are the substitutions generated by
-    anti-unification of \emph{argument} types of the arrow,
-    and $[[uτ1']]$ and $[[uτ2']]$ are the substitutions generated by 
-    anti-unification of \emph{result} types of the arrow.
-    It is important that if ($[[uτ1]]$, $[[uτ2]]$)
-    and ($[[uτ1']]$, $[[uτ2']]$) send some variables to the same pair of types,
-    i.e., $[[ [uτ1]α̂⁻ = [uτ1']β̂⁻]]$ and $[[ [uτ2]α̂⁻ = [uτ2']β̂⁻]]$,
-    then these variables are equal, i.e., $[[α̂⁻ = β̂⁻]]$.
-    This property is guaranteed by \ruleref{\ottdruleAUAULabel}:
-    the name of the placeholder is determined by the pair of 
-    types it is mapped to.
-\end{itemize}
-
-\newpage
-
-\section{Declarative Typing}
-
-In the previous section, we presented the 
-type system together with subtyping specification 
-and the algorithm. In this section, 
-we describe the language under this type system, 
-together with the type inference specification
-and algorithm.
-
-\subsection{Grammar}
-
-First, we define the syntax of the language.
-The language combines System F with call-by-push-value 
-style. 
-\begin{definition}[Language Grammar]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottc\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottv\ottinterrule
-    }
-  \end{multicols}
-\end{definition}
-Notice that the language does not have
-first-class applications: instead, 
-we use applicative let bindings---
-constructions that bind a result of
-a fully applied function to a (positive) 
-variable.
-In the call-by-push-value paradigm, 
-it corresponds to monadic bind or do-notation. 
-Typewise, these let-binders come in two forms:
-annotated and unannotated. The annotated let-binders
-$[[let x:iP = v(args); c]]$ 
-requires the application to infer the annotated
-$[[iP]]$, whereas the unannotated 
-$[[let x = v(args); c]]$ 
-is used when the inferred type is unique.
-
-A computation of a polymorphic type is constructed 
-using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
-is implicit. Conversely, the existential types
-are constructed implicitly and eliminated 
-using the standard unpack mechanism:
-$[[let∃ (nas, x) = v; c]]$.
-
-Another dual pair of constructions are 
-$[[return v]]$ and $[[{c}]]$. 
-The former allows us to embed a value in a pure 
-computations. The latter, on the contrary,
-encapsulates a thunk of computation in a value.
-
-Finally, the language has a number standard constructions:
-lambda-abstractions $[[λx:iP.c]]$,
-standard let-bindings $[[let x = v; c]]$,
-and type annotations that can be added to any value or computation:
-$[[(v:iP)]]$ and $[[(c:iN)]]$.
-
-\subsection{Declarative Type Inference}
-
-Next, we define the specification of the type 
-inference for our language.
-The judgement $[[Γ ; Φ ⊢ v : iP]]$
-means that under the type context $[[Γ]]$ and
-the variable context $[[Φ]]$, for the value $[[v]]$,
-type $[[iP]]$ is inferrable. The negative form of the judgement is symmetric.
-The application inference judgement $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ means that if
-a head of type $[[iN]]$ is applied to list of values $[[args]]$,
-then the resulting computation can be typed as $[[iM]]$.
-
-\begin{definition}[Variable Context]
-  The variable typing context $[[Φ]]$
-  is represented by a set of entries of the form
-  $[[x : iP]]$. 
-\end{definition}
-
-\begin{definition}[Declarative Type Inference]
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnDTNInf{}
-  \ottdefnDTPInf{}
-  \ottdefnDTSpinInf{}
-  \end{multicols}
-\end{definition}
-
-Let us discuss selected rules of the declarative system:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTVarLabel}
-    says that the type of a variable is inferred from the context.
-  \item \ruleref{\ottdruleDTThunkLabel} says
-    that the type of a thunk is inferred by shifting up the type of the 
-    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
-    infers the type of a return by shifting down the type of the
-    contained value.
-  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
-    They allow the inferred type to be refined by annotating it with a supertype.
-  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
-    mean that th declarative system allows to infer any type from the equivalence class.
-  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
-    and its first premise inferring the existential type of the value being unpacked.
-    It is important however that the inferred existential type is normalized. 
-    This is because there might be multiple equivalent existential types 
-    with different order or even number of quantified variables, 
-    and to bind them, the algorithm needs to fix the canonical one.
-  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
-    with annotated let-bindings. The first premise infers the type of the head of the application,
-    which must be a thunked computation. Then if after applying it 
-    to tha arguments, the resulting type can be equated to the annotated one,
-    we infer the body of the let-binding in the context extended with the bound variable.
-  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
-    but it is used when the type of the application is unique, and thus, the annotation
-    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
-    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
-\end{itemize}
-
-Let us discuss the rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTEmptyAppLabel} 
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head. 
-    However, we relax this specification by allowing it to 
-    infer any other equivalent type. 
-    The relaxation of this rule is enough to guarantee 
-    this property for the whole judgement:
-    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
-    $[[iM']]$.
-  \item \ruleref{\ottdruleDTArrowAppLabel}
-    is where the application type is inferred: 
-    if the head has an arrow type $[[iQ → iN]]$,
-     we are allowed to apply it as soon as 
-    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
-  \item \ruleref{\ottdruleDTForallAppLabel}
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    If we are applying a polymorphic computation, 
-    we can instantiate its quantified variables with any types,
-    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
-\end{itemize}
-
-\section{Algorithmic Typing}
-
-\subsection{Algorithmic Type Inference}
-\begin{algorithm}
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnATNInf{}
-
-  \columnbreak
-  \ottdefnATPInf{}
-  \ottdefnATSpinInf{}
-  \end{multicols}
-\end{algorithm}
-
-\subsection{Constraint Singularity}
-
-\begin{algorithm}
-  \hfill
-  \begin{multicols}{2}
-    \ottdefnSINGscEN{}
-
-    \columnbreak
-    \ottdefnSINGscEP{}
-  \end{multicols}
-\end{algorithm}
-
-\newpage
-
-\section{Properties of the Declarative Type System}
-
-\subsection{Type Well-formedness}
-\input{../_build/_gen/tex/wf-lemmas.tex}
-
-\subsection{Substitution}
-\input{../_build/_gen/tex/subst-lemmas.tex}
-
-\subsection{Declarative Subtyping}
-\input{../_build/_gen/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Equivalence}
-\label{sec:decl-equiv-lemmas}
-\input{../_build/_gen/tex/equiv-lemmas.tex}
-
-\subsection{Variable Ordering}
-\input{../_build/_gen/tex/ord-lemmas.tex}
-
-\subsection{Normaliztaion}
-\input{../_build/_gen/tex/norm-lemmas.tex}
-
-\section{Properties of the Algorithmic Type System}
-
-\subsection{Algorithmic Type Well-formedness}
-\input{../_build/_gen/tex/wf-algo-lemmas.tex}
-
-\subsection{Substitution}
-\input{../_build/_gen/tex/subst-algo-lemmas.tex}
-
-\subsection{Normalization}
-\input{../_build/_gen/tex/norm-alg-lemmas.tex}
-
-\subsection{Equivalence}
-\input{../_build/_gen/tex/equiv-alg-lemmas.tex}
-
-\subsection{Unification Constraint Merge}
-\input{../_build/_gen/tex/unif-constraint-merge-lemmas.tex}
-
-\subsection{Unification}
-\input{../_build/_gen/tex/unification-lemmas.tex}\
-
-\subsection{Anti-unification}
-\input{../_build/_gen/tex/au-lemmas.tex}
-
-\subsection{Upper Bounds}
-\label{sec:alg-upper-bounds-proofs}
-\input{../_build/_gen/tex/lub-lemmas.tex}
-
-\subsection{Upgrade}
-\label{sec:upgrade-lemmas}
-\input{../_build/_gen/tex/upgrade-lemmas.tex}
-
-\subsection{Constraint Satisfaction}
-\input{../_build/_gen/tex/constraint-sat-lemmas.tex}
-
-\subsection{Positive Subtyping}
-\input{../_build/_gen/tex/alg-pos-subtyping-lemmas.tex}
-
-\subsection{Subtyping Constraint Merge}
-\input{../_build/_gen/tex/constraint-merge-lemmas.tex}
-
-\subsection{Negative Subtyping}
-\input{../_build/_gen/tex/alg-neg-subtyping-lemmas.tex}
-
-\section{Properties of the Declarative Typing}
-\input{../_build/_gen/tex/decl-typing-lemmas.tex}
-
-\section{Properties of the Algorithmic Typing}
-
-\subsection{Singularity}
-\input{../_build/_gen/tex/singularity-lemmas.tex}
-
-\subsection{Correctness of the Typing Algorithm}
-\input{../_build/_gen/tex/alg-typing-lemmas.tex}
-
-\printbibliography
-
-\end{document}

### 2023-09-29 proofs/main.tex
Resolve reference issues

@@
 type well-formedness (\cref{alg:wf}), 
-and normalization (\cref{alg:alg:type-nf}) judgements.
+and normalization (\cref{alg:type-nf}).
 \begin{itemize}
@@
 normalization (\cref{alg:type-nf}),
-and constraint singularity which will be defined later in \cref{todo}.
+and constraint singularity which will be defined later in 
+\cref{sec:constraint-singularity}.
 It also relies on basic set operations and the ability to 
@@
 \subsection{Constraint Singularity}
+\label{sec:constraint-singularity}
 

### 2023-09-29 proofs/tex/decl-subtyping-lemmas.tex
Resolve reference issues

@@
       and hence, the induction hypothesis applies
-      ($[[Γ1, pbs ⊢ σ : Γ1, pbs]]$ by  \label{lemma:subst-domain-weakening})
+      ($[[Γ1, pbs ⊢ σ : Γ1, pbs]]$ by \cref{lemma:subst-domain-weakening})
       :

### 2023-09-29 proofs/tex/equiv-lemmas.tex
Resolve reference issues

@@
 
-  \label{lemma:poly-types-equivalence}
       Notice that the application of bijection $[[μ]]$ to $[[iM']]$ does
@@
 \begin{lemma}[Substitution preserves declarative equivalence]
-  \label{corollary:subst-pres-equiv}
+  \label{corollary:subst-pres-decl-equiv}
   Suppose that $[[σ]]$ is a substitution. Then

### 2023-09-29 proofs/tex/upgrade-lemmas.tex
Resolve reference issues

@@
     (since $[[pnbs/pnas]]$ can be specified as 
-    $[[Δ,pnbs,pncs ⊢ pnbs/pnas : Δ, pnas, pncs]]$ by \label{lemma:subst-domain-weakening})
+    $[[Δ,pnbs,pncs ⊢ pnbs/pnas : Δ, pnas, pncs]]$ by \cref{lemma:subst-domain-weakening})
     obtain $[[Δ, pnas, pncs ⊢ [pnas/pnbs]iQ ≥ iP]]$.

### 2023-10-01 oopsla24/paper.tex
Makefiles
Fix structure
Restructure, add biblio

@@
+\PassOptionsToPackage{prologue,dvipsnames}{xcolor}
+
 %%
@@
 % \documentclass[acmsmall,screen,review]{acmart}
-\documentclass[acmsmall,review,anonymous]{acmart}
+\documentclass[acmsmall,natbib=false,review,anonymous]{acmart}
 
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
+\input{../prelude.tex}
+
+\newcommand{\ottDir}{../ott/_gen}
+\newcommand{\genDir}{_gen}
+\input{\ottDir/defs.tex}
+\input{\ottDir/renew-ott.tex}
+
+\usepackage{../bibliography}
+
+% \RequirePackage[
+%       datamodel=acmdatamodel,
+%       style=acmauthoryear,backend=biber
+%       ]{biblatex}
+
+% \addbibresource{../biblio.bib}
 
@@
 \section{Introduction}
+
+$[[iN = iM]]$
+
 \input{tex/01-introduction.tex}
@@
 
-\subsection{Template Styles}
+\cite{botlanMLFRaisingML2003}
+\cite{dunfieldBidirectionalTyping2020}
 
-The primary parameter given to the ``\verb|acmart|'' document class is
-the {\itshape template style} which corresponds to the kind of publication
-or SIG publishing the work. This parameter is enclosed in square
-brackets and is a part of the {\verb|documentclass|} command:
-\begin{verbatim}
-  \documentclass[STYLE]{acmart}
-\end{verbatim}
-
-Journals use one of three template styles. All but three ACM journals
-use the {\verb|acmsmall|} template style:
-\begin{itemize}
-\item {\texttt{acmsmall}}: The default journal template style.
-\item {\texttt{acmlarge}}: Used by JOCCH and TAP.
-\item {\texttt{acmtog}}: Used by TOG.
-\end{itemize}
-
-The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
-\begin{itemize}
-\item {\texttt{acmconf}}: The default proceedings template style.
-\item{\texttt{sigchi}}: Used for SIGCHI conference articles.
-\item{\texttt{sigplan}}: Used for SIGPLAN conference articles.
-\end{itemize}
-
-\subsection{Template Parameters}
-
-In addition to specifying the {\itshape template style} to be used in
-formatting your work, there are a number of {\itshape template parameters}
-which modify some part of the applied template style. A complete list
-of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\texttt{anonymous,review}}: Suitable for a ``double-blind''
-  conference submission. Anonymizes the work and includes line
-  numbers. Use with the \texttt{\acmSubmissionID} command to print the
-  submission's unique ID on each page of the work.
-\item{\texttt{authorversion}}: Produces a version of the work suitable
-  for posting by the author.
-\item{\texttt{screen}}: Produces colored hyperlinks.
-\end{itemize}
-
-This document uses the following string as the first command in the
-source file:
-\begin{verbatim}
-\documentclass[acmsmall,screen,review]{acmart}
-\end{verbatim}
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting
-margins, typeface sizes, line spacing, paragraph and list definitions,
-and the use of the \verb|\vspace| command to manually adjust the
-vertical spacing between elements of your work --- is not allowed.
-
-{\bfseries Your document will be returned to you for revision if
-  modifications are discovered.}
-
-\section{Typefaces}
-
-The ``\verb|acmart|'' document class requires the use of the
-``Libertine'' typeface family. Your \TeX\ installation should include
-this set of packages. Please do not substitute other typefaces. The
-``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
-as they will override the built-in typeface families.
-
-\section{Title Information}
-
-The title of your work should use capital letters appropriately -
-\url{https://capitalizemytitle.com/} has useful rules for
-capitalization. Use the {\verb|title|} command to define the title of
-your work. If your work has a subtitle, define it with the
-{\verb|subtitle|} command.  Do not insert line breaks in your title.
-
-If your title is lengthy, you must define a short version to be used
-in the page headers, to prevent overlapping text. The \verb|title|
-command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
-
-\section{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata
-identification.  As an exception, multiple authors may share one
-affiliation. Authors' names should not be abbreviated; use full first
-names wherever possible. Include authors' e-mail addresses whenever
-possible.
-
-Grouping authors' names or e-mail addresses, or providing an ``e-mail
-alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
-
-The \verb|authornote| and \verb|authornotemark| commands allow a note
-to apply to multiple authors --- for example, if the first two authors
-of an article contributed equally to the work.
-
-If your author list is lengthy, you must define a shortened version of
-the list of authors to be used in the page headers, to prevent
-overlapping text. The following command should be placed just after
-the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all
-of the authors' names, which may result in overlapping text in the
-page headers.
-
-The article template's documentation, available at
-\url{https://www.acm.org/publications/proceedings-template}, has a
-complete explanation of these commands and tips for their effective
-use.
-
-Note that authors' addresses are mandatory for journal articles.
-
-\section{Rights Information}
-
-Authors of any work published by ACM will need to complete a rights
-form. Depending on the kind of work, and the rights management choice
-made by the author, this may be copyright transfer, permission,
-license, or an OA (open access) agreement.
-
-Regardless of the rights management choice, the author will receive a
-copy of the completed rights form once it has been submitted. This
-form contains \LaTeX\ commands that must be copied into the source
-document. When the document source is compiled, these commands and
-their parameters add formatted text to several areas of the final
-document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
-
-Rights information is unique to the work; if you are preparing several
-works for an event, make sure to use the correct set of commands with
-each of the works.
-
-The ACM Reference Format text is required for all articles over one
-page in length, and is optional for one-page articles (abstracts).
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful
-taxonomic tools for you to help readers find your work in an online
-search.
-
-The ACM Computing Classification System ---
-\url{https://www.acm.org/publications/class-2012} --- is a set of
-classifiers and concepts that describe the computing
-discipline. Authors can select entries from this classification
-system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
-commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases
-of the authors' choosing, providing a more flexible way of describing
-the research being presented.
-
-CCS concepts and user-defined keywords are required for for all
-articles over two pages in length, and are optional for one- and
-two-page articles (or abstracts).
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands:
-\verb|section|, \verb|subsection|, \verb|subsubsection|, and
-\verb|paragraph|. They should be numbered; do not remove the numbering
-from the commands.
-
-Simulating a sectioning command by setting the first word or words of
-a paragraph in boldface or italicized text is {\bfseries not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
-package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
-high-quality tables.
-
-Table captions are placed {\itshape above} the table.
-
-Because tables cannot be split across pages, the best placement for
-them is typically the top of the page nearest their initial cite.  To
-ensure this proper ``floating'' placement of tables, use the
-environment \textbf{table} to enclose the table's contents and the
-table caption.  The contents of the table itself must go in the
-\textbf{tabular} environment, to be aligned properly in rows and
-columns, with the desired horizontal and vertical rules.  Again,
-detailed instructions on \textbf{tabular} material are found in the
-\textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which
-Table~\ref{tab:freq} is included in the input file; compare the
-placement of the table here with the table in the printed output of
-this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's
-live area, use the environment \textbf{table*} to enclose the table's
-contents and the table caption.  As with a single-column table, this
-wide table will ``float'' to a location deemed more
-desirable. Immediately following this sentence is the point at which
-Table~\ref{tab:commands} is included in the input file; again, it is
-instructive to compare the placement of the table here with the table
-in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-Always use midrule to separate table header rows from data rows, and
-use it only for this purpose. This enables assistive technologies to
-recognise table headers and support their users in navigating tables
-more easily.
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of the three are
-discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an inline or
-in-text formula.  It is produced by the \textbf{math} environment,
-which can be invoked with the usual
-\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
-the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
-and structures, from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
-examples of in-text equations in context. Notice how this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols and
-structures available in \LaTeX\@; this section will just give a couple
-of examples of display equations in context.  First, consider the
-equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures. One or
-more images can be placed within a figure. If your figure contains
-third-party material, you must clearly identify it as such, as shown
-in the example below.
-\begin{figure}[h]
-  \centering
-  % \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
-    Ewing, Inc. [Public domain], via Wikimedia
-    Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{A woman and a girl in white dresses sit in an open car.}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to
-the reader.
-
-Figure captions are placed {\itshape below} the figure.
-
-Every figure should also have a figure description unless it is purely
-decorative. These descriptions convey what’s in the image to someone
-who cannot see it. They are also used by search engine crawlers for
-indexing images, and when images cannot be loaded.
-
-A figure description must be unformatted plain text less than 2000
-characters long (including spaces).  {\bfseries Figure descriptions
-  should not repeat the figure caption – their purpose is to capture
-  important information that is not already provided in the caption or
-  the main text of the paper.} For figures that convey important and
-complex new information, a short text description may not be
-adequate. More complex alternative descriptions can be placed in an
-appendix and referenced in a short figure description. For example,
-provide a data table capturing the information in a bar chart, or a
-structured list representing a graph.  For additional information
-regarding how best to write figure descriptions and why doing this is
-so important, please see
-\url{https://www.acm.org/publications/taps/describing-figures/}.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that
-are placed after all author and affiliation information, and before
-the body of the article, spanning the page. If you wish to have such a
-figure in your article, place the command immediately before the
-\verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's
-references is strongly recommended. Authors' names should be complete
---- use full first names (``Donald E. Knuth'') not initials
-(``D. E. Knuth'') --- and the salient identifying features of a
-reference should be included: title, year, volume, number, pages,
-article DOI, etc.
-
-The bibliography is included in your source document with these two
-commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
-  \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
-suffix, of the \BibTeX\ file.
-
-Citations and references are numbered by default. A small number of
-ACM publications have citations and references formatted in the
-``author year'' style; for these exceptions, please include this
-command in the {\bfseries preamble} (before the command
-``\verb|\begin{document}|'') of your \LaTeX\ source:
-\begin{verbatim}
-  \citestyle{acmauthoryear}
-\end{verbatim}
-
-  Some examples.  A paginated journal article \cite{Abril07}, an
-  enumerated journal article \cite{Cohen07}, a reference to an entire
-  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
-  monograph/whole book in a series (see 2a in spec. document)
-  \cite{Harel79}, a divisible-book such as an anthology or compilation
-  \cite{Editor00} followed by the same example, however we only output
-  the series if the volume number is given \cite{Editor00a} (so
-  Editor00a's series should NOT be present since it has no vol. no.),
-  a chapter in a divisible book \cite{Spector90}, a chapter in a
-  divisible book in a series \cite{Douglass98}, a multi-volume work as
-  book \cite{Knuth97}, a couple of articles in a proceedings (of a
-  conference, symposium, workshop for example) (paginated proceedings
-  article) \cite{Andler79, Hagerup1993}, a proceedings article with
-  all possible elements \cite{Smith10}, an example of an enumerated
-  proceedings article \cite{VanGundy07}, an informally published work
-  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
-    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
-  master's thesis: \cite{anisi03}, an online document / world wide web
-  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
-  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
-  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
-  publication \cite{rous08}, 'YYYYb'-test for prolific author
-  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
-  contain 'duplicate' DOI and URLs (some SIAM articles)
-  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
-  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
-  couple of citations with DOIs:
-  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
-  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
-  Artifacts: \cite{R} and \cite{UMassCitations}.
-
-\section{Acknowledgments}
-
-Identification of funding sources and other support, and thanks to
-individuals and groups that assisted in the research and the
-preparation of the work should be included in an acknowledgment
-section, which is placed just before the reference section in your
-document.
-
-This section has a special environment:
-\begin{verbatim}
-  \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected
-during the article metadata extraction phase, and to ensure
-consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the
-``\verb|\end{document}|'' command at the conclusion of your source
-document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-  \appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not
-numbered. This document has two appendices, demonstrating the section
-and subsection identification method.
-
-\section{Multi-language papers}
-
-Papers may be written in languages other than English or include
-titles, subtitles, keywords and abstracts in different languages (as a
-rule, a paper in a language other than English should include an
-English title and an English abstract).  Use \verb|language=...| for
-every language used in the paper.  The last language indicated is the
-main language of the paper.  For example, a French paper with
-additional titles and abstracts in English and German may start with
-the following command
-\begin{verbatim}
-\documentclass[sigconf, language=english, language=german,
-               language=french]{acmart}
-\end{verbatim}
-
-The title, subtitle, keywords and abstract will be typeset in the main
-language of the paper.  The commands \verb|\translatedXXX|, \verb|XXX|
-begin title, subtitle and keywords, can be used to set these elements
-in the other languages.  The environment \verb|translatedabstract| is
-used to set the translation of the abstract.  These commands and
-environment have a mandatory first argument: the language of the
-second argument.  See \verb|sample-sigconf-i13n.tex| file for examples
-of their usage.
-
-\section{SIGCHI Extended Abstracts}
-
-The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
-not in Word) produces a landscape-orientation formatted article, with
-a wide left margin. Three environments are available for use with the
-``\verb|sigchi-a|'' template style, and produce formatted output in
-the margin:
-\begin{description}
-\item[\texttt{sidebar}:]  Place formatted text in the margin.
-\item[\texttt{marginfigure}:] Place a figure in the margin.
-\item[\texttt{margintable}:] Place a table in the margin.
-\end{description}
-
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Research Methods}
-
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
-malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
-sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
-vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
-lacinia dolor. Integer ultricies commodo sem nec semper.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
-ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
-ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
-eros. Vivamus non purus placerat, scelerisque diam eu, cursus
-ante. Etiam aliquam tortor auctor efficitur mattis.
-
-\section{Online Resources}
-
-Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
-pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
-enim maximus. Vestibulum gravida massa ut felis suscipit
-congue. Quisque mattis elit a risus ultrices commodo venenatis eget
-dui. Etiam sagittis eleifend elementum.
-
-Nam interdum magna at lectus dignissim, ac dignissim lorem
-rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
-massa et mattis lacinia.
+\printbibliography
 
 \end{document}
+
 \endinput

### 2023-10-04 oopsla24/tex/03-declarative-system.tex
Restructure, add biblio
Restructure

@@
 
-\begin{definition}[Declarative Types]
-  \hfill
+The type syntax of \fexists is given in \cref{fig:declarative-types}.
+The types of \fexists are stratified into two syntactic 
+categories (polarities): positive and negative,  
+similarly to the \CBPV system \cite{levy2006:cbpv}.
+The negative types represent computations, and the positive types represent values:
+\begin{itemize}
+\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
+\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
+\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
+  a list of positive type variables 
+  $[[pas]]$. The polarities are chosen to follow the definition of functions.
+\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
+  which at the term-level corresponds to a pure computation $[[return v]]$.
+\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
+\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
+  binds negative variables in a positive type $[[iP]]$. 
+\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
+  thunks a computation, which at the term-level corresponds to $[[ {c} ]]$.
+\end{itemize}
+
+\begin{figure}[h]
   \begin{multicols}{2}
@@
   \end{multicols}
-\end{definition}
+  \caption{Declarative Types of \fexists}
+  \label{fig:declarative-types}
+\end{figure}

### 2023-10-04 oopsla24/tex/03-declarative-system.tex
add plan

@@
 
-The type syntax of \fexists is given in \cref{fig:declarative-types}.
-The types of \fexists are stratified into two syntactic 
+The types of \fexists are given in \cref{fig:declarative-types}.
+They are stratified into two syntactic 
 categories (polarities): positive and negative,  
@@
 \end{figure}
+
+\paragraph{Definitional Equalities}
+For simplicity, we assume alpha-equivalent terms equal.
+This way, we assume that substitutions do not capture bound variables.
+Besides, we equate
+$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
+as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
+and lift these equations transitively and congruently 
+to the whole system.

### 2023-10-05 oopsla24/paper.tex
Restructure
WIP

@@
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+  {\color{red}CHANGE!!}
+  This paper addresses the challenging problem of type inference for
+  Impredicative System F with existential types, a critical aspect of many
+  programming languages. While System F serves as the basis for type systems in
+  numerous languages, existing type inference techniques for Impredicative
+  System F are undecidable due to the presence of existential ($[[∃]]$) 
+  and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
+  sub-optimal. This paper presents novel contributions in the form of a local
+  type inference algorithm for Impredicative System F with existential types.
+  The algorithm introduces innovative techniques, such as a unique combination
+  of unification and anti-unification, a full correctness proof, and the use of
+  control structures inspired by \CBPV. Additionally, the paper discusses a type
+  inference framework that allows the algorithm to be applied to different type
+  systems, offering insights into the under-researched area of impredicative
+  existential type inference. 
 \end{abstract}
@@
 
-$[[iN = iM]]$
-
-\input{tex/01-introduction.tex}
+\input{\genDir/tex/01-introduction.tex}
 
 \section{Overview}
-\input{tex/02-overview.tex}
+\input{\genDir/tex/02-overview.tex}
 
 \section{Declarative System}
-\input{tex/03-declarative-system.tex}
+\input{\genDir/tex/03-declarative-system.tex}
 
 \section{Algorithm}
-\input{tex/04-algorithm.tex}
+\input{\genDir/tex/04-algorithm.tex}
 
 \section{Proof}
-\input{tex/05-proof.tex}
+\input{\genDir/tex/05-proof.tex}
 
 \section{Extensions}
-\input{tex/06-extensions.tex}
+\input{\genDir/tex/06-extensions.tex}
 
 \section{Conclusion}
-\input{tex/07-conclusion.tex}
+\input{\genDir/tex/07-conclusion.tex}
 
@@
 
+%% Bibliography
+%% \balance %% necessary for two-column formats
 \printbibliography
+
+\lstset{basicstyle=\ttfamily\scriptsize,keywordstyle=\color{black}\bfseries\underbar,tabsize=8,showstringspaces=false,frame=L,label= ,caption= ,captionpos=b,numbers=none}
+
+\clearpage
+\appendix
+% \input{./tex/appendix/recovery.tex}
+% \input{./tex/appendix/drf.tex}
 

### 2023-10-09 oopsla24/tex/02-overview.tex
WIP

--- 2023-10-04 oopsla24/tex/03-declarative-system.tex
@@
-\subsection{The Language}
+\subsection{What types do we infer?}
+
+\subsection{The Language of Types}
 
@@
     }
-
     \ottgrammartabular{
@@
 to the whole system.
+
+\subsection{The Language of Terms}
+
+In \cref{fig:declarative-terms}, we define the the language of terms of 
+\fexists. The language combines \systemf with the \CBPV approach.
+
+\begin{itemize}
+    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
+        \ilyam{why no negatives?}
+    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
+        or suspended computation;
+    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
+        positive and negative terms;
+    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
+    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
+        are standard lambda abstractions. Notice that we require
+        the type annotation for the argument of $[[λ]]$;
+    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
+        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
+    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
+        $[[let x = v ( args ) ; c]]$ operate similarly to 
+        the bind of a monad: they take a suspended computation $[[v]]$,
+        apply it to a list of arguments, bind the result 
+        (which is expected to be pure) to a variable $[[x]]$,
+        and continue with a computation $[[c]]$.
+        If the resulting type of the application is unique, 
+        one can omit the type annotation, as in the second form:
+        it will be inferred by the algorithm;
+    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
+        is the standard unpack of an existential type:
+        expecting $[[v]]$ to be an existential type,
+        it binds the packed negative types to a list of 
+        variables $[[nas]]$, binds the body of the existential
+        to $[[x]]$, and continues with a computation $[[c]]$.
+\end{itemize}
+
+\paragraph{Missing constructors}
+Notice that the language does not have first-class applications: 
+their role is played by the applicative let forms, binding 
+the result of a \emph{fully applied} function to a variable.
+Also notice that the language does not have a type application 
+(i.e. the eliminator of $[[∀]]$) and dually, 
+it does not have \pack (i.e. the constructor of $[[∃]]$).
+This is because the instantiation of polymorphic and existential 
+types is inferred by the algorithm.  \ilyam{refer to the extension chapter}
+
+\begin{figure}[h]
+  \begin{multicols}{2}
+    \ottgrammartabular{
+      \ottc\ottinterrule
+    }
+
+    \ottgrammartabular{
+      \ottv\ottinterrule
+    }
+  \end{multicols}
+  \caption{Declarative Terms of \fexists}
+  \label{fig:declarative-terms}
+\end{figure}
+
+\subsection{The key ideas of the algorithm}

### 2023-10-09 oopsla24/tex/03-declarative-system.tex

--- 2023-09-28 oopsla24/tex/03-declarative-system.tex
@@
-\subsection{The Language}
+The declarative system serves as a specification of the
+type inference algorithm. It consists of two main parts:
+the subtyping and the type inference.
 
-\begin{definition}[Declarative Types]
+\subsection{Subtyping}
+It is represented by a set of inference rules shown in
+\cref{fig:declarative-subtyping}.
+
+\begin{figure}[h]
+  \begin{multicols}{2}
+    \ottdefnDOneNsub{}
+
+    \ottdefnDOnePsup{}
+  \end{multicols}
+  \hfill\\
+  \begin{multicols}{2}
+    \ottdefnDOneNeq{}
+
+    \ottdefnDOnePeq{}
+  \end{multicols}
+  \caption{Declarative Subtyping}
+  \label{fig:declarative-subtyping}
+\end{figure}
+
+\paragraph{Invariance of Shifts}
+An important restriction that we put on the subtyping system is
+that the subtyping on shifted types requires their equivalence,
+as shown in \ruleref{\ottdruleDOneShiftDLabel} and
+\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
+invariants makes the system equivalent to \systemf, 
+and thus, undecidable. 
+However, \ruleref{\ottdruleDOneShiftDLabel} might be
+relaxed to a covariant form, as we will discuss in \cref{todo}.
+
+As discussed in \cref{todo}, 
+the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
+\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
+For convenience of representation, we compose the left-hand side rule and
+the right-hand side rule into one, and use substitution $[[σ]]$ to 
+represent instantiation.
+
+The substitution application is defined 
+in a standard way, avoiding capture of bound variables,
+and preserving the variables that are out of the substitution domain.
+The domain and the range of the substitutions are
+specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
+For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
+$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
+well-formed in $[[Γ, pbs]]$.
+
+\subsection{Type Inference}
+
+\begin{figure}[h]
+  \ottdefnDTNInf{}
+
   \hfill
+
   \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottiN\ottinterrule
-    }
+  \ottdefnDTPInf{}
+  \\
+  \ottdefnDTSpinInf{}
+  \end{multicols}
+  \hfill
 
-    \ottgrammartabular{
-      \ottiP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
+  \caption{Declarative Subtyping}
+  \label{fig:declarative-inference}
+\end{figure}
+
+\subsection{Properties of the Declarative System}
+
+Now, we present selected properties of the declarative system,
+which are important for the correctness of the algorithm.
+
+\begin{lemma}
+  Variables do not have proper subtypes and supertypes
+\end{lemma}
+
+\begin{lemma}
+  Free Variable propagation: $[[Γ ⊢ iN ≤ iM]]$ 
+  implies $[[fv(iN) ⊆ fv(iM)]]$.
+\end{lemma}
+
+\begin{lemma}
+  Subtyping is Reflexive and Transitive, and is preserved by substitution
+\end{lemma}
+
+\begin{lemma}
+  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) ≈ nf(iM)]]$.
+\end{lemma}
+
+\begin{lemma}[Characterization of Supertypes]
+\end{lemma}

### 2023-10-11 oopsla24/tex/03-declarative-system.tex
Wip: declarative system

@@
 
-\paragraph{Invariance of Shifts}
-An important restriction that we put on the subtyping system is
-that the subtyping on shifted types requires their equivalence,
-as shown in \ruleref{\ottdruleDOneShiftDLabel} and
-\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
-invariants makes the system equivalent to \systemf, 
-and thus, undecidable. 
-However, \ruleref{\ottdruleDOneShiftDLabel} might be
-relaxed to a covariant form, as we will discuss in \cref{todo}.
-
-As discussed in \cref{todo}, 
+\paragraph{Quantifiers} As discussed in \cref{todo}, 
 the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
@@
 
-The substitution application is defined 
+\paragraph{Substitutions} The substitution application is defined 
 in a standard way, avoiding capture of bound variables,
@@
 
+\paragraph{Invariance of the Shifts}
+An important restriction that we put on the subtyping system is
+that the subtyping on shifted types requires their equivalence,
+as shown in \ruleref{\ottdruleDOneShiftDLabel} and
+\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
+invariants make the system equivalent to \systemf, 
+and thus, undecidable. 
+However, \ruleref{\ottdruleDOneShiftDLabel} might be
+relaxed to a covariant form, as we will discuss in \cref{todo}.
+
 \subsection{Type Inference}
+
+The declarative type inference system is shown in \cref{fig:declarative-inference}.
+The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
+``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
+the term $[[v]]$ is allowed to have the type $[[iP]]$'',
+where $[[Φ]]$---the variable context---is defined standardly as
+a set of pairs of the form $[[x : iP]]$. 
+The negative typing judgment is read similarly.
+To infer the resulting type of an application, the
+applicative let rules refer to the \emph{Application typing} judgment.
+It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
+which is read as ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
+the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
+is allowed to have the type $[[iM]]$''.
 
@@
 
+\paragraph{Non-bidirectionality}
+  Notice that all the rules in our system are inferring.
+
+\paragraph{Typing up to equivalence}
+  As discussed in \cref{todo}, the subtyping, as a preorder, 
+  induces a non-trivial equivalence relation on types. 
+  The system must not distinguish between equivalent types,
+  and thus, type inference must be defined up to equivalence. 
+  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
+  and \ruleref{\ottdruleDTNEquivLabel}.
+  The application typing is defined up to equivalence as well,
+  although it has no special rules for that.
+
+\paragraph{Unpack}
+  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
+  First, it infers the normalized type of the existential package.
+  The normalization is required to fix the order of the quantifying variables
+  to bind them. After the bind, the rule infers the type of the body 
+  and checks that it does not use the bound variables so that they do not 
+  escape the scope.
+
+\paragraph{Applicative Let Binders}
+  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
+  infer the type of the applicative let binders.
+  Both of them infer the type of the head $[[v]]$ 
+  and invoke the application typing to infer the type of the application 
+  before recursing on the body of the let binder.
+  The difference is that the former rule is for the 
+  \emph{unannotated} let binder, and thus
+  it requires the resulting type of application to be unique
+  (up to equivalence), so that the type of the bound variable $[[x]]$
+  is known before it is put into the context.
+  The latter rule is for the \emph{annotated} binder,
+  and thus, the type of the bound $[[x]]$ is given, 
+  however, the rule must check that this type is a
+  a supertype of the inferred type of the application. 
+  This check is done by invoking the subtyping judgment
+  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
+  This judgment is more restrictive than checking bare 
+  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
+  to make the algorithm complete as it allows us to preserve
+  certain invariants (see \cref{todo}).
+  In \cref{todo} we discuss how this restriction can be relaxed
+  together with invariant shift subtyping.
+
+\paragraph{Application of a polymorphic type $[[∀]]$}
+  The complexity of the system is hidden in the rules, 
+  whose output type is not immediately defined by their input
+  and the output of their premises 
+  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
+  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
+  the instantiation of the quantifying variables is
+  taken out of thin air. The algorithm we present in \cref{todo} 
+  delays this instantiation until more information about it 
+  (in particular, typing constraints) is collected.
+
+\paragraph{Application of an Arrow Type}
+  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
+  This is where the subtyping is used to check that the type of the argument
+  is convertible to (a subtype of) the type of the function parameter.
+  In the algorithm (\cref{todo}), this subtyping check will provide the constraints
+  we need to resolve the delayed instantiations of the quantifying variables.
+
+\paragraph{Annotations}
+  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
+  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
+  inferred type is a subtype of the annotation type.
+
 \subsection{Properties of the Declarative System}
 
-Now, we present selected properties of the declarative system,
+Now we present selected properties of the declarative system,
 which are important for the correctness of the algorithm.
 
-\begin{lemma}
-  Variables do not have proper subtypes and supertypes
-\end{lemma}
+As we discussed in \cref{todo}, the equivalence induced by the subtyping is 
+not trivial. To deal with it without modifying standard operations such as unification, 
+we use normalization. This way, it is important for the normalization to be 
+sound and complete \wrt the equivalence:
 
-\begin{lemma}
-  Free Variable propagation: $[[Γ ⊢ iN ≤ iM]]$ 
-  implies $[[fv(iN) ⊆ fv(iM)]]$.
-\end{lemma}
+\begin{property}[Correctness of normalization]
+  For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
+  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$.
+\end{property}
 
-\begin{lemma}
-  Subtyping is Reflexive and Transitive, and is preserved by substitution
-\end{lemma}
+Another property that we extensively use is that the subtyping is reflexive and transitive. 
+Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.
 
-\begin{lemma}
-  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) ≈ nf(iM)]]$.
-\end{lemma}
+\begin{property}[Subtyping is well-defined]
+  The subtyping is a preorder and is preserved under substitutions. 
+\end{property}
 
-\begin{lemma}[Characterization of Supertypes]
-\end{lemma}
+However, the greatest lower bound of two positive types is not always defined.
+Let us consider a positive type $[[iP]]$,
+and two different supertypes $[[iP1]]$ and $[[iP2]]$
+(\ie  $[[· ⊢ iP1 ≥ iP]]$, $[[· ⊢ iP2 ≥ iP]]$, and $[[nf(iP1) ≠ nf(iP2)]]$.
+Let us also consider an arbitrary negative type $[[iN]]$.

### 2023-10-12 oopsla24/tex/03-declarative-system.tex
wip

@@
 
-\paragraph{Quantifiers} As discussed in \cref{todo}, 
-the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
-\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
-For convenience of representation, we compose the left-hand side rule and
-the right-hand side rule into one, and use substitution $[[σ]]$ to 
-represent instantiation.
-
-\paragraph{Substitutions} The substitution application is defined 
-in a standard way, avoiding capture of bound variables,
-and preserving the variables that are out of the substitution domain.
-The domain and the range of the substitutions are
-specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
-For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
+\paragraph{Quantifiers}  
+Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
+\ruleref{\ottdruleDOneExistsLabel} specify 
+the subtyping between top-level quantified types.
+Usually, the polymorphic subtyping is represented by two rules
+introducing quantifiers to the left and to the right-hand side of the subtyping.
+For conciseness of representation, we compose these rules into one.
+First, our rule extends context $[[Γ]]$ with the quantified variables 
+from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
+as these variables must remain abstract.
+Second, it verifies that the left-hand side quantifiers
+($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
+The instantiation is modeled by substitution $[[σ]]$.
+The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
+For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
 $[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
 well-formed in $[[Γ, pbs]]$.
+This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
+$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.
 
-\paragraph{Invariance of the Shifts}
+\paragraph{Invariant Shifts}
 An important restriction that we put on the subtyping system is
@@
 and thus, undecidable. 
-However, \ruleref{\ottdruleDOneShiftDLabel} might be
-relaxed to a covariant form, as we will discuss in \cref{todo}.
+However, after certain changes 
+\ruleref{\ottdruleDOneShiftDLabel} can be relaxed to the covariant form,
+ as we will discuss in \cref{todo}.
+
+\paragraph{Functions}
+Standardly, the subtyping of function types is covariant in the return type
+and contravariant in the argument type.
+
+\paragraph{Variables}
+The subtyping of variables is defined reflexively,
+which is enough to ensure the reflexivity of subtyping in general.
+The algorithm will use the fact that the subtypes of a variable 
+coincide with its supertypes, which however is not true for an
+arbitrary type.
+
+A property that we extensively use is that the subtyping is reflexive and transitive. 
+Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.
+
+\begin{property}[Subtyping is well-defined]
+  The subtyping forms a preorder and is preserved under substitutions. 
+\end{property}
+
+\subsection{Equivalence and Normalization}
+
+The subtyping relation forms a preorder on types,
+and thus, it induces an equivalence relation \aka \emph{bicoercibility}.
+The declarative specification of subtyping must be defined up to this equivalence.
+Moreover, the algorithms we use must withstand changes in input types within the equivalence class.
+To deal with this, we use normalization---a function that uniformly selects a representative of the equivalence class.
+
+Using normalization gives us two benefits:
+\begin{enumerate*}
+  \item [(i)] we do not need to modify significantly standard operations such as unification to withstand non-trivial equivalence, and
+  \item [(ii)] once the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
+    while the rest of the algorithm remains the same.
+\end{enumerate*}
+
+\begin{figure}[h]
+  \label{alg:type-nf}
+  \hfill
+
+  \begin{multicols}{2}
+  \ottdefnNrmNNorm{}
+  \columnbreak\\
+  \ottdefnNrmPNorm{}
+  \end{multicols}
+  \caption{Type Normalization}
+  \label{fig:type-nf}
+\end{figure}
 
@@
 
-Another property that we extensively use is that the subtyping is reflexive and transitive. 
-Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.
-
-\begin{property}[Subtyping is well-defined]
-  The subtyping is a preorder and is preserved under substitutions. 
-\end{property}
-
 However, the greatest lower bound of two positive types is not always defined.

### 2023-10-12 oopsla24/tex/01-introduction.tex
some narrative for the intro

@@
-ACM's consolidated article template, introduced in 2017, provides a
-consistent \LaTeX\ style for use across ACM publications, and
-incorporates accessibility and metadata-extraction functionality
-necessary for future Digital Library endeavors. Numerous ACM and
-SIG-specific \LaTeX\ templates have been examined, and their unique
-features incorporated into this single new template.
+% How we should frame this:
 
-If you are new to publishing with ACM, this document is a valuable
-guide to the process of preparing your work for publication. If you
-have published with ACM before, this document provides insight and
-instruction into more recent changes to the article template.
+\begin{itemize}
+\item There has been very little work on type inference for existentials. 
+\item The state of the art in languages like Ocaml and Haskell is the
+  Odersky-L\"{a}ufer algorithm, which makes existentials a
+  second-class construct tied to datatypes.
+\item More recently, there's the existential crisis paper, which
+  showed how to extend ML-style inference with some support for
+  existentials. However, it is (a) predicative-only, (b) restricted
+  for $\forall\exists$ types, and (c) does not have a declarative
+  specification. (In fact, the authors comment on the need for such.)
+\item In this paper, we lift all of these restrictions. We give a
+  local type inference algorithm which supports first-class
+  existentials, for impredicative types, with no restrictions on
+  nested quantifers, and we have declarative specification we prove
+  our algorithm sound and complete with respect to.
+\item Developing the algorithm required us to break some of the
+  fundamental invariants of HM-style typin, and as a result, our
+  algorithm needs to mix unification and anti-unification, and works
+  over a call-by-push-value metalanguage to control how quantifiers
+  are instantiated and things like function arities.
+\item Unfortunately, local type inference does not infer types anywhere
+  nearly as comprehensively as full Damas-Milner type inference. We
+  do guarantee that all type applications (for $\forall$-elimination) 
+  and all packs (for $\exists$-introdution) are inferred, but functions
+  and some let-bindings still need annotations on their arguments.
+\item The original local type inference paper combined local type inference
+  with bidirectional typechecking to minimize the number of needed
+  annotations, but we show how existential types complicate the integration
+  of bidirectionality with local type inference, and we explore the design space
+  to show how the same scheme could be applied to work with different
+  type systems. 
+\end{itemize}
 
-The ``\verb|acmart|'' document class can be used to prepare articles
-for any ACM publication --- conference or journal, and for any stage
-of publication, from review to final ``camera-ready'' copy, to the
-author's own version, with {\itshape very} few changes to the source.
+\nk{We need to explain what local type inference is, and how it is and isn't
+  related to bidirectional typechecking.}

### 2023-10-17 oopsla24/tex/02-overview.tex
Wip: declarative system
wip

@@
 \item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
-  which at the term-level corresponds to a pure computation $[[return v]]$.
+  which at the term level corresponds to a pure computation $[[return v]]$.
 \item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
@@
 \item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
-  thunks a computation, which at the term-level corresponds to $[[ {c} ]]$.
+  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
 \end{itemize}
@@
       \ottiN\ottinterrule
-    }
+    }\\
     \ottgrammartabular{
@@
 \paragraph{Definitional Equalities}
-For simplicity, we assume alpha-equivalent terms equal.
+For simplicity, we assume that alpha-equivalent terms are equal.
 This way, we assume that substitutions do not capture bound variables.
@@
 
-In \cref{fig:declarative-terms}, we define the the language of terms of 
+In \cref{fig:declarative-terms}, we define the language of terms of 
 \fexists. The language combines \systemf with the \CBPV approach.
@@
     \item [$+$] $[[x]]$ denotes a (positive) term variable; 
-        \ilyam{why no negatives?}
+        \ilyam{why no negatives? Following CBPV}
     \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 

### 2023-10-17 oopsla24/tex/03-declarative-system.tex
wip

@@
+\label{sec:declarative-system}
+
 The declarative system serves as a specification of the
@@
 ($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively.
-The instantiation is modeled by substitution $[[σ]]$.
+
+The instantiation of quantifiers is modeled by substitution $[[σ]]$.
 The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
@@
 However, after certain changes 
-\ruleref{\ottdruleDOneShiftDLabel} can be relaxed to the covariant form,
+\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
  as we will discuss in \cref{todo}.
@@
 
-A property that we extensively use is that the subtyping is reflexive and transitive. 
-Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.
+\subsubsection{Properties of the Declarative Subtyping}
+A property that we extensively use is that the subtyping is reflexive and transitive,
+and agrees with substitution.
 
-\begin{property}[Subtyping is well-defined]
-  The subtyping forms a preorder and is preserved under substitutions. 
+\begin{property}[Subtyping forms a preorder]
+  Let us say that two types $[[iN1]]$ and $[[iN2]]$ are in the subtyping relation if there 
+  exists a context $[[Γ]]$ such that $[[Γ ⊢ iN1 ≤ iN2]]$;
+  symmetrically, two types $[[iP1]]$ and $[[iP2]]$ are in the subtyping relation
+  if there exists $[[Γ]]$ such that $[[G ⊢ iP2 ≥ iP1]]$.
+  Then the subtyping relation defined this way is reflexive and transitive.
 \end{property}
 
+\begin{property}[Subtyping agrees with substitution]
+  Suppose that  $[[σ]]$ is a substitution such that $[[Γ2 ⊢ σ : Γ1]]$. 
+  Then
+  \begin{enumerate}
+    \item [$-$] $[[Γ1 ⊢ iN ≤ iM]]$ implies $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$, and
+    \item [$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ implies $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$.
+  \end{enumerate}
+\end{property}
+
+Moreover, any two \emph{positive} types have the least upper bound, which makes
+the positive subtyping semilattice.
+The positive least upper bound can be found algorithmically,
+which we will discuss in the next section.
+
+\begin{property}[Positive Least Upper Bound exists]
+  Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
+  well-formed in $[[Γ]]$.
+  Then there exists the least common supertype--- a type $[[iP]]$ such that
+  \begin{itemize}
+    \item $[[Γ ⊢ iP ≥ iP1]]$ and $[[Γ ⊢ iP ≥ iP2]]$, and 
+    \item for any $[[iQ]]$ such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
+      $[[Γ ⊢ iQ ≥ iP]]$.
+  \end{itemize}
+\end{property}
+
+\paragraph{Negative Greatest Lower Bound does not exist}
+However, the symmetric construction---the greatest lower bound of two negative types
+does not always exist. Let us consider the following counterexample.
+Let us consider the following types: 
+\begin{itemize}
+  \item $[[iN]]$ and $[[iQ]]$ are arbitrary closed types, 
+  \item $[[iP]]$, $[[iP1]]$, and $[[iP2]]$ are non-equivalent closed types 
+    such that $[[· ⊢ iP1 ≥ iP]]$ and $[[· ⊢ iP2 ≥ iP]]$, and 
+    none of the types is equivalent to $[[iQ]]$.
+\end{itemize}
+What is the greatest common subtype of  
+$[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
+One of the common subtypes is $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$,
+which, however is not the greatest one.
+
+One can find two greater candidates:
+$[[iM1 = ∀α⁺,β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and $[[iM2 = ∀α⁺,β⁺. β⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$.
+Instantiating $[[α⁺]]$ and $[[β⁺]]$ with $[[iQ]]$ ensures 
+that both of these types are subtypes of $[[iQ → ↓↑iQ → ↓↑iQ → iN]]$;
+instantiating $[[α⁺]]$ with $[[iP1]]$ and $[[β⁺]]$ with $[[iP2]]$
+demonstrates the subtyping with $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$,
+as $[[iP]]$ is a subtype of both $[[iP1]]$ and $[[iP2]]$.
+
+By analyzing the inference rules, we can prove that
+both $[[iM1]]$ and $[[iM2]]$ are maximal common subtypes.
+Since $[[iM1]]$ and $[[iM2]]$ are not equivalent,
+it means that none of them is the greatest.
+
 \subsection{Equivalence and Normalization}
+\label{sec:decl-equivalence}
 
 The subtyping relation forms a preorder on types,
-and thus, it induces an equivalence relation \aka \emph{bicoercibility}.
+and thus, it induces an equivalence relation \aka bicoercibility 
+\cite{tiuryn1995:bicoercibility}.
 The declarative specification of subtyping must be defined up to this equivalence.
 Moreover, the algorithms we use must withstand changes in input types within the equivalence class.
-To deal with this, we use normalization---a function that uniformly selects a representative of the equivalence class.
+To deal with non-trivial equivalence, 
+we use normalization---a function that uniformly selects a representative of the equivalence class.
 
@@
   \item [(i)] we do not need to modify significantly standard operations such as unification to withstand non-trivial equivalence, and
-  \item [(ii)] once the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
+  \item [(ii)] if the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
     while the rest of the algorithm remains the same.
 \end{enumerate*}
+
+In our system, equivalence is reacher than equality. Specifically,
+\begin{enumerate}
+  \item[(ii)] one can introduce redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
+  but not equal to $[[∀α⁺.↑α⁺]]$;
+  \item[(i)] one can reorder quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
+  $[[∀α⁺,β⁺.β⁺ → α⁺ → γ⁻]]$; 
+  \item[(iii)] the transformations (i) and (ii) can happen at any
+    position in the type.
+\end{enumerate}
+
+It turns out that the transformations (i-iii) are complete, 
+in the sense that they generate the whole equivalence class.
+This way, to normalize the type, one must
+\begin{enumerate}
+  \item [(i)] remove the redundant quantifiers,
+  \item [(ii)] reorder the quantifiers to the canonical order,
+  \item [(iii)] do the procedures (i) and (ii) recursively on the subterms.
+\end{enumerate}
+
+The normalization algorithm is shown in \cref{fig:type-nf}.
+The steps (i-ii) are implemented by the ordering function, which takes a set of variables $[[varset]]$
+and a type and returns a list of variables from $[[varset]]$ that occur in the type in the order of their
+first occurrence. Its formal definition can be found in \cref{todo}.
 
@@
   \ottdefnNrmNNorm{}
+  \\
+  $[[ord varset in iN]]$ returns a list of variables
+  $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
   \columnbreak\\
   \ottdefnNrmPNorm{}
+  \\
+  $[[ord varset in iP]]$ returns a list of variables
+  $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
   \end{multicols}
-  \caption{Type Normalization}
+
+  \caption{Type Normalization Procedure} 
   \label{fig:type-nf}
@@
 
+For the normalization procedure, 
+we prove soundness and completeness \wrt the equivalence relation.
+\begin{property}[Correctness of normalization]
+  \hfill \\
+  \begin{itemize}
+    \item[$-$] For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
+  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$;
+    \item[$+$] analogously, for $[[iP]]$ and $[[iQ]]$ well-formed in $[[Γ]]$,
+  $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
+  \end{itemize}
+\end{property}
+
 \subsection{Type Inference}
 
-The declarative type inference system is shown in \cref{fig:declarative-inference}.
+The declarative specification of the type inference is shown in 
+\cref{fig:declarative-inference}.
 The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
@@
 The negative typing judgment is read similarly.
-To infer the resulting type of an application, the
-applicative let rules refer to the \emph{Application typing} judgment.
+
+The \emph{Application typing} judgment
+infers the type of the application of a function to a list of arguments.
 It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
-which is read as ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
+which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
 the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
@@
 \begin{figure}[h]
-  \ottdefnDTNInf{}
-
-  \hfill
 
   \begin{multicols}{2}
+  \ottdefnDTNInf{}
+  \columnbreak\\
   \ottdefnDTPInf{}
-  \\
   \ottdefnDTSpinInf{}
@@
 
-\paragraph{Non-bidirectionality}
-  Notice that all the rules in our system are inferring.
+Let us discuss the rules of the declarative system in more detail.
 
-\paragraph{Typing up to equivalence}
-  As discussed in \cref{todo}, the subtyping, as a preorder, 
-  induces a non-trivial equivalence relation on types. 
-  The system must not distinguish between equivalent types,
-  and thus, type inference must be defined up to equivalence. 
-  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
-  and \ruleref{\ottdruleDTNEquivLabel}.
-  The application typing is defined up to equivalence as well,
-  although it has no special rules for that.
+\paragraph{Variables}
+  Rule \ruleref{\ottdruleDTVarLabel} allows to infer
+  the type of a variable from the context. 
+  In literature can be found another version of this rule,
+  that enables inferring a type \emph{equivalent}
+  to the type from the context. 
+  In our case, the inference of equivalent types
+  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.
+
+\paragraph{Annotations}
+  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
+  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
+  inferred type is a subtype of the annotation type.
+
+\paragraph{Abstractions}
+  The typing of lambda abstraction is standard. 
+  Rule \ruleref{\ottdruleDTtLamLabel} first checks
+  that the given type annotating the argument is well-formed,
+  and then infers the type of the body in the extended context.
+  As a result, it returns an arrow type of function from the
+  annotated type of the argument to the type of the body.
+  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
+  It extends the type context with the quantifying variable $[[pa]]$ and 
+  infers the type of the body. As a result, it returns a polymorphic type
+  quantifying the abstracted variable $[[pa]]$ over the type of the body.
+
+\paragraph{Return and Thunk}
+  Rules \ruleref{\ottdruleDTReturnLabel} 
+  and \ruleref{\ottdruleDTThunkLabel}
+  add the corresponding shifts to the type of the body.
 
@@
 
-\paragraph{Application of a polymorphic type $[[∀]]$}
+\paragraph{Typing up to Equivalence}
+  As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
+  induces a non-trivial equivalence relation on types. 
+  The system must not distinguish between equivalent types,
+  and thus, type inference must be defined up to equivalence. 
+  For this purpose, we use rules \ruleref{\ottdruleDTPEquivLabel}  
+  and \ruleref{\ottdruleDTNEquivLabel}.
+  They allow one to replace the inferred type with an equivalent one.
+
+\paragraph{Application to an Empty List of Arguments}
+  The base case of the application type inference is 
+  represented by rule \ruleref{\ottdruleDTEmptyAppLabel}.
+  If the head of the type $[[iN]]$ is applied to no arguments, 
+  the type of the result is allowed to be $[[iN]]$ or any 
+  equivalent type. We need to relax this rule up to equivalence
+  to ensure the corresponding property globally:
+  the inferred application type can be replaced with an equivalent one.
+  Alternatively, we could have added a separate rule similar to 
+  \ruleref{\ottdruleDTPEquivLabel}, however, 
+  the local relaxation is sufficient to prove the global property.
+
+\paragraph{Application of a Polymorphic Type $[[∀]]$}
   The complexity of the system is hidden in the rules, 
@@
   the instantiation of the quantifying variables is
-  taken out of thin air. The algorithm we present in \cref{todo} 
+  not known a priori. The algorithm we present in \cref{todo} 
   delays this instantiation until more information about it 
   (in particular, typing constraints) is collected.
+
+  To ensure the priority of application between this rule and 
+  \ruleref{\ottdruleDTEmptyAppLabel}, we also check that 
+  the list of arguments is not empty.
 
@@
 
-\paragraph{Annotations}
-  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
-  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
-  inferred type is a subtype of the annotation type.
+\subsubsection{Declarative Typing Properties}
+  An important property that the declarative system has is
+  that the declarative specification is correctly defined for
+  equivalence classes.
 
-\subsection{Properties of the Declarative System}
-
-Now we present selected properties of the declarative system,
-which are important for the correctness of the algorithm.
-
-As we discussed in \cref{todo}, the equivalence induced by the subtyping is 
-not trivial. To deal with it without modifying standard operations such as unification, 
-we use normalization. This way, it is important for the normalization to be 
-sound and complete \wrt the equivalence:
-
-\begin{property}[Correctness of normalization]
-  For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
-  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$.
+\begin{property}[Declarative Typing is Defined up to Equivalence]
+  Let us assume that $[[Γ ⊢ Φ1 ≈ Φ2]]$, \ie  
+  the corresponding types assigned by $[[Φ1]]$ and
+  $[[Φ2]]$ are equivalent in $[[Γ]]$.
+  Also, let us assume that 
+  $[[Γ ⊢ iN1 ≈ iN2]]$, $[[Γ ⊢ iP1 ≈ iP2]]$,
+  and $[[Γ ⊢ iM1 ≈ iM2]]$. Then
+  \begin{itemize}
+    \item [$-$] $[[Γ ; Φ1 ⊢ c : iN1]]$ holds if and only if $[[Γ ; Φ2 ⊢ c : iN2]]$,
+    \item [$+$] $[[Γ ; Φ1 ⊢ v : iP1]]$ holds if and only if $[[Γ ; Φ2 ⊢ v : iP2]]$, and
+    \item [$\bullet$] $[[Γ; Φ1 ⊢ iN1 ● args ⇒> iM1]]$ holds if and only if $[[Γ; Φ2 ⊢ iN2 ● args ⇒> iM2]]$.
+  \end{itemize}
 \end{property}
 
-However, the greatest lower bound of two positive types is not always defined.
-Let us consider a positive type $[[iP]]$,
-and two different supertypes $[[iP1]]$ and $[[iP2]]$
-(\ie  $[[· ⊢ iP1 ≥ iP]]$, $[[· ⊢ iP2 ≥ iP]]$, and $[[nf(iP1) ≠ nf(iP2)]]$.
-Let us also consider an arbitrary negative type $[[iN]]$.
+\ilyam{Other properties?}

### 2023-10-17 oopsla24/tex/04-algorithm.tex
wip

--- 2023-09-28 proofs/main.tex
@@
-\UseRawInputEncoding
-% vim: ft=tex
-\documentclass[a4,natbib=false]{article}
-\usepackage[a4paper, total={8in, 10in}]{geometry}
-\usepackage{hyperref}
-\usepackage{mathpartir}
+        % - Justify Adding α̂± 
+        % - algorithmic type contexts Ξ and constraint context Θ
+        % - Describe how wf, nf,... are changed
+        % - Constraints -- important part of the algorithm
 
-\input{prelude.tex}
+        % - Dependency Graph
 
-\newcommand{\ottDir}{../ott/_gen}
-\newcommand{\genDir}{_gen}
-\input{\ottDir/included.tex}
+In this section, we present the algorithmization of the system described before
+Shadowing the declarative system, the algorithm has two main parts: 
+the subtyping and the type inference, which we discuss in this section one after another. 
+First, let us discuss the syntax of the algorithmic system.
 
-% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
-% \newcommand{\ottdrulename}[1]{\textsc{#1}}
+\begin{description}
+        \item[Positive Algorithmic Variables] $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots 
+        \item[Negative Algorithmic Variables] $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots
+        \item[Positive Algorithmic Types] $[[uP]] = \dots \mid [[α̂⁺]]$  
+        \item[Negative Algorithmic Types] $[[uN]] = \dots \mid [[α̂⁻]]$
+        \item[Algorithmic Type Context] $[[Ξ]] = \text{set of Algorithmic Variables}$
+\end{description}
 
-% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
-%   }\ruleLabel{#4}{#4}}
+\paragraph{Algorithmic Variables}
+Both subtyping and the inference algorithms are represented by sets of inference rules, 
+to simplify the soundness and completeness proofs \wrt the declarative specification. 
+The terms these rules manipulate we call \emph{algorithmic}. They extend the 
+previously defined declarative terms and types by adding \emph{algorithmic type variables}
+(\aka existential variables). The algorithmic variables represent unknown types, 
+which cannot be inferred immediately but are promised to be instantiated
+as the algorithm proceeds.
 
-% \renewcommand{\ottdrule}[4][]{%
-%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
-%   \mpr@label{\textsc{#1}}%
-% }
-% \DeclareDocumentCommand \redrule {m m o}{%
-%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
-%   \mpr@label{\textsc{#3}}
-% }
+We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
+normal variables $[[α]]$, $[[β]]$, \dots and also to smooth out the transition
+from the declarative to the algorithmic system, when we replace the quantified variables $[[pas]]$ 
+that we are not able to instantiate with their algorithmic counterpart $[[puas]]$.
+The procedure of replacing declarative variables with algorithmic ones we call 
+\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$.
 
-% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}
+\paragraph{Algorithmic Types}
+The syntax of  algorithmic types is the syntax of declarative types extended
+with algorithmic type variables: we add positive algorithmic variables $[[α̂⁺]]$ 
+to the positive types, and negative algorithmic variables $[[α̂⁻]]$ to the negative types.
+Notice that these variables cannot be abstracted by the quantifiers $[[∀]]$ and $[[∃]]$.
+We denote the algorithmic types
 
-% \renewcommand{\ottdruleEOneNVarName}[0]{foo }
+\paragraph{Algorithmic Contexts and Well-formedness}
+To specify well-formedness, we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
+Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
+judgment of algorithmic terms defined as expected: in addition to the declarative definition
+\cref{todo}, we also check that each algorithmic variable is in the context $[[Ξ]]$.
 
-% \makeatletter
-% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
-%   }{
-%     % \protected@edef\@currentlabelname{foo}
-%     \def\@currentlabelname{foo}%
-%     \phantomsection
-%     \label{boo}
-%     \alpha ^{-}   \eqEOne   \alpha ^{-}
-%   }{%
-%     {\ottdruleEOneNVarName}{}%
-%   } }
-% \makeatother
-
-\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
-\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
-
-\newcommand{\ruleref}[1]{Rule \nameref{#1}}
-
-% ord varset in uN = varset'
-
-\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}
-
-% ord varset in uP = varset'
-
-\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}
-
-% nf(N) = M
-\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}
-
-% nf(P) = Q
-\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}
-
-% N ≈ M
-
-\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}
-
-% P ≈ Q
-\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}
-
-% G ⊢ N ≤1 M
-
-\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
-\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}
-
-% G ⊢ P ≥1 Q
-\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
-\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
-\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}
-
-% G ⊢ N ≈1 M
-\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}
-
-% G ⊢ P ≈1 Q
-\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}
-
-% G ⊨ iP1 ∨ iP2 = iQ
-\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
-\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
-\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
-\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}
-
-% G ; Θ ⊨ uN ≤ iM ⫤ us
-\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
-\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
-\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
-\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}
-
-% G ; Θ ⊨ iP ≥ uQ ⫤ us
-\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
-\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
-\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
-\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}
-
-% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
-\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}
-
-% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
-\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
-\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}
-
-% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
-\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
-\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}
-
-% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
-\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
-
-% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
-\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUAUName}[0]{(AU)}
-
-% Γ ⊢ iN 
-\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ⊢ iP
-\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uN
-\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uP
-\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ⊢ iP : scE
-\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
-\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}
-
-% Γ ⊢ iN : scE
-\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}
-
-% Γ ; Φ ⊢ v : iP 
-\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ c : iN 
-\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ iN ● args ⇒> iM 
-\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ v : iP  
-\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ c : iN 
-\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-
-% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
-\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% scE1 singular with iP
-\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}
-
-% scE1 singular with iN
-\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}
-
-\begin{document}
-
-\tableofcontents
-
-\newpage
-
-\section{Declarative Type Systems}
-
-\subsection{Grammar}
-We assume that there is an infinite set of positive and 
-negative \emph{type} variables. Positive type variables are denoted as 
-$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
-Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
-We assume there is an infinite set of \emph{term} variables,
-which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
-A list of objects (variables, types or terms) is denoted by
-an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
-$[[nbs]]$ is a list of negative type variables, 
-$[[args]]$ is a list of values, which are arguments of a function.
-$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
-in a type $[[iP]]$ and $[[iN]]$, respectively.
-
-\bigskip
-
-% \ottgrammartabular{
-%   \ottiP\ottinterrule
-%   \ottiN\ottinterrule
-% }
-
-\begin{definition}[Declarative Types]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottiN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottiP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Equalities}
-For simplicity, we assume alpha-equivalent terms equal.
-This way, we assume that substitutions do not capture bound variables.
-Besides, we equate
-$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
-and lift these equations transitively and congruently 
-to the whole system.
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Declarative Type Context]
-  \hfill \\
-  Declarative type context $[[Γ]]$ is represented by a set of 
-  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
-  union of two contexts $[[Γ1 ∪ Γ2]]$.
-\end{definition}
-
-$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
-which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
-(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
-
-Notice that checking the well-formedness of a type
-is an \emph{algorithmic} procedure, in which 
-both the context and the type are considered inputs.
-In other words, it is syntax-directed and mode-correct 
-(according to \cite{dunfieldBidirectionalTyping2020}), 
-which means that
-
-\begin{algorithm}[Type Well-formedness]
-  \label{alg:wf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFTNiWF{}
-  \columnbreak
-
-  \ottdefnWFTPiWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-\subsection{Substitutions}
-
-\begin{definition}[Substitution]
-  Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions form variables to types, preserving the polarity. 
-\end{definition}
-
-\begin{algorithm}[Substitution Application]
-  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
-  It is defined naturally as follows:
-    \begin{multicols}{2}
-      \begin{itemize}
-        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
-        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
-        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
-        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
-        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
-        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
-        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
-          that is the variable capture never happens);
-      \end{itemize}
-    \end{multicols}
-\end{algorithm}
-
-\begin{definition}[Substitution Signature]
-  The signature $[[Γ' ⊢ σ : Γ]]$ means that
-  \begin{enumerate}
-    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
-  \end{enumerate}
-\end{definition}
-
-A substitution can be restricted to a set of variables. 
-The restricted substitution is define as expected. 
-\begin{definition}[Subsitution Restriction]
-  The specification $[[σ  | varset]]$ is defined as
-  a function such that 
-  \begin{enumerate}
-    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
-    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
-  \end{enumerate}
-\end{definition}
-
-Two substitutions can be composed in two ways:
-$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
-while $[[σ2 <=< σ1]]$
-depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
-$[[σ2]]$ to its results on the domain.
-\begin{definition}[Substitution Composition]
-  $[[σ2 ○ σ1]]$ is defined as a function such that
-  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
-\end{definition}
-
-\begin{definition}[Monadic Substitution Composition]
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
-\end{definition}
-Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
-specification of $[[σ1]]$, which is not unique. 
-However, we assume that the used specification clear from the 
-context of the proof.
-
-\begin{definition}[Equivalent Substitutions]
-  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
-  indicates that on the domain $[[Γ]]$, 
-  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
-\end{definition}
-
-Sometimes it is convenient to construct substitution 
-explicitly mapping each variable from a list (or a set)
-to a type. Such substitutions are denoted as $[[iPs / pas]]$
-and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
-the corresponding types.
-\begin{definition}[Explicit Substitution]
-  \hfill
-  \begin{itemize}
-    \item [$-$]
-      Suppose that $[[nas]]$ is a list of negative type variables,
-      and $[[iNs]]$ is a list of negative types of the same length.
-      Then $[[iNs / nas]]$ denotes a substitution such that 
-      \begin{enumerate}
-        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
-        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
-      \end{enumerate}
-    \item [$+$]
-      Positive explicit substitution $[[iPs / pas]]$
-      is defined symmetrically.
-  \end{itemize}
-\end{definition}
-
-\subsection{Declarative Subtyping}
-Subtyping is one of the key mechanism of our system. 
-It realizes the polymorphism: abstract $[[∀]]$ and 
-$[[∃]]$ types can be used where concrete types are expected,
-exactly because the subtyping relation between them.
-
-\begin{definition} 
-  \label{def:subDOne}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNsub{}
-    \columnbreak\\
-    \ottdefnDOnePsup{}
-  \end{multicols}
-
-  \begin{multicols}{2}
-    \ottdefnDOneNeq{}
-    \columnbreak\\
-    \ottdefnDOnePeq{}
-  \end{multicols}
-\end{definition}
-
-The following observations about the declarative subtyping are worth noting:
-\begin{itemize}
-  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
-    make the subtyping reflexive on variables (and further, on any type).
-  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
-    resulting type and contravariant on the argument type.
-  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
-    the subtyping is \emph{invariant} for shifts. 
-    This way, the subtyping of shifted types in one direction implies the subtyping
-    in the opposite direction.
-    Although this rule restricts the
-    subtyping relation, it makes the system decidable.
-  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
-    non-algorithmic rules: the substitution for the quantified variable is
-    not specified, those, these rules `drive' the subtyping relation.
-\end{itemize}
-
-In the next section, we present the sound and complete algorithm
-checking whether one type is a subtype of another according to \cref{def:subDOne}.
-
-\section{Algorithmic Type System}
-
-\subsection{Grammar}
-
-In the algorithmic system, we extend the grammar of types
-by adding positive and negative \emph{algorithmic variables}
-($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
-They represent the unknown types, which will be inferred by the algorithm.
-This way, we add two base cases to the grammar of 
-positive and negative types, and use highlight to denote that the type
-can potentially contain algorithmic variables.
-
-\begin{definition}[Algorithmic Types]
-  \label{def:algo-types}
-  \hfill\\
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottuN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottuP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Fresh Variable Selection}
-\label{sec:fresh-selection}
-Both the subtyping and the type inference algorithm
-rely on the ability to select fresh, unused variables.
-For a set of variables $[[varset]]$, it is indicated as 
-$[[varset are fresh]]$ in the inference rules.
-We assume that the selection subroutine always succeeds and is 
-deterministic. In other words, whenever it is called in 
-an algorithmic inference rule, it returns the same result, 
-uniquely determined by the input of this rule.
-
-\subsection{Variable Algorithmization}
-\label{sec:variable-algorithmization}
-
-In several places of our algorithm, in particular, during
-algorithmic subtyping,
-we turn a declarative type into the algorithmic one
-via replacing certain type variables with fresh algorithmic variables.
-We call this procedure \emph{variable algorithmization}, and define it as follows.
-
-\begin{definition}[Variable Algorithmization]
-  Suppose that $[[nas]]$ is a list of negative type variables
-  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
-  in a type for $[[αî⁻ ∊ {nuas}]]$.
-\end{definition}
-
-Conversely, we have the opposite procedure turning algorithmic type variables
-into declarative type variables via \emph{dealgorithmization}.
-
-\begin{definition}[Variable Dealgorithmization]
-  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
-  and $[[nas]]$ is a list of negative type variables of the same length. 
-  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
-  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
-\end{definition}
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
-  \hfill \\
-  Algorithmic type context $[[Ξ]]$ is represented by a set of 
-  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
-  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
-\end{definition}
-
-$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
-that the algorithmic type is well-formed in the contexts
-$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
-of the type is contained in $[[Ξ]]$, and each free declarative type variable
-of the type is contained in $[[Γ]]$.
-
-\begin{algorithm}[Algorithmic Type Well-formedness]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFATNauWF{}
-  \columnbreak
-
-  \ottdefnWFATPauWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-Algorithmic Type Context are used in the unification algorithm.
-In the subtyping algorithm, 
-the context needs to remember additional information.
-In the subtyping context, each algorithmic variable is associated with a
-context it must be instantiated in 
-(i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
-  \hfill \\
-  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
-  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
-  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
-  and $[[Γ]]$ is a context in which they must be instantiated.
-  We assume that no two entries associating the same variable
-  appear in $[[Θ]]$.
-
-  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
-  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
-  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
-\end{definition}
-
-\subsection{Subsitutions}
-
-Substitution that operates on algorithmic type variables is denoted as
-$[[uσ]]$. It is defined as a total function from algorithmic 
-type variables to \emph{declarative} types, preserving the polarity.
-
-The signature $[[Θ ⊢ uσ : Ξ]]$ means that
-$[[Ξ ⊆ dom(Θ)]]$ and 
-$[[uσ]]$ maps each algorithmic variable 
-from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
-and for each variable not appearing in $[[dom(Θ)]]$, 
-it acts as identity.
-
-\begin{definition}[Signature of Algorithmic Substitution]
-  \label{def:algo-subst-sig}
-  \hfill
-  \begin{itemize}
-    \item $[[Θ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$,
-          there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
-          and $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-    \item $[[Γ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-  \end{itemize}
-\end{definition}
-
-In the anti-unification algorithm, we use another kind of substitution.
-In contrast to algorithmic substitution $[[uσ]]$,
-it allows mapping algorithmic variables to
-\emph{algorithmic} types.
-Additionally, anti-unification substitution is restricted to the
-\emph{negative} segment of the language.
-Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a
-
-The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
-in which the results of an anti-unification substitution 
-are formed, is fixed for this substitution.
-This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
-variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.
-
-\begin{definition}[Signature of Anti-unification substitution]
-  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
-  \begin{enumerate}
-    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
-    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
-  \end{enumerate}
-\end{definition}
-
-\subsection{Equivalence and Normalization}
-\label{sec:equivalence-normalization}
-
-The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
-there are types that are subtypes of each other but not equal. 
-For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
-and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
-although these types are not alpha-equivalent.
-For the subtyping algorithm, it is crucial to be able to check whether
-two types are equivalent, without checking mutual subtyping. 
-For this purpose we define the normalization procedure, 
-which allows us to uniformly choose the representative type of the equivalence class.
-This way, the equivalence checking is reduced to normalization and equality checking.
-
-For clarification of the proofs and better understanding of the system, 
-we introduce an intermediate relation---\emph{declarative equivalence}. 
-As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
-this relation is equivalent to the subtyping-induced equivalence, but does not 
-depend on it. Although this relation is not defined algorithmically, 
-it gives the intuition of what types our system considers equivalent.
-Specifically, in addition to \emph{alpha-equivalence}, 
-our system allows for \emph{reordering of adjacent quantifiers},
-and \emph{introduction/elimination of unused quantifiers}.
-
-The non-trivial rules of the declarative equivalence are
-\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
-Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
-the recursive call on the body of the quantified type. 
-It will be covered formally in \cref{sec:decl-equiv-lemmas}.
-
-\begin{definition}[Declarative Type Equivalence]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnEOneNeq{}
-  \columnbreak\\
-  \ottdefnEOnePeq{}
-  \end{multicols}
-
-\end{definition}
-
-As the equivalence includes arbitrary reordering of quantified variables,
-the normalization procedure is needed to choose the canonical order.
-For this purpose, we introduce an auxiliary procedure---variable ordering. 
-Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
-in the order they appear in $[[iN]]$.
-
-\begin{algorithm}[Variable Ordering]
-  \label{alg:var-ordering}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnONVar{}
-  \columnbreak\\
-  \ottdefnOPVar{}
-  \end{multicols}
-
-  Analogously, the variable can be ordered in 
-  an \emph{algorithmic} type ($[[ord varset in uP]]$ and 
-  $[[ord varset in uN]]$). In these cases, we treat the algorithmic variables
-  as if they were declarative variables.
-
-\end{algorithm}
-
-Next, we use the variable ordering in the normalization procedure. 
-Specifically, normalization recursively traverses the type, 
-and for each quantified case reorders the quantified variables in a 
-canonical order dictated by \cref{alg:var-ordering}, removing unused ones.
-
-\begin{algorithm}[Type Normalization]
-  \label{alg:type-nf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnNrmNNorm{}
-  \columnbreak\\
-  \ottdefnNrmPNorm{}
-  \end{multicols}
-
-  Analogously,  we define normalization of algorithmic types 
-  by adding base cases:
-
-  \begin{multicols}{2}
-  \ottdefnNrmuNNorm{}
-  \columnbreak\\
-  \ottdefnNrmuPNorm{}
-  \end{multicols}
-
-\end{algorithm}
-
-\Cref{lemma:subt-equiv-algorithmization}
-demonstrates that the equivalence of types is the same
-as the equality of their normal forms.
-\begin{theorempreview}[Correctness of Normalization]
-  Assuming the types are well-formed in $[[Γ]]$, 
-  \begin{itemize}
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]]$ if and only if $[[nf(iN) = nf(iM)]]$;
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]]$ if and only if $[[nf(iP) = nf(iQ)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\begin{algorithm}[Substitution Normalization]
-  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
-  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
-\end{algorithm}
-
-The rest of this chapter is devoted to the
-central algorithm of the type system---the subtyping algorithm. 
-\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
-The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
-$A$ uses $B$ as a sub-procedure.
+\paragraph
 
@@
   \begin{tikzpicture}
-    [>={Stealth[scale=2]},node distance=2.6cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
+    [>={Stealth[scale=2]},node distance=2.4cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
 
@@
     \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[left=2cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[right=2cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
+    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
     \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
@@
 
-\subsection{Subtyping}
-\label{sec:subtyping}
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Subtyping Algorithm}
+Hello Subtyping
+        % - fig
+        % - it takes ... and returns constraint
+        % - explain the rules
+        % - - ↑/↓ invoke unification up-to-equivalence, 
+        %     which is implemented by unification of the normal forms up-to-equality.
+        % - - → makes two recursive calls, and thus, it merges the resulting constraints
+        % - - ∀/∃ introduce new variables as discussed
+        % - - the base case calls the upgrade
 
-Now, we present the subtyping algorithm itself.
-Although the algorithm is presented as a single procedure,
-is important for the structure of the proof that the positive subtyping algorithm
-does not invoke the negative one. This way, the correctness of the positive 
-subtyping will be proved independently, and used afterwards to prove the
-correctness of the negative subtyping.
+    \begin{figure}[h]
+        \hfill\\
+        \ottdefnANsub{}
+        \ottdefnAPsup{}
+        \caption{Subtyping Algorithm}
+        \label{fig:alg-subtyping}
+    \end{figure}
 
-\begin{algorithm}[Subtyping]
-  \label{alg:subtyping}
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Unification}
+
+Hello Unification
+
+\begin{figure}[h]
   \hfill
+  \begin{multicols}{2}
+  \ottdefnUNUnif{}
+  \columnbreak\\
+  \ottdefnUPUnif{}
+  \end{multicols}
+  \caption{Unification Algorithm}
+  \label{fig:unification}
+\end{figure}
 
-  \begin{multicols}{2}
-  \ottdefnANsub{}
-  \columnbreak\\
-  \ottdefnAPsup{}
-  \end{multicols}
-\end{algorithm}
-
-The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
-the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
-must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
-and $[[uP]]$ and $[[iQ]]$ for the positive case. 
-As one of the invariants, we require
-$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
-The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
-which will be discussed in the next section.
-
-Let us overview the inference rules of the subtyping algorithm.
-\begin{itemize}
-  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
-    are the base cases. They copy the corresponding declarative rules and
-    ensure the reflexivity.
-  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
-    subtyping constraints. In this case, we must ensure
-    that the resulting constrains guarantees that the instantiation of 
-    $[[â⁺]]$ is a supertype of $[[iP]]$.
-    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
-    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
-    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
-    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
-
-    Notice that this rule does not have a negative counterpart. This is
-    because one one of the important invariants of the algorithm: 
-    in the negative subtyping, only positive algorithmic variables
-    can occur in the types.
-
-  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
-    \emph{shift} rules. According to the declarative system,
-    shifted subtyping requires equivalence. In presence of the algorithmic 
-    variables, it means that the left and the right-hand sides of the subtyping
-    must be unified. Hence, the shift rules invoke the unification algorithm, 
-    which will be discussed in \cref{sec:unification}. The unification 
-    returns the minimal set of constraints $[[UC]]$, which is necessary
-    and sufficient for the subtyping.
-
-  \item \ruleref{\ottdruleAArrowLabel}.
-    In this case, the algorithm makes two calls:
-    a recursive call to the negative subtyping algorithm for the argument types,
-    and a call to the positive subtyping algorithm for the result types.
-    After that, the resulting constraints are merged using the
-    \emph{subtyping constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
-    are symmetric. These are the only places where 
-    the algorithmic variables are introduced.
-    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
-    of the quantified variables: these variables are replaced by 
-    fresh algorithmic variables in the body of the quantified type,
-    the algorithmic variables are added to the subtyping context $[[Θ]]$,
-    after that the recursive call is made. Notice that the declarative context
-    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
-    which matches the declarative system.
-\end{itemize}
-
-Then soundness lemma (\cref{lemma:pos-subt-soundness,lemma:neg-subt-soundness}) 
-and completeness (\cref{lemma:pos-subt-completeness,lemma:neg-subt-completeness})
-of the algorithm together give us the following simplified theorem:
-
-\begin{theorempreview}[Correctness of subtyping algorithm]
-  \hfill
-  \begin{itemize}
-    \item [$-$]  $[[ Γ ; · ⊨ uN ≤ iM ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iN ≤ iM ]]$;
-    \item [$+$] $[[ Γ ; · ⊨ uP ≥ iQ ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iP ≥ iQ ]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraints}
-
-Unification and subtyping algorithms are based on the constraint generation.
-The constraints are represented by set of constraint entries.
-
-\begin{definition}[Unification Constraint]
-  \hfill
-  \begin{description}
-    \item[unification entry] (denoted as $[[ucE]]$) is an expression of shape 
-      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
-    \item[unification constraint] (denoted as $[[UC]]$) is a set of 
-      unification constraint entries.
-      We denote $\{[[α̂±]] \mid [[ucE ∊ UC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(UC)]]$.
-  \end{description}
-\end{definition}
-
-However, in the subtyping, we need to consider more general
-kind of constraints. Specifically,
-subtyping constraint entries can restrict a variable
-not only to be equivalent to a certain type, but
-also to be a supertype of a positive type.
-
-\begin{definition}[Subtyping Constraint]
-  \hfill
-  \begin{description}
-    \item[subtyping entry] (denoted as $[[scE]]$) is an expression of shape 
-      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
-    \item[subtyping constraint] (denoted as $[[SC]]$) is a set of subtyping constraint entries.
-      We denote $\{[[α̂±]] \mid [[scE ∊ SC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(SC)]]$.
-  \end{description}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint Entry]
-  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
-  its associated type is well-formed in $[[Γ]]$.
-  \begin{itemize}
-    \item[] $[[Γ ⊢ pua :≥ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ pua :≈ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ nua :≈ iN]]$ ~iff~ $[[Γ ⊢ iN]]$.
-  \end{itemize}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  subtyping context $[[Θ]]$ if all its entries are well-formed in
-  the corresponding elements of $[[Θ]]$.
-  More formally, 
-  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
-  such that $[[scE]]$ restricts $[[α̂±]]$,
-  we have $[[Θ(α̂±) ⊢ scE]]$.
-
-  We write $[[Θ ⊢ SC : Ξ]]$ to denote
-  that $[[Θ ⊢ SC]]$ and $[[dom(SC) = Ξ]]$.
-
-  $[[Θ ⊢ UC]]$ and $[[Θ ⊢ UC : Ξ]]$ are defined analogously.
-\end{definition}
-
-\subsubsection{Constraint Satisfaction}
-
-A constraint entry restricts a type that can be assigned to a variable.
-We say that a type satisfies a constraint entry if it can be assigned
-to the variable restricted by the entry.
-
-\begin{definition}[Type Satisfying a Constraint Entry]
-  \hfill\\
-  \begin{multicols}{2}
-  \ottdefnSATSCEN{}
-  \columnbreak\\
-  \ottdefnSATSCEP{}
-  \end{multicols}
-\end{definition}
-
-We say that a substitution satisfies a constraint---a set of constraint 
-entries if each entry is satisfied by the type assigned to the variable
-by the substitution.
-
-\begin{definition}[Substitution Satisfying a Constraint]
-  We write $[[Θ ⊢ uσ : SC]]$ to denote that
-  a substitution $[[uσ]]$ satisfies a constraint $[[SC]]$ in a context $[[Θ]]$.
-  It presumes that $[[Θ ⊢ SC]]$ and 
-   means that for any $[[ucE]] \in [[SC]]$, if $[[ucE]]$ restricts $[[α̂±]]$,
-  then $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$.
-
-  Unification constraint satisfaction $[[Θ ⊢ uσ : UC]]$ 
-  is defined analogously as a special case of subtyping constraint satisfaction.
-
-\end{definition}
-
-Notice that $[[Θ ⊢ uσ : SC]]$ does not 
-imply the signature $[[Θ ⊢ uσ : dom(SC)]]$, because 
-the latter also specifies $[[uσ]]$ outside of the domain $[[dom(SC)]]$
-(see \cref{def:algo-subst-sig}).
-
-\subsubsection{Constraint Merge}
-\label{sec:constraint-merge}
-
-In this section, define the least upper bound 
-for constraints, which we call \emph{merge}.
-Intuitively, the merge of two constraints is the least
-constraint such that any substitution satisfying both constraints
-satisfies the merge as well.
-First, we define the merge of entries,
-and then extend it to the set of entries.
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Constraint Merge}
 
@@
 can be merged in the following way:
-\begin{algorithm}[Merge of Matching Constraint Entries]
-  \label{definition:merge-matching-entries}
-   \hfill
 
+\begin{figure}[h]
   \ottdefnSCME\\
-\end{algorithm}
+
+  \label{fig:merge-entries}
+  \caption{Merge Matching Constraint Entries}
+\end{figure}
 
@@
 
-Unification constraint entries are a special case of subtyping constraint
-entries. They are merged using the same algorithm 
-(\cref{definition:merge-matching-entries}).
-Notice that the merge of two matching unification constraint entries
-is a unification constraint entry.
-\begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
-  \label{lemma:merge-matching-entries-welldef}
-  Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
-  are unification constraint entries. 
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  $[[Γ ⊢ ucE1 & ucE2 = ucE]]$
-  according to \cref{definition:merge-matching-entries},
-  is a unification constraint entry.
-\end{lemma}
-\begin{proof}
-  Since $[[ucE1]]$ and $[[ucE2]]$ are matching unification constraint entries,
-  they have the shape $([[pua :≈ iP1]], [[pua :≈ iP2]])$ or
-  $([[nua :≈ iN1]], [[nua :≈ iN2]])$.
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  can only be defined by \ruleref{\ottdruleSCMEPEqEqLabel} or
-  \ruleref{\ottdruleSCMENEqEqLabel}.
-  In both cases the result, if it exists, 
-  is a unification constraint entry:
-  in the first case, the result has shape $[[pua :≈ iP1]]$,
-  in the second case, the result has shape $[[nua :≈ iN1]]$.
-\end{proof}
-
-% Notice that in case of equivalence, the assigned types
-% must be equal (i.e. alpha-equivalent) to be merged. This is because
-% the unification algorithm assumes that every type is normalized,
-% and hence, equivalence is alpha-equivalence 
-% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
-
-\begin{algorithm}[Merge of Subtyping Constraints]
-  \label{definition:merge-subtyping-constraints}
+\begin{figure}[h]
   Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
@@
   \end{itemize}
-\end{algorithm}
 
-Unification constraints can be considered 
-as a special case of subtyping constraints,
-and the merge of unification constraints
-is defined as the merge of subtyping constraints.
-Then it is easy to see that the merge of two 
-unification constraints is a unification constraint.
+  \label{fig:merge-subtyping-constraints}
+  \caption{Constraint Merge}
+\end{figure}
 
-\begin{lemma}[Merge of Unification Constraints is well-defined]
-  Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
-  are unification constraints. 
-  Then the merge of $[[UC1]]$ and $[[UC2]]$ 
-  $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$
-  according to \cref{definition:merge-subtyping-constraints},
-  is a unification constraint.
-\end{lemma}
-\begin{proof}
-  $[[UC]]$ consists of unmatched entries of $[[UC1]]$ and $[[UC2]]$,
-  which are \emph{unification} constraint entries by assumption,
-  and merge of matching entries, which also are  
-  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-welldef}.
-\end{proof}
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Type Upgrade and the Least Upper Bounds}
 
-\Cref{lemma:merge-soundness,lemma:merge-completeness} 
-show the correctness and initiality of the merge operation,
-which can be expressed in the following simplified theorem:
-\begin{theorempreview}[Correctness of Constraint Merge]
-  A substitution $[[uσ]]$ satisfying both constraints
-  $[[SC1]]$ and $[[SC2]]$ 
-  if and only if it satisfies their merge.
-\end{theorempreview}
+\begin{figure}[h]
+  \ottdefnLUBNsub{}
+  \caption{Least Upper Bound Algorithm}
+  \label{fig:lub}
+\end{figure}
 
-The unification constraint merge satisfies the same theorem,
-however, because the merge of unification constraint entries 
-$[[ucE1]]$ and $[[ucE2]]$ always results in one of them, 
-a stronger soundness property holds (see \cref{lemma:unif-merge-soundness}):
-\begin{theorempreview}[Soundness of Unification Constraint Merge]
-  If $[[Θ ⊢ UC1 & UC2 = UC]]$ then $[[UC = UC1 ∪ UC2]]$.
-\end{theorempreview}
-
-\subsection{Unification}
-\label{sec:unification}
-
-The subtyping algorithm calls the following subtask:
-given two algorithmic types, we need to find the most general substitution 
-for the algorithmic variables in these types, such that the resulting 
-types are equivalent. This problem is known as \emph{unification}.
-
-In our case, the unification is restricted in the following way:
-first, before unifying the types, we normalize them, which 
-allows us to reduce (non-trivial) equivalence to (trivial) equality;
-second, we preserve invariants which guarantee that
-one side of the unification is always declarative, which in fact, 
-reduces the unification to the \emph{matching} problem.
-
-The unification procedure
-returns a set of minimal constraints,
-that must be satisfied by a substitution
-unifying the input types.
-
-\begin{algorithm}[Unification]
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnUNUnif{}
-  \columnbreak\\
-  \ottdefnUPUnif{}
-  \end{multicols}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
-    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
-    adding the quantified variables to the context $[[Γ]]$.
-    Notice that $[[Θ]]$, which specifies
-    the contexts in which the algorithmic variables must be instantiated,
-    is not changed.
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
-    are the base cases. 
-    Since the sides are equal and free from algorithmic variables,
-    the unification returns an empty constraint. 
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
-    are symmetric cases constructing the constraints. 
-    When an algorithmic variable is unified with a type, 
-    we must check that the type is well-formed in the required context,
-    and if it is, we return a constraint restricting the variable
-    to be equivalent to that type.
-  \item \ruleref{\ottdruleUArrowLabel}.
-    In this case, the algorithm makes two recursive calls:
-    it unifies the arguments and the results of the arrows.
-    After that, the resulting constraints are merged using the
-    \emph{unification constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
-    \emph{unification} constraints, not arbitrary \emph{subtyping} 
-    constraints: it is important for modularizing the proofs, 
-    since the properties of the \emph{unification} constraint merge
-    can be proved independently from the \emph{subtyping} constraint merge.
-\end{itemize}
-
-\subsection{Least Upper Bound}
-\label{sec:lub}
-
-In this section, we present
-the algorithm finding the least common supertype of two positive types. 
-It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
-and indirectly, through the type upgrade by positive subtyping
-(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
-intuitive part of the algorithm, and its correctness will be covered
-in \cref{sec:alg-upper-bounds-proofs}.
-
-\begin{algorithm}[The Least Upper Bound Algorithm]
-  \hfill\\
-  \ottdefnLUBNsub{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleLUBVarLabel}
-    The base case is trivial: 
-    the least upper bound of to equal variables
-    is the variable itself.
-  \item \ruleref{\ottdruleLUBShiftLabel}
-    In case both sides of the least upper bound are shifted,
-    the algorithm needs to find the anti-unifier of them. 
-    Intuitively, this is because in general, the upper bounds of
-    $[[↓iN]]$ are $[[∃nas.iP]]$ such that 
-    $[[nas]]$ can be instantiated with some $[[iMs]]$ so that
-    $[[ Γ ⊢ [iMs/nas]iP ≈ ↓iN ]]$ (see \cref{lemma:shape-of-supertypes}).
-  \item \ruleref{\ottdruleLUBExistsLabel}
-    In this case, we move the quantified variables to the context $[[Γ]]$, 
-    and make a recursive call. 
-    It is important to make sure that $[[nas]]$ and $[[nbs]]$ are disjoint.
-    In this case, it is guaranteed that the resulting 
-    $[[fv(iQ)]]$ will be free of $[[nas]]$ and $[[nbs]]$,
-    and thus, the resulting type will be a supertype of both sides
-    (it will be discussed in \cref{lemma:shape-of-supertypes}).
-\end{itemize}
-
-In the positive subtyping algorithm (\cref{sec:subtyping}),
-\ruleref{\ottdruleAPUVarLabel} 
-generates a restriction of a variable $[[α̂⁺]]$.
-On the one hand, this restriction must imply 
-$[[â⁺ :≥ iP]]$ for the subtyping to hold.
-On the other hand, the type used in this restriction 
-must be well-formed in a potentially stronger (smaller) 
-context than $[[iP]]$.
-
-To resolve this problem, we define the \emph{upgrade} procedure,
-which for given $[[Δ]]$, $[[pnas]]$, and $[[Δ, pnas ⊢ iP]]$,
-finds $[[Δ ⊢ iQ]]$---the least supertype of $[[iP]]$ 
-among the types well-formed in $[[Δ]]$.
-
-The trick is to make sure that the `forbidden' variables
-$[[pnas]]$ are not used explicitly in the supertypes
-of $[[iP]]$. For this purpose, we
-construct new types $[[iP1]]$ and $[[iP2]]$,
-in each of them replacing the forbidden variables
-with fresh variables $[[pnbs]]$ and $[[pncs]]$,
-and then find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
-It turns out that this renaming forces the common types of 
-$[[iP1]]$ and $[[iP2]]$ to be agnostic to $[[pnas]]$,
-and thus, the supertypes of $[[iP]]$ well-formed in $[[Δ]]$
-are exactly the common supertypes of $[[iP1]]$ and $[[iP2]]$.
-These properties are considered in more details in \cref{sec:upgrade-lemmas}.
-
-\begin{algorithm}[Type Upgrade]
+\begin{figure}[h]
     \hfill\\
     \ottdefnLUBUp{}
-\end{algorithm}
+    \caption{Type Upgrade Algorithm}
+    \label{fig:type-upgrade}
+\end{figure}
 
-\begin{paragraph}{Note on the Greatest Lower Bound}
-  In contrast to the least upper bound, 
-  the general greatest lower bound does not exist in our system.
-  For instance, consider a positive type $[[iP]]$, 
-  together with its non-equivalent
-  supertypes $[[iP1]]$ and $[[iP2]] \not\simeq [[iP1]]$
-  (for example, $[[iP = ↓↑↓γ⁻]]$, $[[iP1 = ∃α⁻.↓↑↓α⁻]]$, 
-  and $[[iP2 = ∃α⁻.↓α⁻]]$).
-  Then for arbitrary $[[iQ]]$ and $[[iN]]$, 
-  let us consider the common subtypes of 
-  $A = [[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $B = [[iP → ↓↑iP1 → ↓↑iP2 → iN]]$.
-  It is easy to see that $[[∀α⁺.∀β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and 
-  $[[∀α⁺.∀β⁺. α⁺ → ↓↑β⁺ → ↓↑α⁺ → iN]]$ are
-  both \emph{maximal} common subtypes of $A$ and $B$,
-  and since they are not equivalent, none of them is 
-  the \emph{greatest} one.
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Anti-Unification}
 
-  However, we designed the subtyping system in such a way 
-  that the greatest lower bound is not needed:
-  the negative variables are always `protected'
-  by \emph{invariant} shifts ($[[↑]]$ and $[[↓]]$), 
-  and thus, the algorithm can only require
-  a substitution of a negative variable to be 
-  \emph{equivalent} to some type but never 
-  to be a \emph{subtype}.
-\end{paragraph}
-
-\subsection{Anti-unification}
-\label{sec:antiunification}
-
-Next, we define the anti-unification procedure,
-also known as the \emph{most specific generalization}.
-As an input, it takes two declarative types
-(e.g., in the positive case $[[iP1]]$ and $[[iP2]]$)
-and a context $[[Γ]]$.
-and returns a type $[[uQ]]$---the generalizer,
-containing negative placeholders 
-(represented by algorithmic variables) from 
-$[[Ξ]]$ and two substitutions $[[uτ1]]$ and $[[uτ2]]$.
-The substitutions replace the placeholders with 
-declarative types well-formed in $[[Γ]]$,
-such that $[[ [uτ1]uQ = iP1 ]]$ and $[[ [uτ2]uQ = iP2 ]]$.
-Moreover, the algorithm guarantees that 
-$[[uQ]]$ is the most specific type with this property: 
-any other generalizer can be turned into $[[uQ]]$ by some substitution
-$[[uρ]]$.
-
-It is important to note the differences between 
-the standard anti-unification and our version.
-First, we only allow the placeholders at \emph{negative} positions,
-which means, for example, that $[[α⁺]]$ and $[[β⁺]]$ cannot be
-generalized. Second, the generated pair of substitutions 
-$[[uτ1]]$ and $[[uτ2]]$ must replace the placeholders with 
-types well-formed in a specified context $[[Γ]]$.
-
-The anti-unification algorithm assumes that the input types 
-are normalized. This way, anti-unification up-to-equality rather than 
- anti-unification up-to-equivalence is sufficient.
-
-\begin{algorithm}[Anti-unification]
-  \hfill
-
+\begin{figure}[h]
     \ottdefnAUAUP{}
@@
 
-\end{algorithm}
+    \caption{Anti-unification Algorithm}
+    \label{fig:anti-unification}
+\end{figure}
 
-\begin{itemize}
-  \item \ruleref{\ottdruleAUPVarLabel} and \ruleref{\ottdruleAUNVarLabel}
-    are the base cases. 
-    In this case, since the input types are equal, 
-    the algorithm returns this type as a generalizer,
-    without generating any placeholders.
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Type Inference}
 
-  \item \ruleref{\ottdruleAUShiftDLabel}, \ruleref{\ottdruleAUShiftULabel},
-    \ruleref{\ottdruleAUForallLabel}, and \ruleref{\ottdruleAUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers.
-    Notice that the algorithm does not add the removed variables to
-    the context $[[Γ]]$. This is because $[[Γ]]$
-    is used to restrict the resulting anti-unification substitutions, 
-    and is fixed throughout the algorithm.
+        % - Overview (Dependency Graph)
+        % *** The Inference Algorithm
+        % *** Constraint Singularity
 
-  \item \ruleref{\ottdruleAUAULabel} is the most important rule, 
-    since it generates the placeholders. 
-    This rule only applies if other negative rules failed.
-    Because of that, the anti-unification procedure is 
-    \emph{not} syntax-directed.
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.3\textwidth}
+        \centering
+        \ottdefnATNInf{}
+    \end{subfigure}
 
-    The generated placeholder is indexed with a pair of 
-    types it is mapped to. It allows the algorithm to 
-    automatically unite the anti-unification solutions 
-    generated by the different branches of 
-    \ruleref{\ottdruleAUArrowLabel}.
+    \begin{subfigure}[b]{0.3\textwidth}
+        \centering
+        \ottdefnATPInf{}
+    \end{subfigure}
 
-    Notice that this rule does not have a positive counterpart,
-    since we only allow negative placeholders.
-
-  \item \ruleref{\ottdruleAUArrowLabel}
-    makes two recursive calls to the anti-unification procedure,
-    and unites the results. Suppose that
-    $[[uτ1]]$  and $[[uτ2]]$ are the substitutions generated by
-    anti-unification of \emph{argument} types of the arrow,
-    and $[[uτ1']]$ and $[[uτ2']]$ are the substitutions generated by 
-    anti-unification of \emph{result} types of the arrow.
-    It is important that if ($[[uτ1]]$, $[[uτ2]]$)
-    and ($[[uτ1']]$, $[[uτ2']]$) send some variables to the same pair of types,
-    i.e., $[[ [uτ1]α̂⁻ = [uτ1']β̂⁻]]$ and $[[ [uτ2]α̂⁻ = [uτ2']β̂⁻]]$,
-    then these variables are equal, i.e., $[[α̂⁻ = β̂⁻]]$.
-    This property is guaranteed by \ruleref{\ottdruleAUAULabel}:
-    the name of the placeholder is determined by the pair of 
-    types it is mapped to.
-\end{itemize}
-
-\newpage
-
-\section{Declarative Typing}
-
-In the previous section, we presented the 
-type system together with subtyping specification 
-and the algorithm. In this section, 
-we describe the language under this type system, 
-together with the type inference specification
-and algorithm.
-
-\subsection{Grammar}
-
-First, we define the syntax of the language.
-The language combines System F with call-by-push-value 
-style. 
-\begin{definition}[Language Grammar]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottc\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottv\ottinterrule
-    }
-  \end{multicols}
-\end{definition}
-Notice that the language does not have
-first-class applications: instead, 
-we use applicative let bindings---
-constructions that bind a result of
-a fully applied function to a (positive) 
-variable.
-In the call-by-push-value paradigm, 
-it corresponds to monadic bind or do-notation. 
-Typewise, these let-binders come in two forms:
-annotated and unannotated. The annotated let-binders
-$[[let x:iP = v(args); c]]$ 
-requires the application to infer the annotated
-$[[iP]]$, whereas the unannotated 
-$[[let x = v(args); c]]$ 
-is used when the inferred type is unique.
-
-A computation of a polymorphic type is constructed 
-using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
-is implicit. Conversely, the existential types
-are constructed implicitly and eliminated 
-using the standard unpack mechanism:
-$[[let∃ (nas, x) = v; c]]$.
-
-Another dual pair of constructions are 
-$[[return v]]$ and $[[{c}]]$. 
-The former allows us to embed a value in a pure 
-computations. The latter, on the contrary,
-encapsulates a thunk of computation in a value.
-
-Finally, the language has a number standard constructions:
-lambda-abstractions $[[λx:iP.c]]$,
-standard let-bindings $[[let x = v; c]]$,
-and type annotations that can be added to any value or computation:
-$[[(v:iP)]]$ and $[[(c:iN)]]$.
-
-\subsection{Declarative Type Inference}
-
-Next, we define the specification of the type 
-inference for our language. First, we introduce 
-variable context specifying the types of the variables 
-in the scope of the current rule.
-
-\begin{definition}[Variable Context]
-  The variable typing context $[[Φ]]$
-  is represented by a set of entries of the form
-  $[[x : iP]]$. 
-\end{definition}
-
-The specification is represented by an inference system of
-three mutually recursive judgments:
-positive inference $[[Γ ; Φ ⊢ v : iP]]$,
-negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
-application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
-In the premises, the inference rules also refer to 
-the declarative subtyping (\cref{def:subDOne}),
-type well-formedness (\cref{alg:wf}), 
-and normalization (\cref{alg:alg:type-nf}) judgements.
-\begin{itemize}
-  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
-    means that under the type context $[[Γ]]$ and
-    the variable context $[[Φ]]$, for the value $[[v]]$,
-    type $[[iP]]$ is inferrable. It guarantees that 
-    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
-    the standard sense.
-  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
-    judgement. It means that if a head of type $[[iN]]$ 
-    is applied to list of values $[[args]]$, 
-    then the resulting computation can be typed as $[[iM]]$.
-\end{itemize}
-
-\begin{definition}[Declarative Type Inference]
-  \label{def:declarative-typing}
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnDTNInf{}
-  \ottdefnDTPInf{}
-  \ottdefnDTSpinInf{}
-  \end{multicols}
-\end{definition}
-
-Let us discuss selected rules of the declarative system:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTVarLabel}
-    says that the type of a variable is inferred from the context.
-  \item \ruleref{\ottdruleDTThunkLabel} says
-    that the type of a thunk is inferred by shifting up the type of the 
-    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
-    infers the type of a return by shifting down the type of the
-    contained value.
-  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
-    They allow the inferred type to be refined by annotating it with a supertype.
-  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
-    mean that th declarative system allows to infer any type from the equivalence class.
-  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
-    and its first premise inferring the existential type of the value being unpacked.
-    It is important however that the inferred existential type is normalized. 
-    This is because there might be multiple equivalent existential types 
-    with different order or even number of quantified variables, 
-    and to bind them, the algorithm needs to fix the canonical one.
-  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
-    with annotated let-bindings. The first premise infers the type of the head of the application,
-    which must be a thunked computation. Then if after applying it 
-    to tha arguments, the resulting type can be equated to the annotated one,
-    we infer the body of the let-binding in the context extended with the bound variable.
-  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
-    but it is used when the type of the application is unique, and thus, the annotation
-    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
-    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
-\end{itemize}
-
-Let us discuss the rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTEmptyAppLabel} 
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head. 
-    However, we relax this specification by allowing it to 
-    infer any other equivalent type. 
-    The relaxation of this rule is enough to guarantee 
-    this property for the whole judgement:
-    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
-    $[[iM']]$.
-  \item \ruleref{\ottdruleDTArrowAppLabel}
-    is where the application type is inferred: 
-    if the head has an arrow type $[[iQ → iN]]$,
-     we are allowed to apply it as soon as 
-    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
-  \item \ruleref{\ottdruleDTForallAppLabel}
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    If we are applying a polymorphic computation, 
-    we can instantiate its quantified variables with any types,
-    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
-\end{itemize}
-
-\section{Algorithmic Typing}
-
-Next, we present the type inference algorithm, 
-which is sound and complete with respect to the declarative specification
-(\cref{def:declarative-typing}).
-
-\subsection{Algorithmic Type Inference}
-
-Mirroring the declarative typing, 
-the algorithm is represented by an inference system of three mutually recursive
-judgments:
-\begin{itemize}
-  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
-    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
-    In contrast with the declarative counterparts, they are deterministic,
-    and  guarantee that the inferred type is normalized. 
-  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
-    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
-    Notice that $[[uN]]$ contains algorithmic variables, 
-    which are specified by the context $[[Θ1]]$.
-    Moreover, the inferred type $[[uM]]$ is also algorithmic,
-    and can have several non-equivalent instantiations. To accommodate that, 
-    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
-    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
-    in which the variables must be instantiated, and $[[SC]]$
-    imposes restrictions on the variables. 
-\end{itemize}
-As subroutines, the algorithm calls
-subtyping (\cref{alg:subtyping}),
-type well-formedness (\cref{alg:wf}),
-constraint merge (\cref{sec:constraint-merge}),
-normalization (\cref{alg:type-nf}),
-and constraint singularity which will be defined later in \cref{todo}.
-It also relies on basic set operations and the ability to 
-deterministically choose fresh variables.
-
-\begin{algorithm}
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnATNInf{}
-
-  \columnbreak
-  \ottdefnATPInf{}
+    \begin{subfigure}[b]{0.3\textwidth}
+        \centering
         \ottdefnATSpinInf{}
-  \end{multicols}
-\end{algorithm}
-
-Let us discuss the inference rules of the algorithm:
-\begin{itemize}
-  \item \ruleref{\ottdruleATVarLabel} 
-    infers the type of a variable by looking it up in the context
-    and normalizing the result.
-  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
-    are similar to the declarative rules: they make a recursive call
-    to type the body of the thunk or the return expression and
-    put the shift on top of the result.
-  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
-    are symmetric. They make a recursive call to infer the type of the annotated
-    expression, check that the inferred type is a subtype of the annotation,
-    and return the normalized annotation.
-  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
-    It makes a recursive call to infer the type of the body in the extended context,
-    and returns the corresponding arrow type. Notice that the algorithm also
-    normalizes the result, which is because the annotation type $[[iP]]$
-    is allowed to be non-normalized.
-  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
-    Similarly to the previous case, it makes a recursive call to infer the type
-    of the body in the extended \emph{type} context. 
-    After that, it returns the corresponding universal type. 
-    It is also required to normalize the result, because, 
-    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
-    in which case the $[[∀]]$ must be removed. 
-  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
-    it makes a recursive call to infer the type of the bound value,
-    and then returns the type of the body in the extended context.
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
-    is interpreted as follows.
-    First, it infers the type of the head of the application,
-    ensuring that it is a thunked computation $[[↓iM]]$;
-    after that, it makes a recursive call
-    to the application inference procedure,
-    which returns the algorithmic type, whose
-    instantiation to a declarative type must be associated with the bound variable 
-    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
-    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
-    check whether the instantiation to the annotated type $[[iP]]$ is possible,
-    and if it is, the algorithm infers the type of the body in the extended context,
-    and returns it as the result. 
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
-    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
-    However, since there is no annotation, 
-    instead of checking the instantiation to it, 
-    the algorithm checks that the inferred type
-    $[[↑uQ]]$ is unique.
-    It is the case if all the algorithmic variables of 
-    $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$,
-    which is checked by the combination of
-    $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
-    Together, these two premises guarantee that the only 
-    possible instantiation of $[[↑uQ]]$ is $[[ [uσ]uQ ]]$.
-  \item \ruleref{\ottdruleATUnpackLabel}
-    works in the expected way. First, it infers the 
-    existential type $[[∃nas.iP]]$ of the value being unpacked,
-    and since the type is guaranteed to be normalized, binds 
-    the quantified variables with $[[nas]]$.
-    Then it infers the type of the body in the appropriately extended context,
-    and checks that the inferred type does not depend on $[[nas]]$
-    by checking well-formedness $[[Γ ⊢ iN]]$.
-\end{itemize}
-
-Finally, let us discuss the algorithmic rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleATEmptyAppLabel}
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head,
-    and the algorithm returns it after normalizing.
-  \item \ruleref{\ottdruleATArrowAppLabel}
-    is the main rule of the algorithmic application inference.
-    If the head has an arrow type $[[uQ → uN]]$,
-    we find $[[SC1]]$---the minimal constraint ensuring that 
-    $[[uQ]]$ is a supertype of the first argument's type.
-    Then we make a recursive call applying $[[uN]]$ to the rest of the arguments,
-    and merge the resulting constraint with $[[SC1]]$
-  \item \ruleref{\ottdruleATForallAppLabel},
-    analogously to the declarative case,
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    This is the place where the algorithmic variables are generated.
-    The algorithm simply replaces the quantified variables 
-    $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
-    and makes a recursive call in the extended context. 
-\end{itemize}
-
-The correctness of the algorithm consists of its soundness and 
-completeness, which is by mutual
-induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
-The simplified result is the following.
-\begin{theorempreview}
-  \hfill
-  \begin{itemize}
-    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
-      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
-    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
-      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraint Singularity}
-
-The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
-of the algorithmic typing check wether the constraint $[[SC]]$
-is uniquely defines a substitution satisfying it, and if it does
-returns such a substitution as the result.
-To do that, we define a partial function $[[SC singular with uσ]]$,
-taking a subtyping constraint $[[SC]]$ and returning a substitution 
-$[[uσ]]$---the only possible solution of $[[SC]]$.
-
-First, we define the notion of singularity on constraint entries. 
-$[[scE singular with iP]]$ and $[[scE singular with iN]]$
-are considered partial functions taking a constraint entry $[[scE]]$
-and returning the type satisfying $[[scE]]$ if such a type is unique.
-
-\begin{algorithm}[Singular Constraint Entry]
-  \hfill
-  \begin{multicols}{2}
-    \ottdefnSINGscEN{}
-
-    \columnbreak
-    \ottdefnSINGscEP{}
-  \end{multicols}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSINGNEqLabel} and \ruleref{\ottdruleSINGPEqLabel} 
-    are symmetric. If the constraint entry says that a variable must be equivalent to 
-    a type $T$, then it is evidently singular, and the only (up-to-equivalence) type
-    instantiating this variable could be $T$. This way, we return its normal form. 
-  \item \ruleref{\ottdruleSINGSupVarLabel}
-    implies that the only (normalized) solution of $[[pua :≥ ∃nas.pa]]$ is 
-    $[[pa]]$ (it will be shown in \cref{lemma:var-subt}).
-  \item \ruleref{\ottdruleSINGSupShiftLabel}
-    is perhaps the least obvious rule.
-    In type $[[∃nas.↓iN]]$, if $[[iN]]$ is anything different (non-equivalent) to 
-    $[[nai ∊ {nas}]]$, there are at least 
-    one proper supertype of $[[∃nas.↓iN]]$, since $[[iN]]$ can be abstracted over
-    by an existential quantifier. Otherwise, any supertype of $[[∃nas.↓nai]]$
-    is equivalent to it, and thus, the solution is unique. 
-\end{itemize}
-
-Next, we extrapolate the singularity function on constraints---sets of constraint entries. 
-We require $[[SC]]$ to be a set of singular constraints, and the resulting substitution
-sends each variable from $[[dom(SC)]]$ to the unique type satisfying the corresponding constraint.
-
-\begin{algorithm}
-  $[[SC singular with uσ]]$
-  means that 
-  \begin{enumerate}
-    \item for any positive $[[scE ∊ SC]]$,
-      there exists $[[iP]]$ such that $[[scE singular with iP]]$, 
-      and for any negative $[[scE ∊ SC]]$,
-      there exists $[[iN]]$ such that $[[scE singular with iN]]$;
-    \item $[[uσ]]$ is defined as follows:
-      $$
-      [[ [uσ]β̂⁺ ]]  = 
-          \begin{cases}
-              [[ iP ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁺]] 
-                         \text{ and } [[scE singular with iP]] \\
-              [[ β̂⁺ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-      $$
-      [[ [uσ]β̂⁻ ]]  = 
-          \begin{cases}
-              [[ iN ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁻]] 
-                         \text{ and } [[scE singular with iN]]\\
-              [[ β̂⁻ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-  \end{enumerate} 
-\end{algorithm}
-
-The correctness of the singularity algorithm is formulated as follows:
-\begin{theorempreview}
-  Suppose that $[[SC]]$ is a subtyping constraint.
-  Then $[[SC singular with uσ]]$ holds if and only if 
-  $[[uσ]]$ is the only (up-to-equivalence on $[[dom(SC)]]$) 
-  normalized substitution satisfying $[[SC]]$.
-\end{theorempreview}
-
-\newpage
-
-\section{Properties of the Declarative Type System}
-
-\subsection{Type Well-formedness}
-\input{\genDir/tex/wf-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-lemmas.tex}
-
-\subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Equivalence}
-\label{sec:decl-equiv-lemmas}
-\input{\genDir/tex/equiv-lemmas.tex}
-
-\subsection{Variable Ordering}
-\input{\genDir/tex/ord-lemmas.tex}
-
-\subsection{Normaliztaion}
-\input{\genDir/tex/norm-lemmas.tex}
-
-\section{Properties of the Algorithmic Type System}
-
-\subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-algo-lemmas.tex}
-
-\subsection{Normalization}
-\input{\genDir/tex/norm-alg-lemmas.tex}
-
-\subsection{Equivalence}
-\input{\genDir/tex/equiv-alg-lemmas.tex}
-
-\subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-lemmas.tex}
-
-\subsection{Unification}
-\input{\genDir/tex/unification-lemmas.tex}\
-
-\subsection{Anti-unification}
-\input{\genDir/tex/au-lemmas.tex}
-
-\subsection{Upper Bounds}
-\label{sec:alg-upper-bounds-proofs}
-\input{\genDir/tex/lub-lemmas.tex}
-
-\subsection{Upgrade}
-\label{sec:upgrade-lemmas}
-\input{\genDir/tex/upgrade-lemmas.tex}
-
-\subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-lemmas.tex}
-
-\subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-lemmas.tex}
-
-\subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-lemmas.tex}
-
-\subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-lemmas.tex}
-
-\section{Properties of the Declarative Typing}
-\input{\genDir/tex/decl-typing-lemmas.tex}
-
-\section{Properties of the Algorithmic Typing}
-
-\subsection{Singularity}
-\input{\genDir/tex/singularity-lemmas.tex}
-
-\subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-lemmas.tex}
-
-\printbibliography
-
-\end{document}
+    \end{subfigure}
+    \caption{Algorithmic Type Inferences}
+    \label{fig:type-inference}
+\end{figure}

### 2023-10-17 proofs/main.tex
Fix structure
Restructure, add biblio
Restructure
WIP
wip

@@
-\UseRawInputEncoding
+\PassOptionsToPackage{prologue,dvipsnames}{xcolor}
+
+% \UseRawInputEncoding
 % vim: ft=tex
-\documentclass[a4,natbib=false]{article}
-\usepackage[a4paper, total={8in, 10in}]{geometry}
-\usepackage{hyperref}
-\usepackage{mathpartir}
-
-\input{prelude.tex}
+
+% \documentclass[acmsmall,natbib=false]{article}
+% \usepackage[a4paper, total={8in, 10in}]{geometry}
+
+\documentclass[acmsmall,natbib=false,review,anonymous]{acmart}
+\input{../prelude.tex}
 
@@
 \newcommand{\genDir}{_gen}
-\input{\ottDir/included.tex}
-
-% \renewcommand{\ottdruleEOneNVarName}[0]{ Hello \def\@currentlabelname{FOO} \phantomsection  }
-% \newcommand{\ottdrulename}[1]{\textsc{#1}}
-
-% \renewcommand{\ottdrule}[4][]{{\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}%
-%   }\ruleLabel{#4}{#4}}
-
-% \renewcommand{\ottdrule}[4][]{%
-%   {\displaystyle\frac{\begin{array}{l}#2\end{array}}{#3}\quad\ottdrulename{#4}}%
-%   \mpr@label{\textsc{#1}}%
+\input{\ottDir/defs.tex}
+\input{\ottDir/renew-ott.tex}
+
+\usepackage{../bibliography}
+% \usepackage{amssymb}
+
+\usepackage[matha]{mathabx}
+
+% https://tex.stackexchange.com/questions/85033/colored-symbols/85035#85035
+% \newcommand*{\mathcolor}{}
+% \def\mathcolor#1#{ \mathcoloraux{#1} }
+% \newcommand*{\mathcoloraux}[3]{%
+%   \protect\leavevmode
+%   \begingroup
+%   \color#1{#2}#3%
+%   \endgroup
 % }
-% \DeclareDocumentCommand \redrule {m m o}{%
-%   \inferrule*[vcenter,left={#3:}]{}{#1 \tred #2}
-%   \mpr@label{\textsc{#3}}
-% }
-
-% \renewcommand{\ottdrulename}[2][#1]{\textsc{#1} \label{}}
-
-% \renewcommand{\ottdruleEOneNVarName}[0]{foo }
-
-% \makeatletter
-% \renewcommand{\ottdruleEOneNVar}[1]{\ottdrule[#1]{%
-%   }{
-%     % \protected@edef\@currentlabelname{foo}
-%     \def\@currentlabelname{foo}%
-%     \phantomsection
-%     \label{boo}
-%     \alpha ^{-}   \eqEOne   \alpha ^{-}
-%   }{%
-%     {\ottdruleEOneNVarName}{}%
-%   } }
-% \makeatother
-
-\newcommand{\depth}[1]{\ensuremath{\mathsf{depth}(#1)}}
-\newcommand{\size}[1]{\ensuremath{\mathsf{size}(#1)}}
-
-\newcommand{\ruleref}[1]{Rule \nameref{#1}}
-
-% ord varset in uN = varset'
-
-\renewcommand{\ottdruleONVarInName}[0]{(Var$_{-\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONVarNInName}[0]{(Var$_{-\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleONUVarName}[0]{(UVar$_{-} ^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftUName}[0]{($\uparrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOArrowName}[0]{($\rightarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOForallName}[0]{($\forall^{\text{Ord}}$)}
-
-% ord varset in uP = varset'
-
-\renewcommand{\ottdruleOPVarInName}[0]{(Var$_{+\in}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPVarNInName}[0]{(Var$_{+\notin}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOPUVarName}[0]{(UVar$_{+}^{\text{Ord}}$)}
-\renewcommand{\ottdruleOShiftDName}[0]{($\downarrow^{\text{Ord}}$)}
-\renewcommand{\ottdruleOExistsName}[0]{($\exists^{\text{Ord}}$)}
-
-% nf(N) = M
-\renewcommand{\ottdruleNrmNVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmNUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftUName}[0]{($\uparrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmArrowName}[0]{($\rightarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmForallName}[0]{($\forall^{\text{nf}}$)}
-
-% nf(P) = Q
-\renewcommand{\ottdruleNrmPVarName}[0]{(Var$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmPUVarName}[0]{(UVar$_{-}^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmShiftDName}[0]{($\downarrow^{\text{nf}}$)}
-\renewcommand{\ottdruleNrmExistsName}[0]{($\exists^{\text{nf}}$)}
-
-% N ≈ M
-
-\renewcommand{\ottdruleEOneNVarName}[0]{(Var$_-^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftUName}[0]{($\uparrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneArrowName}[0]{($\rightarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneForallName}[0]{($\forall^{\eqEOne}$)}
-
-% P ≈ Q
-\renewcommand{\ottdruleEOnePVarName}[0]{(Var$_+^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneShiftDName}[0]{($\downarrow^{\eqEOne}$)}
-\renewcommand{\ottdruleEOneExistsName}[0]{($\exists^{\eqEOne}$)}
-
-% G ⊢ N ≤1 M
-
-\renewcommand{\ottdruleDOneNVarName}[0]{(Var$_-^{\subDOne}$)}
-\renewcommand{\ottdruleDOneShiftUName}[0]{($\uparrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneArrowName}[0]{($\rightarrow^{\subDOne}$)}
-\renewcommand{\ottdruleDOneForallName}[0]{($\forall^{\subDOne}$)}
-
-% G ⊢ P ≥1 Q
-\renewcommand{\ottdruleDOnePVarName}[0]{(Var$_+^{\supDOne}$)}
-\renewcommand{\ottdruleDOneShiftDName}[0]{($\downarrow^{\supDOne}$)}
-\renewcommand{\ottdruleDOneExistsName}[0]{($\exists^{\supDOne}$)}
-
-% G ⊢ N ≈1 M
-\renewcommand{\ottdruleDOneNDefName}[0]{($\eqDOne_{-}$)}
-
-% G ⊢ P ≈1 Q
-\renewcommand{\ottdruleDOnePDefName}[0]{($\eqDOne_{+}$)}
-
-% G ⊨ iP1 ∨ iP2 = iQ
-\renewcommand{\ottdruleLUBVarName}[0]{(Var$^{\vee}$)}
-\renewcommand{\ottdruleLUBShiftName}[0]{($\downarrow^{\vee}$)}
-\renewcommand{\ottdruleLUBExistsName}[0]{($\exists^{\vee}$)}
-\renewcommand{\ottdruleLUBUpgradeName}[0]{(Upg)}
-
-% G ; Θ ⊨ uN ≤ iM ⫤ us
-\renewcommand{\ottdruleANVarName}[0]{(Var$_-^{\subA}$)}
-\renewcommand{\ottdruleAShiftUName}[0]{($\uparrow^{\subA}$)}
-\renewcommand{\ottdruleAArrowName}[0]{($\rightarrow^{\subA}$)}
-\renewcommand{\ottdruleAForallName}[0]{($\forall^{\subA}$)}
-
-% G ; Θ ⊨ iP ≥ uQ ⫤ us
-\renewcommand{\ottdruleAPVarName}[0]{(Var$_+^{\supA}$)}
-\renewcommand{\ottdruleAShiftDName}[0]{($\downarrow^{\supA}$)}
-\renewcommand{\ottdruleAExistsName}[0]{($\exists^{\supA}$)}
-\renewcommand{\ottdruleAPUVarName}[0]{(UVar$^{\supA}$)}
-
-% Γ ⊢ scE1 & scE2 = scE3 :: :: E :: 'E'
-\renewcommand{\ottdruleSCMESupSupName}[0]{$([[≥]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMEEqSupName}[0]{$([[≈]]\&^{+}[[≥]])$}
-\renewcommand{\ottdruleSCMESupEqName}[0]{$([[≥]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMEPEqEqName}[0]{$([[≈]]\&^{+}[[≈]])$}
-\renewcommand{\ottdruleSCMENEqEqName}[0]{$([[≈]]\&^{-}[[≈]])$}
-
-% Γ ; Θ ⊨ uN ≈u iM ⫤ UC 
-\renewcommand{\ottdruleUNVarName}[0]{(Var$_{-}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftUName}[0]{($\uparrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUArrowName}[0]{($\rightarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUForallName}[0]{($\forall^{[[≈u]]}$)}
-\renewcommand{\ottdruleUNUVarName}[0]{(UVar$_{-}^{[[≈u]]}$)}
-
-% Γ ; Θ ⊨ uP ≈u iQ ⫤ UC 
-\renewcommand{\ottdruleUPVarName}[0]{(Var$_{+}^{[[≈u]]}$)}
-\renewcommand{\ottdruleUShiftDName}[0]{($\downarrow^{[[≈u]]}$)}
-\renewcommand{\ottdruleUExistsName}[0]{($\exists^{[[≈u]]}$)}
-\renewcommand{\ottdruleUPUVarName}[0]{(UVar$_{+}^{[[≈u]]}$)}
-
-% G ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )
-\renewcommand{\ottdruleAUPVarName}[0]{(Var$_{+}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
-
-% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
-\renewcommand{\ottdruleAUNVarName}[0]{(Var$_{-}^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-\renewcommand{\ottdruleAUAUName}[0]{(AU)}
-
-% Γ ⊢ iN 
-\renewcommand{\ottdruleWFTNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ⊢ iP
-\renewcommand{\ottdruleWFTPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFTExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uN
-\renewcommand{\ottdruleWFATNVarName}[0]{(Var$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATNUVarName}[0]{(UVar$_{-}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftUName}[0]{($\uparrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATArrowName}[0]{($\rightarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATForallName}[0]{($\forall^{\text{WF}}$)}
-
-% Γ ; Ξ ⊢ uP
-\renewcommand{\ottdruleWFATPVarName}[0]{(Var$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATPUVarName}[0]{(UVar$_{+}^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATShiftDName}[0]{($\downarrow^{\text{WF}}$)}
-\renewcommand{\ottdruleWFATExistsName}[0]{($\exists^{\text{WF}}$)}
-
-% Γ ⊢ iP : scE
-\renewcommand{\ottdruleSATSCESupName}[0]{($[[:≥]]_{+}^\text{sat}$)}
-\renewcommand{\ottdruleSATSCEPEqName}[0]{($[[:≈]]_{+}^\text{sat}$)}
-
-% Γ ⊢ iN : scE
-\renewcommand{\ottdruleSATSCENEqName}[0]{($[[:≈]]_{-}^\text{sat}$)}
-
-% Γ ; Φ ⊢ v : iP 
-\renewcommand{\ottdruleDTVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-\renewcommand{\ottdruleDTPEquivName}[0]{($[[≈]]_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ c : iN 
-\renewcommand{\ottdruleDTtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleDTReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleDTAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-\renewcommand{\ottdruleDTNEquivName}[0]{($[[≈]]_-^{\text{inf}}$)}
-
-% Γ ; Φ ⊢ iN ● args ⇒> iM 
-\renewcommand{\ottdruleDTEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleDTForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ v : iP  
-\renewcommand{\ottdruleATVarName}[0]{(Var$^{\text{inf}}$)}
-\renewcommand{\ottdruleATThunkName}[0]{($\{\}^{\text{inf}}$)}
-\renewcommand{\ottdruleATPAnnotName}[0]{(ann$_+^{\text{inf}}$)}
-
-% Γ ; Φ ⊨ c : iN 
-\renewcommand{\ottdruleATtLamName}[0]{($\lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATTLamName}[0]{($\Lambda^{\text{inf}}$)}
-\renewcommand{\ottdruleATReturnName}[0]{(ret$^{\text{inf}}$)}
-\renewcommand{\ottdruleATVarLetName}[0]{(let$^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetName}[0]{(let$_@^{\text{inf}}$)}
-\renewcommand{\ottdruleATAppLetAnnName}[0]{(let$_{:@}^{\text{inf}}$)}
-\renewcommand{\ottdruleATUnpackName}[0]{(let$_{\exists}^{\text{inf}}$)}
-\renewcommand{\ottdruleATNAnnotName}[0]{(ann$_-^{\text{inf}}$)}
-
-% Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC 
-\renewcommand{\ottdruleATEmptyAppName}[0]{($\emptyset_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATArrowAppName}[0]{($\rightarrow_{[[●]][[⇒>]]}^{\text{inf}}$)}
-\renewcommand{\ottdruleATForallAppName}[0]{($\forall_{[[●]][[⇒>]]}^{\text{inf}}$)}
-
-% scE1 singular with iP
-\renewcommand{\ottdruleSINGPEqName}[0]{($[[≈]]_{+}^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupVarName}[0]{($[[:≥]]\alpha^{\text{sing}}$)}
-\renewcommand{\ottdruleSINGSupShiftName}[0]{($[[:≥]][[↓]]^{\text{sing}}$)}
-
-% scE1 singular with iN
-\renewcommand{\ottdruleSINGNEqName}[0]{($[[≈]]_{-}^{\text{sing}}$)}
 
 \begin{document}
+
+\title{The Proofs}
+
+\maketitle
 
@@
       \begin{itemize}
-        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$;
-        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$;
-        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$;
-        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$;
-        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$.
-        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$, 
-        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (here we assume that $[[nas]]$ and $[[pas]]$ are lists of fresh variables, 
-          that is the variable capture never happens);
+        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$
+        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$
+        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$
+        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$
+        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$
+        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$ 
+        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (assuming the variable capture never happens)
       \end{itemize}
@@
     \ottdefnDOneNsub{}
-    \columnbreak\\
+
     \ottdefnDOnePsup{}
   \end{multicols}
+  \hfill
 
@@
     \ottdefnDOneNeq{}
-    \columnbreak\\
+
     \ottdefnDOnePeq{}
@@
 This way, we add two base cases to the grammar of 
-positive and negative types, and use highlight to denote that the type
+positive and negative types and use highlight to denote that the type
 can potentially contain algorithmic variables.
@@
 algorithmic subtyping,
-we turn a declarative type into the algorithmic one
+we turn a declarative type into an algorithmic one
 via replacing certain type variables with fresh algorithmic variables.
@@
 
-  \begin{multicols}{2}
   \ottdefnONVar{}
-  \columnbreak\\
   \ottdefnOPVar{}
-  \end{multicols}
 
@@
 
-  Analogously,  we define normalization of algorithmic types 
-  by adding base cases:
+  Analogously, we define the normalization of algorithmic types by adding base cases:
 
@@
 does not invoke the negative one. This way, the correctness of the positive 
-subtyping will be proved independently, and used afterwards to prove the
+subtyping will be proved independently and used afterwards to prove the
 correctness of the negative subtyping.
@@
   \label{alg:subtyping}
-  \hfill
-
-  \begin{multicols}{2}
+  \hfill\\
   \ottdefnANsub{}
-  \columnbreak\\
   \ottdefnAPsup{}
-  \end{multicols}
 \end{algorithm}
@@
 $[[return v]]$ and $[[{c}]]$. 
-The former allows us to embed a value in a pure 
+The former allows us to embed a value in pure 
 computations. The latter, on the contrary,
@@
 
-Finally, the language has a number standard constructions:
+Finally, the language has several standard constructions:
 lambda-abstractions $[[λx:iP.c]]$,
@@
 inference for our language. First, we introduce 
-variable context specifying the types of the variables 
+variable context specifying the types of variables 
 in the scope of the current rule.
@@
   \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
-    judgement. It means that if a head of type $[[iN]]$ 
+    judgment. It means that if a head of type $[[iN]]$ 
     is applied to list of values $[[args]]$, 
@@
     which must be a thunked computation. Then if after applying it 
-    to tha arguments, the resulting type can be equated to the annotated one,
+    to the arguments, the resulting type can be equated to the annotated one,
     we infer the body of the let-binding in the context extended with the bound variable.
@@
 \begin{algorithm}
-  \hfill
-  \begin{multicols}{2}
+  \hfill\\
   \ottdefnATNInf{}
-
-  \columnbreak
   \ottdefnATPInf{}
   \ottdefnATSpinInf{}
-  \end{multicols}
 \end{algorithm}
@@
 The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
-of the algorithmic typing check wether the constraint $[[SC]]$
+of the algorithmic typing check whether the constraint $[[SC]]$
 is uniquely defines a substitution satisfying it, and if it does
@@
 \begin{algorithm}[Singular Constraint Entry]
-  \hfill
-  \begin{multicols}{2}
+  \hfill\\
+  \ottdefnSINGscEP{}
   \ottdefnSINGscEN{}
-
-    \columnbreak
-    \ottdefnSINGscEP{}
-  \end{multicols}
 \end{algorithm}

### 2023-10-17 proofs/tex/subst-algo-lemmas.tex
wip

@@
   $[[nf(uσ1)|(uv uP) = nf(uσ2)|(uv uP)]]$,
-  and by ${{corollary:subst-subt-equiv-algorithmization}}$,
+  and by \cref{corollary:subst-subt-equiv-algorithmization},
   $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$.

### 2023-10-17 oopsla24/tex/01-introduction.tex
some more intro text

@@
 % How we should frame this:
+
+Over the last half-century, there has been considerable work on developing type inference algorithms for programming languages,  mostly aimed at solving the problem of \emph{type polymorphism}.
+
+That is, in pure polymorphic lambda calculus (system F), the polymorphic type $\forall a.\,A$ has a big lambda $\Lambda a.\,e$ as an introduction form, and an explicit type application $e\,[A]$ as an elimination form.
+This is an extremely simple and compact type system, whose rules fit on a single page, but whose semantics are sophisticated enough to model things like parametricity and representation independence.
+However, System F by itself is unwieldy as a programming language.
+The fact that the universal type $\forall a.\,A$ has explicit eliminations means that programs written using polymorphic types will need to be stuffed to the gills with type annotations explaining how and when to instantiate the quantifiers.
+
+Therefore, most work on type inference has been aimed at handling type instantiations implicitly -- we want to be able to use a polymorphic function like $\mathsf{len} : \forall a.\, \mathsf{List}\,a \to \mathsf{int}$ at any concrete list type without explicitly instantiating the quantifier in $\mathsf{len}$'s type.
+That is, we want to write $\mathsf{len}\,[1,2,3]$ instead of writing $\mathsf{len}\,[\mathsf{int}]\,[1,2,3]$.
+
+The most famous of the algorithms for solving these constraints is the Damas-Hindley-Milner algorithm.
+The idea is that type instantiation induces a subtype ordering: the type $\forall a.\,A$ is a subtype of all of its instantiations.
+So we wish to be free to use the same function $\mathsf{len}$ at many different types such as
+$\mathsf{List}\,\mathsf{int} \to \mathsf{int}$,
+$\mathsf{List}\,\mathsf{bool} \to \mathsf{int}$,
+$\mathsf{List}\,(\mathsf{int \times bool}) \to \mathsf{int}$, and so on.
+However, the subtype relation is nondeterministic: it tells us that whenever we see a polymorphic type $\forall a.\, A$, we know it is a subtype of \emph{any} of its instantiations.
+To turn this into an algorithm, we have to actually make some choices, and DHM works by using \emph{unification}.
+Whenever we would have had to introduce a particular concrete type in the specification, the DHM algorithm introduces a unification variable, and incrementally instantiates this variable as more and more type constraints are observed.
+
+However, the universal quantifier is not the only quantifier!
+Dual to the universal quantifer $\forall$ is the existential quantifier $\exists$.
+Even though existential types have an equal logical status to universal types, they have been studied much less frequently in the literature.
+The most widely-used algorithm for handling existentials is the algorithm of Odersky and Laufer.
+This algorithm treats existentials in a second-class fashion: they are not explicit connectives of the logic, but rather are tied to datatype declarations.
+As a result, packing and unpacking existential types is tied to constructor introductions and constructor eliminations in pattern matches.
 

### 2023-10-24 oopsla24/tex/01-introduction.tex
foo

@@
 As a result, packing and unpacking existential types is tied to constructor introductions and constructor eliminations in pattern matches.
+This allows Damas-Milner inference to continue to work almost unchanged, but it does come at the cost of losing first-class existentials and also of restricting the witness types to monomorphic types.
+
+There has been a limited amount of work on support for existential types as a first-class concept.
+In an unpublished draft, \citet{leijen06} studied an extension of Damas-Milner inference in which type schemes contain alternating universal and existential quantifiers.
+Quantifiers still range over monotypes, and higher-rank polymorphism is not permitted.
+More recently, \citet{dk19} studed type inference for existential types in the context of GADT type inference, which, while still predicative, supported higher-rank types (i.e., quantifiers can occur anywhere inside of a type scheme).
+\citet{existential-crisis} propose a system which is intermediate between prenex and higher-rank polymorphism, but which
+
+All of these papers are restricted to \emph{predicative} quantification, whereas existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with
 

### 2023-10-24 oopsla24/tex/04-algorithm.tex
WIP: Algorithm

@@
 the subtyping and the type inference, which we discuss in this section one after another.
+
+\subsection{Algorithmic Syntax}
+
 First, let us discuss the syntax of the algorithmic system.
@@
         \item[Negative Algorithmic Types] $[[uN]] = \dots \mid [[α̂⁻]]$
-        \item[Algorithmic Type Context] $[[Ξ]] = \text{set of Algorithmic Variables}$
+        \item[Algorithmic Type Context] $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$
+        where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct 
 \end{description}
@@
 previously defined declarative terms and types by adding \emph{algorithmic type variables}
-(\aka existential variables). The algorithmic variables represent unknown types, 
+(\aka unification variables). The algorithmic variables represent unknown types, 
 which cannot be inferred immediately but are promised to be instantiated
@@
 from the declarative to the algorithmic system, when we replace the quantified variables $[[pas]]$ 
-that we are not able to instantiate with their algorithmic counterpart $[[puas]]$.
+with their algorithmic counterpart $[[puas]]$.
 The procedure of replacing declarative variables with algorithmic ones we call 
@@
 \paragraph{Algorithmic Contexts and Well-formedness}
-To specify well-formedness, we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
+To specify when algorithmic types are well-formed, 
+we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
 Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
@@
 \cref{todo}, we also check that each algorithmic variable is in the context $[[Ξ]]$.
+\begin{figure}[h]
+\begin{multicols}{2}
+  $\vcenter{\hbox{\vdots}}$\\
+  \ottusedrule{\ottdruleWFATPUVar{}}
+  \columnbreak\\
+  $\vcenter{\hbox{\vdots}}$\\
+  \ottusedrule{\ottdruleWFATNUVar{}}
+\end{multicols}
+\label{fig:algo-wf}
+\caption{Well-formedness of Algorithmic Types}
+\end{figure}
 
-\paragraph
+\paragraph{Algorithmic Normalization}
+Similarly to well-formedness, the normalization of algorithmic types is defined
+by extending the declarative definition trivially with the algorithmic variables.
+
+\begin{figure}[h]
+\begin{multicols}{2}
+  $\vcenter{\hbox{\vdots}}$\\
+  \ottusedrule{\ottdruleNrmPUVar{}}
+  \columnbreak\\
+  $\vcenter{\hbox{\vdots}}$\\
+  \ottusedrule{\ottdruleNrmNUVar{}}
+\end{multicols}
+\label{fig:algo-nf}
+\caption{Normalization of Algorithmic Types}
+\end{figure}
+
+\subsection{Type Constraints}
+As the algorithm proceeds, it accumulates the information 
+about the algorithmic type variables in the form of \emph{constraints}.
+In our system, the constraints can be of two kinds: 
+\emph{subtyping constraints} and \emph{unification constraints}.
+The subtyping constraint can only have a positive shape $[[α̂⁺ :≥ iP]]$, \ie it 
+restricts a positive algorithmic variable to be a 
+supertype of a certain declarative type---this is one of the invariants that we 
+preserve in the algorithm.
+The unification constraint can have either a positive form $[[α̂⁺ :≈ iP]]$ or
+a negative form $[[α̂⁻ :≈ iN]]$, however, the right-hand side of the constraint
+cannot contain algorithmic type variables.
+The set of constraints is denoted as $[[SC]]$. 
+We assume that the constraints in $[[SC]]$ restrict distinct variables.
+
+We separately define $[[UC]]$ as a set consisting of unification constraints only.
+This is done to simplify the representation of the algorithm. 
+The unification, which we use as a subroutine of the subtyping algorithm,
+can only produce unification constraints.
+A set of unification constraints can be resolved in a simpler way than
+a general constraint set. This way, 
+the separation of the unification constraint resolution into a separate
+procedure allows us to better decompose the structure of the algorithm,
+and thus, simplify the inductive proofs.
+
+  \begin{figure}[h]
+    \begin{multicols}{2}
+      \ottgrammartabular { 
+      \ottscE 
+      \ottinterrule
+      \ottruleheadOneLine
+        {[[SC]]}{::=} {\ottcom{Constraint Set}}
+        {\{[[scE1]], \dots, [[scEn]]\}}\ottprodnewline
+      }
+
+    \columnbreak
+
+      \ottgrammartabular { 
+      \ottucE 
+      \ottprodnewline
+      \ottinterrule
+      \ottruleheadOneLine
+        {[[UC]]}{::=} {\ottcom{Unification Constraint Set}}
+        {\{[[ucE1]], \dots, [[ucEn]]\}}\ottprodnewline
+      }
+    \end{multicols}
+
+    \label{fig:syntax-e-sc}
+    \caption{Constraint Entries and Sets}
+  \end{figure}
+
+\paragraph{Constraint Contexts}
+When one instantiates an algorithmic variable, 
+they may only use type variables available in its scope.
+As such, each algorithmic variable must remember the context at the moment when 
+it was introduced. In our algorithm, this information is represented by
+a \emph{constraint context} $[[Θ]]$---a set of pairs associating 
+algorithmic variables and declarative contexts.
+
+  \begin{figure}[h]
+    \begin{center}
+      $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$
+    \end{center}
+    \label{fig:constraint-context}
+    \caption{Constraint Contexts}
+  \end{figure}
+
+\paragraph{Auxiliary Functions}
+We define $[[dom(SC)]]$---a domain of a constraint set $[[SC]]$ as a set of algorithmic variables
+that it restricts. Similarly, we define $[[dom(Θ)]]$---a domain of constraint context
+as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
+We write $[[Θ(α̂±)]]$ to denote the context associated with $[[α̂±]]$ in $[[Θ]]$.
+
+%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%
+\subsection{Subtyping Algorithm}
+
+  For convenience and scalability, 
+  we decompose the subtyping algorithm 
+  into several procedures. \Cref{fig:alg-subtyping-graph}
+  shows these procedures and the dependencies between them:
+  arrows denote the invocation of one procedure from another.
+  The label <<$\ottkw{nf}$>> annotating arrows means that the calling
+  procedure normalizes the input before passing it to the callee.
 
@@
     \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
+    \node[left=1.5cm of 3] (2) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[right=1.5cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
     \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
@@
     \draw[->] (1) to (3);
-    \draw[->] (1) to (5);
+    \draw[->] (1) to node[above, draw=none]{$\ottkw{nf}$} (5);
     \draw[->] (2) to (3);
@@
     \draw[->] (3) to (4);
-    \draw[->] (3) to (5);
+    \draw[->] (3) to node[above, draw=none]{$\ottkw{nf}$} (5);
     \draw[->] (4) to (6);
     \draw[->] (5) to (8);
-    \draw[->] (6) to (7);
+    \draw[->] (6) to node[left, draw=none]{$\ottkw{nf}$} (7);
   \end{tikzpicture}  
   \caption{Dependency graph of the subtyping algorithm}
-  \label{fig:subtyping-algo}
+  \label{fig:alg-subtyping-graph}
 \end{figure}
 
-%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%
-\subsection{Subtyping Algorithm}
-Hello Subtyping
-        % - fig
-        % - it takes ... and returns constraint
-        % - explain the rules
-        % - - ↑/↓ invoke unification up-to-equivalence, 
-        %     which is implemented by unification of the normal forms up-to-equality.
-        % - - → makes two recursive calls, and thus, it merges the resulting constraints
-        % - - ∀/∃ introduce new variables as discussed
-        % - - the base case calls the upgrade
+In the remainder of this section, we will delve into each of these procedures in
+detail, following the top-down order of the dependency graph. First, 
+we present the subtyping algorithm itself.
+
+As an input, the subtyping algorithm takes
+a type context $[[Γ]]$, a constraint context $[[Θ]]$,
+and two types of the corresponding polarity:
+$[[uN]]$ and  $[[iM]]$ for the negative subtyping, and
+$[[uP]]$ and  $[[iQ]]$ for the positive subtyping.
+We assume the second type ($[[iM]]$ and $[[iQ]]$) to be 
+declarative (with no algorithmic variables) and well-formed in $[[Γ]]$,
+but the first type ($[[uN]]$ and $[[uP]]$) may contain algorithmic variables,
+whose instantiation contexts are specified by $[[Θ]]$.
+
+Notice that the shape of the input types uniquely determines the
+applied subtyping rule.  If the subtyping is successful, it returns
+a set of constraints $[[SC]]$ restricting the algorithmic 
+variables of the first type. If the subtyping does not hold, 
+there will be no inference tree with such inputs.
 
@@
   \hfill\\
+  \begin{multicols}{2}
     \ottdefnANsub{}
+    \columnbreak\\
     \ottdefnAPsup{}
+  \end{multicols}
   \caption{Subtyping Algorithm}
@@
 \end{figure}
+
+The rules of the subtyping algorithm bijectively correspond to the rules of the declarative
+system. Let us discuss them in detail.
+
+\paragraph{Variables} Rules \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel}
+say that if both of the input types are equal declarative variables,
+they are subtypes of each other, with no constraints (as there are no algorithmic variables).
+
+\paragraph{Shifts} Rules \ruleref{\ottdruleAShiftDLabel} and
+\ruleref{\ottdruleAShiftULabel} cover the downshift and the upshift cases,
+respectively. If the input types are constructed by shifts, then the subtyping
+can only hold if they are equivalent. This way, the algorithm must find the
+instantiations of the algorithmic variables on the left-hand side, which make it
+equivalent to the right-hand side. For this purpose, the algorithm invokes the
+unification procedure \cref{todo} preceded by normalization of the input types.
+It returns the resulting constraints given by the unification algorithm.
+
+\paragraph{Quantifiers}  
+Rules \ruleref{\ottdruleAForallLabel} and 
+\ruleref{\ottdruleAExistsLabel} are symmetric. 
+According to the declarative specification, 
+the quantified variables on the left-hand side must be instantiated 
+with types, which, however, are not known in advance.
+We deal with it by algorithmization (\cref{todo})
+of the quantified variables:
+we introduce fresh algorithmic variables
+$[[puas]]$ or $[[nuas]]$,
+put them into the constraint context $[[Θ]]$
+(specifying that they must be instantiated in the extended context
+$[[Γ, pbs]]$ or $[[Γ, nbs]]$) and substitute the quantified variables
+for them in the input type.
+
+After that, the algorithm proceeds with the recursive call, returning
+constraints $[[SC]]$. As the output, the algorithm removes the freshly
+introduced algorithmic variables from the constraint context, This operation is
+sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
+instantiation of the freshly introduced algorithmic variables is not important,
+as they do not occur in the input types.
+
+\paragraph{Functions}
+To infer the subtyping of the function types, the algorithm
+makes two calls: 
+\begin{enumerate*}
+  \item[(i)] a recursive call ensuring the subtyping of the result types, and
+  \item[(ii)] a call to positive subtyping (or rather super-typing) on the argument types.
+\end{enumerate*}
+The resulting constraints are merged (using a special procedure defined later in \cref{todo})
+and returned as the output.
+
+\paragraph{Algorithmic Variable}
+If one of the sides of the subtyping is a unification variable,
+the algorithm must create a new restriction. 
+Because the right-hand side of the subtyping is always declarative,
+it is only the left-hand side that can be a unification variable.
+Moreover, another invariant we preserve prevents the negative
+algorithmic variables from occurring in types during the
+ negative subtyping algorithm. It means
+that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]] [[iP]]$,
+which is covered by \ruleref{\ottdruleAPUVarLabel}.
+
+The potential problem here is that the type $[[iP]]$
+might be not well-formed in the context required for $[[α̂⁺]]$ by $[[Θ]]$, 
+because this context might be smaller than the current context $[[Γ]]$.
+As we wish the resulting constraint set to be sound \wrt $[[Θ]]$,
+we cannot simply put $[[α̂⁺ :≥ iP]]$ into the output. 
+Prior to that, we update the type $[[iP]]$ to its lowest supertype $[[iQ]]$
+well-formed in $[[Θ(α̂⁺)]]$. It is done by the \emph{upgrade} procedure,
+which we discuss in detail in \cref{todo}.
+
+\vspace{\baselineskip}
+% \indent
+To summarize, the subtyping algorithm uses the following additional subroutines:
+\begin{enumerate*}[noitemsep]
+  \item[(i)] rules \ruleref{\ottdruleAShiftDLabel} and
+    \ruleref{\ottdruleAShiftULabel} invoke the \emph{unification} algorithm
+    to equate the input types;
+  \item[(ii)] rule \ruleref{\ottdruleAArrowLabel} \emph{merges} the constraints
+    produced by the recursive calls on the result and the argument types; and
+  \item[(iii)] rule \ruleref{\ottdruleAPUVarLabel} \emph{upgrades} the input type
+    to its least supertype well-formed in the context required by the
+    algorithmic variable.
+\end{enumerate*}
+The following sections discuss these additional procedures in detail.
 
@@
 
-Hello Unification
+As an input the unification context 
+takes a type context $[[Γ]]$, a constraint context $[[Θ]]$,
+and two types of the required polarity:
+$[[uN]]$ and  $[[iM]]$ for the negative unification, and
+$[[uP]]$ and  $[[iQ]]$ for the positive unification.
+It is assumed that only the left-hand side type may contain algorithmic variables,
+this way, the left-hand side is well-formed as an algorithmic type in $[[Γ]]$ and 
+$[[Θ]]$, whereas the right-hand side is well-formed declaratively in $[[Γ]]$.
+
+Since only the left-hand side may contain algorithmic variables,
+that the unification instantiates, we could have called this procedure \emph{matching}.
+However, in \cref{todo} we will discuss several modifications of the 
+type system, where this invariant is not preserved, and thus, this procedure
+becomes a genuine first-order pattern unification \cite{miller1991:unification}.
+
+As the output, the unification algorithm returns the weakest
+set of unification constraints $[[UC]]$ such that any instantiation 
+satisfying these constraints unifies the input types.
 
@@
 
+The algorithm works as one might expect:
+if both sides are formed by constructors, 
+it is required that the constructors are the same, and the
+types unify recursively. If one of the sides
+is a unification variable (in our case it can only be the left-hand side),
+we create a new unification constraint restricting it to be equal to the other side.
+Let us discuss the rules that implement this strategy.
+
+\paragraph{Variables}
+  The variable rules \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
+  are trivial: as the input types do not have algorithmic variables, and are already equal, 
+  the unification returns no constraints.
+
+\paragraph{Shifts}
+  The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
+  require the input types to be formed by the same shift constructor. 
+  They remove this constructor, unify the types recursively, and return the resulting
+  set of constraints.
+
+\paragraph{Quantifiers} 
+  Similarly, the quantifier rules \ruleref{\ottdruleUForallLabel} and \ruleref{\ottdruleUExistsLabel}
+  require the quantifier variables on the left-hand side and the right-hand side to be the same.
+  This requirement is complete because we assume the input types of the unification 
+  to be normalized, and thus, the equivalence implies alpha-equivalence. 
+  In the implementation of this rule, an alpha-renaming might be needed to ensure 
+  that the quantified variables are the same, however, we omit it for brevity.
+
+\paragraph{Functions}
+  Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
+  First, it unifies the argument types and their result types recursively
+  Then it merges the resulting constraints using the constraint merge procedure (\cref{todo}).
+
+  Notice that the resulting constraints can only have \emph{unification} entries.
+  It means that they can be merged in a simpler way than general constraints.
+  In particular, the merging procedure does not call any of the subroutines discussed here,
+  but rather simply checks the matching constraint entries for equality.
+
+\paragraph{Algorithmic Variable}
+  Finally, if the left-hand side of the unification is an algorithmic variable,
+  \ruleref{\ottdruleUNVarLabel} or \ruleref{\ottdruleUPVarLabel} is applied. 
+  It simply checks that the right-hand side type is well-formed in the required
+  constraint context, and returns a newly created constraint restricting the variable
+  to be equal to the right-hand side type.
+
+\vspace{\baselineskip}
+As one can see, the unification procedure is standard, 
+except that it makes sure that the resulting instantiations agree with the input
+constraint context $[[Θ]]$.  As a subroutine, the unification algorithm only uses the (unification) 
+constraint merge procedure and the well-formedness checking.
+
 %%%%%%%%%%%%%%%%%
@@
 
-\begin{definition} [Matching Entries]
-  We call two unification constraint entries 
-  or two subtyping constraint entries matching 
-  if they are restricting the same unification variable.
-\end{definition}
+The merge procedure that we discuss in this section, allows one to 
+combine two constraint sets into one. It might seem that 
+just taking the union of the two sets would be enough, however,
+we require the constraint sets to have a certain structure,
+in particular not to have two entries restricting the same algorithmic
+ variable---we call such entries \emph{matching}.
+The matching entries must be combined into one constraint entry, that 
+would represent their conjunction. 
+This way, to merge two constraint sets, we unite the entries of two
+sets, and then merge the matching pairs.
 
-Two matching entries formed in the same context $[[Γ]]$ 
-can be merged in the following way:
+\paragraph{Merging Matching Constraint Entreies}
+Two \emph{matching} entries formed in the same context $[[Γ]]$ 
+can be merged as shown in \cref{fig:merge-entries}.
+Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
+The result of the merge $[[scE1 & scE2]]$ must be the 
+weakest entry which implies both  $[[scE1]]$ and $[[scE2]]$.
 
@@
   \ottdefnSCME\\
-
+  \caption{Merge of Matching Constraint Entries}
   \label{fig:merge-entries}
-  \caption{Merge Matching Constraint Entries}
 \end{figure}
 
+Suppose that one of the input entries, say $[[scE1]]$, is a unification 
+constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
+(up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
+is implied by $[[scE1]]$.
 \begin{itemize}
-  \item \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}
-    are symmetric cases. To merge two matching entries restricting
-    a variable to be equivalent to certain types, we check
-    that these types are equivalent to each other.
-    To do so, it suffices to check for \emph{equality} of their normal forms,
-    as discussed in \cref{sec:equivalence-normalization}. 
-    After that, we return the left-hand entry.
-
-  \item \ruleref{\ottdruleSCMEEqSupLabel} and \ruleref{\ottdruleSCMESupEqLabel}
-    are also symmetric. 
-    In this case,
-    since one of the entries requires the variable to be equal to 
-    a type, the resulting entry must also imply that.
-    However, for the soundness, it is needed to ensure that
-    the equating restriction is stronger than the subtyping restriction.
-    For this purpose, the premise invokes the positive subtyping.
-
-  \item \ruleref{\ottdruleSCMESupSupLabel} 
-    In this case, we find the least upper bound of the types from the input
-    restrictions, 
-    and as the output, restrict the variable to be a supertype of the result.
-    The least upper bound procedure will be discussed in \cref{sec:lub}.
+  \item If $[[scE2]]$ is also a restricting entry, then the types on the right-hand side
+    of $[[scE1]]$ and $[[scE2]]$ must be equivalent,
+    as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
+  \item If $[[scE2]]$ is a supertype restriction $[[α̂⁺ :≥ iP]]$,
+    the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
+    The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
 \end{itemize}
 
+If both input entries are supertype restrictions: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
+then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
+The least upper bound---$[[iP ∨ iQ]]$ is the least supertype of both $[[iP]]$ and $[[iQ]]$,
+and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
+$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm for finding the least upper bound
+is discussed in \cref{todo}.
+
+\paragraph{Merging Constraint Sets}
+  The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
+  As discussed, the result of merge $[[SC1]]$ and $[[SC2]]$ consists of three parts: 
+  \begin{enumerate*}
+    \item[(i)] the entries of $[[SC1]]$ that do not match any entry of $[[SC2]]$;
+    \item[(ii)] the entries of $[[SC2]]$ that do not match any entry of $[[SC1]]$; and
+    \item[(iii)] the merge (\cref{fig:merge-entries}) of matching entries.
+  \end{enumerate*}
+
 \begin{figure}[h]
-  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
+  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.\\
   Then $[[Θ ⊢ SC1 & SC2 = SC]]$
-  defines a set such that $[[ucE]] \in [[SC]]$ iff either
+  defines a set of constraints $[[SC]]$ such that $[[scE]] \in [[SC]]$ iff either:
   \begin{itemize}
-    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
-    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
-    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
-      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
-      $[[α̂±]]$. 
+    \item $[[scE]] \in [[SC1]]$ and there is no matching $[[scE']] \in [[SC2]]$; or
+    \item $[[scE]] \in [[SC2]]$ and there is no matching $[[scE']] \in [[SC1]]$; or
+    \item $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ for some $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$
+      such that $[[scE1]]$ and $[[scE2]]$ both restrict variable $[[α̂±]]$. 
   \end{itemize}
 
+  \caption{Constraint Merge}
   \label{fig:merge-subtyping-constraints}
-  \caption{Constraint Merge}
 \end{figure}
+
+As shown in \cref{fig:merge-entries}, the merging procedure relies 
+substantially on the least upper bound algorithm.
+In the next section, we discuss this algorithm in detail,
+together with the upgrade procedure, selecting the least supertype 
+ell-formed in a given context.
 
@@
 
+Both type upgrade and the least upper bound algorithms are used
+to find a minimal supertype under certain conditions. 
+For a given type $[[iP]]$ well-formed in $[[Γ]]$, the \emph{upgrade} operation 
+finds the least among those supertypes of $[[iP]]$ that are well-formed
+in a smaller context $[[Δ ⊆ Γ]]$.
+For given two types $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
+the \emph{least upper bound} operation finds the least among
+common supertypes of $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$.
+These algorithms are shown in \cref{fig:type-upgrade}.
+
 \begin{figure}[h]
+  \begin{multicols}{2}
+    \ottdefnLUBUp{}
+    \columnbreak\\
     \ottdefnLUBNsub{}
-  \caption{Least Upper Bound Algorithm}
-  \label{fig:lub}
+  \end{multicols}
+  \caption{Type Upgrade and Leas Upper Bound Algorithms}
+  \label{fig:type-upgrade}
 \end{figure}
 
-\begin{figure}[h]
-    \hfill\\
-    \ottdefnLUBUp{}
-    \caption{Type Upgrade Algorithm}
-    \label{fig:type-upgrade}
-\end{figure}
+\paragraph{The Type Upgarde}
+The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
+It exploits the idea that the free variables of a positive type $[[iQ]]$
+cannot disappear in its subtypes. It means that if 
+a type $[[iP]]$ has free variables not occurring 
+in $[[iP']]$, then any common supertype of $[[iP]]$
+and $[[iP']]$ must not contain these variables either.
+This way, any supertype of $[[iP]]$
+not containing certain variables $[[pnas]]$ must also be 
+a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
+and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
+does not contain $[[pnas]]$ nor $[[pnbs]]$.
+
+This way, to find the least supertype of $[[iP]]$ well-formed in $[[Δ]] = [[Γ \ {pnas}]]$
+(\ie not containing $[[pnas]]$), we can do the following.
+First, construct a new type $[[iP']]$ by renaming $[[pnas]]$ in $[[iP]]$ to fresh $[[pnbs]]$,
+and second, find \emph{the least upper bound} of $[[iP]]$ and $[[iP']]$ in the appropriate
+context. However, for reasons of symmetry, in rule
+\ruleref{\ottdruleLUBUpgradeLabel} we employ a different but equivalent approach:
+we create \emph{two} types $[[iP1]]$ and $[[iP2]]$ constructed by renaming $[[pnas]]$ in $[[iP]]$
+to fresh disjoint variables $[[pnbs]]$ and $[[pncs]]$ respectively, and then 
+find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
+
+\paragraph{The Least Upper Bound}
+The Least Upper Bound algorithm we use operates on \emph{positive}
+types. This way, the inference rules of the algorithm
+analyze the three possible shapes of the input types:
+a variable type, an existential type, and a shifted computation.
+
+ Rule \ruleref{\ottdruleLUBExistsLabel} covers the case when 
+ at least one of the input types is an existential type.
+ In this case, we can simply move the existential quantifiers
+ from both sides to the context, and make a tail-recursive call.
+ However, it is important to make sure that 
+ the quantified variables $[[nas]]$ and $[[nbs]]$ are disjoint
+ (\ie alpha-renaming might be required in the implementation).
+
+ Rule \ruleref{\ottdruleLUBVarLabel} applies when 
+ both sides are variables. In this case,
+ the common supertype only exists if these variables are 
+ the same. And if they are, the common supertypes
+ must be equivalent to this variable.
+
+Rule \ruleref{\ottdruleLUBShiftLabel} is the most 
+interesting. If both sides are not quantified, and one of the sides is 
+a shift, so must be the other side. 
+However, the set of common upper bounds is not trivial in this case.
+For example, $[[↓(β⁺ → γ1⁻)]]$ and $[[↓(β⁺ → γ2⁻)]]$ have
+two non-equivalent common supertypes: 
+$[[∃α⁻.↓α⁻]]$ 
+(by instantiating $[[α⁻]]$ with $[[β⁺ → γ1⁻]]$ and $[[β⁺ → γ2⁻]]$ respectively)
+and 
+$[[∃α⁻.↓(β⁺ → α⁻)]]$ 
+(by instantiating $[[α⁻]]$ with $[[γ1⁻]]$ and $[[γ2⁻]]$ respectively).
+As one can see, the second supertype $[[∃α⁻.↓(β⁺ → α⁻)]]$ is the least among them
+because it abstracts over a `deeper' negative subexpression.
+
+In general, we must 
+\begin{itemize*}
+  \item[(i)] find the most detailed pattern (a type with `holes' at negative positions) 
+    that matches both sides, and 
+  \item[(ii)] abstract over the `holes' by existential quantifiers.
+\end{itemize*}
+The algorithm that finds the most detailed common pattern is called \emph{anti-unification}.
+As output, it returns $[[(Ξ, uP, aus1, aus2)]]$, where important for us is
+$[[uP]]$---the pattern and $[[Ξ]]$---the set of `holes' represented by negative algorithmic variables.
+We discuss the anti-unification algorithm in detail in the following section.
 
@@
     \ottdefnAUAUP{}
+
+    \vspace{\baselineskip}
+
     \ottdefnAUAUN{}
-
-    \caption{Anti-unification Algorithm}
+    \caption{Anti-Unification Algorithm}
     \label{fig:anti-unification}
@@
 \begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.3\textwidth}
-        \centering
   \ottdefnATNInf{}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.3\textwidth}
-        \centering
-        \ottdefnATPInf{}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.3\textwidth}
-        \centering
-        \ottdefnATSpinInf{}
-    \end{subfigure}
-    \caption{Algorithmic Type Inferences}
+  \caption{Algorithmic Negative Type Inferences}
   \label{fig:type-inference}
 \end{figure}
+
+\begin{figure}
+  \ottdefnATPInf{}
+  \caption{Algorithmic Positive Type Inferences}
+  \label{fig:type-inference-pos}
+\end{figure}
+
+\begin{figure}
+  \ottdefnATSpinInf{}
+  \caption{Algorithmic Application Type Inferences}
+  \label{fig:type-inference-app}
+\end{figure}

### 2023-10-24 proofs/main.tex
WIP: Algorithm

@@
     are the base cases. They copy the corresponding declarative rules and
-    ensure the reflexivity.
+    ensure reflexivity.
   \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
     subtyping constraints. In this case, we must ensure
-    that the resulting constrains guarantees that the instantiation of 
+    that the resulting constraints guarantee that the instantiation of 
     $[[â⁺]]$ is a supertype of $[[iP]]$.
@@
     Notice that this rule does not have a negative counterpart. This is
-    because one one of the important invariants of the algorithm: 
+    because one of the important invariants of the algorithm: 
     in the negative subtyping, only positive algorithmic variables
@@
     \emph{shift} rules. According to the declarative system,
-    shifted subtyping requires equivalence. In presence of the algorithmic 
+    shifted subtyping requires equivalence. In the presence of the algorithmic 
     variables, it means that the left and the right-hand sides of the subtyping
@@
     the algorithmic variables are added to the subtyping context $[[Θ]]$,
-    after that the recursive call is made. Notice that the declarative context
+    after that, the recursive call is made. Notice that the declarative context
     $[[Γ]]$ is extended by the quantified variables from the right-hand side,
@@
 
-Unification and subtyping algorithms are based on the constraint generation.
-The constraints are represented by set of constraint entries.
+Unification and subtyping algorithms are based on constraint generation.
+The constraints are represented by a set of constraint entries.
 
@@
   \label{definition:merge-subtyping-constraints}
-  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
+  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.\\
   Then $[[Θ ⊢ SC1 & SC2 = SC]]$
-  defines a set such that $[[ucE]] \in [[SC]]$ iff either
-  \begin{itemize}
-    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
-    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
-    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
-      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
-      $[[α̂±]]$. 
+  defines a set of constraints $[[SC]]$ such that $[[scE]] \in [[SC]]$ iff either:
+  \begin{itemize}
+    \item $[[scE]] \in [[SC1]]$ and there is no matching $[[scE']] \in [[SC2]]$; or
+    \item $[[scE]] \in [[SC2]]$ and there is no matching $[[scE']] \in [[SC1]]$; or
+    \item $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ for some $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$
+      such that $[[scE1]]$ and $[[scE2]]$ both restrict variable $[[α̂±]]$. 
   \end{itemize}

### 2023-10-30 oopsla24/tex/04-algorithm.tex
Finish the Algorithmic section

@@
   \begin{multicols}{2}
-    \ottdefnANsub{}
+    \ottdefnANsubLabeled{}
     \columnbreak\\
-    \ottdefnAPsup{}
+    \ottdefnAPsupLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-  \ottdefnUNUnif{}
+  \ottdefnUNUnifLabeled{}
   \columnbreak\\
-  \ottdefnUPUnif{}
+  \ottdefnUPUnifLabeled{}
   \end{multicols}
@@
 \begin{figure}[h]
-  \ottdefnSCME\\
+  \ottdefnSCMELabeled\\
   \caption{Merge of Matching Constraint Entries}
@@
   \begin{multicols}{2}
-    \ottdefnLUBUp{}
+    \ottdefnLUBUpLabeled{}
     \columnbreak\\
-    \ottdefnLUBNsub{}
+    \ottdefnLUBNsubLabeled{}
   \end{multicols}
@@
 
+The anti-unification algorithm \cite{todo}, 
+is a procedure dual to unification
+For two given (potentially different) expressions, 
+it finds the most specific generalizer---the most detailed pattern that 
+matches both of the input expressions. As evidence, it can also return 
+two substitutions that instantiate the `holes' of the pattern to the input expressions.
+
+In our case, we have to be more demanding on the anti-unification algorithm.
+Since we use it to construct an existential type, whose (negative) quantified variables
+can only be instantiated with negative types, we must make sure that the pattern 
+has `holes' only at negative positions. Moreover, we must make sure that
+the resulting substitutions for the `holes' are well-formed in the initial contest, 
+and do not contain variables bound later. For example, 
+the anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and
+$[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$
+must result in a `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, 
+with a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$.
+But it cannot be more specific such as
+$[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole cannot be positive)
+or $[[∀β⁺.γ̂⁻]]$ 
+(since the instantiation cannot capture the bound variable $[[β⁺]]$).
+
+The algorithm that finds the most specific generalizer of two types
+under required conditions is given in \cref{fig:anti-unification}.
+It consists of two mutually recursive procedures:
+the positive and the negative anti-unification. 
+As the positive and the negative anti-unification procedures
+are symmetric in their interface, let us discuss how to read
+the positive judgment.
+
+The positive anti-unification judgment has form
+$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$.
+As an input, it takes a context $[[Γ]]$, in which the `holes'
+instantiations must be well-formed, 
+and two positive types: $[[iP1]]$ and $[[iP2]]$;
+it returns a tuple of four components:
+$[[Ξ]]$---a set of `holes' represented by negative algorithmic variables,
+$[[uQ]]$---a pattern represented as a positive algorithmic type,
+whose algorithmic variables are in $[[Ξ]]$,
+and two substitutions $[[aus1]]$ and $[[aus2]]$
+instantiating the variables from $[[Ξ]]$ such that
+$[[ [aus1]uQ = iP1 ]]$ and $[[ [aus2]uQ = iP2 ]]$.
+
 \begin{figure}[h]
-    \ottdefnAUAUP{}
-
-    \vspace{\baselineskip}
-
-    \ottdefnAUAUN{}
+    \ottdefnAUAUPLabeled{}
+    \hfill\\
+    \hfill\\
+    \ottdefnAUAUNLabeled{}
     \caption{Anti-Unification Algorithm}
@@
 \end{figure}
+
+At the high level, the algorithm scheme follows
+the standard approach \cite{todo} consisting of two principles:
+\begin{enumerate}
+  \item[(i)] if the input terms start with the same constructor,
+    we anti-unify the corresponding parts recursively and 
+    unite the results. This principle is followed by 
+    all the rules except \ruleref{\ottdruleAUAULabel},
+    which works as follows:
+  \item[(ii)] if the first principle does not apply to the input terms $[[iN]]$
+    and $[[iM]]$ (for instance, if they have different outer constructors), the
+    anti-unification algorithm returns a `hole' such that one substitution maps
+    it to $[[iN]]$ and the other maps it to $[[iM]]$. The name of this `hole'
+    should have a name uniquely defined by the pair $([[iN]], [[iM]])$, so
+    that it automatically merges with other `holes' mapped to the same
+    pair of types, and thus, the initiality of the generalizer is ensured. 
+\end{enumerate}
+
+Let us discuss the specific rules of the algorithm in detail.
+
+\paragraph{Variables} 
+  Rules \ruleref{\ottdruleAUPVarLabel} and
+  \ruleref{\ottdruleAUNVarLabel} generalize two equal variables.
+  In this case, the resulting pattern is the variable itself,
+  and no `holes' are needed.
+
+\paragraph{Shifts} 
+  Rules \ruleref{\ottdruleAUShiftDLabel} and
+  \ruleref{\ottdruleAUShiftULabel} operate by congruence:
+  they anti-unify the bodies of the shifts recursively and add
+  the shift constructor back to the resulting pattern.
+
+\paragraph{Quantifiers}
+  Rules \ruleref{\ottdruleAUForallLabel} and
+  \ruleref{\ottdruleAUExistsLabel} are symmetric. 
+  They generalize two quantified types congruently, 
+  similarly to the shift rules. 
+  However, we also require that the quantified variables
+  are fresh, and that the left-hand side variables are 
+  equal to the corresponding variables on the right-hand side.
+  To ensure it, alpha-renaming might be required in the
+  implementation.
+
+  Notice that the context $[[Γ]]$ is \emph{not} extended with 
+  the quantified variables. In this algorithm, $[[Γ]]$ 
+  does not play the role of a current typing context, but rather
+  a snapshot of a context at the moment of calling the anti-unification,
+  \ie the context in which the instantiations of the `holes' 
+  must be well-formed.
+
+\paragraph{Functions}
+  Rule \ruleref{\ottdruleAUArrowLabel} congruently generalizes two function types. 
+  An arrow type is the only binary constructor, 
+  and thus, it is the only rule where the union of the anti-unification results is substantial.
+  The interesting is the case when the resulting generalization of the 
+  input types and the resulting generalization of the output types 
+  have `holes' mapped to the same pair of types. 
+  In this case, the algorithm must merge the `holes' into one.
+  For example, the anti-unification of 
+  $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$
+  must result in $[[↓γ̂⁻ → γ̂⁻]]$,
+  rather than $[[↓γ1̂⁻ → γ2̂⁻]]$.
+
+  In our representation of the anti-unification algorithm, this `merge' happens
+  automatically:
+  following the rule \ruleref{\ottdruleAUAULabel},
+  the name of the `hole' is uniquely defined by the pair of types it is mapped to.  
+  Specifically, when anti-unifying $[[↓α⁻ → α⁻]]$ and $[[↓β⁻ → β⁻]]$ our algorithm returns 
+  $[[↓α̂⁻_{α⁻, β⁻} → α̂⁻_{α⁻, β⁻}]]$, that is a renaming of $[[↓γ̂⁻ → γ̂⁻]]$.
+
+  This way, as the output the rule returns the following tuple:
+  \begin{itemize}
+    \item $[[Ξ1 ∪ Ξ2]]$---a simple union of the sets of `holes' 
+      returned from by the recursive calls, 
+    \item $[[uQ → uM]]$---the resulting pattern 
+      constructed from the patterns returned recursively.
+    \item $[[aus1 ∪ aus1']]$ and $[[aus2 ∪ aus'2]]$
+      --- a union (in a relational sense)
+      of the substitutions returned by the recursive calls. 
+      It is worth noting that the union is well-defined because
+      the result of the substation on a `hole' is determined by the 
+      name of the `hole'.
+  \end{itemize}
+
+\paragraph{The Anti-Unification Rule}
+  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
+  algorithm. If the congruent rules are not applicable, 
+  it means that the input types have a substantially different structure,
+  and thus, the only option is to create a `hole'. 
+  There are three important aspects of this rule that we would like to discuss.
+
+  First, as mentioned earlier, the freshly created `hole' has a name
+  that is uniquely defined by the pair of input types.
+  It is ensured by the following invariant:
+  all the `holes' in the algorithm have name $[[α̂⁻]]$ 
+  indexed by the pair of negative types it is mapped to.
+  This way, the returning set of `holes' is a singleton set
+  $\{ [[ â⁻_{iN, iM} ]]\}$; the resulting 
+  pattern is the `hole' $[[â⁻_{iN, iM}]]$,
+  and the mappings simply send it to the corresponding types:
+  $[[â⁻_{iN, iM} ↦ iN]]$ and $[[â⁻_{iN, iM} ↦ iM]]$.
+
+  Second, this rule is only applicable to negative types, 
+  moreover, the input types are checked to be well-formed 
+  in the outer context $[[Γ]]$. This is required by the 
+  usage of anti-unification: we 
+  call it to build an existential type that would be 
+  an upper bound of two input types via abstracting
+  some of their subexpressions under existential quantifiers.
+  The existentials quantify over \emph{negative} variables,
+  and they must be instantiated in the context
+  available at that moment.
+
+  Third, the rule is only applicable if all other rules fail. 
+  Notice that it could happen even when the input types have matching constructors. 
+  For example, the generalizer of $[[↑α⁺]]$ and
+  $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$),
+  rather than $[[↑γ̂⁺]]$. This way, the algorithm must try to 
+  apply the congruent rules first, and only if they fail,
+  apply \ruleref{\ottdruleAUAULabel}.
+  This principle makes the inference system not syntax-directed:
+  it is not known a priori (before the recursive call) 
+  whether the corresponding congruent rule or rule \ruleref{\ottdruleAUAULabel} 
+  will be applied.
 
@@
 
-        % - Overview (Dependency Graph)
-        % *** The Inference Algorithm
-        % *** Constraint Singularity
+Finally, we present the type inference algorithm. 
+Similarly to the subtyping algorithm, it structurally corresponds to the declarative
+inference specification, meaning that most of the algorithmic rules have
+declarative counterparts, with respect to which they are sound and complete.
+
+This way, the inference algorithm also consists of three mutually recursive procedures: 
+the positive type inference, the negative type inference, and the
+application type inference.
+
+The positive and the negative type inference judgments have symmetric forms:
+$[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ c : iN]]$. Both of these algorithms
+take as an input typing context $[[Γ]]$, a variable context $[[Φ]]$, and 
+a term (a value or a computation) taking its type variables from $[[Γ]]$, 
+and term variables from $[[Φ]]$. As an output, they return a type
+of the given term, which we guarantee to be normalized.
+
+The application type inference judgment has form 
+$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
+As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
+and a constraint context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
+a list of arguments (terms) $[[args]]$ the head is applied to.
+The head may contain algorithmic variables specified by $[[Θ1]]$, 
+in other words, $[[Γ; dom(Θ1) ⊢ uN]]$.
+As a result, the application inference judgment returns 
+$[[uM]]$---a normalized type of the result of the application.
+Type $[[uM]]$ may contain new algorithmic variables, and thus, 
+the judgment also returns $[[Θ2]]$---an updated constraint context
+and $[[SC]]$---a set of subtyping constraints.
+Together $[[Θ2]]$  and $[[SC]]$ specify how the algorithmic variables 
+must be instantiated.
 
 \begin{figure}
-  \ottdefnATNInf{}
-  \caption{Algorithmic Negative Type Inferences}
+  \ottdefnATPInfLabeled{}
+  \hfill \\
+  \ottdefnATNInfLabeled{}
+  \hfill \\
+  \ottdefnATSpinInfLabeled{}
+  \caption{Algorithmic Type Inference}
   \label{fig:type-inference}
@@
 
-\begin{figure}
-  \ottdefnATPInf{}
-  \caption{Algorithmic Positive Type Inferences}
-  \label{fig:type-inference-pos}
-\end{figure}
+The inference rules are shown in
+\cref{fig:type-inference}.
+Next, we discuss them in detail.
 
-\begin{figure}
-  \ottdefnATSpinInf{}
-  \caption{Algorithmic Application Type Inferences}
-  \label{fig:type-inference-app}
-\end{figure}
+\paragraph{Variables}
+  Rule \ruleref{\ottdruleATVarLabel} 
+  infers the type of a positive variable by looking it up in the 
+  term variable context and normalizing the result.
+
+\paragraph{Annotations}
+  Rules \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
+  are symmetric.
+  First, they check that the annotated type is well-formed in the 
+  given context $[[Γ]]$. Then they make a recursive call to infer the 
+  type of annotated expression, check that the inferred type is a subtype of 
+  the annotation, and return the normalized annotation.
+
+\paragraph{Abstractions}
+  Rule \ruleref{\ottdruleATtLamLabel} infers the type of a lambda abstraction.
+  It checks the well-formedness of the annotation $[[iP]]$,
+  makes a recursive call to infer the type of the body in the extended context, 
+  and returns the corresponding arrow type.
+  Since the annotation $[[iP]]$ is allowed to be non-normalized,
+  the rule also normalizes the resulting type.
+
+  Rule \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
+  Similarly to the previous case, it makes a recursive call to infer the type
+  of the body in the extended \emph{type} context. 
+  After that, it returns the corresponding universal type. 
+  It is also required to normalize the result.  
+  For instance, if $[[α⁺]]$ does not occur in the body of the lambda,
+  the corresponding $[[∀]]$ will be removed.
+
+\paragraph{Return and Thunk}
+  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
+  are similar to the declarative rules: they make a recursive call
+  to type the body of the thunk or the return expression and
+  put the shift on top of the result.
+
+\paragraph{Unpack}
+  Rule \ruleref{\ottdruleATUnpackLabel}
+  allows one to unpack an existential type.
+  First, it infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
+  and since the type is guaranteed to be normalized, binds 
+  the quantified variables with $[[nas]]$.
+  Then it infers the type of the body in the appropriately extended context
+  and checks that the inferred type does not depend on $[[nas]]$
+  by checking well-formedness $[[Γ ⊢ iN]]$.
+
+\paragraph{Let Binders}
+  Rule \ruleref{\ottdruleATVarLetLabel}
+  represents the type inference of a standard let binder.
+  It infers the type of the bound value $[[v]]$, and
+  makes a recursive call to infer the type of the body in the extended context.
+
+  Rule \ruleref{\ottdruleATAppLetAnnLabel}
+  infers a type of \emph{annotated} applicative let binder.
+  First, it infers the type of the head of the application,
+  ensuring that it is a \emph{thunked computation} $[[↓iM]]$.
+  After that, it makes a recursive call
+  to the application inference procedure,
+  returning an algorithmic type $[[↑uQ]]$, 
+  that must be instantiated to a subtype of 
+  the annotation $[[↑uP]]$.
+  Then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
+  together with $[[Θ ⊢ SC1 & SC2 = SC]]$
+  check whether the instantiation to the annotated type $[[↑iP]]$ is possible,
+  and if it is, the algorithm infers the type of the body in the extended context,
+  and returns it as the result.
+
+  Rule \ruleref{\ottdruleATAppLetAnnLabel}
+  works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
+  However, since there is no annotation, 
+  instead of checking
+  that the inferred type $[[↑uQ]]$ 
+  can be a subtype of the annotation,
+  the algorithm checks that $[[↑uQ]]$ is unique. 
+  As we prove, uniqueness means that all the algorithmic variables of 
+  $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$.
+  This way, uniqueness is checked by the combination of
+  $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
+  Together, these two premises guarantee that the only 
+  possible instantiation of $[[uQ]]$ is $[[ [uσ]uQ ]]$.
+
+\paragraph{Application to an Empty List of Arguments}
+  Rule \ruleref{\ottdruleATEmptyAppLabel}
+  is the base case of application inference. 
+  If the list of applied arguments is empty, 
+  the inferred type is the type of the head,
+  and the algorithm returns it after normalizing.
+
+\paragraph{Application of a Polymorphic Type $[[∀]]$}
+  Rule \ruleref{\ottdruleATForallAppLabel},
+  analogously to the declarative case,
+  is the rule ensuring the implicit elimination of the universal quantifiers. 
+  This is the place where the algorithmic variables are introduced.
+  The algorithm simply replaces the quantified variables 
+  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
+  and makes a recursive call in the extended context.
+
+  To ensure that this step does not cause infinite recursion, 
+  we also check that the head type has at least one 
+  $[[∀]]$-quantifier. Also, to force the algorithm to
+  apply rule \ruleref{\ottdruleATEmptyAppLabel} when  
+  there are no arguments, we require $[[args ≠ ·]]$.
+
+\paragraph{Application of an Arrow Type}
+  Rule \ruleref{\ottdruleATArrowAppLabel}
+  is the main rule of algorithmic application inference.
+  It is applied when the head has an arrow type $[[uQ → uN]]$.
+  First, it infers the type of the first argument $[[v]]$,
+  and then, calling the algorithmic subtyping,
+  finds $[[SC1]]$---the minimal constraint ensuring that 
+  $[[uQ]]$ is a supertype of the type of $[[v]]$.
+  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
+  and merges the resulting constraint with $[[SC1]]$.

### 2023-10-31 oopsla24/tex/03-declarative-system.tex
WIP: Algorithm
Finish the Algorithmic section
Corrections

@@
   \begin{multicols}{2}
-    \ottdefnDOneNsub{}
+    \ottdefnDOneNsubLabeled{}
 
-    \ottdefnDOnePsup{}
+    \ottdefnDOnePsupLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-    \ottdefnDOneNeq{}
+    \ottdefnDOneNeqLabeled{}
 
-    \ottdefnDOnePeq{}
+    \ottdefnDOnePeqLabeled{}
   \end{multicols}
@@
 \subsubsection{Properties of the Declarative Subtyping}
-A property that we extensively use is that the subtyping is reflexive and transitive,
+\label{sec:decl-subtyping-properties}
+
+A property that is important for the subtyping algorithm, 
+in particular for the type \emph{upgrade} procedure (\cref{todo}),
+is the preservation of free variables by subtyping.
+Informally, it says that the free variables
+of a positive type cannot disappear in its subtypes,
+and the free variables of a negative type
+cannot disappear in its supertypes.
+
+\begin{property}[Subtyping Preserves Free Variables]
+  \label{prop:subtyping-preserves-fv}
+  Let us assume that all the mentioned types are well-formed in $[[Γ]]$. Then
+  $[[Γ ⊢ iN1 ≤ iN2]]$ implies $[[fv(iN1) ⊆ fv(iN2)]]$,
+  and $[[Γ ⊢ iP1 ≥ iP2]]$ implies $[[fv(iP1) ⊆ fv(iP2)]]$.
+\end{property}
+
+Another property that we extensively use is that the subtyping is reflexive and transitive,
 and agrees with substitution.
@@
 \begin{figure}[h]
-  \label{alg:type-nf}
-  \hfill
 
   \begin{multicols}{2}
-  \ottdefnNrmNNorm{}
+  \ottdefnNrmNNormLabeled{}
   \\
@@
   \columnbreak\\
-  \ottdefnNrmPNorm{}
+  \ottdefnNrmPNormLabeled{}
   \\
@@
 \begin{figure}[h]
-
-  \begin{multicols}{2}
-  \ottdefnDTNInf{}
-  \columnbreak\\
-  \ottdefnDTPInf{}
-  \ottdefnDTSpinInf{}
-  \end{multicols}
-  \hfill
-
-  \caption{Declarative Subtyping}
+  \ottdefnsDTLabeled
+  \caption{Declarative Inference}
   \label{fig:declarative-inference}

### 2023-10-31 oopsla24/tex/04-algorithm.tex
Corrections

@@
-        % - Justify Adding α̂± 
-        % - algorithmic type contexts Ξ and constraint context Θ
-        % - Describe how wf, nf,... are changed
-        % - Constraints -- important part of the algorithm
-
-        % - Dependency Graph
-
-In this section, we present the algorithmization of the system described before
-Shadowing the declarative system, the algorithm has two main parts: 
-the subtyping and the type inference, which we discuss in this section one after another.
+In this section, we present the algorithmization of the declarative system described above.
+The algorithmic system follows the structure of the declarative specification closely.
+First, it is also given by a set of inference rules, which, however,
+must be mode-correct (\cite{dunfield2020:bidirectional}), \ie
+the output of each rule is always uniquely defined by its input.
+And second, most of the declarative rules 
+(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
+have a unique algorithmic counterpart, 
+which simplifies reasoning about the algorithm and its correctness proofs.
 
@@
 
-\begin{description}
-        \item[Positive Algorithmic Variables] $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots 
-        \item[Negative Algorithmic Variables] $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots
-        \item[Positive Algorithmic Types] $[[uP]] = \dots \mid [[α̂⁺]]$  
-        \item[Negative Algorithmic Types] $[[uN]] = \dots \mid [[α̂⁻]]$
-        \item[Algorithmic Type Context] $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$
-        where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct 
-\end{description}
+\begin{figure}[h]
+  \begin{multicols}{2}
+      Negative Algorithmic Variables\\
+      $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
+      \columnbreak
+      Positive Algorithmic Variables\\
+      $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
+  \end{multicols}
+  \hfill\\
+  \begin{multicols}{2}
+      \ottuNShort
+      \columnbreak
+      \ottuPShort
+  \end{multicols}
+  \hfill\\
+  Algorithmic Type Context\\
+   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
+  \hfill\\
+  Constraint Type Context\\
+   $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
+  \caption{Algorithmic Syntax}
+  \label{fig:algo-syntax}
+\end{figure}
 
 \paragraph{Algorithmic Variables}
-Both subtyping and the inference algorithms are represented by sets of inference rules, 
-to simplify the soundness and completeness proofs \wrt the declarative specification. 
-The terms these rules manipulate we call \emph{algorithmic}. They extend the 
-previously defined declarative terms and types by adding \emph{algorithmic type variables}
-(\aka unification variables). The algorithmic variables represent unknown types, 
+To design a mode-correct inference system, we slightly modify the language we operate on.
+The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
+They extend the previously defined declarative terms and types by adding 
+\emph{algorithmic type variables} (\aka unification variables). 
+The algorithmic variables represent unknown types, 
 which cannot be inferred immediately but are promised to be instantiated
@@
 We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
-normal variables $[[α]]$, $[[β]]$, \dots and also to smooth out the transition
-from the declarative to the algorithmic system, when we replace the quantified variables $[[pas]]$ 
+normal variables $[[α⁺]]$, $[[β⁻]]$.
+In a few places, we replace the quantified variables $[[pas]]$ 
 with their algorithmic counterpart $[[puas]]$.
@@
 \paragraph{Algorithmic Types}
-The syntax of  algorithmic types is the syntax of declarative types extended
-with algorithmic type variables: we add positive algorithmic variables $[[α̂⁺]]$ 
-to the positive types, and negative algorithmic variables $[[α̂⁻]]$ to the negative types.
-Notice that these variables cannot be abstracted by the quantifiers $[[∀]]$ and $[[∃]]$.
-We denote the algorithmic types
+The syntax of algorithmic types extends the declarative syntax by adding
+algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
+to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
+syntax of negative types. All the constructors of the system can be applied 
+to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
+quantifiers $[[∀]]$ and $[[∃]]$.
 
@@
 Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
-judgment of algorithmic terms defined as expected: in addition to the declarative definition
-\cref{todo}, we also check that each algorithmic variable is in the context $[[Ξ]]$.
+judgment of algorithmic terms defined as expected. 
+Informally, they check that all free declarative variables are in $[[Γ]]$, 
+and all free algorithmic variables are in $[[Ξ]]$.
+In addition to the rules repeating the declarative definition \cref{todo}, 
+we have two base cases for the algorithmic variables:
+ \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf})
+
 \begin{figure}[h]
@@
   $\vcenter{\hbox{\vdots}}$\\
-  \ottusedrule{\ottdruleWFATPUVar{}}
+  \ottusedrule{\ottdruleWFATPUVarLabeled{}}
   \columnbreak\\
   $\vcenter{\hbox{\vdots}}$\\
-  \ottusedrule{\ottdruleWFATNUVar{}}
+  \ottusedrule{\ottdruleWFATNUVarLabeled{}}
 \end{multicols}
@@
 Similarly to well-formedness, the normalization of algorithmic types is defined
-by extending the declarative definition trivially with the algorithmic variables.
+by extending the declarative definition with the algorithmic variables.
+To the rules repeating the declarative normalization, we add
+rules saying that normalization is trivial on algorithmic variables (see \cref{fig:algo-nf}).
 
@@
 The set of constraints is denoted as $[[SC]]$. 
-We assume that the constraints in $[[SC]]$ restrict distinct variables.
+We assume that each algorithmic variable can be restricted by at most one constraint.
 
@@
 This is done to simplify the representation of the algorithm. 
-The unification, which we use as a subroutine of the subtyping algorithm,
+The unification algorithm, which we use as a subroutine of the subtyping algorithm,
 can only produce unification constraints.
@@
 algorithmic variables and declarative contexts.
-
-  \begin{figure}[h]
-    \begin{center}
-      $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$
-    \end{center}
-    \label{fig:constraint-context}
-    \caption{Constraint Contexts}
-  \end{figure}
 
@@
 \ruleref{\ottdruleAExistsLabel} are symmetric. 
-According to the declarative specification, 
-the quantified variables on the left-hand side must be instantiated 
-with types, which, however, are not known in advance.
-We deal with it by algorithmization (\cref{todo})
-of the quantified variables:
-we introduce fresh algorithmic variables
+Declaratively, the quantified variables on the left-hand side must 
+be instantiated with types, which are not known beforehand.
+We address this problem by algorithmization (\cref{todo})
+of the quantified variables.
+The rule introduces fresh algorithmic variables
 $[[puas]]$ or $[[nuas]]$,
-put them into the constraint context $[[Θ]]$
+puts them into the constraint context $[[Θ]]$
 (specifying that they must be instantiated in the extended context
@@
 
-After that, the algorithm proceeds with the recursive call, returning
+After algorithmization of the quantified variables, 
+the algorithm proceeds with the recursive call, returning
 constraints $[[SC]]$. As the output, the algorithm removes the freshly
@@
 If one of the sides of the subtyping is a unification variable,
-the algorithm must create a new restriction. 
+the algorithm must create a new constraint. 
 Because the right-hand side of the subtyping is always declarative,
@@
 
-The merge procedure that we discuss in this section, allows one to 
-combine two constraint sets into one. It might seem that 
-just taking the union of the two sets would be enough, however,
-we require the constraint sets to have a certain structure,
-in particular not to have two entries restricting the same algorithmic
- variable---we call such entries \emph{matching}.
-The matching entries must be combined into one constraint entry, that 
+In this section, we discuss the constraint merging procedure.
+It allows one to combine two constraint sets into one. 
+A simple union of two constraint sets is not sufficient, 
+since the resulting set must not contain two entries restricting the 
+same algorithmic variable---we call such entries \emph{matching}.
+The matching entries must be combined into \emph{one} constraint entry, that 
 would represent their conjunction. 
@@
     as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
-  \item If $[[scE2]]$ is a supertype restriction $[[α̂⁺ :≥ iP]]$,
+  \item If $[[scE2]]$ is a supertype constraint $[[α̂⁺ :≥ iP]]$,
     the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
@@
 
-If both input entries are supertype restrictions: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
+If both input entries are supertype constraints: $[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$,
 then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
@@
 It exploits the idea that the free variables of a positive type $[[iQ]]$
-cannot disappear in its subtypes. It means that if 
+cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
+It means that if 
 a type $[[iP]]$ has free variables not occurring 

### 2023-10-31 oopsla24/tex/01-introduction.tex
fooo

@@
 More recently, \citet{dk19} studed type inference for existential types in the context of GADT type inference, which, while still predicative, supported higher-rank types (i.e., quantifiers can occur anywhere inside of a type scheme).
-\citet{existential-crisis} propose a system which is intermediate between prenex and higher-rank polymorphism, but which
+\citet{existential-crisis} propose a system which only permits types of the form forall-exists, but which permits projective elimination in the style of ML modules.
 
-All of these papers are restricted to \emph{predicative} quantification, whereas existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with
+All of these papers are restricted to \emph{predicative} quantification, where quantifiers can only be instantiated with monotypes (i.e., types without any occurences of quantifiers).
+However, existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with arbitrary types, specifically including types containing quantifiers.
 
+Historically, inference for impredicative quantification has been neglected, due to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that not only is full type inference for System F undecidable, but even that the subtyping relation induced by type instantiation is undecidable.
+However, in recent years, interest in impredicative inference has revived (for example, the work of \citet{serrano-2020}), with a focus on avoiding the undecidabilty barriers by doing \emph{partial} type inference.
+That is, we no longer try to do full type inference, but rather accept that the programmer will need to write annotations in cases where inference would be too difficult.
+Unfortunately, it is often difficult to give a specification for what the partial algorithm does -- for example, \citet{existential-crisis} observe that their algorithm lacks a declarative specification, and explain why existential types make this particularly difficult to do.
+
+One especially well-behaved form of partial type inference is \emph{local type inference}, introduced by \citet{pierce-local-2000}.
+In this approach, quantifiers in a polymorphic function are instantiated using only the type information available in the arguments at each call site.
+While this infers fewer quantifiers than global algorithms such as Damas-Milner can, it has the twin benefits that it is easy to give a mathematical specification to, and that failures of type inference are easily explained in terms of local features of the call site.
+
+In this paper, we extend the local type inference algorithm to a language with both universal and existential quantifiers, which can both be instantiated impredicatively.
+This combination of features broke a number of the invariants which traditional type inference algorithms depend on, and required us to invent new algorithms which combine both unification and (surprisingly) anti-unification.
+
+\paragraph{Contributions} Our contributions are as follows: 
 \begin{itemize}
-\item There has been very little work on type inference for existentials. 
-\item The state of the art in languages like Ocaml and Haskell is the
-  Odersky-L\"{a}ufer algorithm, which makes existentials a
-  second-class construct tied to datatypes.
-\item More recently, there's the existential crisis paper, which
-  showed how to extend ML-style inference with some support for
-  existentials. However, it is (a) predicative-only, (b) restricted
-  for $\forall\exists$ types, and (c) does not have a declarative
-  specification. (In fact, the authors comment on the need for such.)
-\item In this paper, we lift all of these restrictions. We give a
-  local type inference algorithm which supports first-class
-  existentials, for impredicative types, with no restrictions on
-  nested quantifers, and we have declarative specification we prove
-  our algorithm sound and complete with respect to.
-\item Developing the algorithm required us to break some of the
-  fundamental invariants of HM-style typin, and as a result, our
-  algorithm needs to mix unification and anti-unification, and works
-  over a call-by-push-value metalanguage to control how quantifiers
-  are instantiated and things like function arities.
-\item Unfortunately, local type inference does not infer types anywhere
-  nearly as comprehensively as full Damas-Milner type inference. We
-  do guarantee that all type applications (for $\forall$-elimination) 
-  and all packs (for $\exists$-introdution) are inferred, but functions
-  and some let-bindings still need annotations on their arguments.
+\item We give a declarative type system (again, based on call-by-push-value) to serve as a specification of our algorithm, and we we prove our algorithm is sound and complete with respect to the declarative type system.
+The specification makes it easy to see that all type applications (for $\forall$-elimination) and all packs (for $\exists$-introdution) are inferred.
+
+\item We give a local type inference algorithm which supports both first-class existential and universal quantifiers, both of which can be instantiated impredicatively.
+To evade the undecidability results surrounding type inference for System F, we work in a variant of call-by-push-value, which lets us formulate a subtyping relation which is still decidable.
+
+\item Our algorithm breaks some of the fundamental invariants of
+  HM-style type inference. As a result, it needs to mix unification
+  and anti-unification, uses the call-by-push-value structure to
+  control function arities and how quantifiers can be instantiated.
+
 \item The original local type inference paper combined local type inference

### 2023-11-13 figs/subtyping-dependency.tex
Restructure proofs

--- 2023-10-17 proofs/main.tex
@@
-\PassOptionsToPackage{prologue,dvipsnames}{xcolor}
-
-% \UseRawInputEncoding
-% vim: ft=tex
-
-% \documentclass[acmsmall,natbib=false]{article}
-% \usepackage[a4paper, total={8in, 10in}]{geometry}
-
-\documentclass[acmsmall,natbib=false,review,anonymous]{acmart}
-\input{../prelude.tex}
-
-\newcommand{\ottDir}{../ott/_gen}
-\newcommand{\genDir}{_gen}
-\input{\ottDir/defs.tex}
-\input{\ottDir/renew-ott.tex}
-
-\usepackage{../bibliography}
-% \usepackage{amssymb}
-
-\usepackage[matha]{mathabx}
-
-% https://tex.stackexchange.com/questions/85033/colored-symbols/85035#85035
-% \newcommand*{\mathcolor}{}
-% \def\mathcolor#1#{ \mathcoloraux{#1} }
-% \newcommand*{\mathcoloraux}[3]{%
-%   \protect\leavevmode
-%   \begingroup
-%   \color#1{#2}#3%
-%   \endgroup
-% }
-
-\begin{document}
-
-\title{The Proofs}
-
-\maketitle
-
-\tableofcontents
-
-\newpage
-
-\section{Declarative Type Systems}
-
-\subsection{Grammar}
-We assume that there is an infinite set of positive and 
-negative \emph{type} variables. Positive type variables are denoted as 
-$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
-Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
-We assume there is an infinite set of \emph{term} variables,
-which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
-A list of objects (variables, types or terms) is denoted by
-an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
-$[[nbs]]$ is a list of negative type variables, 
-$[[args]]$ is a list of values, which are arguments of a function.
-$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
-in a type $[[iP]]$ and $[[iN]]$, respectively.
-
-\bigskip
-
-% \ottgrammartabular{
-%   \ottiP\ottinterrule
-%   \ottiN\ottinterrule
-% }
-
-\begin{definition}[Declarative Types]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottiN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottiP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Equalities}
-For simplicity, we assume alpha-equivalent terms equal.
-This way, we assume that substitutions do not capture bound variables.
-Besides, we equate
-$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
-and lift these equations transitively and congruently 
-to the whole system.
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Declarative Type Context]
-  \hfill \\
-  Declarative type context $[[Γ]]$ is represented by a set of 
-  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
-  union of two contexts $[[Γ1 ∪ Γ2]]$.
-\end{definition}
-
-$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
-which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
-(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
-
-Notice that checking the well-formedness of a type
-is an \emph{algorithmic} procedure, in which 
-both the context and the type are considered inputs.
-In other words, it is syntax-directed and mode-correct 
-(according to \cite{dunfieldBidirectionalTyping2020}), 
-which means that
-
-\begin{algorithm}[Type Well-formedness]
-  \label{alg:wf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFTNiWF{}
-  \columnbreak
-
-  \ottdefnWFTPiWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-\subsection{Substitutions}
-
-\begin{definition}[Substitution]
-  Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions form variables to types, preserving the polarity. 
-\end{definition}
-
-\begin{algorithm}[Substitution Application]
-  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
-  It is defined naturally as follows:
-    \begin{multicols}{2}
-      \begin{itemize}
-        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$
-        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$
-        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$
-        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$
-        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$
-        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$ 
-        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (assuming the variable capture never happens)
-      \end{itemize}
-    \end{multicols}
-\end{algorithm}
-
-\begin{definition}[Substitution Signature]
-  The signature $[[Γ' ⊢ σ : Γ]]$ means that
-  \begin{enumerate}
-    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
-  \end{enumerate}
-\end{definition}
-
-A substitution can be restricted to a set of variables. 
-The restricted substitution is define as expected. 
-\begin{definition}[Subsitution Restriction]
-  The specification $[[σ  | varset]]$ is defined as
-  a function such that 
-  \begin{enumerate}
-    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
-    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
-  \end{enumerate}
-\end{definition}
-
-Two substitutions can be composed in two ways:
-$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
-while $[[σ2 <=< σ1]]$
-depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
-$[[σ2]]$ to its results on the domain.
-\begin{definition}[Substitution Composition]
-  $[[σ2 ○ σ1]]$ is defined as a function such that
-  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
-\end{definition}
-
-\begin{definition}[Monadic Substitution Composition]
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
-\end{definition}
-Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
-specification of $[[σ1]]$, which is not unique. 
-However, we assume that the used specification clear from the 
-context of the proof.
-
-\begin{definition}[Equivalent Substitutions]
-  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
-  indicates that on the domain $[[Γ]]$, 
-  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
-\end{definition}
-
-Sometimes it is convenient to construct substitution 
-explicitly mapping each variable from a list (or a set)
-to a type. Such substitutions are denoted as $[[iPs / pas]]$
-and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
-the corresponding types.
-\begin{definition}[Explicit Substitution]
-  \hfill
-  \begin{itemize}
-    \item [$-$]
-      Suppose that $[[nas]]$ is a list of negative type variables,
-      and $[[iNs]]$ is a list of negative types of the same length.
-      Then $[[iNs / nas]]$ denotes a substitution such that 
-      \begin{enumerate}
-        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
-        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
-      \end{enumerate}
-    \item [$+$]
-      Positive explicit substitution $[[iPs / pas]]$
-      is defined symmetrically.
-  \end{itemize}
-\end{definition}
-
-\subsection{Declarative Subtyping}
-Subtyping is one of the key mechanism of our system. 
-It realizes the polymorphism: abstract $[[∀]]$ and 
-$[[∃]]$ types can be used where concrete types are expected,
-exactly because the subtyping relation between them.
-
-\begin{definition} 
-  \label{def:subDOne}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNsub{}
-
-    \ottdefnDOnePsup{}
-  \end{multicols}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNeq{}
-
-    \ottdefnDOnePeq{}
-  \end{multicols}
-\end{definition}
-
-The following observations about the declarative subtyping are worth noting:
-\begin{itemize}
-  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
-    make the subtyping reflexive on variables (and further, on any type).
-  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
-    resulting type and contravariant on the argument type.
-  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
-    the subtyping is \emph{invariant} for shifts. 
-    This way, the subtyping of shifted types in one direction implies the subtyping
-    in the opposite direction.
-    Although this rule restricts the
-    subtyping relation, it makes the system decidable.
-  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
-    non-algorithmic rules: the substitution for the quantified variable is
-    not specified, those, these rules `drive' the subtyping relation.
-\end{itemize}
-
-In the next section, we present the sound and complete algorithm
-checking whether one type is a subtype of another according to \cref{def:subDOne}.
-
-\section{Algorithmic Type System}
-
-\subsection{Grammar}
-
-In the algorithmic system, we extend the grammar of types
-by adding positive and negative \emph{algorithmic variables}
-($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
-They represent the unknown types, which will be inferred by the algorithm.
-This way, we add two base cases to the grammar of 
-positive and negative types and use highlight to denote that the type
-can potentially contain algorithmic variables.
-
-\begin{definition}[Algorithmic Types]
-  \label{def:algo-types}
-  \hfill\\
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottuN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottuP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Fresh Variable Selection}
-\label{sec:fresh-selection}
-Both the subtyping and the type inference algorithm
-rely on the ability to select fresh, unused variables.
-For a set of variables $[[varset]]$, it is indicated as 
-$[[varset are fresh]]$ in the inference rules.
-We assume that the selection subroutine always succeeds and is 
-deterministic. In other words, whenever it is called in 
-an algorithmic inference rule, it returns the same result, 
-uniquely determined by the input of this rule.
-
-\subsection{Variable Algorithmization}
-\label{sec:variable-algorithmization}
-
-In several places of our algorithm, in particular, during
-algorithmic subtyping,
-we turn a declarative type into an algorithmic one
-via replacing certain type variables with fresh algorithmic variables.
-We call this procedure \emph{variable algorithmization}, and define it as follows.
-
-\begin{definition}[Variable Algorithmization]
-  Suppose that $[[nas]]$ is a list of negative type variables
-  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
-  in a type for $[[αî⁻ ∊ {nuas}]]$.
-\end{definition}
-
-Conversely, we have the opposite procedure turning algorithmic type variables
-into declarative type variables via \emph{dealgorithmization}.
-
-\begin{definition}[Variable Dealgorithmization]
-  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
-  and $[[nas]]$ is a list of negative type variables of the same length. 
-  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
-  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
-\end{definition}
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
-  \hfill \\
-  Algorithmic type context $[[Ξ]]$ is represented by a set of 
-  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
-  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
-\end{definition}
-
-$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
-that the algorithmic type is well-formed in the contexts
-$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
-of the type is contained in $[[Ξ]]$, and each free declarative type variable
-of the type is contained in $[[Γ]]$.
-
-\begin{algorithm}[Algorithmic Type Well-formedness]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFATNauWF{}
-  \columnbreak
-
-  \ottdefnWFATPauWF{}
-  \end{multicols}
-
-\end{algorithm}
-
-Algorithmic Type Context are used in the unification algorithm.
-In the subtyping algorithm, 
-the context needs to remember additional information.
-In the subtyping context, each algorithmic variable is associated with a
-context it must be instantiated in 
-(i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
-  \hfill \\
-  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
-  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
-  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
-  and $[[Γ]]$ is a context in which they must be instantiated.
-  We assume that no two entries associating the same variable
-  appear in $[[Θ]]$.
-
-  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
-  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
-  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
-\end{definition}
-
-\subsection{Subsitutions}
-
-Substitution that operates on algorithmic type variables is denoted as
-$[[uσ]]$. It is defined as a total function from algorithmic 
-type variables to \emph{declarative} types, preserving the polarity.
-
-The signature $[[Θ ⊢ uσ : Ξ]]$ means that
-$[[Ξ ⊆ dom(Θ)]]$ and 
-$[[uσ]]$ maps each algorithmic variable 
-from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
-and for each variable not appearing in $[[dom(Θ)]]$, 
-it acts as identity.
-
-\begin{definition}[Signature of Algorithmic Substitution]
-  \label{def:algo-subst-sig}
-  \hfill
-  \begin{itemize}
-    \item $[[Θ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$,
-          there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
-          and $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-    \item $[[Γ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-  \end{itemize}
-\end{definition}
-
-In the anti-unification algorithm, we use another kind of substitution.
-In contrast to algorithmic substitution $[[uσ]]$,
-it allows mapping algorithmic variables to
-\emph{algorithmic} types.
-Additionally, anti-unification substitution is restricted to the
-\emph{negative} segment of the language.
-Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a
-
-The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
-in which the results of an anti-unification substitution 
-are formed, is fixed for this substitution.
-This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
-variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.
-
-\begin{definition}[Signature of Anti-unification substitution]
-  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
-  \begin{enumerate}
-    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
-    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
-  \end{enumerate}
-\end{definition}
-
-\subsection{Equivalence and Normalization}
-\label{sec:equivalence-normalization}
-
-The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
-there are types that are subtypes of each other but not equal. 
-For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
-and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
-although these types are not alpha-equivalent.
-For the subtyping algorithm, it is crucial to be able to check whether
-two types are equivalent, without checking mutual subtyping. 
-For this purpose we define the normalization procedure, 
-which allows us to uniformly choose the representative type of the equivalence class.
-This way, the equivalence checking is reduced to normalization and equality checking.
-
-For clarification of the proofs and better understanding of the system, 
-we introduce an intermediate relation---\emph{declarative equivalence}. 
-As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
-this relation is equivalent to the subtyping-induced equivalence, but does not 
-depend on it. Although this relation is not defined algorithmically, 
-it gives the intuition of what types our system considers equivalent.
-Specifically, in addition to \emph{alpha-equivalence}, 
-our system allows for \emph{reordering of adjacent quantifiers},
-and \emph{introduction/elimination of unused quantifiers}.
-
-The non-trivial rules of the declarative equivalence are
-\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
-Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
-the recursive call on the body of the quantified type. 
-It will be covered formally in \cref{sec:decl-equiv-lemmas}.
-
-\begin{definition}[Declarative Type Equivalence]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnEOneNeq{}
-  \columnbreak\\
-  \ottdefnEOnePeq{}
-  \end{multicols}
-
-\end{definition}
-
-As the equivalence includes arbitrary reordering of quantified variables,
-the normalization procedure is needed to choose the canonical order.
-For this purpose, we introduce an auxiliary procedure---variable ordering. 
-Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
-in the order they appear in $[[iN]]$.
-
-\begin{algorithm}[Variable Ordering]
-  \label{alg:var-ordering}
-  \hfill
-
-  \ottdefnONVar{}
-  \ottdefnOPVar{}
-
-  Analogously, the variable can be ordered in 
-  an \emph{algorithmic} type ($[[ord varset in uP]]$ and 
-  $[[ord varset in uN]]$). In these cases, we treat the algorithmic variables
-  as if they were declarative variables.
-
-\end{algorithm}
-
-Next, we use the variable ordering in the normalization procedure. 
-Specifically, normalization recursively traverses the type, 
-and for each quantified case reorders the quantified variables in a 
-canonical order dictated by \cref{alg:var-ordering}, removing unused ones.
-
-\begin{algorithm}[Type Normalization]
-  \label{alg:type-nf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnNrmNNorm{}
-  \columnbreak\\
-  \ottdefnNrmPNorm{}
-  \end{multicols}
-
-  Analogously, we define the normalization of algorithmic types by adding base cases:
-
-  \begin{multicols}{2}
-  \ottdefnNrmuNNorm{}
-  \columnbreak\\
-  \ottdefnNrmuPNorm{}
-  \end{multicols}
-
-\end{algorithm}
-
-\Cref{lemma:subt-equiv-algorithmization}
-demonstrates that the equivalence of types is the same
-as the equality of their normal forms.
-\begin{theorempreview}[Correctness of Normalization]
-  Assuming the types are well-formed in $[[Γ]]$, 
-  \begin{itemize}
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]]$ if and only if $[[nf(iN) = nf(iM)]]$;
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]]$ if and only if $[[nf(iP) = nf(iQ)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\begin{algorithm}[Substitution Normalization]
-  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
-  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
-\end{algorithm}
-
-The rest of this chapter is devoted to the
-central algorithm of the type system---the subtyping algorithm. 
-\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
-The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
-$A$ uses $B$ as a sub-procedure.
-
-\begin{figure}[h]
-  \centering
   \begin{tikzpicture}
@@
   \end{tikzpicture}
-  \caption{Dependency graph of the subtyping algorithm}
-  \label{fig:subtyping-algo}
-\end{figure}
-
-\subsection{Subtyping}
-\label{sec:subtyping}
-
-Now, we present the subtyping algorithm itself.
-Although the algorithm is presented as a single procedure,
-is important for the structure of the proof that the positive subtyping algorithm
-does not invoke the negative one. This way, the correctness of the positive 
-subtyping will be proved independently and used afterwards to prove the
-correctness of the negative subtyping.
-
-\begin{algorithm}[Subtyping]
-  \label{alg:subtyping}
-  \hfill\\
-  \ottdefnANsub{}
-  \ottdefnAPsup{}
-\end{algorithm}
-
-The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
-the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
-must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
-and $[[uP]]$ and $[[iQ]]$ for the positive case. 
-As one of the invariants, we require
-$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
-The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
-which will be discussed in the next section.
-
-Let us overview the inference rules of the subtyping algorithm.
-\begin{itemize}
-  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
-    are the base cases. They copy the corresponding declarative rules and
-    ensure the reflexivity.
-  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
-    subtyping constraints. In this case, we must ensure
-    that the resulting constrains guarantees that the instantiation of 
-    $[[â⁺]]$ is a supertype of $[[iP]]$.
-    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
-    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
-    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
-    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
-
-    Notice that this rule does not have a negative counterpart. This is
-    because one one of the important invariants of the algorithm: 
-    in the negative subtyping, only positive algorithmic variables
-    can occur in the types.
-
-  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
-    \emph{shift} rules. According to the declarative system,
-    shifted subtyping requires equivalence. In presence of the algorithmic 
-    variables, it means that the left and the right-hand sides of the subtyping
-    must be unified. Hence, the shift rules invoke the unification algorithm, 
-    which will be discussed in \cref{sec:unification}. The unification 
-    returns the minimal set of constraints $[[UC]]$, which is necessary
-    and sufficient for the subtyping.
-
-  \item \ruleref{\ottdruleAArrowLabel}.
-    In this case, the algorithm makes two calls:
-    a recursive call to the negative subtyping algorithm for the argument types,
-    and a call to the positive subtyping algorithm for the result types.
-    After that, the resulting constraints are merged using the
-    \emph{subtyping constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
-    are symmetric. These are the only places where 
-    the algorithmic variables are introduced.
-    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
-    of the quantified variables: these variables are replaced by 
-    fresh algorithmic variables in the body of the quantified type,
-    the algorithmic variables are added to the subtyping context $[[Θ]]$,
-    after that the recursive call is made. Notice that the declarative context
-    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
-    which matches the declarative system.
-\end{itemize}
-
-Then soundness lemma (\cref{lemma:pos-subt-soundness,lemma:neg-subt-soundness}) 
-and completeness (\cref{lemma:pos-subt-completeness,lemma:neg-subt-completeness})
-of the algorithm together give us the following simplified theorem:
-
-\begin{theorempreview}[Correctness of subtyping algorithm]
-  \hfill
-  \begin{itemize}
-    \item [$-$]  $[[ Γ ; · ⊨ uN ≤ iM ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iN ≤ iM ]]$;
-    \item [$+$] $[[ Γ ; · ⊨ uP ≥ iQ ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iP ≥ iQ ]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraints}
-
-Unification and subtyping algorithms are based on the constraint generation.
-The constraints are represented by set of constraint entries.
-
-\begin{definition}[Unification Constraint]
-  \hfill
-  \begin{description}
-    \item[unification entry] (denoted as $[[ucE]]$) is an expression of shape 
-      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
-    \item[unification constraint] (denoted as $[[UC]]$) is a set of 
-      unification constraint entries.
-      We denote $\{[[α̂±]] \mid [[ucE ∊ UC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(UC)]]$.
-  \end{description}
-\end{definition}
-
-However, in the subtyping, we need to consider more general
-kind of constraints. Specifically,
-subtyping constraint entries can restrict a variable
-not only to be equivalent to a certain type, but
-also to be a supertype of a positive type.
-
-\begin{definition}[Subtyping Constraint]
-  \hfill
-  \begin{description}
-    \item[subtyping entry] (denoted as $[[scE]]$) is an expression of shape 
-      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
-    \item[subtyping constraint] (denoted as $[[SC]]$) is a set of subtyping constraint entries.
-      We denote $\{[[α̂±]] \mid [[scE ∊ SC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(SC)]]$.
-  \end{description}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint Entry]
-  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
-  its associated type is well-formed in $[[Γ]]$.
-  \begin{itemize}
-    \item[] $[[Γ ⊢ pua :≥ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ pua :≈ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ nua :≈ iN]]$ ~iff~ $[[Γ ⊢ iN]]$.
-  \end{itemize}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  subtyping context $[[Θ]]$ if all its entries are well-formed in
-  the corresponding elements of $[[Θ]]$.
-  More formally, 
-  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
-  such that $[[scE]]$ restricts $[[α̂±]]$,
-  we have $[[Θ(α̂±) ⊢ scE]]$.
-
-  We write $[[Θ ⊢ SC : Ξ]]$ to denote
-  that $[[Θ ⊢ SC]]$ and $[[dom(SC) = Ξ]]$.
-
-  $[[Θ ⊢ UC]]$ and $[[Θ ⊢ UC : Ξ]]$ are defined analogously.
-\end{definition}
-
-\subsubsection{Constraint Satisfaction}
-
-A constraint entry restricts a type that can be assigned to a variable.
-We say that a type satisfies a constraint entry if it can be assigned
-to the variable restricted by the entry.
-
-\begin{definition}[Type Satisfying a Constraint Entry]
-  \hfill\\
-  \begin{multicols}{2}
-  \ottdefnSATSCEN{}
-  \columnbreak\\
-  \ottdefnSATSCEP{}
-  \end{multicols}
-\end{definition}
-
-We say that a substitution satisfies a constraint---a set of constraint 
-entries if each entry is satisfied by the type assigned to the variable
-by the substitution.
-
-\begin{definition}[Substitution Satisfying a Constraint]
-  We write $[[Θ ⊢ uσ : SC]]$ to denote that
-  a substitution $[[uσ]]$ satisfies a constraint $[[SC]]$ in a context $[[Θ]]$.
-  It presumes that $[[Θ ⊢ SC]]$ and 
-   means that for any $[[ucE]] \in [[SC]]$, if $[[ucE]]$ restricts $[[α̂±]]$,
-  then $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$.
-
-  Unification constraint satisfaction $[[Θ ⊢ uσ : UC]]$ 
-  is defined analogously as a special case of subtyping constraint satisfaction.
-
-\end{definition}
-
-Notice that $[[Θ ⊢ uσ : SC]]$ does not 
-imply the signature $[[Θ ⊢ uσ : dom(SC)]]$, because 
-the latter also specifies $[[uσ]]$ outside of the domain $[[dom(SC)]]$
-(see \cref{def:algo-subst-sig}).
-
-\subsubsection{Constraint Merge}
-\label{sec:constraint-merge}
-
-In this section, define the least upper bound 
-for constraints, which we call \emph{merge}.
-Intuitively, the merge of two constraints is the least
-constraint such that any substitution satisfying both constraints
-satisfies the merge as well.
-First, we define the merge of entries,
-and then extend it to the set of entries.
-
-\begin{definition} [Matching Entries]
-  We call two unification constraint entries 
-  or two subtyping constraint entries matching 
-  if they are restricting the same unification variable.
-\end{definition}
-
-Two matching entries formed in the same context $[[Γ]]$ 
-can be merged in the following way:
-\begin{algorithm}[Merge of Matching Constraint Entries]
-  \label{definition:merge-matching-entries}
-   \hfill
-
-  \ottdefnSCME\\
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}
-    are symmetric cases. To merge two matching entries restricting
-    a variable to be equivalent to certain types, we check
-    that these types are equivalent to each other.
-    To do so, it suffices to check for \emph{equality} of their normal forms,
-    as discussed in \cref{sec:equivalence-normalization}. 
-    After that, we return the left-hand entry.
-
-  \item \ruleref{\ottdruleSCMEEqSupLabel} and \ruleref{\ottdruleSCMESupEqLabel}
-    are also symmetric. 
-    In this case,
-    since one of the entries requires the variable to be equal to 
-    a type, the resulting entry must also imply that.
-    However, for the soundness, it is needed to ensure that
-    the equating restriction is stronger than the subtyping restriction.
-    For this purpose, the premise invokes the positive subtyping.
-
-  \item \ruleref{\ottdruleSCMESupSupLabel} 
-    In this case, we find the least upper bound of the types from the input
-    restrictions, 
-    and as the output, restrict the variable to be a supertype of the result.
-    The least upper bound procedure will be discussed in \cref{sec:lub}.
-\end{itemize}
-
-Unification constraint entries are a special case of subtyping constraint
-entries. They are merged using the same algorithm 
-(\cref{definition:merge-matching-entries}).
-Notice that the merge of two matching unification constraint entries
-is a unification constraint entry.
-\begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
-  \label{lemma:merge-matching-entries-welldef}
-  Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
-  are unification constraint entries. 
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  $[[Γ ⊢ ucE1 & ucE2 = ucE]]$
-  according to \cref{definition:merge-matching-entries},
-  is a unification constraint entry.
-\end{lemma}
-\begin{proof}
-  Since $[[ucE1]]$ and $[[ucE2]]$ are matching unification constraint entries,
-  they have the shape $([[pua :≈ iP1]], [[pua :≈ iP2]])$ or
-  $([[nua :≈ iN1]], [[nua :≈ iN2]])$.
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  can only be defined by \ruleref{\ottdruleSCMEPEqEqLabel} or
-  \ruleref{\ottdruleSCMENEqEqLabel}.
-  In both cases the result, if it exists, 
-  is a unification constraint entry:
-  in the first case, the result has shape $[[pua :≈ iP1]]$,
-  in the second case, the result has shape $[[nua :≈ iN1]]$.
-\end{proof}
-
-% Notice that in case of equivalence, the assigned types
-% must be equal (i.e. alpha-equivalent) to be merged. This is because
-% the unification algorithm assumes that every type is normalized,
-% and hence, equivalence is alpha-equivalence 
-% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
-
-\begin{algorithm}[Merge of Subtyping Constraints]
-  \label{definition:merge-subtyping-constraints}
-  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.
-  Then $[[Θ ⊢ SC1 & SC2 = SC]]$
-  defines a set such that $[[ucE]] \in [[SC]]$ iff either
-  \begin{itemize}
-    \item $[[ucE]] \in [[SC1]]$ and there is no matching $[[ucE']] \in [[SC2]]$; or
-    \item $[[ucE]] \in [[SC2]]$ and there is no matching $[[ucE']] \in [[SC1]]$; or
-    \item $[[Θ(α̂±) ⊢ ucE1 & ucE2 = ucE]]$ for some $[[ucE1]] \in [[SC1]]$ and $[[ucE2]] \in [[SC2]]$
-      such that $[[ucE1]]$ matches with $[[ucE2]]$ restricting variable
-      $[[α̂±]]$. 
-  \end{itemize}
-\end{algorithm}
-
-Unification constraints can be considered 
-as a special case of subtyping constraints,
-and the merge of unification constraints
-is defined as the merge of subtyping constraints.
-Then it is easy to see that the merge of two 
-unification constraints is a unification constraint.
-
-\begin{lemma}[Merge of Unification Constraints is well-defined]
-  Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
-  are unification constraints. 
-  Then the merge of $[[UC1]]$ and $[[UC2]]$ 
-  $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$
-  according to \cref{definition:merge-subtyping-constraints},
-  is a unification constraint.
-\end{lemma}
-\begin{proof}
-  $[[UC]]$ consists of unmatched entries of $[[UC1]]$ and $[[UC2]]$,
-  which are \emph{unification} constraint entries by assumption,
-  and merge of matching entries, which also are  
-  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-welldef}.
-\end{proof}
-
-\Cref{lemma:merge-soundness,lemma:merge-completeness} 
-show the correctness and initiality of the merge operation,
-which can be expressed in the following simplified theorem:
-\begin{theorempreview}[Correctness of Constraint Merge]
-  A substitution $[[uσ]]$ satisfying both constraints
-  $[[SC1]]$ and $[[SC2]]$ 
-  if and only if it satisfies their merge.
-\end{theorempreview}
-
-The unification constraint merge satisfies the same theorem,
-however, because the merge of unification constraint entries 
-$[[ucE1]]$ and $[[ucE2]]$ always results in one of them, 
-a stronger soundness property holds (see \cref{lemma:unif-merge-soundness}):
-\begin{theorempreview}[Soundness of Unification Constraint Merge]
-  If $[[Θ ⊢ UC1 & UC2 = UC]]$ then $[[UC = UC1 ∪ UC2]]$.
-\end{theorempreview}
-
-\subsection{Unification}
-\label{sec:unification}
-
-The subtyping algorithm calls the following subtask:
-given two algorithmic types, we need to find the most general substitution 
-for the algorithmic variables in these types, such that the resulting 
-types are equivalent. This problem is known as \emph{unification}.
-
-In our case, the unification is restricted in the following way:
-first, before unifying the types, we normalize them, which 
-allows us to reduce (non-trivial) equivalence to (trivial) equality;
-second, we preserve invariants which guarantee that
-one side of the unification is always declarative, which in fact, 
-reduces the unification to the \emph{matching} problem.
-
-The unification procedure
-returns a set of minimal constraints,
-that must be satisfied by a substitution
-unifying the input types.
-
-\begin{algorithm}[Unification]
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnUNUnif{}
-  \columnbreak\\
-  \ottdefnUPUnif{}
-  \end{multicols}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
-    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
-    adding the quantified variables to the context $[[Γ]]$.
-    Notice that $[[Θ]]$, which specifies
-    the contexts in which the algorithmic variables must be instantiated,
-    is not changed.
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
-    are the base cases. 
-    Since the sides are equal and free from algorithmic variables,
-    the unification returns an empty constraint. 
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
-    are symmetric cases constructing the constraints. 
-    When an algorithmic variable is unified with a type, 
-    we must check that the type is well-formed in the required context,
-    and if it is, we return a constraint restricting the variable
-    to be equivalent to that type.
-  \item \ruleref{\ottdruleUArrowLabel}.
-    In this case, the algorithm makes two recursive calls:
-    it unifies the arguments and the results of the arrows.
-    After that, the resulting constraints are merged using the
-    \emph{unification constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
-    \emph{unification} constraints, not arbitrary \emph{subtyping} 
-    constraints: it is important for modularizing the proofs, 
-    since the properties of the \emph{unification} constraint merge
-    can be proved independently from the \emph{subtyping} constraint merge.
-\end{itemize}
-
-\subsection{Least Upper Bound}
-\label{sec:lub}
-
-In this section, we present
-the algorithm finding the least common supertype of two positive types. 
-It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
-and indirectly, through the type upgrade by positive subtyping
-(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
-intuitive part of the algorithm, and its correctness will be covered
-in \cref{sec:alg-upper-bounds-proofs}.
-
-\begin{algorithm}[The Least Upper Bound Algorithm]
-  \hfill\\
-  \ottdefnLUBNsub{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleLUBVarLabel}
-    The base case is trivial: 
-    the least upper bound of to equal variables
-    is the variable itself.
-  \item \ruleref{\ottdruleLUBShiftLabel}
-    In case both sides of the least upper bound are shifted,
-    the algorithm needs to find the anti-unifier of them. 
-    Intuitively, this is because in general, the upper bounds of
-    $[[↓iN]]$ are $[[∃nas.iP]]$ such that 
-    $[[nas]]$ can be instantiated with some $[[iMs]]$ so that
-    $[[ Γ ⊢ [iMs/nas]iP ≈ ↓iN ]]$ (see \cref{lemma:shape-of-supertypes}).
-  \item \ruleref{\ottdruleLUBExistsLabel}
-    In this case, we move the quantified variables to the context $[[Γ]]$, 
-    and make a recursive call. 
-    It is important to make sure that $[[nas]]$ and $[[nbs]]$ are disjoint.
-    In this case, it is guaranteed that the resulting 
-    $[[fv(iQ)]]$ will be free of $[[nas]]$ and $[[nbs]]$,
-    and thus, the resulting type will be a supertype of both sides
-    (it will be discussed in \cref{lemma:shape-of-supertypes}).
-\end{itemize}
-
-In the positive subtyping algorithm (\cref{sec:subtyping}),
-\ruleref{\ottdruleAPUVarLabel} 
-generates a restriction of a variable $[[α̂⁺]]$.
-On the one hand, this restriction must imply 
-$[[â⁺ :≥ iP]]$ for the subtyping to hold.
-On the other hand, the type used in this restriction 
-must be well-formed in a potentially stronger (smaller) 
-context than $[[iP]]$.
-
-To resolve this problem, we define the \emph{upgrade} procedure,
-which for given $[[Δ]]$, $[[pnas]]$, and $[[Δ, pnas ⊢ iP]]$,
-finds $[[Δ ⊢ iQ]]$---the least supertype of $[[iP]]$ 
-among the types well-formed in $[[Δ]]$.
-
-The trick is to make sure that the `forbidden' variables
-$[[pnas]]$ are not used explicitly in the supertypes
-of $[[iP]]$. For this purpose, we
-construct new types $[[iP1]]$ and $[[iP2]]$,
-in each of them replacing the forbidden variables
-with fresh variables $[[pnbs]]$ and $[[pncs]]$,
-and then find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
-It turns out that this renaming forces the common types of 
-$[[iP1]]$ and $[[iP2]]$ to be agnostic to $[[pnas]]$,
-and thus, the supertypes of $[[iP]]$ well-formed in $[[Δ]]$
-are exactly the common supertypes of $[[iP1]]$ and $[[iP2]]$.
-These properties are considered in more details in \cref{sec:upgrade-lemmas}.
-
-\begin{algorithm}[Type Upgrade]
-  \hfill\\
-  \ottdefnLUBUp{}
-\end{algorithm}
-
-\begin{paragraph}{Note on the Greatest Lower Bound}
-  In contrast to the least upper bound, 
-  the general greatest lower bound does not exist in our system.
-  For instance, consider a positive type $[[iP]]$, 
-  together with its non-equivalent
-  supertypes $[[iP1]]$ and $[[iP2]] \not\simeq [[iP1]]$
-  (for example, $[[iP = ↓↑↓γ⁻]]$, $[[iP1 = ∃α⁻.↓↑↓α⁻]]$, 
-  and $[[iP2 = ∃α⁻.↓α⁻]]$).
-  Then for arbitrary $[[iQ]]$ and $[[iN]]$, 
-  let us consider the common subtypes of 
-  $A = [[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $B = [[iP → ↓↑iP1 → ↓↑iP2 → iN]]$.
-  It is easy to see that $[[∀α⁺.∀β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and 
-  $[[∀α⁺.∀β⁺. α⁺ → ↓↑β⁺ → ↓↑α⁺ → iN]]$ are
-  both \emph{maximal} common subtypes of $A$ and $B$,
-  and since they are not equivalent, none of them is 
-  the \emph{greatest} one.
-
-  However, we designed the subtyping system in such a way 
-  that the greatest lower bound is not needed:
-  the negative variables are always `protected'
-  by \emph{invariant} shifts ($[[↑]]$ and $[[↓]]$), 
-  and thus, the algorithm can only require
-  a substitution of a negative variable to be 
-  \emph{equivalent} to some type but never 
-  to be a \emph{subtype}.
-\end{paragraph}
-
-\subsection{Anti-unification}
-\label{sec:antiunification}
-
-Next, we define the anti-unification procedure,
-also known as the \emph{most specific generalization}.
-As an input, it takes two declarative types
-(e.g., in the positive case $[[iP1]]$ and $[[iP2]]$)
-and a context $[[Γ]]$.
-and returns a type $[[uQ]]$---the generalizer,
-containing negative placeholders 
-(represented by algorithmic variables) from 
-$[[Ξ]]$ and two substitutions $[[uτ1]]$ and $[[uτ2]]$.
-The substitutions replace the placeholders with 
-declarative types well-formed in $[[Γ]]$,
-such that $[[ [uτ1]uQ = iP1 ]]$ and $[[ [uτ2]uQ = iP2 ]]$.
-Moreover, the algorithm guarantees that 
-$[[uQ]]$ is the most specific type with this property: 
-any other generalizer can be turned into $[[uQ]]$ by some substitution
-$[[uρ]]$.
-
-It is important to note the differences between 
-the standard anti-unification and our version.
-First, we only allow the placeholders at \emph{negative} positions,
-which means, for example, that $[[α⁺]]$ and $[[β⁺]]$ cannot be
-generalized. Second, the generated pair of substitutions 
-$[[uτ1]]$ and $[[uτ2]]$ must replace the placeholders with 
-types well-formed in a specified context $[[Γ]]$.
-
-The anti-unification algorithm assumes that the input types 
-are normalized. This way, anti-unification up-to-equality rather than 
- anti-unification up-to-equivalence is sufficient.
-
-\begin{algorithm}[Anti-unification]
-  \hfill
-
-  \ottdefnAUAUP{}
-  \ottdefnAUAUN{}
-
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleAUPVarLabel} and \ruleref{\ottdruleAUNVarLabel}
-    are the base cases. 
-    In this case, since the input types are equal, 
-    the algorithm returns this type as a generalizer,
-    without generating any placeholders.
-
-  \item \ruleref{\ottdruleAUShiftDLabel}, \ruleref{\ottdruleAUShiftULabel},
-    \ruleref{\ottdruleAUForallLabel}, and \ruleref{\ottdruleAUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers.
-    Notice that the algorithm does not add the removed variables to
-    the context $[[Γ]]$. This is because $[[Γ]]$
-    is used to restrict the resulting anti-unification substitutions, 
-    and is fixed throughout the algorithm.
-
-  \item \ruleref{\ottdruleAUAULabel} is the most important rule, 
-    since it generates the placeholders. 
-    This rule only applies if other negative rules failed.
-    Because of that, the anti-unification procedure is 
-    \emph{not} syntax-directed.
-
-    The generated placeholder is indexed with a pair of 
-    types it is mapped to. It allows the algorithm to 
-    automatically unite the anti-unification solutions 
-    generated by the different branches of 
-    \ruleref{\ottdruleAUArrowLabel}.
-
-    Notice that this rule does not have a positive counterpart,
-    since we only allow negative placeholders.
-
-  \item \ruleref{\ottdruleAUArrowLabel}
-    makes two recursive calls to the anti-unification procedure,
-    and unites the results. Suppose that
-    $[[uτ1]]$  and $[[uτ2]]$ are the substitutions generated by
-    anti-unification of \emph{argument} types of the arrow,
-    and $[[uτ1']]$ and $[[uτ2']]$ are the substitutions generated by 
-    anti-unification of \emph{result} types of the arrow.
-    It is important that if ($[[uτ1]]$, $[[uτ2]]$)
-    and ($[[uτ1']]$, $[[uτ2']]$) send some variables to the same pair of types,
-    i.e., $[[ [uτ1]α̂⁻ = [uτ1']β̂⁻]]$ and $[[ [uτ2]α̂⁻ = [uτ2']β̂⁻]]$,
-    then these variables are equal, i.e., $[[α̂⁻ = β̂⁻]]$.
-    This property is guaranteed by \ruleref{\ottdruleAUAULabel}:
-    the name of the placeholder is determined by the pair of 
-    types it is mapped to.
-\end{itemize}
-
-\newpage
-
-\section{Declarative Typing}
-
-In the previous section, we presented the 
-type system together with subtyping specification 
-and the algorithm. In this section, 
-we describe the language under this type system, 
-together with the type inference specification
-and algorithm.
-
-\subsection{Grammar}
-
-First, we define the syntax of the language.
-The language combines System F with call-by-push-value 
-style. 
-\begin{definition}[Language Grammar]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottc\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottv\ottinterrule
-    }
-  \end{multicols}
-\end{definition}
-Notice that the language does not have
-first-class applications: instead, 
-we use applicative let bindings---
-constructions that bind a result of
-a fully applied function to a (positive) 
-variable.
-In the call-by-push-value paradigm, 
-it corresponds to monadic bind or do-notation. 
-Typewise, these let-binders come in two forms:
-annotated and unannotated. The annotated let-binders
-$[[let x:iP = v(args); c]]$ 
-requires the application to infer the annotated
-$[[iP]]$, whereas the unannotated 
-$[[let x = v(args); c]]$ 
-is used when the inferred type is unique.
-
-A computation of a polymorphic type is constructed 
-using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
-is implicit. Conversely, the existential types
-are constructed implicitly and eliminated 
-using the standard unpack mechanism:
-$[[let∃ (nas, x) = v; c]]$.
-
-Another dual pair of constructions are 
-$[[return v]]$ and $[[{c}]]$. 
-The former allows us to embed a value in pure 
-computations. The latter, on the contrary,
-encapsulates a thunk of computation in a value.
-
-Finally, the language has several standard constructions:
-lambda-abstractions $[[λx:iP.c]]$,
-standard let-bindings $[[let x = v; c]]$,
-and type annotations that can be added to any value or computation:
-$[[(v:iP)]]$ and $[[(c:iN)]]$.
-
-\subsection{Declarative Type Inference}
-
-Next, we define the specification of the type 
-inference for our language. First, we introduce 
-variable context specifying the types of variables 
-in the scope of the current rule.
-
-\begin{definition}[Variable Context]
-  The variable typing context $[[Φ]]$
-  is represented by a set of entries of the form
-  $[[x : iP]]$. 
-\end{definition}
-
-The specification is represented by an inference system of
-three mutually recursive judgments:
-positive inference $[[Γ ; Φ ⊢ v : iP]]$,
-negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
-application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
-In the premises, the inference rules also refer to 
-the declarative subtyping (\cref{def:subDOne}),
-type well-formedness (\cref{alg:wf}), 
-and normalization (\cref{alg:type-nf}).
-\begin{itemize}
-  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
-    means that under the type context $[[Γ]]$ and
-    the variable context $[[Φ]]$, for the value $[[v]]$,
-    type $[[iP]]$ is inferrable. It guarantees that 
-    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
-    the standard sense.
-  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
-    judgment. It means that if a head of type $[[iN]]$ 
-    is applied to list of values $[[args]]$, 
-    then the resulting computation can be typed as $[[iM]]$.
-\end{itemize}
-
-\begin{definition}[Declarative Type Inference]
-  \label{def:declarative-typing}
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnDTNInf{}
-  \ottdefnDTPInf{}
-  \ottdefnDTSpinInf{}
-  \end{multicols}
-\end{definition}
-
-Let us discuss selected rules of the declarative system:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTVarLabel}
-    says that the type of a variable is inferred from the context.
-  \item \ruleref{\ottdruleDTThunkLabel} says
-    that the type of a thunk is inferred by shifting up the type of the 
-    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
-    infers the type of a return by shifting down the type of the
-    contained value.
-  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
-    They allow the inferred type to be refined by annotating it with a supertype.
-  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
-    mean that th declarative system allows to infer any type from the equivalence class.
-  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
-    and its first premise inferring the existential type of the value being unpacked.
-    It is important however that the inferred existential type is normalized. 
-    This is because there might be multiple equivalent existential types 
-    with different order or even number of quantified variables, 
-    and to bind them, the algorithm needs to fix the canonical one.
-  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
-    with annotated let-bindings. The first premise infers the type of the head of the application,
-    which must be a thunked computation. Then if after applying it 
-    to the arguments, the resulting type can be equated to the annotated one,
-    we infer the body of the let-binding in the context extended with the bound variable.
-  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
-    but it is used when the type of the application is unique, and thus, the annotation
-    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
-    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
-\end{itemize}
-
-Let us discuss the rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTEmptyAppLabel} 
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head. 
-    However, we relax this specification by allowing it to 
-    infer any other equivalent type. 
-    The relaxation of this rule is enough to guarantee 
-    this property for the whole judgement:
-    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
-    $[[iM']]$.
-  \item \ruleref{\ottdruleDTArrowAppLabel}
-    is where the application type is inferred: 
-    if the head has an arrow type $[[iQ → iN]]$,
-     we are allowed to apply it as soon as 
-    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
-  \item \ruleref{\ottdruleDTForallAppLabel}
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    If we are applying a polymorphic computation, 
-    we can instantiate its quantified variables with any types,
-    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
-\end{itemize}
-
-\section{Algorithmic Typing}
-
-Next, we present the type inference algorithm, 
-which is sound and complete with respect to the declarative specification
-(\cref{def:declarative-typing}).
-
-\subsection{Algorithmic Type Inference}
-
-Mirroring the declarative typing, 
-the algorithm is represented by an inference system of three mutually recursive
-judgments:
-\begin{itemize}
-  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
-    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
-    In contrast with the declarative counterparts, they are deterministic,
-    and guarantee that the inferred type is normalized. 
-  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
-    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
-    Notice that $[[uN]]$ contains algorithmic variables, 
-    which are specified by the context $[[Θ1]]$.
-    Moreover, the inferred type $[[uM]]$ is also algorithmic,
-    and can have several non-equivalent instantiations. To accommodate that, 
-    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
-    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
-    in which the variables must be instantiated, and $[[SC]]$
-    imposes restrictions on the variables. 
-\end{itemize}
-As subroutines, the algorithm calls
-subtyping (\cref{alg:subtyping}),
-type well-formedness (\cref{alg:wf}),
-constraint merge (\cref{sec:constraint-merge}),
-normalization (\cref{alg:type-nf}),
-and constraint singularity which will be defined later in 
-\cref{sec:constraint-singularity}.
-It also relies on basic set operations and the ability to 
-deterministically choose fresh variables.
-
-\begin{algorithm}
-  \hfill\\
-  \ottdefnATNInf{}
-  \ottdefnATPInf{}
-  \ottdefnATSpinInf{}
-\end{algorithm}
-
-Let us discuss the inference rules of the algorithm:
-\begin{itemize}
-  \item \ruleref{\ottdruleATVarLabel} 
-    infers the type of a variable by looking it up in the context
-    and normalizing the result.
-  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
-    are similar to the declarative rules: they make a recursive call
-    to type the body of the thunk or the return expression and
-    put the shift on top of the result.
-  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
-    are symmetric. They make a recursive call to infer the type of the annotated
-    expression, check that the inferred type is a subtype of the annotation,
-    and return the normalized annotation.
-  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
-    It makes a recursive call to infer the type of the body in the extended context,
-    and returns the corresponding arrow type. Notice that the algorithm also
-    normalizes the result, which is because the annotation type $[[iP]]$
-    is allowed to be non-normalized.
-  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
-    Similarly to the previous case, it makes a recursive call to infer the type
-    of the body in the extended \emph{type} context. 
-    After that, it returns the corresponding universal type. 
-    It is also required to normalize the result, because, 
-    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
-    in which case the $[[∀]]$ must be removed. 
-  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
-    it makes a recursive call to infer the type of the bound value,
-    and then returns the type of the body in the extended context.
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
-    is interpreted as follows.
-    First, it infers the type of the head of the application,
-    ensuring that it is a thunked computation $[[↓iM]]$;
-    after that, it makes a recursive call
-    to the application inference procedure,
-    which returns the algorithmic type, whose
-    instantiation to a declarative type must be associated with the bound variable 
-    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
-    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
-    check whether the instantiation to the annotated type $[[iP]]$ is possible,
-    and if it is, the algorithm infers the type of the body in the extended context,
-    and returns it as the result. 
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
-    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
-    However, since there is no annotation, 
-    instead of checking the instantiation to it, 
-    the algorithm checks that the inferred type
-    $[[↑uQ]]$ is unique.
-    It is the case if all the algorithmic variables of 
-    $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$,
-    which is checked by the combination of
-    $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
-    Together, these two premises guarantee that the only 
-    possible instantiation of $[[↑uQ]]$ is $[[ [uσ]uQ ]]$.
-  \item \ruleref{\ottdruleATUnpackLabel}
-    works in the expected way. First, it infers the 
-    existential type $[[∃nas.iP]]$ of the value being unpacked,
-    and since the type is guaranteed to be normalized, binds 
-    the quantified variables with $[[nas]]$.
-    Then it infers the type of the body in the appropriately extended context,
-    and checks that the inferred type does not depend on $[[nas]]$
-    by checking well-formedness $[[Γ ⊢ iN]]$.
-\end{itemize}
-
-Finally, let us discuss the algorithmic rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleATEmptyAppLabel}
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head,
-    and the algorithm returns it after normalizing.
-  \item \ruleref{\ottdruleATArrowAppLabel}
-    is the main rule of the algorithmic application inference.
-    If the head has an arrow type $[[uQ → uN]]$,
-    we find $[[SC1]]$---the minimal constraint ensuring that 
-    $[[uQ]]$ is a supertype of the first argument's type.
-    Then we make a recursive call applying $[[uN]]$ to the rest of the arguments,
-    and merge the resulting constraint with $[[SC1]]$
-  \item \ruleref{\ottdruleATForallAppLabel},
-    analogously to the declarative case,
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    This is the place where the algorithmic variables are generated.
-    The algorithm simply replaces the quantified variables 
-    $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
-    and makes a recursive call in the extended context. 
-\end{itemize}
-
-The correctness of the algorithm consists of its soundness and 
-completeness, which is by mutual
-induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
-The simplified result is the following.
-\begin{theorempreview}
-  \hfill
-  \begin{itemize}
-    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
-      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
-    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
-      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraint Singularity}
-\label{sec:constraint-singularity}
-
-The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
-of the algorithmic typing check whether the constraint $[[SC]]$
-is uniquely defines a substitution satisfying it, and if it does
-returns such a substitution as the result.
-To do that, we define a partial function $[[SC singular with uσ]]$,
-taking a subtyping constraint $[[SC]]$ and returning a substitution 
-$[[uσ]]$---the only possible solution of $[[SC]]$.
-
-First, we define the notion of singularity on constraint entries. 
-$[[scE singular with iP]]$ and $[[scE singular with iN]]$
-are considered partial functions taking a constraint entry $[[scE]]$
-and returning the type satisfying $[[scE]]$ if such a type is unique.
-
-\begin{algorithm}[Singular Constraint Entry]
-  \hfill\\
-  \ottdefnSINGscEP{}
-  \ottdefnSINGscEN{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSINGNEqLabel} and \ruleref{\ottdruleSINGPEqLabel} 
-    are symmetric. If the constraint entry says that a variable must be equivalent to 
-    a type $T$, then it is evidently singular, and the only (up-to-equivalence) type
-    instantiating this variable could be $T$. This way, we return its normal form. 
-  \item \ruleref{\ottdruleSINGSupVarLabel}
-    implies that the only (normalized) solution of $[[pua :≥ ∃nas.pa]]$ is 
-    $[[pa]]$ (it will be shown in \cref{lemma:var-subt}).
-  \item \ruleref{\ottdruleSINGSupShiftLabel}
-    is perhaps the least obvious rule.
-    In type $[[∃nas.↓iN]]$, if $[[iN]]$ is anything different (non-equivalent) to 
-    $[[nai ∊ {nas}]]$, there are at least 
-    one proper supertype of $[[∃nas.↓iN]]$, since $[[iN]]$ can be abstracted over
-    by an existential quantifier. Otherwise, any supertype of $[[∃nas.↓nai]]$
-    is equivalent to it, and thus, the solution is unique. 
-\end{itemize}
-
-Next, we extrapolate the singularity function on constraints---sets of constraint entries. 
-We require $[[SC]]$ to be a set of singular constraints, and the resulting substitution
-sends each variable from $[[dom(SC)]]$ to the unique type satisfying the corresponding constraint.
-
-\begin{algorithm}
-  $[[SC singular with uσ]]$
-  means that 
-  \begin{enumerate}
-    \item for any positive $[[scE ∊ SC]]$,
-      there exists $[[iP]]$ such that $[[scE singular with iP]]$, 
-      and for any negative $[[scE ∊ SC]]$,
-      there exists $[[iN]]$ such that $[[scE singular with iN]]$;
-    \item $[[uσ]]$ is defined as follows:
-      $$
-      [[ [uσ]β̂⁺ ]]  = 
-          \begin{cases}
-              [[ iP ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁺]] 
-                         \text{ and } [[scE singular with iP]] \\
-              [[ β̂⁺ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-      $$
-      [[ [uσ]β̂⁻ ]]  = 
-          \begin{cases}
-              [[ iN ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁻]] 
-                         \text{ and } [[scE singular with iN]]\\
-              [[ β̂⁻ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-  \end{enumerate} 
-\end{algorithm}
-
-The correctness of the singularity algorithm is formulated as follows:
-\begin{theorempreview}
-  Suppose that $[[SC]]$ is a subtyping constraint.
-  Then $[[SC singular with uσ]]$ holds if and only if 
-  $[[uσ]]$ is the only (up-to-equivalence on $[[dom(SC)]]$) 
-  normalized substitution satisfying $[[SC]]$.
-\end{theorempreview}
-
-\newpage
-
-\section{Properties of the Declarative Type System}
-
-\subsection{Type Well-formedness}
-\input{\genDir/tex/wf-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-lemmas.tex}
-
-\subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Equivalence}
-\label{sec:decl-equiv-lemmas}
-\input{\genDir/tex/equiv-lemmas.tex}
-
-\subsection{Variable Ordering}
-\input{\genDir/tex/ord-lemmas.tex}
-
-\subsection{Normaliztaion}
-\input{\genDir/tex/norm-lemmas.tex}
-
-\section{Properties of the Algorithmic Type System}
-
-\subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-algo-lemmas.tex}
-
-\subsection{Normalization}
-\input{\genDir/tex/norm-alg-lemmas.tex}
-
-\subsection{Equivalence}
-\input{\genDir/tex/equiv-alg-lemmas.tex}
-
-\subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-lemmas.tex}
-
-\subsection{Unification}
-\input{\genDir/tex/unification-lemmas.tex}\
-
-\subsection{Anti-unification}
-\input{\genDir/tex/au-lemmas.tex}
-
-\subsection{Upper Bounds}
-\label{sec:alg-upper-bounds-proofs}
-\input{\genDir/tex/lub-lemmas.tex}
-
-\subsection{Upgrade}
-\label{sec:upgrade-lemmas}
-\input{\genDir/tex/upgrade-lemmas.tex}
-
-\subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-lemmas.tex}
-
-\subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-lemmas.tex}
-
-\subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-lemmas.tex}
-
-\subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-lemmas.tex}
-
-\section{Properties of the Declarative Typing}
-\input{\genDir/tex/decl-typing-lemmas.tex}
-
-\section{Properties of the Algorithmic Typing}
-
-\subsection{Singularity}
-\input{\genDir/tex/singularity-lemmas.tex}
-
-\subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-lemmas.tex}
-
-\printbibliography
-
-\end{document}

### 2023-11-13 proofs/main.tex
Finish the Algorithmic section
Restructure proofs

@@
 
-\usepackage[matha]{mathabx}
+% \usepackage[matha]{mathabx}
 
@@
 \subsection{Equalities}
-For simplicity, we assume alpha-equivalent terms equal.
+For simplicity, we assume alpha-equivalent terms are equal.
 This way, we assume that substitutions do not capture bound variables.
@@
 $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
-which, in fact, means that each free type variable of the type is contained in $[[Γ]]$
+which means that each free type variable of the type is contained in $[[Γ]]$
 (it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
@@
   \begin{multicols}{2}
-  \ottdefnWFTNiWF{}
+  \ottdefnWFTNiWFLabeled{}
   \columnbreak
 
-  \ottdefnWFTPiWF{}
+  \ottdefnWFTPiWFLabeled{}
   \end{multicols}
@@
   Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions form variables to types, preserving the polarity. 
+  are represented by total functions from variables to types, preserving the polarity. 
 \end{definition}
@@
 \subsection{Declarative Subtyping}
-Subtyping is one of the key mechanism of our system. 
+Subtyping is one of the key mechanisms of our system. 
 It realizes the polymorphism: abstract $[[∀]]$ and 
 $[[∃]]$ types can be used where concrete types are expected,
-exactly because the subtyping relation between them.
+exactly because of the subtyping relation between them.
 
@@
   \begin{multicols}{2}
-    \ottdefnDOneNsub{}
-
-    \ottdefnDOnePsup{}
+    \ottdefnDOneNsubLabeled{}
+
+    \ottdefnDOnePsupLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-    \ottdefnDOneNeq{}
-
-    \ottdefnDOnePeq{}
+    \ottdefnDOneNeqLabeled{}
+
+    \ottdefnDOnePeqLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-  \ottdefnWFATNauWF{}
+  \ottdefnWFATNauWFLabeled{}
   \columnbreak
 
-  \ottdefnWFATPauWF{}
+  \ottdefnWFATPauWFLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-  \ottdefnEOneNeq{}
+  \ottdefnEOneNeqLabeled{}
   \columnbreak\\
-  \ottdefnEOnePeq{}
+  \ottdefnEOnePeqLabeled{}
   \end{multicols}
@@
 
-  \ottdefnONVar{}
-  \ottdefnOPVar{}
+  \ottdefnONVarLabeled{}
+  \ottdefnOPVarLabeled{}
 
@@
   \begin{multicols}{2}
-  \ottdefnNrmNNorm{}
+  \ottdefnNrmNNormLabeled{}
   \columnbreak\\
-  \ottdefnNrmPNorm{}
+  \ottdefnNrmPNormLabeled{}
   \end{multicols}
@@
   \begin{multicols}{2}
-  \ottdefnNrmuNNorm{}
+  \ottdefnNrmuNNormLabeled{}
   \columnbreak\\
-  \ottdefnNrmuPNorm{}
+  \ottdefnNrmuPNormLabeled{}
   \end{multicols}
@@
   \hfill\\
-  \ottdefnANsub{}
-  \ottdefnAPsup{}
+  \ottdefnANsubLabeled{}
+  \ottdefnAPsupLabeled{}
 \end{algorithm}
@@
   \begin{multicols}{2}
-  \ottdefnSATSCEN{}
+  \ottdefnSATSCENLabeled{}
   \columnbreak\\
-  \ottdefnSATSCEP{}
+  \ottdefnSATSCEPLabeled{}
   \end{multicols}
@@
 
-  \ottdefnSCME\\
+  \ottdefnSCMELabeled\\
 \end{algorithm}
@@
   \hfill\\
-  \ottdefnLUBNsub{}
+  \ottdefnLUBNsubLabeled{}
 \end{algorithm}
@@
   \hfill\\
-  \ottdefnLUBUp{}
+  \ottdefnLUBUpLabeled{}
 \end{algorithm}
@@
 
-  \ottdefnAUAUP{}
-  \ottdefnAUAUN{}
+  \ottdefnAUAUPLabeled{}
+  \ottdefnAUAUNLabeled{}
 
@@
   \label{def:declarative-typing}
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnDTNInf{}
-  \ottdefnDTPInf{}
-  \ottdefnDTSpinInf{}
-  \end{multicols}
-\end{definition}
-
-Let us discuss selected rules of the declarative system:
+  \ottdefnsDTLabeled
+\end{definition}
+
+Let us discuss the selected rules of the declarative system:
 \begin{itemize}
@@
   \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
-    mean that th declarative system allows to infer any type from the equivalence class.
+    mean that the declarative system allows us to infer any type from the equivalence class.
   \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
-    and its first premise inferring the existential type of the value being unpacked.
+    and its first premise infers the existential type of the value being unpacked.
     It is important however that the inferred existential type is normalized. 
     This is because there might be multiple equivalent existential types 
-    with different order or even number of quantified variables, 
+    with a different order or even number of quantified variables, 
     and to bind them, the algorithm needs to fix the canonical one.
@@
   \hfill\\
-  \ottdefnATNInf{}
-  \ottdefnATPInf{}
-  \ottdefnATSpinInf{}
+  \ottdefnATNInfLabeled{}
+  \ottdefnATPInfLabeled{}
+  \ottdefnATSpinInfLabeled{}
 \end{algorithm}
@@
   \item \ruleref{\ottdruleATArrowAppLabel}
-    is the main rule of the algorithmic application inference.
+    is the main rule of algorithmic application inference.
     If the head has an arrow type $[[uQ → uN]]$,
@@
   \hfill\\
-  \ottdefnSINGscEP{}
-  \ottdefnSINGscEN{}
+  \ottdefnSINGscEPLabeled{}
+  \ottdefnSINGscENLabeled{}
 \end{algorithm}
@@
 
+\section{Theorem Statements: Declarative}
+\subsection{Type Well-Formedness}
+\input{\genDir/tex/wf-statements.tex}
+
+\subsection{Substitution}
+\input{\genDir/tex/subst-statements.tex}
+
+\subsection{Declarative Subtyping}
+\input{\genDir/tex/decl-subtyping-statements.tex}
+
+\subsection{Equivalence}
+\label{sec:decl-equiv-statements}
+\input{\genDir/tex/equiv-statements.tex}
+
+\subsection{Variable Ordering}
+\input{\genDir/tex/ord-statements.tex}
+
+\subsection{Normaliztaion}
+\input{\genDir/tex/norm-statements.tex}
+
+\section{Declarative Typing}
+\input{\genDir/tex/decl-typing-statements.tex}
+
+\section{Theorem Statements: Algorithmic}
+
+\subsection{Algorithmic Type Well-formedness}
+\input{\genDir/tex/wf-algo-statements.tex}
+
+\subsection{Algorithmic Substitution}
+\input{\genDir/tex/subst-algo-statements.tex}
+
+\subsection{Algorithmic Normalization}
+\input{\genDir/tex/norm-alg-statements.tex}
+
+\subsection{Algorithmic Equivalence}
+\input{\genDir/tex/equiv-alg-statements.tex}
+
+\subsection{Unification Constraint Merge}
+\input{\genDir/tex/unif-constraint-merge-statements.tex}
+
+\subsection{Unification}
+\input{\genDir/tex/unification-statements.tex}
+
+\subsection{Anti-unification}
+\input{\genDir/tex/au-statements.tex}
+
+\subsection{Upper Bounds}
+\label{sec:alg-upper-bounds-statements}
+\input{\genDir/tex/lub-statements.tex}
+
+\subsection{Upgrade}
+\label{sec:upgrade-statements}
+\input{\genDir/tex/upgrade-statements.tex}
+
+\subsection{Constraint Satisfaction}
+\input{\genDir/tex/constraint-sat-statements.tex}
+
+\subsection{Positive Subtyping}
+\input{\genDir/tex/alg-pos-subtyping-statements.tex}
+
+\subsection{Subtyping Constraint Merge}
+\input{\genDir/tex/constraint-merge-statements.tex}
+
+\subsection{Negative Subtyping}
+\input{\genDir/tex/alg-neg-subtyping-statements.tex}
+
+\subsection{Singularity}
+\input{\genDir/tex/singularity-statements.tex}
+
+\subsection{Correctness of the Typing Algorithm}
+\input{\genDir/tex/alg-typing-statements.tex}
+
 \section{Properties of the Declarative Type System}
 
-\subsection{Type Well-formedness}
+\subsection{Type Well-Formedness}
 \input{\genDir/tex/wf-lemmas.tex}

### 2023-11-13 proofs/tex/alg-neg-subtyping-lemmas.tex
Restructure proofs

@@
-\begin{observation}[Negative Algorithmic Subtyping is Deterministic]
-    \label{obs:neg-subt-deterministic}
-    For fixed $[[Γ]]$, $[[Θ]]$, $[[iM]]$, and $[[uN]]$, 
-    if $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ 
-    and $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC']]$
-    then $[[SC]]$ = $[[SC']]$.
-\end{observation}
+\obsNegSubtDeterministic*
 \begin{proof}
@@
 
-\begin{lemma}[Soundness of Negative Subtyping] \label{lemma:neg-subt-soundness}
-        If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
-        $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
-        $[[Θ ⊢ SC : uv(uN)]]$ and 
-        for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
-        $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
-\end{lemma}
+\lemNegSubtSoundness*
 \begin{proof}
@@
 
-\begin{lemma}[Completeness of the Negative Subtyping]
-    \label{lemma:neg-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
-    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
-    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
-    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
-\end{lemma}
+\lemNegSubtCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/alg-neg-subtyping-statements.tex
Restructure proofs

\begin{restatable}[Negative Algorithmic Subtyping is Deterministic]{observation}{obsNegSubtDeterministic}
    \label{obs:neg-subt-deterministic}
    For fixed $[[Γ]]$, $[[Θ]]$, $[[iM]]$, and $[[uN]]$, 
    if $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ 
    and $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC']]$
    then $[[SC]]$ = $[[SC']]$.
\end{restatable}

\begin{restatable}[Soundness of Negative Subtyping]{lemma}{lemNegSubtSoundness}
    \label{lemma:neg-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
    $[[Θ ⊢ SC : uv(uN)]]$ and 
    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{restatable}

\begin{restatable}[Completeness of the Negative Subtyping]{lemma}{lemNegSubtCompleteness}
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{restatable}

### 2023-11-13 proofs/tex/alg-pos-subtyping-lemmas.tex
Restructure proofs

@@
-\begin{observation}[Positive Subtyping is Deterministic]
-    \label{obs:pos-subt-deterministic}
-    For fixed $[[Γ]]$, $[[Θ]]$, $[[uP]]$, and $[[iQ]]$,
-    if $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC']]$
-    then $[[SC]] = [[SC']]$.
-\end{observation}
+\obsPosSubtDeterministic*
 \begin{proof}
@@
 
-\begin{lemma}[Soundness of the Positive Subtyping] \label{lemma:pos-subt-soundness}
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
-    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
-    then $[[Θ ⊢ SC : uv uP]]$ and
-    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
-    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
-\end{lemma}
+\lemPosSubtSoundness*
 \begin{proof} 
@@
 
-\begin{lemma}[Completeness of the Positive Subtyping] \label{lemma:pos-subt-completeness}
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
-    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
-    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
-\end{lemma}
+\lemPosSubtCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/alg-pos-subtyping-statements.tex
Restructure proofs

\begin{restatable}[Positive Subtyping is Deterministic]{observation}{obsPosSubtDeterministic}
    \label{obs:pos-subt-deterministic}
    For fixed $[[Γ]]$, $[[Θ]]$, $[[uP]]$, and $[[iQ]]$,
    if $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC']]$
    then $[[SC]] = [[SC']]$.
\end{restatable}

\begin{restatable}[Soundness of the Positive Subtyping]{lemma}{lemPosSubtSoundness} \label{lemma:pos-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{restatable}

\begin{restatable}[Completeness of the Positive Subtyping]{lemma}{lemPosSubtCompleteness} \label{lemma:pos-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{restatable}

### 2023-11-13 proofs/tex/au-lemmas.tex
Restructure proofs

@@
-\begin{observation}[Anti-unification algorithm is deterministic]
-    \label{obs:au-deterministic}
-    \hfill\\
-\begin{itemize}
-    \item [$+$] If $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$ and 
-    $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ', uQ', aus'1, aus'2)]]$,
-    then $[[Ξ]] = [[Ξ']]$, $[[uQ]] = [[uQ']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
-    \item [$-$] If $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$ and
-    $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ', uM', aus'1, aus'2)]]$,
-    then $[[Ξ]] = [[Ξ']]$, $[[uM]] = [[uM']]$, $[[aus1]] = [[aus'1]]$, and $[[aus2]] = [[aus'2]]$.
-\end{itemize}
-\end{observation}
+\obsAuDeterministic*
 \begin{proof}
@@
 
-\begin{observation} \label{obs:names-defined-by-mapping}
-    Names of the anti-unification variables are uniquely defined by
-    the types they are mapped to by the resulting substitutions.
-
-    \begin{itemize}
-        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
-        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-        then for any $[[β̂⁻]] \in [[Ξ]]$,
-        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
-        \item [$-$]  Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
-        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
-        then for any $[[β̂⁻]] \in [[Ξ]]$,
-        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$
-    \end{itemize}
-\end{observation}
+\obsNamesDefinedByMapping*
 \begin{proof}
@@
 
-\begin{lemma}[Soundness of Anti-Unification] \label{lemma:au-soundness}
-    \hfill
-    \begin{itemize}
-        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
-        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-        then 
-        \begin{enumerate}
-            \item $[[Γ ; Ξ ⊢ uQ]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification substitutions, and
-            \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
-        \end{enumerate}
-
-        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
-        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
-        then
-        \begin{enumerate}
-            \item $[[Γ ; Ξ ⊢ uM]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification substitutions, and
-            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
-        \end{enumerate}
-    \end{itemize}
-\end{lemma}
+\lemmaAuSoundness*
 \begin{proof}
@@
 
-% \renewcommand{\ottdruleAUPVarName}[0]{(Var$^{+[[≈au]]}$)}
-% \renewcommand{\ottdruleAUShiftDName}[0]{($\downarrow^{[[≈au]]}$)}
-% \renewcommand{\ottdruleAUExistsName}[0]{($\exists^{[[≈au]]}$)}
-
-% G ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )
-% \renewcommand{\ottdruleAUNVarName}[0]{(Var$^{-[[≈au]]}$)}
-% \renewcommand{\ottdruleAUShiftUName}[0]{($\uparrow^{[[≈au]]}$)}
-% \renewcommand{\ottdruleAUForallName}[0]{($\forall^{[[≈au]]}$)}
-% \renewcommand{\ottdruleAUArrowName}[0]{($\rightarrow^{[[≈au]]}$)}
-% \renewcommand{\ottdruleAUAUName}[0]{(AU$^{-}$)}
-
-\begin{lemma}[Completeness of Anti-Unification] \label{lemma:au-completeness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] 
-            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
-            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uQ']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
-                are anti-unification substitutions, and
-                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
-            \end{enumerate}
-
-            Then the anti-unification algorithm terminates, that is there exists
-            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-
-        \item [$-$] 
-            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
-            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
-                are anti-unification substitutions, and
-                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
-            \end{enumerate}
-
-            Then the anti-unification algorithm succeeds, that is 
-            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
-            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
-    \end{itemize}
-\end{lemma}
+\lemmaAuCompleteness*
 \begin{proof}
@@
 
-\begin{lemma}[Initiality of Anti-Unification] \label{lemma:au-initial}
-    \hfill
-    \begin{itemize}
-        \item [$+$] 
-            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
-            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
-            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
-            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
-            if 
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uQ']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
-                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
-            \end{enumerate}
-            then there exists $[[ausr]]$ such that
-            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
-            Moreover, $[[ [ausr]β̂⁻]]$ can
-            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
-            $[[Γ]]$.
-        \item [$-$] 
-            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
-            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
-            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
-            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
-            if
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
-                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
-            \end{enumerate}
-            then there exists $[[ausr]]$ such that
-            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
-            Moreover, $[[ [ausr]β̂⁻]]$ can
-            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
-            $[[Γ]]$.
-    \end{itemize}
-\end{lemma}
+\lemmaAuInitial*
 \begin{proof}

### 2023-11-13 proofs/tex/constraint-merge-lemmas.tex
Restructure proofs

@@
-\begin{observation}[Constraint Entry Merge is Deterministic]
-    \label{obs:entry-merge-deterministic}
-    For fixed $[[Γ]]$, $[[scE1]]$, $[[scE2]]$,
-    if $[[Γ ⊢ scE1 & scE2 = scE]]$ and $[[Γ ⊢ scE1 & scE2 = scE']]$ 
-    then $[[scE]] = [[scE']]$.
-\end{observation}
+\obsEntryMergeDeterministic*
 \begin{proof}
@@
 
-\begin{observation}[Subgyping Constraint Merge is Deterministic]
-    \label{obs:subt-merge-deterministic}
-    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$
-    If $[[Θ ⊢ SC1 & SC2 = UC]]$ and $[[Θ ⊢ SC1 & SC2 = UC']]$ are defined then 
-    $[[SC]] = [[SC']]$.
-\end{observation}
+\obsSubtMergeDeterministic*
 \begin{proof}
@@
 
-\begin{lemma} [Soundness of Constraint Entry Merge]
-    \label{lemma:entry-merge-soundness}
-    For a fixed context $[[Γ]]$,
-    suppose that  $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$. 
-    If $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined then
-    \begin{enumerate}
-        \item $[[Γ ⊢ scE]]$
-        \item For any $[[Γ ⊢ iP]]$, $[[Γ ⊢ iP : scE]]$ implies $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ 
-    \end{enumerate}
-\end{lemma}
+\lemEntryMergeSoundness*
 \begin{proof}
@@
 
-\begin{lemma} [Soundness of Constraint Merge] \label{lemma:merge-soundness}
-    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
-    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
-    Then 
-    \begin{enumerate}
-        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
-        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
-            $[[ Θ ⊢ uσ : SC ]]$
-            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
-    \end{enumerate}
-\end{lemma}
+\lemMergeSoundness*
 \begin{proof}
@@
 
-\begin{lemma} [Completeness of Constraint Entry Merge]
-    \label{lemma:entry-merge-completeness}
-    For a fixed context $[[Γ]]$,
-    suppose that $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$ are matching constraint entries.
-    \begin{itemize}
-        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$,
-        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iP : scE]]$.
-        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : scE1]]$ and $[[Γ ⊢ iN : scE2]]$,
-        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iN : scE]]$.
-    \end{itemize}
-\end{lemma}
+\lemEntryMergeCompleteness*
 \begin{proof}
@@
 
-\begin{lemma} [Completeness of Constraint Merge] 
-    \label{lemma:merge-completeness}
-    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
-    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
-    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
-    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
-\end{lemma}
+\lemMergeCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/constraint-merge-statements.tex
Restructure proofs

\begin{restatable}[Positive Subtyping is Deterministic]{observation}{obsPosSubtDeterministic}
    \label{obs:pos-subt-deterministic}
    For fixed $[[Γ]]$, $[[Θ]]$, $[[uP]]$, and $[[iQ]]$,
    if $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC']]$
    then $[[SC]] = [[SC']]$.
\end{restatable}

\begin{restatable}[Constraint Entry Merge is Deterministic]{observation}{obsEntryMergeDeterministic}
    \label{obs:entry-merge-deterministic}
    For fixed $[[Γ]]$, $[[scE1]]$, $[[scE2]]$,
    if $[[Γ ⊢ scE1 & scE2 = scE]]$ and $[[Γ ⊢ scE1 & scE2 = scE']]$ 
    then $[[scE]] = [[scE']]$.
\end{restatable}

\begin{restatable}[Subtyping Constraint Merge is Deterministic]{observation}{obsSubtMergeDeterministic}
    \label{obs:subt-merge-deterministic}
    Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$
    If $[[Θ ⊢ SC1 & SC2 = SC]]$ and $[[Θ ⊢ SC1 & SC2 = SC']]$ are defined then 
    $[[SC]] = [[SC']]$.
\end{restatable}

\begin{restatable}[Soundness of Constraint Entry Merge]{lemma}{lemEntryMergeSoundness}
    \label{lemma:entry-merge-soundness}
    For a fixed context $[[Γ]]$,
    suppose that  $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$. 
    If $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined then
    \begin{enumerate}
        \item $[[Γ ⊢ scE]]$
        \item For any $[[Γ ⊢ iP]]$, $[[Γ ⊢ iP : scE]]$ implies $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$ 
    \end{enumerate}
\end{restatable}

\begin{restatable}[Soundness of Constraint Merge]{lemma}{lemMergeSoundness} 
    \label{lemma:merge-soundness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
    Then 
    \begin{enumerate}
        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
            $[[ Θ ⊢ uσ : SC ]]$
            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
    \end{enumerate}
\end{restatable}

\begin{restatable}[Completeness of Constraint Entry Merge]{lemma}{lemEntryMergeCompleteness}
    \label{lemma:entry-merge-completeness}
    For a fixed context $[[Γ]]$,
    suppose that $[[Γ ⊢ scE1]]$ and $[[Γ ⊢ scE2]]$ are matching constraint entries.
    \begin{itemize}
        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : scE1]]$ and $[[Γ ⊢ iP : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iP : scE]]$.
        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : scE1]]$ and $[[Γ ⊢ iN : scE2]]$,
        $[[Γ ⊢ scE1 & scE2 = scE]]$ is defined and $[[Γ ⊢ iN : scE]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of Constraint Merge]{lemma}{lemMergeCompleteness} 
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
\end{restatable}

### 2023-11-13 proofs/tex/constraint-sat-statements.tex
Restructure proofs

\begin{restatable}[Any constraint is satisfiable]{lemma}{lemmaConstraintSat}
    \label{lemma:constraint-sat}
    Suppose that $[[Θ ⊢ SC]]$
    and $[[Ξ]]$ is a set such that
    $[[dom(SC)]] \subseteq [[Ξ]] \subseteq [[dom(Θ)]]$.
    Then there exists 
    $[[uσ]]$ such that 
    $[[ Θ  ⊢ uσ : Ξ ]]$ and
    $[[ Θ  ⊢ uσ : SC ]]$.
\end{restatable}

\begin{restatable}[Constraint Entry Satisfaction is Stable under Equivalence]{lemma}{lemmaEntrySatEquiv}
    \label{lemma:entry-sat-equiv}
    \begin{itemize}
        \item [$-$] If $[[Γ ⊢ iN1 : scE]]$ and $[[Γ ⊢ iN1 ≈ iN2]]$ then $[[Γ ⊢ iN2 : scE]]$.
        \item [$+$] If $[[Γ ⊢ iP1 : scE]]$ and $[[Γ ⊢ iP1 ≈ iP2]]$ then $[[Γ ⊢ iP2 : scE]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Constraint Satisfaction is stable under Equivalence]{corollary}{corollarySatEquiv}
    \label{corollary:sat-equiv}
    \hfill\\
    If $[[ Θ ⊢ uσ1 : SC ]]$ and $[[Θ ⊢ uσ1 ≈ uσ2 : dom(SC) ]]$ then $[[ Θ ⊢ uσ2 : SC ]]$;\\
    if $[[ Θ ⊢ uσ1 : UC ]]$ and $[[Θ ⊢ uσ1 ≈ uσ2 : dom(SC) ]]$ then $[[ Θ ⊢ uσ2 : UC ]]$.
\end{restatable}

\begin{restatable}[Normalization preserves Constraint Satisfaction]{corollary}{corollaryNfSat}
    \label{corollary:nf-sat}
    \hfill\\
    If $[[Θ ⊢ uσ : SC]]$ then $[[Θ  ⊢ nf(uσ) : SC]]$;\\
    if $[[Θ ⊢ uσ : UC]]$ then $[[Θ  ⊢ nf(uσ) : UC]]$.
\end{restatable}

### 2023-11-13 proofs/tex/decl-subtyping-lemmas.tex
Restructure proofs

@@
-\begin{lemma}[Free Variable Propagation] \label{lemma:fv-propagation}
-  In the judgments of negative subtyping or positive supertyping,
-  free variables propagate left-to-right. For a context $[[Γ]]$,
-  \begin{itemize}
-    \item $-$ if $[[Γ ⊢ iN ≤ iM]]$ then $[[fv(iN)]] \subseteq [[fv(iM)]]$
-    \item $+$ if $[[Γ ⊢ iP ≥ iQ]]$ then $[[fv(iP)]] \subseteq [[fv(iQ)]]$
-  \end{itemize}
-\end{lemma}
+\lemmaFvPropagation*
 \begin{proof}
@@
 
-\begin{corollary}[Free Variables of mutual subtypes] \label{corollary:fv-mut-sub}
-  \hfill
-  \begin{itemize}
-    \item [$-$] If $[[Γ ⊢ iN ≈ iM]]$ then $[[fv iN]] = [[fv iM]]$, 
-    \item [$+$] If $[[Γ ⊢ iP ≈ iQ]]$ then $[[fv iP]] = [[fv iQ]]$
-  \end{itemize}
-\end{corollary}
+\corollaryFvMutSub*
 
-\begin{lemma}[Decomposition of quantifier rules]
-  \label{lemma:quant-rule-decomposition}
-  Assuming that $[[pas]]$, $[[pbs]]$, $[[nas]]$, and $[[nas]]$ are disjoint from $[[Γ]]$,
-  \begin{itemize}
-    \item [$-_{R}$] $[[Γ ⊢ iN ≤ ∀pbs.iM]]$ holds if and only if $[[Γ, pbs ⊢ iN ≤ iM]]$;
-    \item [$+_{R}$] $[[Γ ⊢ iP ≥ ∃nbs.iQ]]$ holds if and only if $[[Γ, nbs ⊢ iP ≥ iQ]]$;
-    \item [$-_{L}$] suppose $[[iM]] \neq [[∀]]\dots$
-      then $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$
-      for some $[[Γ ⊢ iPs]]$;
-    \item [$+_{L}$] suppose $[[iQ]] \neq [[∃]]\dots$
-      then $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ [iNs/nas]iP ≥ iQ]]$
-      for some $[[Γ ⊢ iNs]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaMutSubTypesWfEquiv*
+\begin{proof}
+  From \cref{lemma:wf-ctxt-equiv,corollary:fv-mut-sub}.
+\end{proof}
+
+\lemmaQuantRuleDecomposition*
 \begin{proof}
@@
 
-\begin{corollary}[Redundant quantifier elimination]
-  \label{corollary:red-quant-elim}
-  \hfill
-  \begin{itemize}
-    \item [$-_{L}$] Suppose that $[[ {pas} ∩ fv(iN) = ∅]]$ then 
-      $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
-    \item [$-_{R}$] Suppose that $[[ {pas} ∩ fv(iM) = ∅]]$ then 
-      $[[Γ ⊢ iN ≤ ∀pas.iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
-    \item [$+_{L}$] Suppose that $[[ {nas} ∩ fv(iP) = ∅]]$ then
-      $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
-    \item [$+_{R}$] Suppose that $[[ {nas} ∩ fv(iQ) = ∅]]$ then 
-      $[[Γ ⊢ iP ≥ ∃nas.iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
-  \end{itemize}
-\end{corollary}
+\corollaryRedQuantElim*
 \begin{proof}
@@
 
-\begin{lemma}[Subtypes and supertypes of a variable]
-  \label{lemma:var-subt}
-  Assuming $[[Γ ⊢  α⁻]]$, $[[Γ ⊢ α⁺]]$, $[[Γ ⊢ iN]]$, and $[[Γ ⊢ iP]]$,
-  \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ iP ≥ ∃nas.α⁺]]$ or $[[Γ ⊢ ∃nas.α⁺ ≥ iP ]]$ then $[[iP]] = [[∃nbs.α⁺]]$ (for some potentially empty $[[nbs]]$)
-  \item[$-$] if $[[Γ ⊢ iN ≤ ∀pas.α⁻]]$ or $[[Γ ⊢ ∀pas.α⁻ ≤ iN ]]$ then $[[iN]] = [[∀pbs.α⁻]]$ (for some potentially empty $[[pbs]]$)
-  \end{itemize}
-\end{lemma}
+\lemmaVarSubt*
 \begin{proof}
@@
 
-\begin{corollary}[Variables have no proper subtypes and supertypes]
-  \label{corollary:vars-no-proper-subtypes}
-  Assuming that all mentioned types are well-formed in $[[Γ]]$,
-  \begin{align*}
-    [[Γ ⊢ iP ≥ α⁺]] ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
-    [[Γ ⊢ α⁺≥ iP]]  ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
-    [[Γ ⊢ iN ≤ α⁻]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
-    [[Γ ⊢ α⁻ ≤ iN]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
-  \end{align*}
-\end{corollary}
+\corollaryVarsNoProperSubtypes*
 \begin{proof}
@@
 
-\begin{lemma}[Subtyping context irrelevance]
-  \label{lemma:subt-ctxt-irrelevance}
-  Suppose that all the mentioned types are well-formed in $[[Γ1]]$ and $[[Γ2]]$. 
-  Then 
-  \begin{enumerate}
-    \item[$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ is equivalent to $[[Γ2 ⊢ iP ≥ iQ]]$;
-    \item[$-$] $[[Γ1 ⊢ iN ≤ iM]]$ is equivalent to $[[Γ2 ⊢ iN ≤ iM]]$.
-  \end{enumerate}
-\end{lemma}
+\lemmaSubtCtxtIrrelevance*
 \begin{proof}
@@
 
-  Suppose that the judgement is inferred by \ruleref{\ottdruleDOneExistsLabel}.
+  Suppose that the judgment is inferred by \ruleref{\ottdruleDOneExistsLabel}.
   Then we are proving that $[[Γ1 ⊢ ∃nas.iP ≥ ∃nbs.iQ]]$ implies 
@@
     \item $[[Γ1, nbs ⊢ [σ0]iP ≥ iQ ]]$, 
-      notice that the tree inferring this judgement 
+      notice that the tree inferring this judgment 
       is the same tree inferring $[[Γ1, nbs ⊢ [σ]iP ≥ iQ ]]$ 
@@
 
-\begin{lemma}[Weakening of subtyping context]
-  \label{lemma:subt-weakening}
-  Suppose $[[Γ1]]$ and $[[Γ2]]$ are contexts 
-  and $[[ Γ1 ⊆ Γ2 ]]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ implies $[[Γ2 ⊢ iP ≥ iQ]]$;
-    \item [$-$] $[[Γ1 ⊢ iN ≤ iM]]$ implies $[[Γ2 ⊢ iN ≤ iM]]$.
-  \end{enumerate}
-\end{lemma}
+\lemmaSubtWeakening*
 \begin{proof}
+  By straightforward induction on the subtyping derivation.
+  The polymorphic cases follow from \cref{lemma:subst-range-weakening}.
 \end{proof}
 
-\begin{lemma}[Reflexivity of subtyping] \label{lemma:subtyping-reflexivity}
-  Assuming all the types are well-formed in $[[Γ]]$,
-  \begin{itemize}
-    \item [$-$] $[[Γ ⊢ iN ≤ iN]]$
-    \item [$+$] $[[Γ ⊢ iP ≥ iP]]$
-  \end{itemize}
-\end{lemma}
+\lemmaSubtypingReflexivity*
 \begin{proof}
@@
 
-\begin{lemma}[Substitution preserves subtyipng]
-  \label{lemma:subst-pres-subt}
-  Suppose that all mentioned types are well-formed in $[[Γ1]]$,
-  and $[[σ]]$ is a substitution $[[Γ2 ⊢ σ : Γ1]]$.
-  \begin{itemize}
-    \item [$-$] If $[[Γ1 ⊢ iN ≤ iM]]$ then $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$.
-    \item [$+$] If $[[Γ1 ⊢ iP ≥ iQ]]$ then $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaSubstPresSubt*
 \begin{proof}
@@
 
-\begin{corollary}[Substitution preserves subtyping induced equivalence]
-  \label{corollary:subst-pres-equiv}
-  Suppose that $[[Γ ⊢ σ : Γ1]]$. Then
-    \begin{itemize}
-      \item[$+$] if $[[Γ1 ⊢ iP]]$,~ $[[Γ1 ⊢ iQ]]$,~ and $[[Γ1 ⊢ iP ≈ iQ]]$ ~ 
-        then $[[Γ ⊢ [σ]iP ≈ [σ]iQ]]$
-      \item[$-$] if $[[Γ1 ⊢ iN]]$,~ $[[Γ1 ⊢ iM]]$,~ and $[[Γ1 ⊢ iN ≈ iM]]$ ~ 
-        then $[[Γ ⊢ [σ]iN ≈ [σ]iM]]$
-    \end{itemize}
-\end{corollary}
+\corollarySubstPresEquiv*
 
-\begin{lemma}[Transitivity of subtyping] 
-  \label{lemma:subtyping-transitivity}
-  Assuming the types are well-formed in $[[Γ]]$,
-  \begin{itemize}
-    \item[$-$] if $[[Γ ⊢ iN1 ≤ iN2]]$ and $[[Γ ⊢ iN2 ≤ iN3]]$ then $[[Γ ⊢ iN1 ≤ iN3]]$,
-    \item[$+$] if $[[Γ ⊢ iP1 ≥ iP2]]$ and $[[Γ ⊢ iP2 ≥ iP3]]$ then $[[Γ ⊢ iP1 ≥ iP3]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaSubtypingTransitivity*
 \begin{proof}
@@
       \cref{lemma:var-subt,corollary:red-quant-elim} are agnostic to the 
-      side on which the the quantifiers occur, and thus, 
+      side on which the quantifiers occur, and thus, 
       the proof stays the same. 
@@
   \end{caseof}
-  After that we consider all the 
-  analogous positive cases, and prove them symmetrically.
+  After that, we consider all the 
+  analogous positive cases and prove them symmetrically.
 \end{proof}
 
-\begin{corollary}[Transitivity of equivalence] \label{corollary:equivalence-transitivity}
-  Assuming the types are well-formed in $[[Γ]]$,
-  \begin{itemize}
-    \item[$-$] if $[[Γ ⊢ iN1 ≈ iN2]]$ and $[[Γ ⊢ iN2 ≈ iN3]]$ then $[[Γ ⊢ iN1 ≈ iN3]]$,
-    \item[$+$] if $[[Γ ⊢ iP1 ≈ iP2]]$ and $[[Γ ⊢ iP2 ≈ iP3]]$ then $[[Γ ⊢ iP1 ≈ iP3]]$.
-  \end{itemize}
-\end{corollary}
+\corollaryEquivalenceTransitivity*

### 2023-11-13 proofs/tex/decl-subtyping-statements.tex
Restructure proofs

\begin{restatable}[Free Variable Propagation]{lemma}{lemmaFvPropagation}
    \label{lemma:fv-propagation}
    In the judgments of negative subtyping or positive supertyping,
    free variables propagate left to right. For a context $[[Γ]]$,
    \begin{itemize}
        \item [$-$] if $[[Γ ⊢ iN ≤ iM]]$ then $[[fv(iN)]] \subseteq [[fv(iM)]]$
        \item [$+$] if $[[Γ ⊢ iP ≥ iQ]]$ then $[[fv(iP)]] \subseteq [[fv(iQ)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Free Variables of mutual subtypes]{corollary}{corollaryFvMutSub}
    \label{corollary:fv-mut-sub}
    \hfill
    \begin{itemize}
        \item [$-$] If $[[Γ ⊢ iN ≈ iM]]$ then $[[fv iN]] = [[fv iM]]$, 
        \item [$+$] If $[[Γ ⊢ iP ≈ iQ]]$ then $[[fv iP]] = [[fv iQ]]$
    \end{itemize}
\end{restatable}

\begin{restatable}{corollary}{lemmaMutSubTypesWfEquiv}
    \label{lemma:mut-sub-types-wf-equiv}
    Suppose that all the types below are well-formed in $[[Γ]]$ and
    $[[Γ' ⊆ Γ]]$. Then
    \begin{itemize}
    \item[$+$] $[[Γ ⊢ iP ≈ iQ]]$ implies $[[Γ' ⊢ iP]] \iff [[Γ' ⊢ iQ]]$
    \item[$-$] $[[Γ ⊢ iN ≈ iM]]$ implies $[[Γ' ⊢ iN]] \iff [[Γ' ⊢ iM]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Decomposition of quantifier rules]{lemma}{lemmaQuantRuleDecomposition}
    \label{lemma:quant-rule-decomposition}
    Assuming that $[[pas]]$, $[[pbs]]$, $[[nas]]$, and $[[nas]]$ are disjoint from $[[Γ]]$,
    \begin{itemize}
        \item [$-_{R}$] $[[Γ ⊢ iN ≤ ∀pbs.iM]]$ holds if and only if $[[Γ, pbs ⊢ iN ≤ iM]]$;
        \item [$+_{R}$] $[[Γ ⊢ iP ≥ ∃nbs.iQ]]$ holds if and only if $[[Γ, nbs ⊢ iP ≥ iQ]]$;
        \item [$-_{L}$] suppose $[[iM]] \neq [[∀]]\dots$
            then $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ [iPs/pas]iN ≤ iM]]$
            for some $[[Γ ⊢ iPs]]$;
        \item [$+_{L}$] suppose $[[iQ]] \neq [[∃]]\dots$
            then $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ [iNs/nas]iP ≥ iQ]]$
            for some $[[Γ ⊢ iNs]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Redundant quantifier elimination]{corollary}{corollaryRedQuantElim}
    \label{corollary:red-quant-elim}
    \hfill
    \begin{itemize}
        \item [$-_{L}$] Suppose that $[[ {pas} ∩ fv(iN) = ∅]]$ then 
            $[[Γ ⊢ ∀pas.iN ≤ iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
        \item [$-_{R}$] Suppose that $[[ {pas} ∩ fv(iM) = ∅]]$ then 
            $[[Γ ⊢ iN ≤ ∀pas.iM]]$ holds if and only if $[[Γ ⊢ iN ≤ iM]]$;
        \item [$+_{L}$] Suppose that $[[ {nas} ∩ fv(iP) = ∅]]$ then
            $[[Γ ⊢ ∃nas.iP ≥ iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
        \item [$+_{R}$] Suppose that $[[ {nas} ∩ fv(iQ) = ∅]]$ then 
            $[[Γ ⊢ iP ≥ ∃nas.iQ]]$ holds if and only if $[[Γ ⊢ iP ≥ iQ]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Subtypes and supertypes of a variable]{lemma}{lemmaVarSubt}
    \label{lemma:var-subt}
    Assuming $[[Γ ⊢  α⁻]]$, $[[Γ ⊢ α⁺]]$, $[[Γ ⊢ iN]]$, and $[[Γ ⊢ iP]]$,
    \begin{itemize}
    \item[$+$] if $[[Γ ⊢ iP ≥ ∃nas.α⁺]]$ or $[[Γ ⊢ ∃nas.α⁺ ≥ iP ]]$ then $[[iP]] = [[∃nbs.α⁺]]$ (for some potentially empty $[[nbs]]$)
    \item[$-$] if $[[Γ ⊢ iN ≤ ∀pas.α⁻]]$ or $[[Γ ⊢ ∀pas.α⁻ ≤ iN ]]$ then $[[iN]] = [[∀pbs.α⁻]]$ (for some potentially empty $[[pbs]]$)
    \end{itemize}
\end{restatable}

\begin{restatable}[Variables have no proper subtypes and supertypes]{corollary}{corollaryVarsNoProperSubtypes}
    \label{corollary:vars-no-proper-subtypes}
    Assuming that all mentioned types are well-formed in $[[Γ]]$,
    \begin{align*}
        [[Γ ⊢ iP ≥ α⁺]] ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
        [[Γ ⊢ α⁺≥ iP]]  ~ &\iff ~ [[iP = ∃nbs.α⁺]]  ~ \iff ~ [[Γ ⊢ iP ≈ α⁺]] ~ \iff ~ [[iP ≈ α⁺]]\\
        [[Γ ⊢ iN ≤ α⁻]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
        [[Γ ⊢ α⁻ ≤ iN]] ~ &\iff ~ [[iN = ∀pbs.α⁻]]  ~ \iff ~ [[Γ ⊢ iN ≈ α⁻]] ~ \iff ~ [[iN ≈ α⁻]]\\
    \end{align*}
\end{restatable}

\begin{restatable}[Subtyping context irrelevance]{lemma}{lemmaSubtCtxtIrrelevance}
    \label{lemma:subt-ctxt-irrelevance}
    Suppose that all the mentioned types are well-formed in $[[Γ1]]$ and $[[Γ2]]$. 
    Then 
    \begin{enumerate}
        \item[$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ is equivalent to $[[Γ2 ⊢ iP ≥ iQ]]$;
        \item[$-$] $[[Γ1 ⊢ iN ≤ iM]]$ is equivalent to $[[Γ2 ⊢ iN ≤ iM]]$.
    \end{enumerate}
\end{restatable}

\begin{restatable}[Weakening of subtyping context]{lemma}{lemmaSubtWeakening}
    \label{lemma:subt-weakening}
    Suppose $[[Γ1]]$ and $[[Γ2]]$ are contexts 
    and $[[ Γ1 ⊆ Γ2 ]]$.
    Then 
    \begin{enumerate}
        \item [$+$] $[[Γ1 ⊢ iP ≥ iQ]]$ implies $[[Γ2 ⊢ iP ≥ iQ]]$;
        \item [$-$] $[[Γ1 ⊢ iN ≤ iM]]$ implies $[[Γ2 ⊢ iN ≤ iM]]$.
    \end{enumerate}
\end{restatable}

\begin{restatable}[Reflexivity of subtyping]{lemma}{lemmaSubtypingReflexivity}
    \label{lemma:subtyping-reflexivity}
    Assuming all the types are well-formed in $[[Γ]]$,
    \begin{itemize}
        \item [$-$] $[[Γ ⊢ iN ≤ iN]]$
        \item [$+$] $[[Γ ⊢ iP ≥ iP]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Substitution preserves subtyipng]{lemma}{lemmaSubstPresSubt}
    \label{lemma:subst-pres-subt}
    Suppose that all mentioned types are well-formed in $[[Γ1]]$,
    and $[[σ]]$ is a substitution $[[Γ2 ⊢ σ : Γ1]]$.
    \begin{itemize}
        \item [$-$] If $[[Γ1 ⊢ iN ≤ iM]]$ then $[[Γ2 ⊢ [σ]iN ≤ [σ]iM]]$.
        \item [$+$] If $[[Γ1 ⊢ iP ≥ iQ]]$ then $[[Γ2 ⊢ [σ]iP ≥ [σ]iQ]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Substitution preserves subtyping induced equivalence]{corollary}{corollarySubstPresEquiv}
    \label{corollary:subst-pres-equiv}
    Suppose that $[[Γ ⊢ σ : Γ1]]$. Then
        \begin{itemize}
            \item[$+$] if $[[Γ1 ⊢ iP]]$,~ $[[Γ1 ⊢ iQ]]$,~ and $[[Γ1 ⊢ iP ≈ iQ]]$ ~ 
                then $[[Γ ⊢ [σ]iP ≈ [σ]iQ]]$
            \item[$-$] if $[[Γ1 ⊢ iN]]$,~ $[[Γ1 ⊢ iM]]$,~ and $[[Γ1 ⊢ iN ≈ iM]]$ ~ 
                then $[[Γ ⊢ [σ]iN ≈ [σ]iM]]$
        \end{itemize}
\end{restatable}

\begin{restatable}[Transitivity of subtyping]{lemma}{lemmaSubtypingTransitivity}
        \label{lemma:subtyping-transitivity}
        Assuming the types are well-formed in $[[Γ]]$,
        \begin{itemize}
            \item[$-$] if $[[Γ ⊢ iN1 ≤ iN2]]$ and $[[Γ ⊢ iN2 ≤ iN3]]$ then $[[Γ ⊢ iN1 ≤ iN3]]$,
            \item[$+$] if $[[Γ ⊢ iP1 ≥ iP2]]$ and $[[Γ ⊢ iP2 ≥ iP3]]$ then $[[Γ ⊢ iP1 ≥ iP3]]$.
        \end{itemize}
\end{restatable}

\begin{restatable}[Transitivity of equivalence]{corollary}{corollaryEquivalenceTransitivity}
    \label{corollary:equivalence-transitivity}
    Assuming the types are well-formed in $[[Γ]]$,
    \begin{itemize}
        \item[$-$] if $[[Γ ⊢ iN1 ≈ iN2]]$ and $[[Γ ⊢ iN2 ≈ iN3]]$ then $[[Γ ⊢ iN1 ≈ iN3]]$,
        \item[$+$] if $[[Γ ⊢ iP1 ≈ iP2]]$ and $[[Γ ⊢ iP2 ≈ iP3]]$ then $[[Γ ⊢ iP1 ≈ iP3]]$.
    \end{itemize}
\end{restatable}

### 2023-11-13 proofs/tex/decl-typing-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Restructure proofs

@@
-\begin{lemma} \label{lemma:app-inf-equ-stable}
-    If $[[Γ; Φ ⊢ iN1 ● args ⇒> iM]]$ and $[[Γ ⊢ iN1 ≈ iN2]]$ 
-    then $[[Γ; Φ ⊢ iN2 ● args ⇒> iM]]$.
-\end{lemma}
+\begin{definition}[Number of prenex quantifiers]
+    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
+    \begin{itemize}
+        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
+        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
+    \end{itemize}
+\end{definition}
+
+\begin{definition}[Size of a Declarative Judgement]
+    \label{def:decl-typing-size}
+    For a declarative typing judgment $J$
+    let us define a metrics $\size{J}$ as a pair of numbers 
+    in the following way:
+    \begin{itemize}
+        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
+            (\size{[[args]]}, \npq{[[iN]]})$)
+    \end{itemize}
+    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
+    syntax tree of the term $[[v]]$ or $[[c]]$
+    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
+\end{definition}
+
+\begin{definition}[Number of Equivalence Nodes]
+    For a tree $T$ inferring
+    a declarative typing judgment,
+    let us a function $\eqNodes{T}$
+    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
+    \ruleref{\ottdruleDTNEquivLabel}.
+\end{definition}
+
+\begin{definition}[Metric]
+    \label{def:decl-typing-metric}
+    For a tree $T$ inferring
+    a declarative typing judgment $J$,
+    let us define a metric $\metric{T}$
+    as a pair $(\size{J}, \eqNodes{T})$.
+\end{definition}
+
+\lemmaAppInfEquStable*
 \begin{proof}
@@
                 \item $[[Γ ⊢ iQ1 ≥ iP]]$,
-                    and then by transitivity \cref{corollary:subtyping-transitivity},
+                    and then by transitivity \cref{lemma:subtyping-transitivity},
                     $[[Γ ⊢ iQ2 ≥ iP]]$;
@@
 
-\newcommand{\pureSize}[1]{\ensuremath{\mathsf{pure\_size}(#1)}}
-\newcommand{\metric}[1]{\ensuremath{\mathsf{metric}(#1)}}
-\newcommand{\eqNodes}[1]{\ensuremath{\mathsf{eq\_nodes}(#1)}}
-\newcommand{\npq}[1]{\ensuremath{\mathsf{npq}(#1)}}
-
-\begin{definition}[Number of prenex quantifiers]
-    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
-    \begin{itemize}
-        \item [$+$] $\npq{[[∃nas.iP]]} = |nas|$, if $[[iP ≠ ∃nbs.iP']]$,
-        \item [$-$] $\npq{[[∀pas.iN]]} = |pas|$, if $[[iN ≠ ∀pbs.iN']]$.
-    \end{itemize}
-\end{definition}
-
-\begin{definition}[Size of a Declarative Judgement]
-    \label{def:decl-typing-size}
-    For a declarative typing judgement $J$
-    let us define a metrics $\size{J}$ as a pair of numbers 
-    in the following way:
-    \begin{itemize}
-        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
-        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
-        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
-            (\size{[[args]]}, \npq{[[iN]]})$)
-    \end{itemize}
-    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
-    syntax tree of the term $[[v]]$ or $[[c]]$
-    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
-\end{definition}
-
-\begin{definition}[Number of Equivalence Nodes]
-    For a tree $T$ inferring
-    a declarative typing judgement,
-    let us a function $\eqNodes{T}$
-    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
-    \ruleref{\ottdruleDTNEquivLabel}.
-\end{definition}
-
-\begin{definition}[Metric]
-    \label{def:decl-typing-metric}
-    For a tree $T$ inferring
-    a declarative typing judgement $J$,
-    let us define a metric $\metric{T}$
-    as a pair $(\size{J}, \eqNodes{T})$.
-\end{definition}
-
-\begin{lemma}[Declarative typing is preserved under context equivalence]
-    \label{lemma:decl-typing-context-equiv}
-    Assuming $[[Γ ⊢ Φ1]]$, $[[Γ ⊢ Φ2]]$, and $[[Γ ⊢ Φ1 ≈ Φ2]]$:
-    \begin{itemize}
-        \item [$+$] 
-            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ v : iP]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ v : iP]]$.
-        \item [$-$] 
-            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ c : iN]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ c : iN]]$.
-        \item [$\bullet$] 
-            for any tree $T_1$ inferring $[[Γ ; Φ1 ⊢ iN ● args ⇒> iM]]$, 
-            there exists a tree $T_2$ inferring $[[Γ ; Φ2 ⊢ iN ● args ⇒> iM]]$.
-    \end{itemize}
-\end{lemma}
+\lemmaDeclTypingContextEquiv*
 \begin{proof}

### 2023-11-13 proofs/tex/equiv-alg-lemmas.tex
Add deterinicity lemmas, get rid of (??)
Restructure proofs

@@
-\begin{lemma}[Algorithmization of declarative equivalence]
-  \label{lemma:decl-equiv-algorithmization}
-  Declarative equivalence is the equality of normal forms. 
-  \begin{itemize}
-    \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
-    \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof} \hfill
-  \begin{itemize}
-    \item[$+$] Let us prove both directions separately.
-    \begin{itemize}
-      \item[$\Rightarrow$] 
-        exactly by \cref{lemma:normalization-completeness},
-      \item[$\Leftarrow$] 
-        from \cref{lemma:normalization-soundness}, we know
-        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
-        $[[iP ≈ iQ]]$.
-    \end{itemize}
-    \item[$-$] The proof is exactly the same.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Algorithmic type well-formedness is invariant under equivalence]
-  \label{lemma:wf-equiv-algo}
-  Mutual subtyping implies declarative equivalence.
-  \begin{itemize}
-  \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
-  \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
-  \end{itemize}
-\end{lemma}
+\lemmaWfEquivAlgo*
 \begin{proof}
@@
 
-\begin{corollary}[Normalization preserves well-formedness of algorithmic types]
-  \label{corollary:wf-nf-algo}
-  \begin{itemize}
-    \item[$+$] $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ nf(uP)]]$,
-    \item[$-$] $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ nf(uN)]]$
-  \end{itemize}
-\end{corollary}
+\corollaryWfNfAlgo*
 \begin{proof}
@@
 
-\begin{corollary}[Normalization preserves the signature of the algorithmic substitution]
-  \label{corollary:norm-subst-sig-algo}
-  \hfill \\
-  $[[Θ ⊢ uσ : Ξ]] \iff [[Θ ⊢ nf(uσ) : Ξ]]$,
-  $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
-\end{corollary}
+\corollaryNormSubstSigAlgo*
 \begin{proof}
@@
 
-\begin{lemma}[Algorithmic substitution equivalence becomes equality after normalization]
-  \label{corollary:subst-subt-equiv-algorithmization}
-  Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
-  and $[[Ξ ⊆ Ξ']]$. 
-  Then $[[Θ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
-\end{lemma}
+\corollarySubstSubtEquivAlg*
 \begin{proof}

### 2023-11-13 proofs/tex/equiv-alg-statements.tex
Restructure proofs

\begin{restatable}[Algorithmic type well-formedness is invariant under equivalence]{lemma}{lemmaWfEquivAlgo}
    \label{lemma:wf-equiv-algo}
    Mutual subtyping implies declarative equivalence.
    \begin{itemize}
    \item[$+$] if $[[uP ≈ uQ]]$ then $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ uQ]]$,
    \item[$-$] if $[[uN ≈ uM]]$ then $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ uM]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves well-formedness of algorithmic types]{corollary}{corollaryWfNfAlgo}
    \label{corollary:wf-nf-algo}
    \hfill\\
    \begin{itemize}
        \item[$+$] $[[Γ; Ξ ⊢ uP]] \iff [[Γ; Ξ ⊢ nf(uP)]]$,
        \item[$-$] $[[Γ; Ξ ⊢ uN]] \iff [[Γ; Ξ ⊢ nf(uN)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves the signature of the algorithmic substitution]{corollary}{corollaryNormSubstSigAlgo}
    \label{corollary:norm-subst-sig-algo}
    \hfill\\
    $[[Θ ⊢ uσ : Ξ]] \iff [[Θ ⊢ nf(uσ) : Ξ]]$,
    $[[Γ ⊢ uσ : Ξ]] \iff [[Γ ⊢ nf(uσ) : Ξ]]$.
\end{restatable}

\begin{restatable}[Algorithmic substitution equivalence becomes equality after normalization]{corollary}{corollarySubstSubtEquivAlg}
    \label{corollary:subst-subt-equiv-algorithmization}
    Suppose that $[[Θ ⊢ uσ1 : Ξ']]$ and $[[Θ ⊢ uσ2 : Ξ']]$ are algorithmic substitutions
    and $[[Ξ ⊆ Ξ']]$. 
    Then $[[Θ ⊢ uσ1 ≈ uσ2 : Ξ]] \iff [[nf(uσ1)|Ξ = nf(uσ2)|Ξ]]$.
\end{restatable}

### 2023-11-13 proofs/tex/equiv-lemmas.tex
wip
Restructure proofs

@@
-\begin{lemma} [Declarative Equivalence is invariant under bijections]
-  \label{lemma:decl-equiv-bijection}
-  Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
-  \begin{itemize}
-    \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
-      and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
-      with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
-    \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
-      and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
-      with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaDeclEquivBijection*
 \begin{proof}
@@
 
-\begin{lemma}[Declarative equivalence is transitive]
-  \hfill
-  \label{lemma:decl-equiv-transitivity}
-  \begin{itemize}
-    \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
-    \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaEquivFV*
+\begin{proof}
+  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
+  The base cases 
+  (\ruleref{\ottdruleEOneNVarLabel} and \ruleref{\ottdruleEOnePVarLabel})
+  are trivial.
+  So are \ruleref{\ottdruleEOneShiftULabel}, 
+  \ruleref{\ottdruleEOneShiftDLabel}, and
+  \ruleref{\ottdruleEOneArrowLabel}, where
+  the required property follows from the induction hypothesis.
+
+  Let us consider the case when the equivalence is formed by
+  \ruleref{\ottdruleEOneForallLabel}, 
+  that is the equivalence has a shape 
+  $[[∀pas.iN ≈ ∀pbs.iM]]$, 
+  and by inversion,
+  there is a bijection 
+  $[[mu : ({pbs} ∩ fv iM) ↔ ({pas} ∩ fv iN)]]$
+  such that $[[iN ≈ [mu] iM]]$, which 
+  by the induction hypothesis means
+  $[[ fv iN = fv [mu]iM ]] = [[ [mu] fv iM]] $.
+
+  Let us ensure by alpha-equivalence that 
+  $[[pas]]$ is disjoint from $[[fv iM]]$.
+  Then $[[ (fv ∀pbs.iM) \ {pas}]] = [[fv ∀pbs.iM]]$.
+  Then we apply the following chain of equalities:
+  $[[fv ∀pas.iN = fv iN \ {pas}]] = 
+  [[ ([mu] fv iM) \ {pas}]] = 
+  [[ [mu] (fv ∀pbs.iM ∪ ({pbs} ∩ fv iM)) \ {pas}]] = 
+  [[ ([mu] fv ∀pbs.iM ∪ [mu]({pbs} ∩ fv iM)) \ {pas}]] = 
+  [[ ([mu] fv ∀pbs.iM) \ {pas}]] = 
+  [[ (fv ∀pbs.iM) \ {pas}]] = 
+  [[  fv ∀pbs.iM ]]$.
+
+  Symmetrically, we prove the case when the equivalence is formed by
+  \ruleref{\ottdruleEOneExistsLabel}.
+\end{proof}
+
+\lemmaDeclEquivTransitivity*
 \begin{proof}
@@
 
-\begin{lemma}[Type well-formedness is invariant under equivalence]
-  \label{lemma:wf-equiv}
-  Mutual subtyping implies declarative equivalence.
-  \begin{itemize}
-  \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
-  \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
-  \end{itemize}
-\end{lemma}
+\lemmaWfEquiv*
 \begin{proof}
@@
       and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
+
       $
@@
 
-\begin{corollary}[Normalization preserves well-formedness]
-  \label{corollary:wf-nf}
-  \hfill
-  \begin{itemize}
-  \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
-  \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
-\end{proof}
-
-\begin{corollary}[Normalization preserves well-formedness of substitution]
-  \label{corollary:wf-s-nf}
-  \hfill \\
-   $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
-\end{corollary}
-\begin{proof}
-  Let us prove the forward direction.
-  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
-  By the definition of substitution normalization,
-  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf},
-  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
-  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
-
-  The backward direction is proved analogously.
-\end{proof}
-
-\begin{lemma}[Normalization preserves substitution signature]
-  \label{lemma:norm-subst-sig}
-  Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
-  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
-\end{lemma}
-\begin{proof}
-  Suppose that $[[α± ∊ Γ1]]$. 
-  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
-  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.
-
-  Suppose that $[[α± ∉ Γ1]]$. 
-  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
-  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
-\end{proof}
-
-\begin{lemma}[Soundness of equivalence]
-  \label{lemma:equiv-soundness}
-  Declarative equivalence implies mutual subtyping.
-  \begin{itemize}
-    \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
-    \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaEquivSoundness*
 \begin{proof}
@@
 
-\begin{corollary}[Normalization is sound w.r.t. subtyping-induced equivalence] 
-  \label{corollary:nf-sound-wrt-subt-equiv}
-  \hfill
-  \begin{itemize}
-    \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
-    \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.
-\end{proof}
-
-\begin{corollary}[Normalization preserves subtyping] 
-  \label{corollary:nf-pres-subt}
-  Assuming all the types are well-formed in context $[[Γ]]$,
-  \begin{itemize}
-    \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
-    \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  \hfill
-  \begin{itemize}
-    \item [$+$]  
-    \begin{itemize}
-      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
-        By \cref{corollary:nf-sound-wrt-subt-equiv},
-        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
-        in particular, by inversion, 
-        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
-        Then by transitivity of subtyping 
-        (\cref{lemma:subtyping-transitivity}), 
-        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
-      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
-        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
-        and inversion, 
-        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
-        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
-    \end{itemize}
-    \item [$-$] The negative case is proved symmetrically.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Subtyping induced by disjoint substitutions]
-  \label{lemma:subt-ind-disj-subst}
-  If two disjoint substitutions induce subtyping, they are degenerate (so is the
-  subtyping).
-  Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
-  where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
-  \begin{itemize}
-  \item[$-$] assuming $[[Γ ⊢ iN]]$,~
-    $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id : fv iN]]$
-  \item[$+$] assuming $[[Γ ⊢ iP]]$,~
-    $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id : fv iP]]$
-  \end{itemize}
-\end{lemma}
+\lemmaSubtIndDisjSubst*
 \begin{proof}
@@
 
-\begin{corollary}[Substitution cannot induce proper subtypes or supertypes] \label{corollary:subst-proper-subt}
-  Assuming all mentioned types are well-formed in $[[Γ]]$ and $[[σ]]$ is a
-  substitution $[[Γ ⊢ σ : Γ]]$,
-  \begin{align*}
-    [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
-    [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
-    [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
-    [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
-                          \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
-  \end{align*}
-\end{corollary}
+\corollarySubstProperSubt*
 
-\begin{lemma} \label{lemma:mutual-subst-subtyping}
-  Assuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
-  are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
-  \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
-    then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
-  \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
-    then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
-    $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaMutualSubstSubtyping*
 \begin{proof}
@@
 
-\begin{lemma}[Equivalent substitution act equivalently]
-  \label{lemma:equiv-subst-on-same-term}
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$ and $[[Γ' ⊢ σ2 : Γ]]$
-  are substitutions equivalent on their domain, that is $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
-  Then
-  \begin{itemize}
-    \item[$+$] for any $[[Γ ⊢ iP]]$, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$;
-    \item[$-$] for any $[[Γ ⊢ iN]]$, $[[Γ' ⊢ [σ1]iN ≈ [σ2]iN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaEquivSubstOnSameTerm*
 \begin{proof}
@@
 
-\begin{lemma}[Equivalence of polymorphic types]
-  \label{lemma:poly-types-equivalence}
-  \hfill
-  \begin{itemize}
-    \item[$-$] For $[[Γ ⊢ ∀pas.iN]]$ and $[[Γ ⊢ ∀pbs.iM]]$,\\ if $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$
-    then there exists a bijection $[[μ : {pbs} ∩ fv iM ↔ {pas} ∩ fv iN]]$
-    such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iM ]]$,
-    \item[$+$] For $[[Γ ⊢ ∃nas.iP]]$ and $[[Γ ⊢ ∃nbs.iQ]]$,\\  if $[[Γ ⊢ ∃nas.iP ≈ ∃nbs.iQ ]]$
-    then there exists a bijection $[[μ : {nbs} ∩ fv iQ ↔ {nas} ∩ fv iP]]$
-    such that $[[ Γ, nbs ⊢ iP ≈ [Sub μ] iQ ]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaPolyTypesEquivalence*
 \begin{proof}
@@
 
-\begin{lemma}[Completeness of equivalence] \label{lemma:equiv-completeness}
-  Mutual subtyping implies declarative equivalence.
-  Assuming all the types below are well-formed in $[[Γ]]$: 
-  \begin{itemize}
-  \item[$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[iP ≈ iQ]]$,
-  \item[$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[iN ≈ iM]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaEquivCompleteness*
 \begin{proof}
@@
 \end{proof}
-
-\begin{corollary}[Normalization is complete w.r.t. subtyping-induced equivalence]
-  \label{corollary:nf-complete-wrt-subt-equiv}
-  Assuming all the types below are well-formed in $[[Γ]]$:
-  \begin{itemize}
-    \item [$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[nf(iP) = nf(iQ)]]$,
-    \item [$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{corollary}  
-\begin{proof}
-  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
-\end{proof}
-
-\begin{lemma}[Algorithmization of subtyping-induced equivalence]
-  \label{lemma:subt-equiv-algorithmization}
-  Mutual subtyping is the equality of normal forms.
- Assuming all the types below are well-formed in $[[Γ]]$:
-  \begin{itemize}
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-  Let us prove the positive case, the negative case is symmetric.
-  We prove both directions of $\iff$ separately:
-  \begin{itemize}
-    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
-    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
-  \end{itemize}
-\end{proof}
-
-\begin{lemma}[Substitution preserves declarative equivalence]
-  \label{corollary:subst-pres-decl-equiv}
-  Suppose that $[[σ]]$ is a substitution. Then
-    \begin{itemize}
-      \item[$+$] $[[iP ≈ iQ]]$ implies $[[ [σ]iP ≈ [σ]iQ ]]$
-      \item[$-$] $[[iN ≈ iM]]$ implies $[[ [σ]iN ≈ [σ]iM ]]$
-  \end{itemize}
-\end{lemma}
-\begin{proof}
-
-  $
-  \begin{aligned}[t] 
-    [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
-                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
-                  &\Rightarrow [[ [nf(σ)]nf(iP) = [nf(σ)]nf(iQ)]]\\
-                  &\Rightarrow [[ nf([σ]iP) = nf([σ]iQ)]]
-                  && \text{by \cref{lemma:norm-subst-distr}}\\ 
-                  &\Rightarrow        [[ [σ]iP ≈ [σ]iQ ]]
-                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
-  \end{aligned} 
-  $
-\end{proof}

### 2023-11-13 proofs/tex/equiv-statements.tex
Restructure proofs

\begin{restatable}[Declarative Equivalence is invariant under bijections]{lemma}{lemmaDeclEquivBijection}
    \label{lemma:decl-equiv-bijection}
    Suppose $[[μ]]$ is a bijection $[[μ : varset1 ↔ varset2]]$, then
    \begin{itemize}
        \item[$+$] $[[iP1 ≈ iP2]]$ implies $[[ [μ]iP1 ≈ [μ]iP2 ]]$,
            and there exists an inference tree of $[[ [μ]iP1 ≈ [μ]iP2 ]]$
            with the same shape as the one inferring $[[iP1 ≈ iP2]]$;
        \item[$-$] $[[iN1 ≈ iN2]]$ implies $[[ [μ]iN1 ≈ [μ]iN2 ]]$, 
            and there exists an inference tree of $[[ [μ]iN1 ≈ [μ]iN2 ]]$
            with the same shape as the one inferring $[[iN1 ≈ iN2]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}{lemma}{lemmaEquivFV}
    \label{lemma:equiv-fv}
    The set of free variables is invariant under equivalence.
    \begin{itemize}
    \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN = fv iM]]$ (as sets)
    \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP = fv iQ]]$ (as sets)
    \end{itemize}
\end{restatable}

\begin{restatable}[Declarative equivalence is transitive]{lemma}{lemmaDeclEquivTransitivity}
    \hfill
    \label{lemma:decl-equiv-transitivity}
    \begin{itemize}
        \item[$+$] if $[[iP1 ≈ iP2]]$ and $[[iP2 ≈ iP3]]$ then $[[iP1 ≈ iP3]]$,
        \item[$-$] if $[[iN1 ≈ iN2]]$ and $[[iN2 ≈ iN3]]$ then $[[iN1 ≈ iN3]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Type well-formedness is invariant under equivalence]{lemma}{lemmaWfEquiv}
    \label{lemma:wf-equiv}
    Mutual subtyping implies declarative equivalence.
    \begin{itemize}
    \item[$+$] if $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$,
    \item[$-$] if $[[iN ≈ iM]]$ then $[[Γ ⊢ iN]] \iff [[Γ ⊢ iM]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Soundness of equivalence]{lemma}{lemmaEquivSoundness}
    \label{lemma:equiv-soundness}
    Declarative equivalence implies mutual subtyping.
    \begin{itemize}
        \item[$+$] if $[[Γ ⊢ iP]]$, $[[Γ ⊢ iQ]]$, and $[[iP ≈ iQ]]$ then $[[Γ ⊢ iP ≈ iQ]]$,
        \item[$-$] if $[[Γ ⊢ iN]]$, $[[Γ ⊢ iM]]$, and $[[iN ≈ iM]]$ then $[[Γ ⊢ iN ≈ iM]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Subtyping induced by disjoint substitutions]{lemma}{lemmaSubtIndDisjSubst}
    \label{lemma:subt-ind-disj-subst}
    Suppose that $[[Γ ⊢ σ1 : Γ1]]$ and $[[Γ ⊢ σ2 : Γ1]]$,
    where $[[Γi ⊆ Γ]]$ and $[[Γ1 ∩ Γ2= ∅]]$. Then
    \begin{itemize}
    \item[$-$] assuming $[[Γ ⊢ iN]]$,~
        $[[Γ ⊢ [σ1]iN ≤ [σ2]iN]]$ implies $[[Γ ⊢ σi ≈ id : fv iN]]$
    \item[$+$] assuming $[[Γ ⊢ iP]]$,~
        $[[Γ ⊢ [σ1]iP ≥ [σ2]iP]]$ implies $[[Γ ⊢ σi ≈ id : fv iP]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Substitution cannot induce proper subtypes or supertypes]{corollary}{corollarySubstProperSubt}
    \label{corollary:subst-proper-subt}
    Assuming all mentioned types are well-formed in $[[Γ]]$ and $[[σ]]$ is a
    substitution $[[Γ ⊢ σ : Γ]]$,
    \begin{align*}
        [[Γ ⊢ [σ]iN ≤ iN]] ~&\Rightarrow~ [[Γ ⊢ [σ]iN ≈ iN]]
                                                    \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
        [[Γ ⊢ iN ≤ [σ]iN]] ~&\Rightarrow~ [[Γ ⊢ iN ≈ [σ]iN]]
                                                    \text{ and } [[Γ ⊢ σ ≈ id :  fv iN]] \\
        [[Γ ⊢ [σ]iP ≥ iP]] ~&\Rightarrow~ [[Γ ⊢ [σ]iP ≈ iP]]
                                                    \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
        [[Γ ⊢ iP ≥ [σ]iP]] ~&\Rightarrow~ [[Γ ⊢ iP ≈ [σ]iP]]
                                                    \text{ and } [[Γ ⊢ σ ≈ id :  fv iP]] \\
    \end{align*}
\end{restatable}

\begin{restatable}[Mutual substitution and subtyping]{lemma}{lemmaMutualSubstSubtyping}
    \label{lemma:mutual-subst-subtyping}
    Assuming that the mentioned types ($[[iP]]$, $[[iQ]]$, $[[iN]]$, and $[[iM]]$)
    are well-formed in $[[Γ]]$ and that the substitutions ($[[σ1]]$ and $[[σ2]]$) have signature $[[Γ ⊢ σi : Γ]]$,
    \begin{itemize}
    \item[$+$] if $[[Γ ⊢ [σ1] iP ≥ iQ]]$ and $[[Γ ⊢ [σ2] iQ ≥ iP]]$\\
        then there exists a bijection $[[μ : fv iP ↔ fv iQ]]$ such that
        $[[Γ ⊢ σ1 ≈ Sub μ :  fv iP]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iQ]]$;
    \item[$-$] if $[[Γ ⊢ [σ1] iN ≤ iM]]$ and $[[Γ ⊢ [σ2] iN ≤ iM]]$\\
        then there exists a bijection $[[μ : fv iN ↔ fv iM]]$ such that
        $[[Γ ⊢ σ1 ≈ Sub μ :  fv iN]]$ and $[[Γ ⊢ σ2 ≈ Sub μ-1 :  fv iM]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Equivalent substitution act equivalently]{lemma}{lemmaEquivSubstOnSameTerm}
    \label{lemma:equiv-subst-on-same-term}
    Suppose that $[[Γ' ⊢ σ1 : Γ]]$ and $[[Γ' ⊢ σ2 : Γ]]$
    are substitutions equivalent on their domain, that is $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$.
    Then
    \begin{itemize}
        \item[$+$] for any $[[Γ ⊢ iP]]$, $[[Γ' ⊢ [σ1]iP ≈ [σ2]iP]]$;
        \item[$-$] for any $[[Γ ⊢ iN]]$, $[[Γ' ⊢ [σ1]iN ≈ [σ2]iN]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Equivalence of polymorphic types]{lemma}{lemmaPolyTypesEquivalence}
    \label{lemma:poly-types-equivalence}
    \hfill
    \begin{itemize}
        \item[$-$] For $[[Γ ⊢ ∀pas.iN]]$ and $[[Γ ⊢ ∀pbs.iM]]$,\\ if $[[Γ ⊢ ∀pas.iN ≈ ∀pbs.iM ]]$
        then there exists a bijection $[[μ : {pbs} ∩ fv iM ↔ {pas} ∩ fv iN]]$
        such that $[[ Γ, pas ⊢ iN ≈ [Sub μ] iM ]]$,
        \item[$+$] For $[[Γ ⊢ ∃nas.iP]]$ and $[[Γ ⊢ ∃nbs.iQ]]$,\\  if $[[Γ ⊢ ∃nas.iP ≈ ∃nbs.iQ ]]$
        then there exists a bijection $[[μ : {nbs} ∩ fv iQ ↔ {nas} ∩ fv iP]]$
        such that $[[ Γ, nbs ⊢ iP ≈ [Sub μ] iQ ]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of Equivalence]{lemma}{lemmaEquivCompleteness} 
    \label{lemma:equiv-completeness}
    Mutual subtyping implies declarative equivalence.
    Assuming all the types below are well-formed in $[[Γ]]$: 
    \begin{itemize}
    \item[$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[iP ≈ iQ]]$,
    \item[$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[iN ≈ iM]]$.
    \end{itemize}
\end{restatable}

### 2023-11-13 proofs/tex/norm-lemmas.tex
Restructure proofs

--- 2023-09-05 proofs/tex/norm-lemmas.tex
@@
-\begin{lemma}
-  \label{lemma:equiv-fv}
-  Set of free variables is invariant under equivalence.
-  \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[fv iN]] \equiv [[fv iM]]$ (as sets)
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[fv iP]] \equiv [[fv iQ]]$ (as sets)
-  \end{itemize}
-\end{lemma}
+\obsNormDeterministic*
 \begin{proof}
-  Straightforward mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$
+  By straightforward induction using \cref{obs:ord-deterministic}.
 \end{proof}
 
-\begin{lemma}
-  \label{lemma:fv-nf}
-  Free variables are not changed by the normalization
-  \begin{itemize}
-  \item[$-$] $[[fv iN]] \equiv [[fv nf(iN)]]$
-  \item[$+$] $[[fv iP]] \equiv [[fv nf(iP)]]$
-  \end{itemize}
-\end{lemma}
+\lemmaFvNf*
 \begin{proof}
-  By straightforward induction on $[[iN]]$ and mutually on $[[iP]]$.
+  By mutual induction on $[[iN]]$ and $[[iP]]$.
+  The base cases 
+  (\ruleref{\ottdruleNrmNVarLabel} and \ruleref{\ottdruleNrmPVarLabel})
+  are trivial; the congruent cases
+  (\ruleref{\ottdruleNrmShiftULabel},
+  \ruleref{\ottdruleNrmShiftDLabel}, and
+  \ruleref{\ottdruleNrmArrowLabel}) are proved by the induction hypothesis.
+
+  Let us consider the case when the term is formed by $[[∀]]$,
+  that is the normalization judgment has a shape 
+  $[[nf(∀pas.iN) = ∀pas'.iN']]$,
+  where by inversion $[[nf(iN) = iN']]$
+  and $[[ord {pas} in iN' = pas']]$.
+  By the induction hypothesis, $[[fv iN = fv iN']]$.
+  Since $[[fv(∀pas.iN) = fv iN \ {pas}]]$,
+  and $[[fv(∀pas'.iN') = fv iN' \ {pas'}]]$,
+  it is left to show that $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
+  By \cref{lemma:ord-completeness}, 
+  $[[{pas'}]] = [[{pas} ∩ fv iN']] = [[{pas} ∩ fv iN]]$.
+  Then
+  $[[fv iN \ {pas} = fv iN \ ({pas} ∪ fv iN)]]$ by
+  set-theoretic properties, and
+  thus, $[[fv iN \ {pas} = fv iN \ {pas'}]]$.
+
+  The case when the term is positive and formed by $[[∃]]$ is symmetric.
 \end{proof}
 
-\begin{lemma}[Soundness of normalization]
-  \label{lemma:normalization-soundness}
-  \hfill
-  \begin{itemize}
-    \item[$-$] $[[iN ≈ nf(iN)]]$
-    \item[$+$] $[[iP ≈ nf(iP)]]$
-  \end{itemize}
-\end{lemma}
+\lemmaNormSoundness*
 \begin{proof}
@@
 
-\begin{corollary}[Normalization preserves ordering]
-  \label{corollary:normalization-ord}
-  For any $[[varset]]$,
+\corollaryWfNf*
+\begin{proof}
+  Immediately from \cref{lemma:wf-equiv,lemma:normalization-soundness}.
+\end{proof}
+
+\corollaryWfSNf*
+\begin{proof}
+  Let us prove the forward direction.
+  Suppose that $[[α± ∊ Γ1]]$.  Let us show that $[[Γ2 ⊢ [nf(σ)]α±]]$.
+  By the definition of substitution normalization,
+  $[[ [nf(σ)]α± = nf([σ]α±) ]]$. Then by \cref{corollary:wf-nf},
+  to show $[[Γ2 ⊢ nf([σ]α±)]]$, it suffices to show $[[Γ2 ⊢ [σ]α±]]$,
+  which holds by the assumption $[[Γ2 ⊢ σ : Γ1]]$.
+
+  The backward direction is proved analogously.
+\end{proof}
+
+\lemmaNormSubstSig*
+\begin{proof}
+  Suppose that $[[α± ∊ Γ1]]$. 
+  Then by \cref{corollary:wf-nf}, $[[Γ2 ⊢ nf([σ]α±)]] = [[ [nf(σ)]α± ]]$ 
+  is equivalent to $[[Γ2 ⊢ [σ]α±]]$.
+
+  Suppose that $[[α± ∉ Γ1]]$. 
+  $[[Γ2 ⊢ σ : Γ1]]$ means that $[[ [σ]α± = α± ]]$, 
+  and then $[[ [nf(σ)]α± ]] = [[nf([σ]α±)]] = [[nf(α±)]] = [[α±]]$.
+\end{proof}
+
+\corollaryNfSoundWrtSubtEquiv*
+\begin{proof}
+  Immediately from \cref{lemma:normalization-soundness,corollary:wf-nf,lemma:equiv-soundness}.  
+\end{proof}
+
+\corollaryNfPresSubt*
+\begin{proof}
+  \hfill
   \begin{itemize}
-  \item[$-$] $[[ord varset in nf(iN)]] = [[ord varset in iM]]$
-  \item[$+$] $[[ord varset in nf(iP)]] = [[ord varset in iQ]]$
+    \item [$+$]  
+    \begin{itemize}
+      \item [$\Rightarrow$] Let us assume $[[Γ ⊢ iP ≥ iQ]]$.
+        By \cref{corollary:nf-sound-wrt-subt-equiv},
+        $[[Γ ⊢ iP ≈ nf(iP)]]$ and $[[Γ ⊢ iQ ≈ nf(iQ)]]$, 
+        in particular, by inversion, 
+        $[[Γ ⊢ nf(iP) ≥ iP]]$ and $[[Γ ⊢ iQ ≥ nf(iQ)]]$.
+        Then by transitivity of subtyping 
+        (\cref{lemma:subtyping-transitivity}), 
+        $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
+      \item [$\Leftarrow$] Let us assume $[[Γ ⊢ nf(iP) ≥ nf(iQ)]]$.
+        Also by \cref{corollary:nf-sound-wrt-subt-equiv}
+        and inversion, 
+        $[[Γ ⊢ iP ≥ nf(iP)]]$ and $[[Γ ⊢ nf(iQ) ≥ iQ]]$.
+        Then by the transitivity, $[[Γ ⊢ iP ≥ iQ]]$.
     \end{itemize}
-\end{corollary}
+    \item [$-$] The negative case is proved symmetrically.
+  \end{itemize}
+\end{proof}
+
+\corollaryNormPreservesOrdering*
 \begin{proof}
@@
 
-\begin{lemma}[Distributivity of normalization over substitution]
-  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
-  Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
-    \begin{itemize}
-      \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
-      \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
-  \end{itemize}
-  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
-\end{lemma}
+\lemmaNormSubstDistr*
 \begin{proof}
@@
 
-     To show alpha-equivalence of 
+     To show the alpha-equivalence of 
      $[[ [nf(σ)] ∀pas'.nf(iM) ]]$ and $[[ ∀pas'.[nf(σ)]nf(iM) ]]$,
@@
 
-\begin{corollary}[Commutativity of normalization and renaming]
-  \label{lemma:norm-subst-commute} Normalization of a term commutes with renaming.
-  Suppose that $\mu$ is a bijection between two sets of variables
-  $\mu : A \leftrightarrow B$. Then
-  \begin{itemize}
-    \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
-    \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
-  \end{itemize}
-\end{corollary}
+\lemmaNormSubstCommute*
 \begin{proof}
-  Immediately from \cref{lemma:norm-subst-distr},
-  after noticing that $[[nf(mu)]] = [[mu]]$.
+  Immediately from \cref{lemma:norm-subst-distr}, after noticing that $[[nf(mu)]] = [[mu]]$.
 \end{proof}
 
-\begin{lemma}[Completeness of quantified normalization]
-  \label{lemma:normalization-completeness}
-  Normalization returns the same representative for equivalent types.
-
-  \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$
-  \end{itemize}
-\end{lemma}
-
+\lemmaNormalizationCompleteness*
 \begin{proof}
@@
 
-\begin{lemma}[Idempotence of normalization]
-  \label{lemma:norm-idemp}
-  Normalization is idempotent
+\lemmaDeclEquivAlg*
+\begin{proof} \hfill
   \begin{itemize}
-  \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
-  \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
+    \item[$+$] Let us prove both directions separately.
+    \begin{itemize}
+      \item[$\Rightarrow$] 
+        exactly by \cref{lemma:normalization-completeness},
+      \item[$\Leftarrow$] 
+        from \cref{lemma:normalization-soundness}, we know
+        $[[iP ≈ nf(iP)]] = [[nf(iQ) ≈ iQ]]$, then by transitivity (\cref{lemma:decl-equiv-transitivity}),
+        $[[iP ≈ iQ]]$.
     \end{itemize}
-\end{lemma}
+    \item[$-$] For the negative case, the proof is the same.
+  \end{itemize}
+\end{proof}
+
+\corollaryNfCompleteWrtSubtEquiv*
+\begin{proof}
+  Immediately from \cref{lemma:equiv-completeness,lemma:normalization-completeness}.
+\end{proof}
+
+\lemmaNormIdemp*
 \begin{proof}
@@
 
-\begin{lemma}
-  \label{lemma:normal-after-subst}
-  The result of a substitution is normalized if and only if the initial type and
-  the substitution are normalized.
-
-  Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
-  $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
-  $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
-  \begin{itemize}
-  \item[$+$]
-    $[[ [σ]iP  ]] \text{ is normal} \iff
-    \begin{cases}
-      [[σ|fv(iP)]] &\text{is normal}\\
-      [[iP]]       &\text{is normal}\\
-    \end{cases} $
-  \item[$-$]
-    $[[ [σ]iN  ]] \text{is normal} \iff
-    \begin{cases}
-      [[σ|fv(iN)]] &\text{is normal}\\
-      [[iN]]       &\text{is normal}\\
-    \end{cases} $
-  \end{itemize}
-\end{lemma}
+\lemmaNormalAfterSubst*
 \begin{proof}
@@
 \end{proof}
+
+\lemmaSubtInducedEquivAlg*
+\begin{proof}
+  Let us prove the positive case, the negative case is symmetric.
+  We prove both directions of $\iff$ separately:
+  \begin{itemize}
+    \item [$\Rightarrow$] exactly \cref{corollary:nf-complete-wrt-subt-equiv};
+    \item [$\Leftarrow$] by \cref{lemma:decl-equiv-algorithmization,lemma:equiv-soundness}.
+  \end{itemize}
+\end{proof}
+
+\corSubstPresDeclEquiv*
+\begin{proof}
+  $
+  \begin{aligned}[t] 
+    [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
+                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
+                  &\Rightarrow [[ [nf(σ)]nf(iP) = [nf(σ)]nf(iQ)]]\\
+                  &\Rightarrow [[ nf([σ]iP) = nf([σ]iQ)]]
+                  && \text{by \cref{lemma:norm-subst-distr}}\\ 
+                  &\Rightarrow        [[ [σ]iP ≈ [σ]iQ ]]
+                  && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
+  \end{aligned} 
+  $
+\end{proof}

### 2023-11-13 proofs/tex/norm-statements.tex
Restructure proofs

\begin{restatable}[Normalization is deterministic]{observation}{obsNormDeterministic}
    \label{obs:norm-deterministic}
    If $[[nf(iN) = iM]]$ and $[[nf(iN) = iM']]$ then $[[iM = iM']]$.
    If $[[nf(iP) = iQ]]$ and $[[nf(iP) = iQ']]$ then $[[iQ = iQ']]$.
    This way, we can use normalization as a function.
\end{restatable}

\begin{restatable}{lemma}{lemmaFvNf}
    \label{lemma:fv-nf}
    Free variables are not changed by the normalization
    \begin{itemize}
    \item[$-$] $[[fv iN = fv nf(iN)]]$
    \item[$+$] $[[fv iP = fv nf(iP)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Soundness of normalization]{lemma}{lemmaNormSoundness}
    \label{lemma:normalization-soundness}
    \hfill
    \begin{itemize}
        \item[$-$] $[[iN ≈ nf(iN)]]$
        \item[$+$] $[[iP ≈ nf(iP)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves well-formedness]{corollary}{corollaryWfNf}
    \label{corollary:wf-nf}
    \hfill
    \begin{itemize}
    \item[$+$] $[[Γ ⊢ iP]] \iff [[Γ ⊢ nf(iP)]]$,
    \item[$-$] $[[Γ ⊢ iN]] \iff [[Γ ⊢ nf(iN)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves well-formedness of substitution]{corollary}{corollaryWfSNf}
    \label{corollary:wf-s-nf}
    \hfill \\
     $[[Γ2 ⊢ σ : Γ1]] \iff [[Γ2 ⊢ nf(σ) : Γ1]]$
\end{restatable}

\begin{restatable}[Normalization preserves substitution signature]{lemma}{lemmaNormSubstSig}
    \label{lemma:norm-subst-sig}
    Suppose that $[[σ]]$ is a substitution, $[[Γ1]]$ and $[[Γ2]]$ are contexts.
    Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ2 ⊢ nf(σ) : Γ1]]$.
\end{restatable}

\begin{restatable}[Normalization is sound w.r.t. subtyping-induced equivalence]{corollary}{corollaryNfSoundWrtSubtEquiv}
    \label{corollary:nf-sound-wrt-subt-equiv}
    \hfill
    \begin{itemize}
        \item [$+$] if $[[Γ ⊢ iP]]$ then $[[Γ ⊢ iP ≈ nf(iP)]]$,
        \item [$-$] if $[[Γ ⊢ iN]]$ then $[[Γ ⊢ iN ≈ nf(iN)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves subtyping]{corollary}{corollaryNfPresSubt}
    \label{corollary:nf-pres-subt}
    Assuming all the types are well-formed in context $[[Γ]]$,
    \begin{itemize}
        \item [$+$] $[[Γ ⊢ iP ≥ iQ]] \iff [[Γ ⊢ nf(iP) ≥ nf(iQ)]]$,
        \item [$-$] $[[Γ ⊢ iN ≤ iM]] \iff [[Γ ⊢ nf(iN) ≤ nf(iM)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Normalization preserves ordering]{corollary}{corollaryNormPreservesOrdering}
    \label{corollary:normalization-ord}
    For any $[[varset]]$,
    \begin{itemize}
    \item[$-$] $[[ord varset in nf(iN)]] = [[ord varset in iM]]$
    \item[$+$] $[[ord varset in nf(iP)]] = [[ord varset in iQ]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Distributivity of normalization over substitution]{lemma}{lemmaNormSubstDistr}
    \label{lemma:norm-subst-distr} 
    Normalization of a term distributes over substitution.
    Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
        \begin{itemize}
            \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
            \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
    \end{itemize}
    where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
\end{restatable}

\begin{restatable}[Commutativity of normalization and renaming]{corollary}{lemmaNormSubstCommute}
    \label{lemma:norm-subst-commute} 
    Normalization of a term commutes with renaming.
    Suppose that $\mu$ is a bijection between two sets of variables
    $\mu : A \leftrightarrow B$. Then
    \begin{itemize}
        \item[$-$] $[[nf([mu]iN)]] = [[ [mu] nf(iN) ]]$
        \item[$+$] $[[nf([mu]iP)]] = [[ [mu] nf(iP) ]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of Normalization \wrt Declarative Equivalence]{lemma}{lemmaNormalizationCompleteness}
    \label{lemma:normalization-completeness}
    Normalization returns the same representative for equivalent types.

    \begin{itemize}
        \item[$-$] If $[[iN ≈ iM]]$ then $[[nf(iN)]] = [[nf(iM)]]$,
        \item[$+$] if $[[iP ≈ iQ]]$ then $[[nf(iP)]] = [[nf(iQ)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Algorithmization of Declarative Equivalence]{lemma}{lemmaDeclEquivAlg}
    \label{lemma:decl-equiv-algorithmization}
    Declarative equivalence is the equality of normal forms. 
    \begin{itemize}
        \item[$+$] $[[iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
        \item[$-$] $[[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of Normalization \wrt Subtyping-Induced Equivalence]{corollary}{corollaryNfCompleteWrtSubtEquiv}
    \label{corollary:nf-complete-wrt-subt-equiv}
    Assuming all the types below are well-formed in $[[Γ]]$:
    \begin{itemize}
        \item [$+$] if $[[Γ ⊢ iP ≈ iQ]]$ then $[[nf(iP) = nf(iQ)]]$,
        \item [$-$] if $[[Γ ⊢ iN ≈ iM]]$ then $[[nf(iN) = nf(iM)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Idempotence of normalization]{lemma}{lemmaNormIdemp}
    \label{lemma:norm-idemp}
    Normalization is idempotent
    \begin{itemize}
    \item[$-$] $[[nf(nf(iN))]] = [[nf(iN)]]$
    \item[$+$] $[[nf(nf(iP))]] = [[nf(iP)]]$
    \end{itemize}
\end{restatable}

\begin{restatable}{lemma}{lemmaNormalAfterSubst}
    \label{lemma:normal-after-subst}
    The result of a substitution is normalized if and only if the initial type and
    the substitution are normalized.

    Suppose that $[[σ]]$ is a substitution  $[[Γ2 ⊢ σ : Γ1]]$,
    $[[iP]]$ is a positive type ($[[Γ1 ⊢ iP]]$),
    $[[iN]]$ is a negative type ($[[Γ1 ⊢ iN]]$). Then
    \begin{itemize}
    \item[$+$]
        $[[ [σ]iP  ]] \text{ is normal} \iff
        \begin{cases}
            [[σ|fv(iP)]] &\text{is normal}\\
            [[iP]]       &\text{is normal}\\
        \end{cases} $
    \item[$-$]
        $[[ [σ]iN  ]] \text{is normal} \iff
        \begin{cases}
            [[σ|fv(iN)]] &\text{is normal}\\
            [[iN]]       &\text{is normal}\\
        \end{cases} $
    \end{itemize}
\end{restatable}

\begin{restatable}[Algorithmization of subtyping-induced equivalence]{lemma}{lemmaSubtInducedEquivAlg}
    \label{lemma:subt-equiv-algorithmization}
    Mutual subtyping is the equality of normal forms.
 Assuming all the types below are well-formed in $[[Γ]]$:
    \begin{itemize}
        \item [$+$] $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$,
        \item [$-$] $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Substitution preserves declarative equivalence]{corollary}{corSubstPresDeclEquiv}
    \label{corollary:subst-pres-decl-equiv}
    Suppose that $[[σ]]$ is a substitution. Then
        \begin{itemize}
            \item[$+$] $[[iP ≈ iQ]]$ implies $[[ [σ]iP ≈ [σ]iQ ]]$
            \item[$-$] $[[iN ≈ iM]]$ implies $[[ [σ]iN ≈ [σ]iM ]]$
        \end{itemize}
\end{restatable}

### 2023-11-13 proofs/tex/ord-lemmas.tex
WIP: add comments
Restructure proofs

@@
-\begin{observation}[Ordering is deterministic]
-  \label{obs:ord-deterministic}
-  If $[[ord varset in iN = ordVars1]]$ and $[[ord varset in iN = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
-  If $[[ord varset in iP = ordVars1]]$ and $[[ord varset in iP = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
-  This way, we can use $[[ord varset in iN]]$ and as a function on $[[iN]]$,
-  and $[[ord varset in iP]]$ as a function on $[[iP]]$.
-\end{observation}
+\obsOrdDeterministic*
 \begin{proof}
@@
 
-\begin{lemma}[Soundness of variable ordering]
-  \label{lemma:ord-soundness}
-  Variable ordering extracts used free variables.
-  \begin{itemize}
-    \item[$-$] $[[ {ord varset in iN} ]] = [[varset ∩ fv iN]]$ (as sets)
-    \item[$+$] $[[ {ord varset in iP} ]] = [[varset ∩ fv iP]]$ (as sets)
-  \end{itemize}
-\end{lemma}
+\lemOrdSoundness*
 \begin{proof}
@@
       which holds since $[[varset ∩ {pas} = ∅]]$.
+  \end{caseof}
 \end{proof}
 
-\begin{corollary}[Additivity of ordering]
-  \label{corollary:ord-additivity}
-  Variable ordering is additive (in terms of set union) with respect to its first argument.
-  \begin{itemize}
-    \item[$-$] $[[ {ord (varset1 ∪ varset2) in iN} 
-                = 
-                {ord varset1 in iN} ∪ {ord varset2 in iN}]]$ (as sets)
-    \item[$+$] $[[{ord (varset1 ∪ varset2) in iP}
-                =
-                {ord varset1 in iP} ∪ {ord varset2 in iP}]]$ (as sets)
+\corOrdAdditivity*
 
-  \end{itemize}
-\end{corollary}
-
-\begin{lemma}[Weakening of ordering]
-  \label{corollary:ord-weakening}
-  Only used variables matter in the first argument of the ordering,
-  \begin{itemize}
-    \item[$-$] $[[ ord (varset ∩ fv iN) in iN ]] = [[ ord varset in iN ]]$
-    \item[$+$] $[[ ord (varset ∩ fv iP) in iP ]] = [[ ord varset in iP ]]$
-  \end{itemize}
-\end{lemma}
+\lemOrdWeakening*
 \begin{proof}
@@
 
-\begin{corollary}[Idempotency of ordering]
-  \label{corollary:ord-idemp}
-  \hfill
-  \begin{itemize}
-    \item[$-$] If $[[ ord varset in iN = ordVars ]]$ then 
-      $[[ ord {ordVars} in iN = ordVars ]]$,
-    \item[$+$] If $[[ ord varset in iP = ordVars ]]$ then 
-      $[[ ord {ordVars} in iP = ordVars ]]$;
-  \end{itemize}
-\end{corollary}
+\corOrdIdemp*
 \begin{proof}
@@
 
-Next we make a set-theoretical observation
+Next, we make a set-theoretical observation
 that will be useful further.
 In general, any injective function (its image)
-distributes over set intersection.
-However, for convenience we allow the bijections
+distributes over the set intersection.
+However, for convenience, we allow the bijections
 on variables to be applied
@@
 
-\begin{definition} [Collision free bijection]
+\begin{definition} [Collision-free bijection]
   We say that a bijection $\mu : A \leftrightarrow B$ between sets of
@@
 
-\begin{lemma}[Distributivity of renaming over variable ordering]
-  \label{lemma:distr-mu-ord}
-  Suppose that $\mu$ is a bijection between two sets of variables
-  $\mu : A \leftrightarrow B$.
-
-  \begin{itemize}
-  \item[$-$]
-    If $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$ then
-    $[[ [mu] (ord varset in iN) ]] = [[ord ([mu] varset) in [mu] iN ]]$
-  \item[$+$]
-    If $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iP]]$ then
-    $[[ [mu] (ord varset in iP) ]] = [[ord ([mu] varset) in [mu] iP ]]$
-  \end{itemize}
-\end{lemma}
-
+\lemDistrMuOrd*
 \begin{proof}
@@
 
-\begin{lemma}[Ordering is not affected by independent substitutions]
-  \label{lemma:ord-sigma}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
-  taking free variables from $[[Γ2]]$, and $[[varset]]$ is a set of variables
-  disjoint with both $[[Γ1]]$ and $[[Γ2]]$, 
-  $[[iN]]$ and $[[iP]]$ are types. Then
-    \begin{itemize}
-  \item[$-$] $[[ ord varset in [σ]iN ]] = [[ord varset in iN ]]$
-  \item[$+$] $[[ ord varset in [σ]iP ]] = [[ord varset in iP ]]$
-  \end{itemize}
-\end{lemma}
+\lemOrdSigma*
 \begin{proof}
@@
                                    &= [[ ord varset in ([σ]iP → [σ]iM) ]]
-                                   && \text{by the definition of substitution}\\
-                                   &= [[ ord varset in [σ]iP, (ord varset in [σ]iM \ {ord varset in [σ]iP}) ]]
-                                   && \text{by the definition of ordering}\\
-                                   &= [[ ord varset in iP, (ord varset in iM \ {ord varset in iP}) ]]
-                                   && \text{by the induction hypothesis}\\
+                                   && \text{def. of substitution}\\
+                                   &= [[ ord varset in [σ]iP]],\\
+                                   &\phantom{=} ~ [[(ord varset in [σ]iM \ {ord varset in [σ]iP}) ]]
+                                   && \text{def. of ordering}\\
+                                   &= [[ ord varset in iP]],\\
+                                   &\phantom{=} ~ [[(ord varset in iM \ {ord varset in iP}) ]]
+                                   && \text{the induction hypothesis}\\
                                    &= [[ ord varset in iP → iM ]]
-                                   && \text{by the definition of ordering}\\
+                                   && \text{def. of ordering}\\
                                    &= [[ ord varset in iN ]]
@@
 
-\begin{lemma}[Completeness of variable ordering]
-  \label{lemma:ord-completeness}
-  Variable ordering is invariant under equivalence. For arbitrary $[[varset]]$,
-   \begin{itemize}
-  \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in iN]] = [[ord varset in iM]]$ (as lists)
-  \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in iP]] = [[ord varset in iQ]]$ (as lists)
-  \end{itemize}
-\end{lemma}
+\lemOrdCompleteness*
 \begin{proof}
@@
       and by inversion, $[[iP ≈ iQ]]$ and $[[iN ≈ iM]]$.
-      They by the induction hypothesis,
+      Then by the induction hypothesis,
       $[[ord varset in iP]] = [[ord varset in iQ]]$ 

### 2023-11-13 proofs/tex/ord-statements.tex
Restructure proofs

\begin{restatable}[Ordering is deterministic]{observation}{obsOrdDeterministic}
    \label{obs:ord-deterministic}
    If $[[ord varset in iN = ordVars1]]$ and $[[ord varset in iN = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
    If $[[ord varset in iP = ordVars1]]$ and $[[ord varset in iP = ordVars2]]$ then $[[ordVars1 = ordVars2]]$.
    This way, we can use $[[ord varset in iN]]$ and as a function on $[[iN]]$,
    and $[[ord varset in iP]]$ as a function on $[[iP]]$.
\end{restatable}

\begin{restatable}[Soundness of variable ordering]{lemma}{lemOrdSoundness}
  \label{lemma:ord-soundness}
  Variable ordering extracts used free variables.
  \begin{itemize}
    \item[$-$] $[[ {ord varset in iN} ]] = [[varset ∩ fv iN]]$ (as sets)
    \item[$+$] $[[ {ord varset in iP} ]] = [[varset ∩ fv iP]]$ (as sets)
  \end{itemize}
\end{restatable}

\begin{restatable}[Additivity of ordering]{corollary}{corOrdAdditivity}
  \label{corollary:ord-additivity}
  Variable ordering is additive (in terms of set union) with respect to its first argument.
  \begin{itemize}
    \item[$-$] $[[ {ord (varset1 ∪ varset2) in iN} 
                = 
                {ord varset1 in iN} ∪ {ord varset2 in iN}]]$ (as sets)
    \item[$+$] $[[{ord (varset1 ∪ varset2) in iP}
                =
                {ord varset1 in iP} ∪ {ord varset2 in iP}]]$ (as sets)

  \end{itemize}
\end{restatable}

\begin{restatable}[Weakening of ordering]{lemma}{lemOrdWeakening}
  \label{corollary:ord-weakening}
  Only used variables matter in the first argument of the ordering,
  \begin{itemize}
    \item[$-$] $[[ ord (varset ∩ fv iN) in iN ]] = [[ ord varset in iN ]]$
    \item[$+$] $[[ ord (varset ∩ fv iP) in iP ]] = [[ ord varset in iP ]]$
  \end{itemize}
\end{restatable}

\begin{restatable}[Idempotency of ordering]{corollary}{corOrdIdemp}
  \label{corollary:ord-idemp}
  \hfill
  \begin{itemize}
    \item[$-$] If $[[ ord varset in iN = ordVars ]]$ then 
      $[[ ord {ordVars} in iN = ordVars ]]$,
    \item[$+$] If $[[ ord varset in iP = ordVars ]]$ then 
      $[[ ord {ordVars} in iP = ordVars ]]$;
  \end{itemize}
\end{restatable}

\begin{restatable}[Distributivity of renaming over variable ordering]{lemma}{lemDistrMuOrd}
  \label{lemma:distr-mu-ord}
  Suppose that $\mu$ is a bijection between two sets of variables
  $\mu : A \leftrightarrow B$.

  \begin{itemize}
  \item[$-$]
    If $[[mu]]$ is collision-free on $[[varset]]$ and $[[fv iN]]$ then
    $[[ [mu] (ord varset in iN) ]] = [[ord ([mu] varset) in [mu] iN ]]$
  \item[$+$]
    If $[[mu]]$ is collision-free on $[[varset]]$ and $[[fv iP]]$ then
    $[[ [mu] (ord varset in iP) ]] = [[ord ([mu] varset) in [mu] iP ]]$
  \end{itemize}
\end{restatable}

\begin{restatable}[Ordering is not affected by independent substitutions]{lemma}{lemOrdSigma}
    \label{lemma:ord-sigma}
    Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
    taking free variables from $[[Γ2]]$, and $[[varset]]$ is a set of variables
    disjoint with both $[[Γ1]]$ and $[[Γ2]]$, 
    $[[iN]]$ and $[[iP]]$ are types. Then
        \begin{itemize}
    \item[$-$] $[[ ord varset in [σ]iN ]] = [[ord varset in iN ]]$
    \item[$+$] $[[ ord varset in [σ]iP ]] = [[ord varset in iP ]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of variable ordering]{lemma}{lemOrdCompleteness}
    \label{lemma:ord-completeness}
    Variable ordering is invariant under equivalence. For arbitrary $[[varset]]$,
     \begin{itemize}
    \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in iN]] = [[ord varset in iM]]$ (as lists)
    \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in iP]] = [[ord varset in iQ]]$ (as lists)
    \end{itemize}
\end{restatable}

### 2023-11-13 proofs/tex/singularity-lemmas.tex
Resolve reference issues
Restructure proofs

@@
-\begin{lemma}[Soundness of Entry Singularity]
-    \label{lemma:entry-singularity-soundness}
-    \begin{itemize}
-        \item [$+$] Suppose $[[scE singular with iP]]$ for $[[iP]]$ well-formed in $[[Γ]]$.
-            Then $[[ Γ ⊢ iP : scE ]]$
-            and for any $[[ Γ ⊢ iP']]$ such that $[[Γ ⊢ iP' : scE]]$, $[[Γ ⊢ iP' ≈ iP]]$;
-        \item [$-$] Suppose $[[scE singular with iN]]$ for $[[iN]]$ well-formed in $[[Γ]]$.
-            Then $[[ Γ ⊢ iN : scE ]]$
-            and for any $[[ Γ ⊢ iN']]$ such that $[[Γ ⊢ iN' : scE]]$, $[[Γ ⊢ iN' ≈ iN]]$.
-    \end{itemize}
-\end{lemma}
+\lemEntrySingularitySoundness*
 \begin{proof}
@@
 
-\begin{lemma}[Completeness of Entry Singularity]
-    \label{lemma:entry-singularity-completeness}
-    \hfill
-    \begin{itemize}
-        \item [$-$] Suppose that there exists $[[iN]]$ well-formed in $[[Γ]]$ 
-            such that for any $[[iN']]$ well-formed in $[[Γ]]$,
-            $[[Γ ⊢ iN' : scE]]$ implies $[[Γ ⊢ iN' ≈ iN]]$. 
-            Then $[[scE singular with nf(iN)]]$.
-        \item [$+$] Suppose that there exists $[[iP]]$ well-formed in $[[Γ]]$ 
-            such that for any $[[iP']]$ well-formed in $[[Γ]]$,
-            $[[Γ ⊢ iP' : scE]]$ implies $[[Γ ⊢ iP' ≈ iP]]$. 
-            Then $[[scE singular with nf(iP)]]$ .
-    \end{itemize}
-\end{lemma}
+\lemEntrySingularityCompleteness*
 \begin{proof}
@@
 
-\begin{lemma}[Soundness of Singularity]
-    \label{lemma:singularity-soundness}
-    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
-    Then $[[ Θ ⊢ uσ : Ξ ]]$,
-     $[[ Θ ⊢ uσ : SC ]]$ and for any 
-    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
-    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
-\end{lemma}
+\lemSingularitySoundness*
 \begin{proof}
@@
 
-\begin{observation}[Singularity is Deterministiic]
-    For a fixed $[[SC]]$ such that $[[Θ ⊢ SC : Ξ]]$,
-    if $[[SC singular with uσ]]$ and $[[SC singular with uσ']]$,
-    then $[[uσ = uσ']]$.
-\end{observation}
+\obsSingularityDeterministic*
 \begin{proof}
-    By \label{lemma:singularity-soundness},
+    By \cref{lemma:singularity-soundness},
     $[[Θ ⊢ uσ : Ξ]]$ and $[[Θ ⊢ uσ' : Ξ]]$.
@@
 
-\begin{lemma}[Completeness of Singularity]
-    \label{lemma:singularity-completeness}
-    For a given $[[Θ ⊢ SC]]$,
-    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
-    on $[[Ξ]] \supseteq [[dom(SC)]]$.
-    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
-    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
-    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
-    Then 
-    \begin{itemize}
-        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
-        \item $[[Ξ]] = [[dom(SC)]]$.
-    \end{itemize} 
-\end{lemma}
+\lemSingularityCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/singularity-statements.tex
Restructure proofs

\begin{restatable}[Soundness of Entry Singularity]{lemma}{lemEntrySingularitySoundness}
    \label{lemma:entry-singularity-soundness}
    \hfill
    \begin{itemize}
        \item [$+$] Suppose $[[scE singular with iP]]$ for $[[iP]]$ well-formed in $[[Γ]]$.
            Then $[[ Γ ⊢ iP : scE ]]$
            and for any $[[ Γ ⊢ iP']]$ such that $[[Γ ⊢ iP' : scE]]$, $[[Γ ⊢ iP' ≈ iP]]$;
        \item [$-$] Suppose $[[scE singular with iN]]$ for $[[iN]]$ well-formed in $[[Γ]]$.
            Then $[[ Γ ⊢ iN : scE ]]$
            and for any $[[ Γ ⊢ iN']]$ such that $[[Γ ⊢ iN' : scE]]$, $[[Γ ⊢ iN' ≈ iN]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Completeness of Entry Singularity]{lemma}{lemEntrySingularityCompleteness}
    \label{lemma:entry-singularity-completeness}
    \hfill
    \begin{itemize}
        \item [$-$] Suppose that there exists $[[iN]]$ well-formed in $[[Γ]]$ 
            such that for any $[[iN']]$ well-formed in $[[Γ]]$,
            $[[Γ ⊢ iN' : scE]]$ implies $[[Γ ⊢ iN' ≈ iN]]$. 
            Then $[[scE singular with nf(iN)]]$.
        \item [$+$] Suppose that there exists $[[iP]]$ well-formed in $[[Γ]]$ 
            such that for any $[[iP']]$ well-formed in $[[Γ]]$,
            $[[Γ ⊢ iP' : scE]]$ implies $[[Γ ⊢ iP' ≈ iP]]$. 
            Then $[[scE singular with nf(iP)]]$ .
    \end{itemize}
\end{restatable}

\begin{restatable}[Soundness of Singularity]{lemma}{lemSingularitySoundness}
    \label{lemma:singularity-soundness}
    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
    Then $[[ Θ ⊢ uσ : Ξ ]]$,
     $[[ Θ ⊢ uσ : SC ]]$ and for any 
    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
\end{restatable}

\begin{restatable}[Singularity is Deterministic]{observation}{obsSingularityDeterministic}
    For a fixed $[[SC]]$ such that $[[Θ ⊢ SC : Ξ]]$,
    if $[[SC singular with uσ]]$ and $[[SC singular with uσ']]$,
    then $[[uσ = uσ']]$.
\end{restatable}

\begin{restatable}[Completeness of Singularity]{lemma}{lemSingularityCompleteness}
    \label{lemma:singularity-completeness}
    For a given $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[Ξ]] \supseteq [[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
    Then 
    \begin{itemize}
        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
        \item $[[Ξ]] = [[dom(SC)]]$.
    \end{itemize} 
\end{restatable}

### 2023-11-13 proofs/tex/subst-algo-lemmas.tex
Restructure proofs

@@
-\begin{lemma}[Algorithmic Substitution Strengthening]
-  \label{lemma:subst-restr-uv}
-  Restricting the substitution to the algorithmic 
-  variables of the
-  substitution subject does not affect the result.
-  Suppose that $[[uσ]]$ is an algorithmic substitution,
-  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
- \begin{itemize}
-    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
-    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv uN]uN ]]$
-  \end{itemize}
-\end{lemma}
+\lemmaSubstRestrUV*
 \begin{proof}
@@
 
-\begin{lemma}[Substitutions equal on the algorithmic variables]
-  \label{lemma:subst-eq-algovar}
-  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
-  normalized substitutions of signature $[[Θ ⊢ uσi : Ξ]]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a normalized type $[[Γ; Ξ ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
-      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
-    \item [$-$] for a normalized type $[[Γ; Ξ ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
-      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
-  \end{enumerate}
-\end{lemma}
+\lemmaSubstEqAlgVar*
 \begin{proof}
@@
 
-\begin{corollary}[Substitutions equivalent on the algorithmic variables]
-  \label{lemma:subst-equiv-algovar}
-  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
-  of signature $[[Θ ⊢ uσi : Ξ]]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a type $[[Γ; Ξ ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
-    \item [$-$] for a type $[[Γ; Ξ ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
-      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
-  \end{enumerate}
-\end{corollary}
+\corollarySubstEquivAlgVar*
 \begin{proof}

### 2023-11-13 proofs/tex/subst-lemmas.tex
Restructure
Restructure proofs

@@
-\begin{lemma}[Substitution strengthening]
-  \label{lemma:subst-restr-fv}
-  Restricting the substitution to the free variables of the
-  substitution subject does not affect the result.
-  Suppose that $[[σ]]$ is a substitution,
-  $[[iP]]$ and $[[iN]]$ are types. Then
- \begin{itemize}
-    \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
-    \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
-  \end{itemize}
-\end{lemma}
+\lemmaSubstRestrFv*
 \begin{proof}
@@
 
-\begin{lemma}[Signature of a restricted substitution]
-  \label{lemma:subst-restr-sig}
-  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset :  Γ1 ∩ varset]]$.
-\end{lemma}
+\lemmaSubstRestrSig*
 \begin{proof}
@@
 
-\begin{lemma}
-  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. 
-  Then if $[[varset]]$ is disjoint from $[[Γ1]]$, 
-  then $[[ σ|varset]] = [[ id ]]$. 
-\end{lemma}
+\lemmaSubstDisjoint*
 \begin{proof}
@@
 
-\begin{corollary}[Application of a disjoint substitution]
-  \label{corollary:subst-disj}
-  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-    \item [$+$] if $[[Γ1 ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
-    \item [$-$] if $[[Γ1 ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
-  \end{itemize}
-\end{corollary}
+\corollarySubstDisj*
 
-\begin{lemma}[Substitution range weakening]
-  \label{lemma:subst-range-weakening}
-  Suppose that $[[Γ2 ⊆ Γ'2]]$ are contexts and 
-  $[[σ]]$ is a substitution.
-  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ'2 ⊢ σ : Γ1]]$.
-\end{lemma}
+\lemmaSubstRangeWeak*
 \begin{proof}
@@
 
-\begin{lemma}[]
-  Suppose that $[[Γ' ⊆ Γ]]$,
-  $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
-  Then 
-  \begin{enumerate}
-    \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$;
-    \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
-    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$.
-  \end{enumerate}
-\end{lemma}
+\lemmaSubstEquiv*
 \begin{proof}
@@
 
-\begin{lemma}[Substitution composition well-formedness]
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
-  then $[[Γ'1, Γ'2 ⊢ σ2 ○ σ1 : Γ1, Γ2]]$.
-\end{lemma}
+\lemmaSubstCompWell*
+\begin{proof}
+    By definition of composition.
+\end{proof}
 
-\begin{lemma}[Substitution monadic composition well-formedness]
-  \label{lemma:subst-monad-composition-wf}
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
-  then $[[Γ'2 ⊢ σ2 <=< σ1 : Γ1]]$.
-\end{lemma}
+\lemmaSubstMonadCompWell*
+\begin{proof}
+  By definition of monadic composition. 
+\end{proof}
 
-\begin{lemma}[Substitution composition]
-  \label{lemma:subst-composition}
-    If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
-    $[[Γ1 ∩ Γ'2 = ∅ ]]$ and $[[ Γ1 ∩ Γ2 = ∅ ]]$ then 
-    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
-\end{lemma}
+\lemmaSubstComp*
+\begin{proof}
+  \hfill
+  \begin{enumerate}
+    \item Suppose that $[[α± ∊ Γ1]]$ then $[[α± ∉ Γ2]]$ , and thus, 
+      $[[ [(σ2 <=< σ1) ○ σ2]α± = [(σ2 <=< σ1)]α± ]] = [[ [σ2][σ1]α± = [(σ2 ○ σ1)]α± ]]$.
+    \item Suppose that $[[α± ∉ Γ1]]$ then $[[ [(σ2 ○ σ1)]α± = [σ2]α± ]]$.  Then 
+    \begin{enumerate}
+      \item if $[[α± ∉ Γ2]]$ then $[[ [σ2]α± = α± ]]$ and 
+        $[[ [(σ2 <=< σ1) ○ σ2]α± = [(σ2 <=< σ1)][σ2]α±]] = [[ [σ2 <=< σ1]α± = α± ]]$
+      \item if  $[[α± ∊ Γ2]]$ then $[[ Γ2' ⊢ [σ2]α±  ]]$, and hence, 
+        $[[ [(σ2 <=< σ1) ○ σ2]α±  = [(σ2 <=< σ1)][σ2]α±]] = [[ [σ2]α± ]]$
+        by definition of monadic composition, since 
+        none of the free variables of  $[[ [σ2]α± ]]$ is in $[[Γ1]]$.
+    \end{enumerate}
+  \end{enumerate}
+\end{proof}
 
-\begin{corollary}[Substitution composition commutativity]
-  \label{corollary:subst-composition-commutativity}
-  If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
-  $[[ Γ1 ∩ Γ2 = ∅ ]]$, 
-  $[[ Γ1 ∩ Γ'2 = ∅ ]]$, and
-  $[[ Γ'1 ∩ Γ2 = ∅ ]]$ then 
-  $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
-\end{corollary}
+\corollarySubstCompCumm*
 \begin{proof}
@@
 
-\begin{lemma}[Substitution domain weakening]
-  \label{lemma:subst-domain-weakening}
-  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2, Γ' ⊢ σ : Γ1, Γ']]$
-\end{lemma}
+\lemmaSubstDomWeak*
 \begin{proof}
@@
   and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
-  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
+  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ' ⊆ Γ2, Γ']]$, 
   and thus, $[[Γ2, Γ' ⊢ α± ]]$.
@@
 
-\begin{lemma}[Free variables after substitution]
-  \label{lemma:subst-fv}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, then
-  \begin{enumerate}
-    \item [$+$] for a type $[[iP]]$,
-      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-     $[[fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$;
-    \item [$-$] for a type $[[iN]]$, 
-      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
-      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
-  \end{enumerate}
-\end{lemma}
+\lemmaSubstFv*
 \begin{proof}
@@
 
-\begin{lemma}[Free variables of a variable image]
-  \label{lemma:subst-fv-image}
-  Suppose that $[[σ]]$ is an arbitrary substitution, 
-  Then 
-  \begin{itemize}
-    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
-    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaSubstFvImage*
 \begin{proof}
@@
 
-  Let us suppose that te type is $[[∃nas.iQ]]$.
+  Let us suppose that the type is $[[∃nas.iQ]]$.
   Then $[[α± ∊ fv(∃nas.iQ)]]$ means

### 2023-11-13 proofs/tex/subst-statements.tex
Restructure proofs

\begin{restatable}[Substitution strengthening]{lemma}{lemmaSubstRestrFv}
    \label{lemma:subst-restr-fv}
    Restricting the substitution to the free variables of the
    substitution subject does not affect the result.
    Suppose that $[[σ]]$ is a substitution,
    $[[iP]]$ and $[[iN]]$ are types. Then
    \begin{itemize}
        \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
        \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
    \end{itemize}
\end{restatable}

\begin{restatable}[Signature of a restricted substitution]{lemma}{lemmaSubstRestrSig}
    \label{lemma:subst-restr-sig}
    If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset :  Γ1 ∩ varset]]$.
\end{restatable}

\begin{restatable}{lemma}{lemmaSubstDisjoint}
    \label{lemma:subst-disjoint}
    Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. 
    Then if $[[varset]]$ is disjoint from $[[Γ1]]$, 
    then $[[ σ|varset]] = [[ id ]]$. 
\end{restatable}

\begin{restatable}[Application of a disjoint substitution]{corollary}{corollarySubstDisj}
    \label{corollary:subst-disj}
    Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. Then
    \begin{itemize}
        \item [$+$] if $[[Γ1 ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
        \item [$-$] if $[[Γ1 ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
    \end{itemize}
\end{restatable}

\begin{restatable}[Substitution range weakening]{lemma}{lemmaSubstRangeWeak}
    \label{lemma:subst-range-weakening}
    Suppose that $[[Γ2 ⊆ Γ'2]]$ are contexts and 
    $[[σ]]$ is a substitution.
    Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ'2 ⊢ σ : Γ1]]$.
\end{restatable}

\begin{restatable}[Subsitutions Equivalent on Free Variables]{lemma}{lemmaSubstEquiv}
    \label{lemma:subst-equiv}
    Suppose that $[[Γ' ⊆ Γ]]$,
    $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
    Then 
    \begin{enumerate}
        \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
        $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$;
        \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
        $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$.
    \end{enumerate}
\end{restatable}

\begin{restatable}[Substitution composition well-formedness]{lemma}{lemmaSubstCompWell}
    \label{lemma:subst-comp-well}
    If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
    then $[[Γ'1, Γ'2 ⊢ σ2 ○ σ1 : Γ1, Γ2]]$.
\end{restatable}

\begin{restatable}[Substitution monadic composition well-formedness]{lemma}{lemmaSubstMonadCompWell}
    \label{lemma:subst-monad-composition-wf}
    If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
    then $[[Γ'2 ⊢ σ2 <=< σ1 : Γ1]]$.
\end{restatable}

\begin{restatable}[Substitution composition]{lemma}{lemmaSubstComp}
    \label{lemma:subst-composition}
        If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
        $[[Γ1 ∩ Γ'2 = ∅ ]]$ and $[[ Γ1 ∩ Γ2 = ∅ ]]$ then 
        $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
\end{restatable}

\begin{restatable}[Substitution composition commutativity]{corollary}{corollarySubstCompCumm}
    \label{corollary:subst-composition-commutativity}
    If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
    $[[ Γ1 ∩ Γ2 = ∅ ]]$, 
    $[[ Γ1 ∩ Γ'2 = ∅ ]]$, and
    $[[ Γ'1 ∩ Γ2 = ∅ ]]$ then 
    $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
\end{restatable}

\begin{restatable}[Substitution domain weakening]{lemma}{lemmaSubstDomWeak}
    \label{lemma:subst-domain-weakening}
    If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2, Γ' ⊢ σ : Γ1, Γ']]$
\end{restatable}

\begin{restatable}[Free variables after substitution]{lemma}{lemmaSubstFv}
    \label{lemma:subst-fv}
    Suppose that $[[Γ2 ⊢ σ : Γ1]]$, then
    \begin{enumerate}
        \item [$+$] for a type $[[iP]]$,
            the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
         $[[fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$;
        \item [$-$] for a type $[[iN]]$, 
            the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
            $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
    \end{enumerate}
\end{restatable}

\begin{restatable}[Free variables of a variable image]{lemma}{lemmaSubstFvImage}
    \label{lemma:subst-fv-image}
    Suppose that $[[σ]]$ is an arbitrary substitution, 
    Then 
    \begin{itemize}
        \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
        \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
    \end{itemize}
\end{restatable}

### 2023-11-13 proofs/tex/unif-constraint-merge-lemmas.tex
Add Declarative Typing comments
Restructure proofs

@@
-\begin{observation} [Unification Constraint Merge is Deterministic]
-    \label{obs:unif-merge-deterministic}
-    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
-    If $[[Θ ⊢ UC1 & UC2 = UC]]$ and $[[Θ ⊢ UC1 & UC2 = UC']]$ are defined then 
-    $[[UC]] = [[UC']]$.
-\end{observation}
+\obsUnifMergeDet*
 \begin{proof}
@@
 
-\begin{lemma} [Soundness of Unification Constraint Merge]
-    \label{lemma:unif-merge-soundness}
-    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ 
-    are normalized unification constraints.
-    If $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined then
-    $[[UC = UC1 ∪ UC2]]$.
-\end{lemma}
+\lemUnifMergeSoundness*
 \begin{proof}
@@
 
-\begin{corollary}
-    \label{corollary:unif-merge-soundness}
-    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$ 
-    are normalized unification constraints.
-    If $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined then
-    \begin{enumerate}
-        \item $[[Θ ⊢ UC]]$ is normalized unification constraint,
-        \item for any substitution $[[Θ ⊢ uσ : dom(UC)]]$, $[[ Θ  ⊢ uσ : lift UC ]]$ implies 
-        $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[Θ ⊢ uσ : lift UC2]]$.
-    \end{enumerate}
-\end{corollary}
+\corUnifMergeSoundness*
 \begin{proof}
@@
 
-\begin{lemma} [Completeness of Unification Constraint Entry Merge]
-    \label{lemma:unif-entry-merge-completeness}
-    For a fixed context $[[Γ]]$,
-    suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$ are matching constraint entries.
-    \begin{itemize}
-        \item for a type $[[iP]]$ such that $[[Γ ⊢ iP : ucE1]]$ and $[[Γ ⊢ iP : ucE2]]$,
-        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iP : ucE]]$.
-        \item for a type $[[iN]]$ such that $[[Γ ⊢ iN : ucE1]]$ and $[[Γ ⊢ iN : ucE2]]$,
-        $[[Γ ⊢ ucE1 & ucE2 = ucE]]$ is defined and $[[Γ ⊢ iN : ucE]]$.
-    \end{itemize}
-\end{lemma}
+\lemUnifEntryMergeCompleteness*
 \begin{proof}
@@
 
-\begin{lemma} [Completeness of Unification Constraint Merge] 
-    \label{lemma:unif-merge-completeness}
-    Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$.
-    Then for any $[[Ξ]] \supseteq [[dom(UC1) ∪ dom(UC2)]]$ 
-    and substitution $[[Θ ⊢ uσ : Ξ]]$ such that $[[ Θ ⊢ uσ : lift UC1 ]]$ and $[[ Θ   ⊢ uσ : lift UC2 ]]$, 
-    \begin{enumerate}
-        \item $[[Θ ⊢ UC1 & UC2 = UC]]$ is defined and
-        \item $[[ Θ   ⊢ uσ : UC ]]$.
-    \end{enumerate}
-\end{lemma}
+\lemUnifMergeCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/unification-lemmas.tex
Restructure
Restructure proofs

@@
-\begin{observation}[Unification is deterministic]
-    \label{obs:unif-deterministic}
-    \hfill
-    \begin{itemize}
-        \item [$+$] If $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$
-            and $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC']]$ then $[[UC = UC']]$.
-        \item [$-$] If $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$
-            and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$ then $[[UC = UC']]$.
-    \end{itemize}
-\end{observation}
+\obsUnifDet*
 \begin{proof}
@@
     $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ and $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC']]$.
-    Let us consider the positive case, since the negative case is symmetric.
+    Let us consider the positive case only since the negative case is symmetric.
 
@@
 
-\begin{lemma}[Soundness of Unification] \label{lemma:unification-soundness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
-        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
-        $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
-        such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.
-
-        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
-        $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
-        $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
-    \end{itemize}
-\end{lemma}
+\lemmaUnificationSoundness*
 \begin{proof}
@@
             \item $[[Θ ⊢ UC1 : uv uP]]$,
-            \item for any $(Θ  ⊢ lift UC1) ⊢ uσ'$, $[[ [uσ']uP = iQ ]]$.
+            \item for any $[[Θ ⊢ uσ' : lift UC1]]$, $[[ [uσ']uP = iQ ]]$.
         \end{itemize}
@@
 
-\begin{lemma}[Completeness of Unification] \label{lemma:unification-completeness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
-        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
-        suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
-        then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.
-
-        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
-        suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
-        then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
-   \end{itemize}
-\end{lemma}
+\lemmaUnificationCompleteness*
 \begin{proof}

### 2023-11-13 proofs/tex/upgrade-lemmas.tex
Restructure proofs

@@
 
-\begin{observation}[Upgrade is deterministic]
-    \label{obs:upgrade-deterministic}
-    Assuming $[[iP]]$ is well-formed in $[[Γ ⊆ Δ]],$\\
-    if $[[upgrade Γ ⊢ iP to Δ = iQ]]$ and $[[upgrade Γ ⊢ iP to Δ = iQ']]$ are defined 
-    then $[[iQ = iQ']]$.
-\end{observation}
+\obsUpgradeDeterminism*
 \begin{proof}
@@
 
+\lemmaUpgradeSoundness*
 \begin{lemma}[Soundness of Upgrade]\label{lemma:upgrade-soundness}
@@
 
-\begin{lemma}[Completeness and Initiality of Upgrade] \label{lemma:upgrade-completeness}
-    The upgrade returns the least $[[Γ]]$-supertype of $[[iP]]$ well-formed in $[[Δ]]$.
-    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]],$\\
-    For any $[[iQ']]$ such that 
-    \begin{enumerate}
-        \item $[[Δ ⊢ iQ']]$ and
-        \item $[[Γ ⊢ iQ' ≥ iP]]$,
-    \end{enumerate}
-
-    The result of the upgrade algorithm $[[iQ]]$ exists
-    ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
-\end{lemma}
+\lemmaUpgradeCompleteness*
 \begin{proof}
-
     Let us consider fresh (not intersecting with $[[Γ]]$) $[[pnbs]]$ and $[[pncs]]$.

### 2023-11-13 proofs/tex/upgrade-statements.tex
Restructure proofs

\begin{restatable}[Upgrade determinism]{observation}{obsUpgradeDeterminism}
    \label{obs:upgrade-deterministic}
    Assuming $[[iP]]$ is well-formed in $[[Γ ⊆ Δ]],$\\
    if $[[upgrade Γ ⊢ iP to Δ = iQ]]$ and $[[upgrade Γ ⊢ iP to Δ = iQ']]$ are defined 
    then $[[iQ = iQ']]$.
\end{restatable}

\begin{restatable}[Soundness of Upgrade]{lemma}{lemmaUpgradeSoundness}
    \label{lemma:upgrade-soundness}
    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
    if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
    then
    \begin{enumerate}
        \item $[[Δ ⊢ iQ]]$
        \item $[[Γ ⊢ iQ ≥ iP]]$
    \end{enumerate}
\end{restatable}

\begin{restatable}[Completeness and Initiality of Upgrade]{lemma}{lemmaUpgradeCompleteness}
    \label{lemma:upgrade-completeness}
    The upgrade returns the least $[[Γ]]$-supertype of $[[iP]]$ well-formed in $[[Δ]]$.
    Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]],$\\
    For any $[[iQ']]$ such that 
    \begin{enumerate}
        \item $[[Δ ⊢ iQ']]$ and
        \item $[[Γ ⊢ iQ' ≥ iP]]$,
    \end{enumerate}

    The result of the upgrade algorithm $[[iQ]]$ exists
    ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
\end{restatable}

### 2023-11-13 proofs/tex/wf-algo-lemmas.tex
Restructure proofs

@@
-\begin{lemma}[Soundness of algorithmic type well-formedness]
-    \label{lemma:wf-algo-soundness}
-    \hfill
-  \begin{itemize}
-    \item[$+$] if $[[Γ ; Ξ ⊢ uP]]$ then $[[fv(uP) ⊆ Γ]]$ and $[[uv(uP) ⊆ Ξ]]$;
-    \item[$-$] if $[[Γ ; Ξ ⊢ uN]]$ then $[[fv(uN) ⊆ Γ]]$ and $[[uv(uN) ⊆ Ξ]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaWfAlgoSoundness*
 \begin{proof}
@@
 
-\begin{lemma}[Completeness of algorithmic type well-formedness]
-  \label{lemma:wf-algo-ctxt-equiv}
-  In the well-formedness judgment, only used variables matter:
-  \begin{itemize}
-  \item[$+$] if $[[Γ1 ∩ fv iP]] = [[Γ2 ∩ fv iP]]$
-    and $[[Ξ1 ∩ uv uP]] = [[Ξ2 ∩ uv uP]]$ then
-    $[[Γ1 ; Ξ1 ⊢ uP]] \iff [[Γ2 ; Ξ2 ⊢ uP]]$, and
-  \item[$-$] if $[[Γ1 ∩ fv iN]] = [[Γ2 ∩ fv iN]]$
-    and $[[Ξ1 ∩ uv uN]] = [[Ξ2 ∩ uv uN]]$ then
-    $[[Γ1 ; Ξ1 ⊢ uN]] \iff [[Γ2 ; Ξ2 ⊢ uN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaWfAlgoCtxtEquiv*
 \begin{proof}
@@
 
-\begin{lemma}[Variable algorithmization agrees with well-formedness]
-  \label{lemma:var-algo-wf}
-  \hfill
-  \begin{itemize}
-    \item[$+$]  $[[Γ, nas ⊢ iP]]$ implies $[[Γ; {nuas} ⊢ [nuas/nas]iP]]$;
-    \item[$-$]  $[[Γ, nas ⊢ iN]]$ implies $[[Γ; {nuas} ⊢ [nuas/nas]iN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaVarAlgoWf*
 \begin{proof}
@@
 
-\begin{lemma}[Variable dealgorithmization agrees with well-formedness]
-  \label{lemma:var-dealgo-wf}
-  \hfill
-  \begin{itemize}
-    \item[$+$]  $[[Γ; {nuas} ⊢ uP]]$ implies $[[Γ, nas ⊢ [nas/nuas]iP]]$;
-    \item[$-$]  $[[Γ; {nuas} ⊢ uN]]$ implies $[[Γ, nas ⊢ [nas/nuas]iN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaVarDealgoWf*
 \begin{proof}
@@
 
-\begin{corollary}[Well-formedness Algorithmic Context Weakening]
-  \label{lemma:wf-weakening-algo}
-  Suppose that $[[Γ1 ⊆ Γ2]]$,
-  and $[[Ξ1 ⊆ Ξ2]]$. Then
-  \begin{itemize}
-    \item[$+$] if $[[Γ1 ; Ξ1 ⊢ uP]]$ implies $[[Γ2 ; Ξ2 ⊢ uP]]$,
-    \item [$-$] if $[[Γ1 ; Ξ1 ⊢ uN]]$ implies $[[Γ2 ; Ξ2 ⊢ uN]]$.
-  \end{itemize}
-\end{corollary}
+\corollaryWfWeakeningAlgo*
 \begin{proof}

### 2023-11-13 proofs/tex/wf-lemmas.tex
Add Declarative Typing comments
Restructure proofs

@@
-\begin{lemma}[Soundness of type well-formedness]
-    \label{lemma:wf-soundness}
-    \hfill
-
-  \begin{itemize}
-    \item[$+$] if $[[Γ ⊢ iP]]$ then $[[fv(iP) ⊆ Γ]]$;
-    \item[$-$] if $[[Γ ⊢ iN]]$ then $[[fv(iN) ⊆ Γ]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaWfSoundness*
 \begin{proof}
@@
 
-\begin{lemma}[Completeness of type well-formedness]
-  \label{lemma:wf-ctxt-equiv}
-
-  In the well-formedness judgment, only used variables matter:
-  \begin{itemize}
-  \item[$+$] if $[[Γ1 ∩ fv iP]] = [[Γ2 ∩ fv iP]]$ then
-    $[[Γ1 ⊢ iP]] \iff [[Γ2 ⊢ iP]]$,
-  \item[$-$] if $[[Γ1 ∩ fv iN]] = [[Γ2 ∩ fv iN]]$ then
-    $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaWfCompleteness*
 \begin{proof}
@@
 
-\begin{corollary}[Context Strengthening]
-  \label{corollary:wf-ctxt-strengthening}
-  \hfill
-
-  \begin{itemize}
-    \item [$+$] If $[[Γ ⊢ iP]]$ then $[[fv(iP) ⊢ iP]]$;
-    \item [$-$] If $[[Γ ⊢ iN]]$ then $[[fv(iN) ⊢ iN]]$.
-  \end{itemize}
-\end{corollary}
+\corollaryWfCtxtStrengthening*
 \begin{proof}
@@
   \end{itemize}
+\end{proof}
 
-\begin{corollary}[Well-formedness Context Weakening]
-  \label{lemma:wf-weakening}
-  Suppose that $[[Γ1 ⊆ Γ2]]$, then
-  \begin{itemize}
-    \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[Γ2 ⊢ iP]]$,
-    \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[Γ2 ⊢ iN]]$.
-  \end{itemize}
-\end{corollary}
+\lemmaWfWeakening*
 \begin{proof}
@@
 
-\begin{corollary}
-  \label{lemma:mut-sub-types-wf-equiv}
-  Suppose that all the types below are well-formed in $[[Γ]]$ and
-  $[[Γ' ⊆ Γ]]$. Then
-  \begin{itemize}
-  \item[$+$] $[[Γ ⊢ iP ≈ iQ]]$ implies $[[Γ' ⊢ iP]] \iff [[Γ' ⊢ iQ]]$
-  \item[$-$] $[[Γ ⊢ iN ≈ iM]]$ implies $[[Γ' ⊢ iN]] \iff [[Γ' ⊢ iM]]$
-  \end{itemize}
-\end{corollary}
-\begin{proof}
-  From \cref{lemma:wf-ctxt-equiv,corollary:fv-mut-sub}.
-\end{proof}
-
-\begin{lemma}[Well-formedness agrees with substitution]
-  \label{lemma:wf-subst}
-  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
-  \begin{itemize}
-  \item[$+$] $[[Γ, Γ1 ⊢ iP]]$ implies $[[Γ, Γ2 ⊢ [σ]iP]]$, and
-  \item[$-$] $[[Γ, Γ1 ⊢ iN]]$ implies $[[Γ, Γ2 ⊢ [σ]iN]]$.
-  \end{itemize}
-\end{lemma}
+\lemmaWfSubst*
 \begin{proof}

### 2023-11-13 wits24/main.tex
Restructure proofs

\documentclass[sigplan,natbib=false,screen]{acmart}
\settopmatter{printacmref=false}
\usepackage{../bibliography}
\input{../prelude.tex}

\title{Local Type Inference for Polarised System F with Existentials}

\author{Ilya Kaysin}

\begin{document}

\maketitle

Type inference for polymorphic lambda calculi has been extensively
researched, but handling \emph{existential} types remains under-studied.
Our work presents a novel type inference algorithm for a 
variant of System F with impredicative polymorphism and first-class
existential types (F$_{\pm}^{\exists}$).
Describing the algorithm on a low level, we resolve the challenges that 
an implementer would face. In particular, the first-class existentials 
break the standard algorithm invariants, complicate the constraint resolution,
and require anti-unification---a novel for this area technique.

\paragraph{Polarization}
While powerful, the combination of impredicativity and polymorphism 
makes type inference undecidable. To address this, we work in a variant 
of a Call-By-Push-Value system, that stratifies types into positive ($[[iP]]$) 
and negative ($[[iN]]$) kinds. 
The positive types correspond to values (for example, integers, products, 
and existential $[[∃]]$-types),
and the negative types are computations 
(in particular, functions and polymorphic $[[∀]]$-types).
This stratification lets us formulate a typing specification that is decidable.

\paragraph{Algorithm Structure}
The algorithm has four main components: subtyping, unification, anti-unification
and type inference. It introduces novel techniques like mixed-polarity
unification and upgrade-based anti-unification enabled by CBPV's polarized
structure. These differ significantly from traditional Hindley-Milner style
inference. In particular, CBPV's restrictions break some key invariants relied
on by HM algorithms like stability of type variables. This requires mixing
unification and anti-unification in non-standard ways.

\paragraph{Subtyping}
An important component of the algorithm is the subtyping relation.
The polymorphism it induces is driven by instantiation:
$[[∀α⁺.iN]]$ is a subtype of any of the instantiations $[[ [iP/α⁺]iN]]$,
and dually, $[[∃α⁻.iP]]$ is a \emph{super}type of any of the 
instantiations $[[ [iN/α⁻]iP]]$.

When checking for subtyping of $\forall\alpha.P \leqslant Q$
or $\exists\alpha.N \geqslant M$, one must instantiate
$\alpha$ with a type that is not known beforehand. 
Standardly, we handle it by replacing $\alpha$ with a 
\emph{meta}variable $[[α̂⁺]]$, and then we recursively accumulate the 
constraints on it. Typically, the constraints are resolved by unification.
For example, the constraint $[[α̂⁺]]\rightarrow[[iN]]$ is resolved by 
unification to $[[α̂⁺]] = [[iP]]$.

\paragraph{Anti-Unificaiton}
However, in an impredicative system, unification is no longer sufficient.    
For instance, the conjunction of constraints 
$\alpha \geqslant (Int \rightarrow  Int)$ and $\alpha \geqslant (Bool \rightarrow Bool)$
has a solution $\alpha = \exists \beta. \beta \rightarrow \beta$.
This raises the need for \emph{anti-unification}: to find the least upper bound, 
one must find the most concrete pattern that fits both types and then 
quantify over the placeholders by $\exists$.

Our algorithm performs local type inference, instantiating quantifiers based
only on information available at each application site. This provides clear
error messages, modular checking and avoids complex heuristics like those used
by global inference algorithms. The algorithm supports implicit
elimination universal quantifiers ($\forall$) and implicit construction of the 
existential ($\exists$) quantifiers without type annotations. 
We prove the algorithm is sound and complete relative to a declarative specification.

\paragraph{Extensions}

We also explore extensions to the core algorithm.
One is allowing the quantifiers to be bounded: in particular, 
the universally quantified variables can have 
a lower bound: $\forall \alpha \geqslant P. \cdots$ 
limits the types that can be substituted for $\alpha$.
Another is allowing lambda abstractions to omit type annotations,
which requires bidirectionalization of the type system
and complicates constraint resolution.

\paragraph{Conclusion}

\end{document}

### 2023-11-13 proofs/main.tex
Fix margins

@@
 \begin{algorithm}
-  \hfill\\
+  \hfill \\
+  \ottdefnATPInfLabeled{}
+  \hfill \\
   \ottdefnATNInfLabeled{}
-  \ottdefnATPInfLabeled{}
+  \hfill \\
   \ottdefnATSpinInfLabeled{}
@@
 First, we define the notion of singularity on constraint entries. 
-$[[scE singular with iP]]$ and $[[scE singular with iN]]$
+$[[scE singular with iP]]$ and its negative counterpart 
 are considered partial functions taking a constraint entry $[[scE]]$

### 2023-11-13 proofs/tex/alg-pos-subtyping-lemmas.tex
Fix margins

@@
             \item If this variable is a regular variable $[[β±]] \notin [[nas]]$, then 
-                $[[ [iNs/nas][uσ] β± ]] = [[ β± ]] $ and $[[ [uσ'][â⁻*/nas]β± ]] = [[ β± ]]$. 
+                $[[ [iNs/nas][uσ] β± ]] = [[ β± ]] $ and\\ $[[ [uσ'][â⁻*/nas]β± ]] = [[ β± ]]$. 
             \item If this variable is a regular variable $[[αi⁻]] \in [[nas]]$, then 
@@
             Now, let us show that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
-            To do that, we notice that $[[ [uσ0]uP0 ]] = [[ [uσ][σ][nas/â⁻*]uP0 ]]$:
+            To do that, we notice that\\ $[[ [uσ0]uP0 ]] = [[ [uσ][σ][nas/â⁻*]uP0 ]]$:
             let us consider an arbitrary variable appearing freely in $[[uP0]]$:
@@
                 \item if this variable is a algorithmic variable $[[αî⁻]] \in [[â⁻*]]$, then
-                $[[ [uσ0]αî⁻ ]] = [[ [σ]αi⁻ ]]$ and 
+                $[[ [uσ0]αî⁻ ]] = [[ [σ]αi⁻ ]]$ and\\
                 $[[ [uσ][σ][nas/â⁻*]αî⁻ ]] = [[ [uσ][σ]αi⁻ ]] = [[ [σ]αi⁻ ]]$,

### 2023-11-13 proofs/tex/norm-lemmas.tex
Fix margins

@@
      the required equality follows from the congruence of the normalization and
-     substitution, and the induction hypothesis.
+     substitution and the induction hypothesis.
      For example, if $[[iN]] = [[iP → iM]]$ then \\
@@
                         &= [[ nf([σ]iP → [σ]iM) ]]
-                        && \text{By the congruence of substitution} \\
+                        && \text{By congruence of substitution} \\
                         &= [[ nf([σ]iP) → nf([σ]iM) ]]
-                        && \text{By the congruence of normalization, i.e. \ruleref{\ottdruleNrmArrowLabel}} \\
+                        && \text{By congruence of normalization} \\
                         &= [[ [nf(σ)]nf(iP) → [nf(σ)]nf(iM) ]]
-                        && \text{By the induction hypothesis} \\
+                        && \text{By induction hypothesis} \\
                         &= [[ [nf(σ)](nf(iP) → nf(iM)) ]]
-                        && \text{By the congruence of substitution} \\
+                        && \text{By congruence of substitution} \\
                         &= [[ [nf(σ)]nf(iP → iM) ]]
-                        && \text{By the congruence of normalization} \\
+                        && \text{By congruence of normalization} \\
                         &= [[ [nf(σ)]nf(iN) ]]
@@
                             && \text {Where $[[pas']] = [[ ord {pas} in nf(iM)]]
-                               = [[ord {pas} in iM]]$
-                               (the latter is by
+                               = [[ord {pas} in iM]]$}\\
+                           & && 
+                               \text{(the latter is by
                                \cref{corollary:normalization-ord})}\\
@@
                           && \text {Where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]
-                             = [[ord {pas} in [σ]iM]]$ (the latter is by
-                             \cref{corollary:normalization-ord})}\\
+                             = [[ord {pas} in [σ]iM]]$}\\
+                          & && 
+                             \text{(the latter is by \cref{corollary:normalization-ord})}\\
                           &= [[ ∀pas'.nf([σ]iM) ]]
-                          && \text{By \cref{lemma:ord-sigma}, $[[pbs]] = [[pas']]$
+                          && \text{By \cref{lemma:ord-sigma},}\\
+                          & && \text{$[[pbs]] = [[pas']]$
                              since $[[{pas}]]$ is disjoint with $[[Γ1]]$ and
@@
                                                   &= [[ ord [mu] ({pbs} ∩ fv iM) in [mu] nf(iM) ]]
-                                                  && \text{by \cref{lemma:distr-mu-ord}, because
-                                                     $[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]]
+                                                  && \text{by \cref{lemma:distr-mu-ord}, because}\\
+                                                  & && \text{$[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]]
                                                      = \emptyset$}\\
                                                   &
-                                                  && \text{and $[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq
+                                                  && \text{$[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq
                                                      [[{pas} ∩ fv iM]] = \emptyset$} \\
@@
                                                   &= [[ ord ({pas} ∩ fv iN) in nf(iN) ]]
-                                                  && \text{because $\mu$ is a bijection between
-                                                     $[[{pas} ∩ fv iN]]$ and $[[{pbs} ∩ fv iM]]$} \\
+                                                  && \text{because $\mu$ is a bijection between}\\
+                                                  & && \text{$[[{pas} ∩ fv iN]]$ and $[[{pbs} ∩ fv iM]]$} \\
                                                   &= [[ ord {pas} in nf(iN) ]]
@@
       [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
-                                           && \text{by the substitution
+                                           && \text{by substitution
                                               congruence} \\
@@
     \begin{aligned}[t]
-      [[ [σ](∀pas.iM) ]] \text{ is normal} &\iff [[ (∀pas.[σ]iM) ]] \text{ is normal}
+      &[[ [σ](∀pas.iM) ]] \text{ is normal} \\
+       &\iff [[ (∀pas.[σ]iM) ]] \text{ is normal}
                                            && \text{assuming $[[pas]] \cap [[Γ1]] = \emptyset$ and

### 2023-11-13 proofs/tex/ord-lemmas.tex
Fix margins

@@
 
-      Let us show that that $A = B$ and $X = Y$, so the results are equal. 
+      Let us show that $A = B$ and $X = Y$, so the results are equal. 
       By the induction hypothesis and set properties,
@@
       Analogously, 
-      $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iN ]] = 
-       [[ ord varset in iN ]]$.
+      $[[ ord (varset ∩ (fv iP ∪ fv iN)) in iN ]]$ $=$
+      \\ $[[ ord varset in iN ]]$.
     \item If the term has shape $[[∀pas.iN]]$,
@@
   We say that a bijection $\mu : A \leftrightarrow B$ between sets of
-  variables is \textbf{collision free on sets} $P$ and $Q$ if and only if
+  variables is \textbf{collision-free on sets} $P$ and $Q$ if and only if
   \begin{enumerate}
@@
   Suppose that $\mu : A \leftrightarrow B$ is a bijection between two sets of variables,
-  and $\mu$ is collision free on $P$ and $Q$.
+  and $\mu$ is collision-free on $P$ and $Q$.
   Then $\mu(P \cap Q) = \mu(P) \cap \mu(Q)$.
@@
     \item $[[na]] \in A$ and $[[na]] \in [[varset]]$\\ Then
+    \\
       $
@@
                                                              &= [[nb]]
-                                                             && \text{for some $[[nb]] \in B$ (notice that $[[nb]] \in [[ [mu]varset ]]$)} \\
+                                                             && \text{for some $[[nb]] \in B$ (notice $[[nb]] \in [[ [mu]varset ]]$)} \\
                                                              &= [[ ord [mu]varset in nb ]]
@@
        \ruleref{\ottdruleOPVarNInLabel}, because
-       $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$, so
+       $[[mu]]$ is collision-free on $[[varset]]$ and $[[fv iN]]$, so
        $[[fv iN]] \ni [[na]] \notin [[mu]](A \cap [[varset]]) \cup
@@
        by \ruleref{\ottdruleOPVarNInLabel}.
-       To prove that $[[ ord [mu] varset in [mu] na ]] = [[·]]$, we apply
+       To prove that\\ $[[ ord [mu] varset in [mu] na ]] = [[·]]$, we apply
        \ruleref{\ottdruleOPVarNInLabel}. Let us show that
@@
        \item[(ii)]
-         Since $[[mu]]$ is collision free on $[[varset]]$ and $[[fv iN]]$,
+         Since $[[mu]]$ is collision-free on $[[varset]]$ and $[[fv iN]]$,
          $[[mu]] (A \cap [[fv iN]]) \ni [[mu]]([[na]]) \notin [[varset]]$.
@@
      $\begin{aligned}[t]
-        [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in iP → iM) ]] \\
+        & [[ [mu] (ord varset in iN) ]] \\ 
+        &= [[ [mu] (ord varset in iP → iM) ]] \\
         &= [[ [mu] (ordVars1, (ordVars2 \ {ordVars1})) ]]
@@
           && \text{by induction on $[[ordVars2]]$;
-                                         the inductive step is similar to \cref{case:distr-mu-ord:var}.
-                                         Notice that $[[mu]]$ is} \\
-                                      & && \text{collision free on $[[{ordVars1}]]$ and $[[{ordVars2}]]$
-                                           since
+                  the inductive step is similar to 
+                  \cref{case:distr-mu-ord:var}.}\\
+        & && \text{Notice that $[[mu]]$ is collision free on $[[{ordVars1}]]$ and $[[{ordVars2}]]$} \\
+        & && \text{since
           $[[{ordVars1}]] \subseteq [[varset]]$ and
@@
       \end{aligned}$ \\
+      \hfill\\
+      On the other hand, \\
     $\begin{aligned}[t]
-       [[  (ord [mu] varset in [mu]iN) ]] &= [[ (ord [mu] varset in [mu]iP → [mu]iM) ]] \\
-                                     &= [[ (ordVarsb1, (ordVarsb2 \ {ordVarsb1})) ]]
-                                     && \text{where } [[ord [mu] varset in [mu] iP = ordVarsb1]] \text{ and } [[ord [mu] varset in [mu] iM = ordVarsb2]] \\
-                                          & && \text{then by the induction
-                                               hypothesis,
-                                               $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$,
-                                               $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$,
-                                               }\\
+       & [[  ord [mu] varset in [mu]iN ]] \\ 
+       &= [[ ord [mu] varset in [mu]iP → [mu]iM ]] \\
+       &= [[ ordVarsb1, (ordVarsb2 \ {ordVarsb1}) ]]
+       && \text{where } [[ord [mu] varset in [mu] iP = ordVarsb1]] \text{ and }  \\
+       & && [[ord [mu] varset in [mu] iM = ordVarsb2]]; 
+            \text{ then by the induction}\\
+       & && \text{hypothesis, $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$,
+                   $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$}\\
        &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]

### 2023-11-13 proofs/tex/singularity-lemmas.tex
Fix margins

@@
             $
-            \begin{aligned}[h]
-            [[Γ ⊢ iP' ≥ ∃nbs.↓iN1]] &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nf(iN1)]] \text{ where } [[ord {nbs} in iN' = nbs']] 
+            \begin{aligned}
+            [[Γ ⊢ iP' ≥ ∃nbs.↓iN1]] &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nf(iN1)]]\\
+                                    &\phantom{\Rightarrow} \text{(where $[[ord {nbs} in iN' = nbs']]$)}
                                     && \text{by \cref{corollary:nf-pres-subt}} \\
@@
                                     && \text{since $[[ord {nbs} in nf(iN1)]] = [[nbj]]$}\\
-                                   &\Rightarrow [[Γ, nbj ⊢ nf(iP') ≥ ↓nbj]] \text { and } [[nbj ∉ fv(nf(iP'))]]
+                                    &\Rightarrow [[Γ, nbj ⊢ nf(iP') ≥ ↓nbj]] \\ 
+                                    &\phantom{\Rightarrow} \text { and } [[nbj ∉ fv(nf(iP'))]]
                                     && \text{by \cref{lemma:shape-supertypes-norm}}\\
@@
                             \ruleref{\ottdruleSINGSupVarLabel},
-                            can be rewritten as $[[scE singular with nf(iP)]]$.
+                            can be rewritten as\\ $[[scE singular with nf(iP)]]$.
 

### 2023-11-13 proofs/tex/upgrade-lemmas.tex
Fix margins

@@
                   &&\text{since by \cref{lemma:fv-propagation}, 
-                         $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$ and
-                         $[[fv iQ ⊆ fv [pncs/pnas]iP]]$}\\
+                         $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$}\\
+                  & &&  \text{and $[[fv iQ ⊆ fv [pncs/pnas]iP]]$}\\
                   &\subseteq [[ ((fv iP \ {pnas}) ∪ {pnbs}) ∩ ((fv iP \ {pnas}) ∪ {pncs})]]\\

### 2023-11-14 oopsla24/tex/04-algorithm.tex
wip: correctness

@@
 \subsection{Algorithmic Syntax}
+\label{sec:algo-syntax}
 
@@
 \subsection{Subtyping Algorithm}
+\label{sec:subtyping-algorithm}
 
@@
 \subsection{Unification}
+\label{sec:unification}
 
@@
 \subsection{Constraint Merge}
+\label{sec:constraint-merge}
 
@@
 \subsection{Type Upgrade and the Least Upper Bounds}
+\label{sec:lub}
 
@@
 \subsection{Anti-Unification}
+\label{sec:antiunification}
 
@@
 \subsection{Type Inference}
+\label{sec:typing}
 
@@
 the positive type inference, the negative type inference, and the
-application type inference.
+application type inference. As subroutines, the inference algorithm uses
+subtyping, constraint merge, and normalization.  
+The corresponding graph is shown in \cref{fig:alg-typing-graph}.
+
+\begin{figure}[h]
+  \centering
+  \begin{tikzpicture}
+    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
+    % Define nodes
+    \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
+    \node[right=1.4cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
+    \node[right=1.4cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
+    \node[below of=2,xshift=+2cm] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[below of=4] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \node[left=0.75cm of 4] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
+    \draw[->] ([yshift=+.2cm]1.east) to ([yshift=+.2cm]2.west);
+    \draw[->, bend right=20] (3) to (1);
+    \draw[->] (2) to (3);
+    \draw[->] (2) to (4);
+    \draw[->] (3) to (4);
+    \draw[->, bend right] (1.south west) to (5);
+    \draw[->, bend left] (3) to (5);
+    \draw[->] (2) to (6);
+  \end{tikzpicture}  
+  \caption{Dependency graph of the typing algorithm}
+  \label{fig:alg-typing-graph}
+\end{figure}
 

### 2023-11-14 oopsla24/tex/05-proof.tex
wip: correctness

The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.
The soundness means the algorithm will always
produce a typing \emph{allowed} by the declarative system;
Dually, the completeness says that once a term has some type declaratively, the inference algorithm succeeds.

The proof of soundness and completeness result is done gradually
for all the subroutines of the algorithm,
following the 
structure of the algorithm 
(\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
bottom-up. 
\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, the correctness of normalization means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization Correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we construct an intermediate relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

\subsection{Unification Constraint Merge}

\subsection{Unification}

\begin{lemma*}[Soundness of Unification]
    \label{lemma:unification-soundness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
        $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
        such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
        $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
        $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Completeness of Unification]
    \label{lemma:unification-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
        suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
        then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.

        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
        suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
        then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
   \end{itemize}
\end{lemma*}

\subsection{Anti-Unification}

\begin{lemma*}[Soundness of Anti-Unification]
    \label{lemma:au-soundness}
    \hfill
    \begin{itemize}
        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
        then 
        \begin{enumerate}
            \item $[[Γ ; Ξ ⊢ uQ]]$,
            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
            are anti-unification substitutions, and
            \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
        \end{enumerate}

        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
        then
        \begin{enumerate}
            \item $[[Γ ; Ξ ⊢ uM]]$,
            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
            are anti-unification substitutions, and
            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
        \end{enumerate}
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Completeness of Anti-Unification]
    \label{lemma:au-completeness}
    \hfill
    \begin{itemize}
        \item [$+$] 
            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uQ']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$ 
                are anti-unification substitutions, and
                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
            \end{enumerate}

            Then the anti-unification algorithm terminates, that is there exists
            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$

        \item [$-$] 
            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uM']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
                are anti-unification substitutions, and
                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
            \end{enumerate}

            Then the anti-unification algorithm succeeds, that is 
            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Initiality of Anti-Unification]
    \hfill
    \begin{itemize}
        \item [$+$] 
            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
            if 
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uQ']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
            \end{enumerate}
            then there exists $[[ausr]]$ such that
            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
            Moreover, $[[ [ausr]β̂⁻]]$ can
            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
            $[[Γ]]$.
        \item [$-$] 
            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
            if
            \begin{enumerate}
                \item $[[Γ ; Ξ' ⊢ uM']]$,
                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
            \end{enumerate}
            then there exists $[[ausr]]$ such that
            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
            Moreover, $[[ [ausr]β̂⁻]]$ can
            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
            $[[Γ]]$.
    \end{itemize}
\end{lemma*}

\subsection{Least Upper Bound and Upgrade}

    The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
    The \emph{soundness} means that the returned type is indeed a supertype of the given ones;
    the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
    and the \emph{initiality} means that the returned type is the least among common supertypes.

    \begin{lemma*}[Least Upper Bound Soundness]
        For types $[[Γ ⊢ iP1]]$, and $[[Γ ⊢ iP2]]$,
        if $[[Γ ⊨ iP1 ∨ iP2 = iQ]]$ then
        \begin{enumerate}
            \item[(i)]  $[[Γ ⊢ iQ]]$
            \item[(ii)] $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$
        \end{enumerate}
    \end{lemma*}

    \begin{lemma*}[Least Upper Bound Completeness and Initiality]
        For types $[[Γ ⊢ iP1]]$, $[[Γ ⊢ iP2]]$, and $[[Γ ⊢ iQ]]$
        such that $[[Γ ⊢ iQ ≥ iP1]]$ and $[[Γ ⊢ iQ ≥ iP2]]$,
        there exists $[[iQ']]$ s.t. $[[Γ ⊨ iP1 ∨ iP2 = iQ']]$ 
        and $[[Γ ⊢ iQ ≥ iQ']]$.
    \end{lemma*}

    The key observation that allows us to prove these properties is
    the characterization of positive supertypes. 
    The following lemma justifies the cases of the Least Upper Bound algorithm
    (\cref{sec:lub}). In particular, it establishes
    the correspondence between the upper bounds of shifted types $[[↓iM]]$ and
    \emph{patterns} fitting $[[iM]]$ (represented by existential types), which 
    explains the usage of anti-unification as a way to find a common pattern.

    \begin{lemma*}[Characterization of Normalized Supertypes]
        \label{lemma:char-supertypes}
        For a normalized positive type $[[iP]]$ well-fordmed in $[[Γ]]$,
        the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
        \begin{itemize}
            \item if $[[iP]]$ is a variable $[[pb]]$, its only normalized supertype is $[[pb]]$ itself;
            \item if $[[iP]]$ is an existential type $[[ ∃nbs.iP' ]]$ then 
                its $[[Γ]]$-formed supertypes are the $([[G, nbs]])$-formed supertypes of $[[iP']]$ not using $[[nbs]]$;
            \item if $[[iP]]$ is a downshift type $[[↓iM]]$, 
                its supertypes have form $[[∃nas.↓iM']]$ such that there exists
                a $[[Γ]]$-formed instantiation of $[[nas]]$ in $[[↓iM']]$
                that makes $[[↓iM']]$ equal to $[[↓iM]]$, \ie $[[ [iNs/nas] ↓iM' = ↓iM ]]$.
        \end{itemize}
    \end{lemma*}

    Similarly to the Least Upper Bound algorithm, the Upgrade finds the least type among upper bounds
    (this time the set of considered upper bounds consists of supertypes well-formed in a \emph{smaller} context).
    This way, we also use the supertype characterization to prove the following property of the Upgrade algorithm.

    \begin{lemma*}[Upgrade Soundness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        if $[[upgrade Γ ⊢ iP to Δ = iQ]]$
        then
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ]]$
            \item $[[Γ ⊢ iQ ≥ iP]]$
        \end{enumerate*}
    \end{lemma*}

    \begin{lemma*}[Upgrade Completeness]
        Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
        for any $[[iQ']]$ such that $[[iQ']]$ is a $[[Δ]]$-formed upper bound of $[[iP]]$, \ie
        \begin{enumerate*}
            \item $[[Δ ⊢ iQ']]$ and
            \item $[[Γ ⊢ iQ' ≥ iP]]$,
        \end{enumerate*}
        the result of the upgrade algorithm $[[iQ]]$ exists
        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
    \end{lemma*}

\subsection{Positive Subtyping}

\begin{lemma*}[Soundness of the Positive Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
    $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
    then $[[Θ ⊢ SC : uv uP]]$ and
    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of the Positive Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{lemma*}

\subsection{Subtyping Constraint Merge}

\begin{lemma*}[Completeness of Constraint Merge]
    \label{lemma:merge-completeness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
\end{lemma*}

\begin{lemma*}[Soundness of Constraint Merge] 
    \label{lemma:merge-soundness}
    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
    Then 
    \begin{enumerate}
        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
            $[[ Θ ⊢ uσ : SC ]]$
            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
    \end{enumerate}
\end{lemma*}

\subsection{Negative Subtyping}

\begin{lemma*}[Soundness of Negative Subtyping]
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
    $[[Θ ⊢ SC : uv(uN)]]$ and 
    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of the Negative Subtyping]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{lemma*}

\subsection{Singularity}

\begin{lemma*}[Soundness of Singularity]
    \label{lemma:singularity-soundness}
    Suppose $[[Θ ⊢ SC : Ξ]]$, and $[[SC singular with uσ]]$. 
    Then $[[ Θ ⊢ uσ : Ξ ]]$,
     $[[ Θ ⊢ uσ : SC ]]$ and for any 
    $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,
    $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of Singularity]
    \label{lemma:singularity-completeness}
    For a given $[[Θ ⊢ SC]]$,
    suppose that all the substitutions satisfying $[[SC]]$ are equivalent
    on $[[Ξ]] \supseteq [[dom(SC)]]$.
    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
    Then 
    \begin{itemize}
        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
        \item $[[Ξ]] = [[dom(SC)]]$.
    \end{itemize} 
\end{lemma*}

\subsection{Type Inference}

\begin{lemma*}[Soundness of Typing]
    Suppose that $[[Γ ⊢ Φ]]$.
    For an inference tree $T_1$,
    \hfill
    \begin{itemize}
        \item [$+$] If $T_1$ infers $[[Γ; Φ ⊨ v : iP]]$ then $[[Γ ⊢ iP]]$ and $[[Γ; Φ ⊢ v : iP]]$
        \item [$-$] If $T_1$ infers $[[Γ; Φ ⊨ c : iN]]$ then $[[Γ ⊢ iN]]$ and $[[Γ; Φ ⊢ c : iN]]$
        \item  If $T_1$ infers $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$
                for $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢  uN]]$ free from negative algorithmic variables, 
                then
            \begin{enumerate}
                \item $[[Γ ⊢ Θ']]$
                \item $[[Θ ⊆ Θ']]$
                \item $[[Γ; dom(Θ') ⊢  uM]]$
                \item $[[dom(Θ) ∩ uv(uM) ⊆ uv uN]]$
                \item $[[uM]]$ is normalized and free from negative algorithmic variables
                \item $[[Θ'|uv uN ∪ uv uM ⊢ SC]]$
                \item for any $[[ Θ' ⊢ uσ : uv uN ∪ uv uM ]]$,
                    $[[ Θ' ⊢ uσ : SC ]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$
            \end{enumerate}
    \end{itemize}
\end{lemma*}

\begin{lemma*}[Completeness of Typing]
    Suppose that $[[Γ ⊢ Φ]]$.
    For an inference tree $T_1$,
    \begin{itemize}
        \item [$+$] If $T_1$ infers $[[Γ; Φ ⊢ v : iP]]$ then $[[Γ; Φ ⊨ v : nf(iP)]]$        
        \item [$-$] If $T_1$ infers $[[Γ; Φ ⊢ c : iN]]$ then  $[[Γ; Φ ⊨ c : nf(iN)]]$
        \item [$\bullet$] If 
            $T_1$ infers $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$
            and 
            \begin{enumerate}
                \item $[[Γ ⊢ Θ]]$, 
                \item $[[Γ ⊢ iM]]$,
                \item $[[Γ; dom(Θ) ⊢ uN]]$ (free from negative algorithmic variables, that is $[[α̂⁻]] \notin [[uv uN]]$), and
                \item $[[Θ ⊢ uσ : uv(uN)]]$,
            \end{enumerate}
            then there exist $[[uM']]$, $[[Θ']]$, and $[[SC]]$ such that
            \begin{enumerate}
                \item $[[ Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC ]]$ and
                \item for any $[[Θ ⊢ uσ : uv(uN)]]$ and $[[Γ ⊢ iM]]$
                    such that $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$, 
                    there exists $[[uσ']]$ such that 
                    \begin{enumerate}
                        \item $[[Θ' ⊢ uσ' : uv uN ∪ uv uM]]$
                            and $[[Θ' ⊢ uσ' : SC]]$,
                        \item $[[Θ ⊢ uσ' ≈ uσ : uv uN]]$, and 
                        \item $[[Γ ⊢ [uσ']uM' ≈ iM]]$.
                    \end{enumerate}
            \end{enumerate}
    \end{itemize}
\end{lemma*}

\subsection{Determinicity}

### 2023-11-21 oopsla24/tex/05-proof.tex
wip

@@
 
+The formulation of soundness and completeness of \emph{application inference} is more complex.
+Both of them assume that the input head type $[[uN]]$ is free from \emph{negative}
+algorithmic variables (it is achieved by polarization invariants preserved by the inference rules).
+The soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$ is viable. 
+Specifically, that the constraint set $[[SC]]$ provides a sufficient set of restrictions that 
+a substitution $[[uσ]]$ must satisfy to ensure the \emph{declarative} inference of the output type $[[uM]]$, that is 
+$[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$.
+
+The application inference completeness means that if 
+there exists a substitution $[[uσ]]$ and the resulting type $[[iM]]$
+ensuring the declarative inference $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$ then the algorithm succeeds
+and gives the most general result $[[uM0]]$ and $[[SC0]]$.
+The property of `being the most general' is specified in pt. (\ref{point:mostGeneral}).
+Intuitively, it means that for any other solution---substitution $[[uσ]]$ and the resulting type $[[iM]]$,
+if it ensures the declarative inference, then $[[uσ]]$ can be extended in a $[[SC0]]$-complying way to equate 
+$[[uM0]]$ with $[[iM]]$.
+
+\footnotetext[1]{The presented properties hold, 
+            but the actual inductive proof requires strengthening 
+            of the statement and the corresponding theorem is more involved. 
+            See the appendix for details.}
+
+\begin{theorem*}[Soundness of Typing]
+    \label{thm:soundness-typing}
+    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
+    \hfill
+    \begin{itemize}
+        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
+        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
+        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
+            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$,
+            assuming $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
+            \emph{negative} algorithmic variables.
+    \end{itemize}
+\end{theorem*}
+
+\begin{theorem*}[Completeness of Typing]
+    \label{thm:completeness-typing}
+    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
+    \begin{itemize}
+        \item [$+$] $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$,
+        \item [$-$] $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$,
+        \item [$\bullet$] If $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$
+            where 
+            \begin{enumerate*}
+                \item $[[Γ ⊢ Θ]]$, 
+                \item $[[Γ ⊢ iM]]$,
+                \item $[[Γ; dom(Θ) ⊢ uN]]$ (free from \emph{negative} algorithmic variables), and
+                \item $[[Θ ⊢ uσ : uv(uN)]]$,
+            \end{enumerate*}
+            then there exist $[[uM0]]$, $[[Θ0]]$, and $[[SC0]]$ such that
+            \begin{enumerate}
+                \item $[[ Γ; Φ; Θ ⊨ uN ● args ⇒> uM0 ⫤ Θ0; SC0 ]]$ and
+                \item \label{point:mostGeneral} for any other $[[uσ]]$ and $[[iM]]$ 
+                (where $[[Θ ⊢ uσ : uv(uN)]]$ and $[[Γ ⊢ iM]]$)
+                    such that $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$, 
+                    there exists $[[uσ']]$ such that 
+                    \begin{enumerate*}
+                        \item $[[Θ0 ⊢ uσ' : uv uN ∪ uv uM0]]$
+                            and $[[Θ0 ⊢ uσ' : SC0]]$,
+                        \item $[[Θ ⊢ uσ' ≈ uσ : uv uN]]$, and 
+                        \item $[[Γ ⊢ [uσ']uM0 ≈ iM]]$.
+                    \end{enumerate*}
+            \end{enumerate}
+    \end{itemize}
+\end{theorem*}
+
 The proof of soundness and completeness result is done gradually
-for all the subroutines of the algorithm,
-following the 
-structure of the algorithm 
+for all the subroutines,
+following the structure of the algorithm 
 (\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
-bottom-up. 
+bottom-up. Next, we discuss the main of these results.
+
 \subsection{Normalization}
@@
     To prove the correctness of normalization, 
-    we construct an intermediate relation on types---\emph{declarative equivalence}
+    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
     (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
@@
 
-\subsection{Unification Constraint Merge}
+% \subsection{Unification Constraint Merge}
 
-\subsection{Unification}
+% \subsection{Unification}
 
-\begin{lemma*}[Soundness of Unification]
-    \label{lemma:unification-soundness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
-        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
-        if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
-        $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
-        such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.
+% \begin{lemma*}[Soundness of Unification]
+%     \label{lemma:unification-soundness}
+%     \hfill
+%     \begin{itemize}
+%         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that 
+%         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$,\\ 
+%         if $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ then 
+%         $[[Θ ⊢ UC : uv uP]]$ and for any normalized $[[uσ]]$ 
+%         such that $[[ Θ ⊢ uσ : lift UC ]]$, $[[ [uσ]uP = iQ ]]$.
 
-        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
-        if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
-        $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
-        $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
-    \end{itemize}
-\end{lemma*}
+%         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
+%         $[[Γ ; dom(Θ) ⊢ uN]]$ and $[[Γ ⊢ iM]]$,\\
+%         if $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ then 
+%         $[[Θ ⊢ UC : uv uN]]$ and for any normalized $[[uσ]]$ such that
+%         $[[ Θ   ⊢ uσ : lift UC ]]$, $[[ [uσ]uN = iM ]]$.
+%     \end{itemize}
+% \end{lemma*}
 
-\begin{lemma*}[Completeness of Unification]
-    \label{lemma:unification-completeness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
-        $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
-        suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
-        then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.
+% \begin{lemma*}[Completeness of Unification]
+%     \label{lemma:unification-completeness}
+%     \hfill
+%     \begin{itemize}
+%         \item [$+$] For normalized $[[uP]]$ and $[[iQ]]$ such that
+%         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
+%         suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
+%         then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.
 
-        \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
-        $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
-        suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
-        then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
-   \end{itemize}
-\end{lemma*}
+%         \item [$-$] For normalized $[[uN]]$ and $[[iM]]$ such that
+%         $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
+%         suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
+%         then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
+%    \end{itemize}
+% \end{lemma*}
 
@@
 
-\begin{lemma*}[Soundness of Anti-Unification]
+The anti-unifier of the two types is the most specific pattern that matches against both of them. 
+In our setting, the anti-unifiers are restricted further: first, the pattern might only contain placeholders at 
+\emph{negative} positions (because eventually, the placeholders become variables abstracted by the existential quantifier);
+second, the anti-unification is parametrized with a context $[[Γ]]$, in which the pattern instantiations must be well-formed.
+
+This way, the correctness properties of the anti-unification algorithm are refined accordingly.
+The soundness of anti-unification not only ensures that the resulting pattern matches with 
+the input types, but also that the pattern instantiations are well-formed in the corresponding context, and 
+that all the `placeholder' variables are negative. In turn, completeness states that if there exists a solution 
+satisfying the soundness criteria, then the algorithm succeeds.
+
+The correctness properties are formulated by the following lemmas. For brevity, we only provide the statements for 
+the positive case, since the negative case is symmetric.
+
+\begin{lemma*}[Soundness of (Positive) Anti-Unification]
     \label{lemma:au-soundness}
-    \hfill
-    \begin{itemize}
-        \item [$+$]  Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+     Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
     if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
     then 
-        \begin{enumerate}
+    \begin{enumerate*}
         \item $[[Γ ; Ξ ⊢ uQ]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification substitutions, and
+        \item $[[Γ ; · ⊢ aus1 : Ξ]]$ for $i \in \{1,2\}$
+        are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
         \item $[[ [ausi] uQ = iPi ]]$ for $i \in \{1,2\}$.
-        \end{enumerate}
-
-        \item [$-$] Assuming $[[iN1]]$ and $[[iN2]]$ are normalized,
-        if $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$
-        then
-        \begin{enumerate}
-            \item $[[Γ ; Ξ ⊢ uM]]$,
-            \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
-            are anti-unification substitutions, and
-            \item $[[ [ausi] uM = iNi ]]$ for $i \in \{1,2\}$.
-        \end{enumerate}
-    \end{itemize}
+    \end{enumerate*}
 \end{lemma*}
 
-\begin{lemma*}[Completeness of Anti-Unification]
+\begin{lemma*}[Completeness of (Positive) Anti-Unification]
     \label{lemma:au-completeness}
-    \hfill
-    \begin{itemize}
-        \item [$+$] 
-            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
-            there exists $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
-            \begin{enumerate}
+    Assuming that $[[iP1]]$ and $[[iP2]]$ are normalized terms well-formed in $[[Γ]]$
+    and there exist  $[[(Ξ', uQ', aus'1, aus'2)]]$ such that
+    \begin{enumerate*}
         \item $[[Γ ; Ξ' ⊢ uQ']]$,
@@
         \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$.
-            \end{enumerate}
+    \end{enumerate*}
 
     Then the anti-unification algorithm terminates, that is there exists
-            $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
-
-        \item [$-$] 
-            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
-            there exists $[[(Ξ', uM', aus'1, aus'2)]]$ such that
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$,
-                are anti-unification substitutions, and
-                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$.
-            \end{enumerate}
-
-            Then the anti-unification algorithm succeeds, that is 
-            there exists $[[(Ξ, uM, aus1, aus2)]]$ such that
-            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$.
-    \end{itemize}
+    $[[(Ξ, uQ, aus1, aus2)]]$ such that $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$.
 \end{lemma*}
 
+The anti-unification is used to find the least upper bound (LUB).
+To guarantee that the result of the LUB algorithm is indeed the least, 
+we also need to show the corresponding property of anti-unification.
+We prove that the anti-unifier that the algorithm provides 
+is the most specific (or the most `detailed'). Specifically, 
+it means that any other sound anti-unification solution 
+can be `refined' to the result of the algorithm.  
+The `refinement' is represented as an instantiation of the anti-unifier---a 
+substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the placeholders $[[Ξ1]]$ with
+types that themselves can contain placeholders from $[[Ξ2]]$.
+
 \begin{lemma*}[Initiality of Anti-Unification]
-    \hfill
-    \begin{itemize}
-        \item [$+$] 
-            Assume that $[[iP1]]$ and $[[iP2]]$ are normalized, and
-            $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$, 
-            then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
-            any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, i.e.
-            if 
-            \begin{enumerate}
+        Assume that $[[iP1]]$ and $[[iP2]]$ are normalized types well-formed in $[[Γ]]$, 
+        and the anti-unification algorithm succeeds: $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$. 
+        Then $[[(Ξ, uQ, aus1, aus2)]]$ is more specific than
+        any other sound anti-unifier $[[(Ξ', uQ', aus'1, aus'2)]]$, \ie if
+        \begin{enumerate*}
             \item $[[Γ ; Ξ' ⊢ uQ']]$,
@@
             \item $[[ [aus'i] uQ' = iPi ]]$ for $i \in \{1,2\}$
-            \end{enumerate}
-            then there exists $[[ausr]]$ such that
+        \end{enumerate*}
+        then there exists a `refining' substitution $[[ausr]]$ such that
         $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uQ')]]$ and $[[ [ausr] uQ' = uQ ]]$. 
-            Moreover, $[[ [ausr]β̂⁻]]$ can
-            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
-            $[[Γ]]$.
-        \item [$-$] 
-            Assume that $[[iN1]]$ and $[[iN2]]$ are normalized, and
-            $[[Γ ⊨ iN1 ≈au iN2 ⫤ (Ξ, uM, aus1, aus2)]]$, 
-            then $[[(Ξ, uM, aus1, aus2)]]$ is more specific than
-            any other sound anti-unifier $[[(Ξ', uM', aus'1, aus'2)]]$, i.e.
-            if
-            \begin{enumerate}
-                \item $[[Γ ; Ξ' ⊢ uM']]$,
-                \item $[[Γ ; · ⊢ aus'i : Ξ']]$ for $i \in \{1,2\}$, and
-                \item $[[ [aus'i] uM' = iNi ]]$ for $i \in \{1,2\}$
-            \end{enumerate}
-            then there exists $[[ausr]]$ such that
-            $[[Γ ; Ξ ⊢ ausr : (Ξ' | uv uM')]]$ and $[[ [ausr] uM' = uM ]]$.
-            Moreover, $[[ [ausr]β̂⁻]]$ can
-            be uniquely determined by $[[ [aus'1]β̂⁻ ]]$, $[[ [aus'2]β̂⁻ ]]$, and
-            $[[Γ]]$.
-    \end{itemize}
+        % Moreover, $[[ [ausr]β̂⁻]]$ is uniquely determined by $[[ [aus'1]β̂⁻ ]]$, 
+        % $[[ [aus'2]β̂⁻ ]]$, and $[[Γ]]$.
+\end{lemma*}
+
+To prove the correctness properties of the anti-unification algorithm,
+one extra observation is essential. The algorithm relies on the fact that
+in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
+there are no two different placeholders $[[β̂⁻]]$
+mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
+This is important to guarantee that, for example, 
+the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
+$[[↓↑Bool → ↑Bool]]$ is $[[↓â⁻ → â⁻]]$, but not
+(less specific) $[[↓â⁻ → b̂⁻]]$.
+
+\begin{lemma*}[Uniqueness of Anti-unification Variable Names]
+    Names of the anti-unification variables are uniquely defined by
+    the types they are mapped to by the resulting substitutions. 
+    Assuming $[[iP1]]$ and $[[iP2]]$ are normalized,
+        if $[[Γ ⊨ iP1 ≈au iP2 ⫤ (Ξ, uQ, aus1, aus2)]]$
+        then for any $[[β̂⁻]] \in [[Ξ]]$,
+        $[[β̂⁻]] = [[â⁻_{[aus1]β̂⁻, [aus2]β̂⁻}]]$.
 \end{lemma*}
@@
 
-\subsection{Positive Subtyping}
+\subsection{Subtyping}
+
+    As for other properties, the correctness of subtyping means that
+    the algorithm produces a valid result (soundness) 
+    whenever it exists (completeness).
+    In \cref{fig:subtyping-algo}, one can see that the positive and 
+    negative subtyping relations are not \emph{mututally} recursive: 
+    negative subtyping algorithm uses the positive subtyping, 
+    but not vice versa. 
+    Because of that, the inductive proofs of the \emph{positive} subtyping correctness are done independently.
+
+    The soundness of positive subtyping states that
+    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
+    In other words, any substitution satisfying $[[SC]]$, 
+    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
+    Notice that the soundness requires that only the left-hand side input 
+    type ($[[uP]]$) is algorithmic. This is one of the invariants of the algorithm
+    that significantly simplifies the unification and constraint resolution.
 
@@
     then $[[Θ ⊢ SC : uv uP]]$ and
-    for any normalized $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
-    $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
+    for any  $[[uσ]]$ such that $[[ Θ ⊢ uσ : SC ]]$,
+    we have $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$.
 \end{lemma*}
+
+    The completeness of subtyping says that if the substitution 
+    ensuring the declarative subtyping exists, then the algorithm terminates.
 
@@
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
-    Then for any $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
-    there exists $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
+    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv(uP)]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
+    then the subtyping algorithm succeeds: $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
 \end{lemma*}
 
-\subsection{Subtyping Constraint Merge}
-
-\begin{lemma*}[Completeness of Constraint Merge]
-    \label{lemma:merge-completeness}
-    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$.
-    If there exists a substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$ such that 
-    $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$
-    then $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
-\end{lemma*}
-
-\begin{lemma*}[Soundness of Constraint Merge] 
-    \label{lemma:merge-soundness}
-    Suppose that $[[Θ ⊢ SC1 : Ξ1]]$ and $[[Θ ⊢ SC2 : Ξ2]]$ 
-    and $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined.
-    Then 
-    \begin{enumerate}
-        \item $[[Θ ⊢ SC : Ξ1 ∪ Ξ2]]$,
-        \item for any substitution $[[Θ ⊢ uσ : Ξ1 ∪ Ξ2]]$, 
-            $[[ Θ ⊢ uσ : SC ]]$
-            implies $[[ Θ ⊢ uσ : SC1 ]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
-    \end{enumerate}
-\end{lemma*}
-
-\subsection{Negative Subtyping}
-
-\begin{lemma*}[Soundness of Negative Subtyping]
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
-    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
-    $[[Θ ⊢ SC : uv(uN)]]$ and 
-    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
-    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
-\end{lemma*}
+After the correctness properties of the positive subtyping are established,
+they are used to prove the correctness of the negative subtyping.
+The soundness is formulated symmetrical to the positive case,
+however, the completeness requires an additional invariant to be preserved.
+The algorithmic input type $[[uN]]$ must be free from \emph{negative} algorithmic variables.
+In particular, it ensures that the constraint restricting a \emph{negative} algorithmic
+variable will never be generated, and thus, we do not need the 
+Greatest Common Subtype procedure to resolve the constraints.
 
@@
     Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
-    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
+    the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.
 \end{lemma*}
@@
 \subsection{Singularity}
+
+The rule \ruleref{\ottdruleDTAppLetLabel} of the declarative inference
+says that one can omit the type annotation of a let-binding if the inferred 
+type of the bound application is unique. Accordingly, the algorithm,
+that returns the constraint set, must be able to check that
+the set of constraints has a single solution, and also to find this solution. 
+This subroutine is called singularity, and its correctness is formulated as follows.
+
+The soundness states that $[[SC singular with uσ]]$
+implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
+on the domain.  The completeness states that if all the 
+$[[SC]]$-compliant substitutions are equivalent, then the singularity procedure succeeds.
 
@@
     suppose that all the substitutions satisfying $[[SC]]$ are equivalent
-    on $[[Ξ]] \supseteq [[dom(SC)]]$.
-    In other words, suppose that there exists $[[Θ ⊢ uσ1 : Ξ]]$ such that
-    for any $[[Θ ⊢ uσ : Ξ]]$, $[[Θ ⊢ uσ : SC]]$ implies 
-    $[[Θ ⊢ uσ ≈ uσ1 : Ξ]]$.
-    Then 
+    on $[[dom(SC)]]$.
+    In other words, suppose that there exists $[[Θ ⊢ uσ1 : dom(SC)]]$ such that
+    for any $[[Θ ⊢ uσ : dom(SC)]]$, $[[Θ ⊢ uσ : SC]]$ implies 
+    $[[Θ ⊢ uσ ≈ uσ1 : dom(SC)]]$.
+    Then\footnotemark[1] $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
+\end{lemma*}
+
+\subsection{Typing}
+
+Finally, we discuss the proofs of the theorems stated at the beginning of this section.
+
+\paragraph{Determinicity}
+    One of the properties that our proof relies on is the determinicity of the 
+    typing algorithm: the output (the type, the set of restrictions, \etc) 
+    is uniquely determined by the input (the term, the context, \etc).
+    Determinicity is not hard to demonstrate by structural induction: 
+    in every algorithmic inference system, only one inference rule can be applied 
+    for a given input.  However, we need to prove it for \emph{every} 
+    subroutine of the algorithm. Ultimately, it requires 
+    the determinicity of such procedures as \emph{generation of fresh variables}, 
+    which is easy to ensure, but must be taken into account in the implementation.
+
+\begin{lemma*}[Determinicity of the Typing Algorithm]
+    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
     \begin{itemize}
-        \item $[[SC singular with uσ0]]$ for some $[[uσ0]]$ and
-        \item $[[Ξ]] = [[dom(SC)]]$.
+        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
+        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
+        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
+            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
+            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
     \end{itemize}
@@
 
-\subsection{Type Inference}
+\paragraph{Mutual Induction of Soundness and Completeness}
 
-\begin{lemma*}[Soundness of Typing]
-    Suppose that $[[Γ ⊢ Φ]]$.
-    For an inference tree $T_1$,
-    \hfill
-    \begin{itemize}
-        \item [$+$] If $T_1$ infers $[[Γ; Φ ⊨ v : iP]]$ then $[[Γ ⊢ iP]]$ and $[[Γ; Φ ⊢ v : iP]]$
-        \item [$-$] If $T_1$ infers $[[Γ; Φ ⊨ c : iN]]$ then $[[Γ ⊢ iN]]$ and $[[Γ; Φ ⊢ c : iN]]$
-        \item  If $T_1$ infers $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$
-                for $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢  uN]]$ free from negative algorithmic variables, 
-                then
-            \begin{enumerate}
-                \item $[[Γ ⊢ Θ']]$
-                \item $[[Θ ⊆ Θ']]$
-                \item $[[Γ; dom(Θ') ⊢  uM]]$
-                \item $[[dom(Θ) ∩ uv(uM) ⊆ uv uN]]$
-                \item $[[uM]]$ is normalized and free from negative algorithmic variables
-                \item $[[Θ'|uv uN ∪ uv uM ⊢ SC]]$
-                \item for any $[[ Θ' ⊢ uσ : uv uN ∪ uv uM ]]$,
-                    $[[ Θ' ⊢ uσ : SC ]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$
-            \end{enumerate}
-    \end{itemize}
-\end{lemma*}
-
-\begin{lemma*}[Completeness of Typing]
-    Suppose that $[[Γ ⊢ Φ]]$.
-    For an inference tree $T_1$,
-    \begin{itemize}
-        \item [$+$] If $T_1$ infers $[[Γ; Φ ⊢ v : iP]]$ then $[[Γ; Φ ⊨ v : nf(iP)]]$        
-        \item [$-$] If $T_1$ infers $[[Γ; Φ ⊢ c : iN]]$ then  $[[Γ; Φ ⊨ c : nf(iN)]]$
-        \item [$\bullet$] If 
-            $T_1$ infers $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$
-            and 
-            \begin{enumerate}
-                \item $[[Γ ⊢ Θ]]$, 
-                \item $[[Γ ⊢ iM]]$,
-                \item $[[Γ; dom(Θ) ⊢ uN]]$ (free from negative algorithmic variables, that is $[[α̂⁻]] \notin [[uv uN]]$), and
-                \item $[[Θ ⊢ uσ : uv(uN)]]$,
-            \end{enumerate}
-            then there exist $[[uM']]$, $[[Θ']]$, and $[[SC]]$ such that
-            \begin{enumerate}
-                \item $[[ Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC ]]$ and
-                \item for any $[[Θ ⊢ uσ : uv(uN)]]$ and $[[Γ ⊢ iM]]$
-                    such that $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$, 
-                    there exists $[[uσ']]$ such that 
-                    \begin{enumerate}
-                        \item $[[Θ' ⊢ uσ' : uv uN ∪ uv uM]]$
-                            and $[[Θ' ⊢ uσ' : SC]]$,
-                        \item $[[Θ ⊢ uσ' ≈ uσ : uv uN]]$, and 
-                        \item $[[Γ ⊢ [uσ']uM' ≈ iM]]$.
-                    \end{enumerate}
-            \end{enumerate}
-    \end{itemize}
-\end{lemma*}
-
-\subsection{Determinicity}
+The soundness and completeness of type inference cannot be proven separately:
+the inductive proof of one requires another, and vice versa. 
+Soundness proof requires completeness in the case of 
+\ruleref{\ottdruleATAppLetLabel},
+The soundness proof can be considered as a transformation from the algorithmic 
+inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
+the declarative rule requires that the (declaratively) inferred type of the application
+is unique.  To show that, we take an arbitrary application typing tree and apply the typing
+\emph{completeness}. Then the soundness of singularity

### 2023-11-21 oopsla24/tex/05-proof.tex

@@
 \subsection{Singularity}
+\label{sec:singularity-proof}
 
@@
 
-\paragraph{Mutual Induction of Soundness and Completeness}
+\paragraph{Mutuality of the Soundness and Completeness Proofs}
 
-The soundness and completeness of type inference cannot be proven separately:
-the inductive proof of one requires another, and vice versa. 
-Soundness proof requires completeness in the case of 
-\ruleref{\ottdruleATAppLetLabel},
-The soundness proof can be considered as a transformation from the algorithmic 
-inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
-the declarative rule requires that the (declaratively) inferred type of the application
-is unique.  To show that, we take an arbitrary application typing tree and apply the typing
-\emph{completeness}. Then the soundness of singularity
+Typically in our inductive proofs, the soundness is proven before completeness,
+as the completeness requires certain properties of the premise subtrees.
+However, in the case of the typing algorithm, the soundness and completeness proofs
+cannot be separated: the inductive proof of one requires another, and vice versa.
+
+The soundness proof requires completeness in the case of 
+\ruleref{\ottdruleATAppLetLabel}, whose declarative counterpart requires \emph{uniqueness}
+of the inferred application type. To show the uniqueness, we
+need to conclude that any other \emph{declarative} tree 
+must infer the same (or rather equivalent) type. 
+The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proo}),
+however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
+
+This way, these lemmas are proven by mutual induction. 
+Since the declarative and the algorithmic systems are not mutually inductive,
+we must introduce a metric on which the induction is conducted.
+
+\paragraph{Inductive Metric}
+
+Let us extend the declarative typing metric (\cref{def:decl-typing-metric}) 
+to the algorithmic typing.
+
+\begin{definition}[Number of prenex quantifiers]
+    We define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of 
+    prenex quantifiers, i.e.
+    \begin{itemize}
+        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
+        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
+    \end{itemize}
+\end{definition}
+
+\begin{definition}[Size of a Declarative Judgement]
+    \label{def:decl-typing-size}
+    For a declarative typing judgment $J$
+    let us define a metrics $\size{J}$ as a pair of numbers 
+    in the following way:
+    \begin{itemize}
+        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]} = 
+            (\size{[[args]]}, \npq{[[iN]]})$)
+    \end{itemize}
+    where $\size{[[v]]}$ or $\size{[[c]]}$ is the size of the 
+    syntax tree of the term $[[v]]$ or $[[c]]$
+    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$.
+\end{definition}
+
+\begin{definition}[Number of Equivalence Nodes]
+    For a tree $T$ inferring
+    a declarative typing judgment,
+    let us a function $\eqNodes{T}$
+    as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 
+    \ruleref{\ottdruleDTNEquivLabel}.
+\end{definition}
+
+\begin{definition}[Declarative Tree Metric]
+    \label{def:decl-typing-metric}
+    For a tree $T$ inferring
+    a declarative typing judgment $J$,
+    we define a metric $\metric{T}$
+    as a pair $(\size{J}, \eqNodes{T})$.
+\end{definition}
+
+\begin{definition}[Size of an Algorithmic Judgement]
+    For an algorithmic typing judgement $J$
+    let us define a metrics $\size{J}$ as a pair of numbers 
+    in the following way:
+    \begin{itemize}
+        \item [$+$] $\size{[[Γ ; Φ ⊨ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊨ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]} = 
+            (\size{[[args]]}, \npq{[[uN]]})$)
+    \end{itemize}
+\end{definition}
+
+\begin{definition}[Algorithmic Tree Metric]
+    For a tree $T$ inferring an algorithmic typing judgement $J$, we define 
+    $\metric{T}$ as $(\size{J}, 0)$.
+\end{definition}

### 2023-11-27 oopsla24/tex/05-proof.tex
Finish proofs

--- 2023-11-21 oopsla24/tex/05-proof.tex
@@
         \item $[[Γ ; Ξ ⊢ uQ]]$,
-        \item $[[Γ ; · ⊢ aus1 : Ξ]]$ for $i \in \{1,2\}$
+        \item $[[Γ ; · ⊢ ausi : Ξ]]$ for $i \in \{1,2\}$
         are anti-unification substitutions (in particular, $[[Ξ]]$ contains only negative algorithmic variables), and
@@
 (less specific) $[[↓â⁻ → b̂⁻]]$.
+To preserve this property, we arrange the algorithm in such a way 
+that the name of the placeholder is determined by the types it is mapped to.
+The following lemma specifies this observation.
 
@@
         \label{lemma:char-supertypes}
-        For a normalized positive type $[[iP]]$ well-fordmed in $[[Γ]]$,
+        For a normalized positive type $[[iP]]$ well-formed in $[[Γ]]$,
         the set of normalized $[[Γ]]$-formed supertypes of $[[iP]]$ is the following:
@@
     (this time the set of considered upper bounds consists of supertypes well-formed in a \emph{smaller} context).
-    This way, we also use the supertype characterization to prove the following property of the Upgrade algorithm.
+    This way, we also use the supertype characterization to prove the following properties of the Upgrade algorithm.
 
@@
         \end{enumerate*}
-        the result of the upgrade algorithm $[[iQ]]$ exists
-        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and satisfies $[[Δ ⊢ iQ' ≥ iQ]]$.
+        there exists$[[iQ]]$ such that
+        ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and $[[Δ ⊢ iQ' ≥ iQ]]$.
     \end{lemma*}
@@
     Notice that the soundness requires that only the left-hand side input 
-    type ($[[uP]]$) is algorithmic. This is one of the invariants of the algorithm
+    type ($[[uP]]$) can have algorithmic variables. This is one of the invariants of the algorithm
     that significantly simplifies the unification and constraint resolution.
@@
 \begin{lemma*}[Soundness of the Positive Subtyping]
-    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$, $[[Γ ; dom(Θ) ⊢  uP]]$, and 
+    If $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$ and 
     $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$,
@@
 \begin{lemma*}[Completeness of the Negative Subtyping]
-    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
+    Suppose that $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iM]]$ and $[[Γ ; dom(Θ) ⊢ uN]]$
     and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
@@
 
-Finally, we discuss the proofs of the theorems stated at the beginning of this section.
+Finally, we discuss the proofs of the soundness and completeness of type
+inference algorithm that we stated at the beginning of this section.
+There are three subtleties that we will cover that are important 
+for the proof to go through: the determinacy of the algorithm,
+the mutual dependence of the soundness and completeness proofs, and
+the non-trivial inductive metric that we use.
 
-\paragraph{Determinicity}
-    One of the properties that our proof relies on is the determinicity of the 
-    typing algorithm: the output (the type, the set of restrictions, \etc) 
-    is uniquely determined by the input (the term, the context, \etc).
-    Determinicity is not hard to demonstrate by structural induction: 
+\paragraph{Determinacy}
+    One of the properties that our proof relies on is the determinacy of the 
+    typing algorithm: the output (the inferred type)
+    is uniquely determined by the input (the term and the contexts).
+    Determinacy is not hard to demonstrate by structural induction: 
     in every algorithmic inference system, only one inference rule can be applied 
@@
     subroutine of the algorithm. Ultimately, it requires 
-    the determinicity of such procedures as \emph{generation of fresh variables}, 
+    the determinacy of such procedures as \emph{generation of fresh variables}, 
     which is easy to ensure, but must be taken into account in the implementation.
 
-\begin{lemma*}[Determinicity of the Typing Algorithm]
+\begin{lemma*}[Determinacy of the Typing Algorithm]
     Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
@@
 
-\paragraph{Mutual Induction of Soundness and Completeness}
+\paragraph{Mutuality of the Soundness and Completeness Proofs}
 
-The soundness and completeness of type inference cannot be proven separately:
-the inductive proof of one requires another, and vice versa. 
-Soundness proof requires completeness in the case of 
-\ruleref{\ottdruleATAppLetLabel},
-The soundness proof can be considered as a transformation from the algorithmic 
-inference tree to a declarative one. In the case of \ruleref{\ottdruleATAppLetLabel},
-the declarative rule requires that the (declaratively) inferred type of the application
-is unique.  To show that, we take an arbitrary application typing tree and apply the typing
-\emph{completeness}. Then the soundness of singularity
+Typically in our inductive proofs, the soundness is proven before completeness,
+as the completeness requires certain properties of the premise subtrees.
+However, in the case of the typing algorithm, the soundness and completeness proofs
+cannot be separated: the inductive proof of one requires another, and vice versa.
+
+The soundness proof can be viewed as a mapping from an algorithmic tree to a
+declarative one. We show that each algorithmic inference rule can
+be transformed into the corresponding declarative rule, as long as the premises
+are transformed accordingly, and apply the induction principle.
+
+The soundness requires completeness in the case of
+\ruleref{\ottdruleATAppLetLabel}. To prove the soundness, 
+in other words, to transform this rule into its
+declarative counterpart \ruleref{\ottdruleDTAppLetLabel}, one needs
+to prove \emph{uniqueness} of the inferred application type $[[↑iQ]]$. 
+To show the uniqueness, we need to conclude that any other \emph{declarative} tree
+(which is given by the induction hypothesis) infers an equivalent type. 
+The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proo}), however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
+This way, the soundness and completeness proofs are mutually dependent.
+
+\paragraph{Inductive Metric}
+    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
+    Since the declarative and the algorithmic systems do not depend on each other,
+    we must introduce a uniform \emph{metric} on which the induction is conducted.
+    We define the metric gradually, starting from the
+    auxiliary function---the size of a judgment $\size{J}$.
+
+    The size of \emph{declarative} and \emph{algorithmic} judgments is defined as
+    a pair: the first component is the syntactic size of the terms used in the judgment. 
+    The second component depends on the kind of judgment. For regular
+    type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$
+    or $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments
+    $([[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), 
+    it is equal to the number of prenex quantifiers of the head type $[[iN]]$.
+    We need this adjustment to ensure the monotonicity of the metric, since the rules 
+    \ruleref{\ottdruleATForallAppLabel} and \ruleref{\ottdruleDTForallAppLabel}
+    only reduce the quantifiers in the head type but not change the list of arguments.
+
+\begin{definition*}[Judgement Size]
+    \label{def:decl-typing-size}
+    For a declarative or an algorithmic typing judgment $J$, we define a metric $\size{J}$ as a pair of integers in the following way:
+
+    \begin{multicols}{2}
+    \begin{itemize}
+        \item [$+$] $\size{[[Γ ; Φ ⊢ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊢ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ⊢ iN ● args ⇒> iM]]}=$\\ 
+            $(\size{[[args]]}, \npq{[[iN]]})$;
+    \end{itemize}
+    \columnbreak
+    \begin{itemize}[leftmargin=*]
+        \item [$+$] $\size{[[Γ ; Φ ⊨ v : iP]]} = (\size{[[v]]}, 0)$;
+        \item [$-$] $\size{[[Γ ; Φ ⊨ c : iN]]} = (\size{[[c]]}, 0)$;
+        \item [$\bullet$] $\size{[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]} =$\\ 
+             $(\size{[[args]]}, \npq{[[uN]]})$.
+    \end{itemize}
+    \end{multicols}
+
+    Here $\size{[[v]]}$ and $\size{[[c]]}$ is the size of the 
+    syntax tree of the term,
+    and $\size{[[args]]}$ is the sum of sizes of the terms in $[[args]]$;
+    and $\npq{[[iN]]}$ and $\npq{[[iP]]}$ represent the number of 
+    prenex quantifiers, \ie
+        \begin{itemize} \centering
+            \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
+            \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
+        \end{itemize}
+\end{definition*}
+
+Notice that for \emph{algorithmic} inference system, $\size{J}$ decreases in
+all the inductive steps, \ie for each inference rule, the size of the
+premise judgments is strictly less than the size of the conclusion.
+However, the \emph{declarative} inference system has rules
+\ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}, that
+`step to' an equivalent type, and thus, technically, might keep the judgment
+unchanged altogether.
+
+To deal with this issue, we introduce the metric on the entire 
+\emph{inference trees} rather than on judgments, and plug into this metric the
+parameter that certainly decreases in rules \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}---the number of these rules in the inference tree. We
+denote this number as $\eqNodes{T}$.  Then the final metric is defined as a pair in the following way.
+
+\begin{definition*}[Inference Tree Metric]
+    For a tree $T$, inferring a declarative or an algorithmic judgement $J$, we define $\metric{T}$ as follows:
+        \[
+        \metric{T} = \begin{cases}
+        (\size{J}, \eqNodes{T}) & \text{if } J \text{ represents a declarative judgement}, \\
+        (\size{J}, 0) & \text{if } J \text{ represents an algorithmic judgement}.
+        \end{cases}
+        \]
+    Here $\eqNodes{T}$ is the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or \ruleref{\ottdruleDTNEquivLabel}.
+\end{definition*}
+
+This metric is suitable for mutual induction on the soundness and completeness
+of the typing algorithm.  First, it is monotonous \wrt the inference rules,
+and this way, we can always apply the induction hypothesis to premises 
+of each rule. Second, the induction hypothesis is powerful enough,
+so we can use the completeness of the algorithm in the soundness proof,
+where required. For instance, to prove the soundness
+of typing in case of $[[Γ; Φ ⊨ let x = v(args); c' : iN]]$,
+we can assume, that  \emph{completeness}
+holds for a term of shape $[[Γ; Φ ⊢ iM ● args ⇒> iK ]]$, since
+$\size{args} < \size{let x = v(args); c'}$.
+This is exactly what allows us to 
+deal with the case of \ruleref{\ottdruleATAppLetLabel},
+because then we can conclude that 
+the inferred type (of a declarative judgment $[[Γ; Φ ⊢ iM ● args ⇒> ↑[uσ]uQ
+]]$ constructed by the induction hypothesis) is unique.

### 2023-11-27 proofs/tex/decl-typing-lemmas.tex
Finish proofs

@@
     a declarative typing judgment,
-    let us a function $\eqNodes{T}$
+    let us define a function $\eqNodes{T}$
     as the number of nodes in $T$ labeled with \ruleref{\ottdruleDTPEquivLabel} or 

### 2023-11-29 oopsla24/tex/06-extensions.tex
wip: extensions

\label{sec:extensions}

\subsection{Explicit Type Application}

In our system, all type applications are inferred implicitly: the algorithm
automatically instantiates the variables abstracted by $\forall$. 
The implicit type application can be added to the declarative system by the following rule:
$$\ottdruleDTTypeApp{}$$

However, this rule alone would cause ambiguity.
The declarative system does not fix the order of the quantifiers, 
which means that $[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ can be inferred 
as a type of $[[c]]$ interchangeably. But then explicit instantiation
$([[ c[iP] ]])$ would be ambiguous, as it is unclear whether $[[α⁺]]$ or
$[[β⁺]]$ should be instantiated with $[[iP]]$.

\paragraph{Solution 1: Declarative Normalization}
One way to resolve this ambiguity is to fix the order of quantifiers.  
The algorithm already performs the ordering of the quantifiers in the normalization procedure.
This way, we could require the inferred type to be normalized
to specify the order of the quantifiers:
$$\ottdruleDTTypeAppOrdered{}$$

The drawback of this approach is that it 
would cause the `leakage' of the internal algorithmic 
concept of type normalization into the `surface' declarative 
system.

\paragraph{Solution 2: Elementary Type Inference}
An alternative approach to provide explicit type application
 was proposed by \cite{zhao22:elementary}.
In this work, the subtyping relation is restricted 
in such a way that 
$[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ are \emph{not} mutual subtypes 
(as long as $[[α⁺ ∊ fv(iN)]]$ and $[[β⁺ ∊ fv(iN)]]$).
It implies that the order of the quantifiers of the inferred type is unique,
and thus, the explicit type application is unambiguous.

These restrictions can be incorporated into our system by
replacing the polymorphic subtyping rules
\ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel}
with the following stronger versions:
\vspace{-\baselineskip}
\begin{multicols}{2}
$$ \ottdruleDOneEForallR{} $$\\
$$ \ottdruleDOneEForallL{} $$
\end{multicols}

\begin{multicols}{2}
$$ \ottdruleDOneEExistsR{} $$\\
$$ \ottdruleDOneEExistsL{} $$
\end{multicols}
\vspace{\baselineskip}
From the perspective of mutual subtyping,
these changes fix the order of the quantifiers for an equivalence class. 
Moreover, the equivalence degenerates to equality (alpha-equivalence).

To accommodate these changes in the \emph{algorithm}, 
it suffices to 
    (i) replace the normalization procedure with identity: $[[nf(iN)]] \defeq [[iN]]$, $[[nf(iP)]] \defeq [[iP]]$,
    (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel} in the following way:
        $$\ottdruleLUBEExists{}$$ 
    and 
    (iii) replace the subtyping polymorphic rule \ruleref{\ottdruleAForallLabel} 
    by the following pair of rules:
    \begin{multicols}{2}
            $$\ottdruleAForallR{}$$\\
            $$\ottdruleAForallL{}$$
    \end{multicols}

\subsection{Bounded Quantification}

It is possible to smoothly extend the type system with bounded quantifiers.
In particular, we can add lower bounds to polymorphic $[[∀]]$-quantifiers:
$[[∀(pas ≥ iPs). iN]]$ with the expected subtyping specification:

$$ ??? $$

%   Γ, pbs ⊢ σ : {pas}    Γ, pbs ⊢ [σ] iN ≤ iM
%   ---------------------------------------- :: Forall
%   Γ ⊢ ∀(pas ≥ iPs). iN ≤ ∀(pbs ≥ iQs).iM

\subsection{Bidirectionalization and Invariants Relaxation}

The algorithm we provide requires that all lambda functions are annotated. To
augment the system's expressiveness and lessen this requirement, we can employ
bidirectionalization.

However, this approach will disrupt the crucial invariants of the system and
necessitate a more sophisticated constraint solver. Specifically, the generated
constraints may contain algorithmic variables on both sides of the subtyping
relation. If the constraint set only imposes an equivalence restriction, this
task aligns with pattern-unification. Nevertheless, the subtyping restrictions
render the problem undecidable. For instance, ...

### 2024-01-23 oopsla24/paper.tex
WIP: Declarative
Restructure proofs
wip: extensions
wip

@@
 \begin{abstract}
-  {\color{red}CHANGE!!}
-  This paper addresses the challenging problem of type inference for
-  Impredicative System F with existential types, a critical aspect of many
-  programming languages. While System F serves as the basis for type systems in
-  numerous languages, existing type inference techniques for Impredicative
-  System F are undecidable due to the presence of existential ($[[∃]]$) 
-  and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
-  sub-optimal. This paper presents novel contributions in the form of a local
-  type inference algorithm for Impredicative System F with existential types.
-  The algorithm introduces innovative techniques, such as a unique combination
-  of unification and anti-unification, a full correctness proof, and the use of
-  control structures inspired by \CBPV. Additionally, the paper discusses a type
-  inference framework that allows the algorithm to be applied to different type
-  systems, offering insights into the under-researched area of impredicative
-  existential type inference. 
+  % {\color{red}CHANGE!!}
+  % This paper addresses the challenging problem of type inference for
+  % Impredicative System F with existential types, a critical aspect of many
+  % programming languages. While System F serves as the basis for type systems in
+  % numerous languages, existing type inference techniques for Impredicative
+  % System F are undecidable due to the presence of existential ($[[∃]]$) 
+  % and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
+  % sub-optimal. This paper presents novel contributions in the form of a local
+  % type inference algorithm for Impredicative System F with existential types.
+  % The algorithm introduces innovative techniques, such as a unique combination
+  % of unification and anti-unification, a full correctness proof, and the use of
+  % control structures inspired by \CBPV. Additionally, the paper discusses a type
+  % inference framework that allows the algorithm to be applied to different type
+  % systems, offering insights into the under-researched area of impredicative
+  % existential type inference. 
 \end{abstract}
@@
 
-\section{Algorithm}
+\section{The Algorithm}
 \input{\genDir/tex/04-algorithm.tex}
 
-\section{Proof}
+\section{Algorithm Correctness}
 \input{\genDir/tex/05-proof.tex}
 
-\section{Extensions}
+\section{Extensions and Future Work}
 \input{\genDir/tex/06-extensions.tex}
@@
 \input{\genDir/tex/07-conclusion.tex}
-
-\cite{botlanMLFRaisingML2003}
-\cite{dunfieldBidirectionalTyping2020}
 

### 2024-01-23 oopsla24/tex/02-overview.tex
wip: correctness
wip

@@
+\subsection{Translation to \fexists}
+
+The language of \fexists is different from standard \systemf:
+every term and type has either positive or negative polarity.
+The type variables are annotated with their polarities
+(e.g. $[[α⁺]]$ or $[[β⁻]]$), and the types can change polarity
+via the shift operators $[[↑]]$ (positive to negative) and
+$[[↓]]$ (negative to positive).
+
+To establish a correspondence between \fexists and standard unpolarized systems,
+we present a translation between \fexists and \systemf. The translation of
+\systemf to \fexists is given in \cref{fig:translation}.
+
+\begin{figure}
+  \begin{equation*}
+  \begin{aligned}
+    [[|↓iN|]] &= [[|iN|]] \\
+    [[|↑iP|]] &= [[|iP|]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[|α±|]] &= [[a]] \\
+    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
+    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
+  \end{aligned}
+  \end{equation*}
+  \caption{Type depolarization}
+  \label{fig:depolarization}
+\end{figure}
+
+The translation from \systemf to \fexists can be done in two canonical ways
+depending on the desired polarity. The polarity will also determine the 
+execution strategy \ilyam{how?}.
+Moreover, the translation procedure depends on the \emph{polarity} of the
+free type variables. Thus, the translation functions 
+$[[-| fT |-Γ]]$ and $[[+| fT |+Γ]]$ are indexed by a context $[[Γ]]$, 
+containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.
+
+\begin{figure}
+  \begin{equation*}
+  \begin{aligned}
+    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
+    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
+    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
+    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
+    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
+    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
+    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
+    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
+    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
+  \end{aligned}
+  \end{equation*}
+  \caption{Type polarization}
+  \label{fig:polarization}
+\end{figure}
+
 \subsection{What types do we infer?}
+
+Suppose that
+we have the following variables in context $[[Φ]]$:
+
+\begin{align*}
+  [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
+  [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
+  [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
+  [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
+  [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
+\end{align*}
+
+Then
+
+\begin{align*}
+  &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
+  &[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]] \ilyam{(no)}\\
+\end{align*}
 
@@
 
+\paragraph{Type Context and Type Well-formedness}
+
+In the construction of \fexists, a type context (denoted as $[[Γ]]$) is
+represented as a \emph{set} of positive and negative type variables and it is used to
+assert the well-formedness of types. The well-formedness of a type is denoted as
+$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it asserts that all type variables are either
+bound by a quantifier ($[[∀]]$ and $[[∃]]$) or declared in the context $[[Γ]]$.
+The well-formedness checking is an \emph{algorithmic} procedure. As commonly
+done, we represent it as a system of inference rules, that correspond to a
+recursive algorithm taking the context and the type as input.
+
 \subsection{The Language of Terms}
@@
 \begin{itemize}
-    \item [$+$] $[[x]]$ denotes a (positive) term variable; 
-        \ilyam{why no negatives? Following CBPV}
+    \item [$+$] $[[x]]$ denotes a term variable.
+      Following the \CBPV stratification, we only have \emph{positive} (value)
+      term variables;
     \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
@@
 the result of a \emph{fully applied} function to a variable.
-Also notice that the language does not have a type application 
-(i.e. the eliminator of $[[∀]]$) and dually, 
-it does not have \pack (i.e. the constructor of $[[∃]]$).
-This is because the instantiation of polymorphic and existential 
-types is inferred by the algorithm.  \ilyam{refer to the extension chapter}
+Also notice that the language does not have a type application (i.e. the eliminator of $[[∀]]$) and dually, it does not have \pack (i.e. the constructor of $[[∃]]$).
+This is because the instantiation of polymorphic and existential types is inferred by the algorithm. 
+In \cref{sec:extensions}, we discuss the way to modify the system to introduce \emph{explicit} type applications.
 

### 2024-01-23 oopsla24/tex/03-declarative-system.tex
wip

@@
 \ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
- as we will discuss in \cref{todo}.
+thereby increasing the expressiveness of the system. These
+changes are discussed in \cref{sec:weakening-invariant}.
 
@@
 A property that is important for the subtyping algorithm, 
-in particular for the type \emph{upgrade} procedure (\cref{todo}),
+in particular for the type \emph{upgrade} procedure (\cref{sec:lub}),
 is the preservation of free variables by subtyping.
@@
 The normalization algorithm is shown in \cref{fig:type-nf}.
-The steps (i-ii) are implemented by the ordering function, which takes a set of variables $[[varset]]$
-and a type and returns a list of variables from $[[varset]]$ that occur in the type in the order of their
-first occurrence. Its formal definition can be found in \cref{todo}.
+The steps (i-ii) are implemented by the \emph{ordering} function
+$[[ord varset in iN]]$ and $[[ord varset in iP]]$.
+For a set of variables $[[varset]]$, and a type, it 
+returns a list of variables from $[[varset]]$ that occur in the type in 
+the order of their first occurrence.
+For brevity, we omit the formal definition of ordering,
+referring the reader to the appendix.
 
@@
   The normalization is required to fix the order of the quantifying variables
-  to bind them. After the bind, the rule infers the type of the body 
+  to bind them. Alternative approaches that do not require normalization
+  will be discussed in \cref{sec:explicit-type-application}.
+  After the bind, the rule infers the type of the body 
   and checks that it does not use the bound variables so that they do not 
@@
   to make the algorithm complete as it allows us to preserve
-  certain invariants (see \cref{todo}).
-  In \cref{todo} we discuss how this restriction can be relaxed
-  together with invariant shift subtyping.
+  certain invariants throughout the algorithm to ensure
+  the decidability of the produced unification task.
+  In \cref{sec:proof-subtyping}, we discuss these invariants in
+  more detail, and in \cref{sec:weakening-invariant} we suggest 
+  a way to relax them.
 
@@
   the instantiation of the quantifying variables is
-  not known a priori. The algorithm we present in \cref{todo} 
+  not known a priori. The algorithm we present in \cref{sec:algorithm} 
   delays this instantiation until more information about it 
@@
   is convertible to (a subtype of) the type of the function parameter.
-  In the algorithm (\cref{todo}), this subtyping check will provide the constraints
+  In the algorithm (\cref{sec:algorithm}), this subtyping check will provide the constraints
   we need to resolve the delayed instantiations of the quantifying variables.

### 2024-01-23 oopsla24/tex/04-algorithm.tex
wip

@@
+\label{sec:algorithm}
+
 In this section, we present the algorithmization of the declarative system described above.
@@
 and all free algorithmic variables are in $[[Ξ]]$.
-In addition to the rules repeating the declarative definition \cref{todo}, 
-we have two base cases for the algorithmic variables:
- \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf})
+Most of the rules are inherited from the well-formedness of 
+\emph{declarative} types: the declarative variables are
+checked to belong to the context $[[Γ]]$, the quantifiers extend the context,
+type constructors are well-formed congruently.
+As we extend the syntax with algorithmic variables, we also add 
+two base-case rules for them:
+ \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} 
+ (see \cref{fig:algo-wf}).
 
@@
 \end{multicols}
+\caption{Well-formedness of Algorithmic Types}
 \label{fig:algo-wf}
-\caption{Well-formedness of Algorithmic Types}
 \end{figure}
@@
     % Define nodes
-    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
+    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{fig:alg-subtyping})};
+    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{fig:alg-subtyping})};
     \node[left=1.5cm of 3] (2) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
@@
 can only hold if they are equivalent. This way, the algorithm must find the
-instantiations of the algorithmic variables on the left-hand side, which make it
-equivalent to the right-hand side. For this purpose, the algorithm invokes the
-unification procedure \cref{todo} preceded by normalization of the input types.
+instantiations of the algorithmic variables on the left-hand side
+such that these instantiations make the left-hand side and the ride-hand side
+equivalent. For this purpose, the algorithm invokes the
+unification procedure (\cref{sec:unification}) preceded by the normalization of the input types.
 It returns the resulting constraints given by the unification algorithm.
@@
 be instantiated with types, which are not known beforehand.
-We address this problem by algorithmization (\cref{todo})
-of the quantified variables.
+We address this problem by algorithmization 
+of the quantified variables (see \cref{sec:algo-syntax}).
 The rule introduces fresh algorithmic variables
@@
 \end{enumerate*}
-The resulting constraints are merged (using a special procedure defined later in \cref{todo})
+The resulting constraints are merged
+(using a special procedure defined in \cref{sec:constraint-merge})
 and returned as the output.
@@
 well-formed in $[[Θ(α̂⁺)]]$. It is done by the \emph{upgrade} procedure,
-which we discuss in detail in \cref{todo}.
+which we discuss in detail in \cref{sec:lub}.
 
@@
 that the unification instantiates, we could have called this procedure \emph{matching}.
-However, in \cref{todo} we will discuss several modifications of the 
-type system, where this invariant is not preserved, and thus, this procedure
-becomes a genuine first-order pattern unification \cite{miller1991:unification}.
+However, in \cref{sec:weakening-invariant}, we will discuss several modifications of the 
+type system, where this invariant is not preserved, and therefore, this procedure
+requires general first-order pattern unification \cite{miller1991:unification}.
 
@@
   Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
-  First, it unifies the argument types and their result types recursively
-  Then it merges the resulting constraints using the constraint merge procedure (\cref{todo}).
+  First, it unifies the argument types and their result types recursively.
+  Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.
 
@@
   It means that they can be merged in a simpler way than general constraints.
-  In particular, the merging procedure does not call any of the subroutines discussed here,
+  In particular, the merging procedure does not call any of the subtyping subroutines,
   but rather simply checks the matching constraint entries for equality.
@@
 and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
-$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm for finding the least upper bound
-is discussed in \cref{todo}.
+$[[α̂⁺ :≥ iP]]$ and $[[α̂⁺ :≥ iQ]]$. The algorithm finding the least upper bound
+is discussed in \cref{sec:lub}.
 
@@
 
-The anti-unification algorithm \cite{todo}, 
+The anti-unification algorithm \cite{plotkin1970:generalization,reynolds1970:transform}, 
 is a procedure dual to unification
@@
 At the high level, the algorithm scheme follows
-the standard approach \cite{todo} consisting of two principles:
+the standard approach \cite{plotkin1970:generalization}
+represented as a recursive procedure. Specifically, it follows two principles:
 \begin{enumerate}
@@
     anti-unification algorithm returns a `hole' such that one substitution maps
-    it to $[[iN]]$ and the other maps it to $[[iM]]$. The name of this `hole'
+    it to $[[iN]]$ and the other maps it to $[[iM]]$. This `hole'
     should have a name uniquely defined by the pair $([[iN]], [[iM]])$, so
@@
     \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
-    \node[right=1.4cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
-    \node[right=1.4cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
-    \node[below of=2,xshift=+2cm] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[below of=4] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
-    \node[left=0.75cm of 4] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \node[right=1.2cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
+    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
+    \node[below=0.8cm of 2] (7) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
+    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[below=0.3cm of 7] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
     \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
@@
     \draw[->] (2) to (4);
+    \draw[->] (2) to (7);
     \draw[->] (3) to (4);
-    \draw[->, bend right] (1.south west) to (5);
-    \draw[->, bend left] (3) to (5);
+    \draw[->] (1.south west) |- (5);
+    \draw[->] (3) |- (5);
     \draw[->] (2) to (6);
@@
   and merges the resulting constraint with $[[SC1]]$.
+
+\subsection{Constraint Singularity}
+\label{sec:constraint-singularity}
+
+The singularity algorithm is used to check that the inferred type of an
+application is unique. In this case, the annotation in the applicative let
+binder can be omitted. In the algorithmic system, the uniqueness means that the
+\emph{constraint set} produced by the application inference has a single
+substitution satisfying it.
+
+To implement the singularity algorithm, we define a partial function $[[SC
+singular with uσ]]$, taking a subtyping constraint $[[SC]]$ as an argument and
+returning a substitution $[[uσ]]$---the only solution of $[[SC]]$. The
+constraint $[[SC]]$ consists of constraint entries. Thus, we define the
+singularity using the singularity of constraint entries: each entry of $[[SC]]$
+must have a unique instantiation and the resulting substitution $[[uσ]]$ must be
+the combination of these instantiations (see \cref{fig:constraint-singularity}).
+
+\begin{figure}[h]
+  For a constraint set $[[SC]]$, we write $[[SC singular with uσ]]$ if and only if:
+  \begin{itemize}
+    \item for any \emph{positive} constraint entry $[[scE ∊ SC]]$, there exists $[[iP]]$ such that $[[scE singular with iP]]$,
+      and for the variable $[[β̂⁺]]$ restricted by $[[scE]]$, $[[ [uσ]β̂⁺ = iP ]]$;
+    \item the symmetric property holds for all \emph{negative} $[[scE ∊ SC]]$; and
+    \item for any $[[α̂± ∉ dom(SC)]]$, $[[ [uσ]α̂± = α̂± ]]$.
+  \end{itemize}
+  \caption{Constraint Singularity}
+  \label{fig:constraint-singularity}
+\end{figure}
+
+The singularity of constraint \emph{entries} is defined in
+\cref{fig:constraint-entry-singularity}. The \emph{equivalence} entries are
+always singular, as the only possible type satisfying them is the one given in
+the constraint itself, which is reflected in \ruleref{\ottdruleSINGNEqLabel} and
+\ruleref{\ottdruleSINGPEqLabel}. The \emph{subtyping} constraints are trickier.
+As will be discussed in \cref{sec:proof-lub-upgrade}, variables 
+(and equivalent types such as $[[∃nas.pa]]$) do not have 
+\emph{proper} supertypes, and thus, the constraints of a form 
+$[[pua :≥ ∃nas.pa]]$ are singular with the only possible normalized solution $[[pa]]$
+\ruleref{\ottdruleSINGSupVarLabel}. However, if the body of the existential type
+is guarded by a shift $[[∃nas.↓iN]]$, it is singular if and only if $[[iN]]$ is
+equivalent to some $[[αi⁻ ∊ {nas}]]$ bound by the quantifier, see the rule
+\ruleref{\ottdruleSINGSupShiftLabel}.
+
+\begin{figure}[h]
+  \hfill\\
+  \ottdefnSINGLabeled
+  \caption{Singular Constraint Entry}
+  \label{fig:constraint-entry-singularity}
+\end{figure}

### 2024-01-23 oopsla24/tex/05-proof.tex
wip

@@
 \subsection{Least Upper Bound and Upgrade}
+\label{sec:proof-lub-upgrade}
 
@@
 \subsection{Subtyping}
+    \label{sec:proof-subtyping}
 
@@
     whenever it exists (completeness).
-    In \cref{fig:subtyping-algo}, one can see that the positive and 
+    In \cref{fig:alg-subtyping}, one can see that the positive and 
     negative subtyping relations are not \emph{mututally} recursive: 
@@
 (which is given by the induction hypothesis) infers an equivalent type. 
-The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proo}), however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
+The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proof}), 
+however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
 This way, the soundness and completeness proofs are mutually dependent.

### 2024-01-23 oopsla24/tex/06-extensions.tex
wip

@@
 
+In this section, we discuss several extensions to the system and the algorithm.
+Some of them can be incorporated into the system immediately, while others
+are beyond the scope of this work, and thus are left for future research.
+% In particular, \emph{Explicit Type Application}
+
 \subsection{Explicit Type Application}
+\label{sec:explicit-type-application}
 
@@
 The implicit type application can be added to the declarative system by the following rule:
-$$\ottdruleDTTypeApp{}$$
+$$\ottdruleDTTypeAppLabeled{}$$
 
@@
 to specify the order of the quantifiers:
-$$\ottdruleDTTypeAppOrdered{}$$
+$$\ottdruleDTTypeAppOrderedLabeled{}$$
 
@@
 with the following stronger versions:
-\vspace{-\baselineskip}
-\begin{multicols}{2}
-$$ \ottdruleDOneEForallR{} $$\\
-$$ \ottdruleDOneEForallL{} $$
-\end{multicols}
+% \vspace{-\baselineskip}
+% \begin{multicols}{2}
+% $$ \ottdruleDOneEForallR{} $$\\
+% $$ \ottdruleDOneEForallL{} $$
+% \end{multicols}
 
-\begin{multicols}{2}
-$$ \ottdruleDOneEExistsR{} $$\\
-$$ \ottdruleDOneEExistsL{} $$
-\end{multicols}
-\vspace{\baselineskip}
-From the perspective of mutual subtyping,
-these changes fix the order of the quantifiers for an equivalence class. 
-Moreover, the equivalence degenerates to equality (alpha-equivalence).
+% \begin{multicols}{2}
+% $$ \ottdruleDOneEExistsR{} $$\\
+% $$ \ottdruleDOneEExistsL{} $$
+% \end{multicols}
+% \vspace{\baselineskip}
+
+% \vspace{-\baselineskip}
+
+\begin{minipage}{0.5\textwidth}
+$$ \ottdruleDOneEForallLR{} $$
+\end{minipage}%
+\begin{minipage}{0.5\textwidth}
+$$ \ottdruleDOneEExistsLR{} $$
+\end{minipage}
+\vskip 0.5em
+
+According to these rules, if two polymorphic ($[[∀]]$ or $[[∃]]$) types are
+subtypes of each other, they must have the same top-level quantifiers. Moreover,
+the equivalence on types (mutual subtyping) degenerates to \emph{equality} up to
+alpha-conversion.
 
@@
     (i) replace the normalization procedure with identity: $[[nf(iN)]] \defeq [[iN]]$, $[[nf(iP)]] \defeq [[iP]]$,
-    (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel} in the following way:
+        since the new equivalence classes consist of singletons;
+    (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel}
+        so that it requires the quantifiers to be equal (and performs alpha-conversion if necessary):
         $$\ottdruleLUBEExists{}$$ 
+    (iii) replace the subtyping polymorphic rule \ruleref{\ottdruleAForallLabel} 
+    by the following rule:
+        $$\ottdruleAForallLR{}$$
     and
-    (iii) replace the subtyping polymorphic rule \ruleref{\ottdruleAForallLabel} 
-    by the following pair of rules:
-    \begin{multicols}{2}
-            $$\ottdruleAForallR{}$$\\
-            $$\ottdruleAForallL{}$$
-    \end{multicols}
+    (iv) update the existential rule \ruleref{\ottdruleAExistsLabel} symmetrically.
+
+After these changes, the rule \ruleref{\ottdruleDTTypeAppLabel} and its algorithmic counterpart
+can be used to infer the the type of $[[ c[iP] ]]$.
+
+The elementary type inference restricts the expressiveness of the subtyping perhaps too much.
+As mentioned, it forbids the quantifier reordering; besides, 
+as soon as the right-hand side of the subtyping relation is
+polymorphic, it restricts the instantiation of the left-hand side quantifiers, 
+which \emph{disallows}, for example, $[[· ⊢ ∀α⁺.↑α⁺ ≤ ∀α⁺.↑↓↑α⁺]]$.
+On the other hand, the elementary subtyping system
+can be smoothly extended with bounded quantification, which we discuss later.
+
+\paragraph{Solution 3: Labeled Quantifiers}
+
+A compromise solution that resolves the ambiguity of 
+explicit the instantiation without fixing the order of
+quantifiers is using \emph{labeled} quantifiers.
+Let us discuss how to introduce them to the \emph{negative} part of the system,
+since the positive subsystem is symmetric.
+
+Each type abstraction $[[Λl▷α⁺.c]]$ must be annotated with a label $[[l]]$.
+This label propagates to the inferred polymorphic type. This way, each
+polymorphic quantifier $[[ ∀ (li ▷* αi⁺)^I . iN ]]$ is annotated with a label
+$[[li]]$ whose index is taken from the \emph{list} of labels $[[I]]$ associated
+with the group of quantifiers. To instantiate a quantifying variable, one refers
+to it by its label: $[[ c[lj ▷ iP] ]]$, which is expressed by the following
+typing rules:
+
+\begin{minipage}{0.4\textwidth}
+$$\ottdruleDTTLamLbLabeled{}$$
+\end{minipage}%
+\begin{minipage}{0.6\textwidth}
+$$\ottdruleDTTypeAppLbLabeled{}$$
+\end{minipage}
+\vskip 0.5em
+
+The subtyping rule allows the quantifiers to have an arbitrary order. 
+However, similarly to the elementary subtyping, 
+each of the right-hand side quantifiers must have 
+a matching left-hand side quantifier with the same label. These
+matching quantifiers synchronously removed
+(in other words, each variable is abstractly instantiated to itself), 
+and the remaining quantifiers are instantiated as usual by
+a substitution $[[σ]]$.
+$$ \ottdruleDOneLblForallLRLabeled{} $$
+
+The equivalence (mutual subtyping) in this system 
+allows the quantifiers within the same group to be reordered
+together with their labels. For instance,
+$[[∀l▷α⁺.∀m▷β⁺.iN]]$ and $[[∀m▷β⁺.∀l▷α⁺.iN]]$ are subtypes of each other.
+However, the right-hand side quantifiers are removed synchronously with 
+the matching left-hand side quantifiers, which means that
+$[[· ⊢ ∀l▷α⁺.↑α⁺ ≤ ∀l▷α⁺.↑↓↑α⁺]]$ is still not allowed.
+
+\subsection{Weakening of the Subtyping Invariant}
+\label{sec:weakening-invariant}
+
+In order to make the subtyping relation decidable,
+we made the subtyping of shifts \emph{invariant}, which manifests
+in the following rules:
+
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDOneShiftU{} $$
+\end{minipage}%
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDOneShiftD{} $$
+\end{minipage}
+\vskip 0.5em
+
+To make the system more expressive, we can relax these rules. 
+Although making both rules covariant would make the system undecidable,
+it is possible to make the up-shift subtyping covariant while keeping the down-shift
+invariant: 
+$$ \ottdruleDOneShiftURlxLabeled{} $$\\
+However, this change requires an update to the algorithm.
+As one can expect,  
+the corresponding algorithmic rule \ruleref{\ottdruleAShiftULabel}
+must be changed accordingly:
+$$ \ottdruleAShiftURlxLabeled{} $$
+Also, notice that now the algorithmic variables can occur on \emph{both}
+sides of the positive subtyping relation, and thus, a more sophisticated constraint 
+solver is needed. To see what kind of constraints will be generated,
+let us keep the other algorithmic rules unchanged for a moment
+and make several observations.
+
+First, notice that the negative quantification variables still only occur at
+\emph{invariant} positions: the negative variables are generated from the existential
+quantifiers $[[∃α⁻.iP]]$, and the switch from a positive to a negative type in the syntax path
+to $[[α⁻]]$ in $[[iP]]$ is `guarded' by \emph{invariant} down-shift. It means
+that in the algorithm, the negative constraint entries can only be equivalence
+entries ($[[nua :≈ iN]]$) but not subtyping entries.
+
+Second, notice that the positive subtyping is still not dependent on the negative
+subtyping. It means, that any leaf-to-root path in the inference tree is monotonous:
+in the first phase, the negative subtyping rules are applied, and in the second phase, 
+only the positive ones.
+
+The invariant that only the left-hand side of the judgment is algorithmic is not
+preserved. However, the only rule violating this invariant is
+\ruleref{\ottdruleDOneShiftURlxLabel}, and it is placed at the interface between
+negative subtyping and positive subtyping. It means that the \emph{negative}
+algorithmic variables (produced by $[[∃]]$ in the positive phase of the
+inference path) still only occur on the left-hand side of the judgment, and the
+positive algorithmic variables (produced by $[[∀]]$ in the first phase) can
+occur either on the left-hand side or on the right-hand side (if the switch from
+the negative to the positive phase is made by \ruleref{\ottdruleDOneShiftURlxLabel})
+but not both.
+This way, the produced entries will be of the following form:
+\begin{itemize*}
+    \item[(i)] $[[pua :≈ uP]]$, 
+    \item[(ii)] $[[nua :≈ uN]]$, 
+    \item[(iii)] $[[pua :≥ iP]]$, and
+    \item[(iv)] $[[pua :≤ uP]]$ where $[[uP]]$ does not contain positive unification variables.
+\end{itemize*}
+
+We do not present the details of the constraint resolution procedure or how it
+is integrated into the subtyping algorithm, leaving it for future work. However,
+we believe that the system consisting of the mentioned kinds of constraints is solvable
+by the following reasons.
+\begin{enumerate*}
+    \item[(i)] The equivalence entries $[[pua :≈ uP]]$ and $[[nua :≈ uN]]$ are resolved by standard first-order pattern unification techniques.
+    \item[(ii)] The resolution of supertyping entries ($[[pua :≥ iP]]$) is discussed in the original algorithm (\cref{sec:constraint-merge}).
+    \item[(iii)] The new kind of entries $[[pua :≤ uP]]$ can be
+        reduced by case analysis. If $[[uP]]$ is a variable $[[β⁺]]$
+        or a shifted computation $[[↓uN]]$ then 
+        $[[pua :≤ uP]]$ is equivalent to $[[pua :≈ uP]]$;
+        otherwise, $[[pua :≤ ∃nas.↓uN]]$ is equivalently replaced by $[[pua :≈ [nuas/nas]↓uN]]$
+        for freshly created $[[nuas]]$.
+\end{enumerate*}
+
+This way, the algorithm can be extended to support the \emph{covariant} up-shift
+subtyping. This, for example, enriches the relation with such subtypes as $[[· ⊢
+∀α⁺.α⁺→↑α⁺ ≤ ↓↑Int → ↑∃β⁻.↓β⁻]]$, which does not hold in the original system.
+Moreover, this extension also increases the expressiveness of the \emph{type
+inference}. Namely, when inferring a type of an annotated let binding 
+$[[let x:iP = v(args); c]]$, we require $[[↑iP]]$ to be a supertype of 
+$[[↑iQ]]$---the resulting type of the application $[[v]]([[args]])$. 
+Now, it is equivalent to the requirement that $[[iP]]$ is a supertype of $[[iQ]]$
+(without the up-shift), which is more permissive.
+
+In addition, the enhancement of the unification algorithm described above
+makes possible another improvement to the system---bounded quantification.
 
@@
 
-It is possible to smoothly extend the type system with bounded quantifiers.
-In particular, we can add lower bounds to polymorphic $[[∀]]$-quantifiers:
-$[[∀(pas ≥ iPs). iN]]$ with the expected subtyping specification:
+After the weakening of subtyping invariants, 
+it is possible to smoothly extend the \emph{elementary}
+version of the system with bounded quantification.
+In particular, we can add upper bounds to polymorphic $[[∀]]$-quantifiers:
+$[[∀(pa ≤* iP). iN]]$.
 
-$$ ??? $$
+The declarative subtyping rules for bounded quantification are as expected:
+the instantiation of quantified variables must satisfy the corresponding bounds
+($[[Γ, pas ⊢ iQi ≥ [σ]βi⁺]]$ holds for each $i$);
+and the right-hand side quantifiers must be more restrictive than the matching left-hand side
+quantifiers ($[[Γ ⊢ iPi ≥ iP'i]]$ for each $i$):
+$$\ottdruleDOneEForallLRUbLabeled{}$$
 
-%   Γ, pbs ⊢ σ : {pas}    Γ, pbs ⊢ [σ] iN ≤ iM
-%   ---------------------------------------- :: Forall
-%   Γ ⊢ ∀(pas ≥ iPs). iN ≤ ∀(pbs ≥ iQs).iM
+The corresponding algorithmic rule, as before,
+replaces the polymorphic variables $[[β1]]\dots$ with fresh 
+algorithmic variables $[[β1̂⁺]]\dots$, and merges the generated
+constraints with the constraints given in the bounds
+$[[βî⁺ :≤ uQi]]$:
+$$\ottdruleAEForallLRUbLabeled{}$$
 
-\subsection{Bidirectionalization and Invariants Relaxation}
+Notice that the algorithmic rule \ruleref{\ottdruleAEForallLRUbLabel}
+requires the bounding types $[[iPs]]$ and $[[iQs]]$ to be \emph{declarative}.
+Because of that, the generated constraints have shape $[[βî⁺ :≤ iQi]]$,
+\ie the bounding types do not contain algorithmic variables.
+This kind of constraints is covered by the resolution procedure described in
+\cref{sec:weakening-invariant}, and thus, \ruleref{\ottdruleAEForallLRUbLabel} fits 
+into the algorithmic system without changes to the constraint solver.
 
-The algorithm we provide requires that all lambda functions are annotated. To
-augment the system's expressiveness and lessen this requirement, we can employ
-bidirectionalization.
+However, in order for this rule to be complete, 
+the declarative system must be restricted as well.
+To guarantee that no algorithmic variable is introduced in the bounding types,
+we need to forbid the quantifying variables to occur in the bounding types in the declarative system.
+For that purpose, we must include this requirement in the type well-formedness
+(here $[[bv iN]]$ denotes the set of variables freely occurring 
+in the quantifiers of $[[iN]]$ at any depth):
+$$\ottdruleWFTForallUbLabeled{}$$
 
-However, this approach will disrupt the crucial invariants of the system and
-necessitate a more sophisticated constraint solver. Specifically, the generated
-constraints may contain algorithmic variables on both sides of the subtyping
-relation. If the constraint set only imposes an equivalence restriction, this
-task aligns with pattern-unification. Nevertheless, the subtyping restrictions
-render the problem undecidable. For instance, ...
+For brevity, we do not discuss in detail the combination of bounded and
+unbounded quantification in one system. However, this update is straightforward:
+the unbounded quantifiers are treated as the weakest bound, which is trivially
+satisfied. Analogously, it is possible to extend the system with \emph{lower}
+bound $[[∀]]$-quantifiers: $[[∀(pa ≥* iP). iN]]$. 
+Then the generated constraints will change to $[[βî⁺ :≥ iQi]]$, which 
+is also covered by the resolution procedure.
+However, we do not consider bounded \emph{existential} quantification,
+because in this case, the constraint resolution would require us to 
+find the \emph{greatest lower bound} of two negative types, which is not
+well-defined (see \cref{sec:constraint-merge}).
+
+\subsection{Bidirectionalization}
+
+% \subsubsection{Declarative System}
+% \ottdefnDTNSynth{}
+% \ottdefnDTNCheck{}
+% \ottdefnDTNCheckSynth{}
+% \ottdefnDTPSynth{}
+% \ottdefnDTPCheck{}
+% \ottdefnDTPCheckSynth{}
+% \ottdefnDTSpinInfBidir{}
+
+% \subsubsection{Algorithmic System}
+% \ottdefnATNSynth{}
+% \ottdefnATNCheck{}
+% \ottdefnATNCheckSynth{}
+% \ottdefnATPSynth{}
+% \ottdefnATPCheck{}
+% \ottdefnATPCheckSynth{}
+% \ottdefnATSpinInfBidir{}
+
+The algorithm we provide requires that all lambda functions are annotated. 
+This restriction significantly simplifies the type inference by making the 
+terms uniquely define their types. However, it leads to certain redundancies
+in typing, in particular, the type of a lambda expression cannot be inferred 
+from the context it is used in: $[[((λ x . return x) : (Int → ↑Int))]]$
+is not allowed.
+
+The well-known way to incorporate this expressiveness into the system
+is to make the typing bidirectional \cite{dunfield2020:bidirectional}.
+The idea is to split the typing judgment into two kinds:
+\begin{enumerate}
+    \item[(i)] \emph{Synthesis} 
+        judgments are used when the type of a term can be \emph{inferred}
+        based exclusively on the term itself. 
+        Syntactically, we denote synthesizing judgments
+        as $[[Γ ; Φ ⊢ c ⇒ iN]]$.
+    \item[(ii)] \emph{Checking} judgments 
+        assume that the type of a term is \emph{given}
+        from the context, and 
+        it is required to \emph{check} that the given
+        type can be assigned to the term.
+        In checking judgments $[[Γ ; Φ ⊢ c ⇐ iN]]$, 
+        the type $[[iN]]$ is considered as an \emph{input}.
+\end{enumerate}
+
+To bidirectionalize the system, each typing rule 
+must be oriented either to synthesis or to checking (or both).
+In particular, the \emph{annotated} lambda-abstractions are
+synthesizing, and the \emph{unannotated} ones are checking.
+
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDTtLamSynLabeled{} $$
+\end{minipage}
+\hfill
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDTtLamChkLabeled{} $$
+\end{minipage}
+\vskip 0.5em
+
+As common for bidirectional systems with subtyping, 
+we would also need to introduce the \emph{subsumption} rules.
+They allow us to `forget' the information about the type
+by switching to the synthesis mode, as long as the synthesized
+type is more polymorphic than the checked one:
+
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDTNSubLabeled{} $$
+\end{minipage}
+\hfill
+\begin{minipage}{0.5\textwidth}
+    $$ \ottdruleDTPSubLabeled{} $$
+\end{minipage}
+\vskip 0.5em
+
+Most of the original rules will have two bidirectional counterparts,
+one for each mode. The premises of the rules are oriented with respect to the conclusion.
+For instance, we need both checking and inferring versions for $[[return v]]$: 
+    $$ \ottdruleDTReturnChkSynLabeled{} $$
+For some rules, however, it only makes sense to have one mode.
+As mentioned above, the unannotated lambda-abstraction is always checking. 
+On the other hand, an \emph{annotated} lambda, type-level lambda abstraction, 
+or a variable inference is always synthesizing, since
+the required type information is already given in the term:
+
+\begin{minipage} {0.5\textwidth}
+    $$ \ottdruleDTTLamSynLabeled{} $$
+\end{minipage}
+\hfill
+\begin{minipage} {0.5\textwidth}
+    $$ \ottdruleDTVarSynLabeled{} $$
+\end{minipage}
+\vskip 0.5em
+
+Finally, let us discuss the application inference part of the declarative system. 
+Rules \ruleref{\ottdruleDTEmptyAppLabel} and \ruleref{\ottdruleDTForallAppLabel} are 
+unchanged: they simply do not have typing premises to be oriented.
+However, the arrow application rule \ruleref{\ottdruleDTArrowAppLabel}
+infers the result of the application of an arrow $[[iQ → iN]]$ 
+to a list of arguments $[[v , args]]$, 
+and it must make sure that the first argument $[[v]]$ is typeable with the expected $[[iQ]]$.
+As we will discuss further, simply checking $[[Γ; Φ ⊢ v ⇐ iQ]]$ turns out to be too 
+powerful and potentially leads to undecidability. 
+Instead, this check is approximated by the combination of $[[Γ; Φ ⊢ v ⇒ iP]]$
+and $[[Γ ⊢ iQ ≥ iP]]$.
+
+\paragraph{The Algorithm}
+
+    The algorithmic system is bidirectionalized in a similar way. 
+    Each typing premise and the conclusion of the rule is oriented
+    to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
+    with respect to the declarative system.
+
+    For brevity, we omit the details of the bidirectional algorithm.
+    However, let us discuss one especially important rule---arrow application inference. 
+    An intuitive way to bidirectionalize this rule is the following:
+    $$ \ottdruleATArrowAppBidirLabeled{} $$\\
+    % \paragraph{Reaching Undecidability}
+    However, this rule is too allowing. 
+    The judgement $[[Γ ; Φ ⊨ v ⇐ uQ ⫤ SC1]]$
+    requires us to check a term against an \emph{algorithmic} 
+    type $[[uQ]]$. Further, through the
+    lambda function checking \ruleref{\ottdruleDTtLamChkLabel}
+    the algorithmic types infiltrate the type context
+    and then through variable inference and subsumption, 
+    \emph{both} sides of subtyping. 
+    This way, \ruleref{\ottdruleATArrowAppBidirLabel}
+    compromises the important invariant
+    of the subtyping algorithm: 
+    now the same algorithmic variable can occur on \emph{both}
+     sides of the subtyping relation.
+
+    We believe that the relaxation of this invariant brings the 
+    constraint resolution too close to the second-order pattern 
+    unification, which is undecidable \cite{goldfarb81:undecidability}.
+    In particular, the constraint $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable with 
+    $[[α̂⁺]] = [[∃β⁻.↓β⁻]]$, 
+    since by \ruleref{\ottdruleDOneExistsLabel}, 
+    the existential $[[β⁻]]$ can be impredicatively instantiated to $[[↑∃β⁻.↓β⁻]]$.
+
+    The constraints such as $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ are not merely hypothetical. 
+    It occurs, for example, when we infer the type of the following application:
+    $$[[· ; · ⊢ ∀α⁺ . ↓(α⁺ → ↑Int) → ↓↑α⁺ → ↑Int ● {λx.λy.let y' = x(y); return y'} ⇒> ↑Int]]$$
+    For $[[x]]$ to be applicable to $[[y]]$,
+    the type of $[[y]]$---$[[↓↑α⁺]]$ must be a subtype of
+    $[[α⁺]]$---the type expected by $[[x]]$. This way, 
+    this inference is possible if and only if
+    $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ is solvable.
+    Using a similar scheme, we can construct different examples
+    whose resolution significantly relies on second-order pattern unification.
+    Thus, we leave the resolution of this type of constraints beyond the scope of this work.
+
+    % \paragraph{Restricting Inference}
+    Instead, we strengthen
+    \ruleref{\ottdruleATArrowAppBidirLabel}
+    in such a way that it never checks a term against an algorithmic type.
+    Instead, type checking $[[v]]$ against $[[uQ]]$ is replaced by 
+    a more restrictive premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
+    $$ \ottdruleATArrowAppBidirSLabeled{} $$\\
+
+    As one can expect, the declarative rule is also changed accordingly.
+    In terms of practicality, this change disallows
+    implicit checking against a \emph{polymorphic} type:
+    $[[·;· ⊢ λx.return x ⇐ ∀α⁺.α⁺ → ↑α⁺]]$ is not allowed, because
+    it would require adding algorithmic $[[x:α̂⁺]]$ to the context.
+    However, once the instantiation is made explicit or the lambda is explicitly 
+    polymorphic, the checking is allowed:
+     $[[·;· ⊢ λx.return x ⇐ Int → ↑Int]]$ and 
+     $[[·;· ⊢ Λα⁺.λx:α⁺.return x ⇐ ∀α⁺.α⁺ → ↑α⁺]]$ are both valid.

### 2024-01-23 oopsla24/tex/07-conclusion.tex
wip

We have presented a type inference algorithm for an impredicative polymorphic
lambda calculus with existential types and subtyping.
The system is designed in call-by-value paradigm, that allowed us to 
restrict it to a decidable fragment of the language,
that we describe declaratively. We presented the inference algorithm
and proved its soundness and completeness with respect to the specification.
The algorithm combines unification---a standard type inference technique---with
anti-unification---which has not been used in type inference before.

### 2024-01-23 proofs/main.tex
wip

@@
 (according to \cite{dunfieldBidirectionalTyping2020}), 
-which means that
+which means that checking the well-formedness of a type
+can be done recursively by a deterministic algorithm. 
+We will use the well-formedness checking in the inference algorithm, 
+for example, to check that the existential variables do not escape their scope.
 
@@
 
-The singularity algorithm used in  \ruleref{\ottdruleATAppLetAnnLabel}
-of the algorithmic typing check whether the constraint $[[SC]]$
-is uniquely defines a substitution satisfying it, and if it does
-returns such a substitution as the result.
-To do that, we define a partial function $[[SC singular with uσ]]$,
-taking a subtyping constraint $[[SC]]$ and returning a substitution 
+The singularity algorithm checks whether the constraint $[[SC]]$
+uniquely defines the substitution satisfying it, and if it does,
+the algorithm returns this substitution as the result.
+To implement it, we define a partial function $[[SC singular with uσ]]$,
+taking a subtyping constraint $[[SC]]$ as an argument and returning a substitution 
 $[[uσ]]$---the only possible solution of $[[SC]]$.
@@
     is perhaps the least obvious rule.
-    In type $[[∃nas.↓iN]]$, if $[[iN]]$ is anything different (non-equivalent) to 
-    $[[nai ∊ {nas}]]$, there are at least 
-    one proper supertype of $[[∃nas.↓iN]]$, since $[[iN]]$ can be abstracted over
-    by an existential quantifier. Otherwise, any supertype of $[[∃nas.↓nai]]$
-    is equivalent to it, and thus, the solution is unique. 
+    Having a type $[[∃nas.↓iN]]$, suppose that $[[iN]]$ is
+    not equivalent to any just bound variable $[[nai ∊ {nas}]]$.
+    Then the type $[[∃nas.↓iN]]$ has a proper supertype: $[[∃nas.↓na1]]$,
+    and thus the constraint is not singular.
+    Otherwise, if $[[iN]]$ is equivalent to some $[[nai]]$,
+    any supertype of $[[∃nas.↓nai]]$ is equivalent to it, and thus, 
+    the constraint has a unique solution.
 \end{itemize}

### 2024-02-13 oopsla24/tex/02-overview.tex
wip

@@
-\subsection{Translation to \fexists}
+% Imprediicative
+% CBPV, monadic control
+% Local Inference
+% Two ways to translate
 
@@
 
-To establish a correspondence between \fexists and standard unpolarized systems,
-we present a translation between \fexists and \systemf. The translation of
-\systemf to \fexists is given in \cref{fig:translation}.
+Positive expressions correspond to values in the \CBPV system
+\cite{levy2006:cbpv}, and negative expressions correspond to computations.
+The difference between the computations and values is intuitively described 
+in the slogan of \CBPV: ``a value is, a computation does''. 
+In particular, an argument of a function is always a value (\ie positive), 
+and the resulting type, as well as the function itself, 
+is a computation (\ie negative).
 
-\begin{figure}
-  \begin{equation*}
-  \begin{aligned}
-    [[|↓iN|]] &= [[|iN|]] \\
-    [[|↑iP|]] &= [[|iP|]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[|α±|]] &= [[a]] \\
-    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
-    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
-  \end{aligned}
-  \end{equation*}
-  \caption{Type depolarization}
-  \label{fig:depolarization}
-\end{figure}
+\subsection{Examples}
 
-The translation from \systemf to \fexists can be done in two canonical ways
-depending on the desired polarity. The polarity will also determine the 
-execution strategy \ilyam{how?}.
-Moreover, the translation procedure depends on the \emph{polarity} of the
-free type variables. Thus, the translation functions 
-$[[-| fT |-Γ]]$ and $[[+| fT |+Γ]]$ are indexed by a context $[[Γ]]$, 
-containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.
+The standard \systemf types and terms must be polarized to be used in \fexists, 
+we present the embedding formally in \cref{fig:polarization}.
+For instance, the following types and terms of \systemf 
+can be polarized in the given way:
 
-\begin{figure}
-  \begin{equation*}
-  \begin{aligned}
-    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
-    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
-    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
-    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
-    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
-    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
-    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
-    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
-    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
-  \end{aligned}
-  \end{equation*}
-  \caption{Type polarization}
-  \label{fig:polarization}
-\end{figure}
-
-\subsection{What types do we infer?}
-
-Suppose that
-we have the following variables in context $[[Φ]]$:
-
+\begin{figure}[h]
   \begin{align*}
@@
     [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
+    [[autoP]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
+  \end{align*}
+  \caption{Polarization of standard \systemf terms}
+  \label{fig:polarization-examples}
+\end{figure}
+
+Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
+Then the following types are inferrable in \fexists:
+
+\begin{align*}
+  % &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
+  &[[· ; Φ, x1:↓(↓iN → iN), x2:↓(↓iM → iM) ⊢ let x = choose (x1, x2); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]\\
+  &[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]\\
+  &[[· ; Φ ⊢ let x = choose (id, autoP); return x : ↑ ∃γ⁻.↓γ⁻]]\\
 \end{align*}
 
-Then
+The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
+which has certain limitations.
+In particular, if an argument of a function is polymorphic,
+its polymorphic variables are not instantiated by the inference algorithm. 
+For example, one would expect the following to be inferrable:
+$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$.
+However, to infer this, the $[[id]]$ must be instantiated to $[[Int → ↑Int]]$, 
+which requires the information to be propagated from the neighboring branch 
+of the syntax tree ($[[ [two, three, nine] ]]$), \ie bypass the locality.
 
-\begin{align*}
-  &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
-  &[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]] \ilyam{(no)}\\
-\end{align*}
+However, the polymorphic arguments of the function are instantiated, 
+and thus, the annotated version of the same term infers the type successfully:
+$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$.
 
@@
 categories (polarities): positive and negative,  
-similarly to the \CBPV system \cite{levy2006:cbpv}.
-The negative types represent computations, and the positive types represent values:
+similar to the values and computations in the \CBPV system \cite{levy2006:cbpv}.
 \begin{itemize}
@@
 
+\subsection{Relation to \systemf}
+
+To establish a correspondence between \fexists and standard unpolarized systems,
+we present a translation between \fexists and \systemf. The translation of
+\systemf to \fexists is given in \cref{fig:polarization}, and the backward translation
+is given in \cref{fig:depolarization}.
+
+\begin{figure}[h]
+  \begin{equation*}
+  \begin{aligned}
+    [[|↓iN|]] &= [[|iN|]] \\
+    [[|↑iP|]] &= [[|iP|]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[|α±|]] &= [[a]] \\
+    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
+    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
+  \end{aligned}
+  \end{equation*}
+  \caption{Type depolarization}
+  \label{fig:depolarization}
+\end{figure}
+
+The translation from \systemf to \fexists can be done in two canonical ways
+depending on the desired polarity. The polarity will also determine the 
+execution strategy \ilyam{how?}.
+Moreover, the translation procedure depends on the \emph{polarity} of the
+free type variables. Thus, the translation functions 
+$[[-| fT |-Γ]]$ and $[[+| fT |+Γ]]$ are indexed by a context $[[Γ]]$, 
+containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.
+
+\begin{figure}[h]
+  \begin{equation*}
+  \begin{aligned}
+    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
+    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
+    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
+    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
+    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
+    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
+    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
+    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
+    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
+  \end{aligned}
+  \end{equation*}
+  \caption{Type polarization}
+  \label{fig:polarization}
+\end{figure}
+
 \subsection{The key ideas of the algorithm}

### 2024-02-13 oopsla24/tex/03-declarative-system.tex
wip

@@
   before recursing on the body of the let binder.
-  The difference is that the former rule is for the 
-  \emph{unannotated} let binder, and thus
-  it requires the resulting type of application to be unique
-  (up to equivalence), so that the type of the bound variable $[[x]]$
-  is known before it is put into the context.
+
+  The former rule infers the type of an \emph{unannotated} let binder, and thus
+  it requires the resulting type of application to be \emph{the principal type},
+  so that the type we assign to the bound variable $[[x]]$ is determined.
+  In this context, principality means minimality. In other words, 
+  $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
+  any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
+  is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.
+
   The latter rule is for the \emph{annotated} binder,

### 2024-02-13 oopsla24/tex/04-algorithm.tex
wip

@@
     \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
-    \node[below=0.8cm of 2] (7) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
+    \node[below=0.6cm of 2] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
+    \node[below=0.5cm of 7] (8) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
     \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[below=0.3cm of 7] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
     \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \node[below=0.9cm of 6] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
     \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
@@
     \draw[->] (3) to (4);
+    \draw[->] (7) to (8);
     \draw[->] (1.south west) |- (5);
-    \draw[->] (3) |- (5);
+    \draw[->] (3) |- (5.south east);
     \draw[->] (2) to (6);
@@
 
-  Rule \ruleref{\ottdruleATAppLetAnnLabel}
-  works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
-  However, since there is no annotation, 
-  instead of checking
-  that the inferred type $[[↑uQ]]$ 
-  can be a subtype of the annotation,
-  the algorithm checks that $[[↑uQ]]$ is unique. 
-  As we prove, uniqueness means that all the algorithmic variables of 
-  $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$.
-  This way, uniqueness is checked by the combination of
-  $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
-  Together, these two premises guarantee that the only 
-  possible instantiation of $[[uQ]]$ is $[[ [uσ]uQ ]]$.
+  Rule \ruleref{\ottdruleATAppLetLabel}
+  works similarly to \ruleref{\ottdruleATAppLetAnnLabel}.
+  However, since no annotation is given,
+  the algorithm must ensure that the inferred $[[uQ]]$
+  has the `canonical' minimal instantiation.
+  To find it, it makes a call to the minimal instantiation algorithm 
+  (\cref{sec:minimal-instantiation})
+  that finds the substitution that satisfies the inferred constraints $[[SC]]$ and
+  instantiates $[[uQ]]$ to the minimal (among other such instantiations)
+  type $[[ [uσ]uQ ]]$.
 
@@
 
-\subsection{Constraint Singularity}
+\subsection{Minimal Instantiation and Constraint Singularity}
 \label{sec:constraint-singularity}
 
-The singularity algorithm is used to check that the inferred type of an
-application is unique. In this case, the annotation in the applicative let
-binder can be omitted. In the algorithmic system, the uniqueness means that the
-\emph{constraint set} produced by the application inference has a single
-substitution satisfying it.
+Multiple types $[[iM]]$ can be inferred for a type application:
+$[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ but only one \emph{principal} 
+type should be chosen for a variable in an unannotated let binder.
+Declaratively, we require the principal type $[[iP]]$ to be
+minimal among other all types $[[iP']]$ that can be inferred
+for the application $[[Γ ; Φ ⊢ iN ● args ⇒> ↑iP']]$.
+Algorithmically, the inference returns an
+\emph{algorithmic} type $[[uP]]$ 
+together with a set of necessary and sufficient constraints $[[SC]]$
+that any instantiation of $[[uP]]$ must satisfy.
+This way, to guarantee principality, we must minimize the instantiation of $[[uP]]$,
+while ensuring that it satisfies the given constraints $[[SC]]$.
+
+The minimal instantiation judgment `$[[Γ ⊢ uP SC minby uσ]]$'
+means that $[[uσ]]$ is the minimum among all the substitutions satisfying $[[SC]]$,
+that is $[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
+instantiation of $[[uP]]$. The algorithm is given in \cref{fig:minimal-instantiation}.
+First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
+Then it considers two cases. If the type is an algorithmic variable $[[α̂⁺]]$
+restricted by a constraint $[[(α̂⁺ :≥ iP) ∊ SC]]$,
+its minimal instantiation is the type $[[nf(iP)]]$, 
+and \ruleref{\ottdruleSINGPUvarLabel} is applied.
+Otherwise, \ruleref{\ottdruleSINGSingLabel} is applied,
+meaning that all the instantiations of the type must be equivalent.
+The algorithm ensures that by two premises:
+\begin{enumerate*}
+  \item[(i)] checking that all the algorithmic variables of the type are restricted by the constraint set; and
+  \item[(ii)] building the substitution that satisfies the constraint set on these variables 
+    (and simultaneously checking that such substitution is unique up-to-equivalence)
+    using the constraint singularity algorithm.
+\end{enumerate*}
+
+\begin{figure}[h]
+  \hfill\\
+  \ottdefnMININSTLabeled
+  \caption{Minimal Instantiation}
+  \label{fig:minimal-instantiation}
+\end{figure}
+
+The singularity algorithm 
+performs two tasks: it checks that a
+\emph{constraint set} has a single substitution satisfying it, 
+and if it does, it builds this substitution.
 

### 2024-02-13 oopsla24/tex/05-proof.tex
wip

@@
 
-\subsection{Singularity}
+\subsection{Minimal Instantiation and Singularity}
 \label{sec:singularity-proof}
 
-The rule \ruleref{\ottdruleDTAppLetLabel} of the declarative inference
-says that one can omit the type annotation of a let-binding if the inferred 
-type of the bound application is unique. Accordingly, the algorithm,
-that returns the constraint set, must be able to check that
-the set of constraints has a single solution, and also to find this solution. 
-This subroutine is called singularity, and its correctness is formulated as follows.
+\paragraph{Minimal Instantiation}
+Algorithmic typing relies on the \emph{minimal instantiation} procedure.
+For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
+it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
+$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
+This way, the correctness of minimal instantiation is formulated as the following two lemmas.
 
-The soundness states that $[[SC singular with uσ]]$
+\begin{lemma*}[Soundness of Minimal Instantiation]
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
+    Then $[[Γ ⊢ uP SC minby uσ ]]$ implies that 
+    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
+    (\ie $[[Θ ⊢ uσ : SC]]$)
+    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
+    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$ holds
+\end{lemma*}
+
+\begin{lemma*}[Completeness of Minimal Instantiation]
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
+    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
+    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
+    satisfying $[[SC]]$, we have
+    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
+    succeeds, specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
+\end{lemma*}
+
+The soundness is rather straightforwardly proved by induction on the 
+judgment inference tree, relying on the soundness of the singularity procedure.
+The proof of completeness is done by induction on the structure of the type $[[uP]]$.
+It uses the fact that $[[ Γ ⊢ ∃nas.[uσ']uP ≥ ∃nas.[uσ]uP]]$ implies
+$[[ Γ, nas ⊢ [uσ']uP ≥ [uσ]uP]]$.
+
+\paragraph{Singularity}
+The soundness of singularity states that $[[SC singular with uσ]]$
 implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 

### 2024-02-13 proofs/main.tex
wip

@@
   \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
-    but it is used when the type of the application is unique, and thus, the annotation
-    is redundant. Here $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ uniq]]$ means that
-    if also $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$ then $[[Γ ⊢ iQ ≈ iQ']]$
+    bus is used for unannotated let-bindings. In this case, 
+    we require the type application to infer the `canonical' principal type.
+    $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
+    any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
+    is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.
 \end{itemize}
@@
     and returns it as the result. 
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
+  \item \ruleref{\ottdruleATAppLetLabel}
     works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
-    However, since there is no annotation, 
-    instead of checking the instantiation to it, 
-    the algorithm checks that the inferred type
-    $[[↑uQ]]$ is unique.
-    It is the case if all the algorithmic variables of 
-    $[[↑uQ]]$ are sufficiently restricted by $[[SC]]$,
-    which is checked by the combination of
-    $[[uv uQ = dom(SC)]]$ and $[[SC singular with uσ]]$.
-    Together, these two premises guarantee that the only 
-    possible instantiation of $[[↑uQ]]$ is $[[ [uσ]uQ ]]$.
+    However, since there is no annotation to assign the result to,
+    the algorithm must infer the `canonical' principal type.
+    To do that algorithmically, we ensure that 
+    the inferred algorithmic type $[[uQ]]$ is instantiated
+    to the minimal possible type $[[ [uσ]uQ ]]$.
+    The premise $[[Γ ⊢ uQ SC minby uσ]]$
+    provides the minimal instantiation of $[[uQ]]$ \wrt $[[SC]]$. It 
+    guarantees that if we consider all possible substitutions satisfying the inferred 
+    constraints $[[SC]]$, then substitution $[[uσ]]$ will instantiate
+    $[[uQ]]$ to the minimal possible type $[[ [uσ]uQ ]]$.
+    This will be the principal type that we assign to the result of the application 
+    (bound to the variable $[[x]]$)
+    and then we infer the type of the body in the context extended with 
+    the bound variable $[[x : [uσ]uQ]]$.
   \item \ruleref{\ottdruleATUnpackLabel}
-    works in the expected way. First, it infers the 
-    existential type $[[∃nas.iP]]$ of the value being unpacked,
+    first, infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
     and since the type is guaranteed to be normalized, binds 
     the quantified variables with $[[nas]]$.
-    Then it infers the type of the body in the appropriately extended context,
+    Then it infers the type of the body in the appropriately extended context
     and checks that the inferred type does not depend on $[[nas]]$
@@
 \end{theorempreview}
+
+\subsection{Minimal Instantiation}
+The minimal instantiation algorithm is used
+to infer the type of the bound variable in the un-annotated 
+applicative let-binders, as long as there exists a principal
+(minimal) type.  
+Given a positive algorithmic type $[[uP]]$ 
+and a set of constraints $[[SC]]$, it finds the substitution $[[uσ]]$
+respecting $[[SC]]$ (i.e., $[[Θ ⊢ uσ : SC]]$) such that
+it instantiates $[[uP]]$ to the minimal type, in other words 
+for any other substitution $[[uσ']]$ respecting $[[SC]]$,
+we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP]]$.
+
+The minimal instantiation algorithm is defined as follows:
+\begin{algorithm}[Minimal Instantiation]
+  \hfill \\
+  \ottdefnSINGminInstLabeled{}
+\end{algorithm}
 
@@
 
-\subsection{Singularity}
+\subsection{Singularity and Minimal Instantiation}
 \input{\genDir/tex/singularity-statements.tex}
@@
 
-\subsection{Singularity}
+\subsection{Singularity and Minimal Instantiation}
 \input{\genDir/tex/singularity-lemmas.tex}

### 2024-02-13 proofs/tex/decl-typing-lemmas.tex
wip

@@
 
-              And after that, we apply the same rule to infer the required judgement.
+              And after that, we apply the same rule to infer the required judgment.
 
@@
         \item \ruleref{\ottdruleDTtLamLabel}
-            Then we are proving that 
+            Then we prove that 
             $[[Γ ; Φ1 ⊢ λx:iP.c : iP → iN]]$ implies $[[Γ ; Φ2 ⊢ λx:iP.c : iP → iN]]$.
@@
         \item \ruleref{\ottdruleDTVarLetLabel}
-            Then we are proving that 
+            Then we prove that 
             $[[Γ ; Φ1 ⊢ let x = v; c : iN]]$ implies $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$.
@@
         \item \ruleref{\ottdruleDTAppLetLabel}
-            Then we are proving that 
+            Then we prove that 
             $[[Γ ; Φ1 ⊢ let x = v(args); c : iN]]$ implies 
@@
 
-            It is left to show the uniqueness of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.
-            Let us assume that this judgement holds for other $[[iQ']]$, 
+            It is left to show the principality of $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$.
+            Let us assume that this judgment holds for other $[[iQ']]$, 
             i.e.  there exists a tree $T_0$ inferring 
             $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ']]$.
-            Then notice that the induction hypothesis is applicable to
+            Then notice that the induction hypothesis applies to
             $T_0$: the first component of the first component of $\metric{T_0}$
@@
             This way, $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ']]$ holds by the induction hypothesis,
-            but since $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ uniq]]$, we have $[[Γ ⊢ iQ' ≈ iQ]]$.
-
+            but since $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ]]$ is principal, 
+            we have $[[Γ ⊢ iQ' ≥ iQ]]$. 
+            This way, $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ principal]]$.
             Then we infer $[[Γ ; Φ2 ⊢ let x = v(args); c : iN]]$ by \ruleref{\ottdruleDTAppLetLabel}.
@@
         \item \ruleref{\ottdruleDTAppLetAnnLabel}
-            Then we are proving that
+            Then we prove that
             $[[Γ ; Φ1 ⊢ let x:iP = v(args); c : iN]]$ implies

### 2024-02-13 proofs/tex/singularity-lemmas.tex
wip

@@
+\lemMinInstSoundness*
+\begin{proof}
+    We prove it by induction on the inference of $[[Γ ⊢ uP SC minby uσ ]]$.
+    Let us consider the last rule used in the inference.
+    \begin{caseof}
+        \item \ruleref{\ottdruleSINGPUvarLabel}, which means 
+            that the inferred judgment is $[[Γ ⊢  α̂⁺ SC minby (nf(iP) / α̂⁺)]]$, 
+            and by inversion, $[[(α̂⁺ :≥ iP) ∊ SC]]$. Let us show the required properties:
+            \begin{itemize}
+                \item $[[Θ ⊢ (nf(iP) / α̂⁺) : uv α̂⁺]]$ holds trivially;
+                \item $[[Θ ⊢  (nf(iP) / α̂⁺) : SC]]$ holds since
+                    $[[Γ ⊢ nf(iP) : (α̂⁺ :≥ iP)]]$, which is true 
+                    since $[[Γ ⊢ nf(iP) ≥ iP]]$ by the soundness of normalization
+                    (\cref{lemma:normalization-soundness});
+                \item $[[(nf(iP) / α̂⁺)]]$ is normalized trivially;
+                \item let us take an arbitrary $[[Θ ⊢ uσ' : {α̂⁺} ]]$
+                    respecting $[[SC]]$. Since $[[uσ']]$ respects $[[SC]]$,
+                    $[[Γ ⊢ [uσ']α̂⁺ ≥ iP]]$ holds, and then
+                    $[[Γ ⊢ [uσ']α̂⁺ ≥ nf(iP)]]$ holds by the soundness of normalization
+                    and transitivity of subtyping. Finally,
+                    $[[Γ ⊢ [uσ']α̂⁺ ≥ nf(iP)]]$ can be rewritten as
+                    $[[Γ ⊢ [uσ']α̂⁺ ≥ [(nf(iP) / α̂⁺)]α̂⁺]]$.
+            \end{itemize}
+        \item \ruleref{\ottdruleSINGExistsLabel}, which means that 
+            the inferred judgment has form $[[Γ ⊢ ∃nas.uP SC minby uσ ]]$,
+            and by inversion, $[[Γ, nas ⊢ uP SC minby uσ]]$.
+            By applying the induction hypothesis to $[[Γ, nas ⊢ uP SC minby uσ]]$
+            we have 
+            \begin{itemize}
+                \item $[[Θ ⊢ uσ : uv uP]]$, which also means $[[Θ ⊢ uσ : uv ∃nas.uP]]$,
+                \item $[[Θ ⊢ uσ : SC]]$,
+                \item $[[uσ]]$ is normalized, and
+                \item for any other $[[Θ ⊢ uσ' : uv uP ]]$
+                    respecting $[[SC]]$ (\ie $[[Θ ⊢ uσ' : SC]]$), we have
+                    $[[Γ, nas ⊢ [uσ']uP ≥ [uσ]uP ]]$, which immediately implies 
+                    $[[Γ ⊢ [uσ']∃nas.uP ≥ [uσ]∃nas.uP ]]$ (the left-hand side
+                    existential variables are instantiated with the corresponding 
+                    right-hand side existential variables).
+            \end{itemize}
+        \item \ruleref{\ottdruleSINGSingLabel}, which means that
+            the inferred judgment has form $[[Γ ⊢ uP SC minby uσ]]$,
+            and by inversion, $[[uv(uP) ⊆ dom(SC)]]$ and
+            $[[SC|uv(uP) singular with uσ]]$. Let us apply the soundness of singularity
+            (\cref{lemma:singularity-soundness}) to $[[SC|uv(uP) singular with uσ]]$ to
+            obtain the following properties:
+            \begin{itemize}
+                \item $[[ Θ ⊢ uσ : uv(uP) ∩ dom(SC) ]]$, which also means 
+                    $[[Θ ⊢ uσ : uv(uP)]]$,
+                \item $[[Θ ⊢ uσ : SC|uv(uP)]]$,
+                \item $[[uσ]]$ is normalized, and
+                \item for any other $[[Θ ⊢ uσ' : uv(uP) ]]$
+                    respecting $[[SC|uv(uP)]]$, we have
+                    $[[Θ ⊢ uσ' ≈ uσ : uv(uP)]]$. The latter means that
+                    $[[Γ ⊢ [uσ']uP ≈ [uσ]uP]]$, and in particular, 
+                    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP]]$.
+            \end{itemize}
+    \end{caseof}
+
+\end{proof}
+
+\lemMinInstCompleteness*
+\begin{proof}
+    We prove it by induction on $[[uP]]$.
+    \begin{caseof}
+        \item \label{case:min-inst-completeness-uvar} 
+            $[[uP]] = [[α̂⁺]]$. Suppose that $[[α̂⁺ ∉ dom(SC)]]$.
+            Then the instantiation of $[[α̂⁺]]$ is
+            not restricted, and thus, any type can instantiate it.
+            However, among unrestricted instantiations, there is no minimum:
+            any type $[[iP]]$ is \emph{not} a subtype of $[[↓↑iP]]$,
+            which contradicts the assumption.
+            This way, $[[α̂⁺ ∊ dom(SC)]]$.
+
+            If the entry restricting $[[α̂⁺]]$ in $[[SC]]$ is a \emph{subtyping} entry
+            ($[[(α̂⁺ :≥ iQ) ∊ SC]]$), then 
+            we apply \ruleref{\ottdruleSINGPUvarLabel} to infer 
+            $[[Γ ⊢  α̂⁺ SC minby (nf(iQ) / α̂⁺)]]$.
+            It is left to show that $[[ nf(iQ) = nf([uσ]α̂⁺) ]]$.
+            Since $[[Θ ⊢ uσ : SC]]$, and $[[(α̂⁺ :≥ iQ) ∊ SC]]$, 
+            we know that $[[Γ ⊢ [uσ]α̂⁺ ≥ iQ]]$. 
+            On the other hand, let us consider $[[Θ ⊢ uσ' : SC]]$,
+            that copies $[[uσ]]$ on $[[dom(SC)]]$ except $[[α̂⁺]]$,
+            where it is instantiated with $[[iQ]]$.
+            Then $[[Γ ⊢ [uσ']α̂⁺ ≥ [uσ]α̂⁺ ]]$ means $[[Γ ⊢ iQ ≥ [uσ]α̂⁺ ]]$,
+            this way, $[[Γ ⊢ iQ ≈ [uσ]α̂⁺ ]]$, which by \cref{lemma:subt-equiv-algorithmization}
+            means $[[nf(iQ) = nf([uσ]α̂⁺)]]$.
+
+            If the entry restricting $[[α̂⁺]]$ in $[[SC]]$ is an \emph{equivalence} entry
+            ($[[(α̂⁺ :≈ iQ) ∊ SC]]$), then we wish to apply
+            \ruleref{\ottdruleSINGSingLabel}.
+            The first premise $[[uv(α̂⁺) ⊆ dom(SC)]]$  
+            holds by assumption;
+            to infer $[[SC|{α̂⁺} singular with uσ0]]$, 
+            we apply the completeness of singularity (\cref{lemma:singularity-completeness}).
+            It applies because all the substitutions satisfying $[[SC|{α̂⁺}]] = [[(α̂⁺ :≈ iQ)]]$
+            are equivalent on $[[{α̂⁺}]]$ by transitivity of equivalence 
+            (\cref{corollary:equivalence-transitivity}): 
+            the satisfaction of this constraint means that the substitution 
+            sends $[[α̂⁺]]$ to $[[iQ]]$ or an equivalent type.
+            This way, $[[SC|{α̂⁺} singular with uσ0]]$ for some $[[uσ0]]$,
+            which means $[[Γ ⊢ α̂⁺ SC minby uσ0]]$.
+            To show that $[[uσ0 = nf(uσ)]]$ notice that 
+            Since $[[uσ0]]$ is normalized and equivalent to $[[uσ]]$ on $[[{α̂⁺}]]$, 
+            and only has $[[α̂⁺]]$ in its domain
+            (by soundness of singularity, \cref{lemma:singularity-soundness}).
+            This way, $[[Γ ⊢ α̂⁺ SC minby uσ]]$, as required.
+        \item  \label{case:min-inst-completeness-down}
+            $[[uP = ↓uN]]$. Then since
+            $[[Γ ⊢ ↓[uσ']uN ≥ ↓[uσ]uN ]]$ means
+            $[[Γ ⊢ ↓[uσ']uN ≈ ↓[uσ]uN ]]$ by inversion.
+            Then by \cref{lemma:subst-equiv}, 
+            $[[uσ]]$ is equivalent to any other substitution 
+            $[[Θ ⊢ uσ' : uv uN ]]$ satisfying $[[SC|uv uN]]$,
+            hence, the completeness of singularity (\cref{lemma:singularity-completeness})
+            can be applied to conclude that
+            \begin{itemize}
+                \item $[[uv(uN) = dom(SC|uv uN)]]$, then $[[uv(uP) ⊆ dom(SC)]]$,
+                \item $[[SC|uv uN singular with uσ0]]$ for some (normalized) $[[uσ0]]$.
+            \end{itemize}
+            It means $[[Γ ⊢ uP SC minby uσ0 ]]$, and then
+            since $[[uσ0]]$ is normalized and equivalent to $[[uσ0]]$ on $[[uv(uN)]]$,
+            and its domain is $[[uv(uN)]]$, $[[uσ0 = nf(uσ)]]$.
+        % \item $[[uP = ∃nas.uP']]$ 
+        %     Then we wish to apply the induction hypothesis to show 
+        %     $[[Γ, nas ⊢ uP' SC minby nf(uσ)]]$, which would imply 
+        %     $[[Γ ⊢ ∃nas.uP' SC minby nf(uσ)]]$ by \ruleref{\ottdruleMINExistsLabel}.
+        %     Let us demonstrate that the premises of the induction hypothesis hold. 
+        %     \begin{itemize}
+        %         \item  $[[Γ, nas ⊢ Θ]]$ holds by weakening;
+        %         \item  $[[Γ, nas; dom(Θ) ⊢ uP']]$ holds by inversion of $[[Γ; dom(Θ) ⊢ uP]]$;
+        %         \item $[[uv uP' = uv uP]]$, and 
+        %                 $[[Γ ⊢ [uσ']∃nas.uP' ≥ [uσ]∃nas.uP' ]]$ implies
+        %                 $[[Γ, nas ⊢ [uσ']uP' ≥ [uσ]uP' ]]$ by \cref{lemma:subst-equiv}.
+        %     \end{itemize}
+
+        \item $[[uP = ∃nas.β⁺]]$ then 
+            as there are no algorithmic variables in $[[uP]]$, 
+            $[[nf([uσ]uP) = β⁺]]$,
+            and thus, we wish to show that 
+            $[[Γ ⊢ ∃nas.β⁺ SC minby ·]]$.
+            To do so, we apply \ruleref{\ottdruleSINGExistsLabel},
+            and it is left to show that $[[Γ, nas ⊢ β⁺ SC minby · ]]$,
+            which holds vacuously by \ruleref{\ottdruleSINGSingLabel}.
+        \item $[[uP = ∃nas.α̂⁺]]$ then 
+            $[[Γ ⊢ [uσ']∃nas.α̂⁺ ≥ [uσ]∃nas.α̂⁺ ]]$ implies 
+            $[[Γ ⊢ ∃nas.[uσ']α̂⁺ ≥ ∃nas.[uσ]α̂⁺ ]]$ implies 
+            $[[Γ ⊢ [uσ']α̂⁺ ≥ [uσ]α̂⁺ ]]$.
+            It means that $[[uσ]]$ instantiates $[[α̂⁺]]$ to the minimal 
+            type among all the instantiations of $[[α̂⁺]]$ respecting $[[SC]]$.
+            In other words, we can apply the reasoning from 
+            \cref{case:min-inst-completeness-uvar}
+            to conclude that $[[Γ ⊢ α̂⁺ SC minby nf(uσ) ]]$.
+            And then \ruleref{\ottdruleSINGExistsLabel}
+            gives us $[[Γ ⊢ ∃nas.α̂⁺ SC minby nf(uσ) ]]$.
+        \item $[[uP = ∃nas.↓uN]]$ then 
+            $[[Γ ⊢ [uσ']∃nas.↓uN ≥ ∃nas.↓[uσ]uN ]]$ implies 
+            $[[Γ ⊢ ∃nas.↓[uσ']uN ≥ ∃nas.↓[uσ]uN ]]$ implies
+            $[[Γ, nas ⊢ ↓[σ0][uσ']uN ≥ ↓[uσ]uN ]]$ for some $[[σ0]]$ implies
+            $[[Γ, nas ⊢ [σ0][uσ']uN ≈ [uσ]uN ]]$.
+            By \cref{lemma:subst-equiv}, it means in particular
+            that $[[uσ']]$ and $[[uσ]]$ are equivalent on 
+            $[[uv uN]]$. 
+            This way, we can apply the completeness of singularity 
+            (\cref{lemma:singularity-completeness}), and continue as in
+            \cref{case:min-inst-completeness-down}
+            to conclude that $[[Γ, nas ⊢ ↓uN SC minby nf(uσ)]]$.
+            Then by \ruleref{\ottdruleSINGExistsLabel},
+            we have $[[Γ ⊢ ∃nas.↓uN minby nf(uσ)]]$.
+    \end{caseof}
+\end{proof}
+
+\obsMinInstDeterministic*
+\begin{proof}
+    We prove it by induction on $[[Γ ⊢ uP SC minby uσ]]$.
+    It is easy to see that each inference rule is deterministic.
+\end{proof}
+
 \lemEntrySingularitySoundness*

### 2024-02-13 proofs/tex/singularity-statements.tex
wip

@@
+\begin{restatable}[Soundness of Minimal Instantiation]{lemma}{lemMinInstSoundness}
+    \label{lemma:min-inst-soundness}
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
+    If $[[Γ ⊢ uP SC minby uσ ]]$ then
+    \begin{itemize}
+        \item $[[Θ ⊢ uσ : uv uP]]$,
+        \item $[[Θ ⊢ uσ : SC]]$,
+        \item $[[uσ]]$ is normalized, and
+        \item for any other $[[Θ ⊢ uσ' : uv uP ]]$
+            respecting $[[SC]]$ (\ie $[[Θ ⊢ uσ' : SC]]$), we have
+            $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$. 
+    \end{itemize}
+\end{restatable}
+
+\begin{restatable}[Completeness of Minimal Instantiation]{lemma}{lemMinInstCompleteness}
+    \label{lemma:min-inst-completeness}
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
+    $[[Θ ⊢ uσ : uv uP ]]$ respecting $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
+    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
+    respecting $[[SC]]$ ($[[Θ ⊢ uσ' : SC]]$), we have
+    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$. 
+    Then $[[Γ ⊢ uP SC minby nf(uσ) ]]$.
+\end{restatable}
+
+\begin{restatable}[Minimal Instantiation is Deterministic]{observation}{obsMinInstDeterministic}
+    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$.
+    Then $[[Γ ⊢ uP SC minby uσ]]$ and $[[Γ ⊢ uP SC minby uσ']]$ implies $[[uσ = uσ']]$.
+\end{restatable}
+
 \begin{restatable}[Soundness of Entry Singularity]{lemma}{lemEntrySingularitySoundness}
@@
         \item [$+$] Suppose $[[scE singular with iP]]$ for $[[iP]]$ well-formed in $[[Γ]]$.
-            Then $[[ Γ ⊢ iP : scE ]]$
+            Then $[[ Γ ⊢ iP : scE ]]$, $[[iP]]$ is normalized, 
             and for any $[[ Γ ⊢ iP']]$ such that $[[Γ ⊢ iP' : scE]]$, $[[Γ ⊢ iP' ≈ iP]]$;
         \item [$-$] Suppose $[[scE singular with iN]]$ for $[[iN]]$ well-formed in $[[Γ]]$.
-            Then $[[ Γ ⊢ iN : scE ]]$
+            Then $[[ Γ ⊢ iN : scE ]]$, $[[iN]]$ is normalized,
             and for any $[[ Γ ⊢ iN']]$ such that $[[Γ ⊢ iN' : scE]]$, $[[Γ ⊢ iN' ≈ iN]]$.
@@
     Then $[[ Θ ⊢ uσ : Ξ ]]$, 
-     $[[ Θ ⊢ uσ : SC ]]$ and for any 
+     $[[ Θ ⊢ uσ : SC ]]$, $[[uσ]]$ is normalized, and for any 
     $[[uσ']]$ such that $[[Θ ⊢ uσ : SC]]$,

### 2024-03-15 oopsla24/tex/04-algorithm.tex
wip

@@
   to the application inference procedure,
-  returning an algorithmic type $[[↑uQ]]$, 
-  that must be instantiated to a subtype of 
-  the annotation $[[↑uP]]$.
-  Then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
+  returning an algorithmic type $[[iM']]$, 
+  that must be a subtype of the annotation $[[↑uP]]$.
+
+  Then premise $[[Γ; Θ ⊨ iM' ≤ ↑iP ⫤ SC2]]$
   together with $[[Θ ⊢ SC1 & SC2 = SC]]$
-  check whether the instantiation to the annotated type $[[↑iP]]$ is possible,
+  check that $[[iM']]$ can be instantiated to the annotated type $[[↑iP]]$,
   and if it is, the algorithm infers the type of the body in the extended context,

### 2024-03-15 proofs/main.tex
wip

@@
 checking whether one type is a subtype of another according to \cref{def:subDOne}.
+
+\section{Relation to System F}
+
+The types and terms of \fexists have an additional polarization structure.
+Despite that, it is closely related to the well-known \systemf.
+To demonstrate that, we provide two translations in both ways:
+the polarization from \systemf to \fexists and the depolarization from 
+\fexists to \systemf. These translations are done at the level of types and terms,
+and the expected typing preservation properties are proved.
+
+\begin{figure}[h]
+  \ottdefnSFtyLabeled{}
+  \ottdefnSFtyALabeled{}
+  \caption{Typing rules of \systemf}
+  \label{fig:systemf-typing}
+\end{figure}
+
+\begin{figure}[h]
+  \ottdefnSFWFLabeled{}
+\end{figure}
+
+\subsection{Type-level Translation}
+
+\begin{figure}[h]
+  \ottfundefnunpolP{}
+  \ottfundefnunpolN{}
+\end{figure}
+
+\begin{figure}[h]
+  \ottdefnELNLabeled{}
+  \ottdefnELPLabeled{}
+  \caption{Subtyping elaboration from \fexists to \systemf}
+  \label{fig:subt-elaboration}
+\end{figure}
+
+% \lemmaSubtElabSoundness
+
+% \lemmaElabRemovable
+
+% \lemmaDepolSubst
+
+\subsection{Term-level Translation}
+
+The term-level translation is defined not solely for terms, but for typing
+derivations. The reason for that is that the terms of \systemf and \fexists
+contain different typing information that cannot be reconstructed without the
+appropriate derivation tree. For instance, the lambda expressions in \systemf do
+not have the bound variable annotation, while the lambda expressions in \fexists
+do; on the other hand, \fexists has rich subtyping, which cannot be expressed in
+in terms of polymorphic type instantiation of \systemf.
+
+\begin{figure}[h]
+  \ottdefnTELPLabeled{}
+  \caption{Typing elaboration from \fexists to \systemf}
+  \label{fig:typing-elaboration}
+\end{figure}
+
+\begin{figure}[h]
+  \ottdefnTELNLabeled{}
+\end{figure}
+
+\begin{figure}[h]
+  \ottdefnTELAppLabeled{}
+\end{figure}
+
+\begin{figure}[h]
+  \ottdefnSFfPolLabeled{}
+  \caption{Polarization: translation from \systemf to \fexists}
+  \label{fig:systemf-polarization}
+\end{figure}
+
+% \lemmaDepolarSoundness*
+
+% \lemmaPolSubst
+
+% \lemmaPolRemovable
+
+% \lemmaTypPolWf*
+
+% \lemmaPolarSoundness
 
@@
     and to bind them, the algorithm needs to fix the canonical one.
-  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
-    with annotated let-bindings. The first premise infers the type of the head of the application,
+  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to
+    type the \emph{annotated} applicative let-binders. 
+    The first premise infers the type of the head of the application,
     which must be a thunked computation. Then if after applying it 
-    to the arguments, the resulting type can be equated to the annotated one,
+    to the arguments, the resulting type can be instantiated to the annotated one,
     we infer the body of the let-binding in the context extended with the bound variable.
@@
     instantiation to a declarative type must be associated with the bound variable 
-    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
+    $[[x]]$; then premise $[[Γ; Θ ⊨ uM' ≤ ↑iP ⫤ SC2]]$
     together with $[[Θ ⊢ SC1 & SC2 = SC]]$
@@
 The correctness of the algorithm consists of its soundness and 
-completeness, which is by mutual
+completeness, which is proved by mutual
 induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
@@
 
+\section{Relation to \systemf}
+\input{\genDir/tex/translation-statements.tex}
+
 \section{Theorem Statements: Algorithmic}
@@
 
+\section{Relation to \systemf}
+\input{\genDir/tex/translation-proofs.tex}
+
 \section{Properties of the Algorithmic Type System}

### 2024-03-15 proofs/tex/decl-typing-lemmas.tex
wip

@@
             $[[Γ; Φ1 ⊢ v : iP]]$ to obtain $[[Γ; Φ2 ⊢ v : iP]]$ 
-            of the same pure size.
+            of the same size.
 
@@
             Then we infer $[[Γ ; Φ2 ⊢ let x = v; c : iN]]$ by \ruleref{\ottdruleDTVarLetLabel}.
+
+        \item \ruleref{\ottdruleDTCVarLetLabel}
+            Then we prove that 
+            $[[Γ ; Φ1 ⊢ let x : iP = c; c' : iN]]$ implies $[[Γ ; Φ2 ⊢ let x : iP = c; c' : iN]]$.
+
+            First, we apply the induction hypothesis to 
+            $[[Γ; Φ1 ⊢ c : iM]]$ to obtain $[[Γ; Φ2 ⊢ c : iM]]$
+            of the same size.
+
+            Then we apply the induction hypothesis to
+            the equivalent contexts $[[Γ ⊢ Φ1, x:iP ≈ Φ2, x:iP]]$
+            and the premise $[[Γ ; Φ1, x:iP ⊢ c' : iN]]$ to obtain
+            $[[Γ ; Φ2, x:iP ⊢ c' : iN]]$.
+            Then we infer $[[Γ ; Φ2 ⊢ let x : iP = c; c' : iN]]$ by \ruleref{\ottdruleDTCVarLetLabel}.
 
@@
                 \item $[[Γ ; Φ1 ⊢ v : ↓iM]]$ into $[[Γ ; Φ2 ⊢ v : ↓iM]]$,
-                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> ↑iQ]]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> ↑iQ]]$, 
+                \item $[[Γ ; Φ1 ⊢ iM ● args ⇒> iM']]$ into $[[Γ ; Φ2 ⊢ iM ● args ⇒> iM']]$, 
                     and
@@
 
-            Notice that $[[Γ ⊢ iP]]$ and $[[Γ ⊢ ↑iQ ≤ ↑iP]]$ 
+            Notice that $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iM' ≤ ↑iP]]$ 
             do not depend on the variable context, and hold by assumption.

### 2024-04-03 oopsla24/tex/02-overview.tex
add proofs: elaboration

@@
 
-\subsection{Relation to \systemf}
-
-To establish a correspondence between \fexists and standard unpolarized systems,
-we present a translation between \fexists and \systemf. The translation of
-\systemf to \fexists is given in \cref{fig:polarization}, and the backward translation
-is given in \cref{fig:depolarization}.
-
-\begin{figure}[h]
-  \begin{equation*}
-  \begin{aligned}
-    [[|↓iN|]] &= [[|iN|]] \\
-    [[|↑iP|]] &= [[|iP|]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[|α±|]] &= [[a]] \\
-    [[|iP → iN|]] &= [[|iP| → |iN|]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
-    [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
-  \end{aligned}
-  \end{equation*}
-  \caption{Type depolarization}
-  \label{fig:depolarization}
-\end{figure}
-
-The translation from \systemf to \fexists can be done in two canonical ways
-depending on the desired polarity. The polarity will also determine the 
-execution strategy \ilyam{how?}.
-Moreover, the translation procedure depends on the \emph{polarity} of the
-free type variables. Thus, the translation functions 
-$[[-| fT |-Γ]]$ and $[[+| fT |+Γ]]$ are indexed by a context $[[Γ]]$, 
-containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.
-
-\begin{figure}[h]
-  \begin{equation*}
-  \begin{aligned}
-    [[-| a |-Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
-    [[-| a |-Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
-    [[-| fA → fB |-Γ]] &= [[ +|fA|+Γ → -|fB|-Γ  ]] \\
-    [[-| ∀α.fT |-Γ]] &= [[ ∀α⁺.-|fT|-Γ,α⁺ ]] \\
-    [[-| ∃α.fT |-Γ]] &= [[ ↑ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    [[+| a |+Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
-    [[+| a |+Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
-    [[+| fA → fB |+Γ]] &= [[ ↓ +|fA|+Γ → -|fB|-Γ  ]] \\
-    [[+| ∀α.fT |+Γ]] &= [[ ↓∀α⁺.-|fT|-Γ,α⁺ ]] \\
-    [[+| ∃α.fT |+Γ]] &= [[ ∃α⁻.+|fT|+Γ,α⁻ ]] \\
-  \end{aligned}
-  \end{equation*}
-  \caption{Type polarization}
-  \label{fig:polarization}
-\end{figure}
-
 \subsection{The key ideas of the algorithm}

### 2024-04-03 proofs/main.tex
add proofs: elaboration

@@
 checking whether one type is a subtype of another according to \cref{def:subDOne}.
-
-\section{Relation to System F}
-
-The types and terms of \fexists have an additional polarization structure.
-Despite that, it is closely related to the well-known \systemf.
-To demonstrate that, we provide two translations in both ways:
-the polarization from \systemf to \fexists and the depolarization from 
-\fexists to \systemf. These translations are done at the level of types and terms,
-and the expected typing preservation properties are proved.
-
-\begin{figure}[h]
-  \ottdefnSFtyLabeled{}
-  \ottdefnSFtyALabeled{}
-  \caption{Typing rules of \systemf}
-  \label{fig:systemf-typing}
-\end{figure}
-
-\begin{figure}[h]
-  \ottdefnSFWFLabeled{}
-\end{figure}
-
-\subsection{Type-level Translation}
-
-\begin{figure}[h]
-  \ottfundefnunpolP{}
-  \ottfundefnunpolN{}
-\end{figure}
-
-\begin{figure}[h]
-  \ottdefnELNLabeled{}
-  \ottdefnELPLabeled{}
-  \caption{Subtyping elaboration from \fexists to \systemf}
-  \label{fig:subt-elaboration}
-\end{figure}
-
-% \lemmaSubtElabSoundness
-
-% \lemmaElabRemovable
-
-% \lemmaDepolSubst
-
-\subsection{Term-level Translation}
-
-The term-level translation is defined not solely for terms, but for typing
-derivations. The reason for that is that the terms of \systemf and \fexists
-contain different typing information that cannot be reconstructed without the
-appropriate derivation tree. For instance, the lambda expressions in \systemf do
-not have the bound variable annotation, while the lambda expressions in \fexists
-do; on the other hand, \fexists has rich subtyping, which cannot be expressed in
-in terms of polymorphic type instantiation of \systemf.
-
-\begin{figure}[h]
-  \ottdefnTELPLabeled{}
-  \caption{Typing elaboration from \fexists to \systemf}
-  \label{fig:typing-elaboration}
-\end{figure}
-
-\begin{figure}[h]
-  \ottdefnTELNLabeled{}
-\end{figure}
-
-\begin{figure}[h]
-  \ottdefnTELAppLabeled{}
-\end{figure}
-
-\begin{figure}[h]
-  \ottdefnSFfPolLabeled{}
-  \caption{Polarization: translation from \systemf to \fexists}
-  \label{fig:systemf-polarization}
-\end{figure}
-
-% \lemmaDepolarSoundness*
-
-% \lemmaPolSubst
-
-% \lemmaPolRemovable
-
-% \lemmaTypPolWf*
-
-% \lemmaPolarSoundness
 
@@
 \end{itemize}
+
+\section{Relation to System F}
+
+Although based on \systemf, \fexists has an additional polarization structure.
+To demonstrate the relation between these systems we establish translations in
+both ways: the polarization from \systemf to \fexists and the depolarization
+from \fexists to \systemf. These translations are done at the level of types and
+terms, and the expected typing preservation properties are proved.
+
+First, let us agree on the variant of \systemf that we use to establish the
+relation with \fexists. At the type level, we have variables, functional arrows,
+and universal quantifiers, and define existential quantifiers as a syntactic
+sugar using standard encodings. At the term level, we have variables,
+unannotated lambda abstractions, type abstractions, and term-level applications,
+but not type applications (they can be done implicitly); for convenience, we
+introduce multi-argument application, let-blinders, and existential constructors
+and eliminators as syntactic sugar (see \cref{fig:systemf-grammar}).
+
+\begin{figure}
+  \begin{minipage}[t]{0.39\linewidth}
+    \vspace*{0pt}
+    \ottgrammartabular{
+      \ottfT\ottprodnewline
+      \ottprodline{}{}{}{}{}{}\ottprodnewline
+      \ottprodline{}{}{}{}{}{}\ottprodnewline
+      \ottprodline{|}{[[∃as.fT]]}{}{}{}{}\ottprodnewline
+      \ottprodline{}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}{}\ottinterrule
+    }
+  \end{minipage}
+  \begin{minipage}[t]{0.6\linewidth}
+    \vspace*{0pt}
+    \ottgrammartabular{
+      \ottft\ottprodnewline
+      \ottprodline{}{}{}{}{}{}\ottprodnewline
+      \ottprodline{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
+      \ottprodline{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
+      \ottprodline{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
+      \ottprodline{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
+    }
+  \end{minipage}
+  \caption{Grammar of \systemf}
+  \label{fig:systemf-grammar}
+\end{figure}
+
+\begin{figure}
+  \ottdefnSFtyLabeledTwoColumns{}
+  \caption{Typing rules of \systemf}
+  \label{fig:systemf-typing}
+\end{figure}
+
+\begin{observation}
+  The following rules are admissible in \systemf: \hfill
+
+  \begin{minipage}{0.5\textwidth}
+    \ottusedrule{\ottdruleSFtyAUnpackLabeled{}}
+  \end{minipage}
+  \begin{minipage}{0.5\textwidth}
+    \ottusedrule{\ottdruleSFtyAPackLabeled{}}
+    \ottusedrule{\ottdruleSFtyALetLabeled{}}
+  \end{minipage}
+\end{observation}
+
+% \begin{figure}[h]
+%   \ottdefnSFWFLabeled{}
+% \end{figure}
+
+\subsection{Type-level Translation}
+
+The translation between the types is defined in 
+\cref{fig:depolarization} and \cref{fig:polarization}.
+
+The depolarization---translation from \fexists to \systemf---is
+straightforward: it recursively
+\begin{enumerate*}
+  \item [(i)] removes the shift operators $[[↑]]$ and $[[↓]]$, and
+  \item [(ii)] removes the sign annotations from the type variables.
+\end{enumerate*}
+
+The polarization---translation from \systemf to \fexists---is more complex.
+There exist several ways to define it since any \systemf type can
+be polarized either positively or negatively.
+We chose the positive translation: every term of \systemf is translated
+into a value. Although this translation does not minimize the number of
+inserted shifts, it is more straightforward to consistently
+lift it to the level of terms.
+
+\begin{figure}[ht]
+  \begin{minipage}[t]{0.65\linewidth}
+    \centering
+      \begin{minipage}[t]{0.32\linewidth}
+      \ottfundefnunpolP{}
+      \end{minipage}
+      \begin{minipage}[t]{0.32\linewidth}
+      \ottfundefnunpolN{}
+      \end{minipage}
+    \caption{Type Depolarization}
+    \label{fig:depolarization}
+  \end{minipage}
+  \hspace{0.5cm}
+  \begin{minipage}[t]{0.3\linewidth}
+    \centering
+    \ottfundefnpolarP{}
+    \caption{Type Polarization}
+    \label{fig:polarization}
+  \end{minipage}
+\end{figure}
+
+% \lemmaSubtElabSoundness
+
+% \lemmaElabRemovable
+
+% \lemmaDepolSubst
+
+\subsection{Term-level Translation}
+
+The term-level translation is defined not solely for terms, but for typing
+derivations (we call this translation \emph{elaboration}). 
+The reason for that is that the terms of \systemf and \fexists
+contain different typing information that cannot be reconstructed without the
+appropriate derivation tree. For instance, the lambda expressions in \systemf do
+not have the bound variable annotation, while the lambda expressions in \fexists
+do; on the other hand, \fexists has richer subtyping, which cannot be expressed in
+in terms of polymorphic type instantiation of \systemf.
+
+The elaboration \fexists $\leadsto$ \systemf
+annotates each judgment of an \fexists typing derivation tree
+with the corresponding \systemf term. This way, 
+we define elaboration for each kind of judgment:
+subtyping, positive typing, negative typing, and application typing.
+
+\paragraph{From \fexists to \systemf: Subtyping Elaboration}
+The subtyping elaboration is defined in \cref{fig:subt-elaboration}.
+The soundness property that is preserved by the elaboration is stated in
+\cref{lemma:subt-elab-soundness}. Informally, $[[Γ ⊢ iN ≤ iM ⤳ ft]]$
+guarantees that $[[ft]]$ is a \systemf term that represents 
+a conversion (\ie a function) from the depolarized $[[|iN|]]$ to $[[|iM|]]$.
+Symmetrically, $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ implies $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$.
+\begin{itemize}
+  \item Rules \ruleref{\ottdruleELNVarLabel} and \ruleref{\ottdruleELPVarLabel} 
+    are trivial: if the left-hand side and the right-hand side types are the 
+    same, the function converting one to another is the identity.
+  \item Rules \ruleref{\ottdruleELShiftULabel} and \ruleref{\ottdruleELShiftDLabel}
+    are symmetric. Since the shifts are removed during the depolarization,
+    the conversion function is obtained by the recursive call to the 
+    premise representing the subtyping of the required order
+    (the mutual subtyping equivalence is defined as two separate judgments).
+    Notice that the subtyping in the other order is still required
+    so that the removal of the elaboration produces a correct
+    subtyping inference tree in the original \fexists.
+  \item Rule \ruleref{\ottdruleELArrowLabel} allows us to
+    define elaboration between function. Since the subtyping is 
+    contravariant on the argument type and covariant on the result type,
+    we can use the elaboration functions acquired in the premises
+    (their types are $[[|iQ| → |iP|]]$ and $[[|iN| → |iM|]]$)
+    to construct the required elaboration term of type 
+    $[[(|iP| → |iN|) → |iQ| → |iM|]]$.
+  \item Rules 
+    \ruleref{\ottdruleELForallLabel}, and \ruleref{\ottdruleELExistsLabel}
+    are more involved. However, the soundness property is preserved
+    by \cref{lemma:depol-subst} allowing one to distribute
+    depolarization over the substitution whereby
+    translating the instantiation substitution $[[σ]]$ from \fexists to \systemf.
+\end{itemize}
+
+\paragraph{From \fexists to \systemf: Typing Elaboration}
+Using the subtyping elaboration, we define the elaboration for the
+positive/negative typing and application typing in
+\cref{fig:typing-elaboration}.
+The soundness property (\cref{lemma:depolar-soundness})
+guarantees that if the initial term $[[v]]$ has type $[[iP]]$ in \fexists,
+then the result of the elaboration of this judgment $[[ft]]$ has
+type $[[|iP|]]$ in \systemf. The negative typing has a symmetric property.
+As with the subtyping elaboration, 
+the rules in \cref{fig:typing-elaboration} are obtained from 
+the \fexists declarative typing rules (\cref{def:declarative-typing})
+by annotating each judgment with the elaboration term
+so that the typing is preserved.
+
+\begin{figure}[h]
+  \ottdefnELLabeled
+  \caption{Subtyping elaboration from \fexists to \systemf}
+  \label{fig:subt-elaboration}
+\end{figure}
+
+\clearpage % This command will start a new page
+\begin{figure}[t]
+  \ottdefnTELNLabeledTwoColumns{}
+
+  \hfill
+
+  \ottdefnTELPAppLabeled{}
+  \caption{Typing elaboration from \fexists to \systemf}
+  \label{fig:typing-elaboration}
+\end{figure}
+
+\paragraph{From \systemf to \fexists}
+
+The other direction of translation is also represented as an elaboration. The
+inference rules are defined in \cref{fig:systemf-polarization}, and constitute
+the rules of \systemf (\cref{fig:systemf-typing}). annotated with the
+elaboration terms. The elaborated term is chosen in such a way that
+the soundness property (\cref{lemma:polar-soundness}) holds: 
+$[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ implies $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$.
+
+\begin{figure}[h]
+  \ottdefnELfPolLabeled{}
+  \caption{Typing elaboration form from \systemf to \fexists}
+  \label{fig:systemf-polarization}
+\end{figure}
 

### 2024-04-03 proofs/tex/translation-proofs.tex
add proofs: elaboration

\lemmaElabRemovable*
\begin{proof}
  We prove it separately in both directions of the implication.
  both of the implications are proved by simple induction on the judgment.
\end{proof}

\lemmaDepolSubst*
\begin{proof}
  By mutual induction on the type $[[iN]]$ and $[[iP]]$.
\end{proof}

\lemmaSubtElabSoundness*
\begin{proof}
  We prove it by (mutual) induction on $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ and 
  $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$.
  Let us consider the last rule applied to infer this judgment.
  \begin{caseof}
    \item \ruleref{\ottdruleELNVarLabel} 
      Notice that $[[|α⁻| = a]]$. Then
      $[[|Γ| ; · ⊢ λx.x : a → a]]$ immediately by 
      \ruleref{\ottdruleSFtytLamLabel} and \ruleref{\ottdruleSFtyVarLabel}.

    \item \ruleref{\ottdruleELPVarLabel}
      This case is symmetric to the previous one.

    \item \ruleref{\ottdruleELShiftULabel}
      Notice that $[[|↑iP| = |iP|]]$,
      and by induction hypothesis $[[Γ ⊢ iQ ≥ iP ⤳ ft]]$ implies 
      $[[|Γ| ; · ⊢ ft : |iP| → |iQ|]]$.

    \item \ruleref{\ottdruleELShiftDLabel}
      This case is symmetric to the previous one.

    \item \ruleref{\ottdruleELArrowLabel}
      We need to show that 
      $[[|Γ|; · ⊢ λx. λy. ft' (x (ft y)) : (|iP| → |iN|) → |iQ| → |iM|]]$.
      By induction hypothesis applied to the premises, we know that 
      $[[|Γ|; · ⊢ ft : |iQ| → |iP|]]$ and 
      $[[|Γ|; · ⊢ ft' : |iN| → |iM|]]$.
      Then the required typing judgment follows from
      standard rules of \systemf.

    \item \ruleref{\ottdruleELForallLabel}
      We need to show that 
      $[[|Γ|; · ⊢ λx. Λbs. ft x : (∀as.|iN|) → ∀bs.|iM|]]$.
      By the induction hypothesis applied to the premise, 
      we know that $[[|Γ|; · ⊢ ft : |[σ]iN| → |iM|]]$, 
      that we rewrite as $[[|Γ|; · ⊢ ft : [|σ|]|iN| → |iM|]]$
      by \cref{lemma:depol-subst}.
      Then the required typing judgment follows from
      the standard rules of \systemf
      notice that 
      $[[|Γ|, bs; x : ∀as.|iN|  ⊢ x : ∀as.|iN|]]$
      implies 
      $[[|Γ|, bs; x : ∀as.|iN|  ⊢ x : [|σ|]|iN|]]$ by
      \ruleref{\ottdruleSFtyTAppLabel}.

    \item \ruleref{\ottdruleELExistsLabel}
      We need to show that
      $[[|Γ| ; · ⊢  λx. unpack(bs, y) = x; pack (ft y) : ∃as.|iP| : |∃nbs.iQ| → |∃nas.iP| ]]$.
      By definition, 
      $[[|∃nbs.iQ| → |∃nas.iP| = ∃bs.|iQ| → ∃as.|iP|]]$.
      Then by applying \ruleref{\ottdruleSFtytLamLabel} and 
      the admissible rules \ruleref{\ottdruleSFtyAUnpackLabel}
      and \ruleref{\ottdruleSFtyAPackLabel}, 
      it suffices to show
      $[[|Γ|, bs; y : |iQ| ⊢ (ft y) : [|σ|]|iP|]]$
      and that $[[|Γ|, bs ⊢ |σ| : as]]$. 
      The latter follows from $[[Γ, nbs ⊢ σ : {nas}]]$.
      The former holds by \ruleref{\ottdruleSFtyAppLabel} 
      and the induction hypothesis applied to the premise
      ($[[|Γ|; · ⊢ ft : |iQ| → |[σ]iP|]]$, and by \cref{lemma:depol-subst}, $[[|[σ]iP| = [|σ|]|iP|]]$).
  \end{caseof}
\end{proof}

\lemmaDepolarSoundness*
\begin{proof}
  We prove it by (mutual) induction on the derivation of $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ 
  and $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ .

  \begin{caseof}
    \item \ruleref{\ottdruleTELArrowAppLabel}
      Suppose that $[[Γ ; Φ ⊢ iQ → iN ● v , args ⇒> iM ⤳ fe'; fe ft, fargs]]$.
      Then we know 
      \begin{itemize}
        \item $[[Γ ; Φ ⊢ v : iP ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$;
        \item $[[Γ ⊢ iQ ≥ iP ⤳ fe]]$, and then by the soundness of 
          subtyping elaboration (\cref{lemma:subt-elab-soundness}),
          $[[|Γ| ; · ⊢ fe : |iP| → |iQ|]]$;
        \item \label{case:depolar-soundness-app-ih}
          $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe'; fargs]]$, and then by the induction hypothesis, 
          $[[|Γ| ; |Φ|, x:|iN| ⊢ fe' (x ● fargs) : |iM|]]$.
      \end{itemize}
      We wish to show that $[[|Γ| ; |Φ|, f : |iQ| → |iN| ⊢ fe' (f ● (fe ft, fargs)) : |iM|]]$,
      that is $[[|Γ| ; |Φ|, f : |iQ| → |iN| ⊢ fe' (f (fe ft) ● fargs) : |iM|]]$.
      By substitution applied to \cref{case:depolar-soundness-app-ih}, it suffices to show that 
      $[[|Γ| ; |Φ|, f : |iQ| → |iN| ⊢ f (fe ft) : |iM|]]$,
      which is inferred by multiple applications of  \ruleref{\ottdruleSFtyAppLabel} to the
      judgments stated above.
    \item \ruleref{\ottdruleTELForallAppLabel}
      Then we know that 
      $[[Γ ; Φ ⊢ ∀pas.iN ● args ⇒> iM  ⤳ fe ; fargs]]$, and by inversion:
      $[[Γ ; Φ ⊢ [σ]iN ● args ⇒> iM  ⤳ fe ; fargs]]$,
      where $[[σ]]$ is a substitution from $[[pas]]$ to $[[Γ]]$.
      By the induction hypothesis, we know that
      $[[|Γ| ; |Φ|, x : [|σ|]|iN| ⊢ fe (x ● fargs) : |iM|]]$.

      We wish to show that 
      $[[|Γ| ; |Φ|, x : ∀as.|iN| ⊢ fe (x ● fargs) : |iM| ]]$.
      To do that, we can modify the inference tree of 
      $[[|Γ| ; |Φ|, x : [|σ|]|iN| ⊢ fe (x ● fargs) : |iM|]]$
      to infer 
      $[[|Γ| ; |Φ|, x  : ∀as.|iN| ⊢ fe (x ● fargs) : |iM|]]$.
      The only thing we need to change is every time 
      the inference asks for the type of $[[x]]$ in the context
      (\ie in the leaf rule \ruleref{\ottdruleSFtyVarLabel}), 
      we use the combination of \ruleref{\ottdruleSFtyTAppLabel} and \ruleref{\ottdruleSFtyVarLabel}
      instead of only \ruleref{\ottdruleSFtyVarLabel} to infer $[[|Γ| ; |Φ|, x : ∀as.|iN| ⊢ x : [|σ|]|iN| ]]$.

    \item \ruleref{\ottdruleTELEmptyAppLabel}
      Then we know $[[Γ ; Φ ⊢ iN ● · ⇒> iN'  ⤳  fe ; ·]]$.
      By inversion, we have $[[Γ ⊢ iN' ≤ iN ⤳ fe]]$, 
      which by the soundness of subtyping elaboration (\cref{lemma:subt-elab-soundness}),
      means $[[|Γ| ; · ⊢ fe : |iN| → |iN'|]]$.

      Then the desired $[[|Γ| ; |Φ|, x:|iN| ⊢ fe x : |iN'|]]$
      follows from the standard rules of \systemf.

    \item \ruleref{\ottdruleTELVarLabel}, 
      which means we wish to prove that  $[[Γ ; Φ ⊢ x : iP ⤳ x]]$
      implies $[[|Γ| ; |Φ| ⊢ x : |iP|]]$.
      By inversion, $[[x : iP ∊ Φ]]$, and thus, 
      by definition of context depolarization, $[[x : |iP| ∊ |Φ|]]$.
      Then by \ruleref{\ottdruleSFtyVarLabel}, we infer $[[|Γ| ; |Φ| ⊢ x : |iP|]]$.

    \item \ruleref{\ottdruleTELThunkLabel} 
      Then we know that 
      $[[Γ ; Φ ⊢ {c} : ↓iN ⤳ ft]]$ and by the inversion, 
      $[[Γ ; Φ ⊢ c : iN ⤳ ft]]$.
      Then the desired $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ follows from the induction hypothesis.
    \item \ruleref{\ottdruleTELReturnLabel}
      The proof is symmetric to the previous case.

    \item \ruleref{\ottdruleTELPAnnotLabel} 
      Then we know that $[[Γ ; Φ ⊢ (v : iQ) : iQ ⤳ fe ft]]$
      and by inversion: 
      \begin{itemize}
        \item $[[Γ ; Φ ⊢ v : iP ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$;
        \item $[[Γ ⊢ iQ ≥ iP ⤳ fe]]$, and then by the soundness of 
          subtyping elaboration (\cref{lemma:subt-elab-soundness}),
          $[[|Γ| ; · ⊢ fe : |iP| → |iQ|]]$.
      \end{itemize}

      The desired $[[|Γ| ; |Φ| ⊢ fe ft : |iQ|]]$
      follows by \ruleref{\ottdruleSFtyAppLabel} applied to the judgement stated above.

    \item For \ruleref{\ottdruleTELNAnnotLabel}, \ruleref{\ottdruleTELPEquivLabel} and \ruleref{\ottdruleTELNEquivLabel}
      the proof is analogous to the previous case.

    \item \ruleref{\ottdruleTELtLamLabel}
      We know that $[[Γ ; Φ ⊢ λx:iP . c : iP → iN  ⤳ λx.ft]]$ and by inversion:
      $[[Γ ; Φ, x:iP ⊢ c : iN  ⤳ ft]]$, then by the induction hypothesis,
      $[[|Γ| ; |Φ|, x:|iP| ⊢ ft : |iN|]]$.
      By applying \ruleref{\ottdruleSFtytLamLabel} to this judgment, we infer
      the desired $[[|Γ| ; |Φ| ⊢ λx.ft : |iP| → |iN|]]$.

    \item \ruleref{\ottdruleTELTLamLabel}
      We know that $[[Γ ; Φ ⊢ Λα⁺ . c : ∀α⁺.iN  ⤳  Λa.ft]]$ and by inversion:
      $[[Γ, α⁺ ; Φ ⊢ c : iN  ⤳  ft]]$, then by the induction hypothesis,
      $[[|Γ|, a ; |Φ| ⊢ ft : |iN|]]$.
      By applying \ruleref{\ottdruleSFtyTLamLabel} to this judgment, we infer
      $[[|Γ| ; |Φ| ⊢ Λa.ft : ∀a.|iN|]]$, that is 
      $[[|Γ| ; |Φ| ⊢ Λa.ft : |∀α⁺.iN|]]$.

    \item \ruleref{\ottdruleTELCVarLetLabel}
      We know that $[[Γ; Φ ⊢ let x : iP = c; c' : iN  ⤳  let x = (fe ft) ; ft']]$ 
      and by inversion:
      \begin{itemize}
        \item $[[Γ; Φ ⊢ c : iM ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft : |iM|]]$;
        \item $[[Γ ⊢ iM ≤ ↑iP ⤳ fe]]$, and then by the soundness of 
          subtyping elaboration (\cref{lemma:subt-elab-soundness}),
          $[[|Γ| ; · ⊢ fe : |iM| → |iP|]]$;
        \item $[[Γ; Φ, x:iP ⊢ c' : iN ⤳ ft']]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iP| ⊢ ft' : |iN|]]$.
      \end{itemize}
      To infer $[[|Γ| ; |Φ| ⊢ let x = (fe ft) ; ft' : |iN|]]$,
      we apply the admissible \systemf rule \ruleref{\ottdruleSFtyALetLabel}.
      Both premises hold: $[[|Γ| ; |Φ| ⊢ fe ft : |iP|]]$ follows form \ruleref{\ottdruleSFtyAppLabel},
      and $[[|Γ| ; |Φ|, x:|iP| ⊢ ft' : |iN|]]$ holds as stated above.

    \item \ruleref{\ottdruleTELVarLetLabel}
      We know that 
      $[[Γ; Φ ⊢ let x = v; c : iN  ⤳  let x = ft ; ft']]$
      and by inversion:
      \begin{itemize}
        \item $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ implying $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$; and
        \item $[[Γ; Φ, x:iP ⊢ c : iN ⤳ ft']]$ implying $[[|Γ| ; |Φ|, x:|iP| ⊢ ft' : |iN|]]$.
      \end{itemize}
      Then the desired $[[|Γ| ; |Φ| ⊢ let x = ft ; ft' : |iN|]]$
      follows by the admissible \systemf rule \ruleref{\ottdruleSFtyALetLabel}
      applied to the judgments above.

    \item \ruleref{\ottdruleTELAppLetLabel}
      We know that
      $[[Γ ; Φ ⊢ let x = v(args); c : iN ⤳ let x = (fe (ft' ● fargs)) ; ft]]$
      and by inversion, in particular, we have:
      \begin{itemize}
        \item $[[Γ ; Φ ⊢ v : ↓iM ⤳ ft']]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft' : |iM|]]$;
        \item $[[Γ; Φ ⊢ iM ● args ⇒> ↑iQ ⤳ fe ; fargs]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iM| ⊢ fe (x ● fargs) : |iQ|]]$;
        \item $[[Γ; Φ, x:iQ ⊢ c : iN ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iQ| ⊢ ft : |iN|]]$.
      \end{itemize}
      We wish to show that 
      $[[|Γ| ; |Φ| ⊢ let x = (fe (ft' ● fargs)) ; ft : |iN|]]$.
      By applying \ruleref{\ottdruleSFtyALetLabel},
      we reduce this to
      $[[|Γ| ; |Φ| ⊢ fe (ft' ● fargs) : |iQ|]]$,
      which holds by substitution $[[ft']]$ for $[[x]]$ in
      $[[|Γ| ; |Φ|, x:|iM| ⊢ fe (x ● fargs) : |iQ|]]$
      (which is possible since $[[|Γ| ; |Φ| ⊢ ft' : |iM|]]$).
    \item \ruleref{\ottdruleTELAppLetAnnLabel}
      By inversion of 
      $[[Γ ; Φ ⊢ let x:iP = v(args); c : iN ⤳ let x = fe' (fe (ft' ● fargs)); ft]]$, 
      we know that
      \begin{itemize}
        \item $[[Γ ; Φ ⊢ v : ↓iM ⤳ ft']]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft' : |iM|]]$;
        \item $[[Γ; Φ ⊢ iM ● args ⇒> iM' ⤳ fe ; fargs]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iM| ⊢ fe (x ● fargs) : |iM'|]]$. 
          By substitution, it implies $[[|Γ| ; |Φ|⊢ fe (ft' ● fargs) : |iM'|]]$. 
        \item $[[Γ ⊢ iM' ≤ ↑iP ⤳ fe']]$, and then by the soundness of 
          subtyping elaboration (\cref{lemma:subt-elab-soundness}),
          $[[|Γ| ; · ⊢ fe' : |iM'| → |iP|]]$;
        \item $[[Γ; Φ, x:iP ⊢ c : iN ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iP| ⊢ ft : |iN|]]$.
      \end{itemize}

      To infer $[[|Γ| ; |Φ| ⊢ let x = fe' (fe (ft' ● fargs)) ; ft : |iN|]]$,
      we apply \ruleref{\ottdruleSFtyALetLabel},
      so it is left to show that $[[|Γ| ; |Φ| ⊢ fe' (fe (ft' ● fargs)) : |iP|]]$, 
      which follows from \ruleref{\ottdruleSFtyAppLabel} and the judgments stated above.

    \item \ruleref{\ottdruleTELUnpackLabel}
      By inversion of 
      $[[Γ ; Φ ⊢ let∃ (nas, x) = v; c : iN   ⤳  unpack (as, x) = ft; ft']]$, 
      we have:
      \begin{itemize}
        \item $[[Γ ; Φ ⊢ v : ∃nas.iP ⤳ ft]]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ| ⊢ ft : ∃as.|iP|]]$;
        \item $[[Γ; Φ, x:iP ⊢ c : iN ⤳ ft']]$, and then by the induction hypothesis,
          $[[|Γ| ; |Φ|, x:|iP| ⊢ ft' : |iN|]]$; 
        \item $[[Γ ⊢ iN]]$ implying $[[|Γ| ⊢ |iN|]]$.
      \end{itemize}
      To infer the desired
      $[[|Γ| ; |Φ| ⊢ unpack (as, x) = ft; ft' : |iN|]]$,
      we apply \ruleref{\ottdruleSFtyAUnpackLabel}:
      all its premises hold, as noted above.
  \end{caseof}
\end{proof}

\lemmaPolSubst*
\begin{proof}
  We prove it by induction on $[[fT]]$. Each congruent case is proven 
  by congruent rewriting, and applying the induction hypothesis.
  \begin{caseof}
    \item $[[fT = a]]$. Then $[[ +|[fA/a]fT|+ ]] = [[ +|[fA/a]a|+ ]] = [[ +|fA|+ ]]$ and 
    $[[ [+|fA|+ / α⁺] +|fT|+ ]] = [[ [+|fA|+ / α⁺] α⁺ ]] = [[ +|fA|+ ]]$. 
    \item $[[fT = a0]] \neq [[a]]$.
      Then $[[ +|[fA/a]fT|+ ]] = [[ +|[fA/a]a0|+ ]] = [[ +|a0|+ ]] = [[α0⁺]]$ and
      $[[ [+|fA|+ / α⁺] +|fT|+ ]] = [[ [+|fA|+ / α⁺] α0⁺ ]] = [[ +|a0|+ ]]$.
    \item $[[fT = fB1 → fB1]]$. 
      By the induction hypothesis,
      $[[ +|[fA/a]fBi|+ = [+|fA|+ / α⁺] +|fBi|+ ]]$ for $i = 1, 2$.
      Then 
      $[[ +|[fA/a]fT|+ ]] = 
       [[ +|[fA/a](fB1 → fB2)|+ ]] = 
       [[ +|[fA/a]fB1 → [fA/a]fB2|+ ]] = 
       [[ ↓(+|[fA/a]fB1|+ → ↑+|[fA/a]fB2|+) ]] = 
       [[ ↓([+|fA|+ / α⁺] +|fB1|+ → ↑[+|fA|+ / α⁺] +|fB2|+) ]] = 
       [[ [+|fA|+ / α⁺] ↓(+|fB1|+ → ↑+|fB2|+) ]] = 
       [[ [+|fA|+ / α⁺] +|fB1 → fB2|+ ]] = 
       [[ [+|fA|+ / α⁺] +|fT|+ ]]$.
    \item $[[fT = ∀a0.fT0]]$. 
      By the induction hypothesis,
      $[[ +|[fA/a]fT0|+ = [+|fA|+ / α⁺] +|fT0|+ ]]$.
      Then 
      $[[ +|[fA/a]fT|+ ]] = 
       [[ +|[fA/a]∀a0.fT0|+ ]] = 
       [[ +|∀a0.[fA/a]fT0|+ ]] = 
       [[ ↓∀α0⁺.↑+|[fA/a]fT0|+ ]] = 
       [[ ↓∀α0⁺.↑[+|fA|+ / α⁺] +|fT0|+ ]] = 
       [[ [+|fA|+ / α⁺] ↓∀α0⁺.↑+|fT0|+ ]] = 
       [[ [+|fA|+ / α⁺] +|∀a0.fT0|+ ]] = 
       [[ [+|fA|+ / α⁺] +|fT|+ ]]$.
  \end{caseof}
\end{proof}

\lemmaPolRemovable*
\begin{proof}
  We prove it separately in both directions of the implication.

  The erasure of the computation is proved by
   induction on the derivation of $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$.
  Notice that the inference rules of \cref{fig:systemf-typing}
  are obtained by removing the resulting computation from the judgement of the 
  rules of \cref{fig:systemf-polarization}.

  The other direction is proved by straightforward 
  induction on the typing derivation $[[fΓ ; fΦ ⊢ ft : fT]]$.
\end{proof}

\lemmaTypPolWf*
\begin{proof}
  We prove it by induction on $[[fΓ ⊢ fT]]$.
  Let us consider the last rule applied to infer this judgment.
  \begin{caseof}
    \item \ruleref{\ottdruleSFwfVarLabel}
      By inversion, $[[a ∊ fΓ]]$, 
      which implies $[[+|a|+]] = [[α⁺ ∊ +|fΓ|+]]$.
      Then $[[+|fΓ|+ ⊢ +|a|+ ]]$
      by \ruleref{\ottdruleWFTPVarLabel};
    \item \ruleref{\ottdruleSFwfArrowLabel}
      By inversion,  $[[fΓ ⊢ fA]]$ and $[[fΓ ⊢ fB]]$,
      which by the induction hypothesis implies $[[+|fΓ|+ ⊢ +|fA|+]]$ and $[[+|fΓ|+ ⊢ +|fB|+]]$.
      Then $[[+|fΓ|+ ⊢ ↓(+|fA|+ → ↑+|fB|+)]] = [[+|fA → fB|+]]$ by 
      \ruleref{\ottdruleWFTShiftDLabel},
      \ruleref{\ottdruleWFTShiftULabel}, and
      \ruleref{\ottdruleWFTArrowLabel};
    \item \ruleref{\ottdruleSFwfForallLabel}
      By inversion, $[[fΓ, a ⊢ fT]]$,
      which by the induction hypothesis implies $[[+|fΓ|+, α⁺ ⊢ +|fT|+]]$.
      Then $[[+|fΓ|+ ⊢ ∀α⁺.↑+|fT|+]] = [[+|∀a.fT|+]]$ by 
      \ruleref{\ottdruleWFTForallLabel}, and \ruleref{\ottdruleWFTShiftULabel}.
  \end{caseof}
\end{proof}

\lemmaPolarSoundness*
\begin{proof}
  We prove it by induction on $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$.
  Let us consider the last rule applied to infer this judgment.
  \begin{caseof}
    \item \ruleref{\ottdruleELVarLabel},
      in which case
      we wish to prove that if
      $[[ fΓ ; fΦ ⊢ x : fT  ⤳  return x]]$
      then 
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ return x : ↑+|fT|+]]$.

      By inversion of the given judgement, 
      $[[x : fT ∊ fΦ]]$, and thus, 
      $[[x : +|fT|+ ∊ +|fΦ|+]]$.
      We apply \ruleref{\ottdruleDTVarLabel}
      to infer $[[+|fΓ|+ ; +|fΦ|+  ⊢ x : +|fT|+]]$, 
      and then \ruleref{\ottdruleDTReturnLabel}
      to infer $[[+|fΓ|+ ; +|fΦ|+  ⊢ return x : ↑+|fT|+]]$.

    \item \ruleref{\ottdruleELtLamLabel},
      in which case we wish to prove that if
      $[[ fΓ ; fΦ ⊢ λx. ft : fA → fB  ⤳  return {λx : +|fA|+. c}]]$
      then
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ return {λx : +|fA|+. c} : ↑+|fA → fB|+]]$.

      The premise of the given judgment is 
      $[[fΓ; fΦ, x:fA  ⊢ ft : fB  ⤳ c]]$, which by the induction hypothesis
      implies $[[+|fΓ|+ ; +|fΦ, x:fA|+  ⊢ c : ↑+|fB|+]]$,
      which we rewrite as 
      $[[+|fΓ|+ ; +|fΦ|+, x:+|fA|+  ⊢ c : ↑+|fB|+]]$.
      Also, the soundness of \systemf typing (together with \cref{lemma:pol-removable}) 
      imply that the inferred type $[[fA → fB]]$ is well-formed in $[[fΓ]]$,
      which means that $[[+|fA|+]]$ is well-formed in 
      $[[+|fΓ|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
      Then by \ruleref{\ottdruleDTtLamLabel} we infer 
      $[[+|fΓ|+ ; +|fΦ|+  ⊢ λx : +|fA|+. c : +|fA|+ → ↑+|fB|+]]$,
      and then by \ruleref{\ottdruleDTThunkLabel} and \ruleref{\ottdruleDTReturnLabel}, 
      $[[+|fΓ|+ ; +|fΦ|+  ⊢ return {λx : +|fA|+. c} : ↑↓(+|fA|+ → ↑+|fB|+)]]$, 
      which can be rewritten to the required judgment since
      $[[↓(+|fA|+ → ↑+|fB|+) = +|fA → fB|+]]$ by definition.

    \item \ruleref{\ottdruleELTLamLabel}.
      The case is similar to the previous one.
      We wish to prove that if
      $[[ fΓ ; fΦ ⊢ Λa.ft : ∀a.fT  ⤳  return {Λα⁺.c}]]$
      then
      $[[ +|fΓ|+ ; +|fΦ|+ ⊢ return {Λα⁺.c} : ↑+|∀a.fT|+]]$.

      By inversion of the given judgment, we have:
      $[[fΓ, a ; fΦ ⊢ ft : fT  ⤳ c]]$, and by the induction hypothesis,
      $[[+|fΓ|+, α⁺ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$,

      Then we apply \ruleref{\ottdruleDTTLamLabel} to infer
      $[[+|fΓ|+ ; +|fΦ|+  ⊢ Λα⁺.c : ∀α⁺.↑+|fT|+]]$,
      and then 
      \ruleref{\ottdruleDTThunkLabel} with
      \ruleref{\ottdruleDTReturnLabel}
      to infer
      $[[+|fΓ|+ ; +|fΦ|+  ⊢ return {Λα⁺.c} : ↑↓∀α⁺.↑+|fT|+]]$.
      Finally, notice that $[[↓∀α⁺.↑+|fT|+ = +|∀a.fT|+]]$ by definition.

    \item \ruleref{\ottdruleELAppLabel}.
      By inversion of the given judgment, we have:
      \begin{itemize}
        \item $[[fΓ ; fΦ ⊢ ft : fA → fB  ⤳  c]]$,
          and by the induction hypothesis,
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fA → fB|+]]$,
          which is equivalent to
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑↓(+|fA|+ → ↑+|fB|+)]]$.
        \item $[[fΓ ; fΦ ⊢ ft' : fA  ⤳  c']]$, 
          and by the induction hypothesis,
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c' : ↑+|fA|+]]$.
      \end{itemize}
      To infer 
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|fA → fB|+ = c; let x : +|fA|+ = c'; let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$, 
      we gradually apply the corresponding rules of \fexists to construct the required typing inference tree:
      \begin{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+, y:+|fB|+ ⊢ return y : ↑+|fB|+]]$ by 
          \ruleref{\ottdruleDTVarLabel} and \ruleref{\ottdruleDTReturnLabel}.  
        \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+ ⊢ let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$ by 
          \ruleref{\ottdruleDTAppLetAnnLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ +|fB|+]]$ by \cref{lemma:typ-pol-wf-completeness};
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+ ⊢ f : ↓(+|fA|+ → ↑+|fB|+)]]$ by 
              \ruleref{\ottdruleDTVarLabel};
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+ ⊢ (+|fA|+ → ↑+|fB|+) ● x ⇒> ↑+|fB|+]]$
              by \ruleref{\ottdruleDTArrowAppLabel};
            \item $[[+|fΓ|+ ⊢ ↑+|fB|+ ≤ ↑+|fB|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity});
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+, y:+|fB|+ ⊢ return y : ↑+|fB|+]]$ 
            as noted above;
          \end{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+) ⊢ let x:+|fA|+ = c'; let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$ by 
          \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness};
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+) ⊢ c' : ↑+|fA|+]]$
              holds by weakening of $[[+|fΓ|+ ; +|fΦ|+ ⊢ c' : ↑+|fA|+]]$, which in turn holds by the induction hypothesis,
              as noted above;
            \item $[[+|fΓ|+  ⊢ ↑+|fA|+ ≤ ↑+|fA|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity});
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+ ⊢ let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$ as noted above;
          \end{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|fA → fB|+ = c; let x : +|fA|+ = c'; let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$ by 
          \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ +|fA → fB|+]]$ by \cref{lemma:typ-pol-wf-completeness};
            \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fA → fB|+]]$ by the induction hypothesis, as noted above;
            \item $[[+|fΓ|+  ⊢ ↑+|fA → fB|+ ≤ ↑+|fA → fB|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity});
            \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+) ⊢ let x:+|fA|+ = c'; let y:+|fB|+ = f(x); return y : ↑+|fB|+]]$ 
              as noted above (we rewrote $[[+|fA → fB|+]]$ as $[[↓(+|fA|+ → ↑+|fB|+)]]$).
          \end{enumerate}
      \end{enumerate}

    \item \ruleref{\ottdruleELTAppLabel}
      By inversion of this rule, we have
      \begin{itemize}
        \item $[[fΓ ; fΦ ⊢ ft : ∀a.fT  ⤳  c]]$, and thus, 
          $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|∀a.fT|+]]$ by the induction hypothesis; and
        \item $[[fΓ ⊢ fA]]$, which implies $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
      \end{itemize}
      We wish to infer
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|∀a.fT|+ = c; let y : +|[fA/a]fT|+ = f(); return y : ↑+|[fA/a]fT|+]]$.
      First, let us rewrite the polarized types by definition and \cref{lemma:pol-subst}:
      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$.
      Then we prove this judgment gradually by applying the corresponding rules of \fexists:
      \begin{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTVarLabel} and \ruleref{\ottdruleDTReturnLabel}.
        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTAppLetAnnLabel}. Let us show that the 
          required premises hold. 
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ [+|fA|+/α⁺]+|fT|+]]$ holds since $[[+|fΓ|+ ⊢ +|fA|+]]$ and $[[+|fΓ|+ ⊢ +|fT|+]]$ by \cref{lemma:wf-subst}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ f : ↓∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTVarLabel}.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ ∀α⁺.↑+|fT|+ ● · ⇒> ∀α⁺.↑+|fT|+]]$ by \ruleref{\ottdruleDTEmptyAppLabel}.
            \item $[[+|fΓ|+  ⊢ ∀α⁺.↑+|fT|+ ≤ ↑[+|fA|+/α⁺]+|fT|+ ]]$ by \ruleref{\ottdruleDOneForallLabel}:
              $[[+|fΓ|+  ⊢ +|fA|+/α⁺ : {α⁺}]]$ since $[[+|fΓ|+  ⊢ +|fA|+]]$, as noted above.
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+, y : [+|fA|+/α⁺]+|fT|+ ⊢ return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
        \item $[[+|fΓ|+ ; +|fΦ|+  ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
          holds by \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the
          required premises hold.
          \begin{enumerate}
            \item $[[+|fΓ|+ ⊢ ↓∀α⁺.↑+|fT|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
            \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑↓∀α⁺.↑+|fT|+]]$ by the induction hypothesis, as noted above.
            \item $[[+|fΓ|+  ⊢ ↑↓∀α⁺.↑+|fT|+ ≤ ↑↓∀α⁺.↑+|fT|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity}).
            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
          \end{enumerate}
      \end{enumerate}
  \end{caseof}
\end{proof}

### 2024-04-04 oopsla24/tex/02-overview.tex
WIP: algorithm overview and the relation to SystemF

@@
 
-The standard \systemf types and terms must be polarized to be used in \fexists, 
-we present the embedding formally in \cref{fig:polarization}.
-For instance, the following types and terms of \systemf 
-can be polarized in the given way:
+In general, any term and type of \systemf can be embedded into \fexists.
+This embedding---polarization---will be covered in more details in \cref{sec:rel-to-systemf}.
+Now, we discuss a number of examples of standard \systemf terms and types
+and their polarization in \fexists.
 
@@
 \subsection{The key ideas of the algorithm}
+
+The inference algorithm for \fexists is \emph{local}: it has a limited scope of
+the inference information and does not propagate it between far-apart branches
+of the syntax tree. Nevertheless, many difficulties appear in the inference of
+polymorphic and existential types. Next, we discuss the most challenging of
+these difficulties and how they are addressed in the algorithm.
+
+\paragraph{Subtyping} One of the hardest parts of the inference is subtyping,
+because this is where the polymorphic instantiation happens. In particular, it
+is the subtyping relation that allows us to use the polymorphic terms in the
+context where their concrete instantiations are expected: 
+$[[Γ ⊢ ∀α⁺.↑α⁺ ≤ ↑Int]]$.
+
+We are able to concentrate most of the difficulty in the subtyping relation 
+and then use the polarization to make it decidable.
+
+\paragraph{Impredicativity}
+Impredicativity is another difficulty.
+Even with forall, the natural subtyping is undecidable.
+We have to find the restrictions that make it decidable.
+Unification allows us to solve equality constraints, and the LUB has to be solved with anti-unification.
+We have to design the system carefully to maintain decidability.
+
+\paragraph{Least Upper Bound}
+Now the LUB (Least Upper Bound) is not necessarily the syntactic equality.
+Examples, types don't have a LUB.
+How do we decide?
+The answer is anti-unification, which we explain at a high level.
+Now we have both forall and exists, and mixing unification and anti-unification.

### 2024-04-04 oopsla24/tex/03-declarative-system.tex
wip
WIP: algorithm overview and the relation to SystemF

@@
 
-\ilyam{Other properties?}
+\subsection{Relation to \systemf}
+\label{sec:rel-to-systemf}
+
+To establish a correspondence between \fexists and standard unpolarized \systemf,
+we present a translation in both ways: the polarization from \systemf to \fexists
+and the depolarization from \fexists to \systemf.
+
+\paragraph{Type-level Translation}
+
+The type depolarization (\cref{fig:depolarization}) simply forgets the polarization structure of types:
+it removes the shift operators and the polarities of the free type variables.
+
+The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
+ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
+The choice of polarization affects the execution strategy of the program.
+We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
+This choice makes it smoother to lift the translation to the term level, which will be disused next.
+
+\begin{figure}[ht]
+  \begin{minipage}[t]{0.65\linewidth}
+    \centering
+      \begin{minipage}[t]{0.32\linewidth}
+      \ottfundefnunpolP{}
+      \end{minipage}
+      \begin{minipage}[t]{0.32\linewidth}
+      \ottfundefnunpolN{}
+      \end{minipage}
+    \caption{Type Depolarization}
+    \label{fig:depolarization}
+  \end{minipage}
+  \hspace{0.5cm}
+  \begin{minipage}[t]{0.3\linewidth}
+    \centering
+    \ottfundefnpolarP{}
+    \caption{Type Polarization}
+    \label{fig:polarization}
+  \end{minipage}
+\end{figure}
+
+\paragraph{Term-level Translation}
+
+The term-level translations between systems are more complex.
+They are defined as \emph{elaborations} of the typing derivations.
+In other words, the input of the translation is not a term, 
+but a whole typing derivation of this term in the corresponding system.
+For brevity, we omit the formal definition of the term-level translation,
+referring the reader to the appendix (\cref{sec:term-translation}).
+
+We prove the soundness of the translation (\cref{lemma:translation-soundness}),
+which in particular guarantees that the translation preserves typing.
+
+\begin{property}[Soundness of the Translation]
+    \label{lemma:translation-soundness}
+    \begin{align*}
+      & [[Γ ⊢ iN ≤ iM ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iN| → |iM| ]], \\
+      & [[Γ ⊢ iP ≥ iQ ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iQ| → |iP| ]], \\
+      & [[Γ; Φ ⊢ v : iP ⤳ ft]]  \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iP|]], \\
+      & [[Γ; Φ ⊢ c : iN ⤳ ft]] \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iN|]], \\
+      & [[fΓ ; fΦ ⊢ ft : fT ⤳ c]]  \text{ implies }  [[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]].
+    \end{align*}
+\end{property}
+
+% \begin{figure}[h]
+%   \begin{equation*}
+%   \begin{aligned}
+%     [[|↓iN|]] &= [[|iN|]] \\
+%     [[|↑iP|]] &= [[|iP|]] \\
+%   \end{aligned}
+%   \qquad
+%   \begin{aligned}
+%     [[|α±|]] &= [[a]] \\
+%     [[|iP → iN|]] &= [[|iP| → |iN|]] \\
+%   \end{aligned}
+%   \qquad
+%   \begin{aligned}
+%     [[|∀α⁺.iN|]] &= [[∀α.|iN|]] \\
+%     [[|∃α⁻.iP|]] &= [[∃α.|iP|]] \\
+%   \end{aligned}
+%   \end{equation*}
+%   \caption{Type depolarization}
+%   \label{fig:depolarization}
+% \end{figure}
+
+% The translation from \systemf to \fexists can be done in two canonical ways
+% depending on the desired polarity. The polarity will also determine the 
+% execution strategy \ilyam{how?}.
+% Moreover, the translation procedure depends on the \emph{polarity} of the
+% free type variables. Thus, the translation functions 
+% $[[-| fT |-^Γ]]$ and $[[+| fT |+^Γ]]$ are indexed by a context $[[Γ]]$, 
+% containing either $[[α⁺]]$ or $[[α⁻]]$ for each variable $[[α]]$ free in $[[fT]]$.
+
+% \begin{figure}[h]
+%   \begin{equation*}
+%   \begin{aligned}
+%     [[-| a |-^Γ]] &= [[α⁻]]  \text{,~if $[[α⁻ ∊ Γ]]$}\\
+%     [[-| a |-^Γ]] &= [[↑α⁺]] \text{,~if $[[α⁺ ∊ Γ]]$}\\
+%     [[-| fA → fB |-^Γ]] &= [[ +|fA|+^Γ → -|fB|-^Γ  ]] \\
+%     [[-| ∀α.fT |-^Γ]] &= [[ ∀α⁺.-|fT|-^Γ,α⁺ ]] \\
+%     [[-| ∃α.fT |-^Γ]] &= [[ ↑ ∃α⁻.+|fT|+^Γ,α⁻ ]] \\
+%   \end{aligned}
+%   \qquad
+%   \begin{aligned}
+%     [[+| a |+^Γ]] &= [[α⁺]]  \text{,~if $[[α⁺ ∊ Γ]]$}\\
+%     [[+| a |+^Γ]] &= [[↓α⁻]] \text{,~if $[[α⁻ ∊ Γ]]$}\\
+%     [[+| fA → fB |+^Γ]] &= [[ ↓ +|fA|+^Γ → -|fB|-^Γ  ]] \\
+%     [[+| ∀α.fT |+^Γ]] &= [[ ↓∀α⁺.-|fT|-^Γ,α⁺ ]] \\
+%     [[+| ∃α.fT |+^Γ]] &= [[ ∃α⁻.+|fT|+^Γ,α⁻ ]] \\
+%   \end{aligned}
+%   \end{equation*}
+%   \caption{Type polarization}
+%   \label{fig:polarization}
+% \end{figure}

### 2024-04-04 oopsla24/tex/04-algorithm.tex
WIP: algorithm overview and the relation to SystemF

@@
   to the application inference procedure,
-  returning an algorithmic type $[[iM']]$, 
-  that must be a subtype of the annotation $[[↑uP]]$.
-
-  Then premise $[[Γ; Θ ⊨ iM' ≤ ↑iP ⫤ SC2]]$
+  returning an algorithmic type $[[uM']]$, 
+  that must be a subtype of the annotation $[[↑iP]]$.
+
+  Then premise $[[Γ; Θ ⊨ uM' ≤ ↑iP ⫤ SC2]]$
   together with $[[Θ ⊢ SC1 & SC2 = SC]]$
-  check that $[[iM']]$ can be instantiated to the annotated type $[[↑iP]]$,
+  check that $[[uM']]$ can be instantiated to the annotated type $[[↑iP]]$,
   and if it is, the algorithm infers the type of the body in the extended context,
@@
   To find it, it makes a call to the minimal instantiation algorithm 
-  (\cref{sec:minimal-instantiation})
+  (\cref{sec:constraint-singularity})
   that finds the substitution that satisfies the inferred constraints $[[SC]]$ and

### 2024-04-12 oopsla24/tex/02-overview.tex
WIP: examples, overview

--- 2024-04-03 oopsla24/tex/02-overview.tex
@@
 
-The standard \systemf types and terms must be polarized to be used in \fexists, 
-we present the embedding formally in \cref{fig:polarization}.
-For instance, the following types and terms of \systemf 
-can be polarized in the given way:
+In general, any term and type of \systemf can be embedded into \fexists in
+several ways, because at each moment one can choose the positive or negative
+polarity of the subterm. The consistent term-level and type-level embeddings
+(that preserve typing) will be covered in detail in
+\cref{sec:rel-to-systemf}. Now, we discuss several examples of standard
+\systemf terms and types and their polarization in \fexists.
 
@@
   \begin{align*}
-    [[map]] &: [[↓∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺]] \\
-    [[len]] &: [[↓∀α⁺.List α⁺ → ↑Int]] \\
-    [[choose]] &: [[↓∀α⁺.α⁺ → α⁺ → ↑α⁺]] \\
-    [[id]] &: [[↓∀α⁺.α⁺ → ↑α⁺]] \\
-    [[auto]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
-    [[autoP]] &: [[↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺)]]\\
+    [[map]] &: [[↓(∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺)]] \\
+    [[len]] &: [[↓(∀α⁺.List α⁺ → ↑Int)]] \\
+    [[choose]] &: [[↓(∀α⁺.α⁺ → α⁺ → ↑α⁺)]] \\
+    [[id]] &: [[↓(∀α⁺.α⁺ → ↑α⁺)]] \\
+    [[auto]] &: [[↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]
   \end{align*}
@@
 Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
-Then the following types are inferrable in \fexists:
+Then the following types are inferrable in \fexists.
 
-\begin{align*}
-  % &[[· ; Φ ⊢ [two, three, nine] : List Int]]\\
-  &[[· ; Φ, x1:↓(↓iN → iN), x2:↓(↓iM → iM) ⊢ let x = choose (x1, x2); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]\\
-  &[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]\\
-  &[[· ; Φ ⊢ let x = choose (id, autoP); return x : ↑ ∃γ⁻.↓γ⁻]]\\
-\end{align*}
+\paragraph{Application of a Polymorphic Function}
+  A polymorphic function can be applied to a value of a concrete type.
+  In this case, the quantified variables are instantiated to make the application type-check.
+  For instance, the polymorphic $[[id]]$ can be applied to any positive type, 
+  and the resulting type will be inferred by the algorithm:
+  $$[[· ; Φ ⊢ let x = id(fortyTwo); return x : ↑Int]]$$
 
+\paragraph{Polymorphic Self-application}
+  Unlike predicative systems, \fexists
+  permits self-application of polymorphic functions.
+  For instance, $[[id]]$ can be applied to itself
+  resulting in a polymorphic type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$:
+  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$
+
+\paragraph{Non-example: Polymorphic Arguments}
 The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
@@
 In particular, if an argument of a function is polymorphic,
-its polymorphic variables are not instantiated by the inference algorithm. 
+its polymorphic variables are \emph{not} instantiated by the inference algorithm. 
 For example, one would expect the following to be inferrable:
-$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$.
-However, to infer this, the $[[id]]$ must be instantiated to $[[Int → ↑Int]]$, 
+\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
+However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
 which requires the information to be propagated from the neighboring branch 
-of the syntax tree ($[[ [two, three, nine] ]]$), \ie bypass the locality.
+of the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.
 
-However, the polymorphic arguments of the function are instantiated, 
-and thus, the annotated version of the same term infers the type successfully:
-$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$.
+The annotated version of the same term infers the type successfully, 
+since all the polymorphic variables of the arguments are instantiated:
+$$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$
+
+\paragraph{Inferring the Common Supertype}
+  The function $[[choose]]$ takes two arguments of the same
+  polymorphic type. To apply it to two values of different (concrete) types, 
+  the inference algorithm must find the minimal type they can be coerced 
+  to---the least common supertype.
+
+  Let us assume that in $[[Φ]]$, we have two identity-like functions 
+  \mbox{$[[id_iM]] : [[↓(↓iM → iM)]]$} and \mbox{$[[id_iN]] : [[↓(↓iN → iN)]]$}.
+  Then their least common supertype---the existential $[[∃γ⁻.↓(↓γ⁻ → γ⁻)]]$
+  is inferrable:
+  $$[[· ; Φ ⊢ let x = choose (id_iN, id_iM); return x : ↑ ∃γ⁻.↓(↓γ⁻ → γ⁻)]]$$
+
+\paragraph{Inferring a general type}
+  As mentioned before, the local inference does not instantiate the polymorphic
+  types in the argument position.
+  Because of that, the application 
+  \mbox{$[[let x = choose (id, auto); return x ]]$} cannot infer 
+  $[[↑↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]$.
+  However, the inference succeeds and infers a less informative type:
+  $$[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]$$
 
@@
 \begin{figure}[h]
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottiN\ottinterrule
-    }\\
-    \ottgrammartabular{
-      \ottiP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
+  \begin{minipage}[t]{0.6\textwidth}
+    \begin{supertabular}{ll p{0.8\textwidth}}
+      \ottiNInline\\
+    \end{supertabular}
+  \end{minipage}
+  \begin{minipage}[t]{0.35\textwidth}
+    \begin{supertabular}{ll p{0.8\textwidth}}
+      \ottiPInline\\
+    \end{supertabular}
+  \end{minipage}
+
   \caption{Declarative Types of \fexists}
@@
 \begin{figure}[h]
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottc\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottv\ottinterrule
-    }
-  \end{multicols}
+  \begin{supertabular}{ll p{0.8\textwidth}}
+    \ottcInline\\
+    \ottvInline\\
+  \end{supertabular}
   \caption{Declarative Terms of \fexists}
@@
 \subsection{The key ideas of the algorithm}
+
+The inference algorithm for \fexists is \emph{local}: it has a limited scope of
+the inference information and does not propagate it between far-apart branches
+of the syntax tree. Nevertheless, many difficulties appear in the inference of
+polymorphic and existential types. Next, we discuss the most challenging of
+these difficulties and how they are addressed in the algorithm.
+
+\paragraph{Subtyping} Most of the difficulty of the inference algorithm is
+concentrated in the subtyping relation. This is the part of the system where
+most of the polymorphic instantiation happens. In particular, it is the
+subtyping relation that allows us to use the polymorphic terms in the context
+where their concrete instantiations are expected.
+
+To check whether one type is a subtype of another, the algorithm introduces
+\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)--- the
+`placeholders' representing the polymorphic variables whose instantiation is
+postponed. This way, the problem of $[[∀α⁺.↑α⁺ ≤ ↑Int]]$ becomes the problem
+of finding $[[α̂⁺]]$ such that $[[↑α̂⁺ ≤ ↑Int]]$.
+
+\newcommand{\defiff}{%
+  \mathrel{\vbox{\offinterlineskip\ialign{%
+    \hfil##\hfil\cr
+    $\scriptscriptstyle\text{def}$\cr
+    %\noalign{\kern0ex}
+    $\iff$\cr
+}}}}
+
+\paragraph{Equivalence}
+Our system introduces another complexity: the equivalence induced by subtyping
+($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
+straightforward. This relation extends beyond just alpha-equivalent types. As
+the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
+the equivalence relation must permit the permutations of the quantifiers. For
+example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
+equivalence permits the removal/addition of the unused quantifiers: $[[· ⊢
+∀α⁺.↑Int ≈ ↑Int]]$.
+
+In order to manage the complexity of non-trivial equivalence, we implement
+a type normalization algorithm. It effectively reduces the mutual subtyping to
+a simpler alpha-equivalence. The normalization works by recursively
+eliminating unused quantifiers and reordering the remaining ones into a
+canonical permutation.
+
+\paragraph{Least Upper Bound}
+Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
+constraints that need to be resolved. At this moment, another problem arises:
+as one variable can occur in multiple constraints, the resolution becomes non-trivial.
+In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
+we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.
+
+In many type systems, the least upper bound boils down to the trivial syntactic equality.
+Because of that, the set of constraints is typically resolved using \emph{unification}.
+However, this problem is trickier in the presence of existential types. For example, 
+in \fexists, the least upper bound of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[∃α⁻.↓α⁻]]$,
+because the quantified $[[α⁻]]$ can be instantiated to either $[[↑Int]]$ or $[[↑Bool]]$.
+
+To find this least upper bound, we use \emph{anti-unification}---the dual of
+unification. While unification finds \emph{the most general instance} of two
+given patterns, the anti-unification finds \emph{the most detailed} (under the
+restrictions of the system) \emph{pattern} that matches two given types. In the
+example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
+and in \fexists, this unifier is the most specific (in particular because
+$[[↓↑α̂⁺]]$ is not considered as only \emph{negative} variables are
+existentially quantified).
+
+\paragraph{Impredicativity}
+Another difficulty that hinders the constraint resolution arises from the
+impredicativity of the system---the ability to quantify over the types that are 
+themselves polymorphic. In the impredicative polymorphic system (even
+with only $[[∀]]$-quantifiers), the naturally formulated subtyping is
+undecidable, which follows from the undecidability of second-order unification.
+All the more the undecidability is evident in the presence of existential types
+and subtyping constraints. For instance, in \fexists the constraint 
+$[[α̂⁺ :≥ ↓↑α̂⁺]]$ would have a surprising impredicative solution $[[α̂⁺]] = [[∃α⁻.↓α⁻]]$.
+
+To maintain the decidability in the presence of impredicativity, we carefully
+restrict the system using the polarity stratification. One of the main
+constraints that we put on subtyping is the invariance of the shift operators:
+the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
+\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
+still have non-trivial upper bounds, and thus, need to use anti-unification to
+find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and in
+fact, makes the constraint resolution decidable.

### 2024-04-12 oopsla24/tex/03-declarative-system.tex
WIP: examples, overview

@@
 \paragraph{Invariant Shifts}
-An important restriction that we put on the subtyping system is
-that the subtyping on shifted types requires their equivalence,
-as shown in \ruleref{\ottdruleDOneShiftDLabel} and
-\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
-invariants make the system equivalent to \systemf, 
-and thus, undecidable. 
-However, after certain changes 
-\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form,
-thereby increasing the expressiveness of the system. These
-changes are discussed in \cref{sec:weakening-invariant}.
+As mentioned above, an important restriction that we put on the subtyping system
+is that the subtyping on shifted types requires their equivalence, as shown in
+\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}.
+Relaxing both of these invariants make the system equivalent to \systemf, and
+thus, undecidable. However, after certain changes
+\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form, thereby
+increasing the expressiveness of the system. These changes are discussed in
+\cref{sec:weakening-invariant}.
 

### 2024-04-12 oopsla24/tex/04-algorithm.tex
WIP: examples, overview

@@
   Constraint Type Context\\
-   $[[Θ]] ::= \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
+   $[[Θ]] \Coloneqq \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
   \caption{Algorithmic Syntax}
@@
       \ottruleheadOneLine
-        {[[SC]]}{::=} {\ottcom{Constraint Set}}
+        {[[SC]]}{\Coloneqq} {\ottcom{Constraint Set}}
         {\{[[scE1]], \dots, [[scEn]]\}}\ottprodnewline
@@
       \ottruleheadOneLine
-        {[[UC]]}{::=} {\ottcom{Unification Constraint Set}}
+        {[[UC]]}{\Coloneqq} {\ottcom{Unification Constraint Set}}
         {\{[[ucE1]], \dots, [[ucEn]]\}}\ottprodnewline

### 2024-04-15 oopsla24/tex/01-introduction.tex
WIP: examples, overview
Edit Introduction

@@
-% How we should frame this:
+Over the last half-century, there has been considerable work on developing type
+inference algorithms for programming languages,  mostly aimed at solving the
+problem of \emph{type polymorphism}.
 
-Over the last half-century, there has been considerable work on developing type inference algorithms for programming languages,  mostly aimed at solving the problem of \emph{type polymorphism}.
+That is, in pure polymorphic lambda
+calculus---\systemf\cite{girard-system-f,jcr-system-f}---the polymorphic type
+$[[∀a.fA]]$ has a big lambda $[[Λa.fe]]$ as an introduction form, and an
+explicit type application $[[ fe[fA] ]]$ as an elimination form. This is an
+extremely simple and compact type system, whose rules fit on a single page, but
+whose semantics are sophisticated enough to model things like parametricity and
+representation independence. However, System F by itself is unwieldy as a
+programming language. The fact that the universal type $[[∀a.fA]]$ has explicit
+eliminations means that programs written using polymorphic types will need to be
+stuffed to the gills with type annotations explaining how and when to
+instantiate the quantifiers.
 
-That is, in pure polymorphic lambda calculus (system F), the polymorphic type $\forall a.\,A$ has a big lambda $\Lambda a.\,e$ as an introduction form, and an explicit type application $e\,[A]$ as an elimination form.
-This is an extremely simple and compact type system, whose rules fit on a single page, but whose semantics are sophisticated enough to model things like parametricity and representation independence.
-However, System F by itself is unwieldy as a programming language.
-The fact that the universal type $\forall a.\,A$ has explicit eliminations means that programs written using polymorphic types will need to be stuffed to the gills with type annotations explaining how and when to instantiate the quantifiers.
+Therefore, most work on type inference has been aimed at handling type
+instantiations implicitly---we want to be able to use a polymorphic function
+like $[[len : ∀a.List a → Int]]$ at any concrete list type without explicitly
+instantiating the quantifier in $[[len]]$'s type. That is, we want to write
+$[[ len [one,two,three] ]]$ instead of writing
+$[[ len [Int] [one,two,three] ]]$.
 
-Therefore, most work on type inference has been aimed at handling type instantiations implicitly -- we want to be able to use a polymorphic function like $\mathsf{len} : \forall a.\, \mathsf{List}\,a \to \mathsf{int}$ at any concrete list type without explicitly instantiating the quantifier in $\mathsf{len}$'s type.
-That is, we want to write $\mathsf{len}\,[1,2,3]$ instead of writing $\mathsf{len}\,[\mathsf{int}]\,[1,2,3]$.
+The most famous of the algorithms for solving these constraints is the
+Damas-Hindley-Milner (HDM) algorithm
+\cite{hindley69:principal,milner78:theory,damas82:principal}. The idea is that
+type instantiation induces a subtype ordering: the type $[[∀a.fA]]$ is a subtype
+of all of its instantiations. So we wish to be free to use the same function
+$[[len]]$ at many different types such as 
+$[[List Int → Int]]$, $[[List Bool → Int]]$,
+$[[List (Int × Bool) → Int]]$, and so on.
+However, the subtype relation is nondeterministic: it tells us that whenever we
+see a polymorphic type $[[∀a.fA]]$, we know it is a subtype of \emph{any}
+of its instantiations. To turn this into an algorithm, we have to actually make
+some choices, and DHM works by using \emph{unification}. Whenever we would have
+had to introduce a particular concrete type in the specification, the DHM
+algorithm introduces a unification variable, and incrementally instantiates this
+variable as more and more type constraints are observed.
 
-The most famous of the algorithms for solving these constraints is the Damas-Hindley-Milner algorithm.
-The idea is that type instantiation induces a subtype ordering: the type $\forall a.\,A$ is a subtype of all of its instantiations.
-So we wish to be free to use the same function $\mathsf{len}$ at many different types such as
-$\mathsf{List}\,\mathsf{int} \to \mathsf{int}$,
-$\mathsf{List}\,\mathsf{bool} \to \mathsf{int}$,
-$\mathsf{List}\,(\mathsf{int \times bool}) \to \mathsf{int}$, and so on.
-However, the subtype relation is nondeterministic: it tells us that whenever we see a polymorphic type $\forall a.\, A$, we know it is a subtype of \emph{any} of its instantiations.
-To turn this into an algorithm, we have to actually make some choices, and DHM works by using \emph{unification}.
-Whenever we would have had to introduce a particular concrete type in the specification, the DHM algorithm introduces a unification variable, and incrementally instantiates this variable as more and more type constraints are observed.
+However, the universal quantifier is not the only quantifier! Dual to the
+universal quantifier $\forall$ is the existential quantifier $\exists$. Even
+though existential types have an equal logical status to universal types, they
+have been studied much less frequently in the literature. The most widely-used
+algorithm for handling existentials is the algorithm of Odersky and Laufer
+\cite{laufer94:polymorphic}. This algorithm treats existentials in a
+second-class fashion: they are not explicit connectives of the logic but rather
+are tied to datatype declarations. As a result, packing and unpacking
+existential types is tied to constructor introductions and constructor
+eliminations in pattern matches. This allows Damas-Milner inference to continue
+to work almost unchanged, but it does come at the cost of losing first-class
+existentials and also of restricting the witness types to monomorphic types.
 
-However, the universal quantifier is not the only quantifier!
-Dual to the universal quantifer $\forall$ is the existential quantifier $\exists$.
-Even though existential types have an equal logical status to universal types, they have been studied much less frequently in the literature.
-The most widely-used algorithm for handling existentials is the algorithm of Odersky and Laufer.
-This algorithm treats existentials in a second-class fashion: they are not explicit connectives of the logic, but rather are tied to datatype declarations.
-As a result, packing and unpacking existential types is tied to constructor introductions and constructor eliminations in pattern matches.
-This allows Damas-Milner inference to continue to work almost unchanged, but it does come at the cost of losing first-class existentials and also of restricting the witness types to monomorphic types.
+There has been a limited amount of work on support for existential types as a
+first-class concept. In an unpublished draft, \citet{leijen06:first-class}
+studied an extension of Damas-Milner inference in which type schemes contain
+alternating universal and existential quantifiers. Quantifiers still range over
+monotypes, and higher-rank polymorphism is not permitted. More recently,
+\citet{dunfield16:existential} studied type inference for existential types in
+the context of GADT type inference, which, while still predicative, supported
+higher-rank types (i.e., quantifiers can occur anywhere inside of a type
+scheme). \citet{eisenberg21:existential} propose a system which only permits
+types of the form forall-exists, but which permits projective elimination in the
+style of ML modules.
 
-There has been a limited amount of work on support for existential types as a first-class concept.
-In an unpublished draft, \citet{leijen06} studied an extension of Damas-Milner inference in which type schemes contain alternating universal and existential quantifiers.
-Quantifiers still range over monotypes, and higher-rank polymorphism is not permitted.
-More recently, \citet{dk19} studed type inference for existential types in the context of GADT type inference, which, while still predicative, supported higher-rank types (i.e., quantifiers can occur anywhere inside of a type scheme).
-\citet{existential-crisis} propose a system which only permits types of the form forall-exists, but which permits projective elimination in the style of ML modules.
+All of these papers are restricted to \emph{predicative} quantification, where
+quantifiers can only be instantiated with monotypes (i.e., types without any
+occurences of quantifiers). However, existential types in full System F are
+\emph{impredicative}---that is, quantifiers can be instantiated with arbitrary
+types, specifically including types containing quantifiers.
 
-All of these papers are restricted to \emph{predicative} quantification, where quantifiers can only be instantiated with monotypes (i.e., types without any occurences of quantifiers).
-However, existential types in full System F are \emph{impredicative} -- that is, quantifiers can be instantiated with arbitrary types, specifically including types containing quantifiers.
+Historically, inference for impredicative quantification has been neglected, due
+to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that
+not only is full type inference for \systemf undecidable, but even that the
+subtyping relation induced by type instantiation is undecidable. However, in
+recent years, interest in impredicative inference has revived (for example, the
+work of \citet{serrano-2020}), with a focus on avoiding the undecidability
+barriers by doing \emph{partial} type inference. That is, we no longer try to do
+full type inference, but rather accept that the programmer will need to write
+annotations in cases where inference would be too difficult. Unfortunately, it
+is often difficult to give a specification for what the partial algorithm
+does--for example, \citet{eisenberg21:existential} observe that their algorithm
+lacks a declarative specification, and explain why existential types make this
+particularly difficult to do.
 
-Historically, inference for impredicative quantification has been neglected, due to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that not only is full type inference for System F undecidable, but even that the subtyping relation induced by type instantiation is undecidable.
-However, in recent years, interest in impredicative inference has revived (for example, the work of \citet{serrano-2020}), with a focus on avoiding the undecidabilty barriers by doing \emph{partial} type inference.
-That is, we no longer try to do full type inference, but rather accept that the programmer will need to write annotations in cases where inference would be too difficult.
-Unfortunately, it is often difficult to give a specification for what the partial algorithm does -- for example, \citet{existential-crisis} observe that their algorithm lacks a declarative specification, and explain why existential types make this particularly difficult to do.
+One especially well-behaved form of partial type inference is \emph{local type
+inference}, introduced by \citet{pierce2000:local}. In this approach,
+quantifiers in a polymorphic function are instantiated using only the type
+information available in the arguments at each call site. While this infers
+fewer quantifiers than global algorithms such as Damas-Milner can, it has the
+twin benefits that it is easy to give a mathematical specification to, and that
+failures of type inference are easily explained in terms of local features of
+the call site. In fact, many production programming languages such as C\# and
+Java use a form of local type inference, due to the ease of extending this style
+of inference to support OO features.
 
-One especially well-behaved form of partial type inference is \emph{local type inference}, introduced by \citet{pierce-local-2000}.
-In this approach, quantifiers in a polymorphic function are instantiated using only the type information available in the arguments at each call site.
-While this infers fewer quantifiers than global algorithms such as Damas-Milner can, it has the twin benefits that it is easy to give a mathematical specification to, and that failures of type inference are easily explained in terms of local features of the call site.
+In this paper, we extend the local type inference algorithm to a language with
+both universal and existential quantifiers, which can both be instantiated
+impredicatively. This combination of features broke a number of the invariants
+which traditional type inference algorithms depend on, and required us to invent
+new algorithms which combine both unification and (surprisingly)
+anti-unification.
 
-In this paper, we extend the local type inference algorithm to a language with both universal and existential quantifiers, which can both be instantiated impredicatively.
-This combination of features broke a number of the invariants which traditional type inference algorithms depend on, and required us to invent new algorithms which combine both unification and (surprisingly) anti-unification.
+To research the type inference in impredicative polymorphic systems
+in the presence of existential types, we make the following \emph{contributions}.
 
-\paragraph{Contributions} Our contributions are as follows: 
-\begin{itemize}
-\item We give a declarative type system (again, based on call-by-push-value) to serve as a specification of our algorithm, and we we prove our algorithm is sound and complete with respect to the declarative type system.
-The specification makes it easy to see that all type applications (for $\forall$-elimination) and all packs (for $\exists$-introdution) are inferred.
+\paragraph{Type Inference Algorithm} We give a local type inference algorithm
+    which supports both first-class existential and universal quantifiers, both
+    of which can be instantiated impredicatively. To evade the undecidability
+    results surrounding type inference for \systemf, we work in a variant of
+    \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping relation
+    which is still decidable.
 
-\item We give a local type inference algorithm which supports both first-class existential and universal quantifiers, both of which can be instantiated impredicatively.
-To evade the undecidability results surrounding type inference for System F, we work in a variant of call-by-push-value, which lets us formulate a subtyping relation which is still decidable.
+\paragraph{Declarative Specification}
+    We give a declarative type system (again, based on \CBPV) to
+    serve as a specification of our algorithm, and we prove our algorithm is
+    sound and complete with respect to the declarative type system. The
+    specification makes it easy to see that all type applications (for
+    $\forall$-elimination) and all packs (for $\exists$-introdution) are
+    inferred.
 
-\item Our algorithm breaks some of the fundamental invariants of
-  HM-style type inference. As a result, it needs to mix unification
-  and anti-unification, uses the call-by-push-value structure to
-  control function arities and how quantifiers can be instantiated.
+\paragraph{Employment of the Anti-unification}
+    The introduction of impredicative existentials breaks some of the
+    fundamental invariants of HM-style type inference. As a result, it needs to
+    mix unification with \emph{anti-unification}---a dual operation which has
+    been used in the literature on term rewriting systems, but which has never
+    appeared in the context of polymorphic type inference.
 
-\item The original local type inference paper combined local type inference
-  with bidirectional typechecking to minimize the number of needed
-  annotations, but we show how existential types complicate the integration
-  of bidirectionality with local type inference, and we explore the design space
-  to show how the same scheme could be applied to work with different
-  type systems. 
-\end{itemize}
-
-\nk{We need to explain what local type inference is, and how it is and isn't
-  related to bidirectional typechecking.}
+\paragraph{Extrapolation of the Inference Framework to Other Type Systems}
+    We explore the design space further to show how the
+    same type inference scheme could be applied to
+    work with different type systems. In particular,
+    \begin{enumerate*}
+      \item[(i)] in combination with \emph{bidirectional} typechecking---
+        the approach used in the original local type inference paper of \citet{pierce2000:local}
+        that minimizes the number of needed annotations;
+      \item[(ii)] in systems with different subtyping rules, such as  
+        the one used in \citet{zhao22:elementary}---a system that
+        permits explicit type applications;
+      \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
+        That is, one can only instantiate a quantifier with a 
+        type that is a subtype of a given bound.
+    \end{enumerate*}

### 2024-04-16 oopsla24/tex/01-introduction.tex
Edit intro and overview

@@
 
+In this paper, we extend the local type inference algorithm to a language with
+both universal and existential quantifiers, which can both be instantiated
+impredicatively. This extended system is referred to as \fexists. The
+combination of features in \fexists  broke a number of the invariants which
+traditional type inference algorithms depend on and required us to invent
+new algorithms which combine both unification and (surprisingly)
+anti-unification.
+
 To research the type inference in impredicative polymorphic systems
@@
     of which can be instantiated impredicatively. To evade the undecidability
-    results surrounding type inference for \systemf, we work in a variant of
-    \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping relation
-    which is still decidable.
+    results surrounding type inference for \systemf, we introduce \fexists---a
+    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
+    relation which is still decidable.
 

### 2024-04-16 oopsla24/tex/02-overview.tex
Edit intro and overview

@@
 
-Positive expressions correspond to values in the \CBPV system
-\cite{levy2006:cbpv}, and negative expressions correspond to computations.
-The difference between the computations and values is intuitively described 
-in the slogan of \CBPV: ``a value is, a computation does''. 
-In particular, an argument of a function is always a value (\ie positive), 
-and the resulting type, as well as the function itself, 
-is a computation (\ie negative).
+Positive expressions correspond to \emph{values} in the \CBPV system, and
+negative expressions correspond to \emph{computations}. The difference between
+the computations and values is intuitively described in the motto of \CBPV: ``a
+value is, a computation does''. In particular, an argument of a function is
+always a value (\ie positive), and the resulting type, as well as the function
+itself, is a computation (\ie negative). Analogously, the polymorphic
+$\forall$ quantifies a computation over positive types.
 
@@
 
-In general, any term and type of \systemf can be embedded into \fexists in
-several ways, because at each moment one can choose the positive or negative
-polarity of the subterm. The consistent term-level and type-level embeddings
-(that preserve typing) will be covered in detail in
-\cref{sec:rel-to-systemf}. Now, we discuss several examples of standard
-\systemf terms and types and their polarization in \fexists.
+Generally, any term and type of \systemf can be embedded into \fexists in
+multiple ways. This is because at any given point, one can opt for either the
+positive or negative polarity of the subterm. We will delve into the consistent
+term-level and type-level embeddings (\ie preserving typing) in more detail in
+\cref{sec:rel-to-systemf}. \Cref{fig:polarization-examples} showcases the
+polarized versions of the standard \systemf terms. Next, we will discuss several
+examples (using the terms from \cref{fig:polarization-examples}) to illustrate
+what can and cannot be inferred in \fexists.
 
@@
 
-Let us assume that $[[Φ]]$ is a context containing the terms from \cref{fig:polarization-examples}.
-Then the following types are inferrable in \fexists.
-
 \paragraph{Application of a Polymorphic Function}
-  A polymorphic function can be applied to a value of a concrete type.
-  In this case, the quantified variables are instantiated to make the application type-check.
-  For instance, the polymorphic $[[id]]$ can be applied to any positive type, 
-  and the resulting type will be inferred by the algorithm:
-  $$[[· ; Φ ⊢ let x = id(fortyTwo); return x : ↑Int]]$$
+  A polymorphic function can be applied to a value of a concrete type. In this
+  case, the quantified variables are instantiated to make the application
+  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
+  of any positive type, and the resulting type will be inferred by the
+  algorithm:
+  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$
 
 \paragraph{Polymorphic Self-application}
-  Unlike predicative systems, \fexists
-  permits self-application of polymorphic functions.
-  For instance, $[[id]]$ can be applied to itself
-  resulting in a polymorphic type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$:
+  Unlike predicative systems, \fexists permits self-application of polymorphic
+  functions. For instance, $[[id]]$ can be applied to itself resulting in a
+  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
+  turn, can be applied to itself:
   $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$
@@
 \paragraph{Non-example: Polymorphic Arguments}
-The focus of our system is the \emph{local} type inference \cite{pierce2000:local},
-which has certain limitations.
-In particular, if an argument of a function is polymorphic,
-its polymorphic variables are \emph{not} instantiated by the inference algorithm. 
-For example, one would expect the following to be inferrable:
+The focus of our system is the \emph{local} type inference
+\cite{pierce2000:local}, which has certain limitations. In particular, the
+instantiation does not happen when the polymorphic term is in the argument
+position. For example, one would expect the following to be inferrable:
 \mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
 However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
-which requires the information to be propagated from the neighboring branch 
-of the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.
+which requires the information to be propagated from the neighboring branch of
+the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.
 
@@
 categories (polarities): positive and negative,  
-similar to the values and computations in the \CBPV system \cite{levy2006:cbpv}.
-\begin{itemize}
-\item [$-$] $[[na]]$ is a negative type variable, which can be taken from a context or introduced by $[[∃]]$.
-\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
-\item [$-$] a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
-  a list of positive type variables 
-  $[[pas]]$. The polarities are chosen to follow the definition of functions.
-\item [$-$] a shift $[[↑iP]]$ allows a value to be used as a computation, 
-  which at the term level corresponds to a pure computation $[[return v]]$.
-\item [$+$] $[[pa]]$ is a positive type variable, taken from a context or introduced by $[[∀]]$.
-\item [$+$] $[[∃nas.iP]]$, symmetrically to $[[∀]]$, 
-  binds negative variables in a positive type $[[iP]]$. 
-\item [$+$] a shift $[[↓iN]]$, symmetrically to the up-shift, 
-  thunks a computation, which at the term level corresponds to $[[ {c} ]]$.
-\end{itemize}
+similar to the values and computations in the \CBPV system.
 
-\begin{figure}[h]
+\begin{figure}
   \begin{minipage}[t]{0.6\textwidth}
@@
 
+\begin{itemize}
+\item [$\pm$] $[[na]]$ and $[[pa]]$
+  denote negative and positive type variables, respectively.
+  The variables are either introduced by the corresponding quantifiers ($[[∀]]$ and $[[∃]]$)
+  or declared in the type context;
+\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
+\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
+  a list of positive type variables $[[pas]]$;
+\item [$+$] dually, $[[∃nas.iP]]$, binds negative variables in a positive type $[[iP]]$;
+\item [$-$] an upshift $[[↑iP]]$ allows a value to be used as a computation, 
+  which at the term level corresponds to a pure computation $[[return v]]$;
+\item [$+$] dually, a downshift $[[↓iN]]$ thunks a computation, which at the
+  term level corresponds to $[[ {c} ]]$.
+\end{itemize}
+
 \paragraph{Definitional Equalities}
@@
 This way, we assume that substitutions do not capture bound variables.
-Besides, we equate
-$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
-and lift these equations transitively and congruently 
-to the whole system.
+Besides, we assume the quantification is associative by
+(congruently) equating $[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
+as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$.
 
@@
 \subsection{The Language of Terms}
+
+\begin{figure}
+  \begin{supertabular}{ll p{0.8\textwidth}}
+    \ottcInline\\
+    \ottvInline\\
+  \end{supertabular}
+  \caption{Declarative Terms of \fexists}
+  \label{fig:declarative-terms}
+\end{figure}
 
@@
         a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
-    \item [$-$] Applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
-        $[[let x = v ( args ) ; c]]$ operate similarly to 
-        the bind of a monad: they take a suspended computation $[[v]]$,
-        apply it to a list of arguments, bind the result 
-        (which is expected to be pure) to a variable $[[x]]$,
-        and continue with a computation $[[c]]$.
-        If the resulting type of the application is unique, 
-        one can omit the type annotation, as in the second form:
-        it will be inferred by the algorithm;
+    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
+      operates similarly to a monadic bind.
+      It binds the result of a computation $[[c : ↑iP]]$ 
+      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
+    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
+        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
+        computational let, when the first computation is a function application
+        $[[v]] ([[args]])$.
+        They take a suspended computation $[[v]]$, apply it to a list of
+        arguments, bind the result (which is expected to be pure) to a variable
+        $[[x]]$, and continue with a computation $[[c]]$. If the resulting type
+        of the application is unique, one can omit the type annotation, as in
+        the second form: it will be inferred by the algorithm;
     \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
@@
 In \cref{sec:extensions}, we discuss the way to modify the system to introduce \emph{explicit} type applications.
-
-\begin{figure}[h]
-  \begin{supertabular}{ll p{0.8\textwidth}}
-    \ottcInline\\
-    \ottvInline\\
-  \end{supertabular}
-  \caption{Declarative Terms of \fexists}
-  \label{fig:declarative-terms}
-\end{figure}
 

### 2024-04-30 oopsla24/paper.tex
Edit

@@
 \begin{abstract}
-  % {\color{red}CHANGE!!}
+  {\color{red}Change!}
+  Type inference for polymorphic types has been extensively studied, with the
+  Damas-Hindley-Milner algorithm being the most well-known approach for handling
+  parametric polymorphism. However, supporting impredicative existential types
+  has remained an open challenge due to undecidability barriers. This paper
+  presents \fexists, a type system extending the \CBPV calculus with
+  first-class existential and universal quantifiers that can be instantiated
+  impredicatively. We give a local type inference algorithm for \fexists that avoids
+  the undecidability issues through a novel combination of unification and
+  anti-unification techniques. A key contribution is providing a declarative
+  specification of the inference algorithm based on the \fexists type system and
+  proving the algorithm's soundness and completeness. We also explore extending
+  the inference framework to bidirectional typechecking, different subtyping
+  regimes, and bounded polymorphism. This work opens up new avenues for bringing
+  the benefits of impredicative existential types to programming languages
+  through principled type inference.
+
   % This paper addresses the challenging problem of type inference for
@@
 \section{Introduction}
-
 \input{\genDir/tex/01-introduction.tex}

### 2024-04-30 oopsla24/tex/01-introduction.tex
Edit

@@
 extremely simple and compact type system, whose rules fit on a single page, but
-whose semantics are sophisticated enough to model things like parametricity and
-representation independence. However, System F by itself is unwieldy as a
-programming language. The fact that the universal type $[[∀a.fA]]$ has explicit
-eliminations means that programs written using polymorphic types will need to be
-stuffed to the gills with type annotations explaining how and when to
-instantiate the quantifiers.
+whose semantics are advanced enough to accurately represent concepts such as
+parametricity and representation independence. However, \systemf by itself is
+unwieldy as a programming language. The fact that the universal type $[[∀a.fA]]$
+has explicit eliminations means that programs written using polymorphic types
+will need to be stuffed to the gills with type annotations explaining how and
+when to instantiate the quantifiers.
 
@@
 
-The most famous of the algorithms for solving these constraints is the
-Damas-Hindley-Milner (HDM) algorithm
-\cite{hindley69:principal,milner78:theory,damas82:principal}. The idea is that
-type instantiation induces a subtype ordering: the type $[[∀a.fA]]$ is a subtype
-of all of its instantiations. So we wish to be free to use the same function
-$[[len]]$ at many different types such as 
-$[[List Int → Int]]$, $[[List Bool → Int]]$,
-$[[List (Int × Bool) → Int]]$, and so on.
-However, the subtype relation is nondeterministic: it tells us that whenever we
-see a polymorphic type $[[∀a.fA]]$, we know it is a subtype of \emph{any}
-of its instantiations. To turn this into an algorithm, we have to actually make
-some choices, and DHM works by using \emph{unification}. Whenever we would have
-had to introduce a particular concrete type in the specification, the DHM
-algorithm introduces a unification variable, and incrementally instantiates this
-variable as more and more type constraints are observed.
+The way this is typically done is by using a \emph{subtyping} relation, induced
+by the polymorphic instantiation: the type $[[∀a.fA]]$ is a \emph{subtype} of
+all of its instantiations. So we wish to be free to use the same polymorphic
+function $[[len]]$ at many different types such as $[[List Int → Int]]$, $[[List
+Bool → Int]]$, $[[List Int × Bool → Int]]$, and so on. However, the subtyping
+can be used nondeterministically: whenever we see a polymorphic type
+$[[∀a.fA]]$, we know it is a subtype of \emph{any} of its instantiations. To
+turn this into an algorithm, we have to actually make some choices.
+
+The most famous of the algorithms for handling this is the Damas-Hindley-Milner
+(DHM) algorithm \cite{hindley69:principal,milner78:theory,damas82:principal}.
+DHM deals with this choice using \emph{unification}. Whenever we would have had
+to introduce a particular concrete type in the specification, the DHM algorithm
+introduces a \emph{unification variable}, and incrementally instantiates this variable
+as more and more type constraints are observed.
 
@@
 alternating universal and existential quantifiers. Quantifiers still range over
-monotypes, and higher-rank polymorphism is not permitted. More recently,
+monotypes (\ie types with no quantifiers inside of them), and higher-rank 
+polymorphism is not permitted: one cannot instantiate a quantifier with a type
+that itself contains quantifiers.
+More recently,
 \citet{dunfield16:existential} studied type inference for existential types in
@@
 All of these papers are restricted to \emph{predicative} quantification, where
-quantifiers can only be instantiated with monotypes (i.e., types without any
-occurences of quantifiers). However, existential types in full System F are
-\emph{impredicative}---that is, quantifiers can be instantiated with arbitrary
-types, specifically including types containing quantifiers.
+quantifiers can only be instantiated with monotypes. However, existential types
+in full System F are \emph{impredicative}---that is, quantifiers can be
+instantiated with arbitrary types, specifically including types containing
+quantifiers.
 
@@
 work of \citet{serrano-2020}), with a focus on avoiding the undecidability
-barriers by doing \emph{partial} type inference. That is, we no longer try to do
-full type inference, but rather accept that the programmer will need to write
-annotations in cases where inference would be too difficult. Unfortunately, it
-is often difficult to give a specification for what the partial algorithm
-does--for example, \citet{eisenberg21:existential} observe that their algorithm
-lacks a declarative specification, and explain why existential types make this
-particularly difficult to do.
+barriers by doing \emph{partial} type inference. That is, we no longer try to
+infer the type of any term in the language, but rather accept that the
+programmer will need to write annotations in cases where inference would be too
+difficult. Unfortunately, it is often difficult to give a specification for when
+the partial algorithm succeeds---for example, \citet{eisenberg21:existential}
+observe that their algorithm lacks a declarative specification, and explain why
+existential types make this particularly difficult to do.
 
@@
 Java use a form of local type inference, due to the ease of extending this style
-of inference to support OO features.
+of inference to support object-oriented features.
 
@@
 new algorithms which combine both unification and (surprisingly)
-anti-unification.
+the dual---anti-unification.
 
@@
 To research the type inference in impredicative polymorphic systems
-in the presence of existential types, we make the following \emph{contributions}.
+in the presence of existential types, we make the following \emph{\textbf{contributions}}.
 
-\paragraph{Type Inference Algorithm} We give a local type inference algorithm
+\paragraph{Type Inference Algorithm (\cref{sec:algorithm})}
+    We give a local type inference algorithm
     which supports both first-class existential and universal quantifiers, both
@@
 
-\paragraph{Declarative Specification}
+\paragraph{Declarative Specification (\cref{sec:declarative-system,sec:proofs})}
     We give a declarative type system (again, based on \CBPV) to
@@
 
-\paragraph{Employment of the Anti-unification}
+\paragraph{Employment of the Anti-unification (\cref{sec:antiunification})}
     The introduction of impredicative existentials breaks some of the
@@
 
-\paragraph{Extrapolation of the Inference Framework to Other Type Systems}
+\paragraph{Extrapolation of the Inference Framework to Other Type Systems
+          (\cref{sec:extensions})}
     We explore the design space further to show how the
-    same type inference scheme could be applied to
+    same type inference scheme can be applied to
     work with different type systems. In particular,
     \begin{enumerate*}
-      \item[(i)] in combination with \emph{bidirectional} typechecking---
-        the approach used in the original local type inference paper of \citet{pierce2000:local}
-        that minimizes the number of needed annotations;
+      \item[(i)] in combination with \emph{bidirectional} typechecking---the 
+        approach used in the original local type inference paper of 
+        \citet{pierce2000:local} that minimizes the number of needed annotations;
       \item[(ii)] in systems with different subtyping rules, such as  
@@
     \end{enumerate*}
+
+\ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
+\paragraph{\textbf{Structure of this Paper}}
+In \cref{sec:overview}, we discuss the language of the \fexists system, providing
+examples, outlining limitations, and summarizing the main ideas behind the
+algorithm. \Cref{sec:declarative-system} presents a formal definition of the
+declarative specification and its key properties. In \cref{sec:algorithm}, the
+type inference algorithm is introduced, detailing a hierarchy of subroutines;
+upon perusing this section, one could \emph{implement} the type inference
+algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
+explaining its structure and how the declarative system's features make it
+decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
+work directions. Finally, \cref{appendix} contains comprehensive formal
+definitions and proofs of lemmas and theorems used in the paper; in instances
+where the full definitions might not be of immediate interest, we direct the
+reader to the appendix.
+
+% \paragraph{\textbf{Acknowledgments}} We would like to thank Meven
+% Lennon-Bertrand for his extremely helpful comments and suggestions.

### 2024-04-30 oopsla24/tex/02-overview.tex
Edit

@@
-% Imprediicative
-% CBPV, monadic control
-% Local Inference
-% Two ways to translate
+\label{sec:overview}
 
-The language of \fexists is different from standard \systemf:
-every term and type has either positive or negative polarity.
-The type variables are annotated with their polarities
-(e.g. $[[α⁺]]$ or $[[β⁻]]$), and the types can change polarity
-via the shift operators $[[↑]]$ (positive to negative) and
-$[[↓]]$ (negative to positive).
+In the presence of impredicative polymorphism, the standard \systemf has
+undecidable subtyping and type inference. It means that any inference algorithm
+is incomplete: it only infers the types for a \emph{fragment} of \systemf.
+
+To tackle this issue, we restrict \systemf by employing the ideas from the \CBPV
+system \cite{levy2006:cbpv}. We introduce a new language, \fexists, which
+stratifies the syntax of \systemf into positive and negative parts. It provides
+the structure that allows us to restrict the typing specification to make it
+decidable.
+
+Every term and type fo \fexists has either \emph{positive} or \emph{negative}
+polarity. The type variables are annotated with their polarities (e.g. $[[α⁺]]$
+or $[[β⁻]]$), and the types can change polarity via the shift operators $[[↑]]$
+(positive to negative) and $[[↓]]$ (negative to positive).
 
@@
 instantiation does not happen when the polymorphic term is in the argument
-position. For example, one would expect the following to be inferrable:
+position. For example, one could expect the following to be inferrable:
 \mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
@@
   denote negative and positive type variables, respectively.
-  The variables are either introduced by the corresponding quantifiers ($[[∀]]$ and $[[∃]]$)
-  or declared in the type context;
-\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
-\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
-  a list of positive type variables $[[pas]]$;
-\item [$+$] dually, $[[∃nas.iP]]$, binds negative variables in a positive type $[[iP]]$;
-\item [$-$] an upshift $[[↑iP]]$ allows a value to be used as a computation, 
-  which at the term level corresponds to a pure computation $[[return v]]$;
-\item [$+$] dually, a downshift $[[↓iN]]$ thunks a computation, which at the
-  term level corresponds to $[[ {c} ]]$.
+  The variables are either introduced by the corresponding quantifiers 
+  ($[[∀]]$ and $[[∃]]$, respectively) or declared in the type context;
+\item [$-$] a function $[[iP → iN]]$ takes a \emph{positive} input $[[iP]]$ and
+returns a \emph{negative} output $[[iN]]$; the function type itself is
+\emph{negative}, so the `arrow' operator is right-associative: 
+$[[iP → iQ → iN]]$ is equivalent to $[[iP → (iQ → iN)]]$;
+\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$
+  quantifies a negative type $[[iN]]$ over a list of positive type variables $[[pas]]$;
+\item [$+$] dually, $[[∃nas.iP]]$ binds negative variables in a positive type $[[iP]]$;
+\item [$\pm$] an upshift $[[↑iP]]$ and a downshift $[[↓iN]]$ change the polarity of a type
+  from positive to negative and vice versa.
+  At the level of terms, the constructors for shift types are
+  $[[return v]]$ and $[[{c}]]$ (thunked computation), respectively.
 \end{itemize}
 
-\paragraph{Definitional Equalities}
-For simplicity, we assume that alpha-equivalent terms are equal.
-This way, we assume that substitutions do not capture bound variables.
-Besides, we assume the quantification is associative by
-(congruently) equating $[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$.
+\paragraph{Syntactic Conventions}
+We always consider terms and types up to alpha-equivalence, and capture-avoiding
+substitution. Besides, we assume the quantification is associative,
+for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
+represent the same type.
 
@@
 In the construction of \fexists, a type context (denoted as $[[Γ]]$) is
-represented as a \emph{set} of positive and negative type variables and it is used to
-assert the well-formedness of types. The well-formedness of a type is denoted as
-$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it asserts that all type variables are either
-bound by a quantifier ($[[∀]]$ and $[[∃]]$) or declared in the context $[[Γ]]$.
-The well-formedness checking is an \emph{algorithmic} procedure. As commonly
-done, we represent it as a system of inference rules, that correspond to a
-recursive algorithm taking the context and the type as input.
+represented as a \emph{set} of positive and negative type variables and it is
+used to assert the well-formedness of types. The well-formedness (or
+well-scopeness) of a type is denoted as $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it
+asserts that all type variables are either bound by a quantifier ($[[∀]]$ and
+$[[∃]]$) or declared in the context $[[Γ]]$. The well-formedness checking is an
+\emph{algorithmic} procedure. We represent it as a system of inference rules,
+that correspond to a recursive algorithm taking the context and the type as
+input. For brevity, we omit these rules, as they are standard binder scoping
+rules.
 
@@
 
-In \cref{fig:declarative-terms}, we define the language of terms of 
-\fexists. The language combines \systemf with the \CBPV approach.
+In \cref{fig:declarative-terms}, we define the language of terms of \fexists.
+The language is a combination of \systemf and \CBPV constructs: the
+terms are stratified into values and computations; there are return 
+and thunk constructors, as well as several forms of let bindings to
+sequentialize the computation/application with or without type annotation;
 
@@
         They take a suspended computation $[[v]]$, apply it to a list of
-        arguments, bind the result (which is expected to be pure) to a variable
-        $[[x]]$, and continue with a computation $[[c]]$. If the resulting type
+        arguments, bind the (pure) result to a variable
+        $[[x]]$, and continues with a computation $[[c]]$. If the resulting type
         of the application is unique, one can omit the type annotation, as in
@@
     \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
-        is the standard unpack of an existential type:
-        expecting $[[v]]$ to be an existential type,
+        is the standard eliminator of an existential type (unpack):
+        expecting $[[v]]$ to be of an existential type,
         it binds the packed negative types to a list of 
@@
 
-\subsection{The key ideas of the algorithm}
+\subsection{The Key Ideas of the Algorithm}
 
@@
 of the syntax tree. Nevertheless, many difficulties appear in the inference of
-polymorphic and existential types. Next, we discuss the most challenging of
-these difficulties and how they are addressed in the algorithm.
+the combination of polymorphic and existential types. Next, we discuss the most
+challenging of these difficulties and how they are addressed in the algorithm.
 
-\paragraph{Subtyping} Most of the difficulty of the inference algorithm is
-concentrated in the subtyping relation. This is the part of the system where
-most of the polymorphic instantiation happens. In particular, it is the
-subtyping relation that allows us to use the polymorphic terms in the context
-where their concrete instantiations are expected.
+\paragraph{Subtyping} 
+The inference algorithm's complexity mainly lies in the subtyping relation,
+which is, in particular, responsible for polymorphic instantiation. It's this
+subtyping relation that enables us to apply polymorphic terms in situations
+where their specific instantiations are anticipated.
 
@@
 \paragraph{Equivalence}
-Our system introduces another complexity: the equivalence induced by subtyping
+The system introduces another complexity: the equivalence induced by subtyping
 ($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
@@
 the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
-the equivalence relation must permit the permutations of the quantifiers. For
-example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
-equivalence permits the removal/addition of the unused quantifiers: $[[· ⊢
-∀α⁺.↑Int ≈ ↑Int]]$.
+the equivalence relation must permit the \emph{permutations} of the quantifiers.
+For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
+equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
+⊢ ∀α⁺.↑Int ≈ ↑Int]]$.
 
-In order to manage the complexity of non-trivial equivalence, we implement
-a type normalization algorithm. It effectively reduces the mutual subtyping to
-a simpler alpha-equivalence. The normalization works by recursively
-eliminating unused quantifiers and reordering the remaining ones into a
-canonical permutation.
+To handle the complexity of non-trivial equivalence, we employ a type
+normalization algorithm. This approach simplifies mutual subtyping to a more
+manageable alpha-equivalence by eliminating unused quantifiers and reordering
+the rest into a canonical permutation, effectively replacing subtyping-induced
+equivalence on terms with alpha-equivalence on their normal forms.
 
@@
 still have non-trivial upper bounds, and thus, need to use anti-unification to
-find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and in
-fact, makes the constraint resolution decidable.
+find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and 
+as we prove, makes the constraint resolution decidable.

### 2024-04-30 oopsla24/tex/03-declarative-system.tex
Edit

@@
 
-The declarative system serves as a specification of the
-type inference algorithm. It consists of two main parts:
-the subtyping and the type inference.
+In this section, we present the declarative system of \fexists,
+which serves as a specification of the type inference algorithm.
+The declarative system is represented as a set of inference rules
+and consists of two main subsystems: subtyping and type inference.
+First, we present the declarative subtyping rules specifying 
+when one type is a subtype of another; next, we discuss 
+the equivalence relation induced by mutual subtyping;
+finally, we present the type inference rules,
+that refer to the subtyping and equivalence relation.
+We conclude this section by discussing the relation between the
+proposed type system and the standard \systemf.
 
 \subsection{Subtyping}
-It is represented by a set of inference rules shown in
-\cref{fig:declarative-subtyping}.
 
-\begin{figure}[h]
-  \begin{multicols}{2}
+\begin{figure}
+  \begin{minipage}[t]{0.49\textwidth}
     \ottdefnDOneNsubLabeled{}
-
+  \end{minipage}%
+  \begin{minipage}[t]{0.49\textwidth}
     \ottdefnDOnePsupLabeled{}
-  \end{multicols}
+  \end{minipage}
   \hfill\\
-  \begin{multicols}{2}
+  \begin{minipage}[t]{0.49\textwidth}
     \ottdefnDOneNeqLabeled{}
-
+  \end{minipage}%
+  \begin{minipage}[t]{0.49\textwidth}
     \ottdefnDOnePeqLabeled{}
-  \end{multicols}
+  \end{minipage}
   \caption{Declarative Subtyping}
@@
 
+The inference rules representing declarative subtyping are shown in
+\cref{fig:declarative-subtyping}. Let us discuss them in more detail.
+
 \paragraph{Quantifiers}  
@@
 \ruleref{\ottdruleDOneExistsLabel} specify 
-the subtyping between top-level quantified types.
+subtyping between top-level quantified types.
 Usually, the polymorphic subtyping is represented by two rules
-introducing quantifiers to the left and to the right-hand side of the subtyping.
+introducing quantifiers to the left and to the right-hand side of subtyping.
 For conciseness of representation, we compose these rules into one.
@@
 ($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
+This is the point of non-determinism as the instantiation can be chosen arbitrarily
 
@@
 As mentioned above, an important restriction that we put on the subtyping system
-is that the subtyping on shifted types requires their equivalence, as shown in
-\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}.
-Relaxing both of these invariants make the system equivalent to \systemf, and
-thus, undecidable. However, after certain changes
-\ruleref{\ottdruleDOneShiftULabel} can be relaxed to the covariant form, thereby
-increasing the expressiveness of the system. These changes are discussed in
-\cref{sec:weakening-invariant}.
+is that subtyping of shifted types requires their equivalence, as shown in
+\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
+reason for this is that if both of these rules were relaxed to the covariant
+form, the subtyping relation would become equivalent to the standard subtyping
+of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, after
+certain changes \ruleref{\ottdruleDOneShiftULabel} can be relaxed to the
+covariant form, thereby increasing the expressiveness of the system. These
+changes are discussed in \cref{sec:weakening-invariant}.
 
 \paragraph{Functions}
-Standardly, the subtyping of function types is covariant in the return type
+Standardly, subtyping of function types is covariant in the return type
 and contravariant in the argument type.
@@
 \paragraph{Variables}
-The subtyping of variables is defined reflexively,
-which is enough to ensure the reflexivity of subtyping in general.
-The algorithm will use the fact that the subtypes of a variable 
-coincide with its supertypes, which however is not true for an
-arbitrary type.
+Subtyping of variables is defined reflexively, which is enough to ensure the
+reflexivity of subtyping in general. The algorithm---specifically the least
+upper bound procedure---will use the fact that the subtypes of a variable
+coincide with its supertypes (\cref{prop:var-no-subtypes}),
+which however does not hold for arbitrary types.
 
-\subsubsection{Properties of the Declarative Subtyping}
+\subsubsection{Properties of Declarative Subtyping}
 \label{sec:decl-subtyping-properties}
@@
 
-\begin{property}[Subtyping Preserves Free Variables]
+\begin{property}[Subtyping preserves free variables]
   \label{prop:subtyping-preserves-fv}
@@
 
-Another property that we extensively use is that the subtyping is reflexive and transitive,
+\begin{property}[Variable subtyping is trivial]
+  \label{prop:var-no-subtypes}
+  A subtype or a supertype of a variable is equivalent to the variable itself:
+
+  \begin{tabular}{@{}llclcl@{}}
+    $-$ & $[[Γ ⊢ α⁻ ≤ iN]]$ & $\iff$ & $[[Γ ⊢ iN ≤ α⁻]]$ & $\iff$ & $[[iN]] = [[∀pbs.α⁻]]$ \\
+    $+$ & $[[Γ ⊢ α⁺ ≥ iP]]$ & $\iff$ & $[[Γ ⊢ iP ≥ α⁺]]$ & $\iff$ & $[[iP]] = [[∃nbs.α⁺]]$
+  \end{tabular}
+\end{property}
+
+Another property that we extensively use is that subtyping is reflexive and transitive,
 and agrees with substitution.
@@
 \begin{property}[Subtyping forms a preorder]
-  Let us say that two types $[[iN1]]$ and $[[iN2]]$ are in the subtyping relation if there 
-  exists a context $[[Γ]]$ such that $[[Γ ⊢ iN1 ≤ iN2]]$;
-  symmetrically, two types $[[iP1]]$ and $[[iP2]]$ are in the subtyping relation
-  if there exists $[[Γ]]$ such that $[[G ⊢ iP2 ≥ iP1]]$.
-  Then the subtyping relation defined this way is reflexive and transitive.
+  For a fixed context $[[Γ]]$, the negative subtyping relation 
+  $[[Γ ⊢ iN1 ≤ iN2]]$ and the positive subtyping relation 
+  $[[Γ ⊢ iP1 ≥ iP2]]$
+  are reflexive and transitive on types well-formed in $[[Γ]]$.
 \end{property}
@@
 
-Moreover, any two \emph{positive} types have the least upper bound, which makes
-the positive subtyping semilattice.
-The positive least upper bound can be found algorithmically,
-which we will discuss in the next section.
+Moreover, any two \emph{positive} types have a least upper bound, which makes
+positive subtyping a semilattice. The positive least upper bound can be found
+algorithmically, which we will discuss in the next section.
 
-\begin{property}[Positive Least Upper Bound exists]
+\begin{property}[Positive least upper bound exists]
   Suppose that $[[iP1]]$ and $[[iP2]]$ are positive types
   well-formed in $[[Γ]]$.
-  Then there exists the least common supertype--- a type $[[iP]]$ such that
+  Then there exists a least common supertype--- a type $[[iP]]$ such that
   \begin{itemize}
@@
 
-\paragraph{Negative Greatest Lower Bound does not exist}
-However, the symmetric construction---the greatest lower bound of two negative types
-does not always exist. Let us consider the following counterexample.
-Let us consider the following types: 
+\paragraph{Negative greatest lower bound might not exist}
+The symmetric construction---the greatest lower bound of two negative types---does 
+not always exist, as the following counterexample shows.
+Consider the following types: 
 \begin{itemize}
@@
 $[[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $[[iP → ↓↑iP1 → ↓↑iP2 → iN]]$?
-One of the common subtypes is $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$,
-which, however is not the greatest one.
+The type $[[∀α⁺,β⁺,γ⁺. α⁺ → ↓↑β⁺ → ↓↑γ⁺ → iN]]$ is a common subtype,
+however, it is not the greatest one, as it is too general.
 
@@
 By analyzing the inference rules, we can prove that
-both $[[iM1]]$ and $[[iM2]]$ are maximal common subtypes.
-Since $[[iM1]]$ and $[[iM2]]$ are not equivalent,
-it means that none of them is the greatest.
+both $[[iM1]]$ and $[[iM2]]$ are \emph{maximal} common 
+subtypes, \ie there is no common subtype that is greater than them.
+However, $[[iM1]]$ and $[[iM2]]$ are not equivalent,
+which means that none of them is the greatest.
 
@@
 
-In our system, equivalence is reacher than equality. Specifically,
+In our system, equivalence is richer than equality. Specifically,
+while staying within one equivalence class, one can 
+change the type:
 \begin{enumerate}
-  \item[(ii)] one can introduce redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
+  \item[(i)] introduce and remove redundant quantifiers. For example, $[[∀α⁺,β⁺.↑α⁺]]$ is equivalent
   but not equal to $[[∀α⁺.↑α⁺]]$;
-  \item[(i)] one can reorder quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
+  \item[(ii)] reorder adjacent quantifiers. For example, $[[∀α⁺,β⁺.α⁺ → β⁺ → γ⁻]]$ is equivalent but not equal to
   $[[∀α⁺,β⁺.β⁺ → α⁺ → γ⁻]]$; 
-  \item[(iii)] the transformations (i) and (ii) can happen at any
-    position in the type.
+  \item[(iii)] make the transformations (i) and (ii) at any position in the type.
 \end{enumerate}
@@
 The steps (i-ii) are implemented by the \emph{ordering} function
-$[[ord varset in iN]]$ and $[[ord varset in iP]]$.
+`$[[ord varset in iN]]$' and `$[[ord varset in iP]]$'.
 For a set of variables $[[varset]]$, and a type, it 
 returns a list of variables from $[[varset]]$ that occur in the type in 
-the order of their first occurrence.
-For brevity, we omit the formal definition of ordering,
-referring the reader to the appendix.
+the \emph{order of their first occurrence}.
+For brevity, we omit the formal definition of ordering referring the reader to the appendix (\cref{sec:ordering}).
 
@@
   \\
-  $[[ord varset in iN]]$ returns a list of variables
+  $[[ord varset in iN]]$ returns a list of variables from 
   $[[varset ∩ fv(iN)]]$ in the order of their first occurrence in $[[iN]]$
@@
   \\
-  $[[ord varset in iP]]$ returns a list of variables
+  $[[ord varset in iP]]$ returns a list of variables from 
   $[[varset ∩ fv(iP)]]$ in the order of their first occurrence in $[[iP]]$
@@
 \begin{property}[Correctness of normalization]
-  \hfill \\
+  Assuming all types are well-formed in $[[Γ]]$,
   \begin{itemize}
-    \item[$-$] For $[[iN]]$ and $[[iM]]$ well-formed in $[[Γ]]$,
-  $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$;
-    \item[$+$] analogously, for $[[iP]]$ and $[[iQ]]$ well-formed in $[[Γ]]$,
-  $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
+    \item[$-$] $[[Γ ⊢ iN ≈ iM]]$ is equivalent to $[[nf(iN) = nf(iM)]]$, and
+    \item[$+$] $[[Γ ⊢ iP ≈ iQ]]$ is equivalent to $[[nf(iP) = nf(iQ)]]$.
   \end{itemize}
@@
 
-\subsection{Type Inference}
+\subsection{Typing}
 
-The declarative specification of the type inference is shown in 
+The declarative specification of the type inference is given in 
 \cref{fig:declarative-inference}.
@@
 ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
-the term $[[v]]$ is allowed to have the type $[[iP]]$'',
+the term $[[v]]$ is allowed to infer type $[[iP]]$'',
 where $[[Φ]]$---the variable context---is defined standardly as
@@
 infers the type of the application of a function to a list of arguments.
-It has form of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
+It has form $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$, 
 which reads ``under the type context $[[Γ]]$ and variable context $[[Φ]]$,
 the application of a function of type $[[iN]]$ to the list of arguments $[[args]]$
-is allowed to have the type $[[iM]]$''.
+is allowed to infer type $[[iM]]$''.
 
@@
 \paragraph{Variables}
-  Rule \ruleref{\ottdruleDTVarLabel} allows to infer
+  Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
   the type of a variable from the context. 
-  In literature can be found another version of this rule,
+  In literature, one can find another version of this rule,
   that enables inferring a type \emph{equivalent}
@@
 \paragraph{Annotations}
-  Subtyping is also used by the annotation rules \ruleref{\ottdruleDTNAnnotLabel}
-  and \ruleref{\ottdruleDTPAnnotLabel}. The annotation is only valid if the
+  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
+  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
+  The annotation is only valid if the
   inferred type is a subtype of the annotation type.
@@
 
-\paragraph{Return and Thunk}
+\paragraph{Return and thunk}
   Rules \ruleref{\ottdruleDTReturnLabel} 
   and \ruleref{\ottdruleDTThunkLabel}
-  add the corresponding shifts to the type of the body.
+  simply add the corresponding shift constructors to the type of the body.
 
@@
 
-\paragraph{Applicative Let Binders}
+\paragraph{Applicative let binders}
   Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
@@
 
-\paragraph{Typing up to Equivalence}
+\paragraph{Typing up to equivalence}
   As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
@@
 
-\paragraph{Application to an Empty List of Arguments}
+\paragraph{Application to an empty list of arguments}
   The base case of the application type inference is 
@@
 
-\paragraph{Application of a Polymorphic Type $[[∀]]$}
-  The complexity of the system is hidden in the rules, 
-  whose output type is not immediately defined by their input
-  and the output of their premises 
-  (\aka not mode-correct \cite{dunfield2020:bidirectional}).
-  In our typing system, such rule is \ruleref{\ottdruleDTForallAppLabel}: 
-  the instantiation of the quantifying variables is
-  not known a priori. The algorithm we present in \cref{sec:algorithm} 
-  delays this instantiation until more information about it 
+\paragraph{Application of a polymorphic type $[[∀]]$}
+  The complexity of the system lives in the rules whose output type is not
+  immediately defined by their input and the output of their premises (\aka not
+  mode-correct \cite{dunfield2020:bidirectional}). In our typing system,
+  \ruleref{\ottdruleDTForallAppLabel} is such a rule: the instantiation of the
+  quantifying variables is not known a priori. The algorithm we present in
+  \cref{sec:algorithm} delays this instantiation until more information about it
   (in particular, typing constraints) is collected.
@@
 
-\paragraph{Application of an Arrow Type}
-  Another important application rule is \ruleref{\ottdruleDTArrowAppLabel}.
+\paragraph{Application of an arrow type}
+  Another application rule is \ruleref{\ottdruleDTArrowAppLabel}.
   This is where the subtyping is used to check that the type of the argument
@@
 
-\subsubsection{Declarative Typing Properties}
+  \vspace{\baselineskip}
+% \subsubsection{Properties of Declarative Typing}
   An important property that the declarative system has is
@@
 
-\begin{property}[Declarative Typing is Defined up to Equivalence]
+\begin{property}[Declarative typing is defined up to equivalence]
   Let us assume that $[[Γ ⊢ Φ1 ≈ Φ2]]$, \ie  
@@
 The choice of polarization affects the execution strategy of the program.
+\ilyam{How?}
 We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
 This choice makes it smoother to lift the translation to the term level, which will be disused next.
+
+The type-level translation is naturally lifted to the contexts:
+$[[|Γ|]]$ is the context $[[Γ]]$ with every type depolarized,
+and symmetrically, $[[+|fΓ|+]]$ is the context $[[fΓ]]$ with every type polarized.
+The context translation is used to formulate the soundness of the
+ term-level translation, which we discuss next.
 
@@
       \end{minipage}
-    \caption{Type Depolarization}
+    \caption{Type Depolarization \protect\footnotemark}
     \label{fig:depolarization}
@@
   \end{minipage}
+
 \end{figure}
+
+\footnotetext{Here the \systemf existential $[[∃ as . fT]]$ is a syntax sugar for its standard encoding $[[∀b.(∀as.(fT → b)) → b]]$}
 
@@
 
-The term-level translations between systems are more complex.
-They are defined as \emph{elaborations} of the typing derivations.
-In other words, the input of the translation is not a term, 
-but a whole typing derivation of this term in the corresponding system.
-For brevity, we omit the formal definition of the term-level translation,
-referring the reader to the appendix (\cref{sec:term-translation}).
+The term-level translations between the systems are more complex. They are
+defined as \emph{elaborations} of the typing derivations. In other words, the
+input of the translation is not a term, but a whole typing derivation in the
+corresponding system. For example, the subtyping elaboration $[[Γ ⊢ iN ≤ iM ⤳
+ft]]$ constructs a \systemf function $[[ft]]$ of type $[[|iN| → |iM|]]$ that
+witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$; a typing elaboration
+$[[Γ; Φ ⊢ v : iP ⤳ ft]]$ builds a term $[[ft]]$---a \systemf counterpart of
+$[[v]]$---of type $[[|iP|]]$.
 
-We prove the soundness of the translation (\cref{lemma:translation-soundness}),
-which in particular guarantees that the translation preserves typing.
+For brevity, we omit the formal definition of the term-level elaboration,
+referring the reader to the appendix (\cref{sec:term-translation}). We define
+the elaboration in such a way that it preserves the soundness invariants given
+in \cref{lemma:translation-soundness}.
 
-\begin{property}[Soundness of the Translation]
-    \label{lemma:translation-soundness}
-    \begin{align*}
-      & [[Γ ⊢ iN ≤ iM ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iN| → |iM| ]], \\
-      & [[Γ ⊢ iP ≥ iQ ⤳ ft]]  \text{ implies }  [[|Γ| ; · ⊢ ft : |iQ| → |iP| ]], \\
-      & [[Γ; Φ ⊢ v : iP ⤳ ft]]  \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iP|]], \\
-      & [[Γ; Φ ⊢ c : iN ⤳ ft]] \text{ implies }  [[|Γ| ; |Φ| ⊢ ft : |iN|]], \\
-      & [[fΓ ; fΦ ⊢ ft : fT ⤳ c]]  \text{ implies }  [[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]].
-    \end{align*}
-\end{property}
+\begin{table}[h]
+\centering
+\begin{tabular}{rll}
+\hline
+\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
+\hline
+Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
+Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
+Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
+Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
+Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
+\hline
+\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
+\hline
+\end{tabular}
+\caption{Soundness of the Elaboration:
+  the elaboration judgment on the left implies the typing judgment on the right}
+\end{table}
+
+The least obvious aspect of the elaboration soundness is the application typing.
+The elaboration judgment $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ 
+outputs two things: 
+\begin{enumerate*}
+  \item [(i)] $[[fargs]]$---the \systemf counterpart of the argument list $[[args]]$, and 
+  \item [(ii)] $[[fe]]$---a \systemf term that is used for the final cast 
+    of the application $[[(x ● ft *)]]$ to the resulting type $[[|iM|]]$.
+\end{enumerate*}
+This final cast is needed to 
+reflect the \fexists feature of \ruleref{\ottdruleDTEmptyAppLabel}
+that permits the equivalent conversion of the resulting type $[[Γ ⊢ iN ≈ iN']]$,
+as \fexists-equivalent types
+might have different \systemf representations (\ie $[[|iN|]] \neq [[|iN'|]]$).
 

### 2024-04-30 oopsla24/tex/04-algorithm.tex
Edit

@@
 First, it is also given by a set of inference rules, which, however,
-must be mode-correct (\cite{dunfield2020:bidirectional}), \ie
+are mode-correct (\cite{dunfield2020:bidirectional}), \ie
 the output of each rule is always uniquely defined by its input.
@@
 
-\begin{figure}[h]
+\begin{figure}
   \begin{multicols}{2}
@@
   Algorithmic Type Context\\
-   $[[Ξ]] = \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
+   $[[Ξ]] \Coloneqq \{[[α1̂±]], \dots, [[αn̂±]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
   \hfill\\
-  Constraint Type Context\\
+  Instantiation Context\\
    $[[Θ]] \Coloneqq \{[[ α1̂±[Γ1] ]], \dots, [[ αn̂±[Γn] ]]\}$ where $[[α1̂±]], \dots, [[αn̂±]]$ are pairwise distinct \\
@@
 
-\paragraph{Algorithmic Contexts and Well-formedness}
-To specify when algorithmic types are well-formed, 
-we define algorithmic contexts $[[Ξ]]$ as sets of algorithmic variables.
-Then $[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness
-judgment of algorithmic terms defined as expected. 
-Informally, they check that all free declarative variables are in $[[Γ]]$, 
-and all free algorithmic variables are in $[[Ξ]]$.
-Most of the rules are inherited from the well-formedness of 
-\emph{declarative} types: the declarative variables are
-checked to belong to the context $[[Γ]]$, the quantifiers extend the context,
-type constructors are well-formed congruently.
-As we extend the syntax with algorithmic variables, we also add 
-two base-case rules for them:
- \ruleref{\ottdruleWFATPUVarLabel} and \ruleref{\ottdruleWFATNUVarLabel} 
- (see \cref{fig:algo-wf}).
-
-\begin{figure}[h]
+\paragraph{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
+To specify when algorithmic types are well-formed, we define algorithmic
+contexts $[[Ξ]]$ as sets of algorithmic variables. Then
+$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
+algorithmic terms defined as expected. Informally, they check that all free
+declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
+$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
+\emph{declarative} types: the declarative variables are checked to belong to the
+context $[[Γ]]$, the quantifiers extend the context, type constructors are
+well-formed congruently. As we extend the syntax with algorithmic variables, we
+also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
+\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).
+
+\begin{figure}
 \begin{multicols}{2}
@@
 
+\paragraph{Instantiation Context $[[Θ]]$}
+When one instantiates an algorithmic variable, one may only use type variables
+\emph{available in its scope}. As such, each algorithmic variable must remember
+the context at the moment when it was introduced. In our algorithm, this
+information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
+pairs associating algorithmic variables and declarative contexts.
+
 \paragraph{Algorithmic Normalization}
 Similarly to well-formedness, the normalization of algorithmic types is defined
-by extending the declarative definition with the algorithmic variables.
-To the rules repeating the declarative normalization, we add
-rules saying that normalization is trivial on algorithmic variables (see \cref{fig:algo-nf}).
-
-\begin{figure}[h]
+by extending the declarative definition (\cref{fig:type-nf}) with the
+algorithmic variables. To the rules repeating the declarative normalization, we
+add rules saying that normalization is trivial on algorithmic variables.
+
+\begin{figure}
 \begin{multicols}{2}
@@
 \end{multicols}
+\caption{Normalization of Algorithmic Types}
 \label{fig:algo-nf}
-\caption{Normalization of Algorithmic Types}
 \end{figure}
@@
 \subsection{Type Constraints}
-As the algorithm proceeds, it accumulates the information 
-about the algorithmic type variables in the form of \emph{constraints}.
-In our system, the constraints can be of two kinds: 
-\emph{subtyping constraints} and \emph{unification constraints}.
-The subtyping constraint can only have a positive shape $[[α̂⁺ :≥ iP]]$, \ie it 
-restricts a positive algorithmic variable to be a 
-supertype of a certain declarative type---this is one of the invariants that we 
-preserve in the algorithm.
-The unification constraint can have either a positive form $[[α̂⁺ :≈ iP]]$ or
-a negative form $[[α̂⁻ :≈ iN]]$, however, the right-hand side of the constraint
-cannot contain algorithmic type variables.
-The set of constraints is denoted as $[[SC]]$. 
-We assume that each algorithmic variable can be restricted by at most one constraint.
-
-We separately define $[[UC]]$ as a set consisting of unification constraints only.
-This is done to simplify the representation of the algorithm. 
-The unification algorithm, which we use as a subroutine of the subtyping algorithm,
-can only produce unification constraints.
-A set of unification constraints can be resolved in a simpler way than
-a general constraint set. This way, 
-the separation of the unification constraint resolution into a separate
-procedure allows us to better decompose the structure of the algorithm,
-and thus, simplify the inductive proofs.
+Throughout the algorithm's operation, it gathers information about the
+algorithmic type variables, represented as \emph{constraints}. These constraints
+in our system can be either \emph{subtyping constraints} or \emph{unification
+constraints}. As preserved by the algorithm, each subtyping constraint has a
+positive shape $[[α̂⁺ :≥ iP]]$, meaning it confines a positive algorithmic
+variable to be the supertype of a positive declarative type. Unification
+constraints can take a positive $[[α̂⁺ :≈ iP]]$ or negative $[[α̂⁻ :≈ iN]]$
+form, but algorithmic type variables cannot occur on their right-hand side. The
+constraints \emph{set} is denoted as $[[SC]]$, and we presume that each
+algorithmic variable can be restricted by at most one constraint.
+
+We define $[[UC]]$ separately as a set solely containing \emph{unification}
+constraints for simpler algorithm representation. The unification algorithm,
+which we use as a subroutine of the subtyping algorithm, can only produce
+unification constraints. A resolution of unification constraints is simpler than
+that of a general constraint set. This way, this separation allows us to better
+decompose the algorithm's structure, thus simplifying the inductive proofs.
 
@@
 
-\paragraph{Constraint Contexts}
-When one instantiates an algorithmic variable, 
-they may only use type variables available in its scope.
-As such, each algorithmic variable must remember the context at the moment when 
-it was introduced. In our algorithm, this information is represented by
-a \emph{constraint context} $[[Θ]]$---a set of pairs associating 
-algorithmic variables and declarative contexts.
-
 \paragraph{Auxiliary Functions}
-We define $[[dom(SC)]]$---a domain of a constraint set $[[SC]]$ as a set of algorithmic variables
-that it restricts. Similarly, we define $[[dom(Θ)]]$---a domain of constraint context
-as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
-We write $[[Θ(α̂±)]]$ to denote the context associated with $[[α̂±]]$ in $[[Θ]]$.
+We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of algorithmic variables
+that it restricts. Similarly, we define $[[dom(Θ)]]$---the domain of instantiation context
+---as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
+We write $[[Θ(α̂±)]]$ to denote the declarative context associated with $[[α̂±]]$ in $[[Θ]]$.
 
@@
   arrows denote the invocation of one procedure from another.
-  The label <<$\ottkw{nf}$>> annotating arrows means that the calling
-  procedure normalizes the input before passing it to the callee.
+
+  Some of the procedures (in particular, the unification and the
+  anti-unification) assume that the input types are normalized. Therefore, we
+  call the normalization procedure before invoking them, indicating this by the
+  `$\ottkw{nf}$' annotation on the arrows in \cref{fig:alg-subtyping-graph}.
+  Alternatively, one could normalize the input types in the very beginning,
+  as the normalization is preserved throughout the algorithm. However, we
+  delay the normalization to the places where it is required to show that
+  normalization is needed only at these stages to maintain consistent
+  invariants.
 
@@
 As an input, the subtyping algorithm takes
-a type context $[[Γ]]$, a constraint context $[[Θ]]$,
+a type context $[[Γ]]$, an instantiation context $[[Θ]]$,
 and two types of the corresponding polarity:
@@
 $[[puas]]$ or $[[nuas]]$,
-puts them into the constraint context $[[Θ]]$
+puts them into the instantiation context $[[Θ]]$
 (specifying that they must be instantiated in the extended context
@@
 constraints $[[SC]]$. As the output, the algorithm removes the freshly
-introduced algorithmic variables from the constraint context, This operation is
+introduced algorithmic variables from the instantiation context, This operation is
 sound: it is guaranteed that $[[SC]]$ always has a solution, but the specific
@@
 The resulting constraints are merged
-(using a special procedure defined in \cref{sec:constraint-merge})
+using a special procedure defined in \cref{sec:constraint-merge}
 and returned as the output.
 
-\paragraph{Algorithmic Variable}
-If one of the sides of the subtyping is a unification variable,
-the algorithm must create a new constraint. 
-Because the right-hand side of the subtyping is always declarative,
-it is only the left-hand side that can be a unification variable.
-Moreover, another invariant we preserve prevents the negative
-algorithmic variables from occurring in types during the
- negative subtyping algorithm. It means
-that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]] [[iP]]$,
-which is covered by \ruleref{\ottdruleAPUVarLabel}.
-
-The potential problem here is that the type $[[iP]]$
-might be not well-formed in the context required for $[[α̂⁺]]$ by $[[Θ]]$, 
-because this context might be smaller than the current context $[[Γ]]$.
-As we wish the resulting constraint set to be sound \wrt $[[Θ]]$,
-we cannot simply put $[[α̂⁺ :≥ iP]]$ into the output. 
-Prior to that, we update the type $[[iP]]$ to its lowest supertype $[[iQ]]$
-well-formed in $[[Θ(α̂⁺)]]$. It is done by the \emph{upgrade} procedure,
-which we discuss in detail in \cref{sec:lub}.
+\paragraph{Algorithmic variables}
+If one of the sides of the subtyping is a unification variable, the algorithm
+creates a new constraint. Because the right-hand side of the subtyping is always
+declarative, it is only the left-hand side that can be a unification variable.
+Moreover, another invariant we preserve prevents the negative algorithmic
+variables from occurring in types during the negative subtyping algorithm. It
+means that the only possible form of the subtyping here is $[[α̂⁺]] [[≥]]
+[[iP]]$, which is covered by \ruleref{\ottdruleAPUVarLabel}.
+
+The potential problem here is that the type $[[iP]]$ might be not well-formed in
+the instantiation context required for $[[α̂⁺]]$ by $[[Θ]]$ because this
+context might be smaller than the current context $[[Γ]]$. As we wish the
+resulting constraint set to be sound \wrt $[[Θ]]$, we cannot simply put 
+$[[α̂⁺ :≥ iP]]$ into the output. Prior to that, we update the type $[[iP]]$ to its
+lowest supertype $[[iQ]]$ well-formed in $[[Θ(α̂⁺)]]$. It is done by the
+\emph{upgrade} procedure, which we discuss in detail in \cref{sec:lub}.
 
@@
 
-As an input the unification context 
-takes a type context $[[Γ]]$, a constraint context $[[Θ]]$,
-and two types of the required polarity:
-$[[uN]]$ and  $[[iM]]$ for the negative unification, and
-$[[uP]]$ and  $[[iQ]]$ for the positive unification.
-It is assumed that only the left-hand side type may contain algorithmic variables,
-this way, the left-hand side is well-formed as an algorithmic type in $[[Γ]]$ and 
-$[[Θ]]$, whereas the right-hand side is well-formed declaratively in $[[Γ]]$.
-
-Since only the left-hand side may contain algorithmic variables,
-that the unification instantiates, we could have called this procedure \emph{matching}.
-However, in \cref{sec:weakening-invariant}, we will discuss several modifications of the 
-type system, where this invariant is not preserved, and therefore, this procedure
-requires general first-order pattern unification \cite{miller1991:unification}.
-
-As the output, the unification algorithm returns the weakest
-set of unification constraints $[[UC]]$ such that any instantiation 
-satisfying these constraints unifies the input types.
+As an input, the unification context takes a type context $[[Γ]]$, an
+instantiation context $[[Θ]]$, and two types of the required polarity: $[[uN]]$
+and  $[[iM]]$ for the negative unification, and $[[uP]]$ and  $[[iQ]]$ for the
+positive unification. It is assumed that only the left-hand side type may
+contain algorithmic variables. This way, the left-hand side is well-formed as an
+algorithmic type in $[[Γ]]$ and $[[Θ]]$, whereas the right-hand side is
+well-formed declaratively in $[[Γ]]$.
+
+Since only the left-hand side may contain algorithmic variables that the unification instantiates, we could have called this procedure \emph{matching}.
+However, in \cref{sec:weakening-invariant}, we will discuss several
+modifications of the type system, where this invariant is not preserved, and
+therefore, this procedure requires general first-order pattern unification
+\cite{miller1991:unification}.
+
+As the output, the unification algorithm returns \emph{the weakest} set of
+unification constraints $[[UC]]$ such that \emph{any} instantiation satisfying
+these constraints unifies the input types.
 
@@
   The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
-  require the input types to be formed by the same shift constructor. 
+  require the two input types to be formed by the same shift constructor. 
   They remove this constructor, unify the types recursively, and return the resulting
@@
   Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
-  First, it unifies the argument types and their result types recursively.
+  First, it unifies their argument types and their result types recursively.
   Then it merges the resulting constraints using the procedure described in \cref{sec:constraint-merge}.
 
-  Notice that the resulting constraints can only have \emph{unification} entries.
-  It means that they can be merged in a simpler way than general constraints.
-  In particular, the merging procedure does not call any of the subtyping subroutines,
-  but rather simply checks the matching constraint entries for equality.
-
-\paragraph{Algorithmic Variable}
+  Notice that the resulting constraints can only have \emph{unification}
+  entries. It means that they can be merged in a simpler way than general
+  constraints. In particular, the merging procedure does not call any of the
+  subtyping subroutines but rather simply checks the matching constraint entries
+  for equality.
+
+\paragraph{Algorithmic variable}
   Finally, if the left-hand side of the unification is an algorithmic variable,
@@
   It simply checks that the right-hand side type is well-formed in the required
-  constraint context, and returns a newly created constraint restricting the variable
+  instantiation context, and returns a newly created constraint restricting the variable
   to be equal to the right-hand side type.
@@
 \vspace{\baselineskip}
-As one can see, the unification procedure is standard, 
-except that it makes sure that the resulting instantiations agree with the input
-constraint context $[[Θ]]$.  As a subroutine, the unification algorithm only uses the (unification) 
-constraint merge procedure and the well-formedness checking.
+As one can see, the unification procedure is standard; the only peculiarity
+(although it is common for type inference) is that it makes sure that the resulting
+instantiations agree with the input instantiation context $[[Θ]]$.  As a
+subroutine, the unification algorithm only uses the (unification) constraint
+merge procedure and the well-formedness checking.
 
@@
 same algorithmic variable---we call such entries \emph{matching}.
+
 The matching entries must be combined into \emph{one} constraint entry, that
-would represent their conjunction. 
-This way, to merge two constraint sets, we unite the entries of two
-sets, and then merge the matching pairs.
-
-\paragraph{Merging Matching Constraint Entreies}
+would represent their conjunction. This way, to merge two constraint sets, we
+unite the entries of two sets and then merge the matching pairs.
+
+\paragraph{Merging matching constraint entries}
 Two \emph{matching} entries formed in the same context $[[Γ]]$ 
@@
 Suppose that $[[scE1]]$ and $[[scE2]]$ are input entries. 
-The result of the merge $[[scE1 & scE2]]$ must be the 
-weakest entry which implies both  $[[scE1]]$ and $[[scE2]]$.
+The result of the merge $[[scE1 & scE2]]$ is 
+\emph{the weakest entry which implies both $[[scE1]]$ and $[[scE2]]$}.
 
@@
 
-Suppose that one of the input entries, say $[[scE1]]$, is a unification 
+Suppose that one of the input entries, say $[[scE1]]$, is a \emph{unification}
 constraint entry. Then the resulting entry $[[scE1]]$ must coincide with it 
 (up-to-equivalence), and thus, it is only required to check that $[[scE2]]$ 
-is implied by $[[scE1]]$.
+is implied by $[[scE1]]$. We consider two options:
 \begin{itemize}
-  \item If $[[scE2]]$ is also a restricting entry, then the types on the right-hand side
-    of $[[scE1]]$ and $[[scE2]]$ must be equivalent,
-    as given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}.
-  \item If $[[scE2]]$ is a supertype constraint $[[α̂⁺ :≥ iP]]$,
+  \item[(i)] if $[[scE2]]$ is also a \emph{unification} entry, then the types on
+    the right-hand side of $[[scE1]]$ and $[[scE2]]$ must be equivalent, as
+    given by rules \ruleref{\ottdruleSCMEPEqEqLabel} and
+    \ruleref{\ottdruleSCMENEqEqLabel};
+  \item[(ii)] if $[[scE2]]$ is a \emph{supertype} constraint entry $[[α̂⁺ :≥ iP]]$,
     the algorithm must check that the type assigned by $[[scE1]]$ is a supertype of $[[iP]]$.
     The corresponding symmetric rules are \ruleref{\ottdruleSCMESupEqLabel} and \ruleref{\ottdruleSCMEEqSupLabel}.
+    In the premises of these rules, $[[uP]]$ is the same as $[[iP]]$ below, 
+    and $[[uQ]]$ is the same as $[[iQ]]$ below but relaxed to an algorithmic type.
 \end{itemize}
@@
 then their conjunction is $[[α̂⁺ :≥ iP ∨ iQ]]$, as given by \ruleref{\ottdruleSCMESupSupLabel}.
-The least upper bound---$[[iP ∨ iQ]]$ is the least supertype of both $[[iP]]$ and $[[iQ]]$,
+The least upper bound---$[[iP ∨ iQ]]$---is the least supertype of both $[[iP]]$ and $[[iQ]]$,
 and this way, $[[α̂⁺ :≥ iP ∨ iQ]]$ is the weakest constraint entry that implies
@@
 
-\paragraph{Merging Constraint Sets}
+\paragraph{Merging constraint sets}
   The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
@@
 together with the upgrade procedure, selecting the least supertype 
-ell-formed in a given context.
+well-formed \emph{in a given context}.
 
@@
 As an input, it takes three contexts: typing context $[[Γ]]$, a variable context $[[Φ]]$,
-and a constraint context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
+and an instantiation context $[[Θ1]]$. It also takes a head type $[[uN]]$ and 
 a list of arguments (terms) $[[args]]$ the head is applied to.
@@
 Type $[[uM]]$ may contain new algorithmic variables, and thus, 
-the judgment also returns $[[Θ2]]$---an updated constraint context
+the judgment also returns $[[Θ2]]$---an updated instantiation context
 and $[[SC]]$---a set of subtyping constraints.

### 2024-04-30 oopsla24/tex/05-proof.tex
Edit

@@
+\label{sec:proofs}
 The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.

### 2024-05-07 oopsla24/tex/01-introduction.tex
Edit: sections 1-5

@@
 
-One especially well-behaved form of partial type inference is \emph{local type
-inference}, introduced by \citet{pierce2000:local}. In this approach,
+One especially well-behaved form of partial type inference is 
+\emph{local type inference}, introduced by \citet{pierce2000:local}. In this approach,
 quantifiers in a polymorphic function are instantiated using only the type
@@
 both universal and existential quantifiers, which can both be instantiated
-impredicatively. This combination of features broke a number of the invariants
-which traditional type inference algorithms depend on, and required us to invent
-new algorithms which combine both unification and (surprisingly)
-the dual---anti-unification.
-
-In this paper, we extend the local type inference algorithm to a language with
-both universal and existential quantifiers, which can both be instantiated
 impredicatively. This extended system is referred to as \fexists. The
@@
 
-To research the type inference in impredicative polymorphic systems
-in the presence of existential types, we make the following \emph{\textbf{contributions}}.
+To research the type inference in impredicative polymorphic systems in the
+presence of existential types, we make the following
+\emph{\textbf{contributions}}.
 
-\paragraph{Type Inference Algorithm (\cref{sec:algorithm})}
-    We give a local type inference algorithm
-    which supports both first-class existential and universal quantifiers, both
-    of which can be instantiated impredicatively. To evade the undecidability
-    results surrounding type inference for \systemf, we introduce \fexists---a
-    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
-    relation which is still decidable.
+\paragraph{Declarative Type System Specification (\cref{sec:declarative-system,sec:proofs})} 
+    We devise \fexists---a second-order polymorphic type system that supports first-class
+    existential and universal quantifiers, both of which can be instantiated
+    impredicatively. To evade the undecidability results surrounding subtyping
+    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
+    use \emph{polarization} of types and terms to formulate the restrictions
+    that make typing decidable.
 
-\paragraph{Declarative Specification (\cref{sec:declarative-system,sec:proofs})}
-    We give a declarative type system (again, based on \CBPV) to
-    serve as a specification of our algorithm, and we prove our algorithm is
-    sound and complete with respect to the declarative type system. The
-    specification makes it easy to see that all type applications (for
-    $\forall$-elimination) and all packs (for $\exists$-introdution) are
-    inferred.
+    We give a \emph{declarative} specification of local type inference in this
+    system (\ie the typing information is only available at a limited scope)
+    based on \emph{subtyping} relation. The specification makes it easy to see
+    what is and what is not inferrable in the system.
 
-\paragraph{Employment of the Anti-unification (\cref{sec:antiunification})}
-    The introduction of impredicative existentials breaks some of the
-    fundamental invariants of HM-style type inference. As a result, it needs to
-    mix unification with \emph{anti-unification}---a dual operation which has
-    been used in the literature on term rewriting systems, but which has never
-    appeared in the context of polymorphic type inference.
+    To show that our restrictions do not unduly limit the expressiveness of the
+    language, we prove that every \fexists term can be embedded into \systemf
+    and vice versa.
 
-\paragraph{Extrapolation of the Inference Framework to Other Type Systems
-          (\cref{sec:extensions})}
-    We explore the design space further to show how the
-    same type inference scheme can be applied to
-    work with different type systems. In particular,
-    \begin{enumerate*}
-      \item[(i)] in combination with \emph{bidirectional} typechecking---the 
-        approach used in the original local type inference paper of 
-        \citet{pierce2000:local} that minimizes the number of needed annotations;
-      \item[(ii)] in systems with different subtyping rules, such as  
-        the one used in \citet{zhao22:elementary}---a system that
-        permits explicit type applications;
-      \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
-        That is, one can only instantiate a quantifier with a 
-        type that is a subtype of a given bound.
-    \end{enumerate*}
+\paragraph{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
+  inference algorithm implementing the declarative specification. Because our
+  system supports both existential and universal quantifiers, our implementation
+  of the subtyping relation requires the use of \emph{anti-unification}---a dual
+  to unification that finds the most specific generalization rather than the
+  most general unifier. To our knowledge, anit-unification has never been used
+  in the context of polymorphic type inference.
 
-\ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
+  The type inference algorithm itself is a purely local one, in the style of
+  \citet{pierce2000:local}. The locality can be seen from the fact that 
+  none of the unification variables introduced as part of type inference
+  escape the scope of a single function application.
+
+\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
+  We prove that our algorithm is terminating and both sound and complete with
+  respect to the declarative specification. This proof of soundness and completeness
+  of the algorithm required us to introduce an intermediate system in-between
+  the declarative and the algorithmic systems, and furthermore, the
+  soundness and completeness proofs are mutually recursive with each other.
+
+% \paragraph{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
+%     We explore the design space further to show how the
+%     same type inference scheme can be applied to
+%     work with different type systems. In particular,
+%     \begin{enumerate*}
+%       \item[(i)] in combination with \emph{bidirectional} typechecking---the 
+%         approach used in the original local type inference paper of 
+%         \citet{pierce2000:local} that minimizes the number of needed annotations;
+%       \item[(ii)] in systems with different subtyping rules, such as  
+%         the one used in \citet{zhao22:elementary}---a system that
+%         permits explicit type applications;
+%       \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
+%         That is, one can only instantiate a quantifier with a 
+%         type that is a subtype of a given bound.
+%     \end{enumerate*}
+
+% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
 \paragraph{\textbf{Structure of this Paper}}

### 2024-05-07 oopsla24/tex/03-declarative-system.tex
Edit: sections 1-5

@@
 
+\paragraph{Let binders}
+  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
+  binders: we infer the type of the bound value and continue the typing of the
+  computation in the extended context.
+
+  Rule \ruleref{\ottdruleDTCVarLetLabel} associates a variable with a computation.
+  The inferred type for this computation ($[[iM]]$) must be castable to a
+  shifted positive type $[[↑iP]]$, with $[[iP]]$ then assigned to the bound
+  variable $[[x]]$ to type the let binder's body. Like all annotated
+  constructors, we also verify the annotation type $[[iP]]$'s well-formedness.
+
 \paragraph{Applicative let binders}
@@
   This check is done by invoking the subtyping judgment
-  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
-  This judgment is more restrictive than checking bare 
-  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
-  to make the algorithm complete as it allows us to preserve
-  certain invariants throughout the algorithm to ensure
-  the decidability of the produced unification task.
-  In \cref{sec:proof-subtyping}, we discuss these invariants in
-  more detail, and in \cref{sec:weakening-invariant} we suggest 
-  a way to relax them.
+  $[[Γ ⊢ iM' ≤ ↑iP]]$.
+  % This judgment is more restrictive than checking bare 
+  % $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
+  % to make the algorithm complete as it allows us to preserve
+  % certain invariants throughout the algorithm to ensure
+  % the decidability of the produced unification task.
+  % In \cref{sec:proof-subtyping}, we discuss these invariants in
+  % more detail, and in \cref{sec:weakening-invariant} we suggest 
+  % a way to relax them.
 

### 2024-05-07 oopsla24/tex/04-algorithm.tex
Edit: sections 1-5

@@
 
-We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
-normal variables $[[α⁺]]$, $[[β⁻]]$.
-In a few places, we replace the quantified variables $[[pas]]$ 
-with their algorithmic counterpart $[[puas]]$.
+We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish
+them from normal variables $[[α⁺]]$, $[[β⁻]]$. In a few places, we replace the
+quantified variables $[[pas]]$ with their algorithmic counterpart $[[puas]]$.
 The procedure of replacing declarative variables with algorithmic ones we call
-\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$.
+\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$. The
+converse operation---\emph{dealgorithmization}---is denoted as $[[ nas/nuas ]]$ and 
+is used in the least upper bound procedure (\cref{sec:lub}).
 
@@
 
+\paragraph{Algorithmic Substitution}
+We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
+variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
+specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
+in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
+$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
+is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
+acts as the identity on the variables not in $[[Ξ]]$.
+
 \paragraph{Algorithmic Normalization}
@@
 \paragraph{Auxiliary Functions}
-We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of algorithmic variables
-that it restricts. Similarly, we define $[[dom(Θ)]]$---the domain of instantiation context
----as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
-We write $[[Θ(α̂±)]]$ to denote the declarative context associated with $[[α̂±]]$ in $[[Θ]]$.
+\begin{itemize}
+  \item
+    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
+    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
+    domain of instantiation context---as a set of algorithmic variables that
+    $[[Θ]]$ associates with their contexts.  
+  \item We write $[[Θ(α̂±)]]$ to denote the
+    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
+  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
+    $[[uP]]$ respectively.
+  \item We write $[[Γ ⊢ Θ]]$ to denote that each
+    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
+    a subcontext (subset) of $[[Γ]]$.
+\end{itemize}
+
+\paragraph{Constraint well-formedness and satisfaction}
+Suppose that $[[Θ]]$ is an instantiation context. 
+We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
+(denoted as $[[Θ ⊢ SC]]$) 
+if for every every constrain entry $[[scE ∊ SC]]$
+associating a variable $[[α̂±]]$ with a type $[[iP]]$,
+this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.
+
+Substitution $[[uσ]]$ satisfies a constraint \emph{entry} $[[scE]]$ restricting
+algorithmic variable $[[α̂±]]$ if
+$[[ [σ]α̂± ]]$ can be validly substituted for $[[α̂±]]$ in $[[scE]]$ 
+(so that the corresponding equivalence or subtyping holds).
+
+Substitution $[[uσ]]$ satisfies a constraint \emph{set}
+if $[[Θ ⊢ SC]]$ and \emph{each entry} of $[[SC]]$ is satisfied by $[[uσ]]$.
 
@@
   `$\ottkw{nf}$' annotation on the arrows in \cref{fig:alg-subtyping-graph}.
-  Alternatively, one could normalize the input types in the very beginning,
-  as the normalization is preserved throughout the algorithm. However, we
+  Alternatively, one could normalize the input types in the very beginning
+  and preserve the normalization throughout the algorithm. However, we
   delay the normalization to the places where it is required to show that
@@
 As output, it returns $[[(Ξ, uP, aus1, aus2)]]$, where important for us is
-$[[uP]]$---the pattern and $[[Ξ]]$---the set of `holes' represented by negative algorithmic variables.
+$[[uP]]$---the pattern---and $[[Ξ]]$--the set of `holes' represented by negative algorithmic variables.
 We discuss the anti-unification algorithm in detail in the following section.
@@
 
-The anti-unification algorithm \cite{plotkin1970:generalization,reynolds1970:transform}, 
-is a procedure dual to unification
-For two given (potentially different) expressions, 
-it finds the most specific generalizer---the most detailed pattern that 
-matches both of the input expressions. As evidence, it can also return 
-two substitutions that instantiate the `holes' of the pattern to the input expressions.
+The anti-unification algorithm
+\cite{plotkin1970:generalization,reynolds1970:transform}, is a procedure dual to
+unification. For two given (potentially different) expressions, it finds the most
+specific generalizer---the most detailed pattern that matches both of the input
+expressions. As evidence, it can also return two substitutions that instantiate
+the `holes' of the pattern to the input expressions.
 
 In our case, we have to be more demanding on the anti-unification algorithm.
-Since we use it to construct an existential type, whose (negative) quantified variables
-can only be instantiated with negative types, we must make sure that the pattern 
-has `holes' only at negative positions. Moreover, we must make sure that
-the resulting substitutions for the `holes' are well-formed in the initial contest, 
-and do not contain variables bound later. For example, 
-the anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and
-$[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$
-must result in a `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, 
-with a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$.
-But it cannot be more specific such as
-$[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole cannot be positive)
-or $[[∀β⁺.γ̂⁻]]$ 
-(since the instantiation cannot capture the bound variable $[[β⁺]]$).
+Since we use it to construct an existential type, whose (negative) quantified
+variables can only be instantiated with negative types, we must make sure that
+the pattern has `holes' only at negative positions. Moreover, we must make sure
+that the resulting substitutions for the `holes' are well-formed in the context
+from the past---at the moment when the corresponding polymorphic variables were
+introduced---and do not contain variables bound later. For example, the
+anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and $[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$ is a
+singleton `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, with
+a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$. But it
+\emph{cannot} be more specific such as $[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole
+cannot be positive) or $[[∀β⁺.γ̂⁻]]$ (since \emph{the instantiation of $~[[γ̂⁻]]$ cannot capture the
+bound variable $[[β⁺]]$}).
 
@@
 
-  First, as mentioned earlier, the freshly created `hole' has a name
-  that is uniquely defined by the pair of input types.
-  It is ensured by the following invariant:
-  all the `holes' in the algorithm have name $[[α̂⁻]]$ 
-  indexed by the pair of negative types it is mapped to.
-  This way, the returning set of `holes' is a singleton set
-  $\{ [[ â⁻_{iN, iM} ]]\}$; the resulting 
-  pattern is the `hole' $[[â⁻_{iN, iM}]]$,
-  and the mappings simply send it to the corresponding types:
-  $[[â⁻_{iN, iM} ↦ iN]]$ and $[[â⁻_{iN, iM} ↦ iM]]$.
-
-  Second, this rule is only applicable to negative types, 
-  moreover, the input types are checked to be well-formed 
-  in the outer context $[[Γ]]$. This is required by the 
-  usage of anti-unification: we 
-  call it to build an existential type that would be 
-  an upper bound of two input types via abstracting
-  some of their subexpressions under existential quantifiers.
-  The existentials quantify over \emph{negative} variables,
-  and they must be instantiated in the context
+  First, as mentioned earlier, the freshly created `hole' has a name that is
+  uniquely defined by the pair of input types. It is ensured by the following
+  invariant: all the `holes' in the algorithm have name $[[α̂⁻]]$ indexed by the
+  pair of negative types it is mapped to. This way, the returning set of `holes'
+  is a singleton set $\{ [[ â⁻_{iN, iM} ]]\}$; the resulting pattern is the
+  `hole' $[[â⁻_{iN, iM}]]$, and the mappings simply send it to the
+  corresponding types: $[[â⁻_{iN, iM} ↦ iN]]$ and $[[â⁻_{iN, iM} ↦ iM]]$.
+
+  Second, this rule is only applicable to negative types; moreover, the input
+  types are checked to be well-formed in the outer context $[[Γ]]$. This is
+  required by the usage of anti-unification: we call it to build an existential
+  type that would be an upper bound of two input types via abstracting some of
+  their subexpressions under existential quantifiers. The existentials quantify
+  over \emph{negative} variables, and they must be instantiated in the context
   available at that moment.
 
-  Third, the rule is only applicable if all other rules fail. 
-  Notice that it could happen even when the input types have matching constructors. 
-  For example, the generalizer of $[[↑α⁺]]$ and
-  $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$),
-  rather than $[[↑γ̂⁺]]$. This way, the algorithm must try to 
-  apply the congruent rules first, and only if they fail,
-  apply \ruleref{\ottdruleAUAULabel}.
-  This principle makes the inference system not syntax-directed:
-  it is not known a priori (before the recursive call) 
-  whether the corresponding congruent rule or rule \ruleref{\ottdruleAUAULabel} 
-  will be applied.
+  Third, the rule is only applicable if all other rules fail. Notice that it
+  could happen even when the input types have matching constructors. For
+  example, the generalizer of $[[↑α⁺]]$ and $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with
+  mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$), rather than $[[↑γ̂⁺]]$. This
+  way, the algorithm must try to apply the congruent rules first, and only if
+  they fail, apply \ruleref{\ottdruleAUAULabel}. As it is not known a priori
+  whether the congruent rules will be applicable, the implementation must use 
+  backtracking.
 
@@
 
-Finally, we present the type inference algorithm. 
-Similarly to the subtyping algorithm, it structurally corresponds to the declarative
-inference specification, meaning that most of the algorithmic rules have
-declarative counterparts, with respect to which they are sound and complete.
-
-This way, the inference algorithm also consists of three mutually recursive procedures: 
-the positive type inference, the negative type inference, and the
+Finally, we present the type inference algorithm. Similarly to the subtyping
+algorithm, it structurally corresponds to the declarative inference
+specification, meaning that most of the algorithmic rules have declarative
+counterparts, with respect to which they are sound and complete.
+
+This way, the inference algorithm also consists of three mutually recursive
+procedures: the positive type inference, the negative type inference, and the
 application type inference. As subroutines, the inference algorithm uses
-subtyping, constraint merge, and normalization.  
-The corresponding graph is shown in \cref{fig:alg-typing-graph}.
-
-\begin{figure}[h]
+subtyping, constraint merge, and minimal instantiation. The corresponding
+dependency is shown in \cref{fig:alg-typing-graph}.
+
+\begin{figure}
   \centering
@@
     \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
-    \node[right=1.2cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
+    \node[right=1.4cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
     \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
@@
 \paragraph{Let Binders}
-  Rule \ruleref{\ottdruleATVarLetLabel}
-  represents the type inference of a standard let binder.
-  It infers the type of the bound value $[[v]]$, and
-  makes a recursive call to infer the type of the body in the extended context.
+  Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
+  standard let binder. It infers the type of the bound value $[[v]]$, and makes
+  a recursive call to infer the type of the body in the extended context.
+
+  Rule \ruleref{\ottdruleATCVarLetLabel} infers a type of computational let
+  binder. It follows the corresponding declarative rule
+  \ruleref{\ottdruleDTCVarLetLabel} but uses algorithmic judgments instead of
+  declarative ones. It is worth noting that when calling the subtyping 
+  $[[Γ ; · ⊨ uM ≤ ↑iP ⫤ ·]]$, both $[[uM]]$ and $[[iP]]$ are free of algorithmic
+  variables: $[[uM]]$ is a type inferred for $[[c]]$, and $[[iP]]$ is given as
+  an annotation.
 
@@
 
-  To ensure that this step does not cause infinite recursion, 
-  we also check that the head type has at least one 
-  $[[∀]]$-quantifier. Also, to force the algorithm to
-  apply rule \ruleref{\ottdruleATEmptyAppLabel} when  
-  there are no arguments, we require $[[args ≠ ·]]$.
+  To ensure the rule precedence, we also require
+  the head type to have at least one $[[∀]]$-quantifier, 
+  and the list of arguments to be non-empty.
 
@@
   and then, calling the algorithmic subtyping,
-  finds $[[SC1]]$---the minimal constraint ensuring that 
+  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
   $[[uQ]]$ is a supertype of the type of $[[v]]$.
@@
 \emph{algorithmic} type $[[uP]]$ 
-together with a set of necessary and sufficient constraints $[[SC]]$
-that any instantiation of $[[uP]]$ must satisfy.
-This way, to guarantee principality, we must minimize the instantiation of $[[uP]]$,
-while ensuring that it satisfies the given constraints $[[SC]]$.
-
-The minimal instantiation judgment `$[[Γ ⊢ uP SC minby uσ]]$'
-means that $[[uσ]]$ is the minimum among all the substitutions satisfying $[[SC]]$,
-that is $[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
-instantiation of $[[uP]]$. The algorithm is given in \cref{fig:minimal-instantiation}.
+together with a set of \emph{necessary and sufficient} 
+constraints $[[SC]]$ that any instantiation of $[[uP]]$ must satisfy.
+This way, the principal type is the minimal instantiation of $[[uP]]$,
+obtained by a substitution satisfying the given constraints $[[SC]]$.
+
+To find this substitution, we use the minimal instantiation
+algorithm (\cref{fig:minimal-instantiation}). The judgment `$[[Γ ⊢ uP SC minby uσ]]$'
+$[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
+instantiation of $[[uP]]$ that satisfies $[[SC]]$. 
 First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
-Then it considers two cases. If the type is an algorithmic variable $[[α̂⁺]]$
-restricted by a constraint $[[(α̂⁺ :≥ iP) ∊ SC]]$,
-its minimal instantiation is the type $[[nf(iP)]]$, 
-and \ruleref{\ottdruleSINGPUvarLabel} is applied.
-Otherwise, \ruleref{\ottdruleSINGSingLabel} is applied,
-meaning that all the instantiations of the type must be equivalent.
-The algorithm ensures that by two premises:
+Then it considers two cases. 
+\begin{enumerate}
+  \item If the type $[[uP]]$ is an algorithmic variable $[[α̂⁺]]$
+    restricted by a \emph{subtyping} constraint $[[(α̂⁺ :≥ iQ) ∊ SC]]$,
+    its minimal instantiation is the type $[[nf(iQ)]]$, and 
+    \ruleref{\ottdruleSINGPUvarLabel} is applied.
+  \item Otherwise, the type $[[uP]]$ is either a shifted computation, a
+    declarative variable, an unrestricted algorithmic variable, or an
+    algorithmic variable restricted by an \emph{equivalence} constraint. In all
+    of these cases, the minimal instantiation exists if and only if there is
+    only one possible instantiation of $[[uP]]$ that satisfies $[[SC]]$.
+
+    The algorithm finds this instantiation by two premises:
     \begin{enumerate*}
-  \item[(i)] checking that all the algorithmic variables of the type are restricted by the constraint set; and
-  \item[(ii)] building the substitution that satisfies the constraint set on these variables 
-    (and simultaneously checking that such substitution is unique up-to-equivalence)
-    using the constraint singularity algorithm.
+      \item[(i)] checking that all the algorithmic variables of the type are
+      restricted by the constraint set; and
+      \item[(ii)] building the substitution that satisfies the constraint set on
+        these variables (and simultaneously checking that such substitution is
+        unique up to equivalence) using the \emph{constraint singularity algorithm}.
     \end{enumerate*}
+  \end{enumerate}
 
@@
 
-To implement the singularity algorithm, we define a partial function $[[SC
-singular with uσ]]$, taking a subtyping constraint $[[SC]]$ as an argument and
-returning a substitution $[[uσ]]$---the only solution of $[[SC]]$. The
-constraint $[[SC]]$ consists of constraint entries. Thus, we define the
-singularity using the singularity of constraint entries: each entry of $[[SC]]$
-must have a unique instantiation and the resulting substitution $[[uσ]]$ must be
-the combination of these instantiations (see \cref{fig:constraint-singularity}).
-
-\begin{figure}[h]
-  For a constraint set $[[SC]]$, we write $[[SC singular with uσ]]$ if and only if:
-  \begin{itemize}
-    \item for any \emph{positive} constraint entry $[[scE ∊ SC]]$, there exists $[[iP]]$ such that $[[scE singular with iP]]$,
-      and for the variable $[[β̂⁺]]$ restricted by $[[scE]]$, $[[ [uσ]β̂⁺ = iP ]]$;
-    \item the symmetric property holds for all \emph{negative} $[[scE ∊ SC]]$; and
-    \item for any $[[α̂± ∉ dom(SC)]]$, $[[ [uσ]α̂± = α̂± ]]$.
+\begin{figure}
+  \begin{minipage}{0.49\textwidth}
+    `$[[SC singular with uσ]]$' means:
+    \begin{itemize}[leftmargin=*]
+      \item[$+$] for any \emph{positive} constraint entry $[[scE ∊ SC]]$ restricting 
+        a variable $[[β̂⁺]]$, there exists $[[iP]]$ such that $[[scE singular with iP]]$
+        (as defined in \cref{fig:constraint-entry-singularity}),
+        and $[[ [uσ]β̂⁺ = iP ]]$; and
+      \item[$-$] the symmetric property holds for all \emph{negative} $[[scE ∊ SC]]$; and
+      \item[$\notin$] for any $[[α̂± ∉ dom(SC)]]$, $[[ [uσ]α̂± = α̂± ]]$.
     \end{itemize}
-  \caption{Constraint Singularity}
+    \caption{Singular Constraint}
     \label{fig:constraint-singularity}
-\end{figure}
-
-The singularity of constraint \emph{entries} is defined in
-\cref{fig:constraint-entry-singularity}. The \emph{equivalence} entries are
-always singular, as the only possible type satisfying them is the one given in
-the constraint itself, which is reflected in \ruleref{\ottdruleSINGNEqLabel} and
-\ruleref{\ottdruleSINGPEqLabel}. The \emph{subtyping} constraints are trickier.
-As will be discussed in \cref{sec:proof-lub-upgrade}, variables 
-(and equivalent types such as $[[∃nas.pa]]$) do not have 
-\emph{proper} supertypes, and thus, the constraints of a form 
-$[[pua :≥ ∃nas.pa]]$ are singular with the only possible normalized solution $[[pa]]$
-\ruleref{\ottdruleSINGSupVarLabel}. However, if the body of the existential type
-is guarded by a shift $[[∃nas.↓iN]]$, it is singular if and only if $[[iN]]$ is
-equivalent to some $[[αi⁻ ∊ {nas}]]$ bound by the quantifier, see the rule
-\ruleref{\ottdruleSINGSupShiftLabel}.
-
-\begin{figure}[h]
-  \hfill\\
+  \end{minipage}
+  \begin{minipage}{0.48\textwidth}
     \ottdefnSINGLabeled
@@
     \label{fig:constraint-entry-singularity}
-\end{figure}
+  \end{minipage}
+\end{figure}
+
+To implement the singularity algorithm, we define a partial function 
+`$[[SC singular with uσ]]$', taking a subtyping constraint $[[SC]]$ as an argument and
+returning a substitution $[[uσ]]$---the only solution of $[[SC]]$.
+
+The constraint $[[SC]]$ is composed of constraint entries. Therefore, we define
+singularity by combining the singularity of each constraint entry
+(\cref{fig:constraint-singularity}). In order for $[[SC]]$ to be singular, each
+entry must have a unique instantiation, and the resulting substitution $[[uσ]]$
+must be a union of these instantiations. In addition, $[[uσ]]$ must act as an
+identity on the variables not restricted by $[[SC]]$.
+
+The singularity of constraint \emph{entries} is defined in
+\cref{fig:constraint-entry-singularity}. 
+\begin{itemize}
+  \item The \emph{equivalence} entries are
+    always singular, as the only possible type satisfying them is the one given in
+    the constraint itself, which is reflected in rules \ruleref{\ottdruleSINGNEqLabel} and
+    \ruleref{\ottdruleSINGPEqLabel}. 
+  \item The \emph{subtyping} constraints are trickier.
+    As will be discussed in \cref{sec:proof-lub-upgrade}, variables (and equivalent
+    them $[[∃nas.pa]]$) do not have proper supertypes, and thus, the
+    constraints of a form $[[pua :≥ ∃nas.pa]]$ are singular with the only possible
+    normalized solution $[[pa]]$---see rule \ruleref{\ottdruleSINGSupVarLabel}. 
+  \item However, if the body of the existential type is guarded by a shift $[[∃nas.↓iN]]$, it is
+    singular if and only if $[[iN]]$ is equivalent to some $[[αi⁻ ∊ {nas}]]$ bound
+    by the quantifier---see rule \ruleref{\ottdruleSINGSupShiftLabel}.
+\end{itemize}

### 2024-05-07 oopsla24/tex/05-proof.tex
Edit: sections 1-5

@@
 \label{sec:proofs}
-The central results ensuring the correctness of the inference algorithm are its soundness and completeness with respect to the declarative specification.
-The soundness means the algorithm will always
-produce a typing \emph{allowed} by the declarative system;
-Dually, the completeness says that once a term has some type declaratively, the inference algorithm succeeds.
+The central results ensuring the correctness of the inference algorithm are its
+soundness and completeness with respect to the declarative specification. The
+soundness means the algorithm will always produce a typing \emph{allowed} by the
+declarative system; dually, the completeness says that once a term has some type
+declaratively, the inference algorithm succeeds.
 
-The formulation of soundness and completeness of \emph{application inference} is more complex.
-Both of them assume that the input head type $[[uN]]$ is free from \emph{negative}
-algorithmic variables (it is achieved by polarization invariants preserved by the inference rules).
-The soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$ is viable. 
-Specifically, that the constraint set $[[SC]]$ provides a sufficient set of restrictions that 
-a substitution $[[uσ]]$ must satisfy to ensure the \emph{declarative} inference of the output type $[[uM]]$, that is 
+The formal statements of soundness and completeness are given in the theorems
+below. Notice that the theorems also include the soundness and completeness of
+\emph{application inference} (labeled as $\bullet$), which is more complex. As
+such, let us discuss it in more detail.
+
+Both soundness and completeness of application inference assume that the input
+head type $[[uN]]$ is free from \emph{negative} algorithmic variables---it is
+achieved by polarization invariants preserved by the inference rules. The
+soundness states that the output of the algorithm---$[[uM]]$ and $[[SC]]$---is
+viable. Specifically, that the constraint set $[[SC]]$ provides a sufficient set
+of restrictions that any substitution $[[uσ]]$ must satisfy to ensure the
+\emph{declarative} inference of the output type $[[uM]]$, that is 
 $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$.
 
-The application inference completeness means that if 
-there exists a substitution $[[uσ]]$ and the resulting type $[[iM]]$
-ensuring the declarative inference $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$ then the algorithm succeeds
-and gives the most general result $[[uM0]]$ and $[[SC0]]$.
-The property of `being the most general' is specified in pt. (\ref{point:mostGeneral}).
-Intuitively, it means that for any other solution---substitution $[[uσ]]$ and the resulting type $[[iM]]$,
-if it ensures the declarative inference, then $[[uσ]]$ can be extended in a $[[SC0]]$-complying way to equate 
-$[[uM0]]$ with $[[iM]]$.
+The application inference completeness means that if there exists a substitution
+$[[uσ]]$ and the resulting type $[[iM]]$ ensuring the declarative inference
+$[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$ then the algorithm succeeds and gives the most
+general result $[[uM0]]$ and $[[SC0]]$. The property of `being the most general'
+is specified in pt. (\ref{point:mostGeneral}). Intuitively, it means that for
+any other solution---substitution $[[uσ]]$ and the resulting type $[[iM]]$, if
+it ensures the declarative inference, then $[[uσ]]$ can be extended in a
+$[[SC0]]$-complying way to equate $[[uM0]]$ with $[[iM]]$.
 
@@
             of the statement and the corresponding theorem is more involved. 
-            See the appendix for details.}
+            See the appendix (\cref{sec:appendix}) for details.}
 
@@
         \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
-            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$,
-            assuming $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
+            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$.
+            All of it under assumptions that $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
             \emph{negative} algorithmic variables.
@@
     by selecting a representative from each equivalence class.
-    This way, the correctness of normalization means that
+    This way, \emph{the correctness of normalization} means that
     checking for equivalence of two types is the same as checking for equality of their normal forms.
-    \begin{lemma*}[Normalization Correctness]
+    \begin{lemma*}[Normalization correctness]
         Assuming all types are well-formed in $[[Γ]]$, we have
@@
 
-The anti-unifier of the two types is the most specific pattern that matches against both of them. 
-In our setting, the anti-unifiers are restricted further: first, the pattern might only contain placeholders at 
-\emph{negative} positions (because eventually, the placeholders become variables abstracted by the existential quantifier);
-second, the anti-unification is parametrized with a context $[[Γ]]$, in which the pattern instantiations must be well-formed.
+The anti-unifier of the two types is the most specific pattern that matches
+against both of them. In our setting, the anti-unifiers are restricted further:
+first, the pattern might only contain `holes' at \emph{negative} positions
+(because eventually, the `holes' become variables abstracted by the
+existential quantifier); second, the anti-unification is parametrized with a
+context $[[Γ]]$, in which the pattern instantiations must be well-formed.
 
-This way, the correctness properties of the anti-unification algorithm are refined accordingly.
-The soundness of anti-unification not only ensures that the resulting pattern matches with 
-the input types, but also that the pattern instantiations are well-formed in the corresponding context, and 
-that all the `placeholder' variables are negative. In turn, completeness states that if there exists a solution 
-satisfying the soundness criteria, then the algorithm succeeds.
+This way, the correctness properties of the anti-unification algorithm are
+refined accordingly. The soundness of anti-unification not only ensures that the
+resulting pattern matches with the input types, but also that the pattern
+instantiations are well-formed in the corresponding context, and that all the
+`hole' variables are negative. In turn, completeness states that if there
+exists a solution satisfying the soundness criteria, then the algorithm
+succeeds.
 
-The correctness properties are formulated by the following lemmas. For brevity, we only provide the statements for 
-the positive case, since the negative case is symmetric.
+The correctness properties are formulated by the following lemmas. For brevity,
+we only provide the statements for the positive case, since the negative case is
+symmetric.
 
-\begin{lemma*}[Soundness of (Positive) Anti-Unification]
+\begin{lemma*}[Soundness of (positive) anti-unification]
     \label{lemma:au-soundness}
@@
 
-\begin{lemma*}[Completeness of (Positive) Anti-Unification]
+\begin{lemma*}[Completeness of (positive) anti-unification]
     \label{lemma:au-completeness}
@@
 
-The anti-unification is used to find the least upper bound (LUB).
-To guarantee that the result of the LUB algorithm is indeed the least, 
-we also need to show the corresponding property of anti-unification.
-We prove that the anti-unifier that the algorithm provides 
-is the most specific (or the most `detailed'). Specifically, 
-it means that any other sound anti-unification solution 
-can be `refined' to the result of the algorithm.  
+Notice that for the anti-unification substitution $[[aus]]$ we use a separate
+signature specifying the domain and the range. $[[Γ ; Ξ2 ⊢ aus : Ξ1]]$ means
+that $[[aus]]$ maps the `holes' (\ie algorithmic variables) from $[[Ξ1]]$
+to \emph{algorithmic} types well-formed in $[[Γ]]$ and $[[Ξ2]]$. To put it
+formally, $[[Γ ; Ξ2 ⊢ [aus]α̂⁻]]$ for any $[[α̂⁻ ∊ Ξ1]]$.
+Although in the formulation of soundness and completeness, the resulting types are declarative
+(\ie $[[Ξ2]]$ is always empty), we need the possibility of mapping the
+`holes' to types with `holes' to formulate the \emph{initiality} of
+anti-unification.
+
+The initiality shows that the anti-unifier that the algorithm provides is the
+most specific (or the most `detailed'). Specifically, it means that any other
+sound anti-unification solution can be `refined' to the result of the algorithm.  
 The `refinement' is represented as an instantiation of the anti-unifier---a 
-substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the placeholders $[[Ξ1]]$ with
-types that themselves can contain placeholders from $[[Ξ2]]$.
+substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the `holes' from $[[Ξ1]]$ with
+types that themselves can contain `holes' from $[[Ξ2]]$.
 
@@
 in the resulting tuple $[[(Ξ, uQ, aus1, aus2)]]$,
-there are no two different placeholders $[[β̂⁻]]$
+there are no two different `holes' $[[β̂⁻]]$
 mapped to the same pair of types by $[[aus1]]$ and $[[aus2]]$.
-This is important to guarantee that, for example, 
+This is used to ensure that, for example, 
 the anti-unifier of $[[↓↑Int → ↑Int]]$ and 
@@
 To preserve this property, we arrange the algorithm in such a way 
-that the name of the placeholder is determined by the types it is mapped to.
+that the name of the `hole' is determined by the types it is mapped to.
 The following lemma specifies this observation.
@@
     The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
-    The \emph{soundness} means that the returned type is indeed a supertype of the given ones;
+    Its \emph{soundness} means that the returned type is indeed a supertype of the given ones;
     the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
@@
 
-    \begin{lemma*}[Least Upper Bound Soundness]
+    \begin{lemma*}[Least Upper Bound soundness]
         For types $[[Γ ⊢ iP1]]$, and $[[Γ ⊢ iP2]]$,
@@
 
-    \begin{lemma*}[Least Upper Bound Completeness and Initiality]
+    \begin{lemma*}[Least Upper Bound completeness and initiality]
         For types $[[Γ ⊢ iP1]]$, $[[Γ ⊢ iP2]]$, and $[[Γ ⊢ iQ]]$
@@
 
-    The key observation that allows us to prove these properties is
-    the characterization of positive supertypes. 
-    The following lemma justifies the cases of the Least Upper Bound algorithm
-    (\cref{sec:lub}). In particular, it establishes
-    the correspondence between the upper bounds of shifted types $[[↓iM]]$ and
-    \emph{patterns} fitting $[[iM]]$ (represented by existential types), which 
-    explains the usage of anti-unification as a way to find a common pattern.
+    The key observation that allows us to prove these properties is the
+    characterization of positive supertypes. The following lemma justifies the
+    case analysis used in the Least Upper Bound algorithm (in \cref{sec:lub}).
+    In particular, it establishes the correspondence between the upper bounds of
+    shifted types $[[↓iM]]$ and \emph{patterns} fitting $[[iM]]$ (represented by
+    existential types), which explains the usage of anti-unification as a way to
+    find a common pattern.
 
-    \begin{lemma*}[Characterization of Normalized Supertypes]
+    \begin{lemma*}[Characterization of normalized supertypes]
         \label{lemma:char-supertypes}
@@
 
-    \begin{lemma*}[Upgrade Soundness]
+    \begin{lemma*}[Upgrade soundness]
         Assuming $[[iP]]$ is well-formed in $[[Γ = Δ, pnas]]$,
@@
         \begin{enumerate*}
-            \item $[[Δ ⊢ iQ]]$
-            \item $[[Γ ⊢ iQ ≥ iP]]$
+            \item $[[Δ ⊢ iQ]]$ and
+            \item $[[Γ ⊢ iQ ≥ iP]]$.
         \end{enumerate*}
@@
         \end{enumerate*}
-        there exists$[[iQ]]$ such that
+        there exists $[[iQ]]$ such that
         ($[[upgrade Γ ⊢ iP to Δ = iQ]]$) and $[[Δ ⊢ iQ' ≥ iQ]]$.
@@
 
-    As for other properties, the correctness of subtyping means that
-    the algorithm produces a valid result (soundness) 
-    whenever it exists (completeness).
-    In \cref{fig:alg-subtyping}, one can see that the positive and 
-    negative subtyping relations are not \emph{mututally} recursive: 
-    negative subtyping algorithm uses the positive subtyping, 
-    but not vice versa. 
-    Because of that, the inductive proofs of the \emph{positive} subtyping correctness are done independently.
+    As for other properties, the correctness of subtyping means that the
+    algorithm produces a valid result (soundness) whenever it exists
+    (completeness). In the rules defining the subtyping algorithm
+    (\cref{fig:alg-subtyping}), one can see that the positive and negative
+    subtyping relations are not \emph{mututally} recursive: negative subtyping
+    algorithm uses the positive subtyping, but not vice versa. Because of that,
+    the inductive proofs of the \emph{positive} subtyping correctness are done
+    independently.
 
-    The soundness of positive subtyping states that
-    the output constraint $[[SC]]$ provides a sufficient set of restrictions.
-    In other words, any substitution satisfying $[[SC]]$, 
-    ensures the desired declarative subtyping: $[[Γ ⊢ [uσ]uP ≥ iQ]]$.
-    Notice that the soundness requires that only the left-hand side input 
-    type ($[[uP]]$) can have algorithmic variables. This is one of the invariants of the algorithm
-    that significantly simplifies the unification and constraint resolution.
+    The soundness of positive subtyping states that the output constraint
+    $[[SC]]$ provides a sufficient set of restrictions. In other words, any
+    substitution satisfying $[[SC]]$ ensures the desired declarative subtyping:
+    $[[Γ ⊢ [uσ]uP ≥ iQ]]$. Notice that the soundness formulation assumes that
+    only the left-hand side input type ($[[uP]]$) can contain algorithmic
+    variables. This is one of the invariants of the algorithm that significantly
+    simplifies the unification and constraint resolution.
 
-\begin{lemma*}[Soundness of the Positive Subtyping]
+\begin{lemma*}[Soundness of Positive Subtyping]
     If $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$ and 
@@
     The completeness of subtyping says that if the substitution 
-    ensuring the declarative subtyping exists, then the algorithm terminates.
+    ensuring the declarative subtyping exists, then the subtyping algorithm 
+    succeeds (terminates).
 
-\begin{lemma*}[Completeness of the Positive Subtyping]
+\begin{lemma*}[Completeness of Positive Subtyping]
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
@@
 
-After the correctness properties of the positive subtyping are established,
+After the correctness properties of positive subtyping are established,
 they are used to prove the correctness of the negative subtyping.
@@
 
-\begin{lemma*}[Completeness of the Negative Subtyping]
+\begin{lemma*}[Completeness of Negative Subtyping]
     Suppose that $[[Γ ⊢ Θ]]$ and $[[Γ ⊢ iM]]$ and $[[Γ ; dom(Θ) ⊢ uN]]$
@@
 
-\paragraph{Minimal Instantiation}
 Algorithmic typing relies on the \emph{minimal instantiation} procedure.
@@
 \begin{lemma*}[Soundness of Minimal Instantiation]
+    \label{lemma:minimal-instantiation-soundness}
     Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
-    Then $[[Γ ⊢ uP SC minby uσ ]]$ implies that 
+    Then `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
     $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
@@
     and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
-    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$ holds
+    we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.
 \end{lemma*}
@@
     $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
-    succeeds, specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
+    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
 \end{lemma*}
@@
 
-\paragraph{Singularity}
-The soundness of singularity states that $[[SC singular with uσ]]$
+The soundness of \emph{singularity} states that $[[SC singular with uσ]]$
 implies that any substitution satisfying $[[SC]]$ is equivalent to $[[uσ]]$ 
@@
     \label{lemma:singularity-completeness}
-    For a given $[[Θ ⊢ SC]]$,
+    For a given set of constraints $[[Θ ⊢ SC]]$,
     suppose that all the substitutions satisfying $[[SC]]$ are equivalent
@@
     $[[Θ ⊢ uσ ≈ uσ1 : dom(SC)]]$.
-    Then\footnotemark[1] $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
+    Then $[[SC singular with uσ0]]$ for some $[[uσ0]]$.
 \end{lemma*}
@@
 Typically in our inductive proofs, the soundness is proven before completeness,
-as the completeness requires certain properties of the premise subtrees.
-However, in the case of the typing algorithm, the soundness and completeness proofs
-cannot be separated: the inductive proof of one requires another, and vice versa.
+as the completeness requires the premise subtrees to satisfy certain properties
+(which can be given by the soundness). However, in the case of the typing algorithm,
+the soundness and completeness proofs cannot be separated: the inductive proof
+of one requires another, and vice versa.
 
-The soundness proof can be viewed as a mapping from an algorithmic tree to a
-declarative one. We show that each algorithmic inference rule can
-be transformed into the corresponding declarative rule, as long as the premises
-are transformed accordingly, and apply the induction principle.
+The proof of soundness can be viewed as a mapping from an algorithmic tree to a
+declarative one. We show that each algorithmic inference rule can be transformed
+into the corresponding declarative rule, as long as the premises are transformed
+accordingly, and apply the induction principle.
 
 The soundness requires completeness in the case of
-\ruleref{\ottdruleATAppLetLabel}. To prove the soundness, 
-in other words, to transform this rule into its
-declarative counterpart \ruleref{\ottdruleDTAppLetLabel}, one needs
-to prove \emph{uniqueness} of the inferred application type $[[↑iQ]]$. 
-To show the uniqueness, we need to conclude that any other \emph{declarative} tree
-(which is given by the induction hypothesis) infers an equivalent type. 
-The only way to do so is by applying the soundness of singularity (\cref{sec:singularity-proof}), 
-however, first, the declarative tree must be converted to the algorithmic one (by completeness!).
+\ruleref{\ottdruleATAppLetLabel}. To prove the soundness, in other words, to
+transform this rule into its declarative counterpart
+\ruleref{\ottdruleDTAppLetLabel}, one needs to prove the \emph{principality} of
+the inferred application type $[[↑[uσ]uQ]]$. In other words, we need to conclude
+that any other \emph{declarative} tree infers for the application a supertype of
+$[[ [uσ]uQ ]]$.
+
+The only way to do so is by applying the soundness of minimal instantiation
+(\cref{lemma:minimal-instantiation-soundness}). However, first, the declarative
+tree must be converted to the corresponding algorithmic one (by completeness!).
 This way, the soundness and completeness proofs are mutually dependent.
@@
     \ruleref{\ottdruleATForallAppLabel} and \ruleref{\ottdruleDTForallAppLabel}
-    only reduce the quantifiers in the head type but not change the list of arguments.
+    only reduce the quantifiers in the head type but they do not change the list of arguments.
 
@@
 
-To deal with this issue, we introduce the metric on the entire 
-\emph{inference trees} rather than on judgments, and plug into this metric the
-parameter that certainly decreases in rules \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}---the number of these rules in the inference tree. We
-denote this number as $\eqNodes{T}$.  Then the final metric is defined as a pair in the following way.
+To deal with this issue, we introduce the metric on the entire \emph{inference
+trees} rather than on judgments, and plug into this metric the parameter that
+certainly decreases in rules \ruleref{\ottdruleDTNEquivLabel} and
+\ruleref{\ottdruleDTPEquivLabel}---the number of such nodes in the inference
+tree. We denote this number as $\eqNodes{T}$. Then the final metric is defined
+as a pair in the following way.
 

### 2024-05-07 oopsla24/tex/06-extensions.tex
Edit extentions

@@
 automatically instantiates the variables abstracted by $\forall$. 
-The implicit type application can be added to the declarative system by the following rule:
+The \emph{explicit} type application can be added to the declarative system by the following rule:
 $$\ottdruleDTTypeAppLabeled{}$$
 
-However, this rule alone would cause ambiguity.
-The declarative system does not fix the order of the quantifiers, 
-which means that $[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ can be inferred 
-as a type of $[[c]]$ interchangeably. But then explicit instantiation
-$([[ c[iP] ]])$ would be ambiguous, as it is unclear whether $[[α⁺]]$ or
-$[[β⁺]]$ should be instantiated with $[[iP]]$.
+However, this rule alone would cause ambiguity. The declarative system does not
+fix the order of the quantifiers, which means that $[[∀α⁺.∀β⁺.iN]]$ and
+$[[∀β⁺.∀α⁺.iN]]$ can be inferred as a type of $[[c]]$ interchangeably. But that
+would make explicit instantiation $[[ c[iP] ]]$ ambiguous, as it is unclear
+whether $[[α⁺]]$ or $[[β⁺]]$ should be instantiated with $[[iP]]$.
 
@@
 
-The drawback of this approach is that it 
-would cause the `leakage' of the internal algorithmic 
-concept of type normalization into the `surface' declarative 
+The drawback of this approach is that it adds another point where the internal
+algorithmic concept---normalization---is exposed to the `surface' declarative
 system.
@@
 
-According to these rules, if two polymorphic ($[[∀]]$ or $[[∃]]$) types are
+According to these rules, if two quantified ($[[∀]]$ or $[[∃]]$) types are
 subtypes of each other, they must have the same top-level quantifiers. Moreover,
@@
     (i) replace the normalization procedure with identity: $[[nf(iN)]] \defeq [[iN]]$, $[[nf(iP)]] \defeq [[iP]]$,
-        since the new equivalence classes consist of singletons;
+        since the new equivalence classes are singletons;
     (ii) modify the least upper bound polymorphic rule \ruleref{\ottdruleLUBExistsLabel}
@@
 
-\paragraph{Solution 3: Labeled Quantifiers}
+\paragraph{Solution 3: labeled quantifiers}
 
-A compromise solution that resolves the ambiguity of 
-explicit the instantiation without fixing the order of
-quantifiers is using \emph{labeled} quantifiers.
+A compromise solution that resolves the ambiguity of explicit instantiation
+without fixing the order of quantifiers is by using \emph{labeled} quantifiers.
 Let us discuss how to introduce them to the \emph{negative} part of the system,
@@
 
-The subtyping rule allows the quantifiers to have an arbitrary order. 
+The subtyping rule permits an arbitrary order of the quantifiers.
 However, similarly to the elementary subtyping, 
@@
 
-\subsection{Weakening of the Subtyping Invariant}
+\subsection{Weakening of the subtyping invariants}
 \label{sec:weakening-invariant}
@@
 Moreover, this extension also increases the expressiveness of the \emph{type
-inference}. Namely, when inferring a type of an annotated let binding 
-$[[let x:iP = v(args); c]]$, we require $[[↑iP]]$ to be a supertype of 
-$[[↑iQ]]$---the resulting type of the application $[[v]]([[args]])$. 
-Now, it is equivalent to the requirement that $[[iP]]$ is a supertype of $[[iQ]]$
+inference}. Namely, when inferring a type of an annotated let binding $[[let
+x:iP = v(args); c]]$, we require $[[↑iP]]$ to be a supertype of $[[↑iQ]]$---the
+resulting type of the application $[[v]]([[args]])$. In the relaxed system, it
+can be replaced with the requirement that $[[iP]]$ is a supertype of $[[iQ]]$
 (without the up-shift), which is more permissive.
 
-In addition, the enhancement of the unification algorithm described above
+In addition, the refinement of the unification algorithm described above
 makes possible another improvement to the system---bounded quantification.
@@
 For brevity, we do not discuss in detail the combination of bounded and
-unbounded quantification in one system. However, this update is straightforward:
-the unbounded quantifiers are treated as the weakest bound, which is trivially
+unbounded quantification in one system. We believe this update is straightforward:
+the unbounded quantifiers are treated as the weakest constraints, which are trivially
 satisfied. Analogously, it is possible to extend the system with \emph{lower}
@@
 in typing, in particular, the type of a lambda expression cannot be inferred 
-from the context it is used in: $[[((λ x . return x) : (Int → ↑Int))]]$
-is not allowed.
+from the context it is used in: the annotated $[[((λ x . return x) : (Int → ↑Int))]]$
+does not infer $[[Int → ↑Int]]$.
 
@@
     For brevity, we omit the details of the bidirectional algorithm.
-    However, let us discuss one especially important rule---arrow application inference. 
+    Let us discuss one especially important rule---arrow application inference. 
     An intuitive way to bidirectionalize this rule is the following:
-    $$ \ottdruleATArrowAppBidirLabeled{} $$\\
+    $$ \ottdruleATArrowAppBidirLabeled{} $$
     % \paragraph{Reaching Undecidability}
-    However, this rule is too allowing. 
+    However, this rule is too permissive. 
     The judgement $[[Γ ; Φ ⊨ v ⇐ uQ ⫤ SC1]]$
@@
     We believe that the relaxation of this invariant brings the 
-    constraint resolution too close to the second-order pattern 
+    constraint resolution too close to second-order pattern 
     unification, which is undecidable \cite{goldfarb81:undecidability}.
@@
 
-    The constraints such as $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ are not merely hypothetical. 
+    Constraints such as $[[(α̂⁺ :≥ ↓↑α̂⁺)]]$ are not merely hypothetical. 
     It occurs, for example, when we infer the type of the following application:

### 2024-05-07 oopsla24/tex/03-declarative-system.tex
edit (misc)

@@
 form, the subtyping relation would become equivalent to the standard subtyping
-of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, after
-certain changes \ruleref{\ottdruleDOneShiftULabel} can be relaxed to the
-covariant form, thereby increasing the expressiveness of the system. These
-changes are discussed in \cref{sec:weakening-invariant}.
+of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, it is
+suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
+relaxed to the covariant form, thereby increasing the expressiveness of the
+system. These changes are discussed in \cref{sec:weakening-invariant}.
 

### 2024-05-07 oopsla24/tex/04-algorithm.tex
edit (misc)

@@
 
+\paragraph{Equivalent Substitutions}
+In the proofs of the algorithm correctness, 
+we often state or require that two substitutions are equivalent
+on a given set of algorithmic variables.
+We denote it as $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$ meaning that
+for each $[[α̂±]]$ in $[[Ξ]]$, substitutions $[[uσ]]$ and $[[uσ']]$
+map $[[α̂±]]$ to types equivalent in the corresponding context:
+$[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂± ]]$.
+
 \paragraph{Constraint well-formedness and satisfaction}
@@
 
-Substitution $[[uσ]]$ satisfies a constraint \emph{set}
+Substitution $[[uσ]]$ satisfies a constraint \emph{set} $[[SC]]$
 if $[[Θ ⊢ SC]]$ and \emph{each entry} of $[[SC]]$ is satisfied by $[[uσ]]$.

### 2024-05-08 oopsla24/tex/04-algorithm.tex
edit

@@
     normalized solution $[[pa]]$---see rule \ruleref{\ottdruleSINGSupVarLabel}. 
-  \item However, if the body of the existential type is guarded by a shift $[[∃nas.↓iN]]$, it is
-    singular if and only if $[[iN]]$ is equivalent to some $[[αi⁻ ∊ {nas}]]$ bound
-    by the quantifier---see rule \ruleref{\ottdruleSINGSupShiftLabel}.
+  \item However, if the body of the existential type is guarded by a shift
+    $[[∃nas.↓iN]]$, it is singular if and only if $[[iN]]$ is equivalent to some
+    $[[αi⁻ ∊ {nas}]]$ bound by the quantifier---see rule
+    \ruleref{\ottdruleSINGSupShiftLabel}. The completeness of this criterion is
+    justified by the fact that if $[[iN]]$ is \emph{not} equivalent to any
+    $[[αi⁻]]$, then there are two non-equivalent solutions of constraint $[[(pua :≥
+    ∃nas.↓iN)]]$: the trivial $[[∃nas.↓iN]]$ and $[[∃nas.↓a1⁻]]$ (which is a
+    supertype of $[[∃nas.↓iN]]$ since $[[a1⁻]]$ can be instantiated to
+    $[[iN]]$).
+
 \end{itemize}

### 2024-05-09 paper/tex/01-introduction.tex
amend

--- 2024-05-07 oopsla24/tex/01-introduction.tex
@@
 
-\paragraph{Declarative Type System Specification (\cref{sec:declarative-system,sec:proofs})} 
+\paragraph{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
     We devise \fexists---a second-order polymorphic type system that supports first-class
@@
   soundness and completeness proofs are mutually recursive with each other.
+  The central theorems and the key proof ideas are formulated in \cref{sec:proofs}, 
+  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix}).
 
@@
 % \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
-\paragraph{\textbf{Structure of this Paper}}
-In \cref{sec:overview}, we discuss the language of the \fexists system, providing
-examples, outlining limitations, and summarizing the main ideas behind the
-algorithm. \Cref{sec:declarative-system} presents a formal definition of the
-declarative specification and its key properties. In \cref{sec:algorithm}, the
-type inference algorithm is introduced, detailing a hierarchy of subroutines;
-upon perusing this section, one could \emph{implement} the type inference
-algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
-explaining its structure and how the declarative system's features make it
-decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
-work directions. Finally, \cref{appendix} contains comprehensive formal
-definitions and proofs of lemmas and theorems used in the paper; in instances
-where the full definitions might not be of immediate interest, we direct the
-reader to the appendix.
+% \paragraph{\textbf{Structure of this Paper}}
+% In \cref{sec:overview}, we discuss the language of the \fexists system, providing
+% examples, outlining limitations, and summarizing the main ideas behind the
+% algorithm. \Cref{sec:declarative-system} presents a formal definition of the
+% declarative specification and its key properties. In \cref{sec:algorithm}, the
+% type inference algorithm is introduced, detailing a hierarchy of subroutines;
+% upon perusing this section, one could \emph{implement} the type inference
+% algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
+% explaining its structure and how the declarative system's features make it
+% decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
+% work directions. Finally, \cref{appendix} contains comprehensive formal
+% definitions and proofs of lemmas and theorems used in the paper; in instances
+% where the full definitions might not be of immediate interest, we direct the
+% reader to the appendix.
 

### 2024-05-09 paper/tex/02-overview.tex
amend

--- 2024-04-30 oopsla24/tex/02-overview.tex
@@
 To check whether one type is a subtype of another, the algorithm introduces
-\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)--- the
+\emph{algorithmic} type variables ($[[α̂±]], [[β̂±]], \dots$)---the
 `placeholders' representing the polymorphic variables whose instantiation is

### 2024-05-09 paper/tex/03-declarative-system.tex
amend

--- 2024-05-07 oopsla24/tex/03-declarative-system.tex
@@
 
-The normalization algorithm is shown in \cref{fig:type-nf}.
-The steps (i-ii) are implemented by the \emph{ordering} function
-`$[[ord varset in iN]]$' and `$[[ord varset in iP]]$'.
-For a set of variables $[[varset]]$, and a type, it 
-returns a list of variables from $[[varset]]$ that occur in the type in 
-the \emph{order of their first occurrence}.
-For brevity, we omit the formal definition of ordering referring the reader to the appendix (\cref{sec:ordering}).
+The normalization algorithm is shown in \cref{fig:type-nf}. The steps (i-ii) are
+implemented by the \emph{ordering} function `$[[ord varset in iN]]$' and `$[[ord
+varset in iP]]$'. For a set of variables $[[varset]]$, and a type, it returns a
+list of variables from $[[varset]]$ that occur in the type in the \emph{order of
+their first occurrence}. For brevity, we omit the formal definition of ordering
+referring the reader to the appendix (\cref{appendix}).
 
@@
 
-The type polarization (\cref{fig:polarization}) is more involved, as there are multiple
-ways to polarize a type: for instance, a variable $[[α]]$ can be polarized to $[[α⁺]]$ or $[[α⁻]]$.
-The choice of polarization affects the execution strategy of the program.
-\ilyam{How?}
-We chose the positive polarization: every \systemf type is translated to a \emph{positive} type of \fexists.
-This choice makes it smoother to lift the translation to the term level, which will be disused next.
+The type polarization (\cref{fig:polarization}) is more involved, as there are
+multiple ways to polarize a type: for instance, a variable $[[α]]$ can be
+polarized to $[[α⁺]]$ or $[[α⁻]]$. The choice of polarization affects the
+execution strategy of the program. 
+In particular, a functional type can be represented in either positive (thunked, call-by-name) way:
+$[[↓(iP → ↑iQ)]]$ or negative (call-by-value): $[[↓iN → iM]]$.
+We chose the positive polarization: every \systemf type is translated to a
+\emph{positive} type of \fexists. This choice makes it smoother to lift the
+translation to the term level, which will be discussed next.
 
@@
 the elaboration in such a way that it preserves the soundness invariants given
-in \cref{lemma:translation-soundness}.
+in \cref{tab:elaboration-soundness}.
 
@@
   the elaboration judgment on the left implies the typing judgment on the right}
+  \label{tab:elaboration-soundness}
 \end{table}

### 2024-05-09 paper/tex/06-extensions.tex
amend

--- 2024-05-07 oopsla24/tex/06-extensions.tex
@@
 it is possible to smoothly extend the \emph{elementary}
-version of the system with bounded quantification.
+version of the system 
+(the second solution discussed in \cref{sec:explicit-type-application})
+ with bounded quantification.
 In particular, we can add upper bounds to polymorphic $[[∀]]$-quantifiers:

### 2024-06-11 paper/tex/01-introduction.tex
corrections

@@
 DHM deals with this choice using \emph{unification}. Whenever we would have had
-to introduce a particular concrete type in the specification, the DHM algorithm
+to guess a particular concrete type in the specification, the DHM algorithm
 introduces a \emph{unification variable}, and incrementally instantiates this variable
@@
 have been studied much less frequently in the literature. The most widely-used
-algorithm for handling existentials is the algorithm of Odersky and Laufer
+algorithm for handling existentials is the algorithm of Odersky and L{\"a}ufer
 \cite{laufer94:polymorphic}. This algorithm treats existentials in a
@@
 to work almost unchanged, but it does come at the cost of losing first-class
-existentials and also of restricting the witness types to monomorphic types.
+existentials and also of restricting the witness types to monomorphic types
+(\ie types with no quantifiers inside of them).
 
@@
 alternating universal and existential quantifiers. Quantifiers still range over
-monotypes (\ie types with no quantifiers inside of them), and higher-rank 
+monotypes, so higher-rank 
 polymorphism is not permitted: one cannot instantiate a quantifier with a type
@@
 quantifiers can only be instantiated with monotypes. However, existential types
-in full System F are \emph{impredicative}---that is, quantifiers can be
+in full \systemf are \emph{impredicative}---that is, quantifiers can be
 instantiated with arbitrary types, specifically including types containing
@@
 observe that their algorithm lacks a declarative specification, and explain why
-existential types make this particularly difficult to do.
+supporting existential types makes giving a specification particularly
+difficult.
 
@@
 
-To research the type inference in impredicative polymorphic systems in the
-presence of existential types, we make the following
-\emph{\textbf{contributions}}.
+We make the following \emph{\textbf{contributions}}:
 
-\paragraph{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
-    We devise \fexists---a second-order polymorphic type system that supports first-class
+\paragraph*{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
+    We present \fexists---a second-order polymorphic type system that supports first-class
     existential and universal quantifiers, both of which can be instantiated
@@
     system (\ie the typing information is only available at a limited scope)
-    based on \emph{subtyping} relation. The specification makes it easy to see
+    based on a \emph{subtyping} relation. Our specification makes it easy to see
     what is and what is not inferrable in the system.
@@
 
-\paragraph{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
+\paragraph*{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
   inference algorithm implementing the declarative specification. Because our
@@
   to unification that finds the most specific generalization rather than the
-  most general unifier. To our knowledge, anit-unification has never been used
+  most general unifier. To our knowledge, anti-unification has never been used
   in the context of polymorphic type inference.
@@
 
-\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
+\paragraph*{Correctness of the algorithm (\cref{sec:proofs})}
   We prove that our algorithm is terminating and both sound and complete with
@@
 
-% \paragraph{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
+% \paragraph*{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
 %     We explore the design space further to show how the
@@
 % \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
-% \paragraph{\textbf{Structure of this Paper}}
+% \paragraph*{\textbf{Structure of this Paper}}
 % In \cref{sec:overview}, we discuss the language of the \fexists system, providing
@@
 
-% \paragraph{\textbf{Acknowledgments}} We would like to thank Meven
-% Lennon-Bertrand for his extremely helpful comments and suggestions.
+\paragraph*{\textbf{Acknowledgments}} We would like to thank Meven
+Lennon-Bertrand for his extremely helpful comments and suggestions.

### 2024-06-11 paper/tex/02-overview.tex
corrections

@@
 
-In the presence of impredicative polymorphism, the standard \systemf has
+In the presence of impredicative polymorphism, \systemf has
 undecidable subtyping and type inference. It means that any inference algorithm
@@
 
-Every term and type fo \fexists has either \emph{positive} or \emph{negative}
+Every term and type of \fexists has either \emph{positive} or \emph{negative}
 polarity. The type variables are annotated with their polarities (e.g. $[[α⁺]]$
@@
 Positive expressions correspond to \emph{values} in the \CBPV system, and
-negative expressions correspond to \emph{computations}. The difference between
-the computations and values is intuitively described in the motto of \CBPV: ``a
-value is, a computation does''. In particular, an argument of a function is
-always a value (\ie positive), and the resulting type, as well as the function
-itself, is a computation (\ie negative). Analogously, the polymorphic
-$\forall$ quantifies a computation over positive types.
+negative expressions correspond to (potentially effectful) \emph{computations}.
+The difference between the computations and values is intuitively described in
+the motto of \CBPV: ``a value is, a computation does''. In particular, a
+function type always takes a value and returns a computation. Similarly, the
+polymorphic $\forall$ quantifies a computation over positive types.
+
+Whenever a computation is used in a value position (e.g. as an argument to a
+function), it must be \emph{suspended} by a thunk constructor, which at the
+type level corresponds to the downshift operator $[[↓iN]]$.
+Symmetrically, to use a value as a computation, one can \emph{return} it,
+which at the type level is reflected as an up-shift $[[↑iP]]$.
 
@@
 Generally, any term and type of \systemf can be embedded into \fexists in
-multiple ways. This is because at any given point, one can opt for either the
-positive or negative polarity of the subterm. We will delve into the consistent
-term-level and type-level embeddings (\ie preserving typing) in more detail in
-\cref{sec:rel-to-systemf}. \Cref{fig:polarization-examples} showcases the
-polarized versions of the standard \systemf terms. Next, we will discuss several
-examples (using the terms from \cref{fig:polarization-examples}) to illustrate
-what can and cannot be inferred in \fexists.
+multiple ways corresponding to different evaluation strategies of \systemf,
+which we will cover in more detail in \cref{sec:rel-to-systemf}.
+
+\Cref{fig:polarization-examples} illustrates how standard \systemf types are
+polarized when call-by-value evaluation is presumed.
+
+Observe the consistent polarization pattern of types. Consider the type for
+$[[double]]$, which in \systemf would be $[[Int → Int]]$. In polarized form, it
+becomes $[[↓(Int → ↑Int)]]$---a \emph{suspended} function (indicated by the
+downshift) that accepts an integer $[[Int]]$ and yields an integer
+\emph{computation} (indicated by the upshift) $[[↑Int]]$.
+
+Moving on to the examples in \cref{fig:polarization-examples}, we will delve
+into several cases that demonstrate the capabilities of type inference within
+\fexists.
 
@@
   \begin{align*}
+    [[double]] &: [[↓(Int → ↑Int)]] \\
     [[map]] &: [[↓(∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺)]] \\
@@
   \end{align*}
-  \caption{Polarization of standard \systemf terms}
+  \caption{Polarization of \systemf types}
   \label{fig:polarization-examples}
@@
 
-\paragraph{Application of a Polymorphic Function}
+\paragraph*{Application of a Polymorphic Function}
   A polymorphic function can be applied to a value of a concrete type. In this
@@
 
-\paragraph{Polymorphic Self-application}
+\paragraph*{Polymorphic Self-application}
   Unlike predicative systems, \fexists permits self-application of polymorphic
@@
 
-\paragraph{Non-example: Polymorphic Arguments}
+\paragraph*{Non-example: Polymorphic Arguments}
 The focus of our system is the \emph{local} type inference
@@
 
-The annotated version of the same term infers the type successfully, 
-since all the polymorphic variables of the arguments are instantiated:
-$$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$
+% The annotated version of the same term infers the type successfully, 
+% since all the polymorphic variables of the arguments are instantiated:
+% $$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$
 
-\paragraph{Inferring the Common Supertype}
+\paragraph*{Inferring the Common Supertype}
   The function $[[choose]]$ takes two arguments of the same
@@
 
-\paragraph{Inferring a general type}
+\paragraph*{Inferring a general type}
   As mentioned before, the local inference does not instantiate the polymorphic
@@
 
-\paragraph{Syntactic Conventions}
+\paragraph*{Syntactic Conventions}
 We always consider terms and types up to alpha-equivalence, and capture-avoiding
@@
 
-\paragraph{Type Context and Type Well-formedness}
+\paragraph*{Type Context and Type Well-formedness}
 
@@
 
-\paragraph{Missing constructors}
+\paragraph*{Missing constructors}
 Notice that the language does not have first-class applications: 
@@
 of the syntax tree. Nevertheless, many difficulties appear in the inference of
-the combination of polymorphic and existential types. Next, we discuss the most
+the combination of polymorphic and existential types. We now discuss the most
 challenging of these difficulties and how they are addressed in the algorithm.
 
-\paragraph{Subtyping} 
+\paragraph*{Subtyping} 
 The inference algorithm's complexity mainly lies in the subtyping relation,
@@
 
-\paragraph{Equivalence}
+\paragraph*{Equivalence}
 The system introduces another complexity: the equivalence induced by subtyping
@@
 straightforward. This relation extends beyond just alpha-equivalent types. As
-the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
+the neighboring polymorphic quantifiers can be instantiated in arbitrary order.
 the equivalence relation must permit the \emph{permutations} of the quantifiers.
@@
 
-\paragraph{Least Upper Bound}
+\paragraph*{Least Upper Bound}
 Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
@@
 and in \fexists, this unifier is the most specific (in particular because
-$[[↓↑α̂⁺]]$ is not considered as only \emph{negative} variables are
+$[[↓↑α̂⁺]]$ is not considered as a candidate since only \emph{negative} variables are
 existentially quantified).
 
-\paragraph{Impredicativity}
+\paragraph*{Impredicativity}
 Another difficulty that hinders the constraint resolution arises from the
@@
 
-To maintain the decidability in the presence of impredicativity, we carefully
+To maintain decidability in the presence of impredicativity, we carefully
 restrict the system using the polarity stratification. One of the main

### 2024-06-11 paper/tex/03-declarative-system.tex
corrections

@@
 
-\paragraph{Quantifiers}  
+\paragraph*{Quantifiers}  
 Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
@@
 
-\paragraph{Invariant Shifts}
+\paragraph*{Invariant Shifts}
 As mentioned above, an important restriction that we put on the subtyping system
@@
 
-\paragraph{Functions}
+\paragraph*{Functions}
 Standardly, subtyping of function types is covariant in the return type
@@
 
-\paragraph{Variables}
+\paragraph*{Variables}
 Subtyping of variables is defined reflexively, which is enough to ensure the
@@
 
-\paragraph{Negative greatest lower bound might not exist}
+\paragraph*{Negative greatest lower bound might not exist}
 The symmetric construction---the greatest lower bound of two negative types---does 
@@
 
-\begin{figure}[h]
+\begin{figure}[t!]
   \ottdefnsDTLabeled
@@
 
-\paragraph{Variables}
+\paragraph*{Variables}
   Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
@@
 
-\paragraph{Annotations}
+\paragraph*{Annotations}
   The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
@@
 
-\paragraph{Abstractions}
+\paragraph*{Abstractions}
   The typing of lambda abstraction is standard. 
@@
 
-\paragraph{Return and thunk}
+\paragraph*{Return and thunk}
   Rules \ruleref{\ottdruleDTReturnLabel} 
@@
 
-\paragraph{Unpack}
+\paragraph*{Unpack}
   Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
@@
 
-\paragraph{Let binders}
+\paragraph*{Let binders}
   Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
@@
 
-\paragraph{Applicative let binders}
+\paragraph*{Applicative let binders}
   Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
@@
 
-\paragraph{Typing up to equivalence}
+\paragraph*{Typing up to equivalence}
   As discussed in \cref{sec:decl-equivalence}, the subtyping, as a preorder, 
@@
 
-\paragraph{Application to an empty list of arguments}
+\paragraph*{Application to an empty list of arguments}
   The base case of the application type inference is 
@@
 
-\paragraph{Application of a polymorphic type $[[∀]]$}
+\paragraph*{Application of a polymorphic type $[[∀]]$}
   The complexity of the system lives in the rules whose output type is not
@@
 
-\paragraph{Application of an arrow type}
+\paragraph*{Application of an arrow type}
   Another application rule is \ruleref{\ottdruleDTArrowAppLabel}.
@@
 
-\paragraph{Type-level Translation}
+\paragraph*{Type-level Translation}
 
@@
 
-\paragraph{Term-level Translation}
+\paragraph*{Term-level Translation}
 

### 2024-06-11 paper/tex/04-algorithm.tex
corrections

--- 2024-05-08 oopsla24/tex/04-algorithm.tex
@@
 
-\paragraph{Algorithmic Variables}
+\paragraph*{Algorithmic Variables}
 To design a mode-correct inference system, we slightly modify the language we operate on.
@@
 
-\paragraph{Algorithmic Types}
+\paragraph*{Algorithmic Types}
 The syntax of algorithmic types extends the declarative syntax by adding
@@
 
-\paragraph{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
+\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
 To specify when algorithmic types are well-formed, we define algorithmic
@@
 
-\paragraph{Instantiation Context $[[Θ]]$}
+\paragraph*{Instantiation Context $[[Θ]]$}
 When one instantiates an algorithmic variable, one may only use type variables
@@
 
-\paragraph{Algorithmic Substitution}
+\paragraph*{Algorithmic Substitution}
 We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
@@
 
-\paragraph{Algorithmic Normalization}
+\paragraph*{Algorithmic Normalization}
 Similarly to well-formedness, the normalization of algorithmic types is defined
@@
 
-\paragraph{Auxiliary Functions}
+\paragraph*{Auxiliary Functions}
 \begin{itemize}
@@
 
-\paragraph{Equivalent Substitutions}
+\paragraph*{Equivalent Substitutions}
 In the proofs of the algorithm correctness, 
@@
 
-\paragraph{Constraint well-formedness and satisfaction}
+\paragraph*{Constraint well-formedness and satisfaction}
 Suppose that $[[Θ]]$ is an instantiation context. 
@@
 
-\paragraph{Variables} Rules \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel}
+\paragraph*{Variables} Rules \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel}
 say that if both of the input types are equal declarative variables,
@@
 
-\paragraph{Shifts} Rules \ruleref{\ottdruleAShiftDLabel} and
+\paragraph*{Shifts} Rules \ruleref{\ottdruleAShiftDLabel} and
 \ruleref{\ottdruleAShiftULabel} cover the downshift and the upshift cases,
@@
 
-\paragraph{Quantifiers}  
+\paragraph*{Quantifiers}  
 Rules \ruleref{\ottdruleAForallLabel} and 
@@
 
-\paragraph{Functions}
+\paragraph*{Functions}
 To infer the subtyping of the function types, the algorithm
@@
 
-\paragraph{Algorithmic variables}
+\paragraph*{Algorithmic variables}
 If one of the sides of the subtyping is a unification variable, the algorithm
@@
 
-\paragraph{Variables}
+\paragraph*{Variables}
   The variable rules \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
@@
 
-\paragraph{Shifts}
+\paragraph*{Shifts}
   The shift rules \ruleref{\ottdruleUShiftDLabel} and \ruleref{\ottdruleUShiftULabel}
@@
 
-\paragraph{Quantifiers} 
+\paragraph*{Quantifiers} 
   Similarly, the quantifier rules \ruleref{\ottdruleUForallLabel} and \ruleref{\ottdruleUExistsLabel}
@@
 
-\paragraph{Functions}
+\paragraph*{Functions}
   Rule \ruleref{\ottdruleUArrowLabel} unifies two functional types. 
@@
 
-\paragraph{Algorithmic variable}
+\paragraph*{Algorithmic variable}
   Finally, if the left-hand side of the unification is an algorithmic variable,
@@
 
-\paragraph{Merging matching constraint entries}
+\paragraph*{Merging matching constraint entries}
 Two \emph{matching} entries formed in the same context $[[Γ]]$ 
@@
 
-\paragraph{Merging constraint sets}
+\paragraph*{Merging constraint sets}
   The algorithm for merging constraint sets is shown in \cref{fig:merge-subtyping-constraints}.
@@
 
-\paragraph{The Type Upgarde}
+\paragraph*{The Type Upgarde}
 The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
@@
 
-\paragraph{The Least Upper Bound}
+\paragraph*{The Least Upper Bound}
 The Least Upper Bound algorithm we use operates on \emph{positive}
@@
 
-\paragraph{Variables} 
+\paragraph*{Variables} 
   Rules \ruleref{\ottdruleAUPVarLabel} and
@@
 
-\paragraph{Shifts} 
+\paragraph*{Shifts} 
   Rules \ruleref{\ottdruleAUShiftDLabel} and
@@
 
-\paragraph{Quantifiers}
+\paragraph*{Quantifiers}
   Rules \ruleref{\ottdruleAUForallLabel} and
@@
 
-\paragraph{Functions}
+\paragraph*{Functions}
   Rule \ruleref{\ottdruleAUArrowLabel} congruently generalizes two function types. 
@@
 
-\paragraph{The Anti-Unification Rule}
+\paragraph*{The Anti-Unification Rule}
   Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
@@
 
-\paragraph{Variables}
+\paragraph*{Variables}
   Rule \ruleref{\ottdruleATVarLabel} 
@@
 
-\paragraph{Annotations}
+\paragraph*{Annotations}
   Rules \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
@@
 
-\paragraph{Abstractions}
+\paragraph*{Abstractions}
   Rule \ruleref{\ottdruleATtLamLabel} infers the type of a lambda abstraction.
@@
 
-\paragraph{Return and Thunk}
+\paragraph*{Return and Thunk}
   Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
@@
 
-\paragraph{Unpack}
+\paragraph*{Unpack}
   Rule \ruleref{\ottdruleATUnpackLabel}
@@
 
-\paragraph{Let Binders}
+\paragraph*{Let Binders}
   Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
@@
 
-\paragraph{Application to an Empty List of Arguments}
+\paragraph*{Application to an Empty List of Arguments}
   Rule \ruleref{\ottdruleATEmptyAppLabel}
@@
 
-\paragraph{Application of a Polymorphic Type $[[∀]]$}
+\paragraph*{Application of a Polymorphic Type $[[∀]]$}
   Rule \ruleref{\ottdruleATForallAppLabel},
@@
 
-\paragraph{Application of an Arrow Type}
+\paragraph*{Application of an Arrow Type}
   Rule \ruleref{\ottdruleATArrowAppLabel}

### 2024-06-11 paper/tex/05-proof.tex
amend
corrections

--- 2024-05-07 oopsla24/tex/05-proof.tex
@@
 
-\paragraph{Determinacy}
+\paragraph*{Determinacy}
     One of the properties that our proof relies on is the determinacy of the 
@@
 
-\paragraph{Mutuality of the Soundness and Completeness Proofs}
+\paragraph*{Mutuality of the Soundness and Completeness Proofs}
 
@@
 The only way to do so is by applying the soundness of minimal instantiation
-(\cref{lemma:minimal-instantiation-soundness}). However, first, the declarative
+(see the lemma above). However, first, the declarative
 tree must be converted to the corresponding algorithmic one (by completeness!).
@@
 
-\paragraph{Inductive Metric}
+\paragraph*{Inductive Metric}
     The soundness and completeness lemmas are proven by \emph{mutual} induction. 

### 2024-06-11 paper/tex/06-extensions.tex
corrections

@@
 
-\paragraph{Solution 1: Declarative Normalization}
+\paragraph*{Solution 1: Declarative Normalization}
 One way to resolve this ambiguity is to fix the order of quantifiers.  
@@
 
-\paragraph{Solution 2: Elementary Type Inference}
+\paragraph*{Solution 2: Elementary Type Inference}
 An alternative approach to provide explicit type application
@@
 
-\paragraph{Solution 3: labeled quantifiers}
+\paragraph*{Solution 3: labeled quantifiers}
 
@@
 
-\paragraph{The Algorithm}
+\paragraph*{The Algorithm}
 
@@
     $$ \ottdruleATArrowAppBidirLabeled{} $$
-    % \paragraph{Reaching Undecidability}
+    % \paragraph*{Reaching Undecidability}
     However, this rule is too permissive. 
@@
 
-    % \paragraph{Restricting Inference}
+    % \paragraph*{Restricting Inference}
     Instead, we strengthen

### 2024-06-25 paper/tex/02-overview.tex
edit

@@
 
-Moving on to the examples in \cref{fig:polarization-examples}, we will delve
-into several cases that demonstrate the capabilities of type inference within
-\fexists.
+Using the examples provided in \cref{fig:polarization-examples}, we now showcase 
+the type inference capabilities and restrictions of \fexists
 
@@
   A polymorphic function can be applied to a value of a concrete type. In this
-  case, the quantified variables are instantiated to make the application
-  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
-  of any positive type, and the resulting type will be inferred by the
-  algorithm:
+  case, the explicit type application is not required: the inference algorithm
+  automatically instantiates the quantified variables. For instance, the
+  polymorphic $[[len]]$ can be applied to a list of any positive type, and the
+  resulting type will be inferred by the algorithm:
   $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$
 
-\paragraph*{Polymorphic Self-application}
+\paragraph*{Polymorphic Self-application.}
   Unlike predicative systems, \fexists permits self-application of polymorphic

### 2024-06-25 paper/tex/03-declarative-system.tex
edit

@@
 
-\paragraph*{Variables}
+\subparagraph*{Variables}
     Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
@@
     to the type from the context. 
-  In our case, the inference of equivalent types
-  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.
+    In our case, the inference of equivalent types is admissible by \ruleref{\ottdruleDTPEquivLabel}.
 
-\paragraph*{Annotations}
+\subparagraph*{Annotations}
   The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
@@
 
-\paragraph*{Abstractions}
+\subparagraph*{Abstractions}
   The typing of lambda abstraction is standard. 
@@
 
-\paragraph*{Return and thunk}
+\subparagraph*{Return and thunk}
   Rules \ruleref{\ottdruleDTReturnLabel} 
@@
 
-\paragraph*{Unpack}
+\subparagraph*{Unpack}
   Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
@@
 
-\paragraph*{Let binders}
+\subparagraph*{Let binders}
   Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
@@
 
-\paragraph*{Applicative let binders}
+\subparagraph*{Applicative let binders}
   Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
@@
 
-\begin{figure}[ht]
-  \begin{minipage}[t]{0.65\linewidth}
+\begin{figure}[t]
+  \begin{minipage}[t]{0.60\linewidth}
     \centering
-      \begin{minipage}[t]{0.32\linewidth}
+      \begin{minipage}[t]{0.30\linewidth}
       \ottfundefnunpolP{}
       \end{minipage}
-      \begin{minipage}[t]{0.32\linewidth}
+      \hspace{0.5cm}
+      \begin{minipage}[t]{0.30\linewidth}
       \ottfundefnunpolN{}
@@
   \end{minipage}
-  \hspace{0.5cm}
+  \hspace{0.3cm}
   \begin{minipage}[t]{0.3\linewidth}
@@
   \end{minipage}
-
 \end{figure}
@@
 Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
-Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
+Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● fargs) : |iM|]]$ \\
 \hline

### 2024-06-28 paper/tex/02-overview.tex
edit

@@
 
-Observe the consistent polarization pattern of types. Consider the type for
-$[[double]]$, which in \systemf would be $[[Int → Int]]$. In polarized form, it
-becomes $[[↓(Int → ↑Int)]]$---a \emph{suspended} function (indicated by the
-downshift) that accepts an integer $[[Int]]$ and yields an integer
-\emph{computation} (indicated by the upshift) $[[↑Int]]$.
+Observe the consistent pattern of the type polarization. Consider the type for
+$[[double]]$---a function multiplying an input integer by two. In \systemf, it
+would be $[[Int → Int]]$. In call-by-value polarized form, it becomes $[[↓(Int →
+↑Int)]]$---a \emph{suspended} function (indicated by the downshift) that accepts
+an integer $[[Int]]$ and yields an integer \emph{computation} (indicated by the
+upshift) $[[↑Int]]$.
+
+The type of $[[map]]$ is more complex. Similar to $[[double]]$, the
+entire type is a suspended function, hence $[[↓]]$. The function takes two implicit type
+arguments: $[[α⁺]]$ and $[[β⁺]]$, which indicated by the $[[∀]]$s.
+The third argument of $[[map]]$---the applied function---has type $[[α⁺ → ↑β⁺]]$,
+which is suspended by $[[↓]]$ since it is in a (positive) argument position. 
+The last argument---$[[List α⁺]]$---represents a list of $[[α⁺]]$s and does not need
+to be suspended as it is already a value. Finally, the positive return type,
+$[[List β⁺]]$, is lifted to computations using $[[↑]]$.
+
+The types of $[[len]]$---the length function for lists, $[[choose]]$---a
+function that chooses between two input values, and $[[id]]$---the identity
+function, are similarly polarized. A slightly more complex function $[[auto]]$
+takes as an argument and returns something of the same identity type. However,
+since the argument position is positive, the argument type is suspended by
+$[[↓]]$.
+
+The types of $[[len]]$ (the length function for lists), $[[choose]]$ (a function
+that chooses between two input values), and $[[id]]$ (the identity function) are
+polarized in a similar way. Another function, $[[auto]]$, takes an argument and returns
+a function of the same identity type. However, since the argument position is
+positive, the argument type is suspended using $[[↓]]$.
 
@@
 \begin{itemize}
-\item [$\pm$] $[[na]]$ and $[[pa]]$
+\item [$\pm$] $[[pa]]$ and $[[na]]$
   denote negative and positive type variables, respectively.
@@
 \paragraph*{Syntactic Conventions}
-We always consider terms and types up to alpha-equivalence, and capture-avoiding
-substitution. Besides, we assume the quantification is associative,
+We always consider terms and types up to alpha-equivalence, 
+and substitution is capture-avoiding. In addition, we assume the quantification is associative,
 for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
@@
 terms are stratified into values and computations; there are return 
-and thunk constructors, as well as several forms of let bindings to
-sequentialize the computation/application with or without type annotation;
+and thunk constructors, as well as several forms of let bindings
+to sequence computations, bind function applications, 
+and eliminate existentials, all with or without type annotations.
 
@@
         the type annotation for the argument of $[[λ]]$;
-    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
+    \item [$-$] $[[ let x = v ; c]]$ is a standard let form, binding
         a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
@@
         $[[v]] ([[args]])$.
-        They take a suspended computation $[[v]]$, apply it to a list of
+        They take a function $[[v]]$, apply it to a list of
         arguments, bind the (pure) result to a variable
@@
 The inference algorithm's complexity mainly lies in the subtyping relation,
-which is, in particular, responsible for polymorphic instantiation. It's this
+which is responsible for polymorphic instantiation. It is this
 subtyping relation that enables us to apply polymorphic terms in situations
@@
 straightforward. This relation extends beyond just alpha-equivalent types. As
-the neighboring polymorphic quantifiers can be instantiated in arbitrary order.
+the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
 the equivalence relation must permit the \emph{permutations} of the quantifiers.
@@
 
-In many type systems, the least upper bound boils down to the trivial syntactic equality.
+In Damas Milner type inference, the least upper bound boils down to the trivial syntactic equality.
 Because of that, the set of constraints is typically resolved using \emph{unification}.
@@
 example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
-and in \fexists, this unifier is the most specific (in particular because
+and in \fexists, this anti-unifier is the most specific (in particular because
 $[[↓↑α̂⁺]]$ is not considered as a candidate since only \emph{negative} variables are

### 2024-06-28 paper/tex/03-declarative-system.tex
edit

@@
 First, we present the declarative subtyping rules specifying 
-when one type is a subtype of another; next, we discuss 
-the equivalence relation induced by mutual subtyping;
-finally, we present the type inference rules,
+when one type is a subtype of another. Next, we discuss 
+the equivalence relation induced by mutual subtyping.
+Finally, we present the type inference rules,
 that refer to the subtyping and equivalence relation.
@@
 Second, it verifies that the left-hand side quantifiers
-($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
-This is the point of non-determinism as the instantiation can be chosen arbitrarily
+($[[pas]]$ or $[[nas]]$) can be instantiated 
+with an arbitrary substitution to continue subtyping recursively,
+which introduces non-determinism.
 
-The instantiation of quantifiers is modeled by substitution $[[σ]]$.
+The instantiation of quantifiers is modeled by a substitution $[[σ]]$.
 The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
@@
 form, the subtyping relation would become equivalent to the standard subtyping
-of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. However, it is
-suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
-relaxed to the covariant form, thereby increasing the expressiveness of the
-system. These changes are discussed in \cref{sec:weakening-invariant}.
+of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. 
+Relaxations of this condition are discussed in \cref{sec:weakening-invariant}.
+
+% However, it is
+% suggested after certain changes \ruleref{\ottdruleDOneShiftULabel} can be
+% relaxed to the covariant form, thereby increasing the expressiveness of the
+% system.  These changes are discussed in \cref{sec:weakening-invariant}.
 
@@
 
-\subsubsection{Properties of Declarative Subtyping}
+\subsection{Properties of Declarative Subtyping}
 \label{sec:decl-subtyping-properties}
@@
   well-formed in $[[Γ]]$.
-  Then there exists a least common supertype--- a type $[[iP]]$ such that
+  Then there exists a least common supertype---a type $[[iP]]$ such that
   \begin{itemize}
@@
 implemented by the \emph{ordering} function `$[[ord varset in iN]]$' and `$[[ord
-varset in iP]]$'. For a set of variables $[[varset]]$, and a type, it returns a
+varset in iP]]$'. For a set of variables $[[varset]]$ and a type, it returns a
 list of variables from $[[varset]]$ that occur in the type in the \emph{order of
@@
 \cref{fig:declarative-inference}.
+It consists of three main judgments: the positive typing judgment,
+the negative typing judgment, and the application typing judgment.
 The positive typing judgment $[[Γ ; Φ ⊢ v : iP]]$ is read as 
@@
 The negative typing judgment is read similarly.
-
 The \emph{Application typing} judgment
@@
   Rule \ruleref{\ottdruleDTtLamLabel} first checks
-  that the given type annotating the argument is well-formed,
-  and then infers the type of the body in the extended context.
-  As a result, it returns an arrow type of function from the
-  annotated type of the argument to the type of the body.
+  that the given type $[[iP]]$ annotating the argument is well-formed,
+  and then infers the type $[[iN]]$ of the body in the extended context.
+  As a result, it returns an arrow type $[[iP → iN]]$.
   Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
   It extends the type context with the quantifying variable $[[pa]]$ and 
-  infers the type of the body. As a result, it returns a polymorphic type
-  quantifying the abstracted variable $[[pa]]$ over the type of the body.
+  infers the type $[[iN]]$ of the body. As a result, it returns a polymorphic type
+  $[[∀pa.iN]]$.
 
@@
   After the bind, the rule infers the type of the body 
-  and checks that it does not use the bound variables so that they do not 
-  escape the scope.
+  and ensures that the bound variables do not escape the scope.
 
@@
   \cref{sec:algorithm} delays this instantiation until more information about it
-  (in particular, typing constraints) is collected.
+  (in particular, the set of typing constraints) is collected.
 
@@
 execution strategy of the program. 
-In particular, a functional type can be represented in either positive (thunked, call-by-name) way:
-$[[↓(iP → ↑iQ)]]$ or negative (call-by-value): $[[↓iN → iM]]$.
+In particular, a functional type can be represented in either positive (thunked, call-by-value) way:
+$[[↓(iP → ↑iQ)]]$ or negative (call-by-name): $[[↓iN → iM]]$.
 We chose the positive polarization: every \systemf type is translated to a
@@
 ft]]$ constructs a \systemf function $[[ft]]$ of type $[[|iN| → |iM|]]$ that
-witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$; a typing elaboration
+witnesses the given \fexists-subtyping $[[Γ ⊢ iN ≤ iM]]$. A typing elaboration
 $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ builds a term $[[ft]]$---a \systemf counterpart of

### 2024-06-28 paper/tex/04-algorithm.tex
edit

@@
 
-\begin{figure}
-  \begin{multicols}{2}
+\begin{figure}[t]
+
+  \begin{minipage}[t]{0.49\textwidth}
       Negative Algorithmic Variables\\
       $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, \dots\\
-      \columnbreak
+  \end{minipage}%
+  \begin{minipage}[t]{0.49\textwidth}
       Positive Algorithmic Variables\\
       $[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, \dots\\
-  \end{multicols}
+  \end{minipage}
+
   \hfill\\
-  \begin{multicols}{2}
+  \begin{minipage}[t]{0.49\textwidth}
       \ottuNShort
-      \columnbreak
+  \end{minipage}
+  \begin{minipage}[t]{0.49\textwidth}
       \ottuPShort
-  \end{multicols}
+  \end{minipage}
   \hfill\\
@@
 
-\begin{figure}
-\begin{multicols}{2}
-  $\vcenter{\hbox{\vdots}}$\\
+\begin{figure}[t]
+  \begin{minipage}{0.49\textwidth}
+  $$\vcenter{\hbox{\vdots}}$$
   \ottusedrule{\ottdruleWFATPUVarLabeled{}}
-  \columnbreak\\
-  $\vcenter{\hbox{\vdots}}$\\
+  \end{minipage}
+  \begin{minipage}{0.49\textwidth}
+  $$\vcenter{\hbox{\vdots}}$$
   \ottusedrule{\ottdruleWFATNUVarLabeled{}}
-\end{multicols}
-\caption{Well-formedness of Algorithmic Types}
+  \end{minipage}
+\caption{Well-formedness of Algorithmic Types extends Declarative Well-formedness}
 \label{fig:algo-wf}
@@
 
-\begin{figure}
-\begin{multicols}{2}
-  $\vcenter{\hbox{\vdots}}$\\
+\begin{figure}[t]
+\begin{minipage}{0.49\textwidth}
+  $$\vcenter{\hbox{\vdots}}$$
   \ottusedrule{\ottdruleNrmPUVar{}}
-  \columnbreak\\
-  $\vcenter{\hbox{\vdots}}$\\
+\end{minipage}
+\hfill
+\begin{minipage}{0.49\textwidth}
+  $$\vcenter{\hbox{\vdots}}$$
   \ottusedrule{\ottdruleNrmNUVar{}}
-\end{multicols}
-\caption{Normalization of Algorithmic Types}
+\end{minipage}
+\caption{Normalization of Algorithmic Types extends Declarative Normalization}
 \label{fig:algo-nf}
@@
 
-  \begin{figure}[h]
-    \begin{multicols}{2}
+  \begin{figure}[t]
+    \begin{minipage}{0.49\textwidth}
       \ottgrammartabular { 
@@
       }
-
-    \columnbreak
-
+    \end{minipage}
+    \begin{minipage}{0.49\textwidth}
       \ottgrammartabular { 
@@
       }
-    \end{multicols}
-
+    \end{minipage}
     \label{fig:syntax-e-sc}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \centering
@@
 
-In the remainder of this section, we will delve into each of these procedures in
-detail, following the top-down order of the dependency graph. First, 
-we present the subtyping algorithm itself.
+In the remainder of this section, we will discuss each of these procedures in
+detail, following the top-down order of the dependency graph. First, we present
+the subtyping algorithm itself.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \hfill\\
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \hfill
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \ottdefnSCMELabeled\\
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.\\
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \begin{multicols}{2}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
     \ottdefnAUAUPLabeled{}
@@
 
-\begin{figure}
+\begin{figure}[t]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[t]
   \ottdefnATPInfLabeled{}
@@
 
-\begin{figure}
+\begin{figure}[t]
   \begin{minipage}{0.49\textwidth}

### 2024-06-28 paper/tex/04-algorithm.tex
edit

@@
 the output of each rule is always uniquely defined by its input.
-And second, most of the declarative rules 
-(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel})
-have a unique algorithmic counterpart, 
-which simplifies reasoning about the algorithm and its correctness proofs.
+And second, the rules of the algorithmic system mirrors the declarative rules 
+(except for the rules \ruleref{\ottdruleDTPEquivLabel} and \ruleref{\ottdruleDTNEquivLabel}), 
+which simplifies the correctness proof.
 
@@
 
-First, let us discuss the syntax of the algorithmic system.
+First, let us discuss the syntax of the algorithmic system (\cref{fig:algo-syntax}).
 
@@
 \emph{declarative} types: the declarative variables are checked to belong to the
-context $[[Γ]]$, the quantifiers extend the context, type constructors are
+context $[[Γ]]$, the quantifiers extend the context $[[Γ]]$, type constructors are
 well-formed congruently. As we extend the syntax with algorithmic variables, we
@@
 constraints can take a positive $[[α̂⁺ :≈ iP]]$ or negative $[[α̂⁻ :≈ iN]]$
-form, but algorithmic type variables cannot occur on their right-hand side. The
-constraints \emph{set} is denoted as $[[SC]]$, and we presume that each
+form. Algorithmic variables cannot occur in the right-hand side of the
+constraints.  The constraint \emph{set} is denoted as $[[SC]]$, and we presume that each
 algorithmic variable can be restricted by at most one constraint.
@@
 which we use as a subroutine of the subtyping algorithm, can only produce
-unification constraints. A resolution of unification constraints is simpler than
+unification constraints. The resolution of unification constraints is simpler than
 that of a general constraint set. This way, this separation allows us to better
@@
     algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
-    domain of instantiation context---as a set of algorithmic variables that
+    domain of instantiation context $[[Θ]]$---as a set of algorithmic variables that
     $[[Θ]]$ associates with their contexts.  
@@
 (denoted as $[[Θ ⊢ SC]]$) 
-if for every every constrain entry $[[scE ∊ SC]]$
+if for every constrain entry $[[scE ∊ SC]]$
 associating a variable $[[α̂±]]$ with a type $[[iP]]$,

### 2024-07-09 paper/tex/03-declarative-system.tex
edit and prepare for jfp

@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 
@@
 
-\subparagraph*{Variables}
+\paragraph*{Variables}
     Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
@@
 
-\subparagraph*{Annotations}
+\paragraph*{Annotations}
   The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
@@
 
-\subparagraph*{Abstractions}
+\paragraph*{Abstractions}
   The typing of lambda abstraction is standard. 
@@
 
-\subparagraph*{Return and thunk}
+\paragraph*{Return and thunk}
   Rules \ruleref{\ottdruleDTReturnLabel} 
@@
 
-\subparagraph*{Unpack}
+\paragraph*{Unpack}
   Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
@@
 
-\subparagraph*{Let binders}
+\paragraph*{Let binders}
   Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
@@
 
-\subparagraph*{Applicative let binders}
+\paragraph*{Applicative let binders}
   Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}

### 2024-07-09 paper/tex/04-algorithm.tex
edit and prepare for jfp

@@
 
-  For convenience and scalability, 
-  we decompose the subtyping algorithm 
+  We decompose the subtyping algorithm 
   into several procedures. \Cref{fig:alg-subtyping-graph}
@@
 \begin{figure}[t]
+  \begin{minipage}[t]{0.50\textwidth}
+    \ottdefnUNUnifLabeled{}
+  \end{minipage}
   \hfill
-  \begin{multicols}{2}
-  \ottdefnUNUnifLabeled{}
-  \columnbreak\\
+  \begin{minipage}[t]{0.49\textwidth}
     \ottdefnUPUnifLabeled{}
-  \end{multicols}
+  \end{minipage}
   \caption{Unification Algorithm}
@@
 in a smaller context $[[Δ ⊆ Γ]]$.
-For given two types $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
+When given two types $[[iP1]]$ and $[[iP2]]$ well-formed in $[[Γ]]$,
 the \emph{least upper bound} operation finds the least among
@@
 
-\paragraph*{The Type Upgarde}
+\paragraph*{The Type Upgrade}
 The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
@@
 and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
-does not contain $[[pnas]]$ nor $[[pnbs]]$.
+does not contain $[[pnas]]$ or $[[pnbs]]$.
 
@@
 \ruleref{\ottdruleLUBUpgradeLabel} we employ a different but equivalent approach:
-we create \emph{two} types $[[iP1]]$ and $[[iP2]]$ constructed by renaming $[[pnas]]$ in $[[iP]]$
+we create \emph{two} types $[[iP1]]$ and $[[iP2]]$ by renaming $[[pnas]]$ in $[[iP]]$
 to fresh disjoint variables $[[pnbs]]$ and $[[pncs]]$ respectively, and then 
@@
 
-  In our representation of the anti-unification algorithm, this `merge' happens
+  In our representation of the anti-unification algorithm, this `merging' happens
   automatically:
@@
       It is worth noting that the union is well-defined because
-      the result of the substation on a `hole' is determined by the 
+      the result of the substitution on a `hole' is determined by the 
       name of the `hole'.
@@
   way, the algorithm must try to apply the congruent rules first, and only if
-  they fail, apply \ruleref{\ottdruleAUAULabel}. As it is not known a priori
-  whether the congruent rules will be applicable, the implementation must use 
-  backtracking.
+  they fail, apply \ruleref{\ottdruleAUAULabel}.
 
@@
 
-The positive and the negative type inference judgments have symmetric forms:
+The positive and the negative type inference judgments have the symmetric forms:
 $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ c : iN]]$. Both of these algorithms
@@
 
-The application type inference judgment has form 
+The application type inference judgment has the form 
 $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$.
@@
 
-\begin{figure}[t]
+\begin{figure}[t!]
   \ottdefnATPInfLabeled{}
@@
   is the base case of application inference. 
-  If the list of applied arguments is empty, 
+  If the list of arguments is empty, 
   the inferred type is the type of the head,
@@
 To find this substitution, we use the minimal instantiation
-algorithm (\cref{fig:minimal-instantiation}). The judgment `$[[Γ ⊢ uP SC minby uσ]]$'
+algorithm (\cref{fig:minimal-instantiation}). In the judgment "$[[Γ ⊢ uP SC minby uσ]]$",
 $[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 

### 2024-07-09 paper/tex/05-proof.tex
edit and prepare for jfp

@@
     Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
-    \hfill
     \begin{itemize}
@@
 (\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
-bottom-up. Next, we discuss the main of these results.
+bottom-up. Next, we discuss the main results.
 
@@
     This way, \emph{the correctness of normalization} means that
-    checking for equivalence of two types is the same as checking for equality of their normal forms.
+    checking for the equivalence of two types is the same as checking for the equality of their normal forms.
     \begin{lemma*}[Normalization correctness]
@@
     (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
-    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
+    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), 
+    $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
     but explicitly allows quantifier reordering and redundant quantifier removal.
@@
 The `refinement' is represented as an instantiation of the anti-unifier---a 
-substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the `holes' from $[[Ξ1]]$ with
+substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$---replacing the `holes' from $[[Ξ1]]$ with
 types that themselves can contain `holes' from $[[Ξ2]]$.
@@
     the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
-    and the \emph{initiality} means that the returned type is the least among common supertypes.
+    and the \emph{initiality} means that the returned type is the least among the common supertypes.
 
@@
 $[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
-This way, the correctness of minimal instantiation is formulated as the following two lemmas.
+The correctness of minimal instantiation is formulated as the following two lemmas.
 
@@
     Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
-    Then `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
+    Then the judgment `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
     $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
@@
     $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
-    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
+    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ is derivable.
 \end{lemma*}
 
-The soundness is rather straightforwardly proved by induction on the 
+Soundness is rather straightforwardly proved by induction on the 
 judgment inference tree, relying on the soundness of the singularity procedure.
@@
 
-Finally, we discuss the proofs of the soundness and completeness of type
-inference algorithm that we stated at the beginning of this section.
-There are three subtleties that we will cover that are important 
+Finally, we discuss the proofs of the soundness and completeness of the type
+inference algorithm that we described at the beginning of this section.
+There are three subtleties that are important 
 for the proof to go through: the determinacy of the algorithm,
@@
 
+\vspace{\baselineskip}
 \paragraph*{Determinacy}
@@
 \begin{lemma*}[Determinacy of the Typing Algorithm]
-    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
+    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ0]]$. Then 
     \begin{itemize}
@@
         \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
-        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
-            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
+        \item If $[[Γ; Φ; Θ0 ⊨ uN ● args ⇒> uM ⫤ Θ; SC]]$ and 
+            $[[Γ; Φ; Θ0 ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
             $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
@@
 the soundness and completeness proofs cannot be separated: the inductive proof
-of one requires another, and vice versa.
+of one requires the other.
 
@@
 
-    The size of \emph{declarative} and \emph{algorithmic} judgments is defined as
-    a pair: the first component is the syntactic size of the terms used in the judgment. 
-    The second component depends on the kind of judgment. For regular
-    type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$
-    or $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments
-    $([[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), 
-    it is equal to the number of prenex quantifiers of the head type $[[iN]]$.
-    We need this adjustment to ensure the monotonicity of the metric, since the rules 
-    \ruleref{\ottdruleATForallAppLabel} and \ruleref{\ottdruleDTForallAppLabel}
-    only reduce the quantifiers in the head type but they do not change the list of arguments.
+    The size of \emph{declarative} and \emph{algorithmic} judgments is defined
+    as a pair: the first component is the cumulative syntactic size of the terms
+    used in the judgment. The second component depends on the kind of judgment.
+    For regular type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$ or 
+    $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments $([[Γ
+    ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), it
+    is equal to the number of prenex quantifiers of the head type $[[iN]]$. We
+    need this adjustment to ensure the monotonicity of the metric, since the
+    quantifier instantiation rules \ruleref{\ottdruleATForallAppLabel} and
+    \ruleref{\ottdruleDTForallAppLabel} only reduce the quantifiers in the head
+    type but they do not change the list of arguments.
 
@@
 This metric is suitable for mutual induction on the soundness and completeness
-of the typing algorithm.  First, it is monotonous \wrt the inference rules,
+of the typing algorithm.  First, it is monotone \wrt the inference rules,
 and this way, we can always apply the induction hypothesis to premises 

### 2024-07-09 paper/tex/06-extensions.tex
edit and prepare for jfp

@@
 In our system, all type applications are inferred implicitly: the algorithm
-automatically instantiates the variables abstracted by $\forall$. 
-The \emph{explicit} type application can be added to the declarative system by the following rule:
+\emph{Explicit} type application can be added to the declarative system by the following rule:
+
 $$\ottdruleDTTypeAppLabeled{}$$
@@
 
+\vspace{\baselineskip}
 \paragraph*{Solution 1: Declarative Normalization}
@@
 
+\vspace{\baselineskip}
 \paragraph*{Solution 2: Elementary Type Inference}
-An alternative approach to provide explicit type application
+An alternative approach to explicit type application
  was proposed by \cite{zhao22:elementary}.
@@
 
-The elementary type inference restricts the expressiveness of the subtyping perhaps too much.
+Elementary type inference
+is much more restrictive than what we present.
 As mentioned, it forbids the quantifier reordering; besides, 
@@
 
+\vspace{\baselineskip}
 \paragraph*{Solution 3: labeled quantifiers}
@@
 without fixing the order of quantifiers is by using \emph{labeled} quantifiers.
-Let us discuss how to introduce them to the \emph{negative} part of the system,
+Let us discuss how to introduce them to the \emph{negative} part of the system
 since the positive subsystem is symmetric.
@@
 a matching left-hand side quantifier with the same label. These
-matching quantifiers synchronously removed
+matching quantifiers are synchronously removed
 (in other words, each variable is abstractly instantiated to itself), 
@@
 Second, notice that the positive subtyping is still not dependent on the negative
-subtyping. It means, that any leaf-to-root path in the inference tree is monotonous:
+subtyping. It means that any leaf-to-root path in the inference tree is monotone:
 in the first phase, the negative subtyping rules are applied, and in the second phase, 
@@
 we believe that the system consisting of the mentioned kinds of constraints is solvable
-by the following reasons.
+for the following reasons.
 \begin{enumerate*}
@@
 
+\vspace{\baselineskip}
 \paragraph*{The Algorithm}
-
     The algorithmic system is bidirectionalized in a similar way. 
-    Each typing premise and the conclusion of the rule is oriented
+    Each typing premise and the conclusion of the rule are oriented
     to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
@@
     \ruleref{\ottdruleATArrowAppBidirLabel}
-    in such a way that it never checks a term against an algorithmic type.
-    Instead, type checking $[[v]]$ against $[[uQ]]$ is replaced by 
-    a more restrictive premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
+    in such a way that it never checks a term against an algorithmic type. Type
+    checking $[[v]]$ against $[[uQ]]$ is replaced by a more restrictive
+    premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
     $$ \ottdruleATArrowAppBidirSLabeled{} $$\\

### 2024-07-09 proofs/content.tex
edit and prepare for jfp

--- 2024-04-03 proofs/main.tex
@@
-\PassOptionsToPackage{prologue,dvipsnames}{xcolor}
-
-% \UseRawInputEncoding
-% vim: ft=tex
-
-% \documentclass[acmsmall,natbib=false]{article}
-% \usepackage[a4paper, total={8in, 10in}]{geometry}
-
-\documentclass[acmsmall,natbib=false,review,anonymous]{acmart}
-\input{../prelude.tex}
-
-\newcommand{\ottDir}{../ott/_gen}
-\newcommand{\genDir}{_gen}
-\input{\ottDir/defs.tex}
-\input{\ottDir/renew-ott.tex}
-
-\usepackage{../bibliography}
-% \usepackage{amssymb}
-
-% \usepackage[matha]{mathabx}
-
-% https://tex.stackexchange.com/questions/85033/colored-symbols/85035#85035
-% \newcommand*{\mathcolor}{}
-% \def\mathcolor#1#{ \mathcoloraux{#1} }
-% \newcommand*{\mathcoloraux}[3]{%
-%   \protect\leavevmode
-%   \begingroup
-%   \color#1{#2}#3%
-%   \endgroup
-% }
-
-\begin{document}
-
-\title{The Proofs}
-
-\maketitle
-
-\tableofcontents
-
-\newpage
+\newcommand{\proofGenDir}{_gen}
 
@@
 
-Substitution that operates on algorithmic type variables is denoted as
+A substitution that operates on algorithmic type variables is denoted as
 $[[uσ]]$. It is defined as a total function from algorithmic 
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
   \begin{minipage}[t]{0.65\linewidth}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \ottdefnELLabeled
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
   \ottdefnELfPolLabeled{}
@@
 \subsection{Type Well-Formedness}
-\input{\genDir/tex/wf-statements.tex}
+\input{\proofGenDir/tex/wf-statements.tex}
 
 \subsection{Substitution}
-\input{\genDir/tex/subst-statements.tex}
+\input{\proofGenDir/tex/subst-statements.tex}
 
 \subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-statements.tex}
+\input{\proofGenDir/tex/decl-subtyping-statements.tex}
 
@@
 \label{sec:decl-equiv-statements}
-\input{\genDir/tex/equiv-statements.tex}
+\input{\proofGenDir/tex/equiv-statements.tex}
 
 \subsection{Variable Ordering}
-\input{\genDir/tex/ord-statements.tex}
+\input{\proofGenDir/tex/ord-statements.tex}
 
 \subsection{Normaliztaion}
-\input{\genDir/tex/norm-statements.tex}
+\input{\proofGenDir/tex/norm-statements.tex}
 
 \section{Declarative Typing}
-\input{\genDir/tex/decl-typing-statements.tex}
+\input{\proofGenDir/tex/decl-typing-statements.tex}
 
 \section{Relation to \systemf}
-\input{\genDir/tex/translation-statements.tex}
+\input{\proofGenDir/tex/translation-statements.tex}
 
@@
 \subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-statements.tex}
+\input{\proofGenDir/tex/wf-algo-statements.tex}
 
 \subsection{Algorithmic Substitution}
-\input{\genDir/tex/subst-algo-statements.tex}
+\input{\proofGenDir/tex/subst-algo-statements.tex}
 
 \subsection{Algorithmic Normalization}
-\input{\genDir/tex/norm-alg-statements.tex}
+\input{\proofGenDir/tex/norm-alg-statements.tex}
 
 \subsection{Algorithmic Equivalence}
-\input{\genDir/tex/equiv-alg-statements.tex}
+\input{\proofGenDir/tex/equiv-alg-statements.tex}
 
 \subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-statements.tex}
+\input{\proofGenDir/tex/unif-constraint-merge-statements.tex}
 
 \subsection{Unification}
-\input{\genDir/tex/unification-statements.tex}
+\input{\proofGenDir/tex/unification-statements.tex}
 
 \subsection{Anti-unification}
-\input{\genDir/tex/au-statements.tex}
+\input{\proofGenDir/tex/au-statements.tex}
 
@@
 \label{sec:alg-upper-bounds-statements}
-\input{\genDir/tex/lub-statements.tex}
+\input{\proofGenDir/tex/lub-statements.tex}
 
@@
 \label{sec:upgrade-statements}
-\input{\genDir/tex/upgrade-statements.tex}
+\input{\proofGenDir/tex/upgrade-statements.tex}
 
 \subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-statements.tex}
+\input{\proofGenDir/tex/constraint-sat-statements.tex}
 
 \subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-statements.tex}
+\input{\proofGenDir/tex/alg-pos-subtyping-statements.tex}
 
 \subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-statements.tex}
+\input{\proofGenDir/tex/constraint-merge-statements.tex}
 
 \subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-statements.tex}
+\input{\proofGenDir/tex/alg-neg-subtyping-statements.tex}
 
 \subsection{Singularity and Minimal Instantiation}
-\input{\genDir/tex/singularity-statements.tex}
+\input{\proofGenDir/tex/singularity-statements.tex}
 
 \subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-statements.tex}
+\input{\proofGenDir/tex/alg-typing-statements.tex}
 
@@
 \subsection{Type Well-Formedness}
-\input{\genDir/tex/wf-lemmas.tex}
+\input{\proofGenDir/tex/wf-lemmas.tex}
 
 \subsection{Substitution}
-\input{\genDir/tex/subst-lemmas.tex}
+\input{\proofGenDir/tex/subst-lemmas.tex}
 
 \subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-lemmas.tex}
+\input{\proofGenDir/tex/decl-subtyping-lemmas.tex}
 
@@
 \label{sec:decl-equiv-lemmas}
-\input{\genDir/tex/equiv-lemmas.tex}
+\input{\proofGenDir/tex/equiv-lemmas.tex}
 
 \subsection{Variable Ordering}
-\input{\genDir/tex/ord-lemmas.tex}
+\input{\proofGenDir/tex/ord-lemmas.tex}
 
 \subsection{Normaliztaion}
-\input{\genDir/tex/norm-lemmas.tex}
+\input{\proofGenDir/tex/norm-lemmas.tex}
 
 \section{Relation to \systemf}
-\input{\genDir/tex/translation-proofs.tex}
+\input{\proofGenDir/tex/translation-proofs.tex}
 
@@
 \subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-lemmas.tex}
+\input{\proofGenDir/tex/wf-algo-lemmas.tex}
 
 \subsection{Substitution}
-\input{\genDir/tex/subst-algo-lemmas.tex}
+\input{\proofGenDir/tex/subst-algo-lemmas.tex}
 
 \subsection{Normalization}
-\input{\genDir/tex/norm-alg-lemmas.tex}
+\input{\proofGenDir/tex/norm-alg-lemmas.tex}
 
 \subsection{Equivalence}
-\input{\genDir/tex/equiv-alg-lemmas.tex}
+\input{\proofGenDir/tex/equiv-alg-lemmas.tex}
 
 \subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-lemmas.tex}
+\input{\proofGenDir/tex/unif-constraint-merge-lemmas.tex}
 
 \subsection{Unification}
-\input{\genDir/tex/unification-lemmas.tex}\
+\input{\proofGenDir/tex/unification-lemmas.tex}\
 
 \subsection{Anti-unification}
-\input{\genDir/tex/au-lemmas.tex}
+\input{\proofGenDir/tex/au-lemmas.tex}
 
@@
 \label{sec:alg-upper-bounds-proofs}
-\input{\genDir/tex/lub-lemmas.tex}
+\input{\proofGenDir/tex/lub-lemmas.tex}
 
@@
 \label{sec:upgrade-lemmas}
-\input{\genDir/tex/upgrade-lemmas.tex}
+\input{\proofGenDir/tex/upgrade-lemmas.tex}
 
 \subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-lemmas.tex}
+\input{\proofGenDir/tex/constraint-sat-lemmas.tex}
 
 \subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-lemmas.tex}
+\input{\proofGenDir/tex/alg-pos-subtyping-lemmas.tex}
 
 \subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-lemmas.tex}
+\input{\proofGenDir/tex/constraint-merge-lemmas.tex}
 
 \subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-lemmas.tex}
+\input{\proofGenDir/tex/alg-neg-subtyping-lemmas.tex}
 
 \section{Properties of the Declarative Typing}
-\input{\genDir/tex/decl-typing-lemmas.tex}
+\input{\proofGenDir/tex/decl-typing-lemmas.tex}
 
@@
 \subsection{Singularity and Minimal Instantiation}
-\input{\genDir/tex/singularity-lemmas.tex}
+\input{\proofGenDir/tex/singularity-lemmas.tex}
 
 \subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-lemmas.tex}
-
-\printbibliography
-
-\end{document}
+\input{\proofGenDir/tex/alg-typing-lemmas.tex}

### 2024-07-09 proofs/main.tex
WIP: examples, overview
Edit: sections 1-5
edit and prepare for jfp

--- 2024-02-13 proofs/main.tex
@@
 \newcommand{\genDir}{_gen}
+\newcommand{\proofGenDir}{_gen}
 \input{\ottDir/defs.tex}
@@
 
-\section{Declarative Type Systems}
-
-\subsection{Grammar}
-We assume that there is an infinite set of positive and 
-negative \emph{type} variables. Positive type variables are denoted as 
-$[[α⁺]]$, $[[β⁺]]$, $[[γ⁺]]$, etc.
-Negative type variables are denoted as $[[α⁻]]$, $[[β⁻]]$, $[[γ⁻]]$, etc.
-We assume there is an infinite set of \emph{term} variables,
-which are denoted as $[[x]]$, $[[y]]$, $[[z]]$, etc.
-A list of objects (variables, types or terms) is denoted by
-an overline arrow. For instance, $[[pas]]$ is a list of positive type variables, 
-$[[nbs]]$ is a list of negative type variables, 
-$[[args]]$ is a list of values, which are arguments of a function.
-$[[fv(iP)]]$ and $[[fv(iN)]]$ denote the set of free variables 
-in a type $[[iP]]$ and $[[iN]]$, respectively.
-
-\bigskip
-
-% \ottgrammartabular{
-%   \ottiP\ottinterrule
-%   \ottiN\ottinterrule
-% }
-
-\begin{definition}[Declarative Types]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottiN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottiP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Equalities}
-For simplicity, we assume alpha-equivalent terms are equal.
-This way, we assume that substitutions do not capture bound variables.
-Besides, we equate
-$[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
-as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$,
-and lift these equations transitively and congruently 
-to the whole system.
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Declarative Type Context]
-  \hfill \\
-  Declarative type context $[[Γ]]$ is represented by a set of 
-  type variables. The concatenation $[[Γ1, Γ2]]$ means the 
-  union of two contexts $[[Γ1 ∪ Γ2]]$.
-\end{definition}
-
-$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ denote that the type is well-formed in the context $[[Γ]]$,
-which means that each free type variable of the type is contained in $[[Γ]]$
-(it will be shown later in \cref{lemma:wf-soundness,lemma:wf-ctxt-equiv}).
-
-Notice that checking the well-formedness of a type
-is an \emph{algorithmic} procedure, in which 
-both the context and the type are considered inputs.
-In other words, it is syntax-directed and mode-correct 
-(according to \cite{dunfieldBidirectionalTyping2020}), 
-which means that checking the well-formedness of a type
-can be done recursively by a deterministic algorithm. 
-We will use the well-formedness checking in the inference algorithm, 
-for example, to check that the existential variables do not escape their scope.
-
-\begin{algorithm}[Type Well-formedness]
-  \label{alg:wf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFTNiWFLabeled{}
-  \columnbreak
-
-  \ottdefnWFTPiWFLabeled{}
-  \end{multicols}
-
-\end{algorithm}
-
-\subsection{Substitutions}
-
-\begin{definition}[Substitution]
-  Substitutions (denoted as $[[σ]]$) 
-  are represented by total functions from variables to types, preserving the polarity. 
-\end{definition}
-
-\begin{algorithm}[Substitution Application]
-  Substitution application is denoted as $[[ [σ]iP ]]$ and $[[ [σ]iN ]]$.
-  It is defined naturally as follows:
-    \begin{multicols}{2}
-      \begin{itemize}
-        \item[] $[[ [σ]α⁺ ]] = [[σ]] ([[α⁺]])$
-        \item[] $[[ [σ]α⁻ ]] = [[σ]] ([[α⁻]])$
-        \item[] $[[ [σ]↓iN ]] = [[↓[σ]iN]]$
-        \item[] $[[ [σ]↑iP ]] = [[↑[σ]iP]]$
-        \item[] $[[ [σ](iP → iN) ]] = [[ [σ]iP → [σ]iN ]]$
-        \item[] $[[ [σ]∃nas.iQ ]] = [[∃nas.[σ]iQ]]$ 
-        \item[] $[[ [σ]∀pas.iN ]] = [[∀pas.[σ]iN]]$ (assuming the variable capture never happens)
-      \end{itemize}
-    \end{multicols}
-\end{algorithm}
-
-\begin{definition}[Substitution Signature]
-  The signature $[[Γ' ⊢ σ : Γ]]$ means that
-  \begin{enumerate}
-    \item for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ]α± ]]$; and
-    \item for any $[[α± ∉ Γ']], [[ [σ]α± = α± ]]$.
-  \end{enumerate}
-\end{definition}
-
-A substitution can be restricted to a set of variables. 
-The restricted substitution is define as expected. 
-\begin{definition}[Subsitution Restriction]
-  The specification $[[σ  | varset]]$ is defined as
-  a function such that 
-  \begin{enumerate}
-    \item $[[σ|varset]]([[α± ]]) = [[σ]]([[α± ]])$, if $[[α± ]] \in [[varset]]$; and
-    \item $[[σ|varset]]([[α± ]]) = [[α± ]]$, if $[[α± ]] \notin [[varset]]$.
-  \end{enumerate}
-\end{definition}
-
-Two substitutions can be composed in two ways:
-$[[σ2 ○ σ1]]$ corresponds to a consecutive application of $[[σ1]]$ and $[[σ2]]$,
-while $[[σ2 <=< σ1]]$
-depends on a signature of $[[σ1]]$ and modifies $[[σ1]]$ by applying
-$[[σ2]]$ to its results on the domain.
-\begin{definition}[Substitution Composition]
-  $[[σ2 ○ σ1]]$ is defined as a function such that
-  $[[σ2 ○ σ1]]([[α± ]]) = [[σ2]]([[σ1]]([[α± ]]))$.
-\end{definition}
-
-\begin{definition}[Monadic Substitution Composition]
-  Suppose that $[[Γ' ⊢ σ1 : Γ]]$.
-  Then we define $[[σ2 <=< σ1]]$ as $[[(σ2 ○ σ1)|Γ]]$.
-\end{definition}
-Notice that the result of $[[σ2 <=< σ1]]$ depends on the 
-specification of $[[σ1]]$, which is not unique. 
-However, we assume that the used specification clear from the 
-context of the proof.
-
-\begin{definition}[Equivalent Substitutions]
-  The substitution equivalence judgement $[[Γ' ⊢ σ1 ≈ σ2 : Γ]]$ 
-  indicates that on the domain $[[Γ]]$, 
-  the result of $[[σ1]]$ and $[[σ2]]$ are equivalent in context $[[Γ']]$.
-  Formally, for any $[[α± ∊ Γ]], [[ Γ' ⊢ [σ1]α± ≈ [σ2]α± ]]$.
-\end{definition}
-
-Sometimes it is convenient to construct substitution 
-explicitly mapping each variable from a list (or a set)
-to a type. Such substitutions are denoted as $[[iPs / pas]]$
-and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
-the corresponding types.
-\begin{definition}[Explicit Substitution]
-  \hfill
-  \begin{itemize}
-    \item [$-$]
-      Suppose that $[[nas]]$ is a list of negative type variables,
-      and $[[iNs]]$ is a list of negative types of the same length.
-      Then $[[iNs / nas]]$ denotes a substitution such that 
-      \begin{enumerate}
-        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
-        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
-      \end{enumerate}
-    \item [$+$]
-      Positive explicit substitution $[[iPs / pas]]$
-      is defined symmetrically.
-  \end{itemize}
-\end{definition}
-
-\subsection{Declarative Subtyping}
-Subtyping is one of the key mechanisms of our system. 
-It realizes the polymorphism: abstract $[[∀]]$ and 
-$[[∃]]$ types can be used where concrete types are expected,
-exactly because of the subtyping relation between them.
-
-\begin{definition} 
-  \label{def:subDOne}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNsubLabeled{}
-
-    \ottdefnDOnePsupLabeled{}
-  \end{multicols}
-  \hfill
-
-  \begin{multicols}{2}
-    \ottdefnDOneNeqLabeled{}
-
-    \ottdefnDOnePeqLabeled{}
-  \end{multicols}
-\end{definition}
-
-The following observations about the declarative subtyping are worth noting:
-\begin{itemize}
-  \item \ruleref{\ottdruleDOneNVarLabel} and \ruleref{\ottdruleDOnePVarLabel}
-    make the subtyping reflexive on variables (and further, on any type).
-  \item \ruleref{\ottdruleDOneArrowLabel} is standard: the arrow is covariant on the
-    resulting type and contravariant on the argument type.
-  \item \ruleref{\ottdruleDOneShiftDLabel}  and \ruleref{\ottdruleDOneShiftULabel} are non-standard:
-    the subtyping is \emph{invariant} for shifts. 
-    This way, the subtyping of shifted types in one direction implies the subtyping
-    in the opposite direction.
-    Although this rule restricts the
-    subtyping relation, it makes the system decidable.
-  \item \ruleref{\ottdruleDOneForallLabel} and \ruleref{\ottdruleDOneExistsLabel} are the only
-    non-algorithmic rules: the substitution for the quantified variable is
-    not specified, those, these rules `drive' the subtyping relation.
-\end{itemize}
-
-In the next section, we present the sound and complete algorithm
-checking whether one type is a subtype of another according to \cref{def:subDOne}.
-
-\section{Algorithmic Type System}
-
-\subsection{Grammar}
-
-In the algorithmic system, we extend the grammar of types
-by adding positive and negative \emph{algorithmic variables}
-($[[α̂⁺]]$, $[[β̂⁺]]$, $[[γ̂⁺]]$, etc. and $[[α̂⁻]]$, $[[β̂⁻]]$, $[[γ̂⁻]]$, etc.).
-They represent the unknown types, which will be inferred by the algorithm.
-This way, we add two base cases to the grammar of 
-positive and negative types and use highlight to denote that the type
-can potentially contain algorithmic variables.
-
-\begin{definition}[Algorithmic Types]
-  \label{def:algo-types}
-  \hfill\\
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottuN\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottuP\ottinterrule
-    }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Fresh Variable Selection}
-\label{sec:fresh-selection}
-Both the subtyping and the type inference algorithm
-rely on the ability to select fresh, unused variables.
-For a set of variables $[[varset]]$, it is indicated as 
-$[[varset are fresh]]$ in the inference rules.
-We assume that the selection subroutine always succeeds and is 
-deterministic. In other words, whenever it is called in 
-an algorithmic inference rule, it returns the same result, 
-uniquely determined by the input of this rule.
-
-\subsection{Variable Algorithmization}
-\label{sec:variable-algorithmization}
-
-In several places of our algorithm, in particular, during
-algorithmic subtyping,
-we turn a declarative type into an algorithmic one
-via replacing certain type variables with fresh algorithmic variables.
-We call this procedure \emph{variable algorithmization}, and define it as follows.
-
-\begin{definition}[Variable Algorithmization]
-  Suppose that $[[nas]]$ is a list of negative type variables
-  and $[[nuas]]$ is a list of negative algorithmic variables of the same length. 
-  Then $[[ nuas/nas ]]$ is a substitution-like procedure replacing each $[[αi⁻ ∊ {nas}]]$
-  in a type for $[[αî⁻ ∊ {nuas}]]$.
-\end{definition}
-
-Conversely, we have the opposite procedure turning algorithmic type variables
-into declarative type variables via \emph{dealgorithmization}.
-
-\begin{definition}[Variable Dealgorithmization]
-  Suppose that $[[nuas]]$ is a list of negative algorithmic variables
-  and $[[nas]]$ is a list of negative type variables of the same length. 
-  Then $[[ nas/nuas ]]$ is a substitution-like procedure replacing each
-  $[[αî⁻ ∊ {nuas}]]$ in a type for $[[αi⁻ ∊ {nas}]]$.
-\end{definition}
-
-\subsection{Contexts and Well-formedness}
-
-\begin{definition}[Algorithmic Type Context $[[Ξ]]$]
-  \hfill \\
-  Algorithmic type context $[[Ξ]]$ is represented by a set of 
-  \emph{algorithmic} type variables ($[[α̂⁺]]$, $[[α̂⁻]]$, $[[β̂⁺]]$, \dots).
-  The concatenation $[[Ξ1, Ξ2]]$ means the union of two contexts $[[Ξ1 ∪ Ξ2]]$.
-\end{definition}
-
-$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ are used to denote
-that the algorithmic type is well-formed in the contexts
-$[[Γ]]$ and $[[Ξ]]$, which means that each algorithmic variable
-of the type is contained in $[[Ξ]]$, and each free declarative type variable
-of the type is contained in $[[Γ]]$.
-
-\begin{algorithm}[Algorithmic Type Well-formedness]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnWFATNauWFLabeled{}
-  \columnbreak
-
-  \ottdefnWFATPauWFLabeled{}
-  \end{multicols}
-
-\end{algorithm}
-
-Algorithmic Type Context are used in the unification algorithm.
-In the subtyping algorithm, 
-the context needs to remember additional information.
-In the subtyping context, each algorithmic variable is associated with a
-context it must be instantiated in 
-(i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
-  \hfill \\
-  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
-  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
-  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
-  and $[[Γ]]$ is a context in which they must be instantiated.
-  We assume that no two entries associating the same variable
-  appear in $[[Θ]]$.
-
-  $[[dom(Θ)]]$ denotes the set of variables appearing in $[[Θ]]$:
-  $[[dom(Θ)]] = \{ [[α̂±]] \mid [[α̂±[Γ] ]] \in [[Θ]] \}$.
-  If $[[ α̂±[Γ] ]] \in [[Θ]]$, we denote $[[Γ]]$ as $[[Θ(α̂±)]]$.
-\end{definition}
-
-\subsection{Subsitutions}
-
-Substitution that operates on algorithmic type variables is denoted as
-$[[uσ]]$. It is defined as a total function from algorithmic 
-type variables to \emph{declarative} types, preserving the polarity.
-
-The signature $[[Θ ⊢ uσ : Ξ]]$ means that
-$[[Ξ ⊆ dom(Θ)]]$ and 
-$[[uσ]]$ maps each algorithmic variable 
-from $[[Ξ]]$ to a type well-formed in $[[Θ(α̂±)]]$;
-and for each variable not appearing in $[[dom(Θ)]]$, 
-it acts as identity.
-
-\begin{definition}[Signature of Algorithmic Substitution]
-  \label{def:algo-subst-sig}
-  \hfill
-  \begin{itemize}
-    \item $[[Θ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$,
-          there exists $[[Γ]]$ such that $[[ α̂±[Γ] ∊ Θ ]]$
-          and $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-    \item $[[Γ ⊢ uσ : Ξ]]$ means that
-      \begin{enumerate}
-        \item for any $[[α̂± ∊ Ξ]]$, $[[ Γ ⊢ [uσ]α̂± ]]$; 
-        \item for any $[[ α̂± ∉ Ξ]]$, $[[ [uσ]α̂± ]] =  [[ α̂± ]]$.
-      \end{enumerate}
-  \end{itemize}
-\end{definition}
-
-In the anti-unification algorithm, we use another kind of substitution.
-In contrast to algorithmic substitution $[[uσ]]$,
-it allows mapping algorithmic variables to
-\emph{algorithmic} types.
-Additionally, anti-unification substitution is restricted to the
-\emph{negative} segment of the language.
-Anti-unification substitution is denoted as $[[aus]]$ and $[[ausr]]$.a
-
-The pair of contexts $[[Γ]]$ and $[[Ξ]]$,
-in which the results of an anti-unification substitution 
-are formed, is fixed for this substitution.
-This way, $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that $[[aus]]$ maps each negative algorithmic
-variable appearing in $[[Ξ1]]$ to a term well-formed in $[[Γ]]$ and $[[Ξ2]]$.
-
-\begin{definition}[Signature of Anti-unification substitution]
-  $[[Γ; Ξ2 ⊢ aus : Ξ1]]$ means that
-  \begin{enumerate}
-    \item for any $[[ α̂⁻ ∊ Ξ1]]$, $[[ Γ; Ξ2 ⊢ [aus]α̂⁻ ]]$ and
-    \item for any $[[ α̂⁻ ∉ Ξ1]]$, $[[ [aus]α̂⁻ = α̂⁻ ]]$.
-  \end{enumerate}
-\end{definition}
-
-\subsection{Equivalence and Normalization}
-\label{sec:equivalence-normalization}
-
-The subtyping-induced equivalence (\cref{def:subDOne}) is non-trivial:
-there are types that are subtypes of each other but not equal. 
-For example, $[[∀α⁺,β⁺.α⁺ → ↑β⁺]]$ is a subtype and a supertype of $[[∀α⁺,β⁺.β⁺ → ↑α⁺]]$
-and of, for example, $[[∀α⁺,β⁺.β⁺ → ↑∃γ⁻.α⁺]]$, 
-although these types are not alpha-equivalent.
-For the subtyping algorithm, it is crucial to be able to check whether
-two types are equivalent, without checking mutual subtyping. 
-For this purpose we define the normalization procedure, 
-which allows us to uniformly choose the representative type of the equivalence class.
-This way, the equivalence checking is reduced to normalization and equality checking.
-
-For clarification of the proofs and better understanding of the system, 
-we introduce an intermediate relation---\emph{declarative equivalence}. 
-As will be shown in \cref{lemma:equiv-soundness,lemma:equiv-completeness}, 
-this relation is equivalent to the subtyping-induced equivalence, but does not 
-depend on it. Although this relation is not defined algorithmically, 
-it gives the intuition of what types our system considers equivalent.
-Specifically, in addition to \emph{alpha-equivalence}, 
-our system allows for \emph{reordering of adjacent quantifiers},
-and \emph{introduction/elimination of unused quantifiers}.
-
-The non-trivial rules of the declarative equivalence are
-\ruleref{\ottdruleEOneForallLabel} and \ruleref{\ottdruleEOneExistsLabel}.
-Intuitively, the variable bijection $[[μ]]$ reorders the quantifiers before
-the recursive call on the body of the quantified type. 
-It will be covered formally in \cref{sec:decl-equiv-lemmas}.
-
-\begin{definition}[Declarative Type Equivalence]
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnEOneNeqLabeled{}
-  \columnbreak\\
-  \ottdefnEOnePeqLabeled{}
-  \end{multicols}
-
-\end{definition}
-
-As the equivalence includes arbitrary reordering of quantified variables,
-the normalization procedure is needed to choose the canonical order.
-For this purpose, we introduce an auxiliary procedure---variable ordering. 
-Intuitively, $[[ord varset in iN]]$ returns a list of variables from $[[varset]]$
-in the order they appear in $[[iN]]$.
-
-\begin{algorithm}[Variable Ordering]
-  \label{alg:var-ordering}
-  \hfill
-
-  \ottdefnONVarLabeled{}
-  \ottdefnOPVarLabeled{}
-
-  Analogously, the variable can be ordered in 
-  an \emph{algorithmic} type ($[[ord varset in uP]]$ and 
-  $[[ord varset in uN]]$). In these cases, we treat the algorithmic variables
-  as if they were declarative variables.
-
-\end{algorithm}
-
-Next, we use the variable ordering in the normalization procedure. 
-Specifically, normalization recursively traverses the type, 
-and for each quantified case reorders the quantified variables in a 
-canonical order dictated by \cref{alg:var-ordering}, removing unused ones.
-
-\begin{algorithm}[Type Normalization]
-  \label{alg:type-nf}
-  \hfill
-
-  \begin{multicols}{2}
-  \ottdefnNrmNNormLabeled{}
-  \columnbreak\\
-  \ottdefnNrmPNormLabeled{}
-  \end{multicols}
-
-  Analogously, we define the normalization of algorithmic types by adding base cases:
-
-  \begin{multicols}{2}
-  \ottdefnNrmuNNormLabeled{}
-  \columnbreak\\
-  \ottdefnNrmuPNormLabeled{}
-  \end{multicols}
-
-\end{algorithm}
-
-\Cref{lemma:subt-equiv-algorithmization}
-demonstrates that the equivalence of types is the same
-as the equality of their normal forms.
-\begin{theorempreview}[Correctness of Normalization]
-  Assuming the types are well-formed in $[[Γ]]$, 
-  \begin{itemize}
-    \item [$-$] $[[Γ ⊢ iN ≈ iM]]$ if and only if $[[nf(iN) = nf(iM)]]$;
-    \item [$+$] $[[Γ ⊢ iP ≈ iQ]]$ if and only if $[[nf(iP) = nf(iQ)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\begin{algorithm}[Substitution Normalization]
-  For a substitution $[[σ]]$, we define $[[nf(σ)]]$
-  as a substitution that maps $[[α±]]$ into $[[nf([σ]α±)]]$.
-\end{algorithm}
-
-The rest of this chapter is devoted to the
-central algorithm of the type system---the subtyping algorithm. 
-\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
-The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
-$A$ uses $B$ as a sub-procedure.
-
-\begin{figure}[h]
-  \centering
-  \begin{tikzpicture}
-    [>={Stealth[scale=2]},node distance=2.4cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
-
-    % Define nodes
-    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
-    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
-    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
-
-    % Define edges
-    \draw[->] (1) to (2);
-    \draw[->] (1) to (3);
-    \draw[->] (1) to (5);
-    \draw[->] (2) to (3);
-    \draw[->] (2) to (6);
-    \draw[->] (3) to (4);
-    \draw[->] (3) to (5);
-    \draw[->] (4) to (6);
-    \draw[->] (5) to (8);
-    \draw[->] (6) to (7);
-  \end{tikzpicture}  
-  \caption{Dependency graph of the subtyping algorithm}
-  \label{fig:subtyping-algo}
-\end{figure}
-
-\subsection{Subtyping}
-\label{sec:subtyping}
-
-Now, we present the subtyping algorithm itself.
-Although the algorithm is presented as a single procedure,
-is important for the structure of the proof that the positive subtyping algorithm
-does not invoke the negative one. This way, the correctness of the positive 
-subtyping will be proved independently and used afterwards to prove the
-correctness of the negative subtyping.
-
-\begin{algorithm}[Subtyping]
-  \label{alg:subtyping}
-  \hfill\\
-  \ottdefnANsubLabeled{}
-  \ottdefnAPsupLabeled{}
-\end{algorithm}
-
-The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
-the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
-must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
-and $[[uP]]$ and $[[iQ]]$ for the positive case. 
-As one of the invariants, we require
-$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
-The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
-which will be discussed in the next section.
-
-Let us overview the inference rules of the subtyping algorithm.
-\begin{itemize}
-  \item \ruleref{\ottdruleANVarLabel} and \ruleref{\ottdruleAPVarLabel} 
-    are the base cases. They copy the corresponding declarative rules and
-    ensure reflexivity.
-  \item \ruleref{\ottdruleAPUVarLabel} is the only case generating 
-    subtyping constraints. In this case, we must ensure
-    that the resulting constraints guarantee that the instantiation of 
-    $[[â⁺]]$ is a supertype of $[[iP]]$.
-    However, the obvious constraint $[[â⁺ :≥ iP]]$ might be problematic
-    if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
-    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
-    to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
-
-    Notice that this rule does not have a negative counterpart. This is
-    because one of the important invariants of the algorithm: 
-    in the negative subtyping, only positive algorithmic variables
-    can occur in the types.
-
-  \item \ruleref{\ottdruleAShiftDLabel} and \ruleref{\ottdruleAShiftULabel} are the
-    \emph{shift} rules. According to the declarative system,
-    shifted subtyping requires equivalence. In the presence of the algorithmic 
-    variables, it means that the left and the right-hand sides of the subtyping
-    must be unified. Hence, the shift rules invoke the unification algorithm, 
-    which will be discussed in \cref{sec:unification}. The unification 
-    returns the minimal set of constraints $[[UC]]$, which is necessary
-    and sufficient for the subtyping.
-
-  \item \ruleref{\ottdruleAArrowLabel}.
-    In this case, the algorithm makes two calls:
-    a recursive call to the negative subtyping algorithm for the argument types,
-    and a call to the positive subtyping algorithm for the result types.
-    After that, the resulting constraints are merged using the
-    \emph{subtyping constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
-    are symmetric. These are the only places where 
-    the algorithmic variables are introduced.
-    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
-    of the quantified variables: these variables are replaced by 
-    fresh algorithmic variables in the body of the quantified type,
-    the algorithmic variables are added to the subtyping context $[[Θ]]$,
-    after that, the recursive call is made. Notice that the declarative context
-    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
-    which matches the declarative system.
-\end{itemize}
-
-Then soundness lemma (\cref{lemma:pos-subt-soundness,lemma:neg-subt-soundness}) 
-and completeness (\cref{lemma:pos-subt-completeness,lemma:neg-subt-completeness})
-of the algorithm together give us the following simplified theorem:
-
-\begin{theorempreview}[Correctness of subtyping algorithm]
-  \hfill
-  \begin{itemize}
-    \item [$-$]  $[[ Γ ; · ⊨ uN ≤ iM ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iN ≤ iM ]]$;
-    \item [$+$] $[[ Γ ; · ⊨ uP ≥ iQ ⫤ · ]]$ is equivalent to $[[ Γ ⊢ iP ≥ iQ ]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Constraints}
-
-Unification and subtyping algorithms are based on constraint generation.
-The constraints are represented by a set of constraint entries.
-
-\begin{definition}[Unification Constraint]
-  \hfill
-  \begin{description}
-    \item[unification entry] (denoted as $[[ucE]]$) is an expression of shape 
-      $[[pua :≈ iP]]$ or $[[nua :≈ iN]]$;
-    \item[unification constraint] (denoted as $[[UC]]$) is a set of 
-      unification constraint entries.
-      We denote $\{[[α̂±]] \mid [[ucE ∊ UC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(UC)]]$.
-  \end{description}
-\end{definition}
-
-However, in the subtyping, we need to consider more general
-kind of constraints. Specifically,
-subtyping constraint entries can restrict a variable
-not only to be equivalent to a certain type, but
-also to be a supertype of a positive type.
-
-\begin{definition}[Subtyping Constraint]
-  \hfill
-  \begin{description}
-    \item[subtyping entry] (denoted as $[[scE]]$) is an expression of shape 
-      $[[pua :≥ iP]]$, $[[nua :≈ iN]]$, or $[[pua :≈ iP]]$;
-    \item[subtyping constraint] (denoted as $[[SC]]$) is a set of subtyping constraint entries.
-      We denote $\{[[α̂±]] \mid [[scE ∊ SC]] \text{ restricting $[[α̂±]]$ }\}$ 
-      as $[[dom(SC)]]$.
-  \end{description}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint Entry]
-  We say that a constraint entry is well-formed in a context $[[Γ]]$ if
-  its associated type is well-formed in $[[Γ]]$.
-  \begin{itemize}
-    \item[] $[[Γ ⊢ pua :≥ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ pua :≈ iP]]$ ~iff~ $[[Γ ⊢ iP]]$;
-    \item[] $[[Γ ⊢ nua :≈ iN]]$ ~iff~ $[[Γ ⊢ iN]]$.
-  \end{itemize}
-\end{definition}
-
-\begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  subtyping context $[[Θ]]$ if all its entries are well-formed in
-  the corresponding elements of $[[Θ]]$.
-  More formally, 
-  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
-  such that $[[scE]]$ restricts $[[α̂±]]$,
-  we have $[[Θ(α̂±) ⊢ scE]]$.
-
-  We write $[[Θ ⊢ SC : Ξ]]$ to denote
-  that $[[Θ ⊢ SC]]$ and $[[dom(SC) = Ξ]]$.
-
-  $[[Θ ⊢ UC]]$ and $[[Θ ⊢ UC : Ξ]]$ are defined analogously.
-\end{definition}
-
-\subsubsection{Constraint Satisfaction}
-
-A constraint entry restricts a type that can be assigned to a variable.
-We say that a type satisfies a constraint entry if it can be assigned
-to the variable restricted by the entry.
-
-\begin{definition}[Type Satisfying a Constraint Entry]
-  \hfill\\
-  \begin{multicols}{2}
-  \ottdefnSATSCENLabeled{}
-  \columnbreak\\
-  \ottdefnSATSCEPLabeled{}
-  \end{multicols}
-\end{definition}
-
-We say that a substitution satisfies a constraint---a set of constraint 
-entries if each entry is satisfied by the type assigned to the variable
-by the substitution.
-
-\begin{definition}[Substitution Satisfying a Constraint]
-  We write $[[Θ ⊢ uσ : SC]]$ to denote that
-  a substitution $[[uσ]]$ satisfies a constraint $[[SC]]$ in a context $[[Θ]]$.
-  It presumes that $[[Θ ⊢ SC]]$ and 
-   means that for any $[[ucE]] \in [[SC]]$, if $[[ucE]]$ restricts $[[α̂±]]$,
-  then $[[Θ(α̂±) ⊢ [uσ]α̂± : ucE]]$.
-
-  Unification constraint satisfaction $[[Θ ⊢ uσ : UC]]$ 
-  is defined analogously as a special case of subtyping constraint satisfaction.
-
-\end{definition}
-
-Notice that $[[Θ ⊢ uσ : SC]]$ does not 
-imply the signature $[[Θ ⊢ uσ : dom(SC)]]$, because 
-the latter also specifies $[[uσ]]$ outside of the domain $[[dom(SC)]]$
-(see \cref{def:algo-subst-sig}).
-
-\subsubsection{Constraint Merge}
-\label{sec:constraint-merge}
-
-In this section, define the least upper bound 
-for constraints, which we call \emph{merge}.
-Intuitively, the merge of two constraints is the least
-constraint such that any substitution satisfying both constraints
-satisfies the merge as well.
-First, we define the merge of entries,
-and then extend it to the set of entries.
-
-\begin{definition} [Matching Entries]
-  We call two unification constraint entries 
-  or two subtyping constraint entries matching 
-  if they are restricting the same unification variable.
-\end{definition}
-
-Two matching entries formed in the same context $[[Γ]]$ 
-can be merged in the following way:
-\begin{algorithm}[Merge of Matching Constraint Entries]
-  \label{definition:merge-matching-entries}
-   \hfill
-
-  \ottdefnSCMELabeled\\
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSCMEPEqEqLabel} and \ruleref{\ottdruleSCMENEqEqLabel}
-    are symmetric cases. To merge two matching entries restricting
-    a variable to be equivalent to certain types, we check
-    that these types are equivalent to each other.
-    To do so, it suffices to check for \emph{equality} of their normal forms,
-    as discussed in \cref{sec:equivalence-normalization}. 
-    After that, we return the left-hand entry.
-
-  \item \ruleref{\ottdruleSCMEEqSupLabel} and \ruleref{\ottdruleSCMESupEqLabel}
-    are also symmetric. 
-    In this case,
-    since one of the entries requires the variable to be equal to 
-    a type, the resulting entry must also imply that.
-    However, for the soundness, it is needed to ensure that
-    the equating restriction is stronger than the subtyping restriction.
-    For this purpose, the premise invokes the positive subtyping.
-
-  \item \ruleref{\ottdruleSCMESupSupLabel} 
-    In this case, we find the least upper bound of the types from the input
-    restrictions, 
-    and as the output, restrict the variable to be a supertype of the result.
-    The least upper bound procedure will be discussed in \cref{sec:lub}.
-\end{itemize}
-
-Unification constraint entries are a special case of subtyping constraint
-entries. They are merged using the same algorithm 
-(\cref{definition:merge-matching-entries}).
-Notice that the merge of two matching unification constraint entries
-is a unification constraint entry.
-\begin{lemma}[Merge of Matching Unification Constraint Entries is well-defined]
-  \label{lemma:merge-matching-entries-welldef}
-  Suppose that $[[Γ ⊢ ucE1]]$ and $[[Γ ⊢ ucE2]]$
-  are unification constraint entries. 
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  $[[Γ ⊢ ucE1 & ucE2 = ucE]]$
-  according to \cref{definition:merge-matching-entries},
-  is a unification constraint entry.
-\end{lemma}
-\begin{proof}
-  Since $[[ucE1]]$ and $[[ucE2]]$ are matching unification constraint entries,
-  they have the shape $([[pua :≈ iP1]], [[pua :≈ iP2]])$ or
-  $([[nua :≈ iN1]], [[nua :≈ iN2]])$.
-  Then the merge of $[[ucE1]]$ and $[[ucE2]]$ 
-  can only be defined by \ruleref{\ottdruleSCMEPEqEqLabel} or
-  \ruleref{\ottdruleSCMENEqEqLabel}.
-  In both cases the result, if it exists, 
-  is a unification constraint entry:
-  in the first case, the result has shape $[[pua :≈ iP1]]$,
-  in the second case, the result has shape $[[nua :≈ iN1]]$.
-\end{proof}
-
-% Notice that in case of equivalence, the assigned types
-% must be equal (i.e. alpha-equivalent) to be merged. This is because
-% the unification algorithm assumes that every type is normalized,
-% and hence, equivalence is alpha-equivalence 
-% (\cref{corollary:nf-complete-wrt-subt-equiv,corollary:nf-sound-wrt-subt-equiv}).
-
-\begin{algorithm}[Merge of Subtyping Constraints]
-  \label{definition:merge-subtyping-constraints}
-  Suppose that $[[Θ ⊢ SC1]]$ and $[[Θ ⊢ SC2]]$.\\
-  Then $[[Θ ⊢ SC1 & SC2 = SC]]$
-  defines a set of constraints $[[SC]]$ such that $[[scE]] \in [[SC]]$ iff either:
-  \begin{itemize}
-    \item $[[scE]] \in [[SC1]]$ and there is no matching $[[scE']] \in [[SC2]]$; or
-    \item $[[scE]] \in [[SC2]]$ and there is no matching $[[scE']] \in [[SC1]]$; or
-    \item $[[Θ(α̂±) ⊢ scE1 & scE2 = scE]]$ for some $[[scE1]] \in [[SC1]]$ and $[[scE2]] \in [[SC2]]$
-      such that $[[scE1]]$ and $[[scE2]]$ both restrict variable $[[α̂±]]$. 
-  \end{itemize}
-\end{algorithm}
-
-Unification constraints can be considered 
-as a special case of subtyping constraints,
-and the merge of unification constraints
-is defined as the merge of subtyping constraints.
-Then it is easy to see that the merge of two 
-unification constraints is a unification constraint.
-
-\begin{lemma}[Merge of Unification Constraints is well-defined]
-  Suppose that $[[Θ ⊢ UC1]]$ and $[[Θ ⊢ UC2]]$
-  are unification constraints. 
-  Then the merge of $[[UC1]]$ and $[[UC2]]$ 
-  $[[Θ ⊢ lift UC1 & lift UC2 = lift UC]]$
-  according to \cref{definition:merge-subtyping-constraints},
-  is a unification constraint.
-\end{lemma}
-\begin{proof}
-  $[[UC]]$ consists of unmatched entries of $[[UC1]]$ and $[[UC2]]$,
-  which are \emph{unification} constraint entries by assumption,
-  and merge of matching entries, which also are  
-  \emph{unification} constraint entries by \cref{lemma:merge-matching-entries-welldef}.
-\end{proof}
-
-\Cref{lemma:merge-soundness,lemma:merge-completeness} 
-show the correctness and initiality of the merge operation,
-which can be expressed in the following simplified theorem:
-\begin{theorempreview}[Correctness of Constraint Merge]
-  A substitution $[[uσ]]$ satisfying both constraints
-  $[[SC1]]$ and $[[SC2]]$ 
-  if and only if it satisfies their merge.
-\end{theorempreview}
-
-The unification constraint merge satisfies the same theorem,
-however, because the merge of unification constraint entries 
-$[[ucE1]]$ and $[[ucE2]]$ always results in one of them, 
-a stronger soundness property holds (see \cref{lemma:unif-merge-soundness}):
-\begin{theorempreview}[Soundness of Unification Constraint Merge]
-  If $[[Θ ⊢ UC1 & UC2 = UC]]$ then $[[UC = UC1 ∪ UC2]]$.
-\end{theorempreview}
-
-\subsection{Unification}
-\label{sec:unification}
-
-The subtyping algorithm calls the following subtask:
-given two algorithmic types, we need to find the most general substitution 
-for the algorithmic variables in these types, such that the resulting 
-types are equivalent. This problem is known as \emph{unification}.
-
-In our case, the unification is restricted in the following way:
-first, before unifying the types, we normalize them, which 
-allows us to reduce (non-trivial) equivalence to (trivial) equality;
-second, we preserve invariants which guarantee that
-one side of the unification is always declarative, which in fact, 
-reduces the unification to the \emph{matching} problem.
-
-The unification procedure
-returns a set of minimal constraints,
-that must be satisfied by a substitution
-unifying the input types.
-
-\begin{algorithm}[Unification]
-  \hfill
-  \begin{multicols}{2}
-  \ottdefnUNUnifLabeled{}
-  \columnbreak\\
-  \ottdefnUPUnifLabeled{}
-  \end{multicols}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleUShiftULabel}, \ruleref{\ottdruleUShiftDLabel}, 
-    \ruleref{\ottdruleUForallLabel}, and \ruleref{\ottdruleUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers,
-    adding the quantified variables to the context $[[Γ]]$.
-    Notice that $[[Θ]]$, which specifies
-    the contexts in which the algorithmic variables must be instantiated,
-    is not changed.
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel} 
-    are the base cases. 
-    Since the sides are equal and free from algorithmic variables,
-    the unification returns an empty constraint. 
-  \item \ruleref{\ottdruleUNVarLabel} and \ruleref{\ottdruleUPVarLabel}
-    are symmetric cases constructing the constraints. 
-    When an algorithmic variable is unified with a type, 
-    we must check that the type is well-formed in the required context,
-    and if it is, we return a constraint restricting the variable
-    to be equivalent to that type.
-  \item \ruleref{\ottdruleUArrowLabel}.
-    In this case, the algorithm makes two recursive calls:
-    it unifies the arguments and the results of the arrows.
-    After that, the resulting constraints are merged using the
-    \emph{unification constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
-    Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
-    \emph{unification} constraints, not arbitrary \emph{subtyping} 
-    constraints: it is important for modularizing the proofs, 
-    since the properties of the \emph{unification} constraint merge
-    can be proved independently from the \emph{subtyping} constraint merge.
-\end{itemize}
-
-\subsection{Least Upper Bound}
-\label{sec:lub}
-
-In this section, we present
-the algorithm finding the least common supertype of two positive types. 
-It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
-and indirectly, through the type upgrade by positive subtyping
-(\cref{sec:subtyping}). Perhaps, the least upper bound is the least 
-intuitive part of the algorithm, and its correctness will be covered
-in \cref{sec:alg-upper-bounds-proofs}.
-
-\begin{algorithm}[The Least Upper Bound Algorithm]
-  \hfill\\
-  \ottdefnLUBNsubLabeled{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleLUBVarLabel}
-    The base case is trivial: 
-    the least upper bound of to equal variables
-    is the variable itself.
-  \item \ruleref{\ottdruleLUBShiftLabel}
-    In case both sides of the least upper bound are shifted,
-    the algorithm needs to find the anti-unifier of them. 
-    Intuitively, this is because in general, the upper bounds of
-    $[[↓iN]]$ are $[[∃nas.iP]]$ such that 
-    $[[nas]]$ can be instantiated with some $[[iMs]]$ so that
-    $[[ Γ ⊢ [iMs/nas]iP ≈ ↓iN ]]$ (see \cref{lemma:shape-of-supertypes}).
-  \item \ruleref{\ottdruleLUBExistsLabel}
-    In this case, we move the quantified variables to the context $[[Γ]]$, 
-    and make a recursive call. 
-    It is important to make sure that $[[nas]]$ and $[[nbs]]$ are disjoint.
-    In this case, it is guaranteed that the resulting 
-    $[[fv(iQ)]]$ will be free of $[[nas]]$ and $[[nbs]]$,
-    and thus, the resulting type will be a supertype of both sides
-    (it will be discussed in \cref{lemma:shape-of-supertypes}).
-\end{itemize}
-
-In the positive subtyping algorithm (\cref{sec:subtyping}),
-\ruleref{\ottdruleAPUVarLabel} 
-generates a restriction of a variable $[[α̂⁺]]$.
-On the one hand, this restriction must imply 
-$[[â⁺ :≥ iP]]$ for the subtyping to hold.
-On the other hand, the type used in this restriction 
-must be well-formed in a potentially stronger (smaller) 
-context than $[[iP]]$.
-
-To resolve this problem, we define the \emph{upgrade} procedure,
-which for given $[[Δ]]$, $[[pnas]]$, and $[[Δ, pnas ⊢ iP]]$,
-finds $[[Δ ⊢ iQ]]$---the least supertype of $[[iP]]$ 
-among the types well-formed in $[[Δ]]$.
-
-The trick is to make sure that the `forbidden' variables
-$[[pnas]]$ are not used explicitly in the supertypes
-of $[[iP]]$. For this purpose, we
-construct new types $[[iP1]]$ and $[[iP2]]$,
-in each of them replacing the forbidden variables
-with fresh variables $[[pnbs]]$ and $[[pncs]]$,
-and then find the least upper bound of $[[iP1]]$ and $[[iP2]]$.
-It turns out that this renaming forces the common types of 
-$[[iP1]]$ and $[[iP2]]$ to be agnostic to $[[pnas]]$,
-and thus, the supertypes of $[[iP]]$ well-formed in $[[Δ]]$
-are exactly the common supertypes of $[[iP1]]$ and $[[iP2]]$.
-These properties are considered in more details in \cref{sec:upgrade-lemmas}.
-
-\begin{algorithm}[Type Upgrade]
-  \hfill\\
-  \ottdefnLUBUpLabeled{}
-\end{algorithm}
-
-\begin{paragraph}{Note on the Greatest Lower Bound}
-  In contrast to the least upper bound, 
-  the general greatest lower bound does not exist in our system.
-  For instance, consider a positive type $[[iP]]$, 
-  together with its non-equivalent
-  supertypes $[[iP1]]$ and $[[iP2]] \not\simeq [[iP1]]$
-  (for example, $[[iP = ↓↑↓γ⁻]]$, $[[iP1 = ∃α⁻.↓↑↓α⁻]]$, 
-  and $[[iP2 = ∃α⁻.↓α⁻]]$).
-  Then for arbitrary $[[iQ]]$ and $[[iN]]$, 
-  let us consider the common subtypes of 
-  $A = [[iQ → ↓↑iQ → ↓↑iQ → iN]]$ and $B = [[iP → ↓↑iP1 → ↓↑iP2 → iN]]$.
-  It is easy to see that $[[∀α⁺.∀β⁺. α⁺ → ↓↑α⁺ → ↓↑β⁺ → iN]]$ and 
-  $[[∀α⁺.∀β⁺. α⁺ → ↓↑β⁺ → ↓↑α⁺ → iN]]$ are
-  both \emph{maximal} common subtypes of $A$ and $B$,
-  and since they are not equivalent, none of them is 
-  the \emph{greatest} one.
-
-  However, we designed the subtyping system in such a way 
-  that the greatest lower bound is not needed:
-  the negative variables are always `protected'
-  by \emph{invariant} shifts ($[[↑]]$ and $[[↓]]$), 
-  and thus, the algorithm can only require
-  a substitution of a negative variable to be 
-  \emph{equivalent} to some type but never 
-  to be a \emph{subtype}.
-\end{paragraph}
-
-\subsection{Anti-unification}
-\label{sec:antiunification}
-
-Next, we define the anti-unification procedure,
-also known as the \emph{most specific generalization}.
-As an input, it takes two declarative types
-(e.g., in the positive case $[[iP1]]$ and $[[iP2]]$)
-and a context $[[Γ]]$.
-and returns a type $[[uQ]]$---the generalizer,
-containing negative placeholders 
-(represented by algorithmic variables) from 
-$[[Ξ]]$ and two substitutions $[[uτ1]]$ and $[[uτ2]]$.
-The substitutions replace the placeholders with 
-declarative types well-formed in $[[Γ]]$,
-such that $[[ [uτ1]uQ = iP1 ]]$ and $[[ [uτ2]uQ = iP2 ]]$.
-Moreover, the algorithm guarantees that 
-$[[uQ]]$ is the most specific type with this property: 
-any other generalizer can be turned into $[[uQ]]$ by some substitution
-$[[uρ]]$.
-
-It is important to note the differences between 
-the standard anti-unification and our version.
-First, we only allow the placeholders at \emph{negative} positions,
-which means, for example, that $[[α⁺]]$ and $[[β⁺]]$ cannot be
-generalized. Second, the generated pair of substitutions 
-$[[uτ1]]$ and $[[uτ2]]$ must replace the placeholders with 
-types well-formed in a specified context $[[Γ]]$.
-
-The anti-unification algorithm assumes that the input types 
-are normalized. This way, anti-unification up-to-equality rather than 
- anti-unification up-to-equivalence is sufficient.
-
-\begin{algorithm}[Anti-unification]
-  \hfill
-
-  \ottdefnAUAUPLabeled{}
-  \ottdefnAUAUNLabeled{}
-
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleAUPVarLabel} and \ruleref{\ottdruleAUNVarLabel}
-    are the base cases. 
-    In this case, since the input types are equal, 
-    the algorithm returns this type as a generalizer,
-    without generating any placeholders.
-
-  \item \ruleref{\ottdruleAUShiftDLabel}, \ruleref{\ottdruleAUShiftULabel},
-    \ruleref{\ottdruleAUForallLabel}, and \ruleref{\ottdruleAUExistsLabel}
-    are defined congruently. In the shift rules, the algorithm
-    removes the outermost constructor. In the
-    $[[∀]]$ and $[[∃]]$ rules, it removes the quantifiers.
-    Notice that the algorithm does not add the removed variables to
-    the context $[[Γ]]$. This is because $[[Γ]]$
-    is used to restrict the resulting anti-unification substitutions, 
-    and is fixed throughout the algorithm.
-
-  \item \ruleref{\ottdruleAUAULabel} is the most important rule, 
-    since it generates the placeholders. 
-    This rule only applies if other negative rules failed.
-    Because of that, the anti-unification procedure is 
-    \emph{not} syntax-directed.
-
-    The generated placeholder is indexed with a pair of 
-    types it is mapped to. It allows the algorithm to 
-    automatically unite the anti-unification solutions 
-    generated by the different branches of 
-    \ruleref{\ottdruleAUArrowLabel}.
-
-    Notice that this rule does not have a positive counterpart,
-    since we only allow negative placeholders.
-
-  \item \ruleref{\ottdruleAUArrowLabel}
-    makes two recursive calls to the anti-unification procedure,
-    and unites the results. Suppose that
-    $[[uτ1]]$  and $[[uτ2]]$ are the substitutions generated by
-    anti-unification of \emph{argument} types of the arrow,
-    and $[[uτ1']]$ and $[[uτ2']]$ are the substitutions generated by 
-    anti-unification of \emph{result} types of the arrow.
-    It is important that if ($[[uτ1]]$, $[[uτ2]]$)
-    and ($[[uτ1']]$, $[[uτ2']]$) send some variables to the same pair of types,
-    i.e., $[[ [uτ1]α̂⁻ = [uτ1']β̂⁻]]$ and $[[ [uτ2]α̂⁻ = [uτ2']β̂⁻]]$,
-    then these variables are equal, i.e., $[[α̂⁻ = β̂⁻]]$.
-    This property is guaranteed by \ruleref{\ottdruleAUAULabel}:
-    the name of the placeholder is determined by the pair of 
-    types it is mapped to.
-\end{itemize}
-
-\newpage
-
-\section{Declarative Typing}
-
-In the previous section, we presented the 
-type system together with subtyping specification 
-and the algorithm. In this section, 
-we describe the language under this type system, 
-together with the type inference specification
-and algorithm.
-
-\subsection{Grammar}
-
-First, we define the syntax of the language.
-The language combines System F with call-by-push-value 
-style. 
-\begin{definition}[Language Grammar]
-  \hfill
-  \begin{multicols}{2}
-    \ottgrammartabular{
-      \ottc\ottinterrule
-    }
-
-    \ottgrammartabular{
-      \ottv\ottinterrule
-    }
-  \end{multicols}
-\end{definition}
-Notice that the language does not have
-first-class applications: instead, 
-we use applicative let bindings---
-constructions that bind a result of
-a fully applied function to a (positive) 
-variable.
-In the call-by-push-value paradigm, 
-it corresponds to monadic bind or do-notation. 
-Typewise, these let-binders come in two forms:
-annotated and unannotated. The annotated let-binders
-$[[let x:iP = v(args); c]]$ 
-requires the application to infer the annotated
-$[[iP]]$, whereas the unannotated 
-$[[let x = v(args); c]]$ 
-is used when the inferred type is unique.
-
-A computation of a polymorphic type is constructed 
-using $[[Λα⁺ . c]]$, however, the elimination of $[[∀]]$
-is implicit. Conversely, the existential types
-are constructed implicitly and eliminated 
-using the standard unpack mechanism:
-$[[let∃ (nas, x) = v; c]]$.
-
-Another dual pair of constructions are 
-$[[return v]]$ and $[[{c}]]$. 
-The former allows us to embed a value in pure 
-computations. The latter, on the contrary,
-encapsulates a thunk of computation in a value.
-
-Finally, the language has several standard constructions:
-lambda-abstractions $[[λx:iP.c]]$,
-standard let-bindings $[[let x = v; c]]$,
-and type annotations that can be added to any value or computation:
-$[[(v:iP)]]$ and $[[(c:iN)]]$.
-
-\subsection{Declarative Type Inference}
-
-Next, we define the specification of the type 
-inference for our language. First, we introduce 
-variable context specifying the types of variables 
-in the scope of the current rule.
-
-\begin{definition}[Variable Context]
-  The variable typing context $[[Φ]]$
-  is represented by a set of entries of the form
-  $[[x : iP]]$. 
-\end{definition}
-
-The specification is represented by an inference system of
-three mutually recursive judgments:
-positive inference $[[Γ ; Φ ⊢ v : iP]]$,
-negative type inference $[[Γ ; Φ ⊢ c : iN]]$, and
-application type inference $[[Γ ; Φ ⊢ iN ● args ⇒> iM ]]$.
-In the premises, the inference rules also refer to 
-the declarative subtyping (\cref{def:subDOne}),
-type well-formedness (\cref{alg:wf}), 
-and normalization (\cref{alg:type-nf}).
-\begin{itemize}
-  \item $[[Γ ; Φ ⊢ v : iP]]$ (and symmetrically, $[[Γ ; Φ ⊢ c : iN]]$)
-    means that under the type context $[[Γ]]$ and
-    the variable context $[[Φ]]$, for the value $[[v]]$,
-    type $[[iP]]$ is inferrable. It guarantees that 
-    $[[v]]$ is well-formed in $[[Γ]]$ and $[[Φ]]$ in
-    the standard sense.
-  \item $[[ Γ ; Φ ⊢ iN ● args ⇒> iM ]]$ is the application type inference 
-    judgment. It means that if a head of type $[[iN]]$ 
-    is applied to list of values $[[args]]$, 
-    then the resulting computation can be typed as $[[iM]]$.
-\end{itemize}
-
-\begin{definition}[Declarative Type Inference]
-  \label{def:declarative-typing}
-  \ottdefnsDTLabeled
-\end{definition}
-
-Let us discuss the selected rules of the declarative system:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTVarLabel}
-    says that the type of a variable is inferred from the context.
-  \item \ruleref{\ottdruleDTThunkLabel} says
-    that the type of a thunk is inferred by shifting up the type of the 
-    contained computation. Symmetrically, \ruleref{\ottdruleDTReturnLabel}
-    infers the type of a return by shifting down the type of the
-    contained value.
-  \item \ruleref{\ottdruleDTPAnnotLabel} and \ruleref{\ottdruleDTNAnnotLabel} are symmetric.
-    They allow the inferred type to be refined by annotating it with a supertype.
-  \item \ruleref{\ottdruleDTNEquivLabel} and \ruleref{\ottdruleDTPEquivLabel}
-    mean that the declarative system allows us to infer any type from the equivalence class.
-  \item \ruleref{\ottdruleDTUnpackLabel} is standard for existential types,
-    and its first premise infers the existential type of the value being unpacked.
-    It is important however that the inferred existential type is normalized. 
-    This is because there might be multiple equivalent existential types 
-    with a different order or even number of quantified variables, 
-    and to bind them, the algorithm needs to fix the canonical one.
-  \item \ruleref{\ottdruleDTAppLetAnnLabel} allows us to accommodate the applications
-    with annotated let-bindings. The first premise infers the type of the head of the application,
-    which must be a thunked computation. Then if after applying it 
-    to the arguments, the resulting type can be equated to the annotated one,
-    we infer the body of the let-binding in the context extended with the bound variable.
-  \item \ruleref{\ottdruleDTAppLetLabel} is similar to \ruleref{\ottdruleDTAppLetAnnLabel},
-    bus is used for unannotated let-bindings. In this case, 
-    we require the type application to infer the `canonical' principal type.
-    $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ principal]]$ means that
-    any other type $[[iQ']]$ inferrable for the application (\ie $[[Γ ; Φ ⊢ iM ● args ⇒> ↑iQ']]$)
-    is greater than the principal type $[[iQ]]$, \ie $[[Γ ⊢ iQ' ≥ iQ]]$.
-\end{itemize}
-
-Let us discuss the rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleDTEmptyAppLabel} 
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head. 
-    However, we relax this specification by allowing it to 
-    infer any other equivalent type. 
-    The relaxation of this rule is enough to guarantee 
-    this property for the whole judgement:
-    if $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ then 
-    $[[Γ ; Φ ⊢ iN ● args ⇒> iM']]$ for any equivalent 
-    $[[iM']]$.
-  \item \ruleref{\ottdruleDTArrowAppLabel}
-    is where the application type is inferred: 
-    if the head has an arrow type $[[iQ → iN]]$,
-     we are allowed to apply it as soon as 
-    as soon as the first argument has a type, which is a subtype of $[[iQ]]$.
-  \item \ruleref{\ottdruleDTForallAppLabel}
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    If we are applying a polymorphic computation, 
-    we can instantiate its quantified variables with any types,
-    which is expressed by the substitution $[[Γ ⊢ σ : {pas}]]$.
-\end{itemize}
-
-\section{Algorithmic Typing}
-
-Next, we present the type inference algorithm, 
-which is sound and complete with respect to the declarative specification
-(\cref{def:declarative-typing}).
-
-\subsection{Algorithmic Type Inference}
-
-Mirroring the declarative typing, 
-the algorithm is represented by an inference system of three mutually recursive
-judgments:
-\begin{itemize}
-  \item $[[Γ ; Φ ⊨ v : iP]]$ and  $[[Γ ; Φ ⊨ c : iN]]$
-    are the algorithmic versions of $[[Γ ; Φ ⊢ v : iP]]$ and $[[Γ ; Φ ⊢ c : iN]]$.
-    In contrast with the declarative counterparts, they are deterministic,
-    and guarantee that the inferred type is normalized. 
-  \item $[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$
-    is the algorithmization of $[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$.
-    Notice that $[[uN]]$ contains algorithmic variables, 
-    which are specified by the context $[[Θ1]]$.
-    Moreover, the inferred type $[[uM]]$ is also algorithmic,
-    and can have several non-equivalent instantiations. To accommodate that, 
-    the algorithm also returns $[[Θ2]]$ and $[[SC]]$ specifying 
-    the variables used in $[[uM]]$: $[[Θ2]]$ defines the contexts
-    in which the variables must be instantiated, and $[[SC]]$
-    imposes restrictions on the variables. 
-\end{itemize}
-As subroutines, the algorithm calls
-subtyping (\cref{alg:subtyping}),
-type well-formedness (\cref{alg:wf}),
-constraint merge (\cref{sec:constraint-merge}),
-normalization (\cref{alg:type-nf}),
-and constraint singularity which will be defined later in 
-\cref{sec:constraint-singularity}.
-It also relies on basic set operations and the ability to 
-deterministically choose fresh variables.
-
-\begin{algorithm}
-  \hfill \\
-  \ottdefnATPInfLabeled{}
-  \hfill \\
-  \ottdefnATNInfLabeled{}
-  \hfill \\
-  \ottdefnATSpinInfLabeled{}
-\end{algorithm}
-
-Let us discuss the inference rules of the algorithm:
-\begin{itemize}
-  \item \ruleref{\ottdruleATVarLabel} 
-    infers the type of a variable by looking it up in the context
-    and normalizing the result.
-  \item \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
-    are similar to the declarative rules: they make a recursive call
-    to type the body of the thunk or the return expression and
-    put the shift on top of the result.
-  \item \ruleref{\ottdruleATPAnnotLabel} and \ruleref{\ottdruleATNAnnotLabel}
-    are symmetric. They make a recursive call to infer the type of the annotated
-    expression, check that the inferred type is a subtype of the annotation,
-    and return the normalized annotation.
-  \item \ruleref{\ottdruleATtLamLabel} infers the type of a lambda-abstraction.
-    It makes a recursive call to infer the type of the body in the extended context,
-    and returns the corresponding arrow type. Notice that the algorithm also
-    normalizes the result, which is because the annotation type $[[iP]]$
-    is allowed to be non-normalized.
-  \item \ruleref{\ottdruleATTLamLabel} infers the type of a big lambda.
-    Similarly to the previous case, it makes a recursive call to infer the type
-    of the body in the extended \emph{type} context. 
-    After that, it returns the corresponding universal type. 
-    It is also required to normalize the result, because, 
-    for instance, $[[α⁺]]$ might not occur in the body of the lambda,
-    in which case the $[[∀]]$ must be removed. 
-  \item \ruleref{\ottdruleATVarLetLabel} is defined in a standard way:
-    it makes a recursive call to infer the type of the bound value,
-    and then returns the type of the body in the extended context.
-  \item \ruleref{\ottdruleATAppLetAnnLabel}
-    is interpreted as follows.
-    First, it infers the type of the head of the application,
-    ensuring that it is a thunked computation $[[↓iM]]$;
-    after that, it makes a recursive call
-    to the application inference procedure,
-    which returns the algorithmic type, whose
-    instantiation to a declarative type must be associated with the bound variable 
-    $[[x]]$; then premise $[[Γ; Θ ⊨ ↑uQ ≤ ↑iP ⫤ SC2]]$
-    together with $[[Θ ⊢ SC1 & SC2 = SC]]$
-    check whether the instantiation to the annotated type $[[iP]]$ is possible,
-    and if it is, the algorithm infers the type of the body in the extended context,
-    and returns it as the result. 
-  \item \ruleref{\ottdruleATAppLetLabel}
-    works similarly to \ruleref{\ottdruleATAppLetAnnLabel},
-    However, since there is no annotation to assign the result to,
-    the algorithm must infer the `canonical' principal type.
-    To do that algorithmically, we ensure that 
-    the inferred algorithmic type $[[uQ]]$ is instantiated
-    to the minimal possible type $[[ [uσ]uQ ]]$.
-    The premise $[[Γ ⊢ uQ SC minby uσ]]$
-    provides the minimal instantiation of $[[uQ]]$ \wrt $[[SC]]$. It 
-    guarantees that if we consider all possible substitutions satisfying the inferred 
-    constraints $[[SC]]$, then substitution $[[uσ]]$ will instantiate
-    $[[uQ]]$ to the minimal possible type $[[ [uσ]uQ ]]$.
-    This will be the principal type that we assign to the result of the application 
-    (bound to the variable $[[x]]$)
-    and then we infer the type of the body in the context extended with 
-    the bound variable $[[x : [uσ]uQ]]$.
-  \item \ruleref{\ottdruleATUnpackLabel}
-    first, infers the existential type $[[∃nas.iP]]$ of the value being unpacked,
-    and since the type is guaranteed to be normalized, binds 
-    the quantified variables with $[[nas]]$.
-    Then it infers the type of the body in the appropriately extended context
-    and checks that the inferred type does not depend on $[[nas]]$
-    by checking well-formedness $[[Γ ⊢ iN]]$.
-\end{itemize}
-
-Finally, let us discuss the algorithmic rules of the application inference:
-\begin{itemize}
-  \item \ruleref{\ottdruleATEmptyAppLabel}
-    is the base case. If the list of arguments is empty, 
-    the inferred type is the type of the head,
-    and the algorithm returns it after normalizing.
-  \item \ruleref{\ottdruleATArrowAppLabel}
-    is the main rule of algorithmic application inference.
-    If the head has an arrow type $[[uQ → uN]]$,
-    we find $[[SC1]]$---the minimal constraint ensuring that 
-    $[[uQ]]$ is a supertype of the first argument's type.
-    Then we make a recursive call applying $[[uN]]$ to the rest of the arguments,
-    and merge the resulting constraint with $[[SC1]]$
-  \item \ruleref{\ottdruleATForallAppLabel},
-    analogously to the declarative case,
-    is the rule ensuring the implicit elimination of the universal quantifiers. 
-    This is the place where the algorithmic variables are generated.
-    The algorithm simply replaces the quantified variables 
-    $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
-    and makes a recursive call in the extended context. 
-\end{itemize}
-
-The correctness of the algorithm consists of its soundness and 
-completeness, which is by mutual
-induction in \cref{lemma:typing-soundness,lemma:typing-completeness}.
-The simplified result is the following.
-\begin{theorempreview}
-  \hfill
-  \begin{itemize}
-    \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$, 
-      and $[[Γ; Φ ⊢ c : iN]]$ implies $[[Γ; Φ ⊨ c : nf(iN)]]$;
-    \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$, 
-      and $[[Γ; Φ ⊢ v : iP]]$ implies $[[Γ; Φ ⊨ v : nf(iP)]]$.
-  \end{itemize}
-\end{theorempreview}
-
-\subsection{Minimal Instantiation}
-The minimal instantiation algorithm is used
-to infer the type of the bound variable in the un-annotated 
-applicative let-binders, as long as there exists a principal
-(minimal) type.  
-Given a positive algorithmic type $[[uP]]$ 
-and a set of constraints $[[SC]]$, it finds the substitution $[[uσ]]$
-respecting $[[SC]]$ (i.e., $[[Θ ⊢ uσ : SC]]$) such that
-it instantiates $[[uP]]$ to the minimal type, in other words 
-for any other substitution $[[uσ']]$ respecting $[[SC]]$,
-we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP]]$.
-
-The minimal instantiation algorithm is defined as follows:
-\begin{algorithm}[Minimal Instantiation]
-  \hfill \\
-  \ottdefnSINGminInstLabeled{}
-\end{algorithm}
-
-\subsection{Constraint Singularity}
-\label{sec:constraint-singularity}
-
-The singularity algorithm checks whether the constraint $[[SC]]$
-uniquely defines the substitution satisfying it, and if it does,
-the algorithm returns this substitution as the result.
-To implement it, we define a partial function $[[SC singular with uσ]]$,
-taking a subtyping constraint $[[SC]]$ as an argument and returning a substitution 
-$[[uσ]]$---the only possible solution of $[[SC]]$.
-
-First, we define the notion of singularity on constraint entries. 
-$[[scE singular with iP]]$ and its negative counterpart 
-are considered partial functions taking a constraint entry $[[scE]]$
-and returning the type satisfying $[[scE]]$ if such a type is unique.
-
-\begin{algorithm}[Singular Constraint Entry]
-  \hfill\\
-  \ottdefnSINGscEPLabeled{}
-  \ottdefnSINGscENLabeled{}
-\end{algorithm}
-
-\begin{itemize}
-  \item \ruleref{\ottdruleSINGNEqLabel} and \ruleref{\ottdruleSINGPEqLabel} 
-    are symmetric. If the constraint entry says that a variable must be equivalent to 
-    a type $T$, then it is evidently singular, and the only (up-to-equivalence) type
-    instantiating this variable could be $T$. This way, we return its normal form. 
-  \item \ruleref{\ottdruleSINGSupVarLabel}
-    implies that the only (normalized) solution of $[[pua :≥ ∃nas.pa]]$ is 
-    $[[pa]]$ (it will be shown in \cref{lemma:var-subt}).
-  \item \ruleref{\ottdruleSINGSupShiftLabel}
-    is perhaps the least obvious rule.
-    Having a type $[[∃nas.↓iN]]$, suppose that $[[iN]]$ is
-    not equivalent to any just bound variable $[[nai ∊ {nas}]]$.
-    Then the type $[[∃nas.↓iN]]$ has a proper supertype: $[[∃nas.↓na1]]$,
-    and thus the constraint is not singular.
-    Otherwise, if $[[iN]]$ is equivalent to some $[[nai]]$,
-    any supertype of $[[∃nas.↓nai]]$ is equivalent to it, and thus, 
-    the constraint has a unique solution.
-\end{itemize}
-
-Next, we extrapolate the singularity function on constraints---sets of constraint entries. 
-We require $[[SC]]$ to be a set of singular constraints, and the resulting substitution
-sends each variable from $[[dom(SC)]]$ to the unique type satisfying the corresponding constraint.
-
-\begin{algorithm}
-  $[[SC singular with uσ]]$
-  means that 
-  \begin{enumerate}
-    \item for any positive $[[scE ∊ SC]]$,
-      there exists $[[iP]]$ such that $[[scE singular with iP]]$, 
-      and for any negative $[[scE ∊ SC]]$,
-      there exists $[[iN]]$ such that $[[scE singular with iN]]$;
-    \item $[[uσ]]$ is defined as follows:
-      $$
-      [[ [uσ]β̂⁺ ]]  = 
-          \begin{cases}
-              [[ iP ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁺]] 
-                         \text{ and } [[scE singular with iP]] \\
-              [[ β̂⁺ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-      $$
-      [[ [uσ]β̂⁻ ]]  = 
-          \begin{cases}
-              [[ iN ]] & \text{if there is } [[scE]] \in [[dom(SC)]] \text{ restricting } [[β̂⁻]] 
-                         \text{ and } [[scE singular with iN]]\\
-              [[ β̂⁻ ]] & \text{otherwise}  \\
-          \end{cases}
-      $$
-  \end{enumerate} 
-\end{algorithm}
-
-The correctness of the singularity algorithm is formulated as follows:
-\begin{theorempreview}
-  Suppose that $[[SC]]$ is a subtyping constraint.
-  Then $[[SC singular with uσ]]$ holds if and only if 
-  $[[uσ]]$ is the only (up-to-equivalence on $[[dom(SC)]]$) 
-  normalized substitution satisfying $[[SC]]$.
-\end{theorempreview}
-
-\newpage
-
-\section{Theorem Statements: Declarative}
-\subsection{Type Well-Formedness}
-\input{\genDir/tex/wf-statements.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-statements.tex}
-
-\subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-statements.tex}
-
-\subsection{Equivalence}
-\label{sec:decl-equiv-statements}
-\input{\genDir/tex/equiv-statements.tex}
-
-\subsection{Variable Ordering}
-\input{\genDir/tex/ord-statements.tex}
-
-\subsection{Normaliztaion}
-\input{\genDir/tex/norm-statements.tex}
-
-\section{Declarative Typing}
-\input{\genDir/tex/decl-typing-statements.tex}
-
-\section{Theorem Statements: Algorithmic}
-
-\subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-statements.tex}
-
-\subsection{Algorithmic Substitution}
-\input{\genDir/tex/subst-algo-statements.tex}
-
-\subsection{Algorithmic Normalization}
-\input{\genDir/tex/norm-alg-statements.tex}
-
-\subsection{Algorithmic Equivalence}
-\input{\genDir/tex/equiv-alg-statements.tex}
-
-\subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-statements.tex}
-
-\subsection{Unification}
-\input{\genDir/tex/unification-statements.tex}
-
-\subsection{Anti-unification}
-\input{\genDir/tex/au-statements.tex}
-
-\subsection{Upper Bounds}
-\label{sec:alg-upper-bounds-statements}
-\input{\genDir/tex/lub-statements.tex}
-
-\subsection{Upgrade}
-\label{sec:upgrade-statements}
-\input{\genDir/tex/upgrade-statements.tex}
-
-\subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-statements.tex}
-
-\subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-statements.tex}
-
-\subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-statements.tex}
-
-\subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-statements.tex}
-
-\subsection{Singularity and Minimal Instantiation}
-\input{\genDir/tex/singularity-statements.tex}
-
-\subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-statements.tex}
-
-\section{Properties of the Declarative Type System}
-
-\subsection{Type Well-Formedness}
-\input{\genDir/tex/wf-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-lemmas.tex}
-
-\subsection{Declarative Subtyping}
-\input{\genDir/tex/decl-subtyping-lemmas.tex}
-
-\subsection{Equivalence}
-\label{sec:decl-equiv-lemmas}
-\input{\genDir/tex/equiv-lemmas.tex}
-
-\subsection{Variable Ordering}
-\input{\genDir/tex/ord-lemmas.tex}
-
-\subsection{Normaliztaion}
-\input{\genDir/tex/norm-lemmas.tex}
-
-\section{Properties of the Algorithmic Type System}
-
-\subsection{Algorithmic Type Well-formedness}
-\input{\genDir/tex/wf-algo-lemmas.tex}
-
-\subsection{Substitution}
-\input{\genDir/tex/subst-algo-lemmas.tex}
-
-\subsection{Normalization}
-\input{\genDir/tex/norm-alg-lemmas.tex}
-
-\subsection{Equivalence}
-\input{\genDir/tex/equiv-alg-lemmas.tex}
-
-\subsection{Unification Constraint Merge}
-\input{\genDir/tex/unif-constraint-merge-lemmas.tex}
-
-\subsection{Unification}
-\input{\genDir/tex/unification-lemmas.tex}\
-
-\subsection{Anti-unification}
-\input{\genDir/tex/au-lemmas.tex}
-
-\subsection{Upper Bounds}
-\label{sec:alg-upper-bounds-proofs}
-\input{\genDir/tex/lub-lemmas.tex}
-
-\subsection{Upgrade}
-\label{sec:upgrade-lemmas}
-\input{\genDir/tex/upgrade-lemmas.tex}
-
-\subsection{Constraint Satisfaction}
-\input{\genDir/tex/constraint-sat-lemmas.tex}
-
-\subsection{Positive Subtyping}
-\input{\genDir/tex/alg-pos-subtyping-lemmas.tex}
-
-\subsection{Subtyping Constraint Merge}
-\input{\genDir/tex/constraint-merge-lemmas.tex}
-
-\subsection{Negative Subtyping}
-\input{\genDir/tex/alg-neg-subtyping-lemmas.tex}
-
-\section{Properties of the Declarative Typing}
-\input{\genDir/tex/decl-typing-lemmas.tex}
-
-\section{Properties of the Algorithmic Typing}
-
-\subsection{Singularity and Minimal Instantiation}
-\input{\genDir/tex/singularity-lemmas.tex}
-
-\subsection{Correctness of the Typing Algorithm}
-\input{\genDir/tex/alg-typing-lemmas.tex}
+\input{\genDir/content.tex}
 

### 2024-07-09 proofs/tex/decl-subtyping-lemmas.tex
edit and prepare for jfp

@@
           and by inversion, 
-          there exist $[[Γ ⊢ iPs,iPs']]$ 
+          there exist $[[Γ ⊢ iPs]]$ and $[[Γ ⊢ iPs']]$ 
           such that $[[Γ ⊢ [iPs/pas][iPs'/pas']iN' ≤ iM']]$ 

### 2024-07-09 proofs/tex/subst-lemmas.tex
edit and prepare for jfp

@@
       $[[α± ]] \in [[pas]]$, and thus, 
-      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
+      $[[ σ|varset ]] = [[ σ|(varset ∪ {pas}) ]]$
       immediately from the definition.
@@
       [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
-    \item The positive cases are proves symmetrically. 
+    \item The positive cases are proven symmetrically. 
   \end{caseof}

### 2024-07-09 proofs/tex/translation-proofs.tex
edit and prepare for jfp

@@
       We wish to infer
-      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|∀a.fT|+ = c; let y : +|[fA/a]fT|+ = f(); return y : ↑+|[fA/a]fT|+]]$.
+      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : +|∀a.fT|+ = c; let y : +|[fA/a]fT|+ = f(·); return y : ↑+|[fA/a]fT|+]]$.
       First, let us rewrite the polarized types by definition and \cref{lemma:pol-subst}:
-      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$.
+      $[[+|fΓ|+ ; +|fΦ|+ ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$.
       Then we prove this judgment gradually by applying the corresponding rules of \fexists:
@@
           holds by \ruleref{\ottdruleDTVarLabel} and \ruleref{\ottdruleDTReturnLabel}.
-        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
+        \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
           holds by \ruleref{\ottdruleDTAppLetAnnLabel}. Let us show that the 
@@
           \end{enumerate}
-        \item $[[+|fΓ|+ ; +|fΦ|+  ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
+        \item $[[+|fΓ|+ ; +|fΦ|+  ⊢ let f : ↓∀α⁺.↑+|fT|+ = c; let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$
           holds by \ruleref{\ottdruleDTCVarLetLabel}. Let us show that the
@@
             \item $[[+|fΓ|+  ⊢ ↑↓∀α⁺.↑+|fT|+ ≤ ↑↓∀α⁺.↑+|fT|+]]$ by reflexivity (\cref{lemma:subtyping-reflexivity}).
-            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(); return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
+            \item $[[+|fΓ|+ ; +|fΦ|+, f : ↓∀α⁺.↑+|fT|+  ⊢ let y : [+|fA|+/α⁺]+|fT|+ = f(·); return y : ↑[+|fA|+/α⁺]+|fT|+]]$ as noted above.
           \end{enumerate}

### 2024-07-16 paper/tex/01-introduction.tex
WIP: fix references, add appendix

@@
   The central theorems and the key proof ideas are formulated in \cref{sec:proofs}, 
-  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix}).
+  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix,app:proofs}).
 

### 2024-07-16 paper/tex/03-declarative-system.tex
WIP: fix references, add appendix

@@
 their first occurrence}. For brevity, we omit the formal definition of ordering
-referring the reader to the appendix (\cref{appendix}).
+referring the reader to the appendix (\cref{alg:var-ordering}).
 
@@
 For brevity, we omit the formal definition of the term-level elaboration,
-referring the reader to the appendix (\cref{sec:term-translation}). We define
+referring the reader to the appendix (\cref{sec:term-level-translation}). We define
 the elaboration in such a way that it preserves the soundness invariants given

### 2024-07-16 paper/tex/06-extensions.tex
WIP: fix references, add appendix

@@
     premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
-    $$ \ottdruleATArrowAppBidirSLabeled{} $$\\
-
+    $$ \ottdruleATArrowAppBidirSLabeled{} $$
     As one can expect, the declarative rule is also changed accordingly.

### 2024-07-16 proofs/content.tex
WIP: fix references, add appendix

@@
-\newcommand{\proofGenDir}{_gen}
-
-\section{Declarative Type Systems}
-
-\subsection{Grammar}
+% \renewcommand{\proofGenDir}{_gen}
+
+\section{Definitions and Algorithms} 
+\label{appendix}
+
+\subsection{Declarative Types}
+
+\subsubsection{Grammar}
 We assume that there is an infinite set of positive and 
@@
 in a type $[[iP]]$ and $[[iN]]$, respectively.
-
-\bigskip
-
+% \bigskip
 % \ottgrammartabular{
@@
 \begin{definition}[Declarative Types]
-  \hfill
-  \begin{multicols}{2}
+  \hfill\\
+  \begin{minipage}[t]{0.45\textwidth}
     \ottgrammartabular{
@@
     }
-
+  \end{minipage}
+  \hfill
+  \begin{minipage}[t]{0.45\textwidth}
     \ottgrammartabular{
@@
     }
-    \columnbreak
-  \end{multicols}
-\end{definition}
-
-\subsection{Equalities}
+  \end{minipage}
+\end{definition}
+
+\subsubsection{Equalities}
 For simplicity, we assume alpha-equivalent terms are equal.
@@
 
-\subsection{Contexts and Well-formedness}
+\subsubsection{Contexts and Well-formedness}
 
@@
 
-\subsection{Substitutions}
+\subsubsection{Substitutions}
 
@@
 
-\subsection{Declarative Subtyping}
+\subsubsection{Declarative Subtyping}
 Subtyping is one of the key mechanisms of our system. 
@@
 
-\section{Algorithmic Type System}
-
-\subsection{Grammar}
+\subsection{Algorithmic Types}
+
+\subsubsection{Grammar}
 
@@
 
-\subsection{Fresh Variable Selection}
+\subsubsection{Fresh Variable Selection}
 \label{sec:fresh-selection}
@@
 
-\subsection{Variable Algorithmization}
+\subsubsection{Variable Algorithmization}
 \label{sec:variable-algorithmization}
@@
 
-\subsection{Subsitutions}
+\subsubsection{Subsitutions}
 
@@
 
-\subsection{Equivalence and Normalization}
+\subsubsection{Equivalence and Normalization}
 \label{sec:equivalence-normalization}
@@
 central algorithm of the type system---the subtyping algorithm.
-\Cref{fig:subtyping-algo} shows the dependency graph of the subtyping algorithm.
-The nodes represent the algorithmic procedures, and the edge $A \to B$ means that 
-$A$ uses $B$ as a sub-procedure.
-
-\begin{figure}[t]
-  \centering
-  \begin{tikzpicture}
-    [>={Stealth[scale=2]},node distance=2.4cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
-
-    % Define nodes
-    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{sec:subtyping})};
-    \node[left=1cm of 3] (2) {Subtyping Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
-    \node[right=1.4cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
-    \node[below of=3] (4) {Upgrade\\$[[upgrade Γ ⊢ iP to Δ = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
-    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
-
-    % Define edges
-    \draw[->] (1) to (2);
-    \draw[->] (1) to (3);
-    \draw[->] (1) to (5);
-    \draw[->] (2) to (3);
-    \draw[->] (2) to (6);
-    \draw[->] (3) to (4);
-    \draw[->] (3) to (5);
-    \draw[->] (4) to (6);
-    \draw[->] (5) to (8);
-    \draw[->] (6) to (7);
-  \end{tikzpicture}  
-  \caption{Dependency graph of the subtyping algorithm}
-  \label{fig:subtyping-algo}
-\end{figure}
-
-\subsection{Subtyping}
+
+\subsubsection{Subtyping}
 \label{sec:subtyping}
@@
     if  $[[iP]]$ is not well-formed in $[[Θ(â⁺)]]$. For this reason,
-    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:lub})
+    we use the \emph{upgrade} procedure (it will be covered in \cref{sec:a:lub})
     to find the minimal supertype of $[[iP]]$, which is well-formed in $[[Θ(â⁺)]]$.
@@
     must be unified. Hence, the shift rules invoke the unification algorithm, 
-    which will be discussed in \cref{sec:unification}. The unification 
+    which will be discussed in \cref{sec:a:unification}. The unification 
     returns the minimal set of constraints $[[UC]]$, which is necessary
@@
     \emph{subtyping constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
+    which is discussed in \cref{sec:a:constraint-merge}.
   \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
@@
 
-\subsection{Constraints}
+\subsubsection{Constraints}
+\label{sec:a:constraint-merge}
 
@@
 
-\subsubsection{Constraint Satisfaction}
+\paragraph{Constraint Satisfaction}
 
@@
 
-\subsubsection{Constraint Merge}
-\label{sec:constraint-merge}
+\paragraph{Constraint Merge}
 
@@
     and as the output, restrict the variable to be a supertype of the result.
-    The least upper bound procedure will be discussed in \cref{sec:lub}.
+    The least upper bound procedure will be discussed in \cref{sec:a:lub}.
 \end{itemize}
@@
 
-\subsection{Unification}
-\label{sec:unification}
+\subsubsection{Unification}
+\label{sec:a:unification}
 
@@
     \emph{unification constraint merge} procedure, 
-    which is discussed in \cref{sec:constraint-merge}.
+    which is discussed in \cref{sec:a:constraint-merge}.
     Notice that $[[UC1]]$ and $[[UC2]]$ are guaranteed to be
@@
 
-\subsection{Least Upper Bound}
-\label{sec:lub}
+\subsubsection{Least Upper Bound}
+\label{sec:a:lub}
 
@@
 the algorithm finding the least common supertype of two positive types. 
-It is used directly by the constraint merge procedure (\cref{sec:constraint-merge}),
+It is used directly by the constraint merge procedure (\cref{sec:a:constraint-merge}),
 and indirectly, through the type upgrade by positive subtyping
@@
 
-\subsection{Anti-unification}
-\label{sec:antiunification}
+\subsubsection{Anti-unification}
+\label{sec:a:antiunification}
 
@@
 
-\section{Declarative Typing}
-
-In the previous section, we presented the 
-type system together with subtyping specification 
-and the algorithm. In this section, 
-we describe the language under this type system, 
-together with the type inference specification
-and algorithm.
-
-\subsection{Grammar}
-
-First, we define the syntax of the language.
-The language combines System F with call-by-push-value 
-style. 
-\begin{definition}[Language Grammar]
-  \hfill
-  \begin{multicols}{2}
+\subsection{Declarative Typing}
+
+In the previous section, we presented the type system together with the subtyping
+specification and the algorithm. In this section, we describe the language under
+this type system, together with the type inference specification and algorithm.
+
+\subsubsection{Grammar}
+
+The syntax of \fexists terms is given by the following grammar:
+\begin{definition}[Grammar of Terms]
+  \hfill\\
+  \begin{minipage}[t]{0.45\textwidth}
     \ottgrammartabular{
@@
     }
-
+  \end{minipage} \hfill
+  \begin{minipage}[t]{0.45\textwidth}
     \ottgrammartabular{
@@
     }
-  \end{multicols}
-\end{definition}
+  \end{minipage}
+\end{definition}
+
 Notice that the language does not have
@@
 
-\subsection{Declarative Type Inference}
+\subsubsection{Declarative Type Inference}
 
@@
 \begin{definition}[Declarative Type Inference]
+  \hfill\\
   \label{def:declarative-typing}
@@
 
-\section{Relation to System F}
+\subsection{Relation between \fexists and \systemf}
 
@@
 but not type applications (they can be done implicitly); for convenience, we
-introduce multi-argument application, let-blinders, and existential constructors
+introduce multi-argument applications, let-blinders, and existential constructors
 and eliminators as syntactic sugar (see \cref{fig:systemf-grammar}).
@@
 
-\subsection{Type-level Translation}
+\subsubsection{Type-level Translation}
 
@@
 
-\subsection{Term-level Translation}
+\subsubsection{Term-level Translation}
+
+\label{sec:term-level-translation}
 
@@
 
-\paragraph{From \fexists to \systemf: Subtyping Elaboration}
 The subtyping elaboration is defined in \cref{fig:subt-elaboration}.
@@
 
-\paragraph{From \fexists to \systemf: Typing Elaboration}
 Using the subtyping elaboration, we define the elaboration for the
@@
 
-\clearpage % This command will start a new page
+\clearpage
 \begin{figure}[t]
   \ottdefnTELNLabeledTwoColumns{}
-
-  \hfill
 
@@
 \end{figure}
-
-\paragraph{From \systemf to \fexists}
+\clearpage
 
@@
 
-\section{Algorithmic Typing}
+\subsection{Algorithmic Typing}
 
@@
 
-\subsection{Algorithmic Type Inference}
+\subsubsection{Algorithmic Type Inference}
 
@@
 type well-formedness (\cref{alg:wf}),
-constraint merge (\cref{sec:constraint-merge}),
+constraint merge (\cref{sec:a:constraint-merge}),
 normalization (\cref{alg:type-nf}),
 and constraint singularity which will be defined later in 
-\cref{sec:constraint-singularity}.
+\cref{sec:a:constraint-singularity}.
 It also relies on basic set operations and the ability to 
@@
 The simplified result is the following.
-\begin{theorempreview}
+\begin{theorempreview}[Correctness of Algorithmic Typing (simplified)]
   \hfill
@@
 
-\subsection{Minimal Instantiation}
+\subsubsection{Minimal Instantiation}
 The minimal instantiation algorithm is used
@@
 
-\subsection{Constraint Singularity}
-\label{sec:constraint-singularity}
+\subsubsection{Constraint Singularity}
+\label{sec:a:constraint-singularity}
 
@@
 
-\newpage
-
-\section{Theorem Statements: Declarative}
-\subsection{Type Well-Formedness}
+\clearpage
+\section{Theorem Statements}
+
+\subsection{Theorem Statements: Declarative}
+\subsubsection{Type Well-Formedness}
 \input{\proofGenDir/tex/wf-statements.tex}
 
-\subsection{Substitution}
+\subsubsection{Substitution}
 \input{\proofGenDir/tex/subst-statements.tex}
 
-\subsection{Declarative Subtyping}
+\subsubsection{Declarative Subtyping}
 \input{\proofGenDir/tex/decl-subtyping-statements.tex}
 
-\subsection{Equivalence}
+\subsubsection{Equivalence}
 \label{sec:decl-equiv-statements}
@@
 
-\subsection{Variable Ordering}
+\subsubsection{Variable Ordering}
 \input{\proofGenDir/tex/ord-statements.tex}
 
-\subsection{Normaliztaion}
+\subsubsection{Normaliztaion}
 \input{\proofGenDir/tex/norm-statements.tex}
 
-\section{Declarative Typing}
+\subsection{Declarative Typing}
 \input{\proofGenDir/tex/decl-typing-statements.tex}
 
-\section{Relation to \systemf}
+\subsection{Relation to \systemf}
 \input{\proofGenDir/tex/translation-statements.tex}
 
-\section{Theorem Statements: Algorithmic}
-
-\subsection{Algorithmic Type Well-formedness}
+\subsection{Theorem Statements: Algorithmic}
+
+\subsubsection{Algorithmic Type Well-formedness}
 \input{\proofGenDir/tex/wf-algo-statements.tex}
 
-\subsection{Algorithmic Substitution}
+\subsubsection{Algorithmic Substitution}
 \input{\proofGenDir/tex/subst-algo-statements.tex}
 
-\subsection{Algorithmic Normalization}
+\subsubsection{Algorithmic Normalization}
 \input{\proofGenDir/tex/norm-alg-statements.tex}
 
-\subsection{Algorithmic Equivalence}
+\subsubsection{Algorithmic Equivalence}
 \input{\proofGenDir/tex/equiv-alg-statements.tex}
 
-\subsection{Unification Constraint Merge}
+\subsubsection{Unification Constraint Merge}
 \input{\proofGenDir/tex/unif-constraint-merge-statements.tex}
 
-\subsection{Unification}
+\subsubsection{Unification}
 \input{\proofGenDir/tex/unification-statements.tex}
 
-\subsection{Anti-unification}
+\subsubsection{Anti-unification}
 \input{\proofGenDir/tex/au-statements.tex}
 
-\subsection{Upper Bounds}
+\subsubsection{Upper Bounds}
 \label{sec:alg-upper-bounds-statements}
@@
 
-\subsection{Upgrade}
+\subsubsection{Upgrade}
 \label{sec:upgrade-statements}
@@
 
-\subsection{Constraint Satisfaction}
+\subsubsection{Constraint Satisfaction}
 \input{\proofGenDir/tex/constraint-sat-statements.tex}
 
-\subsection{Positive Subtyping}
+\subsubsection{Positive Subtyping}
 \input{\proofGenDir/tex/alg-pos-subtyping-statements.tex}
 
-\subsection{Subtyping Constraint Merge}
+\subsubsection{Subtyping Constraint Merge}
 \input{\proofGenDir/tex/constraint-merge-statements.tex}
 
-\subsection{Negative Subtyping}
+\subsubsection{Negative Subtyping}
 \input{\proofGenDir/tex/alg-neg-subtyping-statements.tex}
 
-\subsection{Singularity and Minimal Instantiation}
+\subsubsection{Singularity and Minimal Instantiation}
 \input{\proofGenDir/tex/singularity-statements.tex}
 
-\subsection{Correctness of the Typing Algorithm}
+\subsubsection{Correctness of the Typing Algorithm}
+\label{sec:alg-typing-statements}
 \input{\proofGenDir/tex/alg-typing-statements.tex}
 
-\section{Properties of the Declarative Type System}
-
-\subsection{Type Well-Formedness}
+\clearpage
+\section{Theorem Proofs}
+\label{app:proofs}
+
+\subsection{Declarative Types}
+
+\subsubsection{Type Well-Formedness}
 \input{\proofGenDir/tex/wf-lemmas.tex}
 
-\subsection{Substitution}
+\subsubsection{Substitution}
 \input{\proofGenDir/tex/subst-lemmas.tex}
 
-\subsection{Declarative Subtyping}
+\subsubsection{Declarative Subtyping}
 \input{\proofGenDir/tex/decl-subtyping-lemmas.tex}
 
-\subsection{Equivalence}
+\subsubsection{Equivalence}
 \label{sec:decl-equiv-lemmas}
@@
 
-\subsection{Variable Ordering}
+\subsubsection{Variable Ordering}
 \input{\proofGenDir/tex/ord-lemmas.tex}
 
-\subsection{Normaliztaion}
+\subsubsection{Normaliztaion}
 \input{\proofGenDir/tex/norm-lemmas.tex}
 
-\section{Relation to \systemf}
+\subsection{Relation to \systemf}
 \input{\proofGenDir/tex/translation-proofs.tex}
 
-\section{Properties of the Algorithmic Type System}
-
-\subsection{Algorithmic Type Well-formedness}
+\subsection{Algorithmic Types}
+
+\subsubsection{Algorithmic Type Well-formedness}
 \input{\proofGenDir/tex/wf-algo-lemmas.tex}
 
-\subsection{Substitution}
+\subsubsection{Substitution}
 \input{\proofGenDir/tex/subst-algo-lemmas.tex}
 
-\subsection{Normalization}
+\subsubsection{Normalization}
 \input{\proofGenDir/tex/norm-alg-lemmas.tex}
 
-\subsection{Equivalence}
+\subsubsection{Equivalence}
 \input{\proofGenDir/tex/equiv-alg-lemmas.tex}
 
-\subsection{Unification Constraint Merge}
+\subsubsection{Unification Constraint Merge}
 \input{\proofGenDir/tex/unif-constraint-merge-lemmas.tex}
 
-\subsection{Unification}
+\subsubsection{Unification}
 \input{\proofGenDir/tex/unification-lemmas.tex}\
 
-\subsection{Anti-unification}
+\subsubsection{Anti-unification}
 \input{\proofGenDir/tex/au-lemmas.tex}
 
-\subsection{Upper Bounds}
+\subsubsection{Upper Bounds}
 \label{sec:alg-upper-bounds-proofs}
@@
 
-\subsection{Upgrade}
+\subsubsection{Upgrade}
 \label{sec:upgrade-lemmas}
@@
 
-\subsection{Constraint Satisfaction}
+\subsubsection{Constraint Satisfaction}
 \input{\proofGenDir/tex/constraint-sat-lemmas.tex}
 
-\subsection{Positive Subtyping}
+\subsubsection{Positive Subtyping}
 \input{\proofGenDir/tex/alg-pos-subtyping-lemmas.tex}
 
-\subsection{Subtyping Constraint Merge}
+\subsubsection{Subtyping Constraint Merge}
 \input{\proofGenDir/tex/constraint-merge-lemmas.tex}
 
-\subsection{Negative Subtyping}
+\subsubsection{Negative Subtyping}
 \input{\proofGenDir/tex/alg-neg-subtyping-lemmas.tex}
 
-\section{Properties of the Declarative Typing}
+\subsection{Declarative Typing}
 \input{\proofGenDir/tex/decl-typing-lemmas.tex}
 
-\section{Properties of the Algorithmic Typing}
-
-\subsection{Singularity and Minimal Instantiation}
+\subsection{Algorithmic Typing}
+
+\subsubsection{Singularity and Minimal Instantiation}
 \input{\proofGenDir/tex/singularity-lemmas.tex}
 
-\subsection{Correctness of the Typing Algorithm}
+\subsubsection{Correctness of the Typing Algorithm}
 \input{\proofGenDir/tex/alg-typing-lemmas.tex}

### 2024-07-16 proofs/tex/alg-pos-subtyping-lemmas.tex
WIP: fix references, add appendix

@@
                 \item $[[ Θ ⊢ nf(uσ) : uv(uP) ]]$ because $[[Θ ⊢ uσ : uv(uP) ]]$ (\cref{corollary:norm-subst-sig-algo})
-                \item $ \begin{aligned}[t]
+                \item \begin{align*}
                         [[ Γ ⊢ [uσ]uN ≈ iM ]] &\Rightarrow [[ [uσ]uN ≈ iM ]]
@@
                                             && \text {by \cref{lemma:norm-subst-distr}}\\
-                        \end{aligned}
-                    $
+                        \end{align*}
             \end{itemize}

### 2024-07-16 proofs/tex/decl-subtyping-lemmas.tex
WIP: fix references, add appendix

@@
     [[fv(iQ) ∪ fv(iM)]] = [[fv(iQ → iM)]]$.
-
-  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$\\
-    $
-    \begin{aligned}[t]
+  \item $[[G ⊢ ∀pas.iN ≤ ∀pbs.iM]]$
+    \begin{align*}
       [[fv ∀pas.iN ]] &\subseteq [[fv ([iPs/pas] iN) ]] ~\setminus~ [[{pbs}]] 
-                      &&   \text{here $[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
+                      &&   \text{$[[{pbs}]]$ is excluded by the premise $[[fv iN ∩ {pbs} = ∅]]$}\\
                       &\subseteq [[fv iM]] ~\setminus~ [[{pbs}]]
-                      &&   \text{by the induction hypothesis, } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
+                      &&   \text{by the i.h., } [[fv ([iPs/pas] iN) ]] \subseteq [[fv iM]] \\
                       &\subseteq [[fv ∀pbs.iM]]
-    \end{aligned}
-    $
+    \end{align*}
   \item The positive cases are symmetric.
@@
       $[[ σ ○ (iQs/pbs) ]]$ unifies $[[iM1']]$ and $[[ [iQs/pbs]iM2']]$:
-      $$
-      \begin{aligned}[t]
+
+      \begin{align*}
         [[ [σ ○ iQs/pbs]iM1' ]] &= [[ [σ][iQs/pbs]iM1' ]]\\
@@
                                 &= [[  [σ ○ iQs/pbs][iQs/pbs]iM2' ]]
-      \end{aligned}
-      $$
+      \end{align*}
+
       This way the induction hypothesis gives us
@@
       Let us rewrite the substitution that we apply to $[[iN']]$:
-      $$
-      \begin{aligned}[t]
+      \begin{align*}
         [[ [σ ○ iQs/pbs ○ iPs/pas]iN' ]] &= [[ [ (σ <=< iQs/pbs) ○ σ ○ iPs/pas]iN' ]]
@@
                                        &= [[ [(((σ <=< iQs/pbs) ○ σ) <=< iPs/pas) ○ σ] iN' ]]
-                                       && \text{Since $[[fv([σ]iN') ∩ {pbs} = ∅]]$}\\
+                                       && [[fv([σ]iN') ∩ {pbs} = ∅]]\\
                                        &= [[ [((σ ○ iQs/pbs) <=< iPs/pas) ○ σ] iN' ]]
@@
                                        &= [[ [(σ ○ iQs/pbs) <=< iPs/pas][σ] iN' ]]
-      \end{aligned}
-      $$
+      \end{align*}
       Notice that $[[(σ ○ iQs/pbs) <=< iPs/pas]]$

### 2024-07-16 proofs/tex/equiv-lemmas.tex
WIP: fix references, add appendix

@@
       and by the induction hypothesis, $[[Γ ⊢ iP]] \iff [[Γ ⊢ iQ]]$ and $[[Γ ⊢ iN']] \iff [[Γ ⊢ iM']]$.
-
-      $
-      \begin{aligned}[t]
+      \begin{align*}
         [[Γ ⊢ iP → iN']]  &\iff [[Γ ⊢ iP]] \text{ and } [[Γ ⊢ iN']] 
@@
                           && \text{by \ruleref{\ottdruleWFTArrowLabel} and inversion}\\
-      \end{aligned}
-      $\\
+      \end{align*}
     \item \ruleref{\ottdruleEOneForallLabel},
@@
     $[[β±]] \in [[fv ∀pas.iM]] = [[fv iM \ {pas}]]$.
-
-    $
-    \begin{aligned}[t]
+    \begin{align*}
       [[β±]] &= [[ [id]β± ]]
@@
              && \text{because $[[{pas} ∩ fv [σ1]β±]] \subseteq [[{pas} ∩ Γ]] = \emptyset$}
-    \end{aligned}
-    $\\
+    \end{align*}
     This way, $[[Γ ⊢ [σ1]β± ≈ β±]]$ for any $[[β±]] \in [[fv ∀pas.iM]]$ and thus,

### 2024-07-16 proofs/tex/norm-lemmas.tex
WIP: fix references, add appendix

@@
      substitution and the induction hypothesis.
-     For example, if $[[iN]] = [[iP → iM]]$ then \\
-     $\begin{aligned}[t]
+     For example, if $[[iN]] = [[iP → iM]]$ then
+     \begin{align*}
         [[nf([σ] iN)]] &= [[ nf([σ] (iP → iM)) ]] \\
@@
                         &= [[ [nf(σ)]nf(iN) ]]
-      \end{aligned}$ \\
-    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute} \\
-      $\begin{aligned}[t]
+      \end{align*}
+    \item $[[iN]] = [[∀ pas.iM]]$ \label{case:norm-subst-commute}
+      \begin{align*}
           [[ [nf(σ)] nf(iN) ]] &= [[ [nf(σ)] nf(∀pas.iM)]] \\
@@
                                \cref{corollary:normalization-ord})}\\
-        \end{aligned}$ \\
+        \end{align*}
 
-      $\begin{aligned}[t]
+      \begin{align*}
          [[ nf([σ]iN) ]] &= [[ nf([σ] ∀pas.iM)]] \\
@@
                           && \text {By the induction hypothesis}\\
-         \end{aligned}$ \\
-
+         \end{align*}
      To show the alpha-equivalence of 
@@
   \begin{caseof}
-  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall} \\
-
-    From the definition of the normalization,
+  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall}
+   From the normalization definition,
     \begin{itemize}
@@
 
-    \item[(ii)] $\begin{aligned}[t] [[ [mu]pbs' ]] &= [[ [mu] ord {pbs} in nf(iM) ]]
+    \item[(ii)] 
+    \begin{align*} 
+      [[ [mu]pbs' ]] &= [[ [mu] ord {pbs} in nf(iM) ]]
                       && \text{by the definition of $[[pbs']]$ } \\
@@
                       && \text{by the definition of $[[pas']]$} \\
-      \end{aligned}$
+      \end{align*}
     \end{enumerate}
@@
 
-  \item $[[iN]] = [[iP → iM]]$\\
-    \label{case:normal-after-subst-arrow}
-    $
-    \begin{aligned}[t]
+  \item $[[iN]] = [[iP → iM]]$ \label{case:normal-after-subst-arrow}
+    \begin{align*}
       [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
@@
                                              \end{cases}
-    \end{aligned}
-    $
+    \end{align*}
   \item $[[iN]] = [[↑iP]]$\\
     By congruence and the inductive hypothesis, similar to \cref{case:normal-after-subst-arrow}
-  \item $[[iN]] = [[∀pas.iM]]$\\
-    $
-    \begin{aligned}[t]
+  \item $[[iN]] = [[∀pas.iM]]$
+    \begin{align*}
       &[[ [σ](∀pas.iM) ]] \text{ is normal} \\
@@
                                               \end{aligned}
-    \end{aligned}
-    $
+    \end{align*}
   \item $[[iP]] = \dots$\\
@@
 \begin{proof}
-  $
-  \begin{aligned}[t] 
+  \begin{align*} 
     [[ iP ≈ iQ ]] &\Rightarrow        [[ nf(iP) = nf(iQ) ]]
@@
                   && \text{by \cref{lemma:subt-equiv-algorithmization}}\\
-  \end{aligned} 
-  $
+  \end{align*} 
 \end{proof}

### 2024-07-16 proofs/tex/ord-lemmas.tex
WIP: fix references, add appendix

@@
       On the other hand, 
-      As a set, $[[{ordVars1} ∪ {ordVars2}]]$
+      as a set $[[{ordVars1} ∪ {ordVars2}]]$
       is equal to $[[ordVars1, (ordVars2 \ {ordVars1})]]$. 
@@
     \begin{caseof}
-    \item $[[na]] \in A$ and $[[na]] \in [[varset]]$\\ Then
-    \\
-      $
-      \begin{aligned}[t] [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in na)]] \\
+    \item $[[na]] \in A$ and $[[na]] \in [[varset]]$. Then
+      \begin{align*} [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in na)]] \\
                                                              &= [[ [mu] na ]]
@@
                                                              &= [[ord [mu] varset in [mu] na ]]
-       \end{aligned}
-       $
+       \end{align*}
      \item $[[na]] \notin A$ and $[[na]] \notin [[varset]]$\\
@@
 
-  \item $[[iN]] = [[↑iP]]$ \\
-    $\begin{aligned}[t]
+  \item $[[iN]] = [[↑iP]]$
+    \begin{align*}
        [[ [mu] (ord varset in iN) ]] &= [[ [mu] (ord varset in ↑iP) ]] \\
@@
                                      &= [[ ord [mu]varset in  [mu]iN ]]
-            \end{aligned}$
+    \end{align*}
 
-   \item $[[iN]] = [[iP → iM]]$  \\
-     $\begin{aligned}[t]
+   \item $[[iN]] = [[iP → iM]]$
+     \begin{align*}
         & [[ [mu] (ord varset in iN) ]] \\ 
@@
           &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
-      \end{aligned}$ \\
+      \end{align*}
       \hfill\\
-      On the other hand, \\
-    $\begin{aligned}[t]
-       & [[  ord [mu] varset in [mu]iN ]] \\ 
-       &= [[ ord [mu] varset in [mu]iP → [mu]iM ]] \\
-       &= [[ ordVarsb1, (ordVarsb2 \ {ordVarsb1}) ]]
-       && \text{where } [[ord [mu] varset in [mu] iP = ordVarsb1]] \text{ and }  \\
-       & && [[ord [mu] varset in [mu] iM = ordVarsb2]]; 
-            \text{ then by the induction}\\
-       & && \text{hypothesis, $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$,
-                   $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$}\\
-       &= [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]
-     \end{aligned}$
+      On the other hand,
+       $[[  ord [mu] varset in [mu]iN ]] = 
+        [[ ord [mu] varset in [mu]iP → [mu]iM ]] = 
+        [[ ordVarsb1, (ordVarsb2 \ {ordVarsb1}) ]] = 
+        [[ [mu] ordVars1, ([mu]ordVars2 \ {[mu]ordVars1}) ]]$,
+        where  $[[ord [mu] varset in [mu] iP = ordVarsb1]]$ 
+        and $[[ord [mu] varset in [mu] iM = ordVarsb2]]$, 
+        then by the induction hypothesis, 
+        $[[ordVarsb1]] = [[ [mu] ordVars1 ]]$, 
+        $[[ordVarsb2]] = [[ [mu] ordVars2 ]]$.
 
-   \item $[[iN]] = [[∀ pas.iM]]$ \\
-     $\begin{aligned}[t]
+   \item $[[iN]] = [[∀ pas.iM]]$
+     \begin{align*}
           [[ [mu] (ord varset in iN) ]] &= [[ [mu] ord varset in ∀pas.iM]] \\
@@
                                         && \text {by the induction hypothesis}\\
-     \end{aligned}$ \\
-     $ 
-     \begin{aligned}[t]
+     \end{align*}
+     \begin{align*}
        [[ (ord [mu] varset in [mu] iN) ]] &= [[ ord [mu] varset in [mu] ∀pas.iM ]] \\
@@
                                           &= [[ ord [mu] varset in [mu] iM ]] \\
-     \end{aligned}
-     $
-
+     \end{align*}
   \end{caseof}
@@
       as a set, $[[ ord varset in [σ]na ]] = [[varset ∩ fv([σ]na)]] \subseteq [[varset ∩ Γ2]] = [[·]]$.
-    \item $[[iN = ∀pas.iM]]$\\
+    \item $[[iN = ∀pas.iM]]$
       We can assume $[[{pas} ∩ Γ1 = ∅]]$
       and $[[{pas} ∩ varset = ∅]]$. Then 
-
-      $\begin{aligned}[t]
+      \begin{align*}[t]
          [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]∀pas.iM ]] \\
@@
                                    &= [[ ord varset in iN ]]
-       \end{aligned}$
-    \item $[[iN = ↑iP]]$\\
-       $\begin{aligned}[t]
+       \end{align*}
+    \item $[[iN = ↑iP]]$
+       \begin{align*}[t]
         [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]↑iP ]] \\
@@
                                    &= [[ ord varset in iN ]]
-       \end{aligned} $
+       \end{align*}
 
-    \item $[[iN = iP → iM]]$\\
-       $ \begin{aligned}
+    \item $[[iN = iP → iM]]$
+       \begin{align*}
         [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ](iP → iM) ]] \\
@@
                                    &= [[ ord varset in iN ]]
-       \end{aligned} $
+       \end{align*}
     \item The proofs of the positive cases are symmetric.

### 2024-07-16 proofs/tex/singularity-lemmas.tex
WIP: fix references, add appendix

@@
             Let us show that it implies $[[Γ ⊢ iP' ≈ ∃α⁻.↓α⁻]]$.
-
-            $
-            \begin{aligned}
+            \begin{align*}
             [[Γ ⊢ iP' ≥ ∃nbs.↓iN1]] &\Rightarrow [[Γ ⊢ nf(iP') ≥ ∃nbs'.↓nf(iN1)]]\\
@@
                                     && \text{by \cref{lemma:shape-supertypes-norm}}\\
-            \end{aligned}
-            $
-
+            \end{align*}
             By \cref{lemma:shape-supertypes-norm}, 

### 2024-07-16 proofs/tex/translation-proofs.tex
WIP: fix references, add appendix

@@
   We prove it by induction on $[[fΓ ⊢ fT]]$.
-  Let us consider the last rule applied to infer this judgment.
+  Let us consider the last rule applied to infer this judgment,
+  or, equivalently, the shape of $[[fΓ ⊢ fT]]$.
   \begin{caseof}
-    \item \ruleref{\ottdruleSFwfVarLabel}
+    \item $[[fΓ ⊢ a]]$.
       By inversion, $[[a ∊ fΓ]]$, 
@@
       by \ruleref{\ottdruleWFTPVarLabel};
-    \item \ruleref{\ottdruleSFwfArrowLabel}
+    \item $[[fΓ ⊢ fA → fB]]$.
       By inversion,  $[[fΓ ⊢ fA]]$ and $[[fΓ ⊢ fB]]$,
@@
       \ruleref{\ottdruleWFTArrowLabel};
-    \item \ruleref{\ottdruleSFwfForallLabel}
+    \item $[[fΓ ⊢ ∀as.fT]]$.
       By inversion, $[[fΓ, a ⊢ fT]]$,

### 2024-07-16 proofs/tex/upgrade-lemmas.tex
WIP: fix references, add appendix

--- 2023-11-13 proofs/tex/upgrade-lemmas.tex
@@
     \end{enumerate}
-
-    $ 
-    \begin{aligned}
+    \begin{align*}
         [[fv iQ]] &\subseteq [[fv [pnbs/pnas]iP ∩ fv [pncs/pnas]iP]]
                   &&\text{since by \cref{lemma:fv-propagation}, 
-                         $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$ and
-                         $[[fv iQ ⊆ fv [pncs/pnas]iP]]$}\\
+                         $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$}\\
+                  & &&  \text{and $[[fv iQ ⊆ fv [pncs/pnas]iP]]$}\\
                   &\subseteq [[ ((fv iP \ {pnas}) ∪ {pnbs}) ∩ ((fv iP \ {pnas}) ∪ {pncs})]]\\
@@
                   &\subseteq [[ {Δ} ]]\\
-    \end{aligned}
-    $\\
+    \end{align*}
     This way, by \cref{lemma:wf-ctxt-equiv}, $[[Δ ⊢ iQ]]$.

### 2024-07-31 paper/tex/02-overview.tex
edit and prepare for jfp
WIP: fix references, add appendix
prettify + minor edits

@@
 which we will cover in more detail in \cref{sec:rel-to-systemf}.
-
 \Cref{fig:polarization-examples} illustrates how standard \systemf types are
@@
 Using the examples provided in \cref{fig:polarization-examples}, we now showcase 
-the type inference capabilities and restrictions of \fexists
+the type inference capabilities and restrictions of \fexists.
 
-\begin{figure}[h]
+\begin{figure}[t]
   \begin{align*}
@@
 
-\paragraph*{Polymorphic Self-application.}
+\paragraph*{Polymorphic Self-application}
   Unlike predicative systems, \fexists permits self-application of polymorphic

### 2024-07-31 paper/tex/04-algorithm.tex
prettify + minor edits

@@
 \begin{figure}[t]
-
   \begin{minipage}[t]{0.49\textwidth}
@@
     declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
-  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
+  \item $[[uv uN]]$ and $[[uv uP]]$ denote the set of free algorithmic variables in $[[uN]]$ and
     $[[uP]]$ respectively.
@@
     % Define nodes
-    \node[] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{fig:alg-subtyping})};
+    \node[xshift=-1cm] (1) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$\\(\cref{fig:alg-subtyping})};
     \node[below of=1] (3) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC]]$\\(\cref{fig:alg-subtyping})};
-    \node[left=1.5cm of 3] (2) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[left=1.7cm of 3] (2) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
     \node[right=1.5cm of 3] (5) {Unification\\ $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$\\ $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$\\(\cref{sec:unification})};
@@
     \node[below of=2] (6) {Least Upper Bound\\$[[Γ ⊨ iP1 ∨ iP2 = iQ]]$\\(\cref{sec:lub})};
-    \node[below of=6] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
-    \node[below of=5] (8) {Unification Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
+    \node[below of=6,xshift=.4cm] (7) {Anti-Unification\\$[[Γ ⊨ iP1 ≈au iP2 ⫤ ( Ξ , uQ , aus1 , aus2 )]]$\\$[[Γ ⊨ iN1 ≈au iN2 ⫤ ( Ξ , uM , aus1 , aus2 )]]$\\(\cref{sec:antiunification})};
+    \node[below of=5] (8) {Unif. Constraint Merge\\$[[Θ ⊢ UC1 & UC2 = UC3]]$\\(\cref{sec:constraint-merge})};
 
@@
     \draw[->] (5) to (8);
-    \draw[->] (6) to node[left, draw=none]{$\ottkw{nf}$} (7);
+    \draw[->] (6) to node[left, draw=none]{$\ottkw{nf}$} ($(7.north) + (-.4cm, 0)$);
   \end{tikzpicture}  
@@
   \begin{tikzpicture}
-    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
+    [>={Stealth[scale=2]},node distance=2.1cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
     % Define nodes
     \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
-    \node[right=1.4cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
-    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
-    \node[below=0.6cm of 2] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
+    \node[right=1.1cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
+    \node[right=0.8cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
+    \node[below=0.6cm of 2, xshift=+0.4cm] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
     \node[below=0.5cm of 7] (8) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
-    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
+    \node[right=0.2cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
     \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
-    \node[below=0.9cm of 6] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
+    \node[below=0.9cm of 6, xshift=+.3cm] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
     \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
@@
     \draw[->] (2) to (3);
-    \draw[->] (2) to (4);
-    \draw[->] (2) to (7);
+    \draw[->] (2) to ([xshift=+.2cm]4.north west);
+    \draw[->] ([xshift=+.4cm]2.south) to (7);
     \draw[->] (3) to (4);
@@
     \draw[->] (1.south west) |- (5);
-    \draw[->] (3) |- (5.south east);
+    \draw[->] ([xshift=+1.2cm]3.south) |- (5.south east);
     \draw[->] (2) to (6);
@@
 \begin{figure}[t]
-  \begin{minipage}{0.49\textwidth}
+  \begin{minipage}{0.43\textwidth}
     `$[[SC singular with uσ]]$' means:
@@
   \end{minipage}
-  \begin{minipage}{0.48\textwidth}
+  \begin{minipage}{0.57\textwidth}
     \ottdefnSINGLabeled

### 2024-07-31 paper/tex/05-proof.tex
WIP: fix references, add appendix
prettify + minor edits

@@
             of the statement and the corresponding theorem is more involved. 
-            See the appendix (\cref{sec:appendix}) for details.}
+            See the appendix (\cref{sec:alg-typing-statements}) for details.}
 
@@
                 \item $[[Γ; dom(Θ) ⊢ uN]]$ (free from \emph{negative} algorithmic variables), and
-                \item $[[Θ ⊢ uσ : uv(uN)]]$,
+                \item $[[Θ ⊢ uσ : uv uN]]$,
             \end{enumerate*}
@@
                 \item \label{point:mostGeneral} for any other $[[uσ]]$ and $[[iM]]$ 
-                (where $[[Θ ⊢ uσ : uv(uN)]]$ and $[[Γ ⊢ iM]]$)
+                (where $[[Θ ⊢ uσ : uv uN]]$ and $[[Γ ⊢ iM]]$)
                     such that $[[Γ; Φ ⊢ [uσ]uN ● args ⇒> iM]]$, 
@@
 %         $[[Γ ; dom(Θ) ⊢ uP]]$ and $[[Γ ⊢ iQ]]$, 
-%         suppose that there exists $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]uP = iQ ]]$,
+%         suppose that there exists $[[Θ ⊢ uσ : uv uP]]$ such that $[[ [uσ]uP = iQ ]]$,
 %         then $[[Γ ; Θ ⊨ uP ≈u iQ ⫤ UC]]$ for some $[[UC]]$.
@@
 %         $[[Γ ; dom(Θ) ⊢  uN]]$ and $[[Γ ⊢ iM]]$,
-%         suppose that there exists $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[ [uσ]uN = iM ]]$,
+%         suppose that there exists $[[Θ ⊢ uσ : uv uN]]$ such that $[[ [uσ]uN = iM ]]$,
 %         then $[[Γ ; Θ ⊨ uN ≈u iM ⫤ UC]]$ for some $[[UC]]$.
@@
     Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iQ]]$ and $[[Γ ; dom(Θ) ⊢  uP]]$.
-    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv(uP)]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
+    Then if there exists $[[uσ]]$ such that $[[Θ ⊢ uσ : uv uP]]$ and $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$,
     then the subtyping algorithm succeeds: $[[Γ; Θ ⊨ uP ≥ iQ ⫤ SC]]$.
@@
     and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
-    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
+    Then for any $[[Θ ⊢ uσ : uv uN]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
     the subtyping algorithm succeeds: $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$.

### 2024-07-31 proofs/content.tex
prettify + minor edits

@@
 
+\pagebreak[4]
 \begin{algorithm}[Type Well-formedness]
   \label{alg:wf}
-  \hfill
-
-  \begin{multicols}{2}
+  \hfill\\
+  \begin{minipage}[t]{0.45\textwidth}
     \ottdefnWFTNiWFLabeled{}
-  \columnbreak
-
+  \end{minipage}
+  \begin{minipage}[t]{0.45\textwidth}
     \ottdefnWFTPiWFLabeled{}
-  \end{multicols}
-
+  \end{minipage}
 \end{algorithm}
@@
 
-Sometimes it is convenient to construct substitution 
-explicitly mapping each variable from a list (or a set)
-to a type. Such substitutions are denoted as $[[iPs / pas]]$
+Sometimes it is convenient to construct a substitution 
+that explicitly maps each variable \emph{from a list} 
+(or a set) to a type.
+Such substitutions are denoted as $[[iPs / pas]]$
 and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
@@
 
-\begin{definition} 
+\begin{definition}[Declarative Subtyping and Equivalence]
   \label{def:subDOne}
-  \hfill
-
-  \begin{multicols}{2}
+  \hfill\\
+  \begin{minipage}[t]{0.45\textwidth}
       \ottdefnDOneNsubLabeled{}
-
+  \end{minipage}
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
       \ottdefnDOnePsupLabeled{}
-  \end{multicols}
-  \hfill
-
-  \begin{multicols}{2}
+  \end{minipage}
+
+  \begin{minipage}[t]{0.45\textwidth}
       \ottdefnDOneNeqLabeled{}
-
+  \end{minipage}
+  \hfill
+  \begin{minipage}[t]{0.45\textwidth}
       \ottdefnDOnePeqLabeled{}
-  \end{multicols}
+  \end{minipage}
 \end{definition}
@@
   \label{def:algo-types}
-  \hfill\\
+  \hfill
   \begin{multicols}{2}
@@
     }
-
+    \columnbreak\\
     \ottgrammartabular{
@@
     }
-    \columnbreak
   \end{multicols}
+  \hfill\\
 \end{definition}
@@
 
-  \begin{multicols}{2}
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnWFATNauWFLabeled{}
-  \columnbreak
-
+  \end{minipage}%
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnWFATPauWFLabeled{}
-  \end{multicols}
-
+  \end{minipage}
 \end{algorithm}
 
-Algorithmic Type Context are used in the unification algorithm.
+Algorithmic Type Contexts are used in the unification algorithm.
 In the subtyping algorithm, 
 the context needs to remember additional information.
-In the subtyping context, each algorithmic variable is associated with a
+In particular, each algorithmic variable is associated with a
 context it must be instantiated in 
 (i.e. the context in which the type replacing the variable must be well-formed).
-This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
-\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
+This association is represented by \emph{instantiation context} $[[Θ]]$.
+\begin{definition}[Instantiation context $[[Θ]]$]
   \hfill \\
-  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
+  Instantiation context $[[Θ]]$ is represented by a set of 
   entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
@@
   and $[[Γ]]$ is a context in which they must be instantiated.
-  We assume that no two entries associating the same variable
-  appear in $[[Θ]]$.
+  We assume that $[[Θ]]$ does not contain multiple entries for the same variable.
 
@@
   \ottdefnONVarLabeled{}
+  \newpage
   \ottdefnOPVarLabeled{}
@@
 
-  \begin{multicols}{2}
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnNrmNNormLabeled{}
-  \columnbreak\\
+  \end{minipage}%
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnNrmPNormLabeled{}
-  \end{multicols}
-
-  Analogously, we define the normalization of algorithmic types by adding base cases:
-
-  \begin{multicols}{2}
+  \end{minipage}
+
+  Analogously, we define the normalization of algorithmic types by adding the following base cases:
+
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnNrmuNNormLabeled{}
-  \columnbreak\\
+  \end{minipage}%
+  \hfill
+  \begin{minipage}[t]{0.48\textwidth}
     \ottdefnNrmuPNormLabeled{}
-  \end{multicols}
+  \end{minipage}
 
@@
 The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
-the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
+the instantiation context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
 must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
@@
     fresh algorithmic variables in the body of the quantified type,
-    the algorithmic variables are added to the subtyping context $[[Θ]]$,
+    the algorithmic variables are added to the instantiation context $[[Θ]]$,
     after that, the recursive call is made. Notice that the declarative context
@@
 \begin{definition}[Well-formed Constraint]
-  We say that a constraint is well-formed in a
-  subtyping context $[[Θ]]$ if all its entries are well-formed in
+  We say that a constraint is well-formed in an 
+  instantiation context $[[Θ]]$ if all its entries are well-formed in
   the corresponding elements of $[[Θ]]$.
@@
 
+\newcommand{\graytxt}[1]{\textcolor{darkgray}{#1}}
+\newcommand{\ottprodlinecomm}[7]{\graytxt{#1}& & \graytxt{$#2$} & \graytxt{$#3$} & \graytxt{$#4 #5$} & \graytxt{$#6$} & \graytxt{$#7$}}
+
 \begin{figure}
-  \begin{minipage}[t]{0.39\linewidth}
-    \vspace*{0pt}
     \ottgrammartabular{
       \ottfT\ottprodnewline
-      \ottprodline{}{}{}{}{}{}\ottprodnewline
-      \ottprodline{}{}{}{}{}{}\ottprodnewline
-      \ottprodline{|}{[[∃as.fT]]}{}{}{}{}\ottprodnewline
-      \ottprodline{}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}{}\ottinterrule
+      \\
+      \ottprodlinecomm{(syntax sugar)}{|}{[[∃as.fT]]}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}\ottinterrule
     }
-  \end{minipage}
-  \begin{minipage}[t]{0.6\linewidth}
-    \vspace*{0pt}
+
     \ottgrammartabular{
       \ottft\ottprodnewline
-      \ottprodline{}{}{}{}{}{}\ottprodnewline
-      \ottprodline{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
-      \ottprodline{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
-      \ottprodline{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
-      \ottprodline{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
+      \\
+      \ottprodlinecomm{(syntax sugar)}{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
+      \ottprodlinecomm{}{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
+      \ottprodlinecomm{}{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
+      \ottprodlinecomm{}{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
     }
-  \end{minipage}
+
   \caption{Grammar of \systemf}
@@
 
+\newpage
 \begin{observation}
-  The following rules are admissible in \systemf: \hfill
-
+  The following rules are admissible in \systemf: \hfill\\
   \begin{minipage}{0.5\textwidth}
@@
 \begin{figure}[t]
-  \begin{minipage}[t]{0.65\linewidth}
+  \begin{minipage}[b]{0.65\linewidth}
     \centering
-      \begin{minipage}[t]{0.32\linewidth}
+      \begin{minipage}[t]{0.4\linewidth}
       \ottfundefnunpolP{}
       \end{minipage}
-      \begin{minipage}[t]{0.32\linewidth}
+      \begin{minipage}[t]{0.4\linewidth}
       \ottfundefnunpolN{}
@@
   \end{minipage}
-  \hspace{0.5cm}
-  \begin{minipage}[t]{0.3\linewidth}
+  \begin{minipage}[b]{0.3\linewidth}
     \centering

### 2024-07-31 proofs/tex/alg-neg-subtyping-lemmas.tex
prettify + minor edits

@@
             This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
-            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv(uN')]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
+            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv uN']]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
             Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
-            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv(uP)]]$
+            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv uP]]$
             and $[[ Θ ⊢ uσ : SC1 ]]$.
@@
             To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness})
-            (notice that $[[Θ ⊢ uσ : uv(uP → uN')]]$ means 
-            $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$).
+            (notice that $[[Θ ⊢ uσ : uv uP → uN']]$ means 
+            $[[Θ ⊢ uσ : uv uP ∪ uv uN']]$).
             This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined and $[[ Θ ⊢ uσ : SC ]]$ holds.

### 2024-07-31 proofs/tex/alg-neg-subtyping-statements.tex
prettify + minor edits

@@
     $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
-    $[[Θ ⊢ SC : uv(uN)]]$ and 
+    $[[Θ ⊢ SC : uv uN]]$ and 
     for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
@@
     and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
-    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
+    Then for any $[[Θ ⊢ uσ : uv uN]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
     there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.

### 2024-07-31 proofs/tex/alg-pos-subtyping-lemmas.tex
prettify + minor edits

@@
         By the soundness of unification (\cref{lemma:unification-soundness}),
-        $[[Θ ⊢ SC : uv(uN)]]$ and for any normalized $[[uσ]]$, $[[ Θ ⊢ uσ : SC ]]$
+        $[[Θ ⊢ SC : uv uN]]$ and for any normalized $[[uσ]]$, $[[ Θ ⊢ uσ : SC ]]$
         implies $[[ [uσ]nf(uN) = nf(iM) ]]$, 
@@
                 \item $[[Γ ⊢ nf(iM)]]$ because $[[Γ ⊢ iM]]$ (\cref{corollary:wf-nf})
-                \item $[[ Θ ⊢ nf(uσ) : uv(uP) ]]$ because $[[Θ ⊢ uσ : uv(uP) ]]$ (\cref{corollary:norm-subst-sig-algo})
+                \item $[[ Θ ⊢ nf(uσ) : uv uP ]]$ because $[[Θ ⊢ uσ : uv uP ]]$ (\cref{corollary:norm-subst-sig-algo})
                 \item \begin{align*}
@@
             $[[ Θ ⊢ uσ : SC1 \ {â⁻*} ]]$,
-            it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0 : uv(uP0)]]$---an extension of $[[uσ]]$ with
-            $[[{â⁻*} ∩ uv(uP0)]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
+            it suffices to build $[[Θ, â⁻*[G, nbs] ⊢ us0 : uv uP0]]$---an extension of $[[uσ]]$ with
+            $[[{â⁻*} ∩ uv uP0]]$ such that $[[Γ, nbs ⊢ [us0]uP0 ≥ iQ]]$.
             Then by the induction hypothesis, $[[Θ, â⁻*[G, nbs] ⊢ us0 : SC1]]$,
@@
                     \begin{cases}
-                    [[ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[{â⁻*} ∩ uv(uP0)]] \\
-                    [[ [uσ]β̂± ]] & \text{if } [[β̂±]] \in [[ uv(uP') ]]
+                    [[ [σ]αi⁻ ]] & \text{if } [[β̂±]] = [[αî⁻]] \in [[{â⁻*} ∩ uv uP0]] \\
+                    [[ [uσ]β̂± ]] & \text{if } [[β̂±]] \in [[ uv uP' ]]
                     \end{cases}
             \]
-            It is easy to see $[[Θ, â⁻*[Γ, nbs] ⊢ uσ0 : uv(uP0)]]$:
-            $[[uv(uP0)]] = [[ uv([â⁻*/nas]uP') ]] = [[{â⁻*} ∩ uv(uP0) ∪ uv(uP')]]$. Then
+            It is easy to see $[[Θ, â⁻*[Γ, nbs] ⊢ uσ0 : uv uP0]]$:
+            $[[uv uP0]] = [[ uv [â⁻*/nas]uP' ]] = [[{â⁻*} ∩ uv uP0 ∪ uv uP']]$. Then
             \begin{enumerate}
-                    \item for $[[αî⁻]] \in [[{â⁻*} ∩ uv(uP0)]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
+                    \item for $[[αî⁻]] \in [[{â⁻*} ∩ uv uP0]]$, $[[ (Θ, â⁻*[Γ, nbs])(αî⁻) ⊢ [uσ0] αî⁻]]$, 
                     i.e. $[[ Γ, nbs ⊢ [σ]αi⁻ ]]$ holds since $[[Γ, nbs ⊢ σ :{nas}]]$,
-                    \item for $[[β̂±]] \in [[ uv(uP') ]] \subseteq [[dom(Θ)]]$, $[[ (Θ, â⁻*[Γ, nbs])(β̂±) ⊢ [uσ0] β̂± ]]$,
-                    i.e. $[[Θ(β̂±) ⊢ [uσ] β̂± ]]$ holds since $[[Θ ⊢ uσ : uv(uP)]]$ and $[[β̂±]] \in [[ uv(uP') ]] = [[uv(uP)]]$.
+                    \item for $[[β̂±]] \in [[ uv uP' ]] \subseteq [[dom(Θ)]]$, $[[ (Θ, â⁻*[Γ, nbs])(β̂±) ⊢ [uσ0] β̂± ]]$,
+                    i.e. $[[Θ(β̂±) ⊢ [uσ] β̂± ]]$ holds since $[[Θ ⊢ uσ : uv uP]]$ and $[[β̂±]] \in [[ uv uP' ]] = [[uv uP]]$.
             \end{enumerate}
@@
                 $[[ [uσ][σ][nas/â⁻*]αî⁻ ]] = [[ [uσ][σ]αi⁻ ]] = [[ [σ]αi⁻ ]]$,
-                \item if this variable is a algorithmic variable $[[β̂±]] \in [[ uv(uP0) \ {â⁻*} ]] = [[uv(uP')]]$, then
+                \item if this variable is a algorithmic variable $[[β̂±]] \in [[ uv uP0 \ {â⁻*} ]] = [[uv uP']]$, then
                 $[[ [uσ0]β̂± ]] = [[ [uσ]β̂± ]]$ and $[[ [uσ][σ][nas/â⁻*]β̂± ]] = [[ [uσ][σ]β̂± ]] = [[ [uσ]β̂± ]]$,

### 2024-07-31 proofs/tex/constraint-merge-lemmas.tex
prettify + minor edits

@@
                 \item By assumption, $[[Γ ⊢ iQ]]$, and hence $[[Γ ⊢ scE]]$.
-                \item Since $[[uv(uQ2) = ∅]]$, 
+                \item Since $[[uv uQ2 = ∅]]$, 
                     $[[Γ;· ⊨ uQ2 ≥ iQ1 ⫤ ·]]$ implies $[[Γ ⊢ iQ2 ≥ iQ1]]$

### 2024-07-31 proofs/tex/norm-lemmas.tex
prettify + minor edits

@@
                         &= [[ nf([σ]iP → [σ]iM) ]]
-                        && \text{By congruence of substitution} \\
+                        && \text{by congruence of substitution} \\
                         &= [[ nf([σ]iP) → nf([σ]iM) ]]
-                        && \text{By congruence of normalization} \\
+                        && \text{by congruence of normalization} \\
                         &= [[ [nf(σ)]nf(iP) → [nf(σ)]nf(iM) ]]
-                        && \text{By induction hypothesis} \\
+                        && \text{by induction hypothesis} \\
                         &= [[ [nf(σ)](nf(iP) → nf(iM)) ]]
-                        && \text{By congruence of substitution} \\
+                        && \text{by congruence of substitution} \\
                         &= [[ [nf(σ)]nf(iP → iM) ]]
-                        && \text{By congruence of normalization} \\
+                        && \text{by congruence of normalization} \\
                         &= [[ [nf(σ)]nf(iN) ]]
@@
                             &= [[ [nf(σ)] ∀pas'.nf(iM) ]]
-                            && \text {Where $[[pas']] = [[ ord {pas} in nf(iM)]]
-                               = [[ord {pas} in iM]]$}\\
-                           & && 
-                               \text{(the latter is by
-                               \cref{corollary:normalization-ord})}\\
+                            && \text {where $[[pas']] = [[ ord {pas} in nf(iM)]]$}\\
+                            & && = [[ord {pas} in iM]]\\
+                            & && \text{(the latter is by \cref{corollary:normalization-ord})}\\
         \end{align*}
@@
                           &= [[ nf(∀pas.[σ]iM) ]]
-                          && \text{Assuming $[[{pas} ∩ Γ1]] = \emptyset$
+                          && \text{assuming $[[{pas} ∩ Γ1]] = \emptyset$
                              and $[[{pas} ∩ Γ2]] = \emptyset$}\\
                           &= [[ ∀pbs.nf([σ]iM) ]]
-                          && \text {Where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]
-                             = [[ord {pas} in [σ]iM]]$}\\
-                          & && 
-                             \text{(the latter is by \cref{corollary:normalization-ord})}\\
+                          && \text {where $[[pbs]] = [[ord {pas} in nf([σ]iM)]]$}\\
+                          & && = [[ord {pas} in [σ]iM]]\\
+                          & && \text{(the latter is by \cref{corollary:normalization-ord})}\\
                           &= [[ ∀pas'.nf([σ]iM) ]]
-                          && \text{By \cref{lemma:ord-sigma},}\\
+                          && \text{by \cref{lemma:ord-sigma},}\\
                           & && \text{$[[pbs]] = [[pas']]$
@@
                           &= [[ ∀pas'.[nf(σ)]nf(iM) ]]
-                          && \text {By the induction hypothesis}\\
+                          && \text {by the induction hypothesis}\\
          \end{align*}
@@
                       && \text{by \cref{lemma:distr-mu-ord}, because}\\
-                      & && \text{$[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]]
-                        = \emptyset$}\\
-                      &
-                      && \text{$[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq
-                        [[{pas} ∩ fv iM]] = \emptyset$} \\
+      &  \omit\rlap{\hspace{5em} $[[{pas} ∩ fv iN]] \cap [[fv nf(iM)]] \subseteq [[{pas} ∩ fv iM ]] = \emptyset$, and}\\ 
+      &  \omit\rlap{\hspace{5em} $[[{pas} ∩ fv iN]] \cap [[({pbs} ∩ fv iM)]] \subseteq [[{pas} ∩ fv iM]] = \emptyset$}\\ 
                       &= [[ ord [mu] ({pbs} ∩ fv iM) in nf(iN) ]]
-                      && \text{since $[[ [mu] nf(iM) ]] = [[nf(iN)]]$ is proved } \\
+                      && \text{as $[[ [mu] nf(iM) ]] = [[nf(iN)]]$} \\
                       &= [[ ord ({pas} ∩ fv iN) in nf(iN) ]]
-                      && \text{because $\mu$ is a bijection between}\\
+                      && \text{$\mu$ is a bijection between}\\
                       & && \text{$[[{pas} ∩ fv iN]]$ and $[[{pbs} ∩ fv iM]]$} \\
@@
       [[ [σ](iP → iM) ]] \text{ is normal} &\iff [[ [σ]iP → [σ]iM ]] \text{ is normal}
-                                           && \text{by substitution
-                                              congruence} \\
+                                           && \text{substitution definition} \\
                                            &\iff
@@
                                              \end{cases}
-                                           && \text{by the induction hypothesis}\\
+                                           && \text{induction hypothesis}\\
                                            &\iff

### 2024-07-31 proofs/tex/ord-lemmas.tex
prettify + minor edits

@@
                                                              &= [[nb]]
-                                                             && \text{for some $[[nb]] \in B$ (notice $[[nb]] \in [[ [mu]varset ]]$)} \\
+                                                             && \text{for some $[[nb]] \in B$}\\
+                                                             & && \text{(notice $[[nb]] \in [[ [mu]varset ]]$)} \\
                                                              &= [[ ord [mu]varset in nb ]]
                                                              && \text{by \ruleref{\ottdruleOPVarInLabel},
-                                                                because $[[nb]] \in [[ [mu]varset ]]$} \\
+                                                                as $[[nb]] \in [[ [mu]varset ]]$} \\
                                                              &= [[ord [mu] varset in [mu] na ]]
@@
           && \text{by induction on $[[ordVars2]]$;
-                  the inductive step is similar to 
+                  the ind. step is similar to 
                   \cref{case:distr-mu-ord:var}.}\\
-        & && \text{Notice that $[[mu]]$ is collision free on $[[{ordVars1}]]$ and $[[{ordVars2}]]$} \\
+        & && \text{notice that $[[mu]]$ is collision free on $[[{ordVars1}]]$ and $[[{ordVars2}]]$} \\
         & && \text{since
@@
                                    &\phantom{=} ~ [[(ord varset in iM \ {ord varset in iP}) ]]
-                                   && \text{the induction hypothesis}\\
+                                   && \text{the ind. hypothesis}\\
                                    &= [[ ord varset in iP → iM ]]

### 2024-07-31 proofs/tex/singularity-lemmas.tex
prettify + minor edits

@@
             the inferred judgment has form $[[Γ ⊢ uP SC minby uσ]]$,
-            and by inversion, $[[uv(uP) ⊆ dom(SC)]]$ and
-            $[[SC|uv(uP) singular with uσ]]$. Let us apply the soundness of singularity
-            (\cref{lemma:singularity-soundness}) to $[[SC|uv(uP) singular with uσ]]$ to
+            and by inversion, $[[uv uP ⊆ dom(SC)]]$ and
+            $[[SC|uv uP singular with uσ]]$. Let us apply the soundness of singularity
+            (\cref{lemma:singularity-soundness}) to $[[SC|uv uP singular with uσ]]$ to
             obtain the following properties:
             \begin{itemize}
-                \item $[[ Θ ⊢ uσ : uv(uP) ∩ dom(SC) ]]$, which also means 
-                    $[[Θ ⊢ uσ : uv(uP)]]$,
-                \item $[[Θ ⊢ uσ : SC|uv(uP)]]$,
+                \item $[[ Θ ⊢ uσ : uv uP ∩ dom(SC) ]]$, which also means 
+                    $[[Θ ⊢ uσ : uv uP]]$,
+                \item $[[Θ ⊢ uσ : SC|uv uP]]$,
                 \item $[[uσ]]$ is normalized, and
-                \item for any other $[[Θ ⊢ uσ' : uv(uP) ]]$
-                    respecting $[[SC|uv(uP)]]$, we have
-                    $[[Θ ⊢ uσ' ≈ uσ : uv(uP)]]$. The latter means that
+                \item for any other $[[Θ ⊢ uσ' : uv uP ]]$
+                    respecting $[[SC|uv uP]]$, we have
+                    $[[Θ ⊢ uσ' ≈ uσ : uv uP]]$. The latter means that
                     $[[Γ ⊢ [uσ']uP ≈ [uσ]uP]]$, and in particular, 
@@
             \ruleref{\ottdruleSINGSingLabel}.
-            The first premise $[[uv(α̂⁺) ⊆ dom(SC)]]$  
+            The first premise $[[uv α̂⁺ ⊆ dom(SC)]]$  
             holds by assumption;
@@
             \begin{itemize}
-                \item $[[uv(uN) = dom(SC|uv uN)]]$, then $[[uv(uP) ⊆ dom(SC)]]$,
+                \item $[[uv uN = dom(SC|uv uN)]]$, then $[[uv uP ⊆ dom(SC)]]$,
                 \item $[[SC|uv uN singular with uσ0]]$ for some (normalized) $[[uσ0]]$.
@@
             It means $[[Γ ⊢ uP SC minby uσ0 ]]$, and then
-            since $[[uσ0]]$ is normalized and equivalent to $[[uσ0]]$ on $[[uv(uN)]]$,
-            and its domain is $[[uv(uN)]]$, $[[uσ0 = nf(uσ)]]$.
+            since $[[uσ0]]$ is normalized and equivalent to $[[uσ0]]$ on $[[uv uN]]$,
+            and its domain is $[[uv uN]]$, $[[uσ0 = nf(uσ)]]$.
         % \item $[[uP = ∃nas.uP']]$ 
@@
             Then by \ruleref{\ottdruleSINGExistsLabel},
-            we have $[[Γ ⊢ ∃nas.↓uN minby nf(uσ)]]$.
+            we have $[[Γ ⊢ ∃nas.↓uN SC minby nf(uσ)]]$.
     \end{caseof}

### 2024-07-31 proofs/tex/subst-algo-lemmas.tex
prettify + minor edits

@@
       $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
-      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
+      which since $[[uv uP → uN = uv uP ∪ uv uN]]$ implies
       $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
@@
       we apply congruence of substitution, equality, and normalization,
-      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
+      then the induction hypothesis, and then the fact that $[[uv ↑uP = uv uP]]$.
   \end{caseof}
@@
   $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
-  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
+  $[[Γ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
   \cref{lemma:subt-equiv-algorithmization}.
@@
   This way, by \cref{lemma:subst-eq-algovar}, 
-  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
+  $[[Γ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies 
   $[[nf(uσ1)|(uv nf(uP)) = nf(uσ2)|(uv nf(uP))]]$.
@@
   Symmetrically, 
-  $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
+  $[[Γ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ implies 
   $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.

### 2024-07-31 proofs/tex/unification-lemmas.tex
prettify + minor edits

@@
         \item $[[uN]] = [[α⁻]]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ : uv(α⁻)]]$ such that $[[ [uσ]α⁻ = iM ]]$.
-            Since $[[uv(α⁻) = ∅]]$, $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv α⁻]]$ such that $[[ [uσ]α⁻ = iM ]]$.
+            Since $[[uv α⁻ = ∅]]$, $[[ [us]α⁻ = iM ]]$ means $[[iM = α⁻]]$.
 
@@
         \item $[[uN]] = [[↑uP]]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uP)]]$ such that $[[ [uσ]↑uP = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv uP]]$ such that $[[ [uσ]↑uP = iM ]]$.
             The latter means $[[ ↑[uσ]uP = iM ]]$, i.e.
@@
         \item $[[uN]] = [[uP → uN']]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uP → uN')]]$ such that $[[ [uσ](uP → uN') = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv uP → uN']]$ such that $[[ [uσ](uP → uN') = iM ]]$.
             The latter means $[[ [uσ]uP → [uσ]uN' = iM ]]$, i.e.
@@
             Let us show that the induction hypothesis is applicable to 
-            $[[Θ ⊢ uσ|uv(uP) : uv(uP)]]$ and $[[ [uσ|uv(uP)]uP = iQ ]]$ 
-            (the latter holds since $[[ [uσ|uv(uP)]uP = [uσ]uP ]]$ by \cref{lemma:subst-restr-uv}),
+            $[[Θ ⊢ uσ|uv uP : uv uP]]$ and $[[ [uσ|uv uP]uP = iQ ]]$ 
+            (the latter holds since $[[ [uσ|uv uP]uP = [uσ]uP ]]$ by \cref{lemma:subst-restr-uv}),
             \begin{itemize}
@@
                 $[[Θ ⊢ lift UC2 : uv uN']]$ (which hold by soundness),
-                and $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$, 
-                which holds since $[[uv(uP) ∪ uv(uN') = uv(uP → uN')]]$.
-            Notice that by soundness, $[[Θ ⊢ uσ|uv(uP) : lift UC1]]$, 
+                and $[[Θ ⊢ uσ : uv uP ∪ uv uN']]$, 
+                which holds since $[[uv uP ∪ uv uN' = uv uP → uN']]$.
+            Notice that by soundness, $[[Θ ⊢ uσ|uv uP : lift UC1]]$, 
             which implies $[[Θ ⊢ uσ : lift UC1]]$.
@@
         \item $[[uN]] = [[∀pas.uN']]$\\
-            Let us take an arbitrary $[[Θ ⊢ uσ : uv(uN')]]$ such that $[[ [uσ]∀pas.uN' = iM ]]$.
+            Let us take an arbitrary $[[Θ ⊢ uσ : uv uN']]$ such that $[[ [uσ]∀pas.uN' = iM ]]$.
             The latter means $[[ ∀pas.[uσ]uN' = iM ]]$, i.e.
@@
             $[[Γ,pas ⊢ iM']]$ follows from inversion of $[[Γ ⊢ ∀pas.iM']]$, and
-            $[[Θ ⊢ uσ : uv(uN')]]$ by assumption.
+            $[[Θ ⊢ uσ : uv uN']]$ by assumption.
 

### 2024-07-31 proofs/tex/upgrade-lemmas.tex
prettify + minor edits

@@
         [[fv iQ]] &\subseteq [[fv [pnbs/pnas]iP ∩ fv [pncs/pnas]iP]]
-                  &&\text{since by \cref{lemma:fv-propagation}, 
+                  &&\text{by \cref{lemma:fv-propagation}, 
                          $[[fv iQ ⊆ fv [pnbs/pnas]iP]]$}\\

### 2024-07-31 proofs/tex/wf-algo-lemmas.tex
prettify + minor edits

@@
   and the symmetric negative case.
-  In this case, $[[uP]] = [[α̂⁺]]$, and $[[uv(uP)]] = [[{α̂⁺} ⊆ Ξ]]$ by inversion; $[[fv(uP)]] = [[∅ ⊆ Γ]]$ vacuously.
+  In this case, $[[uP]] = [[α̂⁺]]$, and $[[uv uP]] = [[{α̂⁺} ⊆ Ξ]]$ by inversion; $[[fv(uP)]] = [[∅ ⊆ Γ]]$ vacuously.
 \end{proof}
@@
   $[[Γ1; Ξ1 ⊢ uP]]$ implies $[[fv(uP) ⊆ Γ1 ⊆ Γ2]]$
-  and $[[uv(uP) ⊆ Ξ1 ⊆ Ξ2]]$,
+  and $[[uv uP ⊆ Ξ1 ⊆ Ξ2]]$,
   and thus, $[[fv(uP)]] = [[fv(uP) ∩ Γ1]] = [[fv(uP) ∩ Γ2]]$, 
-  and $[[uv(uP)]] = [[uv(uP) ∩ Ξ1]] = [[uv(uP) ∩ Ξ2]]$.
+  and $[[uv uP]] = [[uv uP ∩ Ξ1]] = [[uv uP ∩ Ξ2]]$.
   Then by \cref{lemma:wf-algo-ctxt-equiv}, $[[Γ2; Ξ2 ⊢ uP]]$. 

### 2024-07-31 proofs/tex/translation-proofs.tex
fix \cref

@@
       which means that $[[+|fA|+]]$ is well-formed in 
-      $[[+|fΓ|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
+      $[[+|fΓ|+]]$ by \cref{lemma:typ-pol-wf}.
       Then by \ruleref{\ottdruleDTtLamLabel} we infer 
@@
           \begin{enumerate}
-            \item $[[+|fΓ|+ ⊢ +|fB|+]]$ by \cref{lemma:typ-pol-wf-completeness};
+            \item $[[+|fΓ|+ ⊢ +|fB|+]]$ by \cref{lemma:typ-pol-wf};
             \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+), x:+|fA|+ ⊢ f : ↓(+|fA|+ → ↑+|fB|+)]]$ by 
@@
           \begin{enumerate}
-            \item $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness};
+            \item $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf};
             \item $[[+|fΓ|+ ; +|fΦ|+, f:↓(+|fA|+ → ↑+|fB|+) ⊢ c' : ↑+|fA|+]]$
@@
           \begin{enumerate}
-            \item $[[+|fΓ|+ ⊢ +|fA → fB|+]]$ by \cref{lemma:typ-pol-wf-completeness};
+            \item $[[+|fΓ|+ ⊢ +|fA → fB|+]]$ by \cref{lemma:typ-pol-wf};
             \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fA → fB|+]]$ by the induction hypothesis, as noted above;
@@
           $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|∀a.fT|+]]$ by the induction hypothesis; and
-        \item $[[fΓ ⊢ fA]]$, which implies $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
+        \item $[[fΓ ⊢ fA]]$, which implies $[[+|fΓ|+ ⊢ +|fA|+]]$ by \cref{lemma:typ-pol-wf}.
       \end{itemize}
@@
           \begin{enumerate}
-            \item $[[+|fΓ|+ ⊢ ↓∀α⁺.↑+|fT|+]]$ by \cref{lemma:typ-pol-wf-completeness}.
+            \item $[[+|fΓ|+ ⊢ ↓∀α⁺.↑+|fT|+]]$ by \cref{lemma:typ-pol-wf}.
             \item $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑↓∀α⁺.↑+|fT|+]]$ by the induction hypothesis, as noted above.

### 2024-07-31 paper/tex/01-introduction.tex
fix capitalization

@@
 
-\paragraph*{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
+\paragraph*{Declarative type system specification (\cref{sec:overview,sec:declarative-system})} 
     We present \fexists---a second-order polymorphic type system that supports first-class
@@
 
-\paragraph*{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
-  inference algorithm implementing the declarative specification. Because our
-  system supports both existential and universal quantifiers, our implementation
-  of the subtyping relation requires the use of \emph{anti-unification}---a dual
-  to unification that finds the most specific generalization rather than the
-  most general unifier. To our knowledge, anti-unification has never been used
-  in the context of polymorphic type inference.
+\paragraph*{Type inference algorithm (\cref{sec:algorithm})} 
+  We give a type inference algorithm implementing the declarative specification.
+  Because our system supports both existential and universal quantifiers, our
+  implementation of the subtyping relation requires the use of
+  \emph{anti-unification}---a dual to unification that finds the most specific
+  generalization rather than the most general unifier. To our knowledge,
+  anti-unification has never been used in the context of polymorphic type
+  inference.
 

### 2024-07-31 paper/tex/02-overview.tex
fix capitalization

@@
 
-\paragraph*{Application of a Polymorphic Function}
+\paragraph*{Application of a polymorphic function}
   A polymorphic function can be applied to a value of a concrete type. In this
@@
 
-\paragraph*{Polymorphic Self-application}
+\paragraph*{Polymorphic self-application}
   Unlike predicative systems, \fexists permits self-application of polymorphic
@@
 
-\paragraph*{Non-example: Polymorphic Arguments}
+\paragraph*{Non-example: polymorphic arguments}
 The focus of our system is the \emph{local} type inference
@@
 
-\paragraph*{Inferring the Common Supertype}
+\paragraph*{Inferring the common supertype}
   The function $[[choose]]$ takes two arguments of the same
@@
 
-\paragraph*{Syntactic Conventions}
+\paragraph*{Syntactic conventions}
 We always consider terms and types up to alpha-equivalence, 
@@
 
-\paragraph*{Type Context and Type Well-formedness}
+\paragraph*{Type context and type well-formedness}
 
@@
 
-\paragraph*{Least Upper Bound}
+\paragraph*{Least upper bound}
 Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of

### 2024-07-31 paper/tex/03-declarative-system.tex
prettify + minor edits
fix capitalization

@@
 
-\paragraph*{Invariant Shifts}
+\paragraph*{Invariant shifts}
 As mentioned above, an important restriction that we put on the subtyping system
@@
 
-  \vspace{\baselineskip}
+  % \vspace{\baselineskip}
 % \subsubsection{Properties of Declarative Typing}
@@
 
-\paragraph*{Type-level Translation}
+\paragraph*{Type-level translation}
 
@@
 
-\paragraph*{Term-level Translation}
+\paragraph*{Term-level translation}
 
@@
 
-\begin{table}[h]
+\begin{figure}[t]
 \centering
@@
   \label{tab:elaboration-soundness}
-\end{table}
+\end{figure}
 

### 2024-07-31 paper/tex/04-algorithm.tex
fix capitalization

@@
 
-\paragraph*{Algorithmic Variables}
+\paragraph*{Algorithmic variables}
 To design a mode-correct inference system, we slightly modify the language we operate on.
@@
 
-\paragraph*{Algorithmic Types}
+\paragraph*{Algorithmic types}
 The syntax of algorithmic types extends the declarative syntax by adding
@@
 
-\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
+\paragraph*{Algorithmic contexts $[[Ξ]]$ and well-formedness}
 To specify when algorithmic types are well-formed, we define algorithmic
@@
 
-\paragraph*{Instantiation Context $[[Θ]]$}
+\paragraph*{Instantiation context $[[Θ]]$}
 When one instantiates an algorithmic variable, one may only use type variables
@@
 
-\paragraph*{Algorithmic Substitution}
+\paragraph*{Algorithmic substitution}
 We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
@@
 
-\paragraph*{Algorithmic Normalization}
+\paragraph*{Algorithmic normalization}
 Similarly to well-formedness, the normalization of algorithmic types is defined
@@
 
-\paragraph*{Auxiliary Functions}
+\paragraph*{Auxiliary functions}
 \begin{itemize}
@@
 
-\paragraph*{Equivalent Substitutions}
+\paragraph*{Equivalent substitutions}
 In the proofs of the algorithm correctness, 
@@
 
-\paragraph*{The Type Upgrade}
+\paragraph*{Type upgrade}
 The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
@@
 
-\paragraph*{The Least Upper Bound}
+\paragraph*{The least upper bound}
 The Least Upper Bound algorithm we use operates on \emph{positive}
@@
 
-\paragraph*{The Anti-Unification Rule}
+\paragraph*{The anti-unification rule}
   Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
@@
 
-\paragraph*{Return and Thunk}
+\paragraph*{Return and thunk}
   Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
@@
 
-\paragraph*{Let Binders}
+\paragraph*{Let binders}
   Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
@@
 
-\paragraph*{Application to an Empty List of Arguments}
+\paragraph*{Application to an empty list of arguments}
   Rule \ruleref{\ottdruleATEmptyAppLabel}
@@
 
-\paragraph*{Application of a Polymorphic Type $[[∀]]$}
+\paragraph*{Application of a polymorphic type $[[∀]]$}
   Rule \ruleref{\ottdruleATForallAppLabel},
@@
 
-\paragraph*{Application of an Arrow Type}
+\paragraph*{Application of an arrow type}
   Rule \ruleref{\ottdruleATArrowAppLabel}

### 2024-07-31 paper/tex/05-proof.tex
fix capitalization

@@
 
-\paragraph*{Mutuality of the Soundness and Completeness Proofs}
+\paragraph*{Mutuality of the soundness and completeness proofs}
 
@@
 
-\paragraph*{Inductive Metric}
+\paragraph*{Inductive metric}
     The soundness and completeness lemmas are proven by \emph{mutual} induction. 

### 2024-07-31 paper/tex/06-extensions.tex
fix capitalization

@@
 \vspace{\baselineskip}
-\paragraph*{Solution 1: Declarative Normalization}
+\paragraph*{Solution 1: declarative normalization}
 One way to resolve this ambiguity is to fix the order of quantifiers.  
@@
 \vspace{\baselineskip}
-\paragraph*{Solution 2: Elementary Type Inference}
+\paragraph*{Solution 2: elementary type inference}
 An alternative approach to explicit type application
@@
 \vspace{\baselineskip}
-\paragraph*{The Algorithm}
+\paragraph*{The algorithm}
     The algorithmic system is bidirectionalized in a similar way. 

### 2024-07-31 proofs/tex/decl-typing-lemmas.tex
fix capitalization

@@
-\begin{definition}[Number of prenex quantifiers]
+\begin{definition}[Number of Prenex Quantifiers]
     Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.

### 2024-07-31 proofs/tex/ord-lemmas.tex
fix capitalization

@@
 
-\begin{definition} [Collision-free bijection]
+\begin{definition} [Collision-free Bijection]
   We say that a bijection $\mu : A \leftrightarrow B$ between sets of

### 2024-08-04 paper/tex/01-introduction.tex
fix refs

@@
 both universal and existential quantifiers, which can both be instantiated
-impredicatively. This extended system is referred to as \fexists. The
-combination of features in \fexists  broke a number of the invariants which
-traditional type inference algorithms depend on and required us to invent
-new algorithms which combine both unification and (surprisingly)
-anti-unification.
+impredicatively. This extended system is referred to as \fexists. 
+The version of this system without existential quantifiers was studied 
+by \citet{mercer2019:system-f}. However, the addition of existentials in \fexists
+broke a number of the invariants that traditional type inference algorithms depend
+ on and required us to invent new algorithms which combine both unification and 
+ (surprisingly) anti-unification.
 
@@
 % reader to the appendix.
-
-\paragraph*{\textbf{Acknowledgments}} We would like to thank Meven
-Lennon-Bertrand for his extremely helpful comments and suggestions.

### 2024-08-04 proofs/content.tex
fix capitalization
fix refs

@@
-% \renewcommand{\proofGenDir}{_gen}
+% Add prefix to the labels so that they do not conflict with the main paper
+\newcommand{\apppref}{:app}
+\renewcommand{\ruleref}[1]{\nameref{\apppref#1}}
 
@@
 In other words, it is syntax-directed and mode-correct 
-(according to \cite{dunfieldBidirectionalTyping2020}), 
+(according to \citet{dunfield2020:bidirectional}), 
 which means that checking the well-formedness of a type
@@
   \begin{minipage}[t]{0.45\textwidth}
-    \ottdefnWFTNiWFLabeled{}
+    \ottdefnWFTNiWFLabeled[\apppref]{}
   \end{minipage}
   \begin{minipage}[t]{0.45\textwidth}
-    \ottdefnWFTPiWFLabeled{}
+    \ottdefnWFTPiWFLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.45\textwidth}
-      \ottdefnDOneNsubLabeled{}
+      \ottdefnDOneNsubLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.48\textwidth}
-      \ottdefnDOnePsupLabeled{}
+      \ottdefnDOnePsupLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.45\textwidth}
-      \ottdefnDOneNeqLabeled{}
+      \ottdefnDOneNeqLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.45\textwidth}
-      \ottdefnDOnePeqLabeled{}
+      \ottdefnDOnePeqLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnWFATNauWFLabeled{}
+    \ottdefnWFATNauWFLabeled[\apppref]{}
   \end{minipage}%
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnWFATPauWFLabeled{}
+    \ottdefnWFATPauWFLabeled[\apppref]{}
   \end{minipage}
@@
 This association is represented by \emph{instantiation context} $[[Θ]]$.
-\begin{definition}[Instantiation context $[[Θ]]$]
+\begin{definition}[Instantiation Context $[[Θ]]$]
   \hfill \\
@@
   \begin{multicols}{2}
-  \ottdefnEOneNeqLabeled{}
+  \ottdefnEOneNeqLabeled[\apppref]{}
   \columnbreak\\
-  \ottdefnEOnePeqLabeled{}
+  \ottdefnEOnePeqLabeled[\apppref]{}
   \end{multicols}
@@
 
-  \ottdefnONVarLabeled{}
+  \ottdefnONVarLabeled[\apppref]{}
   \newpage
-  \ottdefnOPVarLabeled{}
+  \ottdefnOPVarLabeled[\apppref]{}
 
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnNrmNNormLabeled{}
+    \ottdefnNrmNNormLabeled[\apppref]{}
   \end{minipage}%
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnNrmPNormLabeled{}
+    \ottdefnNrmPNormLabeled[\apppref]{}
   \end{minipage}
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnNrmuNNormLabeled{}
+    \ottdefnNrmuNNormLabeled[\apppref]{}
   \end{minipage}%
@@
   \begin{minipage}[t]{0.48\textwidth}
-    \ottdefnNrmuPNormLabeled{}
+    \ottdefnNrmuPNormLabeled[\apppref]{}
   \end{minipage}
@@
   \hfill\\
-  \ottdefnANsubLabeled{}
-  \ottdefnAPsupLabeled{}
+  \ottdefnANsubLabeled[\apppref]{}
+  \ottdefnAPsupLabeled[\apppref]{}
 \end{algorithm}
@@
   \begin{multicols}{2}
-  \ottdefnSATSCENLabeled{}
+  \ottdefnSATSCENLabeled[\apppref]{}
   \columnbreak\\
-  \ottdefnSATSCEPLabeled{}
+  \ottdefnSATSCEPLabeled[\apppref]{}
   \end{multicols}
@@
 
-  \ottdefnSCMELabeled\\
+  \ottdefnSCMELabeled[\apppref]\\
 \end{algorithm}
@@
   \begin{multicols}{2}
-  \ottdefnUNUnifLabeled{}
+  \ottdefnUNUnifLabeled[\apppref]{}
   \columnbreak\\
-  \ottdefnUPUnifLabeled{}
+  \ottdefnUPUnifLabeled[\apppref]{}
   \end{multicols}
@@
   \hfill\\
-  \ottdefnLUBNsubLabeled{}
+  \ottdefnLUBNsubLabeled[\apppref]{}
 \end{algorithm}
@@
   \hfill\\
-  \ottdefnLUBUpLabeled{}
+  \ottdefnLUBUpLabeled[\apppref]{}
 \end{algorithm}
@@
 
-  \ottdefnAUAUPLabeled{}
-  \ottdefnAUAUNLabeled{}
+  \ottdefnAUAUPLabeled[\apppref]{}
+  \ottdefnAUAUNLabeled[\apppref]{}
 
@@
   \label{def:declarative-typing}
-  \ottdefnsDTLabeled
+  \ottdefnsDTLabeled[\apppref]{}
 \end{definition}
@@
 \begin{figure}
-  \ottdefnSFtyLabeledTwoColumns{}
+  \ottdefnSFtyLabeledTwoColumns[\apppref]{}
   \caption{Typing rules of \systemf}
@@
   \begin{minipage}{0.5\textwidth}
-    \ottusedrule{\ottdruleSFtyAUnpackLabeled{}}
+    \ottusedrule{\ottdruleSFtyAUnpackLabeled[\apppref]{}}
   \end{minipage}
   \begin{minipage}{0.5\textwidth}
-    \ottusedrule{\ottdruleSFtyAPackLabeled{}}
-    \ottusedrule{\ottdruleSFtyALetLabeled{}}
+    \ottusedrule{\ottdruleSFtyAPackLabeled[\apppref]{}}
+    \ottusedrule{\ottdruleSFtyALetLabeled[\apppref]{}}
   \end{minipage}
@@
 \begin{figure}[t]
-  \ottdefnELLabeled
+  \ottdefnELLabeled[\apppref]
   \caption{Subtyping elaboration from \fexists to \systemf}
@@
 \begin{figure}[t]
-  \ottdefnTELNLabeledTwoColumns{}
-
-  \ottdefnTELPAppLabeled{}
+  \ottdefnTELNLabeledTwoColumns[\apppref]{}
+
+  \ottdefnTELPAppLabeled[\apppref]{}
   \caption{Typing elaboration from \fexists to \systemf}
@@
 \begin{figure}[t]
-  \ottdefnELfPolLabeled{}
+  \ottdefnELfPolLabeled[\apppref]{}
   \caption{Typing elaboration form from \systemf to \fexists}
@@
   \hfill \\
-  \ottdefnATPInfLabeled{}
+  \ottdefnATPInfLabeled[\apppref]{}
   \hfill \\
-  \ottdefnATNInfLabeled{}
+  \ottdefnATNInfLabeled[\apppref]{}
   \hfill \\
-  \ottdefnATSpinInfLabeled{}
+  \ottdefnATSpinInfLabeled[\apppref]{}
 \end{algorithm}
@@
   \hfill \\
-  \ottdefnSINGminInstLabeled{}
+  \ottdefnSINGminInstLabeled[\apppref]{}
 \end{algorithm}
@@
   \hfill\\
-  \ottdefnSINGscEPLabeled{}
-  \ottdefnSINGscENLabeled{}
+  \ottdefnSINGscEPLabeled[\apppref]{}
+  \ottdefnSINGscENLabeled[\apppref]{}
 \end{algorithm}

### 2024-08-04 proofs/tex/norm-lemmas.tex
fix refs

@@
   \begin{caseof}
-    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
+    \item{\ruleref{\ottdruleNrmNVarLabel} and \ruleref{\ottdruleNrmPVarLabel}}\\ By
       the corresponding equivalence rules.
-    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
-        and \nameref{\ottdruleNrmArrowLabel}}\\
+    \item{\ruleref{\ottdruleNrmShiftULabel}, \ruleref{\ottdruleNrmShiftDLabel},
+        and \ruleref{\ottdruleNrmArrowLabel}}\\
       By the induction hypothesis and the corresponding congruent equivalence rules.
-    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
+    \item{\ruleref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
       From the induction hypothesis, we
@@
       $[[iN ≈ [mu]iN']]$. Since these sets are equal, we take $\mu = id$.
-    \item{\nameref{\ottdruleNrmExistsLabel}} Same as for \cref{case:norm-soundness:forall}.
+    \item{\ruleref{\ottdruleNrmExistsLabel}} Same as for \cref{case:norm-soundness:forall}.
   \end{caseof}
@@
   \begin{caseof}
-  \item {\nameref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall}
+  \item {\ruleref{\ottdruleEOneForallLabel}} \label{case:ord-completeness:forall}
    From the normalization definition,
@@
     \end{enumerate}
-  \item {\nameref{\ottdruleEOneExistsLabel}} Same as for \cref{case:ord-completeness:forall}.
+  \item {\ruleref{\ottdruleEOneExistsLabel}} Same as for \cref{case:ord-completeness:forall}.
   \item Other rules are congruent, and thus, proved by the corresponding congruent alpha-equivalence rule,
